### 2020-07-30 vertical-thermal-review.Rmd
started .Rmd version of manuscript

---
title: ''
output: word_document
bibliography: references.bib
csl: new-phytologist.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Title:** Thermal sensitivity of canopy versus understory leaves: patterns, mechanisms, and ecological implications

**Authors (so far):** 
Nidhi Vinod^1^
Martijn Slot^2^
Kristina J. Anderson-Teixeira^1,2^*

**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA

2. Smithsonian Tropical Research Institute; Panama, Republic of Panama

*corresponding author: teixeirak@si.edu; +1 540 635 6546

\newpage

## Abstract

\newpage

## Introduction

**As the climate warms, understanding forest responses to temperature is critical.**
Forest canopy microclimate buffering is emerging as important for forest ecology in an era of climate change.

- Forest canopies buffer temperatures and other conditions
     - @zellweger_forest_2020

We’re seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change.
     - @zellweger_forest_2020 
     - (Suggitt et al. 2018, Scheffers et. al 2014)
    - Larger trees suffer more during drought (Bennett et al. 2015) and may be partially influenced by temperature

**However, we lack a systematic understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecology (Fig. 1).**

This review addresses the following questions:

i.	How does the biophysical environment vary with height in forests?

ii.	How do leaf traits vary with height (or between sun and shade leaves) in forests?

iii.	How do biophysical environment and traits combine to affect leaf temperature?

iv.	How does leaf metabolism respond to temperature in canopy and understory settings?

v.	What are the implications of these patterns for the ecology and climate change responses of canopy versus understory trees?

*Our primary interest is the height gradient from the top of the canopy to the understory in forests. However,

![**Figure 1. (schematic of a forest summarizing most important gradients.** Current fig is just a rough illustration of how this might look – a draft figure that KAT had on hand illustrating hypotheses (ignore specific content). We could have a set of arrows for each of the major categories considered here. This would be a key figure, and should be beautifully illustrated—KAT could do a watercolor, or Nidhi could illustrate). ](figures_tables/schematic.png)

## The biophysical environment

![**Figure 2. Vertical gradients in the biophysical environment, from NEON data.** Current figure is old version from Ian McGregor’s in-review paper, showing NEON data from SCBI. We could modify his code to analyze all forested NEON sites. We obviously wouldn’t present this much info per site. ](NEON_height_profiles/NEON_height_profile_temp.png)

## Trait variation

## Leaf temperature

## Leaf metabolism and thermal stress

## Ecology

## Future Questions

## Conclusions

\newpage
## References

### 2020-07-30 vertical-thermal-review.Rmd
worked on transfer to .rmd

@@
 **As the climate warms, understanding forest responses to temperature is critical.**
-Forest canopy microclimate buffering is emerging as important for forest ecology in an era of climate change.
 
-- Forest canopies buffer temperatures and other conditions
+**Forest canopy microclimate buffering is emerging as important for forest ecology in an era of climate change.**
+Forest canopies buffer temperatures and other conditions
      - @zellweger_forest_2020
@@
 
-*Our primary interest is the height gradient from the top of the canopy to the understory in forests. However,
+*Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests.
+However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), we also review studies focused on exposure gradients.*
 
@@
 This would be a key figure, and should be beautifully illustrated—KAT could do a watercolor, or Nidhi could illustrate).
-](figures_tables/schematic.png)
+](schematic/schematic.png)
 
@@
 
+**Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).**
+Most notably, solar radiation decreases along a vertical profile from the top of the canopy to the forest floor (Fig. **2x**).
+- *@bonan_ecological_2016 reviews this and points to appropriate references*
+- (Mau et al. 2018?)
+-	Sunflecks: Leaky et al. 2003?
+
+Wind speeds are also higher at the top of the canopy (Fig. **2x**).
+- *(find some good refs for this/ fill in more specifics)*
+- @mcgregor_tree_nodate
+This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999].
+
+Air temperature, $T_{air}$, is sometimes signficantly buffered by forest canopies (Fig. **2x**).
+Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
+Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama (Rey-Sánchez et al. 2016) and coastal Brazil (Fauset et al. 2018).
+However, similar maximum temperatures have been observed during the dry season in Panama (Rey-Sánchez et al.
+2016) and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+Minimum $T_{air}$ is also buffered by forest canopies under some conditions. 
+Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
+However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama (Rey-Sánchez et al.
+2016) and coastal Brazil (Fauset et al.
+2018), as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+
+Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. **2x**).
+- DETAILS - REFS
+- @mcgregor_tree_nodate
+RH, in combination with temperature, determines vapor pressure deficit, $VPD$
+- $VPD$ is lower in understory of Atlantic forest during wet season (Fauset et al. 2018)
+This means that canopy leaves tend to be exposed to higher *evaporative demand*.
+
+Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory. 
+-	Higher in understory, particularly at dusk (Koike et al. 2001).
+-	Higher in the understory overnight; difference persists during the day but is very small (Yang et al. 1999).
+Differences in concentration are by far most pronounced near ground level. 
+Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient.
+
 ![**Figure 2.
-Vertical gradients in the biophysical environment, from NEON data.** Current figure is old version from Ian McGregor’s in-review paper, showing NEON data from SCBI.
-We could modify his code to analyze all forested NEON sites.
-We obviously wouldn’t present this much info per site.
+Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is old version from Ian McGregor’s in-review paper, showing NEON data from SCBI.
+See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2.
 ](NEON_height_profiles/NEON_height_profile_temp.png)
+
+**The strength of this buffering varies across forests, being influenced by both forest characteristics and the biophysical environment.**
+First, buffering increases with canopy cover.
+- greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
+- greater cover -->  lower max T [@zellweger_seasonal_2019]
+- [@thom_effects_2020]
+*Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
+Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
+SCA species increase T buffering (Zelllweger et al. 2019) [@zellweger_seasonal_2019]
+The strength of buffering also varies with respect to geographic and climatic factors.
+- Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
+- [@davis_microclimatic_2019]
 
@@
 
+![**Figure 4. Temperature sensitivity of tree growth for understory versus canopy trees.** Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)
+
 ## Future Questions

### 2020-07-30 vertical-thermal-review.Rmd
worked on .Rmd

@@
 -	Higher in understory, particularly at dusk (Koike et al. 2001).
--	Higher in the understory overnight; difference persists during the day but is very small (Yang et al. 1999).
+-	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
 Differences in concentration are by far most pronounced near ground level. 
@@
 
+Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig.
+3).
+While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.
+Large leaves can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation with stomata closed.  
+Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
+Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
+However, counteracting this, lower wind speed in understory would reduce latent heat loss.
+Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
+
+![**Figure 3.
+Leaf temperature in response to (a) XXX, (b) XXX (key trait or environmental variables).** Examples to be produced based on energy balance model using the tealeaves R package of @muir_tealeaves_2019: https://github.com/EcoClimLab/vertical-thermal-review/issues/6.
+Current placeholder shows figures from Campbell&Norman.](leaf_energy_balance/C&N_figures.png)
+Aligning with biophysical expectations, field observations have show that leaf temperatures are influenced by the biophysical environment and leaf traits.
+1.	Air T
+2.	Solar radiation
+3.	Leaf traits
+
+When vertical gradients in the biopgysical environment and leaf traits combine to shape leaf energy budgets, $T_{leaf}$ often differs little from $T_{air}$ and between understory and canopy (Fig.
+**2?**).
+For instance... 	(Bolstad et al. 1999).
+Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
+However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.
+-	(Slot et al. 2019) and refs therein
+-	(Fauset et al. 2018)
+Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016).
+
 ## Leaf metabolism and thermal stress

### 2020-07-31 vertical-thermal-review.Rmd
progress on .Rmd

@@
 
-This review addresses the following questions:
+**This review addresses the following questions:**
 
@@
 
-Wind speeds are also higher at the top of the canopy (Fig. **2x**).
+**Wind speeds are also higher at the top of the canopy (Fig. **2x**).**
 - *(find some good refs for this/ fill in more specifics)*
@@
 
-Air temperature, $T_{air}$, is sometimes signficantly buffered by forest canopies (Fig. **2x**).
+**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. **2x**).**
 Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
@@
 2018), as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+*(Add something about risk of freeze.)*
 
-Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. **2x**).
+**Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. **2x**).**
 - DETAILS - REFS
@@
 
-Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory. 
+**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
 -	Higher in understory, particularly at dusk (Koike et al. 2001).
@@
 
+**Many traits vary with height and/or between sun and shade leaves on the same species (Table 1).**
+
 ## Leaf temperature
 
-Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig.
-3).
+**Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
+$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
 While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.
@@
 Current placeholder shows figures from Campbell&Norman.](leaf_energy_balance/C&N_figures.png)
-Aligning with biophysical expectations, field observations have show that leaf temperatures are influenced by the biophysical environment and leaf traits.
+**Aligning with biophysical expectations, field observations have show that leaf temperatures are influenced by the biophysical environment and leaf traits.**
 1.	Air T
@@
 
-When vertical gradients in the biopgysical environment and leaf traits combine to shape leaf energy budgets, $T_{leaf}$ often differs little from $T_{air}$ and between understory and canopy (Fig.
-**2?**).
+**When vertical gradients in the biopgysical environment and leaf traits combine to shape leaf energy budgets, $T_{leaf}$ often differs little from $T_{air}$ and between understory and canopy (Fig.
+**2?**).**
 For instance... 	(Bolstad et al. 1999).
 Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
-However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
 -	(Slot et al. 2019) and refs therein
@@
 
-## Leaf metabolism and thermal stress
+## Leaf metabolism and thermal responses
+
+Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. 1). 
+However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves.
+
+**Photosynthesis is generally higher in sun leaves-- a fact that is well-established (REFS) and observed in numerous field studies.**
+This is primarily driven by the greater light available to sun leaves.
+In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light,  including ....
+The vertical gradient in photosythetic rates is also influenced by...
+
+**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves**
+Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
+14.5; *revisit this!*) (Campbell and Norman 1998).
+However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photoshythesis in sun and shade leaves.
+For 3 species in Panama, (Slot et al. 2019) found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
+Mau et al. found no trend along a height gradient in Puerto Rico, and no significant trend in temperate trees (Mau et al. 2018).
+
+**Similar to photosynthesis, respiration tends to higher in sun leaves, but its temperature sensitivity appears to be similar between between sun and shade leaves.**
+(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. Found 
+higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999).
+(*Martijn has some unpublished data on this that we might include: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
+At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999].
+(*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
+May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
+Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*).
+Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
+
+**VOC production**
+
+**Leaf thermal tolerance (Tcrit/ T50)**
+- define.
+Typical values ~45-50$\circ$C, varying somewhat 
+across latitude/climate (O’sullivan et al. 2017),
+with elevation [@feeley_thermal_2020;  Slot unpublished data],
+with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
+
+**Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
+For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, both measured near ground level (Slot et al. 2019). 
+Similarly, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher $T_{50}$ was observed in more exposed (North-facing) canopy positions (Curtis et al. 2019). 
+We're not aware of any other studies on this.
 
@@
 
-![**Figure 4. Temperature sensitivity of tree growth for understory versus canopy trees.** Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)
+**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology in several ways.** First, vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in understory versus canopy conditions.
+-	*Spring and fall leaf phenology in temperate deciduous forests (Augsburger). 	Mechanism: more moderate conditions (T, wind) in understory (REF).*
+*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes (Condit et al.
+2000).
+One potential mechanism is lower temperature/hydraulic stress in understory*
+
+**Vertical gradients also affect tree growth and survival, and how these respond to climatic variation.**
+There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), and we will not attempt to review those here.
+More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
+Many observations of larger trees suffering more during drought (Bennett et al. 2015).
+One likely mechanism behind this pattern is that the crowns of larger trees exist higher in the vertical profile and therefore are exposed to higher evaporative demand and solar radiation [@mcgregor_tree_nodate].
+Although drought is primarily a hydraulic problem, lack of water --> lower gs --> higher leaf T, so leaves face tradeoffs of water loss vs potentially damaging leafT.
+
+**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
+In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ (Rollinson et al., in review).
+Similarly, our previously unpublished analysis shows that in a broadleaf deciduous forest in Virginia, understory trees in a Mid-Atlantic forest responded more negatively to warmer $T_{air}$ than did canopy trees (Fig.
+**4**; Appendix **S3**).
+This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. **3?**). 
+However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
+**3**).
+It is also possible that competition is accelerated under warmer temperatures.
+
+**Appears to affect community change under warming (Zellweger et al. 2020).**
+
+![**Figure 4. Temperature sensitivity of tree growth for understory versus canopy trees.** Analysis methods described in Appendix S3. Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)
 
@@
 
+- Are patterns (in traits, metabolism, ecology) driven by tree height or exposure?
+
+- CO2 may lead to denser understory (Martijn)
+
 ## Conclusions
+
+## SI files
+
+Appendix S1. Methods for NEON vertical profiles
+
+Appendix S2. Methods for literature review
+
+Appendix S3. Methods for SCBI tree-ring analysis
 

### 2020-07-31 vertical-thermal-review.Rmd
finished transferring content from Word doc to .Rmd/ GitHub issues

@@
 **Many traits vary with height and/or between sun and shade leaves on the same species (Table 1).**
+Here, we focus on traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.
+
+**Table 1.
+Summary of observed variation in thermally-relevant leaf traits with canopy height and/or between sun and shade leaves** ([*GitHub issue #9*](https://github.com/EcoClimLab/vertical-thermal-review/issues/9))
+
+Are traits shaped more by height or light?
+-	 “Height is more important than light in determining leaf morphology in a tropical forest” (Cavaleri et al. 2010)
+- 	(Cavaleri et al. 2008) 
+- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues.
+I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
+
+*Leaf area / characteristic dimension*:
+- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected.
+
+*Max stomatal conductance*
+
+*Thermal time constant (probably not a lot out there)*
+-	Curtis et. al, 2018 – higher in more exposed leaves (heat up slower, cool slower)
+-	Michaeletz has used this in theoretical calculations (Michaeletz et al. 2016 and 2015?
+
+*Carotinoids* 
+-	photoprotective- disseminate heat, acclimate to high T. THese are proportional to irradiance 
+-	antioxidant scavenging function- protect against cellular damage
+- Königer et. al. 1995, and  Matsubara et al. 2009?
+
+*Isoprene production*
+-  Emission as a species trait- Tyeen Taylor, Marielle Smith 
+- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564)
+-  Within species, isporene production scales with light/ T
+
+*Deciduous leaf habit* 
+- (Meakem et al. 2018) and refs therein
 
@@
 3.	Leaf traits 
+-  Michaletz et al have done some interesting work on leaf thermoregulation in relation to various leaf traits.
+I don’t think there is sun vs shade tests, but we could look at how this might translate, or how some of the traits he has calculated might differ between sun and shade leaves.
+For example, he has this thermal time constant parameter that is calculated from leaf mass per area, leaf dry matter content, and some environmental parameters.
+E.g.: DOI: 10.1038/nplants.2016.129, http://dx.doi.org/10.1016/j.tree.2015.09.006
 

### 2020-08-19 vertical-thermal-review.Rmd
meaningless change
another meaningless change
tiny update
added content
Please feel free to edit

@@
 output: word_document
-bibliography: references.bib
+bibliography: references.bib #This can be generated by exporting BibTex file from Zotero
 csl: new-phytologist.csl
@@
 **Forest canopy microclimate buffering is emerging as important for forest ecology in an era of climate change.**
+
+With increase in temperature, we are already seeing that temperate and tropical forest canopies are functioning beyond their photosynthetic threshold (Mau et al.
+2020).
+Other effects of raising temperature on forests are seen in the form of larger trees suffering more during drought (Benett et al.
+2015), increased tree mortality, lower tree growth leading to shorter younger trees (Nate McDowell et al.2020?) and large trees dying twice the rate of small trees (Stovall et al.
+2019).
+Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature (Cornelius Senf et al.
+2018) has shown to increase understory temperatures which raises concern for understory species and forest biodiversity (Zellweger et al.
+2019, Defrenne et al.
+2013).
+
 Forest canopies buffer temperatures and other conditions
-     - @zellweger_forest_2020
+     - @zellweger_forest_2020 (Davis et al. 2019)
 
-We’re seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change.
+We’re seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species (Suggitt et al, Scheffers, Defrenne et al.
+2013).
      - @zellweger_forest_2020 
@@
     - Larger trees suffer more during drought (Bennett et al. 2015) and may be partially influenced by temperature
+
+Understanding thermal sensitivity in vertical tree gradient and height from top of the canopy to understory will provide crucial insights for how raising temperature affects forest ecology and biodiversity.
 

### 2020-08-20 vertical-thermal-review.Rmd
added new line per sentence
@nidhivinod , I find this easier to track, and it all knits together into a continuous paragraph
worked on intro

@@
 **As the climate warms, understanding forest responses to temperature is critical.**
+Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
+This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
+With increase in temperature, we are already seeing that temperate and tropical forest canopies are functioning beyond their photosynthetic threshold (Mau et al. 2020). 
+This, alongside cliamte change-driven disturbances such as drought, may contribute to globally observed phenomena including trends of increasing tree mortality (McDowell et al. 2020; REFS).
 
-**Forest canopy microclimate buffering is emerging as important for forest ecology in an era of climate change.**
-
-With increase in temperature, we are already seeing that temperate and tropical forest canopies are functioning beyond their photosynthetic threshold (Mau et al.
-2020).
-Other effects of raising temperature on forests are seen in the form of larger trees suffering more during drought (Benett et al.
-2015), increased tree mortality, lower tree growth leading to shorter younger trees (Nate McDowell et al.2020?) and large trees dying twice the rate of small trees (Stovall et al.
-2019).
+**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_forest_2020 ; Davis et al.
+2019].**
+Globally, we're seeing larger trees suffering more during drought (Benett et al. 2015; Stovall et al. 2019).
+Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020?). 
 Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature (Cornelius Senf et al.
@@
 
-Forest canopies buffer temperatures and other conditions
-     - @zellweger_forest_2020 (Davis et al. 2019)
-
 We’re seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species (Suggitt et al, Scheffers, Defrenne et al.
@@
      - (Suggitt et al. 2018, Scheffers et. al 2014)
-    - Larger trees suffer more during drought (Bennett et al. 2015) and may be partially influenced by temperature
 

### 2020-08-20 vertical-thermal-review.Rmd
work on intro

@@
 
-**As the climate warms, understanding forest responses to temperature is critical.**
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
 This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
-With increase in temperature, we are already seeing that temperate and tropical forest canopies are functioning beyond their photosynthetic threshold (Mau et al. 2020). 
-This, alongside cliamte change-driven disturbances such as drought, may contribute to globally observed phenomena including trends of increasing tree mortality (McDowell et al. 2020; REFS).
+With increases in temperature, temperate and tropical forest canopies are already functioning beyond their optimal photosynthetic threshold (Mau et al. 2020). 
+This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al. 2020; REFS).
 
-**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_forest_2020 ; Davis et al.
-2019].**
+**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
 Globally, we're seeing larger trees suffering more during drought (Benett et al. 2015; Stovall et al. 2019).

### 2020-08-21 vertical-thermal-review.Rmd
added a couple more lines

@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF).
-This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
+This has increased the severity and frequency of heat waves (Perkins et al.
+2012) which is expected to have profound effects on forest biodiversity and ecosystem function (Feeley et al.
+2020, REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
 With increases in temperature, temperate and tropical forest canopies are already functioning beyond their optimal photosynthetic threshold (Mau et al. 2020). 
@@
 Globally, we're seeing larger trees suffering more during drought (Benett et al. 2015; Stovall et al. 2019).
-Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020?). 
+Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020). 
 Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature (Cornelius Senf et al.
@@
 **Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).**
-Most notably, solar radiation decreases along a vertical profile from the top of the canopy to the forest floor (Fig. **2x**).
+
+Most notably, light conditions vary across vertical tree gradient due to the variation in foliage density that limits light to lower canopy and understory (Mau et.
+al 2018, Fauset et al.
+2017, Niinemets et al.., REFS).
+Solar radiation increases with canopy height (Mau et al.
+2018, M.A.
+Caveleri et al , Niinemets, REFS) and decreases along a vertical profile from the top of the canopy to the forest floor such that the forest floor receives 1% of the above canopy values (Roberts et al.
+1990)
+Radiation is an important variable in determining differences in leaf traits and responses to temperature (Mau et. al 2018, Fauset et al. 2017, Rey Sanchez et al. 2016, Doughty and Gouden 2008)
+(Fig. **2x**).
 - *@bonan_ecological_2016 reviews this and points to appropriate references*
-- (Mau et al. 2018?)
 -	Sunflecks: Leaky et al. 2003?
@@
 - @mcgregor_tree_nodate
-This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999].
+This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999].(Roberts et al. 1990)
 

### 2020-08-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-We’re seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species (Suggitt et al, Scheffers, Defrenne et al.
+We’re seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species (Suggitt et al.
+2018, Scheffers et al.
+2013, Defrenne et al.
 2013).

### 2020-08-21 vertical-thermal-review.Rmd
update manuscript

@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-This has increased the severity and frequency of heat waves (Perkins et al.
-2012) which is expected to have profound effects on forest biodiversity and ecosystem function (Feeley et al.
-2020, REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
+There is also increasing severity and frequency of heat waves (Perkins et al. 2012).
+This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
 With increases in temperature, temperate and tropical forest canopies are already functioning beyond their optimal photosynthetic threshold (Mau et al. 2020). 
@@
 **Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).**
-
 Most notably, light conditions vary across vertical tree gradient due to the variation in foliage density that limits light to lower canopy and understory (Mau et.
 al 2018, Fauset et al.
-2017, Niinemets et al.., REFS).
+2017, Niinemets et al.., REFS) (Fig.
+**2x**).
 Solar radiation increases with canopy height (Mau et al.
@@
 Radiation is an important variable in determining differences in leaf traits and responses to temperature (Mau et. al 2018, Fauset et al. 2017, Rey Sanchez et al. 2016, Doughty and Gouden 2008)
-(Fig. **2x**).
-- *@bonan_ecological_2016 reviews this and points to appropriate references*
+- (*@bonan_ecological_2016 reviews this and points to appropriate references*)
 -	Sunflecks: Leaky et al. 2003?
@@
 **Wind speeds are also higher at the top of the canopy (Fig. **2x**).**
-- *(find some good refs for this/ fill in more specifics)*
-- @mcgregor_tree_nodate
-This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999].(Roberts et al. 1990)
+*(fill in some specifics from the NEON analysis)*
+Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [e.g., REFS; @mcgregor_tree_nodate] and open savannas (Curtis et al.
+2018).
+This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999; Roberts et al. 1990].
 
@@
 Current placeholder shows figures from Campbell&Norman.](leaf_energy_balance/C&N_figures.png)
+
 **Aligning with biophysical expectations, field observations have show that leaf temperatures are influenced by the biophysical environment and leaf traits.**
@@
 
-**When vertical gradients in the biopgysical environment and leaf traits combine to shape leaf energy budgets, $T_{leaf}$ often differs little from $T_{air}$ and between understory and canopy (Fig.
+**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, $T_{leaf}$ often differs little from $T_{air}$ and between understory and canopy (Fig.
 **2?**).**
@@
 Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. 1). 
-However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves.
+However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves (Table 2).
+
+**Table 2.
+Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves** ([*GitHub issue #9*](https://github.com/EcoClimLab/vertical-thermal-review/issues/9))
 

### 2020-08-21 vertical-thermal-review.Rmd
bit of rearranging of content on species-level traits

@@
 *Isoprene production*
--  Emission as a species trait- Tyeen Taylor, Marielle Smith 
-- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564)
--  Within species, isporene production scales with light/ T
-
-*Deciduous leaf habit* 
-- (Meakem et al. 2018) and refs therein
+-  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section)
 
@@
 
+**Isoprene production**
+- high emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
+- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
+-  Within species, isporene production scales with light/ T
+
 **Leaf thermal tolerance (Tcrit/ T50)**
 - define.
-Typical values ~45-50$\circ$C, varying somewhat 
+Typical values ~45-50$^\circ$C, varying somewhat 
 across latitude/climate (O’sullivan et al. 2017),
@@
 One potential mechanism is lower temperature/hydraulic stress in understory*
+- something on isoprene production?
 

### 2020-08-24 scripts/vertical_height_neon.R
Converted to loop to test timing.

@@ -5,6 +5,8 @@
 ## Aug-Sept. 2020
 ######################################################
 library(neonUtilities)
+library(data.table)
+
 library(lubridate)
 library(dplyr)
 library(tidyr)
@@ -23,8 +25,7 @@ library(ggplot2)
 #the function "loadByProduct" will load the data into R and collapse into one df (within a list).
 #it will not download/store anything on the computer, but working with large dfs will run slowly. Hence, it is a good idea to look at the 30min avg first.
 
-dp <- data.table(data = c("2DWSD", "RH", "SAAT", 
-                          "IRBT"),
+dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT"),
                  id = c("DP1.00001.001", "DP1.00098.001",
                         "DP1.00002.001", "DP1.00005.001"),
                  name = c("windSpeedMean", "RHMean",
@@ -53,7 +54,8 @@ sites <- c("BART","BONA","CLBJ","DEJU","DELA","GRSM","GUAN","HARV","JERC",
            "LENO","MLBS","ORNL","OSBS","SCBI","SERC","SJER","SOAP","STEI",
            "TALL","TEAK","TREE","UKFS","UNDE","WREF","YELL")
 
-site_data <- lapply(sites, function(x){
+# site_data <- pbapply::pblapply(sites, function(x){
+for(k in seq(along=sites)){
   vars <- list()
   for (i in seq(along=dp[,name])){
     
@@ -61,10 +63,14 @@ site_data <- lapply(sites, function(x){
     for (j in seq(1:nrow(date))){
       neon_tower <- 
         loadByProduct(dpID=dp[,id][i], 
-                      site=x,
+                      site=k,
                       package="basic", 
                       check.size = FALSE, avg=30,
-                      nCores = 2
+                      nCores = 2,
+                      startdate=paste0(date[,year][j], "-0", 
+                                       date[,month][j]),
+                      enddate = paste0(date[,year][j], "-0", 
+                                       date[,month][j])
                       #(use TRUE outside loop to see 
                       #how big the dowloads are)
                       )
@@ -100,8 +106,9 @@ site_data <- lapply(sites, function(x){
     vars[[i]] <- neon_data_all
   }
   names(vars) <- dp[,name]
-  return(vars)
-})
+  # return(vars)
+}
+)
 names(site_data)
 
 

### 2020-08-24 tables.Rmd
add Table 1

---
title: " "

output: 
  pdf_document:
    fig_caption: yes

header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[table]{font=footnotesize}
                  \captionsetup[figure]{font=footnotesize}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}

bibliography: references.bib #This can be generated by exporting BibTex file from Zotero
csl: new-phytologist.csl

---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
options(knitr.table.format = "latex")
```

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
traits_table <- read.csv("Lit_Review/traits_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(traits_table, caption = "Table 1. (traits review)", format = "latex", booktabs = TRUE,  escape = F)%>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "2cm") %>%
  column_spec(6, width = "4cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
  #NOTE: REFERENCES for this table need to go in the nocite section of header in the main doc!!!
```

### 2020-08-24 vertical-thermal-review.Rmd
update doc
authors
add Table 1

@@
 csl: new-phytologist.csl 
-
+nocite: | 
+  @zellweger_seasonal_2019 
 ---
@@
 
-**Authors (so far):** 
+**Authors (so far, not necessarily in final order):** 
 Nidhi Vinod^1^
 Martijn Slot^2^
+Ian McGregor
+Elsa Ordway
+Marielle Smith
+Tyeen Taylor
 Kristina J. Anderson-Teixeira^1,2^*
@@
 ![**Figure 4. Temperature sensitivity of tree growth for understory versus canopy trees.** Analysis methods described in Appendix S3. Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)
+## Implications for Modeling / Scaling
 

### 2020-08-24 tables.Rmd
started metabolism table
#12

@@
 ```
+
+\newpage
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+metabolism_table <- read.csv("Lit_Review/metabolism_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(metabolism_table, caption = "Table 2. (metabolism review)", format = "latex", booktabs = TRUE,  escape = F)%>%
+  column_spec(1, width = "6cm") %>%
+  column_spec(2, width = "2cm") %>%
+  column_spec(6, width = "4cm") %>%
+  pack_rows("Stomatal conductance", 1, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Photosynthesis", 4, 7, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Respiration", 8, 12, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Thermal tolerance", 13, 13, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+  #NOTE: REFERENCES for this table need to go in the nocite section of header in the main doc!!!
+```

### 2020-08-24 tables.Rmd
update metabolism table
#12

--- 2020-08-24 tables.Rmd
@@
 ```
+
+\newpage
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+metabolism_table <- read.csv("Lit_Review/metabolism_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(metabolism_table, caption = "Table 2. (metabolism review)", format = "latex", booktabs = TRUE,  escape = F)%>%
+  column_spec(1, width = "6cm") %>%
+  column_spec(2, width = "2cm") %>%
+  column_spec(6, width = "4cm") %>%
+  pack_rows("Stomatal conductance", 1, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Photosynthesis", 4, 7, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Respiration", 8, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+  #NOTE: REFERENCES for this table need to go in the nocite section of header in the main doc!!!
+```

### 2020-08-25 scripts/vertical_height_neon.R
Update vertical_height_neon.R

@@ -25,26 +25,29 @@ library(ggplot2)
 #the function "loadByProduct" will load the data into R and collapse into one df (within a list).
 #it will not download/store anything on the computer, but working with large dfs will run slowly. Hence, it is a good idea to look at the 30min avg first.
 
-dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT"),
+dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT", "PARPAR", "SLRNR"),
                  id = c("DP1.00001.001", "DP1.00098.001",
-                        "DP1.00002.001", "DP1.00005.001"),
+                        "DP1.00002.001", "DP1.00005.001",
+                        "DP1.00024.001", "DP1.00023.001"),
                  name = c("windSpeedMean", "RHMean",
-                           "tempSingleMean", "bioTempMean"),
+                          "tempSingleMean", "bioTempMean",
+                          "PARMean", "SWIR"),
                  xlabs = c("Wind speed [m/s]", "RH [%]",
                            "Mean Air Temperature [°C]", 
-                           "Mean Infrared Biological Temperature [°C]"))
+                           "Mean Infrared Biological Temperature [°C]",
+                           "Photosynthetic Active Radiation",
+                           "Shortwave and longwave radiation"))
 
 date <- data.table("year" = c(rep(2016, 2), rep(2017, 2), 
                               rep(2018, 2), rep(2019, 2),
                               rep(2020, 2)),
                    "month" = c(rep(c(1,7), 5)))
 
-date <- data.table("year" = c(rep(2019, 2)),
-                   "month" = c(rep(c(1,7), 1)))
+# date <- data.table("year" = c(rep(2019, 2)),
+#                    "month" = c(rep(c(1,7), 1)))
 
 years <- c("2016", "2017", "2018", "2019", "2020")
 
-years <- c("2019")
 
 #2. Loop through and download data ####
 alldt <- list()
@@ -66,14 +69,14 @@ for(k in seq(along=sites)){
                       site=k,
                       package="basic", 
                       check.size = FALSE, avg=30,
-                      nCores = 2,
-                      startdate=paste0(date[,year][j], "-0", 
-                                       date[,month][j]),
-                      enddate = paste0(date[,year][j], "-0", 
[…]

### 2020-08-25 tables.Rmd

--- 2020-08-24 tables.Rmd
@@
 ```
+
+\newpage
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+metabolism_table <- read.csv("Lit_Review/metabolism_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(metabolism_table, caption = "Table 2. (metabolism review)", format = "latex", booktabs = TRUE,  escape = F)%>%
+  column_spec(1, width = "6cm") %>%
+  column_spec(2, width = "2cm") %>%
+  column_spec(6, width = "4cm") %>%
+  pack_rows("Stomatal conductance", 1, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Photosynthesis", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Respiration", 9, 12, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+  #NOTE: REFERENCES for this table need to go in the nocite section of header in the main doc!!!
+```
+*composite climatic stress variable from canopy temperature, vapour pressure deficit, and relative humidity is higher in lower canopy

### 2020-08-25 tables.Rmd
bit of work on tables & manuscript

@@
 traits_table <- read.csv("Lit_Review/traits_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(traits_table, caption = "Table 1. (traits review)", format = "latex", booktabs = TRUE,  escape = F)%>%
+kable(traits_table, caption = "Table 1.
+Summary of observed variation in thermally-relevant leaf traits with canopy height and/or between sun and shade leaves", format = "latex", booktabs = TRUE,  escape = F)%>%
   column_spec(1, width = "3cm") %>%
@@
 metabolism_table <- read.csv("Lit_Review/metabolism_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(metabolism_table, caption = "Table 2. (metabolism review)", format = "latex", booktabs = TRUE,  escape = F)%>%
+kable(metabolism_table, caption = "Table 2.
+Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves", format = "latex", booktabs = TRUE,  escape = F)%>%
   column_spec(1, width = "6cm") %>%
@@
   pack_rows("Photosynthesis", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Respiration", 9, 12, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Respiration", 9, 14, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("VOC production", 15, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-09-02 vertical-thermal-review.Rmd

@@
 csl: new-phytologist.csl 
-nocite: | 
-  @zellweger_seasonal_2019 
 ---
@@
 
-**Table 1.
-Summary of observed variation in thermally-relevant leaf traits with canopy height and/or between sun and shade leaves** ([*GitHub issue #9*](https://github.com/EcoClimLab/vertical-thermal-review/issues/9))
+![(*For latest version of table, see tables.pdf.*)](table1.png)
+```{r eval=FALSE, echo=FALSE}
+# refs in this table need to go in the nocite field below
+```
+---
+nocite: | 
+  @zellweger_seasonal_2019 
+---
 
@@
 
-**Table 2.
-Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves** ([*GitHub issue #9*](https://github.com/EcoClimLab/vertical-thermal-review/issues/9))
+![(*For latest version of table, see tables.pdf.*)](table2.png)
+```{r eval=FALSE, echo=FALSE}
+# refs in this table need to go in the nocite field below
+```
+---
+nocite: | 
+  @zellweger_seasonal_2019 
+---
+
+*Stomatal conductance*
+
+*Photosynthesis*
 
@@
 
-**Similar to photosynthesis, respiration tends to higher in sun leaves, but its temperature sensitivity appears to be similar between between sun and shade leaves.**
-(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. Found 
-higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999).
-(*Martijn has some unpublished data on this that we might include: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
-At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999].
-(*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
-May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
-Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*).
-Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
-
-**VOC production**
-
-**Isoprene production**
-- high emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
-- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
--  Within species, isporene production scales with light/ T
-
 **Leaf thermal tolerance (Tcrit/ T50)**
@@
 We're not aware of any other studies on this.
+
+*Respiration*
+
+**Similar to photosynthesis, respiration tends to higher in sun leaves, but its temperature sensitivity appears to be similar between between sun and shade leaves.**
+(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. 
+Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999).
+(*Martijn has some unpublished data on this that we might include: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
+At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999].
+(*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
+May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
+Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*).
+Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
+
+*VOC production*
+
+**Isoprene production**
+- high emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
+- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
+-  Within species, isporene production scales with light/ T
 
@@
 ![**Figure 4. Temperature sensitivity of tree growth for understory versus canopy trees.** Analysis methods described in Appendix S3. Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)
+
 ## Implications for Modeling / Scaling

### 2020-09-02 tables.Rmd
Revert "Merge branch 'master' of https://github.com/EcoClimLab/vertical-thermal-review"
This reverts commit b50f5c0414c2a90b5c0724601f1d8a691663771a, reversing
changes made to ebb3b1b1ef329053f33cea450c710318aacc422e.

@@
   column_spec(6, width = "4cm") %>%
+  pack_rows("Leaf anatomy and morphological traits", 1, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Leaf physiology and biochemical traits", 11, 17, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-09-02 NEON_height_profiles/scripts/new.R

@@ -4,15 +4,19 @@
 # R version 3.5.3 - First created August 2020
 ## Aug-Sept. 2020
 ######################################################
+
 #install needed packages
 library(data.table)
 library(readxl)
 
-#
 #1. Download and save neon data ####
 ## this is filtered to only the columns needed for plots 
 ## (mean measurements per each 30 min observation) in order to not have massive files.
 source("scripts/get_neon_data.R")
+
+data <- fread("NEON_height_profiles/forested_NEON_sites.csv")
+data <- data[include==1, ]
+sites <- data[,site]
 
 pbapply::pblapply(sites, function(st){
   full_data <- get_NEON(x=st)
@@ -20,20 +24,7 @@ pbapply::pblapply(sites, function(st){
 })
 
 #2. Organize data for plotting ####
-
-data <- fread("forested_NEON_sites.csv")
-data <- data[include==1, ]
-sites <- data[,site]
-
-#bring in normalized height
-meta <- read_excel("data/site_data/TIS site metadata_20190403_forNOAA.xlsx", sheet=1)
-meta <- data.table(meta)
-meta <- meta[SITE %in% data[,site], 
-             ][,DistZaxsCnpy := as.numeric(DistZaxsCnpy)]
-
-load("data/neon_rdata/SCBItest.Rdata")
-
-meta_site <- meta[SITE == j, ]
+load("neon_rdata/SCBItest.Rdata")
 
 dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT", "PARPAR", "SLRNR"),
                  id = c("DP1.00001.001", "DP1.00098.001",
@@ -65,113 +56,15 @@ for(i in 1:length(dp[,name])){
            colnames(var))
   }
   
-  #get daily min/max by vertical position
-  test <- var[, day := as.Date(startDateTime)
-             ][, .(day_max = max(get(dp[,name][i]), na.rm=TRUE),
-                  day_min = min(get(dp[,name][i]), na.rm=TRUE)),
-               by = .(day, verticalPosition)
-               ][, `:=` (day_max = ifelse(day_max %in% c(-Inf, Inf), NA, day_max),
-                         day_min = ifelse(day_min %in% c(-Inf, Inf), NA, day_min))
-                 ][, `:=` (month_num = month(day),
-                           yr = year(day))
-                   ]
+  var <- var[, day := substr(startDateTime, 1, 10)]
   
-  stat <- test[, `:=` (mmax = mean(test_max, na.rm=TRUE),
[…]

### 2020-09-03 tables.Rmd
Revert "Revert "Merge branch 'master' of https://github.com/EcoClimLab/vertical-thermal-review""
This reverts commit 4ddf503f85c55cd59235728fcdd653f54843201d.

--- 2020-08-25 tables.Rmd
@@
 Summary of observed variation in thermally-relevant leaf traits with canopy height and/or between sun and shade leaves", format = "latex", booktabs = TRUE,  escape = F)%>%
-  column_spec(1, width = "3cm") %>%
+  column_spec(1, width = "7cm") %>%
   column_spec(2, width = "2cm") %>%
+  column_spec(3, width = "2cm") %>%
+  column_spec(5, width = "3cm") %>%
   column_spec(6, width = "4cm") %>%
+  pack_rows("Leaf anatomy and morphological traits", 1, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf physiology and biochemical traits", 11, 17, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-09-03 NEON_height_profiles/scripts/new.R

@@ -4,19 +4,15 @@
 # R version 3.5.3 - First created August 2020
 ## Aug-Sept. 2020
 ######################################################
-
 #install needed packages
 library(data.table)
 library(readxl)
 
+#
 #1. Download and save neon data ####
 ## this is filtered to only the columns needed for plots 
 ## (mean measurements per each 30 min observation) in order to not have massive files.
 source("scripts/get_neon_data.R")
-
-data <- fread("NEON_height_profiles/forested_NEON_sites.csv")
-data <- data[include==1, ]
-sites <- data[,site]
 
 pbapply::pblapply(sites, function(st){
   full_data <- get_NEON(x=st)
@@ -24,7 +20,20 @@ pbapply::pblapply(sites, function(st){
 })
 
 #2. Organize data for plotting ####
-load("neon_rdata/SCBItest.Rdata")
+
+data <- fread("forested_NEON_sites.csv")
+data <- data[include==1, ]
+sites <- data[,site]
+
+#bring in normalized height
+meta <- read_excel("data/site_data/TIS site metadata_20190403_forNOAA.xlsx", sheet=1)
+meta <- data.table(meta)
+meta <- meta[SITE %in% data[,site], 
+             ][,DistZaxsCnpy := as.numeric(DistZaxsCnpy)]
+
+load("data/neon_rdata/SCBItest.Rdata")
+
+meta_site <- meta[SITE == j, ]
 
 dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT", "PARPAR", "SLRNR"),
                  id = c("DP1.00001.001", "DP1.00098.001",
@@ -56,15 +65,113 @@ for(i in 1:length(dp[,name])){
            colnames(var))
   }
   
-  var <- var[, day := substr(startDateTime, 1, 10)]
+  #get daily min/max by vertical position
+  test <- var[, day := as.Date(startDateTime)
+             ][, .(day_max = max(get(dp[,name][i]), na.rm=TRUE),
+                  day_min = min(get(dp[,name][i]), na.rm=TRUE)),
+               by = .(day, verticalPosition)
+               ][, `:=` (day_max = ifelse(day_max %in% c(-Inf, Inf), NA, day_max),
+                         day_min = ifelse(day_min %in% c(-Inf, Inf), NA, day_min))
+                 ][, `:=` (month_num = month(day),
+                           yr = year(day))
+                   ]
   
-  #bring in normalized height
[…]

### 2020-09-03 vertical-thermal-review.Rmd
Revert "Revert "Revert "Merge branch 'master' of https://github.com/EcoClimLab/vertical-thermal-review"""
This reverts commit 046938c584cecbc377ce2656c32446dedca962b1.

--- 2020-08-24 vertical-thermal-review.Rmd
@@
 
-**Table 1.
-Summary of observed variation in thermally-relevant leaf traits with canopy height and/or between sun and shade leaves** ([*GitHub issue #9*](https://github.com/EcoClimLab/vertical-thermal-review/issues/9))
+![(*For latest version of table, see tables.pdf.*)](table1.png)
 
@@
 
-**Table 2.
-Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves** ([*GitHub issue #9*](https://github.com/EcoClimLab/vertical-thermal-review/issues/9))
+![(*For latest version of table, see tables.pdf.*)](table2.png)
+
+*Stomatal conductance*
+
+*Photosynthesis*
 

### 2020-09-03 NEON_height_profiles/scripts/get_neon_data.R

@@ -45,13 +45,9 @@ date <- data.table("year" = c(rep(2015, 2), rep(2016, 2), rep(2017, 2),
 # alldt <- list()
 # plotlist <- list()
 
-data <- fread("NEON_height_profiles/forested_NEON_sites.csv")
-data <- data[include==1, ]
-sites <- data[,site]
-
-# sites <- c("BART","BONA","CLBJ","DEJU","DELA","GRSM","GUAN","HARV","JERC",
-#            "LENO","MLBS","ORNL","OSBS","SCBI","SERC","SJER","SOAP","STEI",
-#            "TALL","TEAK","TREE","UKFS","UNDE","WREF","YELL")
+sites <- c("BART","BONA","CLBJ","DEJU","DELA","GRSM","GUAN","HARV","JERC",
+           "LENO","MLBS","ORNL","OSBS","SCBI","SERC","SJER","SOAP","STEI",
+           "TALL","TEAK","TREE","UKFS","UNDE","WREF","YELL")
 
 get_NEON <- function(x="SCBI"){
   # full <- list()

### 2020-09-08 vertical-thermal-review.Rmd
changed a little bit

@@
 **Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).**
-Most notably, light conditions vary across vertical tree gradient due to the variation in foliage density that limits light to lower canopy and understory (Mau et.
-al 2018, Fauset et al.
-2017, Niinemets et al.., REFS) (Fig.
+
+Most notably, availability of light decreases along a verticle profile from top of the canopy to the forest floor due to variation in foliage density,and increases with tree height (Mau et.
+al 2018, Poorter et al.
+2019, Niinemets et al.
+2015) (Fig.
 **2x**).
-Solar radiation increases with canopy height (Mau et al.
-2018, M.A.
-Caveleri et al , Niinemets, REFS) and decreases along a vertical profile from the top of the canopy to the forest floor such that the forest floor receives 1% of the above canopy values (Roberts et al.
-1990)
-Radiation is an important variable in determining differences in leaf traits and responses to temperature (Mau et. al 2018, Fauset et al. 2017, Rey Sanchez et al. 2016, Doughty and Gouden 2008)
+Light is more available above the canopy compared to the forest floor which receives as little as 1% of the above canopy values (Roberts et al. 1990)
+This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.(Niinemets, 2010, Sack et al. 2016)
 - (*@bonan_ecological_2016 reviews this and points to appropriate references*)

### 2020-09-09 tables.Rmd
fixing up tables

--- 2020-08-25 tables.Rmd
@@
   column_spec(6, width = "4cm") %>%
+  pack_rows("Leaf anatomy and morphological traits", 1, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf biochemical and physiological traits", 11, 17, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
@@
   column_spec(6, width = "4cm") %>%
-  pack_rows("Stomatal conductance", 1, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Photosynthesis", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Respiration", 9, 14, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("VOC production", 15, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Stomatal conductance", 1, 7, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Photosynthesis", 8, 20, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Respiration", 21, 27, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("VOC production", 28, 29, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-09-09 tables.Rmd
one more fix

@@
   pack_rows("Photosynthesis", 8, 20, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Respiration", 21, 27, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("VOC production", 28, 29, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Respiration", 21, 28, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("VOC production", 29, 30, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-09-09 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 csl: new-phytologist.csl
-nocite: | 
-  @zellweger_seasonal_2019 
+
 ---
@@
 **Authors (so far, not necessarily in final order):** 
-Nidhi Vinod^1^
-Martijn Slot^2^
-Ian McGregor
-Elsa Ordway
-Marielle Smith
-Tyeen Taylor
+Nidhi Vinod^1^,
+Martijn Slot^2^,
+Ian McGregor,
+Elsa Ordway,
+Marielle Smith,
+Tyeen Taylor,
 Kristina J. Anderson-Teixeira^1,2^*
@@
 
+```{r eval=FALSE, echo=FALSE}
+# refs in this table need to go in the nocite field below
+```
+---
+nocite: | 
+  @zellweger_seasonal_2019 
+---
+
 Are traits shaped more by height or light?
@@
 ![(*For latest version of table, see tables.pdf.*)](table2.png)
+```{r eval=FALSE, echo=FALSE}
+# refs in this table need to go in the nocite field below
+```
+---
+nocite: | 
+  @zellweger_seasonal_2019 
+---
 

### 2020-09-09 vertical-thermal-review.Rmd
very rough abstract draft

@@
 
-## Abstract
+```{r eval=FALSE, echo=FALSE}
+# based on Helene's Tansley review, and regular NewPhyt guidelines, I think the limit for New Phyt is 200 words. 
+```
+
+## Summary
+
+*(currently way too long for NewPhyt)*
+
+* Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across trees of different size.
+* Microclimatic conditions and foliar traits vary across the vertical canopy gradient, driving gradients in leaf temperatures and metabolism, which in turn affect whole-tree growth and stress responses.
+* While some recent studies have suggested that understory trees are less vulnerable to temperature extremes driven by climatic warming, these gradients remain poorly understood.
+* Here, we synthesize evidence as to how environmental conditions, foliar traits, and plant function vary across gradients of height within the canopy or light environment.
+* We find that *(there is strong directional variation in microenvironment)*.
+* Correspondingly, foliar traits relevant to shaping leaf temperatures and metabolism vary strongly across height or light gradients...
+* The resulting leaf temperatures generally track air temperature and do not vary dramatically across the vertical gradient, except for when hydraulic limitations impede leaf thermal regulation via evaporative cooling.
+In these cases, leaf temperatures of sun-exposed canopy leaves can become significantly elevated above air temperatures and the temperatures of understory leaves.
+* Corresponding, while leaf metabolism generally increases with light and height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
+* The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, tree-ring evidence is emerging that understory trees experience relatively greater stress from hot temperatures.
+* The implications for modeling are ...
+* Additional research will be important for understanding our spotty understanding, particulalry of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
+* In the meantime, our results imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.
 
@@
 v.	What are the implications of these patterns for the ecology and climate change responses of canopy versus understory trees?
+
+vi. What are the implications for modeling/ scaling?
 

### 2020-09-09 vertical-thermal-review.Rmd
condensed draft abstract

--- 2020-09-09 vertical-thermal-review.Rmd
@@
 
-## Abstract
+```{r eval=FALSE, echo=FALSE}
+# based on Helene's Tansley review, and regular NewPhyt guidelines, I think the limit for New Phyt is 200 words. 
+```
+
+## Summary
+
+*(currently way too long for NewPhyt)*
+
+Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across trees of different size.
+Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical canopy gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and stress responses.
+We find that there is strong directional variation in microclimate, with canopy trees experiencing greater evaporative demand than understory trees.
+Foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism vary strongly across height or light gradients.
+The resulting $T_{leaf}$'s generally track air temperature ($T_{air}$) and do not vary dramatically across the vertical gradient, except when hydraulic limitations impede leaf evaporative cooling and $T_{leaf}$'s of sun-exposed canopy leaves become significantly elevated above both $T_{air}$ and understory $T_{leaf}$.
+While leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
+The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, tree-ring evidence is emerging that understory trees experience relatively greater stress from high temperatures.
+Additional research will be important for understanding our spotty understanding, particularly of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
+In the meantime, our results imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.
 

### 2020-09-09 vertical-thermal-review.Rmd
improved abstract

--- 2020-09-09 vertical-thermal-review.Rmd
@@
 
-## Abstract
+```{r eval=FALSE, echo=FALSE}
+# based on Helene's Tansley review, and regular NewPhyt guidelines, I think the limit for New Phyt is 200 words. 
+```
+
+## Summary
+
+*(currently too long for NewPhyt)*
+
+Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
+Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
+Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves.
+However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
+Resulting $T_{leaf}$'s generally track air temperature ($T_{air}$) and do not vary dramatically across the vertical gradient, except when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become significantly elevated above both $T_{air}$ and understory $T_{leaf}$.
+Correspondingly, while leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
+The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
+Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
+In the meantime, our findings imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.
 

### 2020-09-10 vertical-thermal-review.Rmd
update title

@@
 
-**Title:** Thermal sensitivity of canopy versus understory leaves: patterns, mechanisms, and ecological implications
+**Title:** Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications
 
@@
 Martijn Slot^2^,
-Ian McGregor,
+Ian McGregor^3^,
 Elsa Ordway,
@@
 2. Smithsonian Tropical Research Institute; Panama, Republic of Panama
+
+3. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
 

### 2020-09-10 tables.Rmd
cosmetic cleanup of tables

@@
 Summary of observed variation in thermally-relevant leaf traits with canopy height and/or between sun and shade leaves", format = "latex", booktabs = TRUE,  escape = F)%>%
-  column_spec(1, width = "3cm") %>%
+  column_spec(1, width = "4cm") %>%
   column_spec(2, width = "2cm") %>%

### 2020-09-10 leaf_energy_balance/Leaf_to_air_temp_predictions_V7.R
Add files via upload
Marielle's code + data inputs for predicting Tleaf from Tair using tealeaves (Muir 2019) for B2 temperature paper (Smith et al. 2020, Nature Plants)

@@ -1,0 +1,274 @@
+### Code for modelling leaf temps for B2 paper
+###
+
+#--Using tealeaves package by Chris Muir
+# https://github.com/cdmuir/tealeaves/blob/master/README.md
+# https://www.biorxiv.org/content/biorxiv/early/2019/01/24/529487.full.pdf 
+
+#install.packages("devtools")
+#library(devtools)
+#devtools::install_github("cdmuir/tealeaves", force=TRUE)
+#library(magrittr)
+#library(tealeaves)
+require(tealeaves)
+require(ggplot2)
+
+
+#####################--USE FULL DISTRIBUTION OF ENVIRONMENTAL PARAMETERS--#################################
+
+# Import the datasets - just for Topt values for K34 and B2 (28 and 38 degrees, respectively)
+dir <- "C:/Users/Marielle/Documents/PhD/RESEARCH PROJECTS/B2 NEE paper"
+setwd(dir)
+load(file="Tleaf_modelling_params_FULL") 
+# b2.dat and K34.dat; note: par column for K34 has been converted to short-wave incoming radiation
+
+
+############################### FUNCTIONS FOR TLEAF TEMP ESTIMATIONS
+
+#--Configure B2 dataset
+B2dat <- b2.dat
+B2dat$tair <- 38 + 273.15
+B2dat$pressure <- 86
+B2dat$rh <- B2dat$rh/100
+## NOTE: make sure order of columns is: par, rh, wind, tair, pressure
+
+#--Configure K34 dataset
+K34dat <- K34.dat
+K34dat$tair <- 28 + 273.15
+K34dat$pressure <- 99
+K34dat$rh <- K34dat$rh/100
+## NOTE: make sure order of columns is: par, rh, wind, tair, pressure
+
+#--Configure a test dataset -- data from Doughty & Goulden 2008
+DGdat <- data.frame(par=c(250,1500), rh=0.7, wind=c(1,1), tair=c(28.5+273.15,29.5+273.15), pressure=99, leaf.size=0.108, gs=c(1.2,1.05), tleaf.meas=c(30,37.5), scenario=c(2,1))
+DGdat$par <- DGdat$par*0.5 # change PAR to short-wave incoming radiation
+
+
+#--B2 dataset -- with same DISTRIBUTION of PAR as K34 (but same number of values as B2 data)
[…]

### 2020-09-10 vertical-thermal-review.Rmd
worked on references
@NidhiVinod , I switched to using Zotero_references.bib. I also regenerated it from Zotero because for some reason it had a different convention for article names.

@@
 output: word_document
-bibliography: references.bib #This can be generated by exporting BibTex file from Zotero
+bibliography: Zotero_references.bib #This can be generated by exporting BibTex file from Zotero
 csl: new-phytologist.csl
@@
 2018).
-This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999; Roberts et al. 1990].
+This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999; @roberts_stomatal_1990].
 
@@
 Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
-Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama (Rey-Sánchez et al. 2016) and coastal Brazil (Fauset et al. 2018).
-However, similar maximum temperatures have been observed during the dry season in Panama (Rey-Sánchez et al.
-2016) and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018].
+However, similar maximum temperatures have been observed during the dry season in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
 Minimum $T_{air}$ is also buffered by forest canopies under some conditions. 
 Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
-However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama (Rey-Sánchez et al.
-2016) and coastal Brazil (Fauset et al.
-2018), as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
 *(Add something about risk of freeze.)*
@@
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$
-- $VPD$ is lower in understory of Atlantic forest during wet season (Fauset et al. 2018)
+- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018]
 This means that canopy leaves tend to be exposed to higher *evaporative demand*.
@@
 First, buffering increases with canopy cover.
-- greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
-- greater cover -->  lower max T [@zellweger_seasonal_2019]
-- [@thom_effects_2020]
+* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
+* greater cover -->  lower max T [@zellweger_seasonal_2019]
+* [@thom_effects_2020]
 *Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
@@
 The strength of buffering also varies with respect to geographic and climatic factors.
-- Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
-- [@davis_microclimatic_2019]
+* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
+* [@davis_microclimatic_2019]
 
@@
 ```{r eval=FALSE, echo=FALSE}
-# refs in this table need to go in the nocite field below
+# refs in this table need to go in the nocite field below (easiest in alphabetical order)
 ```
@@
 nocite: | 
-  @zellweger_seasonal_2019 
+  @mau_temperate_2018
 ---
@@
 - Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected.
-
-*Max stomatal conductance*
 
@@
 nocite: | 
-  @zellweger_seasonal_2019 
+  @bolstad_foliar_1999; @roberts_stomatal_1990; @slot_photosynthetic_2019 
 ---

### 2020-09-10 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-We’re seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species (Suggitt et al.
+We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species (Suggitt et al.
 2018, Scheffers et al.

### 2020-09-17 vertical-thermal-review.Rmd
incorporating Marielle's edits

@@
 Elsa Ordway,
-Marielle Smith,
+Marielle N. Smith^3^,
 Tyeen Taylor,
@@
 3. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
+
+4. Department of Forestry, Michigan State University, East Lansing, MI 48824, USA
 
@@
 This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
-With increases in temperature, temperate and tropical forest canopies are already functioning beyond their optimal photosynthetic threshold (Mau et al. 2020). 
-This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al. 2020; REFS).
+With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold (Mau et al. 2020, Huang et al. 2019). 
+This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al.
+2020; Allen et al.
+2015, Brienen et al.
+2015).
 
 **Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
-Globally, we're seeing larger trees suffering more during drought (Benett et al. 2015; Stovall et al. 2019).
+Globally, we're seeing larger trees suffering more during drought (Bennett et al. 2015; Stovall et al. 2019).
 Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020). 
@@
 
-Understanding thermal sensitivity in vertical tree gradient and height from top of the canopy to understory will provide crucial insights for how raising temperature affects forest ecology and biodiversity.
+Understanding thermal sensitivity along the vertical canopy profile, from the understory to the top of the canopy will provide crucial insights for how raising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
 
@@
 *Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests.
-However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), we also review studies focused on exposure gradients.*
+However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), and the implications [or utility] for ecosystem models we also review studies focused on exposure gradients.*
 

### 2020-09-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 There is also increasing severity and frequency of heat waves (Perkins et al. 2012).
-This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
-With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold (Mau et al. 2020, Huang et al. 2019). 
+This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
+With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018,Huang et al. 2019]. 
 This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al.
@@
 **Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
-Globally, we're seeing larger trees suffering more during drought (Bennett et al. 2015; Stovall et al. 2019).
+Globally, we're seeing larger trees suffering more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
 Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020). 
-Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature (Cornelius Senf et al.
-2018) has shown to increase understory temperatures which raises concern for understory species and forest biodiversity (Zellweger et al.
-2019, Defrenne et al.
-2013).
+Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature [[@senfCanopyMortalityHas2018] has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellwegerSeasonalDriversUnderstorey2019, @defrenneMicroclimateModeratesPlant2013]
 
-We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species (Suggitt et al.
-2018, Scheffers et al.
-2013, Defrenne et al.
-2013).
-     - @zellweger_forest_2020 
-     - (Suggitt et al. 2018, Scheffers et. al 2014)
+We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellwegerForestMicroclimateDynamics2020, @suggittExtinctionRiskClimate2018, @scheffersMicrohabitatsReduceAnimal2013,@defrenneMicroclimateModeratesPlant2013]
 
@@
 
+Maybe we should start this section with a brief definition of what we mean by "biophysical"?
+E.g.
+the physical structure of the forest canopy (from understory to the canopy top) and associated microclimates, which are in large part determined by the structure of the forest itself.
+]
+
 **Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).**
 
-Most notably, availability of light decreases along a verticle profile from top of the canopy to the forest floor due to variation in foliage density,and increases with tree height (Mau et.
-al 2018, Poorter et al.
-2019, Niinemets et al.
-2015) (Fig.
-**2x**).
-Light is more available above the canopy compared to the forest floor which receives as little as 1% of the above canopy values (Roberts et al. 1990)
-This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.(Niinemets, 2010, Sack et al. 2016)
+Most notably, light conditions vary  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fausetDifferencesLeafThermoregulation2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019]
+(Fig.
+**2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990, Chazdon & Fetcher 1984; Parker et al.
+2019]
+In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions (Chazdon & Fetcher 1984; Tymen et al.
+2017 ; Smith et al.
+2019).
+This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.[@fausetDifferencesLeafThermoregulation2018, @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006, Doughty and Gouden 2008]
+
+In addition, the vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; Smith et al.
+2019 ; Parker et al.
+2019) and following disturbance (e.g., Parker et al.
+2002 ; Almeida et al.
+2016 ; Stark et al.
+2020 in press).
 - (*@bonan_ecological_2016 reviews this and points to appropriate references*)
--	Sunflecks: Leaky et al. 2003?
+-	Sunflecks: Leaky et al.
+2003--from Marielle, "This is another, higher resolution axis of the variability of microenvironments that is probably quite important to address.
+(Also see the papers by Robin Chazdon on temporal variability of light environments).
+Maybe we will want to add these sentences (above) and something about sunflecks to a separate section that specifically addresses sources of variation in forest structure & microenvironments (e.g., vertical, horizontal, seasonal, diurnal, and with disturbance)?"
 
@@
 Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
-Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018].
+Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018, Tymen et al.
+2017]
+"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy"
 However, similar maximum temperatures have been observed during the dry season in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
@@
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$
-- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018]
+- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018, Tymen et al. 2017]
 This means that canopy leaves tend to be exposed to higher *evaporative demand*.

### 2020-09-17 vertical-thermal-review.Rmd
added more references

@@
 * greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
-* greater cover -->  lower max T [@zellweger_seasonal_2019]
-* [@thom_effects_2020]
+* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
+* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015]
 *Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
 Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
-SCA species increase T buffering (Zelllweger et al. 2019) [@zellweger_seasonal_2019]
+Species that have higher shade casting ability increase T buffering (Zelllweger et al. 2019) [@zellweger_seasonal_2019]
 The strength of buffering also varies with respect to geographic and climatic factors.
@@
 ## Trait variation
+
+[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
+Chen et al.
+2020 is a good paper to help frame this .
+They make an important point in that paper that we may also want to make here: that the LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy - hence the importance of this review.
+Keenan & Niinemets 2016  will also be helpful in making this point.]--M N.
+Smith
 
@@
 - Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected. 
+- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015).
+Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
 
@@
 
-**Photosynthesis is generally higher in sun leaves-- a fact that is well-established (REFS) and observed in numerous field studies.**
+**Photosynthesis is generally higher in sun leaves-- a fact that is well-established (Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020) and observed in numerous field studies.**
 This is primarily driven by the greater light available to sun leaves.
@@
 The vertical gradient in photosythetic rates is also influenced by...
+.[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N.
+Smith
 

### 2020-09-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**Similar to photosynthesis, respiration tends to higher in sun leaves, but its temperature sensitivity appears to be similar between between sun and shade leaves.**
+**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
 (Bolstad et al. 1999) did a study at Coweeta, including elev gradient. 
@@
 ## Implications for Modeling / Scaling
+E.g.
+see de Pury & Farquhar, 1997 : though likely an improvement over 'big-leaf' models, models that separate the canopy into sunlit and shaded portions (sun/shade models) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g.
+seasonal shifts in vertical leaf area distributions, Smith et al.
+2019) and functions (e.g.
+thermal responses, as we present in this paper).
+Hence, multi-layered ecosystem models may be necessary for accurately projecting future forest function.
 
@@
 - CO2 may lead to denser understory (Martijn)
+
+- Influence of increasing temperatures on species compositions and feedbacks to future forest microclimates and function?
 

### 2020-09-18 vertical-thermal-review.Rmd
fixed zotero (needed pandoc installation)
small fixes

@@
 This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
-With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018,Huang et al. 2019]. 
+With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, Huang et al. 2019]. 
 This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al.
@@
 Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020). 
-Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature [[@senfCanopyMortalityHas2018] has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellwegerSeasonalDriversUnderstorey2019, @defrenneMicroclimateModeratesPlant2013]
+Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature [[@senfCanopyMortalityHas2018a] has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
 
-We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellwegerForestMicroclimateDynamics2020, @suggittExtinctionRiskClimate2018, @scheffersMicrohabitatsReduceAnimal2013,@defrenneMicroclimateModeratesPlant2013]
+We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellweger_forest_2020, @suggittExtinctionRiskClimate2018, @scheffersMicrohabitatsReduceAnimal2013,@defrenneMicroclimateModeratesPlant2013]
 
@@
 
-Most notably, light conditions vary  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fausetDifferencesLeafThermoregulation2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019]
+Most notably, light conditions vary  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
 (Fig.
-**2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990, Chazdon & Fetcher 1984; Parker et al.
+**2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a, Chazdon & Fetcher 1984; Parker et al.
 2019]

### 2020-09-18 vertical-thermal-review.Rmd
added mcdowell's citation

@@
 With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, Huang et al. 2019]. 
-This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al.
-2020; Allen et al.
-2015, Brienen et al.
+This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015).

### 2020-09-18 vertical-thermal-review.Rmd
.

--- 2020-09-18 vertical-thermal-review.Rmd
@@
 With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, Huang et al. 2019]. 
-This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al.
-2020; Allen et al.
-2015, Brienen et al.
-2015).
+This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+2015].
 
@@
 Globally, we're seeing larger trees suffering more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
-Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020). 
-Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature [[@senfCanopyMortalityHas2018a] has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
+Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020). 
+Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature [@senfCanopyMortalityHas2018a] has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
 
@@
 ## Implications for Modeling / Scaling
+
 E.g.

### 2020-09-21 vertical-thermal-review.Rmd
updated

@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves (Perkins et al. 2012).
+There is also increasing severity and frequency of heat waves (@perkinsIncreasingFrequencyIntensity2012).
 This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
-With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, Huang et al. 2019]. 
+With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].
 This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al.
@@
 Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature [@senfCanopyMortalityHas2018a] has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
+
+The structure and microclimates of global forests are changing due to unprecedented disturbance rates.
+Loss of canopy cover due to forest disturbances (e.g., deforestation, fire, and drought) is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al.
+2020; Stark et al.
+2020 ).
+Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al.
+2020).
 

### 2020-09-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2020-09-18 vertical-thermal-review.Rmd
@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves (Perkins et al. 2012).
+There is also increasing severity and frequency of heat waves (@perkinsIncreasingFrequencyIntensity2012).
 This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
-With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, Huang et al. 2019]. 
-This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+These changes, alongside climate change-driven disturbances such as drought, deforestation and fire, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015].
@@
 **Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
-Globally, we're seeing larger trees suffering more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
+Larger trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. (Nakamura et al. 2017, Ozanne et al. 2003)  
+With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
 Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).
-Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature [@senfCanopyMortalityHas2018a] has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
+
+**The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
+Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Additionally, loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al.
+2020; Stark et al.
+2020 ), this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
+
+(Marielle--"Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al.
+2020)".)
 
@@
 
-Understanding thermal sensitivity along the vertical canopy profile, from the understory to the top of the canopy will provide crucial insights for how raising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
+With shifting forest dynamics, it is becoming more important than ever to understand thermal sensitivity along the vertical canopy profile, from the understory to the top of the canopy, this will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
 

### 2020-09-22 vertical-thermal-review.Rmd
added Lawren Sack to the author list

@@
 Tyeen Taylor,
+Lawren Sack,
 Kristina J. Anderson-Teixeira^1,2^*
@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves (@perkinsIncreasingFrequencyIntensity2012).
+There is also increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012].
 This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
@@
 **Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
-Larger trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. (Nakamura et al. 2017, Ozanne et al. 2003)  
+Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. (Nakamura et al. 2017, Ozanne et al. 2003)  
 With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
@@
 **The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
-Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Additionally, loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al.
+Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al.
 2020; Stark et al.

### 2020-09-24 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ## Trait variation
+
+Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b].
+Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light.
+Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
+2020], leaf thermoregular and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016, @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al.
+2020, REFS)
 

### 2020-09-24 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 -	antioxidant scavenging function- protect against cellular damage
-- Königer et. al. 1995, and  Matsubara et al. 2009?
+- Koniger et. al. 1995, and  Matsubara et al. 2009?
 
@@
 $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
-While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.
+While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.[@fauset_differences_2018]
 Large leaves can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation with stomata closed.  

### 2020-09-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
-Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate. (Nakamura et al. 2017, Ozanne et al. 2003)  
+Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate.
+[@nakamuraForestsTheirCanopies2017, @ozanneBiodiversityMeetsAtmosphere2003]
 With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
@@
 **The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
-Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components (Zellweger et al.
-2020; Stark et al.
-2020 ), this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
+Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
 
@@
 
-Most notably, light conditions vary  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
+Most notably, light conditions, specifically photosynthetically active radiation (PAR) varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
 (Fig.
@@
 2019]
-In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions (Chazdon & Fetcher 1984; Tymen et al.
-2017 ; Smith et al.
-2019).
-This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.[@fausetDifferencesLeafThermoregulation2018, @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006, Doughty and Gouden 2008]
+In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984, @tymenQuantifyingMicroenvironmentalVariation2017, @smithSeasonalDroughtrelatedChanges2019].
+This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.[@fausetDifferencesLeafThermoregulation2018, @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006, @doughtyAreTropicalForests2008]
 
@@
 2018).
-This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999; @roberts_stomatal_1990].
+This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990].
 

### 2020-09-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 * [@davis_microclimatic_2019]
+However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017,  @martinCovarianceSunShade2020a]
 
@@
 nocite: | 
-  @mau_temperate_2018
+  @mau_temperate_2018, @cobleHowVerticalPatterns2016a, @slotPhotosyntheticHeatTolerance2019, @valladaresShadeToleranceKey2008b, @beaumontVerticalGradientsLeaf2009, @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014, @scartazzaInvestigatingEuropeanBeech2016, @martinCovarianceSunShade2020, @bolstad_foliar_1999, @harleyEffectsLightTemperature1996, @kullDistributionLeafPhotosynthetic1998, @kafutiFoliarWoodTraits2020, @duursmaVerticalCanopyGradients2006, @vanwittenbergheVariabilityStomatalConductance2012,
 ---
@@
 
+[@nakamuraForestsTheirCanopies2017]--> modelling ideas?
+
 ## Future Questions

### 2020-09-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 nocite: | 
-  @bolstad_foliar_1999; @roberts_stomatal_1990; @slot_photosynthetic_2019 
+  @bolstad_foliar_1999; @roberts_stomatal_1990; @slot_photosynthetic_2019; @harleyEffectsLightTemperature1996;@niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b;@ishiiExploringRelationshipsCanopy2004; @curtis_intracanopy_2019; @bolstad_foliar_1999
 ---

### 2020-09-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2020-09-28 vertical-thermal-review.Rmd
@@
 nocite: | 
-  @bolstad_foliar_1999; @roberts_stomatal_1990; @slot_photosynthetic_2019 
+  @bolstad_foliar_1999; @roberts_stomatal_1990; @slot_photosynthetic_2019; @harleyEffectsLightTemperature1996; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @ishiiExploringRelationshipsCanopy2004; @curtis_intracanopy_2019; @bolstad_foliar_1999; @kafutiFoliarWoodTraits2020;
 ---

### 2020-09-29 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. **2x**).**
-Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
+Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019, @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
 Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018, Tymen et al.
@@
 Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
-However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
-*(Add something about risk of freeze.)*
+However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate, @nakamuraForestsTheirCanopies2017].
+*(Add something about risk of freeze.)*--> [@nakamuraForestsTheirCanopies2017]
 
@@
 - @mcgregor_tree_nodate
+
+Forest canopies moderate relative humidity in below-canopy microclimates along with temperature.
+In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate [@vonarxSpatiotemporalEffectsForest2012]
+
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$
 - $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018, Tymen et al. 2017]
-This means that canopy leaves tend to be exposed to higher *evaporative demand*.
+This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004]
 
@@
 - Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected. 
-- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015).
+- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
 Max transpiration rate - increases with canopy height (Kenzo et al. 2015).

### 2020-09-29 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Forest canopies moderate relative humidity in below-canopy microclimates along with temperature.
-In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate [@vonarxSpatiotemporalEffectsForest2012]
+In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate.With rising temperatures, this trend in maximum $T_{air} and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012]
 
@@
 -	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
-Differences in concentration are by far most pronounced near ground level. 
+Differences in concentration are by far most pronounced near ground level.[@niinemetsPhotosyntheticAcclimationSimultaneous2004]
 Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient.
@@
 Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
-Species that have higher shade casting ability increase T buffering (Zelllweger et al. 2019) [@zellweger_seasonal_2019]
+Species that have higher shade casting ability such as braodleaved trees increase T buffering (Zelllweger et al. 2019) [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
 The strength of buffering also varies with respect to geographic and climatic factors.
@@
 Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b].
-Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light.
-Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
+Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@@niinemetsWorldwideAnalysisWithincanopy2015] Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
 2020], leaf thermoregular and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016, @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al.
@@
 
+Photosynthesis increases with increasing temperature and saturates at quatum flux density of 400 to 700nm m-2 s-1, beyond this the leaf experiences heat stress, undergoes non-photochemical quenching, photoinhibition and leaf necrosis [@niinemetsWorldwideAnalysisWithincanopy2015]
+
 **While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves**

### 2020-09-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2020-09-29 vertical-thermal-review.Rmd
@@
 **The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
-Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
+Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a, Senfetal.2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
 
@@
 
-**However, we lack a systematic understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecology (Fig. 1).**
+(@niinemetsWithincanopyVariationRate2004a highlights biophysical and photosynthetic gradient interactions along vertical canopy profile, however our current knowledge of this topic is lacking.)
+
+**However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecology (Fig.
+1).**
 
@@
 
-Most notably, light conditions, specifically photosynthetically active radiation (PAR) varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
+Most notably, light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
 (Fig.
@@
 Forest canopies moderate relative humidity in below-canopy microclimates along with temperature.
-In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate [@vonarxSpatiotemporalEffectsForest2012]
+In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate.
+With rising temperatures, this trend in maximum $T_{air} and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012]
 
@@
 -	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
-Differences in concentration are by far most pronounced near ground level. 
+-Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004]
+
 Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient.
@@
 Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
-Species that have higher shade casting ability increase T buffering (Zelllweger et al. 2019) [@zellweger_seasonal_2019]
+Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
 The strength of buffering also varies with respect to geographic and climatic factors.
@@
 
-*Carotinoids* 
+*Carotinoids/Xanthophyll cycle pigments* 
 -	photoprotective- disseminate heat, acclimate to high T. THese are proportional to irradiance 
 -	antioxidant scavenging function- protect against cellular damage
-- Koniger et. al. 1995, and  Matsubara et al. 2009?
+- Koniger et. al. 1995, and  Matsubara et al. 2009
 
@@
 
-**Aligning with biophysical expectations, field observations have show that leaf temperatures are influenced by the biophysical environment and leaf traits.**
+**Aligning with biophysical expectations, field observations have shown that leaf temperatures are influenced by the biophysical environment and leaf traits.**
 1.	Air T
@@
 
+Photosynthesis increases with increasing temperature and saturates at quatum flux density of 400 to 700nm m-2 s-1, beyond this the leaf experiences heat stress; undergoes non-photochemical quenching, photoinhibition and leaf necrosis.
+Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves, especially in conditions of water stress or drought.
+[@niinemetsWorldwideAnalysisWithincanopy2015]
+
 **While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves**
@@
 14.5; *revisit this!*) (Campbell and Norman 1998).
+This observed by earlier studies suggested that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of Topt, Vmax anD Jmax along the gradient [@niinemetsWithincanopyVariationRate2004a, Harley and Baldocchi 1995, Friend, 2001].
 However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photoshythesis in sun and shade leaves.
-For 3 species in Panama, (Slot et al. 2019) found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
+For 3 species in Panama, (Slot et al.
+2019) found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
+Similiar results were found for Vmax and Jmax [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
 Mau et al. found no trend along a height gradient in Puerto Rico, and no significant trend in temperate trees (Mau et al. 2018).

### 2020-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2019) found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
-Similiar results were found for Vmax and Jmax [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
-Mau et al. found no trend along a height gradient in Puerto Rico, and no significant trend in temperate trees (Mau et al. 2018).
+Similiar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
+Mau et al. found no trend along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
 

### 2020-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2018).
-This results in higher boundary layer conductance, *DEFINE*, for canopy leaves [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990].
+This results in higher boundary layer conductance ($g_{a}$) for canopy leaves.
+Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata.
+Large $g_{a}$ for canopy leaves results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990].
 

### 2020-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
+
 Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a, Senfetal.2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]

### 2020-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
+
 Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate.
-[@nakamuraForestsTheirCanopies2017, @ozanneBiodiversityMeetsAtmosphere2003]
-With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015, @stovallTreeHeightExplains2019] 
+[@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003]
+With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. 
 Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).
@@
 
-Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018, @huangAirTemperatureOptima2019].Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a, Senfetal.2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019, @defrenneMicroclimateModeratesPlant2013]
+Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
+Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; Senfetal.2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]
 
@@
 
-We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellweger_forest_2020, @suggittExtinctionRiskClimate2018, @scheffersMicrohabitatsReduceAnimal2013,@defrenneMicroclimateModeratesPlant2013]
+We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellweger_forest_2020; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013; @defrenneMicroclimateModeratesPlant2013]
 
@@
 
-(@niinemetsWithincanopyVariationRate2004a highlights biophysical and photosynthetic gradient interactions along vertical canopy profile, however our current knowledge of this topic is lacking.)
+(@niinemetsWithincanopyVariationRate2004a highlights biophysical and photosynthetic gradient interactions along the vertical canopy profile, however our current knowledge of this topic is lacking.)
 
@@
 
-Most notably, light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018, @niinemetsPackingPhotosyntheticMachinery2009, Parker 1995, @poorterMetaanalysisPlantResponses2019b]
+Most notably, light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
 (Fig.
-**2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a, Chazdon & Fetcher 1984; Parker et al.
+**2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al.
 2019]
-In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984, @tymenQuantifyingMicroenvironmentalVariation2017, @smithSeasonalDroughtrelatedChanges2019].
-This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature.[@fausetDifferencesLeafThermoregulation2018, @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006, @doughtyAreTropicalForests2008]
+In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
+This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature [@fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
 
@@
 Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata.
-Large $g_{a}$ for canopy leaves results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990].
+Large $g_{a}$ for canopy leaves results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a].
 
 **Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. **2x**).**
-Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019, @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
+Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
 Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018, Tymen et al.
-2017]
-"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy"
+2017].
+"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy".
 However, similar maximum temperatures have been observed during the dry season in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
@@
 Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
-However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate, @nakamuraForestsTheirCanopies2017].
+However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
 *(Add something about risk of freeze.)*--> [@nakamuraForestsTheirCanopies2017]
@@
 In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate.
-With rising temperatures, this trend in maximum $T_{air} and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012]
+With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012a]
 
@@
 - $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018, Tymen et al. 2017]
-This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004]
+This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
 
@@
 -	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
--Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004]
+-Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
 
@@
 * [@davis_microclimatic_2019]
-However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017,  @martinCovarianceSunShade2020a]
+However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]
 
@@
 Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b].
-Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@@niinemetsWorldwideAnalysisWithincanopy2015] Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
-2020], leaf thermoregular and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016, @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al.
+Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@niinemetsWorldwideAnalysisWithincanopy2015].
+Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
+2020], leaf thermoregular and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al.
 2020, REFS)
@@
 
-Photosynthesis increases with increasing temperature and saturates at quatum flux density of 400 to 700nm m-2 s-1, beyond this the leaf experiences heat stress; undergoes non-photochemical quenching, photoinhibition and leaf necrosis.
+Photosynthesis increases with increasing temperature and saturates at quatum flux density of 400 to 700nm $m^{-2}$ $s^{-1}$, beyond this the leaf experiences heat stress; non-photochemical quenching, photoinhibition and leaf necrosis.
 Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves, especially in conditions of water stress or drought.
@@
 14.5; *revisit this!*) (Campbell and Norman 1998).
-This observed by earlier studies suggested that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of Topt, Vmax anD Jmax along the gradient [@niinemetsWithincanopyVariationRate2004a, Harley and Baldocchi 1995, Friend, 2001].
+This observed by earlier studies suggested that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of Topt, Vmax anD Jmax along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
 However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photoshythesis in sun and shade leaves.
-For 3 species in Panama, (Slot et al.
-2019) found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
+For 3 species in Panama, [@slotPhotosyntheticHeatTolerance2019] found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
 Similiar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]

### 2020-10-01 vertical-thermal-review.Rmd
added some info

@@
 This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
-These changes, alongside climate change-driven disturbances such as drought, deforestation and fire, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008, @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+These changes, alongside climate change-driven disturbances such as drought, deforestation and fire, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015].
@@
 This results in higher boundary layer conductance ($g_{a}$) for canopy leaves.
-Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata.
-Large $g_{a}$ for canopy leaves results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a].
+Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata;($g_{a}$) also depends on leaf traits, morphology and windspeed.
+Large $g_{a}$ for canopy leaves results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a] thus having important implications for leaf thermal sensitivity.
 
@@
 In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate.
-With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012a]
+With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012]
 
@@
 Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
-2020], leaf thermoregular and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al.
+2020], leaf thermoregulation and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al.
 2020, REFS)
@@
 
-Photosynthesis increases with increasing temperature and saturates at quatum flux density of 400 to 700nm $m^{-2}$ $s^{-1}$, beyond this the leaf experiences heat stress; non-photochemical quenching, photoinhibition and leaf necrosis.
-Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves, especially in conditions of water stress or drought.
+Photosynthesis increases with increasing temperature and irradiance, and saturates at quatum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$, beyond this the leaf experiences heat stress; non-photochemical quenching, photoinhibition and leaf necrosis.
+Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought.
 [@niinemetsWorldwideAnalysisWithincanopy2015]
@@
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5; *revisit this!*) (Campbell and Norman 1998).
-This observed by earlier studies suggested that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of Topt, Vmax anD Jmax along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
+14.5; *revisit this!*) (Campbell and Norman 1998) as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
 However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photoshythesis in sun and shade leaves.
-For 3 species in Panama, [@slotPhotosyntheticHeatTolerance2019] found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
+For 3 species in Panama, @slotPhotosyntheticHeatTolerance2019 found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
 Similiar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]

### 2020-10-01 vertical-thermal-review.Rmd
update

@@
 Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata;($g_{a}$) also depends on leaf traits, morphology and windspeed.
-Large $g_{a}$ for canopy leaves results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a] thus having important implications for leaf thermal sensitivity.
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a] thus having important implications for leaf thermal sensitivity.
 

### 2020-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
-Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; Senfetal.2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]
+Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]
 

### 2020-10-01 tables.Rmd
Update tables.Rmd
added lines

@@
 
-bibliography: references.bib #This can be generated by exporting BibTex file from Zotero
+bibliography: Zotero_references.bib #This can be generated by exporting BibTex file from Zotero
 csl: new-phytologist.csl
@@
   pack_rows("Leaf anatomy and morphological traits", 1, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Leaf biochemical and physiological traits", 11, 17, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf biochemical and physiological traits", 11, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-10-01 vertical-thermal-review.Rmd
update

@@
 However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
-*(Add something about risk of freeze.)*--> [@nakamuraForestsTheirCanopies2017]
+*(Add something about risk of freeze.)* Additionally, canopies also buffer effects of snow through intercepting heavy snow accumulution on below-canopy levels [@nakamuraForestsTheirCanopies2017; @storckMeasurementSnowInterception2002]
 

### 2020-10-05 vertical-thermal-review.Rmd
some changes, as discussed in our call and in preparation for tomorrow's call

@@
 ![**Figure 1.
-(schematic of a forest summarizing most important gradients.** Current fig is just a rough illustration of how this might look – a draft figure that KAT had on hand illustrating hypotheses (ignore specific content).
+(schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** Current version is just a very rough sketch of how this might look – needs to be filled in with content from the review.
 We could have a set of arrows for each of the major categories considered here.
-This would be a key figure, and should be beautifully illustrated—KAT could do a watercolor, or Nidhi could illustrate).
+This would be a key figure, and should be beautifully illustrated—KAT could do a watercolor, or Nidhi could illustrate- and make use of NewPhyt's figure editing service for Tansley reviews).
 ](schematic/schematic.png)
@@
 
-Maybe we should start this section with a brief definition of what we mean by "biophysical"?
-E.g.
-the physical structure of the forest canopy (from understory to the canopy top) and associated microclimates, which are in large part determined by the structure of the forest itself.
-]
+The biophysical environment, here defined to include the physical structure of the forest canopy (from understory to the canopy top) and associated microclimates, is in large part determined by the structure of the forest itself.
+Canopy foliage, which varies across forest types and seasonally, strongly shapes the understory light, wind, etc.
+Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).
 
-**Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).**
+*(paragraph on leaf area distribution, possibly including figure? [see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
+The vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; Smith et al.
+2019 ; Parker et al.
+2019) and following disturbance (e.g., Parker et al.
+2002 ; Almeida et al.
+2016 ; Stark et al.
+2020 in press).
+In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season but excluding periods (in deciduous forests) when no leaves are present.
 
-Most notably, light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
+Light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
 (Fig.
@@
 
-In addition, the vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; Smith et al.
-2019 ; Parker et al.
-2019) and following disturbance (e.g., Parker et al.
-2002 ; Almeida et al.
-2016 ; Stark et al.
-2020 in press).
 - (*@bonan_ecological_2016 reviews this and points to appropriate references*)
@@
 -	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
--Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
+- Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
 
@@
 
+\newpage
+
 ![**Figure 2.
-Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is old version from Ian McGregor’s in-review paper, showing NEON data from SCBI.
+Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is old version from Ian McGregor’s in press paper, showing NEON data from SCBI.
 See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2.
 ](NEON_height_profiles/NEON_height_profile_temp.png)
+
+![This is a draft illustration of contents for the intendend figure.
+We'll pick out one variable-month combination per panel, and show it for all the forested NEON sites.](NEON_height_profiles/NEON_height_profile_temp2.png)
+
+\newpage
 
@@
 -  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section)
+
+*(Paragraph on intra- vs inter-specific variation)*
 
@@
 
-![**Figure 4. Temperature sensitivity of tree growth for understory versus canopy trees.** Analysis methods described in Appendix S3. Data from Helcoski et al. 2019.](SCBI_tree_rings/SCBIfig.png)
+![**Figure 4.
+Temperature sensitivity of tree growth (derived from tree rings) for understory versus canopy trees.** THIS IS A ROUGH PLACEHOLDER-- KAT WILL PLAY WITH POTENTIAL VERSIONS.
+IF OUR IN-PREP TREE RING ANALYSIS METHODS PAPER IS ACCEPTED AHEAD OF THIS, WE CAN HIGHLIGHT RESULTS FROM THAT, WHICH ARE MORE RELIABLE.
+Analysis methods described in Appendix S3.
+OUTLIER HERE IS SHADE-TOLERANT BEECH.
+Data from Helcoski et al.
+2019.](SCBI_tree_rings/SCBIfig.png)
 
-## Implications for Modeling / Scaling
+## Implications for Modeling / Scaling / Future Climate Projections
 
@@
 
-## Future Questions
+*(This section should address inter-specific vs intra-specific variation, as Tables 1 and 2 refer to intra-specific variation, but most models will focus on trait variation across strata, regardless of species).
+
+#### Future Questions
+
+*(KAT thinks this section should probably be dropped/ content integrated elsewhere, most within the implications for modeling/ scaling section)*
 

### 2020-10-06 vertical-thermal-review.Rmd
incorporated some changes from group discussion

@@
 
-v.	What are the implications of these patterns for the ecology and climate change responses of canopy versus understory trees?
-
-vi. What are the implications for modeling/ scaling?
+v.	What are the implications of these patterns for the ecology and climate change responses of canopy versus understory trees? What are the implications for modeling/ scaling? (*nerge these better*)
 
@@
 
-*(paragraph on leaf area distribution, possibly including figure? [see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
+- (*@bonan_ecological_2016 reviews this and points to appropriate references*)
+
+*(paragraph on leaf area distribution, including figure ([separate from or combined with current Fig. 2- see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
 The vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; Smith et al.
@@
 This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature [@fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
-
-- (*@bonan_ecological_2016 reviews this and points to appropriate references*)
 -	Sunflecks: Leaky et al.
@@
 
-## Leaf temperature
+- *Are patterns (in traits, metabolism, ecology) driven by tree height or exposure?*
+
+## Leaf temperature and hydraulics
 
@@
 Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016).
+
+*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves and leaf water content)*
 
@@
 
-## Ecology
+## Scaling & implications
+
+### Distribution of species and traits across the vertical profile
+
+*(This section should address inter-specific vs intra-specific variation, as Tables 1 and 2 refer to intra-specific variation, but most models will focus on trait variation across strata, regardless of species).
 
@@
 - something on isoprene production?
+
+**Appears to affect community change under warming (Zellweger et al. 2020).**
+
+### Scaling to tree level: demography
 
@@
 
-**Appears to affect community change under warming (Zellweger et al. 2020).**
-
 ![**Figure 4.
@@
 
-## Implications for Modeling / Scaling / Future Climate Projections
+### Representation in models
 
@@
 
-*(This section should address inter-specific vs intra-specific variation, as Tables 1 and 2 refer to intra-specific variation, but most models will focus on trait variation across strata, regardless of species).
+*(do we need a section on ecosystem function-- e.g., transpiration, gpp?)*
 
-#### Future Questions
-
-*(KAT thinks this section should probably be dropped/ content integrated elsewhere, most within the implications for modeling/ scaling section)*
-
-- Are patterns (in traits, metabolism, ecology) driven by tree height or exposure?
+### Implications for climate change responses
 

### 2020-10-08 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Ian McGregor^3^,
-Elsa Ordway,
-Marielle N. Smith^3^,
+Elsa M. Ordway^4^,
+Marielle N. Smith^5^,
 Tyeen Taylor,
@@
 
-4. Department of Forestry, Michigan State University, East Lansing, MI 48824, USA
+4. Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA 02138, USA
+
+5. Department of Forestry, Michigan State University, East Lansing, MI 48824, USA
 
@@
 Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
-Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]
+Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has been shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]
 
@@
 
-We're seeing increasing evidence that this impacts the ecology, with potential feedbacks to climate change, especially in protecting cool-acclimated species [@zellweger_forest_2020; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013; @defrenneMicroclimateModeratesPlant2013]
+We're seeing increasing evidence that this impacts ecosystems, with potential feedbacks to climate change, especially in protecting species acclimated to cooler [mean annual] temperatures [@zellweger_forest_2020; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013; @defrenneMicroclimateModeratesPlant2013]
 
@@
 
-**However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecology (Fig.
+**However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecosystems (Fig.
 1).**
@@
 
-i.	How does the biophysical environment vary with height in forests?
+i.	How does the biophysical environment vary with height in forests? ("What aspects of biophysical envt? Microclimate and vertical plant structure/phenology/etc (ok yes, see this in fig 1)"?--Elsa:
+So change to micrometeorological environment? Not biological just physical in this question, right?
 
-ii.	How do leaf traits vary with height (or between sun and shade leaves) in forests?
+ii.	How do leaf traits vary with height (or between sun and shade leaves) in forests?(What types of leaf traits?-E.O)
 
-iii.	How do biophysical environment and traits combine to affect leaf temperature?
+iii. How do vertical changes in biophysical environment conditions and leaf traits combine to affect leaf temperature [and in turn photosynthetic capacity]??
 
@@
 
-v.	What are the implications of these patterns for the ecology and climate change responses of canopy versus understory trees? What are the implications for modeling/ scaling? (*nerge these better*)
+v. What are the implications of these patterns for the ecology ecosystem and climate change responses of canopy versus understory trees?
+
+vi.	What are the implications for modeling/ scaling / What are the implications of these leaf-scale responses at the ecosystem scale [in the context of climate change] / How do these leaf-level responses scale to entire ecosystems(*nerge these better*)
 
@@
 (Fig.
-**2x**)such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al.
-2019]
+**2x**), such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al.
+2019].
 In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
@@
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$
-- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018, Tymen et al. 2017]
+- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017]
 This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
@@
 Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
-2020], leaf thermoregulation and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018, @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, Chen et al.
-2020, REFS)
+2020], leaf thermoregulation and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)
 
@@
 
-## Scaling & implications
+## Ecosystem Scaling & implications
 
@@
 
-### Representation in models
+## Implications for Modeling / Scaling / Future Climate Projections
+
+Elsa--The content below could probably be integrated in the above [Ecosystem Scale Implications] section, with subsections focusing on: 
+-	Suggested implications at the ecosystem scale based on literature
+-	How do models currently integrate or not integrate these concepts and dynamics
+-	What major gaps in knowledge and/or important questions emerge from this review (i.e.
+a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
+-	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?
 
@@
 
+--Variation across horizontal biophysical gradients (climate, soils, etc.)(E.O)
+
 ## Conclusions

### 2020-10-08 tables.Rmd
trichome info

@@
   column_spec(6, width = "4cm") %>%
-  pack_rows("Leaf anatomy and morphological traits", 1, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Leaf biochemical and physiological traits", 11, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf anatomy and morphological traits", 1, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf biochemical and physiological traits", 12, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-10-13 tables.Rmd
Update tables.Rmd

@@
   column_spec(6, width = "4cm") %>%
-  pack_rows("Leaf anatomy and morphological traits", 1, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Leaf biochemical and physiological traits", 12, 22, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf anatomy and morphological traits", 1, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf biochemical and physiological traits", 14, 25, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-10-28 tables.Rmd
updated with PAR absorptance

@@
   pack_rows("Leaf anatomy and morphological traits", 1, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Leaf biochemical and physiological traits", 14, 25, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf biochemical and physiological traits", 14, 29, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-12-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
-Resulting $T_{leaf}$'s generally track air temperature ($T_{air}$) and do not vary dramatically across the vertical gradient, except when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become significantly elevated above both $T_{air}$ and understory $T_{leaf}$.
-Correspondingly, while leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
+Resulting $T_{leaf}$'s exceed air temperature ($T_{air}$) under conditions of high solar radiation, low wind, and low stomatal conductance.
+Differences are most pronounced when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become dramatically elevated above both $T_{air}$ and understory $T_{leaf}$.
+While leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
 The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
@@
 
-**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, $T_{leaf}$ often differs little from $T_{air}$ and between understory and canopy (Fig.
+**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig.
 **2?**).**
-For instance... 	(Bolstad et al. 1999).
+In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999).
 Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
@@
 -	(Fauset et al. 2018)
-Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016).
+Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.
 

### 2020-12-10 vertical-thermal-review_SI.Rmd
set up SI file
@NidhiVinod, @mcgregorian1 , setting up an SI file where we can start filling in methods for the various analyses.

---
title: ''
output: word_document
bibliography: Zotero_references.bib
csl: new-phytologist.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}

#Appendices
appendix_NEON_methods <- "S1"
appendix_tealeaves_methods <- "S2"
appendix_lit_review_methods <- "S3"
appendix_SCBI_treering <- "S4"

appendix_NEON_methods_title<- "Methods for NEON vertical profiles"
appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
appendix_lit_review_methods_title<- "Methods for literature review"
appendix_SCBI_treering_title <- "Methods for SCBI tree-ring analysis"

#SI table numbers
SItable_example = "S1"

#SI table captions
SItable_example_legend <- "Table S1. ...."

#SI figure numbers
SIfig_example = "S1"

#SI figure captions
SIfig_example_legend <- "Figure S1. ...."

```

**Title:** Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

**Authors (so far, not necessarily in final order):** 
Nidhi Vinod^1^,
Martijn Slot^2^,
Ian McGregor^3^,
Elsa M. Ordway^4^,
Marielle N. Smith^5^,
Tyeen Taylor,
Lawren Sack,
Kristina J. Anderson-Teixeira^1,2^*

\newpage

#  Appendix `r appendix_NEON_methods`. `r appendix_NEON_methods_title`

\newpage

#  Appendix `r appendix_tealeaves_methods`. `r appendix_tealeaves_methods_title`

\newpage

#  Appendix `r appendix_lit_review_methods`. `r appendix_lit_review_methods_title`

\newpage

# Appendix `r appendix_SCBI_treering`. `r appendix_SCBI_treering_title`

\newpage
## References

### 2020-12-10 vertical-thermal-review.Rmd
work on auto numbering of figures, appendices, plus some content on biophysical gradients

@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+
+# This chunk is for auto numbering of tables/figures
+
+#figures
+fig_schematic = "1"
+fig_NEON_vertical =  "2"
+fig_NEON_differences  = "3"
+
+# SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)
+
+#Appendices
+appendix_NEON_methods <- "S1"
+appendix_tealeaves_methods <- "S2"
+appendix_lit_review_methods <- "S3"
+appendix_SCBI_treering <- "S4"
+
+appendix_NEON_methods_title<- "Methods for NEON vertical profiles"
+appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
+appendix_lit_review_methods_title<- "Methods for literature review"
+appendix_SCBI_treering_title <- "Methods for SCBI tree-ring analysis"
+
+#SI table numbers
+SItable_example = "S1"
+
+#SI table captions
+SItable_example_legend <- "Table S1. ...."
+
+#SI figure numbers
+SIfig_example = "S1"
+
+#SI figure captions
+SIfig_example_legend <- "Figure S1. ...."
+
 ```
@@
 Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
-Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves.
+Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves, particularly in forests with dense canopies.
 However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
@@
 **However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecosystems (Fig.
-1).**
+` r fig_schematic`).**
 
@@
 
-i.	How does the biophysical environment vary with height in forests? ("What aspects of biophysical envt? Microclimate and vertical plant structure/phenology/etc (ok yes, see this in fig 1)"?--Elsa:
+i.	How does the biophysical environment vary with height in forests?
+("What aspects of biophysical envt?
+Microclimate and vertical plant structure/phenology/etc (ok yes, see this in fig ` r fig_schematic`)"?--Elsa:
 So change to micrometeorological environment? Not biological just physical in this question, right?
@@
 The biophysical environment, here defined to include the physical structure of the forest canopy (from understory to the canopy top) and associated microclimates, is in large part determined by the structure of the forest itself.
+We supplement review of the existing literature with a new analysis of data on vegetation structure and vertical profiles in microclimate from the U.S.
+National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`; Schimel et al.
+2007].
+
 Canopy foliage, which varies across forest types and seasonally, strongly shapes the understory light, wind, etc.
-Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. **2**).
+Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. `r fig_NEON_vertical`).
 
@@
 (Fig.
-**2x**), such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al.
+`r fig_NEON_vertical`), such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al.
 2019].
@@
 
-**Wind speeds are also higher at the top of the canopy (Fig. **2x**).**
+**Wind speeds are also higher at the top of the canopy (Fig. `r fig_NEON_vertical`).**
 *(fill in some specifics from the NEON analysis)*
@@
 
-**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. **2x**).**
+**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
 Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
@@
 
-**Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. **2x**).**
+**Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. `r fig_NEON_vertical`).**
 - DETAILS - REFS
@@
 
-\newpage
+![**Figure 2.
+Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is just one of several+ intended sites.
+See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2.
+](NEON_height_profiles/figures/profile_SCBI.png)
 
-![**Figure 2.
-Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is old version from Ian McGregor’s in press paper, showing NEON data from SCBI.
-See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2.
-](NEON_height_profiles/NEON_height_profile_temp.png)
-
-![This is a draft illustration of contents for the intendend figure.
-We'll pick out one variable-month combination per panel, and show it for all the forested NEON sites.](NEON_height_profiles/NEON_height_profile_temp2.png)
-
-\newpage
-
-**The strength of this buffering varies across forests, being influenced by both forest characteristics and the biophysical environment.**
+**The strength of this buffering varies across forests, being influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
 First, buffering increases with canopy cover.
@@
 
+![**Figure 3.
+Role of canopy permeability in shaping temperature buffering, from NEON data.** Current placeholder figure is Nidhi's first rough analysis.
+See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2.
+](NEON_height_profiles/figures/differences_across_forests_temp.png)
+
 ## Trait variation
@@
 Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
-2020], leaf thermoregulation and thermal sensitivty which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)
+2020], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)
 
@@
 Similarly, our previously unpublished analysis shows that in a broadleaf deciduous forest in Virginia, understory trees in a Mid-Atlantic forest responded more negatively to warmer $T_{air}$ than did canopy trees (Fig.
-**4**; Appendix **S3**).
+**4**; Appendix `r appendix_SCBI_treering`).
 This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. **3?**). 
@@
 IF OUR IN-PREP TREE RING ANALYSIS METHODS PAPER IS ACCEPTED AHEAD OF THIS, WE CAN HIGHLIGHT RESULTS FROM THAT, WHICH ARE MORE RELIABLE.
-Analysis methods described in Appendix S3.
+Analysis methods described in Appendix S4.
 OUTLIER HERE IS SHADE-TOLERANT BEECH.
@@
 
-Appendix S1. Methods for NEON vertical profiles
+Appendix `r appendix_NEON_methods`. `r appendix_NEON_methods_title`
 
-Appendix S2. Methods for literature review
+Appendix `r appendix_tealeaves_methods`. `r appendix_tealeaves_methods_title`
 
-Appendix S3. Methods for SCBI tree-ring analysis
+Appendix `r appendix_lit_review_methods`. `r appendix_lit_review_methods_title`
+
+Appendix `r appendix_SCBI_treering`. `r appendix_SCBI_treering_title`
 

### 2020-12-10 vertical-thermal-review.Rmd
Added more information, and Martijn's comments

@@
 There is also increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012].
-This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
-These changes, alongside climate change-driven disturbances such as drought, deforestation and fire, may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
-2015].
+These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
+
+Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages.
+Overstory trees form canopies that play a crucial role in regulating macro-microclimate [@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
 
@@
 
-Taller trees form canopies that maintain vertically-stratified forest ecosystems, and play a crucial role in regulating macro-microclimate.
-[@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003]
-With increase in temperatures globally, we're seeing that larger trees suffer more during drought [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. 
-Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (McDowell et al.2020).
+The structure of canopies and understories are changing due to unprecedented disturbance rates thus having feedbacks to forest structure and climate change.
+The capacity of buffering largly depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestaion, fire and related disruptions, which may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+2015].With increasing temperatures, temperate and tropical forest canopies will be soon or are already functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].Additionally,larger trees suffer more during drought [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].The increase in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is also associated with reductions in canopy structural complexity, and altered microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
 
-**The structure and microclimates of global forests are changing due to unprecedented disturbance rates.**
-
-Temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
-Loss of canopy cover as a result of increasing temperature and related forest disturbances [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is associated with reductions in canopy structural complexity, and altered microclimates and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], this has been shown to increase understory temperatures which raises concern for understory species and forest biodiversity [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]
+Increase in microclimatic temperatures as a consequence of macroclimate warming is causing thermophilization in plant communities, reshuffling of species acclimated to cooler [mean annual] temperatures in the understory [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013]and reduced tree growth rates, driving a global trend towards disquillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008] with potential feedbacks to climate change.
 
@@
 
-We're seeing increasing evidence that this impacts ecosystems, with potential feedbacks to climate change, especially in protecting species acclimated to cooler [mean annual] temperatures [@zellweger_forest_2020; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013; @defrenneMicroclimateModeratesPlant2013]
+With such changing in forest dynamics, it is becoming more important than ever to understand thermal sensitivity of forests along the vertical canopy profile, from the understory to the top of the canopy, this will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
 
-With shifting forest dynamics, it is becoming more important than ever to understand thermal sensitivity along the vertical canopy profile, from the understory to the top of the canopy, this will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
+(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradient varies along the vertical canopy profile, however our current knowledge of this topic is lacking.)
 
-(@niinemetsWithincanopyVariationRate2004a highlights biophysical and photosynthetic gradient interactions along the vertical canopy profile, however our current knowledge of this topic is lacking.)
-
-**However, we lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecosystems (Fig.
+**We lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecosystems (Fig.
 ` r fig_schematic`).**
@@
 
+Martijn : in this illustration it would also be nice to show that you can have shade leaves of canopy trees in the understory, and shaded plants of understory species.
+
 ## The biophysical environment
@@
 
-Light conditions, specifically photosynthetically active radiation (PAR) strongly varies  with canopy height due to cumulative layers of vegetation that limits light to lower canopy layers [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
-(Fig.
-`r fig_NEON_vertical`), such that the forest floor of understory regions receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al.
-2019].
+Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with canopy height, canopy structure and across species [Koike et al.
+2000] due to cumulative layers of vegetation.
+Generally, upper canopies are more exposed to irradiance and limit light to canopy interiors and lower canopy layers[@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
+(Fig. `r fig_NEON_vertical`).
 In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
-This difference in light regime is an important variable in determining leaf traits along the gradient and consequently their responses to temperature [@fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
+ Tropical forests with dense overstory limit light to understories that receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al.
+2019], temperate forests with broad-leaf overstories generally receive ....
+in the understory [ REF ], whereas conifer forest understories receive [ ] of light [REF]
+This difference in light regime is an important variable that drives plasticity among leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].Upper canopy leaves adapted to high irradiance have different leaf traits and thermal sensitivity strategies compared to deeply shaded understory, canopy interiors, and in canopy gaps.
+This is discussed further below under Leaf Traits.
+
 -	Sunflecks: Leaky et al.
@@
 This results in higher boundary layer conductance ($g_{a}$) for canopy leaves.
-Boundary layer conductance is the thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata;($g_{a}$) also depends on leaf traits, morphology and windspeed.
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a] thus having important implications for leaf thermal sensitivity.
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a] Additionally, boundary layer resistance is proportional to thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata;($g_{a}$) also depends on leaf traits, morphology and windspeed.Generally, larger leaf width increases boundary layer resistance, and thus having important implications for leaf thermal sensitivity.[@michaletzEnergeticCarbonEconomic2016]*double check*
 
@@
 "max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy".
-However, similar maximum temperatures have been observed during the dry season in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+However, similar maximum temperatures have been observed during the dry season a semi-deciduous forest in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
 Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
-Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].
+(Martijn: I think this may have been important; in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air)
+Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point?
+(e.g., greater radiative heat loss in exposed areas resulting in lower Tmin))
 However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
-*(Add something about risk of freeze.)* Additionally, canopies also buffer effects of snow through intercepting heavy snow accumulution on below-canopy levels [@nakamuraForestsTheirCanopies2017; @storckMeasurementSnowInterception2002]
+*(Add something about risk of freeze.)*
 
@@
 In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate.
-With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests [@vonarxSpatiotemporalEffectsForest2012]
+With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests and montane forests  [@vonarxSpatiotemporalEffectsForest2012, @fauset_differences_2018] (to clarify why?
+are these high elevation forests limited by low temperature and low rh?)
 
-RH, in combination with temperature, determines vapor pressure deficit, $VPD$
-- $VPD$ is lower in understory of Atlantic forest during wet season [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017]
-This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential, making canopy leaves more efficient in regulating water use [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
+RH, in combination with temperature, determines vapor pressure deficit, $VPD$.
+Atlantic forests during wet season tended to have higher $T_{air}$, and $VPD$ in upper canopy and canopy gaps than lower in understory where $T_{air}$ and $VPD$ were lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
+This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
 
@@
 
-Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient.
+Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient."If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO2 emitted at the forest floor could increase the vertical CO2 gradient in the future, but this would still only be a short-term effect in the early morning, as the CO2 would quickly diffuse through the canopy.
+Furthermore, increases in mortality of big trees (e.g.
+Senf et al.
+2018) would increase canopy roughness and prevent build-up of CO2 concentration gradients."- MS
 
@@
 **The strength of this buffering varies across forests, being influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
-First, buffering increases with canopy cover.
+First, buffering increases with canopy cover and water availabity.
 * greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
@@
 * [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015] 
+The omega factor (McNaughton and Jarvis, 1983; Jarvis & McNaughton 1986) describes the degree of aerodynamic coupling between the vegetation and the atmospheric boundary layer; the lower the value of omega, the greater the coupling.
+"I wonder whether it would be useful to introduce such 'formal' terms used to describe what is essentially the same as buffering effect of canopies.
+I'd have to read up some more to see if there are papers showing what sort of canopy characteristics influence omega.
+
+https://doi.org/10.1016/S0065-2504(08)60119-1"-M.Slot
+
 *Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
-Taller trees don’t necessarily increase buffering [@zellweger_seasonal_2019]. 
+Taller trees don't necessarily increase buffering [@zellweger_seasonal_2019]. 
 Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
@@
 * Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
-* [@davis_microclimatic_2019]
+* [@davis_microclimatic_2019]--Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperatures [@davis_microclimatic_2019]
 However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]
@@
 ## Trait variation
+
+[when I think of shade leaves in the understory there are three categories; 1.
+shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded.
+2.
+seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy.
+3 understory plants that complete their lifecycle in the shaded understory.
+These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on which category of shade plants we're talking about.
+I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations).
+Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear.
+photosynthesis, and canopy-level carbon uptake.
+They don't have to be treated separately, but it should always be clear which is being referred to.
+And in the consequences for scaling section this is of course also relevant.]
+
+in line with Marielle's comment below; it will be important to distinguish between within plant variation vs across species variation.
+Understory species may have conservative  growth strategies because they are resource (light) limited.
+Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well.
+Their traits will be different from shade leaves of canopy trees as well.
+E.g.
+They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well
+
+Leaf traits are shaped by intra canopy light and height, and show intracanopy plasticity with availability of light.
+Taller trees tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
+Higher irradiance has consequential adaptations in leaf morphology with smaller leaf area index in exposed canopies than shaded lower canopy or canopy interior, and well as having deeper leaf lobes in species such as Quercus.
+R.
+The table above summarizes the variation of different traits based on light regime, however, some of the traits might differ between species along the vertical profile, thus also varying in leaf thermal sensitivity.
+[Sack et al.
+2016]
+
+Additionally, passing light spots in the forest understory often caused by over canopy vegetation and wind speed called sunflecks cause a different type of morphological adaptation in shaded leaves, making them efficient in capturing 50% of light through sunflecks for photosynthesis.
 
@@
 *Leaf area / characteristic dimension*:
-- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected. 
+- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
 - Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
 Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
+Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
+Additionally, increase in leaf area also influences leaf temperature, generally by maintaining greater leaf temperature than leaves with lower leaf area {Wright et al. 2019}
 
@@
 -	Michaeletz has used this in theoretical calculations (Michaeletz et al. 2016 and 2015?
+I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
+In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot
 
@@
 This is primarily driven by the greater light available to sun leaves.
-In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light,  including ....
+In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light,  including higher concentration of Rubsico, multiple layers of palisade parenchyma, and saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
 The vertical gradient in photosythetic rates is also influenced by...
 .[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N.
-Smith
+Smith [agreed.
+Again the importance of differentiating between within-species and among species trait variation.
+Also; steepness of the vertical gradient in light and VPD.
+In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]
 
-Photosynthesis increases with increasing temperature and irradiance, and saturates at quatum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$, beyond this the leaf experiences heat stress; non-photochemical quenching, photoinhibition and leaf necrosis.
+Photosynthesis increases with increasing irradiance, and saturates at quatum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
+Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
+2017; Slot & Winter 2017).Beyond the optimum photosynthesis decreases as a result of stomatal closure (e.g.
+Slot & Winter 2017; Smith et al.
+2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs.
+e.g.
+Sage & Kubien 2007; Varhammer et al.
+2015 ).Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis.
 Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought.
@@
 14.5; *revisit this!*) (Campbell and Norman 1998) as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
-However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photoshythesis in sun and shade leaves.
-For 3 species in Panama, @slotPhotosyntheticHeatTolerance2019 found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
+However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photosnythesis in sun and shade leaves.
+For 3 species in Panama, @hernandezSimilarTemperatureDependence2020 and @slotPhotosyntheticHeatTolerance2019 found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
 Similiar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
-Mau et al. found no trend along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
+Mau et al. found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
 
@@
 Similarly, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher $T_{50}$ was observed in more exposed (North-facing) canopy positions (Curtis et al. 2019). 
-We're not aware of any other studies on this.
+We're not aware of any other studies on this."we may want to argue that we have no a priori reasons to expect thermal tolerance to scale with height per se (or do we?).
+I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot
 

### 2020-12-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The capacity of buffering largly depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestaion, fire and related disruptions, which may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
-2015].With increasing temperatures, temperate and tropical forest canopies will be soon or are already functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].Additionally,larger trees suffer more during drought [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].The increase in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is also associated with reductions in canopy structural complexity, and altered microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
+2015].With increasing temperatures, temperate and tropical forest canopies will be soon or are already functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
+Additionally, larger trees suffer more during drought [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
+The increase in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is also associated with reductions in canopy structural complexity, and altered microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
 
-Increase in microclimatic temperatures as a consequence of macroclimate warming is causing thermophilization in plant communities, reshuffling of species acclimated to cooler [mean annual] temperatures in the understory [@zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013]and reduced tree growth rates, driving a global trend towards disquillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008] with potential feedbacks to climate change.
+Increase in microclimatic temperatures as a consequence of macroclimate warming is causing thermophilization in plant communities, reshuffling of species acclimated to cooler [mean annual] temperatures in the understory [@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013]and reduced tree growth rates, driving a global trend towards disquillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008] with potential feedbacks to climate change.
 
@@
 
-Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with canopy height, canopy structure and across species [Koike et al.
-2000] due to cumulative layers of vegetation.
-Generally, upper canopies are more exposed to irradiance and limit light to canopy interiors and lower canopy layers[@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
+Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with canopy height, canopy structure and across species [@koike_leaf_2001] due to cumulative layers of vegetation.
+Generally, upper canopies are more exposed to irradiance and limit light to canopy interiors, lower canopy layers and understories [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
 (Fig. `r fig_NEON_vertical`).
 In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
- Tropical forests with dense overstory limit light to understories that receive 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; Chazdon & Fetcher 1984; Parker et al.
-2019], temperate forests with broad-leaf overstories generally receive ....
-in the understory [ REF ], whereas conifer forest understories receive [ ] of light [REF]
-This difference in light regime is an important variable that drives plasticity among leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].Upper canopy leaves adapted to high irradiance have different leaf traits and thermal sensitivity strategies compared to deeply shaded understory, canopy interiors, and in canopy gaps.
+ Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [Bartemucci et al.
+2005]
+This difference in light regime along a vertical profile is an important driver of plasticity among leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].Upper canopy leaves adapted to high irradiance have different leaf traits and thermal sensitivity strategies compared to leaves in deeply shaded understory, canopy interiors, and in canopy gaps.
 This is discussed further below under Leaf Traits.
@@
 *(fill in some specifics from the NEON analysis)*
-Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [e.g., REFS; @mcgregor_tree_nodate] and open savannas (Curtis et al.
-2018).
+Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
 This results in higher boundary layer conductance ($g_{a}$) for canopy leaves.
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a] Additionally, boundary layer resistance is proportional to thickness of air layer at the surface of the leaf through which water vapor diffuses after leaving the stomata;($g_{a}$) also depends on leaf traits, morphology and windspeed.Generally, larger leaf width increases boundary layer resistance, and thus having important implications for leaf thermal sensitivity.[@michaletzEnergeticCarbonEconomic2016]*double check*
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; stocks et al.
+2006].
+Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata.
+Generally, larger leaf width increases boundary layer resistance [ implications of thermal sensitivity].[@michaletzEnergeticCarbonEconomic2016]*double check*
 
@@
 They don't have to be treated separately, but it should always be clear which is being referred to.
-And in the consequences for scaling section this is of course also relevant.]
+And in the consequences for scaling section this is of course also relevant.
 
@@
 E.g.
-They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well
+They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
+Those differences in traits are likely to affect their thermal properties as well]--Martijn Slot
 
@@
 Higher irradiance has consequential adaptations in leaf morphology with smaller leaf area index in exposed canopies than shaded lower canopy or canopy interior, and well as having deeper leaf lobes in species such as Quercus.
-R.
+rubra.
 The table above summarizes the variation of different traits based on light regime, however, some of the traits might differ between species along the vertical profile, thus also varying in leaf thermal sensitivity.

### 2020-12-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata.
-Generally, larger leaf width increases boundary layer resistance [ implications of thermal sensitivity].[@michaletzEnergeticCarbonEconomic2016]*double check*
 
@@
 Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
-Additionally, increase in leaf area also influences leaf temperature, generally by maintaining greater leaf temperature than leaves with lower leaf area {Wright et al. 2019}
+Additionally, increase in leaf area also influences leaf temperature by maintaining greater leaf temperature than leaves with lower leaf area {Wright et al. 2019}
+-Boundary layer and thermal sensitivity also depends of leaf width, leaf size and structure, such as lobes, leaf elongation and serration, all dissapate heat more efficiently than non-lobed leaves, wide or non serrated leaves.
+Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;
+Schuepp 1993; Sack et al. 2006].
 

### 2020-12-16 NEON_height_profiles/scripts/verticalProfileFigure.R
Got NEON tower coordinates

@@ -21,6 +21,13 @@ library(ggpubr)
 #   save(full_data, file=paste0(st, "test.Rdata"))
 # })
 
+#1a. Get NEON coordinates ####
+# library(data.table)
+# meta <- fread("data/site_data/NEON_Field_Site_Metadata_20201204.csv")
+# meta <- meta[field_site_id %in% c("BART","BONA","CLBJ","DEJU","DELA","GRSM","GUAN","HARV","JERC","LENO","MLBS","ORNL","OSBS","SCBI","SERC","SJER","SOAP","STEI","TALL","TEAK","TREE","UKFS","UNDE","WREF","YELL"), .(field_site_id, field_latitude, field_longitude, field_utm_easting, field_utm_northing, field_utm_zone)]
+# 
+# fwrite(meta, "data/site_data/neon_coordinates.csv")
+
 #2. Organize data for plotting ####
 
 data <- fread("forested_NEON_sites.csv")

### 2020-12-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-With such changing in forest dynamics, it is becoming more important than ever to understand thermal sensitivity of forests along the vertical canopy profile, from the understory to the top of the canopy, this will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
+With such changing in forest dynamics, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy, this will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
 
@@
  Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [Bartemucci et al.
-2005]
+2005].
 This difference in light regime along a vertical profile is an important driver of plasticity among leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].Upper canopy leaves adapted to high irradiance have different leaf traits and thermal sensitivity strategies compared to leaves in deeply shaded understory, canopy interiors, and in canopy gaps.
@@
 Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
-This results in higher boundary layer conductance ($g_{a}$) for canopy leaves.
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$ [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; stocks et al.
+This results in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [Daudet et al.
+1999].
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is typical of canopy leaves with small leaf sizes making them more advantageous in dissapating heat efficiently [Bauerle et al.
+2011].
+However, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation and low wind conditions [@martin_boundary_1999].
+Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; Stockes et al.
 2006].
-Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata.
+Boundary layer thickness, depth and boundary layer resistance increases with leaf size and effective leaf width, such that the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, therefore allowing small leaves to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 
@@
 
-Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient."If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO2 emitted at the forest floor could increase the vertical CO2 gradient in the future, but this would still only be a short-term effect in the early morning, as the CO2 would quickly diffuse through the canopy.
+Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient.
+
+If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO2 emitted at the forest floor could increase the vertical CO2 gradient in the future, but this would still only be a short-term effect in the early morning, as the CO2 would quickly diffuse through the canopy.
 Furthermore, increases in mortality of big trees (e.g.
 Senf et al.
-2018) would increase canopy roughness and prevent build-up of CO2 concentration gradients."- MS
+2018) would increase canopy roughness and prevent build-up of CO2 concentration gradients.
 
@@
 In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot
+-Effective leaf width defined as the largest circle that can be inscribed in a leaf is an effective predictor of leaf thermal time constant and leaf boundary layer thickness.
+Large leaves have greater effective leaf width which increases their thermal time constant by prolonging leaf cooling time compared to small leaves that have a smaller effective leaf width, and thus small thermal time constant with shorter cooling time.
+Increased leaf width typical in larger leaves was also observed to increase leaf to air temperature difference.
+[@leighInfluenceLeafSize2017]
 

### 2020-12-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012].
+There is also increasing severity and frequency of heat waves that may potentially reduce [forest] biomass acculumation and disrupt photosynthesis and metabolic processes across plant species [O'Sullivan;REF ;@perkinsIncreasingFrequencyIntensity2012].
 These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
-
-Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages.
-Overstory trees form canopies that play a crucial role in regulating macro-microclimate [@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
 
@@
 
-The structure of canopies and understories are changing due to unprecedented disturbance rates thus having feedbacks to forest structure and climate change.
-The capacity of buffering largly depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestaion, fire and related disruptions, which may contribute to increasing tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
-2015].With increasing temperatures, temperate and tropical forest canopies will be soon or are already functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
-Additionally, larger trees suffer more during drought [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
-The increase in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] is also associated with reductions in canopy structural complexity, and altered microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
+Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages.
+Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
+This creates a vertical stratification of biophysical environments which influences leaf temperature and metabolism along the gradient.[ref]
 
-Increase in microclimatic temperatures as a consequence of macroclimate warming is causing thermophilization in plant communities, reshuffling of species acclimated to cooler [mean annual] temperatures in the understory [@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013]and reduced tree growth rates, driving a global trend towards disquillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008] with potential feedbacks to climate change.
+The capacity of buffering largly depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions.
+Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
+Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests.
+Larger trees suffer more during drought, which may contribute to increasing tree mortality in forests around the world [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+2015].
+Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
 
-(Marielle--"Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al.
-2020)".)
+While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming.
+This is leading to thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+Microclimate warming has implications for forest regeneration[@von et al.] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008].
+Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al.
+2020).
+Each of these changes have potential feedbacks to climate change.
 
-With such changing in forest dynamics, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy, this will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
+With such changing forest dynamics, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy influenced by biophysical environmental gradient.This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
 
@@
 
-Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with canopy height, canopy structure and across species [@koike_leaf_2001] due to cumulative layers of vegetation.
-Generally, upper canopies are more exposed to irradiance and limit light to canopy interiors, lower canopy layers and understories [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
+Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with canopy height, canopy structure, across species and forests [@koike_leaf_2001] due to cumulative layers of vegetation.
+Generally, upper canopies are more exposed to irradiance and limit light through shading to canopy interiors, lower canopy layers and understories [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
 (Fig. `r fig_NEON_vertical`).
@@
 2005].
-This difference in light regime along a vertical profile is an important driver of plasticity among leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].Upper canopy leaves adapted to high irradiance have different leaf traits and thermal sensitivity strategies compared to leaves in deeply shaded understory, canopy interiors, and in canopy gaps.
-This is discussed further below under Leaf Traits.
+This difference in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
 
@@
 *(fill in some specifics from the NEON analysis)*
-Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
+Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months.
+Wind speed in dry season of semi-deciduous forests at upper canopy has been observed to be higher than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*]
 This results in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [Daudet et al.
 1999].
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is typical of canopy leaves with small leaf sizes making them more advantageous in dissapating heat efficiently [Bauerle et al.
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of canopy leaves with small leaf sizes making them more advantageous in dissapating heat efficiently [Bauerle et al.
 2011].
-However, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation and low wind conditions [@martin_boundary_1999].
+In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999].
 Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; Stockes et al.
 2006].
-Boundary layer thickness, depth and boundary layer resistance increases with leaf size and effective leaf width, such that the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, therefore allowing small leaves to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
+Boundary layer thickness, depth and boundary layer resistance increases with leaf size and effective leaf width [along the vertical gradient], such that the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, therefore allowing small leaves to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 
 **Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
-Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
-Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018, Tymen et al.
+Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [De Frenne et al.
+2019], and increasing recent evidence of similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
+
+Moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons.
+Tropical, and temperate broadleaved and non-pine conifer forests moderate daytime microclimate more than pine and boreal forests [De Frenne et al.
+2019; @vonarxSpatiotemporalEffectsForest2012]
+
+Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012] and coastal Brazil [@fauset_differences_2018, Tymen et al.
 2017].
-"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig. 4 - again illustrating buffering effect of the canopy".
-However, similar maximum temperatures have been observed during the dry season a semi-deciduous forest in Panama [@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+
+"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig.
+4 - again illustrating buffering effect of the canopy".However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+
 Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
-(Martijn: I think this may have been important; in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air)
 Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point?
@@
 However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
+
 *(Add something about risk of freeze.)*

### 2020-12-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 nocite: | 
-  @mau_temperate_2018, @cobleHowVerticalPatterns2016a, @slotPhotosyntheticHeatTolerance2019, @valladaresShadeToleranceKey2008b, @beaumontVerticalGradientsLeaf2009, @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014, @scartazzaInvestigatingEuropeanBeech2016, @martinCovarianceSunShade2020, @bolstad_foliar_1999, @harleyEffectsLightTemperature1996, @kullDistributionLeafPhotosynthetic1998, @kafutiFoliarWoodTraits2020, @duursmaVerticalCanopyGradients2006, @vanwittenbergheVariabilityStomatalConductance2012,
+  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998
 ---

### 2020-12-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environments which influences leaf temperature and metabolism along the gradient.[ref]
+This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient.[Bertrand et al.
+2011]
 
@@
 While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming.
-This is leading to thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
-Microclimate warming has implications for forest regeneration[@von et al.] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008].
+Temperature is directly linked with plant metabolic processes and growth, thus even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011] This is leading to thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008].
 Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al.
@@
 Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months.
-Wind speed in dry season of semi-deciduous forests at upper canopy has been observed to be higher than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*]
+Wind speed in dry season of a semi-deciduous forests at upper canopy has been observed to be higher than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*]
 This results in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [Daudet et al.
@@
 Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [De Frenne et al.
-2019], and increasing recent evidence of similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
+2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
 
 Moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons.
-Tropical, and temperate broadleaved and non-pine conifer forests moderate daytime microclimate more than pine and boreal forests [De Frenne et al.
-2019; @vonarxSpatiotemporalEffectsForest2012]
+Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures more than pine and boreal forests [De Frenne et al.
+2019; @vonarxSpatiotemporalEffectsForest2012a].
+This relationship is reveresed during the night, with pine forests maintaing warmer below canopy temperature than above [@vonarxSpatiotemporalEffectsForest2012a]
 
@@
 "max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig.
-4 - again illustrating buffering effect of the canopy".However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016] and in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+4 - again illustrating buffering effect of the canopy".However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016], similar results were observed in in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
 
@@
 However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
+
+[Add something about canopy temperature is a better proxy for determining leaf temperature than air temperature; in wet season leaf temperature can be linked to intensity of irradiance and air temperature, however, in dry season, air temperature better determines leaf temperature due to general increase in irradiance intensity and mixing of air movement along the gradient caused by fewer leaves [rey sanchez et al]]
 
@@
 
-Forest canopies moderate relative humidity in below-canopy microclimates along with temperature.
-In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering was observed to decrease daily maximum $T_{air}$ and increase daily minimum RH in below-canopy microclimate.
-With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications on understory seedling growth, especially in high altitude temperate pine forests and montane forests  [@vonarxSpatiotemporalEffectsForest2012, @fauset_differences_2018] (to clarify why?
+Along with temperature, forest canopies moderate relative humidity in below-canopy microclimates.
+In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
+Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter.
+Dry months showed less drastic minimum RH changes in below vs upper canopy  [@vonarxSpatiotemporalEffectsForest2012a].
+
+With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications for understory seedling growth, especially in high altitude temperate pine forests and montane forests  [@vonarxSpatiotemporalEffectsForest2012a, @fauset_differences_2018] (to clarify why?
 are these high elevation forests limited by low temperature and low rh?)

### 2020-12-21 vertical-thermal-review.Rmd
updated

@@
 Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
-Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests.
-Larger trees suffer more during drought, which may contribute to increasing tree mortality in forests around the world [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests [ref].
+Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000].
+Evidence shows that larger trees suffer more than small trees during drought, which may contribute to increasing tree mortality in forests around the world [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015].
@@
 
-Moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons.
-Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures more than pine and boreal forests [De Frenne et al.
+Diurnal moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons.
+Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests.
+Increase in crown density and leaf area index increases decreases light and higher maximum temperature permeability in forests, thus maintaining lower maximum daytime temperature.
+Decrease in crown density and leaf area index such as in pine forests that have sparse stands decreases buffering capacity [De Frenne et al.
 2019; @vonarxSpatiotemporalEffectsForest2012a].
-This relationship is reveresed during the night, with pine forests maintaing warmer below canopy temperature than above [@vonarxSpatiotemporalEffectsForest2012a]
+This relationship is reveresed during the night among all forests, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]
 
-Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012] and coastal Brazil [@fauset_differences_2018, Tymen et al.
+Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal Brazil [@fauset_differences_2018, Tymen et al.
 2017].
@@
 In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
-Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter.
-Dry months showed less drastic minimum RH changes in below vs upper canopy  [@vonarxSpatiotemporalEffectsForest2012a].
+Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
 
@@
 Additionally, increase in leaf area also influences leaf temperature by maintaining greater leaf temperature than leaves with lower leaf area {Wright et al. 2019}
--Boundary layer and thermal sensitivity also depends of leaf width, leaf size and structure, such as lobes, leaf elongation and serration, all dissapate heat more efficiently than non-lobed leaves, wide or non serrated leaves.
+-Boundary layer resistance and thermal sensitivity also depends of leaf width, leaf size and structure.
+Increase in leaf width increases stomatal resistance-- dependent on stomtal activity and boundary layer resistance [@fauset_differences_2018] such as lobes, leaf elongation and serration, all dissapate heat more efficiently than non-lobed leaves, wide or non serrated leaves.
 Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;
@@
 $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
+- As light availability typically increases with tree height, leaf temperature which is largely of function of radiation input, does as well [@michaletzPlantThermoregulationEnergetics2015].
+Direct sunlight in upper canopy leaves increases leaf temperature, [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
 While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.[@fauset_differences_2018]
@@
 Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
+- increase in air temperature, increases VPD, decrease in stomtal conductance during miday leading to decrease in CO2 in leaf = increase in leaf temperature in upper canopy. [@fauset_differences_2018]
 
@@
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
+-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads.
+However, leaf temperatures themselves were moderaed by air temperature.
+[@fauset_differences_2018].
 -	(Slot et al. 2019) and refs therein
@@
 *Stomatal conductance*
+Photosynthesis can be largely constrained by stomatal conductance. 
+-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al.
+2019)
 
@@
 In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light,  including higher concentration of Rubsico, multiple layers of palisade parenchyma, and saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
-The vertical gradient in photosythetic rates is also influenced by...
+The vertical gradient in photosythetic rates is also influenced by available concentration of N and P.
 .[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N.
@@
 **Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
+At increasing temperatures, Photosynthetic System 2 [PS2] is particularly vulnerable.
+Damage to pS2  can lead to decreased electron transport rates and photosynthetic failure.
+Irreversible damage to PS2 is done at 40-60C [Baker 2008; @feeley_thermal_2020]
 (Bolstad et al. 1999) did a study at Coweeta, including elev gradient. 
@@
 May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
-Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*).
+Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig.
+1*)."Higher $Q_{10}$ values in the functional groups of upper canopy where temperatures are the highest may lead to exponential loses of C with increasing global temperatures depending upon the ability of canopy leaves to acclimate" [@cavaleri_foliar_2008]
 Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.

### 2020-12-24 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2005].
-This difference in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008].
+This difference in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient.
+Leaves along a vertical intra-canopy gradient thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; Niinemets and Valladares 2006].
 
@@
 Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests.
-Increase in crown density and leaf area index increases decreases light and higher maximum temperature permeability in forests, thus maintaining lower maximum daytime temperature.
-Decrease in crown density and leaf area index such as in pine forests that have sparse stands decreases buffering capacity [De Frenne et al.
+Increase in crown density and leaf area index decreases light and higher maximum temperature permeability in forests, thus maintaining lower maximum daytime temperature.
+Thus, pine forests have sparse stands resulting in decreased buffering capacity [De Frenne et al.
 2019; @vonarxSpatiotemporalEffectsForest2012a].
-This relationship is reveresed during the night among all forests, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]
+This relationship is reveresed during the night among, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]
 
@@
 
-[when I think of shade leaves in the understory there are three categories; 1.
+Leaf traits are shaped by intra canopy light and height, and show intra-canopy plasticity in accordance to intensity and availability of light.
+Taller trees tend to be exposed to greater irradiance than dense lower canopy and shorter understory trees shaded by overstory.
+Degrees of high irradiance has consequential adaptations in leaf morphology compared to degrees of low irradiance to shaded-ness [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al.
+2018; @niinemetsWorldwideAnalysisWithincanopy2015].
+This dichotomy in leaf traits that vary with light intensity generally differ within intra-canopy, however the degree of intra-canopy plasticity also varies across species and geographically [Sack et al.
+2006; Ishii et al.
+2018].
+
+Adaptions of traits to ranges of sun-exposure or shaded-ness also vary within a forest strata. 
+Understory species may have conservative  growth strategies because they are resource (light) limited.
+Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well.
+Their traits will be different from shade leaves of canopy trees as well.
+E.g.
+They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.
+Additionally, shade leaves in the understory are of three categories; 1.
 shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded.
@@
 3 understory plants that complete their lifecycle in the shaded understory.
-These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on which category of shade plants we're talking about.
-I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations).
+These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.[refs]
+
+Passing light in the forest understory often caused by over-canopy vegetation and wind speed called sunflecks cause a different type of morphological adaptation in shaded leaves that make them efficient in capturing >50% of light through sunflecks for photosynthesis [@chazdonPhotosyntheticLightEnvironments1984]
+
+[I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations).
 Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear.
@@
 They don't have to be treated separately, but it should always be clear which is being referred to.
-And in the consequences for scaling section this is of course also relevant.
+And in the consequences for scaling section this is of course also relevant.]
 
-in line with Marielle's comment below; it will be important to distinguish between within plant variation vs across species variation.
-Understory species may have conservative  growth strategies because they are resource (light) limited.
-Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well.
-Their traits will be different from shade leaves of canopy trees as well.
-E.g.
-They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
-Those differences in traits are likely to affect their thermal properties as well]--Martijn Slot
-
-Leaf traits are shaped by intra canopy light and height, and show intracanopy plasticity with availability of light.
-Taller trees tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
-Higher irradiance has consequential adaptations in leaf morphology with smaller leaf area index in exposed canopies than shaded lower canopy or canopy interior, and well as having deeper leaf lobes in species such as Quercus.
-rubra.
-The table above summarizes the variation of different traits based on light regime, however, some of the traits might differ between species along the vertical profile, thus also varying in leaf thermal sensitivity.
-[Sack et al.
-2016]
-
-Additionally, passing light spots in the forest understory often caused by over canopy vegetation and wind speed called sunflecks cause a different type of morphological adaptation in shaded leaves, making them efficient in capturing 50% of light through sunflecks for photosynthesis.
-
-Leaf traits vary along the vertical canopy profile following the pattern of light availability across the gradient [@poorterMetaanalysisPlantResponses2019b].
-Leaves in the upper canopy that are exposed to more light differ in their traits compared to shaded leaves exposed to less light [@niinemetsWorldwideAnalysisWithincanopy2015].
 Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
@@
 **Many traits vary with height and/or between sun and shade leaves on the same species (Table 1).**
-Here, we focus on traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.
+Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.
 
@@
 Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
-Additionally, increase in leaf area also influences leaf temperature by maintaining greater leaf temperature than leaves with lower leaf area {Wright et al. 2019}
+Leaf temperature increases with leaf area {Wright et al. 2019}
 -Boundary layer resistance and thermal sensitivity also depends of leaf width, leaf size and structure.
-Increase in leaf width increases stomatal resistance-- dependent on stomtal activity and boundary layer resistance [@fauset_differences_2018] such as lobes, leaf elongation and serration, all dissapate heat more efficiently than non-lobed leaves, wide or non serrated leaves.
+Increase in leaf width increases stomatal resistance-- dependent on stomtal activity and boundary layer resistance [@fauset_differences_2018].
+Leaf characteristics such as lobes, leaf elongation and serration, all facilitate in dissapating heat more efficiently.
 Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;

### 2020-12-25 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming.
-Temperature is directly linked with plant metabolic processes and growth, thus even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011] This is leading to thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+Temperature is directly linked with plant metabolic processes and growth, thus even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
+This is leading to thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+Considering frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares] which may have further impacts on understory species ontogeny with rising temperatures.
 Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008].
@@
 These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.[refs]
+
+Cateogorization of sun and shade tolerance from Niinemetes and Valladares, 2006: "The five-level scale used for shade tolerance (1,
+very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%"
 

### 2020-12-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This is leading to thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
-Considering frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares] which may have further impacts on understory species ontogeny with rising temperatures.
+Considering frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares, 2006] which may have further impacts on understory species with rising temperatures.
 Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008].
@@
 
-With such changing forest dynamics, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy influenced by biophysical environmental gradient.This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
+While there is evidence for temperature driven shifts in forest dynamics, leaf-level temperature and metabolism changes influenced by biophysical conditions and warming in overstory and understory is still unclear.Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy influenced by biophysical environmental gradient.This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
 
@@
 
-Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with canopy height, canopy structure, across species and forests [@koike_leaf_2001] due to cumulative layers of vegetation.
-Generally, upper canopies are more exposed to irradiance and limit light through shading to canopy interiors, lower canopy layers and understories [@fauset_differences_2018; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
+Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, across species and forest types [@koike_leaf_2001] due to cumulative layers of vegetation.
+Generally, upper canopies in dense forests are exposed to greater PAR values and limit light to canopy interiors, lower canopy layers and understories.
+Shaded regions under canopy cover experience lower values of PAR in ranges of blue, orange and red.
+The gradient of light is more pronounced under broadleaved forests than conifers[@aussenacInteractionsForestStands2000; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
 (Fig. `r fig_NEON_vertical`).
@@
 Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months.
-Wind speed in dry season of a semi-deciduous forests at upper canopy has been observed to be higher than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*]
-This results in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [Daudet et al.
+Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season of than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*] 
+Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [Daudet et al.
 1999].
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of canopy leaves with small leaf sizes making them more advantageous in dissapating heat efficiently [Bauerle et al.
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that them more advantageous in dissapating heat efficiently than larger leaves[Bauerle et al.
 2011].
@@
 2006].
-Boundary layer thickness, depth and boundary layer resistance increases with leaf size and effective leaf width [along the vertical gradient], such that the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, therefore allowing small leaves to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 
@@
 
-Diurnal moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons.
+Diurnal moderating capacity differs between forests with varying canopy structure, crown density, and seasons.
 Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests.
-Increase in crown density and leaf area index decreases light and higher maximum temperature permeability in forests, thus maintaining lower maximum daytime temperature.
-Thus, pine forests have sparse stands resulting in decreased buffering capacity [De Frenne et al.
+Increase in crown density and leaf area index decreases light and higher maximum temperature permeability into forest understory, thus maintaining greater lower maximum daytime temperature.
+Comparitively open forests such as pine forests have sparse stands resulting in decreased buffering capacity in understory and closer air temperatures to nearby clearings [De Frenne et al.
 2019; @vonarxSpatiotemporalEffectsForest2012a].
-This relationship is reveresed during the night among, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]
+This relationship is reveresed during the night, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]
 
@@
 
-Leaf traits are shaped by intra canopy light and height, and show intra-canopy plasticity in accordance to intensity and availability of light.
-Taller trees tend to be exposed to greater irradiance than dense lower canopy and shorter understory trees shaded by overstory.
-Degrees of high irradiance has consequential adaptations in leaf morphology compared to degrees of low irradiance to shaded-ness [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al.
+Leaf traits are shaped by intra canopy light and height, and show intra-canopy plasticity in accordance with intensity and availability of light.
+Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.Degrees of high irradiance has consequential adaptations in leaf morphology compared to degrees of low irradiance to shaded-ness [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al.
 2018; @niinemetsWorldwideAnalysisWithincanopy2015].
@@
 
-Passing light in the forest understory often caused by over-canopy vegetation and wind speed called sunflecks cause a different type of morphological adaptation in shaded leaves that make them efficient in capturing >50% of light through sunflecks for photosynthesis [@chazdonPhotosyntheticLightEnvironments1984]
+Passing light in the forest understory often caused by over-canopy vegetation and wind speed called sunflecks cause a different type of morphological adaptation in shaded leaves [characterized by low LMA and greater effective leaf width] that make them efficient in capturing >50% of light through sunflecks for photosynthesis [@chazdonPhotosyntheticLightEnvironments1984]
 
@@
 Schuepp 1993; Sack et al. 2006].
+-Boundary layer thickness, depth and boundary layer resistance increases with leaf size and effective leaf width [along the vertical gradient], such that the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, therefore allowing small leaves to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 

### 2020-12-30 tables.Rmd
updated all

@@
   column_spec(6, width = "4cm") %>%
-  pack_rows("Leaf anatomy and morphological traits", 1, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Leaf biochemical and physiological traits", 14, 29, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf anatomy and morphological traits", 1, 16, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf biochemical and physiological traits", 18, 29, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-12-30 vertical-thermal-review.Rmd
updated all

@@
 While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming.
-Temperature is directly linked with plant metabolic processes and growth, thus even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
-This is leading to thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
-Considering frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares, 2006] which may have further impacts on understory species with rising temperatures.
-Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellMechanismsPlantSurvival2008].
+As a result, understories experience a different set of pressures with rising temperature than overstory [Bartholomew et al.
+2020].
+Since temperature is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
+Evidence shows reduced growth of understory trees with warmer temperatures in mesic temperate forests[Rollinson et al.
+2020]along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares, 2006] which may have further impacts on understory species with rising temperatures.
+Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [McDowell et al.
+2020].
 Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al.
@@
 
-While there is evidence for temperature driven shifts in forest dynamics, leaf-level temperature and metabolism changes influenced by biophysical conditions and warming in overstory and understory is still unclear.Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy influenced by biophysical environmental gradient.This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
+While there is evidence for temperature driven shifts in forest dynamics, leaf-level temperature and metabolism changes influenced by biophysical conditions and warming in overstory and understory is still unclear.
+Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy influenced by biophysical environmental gradient.This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
 
@@
 Generally, upper canopies in dense forests are exposed to greater PAR values and limit light to canopy interiors, lower canopy layers and understories.
-Shaded regions under canopy cover experience lower values of PAR in ranges of blue, orange and red.
+Shaded regions under canopy cover experience lower values and quality of PAR in ranges of blue, orange and red.
 The gradient of light is more pronounced under broadleaved forests than conifers[@aussenacInteractionsForestStands2000; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
@@
 nocite: | 
-  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998
+  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; Panditharathna et al.
+2008; Baltzer and Thomas 2005; Kusi and Karasi 2020
 ---

### 2020-12-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming.
-As a result, understories experience a different set of pressures with rising temperature than overstory [Bartholomew et al.
+As a result, understories experience a different set of microclimatic pressures with rising temperature than overstory [Bartholomew et al.
 2020].
 Since temperature is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
-Evidence shows reduced growth of understory trees with warmer temperatures in mesic temperate forests[Rollinson et al.
-2020]along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[Rollinson et al.
+2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
 Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares, 2006] which may have further impacts on understory species with rising temperatures.
@@
 Generally, upper canopies in dense forests are exposed to greater PAR values and limit light to canopy interiors, lower canopy layers and understories.
-Shaded regions under canopy cover experience lower values and quality of PAR in ranges of blue, orange and red.
+Shaded regions under canopy cover experience lower values and quality of PAR [in ranges of blue, orange and red].
 The gradient of light is more pronounced under broadleaved forests than conifers[@aussenacInteractionsForestStands2000; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]

### 2020-12-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming.
-As a result, understories experience a different set of microclimatic pressures with rising temperature than overstory [Bartholomew et al.
-2020].
+As a result, understories experience a different set of microclimatic pressures with rising temperature than overstory [@bartholomewSmallTropicalForest2020].
 Since temperature is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
-Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[Rollinson et al.
-2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[Duque et al.
+2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
 Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares, 2006] which may have further impacts on understory species with rising temperatures.
-Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [McDowell et al.
-2020].
-Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions (Stark et al.
-2020).
+Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellPervasiveShiftsForest2020].
+Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 Each of these changes have potential feedbacks to climate change.
@@
 Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season of than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*] 
-Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [Daudet et al.
-1999].
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that them more advantageous in dissapating heat efficiently than larger leaves[Bauerle et al.
-2011].
+Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [@daudetWindSpeedLeaf1999a].
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that them more advantageous in dissapating heat efficiently than larger leaves[@bauerleSeparatingFoliarPhysiology2011].
 In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999].
-Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; Stockes et al.
-2006].
+Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006].
 
 **Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
-Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [De Frenne et al.
-2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
+Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
 
@@
 2018; @niinemetsWorldwideAnalysisWithincanopy2015].
-This dichotomy in leaf traits that vary with light intensity generally differ within intra-canopy, however the degree of intra-canopy plasticity also varies across species and geographically [Sack et al.
-2006; Ishii et al.
-2018].
+This dichotomy in leaf traits that vary with light intensity generally differ within intra-canopy, however the degree of intra-canopy plasticity also varies across species and geographically [@sackHowStrongIntracanopy2006; @ishiiVariationIntracrownLeaf2018].
 
@@
 
-Cateogorization of sun and shade tolerance from Niinemetes and Valladares, 2006: "The five-level scale used for shade tolerance (1,
+Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006: "The five-level scale used for shade tolerance (1,
 very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%"
@@
 
-Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [Chen et al.
-2020], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)
+Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [@chenLeafEconomicsSpectrum2020a], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)
 

### 2020-12-31 vertical-thermal-review.Rmd
updating references

@@
 Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests [ref].
-Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000].
-Evidence shows that larger trees suffer more than small trees during drought, which may contribute to increasing tree mortality in forests around the world [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a].
+Evidence shows that larger trees suffer during drought resulting in heat stressed canopy-leaves, which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to occur [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015].
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
+However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continues evapotranspiration and consequential buffering of the understory [REF]
 
 While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming.
-As a result, understories experience a different set of microclimatic pressures with rising temperature than overstory [@bartholomewSmallTropicalForest2020].
+As a result, understories experience a different set of microclimatic pressures, e.g limited light and related foliar traits, inter-species competition etc..
+with rising temperature [@bartholomewSmallTropicalForest2020].
 Since temperature is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
-Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[Duque et al.
-2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
-Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [Niinemetes and Valladares, 2006] which may have further impacts on understory species with rising temperatures.
+Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [@niinemetsControversyTraitsConferring2006] which may have further impacts on understory species with rising temperatures.
 Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellPervasiveShiftsForest2020].
@@
 
-Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, across species and forest types [@koike_leaf_2001] due to cumulative layers of vegetation.
+Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, across species and forest types due to cumulative layers of vegetation [@koike_leaf_2001].
 Generally, upper canopies in dense forests are exposed to greater PAR values and limit light to canopy interiors, lower canopy layers and understories.
 Shaded regions under canopy cover experience lower values and quality of PAR [in ranges of blue, orange and red].
-The gradient of light is more pronounced under broadleaved forests than conifers[@aussenacInteractionsForestStands2000; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
+The gradient of light is more pronounced under broadleaved forests than conifers[@aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
 (Fig. `r fig_NEON_vertical`).
 In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
- Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [Bartemucci et al.
-2005].
+ Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
 This difference in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient.
-Leaves along a vertical intra-canopy gradient thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fausetDifferencesLeafThermoregulation2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; Niinemets and Valladares 2006].
+Leaves along a vertical intra-canopy gradient thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; Niinemets and Valladares 2006].
 
@@
 *(fill in some specifics from the NEON analysis)*
-Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [Barnard and Bauerle, 2016; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months.
+Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months.
+
 Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season of than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*] 
 Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that them more advantageous in dissapating heat efficiently than larger leaves[@bauerleSeparatingFoliarPhysiology2011].
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that make them more advantageous in dissapating heat efficiently than larger leaves[@bauerleSeparatingFoliarPhysiology2011].
 In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999].
@@
 **Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
-Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012] and in the northwestern United States [@davis_microclimatic_2019].
+Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019].
+Even though canopy cover buffers maximum air temperatures in the understory, inevitable increase in understory temperatures continue to be observed as a result of global increase in temperatures, however the buffering by canopy cover smoothens this warming in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]
 
@@
 Increase in crown density and leaf area index decreases light and higher maximum temperature permeability into forest understory, thus maintaining greater lower maximum daytime temperature.
-Comparitively open forests such as pine forests have sparse stands resulting in decreased buffering capacity in understory and closer air temperatures to nearby clearings [De Frenne et al.
-2019; @vonarxSpatiotemporalEffectsForest2012a].
+Comparitively open forests such as pine forests have sparse stands resulting in decreased buffering capacity in understory and closer air temperatures to nearby clearings [@@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
 This relationship is reveresed during the night, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]
@@
 
-[Add something about canopy temperature is a better proxy for determining leaf temperature than air temperature; in wet season leaf temperature can be linked to intensity of irradiance and air temperature, however, in dry season, air temperature better determines leaf temperature due to general increase in irradiance intensity and mixing of air movement along the gradient caused by fewer leaves [rey sanchez et al]]
-
 *(Add something about risk of freeze.)*
@@
 
-With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications for understory seedling growth, especially in high altitude temperate pine forests and montane forests  [@vonarxSpatiotemporalEffectsForest2012a, @fauset_differences_2018] (to clarify why?
+With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications for understory seedling growth, especially in high altitude temperate pine forests and montane forests  [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018] (to clarify why?
 are these high elevation forests limited by low temperature and low rh?)

### 2020-12-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages.
-Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
+Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
 This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient.[Bertrand et al.
@@
 
-The capacity of buffering largly depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions.
-Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019].
+The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions.
+Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
 Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests [ref].
-Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a].
-Evidence shows that larger trees suffer during drought resulting in heat stressed canopy-leaves, which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to occur [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
-2015].
+Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+Evidence shows that larger trees suffer during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to occur [@@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
@@
 
-While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming.
-As a result, understories experience a different set of microclimatic pressures, e.g limited light and related foliar traits, inter-species competition etc..
-with rising temperature [@bartholomewSmallTropicalForest2020].
-Since temperature is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
-Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
+While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming [@bertrandCommentForestMicroclimate2020].
+As a result, understories experience a different set of microclimatic pressures with rising temperature in combination with conditions of limited light and related foliar traits, inter-species resource competition etc..
+[@bartholomewSmallTropicalForest2020].
+Since temperature sensitivity is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
+Tree-ring evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with other studies that show thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
 Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [@niinemetsControversyTraitsConferring2006] which may have further impacts on understory species with rising temperatures.
@@
 Atlantic forests during wet season tended to have higher $T_{air}$, and $VPD$ in upper canopy and canopy gaps than lower in understory where $T_{air}$ and $VPD$ were lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
-This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
+This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential.
+[@niinemetsPhotosyntheticAcclimationSimultaneous2004a].
+In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas as understory was insensitive [[@rollinsonClimateSensitivityUnderstory, 2020]
 

### 2021-01-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-Evidence shows that larger trees suffer during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to occur [@@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+Evidence shows that larger trees suffer during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to occur [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
@@
 Tree-ring evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with other studies that show thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
-Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [@niinemetsControversyTraitsConferring2006] which may have further impacts on understory species with rising temperatures.
+Considering continued frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [@niinemetsControversyTraitsConferring2006] which may have further impacts on understory species with rising temperatures.
+
 Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellPervasiveShiftsForest2020].
@@
 
-While there is evidence for temperature driven shifts in forest dynamics, leaf-level temperature and metabolism changes influenced by biophysical conditions and warming in overstory and understory is still unclear.
+While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in overstory and understory is still unclear.
 Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy influenced by biophysical environmental gradient.This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
@@
 In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
+
 Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
@@
 Increase in crown density and leaf area index decreases light and higher maximum temperature permeability into forest understory, thus maintaining greater lower maximum daytime temperature.
-Comparitively open forests such as pine forests have sparse stands resulting in decreased buffering capacity in understory and closer air temperatures to nearby clearings [@@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
+Comparitively open forests such as pine forests have sparse stands resulting in decreased buffering capacity in understory and closer air temperatures to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
 This relationship is reveresed during the night, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]
@@
 
-Leaf traits are shaped by intra canopy light and height, and show intra-canopy plasticity in accordance with intensity and availability of light.
-Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.Degrees of high irradiance has consequential adaptations in leaf morphology compared to degrees of low irradiance to shaded-ness [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al.
+Leaf traits are shaped by intra canopy light and height, and show intra-canopy foliar plasticity in accordance with the intensity and availability of light.
+Degrees of high irradiance gives rise to sun-leaf morphology compared to degrees of low irradiance characterized by shade shapes shade-leaf morphology [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al.
 2018; @niinemetsWorldwideAnalysisWithincanopy2015].
-This dichotomy in leaf traits that vary with light intensity generally differ within intra-canopy, however the degree of intra-canopy plasticity also varies across species and geographically [@sackHowStrongIntracanopy2006; @ishiiVariationIntracrownLeaf2018].
+The degree of intra-canopy plasticity also varies across species and geographically [@sackHowStrongIntracanopy2006; @ishiiVariationIntracrownLeaf2018].
 
-Adaptions of traits to ranges of sun-exposure or shaded-ness also vary within a forest strata. 
-Understory species may have conservative  growth strategies because they are resource (light) limited.
-Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well.
-Their traits will be different from shade leaves of canopy trees as well.
-E.g.
-They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.
+Adaptions of traits to ranges of sun-exposure or shaded-ness vary within a forest strata among trees of different sizes.Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
+Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
 Additionally, shade leaves in the understory are of three categories; 1.
@@
 3 understory plants that complete their lifecycle in the shaded understory.
-These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.[refs]
+These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.[refs, -Martijn Slot]
 
-Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006: "The five-level scale used for shade tolerance (1,
-very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%"
+The variation and quanitity of light falling on leaves is perceived through a network of phytohormones that regulate leaf acclimation to light conditions and consequential strategies to regulate growth [Mathur et al.
+2018].
+Under conditions of low light, understory species may have conservative  growth strategies because they are resource (light) limited which leads to insuffiencient production of ATP for carbon fixation and carbohydrate synthesis
+.Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximum light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well.
+Their traits will be different from shade leaves of canopy trees as well.
+E.g.
+They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.
 
-Passing light in the forest understory often caused by over-canopy vegetation and wind speed called sunflecks cause a different type of morphological adaptation in shaded leaves [characterized by low LMA and greater effective leaf width] that make them efficient in capturing >50% of light through sunflecks for photosynthesis [@chazdonPhotosyntheticLightEnvironments1984]
+Passing light called sunflecks in the forest understory often caused by over-canopy vegetation gaps and wind speed cause a different type of morphological adaptation in shaded leaves [characterized by low LMA and greater effective leaf width] that make them efficient in capturing >50% of light through sunflecks for photosynthesis [@chazdonPhotosyntheticLightEnvironments1984]
 
@@
 I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
+-"Our study suggests that height plays an important role in the leaf traits, in contrast with previous findings (Baranski, 1975; Sack et al., 2006).In Quercus.
+alba and Quercus.
+velutina, loba-tion decreased with increasing tree height.
+Lobation allows air passage and light penetration in the tree canopy (Sack et al., 2006; Semchenko & Zobel, 2007)" [@kusiPlasticLeafMorphology2020]
 

### 2021-01-02 vertical-thermal-review.Rmd
updated

@@
 
-Adaptions of traits to ranges of sun-exposure or shaded-ness vary within a forest strata among trees of different sizes.Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
+Adaptions of traits to ranges of sun-exposure or shaded-ness vary within a forest strata among trees of different sizes.
 Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
@@
 2018].
-Under conditions of low light, understory species may have conservative  growth strategies because they are resource (light) limited which leads to insuffiencient production of ATP for carbon fixation and carbohydrate synthesis
-.Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximum light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well.
-Their traits will be different from shade leaves of canopy trees as well.
-E.g.
+Under conditions of low light, understory species may have conservative  growth strategies because they are resource (light) limited which leads to insuffiencient production of ATP for carbon fixation and carbohydrate synthesis.
+
+Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.As a result the different combination of stressors experienced by taller trees with foliage exposed to greater PAR are different from the combination of stressors experienced by understory shorter trees with shaded/semi-shaded leaves [Niinemetes et al.
+2010]
+"Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximum light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]
+Additionally, increased exposure to light increases the carbon reserve, plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited.
+However, understory shaded/younger trees have comparitively less carbon reserve due to light limitation as a primary stressor, decreased plant biomass, increased leaf biomass, and drecreased evapotransition demands.
+Larger carbon reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature].
+Thus, overall stress sensitivity decreases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]
+
+Their traits will be different from shade leaves of canopy trees as well. E.g.
 They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.

### 2021-01-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-01-01 vertical-thermal-review.Rmd
@@
 
-The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions.
+The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions[@senfCanopyMortalityHas2018a].
 Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
@@
 Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-Evidence shows that larger trees suffer during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to occur [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+Evidence shows that larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
@@
 As a result, understories experience a different set of microclimatic pressures with rising temperature in combination with conditions of limited light and related foliar traits, inter-species resource competition etc..
-[@bartholomewSmallTropicalForest2020].
-Since temperature sensitivity is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011].
+[@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
+Since temperature sensitivity is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
 Tree-ring evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with other studies that show thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
@@
 
-Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season of than in upper canopy wet season [@rey-sanchez_spatial_2016].[*double check*] 
+Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season than in upper canopy wet season [@rey-sanchez_spatial_2016] [*double check*]. 
 Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [@daudetWindSpeedLeaf1999a].
@@
 Diurnal moderating capacity differs between forests with varying canopy structure, crown density, and seasons.
-Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests.
+Tropical, temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests.
 Increase in crown density and leaf area index decreases light and higher maximum temperature permeability into forest understory, thus maintaining greater lower maximum daytime temperature.
@@
 [@niinemetsPhotosyntheticAcclimationSimultaneous2004a].
-In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas as understory was insensitive [[@rollinsonClimateSensitivityUnderstory, 2020]
+In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas as understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
 
@@
 
-Leaf traits are shaped by intra canopy light and height, and show intra-canopy foliar plasticity in accordance with the intensity and availability of light.
-Degrees of high irradiance gives rise to sun-leaf morphology compared to degrees of low irradiance characterized by shade shapes shade-leaf morphology [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al.
+Leaf traits are shaped by intra-canopy light and height, and show intra-canopy foliar plasticity in accordance with the intensity and availability of light.
+Degrees of high irradiance gives rise to sun-leaf morphology compared to degrees of low irradiance that is characterized by shade shapes shade-leaf morphology [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al.
 2018; @niinemetsWorldwideAnalysisWithincanopy2015].
@@
 
-Adaptions of traits to ranges of sun-exposure or shaded-ness vary within a forest strata among trees of different sizes.Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
 Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
@@
 
-The variation and quanitity of light falling on leaves is perceived through a network of phytohormones that regulate leaf acclimation to light conditions and consequential strategies to regulate growth [Mathur et al.
+The variation and quanitity of light falling on leaves is perceived through a network of phytohormones that regulate leaf acclimation to light conditions and consequential strategies to regulate growth and traits.
+Under conditions of low light, understory species may have conservative  growth strategies because they are resource (light) limited which leads to insuffiencient production of ATP for carbon fixation and carbohydrate synthesis[Mathur et al.
 2018].
-Under conditions of low light, understory species may have conservative  growth strategies because they are resource (light) limited which leads to insuffiencient production of ATP for carbon fixation and carbohydrate synthesis
-.Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximum light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well.
-Their traits will be different from shade leaves of canopy trees as well.
-E.g.
+
+Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.As a result the different combination of stressors taller trees with foliage exposed to greater PAR are different from the combination of stressors experienced by understory shorter trees with shaded/semi-shaded leaves [Niinemetes et al.
+2010]
+"Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximum light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]
+Additionally, increased exposure to light increases carbon reserve, plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited, however with available water overall sensitivity to stress decreases.
+However, understory shaded/younger trees have comparitively less carbon reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, with greater overall sensitivitty to stress.
+Larger carbon reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature].
+Thus, overall stress sensitivity decreases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]
+Their traits will be different from shade leaves of canopy trees as well. E.g.
 They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.

### 2021-01-06 vertical-thermal-review.Rmd
attempting to organize paragraphs

@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient.[Bertrand et al.
-2011]
+This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and in turn whole plant performance [Bertrand et al.
+2011].
 
@@
 *(paragraph on leaf area distribution, including figure ([separate from or combined with current Fig. 2- see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
-The vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; Smith et al.
-2019 ; Parker et al.
-2019) and following disturbance (e.g., Parker et al.
-2002 ; Almeida et al.
-2016 ; Stark et al.
-2020 in press).
+The vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
 In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season but excluding periods (in deciduous forests) when no leaves are present.
 
-Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, across species and forest types due to cumulative layers of vegetation [@koike_leaf_2001].
+Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, across species and forest types [@koike_leaf_2001].
 Generally, upper canopies in dense forests are exposed to greater PAR values and limit light to canopy interiors, lower canopy layers and understories.
 Shaded regions under canopy cover experience lower values and quality of PAR [in ranges of blue, orange and red].
-The gradient of light is more pronounced under broadleaved forests than conifers[@aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; Parker 1995; @poorterMetaanalysisPlantResponses2019b]
+The gradient of light is more pronounced under broadleaved forests than conifers[@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
 (Fig. `r fig_NEON_vertical`).
-In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
+
+In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]
 
 Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
-This difference in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient.
-Leaves along a vertical intra-canopy gradient thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; Niinemets and Valladares 2006].
+However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and movements caused by wind speed called sunflecks that are often intermittent and breif.
+Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]
+
+This diffence in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient.
+Leaves along a vertical intra-canopy gradient, as well as in inter-canopies, thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].
 
@@
 Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season than in upper canopy wet season [@rey-sanchez_spatial_2016] [*double check*]. 
-Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with increase in wind velocity [@daudetWindSpeedLeaf1999a].
+Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
 Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that make them more advantageous in dissapating heat efficiently than larger leaves[@bauerleSeparatingFoliarPhysiology2011].
 In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999].
-Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006].
+Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
+[ref]
 
@@
 
-Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal Brazil [@fauset_differences_2018, Tymen et al.
-2017].
-
-"max temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground), see Fig.
-4 - again illustrating buffering effect of the canopy".However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016], similar results were observed in in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+Seasonally, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal Brazil [@fauset_differences_2018, Tymen et al.
+2017].Similarly, maximum temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground, see Fig.
+4 - again illustrating buffering effect of the canopy).However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016], similar results were observed in in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
 
@@
 
-With rising temperatures, this trend in maximum $T_{air}$ and RH would have implications for understory seedling growth, especially in high altitude temperate pine forests and montane forests  [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018] (to clarify why?
-are these high elevation forests limited by low temperature and low rh?)
-
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$.
@@
 [@niinemetsPhotosyntheticAcclimationSimultaneous2004a].
-In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas as understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
+In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
 
 **Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
--	Higher in understory, particularly at dusk (Koike et al. 2001).
--	Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
-- Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
+	Higher in understory, particularly at dusk [@koike_leaf_2001].
+Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
+Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
 
@@
 Taller trees don't necessarily increase buffering [@zellweger_seasonal_2019].
+"They found that taller forests (which coincided with forests of higher LAI) had lower mean and max values of VPD and temperature" [Jucker, T., Hardwick, S.R., Both, S., Elias, D.M., Ewers, R.M., Milodowski, D.T., Swinfield, T.
+and Coomes, D.A., 2018.
+Canopy structure and topography jointly constrain the microclimate of human-modified tropical landscapes.
+Global change biology, 24(11), pp.5243-5258]
+.See Jucker et al. 2018 - max height was highly correlated to both LAI and surface roughness
+
 Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
@@
 
-Leaf traits are shaped by intra-canopy light and height, and show intra-canopy foliar plasticity in accordance with the intensity and availability of light.
-Degrees of high irradiance gives rise to sun-leaf morphology compared to degrees of low irradiance that is characterized by shade shapes shade-leaf morphology [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006] within- and across species [Ishii et al.
-2018; @niinemetsWorldwideAnalysisWithincanopy2015].
-The degree of intra-canopy plasticity also varies across species and geographically [@sackHowStrongIntracanopy2006; @ishiiVariationIntracrownLeaf2018].
+Leaf traits are shaped on the basis of light and height in forest vertical strata, and show different degrees of foliar plasticity in accordance with the intensity and availability of light.
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody species [@niinemetsLeafAgeDependent2016a].
+A network of phytohormones in leaves perceive the intensity of light falling on the surface and thus respond through receptors that enable acclimation or avoidance depending on the light scenario [Casal, 2013].
+As a result, degrees of high irradiance gives rise to variations of sun-leaf morphology compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf morphology [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
+2016].
+Increasing number of studies point towards the importance for incorporating sun and shade in modelling leaf economic spectrum (LES) within-canopy gradient[@chenLeafEconomicsSpectrum2020a] because LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
+However, there is also concern for ambiguity in simplified cateogorization of 'sun' and 'shade' leaves due to variation based on intensities of light within-canopy, across functional groups, in canopy gaps, height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
+
+**Inter-species vertical gradient leaf trait variation**
+
+Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
+As a result the combination of stressors taller trees with foliage exposed to greater PAR and $VPD$ experience are different from the combination of stressors experienced by understory shorter trees with shaded/semi-shaded leaves [@niinemetsResponsesForestTrees2010a]
+"Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximize light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]
 
 Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
-Additionally, shade leaves in the understory are of three categories; 1.
+
+"Additionally, shade leaves in the understory are of three categories; 1.
 shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded.
@@
 3 understory plants that complete their lifecycle in the shaded understory.
-These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.[refs, -Martijn Slot]
+These groups are likely to have different traits, and if we expect'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered".[refs, -Martijn Slot]
+
+Leaf traits have been observed to exhibit greater plasticity on the basis of area than mass 
+within-canopy and across inter-canopies [Keenan and Niinemetes, 2016]
+
+**Inter-species ontogeny and height leaf trait variation**
 
@@
 
-Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.As a result the different combination of stressors taller trees with foliage exposed to greater PAR are different from the combination of stressors experienced by understory shorter trees with shaded/semi-shaded leaves [Niinemetes et al.
-2010]
-"Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximum light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]
 Additionally, increased exposure to light increases carbon reserve, plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited, however with available water overall sensitivity to stress decreases.
-However, understory shaded/younger trees have comparitively less carbon reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, with greater overall sensitivitty to stress.
+Understory shaded/younger trees have comparitively less carbon reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, with greater overall sensitivitty to stress.
 Larger carbon reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature].
@@
 
-Passing light called sunflecks in the forest understory often caused by over-canopy vegetation gaps and wind speed cause a different type of morphological adaptation in shaded leaves [characterized by low LMA and greater effective leaf width] that make them efficient in capturing >50% of light through sunflecks for photosynthesis [@chazdonPhotosyntheticLightEnvironments1984]
+**Intra-canopy on the same species leaf trait variation**
+
+Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which shapes leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+The differences among these traits contribute to leaf heat dissapation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1.
+summaries relevant traits for thermal sensitivity.
 
@@
 And in the consequences for scaling section this is of course also relevant.]
-
-Increasing number of studies point towards the importance of incorporating vertical canopy leaf trait variation in modelling for leaf economic spectrum (LES) [@chenLeafEconomicsSpectrum2020a], leaf thermoregulation and thermal sensitivity which will provide a more accurate representation of within-canopy gradient response to increasing temperatures.[@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016; @niinemetsWorldwideAnalysisWithincanopy2015, @chenLeafEconomicsSpectrum2020, REFS)
-
 [Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
@@
 nocite: | 
-  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; Panditharathna et al.
-2008; Baltzer and Thomas 2005; Kusi and Karasi 2020
+  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @kusiPlasticLeafMorphology2020
 ---

### 2021-01-10 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming [@bertrandCommentForestMicroclimate2020].
-As a result, understories experience a different set of microclimatic pressures with rising temperature in combination with conditions of limited light and related foliar traits, inter-species resource competition etc..
+While forest microclimates are insulated from macroclimatic extremes, they are still experiencing warming trends [@bertrandCommentForestMicroclimate2020].
+As a result, understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, etc.
 [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
-Since temperature sensitivity is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
-Tree-ring evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with other studies that show thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013].
-Considering continued frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [@niinemetsControversyTraitsConferring2006] which may have further impacts on understory species with rising temperatures.
+Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
+Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+Tree-ring evidence in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
+In addition, understory growth phenology is sensitive to warming year round, and especially in spring [Zohnner_Renner 2019].
 
-Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellPervasiveShiftsForest2020].
+Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
+Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards disturbed forests with younger trees [@mcdowellPervasiveShiftsForest2020].
 Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
@@
 
-Leaf traits are shaped on the basis of light and height in forest vertical strata, and show different degrees of foliar plasticity in accordance with the intensity and availability of light.
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody species [@niinemetsLeafAgeDependent2016a].
-A network of phytohormones in leaves perceive the intensity of light falling on the surface and thus respond through receptors that enable acclimation or avoidance depending on the light scenario [Casal, 2013].
-As a result, degrees of high irradiance gives rise to variations of sun-leaf morphology compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf morphology [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
+Leaf traits are shaped on the basis of light and height and show different degrees of foliar plasticity in accordance with the intensity and availability of light which also optimize water uptake, heat and C uptake by leaves in forest vertical strata.
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a].
+A network of sensory photoreceptors in leaves such as phytochromes among others mediate response to PAR, and ratio of red to far red radiation [R:FR] falling on the surface.
+This enables acclimation or avoidance response through development of necessary traits [Casal, 2013].
+As a result, degrees of high irradiance gives rise to variations of sun-leaf traits  compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
 2016].
+
 Increasing number of studies point towards the importance for incorporating sun and shade in modelling leaf economic spectrum (LES) within-canopy gradient[@chenLeafEconomicsSpectrum2020a] because LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in simplified cateogorization of 'sun' and 'shade' leaves due to variation based on intensities of light within-canopy, across functional groups, in canopy gaps, height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
-
-**Inter-species vertical gradient leaf trait variation**
-
-Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
-As a result the combination of stressors taller trees with foliage exposed to greater PAR and $VPD$ experience are different from the combination of stressors experienced by understory shorter trees with shaded/semi-shaded leaves [@niinemetsResponsesForestTrees2010a]
-"Spending their lifecycle in the shade means they have different pressures [such as competition for light or adapting leaf traits to maximize light capture and use efficiency] than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]
-
-Cateogorization of sun and shade tolerance from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
-
+However, there is also concern for ambiguity in simplified cateogorization of 'sun' and 'shade' leaves due to variation based on light within-canopy, across functional groups, in canopy gaps, height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
+Classification for sun and shade tolerance rankings have been highted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
+2018] such as from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
 "Additionally, shade leaves in the understory are of three categories; 1.
@@
 
-Leaf traits have been observed to exhibit greater plasticity on the basis of area than mass 
-within-canopy and across inter-canopies [Keenan and Niinemetes, 2016]
+**Inter-species vertical gradient leaf trait variation**
 
-**Inter-species ontogeny and height leaf trait variation**
+Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
+As a result, sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and more lobated than shade leaves
+[Mathur et al.
+2018].
+However, the combination of stressors taller trees' foliage experience are different from the combination of stressors understory shorter trees's foliage experience.
+For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, wind speed, etc.
+whereas for understory--shade or limitation of light, niche competition, herbivory etc.
+These pressures also contribute to the development of necessary traits that is primarily mediated by light response.Leaves act as windows to plant performance enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plant to respond differently to forcoming stress [@niinemetsResponsesForestTrees2010a]
 
-The variation and quanitity of light falling on leaves is perceived through a network of phytohormones that regulate leaf acclimation to light conditions and consequential strategies to regulate growth and traits.
-Under conditions of low light, understory species may have conservative  growth strategies because they are resource (light) limited which leads to insuffiencient production of ATP for carbon fixation and carbohydrate synthesis[Mathur et al.
+"Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn].
+[@niinemetsResponsesForestTrees2010a].
+Under conditions of low light, understory species may have conservative  growth strategies because they are resource (light) limited which leads to insuffiencient production of ATP for carbon fixation and carbohydrate synthesis [Mathur et al.
 2018].
+Canopy foliar traits will be different from shade leaves of canopy trees as well.
+E.g.They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
+Those differences in traits are likely to affect their thermal properties as well.
 
-Additionally, increased exposure to light increases carbon reserve, plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited, however with available water overall sensitivity to stress decreases.
-Understory shaded/younger trees have comparitively less carbon reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, with greater overall sensitivitty to stress.
-Larger carbon reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature].
-Thus, overall stress sensitivity decreases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]
-Their traits will be different from shade leaves of canopy trees as well. E.g.
-They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Those differences in traits are likely to affect their thermal properties as well.
+However, even with distinct features of sun and shade traits, trait adaptation to light is done in such a way where light use efficiency for photosynthesis is maximized at both high or low light.
+In light limited conditions, shade leaves increase light capture efficiency through increasing their specific leaf area [SLA] and by investing in chlorophyll at a mass basis--"leads to a similar chlorphyll content on an area basis compared with sun leaves.
+This makes shade leaves realize a similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
+1996, Poorter et al.
+2000, Rozendaal et al.
+2006]
+
+With increasing height and age, competion for light decreases as availibity  of light increases in dense forests.
+However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements.
+Early successional species are considered more shade intolerant than late successional species that are more shade tolerant.
+Ontogenic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$) in shade tolerators than in intolerators, greater photosynthetic capacity in low light and greater carbon gain capacities[Niinemets, 2006].Rozendaal et al.
+2006 has more info about this.
 
@@
 Are traits shaped more by height or light?
--	 “Height is more important than light in determining leaf morphology in a tropical forest” (Cavaleri et al. 2010)
+Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
 - (Cavaleri et al. 2008) 
@@
 
+Additionally, increased exposure to light increases non structural carbon reserve (NSC), plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited, however with available water overall sensitivity to stress decreases.
+Understory shaded/younger trees have comparitively less NSC reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, as well as greater investment in light harvesting than in photosynthesis--with greater overall sensitivity to stress and reduced tolerance.
+Larger NSC reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature].
+Thus, overall stress sensitivity decreases and tolerance increases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]
+
+Interaction between degrees of shade x ontogeny x height with rising temperature needs more scietific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
+
 ![**Figure 4.

### 2021-01-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf traits are shaped on the basis of light and height and show different degrees of foliar plasticity in accordance with the intensity and availability of light which also optimize water uptake, heat and C uptake by leaves in forest vertical strata.
+Leaf traits are shaped on the basis of light and height and show different degrees of foliar plasticity in accordance with the intensity and availability of light which also optimize water uptake, heat and C uptake by leaves across forest vertical strata.
 After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a].
@@
 However, there is also concern for ambiguity in simplified cateogorization of 'sun' and 'shade' leaves due to variation based on light within-canopy, across functional groups, in canopy gaps, height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
-Classification for sun and shade tolerance rankings have been highted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
+Classification for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
 2018] such as from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
@@
 As a result, sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and more lobated than shade leaves
-[Mathur et al.
-2018].
+[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].
+
 However, the combination of stressors taller trees' foliage experience are different from the combination of stressors understory shorter trees's foliage experience.
@@
 whereas for understory--shade or limitation of light, niche competition, herbivory etc.
-These pressures also contribute to the development of necessary traits that is primarily mediated by light response.Leaves act as windows to plant performance enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plant to respond differently to forcoming stress [@niinemetsResponsesForestTrees2010a]
+These pressures also contribute to the development of necessary traits that is primarily mediated by light response.
+Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature.
+Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
+Adaptive traits in understory enable light capture efficiency while maintaing larger width leaves with comparitevely fewer stomtal density as a result of available water, lesser $VPD$, and wind speed.
+"Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn].Those differences in traits are likely to affect their thermal properties as well.Leaves act as windows to plant performance enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plant to respond differently to forcoming stress [Rozendaal et al.
+2006; @niinemetsResponsesForestTrees2010a]
 
-"Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn].
-[@niinemetsResponsesForestTrees2010a].
-Under conditions of low light, understory species may have conservative  growth strategies because they are resource (light) limited which leads to insuffiencient production of ATP for carbon fixation and carbohydrate synthesis [Mathur et al.
-2018].
-Canopy foliar traits will be different from shade leaves of canopy trees as well.
-E.g.They tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
-Those differences in traits are likely to affect their thermal properties as well.
-
-However, even with distinct features of sun and shade traits, trait adaptation to light is done in such a way where light use efficiency for photosynthesis is maximized at both high or low light.
+However, even with distinct features of sun and shade traits, trait adaptation to light is done in such a way where light use efficiency for photosynthesis is maximized at both high or low light, while also adapting to other stressors as mentioned above.
 In light limited conditions, shade leaves increase light capture efficiency through increasing their specific leaf area [SLA] and by investing in chlorophyll at a mass basis--"leads to a similar chlorphyll content on an area basis compared with sun leaves.
@@
 Ontogenic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$) in shade tolerators than in intolerators, greater photosynthetic capacity in low light and greater carbon gain capacities[Niinemets, 2006].Rozendaal et al.
-2006 has more info about this.
+2006, Koike et al.
+2001 has more info about this.
 

### 2021-01-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf traits are shaped on the basis of light and height and show different degrees of foliar plasticity in accordance with the intensity and availability of light which also optimize water uptake, heat and C uptake by leaves across forest vertical strata.
+Leaf traits are shaped on the basis of light and height and show different degrees of foliar plasticity in accordance with the intensity and availability of light which also optimizes water uptake, heat and C uptake by leaves across forest vertical strata.
 After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a].

### 2021-01-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
Edits up to L180.

@@
 
-**We lack a systematic, up to date understanding of biophysical and biological patterns across this gradient, how these affect leaf-level processes, and in turn how it affects ecosystems (Fig.
+**We lack a systematic, up to date understanding of biophysical and biological patterns across the vertical canopy gradient in forests, how this gradient affects leaf-level processes, and in turn how it affects ecosystems (Fig.
 ` r fig_schematic`).**
@@
 i.	How does the biophysical environment vary with height in forests?
-("What aspects of biophysical envt?
-Microclimate and vertical plant structure/phenology/etc (ok yes, see this in fig ` r fig_schematic`)"?--Elsa:
-So change to micrometeorological environment? Not biological just physical in this question, right?
 
@@
 
-iii. How do vertical changes in biophysical environment conditions and leaf traits combine to affect leaf temperature [and in turn photosynthetic capacity]??
+iii. How do vertical changes in biophysical environment conditions and leaf traits interact to affect leaf temperature [and in turn photosynthetic capacity]??
 
-iv.	How does leaf metabolism respond to temperature in canopy and understory settings?
+iv.	How does leaf metabolism respond to [air or leaf?] temperature in canopy and understory settings?
 
-v. What are the implications of these patterns for the ecology ecosystem and climate change responses of canopy versus understory trees?
+v. What are the implications of vertical canopy gradient patterns for the ecological and climate change responses of canopy versus understory trees?
 
@@
 
-The biophysical environment, here defined to include the physical structure of the forest canopy (from understory to the canopy top) and associated microclimates, is in large part determined by the structure of the forest itself.
-We supplement review of the existing literature with a new analysis of data on vegetation structure and vertical profiles in microclimate from the U.S.
+The biophysical environment, here defined to include the physical structure of the forest canopy (from understory to the top of the canopy) and associated microclimates, is in large part determined by the structure of the forest itself.
+We supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S.
 National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`; Schimel et al.
@@
 
-Canopy foliage, which varies across forest types and seasonally, strongly shapes the understory light, wind, etc.
+Canopy foliage, which varies across forest types and seasonality, strongly shapes understory light, wind, etc.
 Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. `r fig_NEON_vertical`).
@@
 *(paragraph on leaf area distribution, including figure ([separate from or combined with current Fig. 2- see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
-The vertical structure and hence, microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
-In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season but excluding periods (in deciduous forests) when no leaves are present.
+The vertical structure and corresponding microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
+In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.
 
-Light conditions, specifically photosynthetically active radiation (PAR) varies along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, across species and forest types [@koike_leaf_2001].
-Generally, upper canopies in dense forests are exposed to greater PAR values and limit light to canopy interiors, lower canopy layers and understories.
-Shaded regions under canopy cover experience lower values and quality of PAR [in ranges of blue, orange and red].
-The gradient of light is more pronounced under broadleaved forests than conifers[@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
+Light conditions, specifically changes in photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [@koike_leaf_2001].
+Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory.
+Shaded regions experience lower values and quality of PAR [in ranges of blue, orange and red].
+The light gradient is more pronounced under broadleaf forests than needleleaf forests [@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
 (Fig. `r fig_NEON_vertical`).
 
-In heterogeneous canopies [with high gap fractions], distance from the outer canopy is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gap to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]
+In heterogeneous canopies [with high gap fractions], the distance from the outer canopy [to X?] is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]
 
-Tropical forests with dense overstory limit light to understories and subsequent layers, where tropical forest floors receive approximately 1-2%  of the light incident to the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
-However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and movements caused by wind speed called sunflecks that are often intermittent and breif.
+Tropical forests with densely packed overstory trees limit light tosubsequent canopy layers and forest understories.
+Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
+However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and intermittent light penetration from crown, branch, and leaf movement caused by wind called sunflecks.
 Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]
 
-This diffence in light regime along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity along the gradient.
+This diffence in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity.
 Leaves along a vertical intra-canopy gradient, as well as in inter-canopies, thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].
@@
 *(fill in some specifics from the NEON analysis)*
-Consistent with the results from NEON sites, higher wind speed at greater heights within a forest have previously been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months.
+Consistent with the results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months [only in summer months?].
 
-Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in dry season than in upper canopy wet season [@rey-sanchez_spatial_2016] [*double check*]. 
+Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in the dry season than the wet season [@rey-sanchez_spatial_2016] [*double check*]. 
 Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$, this is more typical of small leaves in upper canopy that make them more advantageous in dissapating heat efficiently than larger leaves[@bauerleSeparatingFoliarPhysiology2011].
-In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in upper canopy can become substantially elevated than $T_{air}$ under high radiation [@martin_boundary_1999].
-Additionally, thickness of air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$.
+This is more typical of small leaves in the upper canopy that make them more efficient at dissipating heat than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
+In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in the upper canopy can become substantially more elevated than $T_{air}$ under high radiation [@martin_boundary_1999].
+Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
 [ref]

### 2021-01-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
Comments and edits up to line 430 / Ecosystem Scaling & Implications Section

@@
 Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
-However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and intermittent light penetration from crown, branch, and leaf movement caused by wind called sunflecks.
+However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and intermittent light penetration from tree, branch, and leaf movement caused by wind called sunflecks.
 Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]
@@
 **Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
-Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019].
-Even though canopy cover buffers maximum air temperatures in the understory, inevitable increase in understory temperatures continue to be observed as a result of global increase in temperatures, however the buffering by canopy cover smoothens this warming in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]
+Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019].
+Intact tropical forests have also been show to exhibit lower peak daytime $T_{air}$ than neighboring tree crop agricultural plantations (S.
+R.
+Hardwick et al., The relationship between leaf area index and microclimate in tropical forest and oil palm plantation: Forest disturbance drives changes in microclimate.
+Agric.
+Meteorol.
+201, 187–195 (2015)).
+Even though canopy cover buffers maximum air temperatures in the understory, inevitable increases in understory temperatures continue to be observed as a result of global temperature increases.
+However, the buffering by canopy cover smooths this warming in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]
 
-Diurnal moderating capacity differs between forests with varying canopy structure, crown density, and seasons.
+Diurnal moderating capacity differs between forests with varying canopy structure, crown density, and seasonality.
 Tropical, temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests.
-Increase in crown density and leaf area index decreases light and higher maximum temperature permeability into forest understory, thus maintaining greater lower maximum daytime temperature.
-Comparitively open forests such as pine forests have sparse stands resulting in decreased buffering capacity in understory and closer air temperatures to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
-This relationship is reveresed during the night, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]
+Increases in crown density and leaf area index decrease light and higher maximum temperature permeability into the forest understory, thus maintaining lower maximum daytime temperatures.
+Comparitively open forests such as pine forests with sparse stands (stems?) resulting in decreased buffering capacity in the understory and air temperatures more similar to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
+This relationship is reveresed during the night, however, with open-canopy pine forests maintaining warmer understory temperatures than above canopy [@vonarxSpatiotemporalEffectsForest2012a]
 
-Seasonally, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal Brazil [@fauset_differences_2018, Tymen et al.
-2017].Similarly, maximum temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground, see Fig.
-4 - again illustrating buffering effect of the canopy).However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016], similar results were observed in in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
+During the wet season, maximum air temperatures have been shown to be higher above forest canopies than below in tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al.
+2017].
+Similarly, in these studies, maximum temperatures were higher while humidity was lower in gaps compared to understory regions (both at 1.5 m above the ground, see Fig.
+4 - again illustrating buffering effect of the canopy).
+However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more (Elsa: replace end of sentence with 'boundary layer turbulence'?) air movement between subcanopy and above-canopy air [@rey-sanchez_spatial_2016].
+Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
 
 Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
-Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point?
+Higher (Elsa: or lower?) minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point?
 (e.g., greater radiative heat loss in exposed areas resulting in lower Tmin))
@@
 
-**Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. `r fig_NEON_vertical`).**
+**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**
 - DETAILS - REFS
@@
 
-Along with temperature, forest canopies moderate relative humidity in below-canopy microclimates.
-In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
+Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates.
+In three European forest types (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
 Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
@@
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$.
-Atlantic forests during wet season tended to have higher $T_{air}$, and $VPD$ in upper canopy and canopy gaps than lower in understory where $T_{air}$ and $VPD$ were lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
+Atlantic forests during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
 This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential.
 [@niinemetsPhotosyntheticAcclimationSimultaneous2004a].
-In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
+In North Eastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
 
 **Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
-	Higher in understory, particularly at dusk [@koike_leaf_2001].
+	Higher in the understory, particularly at dusk [@koike_leaf_2001].
 Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
@@
 
-Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient.
+Given that differences are small during the day when photosynthesis is occuring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
 
-If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO2 emitted at the forest floor could increase the vertical CO2 gradient in the future, but this would still only be a short-term effect in the early morning, as the CO2 would quickly diffuse through the canopy.
+If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e.
+from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy.
 Furthermore, increases in mortality of big trees (e.g.
 Senf et al.
-2018) would increase canopy roughness and prevent build-up of CO2 concentration gradients.
+2018) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
+(Elsa: see these papers on the long-term Harvard Forest soil warming experiments.
+CO2 emissions increased, then decreased, then increased again.
+It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
+- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J.
+and Frey, S.D., 2015.
+Long-term forest soil warming alters microbial communities in temperate forest soils.
+Frontiers in microbiology, 6, p.104.
+- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T.
+and Morrisseau, S., 2002.
+Soil warming and carbon-cycle feedbacks to the climate system.
+Science, 298(5601), pp.2173-2176.
+- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A.
+and Grandy, A.S., 2017.
+Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world.
+Science, 358(6359), pp.101-105.)
 
@@
 
-**The strength of this buffering varies across forests, being influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
+**The strength of this buffering varies across forests, influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
 First, buffering increases with canopy cover and water availabity.
@@
 
-Leaf traits are shaped on the basis of light and height and show different degrees of foliar plasticity in accordance with the intensity and availability of light which also optimizes water uptake, heat and C uptake by leaves across forest vertical strata.
+Leaf traits are shaped on the basis of light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light which also optimizes water uptake, heat and C uptake by leaves across forest vertical strata.
 After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a].
-A network of sensory photoreceptors in leaves such as phytochromes among others mediate response to PAR, and ratio of red to far red radiation [R:FR] falling on the surface.
-This enables acclimation or avoidance response through development of necessary traits [Casal, 2013].
-As a result, degrees of high irradiance gives rise to variations of sun-leaf traits  compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
+A network of sensory photoreceptors in leaves such as phytochromes among others mediate leaf-level responses to PAR, and ratio of red to far red radiation [R:FR] intercepting the leaf surface.
+This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013].
+As a result, degrees of high irradiance gives rise to variation in sun-leaf traits compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes (Elsa: I'd recommend using influence or some other word instead of 'shapes' given that leaf shape is a trait sometimes measured and so it may be confusing in this context) leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
 2016].
 
-Increasing number of studies point towards the importance for incorporating sun and shade in modelling leaf economic spectrum (LES) within-canopy gradient[@chenLeafEconomicsSpectrum2020a] because LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in simplified cateogorization of 'sun' and 'shade' leaves due to variation based on light within-canopy, across functional groups, in canopy gaps, height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
-Classification for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
-2018] such as from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
+(Elsa: it may also be worth mentioning leaf turnover / leaf longevity briefly in the above paragraph in the context of leaf development and trait formation.)
+
+An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) (traits?) within-canopy gradient [@chenLeafEconomicsSpectrum2020a] because LES theory was developed using mainly sun exposed (upper canopy) leaves.
+Given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
+However, there is also concern for ambiguity in a simplified cateogorization of 'sun' and 'shade' leaves due to variation based on within-canopy light, across functional groups, in canopy gaps, wtih height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
+Classification schemes for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
+2018].
+For example, classifications from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
 "Additionally, shade leaves in the understory are of three categories; 1.
@@
 3 understory plants that complete their lifecycle in the shaded understory.
-These groups are likely to have different traits, and if we expect'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered".[refs, -Martijn Slot]
+These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered".[refs, -Martijn Slot]
 
@@
 
-Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory.
-As a result, sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and more lobated than shade leaves
+Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees that are shaded by overstory.
+As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and more lobated than shade leaves
 [Rozendaal et al. 2006, Mathur et al. 2018, more REFS].
 
-However, the combination of stressors taller trees' foliage experience are different from the combination of stressors understory shorter trees's foliage experience.
-For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, wind speed, etc.
+However, the combination of stressors taller trees' foliage experience are different from the combination of stressors understory trees' foliage experience.
+For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.
 whereas for understory--shade or limitation of light, niche competition, herbivory etc.
-These pressures also contribute to the development of necessary traits that is primarily mediated by light response.
+These pressures also contribute to the development of necessary traits that are primarily mediated by light response.
 Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature.
@@
 Adaptive traits in understory enable light capture efficiency while maintaing larger width leaves with comparitevely fewer stomtal density as a result of available water, lesser $VPD$, and wind speed.
-"Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn].Those differences in traits are likely to affect their thermal properties as well.Leaves act as windows to plant performance enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plant to respond differently to forcoming stress [Rozendaal et al.
+"Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn].
+Those differences in traits are likely to affect their thermal properties as well.
+Leaves act as windows to plant performance, enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forcoming stress [Rozendaal et al.
 2006; @niinemetsResponsesForestTrees2010a]
 
-However, even with distinct features of sun and shade traits, trait adaptation to light is done in such a way where light use efficiency for photosynthesis is maximized at both high or low light, while also adapting to other stressors as mentioned above.
-In light limited conditions, shade leaves increase light capture efficiency through increasing their specific leaf area [SLA] and by investing in chlorophyll at a mass basis--"leads to a similar chlorphyll content on an area basis compared with sun leaves.
-This makes shade leaves realize a similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
+However, even with distinct features of sun and shade traits, trait adaptation to light occurs in such a way where light use efficiency for photosynthesis is maximized at both high or low light, while also adapting to other stressors as mentioned above.
+In light limited conditions, shade leaves increase light capture efficiency through increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis--"leads to a similar chlorphyll content on an area basis compared with sun leaves.
+This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
 1996, Poorter et al.
@@
 
-With increasing height and age, competion for light decreases as availibity  of light increases in dense forests.
+With increasing height and age, competition for light decreases as availibity  of light increases in dense forests.
 However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements.
 Early successional species are considered more shade intolerant than late successional species that are more shade tolerant.
-Ontogenic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$) in shade tolerators than in intolerators, greater photosynthetic capacity in low light and greater carbon gain capacities[Niinemets, 2006].Rozendaal et al.
+Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$) in shade tolerators than in intolerators, greater photosynthetic capacity in low light and greater carbon gain capacities[Niinemets, 2006].Rozendaal et al.
 2006, Koike et al.
@@
 
-**Many traits vary with height and/or between sun and shade leaves on the same species (Table 1).**
+**Many traits vary with height and/or between sun and shade leaves of the same species (Table 1).**
 Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.
@@
 Lobation allows air passage and light penetration in the tree canopy (Sack et al., 2006; Semchenko & Zobel, 2007)" [@kusiPlasticLeafMorphology2020]
+- Elsa: see also H.
+Poorter, U.
+Niinemets, L.
+Poorter, I.
+J.
+Wright, R.
+Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
+New Phytol.
+182, 565–588 (2009).
+-- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
 
@@
 -  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section) 
+- Elsa: I think Ty's section on isoprene would be better suited up here
 
@@
 $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
-- As light availability typically increases with tree height, leaf temperature which is largely of function of radiation input, does as well [@michaletzPlantThermoregulationEnergetics2015].
+- As light availability typically increases with tree height, leaf temperature which is largely a function of radiation input, does as well [@michaletzPlantThermoregulationEnergetics2015].
 Direct sunlight in upper canopy leaves increases leaf temperature, [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
@@
 Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
-- increase in air temperature, increases VPD, decrease in stomtal conductance during miday leading to decrease in CO2 in leaf = increase in leaf temperature in upper canopy. [@fauset_differences_2018]
+- increase in air temperature, increases VPD, decrease in stomatal conductance during midday leading to decrease in CO2 in leaf = increase in leaf temperature in upper canopy.
+[@fauset_differences_2018]
 
@@
 -Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads.
-However, leaf temperatures themselves were moderaed by air temperature.
+However, leaf temperatures themselves were moderated by air temperature.
 [@fauset_differences_2018].
@@
 
-*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves and leaf water content)*
+*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves - evaporative cooling - and leaf water content)*
+
+- Elsa: I came across these two studies that could be relevent in the evapotranspiration/temperature feedback section hinted at above: 
+- Staudt, K., Serafimovich, A., Siebicke, L., Pyles, R.D.
+and Falge, E., 2011.
+Vertical structure of evapotranspiration at a forest site (a case study).
+Agricultural and Forest Meteorology, 151(6), pp.709-729.
+- Sellin, A. and Lubenets, K., 2010. Variation of transpiration within a canopy of silver birch: effect of canopy position and daily versus nightly water loss. Ecohydrology, 3(4), pp.467-477.
 
@@
 This is primarily driven by the greater light available to sun leaves.
-In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light,  including higher concentration of Rubsico, multiple layers of palisade parenchyma, and saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
-The vertical gradient in photosythetic rates is also influenced by available concentration of N and P.
+In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, multiple layers of palisade parenchyma, and saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
+The vertical gradient in photosythetic rates is also influenced by available concentration of N and P. (and water availability)
 .[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N.
@@
 Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
-2017; Slot & Winter 2017).Beyond the optimum photosynthesis decreases as a result of stomatal closure (e.g.
+2017; Slot & Winter 2017).
+Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g.
 Slot & Winter 2017; Smith et al.
@@
 Sage & Kubien 2007; Varhammer et al.
-2015 ).Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis.
+2015 ).
+Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis.
 Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought.
@@
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5; *revisit this!*) (Campbell and Norman 1998) as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied there to be a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
+14.5; *revisit this!*) (Campbell and Norman 1998) as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
 However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photosnythesis in sun and shade leaves.
 For 3 species in Panama, @hernandezSimilarTemperatureDependence2020 and @slotPhotosyntheticHeatTolerance2019 found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
-Similiar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
+Similar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
 Mau et al. found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
@@
 with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
+- Elsa: also see: 
+- Perez, T.M. and Feeley, K.J., 2020. Photosynthetic heat tolerances and extreme leaf temperatures. Functional Ecology, 34(11), pp.2236-2245.
+- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.
 
@@
 Similarly, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher $T_{50}$ was observed in more exposed (North-facing) canopy positions (Curtis et al. 2019). 
-We're not aware of any other studies on this."we may want to argue that we have no a priori reasons to expect thermal tolerance to scale with height per se (or do we?).
+We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
 I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot
@@
 Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
+- Elsa: and see Perez, T.M., Socha, A., Tserej, O. and Feeley, K.J., 2020. PSII heat tolerances characterize thermal generalists and the upper limit of carbon assimilation. Plant, Cell & Environment.
 

### 2021-01-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves that may potentially reduce [forest] biomass acculumation and disrupt photosynthesis and metabolic processes across plant species [O'Sullivan;REF ;@perkinsIncreasingFrequencyIntensity2012].
-These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
+There is also increasing severity and frequency of heat waves that will potentially reduce [forest] biomass acculumation and disrupt photosynthesis and metabolic processes across plant species [O'Sullivan;REF ;@perkinsIncreasingFrequencyIntensity2012].
+These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
 
@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and in turn whole plant performance [Bertrand et al.
+This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and with implications for whole plant performance [Bertrand et al.
 2011].
 
-The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions[@senfCanopyMortalityHas2018a].
-Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests [ref].
-Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-Evidence shows that larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
+Increasing temperature has profound impacts on canopy cover, such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Additionally, increased temperature is driving more frequent droughts, causing hydraulic limitation in forests [ref].
+Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continues evapotranspiration and consequential buffering of the understory [REF]
+However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continuOUS evapotranspiration, thereby buffering understory microclimates [REF].
 
-While forest microclimates are insulated from macroclimatic extremes, they are still experiencing warming trends [@bertrandCommentForestMicroclimate2020].
-As a result, understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, etc.
-[@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
+While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
+Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
 Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
-Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
-Tree-ring evidence in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
+Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
 In addition, understory growth phenology is sensitive to warming year round, and especially in spring [Zohnner_Renner 2019].
@@
 While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in overstory and understory is still unclear.
-Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy influenced by biophysical environmental gradient.This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
+Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
+This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
 
@@
 
-Tropical forests with densely packed overstory trees limit light tosubsequent canopy layers and forest understories.
-Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
-However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and intermittent light penetration from tree, branch, and leaf movement caused by wind called sunflecks.
-Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]
+Tropical forests with densely packed overstory trees limit light to subsequent canopy layers and forest understories.
+Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
+However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks.
+Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]
 
-This diffence in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity.
-Leaves along a vertical intra-canopy gradient, as well as in inter-canopies, thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].
+This difference in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity.
+Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].
 
@@
 *(fill in some specifics from the NEON analysis)*
-Consistent with the results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months [only in summer months?].
+Consistent with the results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019]
 
@@
 Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$.
-This is more typical of small leaves in the upper canopy that make them more efficient at dissipating heat than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
-In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in the upper canopy can become substantially more elevated than $T_{air}$ under high radiation [@martin_boundary_1999].
+Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiteny than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
+In low wind conditions, because $g_{a}$ to heat transfer is comparitevely small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; Leigh et al.
+2017].
 Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
@@
 Increases in crown density and leaf area index decrease light and higher maximum temperature permeability into the forest understory, thus maintaining lower maximum daytime temperatures.
-Comparitively open forests such as pine forests with sparse stands (stems?) resulting in decreased buffering capacity in the understory and air temperatures more similar to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
-This relationship is reveresed during the night, however, with open-canopy pine forests maintaining warmer understory temperatures than above canopy [@vonarxSpatiotemporalEffectsForest2012a]
+Comparatively open forests such as pine forests with sparse stands (stems?) resulting in decreased buffering capacity in the understory and air temperatures more similar to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
+This relationship is reversed during the night, however, with open-canopy pine forests maintaining warmer understory temperatures than above canopy [@vonarxSpatiotemporalEffectsForest2012a]
 
@@
 4 - again illustrating buffering effect of the canopy).
-However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more (Elsa: replace end of sentence with 'boundary layer turbulence'?) air movement between subcanopy and above-canopy air [@rey-sanchez_spatial_2016].
+However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
 Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
@@
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$.
-Atlantic forests during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
-This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential.
+Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
+This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constrains and lower osmotic potential.
 [@niinemetsPhotosyntheticAcclimationSimultaneous2004a].
-In North Eastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
+In northeastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
 
@@
 
-Leaf traits are shaped on the basis of light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light which also optimizes water uptake, heat and C uptake by leaves across forest vertical strata.
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a].
-A network of sensory photoreceptors in leaves such as phytochromes among others mediate leaf-level responses to PAR, and ratio of red to far red radiation [R:FR] intercepting the leaf surface.
+Leaf traits are shaped by light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat and C uptake by leaves across forest vertical strata.
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a].
+A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
 This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013].
-As a result, degrees of high irradiance gives rise to variation in sun-leaf traits compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes (Elsa: I'd recommend using influence or some other word instead of 'shapes' given that leaf shape is a trait sometimes measured and so it may be confusing in this context) leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
+As a result, degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also influences leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
 2016].
@@
 
-An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) (traits?) within-canopy gradient [@chenLeafEconomicsSpectrum2020a] because LES theory was developed using mainly sun exposed (upper canopy) leaves.
-Given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in a simplified cateogorization of 'sun' and 'shade' leaves due to variation based on within-canopy light, across functional groups, in canopy gaps, wtih height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
+An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [@chenLeafEconomicsSpectrum2020a], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
+Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
+However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and wtih height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
+
 Classification schemes for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
@@
 For example, classifications from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
-"Additionally, shade leaves in the understory are of three categories; 1.
+
+Additionally, shade leaves in the understory can be broken into three categories; 1.
 shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded.
@@
 3 understory plants that complete their lifecycle in the shaded understory.
-These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered".[refs, -Martijn Slot]
+These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered [refs, -Martijn Slot]
 
@@
 Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees that are shaded by overstory.
-As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and more lobated than shade leaves
+As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and greater lobation than shade leaves
 [Rozendaal et al. 2006, Mathur et al. 2018, more REFS].
 
-However, the combination of stressors taller trees' foliage experience are different from the combination of stressors understory trees' foliage experience.
+However, the combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees.
 For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.
-whereas for understory--shade or limitation of light, niche competition, herbivory etc.
-These pressures also contribute to the development of necessary traits that are primarily mediated by light response.
+whereas for understory leaves are subjected to --shade or limitation of light, niche competition and herbivory.
 Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature.
@@
 Those differences in traits are likely to affect their thermal properties as well.
-Leaves act as windows to plant performance, enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forcoming stress [Rozendaal et al.
+Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [Rozendaal et al.
 2006; @niinemetsResponsesForestTrees2010a]
 
-However, even with distinct features of sun and shade traits, trait adaptation to light occurs in such a way where light use efficiency for photosynthesis is maximized at both high or low light, while also adapting to other stressors as mentioned above.
-In light limited conditions, shade leaves increase light capture efficiency through increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis--"leads to a similar chlorphyll content on an area basis compared with sun leaves.
+However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above.
+In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorphyll content on an area basis compared with sun leaves.
 This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
@@
 
-With increasing height and age, competition for light decreases as availibity  of light increases in dense forests.
+With increasing height and age, competition for light decreases as availability  of light increases in dense forests.
 However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements.
 Early successional species are considered more shade intolerant than late successional species that are more shade tolerant.
-Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$) in shade tolerators than in intolerators, greater photosynthetic capacity in low light and greater carbon gain capacities[Niinemets, 2006].Rozendaal et al.
+Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), , greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [Niinemets, 2006].Rozendaal et al.
 2006, Koike et al.
@@
 Are traits shaped more by height or light?
+
+Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
+However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, where as Q10 and E10 remained contant with height [Cavaleri et al.
+2008].
+In temperate forests, Acer saccharum exhibited greater LMA with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing seaon LMA continued to increase in the upper canopy due to light driven changes [Coble et al.
+2016]
+
+Height-based leaf trait variation in Quercus alba and Quercus velutina showed decrease in lobation of upper canopy sun leaves with increasing height, suggesting an adaptation to hydraulic constraints with increasing height [@kusiPlasticLeafMorphology2020].
+However, Sack et al.
+2006 did not observe trait variation based on height in Quercus spp.
+
 Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
@@
 I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
--"Our study suggests that height plays an important role in the leaf traits, in contrast with previous findings (Baranski, 1975; Sack et al., 2006).In Quercus.
-alba and Quercus.
-velutina, loba-tion decreased with increasing tree height.
-Lobation allows air passage and light penetration in the tree canopy (Sack et al., 2006; Semchenko & Zobel, 2007)" [@kusiPlasticLeafMorphology2020]
 - Elsa: see also H.

### 2021-01-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
Edits to *Ecosystem Scale Implications* section

--- 2021-01-11 vertical-thermal-review.Rmd
@@
 
-Elsa--The content below could probably be integrated in the above [Ecosystem Scale Implications] section, with subsections focusing on: 
--	Suggested implications at the ecosystem scale based on literature
+*Ecosystem Scale Implications*
+(Elsa: I included info on ecosystem function here -- e.g., transpiration, gpp)
+
+Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases.
+As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see Nakamura et al 2017), leaf traits, and photosynthesis .
+Linked to this local biophysical variation is the partitioning of evapotranspiration (ET) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al.
+2007, Staudt et al.
+2011).
+Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity (GPP, Banerjee & Linn 2018).
+It has been demonstrated that net carbon assimilation and transpiration can vary substantially for forest canopies with the same LAI but different leaf area density distributions (e.g., Katul et al 2004 , Banerjee & Linn 2018).
+Further, although transpiration tends to be the dominant component of ET, partitioning is more complex in open canopies where soil evaporation can become more dominant (Baldocchi et al.
+2004, Ma et al.
+2020), likely affecting vertical thermal gradients.
+In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout.
+
+Still, partitioning ET into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
+Eddy covariance techniques, sap flow measurements, and stable isotope analysis offer useful tools, although errors can be large (Williams et al 2004, Herbst et al.
+1996).
+To date, empirical analyses of ET vertical partitioning and interactions with temperature and vegetation structure have be concentrated in midlatitude ecosystems (e.g.
+Haverd et al 2009, 2011, Staudt et al.
+2011, Ringgaard et al.
+2014).
+With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.
+
+An increasing availability of airborne and spaceborne LiDAR remote sensing data offers a promising opportunity for mapping vertical thermal gradients and vegetation structure.
+[Drone based FLIR cams, tower thermal cams (Pau et al 2018), ECOSTRESS].
+To evaluate large-scale ecosystem patterns, high resolution vertically resolved vegetation structure from LiDAR data can be analyzed in combination with collocated field measurements collected using towers (FluxNet REF), canopy cranes (Nakamura et al.
+2017), or by tree climbing (REF – e.g.
+Asner papers, Shenkin et al – Unding Jami’s work).
+Existing research efforts have already revealed incredibly heterogeneity in vegetation structure, diversity, traits, and properties both within and across ecosystems (REFS – Stovall, Schneider, …others).
+
+*Representing Vertical Gradients in Models* 
 -	How do models currently integrate or not integrate these concepts and dynamics
+-	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?
+
+*Knowledge Gaps & Future Research*
 -	What major gaps in knowledge and/or important questions emerge from this review (i.e.
 a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
--	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?
 
@@
 [@nakamuraForestsTheirCanopies2017]--> modelling ideas?
-
-*(do we need a section on ecosystem function-- e.g., transpiration, gpp?)*
 

### 2021-01-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
Minor edits throughout

@@
 Science, 358(6359), pp.101-105.)
+- Nottingham, A.T., Meir, P., Velasquez, E. and Turner, B.L., 2020. Soil carbon loss by experimental warming in a tropical forest. Nature, 584(7820), pp.234-237.
 
@@
 Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases.
-As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see Nakamura et al 2017), leaf traits, and photosynthesis .
+As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see @nakamuraForestsTheirCanopies2017), leaf traits, and photosynthesis .
 Linked to this local biophysical variation is the partitioning of evapotranspiration (ET) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al.
@@
 [Drone based FLIR cams, tower thermal cams (Pau et al 2018), ECOSTRESS].
-To evaluate large-scale ecosystem patterns, high resolution vertically resolved vegetation structure from LiDAR data can be analyzed in combination with collocated field measurements collected using towers (FluxNet REF), canopy cranes (Nakamura et al.
-2017), or by tree climbing (REF – e.g.
+To evaluate large-scale ecosystem patterns, high resolution vertically resolved vegetation structure from LiDAR data can be analyzed in combination with collocated field measurements collected using towers (FluxNet REF), canopy cranes (@nakamuraForestsTheirCanopies2017), or by tree climbing (REF – e.g.
 Asner papers, Shenkin et al – Unding Jami’s work).
@@
 -	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?
-
-*Knowledge Gaps & Future Research*
--	What major gaps in knowledge and/or important questions emerge from this review (i.e.
-a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
-
-E.g.
+- E.g.
 see de Pury & Farquhar, 1997 : though likely an improvement over 'big-leaf' models, models that separate the canopy into sunlit and shaded portions (sun/shade models) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g.
@@
 Hence, multi-layered ecosystem models may be necessary for accurately projecting future forest function.
+
+*Knowledge Gaps & Future Research*
+-	What major gaps in knowledge and/or important questions emerge from this review (i.e.
+a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
+- Not sure where this section should go
 

### 2021-01-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-01-12 vertical-thermal-review.Rmd
@@
 
-Leaf traits are shaped by light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat and C uptake by leaves across forest vertical strata.
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a].
+Leaf traits are shaped by light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat and C uptake by leaves across forest vertical strata [Rosendaal et al.
+2006].
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [Keenan and Niinemets et al.2016; @niinemetsLeafAgeDependent2016a].
 A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
@@
 Early successional species are considered more shade intolerant than late successional species that are more shade tolerant.
-Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), , greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [Niinemets, 2006].Rozendaal et al.
-2006, Koike et al.
+Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [Niinemets, 2006].
+Shade intolerant species experience greater heterogeniety in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H.
+Poorter, U.
+Niinemets, L.
+Poorter, I.
+J.
+Wright, R.
+Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
+New Phytol.
+182, 565–588 (2009).].Whereas, Rozendaal et al.
+2006 found no significant trait differences with ontogeny and adult stature trees.
+Koike et al.
 2001 has more info about this.
@@
 2008].
-In temperate forests, Acer saccharum exhibited greater LMA with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing seaon LMA continued to increase in the upper canopy due to light driven changes [Coble et al.
-2016]
+In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing seaon LMA and $N area$ continued to increase in the upper canopy due to light driven changes [Coble & Cavaleri 2015; Coble et al.
+2016].In addition, carbon isotope composition also increased with height, independent of ligtht [Coble & Cavaleri 2015]
 
-Height-based leaf trait variation in Quercus alba and Quercus velutina showed decrease in lobation of upper canopy sun leaves with increasing height, suggesting an adaptation to hydraulic constraints with increasing height [@kusiPlasticLeafMorphology2020].
+Height-based leaf trait variation in Quercus alba and Quercus velutina showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
 However, Sack et al.
@@
 - Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
+
+Leaf temperature is generally associated to increase with leaf area ($LA$) [ Wright et al.
+2019].
+However, thermal imagery shows that leaf area can be a predictor for leaf temperatures ranges [$Trange$] among a given number of samples, but effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
+With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al.
+2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn traspirational cooling in large leaves [@fauset_differences_2018].
+Because rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
+Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
+2016 found leaf width to be a dominant factor in deteriming the changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
 - Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
 Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
-Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
-Leaf temperature increases with leaf area {Wright et al. 2019}
--Boundary layer resistance and thermal sensitivity also depends of leaf width, leaf size and structure.
-Increase in leaf width increases stomatal resistance-- dependent on stomtal activity and boundary layer resistance [@fauset_differences_2018].
 Leaf characteristics such as lobes, leaf elongation and serration, all facilitate in dissapating heat more efficiently.
@@
 Schuepp 1993; Sack et al. 2006].
--Boundary layer thickness, depth and boundary layer resistance increases with leaf size and effective leaf width [along the vertical gradient], such that the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, therefore allowing small leaves to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 
@@
 -	Michaeletz has used this in theoretical calculations (Michaeletz et al. 2016 and 2015?
-I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
-In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot
+
 -Effective leaf width defined as the largest circle that can be inscribed in a leaf is an effective predictor of leaf thermal time constant and leaf boundary layer thickness.
@@
 [@leighInfluenceLeafSize2017]
+
+[I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
+In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot]
 

### 2021-01-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and with implications for whole plant performance [Bertrand et al.
-2011].
+This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 
@@
 
-Leaf traits are shaped by light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat and C uptake by leaves across forest vertical strata [Rosendaal et al.
+Leaf traits are shaped by light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, hea, C uptake and metabolism by leaves across forest vertical strata [Rosendaal et al.
 2006].
@@
 Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and wtih height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
+However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
 
@@
 
-Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees that are shaded by overstory.
+Leaf traits (functional and physiological) co-vary with their environment to maintain optimum metabolic and temperature under various circumstances [@michaletzEnergeticCarbonEconomic2016].
+Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy compared to shorter understory trees that are shaded by overstory.
 As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and greater lobation than shade leaves
@@
 However, the combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees.
-For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.
+For example, overstory sun leaves in particular experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc..
 whereas for understory leaves are subjected to --shade or limitation of light, niche competition and herbivory.
-Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature.
+Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature for photosynthesis and metabolic processes.
 Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
@@
 Those differences in traits are likely to affect their thermal properties as well.
-Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [Rozendaal et al.
+Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [Rozendaal et al.
 2006; @niinemetsResponsesForestTrees2010a]
@@
 
-Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which shapes leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
 The differences among these traits contribute to leaf heat dissapation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1.
-summaries relevant traits for thermal sensitivity.
+summarizes relevant traits for thermal sensitivity.
 
@@
 2008].
-In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing seaon LMA and $N area$ continued to increase in the upper canopy due to light driven changes [Coble & Cavaleri 2015; Coble et al.
+In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [Coble & Cavaleri 2015; Coble et al.
 2016].In addition, carbon isotope composition also increased with height, independent of ligtht [Coble & Cavaleri 2015]
@@
 2019].
-However, thermal imagery shows that leaf area can be a predictor for leaf temperatures ranges [$Trange$] among a given number of samples, but effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
+However, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
 With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al.
 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn traspirational cooling in large leaves [@fauset_differences_2018].
-Because rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
+Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
-2016 found leaf width to be a dominant factor in deteriming the changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
+2016 found leaf width to be a dominant factor in deterimining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
+Leaf width and characteristic dimension is generally greater in shaded understory leaves along the vertical gradient than exposed upper canopy [ref].
+Additionally, leaf characteristics such as lobes, leaf elongation and serration break the entirity of leaf dimension by enhancing a smaller characteristic dimension, all facilitate in dissapating heat more efficiently.
+Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al.
+2006].
+
 - Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
 Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
-Leaf characteristics such as lobes, leaf elongation and serration, all facilitate in dissapating heat more efficiently.
-Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;
-Schuepp 1993; Sack et al. 2006].
 
@@
 
--Effective leaf width defined as the largest circle that can be inscribed in a leaf is an effective predictor of leaf thermal time constant and leaf boundary layer thickness.
-Large leaves have greater effective leaf width which increases their thermal time constant by prolonging leaf cooling time compared to small leaves that have a smaller effective leaf width, and thus small thermal time constant with shorter cooling time.
-Increased leaf width typical in larger leaves was also observed to increase leaf to air temperature difference.
-[@leighInfluenceLeafSize2017]
-
+Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time.
+It comprises of several underlying traits -- leaf dry matter content, leaf width, water mass, specific heat capacity, leaf geometery, two sided surface area and stomatal conductance [@michaletzEnergeticCarbonEconomic2016].
+Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width.
+Dissections in leaves such as lobes decreases leaf with and thus thermal time constant [@leighInfluenceLeafSize2017].
+Additionally, leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?
 [I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
@@
 
-*Carotinoids/Xanthophyll cycle pigments* 
--	photoprotective- disseminate heat, acclimate to high T. THese are proportional to irradiance 
+*Carotenoids/Xanthophyll cycle pigments*
+
+High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from photoinhibition through D1 protein replacement.
+Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments express on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enable acclimation to high temperature [Mathur et al.
+2018].
+These are found to be proportional to within-canopy light gradient at an inter- and intra-species level.
+Therefore, lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [Niinemets et al.
+1998].
+In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll and lutein and neoxanthin carotenoids [Mastubara et al.
+2009]
+
 -	antioxidant scavenging function- protect against cellular damage

### 2021-01-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continuOUS evapotranspiration, thereby buffering understory microclimates [REF].
+However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration, thereby buffering understory microclimates [REF].
 
@@
 
-Leaf traits (functional and physiological) co-vary with their environment to maintain optimum metabolic and temperature under various circumstances [@michaletzEnergeticCarbonEconomic2016].
+Leaf traits (functional and physiological) co-vary with their environment to maintain optimum temperature for metabolic processes under various environmental circumstances [@michaletzEnergeticCarbonEconomic2016].
 Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy compared to shorter understory trees that are shaded by overstory.
@@
 Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width.
-Dissections in leaves such as lobes decreases leaf with and thus thermal time constant [@leighInfluenceLeafSize2017].
+Dissections in leaves such as lobes decreases leaf width and thus thermal time constant [@leighInfluenceLeafSize2017].
 Additionally, leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?
@@
 High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from photoinhibition through D1 protein replacement.
-Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments express on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enable acclimation to high temperature [Mathur et al.
+Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enable acclimation to high temperature [Mathur et al.
 2018].
 These are found to be proportional to within-canopy light gradient at an inter- and intra-species level.
-Therefore, lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [Niinemets et al.
+Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [Niinemets et al.
 1998].
-In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll and lutein and neoxanthin carotenoids [Mastubara et al.
+In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al.
 2009]
@@
 **Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
-- As light availability typically increases with tree height, leaf temperature which is largely a function of radiation input, does as well [@michaletzPlantThermoregulationEnergetics2015].
-Direct sunlight in upper canopy leaves increases leaf temperature, [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
-While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.[@fauset_differences_2018]
-Large leaves can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation with stomata closed.  
+$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+3) and is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].
+In tropical and temperate forests, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
+
+In addition, leaf size impacts leaf temperature.
+While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018].
+Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.
+
 Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
@@
 However, counteracting this, lower wind speed in understory would reduce latent heat loss.
-Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
-- increase in air temperature, increases VPD, decrease in stomatal conductance during midday leading to decrease in CO2 in leaf = increase in leaf temperature in upper canopy.
-[@fauset_differences_2018]
+Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
+However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday leading to a decrease in leaf CO2 and consequentially increasing leaf temperature in upper canopy [@fauset_differences_2018].
+Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
 

### 2021-01-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf traits, thermoregulation and metabolism along the gradient, and with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 
@@
 Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
-This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
+This will provide a pathway for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
 
@@
 Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time.
-It comprises of several underlying traits -- leaf dry matter content, leaf width, water mass, specific heat capacity, leaf geometery, two sided surface area and stomatal conductance [@michaletzEnergeticCarbonEconomic2016].
+It comprises of several underlying traits -- leaf area, LMA, leaf width, leaf dry matter content and stomatal conductance [@michaletzEnergeticCarbonEconomic2016].
 Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width.
@@
 $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
-3) and is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].
-In tropical and temperate forests, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
+3).
+Leaf traits are regulated to varying biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015]
+$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].
 
@@
 3.	Leaf traits
+
+In tropical and temperate forests, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
+
 -  Michaletz et al have done some interesting work on leaf thermoregulation in relation to various leaf traits.
@@
 **2?**).**
-In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999).
+In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al.
+1999).Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].
 Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).

### 2021-01-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves that will potentially reduce [forest] biomass acculumation and disrupt photosynthesis and metabolic processes across plant species [O'Sullivan;REF ;@perkinsIncreasingFrequencyIntensity2012].
+There is also increasing severity and frequency of heat waves that will potentially reduce [forest] biomass acculumation, and disrupt photosynthesis and metabolic processes across plant species [Sullivan et al.
+2020, @perkinsIncreasingFrequencyIntensity2012].
 These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
@@
 The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
-Increasing temperature has profound impacts on canopy cover, such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Additionally, increased temperature is driving more frequent droughts, causing hydraulic limitation in forests [ref].
-Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Conversly, available water facilitates leaf transpiration through open stomata, however increased relative humdity in tropical regions indirectly caused by increased temperature, reduces transpiration leading to high leaf temperatures in the canopy [Tibbitts 1979; Perez and Feeley 2018].
+
+Frequent droughts related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration, thereby buffering understory microclimates [REF].
+However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates [REF].
 
@@
 
-While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in overstory and understory is still unclear.
-Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
+While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory is still unclear.
+Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
 This will provide a pathway for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
@@
 
-Leaf traits are shaped by light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, hea, C uptake and metabolism by leaves across forest vertical strata [Rosendaal et al.
+Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [Rosendaal et al.
 2006].
@@
 
-Leaf traits (functional and physiological) co-vary with their environment to maintain optimum temperature for metabolic processes under various environmental circumstances [@michaletzEnergeticCarbonEconomic2016].
 Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy compared to shorter understory trees that are shaded by overstory.
@@
 2019].
-However, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
+Thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
 With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al.
@@
 2016 found leaf width to be a dominant factor in deterimining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
-Leaf width and characteristic dimension is generally greater in shaded understory leaves along the vertical gradient than exposed upper canopy [ref].
+Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref].
 Additionally, leaf characteristics such as lobes, leaf elongation and serration break the entirity of leaf dimension by enhancing a smaller characteristic dimension, all facilitate in dissapating heat more efficiently.
 Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al.
-2006].
-
-- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
-Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
+2006].However, Perez and Feeley et al.
+2020 showed only a slight negetive effect on leaf temperature with leaf width.
 
@@
 3).
-Leaf traits are regulated to varying biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015]
+Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf, and thus leaf temperatures are decoupled from air temperature through adaptive physiological traits [@michaletzPlantThermoregulationEnergetics2015; Perez et al.
+2020]
 $T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].
@@
 
-In tropical and temperate forests, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
+In a tropical forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
 
@@
 **2?**).**
-In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al.
-1999).Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].
+[Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
+In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999). 
 Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
+
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
@@
 2019)
+- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
+Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
 
@@
 with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
-- Elsa: also see: 
-- Perez, T.M. and Feeley, K.J., 2020. Photosynthetic heat tolerances and extreme leaf temperatures. Functional Ecology, 34(11), pp.2236-2245.
-- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.
+Contrary to O'Sullivan et al.
+2017's findings, Perez and Feeley et al.
+2020 found that across inter-species sun leaves, higher photosynthetic heat tolerance in leaves lead to hotter leaf temperatures and narrower thermal safety margins and thus may be at a greater risk of thermal damage.
+[- Elsa: also see: 
+- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.]
+
+Supporting the significant negetive relationship between $T50$ and thermal safety margins, photosynthesis [PS2] temperature regulation has been associated leaf thermal tolerance in order to maximize net carbon assimilation.
+However, randomly collected within-canopy leaves of individual mature plants in Florida showed that Tmax is not directly coordinated with high PS2 heat tolerance and does not necessarily increase leaf thermal tolerance [shows no direct correlation]; nor increases carbon assimilation at higher temperature.
+Rather, high PS2 heat tolerance reduces optimum temperature for C assimilation [$Topt$] because high PS2 heat tolerance requires a greater cost of PS2 metabolic activity maintainance thus not contributing to increased C assimilation.
+Species exhibiting low carbon assimilation rate tended to have greater stress tolerance.
+In addition, species with low stomtal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimaltion before thermal limits are reached [Perez et al.
+2020]
 
@@
 I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot
+
+increased irradiance = increased VPD = increased air temp = midday stomatal closure =  decreased latent heat loss = increase leaf temperature = decreased C assimilation = greater PS2 heat tolerance = optimum rate of photosynthesis decreases = exhibiting lower t opt (optimum temp for c assimilation)
+
+incr.humidity = decreased VPD = dec. latent heat loss = limited c assimilation [Perez et al. 2020]
 

### 2021-01-16 vertical-thermal-review.Rmd
First pass at Vertical Gradients in Models section
fyi, this is still incomplete. Here I added an overview of differences in model structure. More to come.

--- 2021-01-14 vertical-thermal-review.Rmd
@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves that will potentially reduce [forest] biomass acculumation and disrupt photosynthesis and metabolic processes across plant species [O'Sullivan;REF ;@perkinsIncreasingFrequencyIntensity2012].
-These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
+There is also increasing severity and frequency of heat waves that may potentially reduce [forest] biomass acculumation and disrupt photosynthesis and metabolic processes across plant species [O'Sullivan;REF ;@perkinsIncreasingFrequencyIntensity2012].
+These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
 
@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient, and in turn whole plant performance [Bertrand et al.
+2011].
 
-The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
-Increasing temperature has profound impacts on canopy cover, such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Additionally, increased temperature is driving more frequent droughts, causing hydraulic limitation in forests [ref].
-Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions[@senfCanopyMortalityHas2018a].
+Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests [ref].
+Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+Evidence shows that larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration, thereby buffering understory microclimates [REF].
+However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continues evapotranspiration and consequential buffering of the understory [REF]
 
-While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
-Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
+While forest microclimates are insulated from macroclimatic extremes, they are still experiencing warming trends [@bertrandCommentForestMicroclimate2020].
+As a result, understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, etc.
+[@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
 Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
-Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
-Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
+Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+Tree-ring evidence in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
 In addition, understory growth phenology is sensitive to warming year round, and especially in spring [Zohnner_Renner 2019].
@@
 While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in overstory and understory is still unclear.
-Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
-This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
+Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy influenced by biophysical environmental gradient.This will provide crucial insights for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].
 
@@
 
-Tropical forests with densely packed overstory trees limit light to subsequent canopy layers and forest understories.
-Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
-However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks.
-Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]
+Tropical forests with densely packed overstory trees limit light tosubsequent canopy layers and forest understories.
+Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
+However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and intermittent light penetration from tree, branch, and leaf movement caused by wind called sunflecks.
+Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]
 
-This difference in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity.
-Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].
+This diffence in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity.
+Leaves along a vertical intra-canopy gradient, as well as in inter-canopies, thus exhibit a variation in leaf trait adapaptions to a range of high light to shaded-ness [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].
 
@@
 *(fill in some specifics from the NEON analysis)*
-Consistent with the results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019]
+Consistent with the results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019] in summer months [only in summer months?].
 
@@
 Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiteny than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
-In low wind conditions, because $g_{a}$ to heat transfer is comparitevely small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; Leigh et al.
-2017].
+Large $g_{a}$ to heat transfer results in smaller difference in $T_{air}$ and $T_{leaf}$.
+This is more typical of small leaves in the upper canopy that make them more efficient at dissipating heat than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
+In low wind conditions, because $g_{a}$ to heat transfer is small enough, $T_{leaf}$ in the upper canopy can become substantially more elevated than $T_{air}$ under high radiation [@martin_boundary_1999].
 Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
@@
 Increases in crown density and leaf area index decrease light and higher maximum temperature permeability into the forest understory, thus maintaining lower maximum daytime temperatures.
-Comparatively open forests such as pine forests with sparse stands (stems?) resulting in decreased buffering capacity in the understory and air temperatures more similar to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
-This relationship is reversed during the night, however, with open-canopy pine forests maintaining warmer understory temperatures than above canopy [@vonarxSpatiotemporalEffectsForest2012a]
+Comparitively open forests such as pine forests with sparse stands (stems?) resulting in decreased buffering capacity in the understory and air temperatures more similar to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
+This relationship is reveresed during the night, however, with open-canopy pine forests maintaining warmer understory temperatures than above canopy [@vonarxSpatiotemporalEffectsForest2012a]
 
@@
 4 - again illustrating buffering effect of the canopy).
-However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
+However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more (Elsa: replace end of sentence with 'boundary layer turbulence'?) air movement between subcanopy and above-canopy air [@rey-sanchez_spatial_2016].
 Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
@@
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$.
-Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
-This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constrains and lower osmotic potential.
+Atlantic forests during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
+This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential.
 [@niinemetsPhotosyntheticAcclimationSimultaneous2004a].
-In northeastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
+In North Eastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
 
@@
 
-Leaf traits are shaped by light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, hea, C uptake and metabolism by leaves across forest vertical strata [Rosendaal et al.
-2006].
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [Keenan and Niinemets et al.2016; @niinemetsLeafAgeDependent2016a].
-A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
+Leaf traits are shaped on the basis of light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light which also optimizes water uptake, heat and C uptake by leaves across forest vertical strata.
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a].
+A network of sensory photoreceptors in leaves such as phytochromes among others mediate leaf-level responses to PAR, and ratio of red to far red radiation [R:FR] intercepting the leaf surface.
 This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013].
-As a result, degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also influences leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
+As a result, degrees of high irradiance gives rise to variation in sun-leaf traits compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes (Elsa: I'd recommend using influence or some other word instead of 'shapes' given that leaf shape is a trait sometimes measured and so it may be confusing in this context) leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
 2016].
@@
 
-An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [@chenLeafEconomicsSpectrum2020a], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
-Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
-
+An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) (traits?) within-canopy gradient [@chenLeafEconomicsSpectrum2020a] because LES theory was developed using mainly sun exposed (upper canopy) leaves.
+Given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
+However, there is also concern for ambiguity in a simplified cateogorization of 'sun' and 'shade' leaves due to variation based on within-canopy light, across functional groups, in canopy gaps, wtih height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
 Classification schemes for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
@@
 For example, classifications from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
-
-Additionally, shade leaves in the understory can be broken into three categories; 1.
+"Additionally, shade leaves in the understory are of three categories; 1.
 shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded.
@@
 3 understory plants that complete their lifecycle in the shaded understory.
-These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered [refs, -Martijn Slot]
+These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered".[refs, -Martijn Slot]
 
@@
 
-Leaf traits (functional and physiological) co-vary with their environment to maintain optimum temperature for metabolic processes under various environmental circumstances [@michaletzEnergeticCarbonEconomic2016].
-Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy compared to shorter understory trees that are shaded by overstory.
-As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and greater lobation than shade leaves
+Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees that are shaded by overstory.
+As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and more lobated than shade leaves
 [Rozendaal et al. 2006, Mathur et al. 2018, more REFS].
 
-However, the combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees.
-For example, overstory sun leaves in particular experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc..
-whereas for understory leaves are subjected to --shade or limitation of light, niche competition and herbivory.
-Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature for photosynthesis and metabolic processes.
+However, the combination of stressors taller trees' foliage experience are different from the combination of stressors understory trees' foliage experience.
+For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.
+whereas for understory--shade or limitation of light, niche competition, herbivory etc.
+These pressures also contribute to the development of necessary traits that are primarily mediated by light response.
+Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature.
 Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
@@
 Those differences in traits are likely to affect their thermal properties as well.
-Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [Rozendaal et al.
+Leaves act as windows to plant performance, enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forcoming stress [Rozendaal et al.
 2006; @niinemetsResponsesForestTrees2010a]
 
-However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above.
-In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorphyll content on an area basis compared with sun leaves.
+However, even with distinct features of sun and shade traits, trait adaptation to light occurs in such a way where light use efficiency for photosynthesis is maximized at both high or low light, while also adapting to other stressors as mentioned above.
+In light limited conditions, shade leaves increase light capture efficiency through increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis--"leads to a similar chlorphyll content on an area basis compared with sun leaves.
 This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
@@
 
-With increasing height and age, competition for light decreases as availability  of light increases in dense forests.
+With increasing height and age, competition for light decreases as availibity  of light increases in dense forests.
 However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements.
 Early successional species are considered more shade intolerant than late successional species that are more shade tolerant.
-Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [Niinemets, 2006].
-Shade intolerant species experience greater heterogeniety in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H.
-Poorter, U.
-Niinemets, L.
-Poorter, I.
-J.
-Wright, R.
-Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
-New Phytol.
-182, 565–588 (2009).].Whereas, Rozendaal et al.
-2006 found no significant trait differences with ontogeny and adult stature trees.
-Koike et al.
+Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$) in shade tolerators than in intolerators, greater photosynthetic capacity in low light and greater carbon gain capacities[Niinemets, 2006].Rozendaal et al.
+2006, Koike et al.
 2001 has more info about this.
@@
 
-Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which shapes leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
 The differences among these traits contribute to leaf heat dissapation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1.
-summarizes relevant traits for thermal sensitivity.
+summaries relevant traits for thermal sensitivity.
 
@@
 Are traits shaped more by height or light?
-
-Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
-However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, where as Q10 and E10 remained contant with height [Cavaleri et al.
-2008].
-In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [Coble & Cavaleri 2015; Coble et al.
-2016].In addition, carbon isotope composition also increased with height, independent of ligtht [Coble & Cavaleri 2015]
-
-Height-based leaf trait variation in Quercus alba and Quercus velutina showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
-However, Sack et al.
-2006 did not observe trait variation based on height in Quercus spp.
-
 Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
@@
 I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
+-"Our study suggests that height plays an important role in the leaf traits, in contrast with previous findings (Baranski, 1975; Sack et al., 2006).In Quercus.
+alba and Quercus.
+velutina, loba-tion decreased with increasing tree height.
+Lobation allows air passage and light penetration in the tree canopy (Sack et al., 2006; Semchenko & Zobel, 2007)" [@kusiPlasticLeafMorphology2020]
 - Elsa: see also H.
@@
 - Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
-
-Leaf temperature is generally associated to increase with leaf area ($LA$) [ Wright et al.
-2019].
-However, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
-With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al.
-2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn traspirational cooling in large leaves [@fauset_differences_2018].
-Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
-Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
-2016 found leaf width to be a dominant factor in deterimining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
-Leaf width and characteristic dimension is generally greater in shaded understory leaves along the vertical gradient than exposed upper canopy [ref].
-Additionally, leaf characteristics such as lobes, leaf elongation and serration break the entirity of leaf dimension by enhancing a smaller characteristic dimension, all facilitate in dissapating heat more efficiently.
-Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al.
-2006].
-
 - Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
 Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
+Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
+Leaf temperature increases with leaf area {Wright et al. 2019}
+-Boundary layer resistance and thermal sensitivity also depends of leaf width, leaf size and structure.
+Increase in leaf width increases stomatal resistance-- dependent on stomtal activity and boundary layer resistance [@fauset_differences_2018].
+Leaf characteristics such as lobes, leaf elongation and serration, all facilitate in dissapating heat more efficiently.
+Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;
+Schuepp 1993; Sack et al. 2006].
+-Boundary layer thickness, depth and boundary layer resistance increases with leaf size and effective leaf width [along the vertical gradient], such that the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, therefore allowing small leaves to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 
@@
 -	Michaeletz has used this in theoretical calculations (Michaeletz et al. 2016 and 2015?
+I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
+In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot
+-Effective leaf width defined as the largest circle that can be inscribed in a leaf is an effective predictor of leaf thermal time constant and leaf boundary layer thickness.
+Large leaves have greater effective leaf width which increases their thermal time constant by prolonging leaf cooling time compared to small leaves that have a smaller effective leaf width, and thus small thermal time constant with shorter cooling time.
+Increased leaf width typical in larger leaves was also observed to increase leaf to air temperature difference.
+[@leighInfluenceLeafSize2017]
 
-Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time.
-It comprises of several underlying traits -- leaf dry matter content, leaf width, water mass, specific heat capacity, leaf geometery, two sided surface area and stomatal conductance [@michaletzEnergeticCarbonEconomic2016].
-Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width.
-Dissections in leaves such as lobes decreases leaf width and thus thermal time constant [@leighInfluenceLeafSize2017].
-Additionally, leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?
-[I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
-In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot]
-
-*Carotenoids/Xanthophyll cycle pigments*
-
-High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from photoinhibition through D1 protein replacement.
-Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enable acclimation to high temperature [Mathur et al.
-2018].
-These are found to be proportional to within-canopy light gradient at an inter- and intra-species level.
-Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [Niinemets et al.
-1998].
-In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al.
-2009]
-
+*Carotinoids/Xanthophyll cycle pigments* 
+-	photoprotective- disseminate heat, acclimate to high T. THese are proportional to irradiance 
 -	antioxidant scavenging function- protect against cellular damage
@@
 **Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
-3) and is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].
-In tropical and temperate forests, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
-
-In addition, leaf size impacts leaf temperature.
-While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018].
-Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.
-
+$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
+- As light availability typically increases with tree height, leaf temperature which is largely a function of radiation input, does as well [@michaletzPlantThermoregulationEnergetics2015].
+Direct sunlight in upper canopy leaves increases leaf temperature, [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
+While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.[@fauset_differences_2018]
+Large leaves can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation with stomata closed.  
 Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
@@
 However, counteracting this, lower wind speed in understory would reduce latent heat loss.
-Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
-However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday leading to a decrease in leaf CO2 and consequentially increasing leaf temperature in upper canopy [@fauset_differences_2018].
-Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
+Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
+- increase in air temperature, increases VPD, decrease in stomatal conductance during midday leading to decrease in CO2 in leaf = increase in leaf temperature in upper canopy.
+[@fauset_differences_2018]
 
@@
 
-Elsa--The content below could probably be integrated in the above [Ecosystem Scale Implications] section, with subsections focusing on: 
--	Suggested implications at the ecosystem scale based on literature
+*Ecosystem Scale Implications*
+(Elsa: I included info on ecosystem function here -- e.g., transpiration, gpp)
+
+Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases.
+As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see @nakamuraForestsTheirCanopies2017), leaf traits, and photosynthesis .
+Linked to this local biophysical variation is the partitioning of evapotranspiration (ET) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al.
+2007, Staudt et al.
+2011).
+Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity (GPP, Banerjee & Linn 2018).
+It has been demonstrated that net carbon assimilation and transpiration can vary substantially for forest canopies with the same LAI but different leaf area density distributions (e.g., Katul et al 2004 , Banerjee & Linn 2018).
+Further, although transpiration tends to be the dominant component of ET, partitioning is more complex in open canopies where soil evaporation can become more dominant (Baldocchi et al.
+2004, Ma et al.
+2020), likely affecting vertical thermal gradients.
+In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout.
+
+Still, partitioning ET into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
+Eddy covariance techniques, sap flow measurements, and stable isotope analysis offer useful tools, although errors can be large (Williams et al 2004, Herbst et al.
+1996).
+To date, empirical analyses of ET vertical partitioning and interactions with temperature and vegetation structure have be concentrated in midlatitude ecosystems (e.g.
+Haverd et al 2009, 2011, Staudt et al.
+2011, Ringgaard et al.
+2014).
+With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.
+
+An increasing availability of airborne and spaceborne LiDAR remote sensing data offers a promising opportunity for mapping vertical thermal gradients and vegetation structure.
+[Drone based FLIR cams, tower thermal cams (Pau et al 2018), ECOSTRESS].
+To evaluate large-scale ecosystem patterns, high resolution vertically resolved vegetation structure from LiDAR data can be analyzed in combination with collocated field measurements collected using towers (FluxNet REF), canopy cranes (@nakamuraForestsTheirCanopies2017), or by tree climbing (REF – e.g.
+Asner papers, Shenkin et al – Unding Jami’s work).
+Existing research efforts have already revealed incredibly heterogeneity in vegetation structure, diversity, traits, and properties both within and across ecosystems (REFS – Stovall, Schneider, …others).
+
+*Representing Vertical Gradients in Models* 
+(@Nidhi & @Krista, feel free to edit/remove any of the above text.
+I'm adding more text than we'll want in the end, but I want to air on the side of more rather than less so that you can make decisions about what content you think best fits the goal of this review.
+At some point, perhaps we should discuss this section further and I can revise based on that.)
+
+Models representing ecosystem dynamics and vegetation demographic processes related to carbon, energy, and water fluxes operate at varying scales and have varying degrees of complexity and spatial resolution, including offline models and models that can be fully coupled with Earth System Models (ESMs).
+Dynamic global vegetation models (DGVMs) are the components of land surface models in ESMs used to predict the global distribution of vegetation types from physiological principles (Cao & Woodward, 1998; Foley et al., 1996; Sitch et al., 2003; Woodward & Lomas, 2004).
+Improved representation of vegetation demography in ESMs, specifically including forest canopy architectural variation in both horizontal and vertical directions, has repeatedly been identified as a critical step toward reduced uncertainty and more realistic representation of biologically mediated feedbacks in modeling future climates (Banerjee & Linn 2018, Evans, 2012; Moorcroft, 2006; Moorcroft, Hurtt, & Pacala, 2001; Purves & Pacala, 2008; Thomas, Brookshire, & Gerber, 2015).
+As models increase in complexity, however, it remains important to consider that model developments and improvements have direct tradeoffs with increasing model complexity and computational costs, and potentially decreased interpretability of model output.
+
+In general, DGVMs fall along a spectrum that ranges from Individual Based Models (IBMs) on one end of the complexity/scale spectrum to big-leaf models on the other end.
+Most climate models incorporate vegetation dynamics using big-leaf models.
+These big-leaf models include most first generation DGVMs that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer.
+This simplification results in greater computational efficiency and thus easier integration into ESMs (Arora & Boer, 2010; Bonan, Levis, Sitch, Vertenstein, & Oleson, 2003; Cox, 2001; Krinner et al., 2005), although it comes at a cost of a lack of representation of many important demographic processes, canopy gap formation, vertical light competition, competitive exclusion, and successional recovery from disturbance (Feeley et al., 2007; Hurtt, Moorcroft, Pacala, & Levin, 1998; Moorcroft et al., 2001; Smith, Prentice, & Sykes, 2001; Stark et al., 2012).
+
+On the other end of the spectrum are IBMs, also referred to as forest gap models, representing vegetation at the level of individual plants (Christoffersen et al., 2016; Fischer et al., 2016; Fyllas et al., 2014; Sato, Ito, & Kohyama, 2007; Shuman, Shugart, & Krankina, 2014; Smith et al., 2001; Bugmann, 2001; Dietze & Latimer, 2011).
+These models are often stochastic and spatially explicit, capturing the spatial variability in the light environment and microclimates at high vertical and horizontal resolution.
+As a result, however, IBMs are computationally expensive and tend to be run at very local scales.
+
+In the middle of this spectrum are cohort-based models (CBMs), which are a compromise between the abstraction of big-leaf models and the computational expense of IBMs.
+Many CBMs represent vegetation as individual plants that are grouped together when they have properties (size, age, functional type) (Haverd et al., 2013; Hurtt et al., 1998; Lischke, Zimmermann, Bolliger, Rickebusch, & L€offler, 2006; Medvigy, Wofsy, Munger, Hollinger, & Moorcroft, 2009; Moorcroft et al., 2001; Scherstjanoi, Kaplan, & Lischke, 2014; Smith et al., 2014; Weng et al., 2015).
+The cohort approach retains the dynamics of IBMs, with reduced computational cost, but removes stochastic processes that can enhance the representation of functional diversity (Fisher et al., 2010).
+There is currently a lot of development of CBMs, most notably are the Ecosystem Demography model version 2 (ED2) and the Functionally Assembled Terrestrial Ecosystem Simulator (FATES), which is coupled with the Energy Exascale Earth System Model (E3SM) (Moorcroft et al.
+2001, Longo et al 2019 Pt 1 & pt 2, Holm et al ??
+for FATES).
+
+-	Still to come: more detailed description of vertical light environment and thermodynamics in ED (and FATES?)
+
 -	How do models currently integrate or not integrate these concepts and dynamics
--	What major gaps in knowledge and/or important questions emerge from this review (i.e.
-a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
 -	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?
-
-E.g.
+- E.g.
 see de Pury & Farquhar, 1997 : though likely an improvement over 'big-leaf' models, models that separate the canopy into sunlit and shaded portions (sun/shade models) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g.
@@
 
-[@nakamuraForestsTheirCanopies2017]--> modelling ideas?
-
-*(do we need a section on ecosystem function-- e.g., transpiration, gpp?)*
+*Knowledge Gaps & Future Research*
+-	What major gaps in knowledge and/or important questions emerge from this review (i.e.
+a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
+- Not sure where this section should go
 

### 2021-01-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf traits, thermoregulation and metabolism along the gradient, and with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 
-The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
+The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
 Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Conversly, available water facilitates leaf transpiration through open stomata, however increased relative humdity in tropical regions indirectly caused by increased temperature, reduces transpiration leading to high leaf temperatures in the canopy [Tibbitts 1979; Perez and Feeley 2018].
+In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humdity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [Tibbitts 1979; Perez and Feeley 2018].
 
-Frequent droughts related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
 Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
@@
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates [REF].
+However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [Perez and Feeley, 2018].
 
@@
 Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
-This will provide a pathway for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
+This will provide a pathway for understanding how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
 
@@
 This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013].
-As a result, degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also influences leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
+Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], ultimately influencing leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
 2016].
@@
 Additionally, shade leaves in the understory can be broken into three categories; 1.
-shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded.
+shade leaves of canopy trees in which the--top of the tree is sun exposed, bottom is shaded.
 2.
@@
 
-Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy compared to shorter understory trees that are shaded by overstory.
-As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and greater lobation than shade leaves
-[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].
+Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy, while shorter understory trees are shaded by overstory.
+As a result of different light regimes, a sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves
+[Rozendaal et al.
+2006, Mathur et al.
+2018, more REFS]."Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintaing a few layers of shade leaves as well" -Martijn
 
-However, the combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees.
-For example, overstory sun leaves in particular experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc..
-whereas for understory leaves are subjected to --shade or limitation of light, niche competition and herbivory.
-Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature for photosynthesis and metabolic processes.
+The combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees.
+For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc..
+whereas understory leaves are subjected to shade or limitation of light, niche competition and herbivory.
+Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heatloss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes.
 Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
-Adaptive traits in understory enable light capture efficiency while maintaing larger width leaves with comparitevely fewer stomtal density as a result of available water, lesser $VPD$, and wind speed.
-"Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn].
+Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to available water, lesser $VPD$, and wind speed.
 Those differences in traits are likely to affect their thermal properties as well.
@@
 New Phytol.
-182, 565–588 (2009).].Whereas, Rozendaal et al.
+182, 565–588 (2009).].Rozendaal et al.
 2006 found no significant trait differences with ontogeny and adult stature trees.
@@
 
-Leaf temperature is generally associated to increase with leaf area ($LA$) [ Wright et al.
+Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al.
 2019].
-Thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
+Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
 With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al.
-2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn traspirational cooling in large leaves [@fauset_differences_2018].
+2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018].
 Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
-2016 found leaf width to be a dominant factor in deterimining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
+2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
 Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref].
-Additionally, leaf characteristics such as lobes, leaf elongation and serration break the entirity of leaf dimension by enhancing a smaller characteristic dimension, all facilitate in dissapating heat more efficiently.
+Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation.
 Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al.
 2006].However, Perez and Feeley et al.
-2020 showed only a slight negetive effect on leaf temperature with leaf width.
+2020 showed only a slight negative effect on leaf temperature with leaf width.
 
@@
 Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width.
-Dissections in leaves such as lobes decreases leaf width and thus thermal time constant [@leighInfluenceLeafSize2017].
-Additionally, leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?
+Dissections in leaves, such as lobes, decrease leaf width and thus thermal time constant [@leighInfluenceLeafSize2017].
+Additionally, a leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?
 [I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
@@
 However, counteracting this, lower wind speed in understory would reduce latent heat loss.
-Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
-However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday leading to a decrease in leaf CO2 and consequentially increasing leaf temperature in upper canopy [@fauset_differences_2018].
+Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
+However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequentially increasing leaf temperature in upper canopy [@fauset_differences_2018].
 Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
@@
 
-Supporting the significant negetive relationship between $T50$ and thermal safety margins, photosynthesis [PS2] temperature regulation has been associated leaf thermal tolerance in order to maximize net carbon assimilation.
-However, randomly collected within-canopy leaves of individual mature plants in Florida showed that Tmax is not directly coordinated with high PS2 heat tolerance and does not necessarily increase leaf thermal tolerance [shows no direct correlation]; nor increases carbon assimilation at higher temperature.
-Rather, high PS2 heat tolerance reduces optimum temperature for C assimilation [$Topt$] because high PS2 heat tolerance requires a greater cost of PS2 metabolic activity maintainance thus not contributing to increased C assimilation.
-Species exhibiting low carbon assimilation rate tended to have greater stress tolerance.
-In addition, species with low stomtal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimaltion before thermal limits are reached [Perez et al.
+Supporting the significant negative relationship between $T50$ and thermal safety margins, photosynthesis [PS2] temperature regulation has been associated with leaf thermal tolerance in order to maximize net carbon assimilation.
+However, randomly collected within-canopy leaves of individual mature plants in Florida showed that Tmax is not directly coordinated with high PS2 heat tolerance and does not necessarily increase leaf thermal tolerance [shows no direct correlation]; nor does it increase carbon assimilation at higher temperature.
+Rather, high PS2 heat tolerance reduces optimum temperature for C assimilation [$Topt$] because high PS2 heat tolerance requires a greater cost of PS2 metabolic activity maintenance, thus not contributing to increased C assimilation.
+Species exhibiting low carbon assimilation rates tended to have greater stress tolerance.
+In addition, species with low stomatal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimilation before thermal limits are reached [Perez et al.
 2020]

### 2021-01-18 NEON_height_profiles/scripts/verticalProfileFigure.R
Reformatted figure #20

@@ -12,8 +12,8 @@ library(ggpubr)
 
 #
 #1. Download and save neon data ####
-## this is filtered to only the columns needed for plots 
-## (mean measurements per each 30 min observation) in order to not have massive files.
+# this is filtered to only the columns needed for plots
+# (mean measurements per each 30 min observation) in order to not have massive files.
 # source("scripts/get_neon_data.R")
 # 
 # pbapply::pblapply(sites, function(st){
@@ -24,7 +24,7 @@ library(ggpubr)
 #1a. Get NEON coordinates ####
 # library(data.table)
 # meta <- fread("data/site_data/NEON_Field_Site_Metadata_20201204.csv")
-# meta <- meta[field_site_id %in% c("BART","BONA","CLBJ","DEJU","DELA","GRSM","GUAN","HARV","JERC","LENO","MLBS","ORNL","OSBS","SCBI","SERC","SJER","SOAP","STEI","TALL","TEAK","TREE","UKFS","UNDE","WREF","YELL"), .(field_site_id, field_latitude, field_longitude, field_utm_easting, field_utm_northing, field_utm_zone)]
+# meta <- meta[field_site_id %in% c("BART","BONA","CLBJ","DEJU","DELA","GRSM","GUAN","HARV","JERC","LENO","MLBS","ORNL","OSBS", "PUUM", "SCBI","SERC","SJER","SOAP","STEI","TALL","TEAK","TREE","UKFS","UNDE","WREF","YELL"), .(field_site_id, field_latitude, field_longitude, field_utm_easting, field_utm_northing, field_utm_zone)]
 # 
 # fwrite(meta, "data/site_data/neon_coordinates.csv")
 
@@ -34,10 +34,17 @@ data <- fread("forested_NEON_sites.csv")
 data <- data[include==1, ]
 sites <- data[,site]
 
+#only focus on 6 core sites
+sites <- c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF")
+clrs <- c("#3399FF", "gold", "grey", "red", "#66CCFF", "#003399")
+
+sites <- c("HARV", "OSBS", "SCBI", "SERC", "WREF")
+clrs <- c("#3399FF", "gold", "red", "#66CCFF", "#003399")
+
 #bring in normalized height
 meta <- read_excel("data/site_data/TIS site metadata_20190403_forNOAA.xlsx", sheet=1)
[…]

### 2021-01-19 vertical-thermal-review.Rmd

@@
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves that will potentially reduce [forest] biomass acculumation, and disrupt photosynthesis and metabolic processes across plant species [Sullivan et al.
-2020, @perkinsIncreasingFrequencyIntensity2012].
+There is also increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] that will potentially reduce [forest] biomass  accumulation, and disrupt photosynthesis and metabolic processes across plant species [@sullivanLongtermThermalSensitivity2020].
 These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
@@
 Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humdity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [Tibbitts 1979; Perez and Feeley 2018].
+In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018]
 
 Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al.
+Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [Perez and Feeley, 2018].
+However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].
 
@@
 Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-In addition, understory growth phenology is sensitive to warming year round, and especially in spring [Zohnner_Renner 2019].
+In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].
 
@@
 However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks.
-Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]
+Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [@waySunflecksTreesForests2012]
 
 This difference in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity.
-Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].
+Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010; @sackHowStrongIntracanopy2006; @niinemetsControversyTraitsConferring2006].
 
@@
 Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiteny than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
-In low wind conditions, because $g_{a}$ to heat transfer is comparitevely small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; Leigh et al.
-2017].
+Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
+In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
 Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
@@
 Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
-This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constrains and lower osmotic potential.
-[@niinemetsPhotosyntheticAcclimationSimultaneous2004a].
+This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints and lower osmotic potential [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 In northeastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
@@
 Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
-Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]
+Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]
 
@@
 Furthermore, increases in mortality of big trees (e.g.
-Senf et al.
-2018) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
+@senfCanopyMortalityHas2018a) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
 (Elsa: see these papers on the long-term Harvard Forest soil warming experiments.
@@
 
-Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
+Species that have higher shade casting ability such as broadleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
 The strength of buffering also varies with respect to geographic and climatic factors.
@@
 
-Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [Rosendaal et al.
-2006].
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [Keenan and Niinemets et al.2016; @niinemetsLeafAgeDependent2016a].
+Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
 A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
-This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013].
-Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], ultimately influencing leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al.
-2016].
+This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013].
+Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].
 
@@
 2006, Mathur et al.
-2018, more REFS]."Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintaing a few layers of shade leaves as well" -Martijn
+2018, more REFS]."Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well" -Martijn
 
@@
 Those differences in traits are likely to affect their thermal properties as well.
-Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [Rozendaal et al.
-2006; @niinemetsResponsesForestTrees2010a]
+Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [@rozendaalPlasticityLeafTraits2006; @niinemetsResponsesForestTrees2010a]
 
 However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above.
-In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorphyll content on an area basis compared with sun leaves.
+In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves.
 This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
 1996, Poorter et al.
-2000, Rozendaal et al.
-2006]
+2000, @rozendaalPlasticityLeafTraits2006]
 
@@
 Early successional species are considered more shade intolerant than late successional species that are more shade tolerant.
-Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [Niinemets, 2006].
-Shade intolerant species experience greater heterogeniety in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H.
+Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [@niinemetsControversyTraitsConferring2006].
+Shade intolerant species experience greater heterogeneity in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H.
 Poorter, U.
@@
 Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf heat dissapation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1.
+The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1.
 summarizes relevant traits for thermal sensitivity.
@@
 Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
-However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, where as Q10 and E10 remained contant with height [Cavaleri et al.
+However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
 2008].
-In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [Coble & Cavaleri 2015; Coble et al.
-2016].In addition, carbon isotope composition also increased with height, independent of ligtht [Coble & Cavaleri 2015]
+In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]
 
 Height-based leaf trait variation in Quercus alba and Quercus velutina showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
-However, Sack et al.
-2006 did not observe trait variation based on height in Quercus spp.
+However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.
 
@@
 Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
-With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al.
+With increase in effective leaf width, boundary layer thickness and resistance increases along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018].
-Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
+Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
@@
 Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation.
-Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al.
-2006].However, Perez and Feeley et al.
+Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].However, Perez and Feeley et al.
 2020 showed only a slight negative effect on leaf temperature with leaf width.
@@
 High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from photoinhibition through D1 protein replacement.
-Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enable acclimation to high temperature [Mathur et al.
-2018].
+Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 These are found to be proportional to within-canopy light gradient at an inter- and intra-species level.
-Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [Niinemets et al.
-1998].
+Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998].
 In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al.
@@
 
-**Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
+**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
 $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 3).
-Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf, and thus leaf temperatures are decoupled from air temperature through adaptive physiological traits [@michaletzPlantThermoregulationEnergetics2015; Perez et al.
-2020]
+Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf, and thus leaf temperatures are decoupled from air temperature through adaptive physiological traits [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
 $T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].
@@
 
-Photosynthesis increases with increasing irradiance, and saturates at quatum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
+Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
 Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
@@
 Species exhibiting low carbon assimilation rates tended to have greater stress tolerance.
-In addition, species with low stomatal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimilation before thermal limits are reached [Perez et al.
-2020]
+In addition, species with low stomatal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimilation before thermal limits are reached [@perezPSIIHeatTolerances2020]
 
@@
 
-incr.humidity = decreased VPD = dec. latent heat loss = limited c assimilation [Perez et al. 2020]
+incr.humidity = decreased VPD = dec. latent heat loss = limited c assimilation [@perezPSIIHeatTolerances2020]
 

### 2021-01-19 vertical-thermal-review.Rmd

@@
 
+[Elsa: do we want a section about research opportunities using remote sensing or is that outside the scope of this review?] 
 An increasing availability of airborne and spaceborne LiDAR remote sensing data offers a promising opportunity for mapping vertical thermal gradients and vegetation structure.
@@
 *Representing Vertical Gradients in Models* 
+[How do models currently integrate or not integrate these concepts and dynamics]
 (@Nidhi & @Krista, feel free to edit/remove any of the above text.
@@
 
-Models representing ecosystem dynamics and vegetation demographic processes related to carbon, energy, and water fluxes operate at varying scales and have varying degrees of complexity and spatial resolution, including offline models and models that can be fully coupled with Earth System Models (ESMs).
-Dynamic global vegetation models (DGVMs) are the components of land surface models in ESMs used to predict the global distribution of vegetation types from physiological principles (Cao & Woodward, 1998; Foley et al., 1996; Sitch et al., 2003; Woodward & Lomas, 2004).
-Improved representation of vegetation demography in ESMs, specifically including forest canopy architectural variation in both horizontal and vertical directions, has repeatedly been identified as a critical step toward reduced uncertainty and more realistic representation of biologically mediated feedbacks in modeling future climates (Banerjee & Linn 2018, Evans, 2012; Moorcroft, 2006; Moorcroft, Hurtt, & Pacala, 2001; Purves & Pacala, 2008; Thomas, Brookshire, & Gerber, 2015).
-As models increase in complexity, however, it remains important to consider that model developments and improvements have direct tradeoffs with increasing model complexity and computational costs, and potentially decreased interpretability of model output.
+Models representing ecosystem dynamics related to carbon, energy, and water fluxes operate at a range of scales and have varying degrees of complexity , and include offline models and models that can be fully coupled with Earth System Models (ESMs).
+Dynamic global vegetation models (DGVMs) comprise the land surface models in ESMs used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks (Cao & Woodward, 1998; Foley et al., 1996; Sitch et al., 2003; Woodward & Lomas, 2004).
+Improved representation of vegetation demographic processes in ESMs, specifically including forest canopy architectural variation in both horizontal and vertical directions, has repeatedly been identified as a critical step toward reduced uncertainty  and more accurate characterization of biologically mediated feedbacks (Banerjee & Linn 2018, Evans, 2012; Moorcroft, 2006; Moorcroft, Hurtt, & Pacala, 2001; Purves & Pacala, 2008; Thomas, Brookshire, & Gerber, 2015).
+Still, as models increase in complexity, model developments and improvements have direct tradeoffs with increased computational cost and potentially decreased interpretability of model output.
 
-In general, DGVMs fall along a spectrum that ranges from Individual Based Models (IBMs) on one end of the complexity/scale spectrum to big-leaf models on the other end.
-Most climate models incorporate vegetation dynamics using big-leaf models.
-These big-leaf models include most first generation DGVMs that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer.
-This simplification results in greater computational efficiency and thus easier integration into ESMs (Arora & Boer, 2010; Bonan, Levis, Sitch, Vertenstein, & Oleson, 2003; Cox, 2001; Krinner et al., 2005), although it comes at a cost of a lack of representation of many important demographic processes, canopy gap formation, vertical light competition, competitive exclusion, and successional recovery from disturbance (Feeley et al., 2007; Hurtt, Moorcroft, Pacala, & Levin, 1998; Moorcroft et al., 2001; Smith, Prentice, & Sykes, 2001; Stark et al., 2012).
-
-On the other end of the spectrum are IBMs, also referred to as forest gap models, representing vegetation at the level of individual plants (Christoffersen et al., 2016; Fischer et al., 2016; Fyllas et al., 2014; Sato, Ito, & Kohyama, 2007; Shuman, Shugart, & Krankina, 2014; Smith et al., 2001; Bugmann, 2001; Dietze & Latimer, 2011).
-These models are often stochastic and spatially explicit, capturing the spatial variability in the light environment and microclimates at high vertical and horizontal resolution.
+In general, DGVMs range from Individual Based Models (IBMs) on one end of the complexity and scale spectrum to big-leaf models on the other end.
+Most climate models incorporate vegetation dynamics using big-leaf models, which include many first generation DGVMs that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer.
+This simplification results in greater computational efficiency and thus easier integration into ESMs (Arora & Boer, 2010; Bonan et al.
+2003; Cox, 2001; Krinner et al., 2005), although it comes at a cost of a lack of representation of important demographic processes, canopy gap formation, vertical light competition, competitive exclusion, and successional recovery from disturbance (Feeley et al., 2007; Hurtt et al.
+1998; Moorcroft et al., 2001; Smith et al.
+2001; Stark et al., 2012).
+On the other end of the spectrum are IBMs, also referred to as forest gap models, representing vegetation at the level of individual plants (Christoffersen et al., 2016; Fischer et al., 2016; Fyllas et al., 2014; Sato et al.
+2007; Shuman et al.
+2014; Smith et al., 2001; Bugmann, 2001; Dietze & Latimer, 2011).
+These models are often stochastic and spatially explicit, capturing spatial variability in the light environment and microclimates at high vertical and horizontal resolution.
 As a result, however, IBMs are computationally expensive and tend to be run at very local scales.
 
-In the middle of this spectrum are cohort-based models (CBMs), which are a compromise between the abstraction of big-leaf models and the computational expense of IBMs.
-Many CBMs represent vegetation as individual plants that are grouped together when they have properties (size, age, functional type) (Haverd et al., 2013; Hurtt et al., 1998; Lischke, Zimmermann, Bolliger, Rickebusch, & L€offler, 2006; Medvigy, Wofsy, Munger, Hollinger, & Moorcroft, 2009; Moorcroft et al., 2001; Scherstjanoi, Kaplan, & Lischke, 2014; Smith et al., 2014; Weng et al., 2015).
-The cohort approach retains the dynamics of IBMs, with reduced computational cost, but removes stochastic processes that can enhance the representation of functional diversity (Fisher et al., 2010).
-There is currently a lot of development of CBMs, most notably are the Ecosystem Demography model version 2 (ED2) and the Functionally Assembled Terrestrial Ecosystem Simulator (FATES), which is coupled with the Energy Exascale Earth System Model (E3SM) (Moorcroft et al.
-2001, Longo et al 2019 Pt 1 & pt 2, Holm et al ??
-for FATES).
+Cohort-based models (CBMs) sit in the middle of this continuum, between the oversimplified vegetation dynamics in big-leaf models and the computational expense of IBMs.
+CBMs represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type (Haverd et al., 2013; Hurtt et al., 1998; Lischke et al.
+2006; Medvigy et al.
+2009; Moorcroft et al., 2001; Scherstjanoi et al.
+2014; Smith et al., 2014; Weng et al., 2015).
+Although this cohort approach serves as a compromise that incurs lower computational costs than IBMs while capturing many of the same dynamics, CBMs do not incorporate the stochastic processes of IBMs that typically allow for greater representation of functional diversity (Fisher et al., 2010).
 
--	Still to come: more detailed description of vertical light environment and thermodynamics in ED (and FATES?)
+To review the vertical light environment and thermodynamics of CBMs in more detail, we focus here on the Ecosystem Demography model version 2 (ED2), a model that is actively undergoing continued development and application across a range of ecosystems and landscapes (Moorcroft et al.
+2001, Longo et al 2019 Pt 1 & pt 2).
+ED2 is also the progenitor of the Functionally Assembled Terrestrial Ecosystem Simulator (FATES), the DGVM coupled with the Energy Exascale Earth System Model (E3SM) (Xu & Christoffersen 2017 ) .
+In ED2, cohorts are tracked across multiple size-classes or individuals within the same plant functional type (PFT).
+Each cohort can encounter multiple light environments within a single climatic grid cell depending on patch-level disturbance history and vertical light competition that modulates interactions between plant traits and resource acquisition.
+Heterogeneity in the light environment and canopy structure across patches within a grid cell gives rise to varying horizontal and vertical micro-environments that differ in temperature, humidity, soil moisture, and soil nutrient conditions (Fisher et al 2018).
+The energy, water, and carbon dioxide cycles are solved separately for each patch in ED2, while fluxes and storage associated with individual plants are solved for each cohort within patches (Longo et al 2019).
+As a result, vegetation structure and functional distributions are emergent properties in the model that depend on plant functional traits and their interactions with abiotic environmental conditions.
+Each plant perceives a unique environment, including incident light, temperature, and vapor pressure deficit that varies across local scales in association with the horizontal and vertical position of each cohort relative to other cohorts, in addition to patch level differences in topographic and edaphic conditions.
+Coexistence of different vegetation types emerges from different environmental niches, either along a successional gradient of light availability or vertical position in the canopy (Moorcroft et al., 2001; Smith et al., 2001; Purves & Pacala, 2008).
 
--	How do models currently integrate or not integrate these concepts and dynamics
--	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?
-- E.g.
-see de Pury & Farquhar, 1997 : though likely an improvement over 'big-leaf' models, models that separate the canopy into sunlit and shaded portions (sun/shade models) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g.
-seasonal shifts in vertical leaf area distributions, Smith et al.
+Temperature is an atmospheric boundary condition driving the ED2 model (Longo et al 2019).
+Each patch in each grid cell is defined by a thermodynamic envelope, comprised of distinct thermodynamic systems for each soil layer, temporary surface water or snow layer, the aboveground component of each cohort, and the canopy air space (Longo et al 2019).
+The  atmosphere above and outside of the canopy air space is referred to as free air and is determined directly from boundary conditions in ED2.
+Patches exchange heat and mass with the free air  and lose water and associated energy through surface and sub-surface runoff.
+The net enthalpy flux of soil, temporary surface water, and vegetation thermodynamic systems in ED2 are exclusively due to associated water fluxes, whereas the eddy flux between the free air and the canopy air space includes both water transport and flux associated with the mixing of air with different temperatures between the canopy air space and free air (Longo et al 2019).
+
+Because most CBMs are spatially implicit within grid cells, direct and diffuse photosynthetically active radiation (PAR) is distributed to cohorts of differing heights using radiative transfer models that depend on designated rules associated with how crown architecture is represented, rather than via direct spatial competition determined by spatially explicit crown locations (Fisher et al.
+2010?).
+The flat-top crown method used in ED and ED2 stacks cohort-layers vertically, and a radiative transfer model is used to determine radiation absorbed by each layer at its midpoint, meaning that each cohort is shaded by all taller cohorts.
+This is distinct from other models that represent crown architecture using a perfect plasticity approximation (PPA) approach, which assumes that tree crowns completely fill canopy gaps through phototropism (stem leaning) and crown filling (Strigul et al.
+2008).
+Discrete canopy layers are comprised of self-organizing crowns that all receive the same incoming radiation.
+Once the canopy layer is filled with tree crowns, successive shorter trees inhabit the subsequent understory layer and are shaded by the trees in the canopy (Fisher et al 2010/2018?).
+Models using the PPA approach differ in number of understory layers (e.g.
+some only have one) (Fisher et al 2018).
+Though an improvement over big-leaf models, models that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g., seasonal shifts in vertical leaf area distributions, Smith et al.
 2019) and functions (e.g.
@@
 Hence, multi-layered ecosystem models may be necessary for accurately projecting future forest function.
+(E.g.
+see de Pury & Farquhar, 1997)
 
 *Knowledge Gaps & Future Research*
+-	What do [current model abilities or inabilities to integrate these concepts] mean for about ability to predict/understand how forest ecosystems will respond in the future?
 -	What major gaps in knowledge and/or important questions emerge from this review (i.e.

### 2021-01-19 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
+With increasing temperature experienced by foliage, photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum.
+Beyond this inflection point, fluxnet global data analysis shows that photosynthesis declines whereas respiration increases exponentially, therefore, moving from carbon sink to carbon source with increasing temperature, independent of water and irradiance [Duffy et al.2021]
+
 *VOC production*

### 2021-01-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ](schematic/schematic.png)
+
+![rough figure 1., needs some suggestions and info to include.] (https://github.com/EcoClimLab/vertical-thermal-review/blob/master/schematic/vertical%20thermal%20sensitivity%20image.png)
 
@@
 
-With increasing temperature experienced by foliage, photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum.
+With temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum.
 Beyond this inflection point, fluxnet global data analysis shows that photosynthesis declines whereas respiration increases exponentially, therefore, moving from carbon sink to carbon source with increasing temperature, independent of water and irradiance [Duffy et al.2021]

### 2021-01-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Even though canopy cover buffers maximum air temperatures in the understory, inevitable increases in understory temperatures continue to be observed as a result of global temperature increases.
-However, the buffering by canopy cover smooths this warming in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]
+However, the buffering by canopy cover smooths this warming in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]
 
@@
 It comprises of several underlying traits -- leaf area, LMA, leaf width, leaf dry matter content and stomatal conductance [@michaletzEnergeticCarbonEconomic2016].
-Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width.
+Typically, at a steady state, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width.
 Dissections in leaves, such as lobes, decrease leaf width and thus thermal time constant [@leighInfluenceLeafSize2017].
@@
 
-High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from photoinhibition through D1 protein replacement.
+High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from hronic photoinhibition through enabling D1 protein replacement.
 Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
@@
 
-In a tropical forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
+In a tropical wet and moist forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
 
@@
 [@fauset_differences_2018].
+- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
+Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
+
 -	(Slot et al. 2019) and refs therein
@@
 
+T50 might not capture thermal stress or sensitivity to increasing temperature in foliage, because the degree at which T50 is observed in leaves is atan extreme temperature not usually experienced by leaves acclimated to different light environment, at a lower foliage thermal stress point  T50 might be irrelevant, but variation to extreme temperature response might be better understood through T50 [@perezPSIIHeatTolerances2020]
+
 increased irradiance = increased VPD = increased air temp = midday stomatal closure =  decreased latent heat loss = increase leaf temperature = decreased C assimilation = greater PS2 heat tolerance = optimum rate of photosynthesis decreases = exhibiting lower t opt (optimum temp for c assimilation)
@@
 - The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
--  Within species, isporene production scales with light/ T
+-  Within species, isporene production scales with light/ T; however regarless of light, isoprene emission increases with T such as in understory shaded leaves?
 

### 2021-01-24 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 There is also increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] that will potentially reduce [forest] biomass  accumulation, and disrupt photosynthesis and metabolic processes across plant species [@sullivanLongtermThermalSensitivity2020].
-These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
+These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS; IPBES report), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
 
@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO2 that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 
@@
 
-Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
+Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+Larger trees suffer disproportionally greater hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
@@
 Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
-Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards disturbed forests with younger trees [@mcdowellPervasiveShiftsForest2020].
+Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominanted by younger trees [@mcdowellPervasiveShiftsForest2020].
 Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
@@
 
-While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory is still unclear.
+While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory are still unclear.
 Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
@@
 
-(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradient varies along the vertical canopy profile, however our current knowledge of this topic is lacking.)
+(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)
 
@@
 Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory.
-Shaded regions experience lower values and quality of PAR [in ranges of blue, orange and red].
+Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red].
 The light gradient is more pronounced under broadleaf forests than needleleaf forests [@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
@@
 This difference in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity.
-Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010; @sackHowStrongIntracanopy2006; @niinemetsControversyTraitsConferring2006].
+Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010; @sackHowStrongIntracanopy2006; @niinemetsControversyTraitsConferring2006; Osnas et al.
+2018].
 
@@
 
-Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in the dry season than the wet season [@rey-sanchez_spatial_2016] [*double check*]. 
 Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
@@
 
-Diurnal moderating capacity differs between forests with varying canopy structure, crown density, and seasonality.
-Tropical, temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests.
+The capacity to moderate diurnal temperature fluctuations differs among forests with varying canopy structure, crown density, and seasonality.
+Tropical, temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures relative to pine and boreal forests.
 Increases in crown density and leaf area index decrease light and higher maximum temperature permeability into the forest understory, thus maintaining lower maximum daytime temperatures.
@@
 RH, in combination with temperature, determines vapor pressure deficit, $VPD$.
-Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
-This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints and lower osmotic potential [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-In northeastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
+Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
+This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
 
@@
 Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
-Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]
+Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]
 
@@
 **The strength of this buffering varies across forests, influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
-First, buffering increases with canopy cover and water availabity.
+First, buffering increases with canopy cover and water availability.
 * greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
@@
 Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heatloss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes.
-Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA.
-Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to available water, lesser $VPD$, and wind speed.
+Additionally, overstory sun leaves tend to have long-lived leaves[*double check*] with high LMA while shade leaves of canopy trees have large leaves with low LMA.
+Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to low levels of photosynthesis in the shade, lesser $VPD$, and wind speed.
 Those differences in traits are likely to affect their thermal properties as well.
@@
 Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [@niinemetsControversyTraitsConferring2006].
+However, the findings from the above review conflicts with  Kitajima 1994 found for tropical seedlings; shade intolerance species grew faster in the sun AND in the shade, but at the cost of higher mortality in the shade.
+Lawren and Karoru had contrasting findinds, and so far appears that there is a consensus by now.
+
 Shade intolerant species experience greater heterogeneity in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H.
@@
 Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al.
-2019].
+2019; Tserej & Feeley 2021].
 Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
@@
 Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation.
-Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].However, Perez and Feeley et al.
-2020 showed only a slight negative effect on leaf temperature with leaf width.
+Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
 
@@
 Additionally, a leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?
+
+Tserej & Feeley 2021 experimentally showed that the rate of leaf warming is negetively correlated with leaf water content per unit leaf area (Lawren mentioned in the meeting as well?)
+
 [I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
@@
 
-High light stimulates the need for photoprotection in leaves in order to protect internal photosystem 2 from hronic photoinhibition through enabling D1 protein replacement.
-Thus leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition through enabling D1 protein replacement.
+Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 These are found to be proportional to within-canopy light gradient at an inter- and intra-species level.
@@
 3).
-Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf, and thus leaf temperatures are decoupled from air temperature through adaptive physiological traits [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
-$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].
+Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
+$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [@michaletzPlantThermoregulationEnergetics2015].
 
@@
 Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
-However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequentially increasing leaf temperature in upper canopy [@fauset_differences_2018].
+However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequentlly increasing leaf temperature in upper canopy [@fauset_differences_2018].
 Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
@@
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
--Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads.
+-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads.
 However, leaf temperatures themselves were moderated by air temperature.
@@
 This is primarily driven by the greater light available to sun leaves.
-In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, multiple layers of palisade parenchyma, and saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
+
+In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma.
+Light saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
 The vertical gradient in photosythetic rates is also influenced by available concentration of N and P. (and water availability)
@@
 
-Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
+Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy.
 Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
@@
 2015 ).
+
 Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis.
-Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought.
-[@niinemetsWorldwideAnalysisWithincanopy2015]
+At increasing temperatures, photosynthetic system II [PSII] is particularly vulnerable.
+Damage to PSII  can lead to decreased electron transport rates and photosynthetic failure.
+Irreversible damage to PSII is done at 40-60C [Baker 2008; @feeley_thermal_2020]
+Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought. [@niinemetsWorldwideAnalysisWithincanopy2015]
 
@@
 14.5; *revisit this!*) (Campbell and Norman 1998) as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
-However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photosnythesis in sun and shade leaves.
+However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
 For 3 species in Panama, @hernandezSimilarTemperatureDependence2020 and @slotPhotosyntheticHeatTolerance2019 found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
@@
 across latitude/climate (O’sullivan et al. 2017),
-with elevation [@feeley_thermal_2020;  Slot unpublished data],
+with elevation [@feeley_thermal_2020;  Slot et al unpublished data],
 with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
-Contrary to O'Sullivan et al.
-2017's findings, Perez and Feeley et al.
-2020 found that across inter-species sun leaves, higher photosynthetic heat tolerance in leaves lead to hotter leaf temperatures and narrower thermal safety margins and thus may be at a greater risk of thermal damage.
+Perez and Feeley et al.
+2020 found that across species sun leaves with higher photosynthetic heat tolerance experienced higher maximum temperatures and had narrower thermal safety margins.
+These species may thus be at a greater risk of thermal damage.
 [- Elsa: also see: 
@@
 
-T50 might not capture thermal stress or sensitivity to increasing temperature in foliage, because the degree at which T50 is observed in leaves is atan extreme temperature not usually experienced by leaves acclimated to different light environment, at a lower foliage thermal stress point  T50 might be irrelevant, but variation to extreme temperature response might be better understood through T50 [@perezPSIIHeatTolerances2020]
+T50 might not capture thermal stress or sensitivity to increasing temperature in foliage, because the degree at which T50 is observed in leaves is at an extreme temperature not usually experienced by leaves acclimated to different light environment, at a lower foliage thermal stress point T50 might be irrelevant, but variation to extreme temperature response might be better understood through T50 [@perezPSIIHeatTolerances2020]
 
@@
 **Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
-At increasing temperatures, Photosynthetic System 2 [PS2] is particularly vulnerable.
-Damage to pS2  can lead to decreased electron transport rates and photosynthetic failure.
-Irreversible damage to PS2 is done at 40-60C [Baker 2008; @feeley_thermal_2020]
 (Bolstad et al. 1999) did a study at Coweeta, including elev gradient. 

### 2021-01-25 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
+Related to something in the NSC paragraph:
+In this section we could perhaps also consider differences in biomass distribution between canopy trees an understory plants, and what that may mean for performance at high temperature.
+What I'm thinking is the temperature sensitivity of respiration that applies to leaves, stems, and roots, and even though thermal acclimation of leaf respiration is common, much less is known about acclimation of stem and root respiration, and incomplete acclimation still increases the respiration load at higher temperature.
+It is possible that an increased respiration load is more disadvantageous for understory plants because 1), I suspect that a greater fraction of their total biomass is life biomass that respires (as opposed dead heartwood in the stems of big trees), and 2), because they operate closer to their light compensation point, so if they lose more carbon in respiration, they would need  to make up for lost carbon in photosynthesis, but might lack sunlight to do so
+
 Interaction between degrees of shade x ontogeny x height with rising temperature needs more scietific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.

### 2021-01-26 vertical-thermal-review.Rmd
knit in new schematic
#27

@@
 
-![**Figure 1.
-(schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** Current version is just a very rough sketch of how this might look – needs to be filled in with content from the review.
-We could have a set of arrows for each of the major categories considered here.
-This would be a key figure, and should be beautifully illustrated—KAT could do a watercolor, or Nidhi could illustrate- and make use of NewPhyt's figure editing service for Tansley reviews).
-](schematic/schematic.png)
-
-![rough figure 1., needs some suggestions and info to include.] (https://github.com/EcoClimLab/vertical-thermal-review/blob/master/schematic/vertical%20thermal%20sensitivity%20image.png)
-
-Martijn : in this illustration it would also be nice to show that you can have shade leaves of canopy trees in the understory, and shaded plants of understory species.
+![**Figure 1. DRAFT schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** . ](schematic/vertical thermal sensitivity image.png)
 

### 2021-01-26 vertical-thermal-review.Rmd
taking stock of/ organizing/ updating figures, including auto numbering
#28 , #29

@@
 
-# This chunk is for auto numbering of tables/figures
+# DISPLAY ITEMS  (limit 8) # This chunk is for auto numbering of tables/figures
+#figures
+fig_schematic = 1
+fig_NEON_vertical =  2
+fig_NEON_differences  = 3
+fig_leaf_T = 4
+fig_treerings= 5 #not sure if we want to include one
 
-#figures
-fig_schematic = "1"
-fig_NEON_vertical =  "2"
-fig_NEON_differences  = "3"
+#tables
+table_leaf_traits=1
+table_leaf_metabolism=2
 
@@
 appendix_lit_review_methods <- "S3"
-appendix_SCBI_treering <- "S4"
 
@@
 appendix_lit_review_methods_title<- "Methods for literature review"
-appendix_SCBI_treering_title <- "Methods for SCBI tree-ring analysis"
 
@@
 **We lack a systematic, up to date understanding of biophysical and biological patterns across the vertical canopy gradient in forests, how this gradient affects leaf-level processes, and in turn how it affects ecosystems (Fig.
-` r fig_schematic`).**
+`r fig_schematic`).**
 
@@
 
-![**Figure 1. DRAFT schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** . ](schematic/vertical thermal sensitivity image.png)
+![**Figure `r fig_schematic`.
+DRAFT schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
+](schematic/vertical thermal sensitivity image.png)
 
@@
 
-*(paragraph on leaf area distribution, including figure ([separate from or combined with current Fig. 2- see GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
+*(paragraph on leaf area distribution, including figure ([GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
 The vertical structure and corresponding microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
@@
 2017].
-Similarly, in these studies, maximum temperatures were higher while humidity was lower in gaps compared to understory regions (both at 1.5 m above the ground, see Fig.
+Similarly, in these studies, maximum temperatures were higher while humidity was lower in gaps compared to understory regions (both at 1.5 m above the ground, see their Fig.
 4 - again illustrating buffering effect of the canopy).
@@
 
-![**Figure 2.
+![**Figure `r fig_NEON_vertical`.
 Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is just one of several+ intended sites.
-See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2.
-](NEON_height_profiles/figures/profile_SCBI.png)
+See [issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) and [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20).
+](NEON_height_profiles/figures/profile_all.png)
 
@@
 
-![**Figure 3.
+![**Figure `r fig_NEON_differences`.
 Role of canopy permeability in shaping temperature buffering, from NEON data.** Current placeholder figure is Nidhi's first rough analysis.
-See issue 2: https://github.com/EcoClimLab/vertical-thermal-review/issues/2.
+See [issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2).
 ](NEON_height_profiles/figures/differences_across_forests_temp.png)
@@
 Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1.
-summarizes relevant traits for thermal sensitivity.
+The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 
@@
 
-**Many traits vary with height and/or between sun and shade leaves of the same species (Table 1).**
+**Many traits vary with height and/or between sun and shade leaves of the same species (Table `r table_leaf_traits`).**
 Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.
 
-![(*For latest version of table, see tables.pdf.*)](table1.png)
+![Table `r table_leaf_traits`. (*For latest version of table, see tables.pdf.*)](table1.png)
 
@@
 $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
-3).
+`r fig_leaf_T`).
 Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
@@
 Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
-However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequentlly increasing leaf temperature in upper canopy [@fauset_differences_2018].
+However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018].
 Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
 
-![**Figure 3.
+![**Figure `r fig_leaf_T`.
 Leaf temperature in response to (a) XXX, (b) XXX (key trait or environmental variables).** Examples to be produced based on energy balance model using the tealeaves R package of @muir_tealeaves_2019: https://github.com/EcoClimLab/vertical-thermal-review/issues/6.
@@
 **When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig.
-**2?**).**
+`r fig_NEON_vertical`).**
 [Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
@@
 
-Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. 1). 
-However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves (Table 2).
+Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
+However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves (Table `r table_leaf_metabolism`).
 
-![(*For latest version of table, see tables.pdf.*)](table2.png)
+![Table `r table_leaf_metabolism`. (*For latest version of table, see tables.pdf.*)](table2.png)
 ```{r eval=FALSE, echo=FALSE}
@@
 **There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
-In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ (Rollinson et al., in review).
-Similarly, our previously unpublished analysis shows that in a broadleaf deciduous forest in Virginia, understory trees in a Mid-Atlantic forest responded more negatively to warmer $T_{air}$ than did canopy trees (Fig.
-**4**; Appendix `r appendix_SCBI_treering`).
-This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. **3?**). 
+In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; Rollinson et al., in press].
+This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
 However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
-**3**).
+`r fig_leaf_T`).
 It is also possible that competition is accelerated under warmer temperatures.
 
-Additionally, increased exposure to light increases non structural carbon reserve (NSC), plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availbility become limited, however with available water overall sensitivity to stress decreases.
-Understory shaded/younger trees have comparitively less NSC reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotransition demands, as well as greater investment in light harvesting than in photosynthesis--with greater overall sensitivity to stress and reduced tolerance.
+Additionally, increased exposure to light increases non structural carbon reserve (NSC), plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availability become limited, however with available water overall sensitivity to stress decreases.
+Understory shaded/younger trees have comparatively less NSC reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotranpiration demands, as well as greater investment in light harvesting than in photosynthesis--with greater overall sensitivity to stress and reduced tolerance.
 Larger NSC reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature].
@@
 
-Interaction between degrees of shade x ontogeny x height with rising temperature needs more scietific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
+Interaction between degrees of shade x ontogeny x height with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
 
-![**Figure 4.
-Temperature sensitivity of tree growth (derived from tree rings) for understory versus canopy trees.** THIS IS A ROUGH PLACEHOLDER-- KAT WILL PLAY WITH POTENTIAL VERSIONS.
-IF OUR IN-PREP TREE RING ANALYSIS METHODS PAPER IS ACCEPTED AHEAD OF THIS, WE CAN HIGHLIGHT RESULTS FROM THAT, WHICH ARE MORE RELIABLE.
-Analysis methods described in Appendix S4.
-OUTLIER HERE IS SHADE-TOLERANT BEECH.
-Data from Helcoski et al.
-2019.](SCBI_tree_rings/SCBIfig.png)
+![**Figure `r fig_treerings`.
+DRAFT tree-rings figure.
+Current figure shows (a) drought responses as a function of height in a temperate broadleaf forest (SCBI), from McGregor et al.
+2020; (b) responses to growing season temperature in eight New England forests, from Rollinson et al.
+2020.
+I'M NOT SURE THIS IS WHAT WE WANT, BUT I THINK SOME SORT OF TREE-RING FIGURE WOULD BE GOOD.
+[ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/tree_rings_draft.png)
 
@@
 
-Appendix `r appendix_SCBI_treering`. `r appendix_SCBI_treering_title`
-
 \newpage

### 2021-01-26 vertical-thermal-review_SI.Rmd
Update vertical-thermal-review_SI.Rmd
taking stock of/ organizing/ updating figures, including auto numbering
#28 , #29

@@
 
+METHODS ON NEON ANALYSIS GO HERE.
+
 \newpage
@@
 
+METHODS ON TEALEAVES GO HERE.
+
 \newpage
@@
 #  Appendix `r appendix_lit_review_methods`. `r appendix_lit_review_methods_title`
+
+METHODS ON LIT REVIEW GO HERE.
 

### 2021-01-26 vertical-thermal-review.Rmd
fig 2 caption

@@
 ![**Figure `r fig_NEON_vertical`.
-Vertical gradients in the biophysical environment, from NEON data.** Current placeholder figure is just one of several+ intended sites.
-See [issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) and [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20).
-](NEON_height_profiles/figures/profile_all.png)
+Vertical gradients in the biophysical environment for 5/6 sites in the National Ecological Observatory Network (NEON)**.
+Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+All variables are normalized relative to the height of the top of the canopy (**DETAILS**) and the conditions at the point of lowest measurement.
+([Issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) and [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)) ](NEON_height_profiles/figures/profile_all.png)
 

### 2021-01-26 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
#31
leaf T figure
#6

@@
 All variables are normalized relative to the height of the top of the canopy (**DETAILS**) and the conditions at the point of lowest measurement.
-([Issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) and [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)) ](NEON_height_profiles/figures/profile_all.png)
+([issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)) ](NEON_height_profiles/figures/profile_all.png)
 
@@
 Role of canopy permeability in shaping temperature buffering, from NEON data.** Current placeholder figure is Nidhi's first rough analysis.
-See [issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2).
+See [issue 31](https://github.com/EcoClimLab/vertical-thermal-review/issues/31).
 ](NEON_height_profiles/figures/differences_across_forests_temp.png)
@@
 
-Height-based leaf trait variation in Quercus alba and Quercus velutina showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
+Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
 However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.
@@
 ![**Figure `r fig_leaf_T`.
-Leaf temperature in response to (a) XXX, (b) XXX (key trait or environmental variables).** Examples to be produced based on energy balance model using the tealeaves R package of @muir_tealeaves_2019: https://github.com/EcoClimLab/vertical-thermal-review/issues/6.
-Current placeholder shows figures from Campbell&Norman.](leaf_energy_balance/C&N_figures.png)
+Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to stomatal conductance ($g_s$), wind speed, and shortwave radiation.** Schematic produced using the tealeaves R package of @muir_tealeaves_2019.
+Current placeholder uses typical conditions for Harvard Forest, but model output deviates from real data (directions right, but Tleaf-Tair too high).
+[ISSUE #6](2020_12_draft).](leaf_energy_balance/2020_12_draft.png)
 

### 2021-01-27 vertical-thermal-review.Rmd
review/ edit intro

@@
 
-Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
-There is also increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] that will potentially reduce [forest] biomass  accumulation, and disrupt photosynthesis and metabolic processes across plant species [@sullivanLongtermThermalSensitivity2020].
-These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS; IPBES report), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008].
+Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [REFS].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (REFS; IPBES report), including *(altered rates of photosynthesis and respiration)* [REF], foliar damage during heat waves [], and reduced growth and elevated mortality during drought [REFS].
+Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
+The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
 
 **Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
-
 Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages.
@@
 
-The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
-Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018]
+~~The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018]~~
 
-Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+~~Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
 Larger trees suffer disproportionally greater hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
@@
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].
+However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].~~
 
-While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
+~~While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
 Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
@@
 Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].
+In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].~~
 
-Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
+~~Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
 Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominanted by younger trees [@mcdowellPervasiveShiftsForest2020].
 Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Each of these changes have potential feedbacks to climate change.
+Each of these changes have potential feedbacks to climate change.~~
 
@@
 - The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
--  Within species, isporene production scales with light/ T; however regarless of light, isoprene emission increases with T such as in understory shaded leaves?
+-  Within species, isporene production scales with light/ T; however regardless of light, isoprene emission increases with T such as in understory shaded leaves?
 
-## Ecosystem Scaling & implications
+## Implications for ecology/ demography
 

### 2021-01-27 vertical-thermal-review.Rmd
bold font on topic sentences (intro)

@@
 
+**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
 Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [REFS].
@@
 
-**We lack a systematic, up to date understanding of biophysical and biological patterns across the vertical canopy gradient in forests, how this gradient affects leaf-level processes, and in turn how it affects ecosystems (Fig.
-`r fig_schematic`).**
-
-**This review addresses the following questions:**
+**Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance (Fig.
+`r fig_schematic`), and the implications for understanding forest ecosystem responses and feedbacks to climate change .**
 

### 2021-01-27 vertical-thermal-review.Rmd
review of "The Biophysical Environment"

@@
 
-The biophysical environment, here defined to include the physical structure of the forest canopy (from understory to the top of the canopy) and associated microclimates, is in large part determined by the structure of the forest itself.
-We supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S.
+**The biophysical environment across the vertical gradient from the forest floor to the the top of the canopy is in large part determined by the structure of the forest itself.** 
+Here, we define the biophysical environment to include the physical structure of the vegetation and associated physical conditions.
+In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S.
 National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`; Schimel et al.
@@
 
-Canopy foliage, which varies across forest types and seasonality, strongly shapes understory light, wind, etc.
-Forest canopies have a buffering effect on multiple aspects of the understory climate (Fig. `r fig_NEON_vertical`).
-
-- (*@bonan_ecological_2016 reviews this and points to appropriate references*)
-
-*(paragraph on leaf area distribution, including figure ([GitHub issue #15](https://github.com/EcoClimLab/vertical-thermal-review/issues/15))*
-The vertical structure and corresponding microenvironments of forest canopies vary seasonally (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019) and following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
+### Foliage density
+**Canopy foliage, which is vertically structured in correspondence with the size structure of trees in the forest, acts as the primary physical barrier between the atmosphere and the forest floor, producing a buffering effect on multiple aspects of the understory conditions (Fig.
+`r fig_NEON_vertical`).**
+Leaf area density.... *description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references)*
+Vertical profiles in leaf area density vary across forests, being...
+*summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
+They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
+*(can we say that there is more loss of leaves higher in the profile during cold/ drought dormant seasons? )*
 In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.
 
-Light conditions, specifically changes in photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [@koike_leaf_2001].
+![**Figure `r fig_NEON_vertical`.
+Vertical gradients in the biophysical environment for 5/6 sites in the National Ecological Observatory Network (NEON)**.
+Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+All variables are normalized relative to the height of the top of the canopy (**DETAILS**) and the conditions at the point of lowest measurement.
+([issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)) ](NEON_height_profiles/figures/profile_all.png)
+
+### Light environments
+**Light conditions, specifically changes in proportion incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
+`r fig_NEON_vertical`; @koike_leaf_2001].**
 Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory.
@@
 
+### Wind speeds 
 **Wind speeds are also higher at the top of the canopy (Fig. `r fig_NEON_vertical`).**
@@
 Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
+*(mention boundary layer conductance here, but move the discussion of leaf temperature to the T_leaf section )* ~~Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
 In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
 Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
-[ref]
+[ref] ~~
 
+### Air temperature 
 **Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
@@
 Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
-Higher (Elsa: or lower?) minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point?
+Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point?
 (e.g., greater radiative heat loss in exposed areas resulting in lower Tmin))
@@
 
+![**Figure `r fig_NEON_differences`.
+Relationship of PAR differences between canopy and understory to air temperature buffering (a) across sites and (b) through time within a single site.** From NEON data.
+Current placeholder figure is Nidhi's first rough analysis.
+See [issue 31](https://github.com/EcoClimLab/vertical-thermal-review/issues/31).
+](NEON_height_profiles/figures/differences_across_forests_temp.png)
+
+### Humidity and evaporative demand
 **Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**
@@
 
-**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
+### Carbon dioxide
+**Carbon dioxide (CO~2~) concentrations tend to be higher in the understory, primarily at night.** 
 	Higher in the understory, particularly at dusk [@koike_leaf_2001].
@@
 Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]
-
-Given that differences are small during the day when photosynthesis is occuring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
-
+Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
 If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e.
 from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy.
-Furthermore, increases in mortality of big trees (e.g.
-@senfCanopyMortalityHas2018a) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
-(Elsa: see these papers on the long-term Harvard Forest soil warming experiments.
+~~Furthermore, increases in mortality of big trees (e.g.
+@senfCanopyMortalityHas2018a) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.~~ (*but make the point about big tree mort elsewhere*)
+
+(*keep discussion of CO2 to what's in the paragraph above.
+Definitely don't want to go beyond a brief mention of warming effects on soil respiration*): (Elsa: see these papers on the long-term Harvard Forest soil warming experiments.
 CO2 emissions increased, then decreased, then increased again.
@@
 
-![**Figure `r fig_NEON_vertical`.
-Vertical gradients in the biophysical environment for 5/6 sites in the National Ecological Observatory Network (NEON)**.
-Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-All variables are normalized relative to the height of the top of the canopy (**DETAILS**) and the conditions at the point of lowest measurement.
-([issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)) ](NEON_height_profiles/figures/profile_all.png)
+*ALL OF THE FOLLOWING CONTENT SHOULD BE WORKED INTO RELEVANT PARAGRAPHS EARLIER IN THIS SECTION:*
 
-**The strength of this buffering varies across forests, influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
+The strength of this buffering varies across forests, influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.
 First, buffering increases with canopy cover and water availability.
@@
 However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]
-
-![**Figure `r fig_NEON_differences`.
-Role of canopy permeability in shaping temperature buffering, from NEON data.** Current placeholder figure is Nidhi's first rough analysis.
-See [issue 31](https://github.com/EcoClimLab/vertical-thermal-review/issues/31).
-](NEON_height_profiles/figures/differences_across_forests_temp.png)
 

### 2021-01-27 vertical-thermal-review.Rmd
Traits section
#33
started editing this/ not finished.

@@
 
+## Summary
+
 ```{r eval=FALSE, echo=FALSE}
-# based on Helene's Tansley review, and regular NewPhyt guidelines, I think the limit for New Phyt is 200 words. 
+#  I think the limit for New Phyt is 200 words. 
 ```
-
-## Summary
 
@@
 
-ii.	How do leaf traits vary with height (or between sun and shade leaves) in forests?(What types of leaf traits?-E.O)
+ii.	How do the leaf traits that most strongly influence leaf temperature and metabolism vary with height (or between sun and shade leaves) in forests?
 
@@
 
-Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].
+**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
 After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
@@
 
-(Elsa: it may also be worth mentioning leaf turnover / leaf longevity briefly in the above paragraph in the context of leaf development and trait formation.)
+*(Elsa: it may also be worth mentioning leaf turnover / leaf longevity briefly in the above paragraph in the context of leaf development and trait formation.)*
+*(Krista: the above is really interesting, and I like the idea of starting off the section by mentioning what controls the formation of leaf traits.
+However, I think it gets into more detail than needed.
+I'd follow Elsa's suggestion above without increasing the length.
+Also, is there work out there on how they're influenced by temperature/ water at the time of formation?
+I'd broaden to include the full microenviroment (without increasing length)*)
 
-An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [@chenLeafEconomicsSpectrum2020a], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
+*(move to modeling section:)* ~~An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [@chenLeafEconomicsSpectrum2020a], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
 Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
+However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].~~
 
-Classification schemes for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
+*(remove:)* ~~Classification schemes for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
 2018].
-For example, classifications from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".
+For example, classifications from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".~~
 
-Additionally, shade leaves in the understory can be broken into three categories; 1.
+*(cut, but keep this distinction in mind as you write:)* ~~Additionally, shade leaves in the understory can be broken into three categories; 1.
 shade leaves of canopy trees in which the--top of the tree is sun exposed, bottom is shaded.
@@
 3 understory plants that complete their lifecycle in the shaded understory.
-These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered [refs, -Martijn Slot]
+These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered [refs, -Martijn Slot]~~
 
-**Inter-species vertical gradient leaf trait variation**
+*(KAT: put intra-specific variation first, as we know a lot more about that. Then, add a brief section in inter-specific variation)*
+
+### Inter-species vertical gradient leaf trait variation
 
@@
 
-**Intra-canopy on the same species leaf trait variation**
+### Intra-canopy on the same species leaf trait variation
 

### 2021-01-27 vertical-thermal-review.Rmd
review of leaf temperature section

@@
 
-## Leaf temperature and hydraulics
+## Leaf temperature
+
+### Biophysical drivers
 
@@
 
-In addition, leaf size impacts leaf temperature.
+In addition, leaf size, which itself varies with height (Table `r table_leaf_traits`), impacts leaf temperature.
 While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018].
@@
 
-**Aligning with biophysical expectations, field observations have shown that leaf temperatures are influenced by the biophysical environment and leaf traits.**
-1.	Air T
-2.	Solar radiation
-3.	Leaf traits
+### In-situ observations
+**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`) and leaf traits (Table `r table_leaf_traits`) together shape in-situ leaf temperatures according to the principles outlined above.** 
+The net effect....
 

### 2021-01-27 vertical-thermal-review.Rmd
review metabolism section
#33

@@
 
-*Stomatal conductance*
+### Stomatal conductance
 Photosynthesis can be largely constrained by stomatal conductance. 
@@
 
-*Photosynthesis*
+### Photosynthesis
 
@@
 
-**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves**
+**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves.**
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
@@
 
-*Respiration*
+### Respiration
 
@@
 
-*VOC production*
+### VOC production
 

### 2021-01-27 vertical-thermal-review.Rmd
reviewed ecology section
#33

@@
 Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). 
-(*Martijn has some unpublished data on this that we might include: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
 At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999].
@@
 
-## Implications for ecology/ demography
+## Implications for demography and species distributions
 
-### Distribution of species and traits across the vertical profile
+**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology in several ways.** 
+Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in understory versus canopy conditions.
 
-*(This section should address inter-specific vs intra-specific variation, as Tables 1 and 2 refer to intra-specific variation, but most models will focus on trait variation across strata, regardless of species).
+### Demography
 
-**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology in several ways.** First, vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in understory versus canopy conditions.
--	*Spring and fall leaf phenology in temperate deciduous forests (Augsburger). 	Mechanism: more moderate conditions (T, wind) in understory (REF).*
-*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes (Condit et al.
-2000).
-One potential mechanism is lower temperature/hydraulic stress in understory*
-- something on isoprene production?
-
-**Appears to affect community change under warming (Zellweger et al. 2020).**
-
-### Scaling to tree level: demography
-
-**Vertical gradients also affect tree growth and survival, and how these respond to climatic variation.**
+**Vertical gradients affect tree growth and survival, and how these respond to climatic variation.**
 There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), and we will not attempt to review those here.
@@
 [ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/tree_rings_draft.png)
+
+### Distribution of species and traits across the vertical profile
+
+*(This section should address inter-specific vs intra-specific variation, as Tables 1 and 2 refer to intra-specific variation, but most models will focus on trait variation across strata, regardless of species).
+
+Ultimately, the way in which vertical gradients influence growth, mortality, and recruitment will shape forest composition across strata. ...
+
+-	*Spring and fall leaf phenology in temperate deciduous forests (Augsburger). 	Mechanism: more moderate conditions (T, wind) in understory (REF).*
+*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes (Condit et al.
+2000).
+One potential mechanism is lower temperature/hydraulic stress in understory*
+- something on isoprene production?
+
+**Appears to affect community change under warming (Zellweger et al. 2020).**
 

### 2021-01-27 vertical-thermal-review.Rmd
review of modelling section
#33

@@
 
-## Implications for demography and species distributions
+## Ecological implications
 
@@
 
-## Implications for Modeling / Scaling / Future Climate Projections
-
-*Ecosystem Scale Implications*
-(Elsa: I included info on ecosystem function here -- e.g., transpiration, gpp)
+### Ecosystem Scale Implications
 
@@
 2011, Ringgaard et al.
-2014).
+2014) *(KAT: but see Kunert et al.
+2017 -not sure if this is equivalent)*.
 With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.
@@
 
-*Representing Vertical Gradients in Models* 
-[How do models currently integrate or not integrate these concepts and dynamics]
-(@Nidhi & @Krista, feel free to edit/remove any of the above text.
+## Implications for Modeling and projecting climate change responses
+
+### Representing Vertical Gradients in Models
+
+*(@Nidhi & @Krista, feel free to edit/remove any of the above text.
 I'm adding more text than we'll want in the end, but I want to air on the side of more rather than less so that you can make decisions about what content you think best fits the goal of this review.
-At some point, perhaps we should discuss this section further and I can revise based on that.)
+At some point, perhaps we should discuss this section further and I can revise based on that.) *
 
@@
 (E.g.
-see de Pury & Farquhar, 1997)
-
-*Knowledge Gaps & Future Research*
--	What do [current model abilities or inabilities to integrate these concepts] mean for about ability to predict/understand how forest ecosystems will respond in the future?
--	What major gaps in knowledge and/or important questions emerge from this review (i.e.
-a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
-- Not sure where this section should go
+see de Pury & Farquhar, 1997).
 
@@
 
+*Where we seem to end up is that we really know very little about differences in temperature responses of metabolism and woody growth across the strata-- there are only a small number of studies, often showing insignificant differences or mixed results.*
+
 ## SI files

### 2021-01-31 NEON_height_profiles/scripts/lad_profs_v_therm_plotting_for_ian_v1.R
Updated script for figure #20

@@ -1,126 +1,189 @@
-
-################ CODE TO PLOT LAD AND LIGHT PROFILES FOR VERTICAL PROFILES TANSLEY REVIEW ##################
+######################################################
+# Purpose: Make LAD light profiles for Tansley Review
+# Developed by: Marielle Smith, edited by Ian McGregor
+# R version 3.6.2 - Edited January 2021
+######################################################
 
 #--Load required packages
-require(ggplot2)
-require(ggsci) # Needed for the scale_color_jco() colour scale
+library(ggplot2)
+library(ggsci) # Needed for the scale_color_jco() colour scale
 
-#--Load the data
-setwd("~/Dropbox (MSU Forest-Atmo)/MSU Forest-Atmo Team Folder/Marielle/Collaborations/Vertical_thermal_review/Codes")
-load("lad_n_light_profs_for_ian_v1.RData") # Loads 3 dataframes: LAD.all (LAD profiles), lgt.all (light transmission profs), prop.sun.all (proportion of sun leaves profiles)
-
-
-#>>---------------------------------------------------------------------------------------------------------------------------
-# PLOTTING
-#>>---------------------------------------------------------------------------------------------------------------------------
-
-########## PLOT THE PROFILES ###########
+load("data/lad_n_light_profs_for_ian_v1.RData") 
+# Loads 3 dataframes: LAD.all (LAD profiles), lgt.all (light transmission profs), prop.sun.all (proportion of sun leaves profiles)
 
 #--LAD scaling factor 
-lad.scale <- 5.5 # I have checked our LAI estimates against published values for SERC, but would be good to check more values 
+lad.scale <- 5.5 # I have checked our LAI estimates against published values for SERC, but would be good to check more values
 
-#--LAD
-plot.dat <- LAD.all
+dfList <- list(LAD = LAD.all, lgt = lgt.all, sun = prop.sun.all)
 
[…]

### 2021-02-01 archive/vertical-thermal-review_Feb1.Rmd
Create vertical-thermal-review_Feb1.Rmd
@NidhiVinod , archiving copy of current version here so that anything I delete/change will be easily retrievable.

--- 2021-01-27 vertical-thermal-review.Rmd
@@
 
-- CO2 may lead to denser understory (Martijn)
-
 - Influence of increasing temperatures on species compositions and feedbacks to future forest microclimates and function?
@@
 
-*Where we seem to end up is that we really know very little about differences in temperature responses of metabolism and woody growth across the strata-- there are only a small number of studies, often showing insignificant differences or mixed results.*
+- across vertical gradients, directional trends in biophysical environment and leaf traits are the rule, suggesting that the physiology and ecology that have these as their underpinnings should also vary
+- a number of studies confirm that this is the case.
+- However, when we ask the critical question of how the temperature sensitivities  of metabolism and woody growth varies across these different environments, we can't say much-- there are only a small number of studies, often showing insignificant differences or mixed results.
+- We lean towards the conclusion that while large canopy trees ware the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
+- Critically, change to the vertical gradient--e.g., through increased mortality of canopy trees--will change processses across forest strata, with potentially dramatic consequences.
+(some of the literature on forest degredation or fragmentation could inform what happens when those gradients change)
 

### 2021-02-01 vertical-thermal-review.Rmd
working on intro

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [REFS].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (REFS; IPBES report), including *(altered rates of photosynthesis and respiration)* [REF], foliar damage during heat waves [], and reduced growth and elevated mortality during drought [REFS].
+Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al.
+2014*].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (REFS; IPBES report), including altered rates of photosynthesis and respiration [REFS], foliar damage during heat waves [*O'Sullivan et al.
+2017*], and reduced growth and elevated mortality during drought [REFS].
 Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
@@
 Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
-This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO2 that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 
-~~The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018]~~
-
-~~Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-Larger trees suffer disproportionally greater hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
-2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
-Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].~~
-
-~~While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
-Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
-Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
-Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
-Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].~~
-
-~~Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
-Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominanted by younger trees [@mcdowellPervasiveShiftsForest2020].
-Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Each of these changes have potential feedbacks to climate change.~~
-
-While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory are still unclear.
-Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
+~~While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory are still unclear.~~ Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
 This will provide a pathway for understanding how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
@@
 
+Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
+Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominanted by younger trees [@mcdowellPervasiveShiftsForest2020].
+Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Each of these changes have potential feedbacks to climate change.
+
 ### Distribution of species and traits across the vertical profile
@@
 
+*Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+Larger trees suffer disproportionally greater hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
+2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
+Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
+However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].*
+
+*While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
+Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
+Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
+Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
+In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].*
+
+*The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
+Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018].*
+
 ## Conclusions
 
-*Where we seem to end up is that we really know very little about differences in temperature responses of metabolism and woody growth across the strata-- there are only a small number of studies, often showing insignificant differences or mixed results.*
+- across vertical gradients, directional trends in biophysical environment and leaf traits are the rule, suggesting that the physiology and ecology that have these as their underpinnings should also vary
+- a number of studies confirm that this is the case.
+- However, when we ask the critical question of how the temperature sensitivities  of metabolism and woody growth varies across these different environments, we can't say much-- there are only a small number of studies, often showing insignificant differences or mixed results.
+- We lean towards the conclusion that while large canopy trees ware the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
+- Critically, change to the vertical gradient--e.g., through increased mortality of canopy trees--will change processses across forest strata, with potentially dramatic consequences.
+(some of the literature on forest degredation or fragmentation could inform what happens when those gradients change)
 

### 2021-02-01 vertical-thermal-review.Rmd
worked on organization

@@
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic understanding of these gradients. 
+*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
+Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
 
-~~While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory are still unclear.~~ Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy.
-This will provide a pathway for understanding how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.
-
-(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)
-
-**Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance (Fig.
-`r fig_schematic`), and the implications for understanding forest ecosystem responses and feedbacks to climate change .**
+**Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance , and the implications for understanding forest ecosystem responses and feedbacks to climate change .**
+We structure the review around six major questions (Fig. `r fig_schematic`).
 
@@
 
-*(KAT: put intra-specific variation first, as we know a lot more about that. Then, add a brief section in inter-specific variation)*
+### Intra-canopy on the same species leaf trait variation
+
+Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+
+[I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations).
+Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear.
+photosynthesis, and canopy-level carbon uptake.
+They don't have to be treated separately, but it should always be clear which is being referred to.
+And in the consequences for scaling section this is of course also relevant.]
+[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
+Chen et al.
+2020 is a good paper to help frame this .
+They make an important point in that paper that we may also want to make here: that the LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy - hence the importance of this review.
+Keenan & Niinemets 2016  will also be helpful in making this point.]--M N.
+Smith
+
+**Many traits vary with height and/or between sun and shade leaves of the same species (Table `r table_leaf_traits`).**
+Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.
+
+![Table `r table_leaf_traits`. (*For latest version of table, see tables.pdf.*)](table1.png)
+
+```{r eval=FALSE, echo=FALSE}
+# refs in this table need to go in the nocite field below (easiest in alphabetical order)
+```
+---
+nocite: | 
+  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @kusiPlasticLeafMorphology2020
+---
+
+#### Morphological traits
+- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
+
+Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al.
+2019; Tserej & Feeley 2021].
+Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
+With increase in effective leaf width, boundary layer thickness and resistance increases along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
+2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018].
+Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
+Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
+2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
+Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref].
+Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation.
+Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
+
+#### Biochemical and physiological traits
+
+##### Carotenoids/Xanthophyll cycle pigments
+
+High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition through enabling D1 protein replacement.
+Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+These are found to be proportional to within-canopy light gradient at an inter- and intra-species level.
+Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998].
+In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al.
+2009]
+
+-	antioxidant scavenging function- protect against cellular damage
+- Koniger et. al. 1995, and  Matsubara et al. 2009
+
+##### Isoprene production
+-  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section) 
+- Elsa: I think Ty's section on isoprene would be better suited up here
+
+#### Leaf lifespan and Deciduousness ("Lifecycle / ecological traits? ")
+
+##### lifespan
+##### Deciduousness
+
+*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes (Condit et al. 2000). *
+*Spring and fall leaf phenology in temperate deciduous forests (Augsburger).*
 
 ### Inter-species vertical gradient leaf trait variation
+
+*(KAT: a lot of this contnet should probably move up the the intro to this section)*
 
@@
 
-### Intra-canopy on the same species leaf trait variation
-
-Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
-
-[I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations).
-Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear.
-photosynthesis, and canopy-level carbon uptake.
-They don't have to be treated separately, but it should always be clear which is being referred to.
-And in the consequences for scaling section this is of course also relevant.]
-[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
-Chen et al.
-2020 is a good paper to help frame this .
-They make an important point in that paper that we may also want to make here: that the LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy - hence the importance of this review.
-Keenan & Niinemets 2016  will also be helpful in making this point.]--M N.
-Smith
-
-**Many traits vary with height and/or between sun and shade leaves of the same species (Table `r table_leaf_traits`).**
-Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.
-
-![Table `r table_leaf_traits`. (*For latest version of table, see tables.pdf.*)](table1.png)
-
-```{r eval=FALSE, echo=FALSE}
-# refs in this table need to go in the nocite field below (easiest in alphabetical order)
-```
----
-nocite: | 
-  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @kusiPlasticLeafMorphology2020
----
-
-Are traits shaped more by height or light?
+### Are traits shaped more by height or light?
 
@@
 -- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
-
-*Leaf area / characteristic dimension*:
-- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
-
-Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al.
-2019; Tserej & Feeley 2021].
-Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
-With increase in effective leaf width, boundary layer thickness and resistance increases along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
-2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018].
-Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
-Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
-2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
-Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref].
-Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation.
-Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
-
-*Thermal time constant (probably not a lot out there)*
--	Curtis et. al, 2018 – higher in more exposed leaves (heat up slower, cool slower)
--	Michaeletz has used this in theoretical calculations (Michaeletz et al. 2016 and 2015?
-
-Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time.
-It comprises of several underlying traits -- leaf area, LMA, leaf width, leaf dry matter content and stomatal conductance [@michaletzEnergeticCarbonEconomic2016].
-Typically, at a steady state, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width.
-Dissections in leaves, such as lobes, decrease leaf width and thus thermal time constant [@leighInfluenceLeafSize2017].
-Additionally, a leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?
-
-Tserej & Feeley 2021 experimentally showed that the rate of leaf warming is negetively correlated with leaf water content per unit leaf area (Lawren mentioned in the meeting as well?)
-
-[I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant.
-In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot]
-
-*Carotenoids/Xanthophyll cycle pigments*
-
-High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition through enabling D1 protein replacement.
-Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-These are found to be proportional to within-canopy light gradient at an inter- and intra-species level.
-Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998].
-In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al.
-2009]
-
--	antioxidant scavenging function- protect against cellular damage
-- Koniger et. al. 1995, and  Matsubara et al. 2009
-
-*Isoprene production*
--  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section) 
-- Elsa: I think Ty's section on isoprene would be better suited up here
-
-*(Paragraph on intra- vs inter-specific variation)*
-
-- *Are patterns (in traits, metabolism, ecology) driven by tree height or exposure?*
 
@@
 
-### Distribution of species and traits across the vertical profile
-
-*(This section should address inter-specific vs intra-specific variation, as Tables 1 and 2 refer to intra-specific variation, but most models will focus on trait variation across strata, regardless of species).
-
-Ultimately, the way in which vertical gradients influence growth, mortality, and recruitment will shape forest composition across strata. ...
-
--	*Spring and fall leaf phenology in temperate deciduous forests (Augsburger). 	Mechanism: more moderate conditions (T, wind) in understory (REF).*
-*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes (Condit et al.
-2000).
-One potential mechanism is lower temperature/hydraulic stress in understory*
-- something on isoprene production?
-
-**Appears to affect community change under warming (Zellweger et al. 2020).**
-
 ### Ecosystem Scale Implications
@@
 With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.
+
+*(here, it would be good to add some content on contributions to C cycling across the vertical gradient.
+If Camille's paper is published ahead of this, that would be the best source.
+Meakem et al.
+2018, New Phyt has a similar analysis for 3 sites.)*
 

### 2021-02-01 vertical-thermal-review.Rmd
organization

@@
 
-## Trait variation
+## Leaf traits
+
+*(KAT: I think we want max 2-3 paragraphs here before getting to the subsection on  intra-specific variation.
+These paragraphs need to cover the content before then, including light vs height and how those relate to the data that have been collected,  setting up the distinction between intra- vs inter-specific variation, and some general intro, which is currently at beginning of inter-specific section.)*
 
@@
 
-### Intra-canopy on the same species leaf trait variation
+### Are traits shaped more by height or light?
+
+Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
+However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
+2008].
+In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]
+
+Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
+However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.
+
+Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
+- (Cavaleri et al. 2008) 
+- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues.
+I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
+- Elsa: see also H.
+Poorter, U.
+Niinemets, L.
+Poorter, I.
+J.
+Wright, R.
+Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
+New Phytol.
+182, 565–588 (2009).
+-- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
+
+### Intra-specific variation
 
@@
 
-[I think that this distinction should be addressed throughout the manuscript, so that it is always clear that the reader knows whether the cited references are about inter- or intra-specific variation, and that we address the different trait axes (Marielle made a comment on this earlier as well, that within species patterns may not follow leaf economics spectrum expectations).
-Ideally we would address the different categories, but if lack of information forces us to focus on just within species or just across species, this should still be made clear.
-photosynthesis, and canopy-level carbon uptake.
-They don't have to be treated separately, but it should always be clear which is being referred to.
-And in the consequences for scaling section this is of course also relevant.]
-[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
+*[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
 Chen et al.
@@
 Keenan & Niinemets 2016  will also be helpful in making this point.]--M N.
-Smith
+Smith*
 
@@
 
-##### Isoprene production
--  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section) 
-- Elsa: I think Ty's section on isoprene would be better suited up here
-
 #### Leaf lifespan and Deciduousness ("Lifecycle / ecological traits? ")
@@
 
-### Inter-species vertical gradient leaf trait variation
+### Inter-specific varaiation
 
@@
 
-### Are traits shaped more by height or light?
-
-Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
-However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
-2008].
-In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]
-
-Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
-However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.
-
-Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
-- (Cavaleri et al. 2008) 
-- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues.
-I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
-- Elsa: see also H.
-Poorter, U.
-Niinemets, L.
-Poorter, I.
-J.
-Wright, R.
-Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
-New Phytol.
-182, 565–588 (2009).
--- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
+##### Isoprene production
+-  Isoprene production *capability* as a trait
 

### 2021-02-01 vertical-thermal-review.Rmd
organization

@@
 **Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance , and the implications for understanding forest ecosystem responses and feedbacks to climate change .**
-We structure the review around six major questions (Fig. `r fig_schematic`).
+We structure the review around five major questions (Fig. `r fig_schematic`).
 
@@
 
-vi.	What are the implications for modeling/ scaling / What are the implications of these leaf-scale responses at the ecosystem scale [in the context of climate change] / How do these leaf-level responses scale to entire ecosystems(*nerge these better*)
+We then consider the implications for modeling/ scaling /climate change responses.
 
@@
 ### Carbon dioxide
-**Carbon dioxide (CO~2~) concentrations tend to be higher in the understory, primarily at night.** 
-	Higher in the understory, particularly at dusk [@koike_leaf_2001].
-Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999].
-Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]
+
+**Atmopsheric carbon dioxide (CO~2~) concentrations tend to be higher near ground level at night, but largely dissipate during the day [@koike_leaf_2001; @yang_spatial_1999].** 
+~~concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]~~ *(KAT: I'm confused by this.
+We want to discuss just atmospheric CO2 here, so I think this doesn't fit)*
 Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
-If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e.
-from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy.
-~~Furthermore, increases in mortality of big trees (e.g.
-@senfCanopyMortalityHas2018a) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.~~ (*but make the point about big tree mort elsewhere*)
-
-(*keep discussion of CO2 to what's in the paragraph above.
-Definitely don't want to go beyond a brief mention of warming effects on soil respiration*): (Elsa: see these papers on the long-term Harvard Forest soil warming experiments.
-CO2 emissions increased, then decreased, then increased again.
-It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
-- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J.
-and Frey, S.D., 2015.
-Long-term forest soil warming alters microbial communities in temperate forest soils.
-Frontiers in microbiology, 6, p.104.
-- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T.
-and Morrisseau, S., 2002.
-Soil warming and carbon-cycle feedbacks to the climate system.
-Science, 298(5601), pp.2173-2176.
-- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A.
-and Grandy, A.S., 2017.
-Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world.
-Science, 358(6359), pp.101-105.)
-- Nottingham, A.T., Meir, P., Velasquez, E. and Turner, B.L., 2020. Soil carbon loss by experimental warming in a tropical forest. Nature, 584(7820), pp.234-237.
+~~If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e.
+from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy.~~ *(this last sentence could be cut for space)*
 
@@
 
-## Implications for Modeling and projecting climate change responses
+## Implications
+
+*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
 
 ### Representing Vertical Gradients in Models
-
-*(@Nidhi & @Krista, feel free to edit/remove any of the above text.
-I'm adding more text than we'll want in the end, but I want to air on the side of more rather than less so that you can make decisions about what content you think best fits the goal of this review.
-At some point, perhaps we should discuss this section further and I can revise based on that.) *
 
@@
 
-### Implications for climate change responses
+### Climate change responses
 
@@
 
+increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness
+
 ## Conclusions

### 2021-02-01 tables.Rmd

@@
   column_spec(2, width = "2cm") %>%
+  column_spec(5, width = "2cm") %>%
   column_spec(6, width = "4cm") %>%
-  pack_rows("Leaf anatomy and morphological traits", 1, 16, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Leaf biochemical and physiological traits", 18, 29, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+  pack_rows("Leaf anatomy and morphological traits", 1, 17, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf biochemical and physiological traits", 18, 32, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
   #NOTE: REFERENCES for this table need to go in the nocite section of header in the main doc!!!

### 2021-02-01 vertical-thermal-review.Rmd
organization

@@
 Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
-Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
+Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree demography and ecosystem function.
 Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves, particularly in forests with dense canopies.
@@
 
-**Here we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance , and the implications for understanding forest ecosystem responses and feedbacks to climate change .**
-We structure the review around five major questions (Fig. `r fig_schematic`).
-
-i.	How does the biophysical environment vary with height in forests?
-
-ii.	How do the leaf traits that most strongly influence leaf temperature and metabolism vary with height (or between sun and shade leaves) in forests?
-
-iii. How do vertical changes in biophysical environment conditions and leaf traits interact to affect leaf temperature [and in turn photosynthetic capacity]??
-
-iv.	How does leaf metabolism respond to [air or leaf?] temperature in canopy and understory settings?
-
-v. What are the implications of vertical canopy gradient patterns for the ecological and climate change responses of canopy versus understory trees?
-
-We then consider the implications for modeling/ scaling /climate change responses.
-
-*Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests.
-However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), and the implications [or utility] for ecosystem models we also review studies focused on exposure gradients.*
+**Here we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
+We focus on five key themes (Fig. `r fig_schematic`):
+(1) the biophysical environment;
+(2) the leaf traits that most strongly influence leaf temperature and metabolism;
+(3) resulting leaf temperature ($T_{leaf}$) and
+(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
+(5) resultant size-structuring of whole-tree and ecosystem function.
+We then consider the implications for modeling /  climate change responses.
 
@@
 These paragraphs need to cover the content before then, including light vs height and how those relate to the data that have been collected,  setting up the distinction between intra- vs inter-specific variation, and some general intro, which is currently at beginning of inter-specific section.)*
+
+*Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests.
+However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), and the implications [or utility] for ecosystem models we also review studies focused on exposure gradients.*
 

### 2021-02-01 vertical-thermal-review.Rmd
remove PAR- T_air gradients figure
#31

@@
 fig_NEON_vertical =  2
-fig_NEON_differences  = 3
-fig_leaf_T = 4
-fig_treerings= 5 #not sure if we want to include one
+#fig_NEON_differences (issue #31)
+fig_leaf_T = 3
+fig_treerings= 4 #not sure if we want to include one
 
@@
 (5) resultant size-structuring of whole-tree and ecosystem function.
-We then consider the implications for modeling /  climate change responses.
+*We then consider the implications for modeling /  climate change responses.*
 
@@
 
-![**Figure `r fig_NEON_differences`.
-Relationship of PAR differences between canopy and understory to air temperature buffering (a) across sites and (b) through time within a single site.** From NEON data.
-Current placeholder figure is Nidhi's first rough analysis.
-See [issue 31](https://github.com/EcoClimLab/vertical-thermal-review/issues/31).
-](NEON_height_profiles/figures/differences_across_forests_temp.png)
-
 ### Humidity and evaporative demand
@@
 
-The strength of this buffering varies across forests, influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.
+The strength of this buffering varies across forests, influenced by both forest characteristics and the biophysical environment.
 First, buffering increases with canopy cover and water availability.

### 2021-02-01 vertical-thermal-review.Rmd
organization

@@
 #SI figure numbers
-SIfig_example = "S1"
+SIfigs_NEON_all = "S1-S#"
 
 #SI figure captions
-SIfig_example_legend <- "Figure S1. ...."
 
@@
 ### Wind speeds 
-**Wind speeds are also higher at the top of the canopy (Fig. `r fig_NEON_vertical`).**
-*(fill in some specifics from the NEON analysis)*
-Consistent with the results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019]
-
-Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-*(mention boundary layer conductance here, but move the discussion of leaf temperature to the T_leaf section )* ~~Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
-In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
-Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
-[ref] ~~
+**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
+Across the **#** forested NEON sites, windspeeds universally increased with height, with daily maximum wind speeds averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
+Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
 
@@
 ##### lifespan
+*Leaf turnover tends to be greater in the sun, so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
+This is bound to affect metabolism, and may affect other aspects of their performance.*
 ##### Deciduousness
@@
 Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
-$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [@michaletzPlantThermoregulationEnergetics2015].
+$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [Fig.
+`r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
+
+Greater upper canopy wind speeds (Fig.
+`r fig_NEON_vertical`) result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
+Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
+In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
+Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
+[ref]
 

### 2021-02-01 vertical-thermal-review.Rmd
Minor edits to draft above line 260

--- 2021-02-01 vertical-thermal-review.Rmd
@@
 Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
-Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree demography and ecosystem function.
+Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
 Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves, particularly in forests with dense canopies.
@@
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic understanding of these gradients. 
+Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic [replace synthetic w comprehensive?] understanding of these gradients.
 *(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
@@
 
-**Here we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
-We focus on five key themes (Fig. `r fig_schematic`):
-(1) the biophysical environment;
-(2) the leaf traits that most strongly influence leaf temperature and metabolism;
-(3) resulting leaf temperature ($T_{leaf}$) and
-(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
-(5) resultant size-structuring of whole-tree and ecosystem function.
-We then consider the implications for modeling /  climate change responses.
+**Here, we review biophysical and biological patterns across the vertical canopy gradient in forests, how these gradients affect leaf-level processes and, in turn, whole plant performance , and the implications for understanding forest ecosystem responses and climate change feedbacks.**
+We structure the review around five major questions (Fig. `r fig_schematic`).
+
+i.	How does the biophysical environment vary with height in forests?
+
+ii.	How do the leaf traits that most strongly influence leaf temperature and metabolism vary with height (or between sun and shade leaves) in forests?
+
+iii. How do vertical changes in biophysical environment conditions and leaf traits interact to affect leaf temperature [and in turn photosynthetic capacity]??
+
+iv.	How does leaf metabolism respond to [air or leaf?] temperature in canopy and understory settings?
+
+v. What are the implications of vertical canopy gradient patterns for the ecological and climate change responses of canopy versus understory trees?
+
+We then consider the implications for ecosystem-scale dynamics and climate change feedbacks, as well as remote sensing and modeling opportunities for future research.
+
+*Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests.
+However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), and the implications [or utility] for ecosystem models we also review studies focused on exposure gradients.*
 
@@
 ### Light environments
-**Light conditions, specifically changes in proportion incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
+**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
 `r fig_NEON_vertical`; @koike_leaf_2001].**
@@
 
-In heterogeneous canopies [with high gap fractions], the distance from the outer canopy [to X?] is a better proxy for light environment than height (Parker 1995) and lower canopy light environments are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]
+In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy [to X?] is a better proxy for light environment than height (Parker 1995).
+Lower canopy light environments are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]
 
-Tropical forests with densely packed overstory trees limit light to subsequent canopy layers and forest understories.
+Tropical forests with densely packed overstory trees limit light reaching subsequent canopy layers and forest understories.
 Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
@@
 
-RH, in combination with temperature, determines vapor pressure deficit, $VPD$.
+RH, in combination with temperature, determines vapor pressure deficit ($VPD$), an important metric expressing the driving force of water loss from a leaf.
 Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
@@
 
-*Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests.
-However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), and the implications [or utility] for ecosystem models we also review studies focused on exposure gradients.*
-
 **Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
 After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
+A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
 This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013].
@@
 2008].
-In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]
+In a temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]
 

### 2021-02-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-02-01 vertical-thermal-review.Rmd
@@
 Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red].
-The light gradient is more pronounced under broadleaf forests than needleleaf forests [@lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b]
-(Fig. `r fig_NEON_vertical`).
+The light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig.
+`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b].
 
@@
 ### Wind speeds 
-**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
-Across the **#** forested NEON sites, windspeeds universally increased with height, with daily maximum wind speeds averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
+**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [**REFS**].**
+Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
 Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
@@
 This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
 
@@
 
+In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
+
 **There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
-In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; Rollinson et al., in press].
+In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
 This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 

### 2021-02-02 vertical-thermal-review.Rmd
Minor edits and comments starting at line 260

--- 2021-02-01 vertical-thermal-review.Rmd
@@
 
-Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
-However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
+Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from changes in light availability.
+However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area-based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
 2008].
-In a temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]
+In a temperate Acer saccharum forest, ~~within-canopy Acer saccharum exhibited greater~~LMA and $N_{area}$ were greater with height early in the growing season as a result of hydrostatic constraints on leaf morphology.
+Later in the growing season, LMA and $N_{area}$ continued to increase in the upper canopy, but due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].
+In addition, carbon isotope composition has also been shown to increase with height, independent of light [@cobleLightAcclimationOptimizes2015]
 
-Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
-However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.
+Height-based leaf trait variation in two tall temperate species, *Quercus alba* and *Quercus velutina*, showed decreases in lobation with increased height, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
+However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus species.
 
@@
 
-Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Several studies point towards ~~common variations and~~differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
 The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
@@
 2019; Tserej & Feeley 2021].
-Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
-With increase in effective leaf width, boundary layer thickness and resistance increases along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
-2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018].
+Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperature ranges [$T_{range}$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
+With increases in effective leaf width, boundary layer thickness and resistance increase along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
+2019; @fauset_differences_2018].
+This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018].
 Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
-Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
+Using leaf energy balance models to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
 2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
-Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref].
+Leaf width and characteristic dimension are generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref].
 Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation.
-Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
+Leaf lobes are deeper and leaf elongation and serration are more pronounced in sun exposed canopies than in the shaded understory~~, as well as leaf elongation and serration are more pronounced in sun exposed canopies~~ [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
 
@@
 High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition through enabling D1 protein replacement.
-Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-These are found to be proportional to within-canopy light gradient at an inter- and intra-species level.
+Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis.
+Xanthophylls are photoprotective pigments comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+These are found to be proportional to within-canopy light gradients at an inter- and intra-species level.
 Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998].
-In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al.
+In neotropical forests, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pools on an area or chlorophyll basis compared to low light leaves that showed greater mass-based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al.
 2009]
@@
 
-##### lifespan
+##### Leaf lifespan [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
 ##### Deciduousness
@@
 whereas understory leaves are subjected to shade or limitation of light, niche competition and herbivory.
-Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heatloss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes.
+Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heat loss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes.
 Additionally, overstory sun leaves tend to have long-lived leaves[*double check*] with high LMA while shade leaves of canopy trees have large leaves with low LMA.
@@
 `r fig_leaf_T`).
-Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
+Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
 $T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [@michaletzPlantThermoregulationEnergetics2015].
@@
 Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
-However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018].
-Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
+However, increases in air temperature typically also increase VPD.
+Resulting decreases in stomatal conductance, typically during midday, lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018].
+Thus, $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
 
@@
 In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma.
-Light saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
+Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
 The vertical gradient in photosythetic rates is also influenced by available concentration of N and P. (and water availability)
@@
 
-## Implications
+## Implications [ & future research directions?]
 
+*Elsa: If emphsaize future research directions, could move the remote sensing opportunities to this section and frame modeling section around opportunities to use models to further explore open questions, scale these findings in space and time, and influence predictions*
 *(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
@@
 
-Models representing ecosystem dynamics related to carbon, energy, and water fluxes operate at a range of scales and have varying degrees of complexity , and include offline models and models that can be fully coupled with Earth System Models (ESMs).
+Models representing ecosystem dynamics related to carbon, energy, and water fluxes operate at a range of scales and have varying degrees of complexity, including offline models and models that can be fully coupled with Earth System Models (ESMs).
 Dynamic global vegetation models (DGVMs) comprise the land surface models in ESMs used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks (Cao & Woodward, 1998; Foley et al., 1996; Sitch et al., 2003; Woodward & Lomas, 2004).

### 2021-02-02 vertical-thermal-review.Rmd
organization/ writing
#33

--- 2021-02-01 vertical-thermal-review.Rmd
@@
 
-*(KAT: I think we want max 2-3 paragraphs here before getting to the subsection on  intra-specific variation.
-These paragraphs need to cover the content before then, including light vs height and how those relate to the data that have been collected,  setting up the distinction between intra- vs inter-specific variation, and some general intro, which is currently at beginning of inter-specific section.)*
+Particularly in dense forests, the upper, outer canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs.
+`r fig_NEON_vertical`, `r SIfigs_NEON_all`).
+Accordingly, there is substantive variation across the gradient in leaf traits (this section), which in turn shape leaf temperatures and metabolism (following sections).
 
-*Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests.
-However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), and the implications [or utility] for ecosystem models we also review studies focused on exposure gradients.*
+#### *(clarify light vs height - 1  short paragraph)*
 
-**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
-This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013].
-Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].
+Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.
+However, the majority of relevant research has focused on exposure gradients near ground level, with by far the most common study type being comparisons of sun and shade leaves. 
+Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light. 
+While light is the primary axis along which leaf traits vary, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig.
+`r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.
+In particular, hydraulic constraints vary with height on a tree (*Couvreur et al., 2018*; REFS).
 
-*(Elsa: it may also be worth mentioning leaf turnover / leaf longevity briefly in the above paragraph in the context of leaf development and trait formation.)*
-*(Krista: the above is really interesting, and I like the idea of starting off the section by mentioning what controls the formation of leaf traits.
-However, I think it gets into more detail than needed.
-I'd follow Elsa's suggestion above without increasing the length.
-Also, is there work out there on how they're influenced by temperature/ water at the time of formation?
-I'd broaden to include the full microenviroment (without increasing length)*)
+*(The following notes need to be condensed/ worked into the paragraph above.
+No need to get very specific (listing species studied, traits)--what we want is generalization.
+The goal is to answer the question, "Do we expect light gradients to be equivalent to height gradients, or is there variation attributable to height that is not captured by light gradients?")*
 
-*(move to modeling section:)* ~~An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [@chenLeafEconomicsSpectrum2020a], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
-Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].~~
+*  However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
+2008].
+In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015].
 
-*(remove:)* ~~Classification schemes for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al.
-2018].
-For example, classifications from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%".~~
-
-*(cut, but keep this distinction in mind as you write:)* ~~Additionally, shade leaves in the understory can be broken into three categories; 1.
-shade leaves of canopy trees in which the--top of the tree is sun exposed, bottom is shaded.
-2.
-seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy.
-3 understory plants that complete their lifecycle in the shaded understory.
-These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered [refs, -Martijn Slot]~~
-
-### Are traits shaped more by height or light?
-
-Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
-However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
-2008].
-In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]
-
-Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
+* Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
 However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.
 
-Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
+* Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
 - (Cavaleri et al. 2008) 
@@
 
-### Intra-specific variation
+#### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
+
+Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
+The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
+However, when it comes to understanding and modeling forest ecosystem function, the relevant scale of variation is interspecific (e.g., *Lamour, Serbin in prep*).
+
+### Intraspecific variation
+
+**Within a species, leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
+This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013].
+Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].
+
+*(Krista: the above paragraaph is really interesting, and I like the idea of starting off the section by mentioning what controls the formation of leaf traits.
+However, I think it gets into more detail than needed.
+Also, is there work out there on how they're influenced by temperature/ water at the time of formation?
+I'd broaden to include the full microenviroment (without increasing length)*)
 
@@
 The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
-
-*[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
-Chen et al.
-2020 is a good paper to help frame this .
-They make an important point in that paper that we may also want to make here: that the LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy - hence the importance of this review.
-Keenan & Niinemets 2016  will also be helpful in making this point.]--M N.
-Smith*
-
-**Many traits vary with height and/or between sun and shade leaves of the same species (Table `r table_leaf_traits`).**
-Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.
 
@@
 
-##### Carotenoids/Xanthophyll cycle pigments
+##### *(Carotenoids/Xanthophyll cycle pigments)*
 
@@
 
-##### lifespan
+##### *(lifespan)*
 *Leaf turnover tends to be greater in the sun, so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
 This is bound to affect metabolism, and may affect other aspects of their performance.*
-##### Deciduousness
+
+##### *(Deciduousness)*
 
@@
 
-### Inter-specific varaiation
+### Interspecific varaiation
 
-*(KAT: a lot of this contnet should probably move up the the intro to this section)*
+*(KAT: a lot of this content should probably move up the the intro to this section)*
 
-Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy, while shorter understory trees are shaded by overstory.
 As a result of different light regimes, a sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves
-[Rozendaal et al.
-2006, Mathur et al.
-2018, more REFS]."Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well" -Martijn
+[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].
+
+ "Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well" -Martijn
 
@@
 2001 has more info about this.
+
+*[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
+Chen et al.
+2020 is a good paper to help frame this .
+They make an important point in that paper that we may also want to make here: that the LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy - hence the importance of this review.
+Keenan & Niinemets 2016  will also be helpful in making this point.]--M N.
+Smith*
+
+An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [@chenLeafEconomicsSpectrum2020a], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
+Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
+However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
 
@@
 
+ if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
+
 ## Conclusions

### 2021-02-02 vertical-thermal-review.Rmd
writing/ organization
#33

@@
 
-#### *(clarify light vs height - 1  short paragraph)*
+#### *(clarify light vs height - 1 paragraph)*
 
@@
 `r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.
-In particular, hydraulic constraints vary with height on a tree (*Couvreur et al., 2018*; REFS).
+In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*], such that gas exchange is strongly constrained by at the tops of the world's tallest trees (~100 m.
+tall *Sequoia sempervirens*; *Ambrose et al.
+2009, 2010*).
+Disentangling the influences of hydraulic constraints, light, and other biophysical drivers that vary with height (Fig.
+`r fig_NEON_vertical`) on leaf form and function remains an important area for research, but is beyond the scope of the current review.
+For the purposes of interpreting the results presented here, we note that patterns across light and height gradients age generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
 
-*(The following notes need to be condensed/ worked into the paragraph above.
+*(The following notes need to be condensed/ worked into the paragraph above or discussion of specific traits.
 No need to get very specific (listing species studied, traits)--what we want is generalization.
@@
 2008].
-In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015].
 
@@
 **Within a species, leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody ~~and herbaceous~~ species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. 
+A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation intercepting the leaf surface.
 This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013].
@@
 #### Biochemical and physiological traits
+
+ In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015].
 

### 2021-02-03 vertical-thermal-review.Rmd
rearrangement
#41

@@
 (1) the biophysical environment;
-(2) the leaf traits that most strongly influence leaf temperature and metabolism;
-(3) resulting leaf temperature ($T_{leaf}$) and
+(3) leaf hydraulics and temperature ($T_{leaf}$);
+(2) the leaf traits most strongly influence leaf temperature and metabolism;
 (4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
-(5) resultant size-structuring of whole-tree and ecosystem function.
-*We then consider the implications for modeling /  climate change responses.*
+(5) size-structuring of whole-tree and ecosystem function.
+*We then consider the implications for understanding forest responses to global change, including  modeling /  climate change responses.*
 
@@
 However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]
+
+## Leaf hydraulics and temperature
+
+### Biophysical drivers
+
+**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
+$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+`r fig_leaf_T`).
+Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
+$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [Fig.
+`r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
+
+Greater upper canopy wind speeds (Fig.
+`r fig_NEON_vertical`) result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
+Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
+In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
+Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
+[ref]
+
+In addition, leaf size, which itself varies with height (Table `r table_leaf_traits`), impacts leaf temperature.
+While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018].
+Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.
+
+Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
+Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
+However, counteracting this, lower wind speed in understory would reduce latent heat loss.
+Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
+However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018].
+Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
+
+![**Figure `r fig_leaf_T`.
+Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to stomatal conductance ($g_s$), wind speed, and shortwave radiation.** Schematic produced using the tealeaves R package of @muir_tealeaves_2019.
+Current placeholder uses typical conditions for Harvard Forest, but model output deviates from real data (directions right, but Tleaf-Tair too high).
+[ISSUE #6](2020_12_draft).](leaf_energy_balance/2020_12_draft.png)
+
+### In-situ observations
+**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`) and leaf traits (Table `r table_leaf_traits`) together shape in-situ leaf temperatures according to the principles outlined above.** 
+The net effect....
+
+In a tropical wet and moist forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
+
+-  Michaletz et al have done some interesting work on leaf thermoregulation in relation to various leaf traits.
+I don’t think there is sun vs shade tests, but we could look at how this might translate, or how some of the traits he has calculated might differ between sun and shade leaves.
+For example, he has this thermal time constant parameter that is calculated from leaf mass per area, leaf dry matter content, and some environmental parameters.
+E.g.: DOI: 10.1038/nplants.2016.129, http://dx.doi.org/10.1016/j.tree.2015.09.006
+
+**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig.
+`r fig_NEON_vertical`).**
+[Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
+In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999). 
+Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
+
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
+-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads.
+However, leaf temperatures themselves were moderated by air temperature.
+[@fauset_differences_2018].
+- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
+Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
+
+-	(Slot et al. 2019) and refs therein
+-	(Fauset et al. 2018)
+Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.
+
+*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves - evaporative cooling - and leaf water content)*
+
+- Elsa: I came across these two studies that could be relevent in the evapotranspiration/temperature feedback section hinted at above: 
+- Staudt, K., Serafimovich, A., Siebicke, L., Pyles, R.D.
+and Falge, E., 2011.
+Vertical structure of evapotranspiration at a forest site (a case study).
+Agricultural and Forest Meteorology, 151(6), pp.709-729.
+- Sellin, A. and Lubenets, K., 2010. Variation of transpiration within a canopy of silver birch: effect of canopy position and daily versus nightly water loss. Ecohydrology, 3(4), pp.467-477.
 
@@
 -  Isoprene production *capability* as a trait
-
-## Leaf temperature
-
-### Biophysical drivers
-
-**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
-`r fig_leaf_T`).
-Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
-$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [Fig.
-`r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
-
-Greater upper canopy wind speeds (Fig.
-`r fig_NEON_vertical`) result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
-In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
-Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
-[ref]
-
-In addition, leaf size, which itself varies with height (Table `r table_leaf_traits`), impacts leaf temperature.
-While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018].
-Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.
-
-Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
-Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
-However, counteracting this, lower wind speed in understory would reduce latent heat loss.
-Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
-However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018].
-Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
-
-![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to stomatal conductance ($g_s$), wind speed, and shortwave radiation.** Schematic produced using the tealeaves R package of @muir_tealeaves_2019.
-Current placeholder uses typical conditions for Harvard Forest, but model output deviates from real data (directions right, but Tleaf-Tair too high).
-[ISSUE #6](2020_12_draft).](leaf_energy_balance/2020_12_draft.png)
-
-### In-situ observations
-**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`) and leaf traits (Table `r table_leaf_traits`) together shape in-situ leaf temperatures according to the principles outlined above.** 
-The net effect....
-
-In a tropical wet and moist forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
-
--  Michaletz et al have done some interesting work on leaf thermoregulation in relation to various leaf traits.
-I don’t think there is sun vs shade tests, but we could look at how this might translate, or how some of the traits he has calculated might differ between sun and shade leaves.
-For example, he has this thermal time constant parameter that is calculated from leaf mass per area, leaf dry matter content, and some environmental parameters.
-E.g.: DOI: 10.1038/nplants.2016.129, http://dx.doi.org/10.1016/j.tree.2015.09.006
-
-**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig.
-`r fig_NEON_vertical`).**
-[Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
-In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999). 
-Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
-
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
--Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads.
-However, leaf temperatures themselves were moderated by air temperature.
-[@fauset_differences_2018].
-- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
-Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
-
--	(Slot et al. 2019) and refs therein
--	(Fauset et al. 2018)
-Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.
-
-*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves - evaporative cooling - and leaf water content)*
-
-- Elsa: I came across these two studies that could be relevent in the evapotranspiration/temperature feedback section hinted at above: 
-- Staudt, K., Serafimovich, A., Siebicke, L., Pyles, R.D.
-and Falge, E., 2011.
-Vertical structure of evapotranspiration at a forest site (a case study).
-Agricultural and Forest Meteorology, 151(6), pp.709-729.
-- Sellin, A. and Lubenets, K., 2010. Variation of transpiration within a canopy of silver birch: effect of canopy position and daily versus nightly water loss. Ecohydrology, 3(4), pp.467-477.
 
@@
 
+## Projecting in time and space
+
+*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
+
+### Remote sensing 
 [Elsa: do we want a section about research opportunities using remote sensing or is that outside the scope of this review?] 
@@
 Existing research efforts have already revealed incredibly heterogeneity in vegetation structure, diversity, traits, and properties both within and across ecosystems (REFS – Stovall, Schneider, …others).
-
-## Implications
-
-*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
 
@@
 
-### Climate change responses
+## Global change responses
+
+### warming
 
@@
 
+ if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
+
+### canopy disturbance
+
 increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness
-
- if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
 

### 2021-02-03 vertical-thermal-review.Rmd
organization/ writing
#41 , #33

--- 2021-02-02 vertical-thermal-review.Rmd
@@
 
-## Introduction
+# Introduction
 
@@
 (1) the biophysical environment;
-(2) the leaf traits that most strongly influence leaf temperature and metabolism;
-(3) resulting leaf temperature ($T_{leaf}$) and
+(3) leaf temperature ($T_{leaf}$);
+(2) the leaf traits most strongly influence leaf temperature and metabolism;
 (4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
-(5) resultant size-structuring of whole-tree and ecosystem function.
-*We then consider the implications for modeling /  climate change responses.*
+(5) size-structuring of whole-tree and ecosystem function.
+*We then consider the implications for scaling across space and time, and understanding forest responses to global change.*
 
@@
 ](schematic/vertical thermal sensitivity image.png)
+
+# Review of vertical gradients
 
@@
 
+## Leaf temperature
+
+### Biophysical drivers
+
+**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
+$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+`r fig_leaf_T`).
+Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
+The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of radiation input versus heat lost to the environment [Fig.
+`r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
+
+![**Figure `r fig_leaf_T`.
+Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) stomatal conductance ($g_s$), and (d) leaf width.** Schematic produced using the tealeaves R package of @muir_tealeaves_2019.
+Current placeholder uses typical conditions for Harvard Forest, but model output deviates from real data (directions right, but Tleaf-Tair too high).
+[ISSUE #6](2020_12_draft).](leaf_energy_balance/2020_12_draft.png)
+
+**(paragraph on radiation as a driver)**
+Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
+
+Greater upper canopy wind speeds (Fig.
+`r fig_NEON_vertical`) result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
+Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{leaf}$ and $T_{air}$. 
+In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
+Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
+[ref]
+
+**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
+Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*. 
+Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. 
+However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO~2~, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018].
+Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.
+
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** 
+*Smaller leaves have higher $g_{a}$ to heat transfer, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
+(verify this statement and reference)*
+While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018]. 
+Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.
+
+### In-situ observations
+
+**Vertical profiles in microclimate (Fig.
+`r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed?
+and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig.
+`r fig_leaf_T`).**
+While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
+Under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
+For instance, during the dry season in a tropical moist forest in Panama...
+(Rey-Sánchez et al.
+2016).
+Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al.
+2016).
+At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.
+
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
+-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads.
+However, leaf temperatures themselves were moderated by air temperature.
+[@fauset_differences_2018].
+- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
+Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
+- In a tropical wet and moist forests of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
+-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al.
+2019)
+
+-	(Slot et al. 2019) and refs therein
+
+- Elsa: I came across these two studies that could be relevent in the evapotranspiration/temperature feedback section hinted at above: 
+- Staudt, K., Serafimovich, A., Siebicke, L., Pyles, R.D.
+and Falge, E., 2011.
+Vertical structure of evapotranspiration at a forest site (a case study).
+Agricultural and Forest Meteorology, 151(6), pp.709-729.
+- Sellin, A. and Lubenets, K., 2010. Variation of transpiration within a canopy of silver birch: effect of canopy position and daily versus nightly water loss. Ecohydrology, 3(4), pp.467-477.
+
 ## Leaf traits
@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
-Accordingly, there is substantive variation across the gradient in leaf traits (this section), which in turn shape leaf temperatures and metabolism (following sections).
+Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism and carbon balance (following section).
+It is unsuprising, then, that leaf traits vary dramatically across vertical gradients.
 
@@
 ##### *(lifespan)*
-*Leaf turnover tends to be greater in the sun, so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
+*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
 This is bound to affect metabolism, and may affect other aspects of their performance.*
@@
 
-## Leaf temperature
-
-### Biophysical drivers
-
-**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
-`r fig_leaf_T`).
-Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
-$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [Fig.
-`r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
-
-Greater upper canopy wind speeds (Fig.
-`r fig_NEON_vertical`) result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
-In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
-Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
-[ref]
-
-In addition, leaf size, which itself varies with height (Table `r table_leaf_traits`), impacts leaf temperature.
-While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018].
-Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.
-
-Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
-Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
-However, counteracting this, lower wind speed in understory would reduce latent heat loss.
-Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
-However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018].
-Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
-
-![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to stomatal conductance ($g_s$), wind speed, and shortwave radiation.** Schematic produced using the tealeaves R package of @muir_tealeaves_2019.
-Current placeholder uses typical conditions for Harvard Forest, but model output deviates from real data (directions right, but Tleaf-Tair too high).
-[ISSUE #6](2020_12_draft).](leaf_energy_balance/2020_12_draft.png)
-
-### In-situ observations
-**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`) and leaf traits (Table `r table_leaf_traits`) together shape in-situ leaf temperatures according to the principles outlined above.** 
-The net effect....
-
-In a tropical wet and moist forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
-
--  Michaletz et al have done some interesting work on leaf thermoregulation in relation to various leaf traits.
-I don’t think there is sun vs shade tests, but we could look at how this might translate, or how some of the traits he has calculated might differ between sun and shade leaves.
-For example, he has this thermal time constant parameter that is calculated from leaf mass per area, leaf dry matter content, and some environmental parameters.
-E.g.: DOI: 10.1038/nplants.2016.129, http://dx.doi.org/10.1016/j.tree.2015.09.006
-
-**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig.
-`r fig_NEON_vertical`).**
-[Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
-In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999). 
-Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
-
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
--Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads.
-However, leaf temperatures themselves were moderated by air temperature.
-[@fauset_differences_2018].
-- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
-Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
-
--	(Slot et al. 2019) and refs therein
--	(Fauset et al. 2018)
-Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.
-
-*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves - evaporative cooling - and leaf water content)*
-
-- Elsa: I came across these two studies that could be relevent in the evapotranspiration/temperature feedback section hinted at above: 
-- Staudt, K., Serafimovich, A., Siebicke, L., Pyles, R.D.
-and Falge, E., 2011.
-Vertical structure of evapotranspiration at a forest site (a case study).
-Agricultural and Forest Meteorology, 151(6), pp.709-729.
-- Sellin, A. and Lubenets, K., 2010. Variation of transpiration within a canopy of silver birch: effect of canopy position and daily versus nightly water loss. Ecohydrology, 3(4), pp.467-477.
-
 ## Leaf metabolism and thermal responses
@@
 Photosynthesis can be largely constrained by stomatal conductance.
--"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al.
-2019)
+
 - Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
@@
 
-**Photosynthesis is generally higher in sun leaves-- a fact that is well-established (Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020) and observed in numerous field studies.**
+**Photosynthesis is generally higher in sun leaves-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al.
+2015 ; Slot et al.
+2019; Chen et al.
+2020).**
 This is primarily driven by the greater light available to sun leaves.
@@
 Light saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
-The vertical gradient in photosythetic rates is also influenced by available concentration of N and P. (and water availability)
-.[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N.
+The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)
+*[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N.
 Smith [agreed.
@@
 Also; steepness of the vertical gradient in light and VPD.
-In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]
+In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*
 
 Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy.
+
 Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
@@
 
+**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves.**
+Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
+14.5 in Campbell and Norman 1998; *revisit this!*) *as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].* *(KAT: I'm confused by the second part of this sentence.
+Are these studies, or theoretical expectations?.)*
+However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
+For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [@slotPhotosyntheticHeatTolerance2019 ; @hernandezSimilarTemperatureDependence2020].
+Similar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
+Similarly, Mau et al. found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
+
+*(KAT: let's cut the portion on T50 down to one or at most two par:)*
+
 Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis.
@@
 
-**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves.**
-Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5; *revisit this!*) (Campbell and Norman 1998) as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
-However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
-For 3 species in Panama, @hernandezSimilarTemperatureDependence2020 and @slotPhotosyntheticHeatTolerance2019 found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant.
-Similar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
-Mau et al. found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
-
 **Leaf thermal tolerance (Tcrit/ T50)**
-- define.
+*(define)*
 Typical values ~45-50$^\circ$C, varying somewhat 
@@
 
-## Ecological implications
+## Size-structuring within the ecosystem
 
-**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology in several ways.** 
+**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology and ecosystem ecology in several ways.**
 Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in understory versus canopy conditions.
@@
 
-**Vertical gradients affect tree growth and survival, and how these respond to climatic variation.**
-There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), and we will not attempt to review those here.
+**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
+There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)*
+
 More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
@@
 Although drought is primarily a hydraulic problem, lack of water --> lower gs --> higher leaf T, so leaves face tradeoffs of water loss vs potentially damaging leafT.
-
-In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]
 
@@
 `r fig_leaf_T`).
-It is also possible that competition is accelerated under warmer temperatures.
+It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
 
@@
 
-Related to something in the NSC paragraph:
-In this section we could perhaps also consider differences in biomass distribution between canopy trees an understory plants, and what that may mean for performance at high temperature.
+*In this section we could perhaps also consider differences in biomass distribution between canopy trees an understory plants, and what that may mean for performance at high temperature.
 What I'm thinking is the temperature sensitivity of respiration that applies to leaves, stems, and roots, and even though thermal acclimation of leaf respiration is common, much less is known about acclimation of stem and root respiration, and incomplete acclimation still increases the respiration load at higher temperature.
-It is possible that an increased respiration load is more disadvantageous for understory plants because 1), I suspect that a greater fraction of their total biomass is life biomass that respires (as opposed dead heartwood in the stems of big trees), and 2), because they operate closer to their light compensation point, so if they lose more carbon in respiration, they would need  to make up for lost carbon in photosynthesis, but might lack sunlight to do so
+It is possible that an increased respiration load is more disadvantageous for understory plants because 1), I suspect that a greater fraction of their total biomass is life biomass that respires (as opposed dead heartwood in the stems of big trees), and 2), because they operate closer to their light compensation point, so if they lose more carbon in respiration, they would need  to make up for lost carbon in photosynthesis, but might lack sunlight to do so*
 
@@
 
-Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
-Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominanted by younger trees [@mcdowellPervasiveShiftsForest2020].
-Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Each of these changes have potential feedbacks to climate change.
+### C and water flux
 
-### Ecosystem Scale Implications
+*(here, it would be good to add some content on contributions to C cycling across the vertical gradient.
+If Camille's paper is published ahead of this, that would be the best source.
+Meakem et al.
+2018, New Phyt has a similar analysis for 3 sites.)*
 
@@
 As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see @nakamuraForestsTheirCanopies2017), leaf traits, and photosynthesis .
-Linked to this local biophysical variation is the partitioning of evapotranspiration (ET) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al.
+Linked to this local biophysical variation is the partitioning of evapotranspiration ($ET$) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al.
 2007, Staudt et al.
 2011).
-Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity (GPP, Banerjee & Linn 2018).
+Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity ($GPP$, Banerjee & Linn 2018).
 It has been demonstrated that net carbon assimilation and transpiration can vary substantially for forest canopies with the same LAI but different leaf area density distributions (e.g., Katul et al 2004 , Banerjee & Linn 2018).
@@
 
-Still, partitioning ET into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
+Still, partitioning $ET$ into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
 Eddy covariance techniques, sap flow measurements, and stable isotope analysis offer useful tools, although errors can be large (Williams et al 2004, Herbst et al.
@@
 
-*(here, it would be good to add some content on contributions to C cycling across the vertical gradient.
-If Camille's paper is published ahead of this, that would be the best source.
-Meakem et al.
-2018, New Phyt has a similar analysis for 3 sites.)*
+# Implications
 
-[Elsa: do we want a section about research opportunities using remote sensing or is that outside the scope of this review?] 
+## Scaling across space and time
+
+*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
+
+### Remote sensing
+
 An increasing availability of airborne and spaceborne LiDAR remote sensing data offers a promising opportunity for mapping vertical thermal gradients and vegetation structure.
@@
 
+ if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
+
+### canopy disturbance
+
+Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
+Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
+Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Each of these changes have potential feedbacks to climate change.
+
 increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness
-
- if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
 

### 2021-02-03 vertical-thermal-review.Rmd
organization/ writing/ added draft modeling schematic
#41 , #33 , #42

@@
 fig_treerings= 4 #not sure if we want to include one
+fig_model=5
 
@@
 (5) size-structuring of whole-tree and ecosystem function.
-*We then consider the implications for scaling across space and time, and understanding forest responses to global change.*
+*We then consider the implications for understanding forest responses to global change, including scaling across space and time.*
 
@@
 DRAFT schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
-](schematic/vertical thermal sensitivity image.png)
+](schematics/vertical thermal sensitivity image.png)
 
@@
 
+Having established how physical conditions and biological form and function vary across vertical gradients, we now turn attention to the implications of these patterns for our understanding of how forest ecosystems may be impacted by global change, and our ability to project this across space and time.
+
+## Global change responses
+
+### warming
+
+- Influence of increasing temperatures on species compositions and feedbacks to future forest microclimates and function?
+
+*Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+Larger trees suffer disproportionally greater hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
+2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
+Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
+However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].*
+
+*While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
+Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
+Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
+Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
+In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].*
+
+*The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
+Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018].*
+
+ if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
+
+### canopy disturbance
+
+--Variation across horizontal biophysical gradients (climate, soils, etc.)(E.O)
+
+*(there's a lot of literature on canopy disturbance impacts--tap into that in this section)*
+
+Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
+Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
+Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Each of these changes have potential feedbacks to climate change.
+
+increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness
+
 ## Scaling across space and time
 
-*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
+*Scaling our understanding across space and time requires remote sensing and models.*
 
@@
 
-Models representing ecosystem dynamics related to carbon, energy, and water fluxes operate at a range of scales and have varying degrees of complexity , and include offline models and models that can be fully coupled with Earth System Models (ESMs).
+Models are needed to predict future ecosystem dynamics, which remain one of the largest sources of uncertainty in Earth System Model (ESM) projections of the future of global carbon cycling and climate change (*Friedlingstein et al.
+2006*).
 Dynamic global vegetation models (DGVMs) comprise the land surface models in ESMs used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks (Cao & Woodward, 1998; Foley et al., 1996; Sitch et al., 2003; Woodward & Lomas, 2004).
+DGVMs operate at a range of scales and have varying degrees of complexity, ranging from detailed individual-based models (*a.k.a.* forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal resolution (Christoffersen et al., 2016; Fischer et al., 2016; Fyllas et al., 2014; Sato et al.
+2007; Shuman et al.
+2014; Smith et al., 2001; Bugmann, 2001; Dietze & Latimer, 2011), to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer (Fig.
+`r fig_model`).
+This simplification results in greater computational efficiency and thus easier integration into ESMs (Arora & Boer, 2010; Bonan et al.
+2003; Cox, 2001; Krinner et al., 2005), although it comes at a cost of a lack of representation of important demographic processes, canopy gap formation, vertical light competition, competitive exclusion, and successional recovery from disturbance (Feeley et al., 2007; Hurtt et al.
+1998; Moorcroft et al., 2001; Smith et al.
+2001; Stark et al., 2012).
 Improved representation of vegetation demographic processes in ESMs, specifically including forest canopy architectural variation in both horizontal and vertical directions, has repeatedly been identified as a critical step toward reduced uncertainty  and more accurate characterization of biologically mediated feedbacks (Banerjee & Linn 2018, Evans, 2012; Moorcroft, 2006; Moorcroft, Hurtt, & Pacala, 2001; Purves & Pacala, 2008; Thomas, Brookshire, & Gerber, 2015).
-Still, as models increase in complexity, model developments and improvements have direct tradeoffs with increased computational cost and potentially decreased interpretability of model output.
+Still, as models increase in complexity, model developments and improvements have direct tradeoffs with increased computational cost and potentially decreased interpretability of model output, highlighting the need to identify and parsimoniously represent the most essential drivers of forest ecosystem function.
 
-In general, DGVMs range from Individual Based Models (IBMs) on one end of the complexity and scale spectrum to big-leaf models on the other end.
+![**Figure `r fig_model`.
+DRAFT.
+Enrich this by showing 3-4 carefully selected scenarios (e.g., baseline, warming-mesic, warming-drought, canopy disturbance), and maybe a few key processes/ consequences (leaf T, C balance).** [Issue #42](https://github.com/EcoClimLab/vertical-thermal-review/issues/42).
+](schematics/model_schematic.png)
+
+**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.**
+*[how?]*
+
+~~In general, DGVMs range from Individual Based Models (IBMs) on one end of the complexity and scale spectrum to big-leaf models on the other end.
 Most climate models incorporate vegetation dynamics using big-leaf models, which include many first generation DGVMs that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer.
@@
 These models are often stochastic and spatially explicit, capturing spatial variability in the light environment and microclimates at high vertical and horizontal resolution.
-As a result, however, IBMs are computationally expensive and tend to be run at very local scales.
+As a result, however, IBMs are computationally expensive and tend to be run at very local scales.~~
 
-Cohort-based models (CBMs) sit in the middle of this continuum, between the oversimplified vegetation dynamics in big-leaf models and the computational expense of IBMs.
-CBMs represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type (Haverd et al., 2013; Hurtt et al., 1998; Lischke et al.
+**The computationally feasible approach to representing vertical structuring in DVGMs lies in Cohort-based models (CBMs), which sit in the middle of this continuum between the oversimplified vegetation dynamics in big-leaf models and the computational expense of individual-based models.** CBMs represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type (Haverd et al., 2013; Hurtt et al., 1998; Lischke et al.
 2006; Medvigy et al.
@@
 2014; Smith et al., 2014; Weng et al., 2015).
-Although this cohort approach serves as a compromise that incurs lower computational costs than IBMs while capturing many of the same dynamics, CBMs do not incorporate the stochastic processes of IBMs that typically allow for greater representation of functional diversity (Fisher et al., 2010).
+~~Although this cohort approach serves as a compromise that incurs lower computational costs than IBMs while capturing many of the same dynamics, CBMs do not incorporate the stochastic processes of IBMs that typically allow for greater representation of functional diversity (Fisher et al., 2010).~~
+
+*(KAT: I think all of the following content is great, but probably needs to be heavily tightened:)*
 
@@
 
-## Global change responses
-
-### warming
-
-- Influence of increasing temperatures on species compositions and feedbacks to future forest microclimates and function?
-
---Variation across horizontal biophysical gradients (climate, soils, etc.)(E.O)
-
-*Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-Larger trees suffer disproportionally greater hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
-2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
-Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].*
-
-*While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
-Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
-Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
-Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
-Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].*
-
-*The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
-Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018].*
-
- if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
-
-### canopy disturbance
-
-Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
-Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
-Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Each of these changes have potential feedbacks to climate change.
-
-increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness
-
 ## Conclusions

### 2021-02-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
a few changes based on comments from @tyeentaylor
#41

@@
 
-**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves.**
+**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
@@
 Similarly, Mau et al. found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
+
+*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.*
 
@@
 ## Global change responses
+ *(this section is currently just a bunch of text moved from elsewhere/ loose notes)*
 
@@
  if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
+
+*Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth).
+Differential mortality within the canopy structure will alter the profile of metabolism and hence emergent forest function.
+Such alteration to canopy structure may be detectable by lidar and thermal remote sensing.
+If we know how metabolism maps to the thermal profile, then the observed forest structural changes can inform prediction of forest function.*
 

### 2021-02-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Atmopsheric carbon dioxide (CO~2~) concentrations tend to be higher near ground level at night, but largely dissipate during the day [@koike_leaf_2001; @yang_spatial_1999].** 
-~~concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [because of lower photosynthetic rate in shade leaves?] [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]~~ *(KAT: I'm confused by this.
-We want to discuss just atmospheric CO2 here, so I think this doesn't fit)*
 Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
-~~If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e.
-from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy.~~ *(this last sentence could be cut for space)*
 
@@
 2001 has more info about this.
-
-*[Add something about how leaf traits along the vertical canopy profile are different than the expectation from leaf economic spectrum (LES)?
-Chen et al.
-2020 is a good paper to help frame this .
-They make an important point in that paper that we may also want to make here: that the LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy - hence the importance of this review.
-Keenan & Niinemets 2016  will also be helpful in making this point.]--M N.
-Smith*
 

### 2021-03-07 NEON_height_profiles/scripts/lad_profs_v_therm_plotting_for_ian_v1.R
Reformatted data, changed script, changed delta per #20

@@ -21,7 +21,8 @@ for(i in 1:3){
   df <- dfList[[i]]
   
   if(i==1){
-    df$val <- df$lad*lad.scale
+    # df$val <- df$lad*lad.scale #not standardizing as of Feb. 2021
+    df$val <- df$lad
   } else if(i==2){
     df <- df[df$ht>5, ]
     df$val <- df$lgt
@@ -36,14 +37,14 @@ for(i in 1:3){
   }
   
   #normalize the height
+  ## as of Feb. 2021, we are not normalizing the height
   df$ht.norm <- df$ht/(df$max.ht)
   
   #create the plots and send to list
   plotsLAD[[i]] <- local({
-    graph <- ggplot(df, aes(ht.norm, val, color=site)) +
+    graph <- ggplot(df, aes(val, ht.norm, color=site)) +
       geom_path(size=0.7) + 
-      coord_flip() +
-      xlim(0,1) +
+      ylim(0,1) +
       labs(x="", y = "") +
       theme_bw() +
       # theme_classic(16)  + 
@@ -51,13 +52,13 @@ for(i in 1:3){
     
     if(i==1){ #LAD
       graph <- graph +
-        ggtitle(expression(Leaf~area~density~(m^{2}~m^{-3})))
+        xlab(expression(Leaf~area~density~(m^{2}~m^{-3})))
     } else if(i==2){ # light profile
       graph <- graph +
-        ggtitle('Proportion incident light')
+        xlab('Proportion incident light')
     } else if(i==3){ #sun leaves
       graph <- graph +
-        ggtitle("Proportion sun leaves")
+        xlab("Proportion sun leaves")
     } 
     graph <- graph + 
       theme(plot.title = element_text(size=16))

### 2021-03-07 NEON_height_profiles/scripts/verticalProfileFigure.R

@@ -15,7 +15,7 @@ library(ggpubr)
 # this is filtered to only the columns needed for plots
 # (mean measurements per each 30 min observation) in order to not have massive files.
 # source("scripts/get_neon_data.R")
-# 
+# sites <- c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF")
 # pbapply::pblapply(sites, function(st){
 #   full_data <- get_NEON(x=st)
 #   save(full_data, file=paste0(st, "test.Rdata"))
@@ -41,11 +41,16 @@ clrs <- c("#3399FF", "gold", "grey", "red", "#66CCFF", "#003399")
 sites <- c("HARV", "OSBS", "SCBI", "SERC", "WREF")
 clrs <- c("#3399FF", "gold", "red", "#66CCFF", "#003399")
 
-#bring in normalized height
+#bring in normalized height (from Marielle)
 meta <- read_excel("data/site_data/TIS site metadata_20190403_forNOAA.xlsx", sheet=1)
 meta <- data.table(meta)
 meta <- meta[SITE %in% sites, 
-             ][,DistZaxsCnpy := as.numeric(DistZaxsCnpy)]
+             ][,`:=` (DistZaxsCnpy = as.numeric(DistZaxsCnpy),
+                    maxCanHeightMarielle = c(32, 43, 39, 21, 53))]
+normHeightData <- 
+  data.table(site=c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF"),
+                         normH=c(32, 21, 23, 43, 39, 53))
+# HARV=32, OSBS: 21, PUUM: 23, SCBI: 43, SERC: 39, WREF: 53
 
 dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT", "PARPAR"),
                  id = c("DP1.00001.001", "DP1.00098.001",
@@ -68,13 +73,14 @@ for(i in 1:length(dp[,name])){
   allTopHeight <- c()
   allNormHeight <- c()
   for(j in 1:length(sites)){
-    load(paste0("data/neon_rdata/", sites[j], "test.Rdata"))
+    # load(paste0("data/neon_rdata/", sites[j], "test.Rdata"))
+    load(paste0(sites[j], "test.Rdata"))
     meta_site <- meta[SITE == sites[j], ]
     
     #get max vertical height (for use in plots)
     hei <- sapply(full_data, function(x){unique(x[["verticalPosition"]])})
[…]

### 2021-03-08 vertical-thermal-review.Rmd

@@
 
-**Here we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
+**Here, we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
 We focus on five key themes (Fig. `r fig_schematic`):
@@
 ### Light environments
-**Light conditions, specifically changes in proportion incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
+
+**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
 `r fig_NEON_vertical`; @koike_leaf_2001].**
@@
 
+**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
+This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013].
+Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].
+
 ## Leaf traits
@@
 
-*  However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
+Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from changes in light availability.
+However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area-based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
 2008].
+In a temperate Acer saccharum forest, ~~within-canopy Acer saccharum exhibited greater~~LMA and $N_{area}$ were greater with height early in the growing season as a result of hydrostatic constraints on leaf morphology.
+Later in the growing season, LMA and $N_{area}$ continued to increase in the upper canopy, but due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].
+In addition, carbon isotope composition has also been shown to increase with height, independent of light [@cobleLightAcclimationOptimizes2015]
 
-* Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
-However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.
+Height-based leaf trait variation in two tall temperate species, *Quercus alba* and *Quercus velutina*, showed decreases in lobation with increased height, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
+However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus species.
 
@@
 However, when it comes to understanding and modeling forest ecosystem function, the relevant scale of variation is interspecific (e.g., *Lamour, Serbin in prep*).
+
+Several studies point towards differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 
@@
 
-##### *(lifespan)*
+##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
 *Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
@@
 
-##### *(Deciduousness)*
+##### *Deciduousness*
 
@@
 In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma.
-Light saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
+Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
 The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)
@@
 
+## Implications [ & future research directions?]
+
+*Elsa: If emphsaize future research directions, could move the remote sensing opportunities to this section and frame modeling section around opportunities to use models to further explore open questions, scale these findings in space and time, and influence predictions*
+*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
+
 Still, partitioning $ET$ into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
@@
 With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.
-
-# Implications
 

### 2021-03-08 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Leaf energy balance R code

@@ -1,0 +1,603 @@
+####################################
+# Purpose: Create leaf energy balance figure using HARV neon data parameters for a typical overstory
+# and understory, normal and drought scenarios
+# Tleaves R package created by Chris Muir, 2019
+# Developed by: Nidhi Vinod, contact Nidhi Vinod (nidrup@gmail.com)
+# R version 3.5.3 --First created in October 2020
+# Sept-March. 2020
+####################################
+#install packages if needed
+library(magrittr)
+library(tealeaves)
+library(ggplot2)
+library(dplyr)
+library(ggthemes)
+library(moderndive)
+library(ggpubr)
+library(cowplot)
+
+####################################
+#create data frames using HARV NEON data for environmental parameters. Leaf characteristic dimmension in m is measured from Sun and shade Red oak leaves
+#all variables constant except for one independent variable that represents minimum- maximum range
+#Stomtal conductane(gs) value magnitudes obtained from Tleaves supplementary manual for range of values
+#normal represents HARV neon environmental parameters as observed
+#drought represents low gs and low relative humidity
+####################################
+
+# Create a dataframe for overstory and understory, for scenarios: normal and drought
+# 1. Short wave radiation, normal
+#understory
+sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800), 
+                  rh = 0.97, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
+                  gs = 2.0)
+#overstory
+sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800), 
+                  rh = 0.91, wind = 2.88, tair = 298.1, pressure = 99, leaf.size = 0.04, 
+                  gs = 4.0)
+
+sw_l$par <- sw_l$par*0.5#sw
+sw_u$par <- sw_u$par*0.5#sw
+
+#apply Tleaves function (this is repeated througout)
[…]

### 2021-03-08 vertical-thermal-review.Rmd
knit latest

@@
 fig_NEON_vertical =  2
-#fig_NEON_differences (issue #31)
 fig_leaf_T = 3

### 2021-04-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack synthetic [replace synthetic w comprehensive?] understanding of these gradients.
+Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients. 
 *(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
@@
 ### Foliage density
-**Canopy foliage, which is vertically structured in correspondence with the size structure of trees in the forest, acts as the primary physical barrier between the atmosphere and the forest floor, producing a buffering effect on multiple aspects of the understory conditions (Fig.
+**Canopy foliage, which is vertically structured in correspondence with the size, structure of trees in the forest, acts as the primary physical barrier between the atmosphere and the forest floor, producing a buffering effect on multiple aspects of the understory conditions (Fig.
 `r fig_NEON_vertical`).**

### 2021-04-05 tables.Rmd
temporarily stop evaluation of Table 2, which is where knit is failing
#32

@@
 
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
+```{r eval = FALSE, echo=FALSE, warning=FALSE}
 library(knitr)

### 2021-04-05 tables.Rmd
got Table 1 rendering correctly, adjusted location of headers
#32

@@
 kable(traits_table, caption = "Table 1.
-Summary of observed variation in thermally-relevant leaf traits with canopy height and/or between sun and shade leaves", format = "latex", booktabs = TRUE,  escape = F)%>%
+Summary of observed variation in thermally-relevant leaf traits with canopy height and/or between sun and shade leaves", format = "latex", booktabs = TRUE, longtable = TRUE, escape = F)%>%
   column_spec(1, width = "4cm") %>%
   column_spec(2, width = "2cm") %>%
-  column_spec(5, width = "2cm") %>%
+  column_spec(5, width = "1cm") %>%
   column_spec(6, width = "4cm") %>%
-  pack_rows("Leaf anatomy and morphological traits", 1, 17, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Leaf biochemical and physiological traits", 18, 32, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
+  pack_rows("Leaf anatomy and morphological traits", 1, 28, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Leaf biochemical and physiological traits", 29, 55, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options =c("repeat_header"), protect_latex = T, font_size = 7)
   #NOTE: REFERENCES for this table need to go in the nocite section of header in the main doc!!!

### 2021-04-05 tables.Rmd
update Table2 parallel to 1 (doesn't solve problem)
#32

@@
 
-```{r eval = FALSE, echo=FALSE, warning=FALSE}
+```{r eval = T, echo=FALSE, warning=FALSE}
 library(knitr)
@@
 kable(metabolism_table, caption = "Table 2.
-Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves", format = "latex", booktabs = TRUE,  escape = F)%>%
+Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves", format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F)%>%
   column_spec(1, width = "6cm") %>%
@@
   pack_rows("VOC production", 29, 30, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+  kable_styling(latex_options =c("repeat_header"), protect_latex = T, font_size = 7)
   #NOTE: REFERENCES for this table need to go in the nocite section of header in the main doc!!!

### 2021-04-05 tables.Rmd
fix formatting and row packing
#32

@@
   column_spec(2, width = "2cm") %>%
+  column_spec(3, width = "2cm") %>%
+  column_spec(4, width = "2cm") %>%
   column_spec(5, width = "1cm") %>%
-  column_spec(6, width = "4cm") %>%
+  column_spec(6, width = "3cm") %>%
   pack_rows("Leaf anatomy and morphological traits", 1, 28, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
@@
 Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves", format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F)%>%
-  column_spec(1, width = "6cm") %>%
+  column_spec(1, width = "4cm") %>%
   column_spec(2, width = "2cm") %>%
-  column_spec(6, width = "4cm") %>%
-  pack_rows("Stomatal conductance", 1, 7, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Photosynthesis", 8, 20, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Respiration", 21, 28, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("VOC production", 29, 30, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  column_spec(3, width = "2cm") %>%
+  column_spec(4, width = "2cm") %>%
+  column_spec(5, width = "1cm") %>%
+  column_spec(6, width = "3cm") %>%
+  pack_rows("Stomatal conductance", 1, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Photosynthesis", 12, 49, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Respiration", 50, 68, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("VOC production", 69, 72, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   kable_styling(latex_options =c("repeat_header"), protect_latex = T, font_size = 7)

### 2021-04-05 vertical-thermal-review.Rmd
updated leaf energy balance figure in manuscript
@NidhiVinod , FYI. Let's always save the latest version of this figure as leaf_energy_balance.png.  I've deleted older placeholder versions.

#6

@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) stomatal conductance ($g_s$), and (d) leaf width.** Schematic produced using the tealeaves R package of @muir_tealeaves_2019.
-Current placeholder uses typical conditions for Harvard Forest, but model output deviates from real data (directions right, but Tleaf-Tair too high).
-[ISSUE #6](2020_12_draft).](leaf_energy_balance/2020_12_draft.png)
+Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus XXX*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
+[ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/leaf_energy_balance.png)
 

### 2021-04-05 vertical-thermal-review.Rmd
a bit of work on text

@@
 In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S.
-National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`; Schimel et al.
+National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`, Schimel et al.
 2007].
@@
 They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
-*(can we say that there is more loss of leaves higher in the profile during cold/ drought dormant seasons? )*
+In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size claasses (Condit, Meakem, ?any work on this in temperate forests??).
 In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.
@@
 ![**Figure `r fig_NEON_vertical`.
-Vertical gradients in the biophysical environment for 5/6 sites in the National Ecological Observatory Network (NEON)**.
+Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**.
 Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-All variables are normalized relative to the height of the top of the canopy (**DETAILS**) and the conditions at the point of lowest measurement.
+Sites include a tropical broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
 ([issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)) ](NEON_height_profiles/figures/profile_all.png)

### 2021-04-06 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2014*].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (REFS; IPBES report), including altered rates of photosynthesis and respiration [REFS], foliar damage during heat waves [*O'Sullivan et al.
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities], foliar damage during heat waves [*O'Sullivan et al.
 2017*], and reduced growth and elevated mortality during drought [REFS].
@@
 They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
-In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size claasses (Condit, Meakem, ?any work on this in temperate forests??).
+In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
 In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.
@@
 `r fig_NEON_vertical`; @koike_leaf_2001].**
-Generally, upper canopies in dense forests are exposed to greater PAR values and limit light penetration to canopy interiors, lower canopy layers, and the understory.
-Shaded regions experience lower values of PAR with different light quality [in ranges of blue, orange and red].
-The light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig.
+Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor.
+Foliage density mediates light regime, where denser upper canopies exposed to greater PAR values limit light to canopy interiors, lower canopy layers, and the understory.
+This light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig.
 `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b].
 
-In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy [to X?] is a better proxy for light environment than height (Parker 1995).
-Lower canopy light environments are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]
+The amount of light reaching lower canopy, understories and forest floors are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
+Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
+Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute to the light environment of shaded forest regions, in the form of sunflecks.
+Shaded canopy layers depend on sunflecks where receive 10-80% of their photon flux density for photosynthesis in short bursts within minutes of light exposure [@waySunflecksTreesForests2012].
+Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy in inner canopy is a better proxy for light environment than height (Parker 1995).
 
-Tropical forests with densely packed overstory trees limit light reaching subsequent canopy layers and forest understories.
-Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
-However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks.
-Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [@waySunflecksTreesForests2012]
-
-This difference in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity.
-Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010; @sackHowStrongIntracanopy2006; @niinemetsControversyTraitsConferring2006; Osnas et al.
-2018].
-
--	Sunflecks: Leaky et al.
-2003--from Marielle, "This is another, higher resolution axis of the variability of microenvironments that is probably quite important to address.
-(Also see the papers by Robin Chazdon on temporal variability of light environments).
-Maybe we will want to add these sentences (above) and something about sunflecks to a separate section that specifically addresses sources of variation in forest structure & microenvironments (e.g., vertical, horizontal, seasonal, diurnal, and with disturbance)?"
+These light regimes shape foliage at a vertical gradient, which in turn shapes the microenvironment.
+At a foliar level, resulting light-adaptation in the form of sun and shade leaf dichotomy also facilitates plant-adaptation to the light environment in the understory and overstory, ultimately contributing to shaping the microenvironment, forest vertical-profile's thermal sensitivity and their response to warming [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010; @sackHowStrongIntracanopy2006; @niinemetsControversyTraitsConferring2006]
 
@@
 Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
+Increase in tree height also contributes to decrease in understory wind-speed [@CanopyStructureTopography]
 Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
@@
 **Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
-Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019].
-Intact tropical forests have also been show to exhibit lower peak daytime $T_{air}$ than neighboring tree crop agricultural plantations (S.
+Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019].
+
+The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019].
+Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower daytime maximum temperature and vapor pressure deficit ($VPD$), and warmer higher nighttime minimum temperature in the understory, respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
+2021, @davis_microclimatic_2019, @zellweger_seasonal_2019].
+In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$ [S.
 R.
-Hardwick et al., The relationship between leaf area index and microclimate in tropical forest and oil palm plantation: Forest disturbance drives changes in microclimate.
-Agric.
-Meteorol.
-201, 187–195 (2015)).
-Even though canopy cover buffers maximum air temperatures in the understory, inevitable increases in understory temperatures continue to be observed as a result of global temperature increases.
-However, the buffering by canopy cover smooths this warming in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]
+Hardwick et al.
+2015, Junker et al.
+2018].
+Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
+These pine forests have also been observed to maintain warmer nighttime understory temperatures than above canopy, due to radiative heat loss from the soil, caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a].
+Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer.
+As a result, taller trees up to 20m in an an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature.
+However, beyond a certain height, temperature and VPD offset plateaus [Junker et al.
+2018].
+At a species level, broadleaf trees with higher shade casting ability also contribute to greater temperature buffering [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a].
 
-The capacity to moderate diurnal temperature fluctuations differs among forests with varying canopy structure, crown density, and seasonality.
-Tropical, temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures relative to pine and boreal forests.
-Increases in crown density and leaf area index decrease light and higher maximum temperature permeability into the forest understory, thus maintaining lower maximum daytime temperatures.
-Comparatively open forests such as pine forests with sparse stands (stems?) resulting in decreased buffering capacity in the understory and air temperatures more similar to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
-This relationship is reversed during the night, however, with open-canopy pine forests maintaining warmer understory temperatures than above canopy [@vonarxSpatiotemporalEffectsForest2012a]
-
-During the wet season, maximum air temperatures have been shown to be higher above forest canopies than below in tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al.
-2017].
-Similarly, in these studies, maximum temperatures were higher while humidity was lower in gaps compared to understory regions (both at 1.5 m above the ground, see their Fig.
-4 - again illustrating buffering effect of the canopy).
-However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
+Seasonally, the capacity to moderate buffering is greater with water availability.
+During the wet season, tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al.
+2017], have all shown to offset temperatures positively in the understory, with higher maximum air temperatures and relative humidity in above forest canopies and gaps than in below canopies and understories.
+During the dry season of a semi-deciduous tropical forest in Panama, however, maximum temperatures were similar in upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
 Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
 
-Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
-Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point?
-(e.g., greater radiative heat loss in exposed areas resulting in lower Tmin))
+Under colder conditions (?), minimum T_air is also buffered by forest canopies, increasing negetive temperature offset in the understory.
+Across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019] higher minimum temperatures under forest canopies have been observed relative to nearby clearings.
+Greater radiative heat loss in exposed areas results in lower T_min.
 However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
+
+Thus, we see that buffering by canopy cover smooths the effect temperature extreme and fluctuation in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory].
+However, as a result of global rise in temperatures, inevitable increases in understory temperatures are observed.
+[@bertrandCommentForestMicroclimate2020]
 
@@
 
-Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates.
-In three European forest types (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
+Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates, which tends to decrease with increasing maximum temperature.
+In three European forests (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
 Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
@@
 
-*ALL OF THE FOLLOWING CONTENT SHOULD BE WORKED INTO RELEVANT PARAGRAPHS EARLIER IN THIS SECTION:*
+### Geographic and climatic factors
 
-The strength of this buffering varies across forests, influenced by both forest characteristics and the biophysical environment.
-First, buffering increases with canopy cover and water availability.
-* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
-* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
-* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015] 
-The omega factor (McNaughton and Jarvis, 1983; Jarvis & McNaughton 1986) describes the degree of aerodynamic coupling between the vegetation and the atmospheric boundary layer; the lower the value of omega, the greater the coupling.
-"I wonder whether it would be useful to introduce such 'formal' terms used to describe what is essentially the same as buffering effect of canopies.
-I'd have to read up some more to see if there are papers showing what sort of canopy characteristics influence omega.
-
-https://doi.org/10.1016/S0065-2504(08)60119-1"-M.Slot
-
-*Presumably, buffering would also be affected by canopy roughness, which affects turbulent air flow and the canopy boundary layer.*
-Taller trees don't necessarily increase buffering [@zellweger_seasonal_2019].
-"They found that taller forests (which coincided with forests of higher LAI) had lower mean and max values of VPD and temperature" [Jucker, T., Hardwick, S.R., Both, S., Elias, D.M., Ewers, R.M., Milodowski, D.T., Swinfield, T.
-and Coomes, D.A., 2018.
-Canopy structure and topography jointly constrain the microclimate of human-modified tropical landscapes.
-Global change biology, 24(11), pp.5243-5258]
-.See Jucker et al. 2018 - max height was highly correlated to both LAI and surface roughness
-
-Species that have higher shade casting ability such as broadleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
+*Topography and altitude contribute to the heterogeneity of forest light environment, and thus shape microclimates.
+Trees in higher altitudes have been observed to adapt their branching to allow for light to pass through subsequent canopy layers.
+*Canopy structure and topography jointly constrain the microclimate of human-modified tropical landscapes.
+Global change biology, 24(11), pp.5243-5258] .See *Jucker et al.
+2018 - max height was highly correlated to both LAI and surface roughness
 The strength of buffering also varies with respect to geographic and climatic factors.

### 2021-04-06 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig.
-`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b].
+`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b].
 
@@
 2021, @davis_microclimatic_2019, @zellweger_seasonal_2019].
-In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$ [S.
+In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with canopy cover [S.
 R.
@@
 2018].
-Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a].
+Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
 These pine forests have also been observed to maintain warmer nighttime understory temperatures than above canopy, due to radiative heat loss from the soil, caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a].
@@
 
-*Topography and altitude contribute to the heterogeneity of forest light environment, and thus shape microclimates.
-Trees in higher altitudes have been observed to adapt their branching to allow for light to pass through subsequent canopy layers.
-*Canopy structure and topography jointly constrain the microclimate of human-modified tropical landscapes.
-Global change biology, 24(11), pp.5243-5258] .See *Jucker et al.
-2018 - max height was highly correlated to both LAI and surface roughness
+With higher altitude and topography, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions.
+*(do higher altitude forests have lower buffering capacity?
+due to open canopies?)* In a European silver birch forest, trees in higher altitude maintained open canopies with allowing more light through subsequent canopy layers, resulting in similar leaf traits across canopy positions, than in lower altitude trees [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
+Increased light permeability in higher altitude European sparse pine canopies, has also shown to increase understory light conditions along with higher maximum temperatures than lower altitude canopies, thus potentially having implications for seedling recruitment as temperatures rise [@vonarxSpatiotemporalEffectsForest2012a].
+Vertical climate gradients are observed to be steeper within forests than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]*what does this mean?
+can combine above sentences into one(N.V note)* Tree height in combination with topography modifies microclimate conditions.
+In a Boreneo forest, greater canopy height had a stronger influence on buffering temperature on east facing, wind-exposed slopes than on west-facing slopes.
+
 The strength of buffering also varies with respect to geographic and climatic factors.
-* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
-* [@davis_microclimatic_2019]--Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperatures [@davis_microclimatic_2019]
-However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]
+Distance to coast, topographic position, elevation [@zellweger_seasonal_2019].
+Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperature extremes [@davis_microclimatic_2019]
 
@@
 While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
+In open-canopy pine forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially higher than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021, @martin_boundary_1999]
 Under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
-For instance, during the dry season in a tropical moist forest in Panama...
-(Rey-Sánchez et al.
-2016).
-Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al.
-2016).
+For instance, during the dry season in a semi-decidudous tropical moist forest in Panama [@rey-sanchez_spatial_2016].
+Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves [@rey-sanchez_spatial_2016].
 At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.
@@
 **Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
 A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
@@
 -- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
+- Beyond a certain height, height becomes more important than light in determing traits [Ambrose, Sillet, Cavaleri]
 
@@
 **Within a species, leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody ~~and herbaceous~~ species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. 
+After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. 
 A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation intercepting the leaf surface.
@@
 
-- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
+- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015), however stomatal conductance limitation is also greater, and earlier with canopy height [Sanches et al. 2009] 
 Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
+-understory leaves open their stomata wider than upper canopy leaves [Kosugi et al. 2021]
 
@@
 Interaction between degrees of shade x ontogeny x height with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
+Leaf traits vary more on the basis on light environment than ontogenetically [REF]
 

### 2021-04-08 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
-In open-canopy pine forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially higher than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021, @martin_boundary_1999]
+In open-canopy needle-leaf forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially higher than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021, @martin_boundary_1999, @zweifelMiddayStomatalClosure2002]
 Under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
@@
 Interaction between degrees of shade x ontogeny x height with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
-Leaf traits vary more on the basis on light environment than ontogenetically [REF]
+Leaf traits vary more on the basis on light environment than ontogenetically [@liuPlantSizeEffects2010]
 

### 2021-04-08 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2014*].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities], foliar damage during heat waves [*O'Sullivan et al.
-2017*], and reduced growth and elevated mortality during drought [REFS].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
 Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
@@
 Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute to the light environment of shaded forest regions, in the form of sunflecks.
-Shaded canopy layers depend on sunflecks where receive 10-80% of their photon flux density for photosynthesis in short bursts within minutes of light exposure [@waySunflecksTreesForests2012].
-Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy in inner canopy is a better proxy for light environment than height (Parker 1995).
+Shaded canopy layers depend on sunflecks, where they receive 10-80% of their photon flux density for photosynthesis in short bursts within minutes of light exposure [@waySunflecksTreesForests2012].
+Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
 
 These light regimes shape foliage at a vertical gradient, which in turn shapes the microenvironment.
-At a foliar level, resulting light-adaptation in the form of sun and shade leaf dichotomy also facilitates plant-adaptation to the light environment in the understory and overstory, ultimately contributing to shaping the microenvironment, forest vertical-profile's thermal sensitivity and their response to warming [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010; @sackHowStrongIntracanopy2006; @niinemetsControversyTraitsConferring2006]
+At a foliar level, resulting light-adaptation in the form of sun and shade leaf dichotomy also facilitates plant-adaptation to the light environment in the understory and the overstory, ultimately contributing to shaping the microenvironment, forest vertical-profile's thermal sensitivity and their response to warming  [@fauset_differences_2018; @millerOnlySunlitLeaves2021; @michaletzPlantThermoregulationEnergetics2015; *Defrenne et al.
+2021*]
 
 ### Wind speeds 
-**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [**REFS**].**
+**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
 Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
-Increase in tree height also contributes to decrease in understory wind-speed [@CanopyStructureTopography]
+Understory wind-speed is also substantially reduced with increased canopy height [@juckerCanopyStructureTopography2018]
 Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
@@
 The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019].
-Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower daytime maximum temperature and vapor pressure deficit ($VPD$), and warmer higher nighttime minimum temperature in the understory, respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
+Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
 2021, @davis_microclimatic_2019, @zellweger_seasonal_2019].
@@
 Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
-These pine forests have also been observed to maintain warmer nighttime understory temperatures than above canopy, due to radiative heat loss from the soil, caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a].
+During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to radiative heat loss from the soil caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a].
 Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer.
 As a result, taller trees up to 20m in an an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature.
-However, beyond a certain height, temperature and VPD offset plateaus [Junker et al.
-2018].
+However, beyond a certain height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018].
 At a species level, broadleaf trees with higher shade casting ability also contribute to greater temperature buffering [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a].
@@
 During the wet season, tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al.
-2017], have all shown to offset temperatures positively in the understory, with higher maximum air temperatures and relative humidity in above forest canopies and gaps than in below canopies and understories.
-During the dry season of a semi-deciduous tropical forest in Panama, however, maximum temperatures were similar in upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
+2017], have all shown to offset temperatures positively in the understory, with higher maximum air temperatures and relative humidity above forest canopies and gaps than in below canopies and understories.
+During the dry season of a semi-deciduous tropical forest in Panama, however, maximum temperatures were similar in the upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
 Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
 
-Under colder conditions (?), minimum T_air is also buffered by forest canopies, increasing negetive temperature offset in the understory.
-Across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019] higher minimum temperatures under forest canopies have been observed relative to nearby clearings.
-Greater radiative heat loss in exposed areas results in lower T_min.
+Under colder conditions (?), minimum $T_{air}$ [$T_{min}$] is also buffered by forest canopies, increasing negetive temperature offset in the understory.
+Across Europe [@zellweger_seasonal_2019; Defrenne et al.
+2021] and in the northwestern United States [@davis_microclimatic_2019] higher minimum temperatures under forest canopies have been observed relative to nearby clearings.
+Greater radiative heat loss in exposed areas results in lower $T_{min}$.
 However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
 
-Thus, we see that buffering by canopy cover smooths the effect temperature extreme and fluctuation in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory].
-However, as a result of global rise in temperatures, inevitable increases in understory temperatures are observed.
+Thus, we see that buffering by canopy cover smooths the affect temperature extremes and fluctuation in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory; Defrenne et al.
+2021].
+However, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed.
 [@bertrandCommentForestMicroclimate2020]
@@
 *(do higher altitude forests have lower buffering capacity?
-due to open canopies?)* In a European silver birch forest, trees in higher altitude maintained open canopies with allowing more light through subsequent canopy layers, resulting in similar leaf traits across canopy positions, than in lower altitude trees [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
+due to open canopies?)* In a European silver birch forest, trees in higher altitude maintain open canopies allowing more light through subsequent canopy layers, resulting in similar leaf traits across canopy positions, than in lower altitude trees [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
 Increased light permeability in higher altitude European sparse pine canopies, has also shown to increase understory light conditions along with higher maximum temperatures than lower altitude canopies, thus potentially having implications for seedling recruitment as temperatures rise [@vonarxSpatiotemporalEffectsForest2012a].
@@
 
-The strength of buffering also varies with respect to geographic and climatic factors.
-Distance to coast, topographic position, elevation [@zellweger_seasonal_2019].
-Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperature extremes [@davis_microclimatic_2019]
+The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
+Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperature extremes [@davis_microclimatic_2019; **REF**]
 
@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus XXX*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
 [ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/leaf_energy_balance.png)
@@
 Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
-[ref]
+[*REF*]
 
@@
 **In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** 
-*Smaller leaves have higher $g_{a}$ to heat transfer, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011].
-(verify this statement and reference)*
+*Smaller leaves have higher $g_{a}$ to heat transfer, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].  
 While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018]. 
@@
 2008].
-In a temperate Acer saccharum forest, ~~within-canopy Acer saccharum exhibited greater~~LMA and $N_{area}$ were greater with height early in the growing season as a result of hydrostatic constraints on leaf morphology.
+In a temperate Acer saccharum forest, LMA and $N_{area}$ were greater with height early in the growing season as a result of hydrostatic constraints on leaf morphology.
 Later in the growing season, LMA and $N_{area}$ continued to increase in the upper canopy, but due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].
@@
 -- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
-- Beyond a certain height, height becomes more important than light in determing traits [Ambrose, Sillet, Cavaleri]
+- Beyond a certain height, light becomes less important than height in determing traits [Ambrose, Sillet, Cavaleri]
 
@@
 Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation.
-Leaf lobes are deeper and leaf elongation and serration are more pronounced in sun exposed canopies than in the shaded understory~~, as well as leaf elongation and serration are more pronounced in sun exposed canopies~~ [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
+Leaf lobes are deeper and leaf elongation and serration are more pronounced in sun exposed canopies than in the shaded understory [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
 

### 2021-04-09 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-With higher altitude and topography, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions.
-*(do higher altitude forests have lower buffering capacity?
-due to open canopies?)* In a European silver birch forest, trees in higher altitude maintain open canopies allowing more light through subsequent canopy layers, resulting in similar leaf traits across canopy positions, than in lower altitude trees [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
-Increased light permeability in higher altitude European sparse pine canopies, has also shown to increase understory light conditions along with higher maximum temperatures than lower altitude canopies, thus potentially having implications for seedling recruitment as temperatures rise [@vonarxSpatiotemporalEffectsForest2012a].
-Vertical climate gradients are observed to be steeper within forests than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]*what does this mean?
-can combine above sentences into one(N.V note)* Tree height in combination with topography modifies microclimate conditions.
-In a Boreneo forest, greater canopy height had a stronger influence on buffering temperature on east facing, wind-exposed slopes than on west-facing slopes.
-
-The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
+With higher altitude, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions.
+In European forests, buffering capacity is stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a].
+Trees in European high altitude forests have been shown to maintain sparser stands [@vonarxSpatiotemporalEffectsForest2012a] and open canopies with greater light interception through canopy layers, and therefore to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
+While increase in tree height (in closed-forests) increases buffering in the understory [@juckerCanopyStructureTopography2018], in timberline and temperate mountain forests in China, tree height decreased with altitude, where as tree growth sensitivity (based on tree-rings) to climate change increased.
+Decrease in temperature altitudinally associated with increased relative humidity [Liang et al.
+2019] and understory light interception might have implications to photosynthetic functioning [Sellin et al.
+2013], and high altitude pine forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
+*Vertical climate gradients are observed to be steeper within forests than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a] what does this mean?
+can combine above sentences into one(N.V note)* The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
 Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperature extremes [@davis_microclimatic_2019; **REF**]

### 2021-04-09 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 In European forests, buffering capacity is stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a].
-Trees in European high altitude forests have been shown to maintain sparser stands [@vonarxSpatiotemporalEffectsForest2012a] and open canopies with greater light interception through canopy layers, and therefore to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
-While increase in tree height (in closed-forests) increases buffering in the understory [@juckerCanopyStructureTopography2018], in timberline and temperate mountain forests in China, tree height decreased with altitude, where as tree growth sensitivity (based on tree-rings) to climate change increased.
-Decrease in temperature altitudinally associated with increased relative humidity [Liang et al.
-2019] and understory light interception might have implications to photosynthetic functioning [Sellin et al.
-2013], and high altitude pine forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
-*Vertical climate gradients are observed to be steeper within forests than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a] what does this mean?
-can combine above sentences into one(N.V note)* The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
-Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperature extremes [@davis_microclimatic_2019; **REF**]
+Trees in European high altitude forests have shown to maintain sparser stands [@vonarxSpatiotemporalEffectsForest2012a] and open canopies with greater light interception through canopy layers, and to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
+While increase in tree height (in closed-forests) increases buffering in the understory [@juckerCanopyStructureTopography2018], in timberline and temperate humid-mountain forests in China, tree height decreased with
+altitude, whereas tree growth sensitivity (based on tree-rings) to climate change increased [@liangForestTypeHeight2019].
+Less capacity to buffer climatic conditions in open, high altitude forests might thus have implications to forest regeneration [@vonarxSpatiotemporalEffectsForest2012a, @fauset_differences_2018].The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
+Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davis_microclimatic_2019; @macekMaximumAirTemperature2019]
 

### 2021-04-09 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Hardwick et al.
-2015, Junker et al.
-2018].
+2015; @juckerCanopyStructureTopography2018].
 Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
@@
 
-Thus, we see that buffering by canopy cover smooths the affect temperature extremes and fluctuation in the understory, thereby exhibiting an understory temperature lag [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory; Defrenne et al.
-2021].
-However, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed.
+Thus, we see that buffering by canopy cover smooths the affect temperature extremes and fluctuation in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory; Defrenne et al.
+2021], however, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed.
 [@bertrandCommentForestMicroclimate2020]

### 2021-04-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The amount of light reaching lower canopy, understories and forest floors are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
-Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
+Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]and needle-leaf forests receive 17-34% of incoming sunlight [@baldocchiSeasonalVariationEnergy1997].
 Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute to the light environment of shaded forest regions, in the form of sunflecks.

### 2021-04-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-04-09 vertical-thermal-review.Rmd
@@
 2014*].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
 Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
@@
 The amount of light reaching lower canopy, understories and forest floors are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
-Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006].
+Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]and needle-leaf forest floors receive 17-34% of incoming sunlight [@baldocchiSeasonalVariationEnergy1997].
 Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute to the light environment of shaded forest regions, in the form of sunflecks.
-Shaded canopy layers depend on sunflecks, where they receive 10-80% of their photon flux density for photosynthesis in short bursts within minutes of light exposure [@waySunflecksTreesForests2012].
+Shaded canopy layers rely on sunflecks for photosynthesis, where they receive 10-80% of their photon flux density in short bursts within minutes of light exposure [@waySunflecksTreesForests2012].
 Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
@@
 These light regimes shape foliage at a vertical gradient, which in turn shapes the microenvironment.
-At a foliar level, resulting light-adaptation in the form of sun and shade leaf dichotomy also facilitates plant-adaptation to the light environment in the understory and the overstory, ultimately contributing to shaping the microenvironment, forest vertical-profile's thermal sensitivity and their response to warming  [@fauset_differences_2018; @millerOnlySunlitLeaves2021; @michaletzPlantThermoregulationEnergetics2015; *Defrenne et al.
+At a foliar level, resulting light-adaptation in the form of sun and shade leaf dichotomy also facilitates plant-adaptation to the light environment in the understory and the overstory, ultimately contributing to shaping the microenvironment, forest vertical-profile's thermal sensitivity and their response to warming  [@fauset_differences_2018; @millerOnlySunlitLeaves2021; @michaletzPlantThermoregulationEnergetics2015; *De Frenne et al.
 2021*]
@@
 Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
-Understory wind-speed is also substantially reduced with increased canopy height [@juckerCanopyStructureTopography2018]
+Understory wind-speed is also substantially reduced with increased canopy height [@juckerCanopyStructureTopography2018].
 Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
@@
 Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
-2021, @davis_microclimatic_2019, @zellweger_seasonal_2019].
+2021; @davis_microclimatic_2019; @zellweger_seasonal_2019].
 In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with canopy cover [S.
@@
 2015; @juckerCanopyStructureTopography2018].
-Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
+Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies.
+They maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
 During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to radiative heat loss from the soil caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a].
@@
 During the wet season, tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al.
-2017], have all shown to offset temperatures positively in the understory, with higher maximum air temperatures and relative humidity above forest canopies and gaps than in below canopies and understories.
+2017], have all shown to offset temperatures positively in the understory, with lower maximum air temperatures and relative humidity below canopies and understories than above forest canopies and gaps.
 During the dry season of a semi-deciduous tropical forest in Panama, however, maximum temperatures were similar in the upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
@@
 Thus, we see that buffering by canopy cover smooths the affect temperature extremes and fluctuation in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory; Defrenne et al.
-2021], however, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed.
+2021],though, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed.
 [@bertrandCommentForestMicroclimate2020]
@@
 altitude, whereas tree growth sensitivity (based on tree-rings) to climate change increased [@liangForestTypeHeight2019].
-Less capacity to buffer climatic conditions in open, high altitude forests might thus have implications to forest regeneration [@vonarxSpatiotemporalEffectsForest2012a, @fauset_differences_2018].The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
+Less capacity to buffer climatic conditions in open, high altitude forests might thus have implications to forest regeneration [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
 Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davis_microclimatic_2019; @macekMaximumAirTemperature2019]
@@
 Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
-The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of radiation input versus heat lost to the environment [Fig.
+The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy as radiation input versus heat lost to the environment [Fig.
 `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
@@
 While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
-In open-canopy needle-leaf forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially higher than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021, @martin_boundary_1999, @zweifelMiddayStomatalClosure2002]
-Under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
+In open-canopy needle-leaf forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially warmer than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]
+In closed-forests, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
 For instance, during the dry season in a semi-decidudous tropical moist forest in Panama [@rey-sanchez_spatial_2016].
@@
 - In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
-Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
+Additionally, with increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
 - In a tropical wet and moist forests of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]

### 2021-04-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2015; @juckerCanopyStructureTopography2018].
-Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies.
-They maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
-During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to radiative heat loss from the soil caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a].
+Pine or needle-leaf forests, on the other hand, that are open-forests with sparse stands have lower buffering capacity, greater vertical air mixing and greater light permeability through canopies.
+They maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999;].
+During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
 Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer.
@@
 However, beyond a certain height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018].
-At a species level, broadleaf trees with higher shade casting ability also contribute to greater temperature buffering [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a].
 
@@
 During the wet season, tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al.
-2017], have all shown to offset temperatures positively in the understory, with lower maximum air temperatures and relative humidity below canopies and understories than above forest canopies and gaps.
+2017], have all shown to offset temperatures positively in the understory, with lower maximum air temperatures and higher relative humidity below canopies and understories than above forest canopies and gaps.
 During the dry season of a semi-deciduous tropical forest in Panama, however, maximum temperatures were similar in the upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
@@
 Thus, we see that buffering by canopy cover smooths the affect temperature extremes and fluctuation in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory; Defrenne et al.
-2021],though, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed.
-[@bertrandCommentForestMicroclimate2020]
+2021],though, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed [@bertrandCommentForestMicroclimate2020].
 
@@
 Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates, which tends to decrease with increasing maximum temperature.
-In three European forests (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
-Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
+In three European forests (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.RH moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
 
@@
 With higher altitude, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions.
-In European forests, buffering capacity is stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a].
-Trees in European high altitude forests have shown to maintain sparser stands [@vonarxSpatiotemporalEffectsForest2012a] and open canopies with greater light interception through canopy layers, and to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
-While increase in tree height (in closed-forests) increases buffering in the understory [@juckerCanopyStructureTopography2018], in timberline and temperate humid-mountain forests in China, tree height decreased with
-altitude, whereas tree growth sensitivity (based on tree-rings) to climate change increased [@liangForestTypeHeight2019].
-Less capacity to buffer climatic conditions in open, high altitude forests might thus have implications to forest regeneration [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
+In European forests, buffering capacity was stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a].
+In a few European high altitude forests, trees are sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] and have open canopies with greater light interception through canopy layers, and to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
+While increase in tree height (in closed-forests) typically increases buffering in the understory [@juckerCanopyStructureTopography2018], in timberline and temperate humid-mountain forests in China, tree height decreased with altitude, and tree growth sensitivity (based on tree-rings) to climate change increased [@liangForestTypeHeight2019].
+Less capacity to buffer climatic conditions in open, high altitude forests might thus have implications to forest regeneration [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].
+
+The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
 Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davis_microclimatic_2019; @macekMaximumAirTemperature2019]
@@
 
-**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+**Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
+Leaves try to maintain metabolic optima in order to maximize carbon assimilation, in a given circumstance, through the adaptation of their leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+$T_{leaf}$ and leaf thermoregulation, thus, is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 `r fig_leaf_T`).
-Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]
-The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy as radiation input versus heat lost to the environment [Fig.
+The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost to the environment [Fig.
 `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
@@
 **(paragraph on radiation as a driver)**
-Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
+Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$ [Fauset et al.
+2018]
 
@@
 In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
-Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy?
-[*REF*]
+Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata.
+This tends to be greater in the understory than in the canopy [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]
 
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
+Leaf temperature, is also influenced by stomtal conductance and leaf water content [*Koch et al.
+1994*].
 Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*.
 Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
-However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO~2~, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018].
-Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.
+However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.
+This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [*Koch et al.
+1994*; @fauset_differences_2018].
+On the other, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
 
@@
 `r fig_leaf_T`).**
+Vertical gradient in leaf processes and leaf temperature is influenced by vertical microclimate profile.
 While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
-In open-canopy needle-leaf forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially warmer than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]
-In closed-forests, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
-For instance, during the dry season in a semi-decidudous tropical moist forest in Panama [@rey-sanchez_spatial_2016].
-Sun leaves can also be cooled relative to $T_{air}$ more than shade leaves [@rey-sanchez_spatial_2016].
+
+In open forests, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing, heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+Greater heat stress and $T_{air}$ in the lower canopy is also observed in an Australian open desert forest [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground
+[@hadleyInfluenceKrummholzMat1987].
+Similarly, during the dry season in a semi-decidudous tropical moist forest in Panama, sun leaves can also be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
+Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
+`r fig_NEON_vertical`).
+
 At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.
@@
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
--Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads.
-However, leaf temperatures themselves were moderated by air temperature.
-[@fauset_differences_2018].
+In closed-canopy forests, large leaf to air temperature differences and high leaf temperatures are found in upper canopies of tropical forests such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest in Panama in wet season [@rey-sanchez_spatial_2016], and in a temperate forest [@mauTemperateTropicalForest2018a].
+
 - In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
@@
 
-**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
+**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, for optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
 After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].

### 2021-04-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 In closed-canopy forests, large leaf to air temperature differences and high leaf temperatures are found in upper canopies of tropical forests such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest in Panama in wet season [@rey-sanchez_spatial_2016], and in a temperate forest [@mauTemperateTropicalForest2018a].
-
 - In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
@@
 
--	(Slot et al. 2019) and refs therein
+## Leaf traits
 
-- Elsa: I came across these two studies that could be relevent in the evapotranspiration/temperature feedback section hinted at above: 
-- Staudt, K., Serafimovich, A., Siebicke, L., Pyles, R.D.
-and Falge, E., 2011.
-Vertical structure of evapotranspiration at a forest site (a case study).
-Agricultural and Forest Meteorology, 151(6), pp.709-729.
-- Sellin, A. and Lubenets, K., 2010. Variation of transpiration within a canopy of silver birch: effect of canopy position and daily versus nightly water loss. Ecohydrology, 3(4), pp.467-477.
-
-**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, for optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation [R:FR] intercepting the leaf surface.
-This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013].
-Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].
-
-## Leaf traits
+**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] in accordance with the intensity and availability of light for optimizing C uptake, water uptake, heat and metabolism across forest vertical strata, thus show different degrees foliar plasticity [@rozendaalPlasticityLeafTraits2006]**
 
@@
 
-**Within a species, leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
 After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. 
@@
 
+Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+
 *(Krista: the above paragraaph is really interesting, and I like the idea of starting off the section by mentioning what controls the formation of leaf traits.
@@
 I'd broaden to include the full microenviroment (without increasing length)*)
-
-Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 

### 2021-04-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Leaves try to maintain metabolic optima in order to maximize carbon assimilation, in a given circumstance, through the adaptation of their leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-$T_{leaf}$ and leaf thermoregulation, thus, is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+$T_{leaf}$ and leaf thermoregulation, thus, is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 `r fig_leaf_T`).
@@
 **(paragraph on radiation as a driver)**
-Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$ [Fauset et al.
+Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$.
+On the other hand, under conditions constant radiation, increase in $T_{air}$ can induce evaporative cooliing [Drake et al.
+2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al.
 2018]
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
-Leaf temperature, is also influenced by stomtal conductance and leaf water content [*Koch et al.
+Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance and leaf water content [@fauset_differences_2018; *Koch et al.
 1994*].
@@
 `r fig_leaf_T`).**
-Vertical gradient in leaf processes and leaf temperature is influenced by vertical microclimate profile.
+Leaf processes and leaf temperature at a vertical gradient are influenced by vertical microclimate conditions.
 While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
 
-In open forests, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing, heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+Particularly, in open forests, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing, heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
 Greater heat stress and $T_{air}$ in the lower canopy is also observed in an Australian open desert forest [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground
@@
 
-**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] in accordance with the intensity and availability of light for optimizing C uptake, water uptake, heat and metabolism across forest vertical strata, thus show different degrees foliar plasticity [@rozendaalPlasticityLeafTraits2006]**
+**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light, in order to optimize C uptake, water uptake, heat and metabolism across forest vertical strata, thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015]**
 
@@
 2009, 2010*).
+Carbon isotope composition has also been shown to increase with height, independent of light [@cobleLightAcclimationOptimizes2015]
+
+Height-based leaf trait variation in two tall temperate species, *Quercus alba* and *Quercus velutina*, showed decreases in lobation with increased height, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
+However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus species.
+
 Disentangling the influences of hydraulic constraints, light, and other biophysical drivers that vary with height (Fig.
@@
 
+- Beyond a certain height, light becomes less important than height in determing traits [Ambrose, Sillet, Cavaleri]
+
+- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues.
+I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
+
 *(The following notes need to be condensed/ worked into the paragraph above or discussion of specific traits.
@@
 The goal is to answer the question, "Do we expect light gradients to be equivalent to height gradients, or is there variation attributable to height that is not captured by light gradients?")*
-
-Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from changes in light availability.
-However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area-based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al.
-2008].
-In a temperate Acer saccharum forest, LMA and $N_{area}$ were greater with height early in the growing season as a result of hydrostatic constraints on leaf morphology.
-Later in the growing season, LMA and $N_{area}$ continued to increase in the upper canopy, but due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].
-In addition, carbon isotope composition has also been shown to increase with height, independent of light [@cobleLightAcclimationOptimizes2015]
-
-Height-based leaf trait variation in two tall temperate species, *Quercus alba* and *Quercus velutina*, showed decreases in lobation with increased height, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
-However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus species.
-
-* Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
-- (Cavaleri et al. 2008) 
-- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues.
-I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
-- Elsa: see also H.
-Poorter, U.
-Niinemets, L.
-Poorter, I.
-J.
-Wright, R.
-Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
-New Phytol.
-182, 565–588 (2009).
--- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
-- Beyond a certain height, light becomes less important than height in determing traits [Ambrose, Sillet, Cavaleri]
 
@@
 However, when it comes to understanding and modeling forest ecosystem function, the relevant scale of variation is interspecific (e.g., *Lamour, Serbin in prep*).
-
-Several studies point towards differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 
@@
 *(KAT: a lot of this content should probably move up the the intro to this section)*
+
+- Elsa: see also H.
+Poorter, U.
+Niinemets, L.
+Poorter, I.
+J.
+Wright, R.
+Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
+New Phytol.
+182, 565–588 (2009).
+-- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
 

### 2021-04-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute to the light environment of shaded forest regions, in the form of sunflecks.
-Shaded canopy layers rely on sunflecks for photosynthesis, where they receive 10-80% of their photon flux density in short bursts within minutes of light exposure [@waySunflecksTreesForests2012].
+Shaded canopy layers rely on sunflecks for photosynthesis.
+In some cases, they receive 10-80% of their photon flux density or 1/3 of their carbon gain from short light bursts within minutes of exposure [@waySunflecksTreesForests2012].
 Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
@@
 2021; @davis_microclimatic_2019; @zellweger_seasonal_2019].
-In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with canopy cover [S.
-R.
-Hardwick et al.
-2015; @juckerCanopyStructureTopography2018].
-Pine or needle-leaf forests, on the other hand, that are open-forests with sparse stands have lower buffering capacity, greater vertical air mixing and greater light permeability through canopies.
-They maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999;].
-During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
+In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+On the other hand, open-forests with sparse stands, such as needle-leaf forests, have lower buffering capacity, greater vertical air mixing and greater light permeability through canopy layers.
+They maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
+During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
 Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer.
-As a result, taller trees up to 20m in an an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature.
-However, beyond a certain height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018].
+As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature.
+However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].
 
@@
 In a few European high altitude forests, trees are sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] and have open canopies with greater light interception through canopy layers, and to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
-While increase in tree height (in closed-forests) typically increases buffering in the understory [@juckerCanopyStructureTopography2018], in timberline and temperate humid-mountain forests in China, tree height decreased with altitude, and tree growth sensitivity (based on tree-rings) to climate change increased [@liangForestTypeHeight2019].
-Less capacity to buffer climatic conditions in open, high altitude forests might thus have implications to forest regeneration [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].
+Less capacity to buffer climatic conditions in open, high altitude forests might have implications to forest regeneration and climate change [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].
+This is also observed in  timberline and temperate humid-mountain forests in China, where tree height decreased with altitude, and tree growth sensitivity (based on tree-rings) increased [@liangForestTypeHeight2019].
+Whereas, increase in tree height in closed-forests, typically increases buffering in the understory [@juckerCanopyStructureTopography2018].
 
@@
 **Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-Leaves try to maintain metabolic optima in order to maximize carbon assimilation, in a given circumstance, through the adaptation of their leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-$T_{leaf}$ and leaf thermoregulation, thus, is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+Leaves (try to) maintain metabolic optima in order to maximize carbon assimilation, in a given circumstance, through the adaptation of their leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].Thus, $T_{leaf}$ and leaf thermoregulation, is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 `r fig_leaf_T`).
@@
 Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$.
-On the other hand, under conditions constant radiation, increase in $T_{air}$ can induce evaporative cooliing [Drake et al.
+On the other hand, under conditions of constant radiation, increase in $T_{air}$ can induce evaporative cooling [Drake et al.
 2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al.
@@
 Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{leaf}$ and $T_{air}$. 
-In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
+In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
 Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata.
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
-Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance and leaf water content [@fauset_differences_2018; *Koch et al.
+Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance and leaf water content [@fauset_differences_2018; *Koch et al.
 1994*].
@@
 
-Particularly, in open forests, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing, heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
-Greater heat stress and $T_{air}$ in the lower canopy is also observed in an Australian open desert forest [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground
+Particularly, in open forests with sparse distribution of trees, such as in some needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing, heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+Greater heat stress and $T_{air}$ in the lower canopy is also observed in Australian desert trees [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground
 [@hadleyInfluenceKrummholzMat1987].
-Similarly, during the dry season in a semi-decidudous tropical moist forest in Panama, sun leaves can also be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
+Similarly, during the dry season in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
 Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
@@
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-In closed-canopy forests, large leaf to air temperature differences and high leaf temperatures are found in upper canopies of tropical forests such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest in Panama in wet season [@rey-sanchez_spatial_2016], and in a temperate forest [@mauTemperateTropicalForest2018a].
+In closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies of tropical forests such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and in a temperate forest [@mauTemperateTropicalForest2018a].
 - In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
@@
 
-**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light, in order to optimize C uptake, water uptake, heat and metabolism across forest vertical strata, thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015]**
+**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light, in order to optimize C uptake, water use, thermoregulation and metabolism across forest vertical strata, and thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015; *REFs*]**
 
@@
 
-Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.
-However, the majority of relevant research has focused on exposure gradients near ground level, with by far the most common study type being comparisons of sun and shade leaves. 
 Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
+Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.However, the majority of relevant research has focused on exposure gradients near ground level, with by far the most common study type being comparisons of sun and shade leaves.
 While light is the primary axis along which leaf traits vary, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig.
 `r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.
-In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*], such that gas exchange is strongly constrained by at the tops of the world's tallest trees (~100 m.
+
+In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m.
 tall *Sequoia sempervirens*; *Ambrose et al.
 2009, 2010*).
-Carbon isotope composition has also been shown to increase with height, independent of light [@cobleLightAcclimationOptimizes2015]
-
-Height-based leaf trait variation in two tall temperate species, *Quercus alba* and *Quercus velutina*, showed decreases in lobation with increased height, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020].
-However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus species.
+Under constant light conditions (photon flux density), taller trees (>35m) have lower canopy stomatal conductance compared to shorter trees  [@schaferEffectTreeHeight2000; REF].
+Adaptation to increasing height and hydraulic contraints show that leaf size [@chinLeafAcclimationLight2017] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) has shown to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a].
 
@@
 For the purposes of interpreting the results presented here, we note that patterns across light and height gradients age generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
-
-- Beyond a certain height, light becomes less important than height in determing traits [Ambrose, Sillet, Cavaleri]
 
@@
 
-After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. 
-A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation intercepting the leaf surface.
-This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013].
-Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].
+Leaf traits in woody species take approximately 30-60 days to stabilize to integrated light after formation [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+As a result, degrees of high to low irradiance gives rise to a dichotomy of sun and shade leaf traits.
+Typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves [Rozendaal et al.
+2006, Mathur et al.
+2018, *REFS*].
+Within-species, several studies point towards these common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient based on adaptation to primarily light conditions, and as well biophysical gradient [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 
-Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient (???), because LES theory was developed using mainly sun exposed (upper canopy) leaves.
+Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
+However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018) as well as geographically (Ishii et al., 2018).
 
@@
 
-As a result of different light regimes, a sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves
-[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].
-
  "Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well" -Martijn
@@
 2001 has more info about this.
-
-An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [@chenLeafEconomicsSpectrum2020a], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
-Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018].
 

### 2021-04-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]and needle-leaf forest floors receive 17-34% of incoming sunlight [@baldocchiSeasonalVariationEnergy1997].
-Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute to the light environment of shaded forest regions, in the form of sunflecks.
-Shaded canopy layers rely on sunflecks for photosynthesis.
-In some cases, they receive 10-80% of their photon flux density or 1/3 of their carbon gain from short light bursts within minutes of exposure [@waySunflecksTreesForests2012].
+Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute light in the form sunflecks--around 10-80% of of photon flux density-- to shaded forest regions that receive 1/3 of their carbon  during photosynthesis through these short light exposures [@waySunflecksTreesForests2012].
 Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
@@
 On the other hand, open-forests with sparse stands, such as needle-leaf forests, have lower buffering capacity, greater vertical air mixing and greater light permeability through canopy layers.
-They maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
+For this reason, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
 During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
 Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer.
-As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature.
+As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory.
 However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].
@@
 
-Under colder conditions (?), minimum $T_{air}$ [$T_{min}$] is also buffered by forest canopies, increasing negetive temperature offset in the understory.
+Under colder conditions, minimum $T_{air}$ [$T_{min}$] is also buffered by forest canopies, increasing negetive temperature offset in the understory.
 Across Europe [@zellweger_seasonal_2019; Defrenne et al.
@@
 With higher altitude, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions.
-In European forests, buffering capacity was stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a].
-In a few European high altitude forests, trees are sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] and have open canopies with greater light interception through canopy layers, and to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
+In Switzerland forests, buffering capacity was stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a], partly because trees here were sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] and in other high-altitude forests, trees have open canopies with greater light interception through canopy layers [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
 Less capacity to buffer climatic conditions in open, high altitude forests might have implications to forest regeneration and climate change [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].
 This is also observed in  timberline and temperate humid-mountain forests in China, where tree height decreased with altitude, and tree growth sensitivity (based on tree-rings) increased [@liangForestTypeHeight2019].
-Whereas, increase in tree height in closed-forests, typically increases buffering in the understory [@juckerCanopyStructureTopography2018].
+Whereas in closed-forests, increase in tree height, typically increases buffering in the understory [@juckerCanopyStructureTopography2018].
 
@@
 **Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-Leaves (try to) maintain metabolic optima in order to maximize carbon assimilation, in a given circumstance, through the adaptation of their leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].Thus, $T_{leaf}$ and leaf thermoregulation, is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+Leaves maximize carbon assimilation by maintaining metabolic optima, in a given circumstance, through the adaptation of their leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 `r fig_leaf_T`).
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
-Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance and leaf water content [@fauset_differences_2018; *Koch et al.
+Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; *Koch et al.
 1994*].
@@
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-In closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies of tropical forests such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and in a temperate forest [@mauTemperateTropicalForest2018a].
-- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height.
+Especially in closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies compared to lower canopies of tropical forests, that are substanially buffered such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
+
+In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites with $T_{leaf max}$ exceeding $T_{opt}$ at different canopy heights and temperatures in all forests, with no significant relationship between $T_{opt}$ and height.
 Additionally, with increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
@@
 
-**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light, in order to optimize C uptake, water use, thermoregulation and metabolism across forest vertical strata, and thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015; *REFs*]**
+**Leaf thermoregulation largely depends on leaf traits.
+Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light, in order to optimize C uptake, water uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015; *REFs*]**
 
@@
 The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+
+However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above.
+In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves.
+This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
+1996, Poorter et al.
+2000, @rozendaalPlasticityLeafTraits2006]
 
@@
  "Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well" -Martijn
-
-The combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees.
-For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc..
-whereas understory leaves are subjected to shade or limitation of light, niche competition and herbivory.
-Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heat loss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes.
-Additionally, overstory sun leaves tend to have long-lived leaves[*double check*] with high LMA while shade leaves of canopy trees have large leaves with low LMA.
-Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to low levels of photosynthesis in the shade, lesser $VPD$, and wind speed.
-Those differences in traits are likely to affect their thermal properties as well.
-Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [@rozendaalPlasticityLeafTraits2006; @niinemetsResponsesForestTrees2010a]
-
-However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above.
-In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves.
-This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
-1996, Poorter et al.
-2000, @rozendaalPlasticityLeafTraits2006]
 

### 2021-04-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2009, 2010*).
-Under constant light conditions (photon flux density), taller trees (>35m) have lower canopy stomatal conductance compared to shorter trees  [@schaferEffectTreeHeight2000; REF].
-Adaptation to increasing height and hydraulic contraints show that leaf size [@chinLeafAcclimationLight2017] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) has shown to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a].
+Under constant light conditions (photon flux density), increase in height (>35m) showed decrease in canopy stomatal conductance, along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
+Adaptations to increasing height and hydraulic contraints show that leaf size [@chinLeafAcclimationLight2017] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) tends to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].
 
@@
 As a result, degrees of high to low irradiance gives rise to a dichotomy of sun and shade leaf traits.
-Typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves [Rozendaal et al.
-2006, Mathur et al.
-2018, *REFS*].
-Within-species, several studies point towards these common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient based on adaptation to primarily light conditions, and as well biophysical gradient [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Trait adaptation maximizes light use efficiency for photosynthesis at differing light levels and conditions mentioned above.
+Therefore, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves, where as shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves.
+This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
+1996, Poorter et al.
+2000; @rozendaalPlasticityLeafTraits2006; Mathur et al.
+2018].
+Within-species, several studies point towards these common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
 The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 
-However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above.
-In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves.
-This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
-1996, Poorter et al.
-2000, @rozendaalPlasticityLeafTraits2006]
-
-An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient (???), because LES theory was developed using mainly sun exposed (upper canopy) leaves.
+An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [Chen et al.
+2019], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
 Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
-However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018) as well as geographically (Ishii et al., 2018).
+However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018)
 
@@
 ### Interspecific varaiation
+
+Across ontogeny, while canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits, the difference has been largely attriuted to changes in light conditions [Rosendaal et al.
+2006], therefore maintaing similar responses to changes in light with increasing height as highlight within species.
+Under open-light conditions for canopy and understory,in large gaps and open-grown conditons,
+evidence shows that canopy trees showed significant increase in LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings
+[Thomas and Winner, 2002]. However, plasticity to light has been shown difference between species [Houter and Pons 2012]
 

### 2021-04-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019].
-Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
-2021; @davis_microclimatic_2019; @zellweger_seasonal_2019].
+Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
+2021; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009].
 In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-On the other hand, open-forests with sparse stands, such as needle-leaf forests, have lower buffering capacity, greater vertical air mixing and greater light permeability through canopy layers.
-For this reason, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999].
-During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
+On the other hand, open-forests with sparse stands, have lower buffering capacity, greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [Rambo and NOrth, 2009].
+For this reason, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; Rambo and North, 2009].
+During the night, pine forests have been observed to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
 Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer.
@@
 Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates, which tends to decrease with increasing maximum temperature.
-In three European forests (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.RH moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
+In three European forests (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
+RH moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
 
@@
 2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al.
-2018]
+2018].
+Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$, whereas open-forests, with greater light interception to forest floors, in come cases, reflect a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]
 
@@
 Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*.
-Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.
+Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al.
+2018].
 However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.
@@
 1994*; @fauset_differences_2018].
-On the other, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
+On the other hand, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
 
@@
 While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018].
-Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.
+Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation or $T_{air}$ with stomata closed [@leighInfluenceLeafSize2017].
 
@@
 `r fig_leaf_T`).**
-Leaf processes and leaf temperature at a vertical gradient are influenced by vertical microclimate conditions.
 While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
 
-Particularly, in open forests with sparse distribution of trees, such as in some needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing, heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+Particularly, in open forests with sparse distribution of trees, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
 Greater heat stress and $T_{air}$ in the lower canopy is also observed in Australian desert trees [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground
@@
 Under constant light conditions (photon flux density), increase in height (>35m) showed decrease in canopy stomatal conductance, along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
-Adaptations to increasing height and hydraulic contraints show that leaf size [@chinLeafAcclimationLight2017] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) tends to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].
+Adaptations to increasing height and hydraulic contraints show that leaf size [Kaare H.
+Jensen and Maciej A.
+Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) tends to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].
 
@@
 2019; Tserej & Feeley 2021].
-Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperature ranges [$T_{range}$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017].
-With increases in effective leaf width, boundary layer thickness and resistance increase along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
+Yet, thermal imagery shows that effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant than leaf area[@leighInfluenceLeafSize2017].
+With increases in effective leaf width, thermal time constant also increases due to increase in boundary layer thickness and resistance, along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
 2019; @fauset_differences_2018].
-This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018].
-Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
-Using leaf energy balance models to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al.
-2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$.
-Leaf width and characteristic dimension are generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref].
-Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation.
-Leaf lobes are deeper and leaf elongation and serration are more pronounced in sun exposed canopies than in the shaded understory [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
+Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves.
+Small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
+Evidence shows that leaf temperature increased with effective leaf width [fauset et al.
+2018].
+However, leaf dissection such as lobes, leaf elongation and serration, enhancing a smaller characteristic dimension, thus facilitate efficient heat dissipation, and are evidently characteristics of sun exposed leaves than shaded leaves along a vertical gradient [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
+Other morphological characteristics such as increased petiole length, characteristic of leaves growing in high-light conditions and in exposed canopy positions, also contribute to heat dissapation [Niinemetes]
 
@@
 
-High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition through enabling D1 protein replacement.
-Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis.
-Xanthophylls are photoprotective pigments comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition.
+Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments, which are photoprotective pigments comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 These are found to be proportional to within-canopy light gradients at an inter- and intra-species level.
@@
 -	antioxidant scavenging function- protect against cellular damage
-- Koniger et. al. 1995, and  Matsubara et al. 2009
 
@@
 ##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
-*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
+*Leaf lifespan generally increases with LMA.
+LMA and leaflifespan is greater with available light, and largely depends on cost of carbon investment in a leaf, which varies across species [**double check**] Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
 This is bound to affect metabolism, and may affect other aspects of their performance.*
@@
 
-Across ontogeny, while canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits, the difference has been largely attriuted to changes in light conditions [Rosendaal et al.
-2006], therefore maintaing similar responses to changes in light with increasing height as highlight within species.
-Under open-light conditions for canopy and understory,in large gaps and open-grown conditons,
-evidence shows that canopy trees showed significant increase in LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings
-[Thomas and Winner, 2002]. However, plasticity to light has been shown difference between species [Houter and Pons 2012]
-
-*(KAT: a lot of this content should probably move up the the intro to this section)*
-
-- Elsa: see also H.
-Poorter, U.
-Niinemets, L.
-Poorter, I.
-J.
-Wright, R.
-Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
-New Phytol.
-182, 565–588 (2009).
--- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
-
- "Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well" -Martijn
+Across ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits, however, the difference has been largely attributed to changes in light conditions rather than ontogeny [Rosendaal et al.
+2006], therefore maintaing similar responses to changes in light with increasing height such as in within-species.
+Under constant light conditions of open-grown canopy and understory, for example in large gaps, evidence shows that canopy trees showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings, attributing the differences in traits to ontogeny [Thomas and Winner, 2002; Houter and Pons 2012].
+On the basis of phenology, temperate understory trees have been observed to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al.
+2013], however, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renne, 2019]
 
@@
 Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [@niinemetsControversyTraitsConferring2006].
-However, the findings from the above review conflicts with  Kitajima 1994 found for tropical seedlings; shade intolerance species grew faster in the sun AND in the shade, but at the cost of higher mortality in the shade.
-Lawren and Karoru had contrasting findinds, and so far appears that there is a consensus by now.
+However, the findings from the above review conflicts with  Kitajima 1994 who found for tropical seedlings; shade intolerance species grew faster in the sun and in the shade, but at the cost of higher mortality in the shade.
+Lawren and Karoru had contrasting findinds, and so far appears that there is no a consensus?
 
@@
 
+*(KAT: a lot of this content should probably move up the the intro to this section)*
+
 ##### Isoprene production
@@
 Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
-However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves (Table `r table_leaf_metabolism`).
+However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves and vertical gradient (Table `r table_leaf_metabolism`).
 
@@
 
-- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015), however stomatal conductance limitation is also greater, and earlier with canopy height [Sanches et al. 2009] 
-Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
--understory leaves open their stomata wider than upper canopy leaves [Kosugi et al. 2021]
+- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015), however stomatal conductance limitation is also greater, and earlier with greater canopy height [Sanches et al. 2009] 
+Max transpiration rate - increases with canopy height (Kenzo et al. 2015), and quickly decreases with increasing VPD at greater heights [Scaffer].
+-understory leaves open their stomata wider than upper canopy leaves [Kosugi et al.
+2021].
+Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger].
+Stomatal opening is greater in understory compared to canopy
 
@@
 This is primarily driven by the greater light available to sun leaves.
+Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy.
+Therefore, in dense canopies and across species, photosynthesis is greater at canopy tops where light is most available, compared to middle and lower canopy [REF].
+However, light acclimated shade leaves have greater light use efficiency and therefore are observed to maintain higher photosynthetic capacity on a mass-basis?
 
@@
 The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)
+
+Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
+2017; Slot & Winter 2017).
+Beyond the optimum, photosynthesis decreases as a result of stomatal closure, accompanied by an increase in leaf temperature in dense upper canopies tends to decrease phtotosynthesis [REF](e.g.
+Slot & Winter 2017; Smith et al.
+2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs.
+e.g.
+Sage & Kubien 2007; Varhammer et al.
+2015 ).
+
+Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis.
+At increasing temperatures, photosynthetic system II [PSII] is particularly vulnerable.
+Damage to PSII  can lead to decreased electron transport rates and photosynthetic failure.
+Irreversible damage to PSII is done at 40-60C [Baker 2008; @feeley_thermal_2020]
+Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought. [@niinemetsWorldwideAnalysisWithincanopy2015]
+
 *[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N.
@@
 In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*
-
-Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy.
-
-Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
-2017; Slot & Winter 2017).
-Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g.
-Slot & Winter 2017; Smith et al.
-2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs.
-e.g.
-Sage & Kubien 2007; Varhammer et al.
-2015 ).
 
@@
 
-Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis.
-At increasing temperatures, photosynthetic system II [PSII] is particularly vulnerable.
-Damage to PSII  can lead to decreased electron transport rates and photosynthetic failure.
-Irreversible damage to PSII is done at 40-60C [Baker 2008; @feeley_thermal_2020]
-Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought. [@niinemetsWorldwideAnalysisWithincanopy2015]
-
 **Leaf thermal tolerance (Tcrit/ T50)**
@@
 Typical values ~45-50$^\circ$C, varying somewhat 
-across latitude/climate (O’sullivan et al. 2017),
+across latitude/climate (O' sullivan et al. 2017),
 with elevation [@feeley_thermal_2020;  Slot et al unpublished data],
 with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
+Tcrit is greater for buffered leaves that maintain greater effective leaf with [Slot et al. 2021]
 Perez and Feeley et al.
@@
 **Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
-(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. 
+(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. Conflicting $Q_10$ findings shows no definite trend.
 Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). 
@@
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
+[Increase in humidity?]
 
@@
 
-Interaction between degrees of shade x ontogeny x height with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
-Leaf traits vary more on the basis on light environment than ontogenetically [@liuPlantSizeEffects2010]
+Interaction between degrees of shade x ontogeny with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
 

### 2021-04-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$.
-On the other hand, under conditions of constant radiation, increase in $T_{air}$ can induce evaporative cooling [Drake et al.
+On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ can induce evaporative cooling [Drake et al.
 2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al.
@@
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Especially in closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies compared to lower canopies of tropical forests, that are substanially buffered such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
+Especially in closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies compared to lower canopies of tropical forests that are substanially buffered, such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
 
@@
 
+#### *(clarify light vs height - 1 paragraph)*
+
+Leaf traits in woody species take approximately 30-60 days to stabilize to integrated light after formation [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+As a result, degrees of high to low irradiance gives rise to a dichotomy of sun and shade leaf traits.
+Trait adaptation maximizes light use efficiency for photosynthesis at differing light levels and conditions mentioned above, while maintaing a balance between thermoregulation and water status [**REF**].
+
 Particularly in dense forests, the upper, outer canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs.
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
-Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism and carbon balance (following section).
+Exposed, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves, whereas shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis.
+"This enables shade leaves to realize similar light capture, under low light, to sun leaves at a lower biomass investment" [Chazdon et al.
+1996, Poorter et al.
+2000; @rozendaalPlasticityLeafTraits2006; Mathur et al.
+2018].
+Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) and carbon balance.
 It is unsuprising, then, that leaf traits vary dramatically across vertical gradients.
 
-#### *(clarify light vs height - 1 paragraph)*
-
-Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
+While light based acclimation in leaf traits is well established, research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
 Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.However, the majority of relevant research has focused on exposure gradients near ground level, with by far the most common study type being comparisons of sun and shade leaves.
@@
 
-Leaf traits in woody species take approximately 30-60 days to stabilize to integrated light after formation [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-As a result, degrees of high to low irradiance gives rise to a dichotomy of sun and shade leaf traits.
-Trait adaptation maximizes light use efficiency for photosynthesis at differing light levels and conditions mentioned above.
-Therefore, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves, where as shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves.
-This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al.
-1996, Poorter et al.
-2000; @rozendaalPlasticityLeafTraits2006; Mathur et al.
-2018].
 Within-species, several studies point towards these common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].

### 2021-04-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2018].
-Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$, whereas open-forests, with greater light interception to forest floors, in come cases, reflect a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]
+Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$, whereas open-forests, with greater light interception to forest floors, in come cases, exhibit a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]
 
@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
-Exposed, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves, whereas shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis.
+Exposed, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves.
+These traits also facilitates heat dissapation [Vogel] in sun leaves compared to shade.
+Whereas shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis.
 "This enables shade leaves to realize similar light capture, under low light, to sun leaves at a lower biomass investment" [Chazdon et al.
@@
 2000; @rozendaalPlasticityLeafTraits2006; Mathur et al.
-2018].
+2018], however might not be most suitable in dissapating heat [Vogel].
 Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) and carbon balance.
@@
 
-Within-species, several studies point towards these common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+Within-individuals in dense forests, canopy positions are exposed to different light and biophysical conditions in the upper, middle and lower canopy.
+Thus leaf traits vary accordingly.
+Whereas in open-forests or under consistent canopy light conditions, leaf traits vary on the basis of light exposure.
+In the former condition, several studies point towards common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 
@@
 Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al.
-2019; Tserej & Feeley 2021].
-Yet, thermal imagery shows that effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant than leaf area[@leighInfluenceLeafSize2017].
+2019; Tserej & Feeley 2021], however effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant than leaf area [@leighInfluenceLeafSize2017].
 With increases in effective leaf width, thermal time constant also increases due to increase in boundary layer thickness and resistance, along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
 2019; @fauset_differences_2018].
-Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves.
-Small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
+Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves that are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
 Evidence shows that leaf temperature increased with effective leaf width [fauset et al.
@@
 However, leaf dissection such as lobes, leaf elongation and serration, enhancing a smaller characteristic dimension, thus facilitate efficient heat dissipation, and are evidently characteristics of sun exposed leaves than shaded leaves along a vertical gradient [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
-Other morphological characteristics such as increased petiole length, characteristic of leaves growing in high-light conditions and in exposed canopy positions, also contribute to heat dissapation [Niinemetes]
+Other morphological characteristics in exposed canopy positions such as increased petiole length, cuticle thickness, trichomes, stomatal density also enhance heat dissapation compared to shade leaves [Niinemetes]
 
@@
 *Leaf lifespan generally increases with LMA.
-LMA and leaflifespan is greater with available light, and largely depends on cost of carbon investment in a leaf, which varies across species [**double check**] Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
+LMA and leaflifespan is greater with available light, and largely depends on the cost of carbon investment in a leaf, which varies across species sun vs.
+shade leaves [**double check**] Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
 This is bound to affect metabolism, and may affect other aspects of their performance.*
@@
 
-Across ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits, however, the difference has been largely attributed to changes in light conditions rather than ontogeny [Rosendaal et al.
-2006], therefore maintaing similar responses to changes in light with increasing height such as in within-species.
+Across ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits in their biophysical environment, however, the difference has been largely attributed to changes in light conditions rather than ontogeny itself [Rosendaal et al.
+2006], therefore maintaing similar responses to changes in light with increasing height, as observed in within-species gradient.
 Under constant light conditions of open-grown canopy and understory, for example in large gaps, evidence shows that canopy trees showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings, attributing the differences in traits to ontogeny [Thomas and Winner, 2002; Houter and Pons 2012].
-On the basis of phenology, temperate understory trees have been observed to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al.
-2013], however, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renne, 2019]
+On the basis of phenology, temperate understory trees tended to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al.
+2013], and, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renne, 2019]
 

### 2021-04-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-Leaves maximize carbon assimilation by maintaining metabolic optima, in a given circumstance, through the adaptation of their leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+Maintaining optimum leaf temperature in relation to the biophysical environment, maximizes carbon assimilation by maintaining metabolic optima, in a given circumstance.
+This is done through the adaptation of leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 `r fig_leaf_T`).
-The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost to the environment [Fig.
+The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
 `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
@@
 2018].
-Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$, whereas open-forests, with greater light interception to forest floors, in come cases, exhibit a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]
+Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$ compared to lower canopy or understory, whereas open-forests, with greater light interception to forest floors, in some cases, exhibit a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]
 
@@
 **Leaf thermoregulation largely depends on leaf traits.
-Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light, in order to optimize C uptake, water uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015; *REFs*]**
+Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light and water demand, in order to optimize C uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015; Zweiniecki et al 2004]**
 
@@
 
-Leaf traits in woody species take approximately 30-60 days to stabilize to integrated light after formation [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-As a result, degrees of high to low irradiance gives rise to a dichotomy of sun and shade leaf traits.
-Trait adaptation maximizes light use efficiency for photosynthesis at differing light levels and conditions mentioned above, while maintaing a balance between thermoregulation and water status [**REF**].
+Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [Zweineicki et al.
+2004].
+Degrees of direct to diffused light and consequental water demands shape sun vs.
+shade leaves and their necessary vascular architecture [Casal et al.
+2011, Zweineicki et al.
+2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+Therefore, exposed sun leaves are typically smaller in size and width, thicker, hairier and have greater lobation, and buffer biophysical conditions for shade leaves.
+Whereas shade leaves adapt to diffused light and stable microenvironment through investing in light capture efficiency by expanding specific leaf area (SLA) [Sack et al.
+2006; Chazdon et al.
+1996, Poorter et al.
+2000; @rozendaalPlasticityLeafTraits2006; Mathur et al.
+2018].
+Respective leaf traits also play a role in heat dissapation where exposed sun leaves are more efficient in dissapting heat than buffered shade leaves [Vogel].
 
@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
-Exposed, typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves.
-These traits also facilitates heat dissapation [Vogel] in sun leaves compared to shade.
-Whereas shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis.
-"This enables shade leaves to realize similar light capture, under low light, to sun leaves at a lower biomass investment" [Chazdon et al.
-1996, Poorter et al.
-2000; @rozendaalPlasticityLeafTraits2006; Mathur et al.
-2018], however might not be most suitable in dissapating heat [Vogel].
 Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) and carbon balance.
@@
 
-Within-individuals in dense forests, canopy positions are exposed to different light and biophysical conditions in the upper, middle and lower canopy.
-Thus leaf traits vary accordingly.
-Whereas in open-forests or under consistent canopy light conditions, leaf traits vary on the basis of light exposure.
-In the former condition, several studies point towards common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+Within-individuals exposed canopies subjected to harsher light and biophysical conditions, substanially buffer inner and lower canopies.
+Thus leaf traits in canopy positions vary accordingly [Casal et al.
+2011].
+Several studies point towards common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+The differences among these traits contribute to leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 
@@
 #### Morphological traits
-- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
 
-Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al.
-2019; Tserej & Feeley 2021], however effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant than leaf area [@leighInfluenceLeafSize2017].
-With increases in effective leaf width, thermal time constant also increases due to increase in boundary layer thickness and resistance, along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
+Along the vertical gradient, leaf temperature increases with effective leaf width and leaf size [Fauset et al.
+2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
 2019; @fauset_differences_2018].
 Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves that are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
-Evidence shows that leaf temperature increased with effective leaf width [fauset et al.
-2018].
-However, leaf dissection such as lobes, leaf elongation and serration, enhancing a smaller characteristic dimension, thus facilitate efficient heat dissipation, and are evidently characteristics of sun exposed leaves than shaded leaves along a vertical gradient [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
-Other morphological characteristics in exposed canopy positions such as increased petiole length, cuticle thickness, trichomes, stomatal density also enhance heat dissapation compared to shade leaves [Niinemetes]
+Exposed leaves reduce light exposures, and thus leaf temperature, by maintaining vertical leaf angles, or through temporary wilting [Niinemetes]1998.
+Other trait adaptions to dissapting heat such as lobes, serration, longer petiole length and stomtal density are evidently characteristics of sun exposed leaves than shaded leaves, help maintain a lower leaf temperature along a vertical gradient [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]
 
@@
 
- In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015].
+-Mitsutoshi kitao 2018: Canopy nitrogen distribution is optimized to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks.
+P uptake cannoy keep up wth increased growth at higher light [Poorter et al.
+2019]
+ In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].
 
@@
 
-High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition.
-Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments, which are photoprotective pigments comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-These are found to be proportional to within-canopy light gradients at an inter- and intra-species level.
-Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998].
-In neotropical forests, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pools on an area or chlorophyll basis compared to low light leaves that showed greater mass-based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al.
-2009]
-
--	antioxidant scavenging function- protect against cellular damage
+Proportional to vertical light gradients at an inter- and intra-species level, sun exposed leaves have greater pools of xanthophyll pigments, comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] than in shaded leaves.
+[$VAZ$]--an antioxidant scavenging function--protects against cellular damage by converting excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 
@@
 
-Across ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits in their biophysical environment, however, the difference has been largely attributed to changes in light conditions rather than ontogeny itself [Rosendaal et al.
-2006], therefore maintaing similar responses to changes in light with increasing height, as observed in within-species gradient.
-Under constant light conditions of open-grown canopy and understory, for example in large gaps, evidence shows that canopy trees showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings, attributing the differences in traits to ontogeny [Thomas and Winner, 2002; Houter and Pons 2012].
+Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits in their biophysical environment, however, the differences have been largely attributed to adaption to light conditions rather than ontogeny itself, therefore maintaing similar responses to changes in light with increasing height, as observed in within-species gradient [Rosendaal et al.
+2006; Thomas and Winner, 2002; Houter and Pons 2012].
+However, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, evidence shows that canopy trees showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings, attributing the differences in traits to tree-size ontogeny [Thomas and Winner, 2002; Houter and Pons 2012] which also shapes leaf temperature.
+Across different species, while common adaptations to light and water are observed, inherent traits differences such as leaf shape, size and width, stomtal conductance levels influence leaf temperatures.
+Therefore, some species might be able to tolerate greater heat stress at a vertical gradient than others [[Fauset et al.
+2018].
+
 On the basis of phenology, temperate understory  trees tended to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al.
-2013], and, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renne, 2019]
-
-With increasing height and age, competition for light decreases as availability  of light increases in dense forests.
-However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements.
-Early successional species are considered more shade intolerant than late successional species that are more shade tolerant.
-Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [@niinemetsControversyTraitsConferring2006].
-However, the findings from the above review conflicts with  Kitajima 1994 who found for tropical seedlings; shade intolerance species grew faster in the sun and in the shade, but at the cost of higher mortality in the shade.
+2013], and, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renner, 2019].
+However, May spring heat waves can negetively impact partially expanded canopy leaves by inducing leaf shedding, more than saplings and understory [Filewood and Thomas, 2013].
+Ontogenetic differences also suggests shade tolerators have lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities than in intolerators [@niinemetsControversyTraitsConferring2006] reflecting shade adaptation and avoidance strategies [Casal, 2013].
+While shade intolerant species experience greater heterogeneity in light during ontogeny [poorter et al.
+2013].
+However, these findings conflict with  Kitajima 1994 who found for tropical seedlings; shade intolerance species grew faster in the sun and in the shade, but at the cost of higher mortality in the shade.
 Lawren and Karoru had contrasting findinds, and so far appears that there is no a consensus?
-
-Shade intolerant species experience greater heterogeneity in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H.
-Poorter, U.
-Niinemets, L.
-Poorter, I.
-J.
-Wright, R.
-Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis.
-New Phytol.
-182, 565–588 (2009).].Rozendaal et al.
-2006 found no significant trait differences with ontogeny and adult stature trees.
-Koike et al.
-2001 has more info about this.
+How will leaf temperature vary between the two?
 
@@
 
-- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015), however stomatal conductance limitation is also greater, and earlier with greater canopy height [Sanches et al. 2009] 
+- Max stomatal conductance - increases with canopy height (Kenzo et al.
+2015), however stomatal limitation is steeper also greater with hydraulic limitation, in sun leaves and in exposed canopies [Sanches et al.
+2009; Hernandez]
 Max transpiration rate - increases with canopy height (Kenzo et al. 2015), and quickly decreases with increasing VPD at greater heights [Scaffer].
@@
 
+Leaf temperature reflects photosynthetic optimum temperature [**REF**].
 Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
@@
 Damage to PSII  can lead to decreased electron transport rates and photosynthetic failure.
-Irreversible damage to PSII is done at 40-60C [Baker 2008; @feeley_thermal_2020]
-Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought. [@niinemetsWorldwideAnalysisWithincanopy2015]
+Irreversible damage to PSII is done at 40-60C [Baker 2008; @feeley_thermal_2020].
+Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought.
+[@niinemetsWorldwideAnalysisWithincanopy2015]
 

### 2021-04-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$.
-On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ can induce evaporative cooling [Drake et al.
+On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ induces evaporative cooling [Drake et al.
 2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al.
 2018].
-Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$ compared to lower canopy or understory, whereas open-forests, with greater light interception to forest floors, in some cases, exhibit a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]
+Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore tend to maintain greater $T_{leaf}$ than $T_{air}$ compared to lower canopy or understory, whereas open-forests, with greater light interception to forest floors, in some cases, exhibit a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]
 
@@
 
-Particularly, in open forests with sparse distribution of trees, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
-Greater heat stress and $T_{air}$ in the lower canopy is also observed in Australian desert trees [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground
-[@hadleyInfluenceKrummholzMat1987].
-Similarly, during the dry season in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
+Particularly, in open forests with sparse distribution of trees, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+Greater heat stress and $T_{air}$ in the lower canopy is also observed in Australian desert trees [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground [@hadleyInfluenceKrummholzMat1987].
+Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
+In other dry conditions, inner canopy can be warmer than above canopy and air temperature due to greater relative humidity [Dietz et al.
+2007].
 Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
@@
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Especially in closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies compared to lower canopies of tropical forests that are substanially buffered, such as in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
+Especially in closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies compared to lower canopies, caused by midday stomatal closure that reduces transpirational cooling [Slot et al.
+2019].
+This is seen in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
 
-In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites with $T_{leaf max}$ exceeding $T_{opt}$ at different canopy heights and temperatures in all forests, with no significant relationship between $T_{opt}$ and height.
-Additionally, with increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
-- In a tropical wet and moist forests of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
--"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al.
-2019)
+In three forest types- US temperate deciduous, Puerto Rico tropical wet and tropical moist, $T_{leaf_{max}}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in the upper canopies of all three forests, along with increasing $T_{leaf}$ - $T_{air}$  with height.
+$T_{leaf}$ - $T_{air}$ was observed to be greatest in temperate deciduous forest compared to tropical moist and wet.
+Greater light input in exposed upper canopy leaves can sometimes elevate leaf temperature $1-7{^\circ}C$ above air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [Doughty and Gouden, 2008; @mauTemperateTropicalForest2018a]
 
@@
 Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves that are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
-Exposed leaves reduce light exposures, and thus leaf temperature, by maintaining vertical leaf angles, or through temporary wilting [Niinemetes]1998.
+Exposed leaves reduce light exposures, and thus leaf temperature, by maintaining vertical leaf angles, or through temporary wilting [Ishida et al.
+1999; Niinemetes, 1998].
 Other trait adaptions to dissapting heat such as lobes, serration, longer petiole length and stomtal density are evidently characteristics of sun exposed leaves than shaded leaves, help maintain a lower leaf temperature along a vertical gradient [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]
@@
 
--Mitsutoshi kitao 2018: Canopy nitrogen distribution is optimized to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks.
-P uptake cannoy keep up wth increased growth at higher light [Poorter et al.
+Canopy nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks, and could mediate sun vs.
+shade leaf response to heatflecks [Kitao et al.
+2018].
+Therefore, during growing season, nitrogen per area [$N_{area}$], along with LMA, leaf soluble sugars, starch and non-structural carbons increase with height and light, typically greater in exposed canopy positions or sun leaves compared to shade [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Liu et al.
+2010].
+However, increase in temperature in upper canopy, shows a decrease in LMA and $N_{area}$ [Carter and Cavaleri, 2014] and in leaf soluble sugars, starch and non-structural carbons [Li et al.2016 **double check**].
+Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up [Poorter et al.
 2019]
- In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].
 
@@
 
-Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits in their biophysical environment, however, the differences have been largely attributed to adaption to light conditions rather than ontogeny itself, therefore maintaing similar responses to changes in light with increasing height, as observed in within-species gradient [Rosendaal et al.
+Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in fundamental light adapted leaf traits in their biophysical environment as observed and similar to within-species gradient [Rosendaal et al.
 2006; Thomas and Winner, 2002; Houter and Pons 2012].
-However, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, evidence shows that canopy trees showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings, attributing the differences in traits to tree-size ontogeny [Thomas and Winner, 2002; Houter and Pons 2012] which also shapes leaf temperature.
-Across different species, while common adaptations to light and water are observed, inherent traits differences such as leaf shape, size and width, stomtal conductance levels influence leaf temperatures.
-Therefore, some species might be able to tolerate greater heat stress at a vertical gradient than others [[Fauset et al.
+Under constant light conditions of open-grown canopy and understory trees, for example in large gaps, ontogeny differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [Thomas and Winner, 2002; Houter and Pons 2012].
+Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels influence leaf temperatures.
+Therefore, some species might more susceptible to heat stress at a vertical gradient than others [Fauset et al.
 2018].
 
-On the basis of phenology, temperate understory  trees tended to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al.
-2013], and, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renner, 2019].
-However, May spring heat waves can negetively impact partially expanded canopy leaves by inducing leaf shedding, more than saplings and understory [Filewood and Thomas, 2013].
-Ontogenetic differences also suggests shade tolerators have lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities than in intolerators [@niinemetsControversyTraitsConferring2006] reflecting shade adaptation and avoidance strategies [Casal, 2013].
+Ontogenetic differences also suggests shade tolerators have lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities than in intolerators [@niinemetsControversyTraitsConferring2006].
 While shade intolerant species experience greater heterogeneity in light during ontogeny [poorter et al.
@@
 ### Stomatal conductance
-Photosynthesis can be largely constrained by stomatal conductance.
 
-- Max stomatal conductance - increases with canopy height (Kenzo et al.
-2015), however stomatal limitation is steeper also greater with hydraulic limitation, in sun leaves and in exposed canopies [Sanches et al.
-2009; Hernandez]
-Max transpiration rate - increases with canopy height (Kenzo et al. 2015), and quickly decreases with increasing VPD at greater heights [Scaffer].
--understory leaves open their stomata wider than upper canopy leaves [Kosugi et al.
+While exposed sun leaves are equiped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well.
+Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [Kenzo et al.
+2015; Sanches et al.
+2009].
+During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [Sanches et al.
+2009; Zweinniecki et al.
+2004].
+Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves; therefore decreases with height [Sanches et al.
+2009; Hernandez].
+Similarly, max transpiration rate increases with canopy height [Kenzo et al.
+2015].
+Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [Kosugi et al.
 2021].
-Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger].
-Stomatal opening is greater in understory compared to canopy
 
@@
 
-**Photosynthesis is generally higher in sun leaves-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al.
+**Photosynthesis is generally higher in sun leaves and exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al.
 2015 ; Slot et al.
@@
 This is primarily driven by the greater light available to sun leaves.
-Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy.
-Therefore, in dense canopies and across species, photosynthesis is greater at canopy tops where light is most available, compared to middle and lower canopy [REF].
-However, light acclimated shade leaves have greater light use efficiency and therefore are observed to maintain higher photosynthetic capacity on a mass-basis?
+Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
 
 In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma.
-Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
-The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)
+Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P.
+(and water availability)
 
-Leaf temperature reflects photosynthetic optimum temperature [**REF**].
+Leaf temperature beyond photosynthetic optimum temperature, declines photosynthesis [**REF**].
 Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
 2017; Slot & Winter 2017).
-Beyond the optimum, photosynthesis decreases as a result of stomatal closure, accompanied by an increase in leaf temperature in dense upper canopies tends to decrease phtotosynthesis [REF](e.g.
+Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g.
 Slot & Winter 2017; Smith et al.
@@
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*) *as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].* *(KAT: I'm confused by the second part of this sentence.
-Are these studies, or theoretical expectations?.)*
+14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [Niinemetes et al.
+1999b; Niinemetes and Valladares, 2004].
 However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
 For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [@slotPhotosyntheticHeatTolerance2019 ; @hernandezSimilarTemperatureDependence2020].
-Similar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
-Similarly, Mau et al. found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
+Similar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020].
+Similarly, Mau et al.
+2018 found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
 
@@
 *(define)*
-Typical values ~45-50$^\circ$C, varying somewhat 
-across latitude/climate (O' sullivan et al. 2017),
-with elevation [@feeley_thermal_2020;  Slot et al unpublished data],
-with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
-Tcrit is greater for buffered leaves that maintain greater effective leaf with [Slot et al. 2021]
+Typical values ~45-50$^\circ$C, varying somewhat across latitude/climate (O' sullivan et al.
+2017), with elevation [@feeley_thermal_2020;  Slot et al unpublished data] and with leaf traits [Sastry et al.
+2018, @zhang_photosynthetic_2012].
+Tcrit is greater for buffered leaves that maintain greater effective leaf with [Slot et al.
+2021]
 Perez and Feeley et al.
@@
 For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, both measured near ground level (Slot et al. 2019). 
-Similarly, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher $T_{50}$ was observed in more exposed (North-facing) canopy positions (Curtis et al. 2019). 
+However, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher climatic heat stress and thus $T_{50}$ was observed in the lower canopy of North-facing canopy positions (Curtis et al.
+2019).
 We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
 I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot
-
-T50 might not capture thermal stress or sensitivity to increasing temperature in foliage, because the degree at which T50 is observed in leaves is at an extreme temperature not usually experienced by leaves acclimated to different light environment, at a lower foliage thermal stress point T50 might be irrelevant, but variation to extreme temperature response might be better understood through T50 [@perezPSIIHeatTolerances2020]
 
@@
 Although drought is primarily a hydraulic problem, lack of water --> lower gs --> higher leaf T, so leaves face tradeoffs of water loss vs potentially damaging leafT.
+
+On the basis of phenology, temperate understory  trees tend to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al.
+2013], and, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renner, 2019].
+However, short-term May spring heat waves have shown to negetively impact partially expanded canopy leaves by inducing leaf shedding, more than understory saplings [Filewood and Thomas, 2013].
 

### 2021-04-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-While exposed sun leaves are equiped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well.
+While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well.
 Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [Kenzo et al.
@@
 However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
-For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [@slotPhotosyntheticHeatTolerance2019 ; @hernandezSimilarTemperatureDependence2020].
-Similar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020].
-Similarly, Mau et al.
-2018 found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].
+For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
+2019 ; @hernandezSimilarTemperatureDependence2020].
+Similar results were found for $V_{max}$ and $J_{max}$ [Slot et al.
+2019; @hernandezSimilarTemperatureDependence2020]and along a vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; Miller et al.
+2020], and no significant trend in temperate trees [@mau_temperate_2018].
 
@@
 For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, both measured near ground level (Slot et al. 2019). 
-However, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher climatic heat stress and thus $T_{50}$ was observed in the lower canopy of North-facing canopy positions (Curtis et al.
+However, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher climatic heat stress and thus $T_{50}$ was observed in the lower north-facing canopy positions (Curtis et al.
 2019).

### 2021-04-22 vertical-thermal-review.Rmd
Merge branch 'master' of https://github.com/EcoClimLab/vertical-thermal-review

@@
 2004].
-Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves; therefore decreases with height [Sanches et al.
+Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [Sanches et al.
 2009; Hernandez].
@@
 **Leaf thermal tolerance (Tcrit/ T50)**
-*(define)*
-Typical values ~45-50$^\circ$C, varying somewhat across latitude/climate (O' sullivan et al.
-2017), with elevation [@feeley_thermal_2020;  Slot et al unpublished data] and with leaf traits [Sastry et al.
-2018, @zhang_photosynthetic_2012].
-Tcrit is greater for buffered leaves that maintain greater effective leaf with [Slot et al.
-2021]
-Perez and Feeley et al.
-2020 found that across species sun leaves with higher photosynthetic heat tolerance experienced higher maximum temperatures and had narrower thermal safety margins.
-These species may thus be at a greater risk of thermal damage.
-[- Elsa: also see: 
-- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.]
+$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [Slot et al.
+2020].
+Leaf thermal thresholds strongly reflect species adaptation to their microenvironment rather than macroclimate conditions [@feeley_thermal_2020; Slot et al.
+2020; Perez and Feely 2020a] and phylogeny [Slot et al.
+2020; Perez and Feeley 2020a].
 
-Supporting the significant negative relationship between $T50$ and thermal safety margins, photosynthesis [PS2] temperature regulation has been associated with leaf thermal tolerance in order to maximize net carbon assimilation.
-However, randomly collected within-canopy leaves of individual mature plants in Florida showed that Tmax is not directly coordinated with high PS2 heat tolerance and does not necessarily increase leaf thermal tolerance [shows no direct correlation]; nor does it increase carbon assimilation at higher temperature.
-Rather, high PS2 heat tolerance reduces optimum temperature for C assimilation [$Topt$] because high PS2 heat tolerance requires a greater cost of PS2 metabolic activity maintenance, thus not contributing to increased C assimilation.
-Species exhibiting low carbon assimilation rates tended to have greater stress tolerance.
-In addition, species with low stomatal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimilation before thermal limits are reached [@perezPSIIHeatTolerances2020]
+Typical values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al.
+2017).
+Across species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  Slot et al 2020; O' sullivan et al.
+2017] and varies with leaf traits [Sastry et al.
+2018, @zhang_photosynthetic_2012; Slot et al.
+2020] and their thermal time constant.
+Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$.
+On the other hand, leaves of species with greater construction costs, observed through LMA, had greater $T_{50}$ compared to species with low LMA [Slot et al.
+2020].
+However, higher photosynthetic heat tolerance does not necessary increase leaf thermal safety margins, rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum temperatures, low optimum temperature for C assimilation [$T_{opt}$] and had narrower thermal safety margins [Perez and Feeley, 2020]
 
 **Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
-For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, both measured near ground level (Slot et al. 2019). 
-However, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher climatic heat stress and thus $T_{50}$ was observed in the lower north-facing canopy positions (Curtis et al.
+Thermal tolerance $T_{crit}$ and $T_{50}$ have been predominantly measured on sun leaves; thus there remains limited information on leaves with varying light exposures and vertical height gradient.
+
+For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, though the results weren't significant and both were measured near ground level [Slot et al.
+2019].
+In tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions had higher heat tolerance (by $2{^\circ}C$) than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
+In both species photosynthetic heat tolerance increased from June to August.
+In other cases, leaf thermal tolerance might reflect microclimatic stress along a vertical gradient.
+For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions (Curtis et al.
 2019).
-We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
-I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot
+Greater $T_{50}$ in lower north-position correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [Curtis et al.
+2018].
 
-increased irradiance = increased VPD = increased air temp = midday stomatal closure =  decreased latent heat loss = increase leaf temperature = decreased C assimilation = greater PS2 heat tolerance = optimum rate of photosynthesis decreases = exhibiting lower t opt (optimum temp for c assimilation)
-
-incr.humidity = decreased VPD = dec. latent heat loss = limited c assimilation [@perezPSIIHeatTolerances2020]
+**We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
+I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot**
 
@@
 Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
-- Elsa: and see Perez, T.M., Socha, A., Tserej, O. and Feeley, K.J., 2020. PSII heat tolerances characterize thermal generalists and the upper limit of carbon assimilation. Plant, Cell & Environment.
 

### 2021-04-22 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor.
-Foliage density mediates light regime, where denser upper canopies exposed to greater PAR values limit light to canopy interiors, lower canopy layers, and the understory.
+Foliage LAI and density mediate light regime, where denser upper canopies exposed to greater PAR values limit light to canopy interiors, lower canopy layers, and the understory.
 This light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig.
@@
 In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-On the other hand, open-forests with sparse stands, have lower buffering capacity, greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [Rambo and NOrth, 2009].
+On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [Rambo and NOrth, 2009].
 For this reason, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; Rambo and North, 2009].
-During the night, pine forests have been observed to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
+During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
 Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer.
@@
 However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
-
-Thus, we see that buffering by canopy cover smooths the affect temperature extremes and fluctuation in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory; Defrenne et al.
-2021],though, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed [@bertrandCommentForestMicroclimate2020].
 
@@
 With higher altitude, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions.
-In Switzerland forests, buffering capacity was stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a], partly because trees here were sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] and in other high-altitude forests, trees have open canopies with greater light interception through canopy layers [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
+In Switzerland forests, buffering capacity was stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a], partly because trees in these high-altitude forests are sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] or have open canopies with greater light interception through canopy layers [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
 Less capacity to buffer climatic conditions in open, high altitude forests might have implications to forest regeneration and climate change [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].
 This is also observed in  timberline and temperate humid-mountain forests in China, where tree height decreased with altitude, and tree growth sensitivity (based on tree-rings) increased [@liangForestTypeHeight2019].
-Whereas in closed-forests, increase in tree height, typically increases buffering in the understory [@juckerCanopyStructureTopography2018].
 
@@
 On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ induces evaporative cooling [Drake et al.
-2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al.
+2018] or under high $T_{air}$ and stomtal limitation, can cause leaf heat stress [Fauset et al.
 2018].
-Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore tend to maintain greater $T_{leaf}$ than $T_{air}$ compared to lower canopy or understory, whereas open-forests, with greater light interception to forest floors, in some cases, exhibit a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]
 
@@
 Particularly, in open forests with sparse distribution of trees, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
-Greater heat stress and $T_{air}$ in the lower canopy is also observed in Australian desert trees [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground [@hadleyInfluenceKrummholzMat1987].
+Greater heat stress and $T_{air}$ in the lower canopy closer is also observed in Australian desert trees [Curtis et al.2018] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987].
 Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
@@
 
-In three forest types- US temperate deciduous, Puerto Rico tropical wet and tropical moist, $T_{leaf_{max}}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in the upper canopies of all three forests, along with increasing $T_{leaf}$ - $T_{air}$  with height.
+In three forest types- US temperate deciduous, Puerto Rico tropical wet and tropical moist, $T_{leaf_{max}}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in the upper canopies of all three forests, with increasing $T_{leaf}$ - $T_{air}$  with height.
 $T_{leaf}$ - $T_{air}$ was observed to be greatest in temperate deciduous forest compared to tropical moist and wet.
@@
 
-Within-individuals exposed canopies subjected to harsher light and biophysical conditions, substanially buffer inner and lower canopies.
+Within-individuals exposed substanially buffer inner and lower canopies from light and biophysical conditions.
 Thus leaf traits in canopy positions vary accordingly [Casal et al.
@@
 
-Along the vertical gradient, leaf temperature increases with effective leaf width and leaf size [Fauset et al.
-2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance [@leighInfluenceLeafSize2017; Wright et al.
+Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration--among other traits, decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air, that lowers leaf temperature.
+[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
+Mechanisms such as vertical leaf angles, temporary wilting reduce surface light exposure and help maintain leaf temperature in exposed canopies [Ishida et al.
+1999; Niinemetes, 1998].
+Hydraulic traits, for example leaf vein and stomtal density, help maintain a balance between work and life, photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al.
+2010].
+On the other hand, leaf temperature increases with effective leaf width and leaf size [Fauset et al.
+2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance, and lower convective cooling [@leighInfluenceLeafSize2017; Wright et al.
 2019; @fauset_differences_2018].
-Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves that are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017].
-Exposed leaves reduce light exposures, and thus leaf temperature, by maintaining vertical leaf angles, or through temporary wilting [Ishida et al.
-1999; Niinemetes, 1998].
-Other trait adaptions to dissapting heat such as lobes, serration, longer petiole length and stomtal density are evidently characteristics of sun exposed leaves than shaded leaves, help maintain a lower leaf temperature along a vertical gradient [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]
+Therefore, shade leaves that typically have greater leaf size and effective leaf width compared to sun leaves, remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017].
 
@@
 2010].
-However, increase in temperature in upper canopy, shows a decrease in LMA and $N_{area}$ [Carter and Cavaleri, 2014] and in leaf soluble sugars, starch and non-structural carbons [Li et al.2016 **double check**].
+However, increase in temperature in upper canopy, has been shown to decrease LMA and $N_{area}$ [Carter and Cavaleri, 2014] and leaf soluble sugars, starch and non-structural carbons [Li et al.2016 **double check**].
 Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up [Poorter et al.
@@
 
-Proportional to vertical light gradients at an inter- and intra-species level, sun exposed leaves have greater pools of xanthophyll pigments, comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] than in shaded leaves.
+Proportional to vertical light gradients at an intra- and inter-species level, sun exposed leaves have greater pools of xanthophyll pigments, comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] than in shaded leaves.
 [$VAZ$]--an antioxidant scavenging function--protects against cellular damage by converting excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
@@
 
-Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in fundamental light adapted leaf traits in their biophysical environment as observed and similar to within-species gradient [Rosendaal et al.
+Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment as observed and similar to within-species gradient [Rosendaal et al.
 2006; Thomas and Winner, 2002; Houter and Pons 2012].
@@
 2019 ; @hernandezSimilarTemperatureDependence2020].
-Similar results were found for $V_{max}$ and $J_{max}$ [Slot et al.
-2019; @hernandezSimilarTemperatureDependence2020]and along a vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; Miller et al.
+Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ [Slot et al.
+2019; @hernandezSimilarTemperatureDependence2020],as well in vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; Miller et al.
 2020], and no significant trend in temperate trees [@mau_temperate_2018].
@@
 2017).
-Across species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  Slot et al 2020; O' sullivan et al.
+Across forest species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  Slot et al 2020; O' sullivan et al.
 2017] and varies with leaf traits [Sastry et al.
 2018, @zhang_photosynthetic_2012; Slot et al.
-2020] and their thermal time constant.
+2020].
 Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$.
@@
 2020].
-However, higher photosynthetic heat tolerance does not necessary increase leaf thermal safety margins, rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum temperatures, low optimum temperature for C assimilation [$T_{opt}$] and had narrower thermal safety margins [Perez and Feeley, 2020]
+However, higher photosynthetic heat tolerance does not necessary increase leaf thermal safety margins, rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and had narrower thermal safety margins [Perez and Feeley, 2020]
 
@@
 In both species photosynthetic heat tolerance increased from June to August.
-In other cases, leaf thermal tolerance might reflect microclimatic stress along a vertical gradient.
+
+In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient.
 For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions (Curtis et al.
@@
 Greater $T_{50}$ in lower north-position correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [Curtis et al.
-2018].
+2019].
 
-**We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
-I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot**
+We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
+I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot*
 

### 2021-04-22 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2018].
-Therefore, during growing season, nitrogen per area [$N_{area}$], along with LMA, leaf soluble sugars, starch and non-structural carbons increase with height and light, typically greater in exposed canopy positions or sun leaves compared to shade [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Liu et al.
-2010].
-However, increase in temperature in upper canopy, has been shown to decrease LMA and $N_{area}$ [Carter and Cavaleri, 2014] and leaf soluble sugars, starch and non-structural carbons [Li et al.2016 **double check**].
-Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up [Poorter et al.
+Therefore, during growing season, nitrogen per area [$N_{area}$], along with LMA, leaf soluble sugars, starch and non-structural carbons increase with height and light, typically greater in exposed canopy positions or sun leaves compared to shade [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
+2014].
+Experiemental increase in upper canopy temperature, shows that LMA and $N_{area}$ tend to decrease [Carter and Cavaleri, 2014].
+Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up and thus shows variable results [Poorter et al.
 2019]
@@
 **Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
-(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. Conflicting $Q_10$ findings shows no definite trend.
-Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). 
+
+While dark and light respiration tends to increase with light exposure, and therefore along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration, $Q_{10}$, shows conflicting findings with no definite trend.
+
+In two forests, north-eastern oak dominated-deciduous forest in the US and conifer-dominated temperate New Zealand rainforest, across species $Q_{10}$ was found to be greater in upper-canopy leaves than in lower-canopy leaves, where values differed between species [Turnbull et al.
+2003].
+Similarly, in sweetgum (*Liquidambara styraciflua*), upper canopy sun leaves showed greater $Q_{10}$ than lower within-canopy shade leaves [Harley et al.
+1996].
+However, in other feild studies, $Q_{10}$ didnot vary across species vertical profiles and light exposure such as in a lowland tropical Australian forest study, and across temperate trees at Coweeta [@bolstad_foliar_1999] $Q_{10}$.
 At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999].
 (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
-May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
+May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*).
 Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig.
-1*)."Higher $Q_{10}$ values in the functional groups of upper canopy where temperatures are the highest may lead to exponential loses of C with increasing global temperatures depending upon the ability of canopy leaves to acclimate" [@cavaleri_foliar_2008]
-Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
+1*).
+@bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
 

### 2021-04-23 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-Maintaining optimum leaf temperature in relation to the biophysical environment, maximizes carbon assimilation by maintaining metabolic optima, in a given circumstance.
+Maintaining optimum leaf temperature for metabolic optima in relation to the biophysical environment, maximizes carbon assimilation in a given circumstance.
 This is done through the adaptation of leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
@@
 
-Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration--among other traits, decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air, that lowers leaf temperature.
+Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration--among other traits, decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air.
 [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
-Mechanisms such as vertical leaf angles, temporary wilting reduce surface light exposure and help maintain leaf temperature in exposed canopies [Ishida et al.
+Vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al.
 1999; Niinemetes, 1998].
-Hydraulic traits, for example leaf vein and stomtal density, help maintain a balance between work and life, photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al.
+Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance between work and life, photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al.
 2010].

### 2021-04-26 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2020].
-However, higher photosynthetic heat tolerance does not necessary increase leaf thermal safety margins, rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and had narrower thermal safety margins [Perez and Feeley, 2020]
+However, higher photosynthetic heat tolerance does not necessary increase leaf thermal safety margins.
+Rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had narrower thermal safety margins [Perez and Feeley, 2020.]
 
@@
 
-While dark and light respiration tends to increase with light exposure, and therefore along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration, $Q_{10}$, shows conflicting findings with no definite trend.
+While dark and light respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration, $Q_{10}$, shows conflicting findings with no definite trend.
 

### 2021-04-26 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019].
-Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
+The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019].
+Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
 2021; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009].
-In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [Rambo and NOrth, 2009].
-For this reason, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; Rambo and North, 2009].
+In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [Rambo and North, 2009].
+Therefore, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; Rambo and North, 2009].
 During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
@@
 
-Within-individuals exposed substanially buffer inner and lower canopies from light and biophysical conditions.
+Within-individuals exposed canopies substanially buffer inner and lower canopies from light and biophysical conditions.
 Thus leaf traits in canopy positions vary accordingly [Casal et al.
@@
 
-Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration--among other traits, decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air.
+Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration-- decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air.
 [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
@@
 1999; Niinemetes, 1998].
-Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance between work and life, photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al.
+Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al.
 2010].
@@
 Canopy nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks, and could mediate sun vs.
-shade leaf response to heatflecks [Kitao et al.
+shade leaf response to heatflecks under available soil nitrogen content [Kitao et al.
 2018].
@@
 2014].
-Experiemental increase in upper canopy temperature, shows that LMA and $N_{area}$ tend to decrease [Carter and Cavaleri, 2014].
+Experiemental increase in upper canopy temperature, shows that LMA and $N_{area}$ tend to decrease [Carter and Cavaleri, 2014] due to increase in respiration **?** [**REF**].
 Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up and thus shows variable results [Poorter et al.
@@
 
-Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment as observed and similar to within-species gradient [Rosendaal et al.
+Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [Rosendaal et al.
 2006; Thomas and Winner, 2002; Houter and Pons 2012].
-Under constant light conditions of open-grown canopy and understory trees, for example in large gaps, ontogeny differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [Thomas and Winner, 2002; Houter and Pons 2012].
+Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [Thomas and Winner, 2002; Houter and Pons 2012].
 Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels influence leaf temperatures.
@@
 In tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions had higher heat tolerance (by $2{^\circ}C$) than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
-In both species photosynthetic heat tolerance increased from June to August.
+In both species photosynthetic heat tolerance increased from June to August [Knapp et al.
+1994]
 
@@
 1996].
-However, in other feild studies, $Q_{10}$ didnot vary across species vertical profiles and light exposure such as in a lowland tropical Australian forest study, and across temperate trees at Coweeta [@bolstad_foliar_1999] $Q_{10}$.
+However, in other feild studies, $Q_{10}$ didnot vary across species vertical profiles and light exposure such as in a lowland tropical Australian forest study [Weerasinghe et al.
+2014], and across temperate trees at Coweeta [@bolstad_foliar_1999] $Q_{10}$.
 At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999].
@@
 On the basis of phenology, temperate understory  trees tend to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al.
-2013], and, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renner, 2019].
-However, short-term May spring heat waves have shown to negetively impact partially expanded canopy leaves by inducing leaf shedding, more than understory saplings [Filewood and Thomas, 2013].
+2013], and, increase in annual temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renner, 2019].
+However, short-term May spring heat waves have shown to negetively impact partially-expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than in understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].
 

### 2021-04-26 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019].
-Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al.
-2021; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009].
+Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
+2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009].
 In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [Rambo and North, 2009].
-Therefore, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; Rambo and North, 2009].
+On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009].
+Therefore, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009].
 During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
@@
 Under colder conditions, minimum $T_{air}$ [$T_{min}$] is also buffered by forest canopies, increasing negetive temperature offset in the understory.
-Across Europe [@zellweger_seasonal_2019; Defrenne et al.
-2021] and in the northwestern United States [@davis_microclimatic_2019] higher minimum temperatures under forest canopies have been observed relative to nearby clearings.
+Across Europe [@zellweger_seasonal_2019; *Defrenne et al.
+2021*] and in the northwestern United States [@davis_microclimatic_2019] higher minimum temperatures under forest canopies have been observed relative to nearby clearings.
 Greater radiative heat loss in exposed areas results in lower $T_{min}$.
@@
 Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$.
-On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ induces evaporative cooling [Drake et al.
-2018] or under high $T_{air}$ and stomtal limitation, can cause leaf heat stress [Fauset et al.
-2018].
+On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ induces evaporative cooling [@cavaleriColdbloodedForestsWarming2020] or under high $T_{air}$ and stomtal limitation, can cause leaf heat stress [@fauset_differences_2018].
 
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
-Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; *Koch et al.
-1994*].
+Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
 Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*.
@@
 However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.
-This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [*Koch et al.
-1994*; @fauset_differences_2018].
+This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [@kochDiurnalPatternsLeaf1994; @fauset_differences_2018].
 On the other hand, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
@@
 Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
-In other dry conditions, inner canopy can be warmer than above canopy and air temperature due to greater relative humidity [Dietz et al.
-2007].
+In other dry conditions, inner canopy can be warmer than above canopy and air temperature due to greater relative humidity [@dietzVerticalPatternsDuration2007].
 Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
@@
 2019].
-This is seen in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; Miller; ], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
+This is seen in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
 
@@
 $T_{leaf}$ - $T_{air}$ was observed to be greatest in temperate deciduous forest compared to tropical moist and wet.
-Greater light input in exposed upper canopy leaves can sometimes elevate leaf temperature $1-7{^\circ}C$ above air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [Doughty and Gouden, 2008; @mauTemperateTropicalForest2018a]
+Greater light input in exposed upper canopy leaves can sometimes elevate leaf temperature $1-7{^\circ}C$ above air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]
 
@@
 **Leaf thermoregulation largely depends on leaf traits.
-Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light and water demand, in order to optimize C uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [michaletzPlantThermoregulationEnergetics2015; Zweiniecki et al 2004]**
+Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light and water demand, in order to optimize C uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
 
@@
 
-Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [Zweineicki et al.
-2004].
+Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004].
 Degrees of direct to diffused light and consequental water demands shape sun vs.
-shade leaves and their necessary vascular architecture [Casal et al.
-2011, Zweineicki et al.
-2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+shade leaves and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
 Therefore, exposed sun leaves are typically smaller in size and width, thicker, hairier and have greater lobation, and buffer biophysical conditions for shade leaves.
-Whereas shade leaves adapt to diffused light and stable microenvironment through investing in light capture efficiency by expanding specific leaf area (SLA) [Sack et al.
-2006; Chazdon et al.
-1996, Poorter et al.
-2000; @rozendaalPlasticityLeafTraits2006; Mathur et al.
-2018].
-Respective leaf traits also play a role in heat dissapation where exposed sun leaves are more efficient in dissapting heat than buffered shade leaves [Vogel].
+Whereas shade leaves adapt to diffused light and stable microenvironment through investing in light capture efficiency by expanding specific leaf area (SLA) [@zwienieckiHydraulicLimitationsImposed2004; @rozendaalPlasticityLeafTraits2006; @mathurPhotosyntheticEfficiencySun2018a].
+Respective leaf traits also play a role in heat dissapation where exposed sun leaves are more efficient in dissapting heat than buffered shade leaves [@vogelConvectiveCoolingLow1970].
 
@@
 Within-individuals exposed canopies substanially buffer inner and lower canopies from light and biophysical conditions.
-Thus leaf traits in canopy positions vary accordingly [Casal et al.
-2011].
+Thus leaf traits in canopy positions vary accordingly [@casasSunShadeLeaves2011].
 Several studies point towards common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
@@
 2019], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
-Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016].
+Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016].
 However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018)
@@
 Vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al.
-1999; Niinemetes, 1998].
+1999; @niinemetsAdjustmentFoliageStructure1998].
 Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al.
@@
 Canopy nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks, and could mediate sun vs.
-shade leaf response to heatflecks under available soil nitrogen content [Kitao et al.
-2018].
+shade leaf response to heatflecks under available soil nitrogen content [@kitaoCanopyNitrogenDistribution2018].
 Therefore, during growing season, nitrogen per area [$N_{area}$], along with LMA, leaf soluble sugars, starch and non-structural carbons increase with height and light, typically greater in exposed canopy positions or sun leaves compared to shade [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
 2014].
-Experiemental increase in upper canopy temperature, shows that LMA and $N_{area}$ tend to decrease [Carter and Cavaleri, 2014] due to increase in respiration **?** [**REF**].
-Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up and thus shows variable results [Poorter et al.
-2019]
+Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up and thus shows variable results [@poorterMetaanalysisPlantResponses2019a]
 
@@
 
-Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [Rosendaal et al.
-2006; Thomas and Winner, 2002; Houter and Pons 2012].
-Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [Thomas and Winner, 2002; Houter and Pons 2012].
+Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [@@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [@@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
 Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels influence leaf temperatures.
-Therefore, some species might more susceptible to heat stress at a vertical gradient than others [Fauset et al.
-2018].
+Therefore, some species might more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
 
@@
 While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well.
-Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [Kenzo et al.
-2015; Sanches et al.
-2009].
-During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [Sanches et al.
-2009; Zweinniecki et al.
-2004].
-Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [Sanches et al.
-2009; Hernandez].
-Similarly, max transpiration rate increases with canopy height [Kenzo et al.
-2015].
-Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [Kosugi et al.
-2021].
+Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
+During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004].
+Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
+Similarly, max transpiration rate increases with canopy height [@kosugiSeasonalFluctuationsTemperature2006].
+Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].
 
@@
 Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
-2017; Slot & Winter 2017).
+2017; @slotSituTemperatureResponse2017).
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g.
-Slot & Winter 2017; Smith et al.
+@slotSituTemperatureResponse2017; Smith et al.
 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs.
@@
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [Niinemetes et al.
-1999b; Niinemetes and Valladares, 2004].
+14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
@@
 2019 ; @hernandezSimilarTemperatureDependence2020].
-Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ [Slot et al.
-2019; @hernandezSimilarTemperatureDependence2020],as well in vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; Miller et al.
-2020], and no significant trend in temperate trees [@mau_temperate_2018].
+Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020],as well in vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and no significant trend in temperate trees [@mau_temperate_2018].
 
@@
 2017).
-Across forest species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  Slot et al 2020; O' sullivan et al.
-2017] and varies with leaf traits [Sastry et al.
-2018, @zhang_photosynthetic_2012; Slot et al.
-2020].
+Across forest species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017] and varies with leaf traits [Sastry et al.
+2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020].
 Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$.
-On the other hand, leaves of species with greater construction costs, observed through LMA, had greater $T_{50}$ compared to species with low LMA [Slot et al.
-2020].
-However, higher photosynthetic heat tolerance does not necessary increase leaf thermal safety margins.
-Rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had narrower thermal safety margins [Perez and Feeley, 2020.]
+On the other hand, leaves of species with greater construction costs, observed through LMA, had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020].
+However, higher photosynthetic heat tolerance does not necessary increase the breadth of leaf thermal safety margins.
+Rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$) [@perezPhotosyntheticHeatTolerances2020]
 
@@
 
-For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, though the results weren't significant and both were measured near ground level [Slot et al.
-2019].
+For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, though the results weren't significant and both were measured near ground level [@slotPhotosyntheticHeatTolerance2019a].
 In tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions had higher heat tolerance (by $2{^\circ}C$) than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
@@
 2019).
-Greater $T_{50}$ in lower north-position correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [Curtis et al.
-2019].
+Greater $T_{50}$ in lower north-position correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [@curtis_intracanopy_2019].
 
@@
 
-In two forests, north-eastern oak dominated-deciduous forest in the US and conifer-dominated temperate New Zealand rainforest, across species $Q_{10}$ was found to be greater in upper-canopy leaves than in lower-canopy leaves, where values differed between species [Turnbull et al.
-2003].
-Similarly, in sweetgum (*Liquidambara styraciflua*), upper canopy sun leaves showed greater $Q_{10}$ than lower within-canopy shade leaves [Harley et al.
-1996].
-However, in other feild studies, $Q_{10}$ didnot vary across species vertical profiles and light exposure such as in a lowland tropical Australian forest study [Weerasinghe et al.
-2014], and across temperate trees at Coweeta [@bolstad_foliar_1999] $Q_{10}$.
+In two forests, north-eastern oak dominated-deciduous forest in the US and conifer-dominated temperate New Zealand rainforest, across species $Q_{10}$ was found to be greater in upper-canopy leaves than in lower-canopy leaves, where values differed between species [@turnbullScalingFoliarRespiration2003].
+Similarly, in sweetgum (*Liquidambara styraciflua*), upper canopy sun leaves showed greater $Q_{10}$ than lower within-canopy shade leaves [@harleyEffectsLightTemperature1996].
+However, in other feild studies, $Q_{10}$ didnot vary across species vertical profiles and light exposure such as in a lowland tropical Australian forest study [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta [@bolstad_foliar_1999] $Q_{10}$.
 At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999].
@@
 With temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum.
-Beyond this inflection point, fluxnet global data analysis shows that photosynthesis declines whereas respiration increases exponentially, therefore, moving from carbon sink to carbon source with increasing temperature, independent of water and irradiance [Duffy et al.2021]
+Beyond this inflection point, fluxnet global data analysis shows that photosynthesis declines whereas respiration increases exponentially, therefore, moving from carbon sink to carbon source with increasing temperature, independent of water and irradiance [@duffyHowCloseAre2021]
 

### 2021-05-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009].
-In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+Similarly, when compared to neighboring tree crop agricultural plantations, intact tropical forests maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
 On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009].
@@
 While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well.
-Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
+During the day, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
 During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004].

### 2021-05-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 nocite: | 
-  @bolstad_foliar_1999; @roberts_stomatal_1990; @slot_photosynthetic_2019; @harleyEffectsLightTemperature1996; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @ishiiExploringRelationshipsCanopy2004; @curtis_intracanopy_2019; @bolstad_foliar_1999; @kafutiFoliarWoodTraits2020;
+ @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007;Carter and Cavaleri 2018; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; Hamerlynck and Knapp 1994; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al.
+1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; Harris and Medina 2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @athertonSpatialVariationLeaf2017
 ---

### 2021-05-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 nocite: |
-  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @kusiPlasticLeafMorphology2020
+@mau_temperate_2018; @cobleLightDrivesVertical2014; @@sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; Liakoura 1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; Ishida et al.
+1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; Koniger et al.
+1995; Mastubara et al.
+2009; Harris and Medina 2013; Hansen et al.
+2001; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000
 ---

### 2021-05-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 nocite: |
-@mau_temperate_2018; @cobleLightDrivesVertical2014; @@sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; Liakoura 1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; Ishida et al.
+@mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; Liakoura 1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; Ishida et al.
 1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; Koniger et al.

### 2021-05-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-05-03 vertical-thermal-review.Rmd
@@
 nocite: |
-  @mau_temperate_2018; @hernandezSimilarTemperatureDependence2020; @cobleHowVerticalPatterns2016a; @slotPhotosyntheticHeatTolerance2019; @valladaresShadeToleranceKey2008b; @beaumontVerticalGradientsLeaf2009; @sackHowStrongIntracanopy2006,@cobleLightDrivesVertical2014; @scartazzaInvestigatingEuropeanBeech2016; @martinCovarianceSunShade2020; @bolstad_foliar_1999; @harleyEffectsLightTemperature1996; @kullDistributionLeafPhotosynthetic1998; @kafutiFoliarWoodTraits2020; @duursmaVerticalCanopyGradients2006; @cobleLightDrivesVertical2014a; @vanwittenbergheVariabilityStomatalConductance2012; @poorterLeafOpticalProperties1995; @poorterMetaanalysisPlantResponses2019b; @cavaleri_foliar_2008; @niinemetsAcclimationHighIrradiance1998; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @kusiPlasticLeafMorphology2020
+@mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; Ishida et al.
+1999; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; Harris and Medina 2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000
 ---

### 2021-05-04 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019].
-Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
+Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively relative to pine and boreal forests, and compared to nearby clearings.
+These forests maintain cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively.
+[@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009].
 Similarly, when compared to neighboring tree crop agricultural plantations, intact tropical forests maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009].
-Therefore, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009].
-During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
-Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer.
-As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory.
-However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].
+On the other hand, open-forests with sparse stands, have lower buffering capacity with greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009].
+Therefore, they maintain higher daytime maximum temperature and light availability in the understory that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009].
+During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil from daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
+Buffering is also affected by Canopy roughness, turbulent air flow and the canopy boundary layer.
+As a result, taller trees upto 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory.
+However, beyond a certain tree height, temperature and VPD offset plateaud [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].
 
@@
 2017], have all shown to offset temperatures positively in the understory, with lower maximum air temperatures and higher relative humidity below canopies and understories than above forest canopies and gaps.
-During the dry season of a semi-deciduous tropical forest in Panama, however, maximum temperatures were similar in the upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
+However, during the dry season of a semi-deciduous tropical forest in Panama, maximum temperatures were similar in the upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
 Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
@@
 With higher altitude, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions.
-In Switzerland forests, buffering capacity was stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a], partly because trees in these high-altitude forests are sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] or have open canopies with greater light interception through canopy layers [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
+In Switzerland forests, buffering capacity has shown to be stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a], partly because trees in these high-altitude forests are sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] or have open canopies with greater light interception through canopy layers [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
 Less capacity to buffer climatic conditions in open, high altitude forests might have implications to forest regeneration and climate change [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].
@@
 **Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-Maintaining optimum leaf temperature for metabolic optima in relation to the biophysical environment, maximizes carbon assimilation in a given circumstance.
-This is done through the adaptation of leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+For leaves, maintaining optimum leaf temperature for metabolic optima in relation to the biophysical environment, maximizes carbon assimilation in a given circumstance.
+This is done through the adaptation of leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 `r fig_leaf_T`).
@@
 `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
+When leaves are cooler than air at high temperature and warmer than air at low temperature, they considered to exhibit limited homeothermy.
+When leaves are drastically warmer than $T_{air}$ such as in the upper canopy, this is considered megathermy.
+In some cases, $T_{leaf}$ can be equalled to $T_{air}$, showing poikiolothermic behavior [Cavaleri, 2020].
 
@@
 Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$.
-On the other hand, under conditions of constant radiation and water availability, increase in $T_{air}$ induces evaporative cooling [@cavaleriColdbloodedForestsWarming2020] or under high $T_{air}$ and stomtal limitation, can cause leaf heat stress [@fauset_differences_2018].
+On the other hand, under conditions of constant radiation and water availability, evaporative cooling sustains $T_{leaf}$ under increased $T_{air}$ [@cavaleriColdbloodedForestsWarming2020] until stomtal limitation impedes cooling, which causes leaf heat stress [@fauset_differences_2018].
 
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
-Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
+Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
 Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*.
@@
 `r fig_leaf_T`).**
+Canopy buffering capacity, as well vertical biophysical gradient, affects vertical distribution of leaf temperature.
 While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
 
-Particularly, in open forests with sparse distribution of trees, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
-Greater heat stress and $T_{air}$ in the lower canopy closer is also observed in Australian desert trees [Curtis et al.2018] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987].
+Particularly, in open forests with sparse distribution of trees and lower buffering capacity, as observed in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+Greater heat stress and $T_{air}$ in the lower canopy is also observed in open Australian desert trees [Curtis et al.2018] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987], this is analogous to greenhouse effect caused under dense lower canopies.
 Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
-In other dry conditions, inner canopy can be warmer than above canopy and air temperature due to greater relative humidity [@dietzVerticalPatternsDuration2007].
+In other tropical dry conditions, inner canopy can be warmer than above canopy and air temperature due to conditions of greater relative humidity [@dietzVerticalPatternsDuration2007].
 Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
@@
 Canopy nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks, and could mediate sun vs.
-shade leaf response to heatflecks under available soil nitrogen content [@kitaoCanopyNitrogenDistribution2018].
+shade leaf response to heatflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018].
 Therefore, during growing season, nitrogen per area [$N_{area}$], along with LMA, leaf soluble sugars, starch and non-structural carbons increase with height and light, typically greater in exposed canopy positions or sun leaves compared to shade [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
@@
 
-Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [@@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [@@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels influence leaf temperatures.
-Therefore, some species might more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
+Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels and therefore influence leaf temperatures.As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
 
@@
 
-Typical values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al.
+Across forest species, typical values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al.
 2017).
-Across forest species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017] and varies with leaf traits [Sastry et al.
+$T_{50}$ and $T_{crit}$ decreases with elevation, as seen in tropical forests in Panama and Columbia and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017] and varies with leaf traits [Sastry et al.
 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020].
@@
 In tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions had higher heat tolerance (by $2{^\circ}C$) than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
-In both species photosynthetic heat tolerance increased from June to August [Knapp et al.
+In both species photosynthetic heat tolerance increased seasonally from June to August [Knapp et al.
 1994]
@@
 In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient.
-For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions (Curtis et al.
-2019).
-Greater $T_{50}$ in lower north-position correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [@curtis_intracanopy_2019].
+For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions.
+Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [@curtis_intracanopy_2019].
 
@@
 
-While dark and light respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration, $Q_{10}$, shows conflicting findings with no definite trend.
+While dark and light respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration--$Q_{10}$--shows conflicting findings with no definite trend.
 

### 2021-05-04 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
+*from Martijn in Tleaves figure: One thing that might be more relevant in tropical than in seasonal temperate forests, is the issue I mentioned before about the difference between shade leaves of canopy trees and leaves of understory species.
+The understory species in the tropics (the ones that won't necessarily make it to the canopy, and can reproduce in the understory) appear to invest more in leaf longevity than in light harvesting per se, so even though they are in the shade, they do not have large leaves with low LMA.
+This would be a deviation on, or a special case of, the understory parameters you are using now.
+I don't think we necessarily need to incorporate that in this figure, but I think it would be useful to be able to say that for understory species with small, high LMA leaves the effects of windspeed, radiation, RH, etc, are greater/smaller than for the understory species parameters in the figure.
+DOes this make sense?
+Is this relevant at all for seasonal forests?
+The only thing that comes to mind is Ilex, but maybe there are other species?*
+
  *(KAT: a lot of this content should probably move up the the intro to this section)*

### 2021-05-04 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -4,7 +4,7 @@
 # Tleaves R package created by Chris Muir, 2019
 # Developed by: Nidhi Vinod, contact Nidhi Vinod (nidrup@gmail.com)
 # R version 3.5.3 --First created in October 2020
-# Sept-March. 2020
+# Sept-August 2021
 ####################################
 #install packages if needed
 library(magrittr)
@@ -18,10 +18,10 @@ library(cowplot)
 
 ####################################
 #create data frames using HARV NEON data for environmental parameters. Leaf characteristic dimmension in m is measured from Sun and shade Red oak leaves
-#all variables constant except for one independent variable that represents minimum- maximum range
-#Stomtal conductane(gs) value magnitudes obtained from Tleaves supplementary manual for range of values
+#all variables are constant except for the independent variable that represents minimum- maximum range
 #normal represents HARV neon environmental parameters as observed
-#drought represents low gs and low relative humidity
+#Stomtal conductane(gs) values for typical condition obtained from Tleaves supplementary manual for range of minimum and maximum values
+#drought condition is representative of gs values from Cavendar-Bares and Bazzaz, 2000 and typically low relative humidity
 ####################################
 
 # Create a dataframe for overstory and understory, for scenarios: normal and drought

### 2021-05-04 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -21,7 +21,7 @@ library(cowplot)
 #all variables are constant except for the independent variable that represents minimum- maximum range
 #normal represents HARV neon environmental parameters as observed
 #Stomtal conductane(gs) values for typical condition obtained from Tleaves supplementary manual for range of minimum and maximum values
-#drought condition is representative of gs values from Cavendar-Bares and Bazzaz, 2000 and typically low relative humidity
+#drought condition is representative of gs values from Cavendar-Bares and Bazzaz, 2000 and typically low relative humidity, with other variables similar to normal condition
 ####################################
 
 # Create a dataframe for overstory and understory, for scenarios: normal and drought

### 2021-05-04 vertical-thermal-review_SI.Rmd
leaf T figure
#6
Update vertical-thermal-review_SI.Rmd

@@
 
+see [Issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) , [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)
+
 \newpage
@@
 METHODS ON TEALEAVES GO HERE.
+
+Energy balance for typical overstory sun leaf and understory shade leaf were parameterized with biophysical variables from Harvard NEON data documented here: (https://github.com/EcoClimLab/vertical-thermal-review/blob/master/NEON_height_profiles/HARV_neon_stats.csv).
+Leaf trait measurements for effective leaf width was measured on ~Quercus rubra~ sun and shade leaf, stomatal conductance measurements were refered from Tleaves typical sun and shade measurements [@muir_tealeaves_2019] and as well @cavender-baresChangesDroughtResponse2000.
+Energy balance for biophysical variables in drought scenario were parameterized using Harvard NEON data as well.
+Leaf trait measurements for effective leaf width is the same as typical scenario, stomatal conductance was referred from @cavender-baresChangesDroughtResponse2000.
+In each variable visual, all variables are constant except for the independent variable that represents minimum - maximum range.
 

### 2021-05-04 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -368,7 +368,7 @@ c1<-ggplot(gs)+
   geom_smooth(aes(x = gs, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = gs, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = gs, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
-  ylab("TLeaf - Tair*")+xlab("stomal conductance")
+  ylab("TLeaf - Tair*")+xlab("stomatal conductance")
 c1
 
 
@@ -575,8 +575,8 @@ annotate_figure(
   bottom = text_grob("constants from harvard NEON data 
                     \n overstory- sw: 871, tair: 298.1 [k], ws: 2.88, gs: 4.0 µmol m-2 s-1 Pa-1, leaf size: 0.04m, rh: 0.91, 
                      \n understory- sw: 102, tair: 296.1 [k], ws: 0.24, gs: 2.0 µmol m-2 s-1 Pa-1, leaf size: 0.10m, rh: 0.97
-                     \n drought overstory- sw: 871, tair: 298.1 [k], ws: 2.88, gs: 2.5 µmol m-2 s-1 Pa-1, leaf size: 0.04m, rh: 0.31, 
-                     \n drought understory- sw: 102, tair: 296.1 [k], ws: 0.24, gs: 1.0 µmol m-2 s-1 Pa-1, leaf size: 0.10m, rh: 0.37 
+                     \n drought overstory (dotted line)- sw: 871, tair: 298.1 [k], ws: 2.88, gs: 2.5 µmol m-2 s-1 Pa-1, leaf size: 0.04m, rh: 0.31, 
+                     \n drought understory (dotted line)- sw: 102, tair: 296.1 [k], ws: 0.24, gs: 1.0 µmol m-2 s-1 Pa-1, leaf size: 0.10m, rh: 0.37 
                      \n *TLeaf - Tair not standardized", 
                      lineheight = 0.7, color = "black", size = 8))
 

### 2021-05-05 vertical-thermal-review.Rmd
small edits

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al.
+Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al.
 2014*].
@@
 
-**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
+**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
 Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
@@
 
-The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019].
-Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively relative to pine and boreal forests, and compared to nearby clearings.
-These forests maintain cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively.
-[@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
+The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davisMicroclimaticBufferingForests2019].
+Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively where they maintain cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relatively more than pine and boreal forests, and nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009].
@@
 Therefore, they maintain higher daytime maximum temperature and light availability in the understory that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009].
-During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil from daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
-Buffering is also affected by Canopy roughness, turbulent air flow and the canopy boundary layer.
+During the night, pine forests can maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
+
+Buffering is also affected by canopy roughness, turbulent air flow and the canopy boundary layer.
 As a result, taller trees upto 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory.
@@
 The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
-Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davis_microclimatic_2019; @macekMaximumAirTemperature2019]
+Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
 
@@
 **Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-For leaves, maintaining optimum leaf temperature for metabolic optima in relation to the biophysical environment, maximizes carbon assimilation in a given circumstance.
+Leaves adapt to their biophysical environment.
+Maintaining optimum leaf temperature for metabolic optima in relation to their biophysical environment maximizes carbon assimilation in a given circumstance.
 This is done through the adaptation of leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
@@
 When leaves are cooler than air at high temperature and warmer than air at low temperature, they considered to exhibit limited homeothermy.
-When leaves are drastically warmer than $T_{air}$ such as in the upper canopy, this is considered megathermy.
-In some cases, $T_{leaf}$ can be equalled to $T_{air}$, showing poikiolothermic behavior [Cavaleri, 2020].
+When leaves are drastically warmer than $T_{air}$ such as sometimes in the upper canopy, this is considered megathermy.
+In some cases, $T_{leaf}$ can be equalled to $T_{air}$, showing poikiolothermic behavior [Cavaleri, 2020; @drakeNoEvidenceHomeostatic2020].
 
@@
 Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$.
-On the other hand, under conditions of constant radiation and water availability, evaporative cooling sustains $T_{leaf}$ under increased $T_{air}$ [@cavaleriColdbloodedForestsWarming2020] until stomtal limitation impedes cooling, which causes leaf heat stress [@fauset_differences_2018].
+On the other hand, under conditions of constant radiation and water availability, evaporative cooling sustains $T_{leaf}$ under increased $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
 
@@
 Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*.
-Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al.
+Under hot conditions, leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al.
 2018].
-However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.
-This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [@kochDiurnalPatternsLeaf1994; @fauset_differences_2018].
-On the other hand, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
+However, during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.
+This is caused by increase in air temperature and VPD driving greater leaf water demand than water transport, which limits stomatal conductance along with decreased leaf water content, therefore limiting heat dissapation [@kochDiurnalPatternsLeaf1994; @fauset_differences_2018].
+On the other hand, buffered leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
 
@@
 `r fig_leaf_T`).**
-Canopy buffering capacity, as well vertical biophysical gradient, affects vertical distribution of leaf temperature.
+Canopy buffering capacity, as well vertical biophysical gradient, affect vertical distribution of leaf temperature.
 While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
@@
 Particularly, in open forests with sparse distribution of trees and lower buffering capacity, as observed in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
-Greater heat stress and $T_{air}$ in the lower canopy is also observed in open Australian desert trees [Curtis et al.2018] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987], this is analogous to greenhouse effect caused under dense lower canopies.
+Greater heat stress and $T_{air}$ in the lower canopy is also observed in open Australian desert trees [@curtis_intracanopy_2019] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987], this is analogous to greenhouse effect caused under dense lower canopies.
 Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
@@
 Greater light input in exposed upper canopy leaves can sometimes elevate leaf temperature $1-7{^\circ}C$ above air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]
+
+**Stability of $T_{leaf}$ varies with fluctuation in biophysical conditions**
+Biophysical environment fluctuation dictates how stable $T_{leaf}$ is, where stable biophysical conditions, for eg.
+low wind speed and consistent $T_{air}$ can prolong $T_{leaf}$, consistently reducing stomatal conductance and photosynthesis while increasing $T_{leaf}$, and therefore reducing chances for leaf recovery.
+However, increase in biophysical fluctuations make $T_{leaf}$ unstable as seen in upper canopies where with available water unstable $T_{leaf}$ can be more manageable to peak irradiance and $T_{air}$ than in stable buffered canopy regions [**REF**]
 
@@
 
-- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues.
-I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
-
 *(The following notes need to be condensed/ worked into the paragraph above or discussion of specific traits.
@@
 
-An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient [Chen et al.
-2019], because LES theory was developed using mainly sun exposed (upper canopy) leaves.
-Given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016].
-However, there is also concern for ambiguity in a simplified categorization of 'sun' and 'shade' leaves, due to variation based on within-canopy light, across functional groups, in canopy gaps, and with height and ontogeny (Niinemets et al., 2015; Ishii et al., 2018)
+Additionally, given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016], such as by incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient, partly because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020].
 
@@
 
-Along the vertical gradient, exposed leaves are better equipped to dissapate heat, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration-- decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air.
+Along the vertical gradient, exposed leaves are better equipped to dissapate heat than shaded leaves, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration-- decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air.
 [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
-Vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al.
+Mechanisms such vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al.
 1999; @niinemetsAdjustmentFoliageStructure1998].
@@
 2010].
-On the other hand, leaf temperature increases with effective leaf width and leaf size [Fauset et al.
-2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance, and lower convective cooling [@leighInfluenceLeafSize2017; Wright et al.
+On the other hand, leaf temperature increases with effective leaf width and leaf size [@fauset_differences_2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance, and lower convective cooling [@leighInfluenceLeafSize2017; Wright et al.
 2019; @fauset_differences_2018].
 Therefore, shade leaves that typically have greater leaf size and effective leaf width compared to sun leaves, remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017].
+Additionally, understory leaves tend to open their stomata wider (stomatal apeture) than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].
 
@@
 
-Ontogenetic differences also suggests shade tolerators have lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities than in intolerators [@niinemetsControversyTraitsConferring2006].
-While shade intolerant species experience greater heterogeneity in light during ontogeny [poorter et al.
-2013].
-However, these findings conflict with  Kitajima 1994 who found for tropical seedlings; shade intolerance species grew faster in the sun and in the shade, but at the cost of higher mortality in the shade.
-Lawren and Karoru had contrasting findinds, and so far appears that there is no a consensus?
-How will leaf temperature vary between the two?
-
 *from Martijn in Tleaves figure: One thing that might be more relevant in tropical than in seasonal temperate forests, is the issue I mentioned before about the difference between shade leaves of canopy trees and leaves of understory species.
@@
 While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well.
-During the day, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
-During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004].
-Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
-Similarly, max transpiration rate increases with canopy height [@kosugiSeasonalFluctuationsTemperature2006].
-Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].
+During the day, max transpiration rate and stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
+During midday, stomtal closure occurs when evaporative demand exceeds water transport capacity in exposed leaves thereby increasing $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004].
+Thus, stomatal limitation is greater and steeper in exposed canopies than shaded leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
 
@@
 In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma.
-Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P.
-(and water availability)
+Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P (and water availability).
 
-Leaf temperature beyond photosynthetic optimum temperature, declines photosynthesis [**REF**].
+Leaf temperature beyond photosynthetic optimum temperature declines photosynthesis [**REF**].
 Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
@@
 
-While dark and light respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration--$Q_{10}$--shows conflicting findings with no definite trend.
+While respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration--$Q_{10}$--shows conflicting findings with no definite trend.
 

### 2021-05-05 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R

@@ -161,7 +161,7 @@ a1 <-ggplot(data = sw)+
   scale_linetype_manual(name = "canopy position",
                         labels = c("overstory", "understory", "drought overstory", "drought understory"),
                         values = c( 1, 1, 2, 2), guide = "legend")+
-  ylab("TLeaf - Tair*")+xlab("short wave radiation")
+  ylab("TLeaf - Tair [k]*")+xlab("short wave radiation (W/m^2)")
 
 a1 
 
@@ -272,7 +272,7 @@ b1<-ggplot(ws)+
   geom_smooth(aes(x = wind, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = wind, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = wind, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
-  ylab("TLeaf - Tair*")+xlab("windspeed")
+  ylab("TLeaf - Tair [k]*")+xlab("windspeed (ms-1)")
 
 b1
 
@@ -310,7 +310,7 @@ b<-ggplot(ws)+
   geom_smooth(aes(x = wind, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = wind, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = wind, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
-  ylab("TLeaf - Tair*")+xlab("windspeed")+scale_y_continuous(breaks = c(-2, 0, 2))
+  ylab("TLeaf - Tair*")+xlab("windspeed (ms-1)")+scale_y_continuous(breaks = c(-2, 0, 2))
 b
 
 
@@ -368,7 +368,7 @@ c1<-ggplot(gs)+
   geom_smooth(aes(x = gs, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = gs, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = gs, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
[…]

### 2021-05-05 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 low wind speed and consistent $T_{air}$ can prolong $T_{leaf}$, consistently reducing stomatal conductance and photosynthesis while increasing $T_{leaf}$, and therefore reducing chances for leaf recovery.
-However, increase in biophysical fluctuations make $T_{leaf}$ unstable as seen in upper canopies where with available water unstable $T_{leaf}$ can be more manageable to peak irradiance and $T_{air}$ than in stable buffered canopy regions [**REF**]
+However, increase in biophysical fluctuations, for eg.
+wind speed, cloud mediated irradiance fluctuation and $T_{air}$  make $T_{leaf}$ unstable as seen in upper canopies where with available water unstable $T_{leaf}$ can be more manageable at peak irradiance and $T_{air}$ than in stable buffered canopy regions [**REF**] ~this understanding needs more help from co-authors, should this be included?
+(thermal time constant?)~
 
@@
 
-*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes (Condit et al. 2000). *
-*Spring and fall leaf phenology in temperate deciduous forests (Augsburger).*
+Deciduousness as a leaf trait is an adaptation to water stress therefore can influence biophysical gradient and leaf temperature distribution during leaf-loss [@rey-sanchez_spatial_2016].
+*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes that also tend to lose their leaves than understory species that are seldom deciduous in the tropics.
+Deciduous species are also more prominent in drier sites compared to wet sites [@conditQuantifyingDeciduousnessTropical2000b]*
+
+*Spring and fall leaf phenology in temperate deciduous forests (Augsburger).* Leaf emergence in deciduous forests is increasing delayed with height-ontogeny where understory species leaf-out earlier than canopy trees [Vitasse et al.
+2013; @seiwaChangesLeafPhenology1999].
 
@@
 Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits show significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
 Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels and therefore influence leaf temperatures.As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].

### 2021-05-06 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-~~In general, DGVMs range from Individual Based Models (IBMs) on one end of the complexity and scale spectrum to big-leaf models on the other end.
-Most climate models incorporate vegetation dynamics using big-leaf models, which include many first generation DGVMs that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer.
-This simplification results in greater computational efficiency and thus easier integration into ESMs (Arora & Boer, 2010; Bonan et al.
-2003; Cox, 2001; Krinner et al., 2005), although it comes at a cost of a lack of representation of important demographic processes, canopy gap formation, vertical light competition, competitive exclusion, and successional recovery from disturbance (Feeley et al., 2007; Hurtt et al.
-1998; Moorcroft et al., 2001; Smith et al.
-2001; Stark et al., 2012).
-On the other end of the spectrum are IBMs, also referred to as forest gap models, representing vegetation at the level of individual plants (Christoffersen et al., 2016; Fischer et al., 2016; Fyllas et al., 2014; Sato et al.
-2007; Shuman et al.
-2014; Smith et al., 2001; Bugmann, 2001; Dietze & Latimer, 2011).
-These models are often stochastic and spatially explicit, capturing spatial variability in the light environment and microclimates at high vertical and horizontal resolution.
-As a result, however, IBMs are computationally expensive and tend to be run at very local scales.~~
-
 **The computationally feasible approach to representing vertical structuring in DVGMs lies in Cohort-based models (CBMs), which sit in the middle of this continuum between the oversimplified vegetation dynamics in big-leaf models and the computational expense of individual-based models.** CBMs represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type (Haverd et al., 2013; Hurtt et al., 1998; Lischke et al.
@@
 2014; Smith et al., 2014; Weng et al., 2015).
-~~Although this cohort approach serves as a compromise that incurs lower computational costs than IBMs while capturing many of the same dynamics, CBMs do not incorporate the stochastic processes of IBMs that typically allow for greater representation of functional diversity (Fisher et al., 2010).~~
-
 *(KAT: I think all of the following content is great, but probably needs to be heavily tightened:)*

### 2021-05-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor.
-Foliage LAI and density mediate light regime, where denser upper canopies exposed to greater PAR values limit light to canopy interiors, lower canopy layers, and the understory.
-This light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig.
-`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b].
-
-The amount of light reaching lower canopy, understories and forest floors are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019].
-Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]and needle-leaf forest floors receive 17-34% of incoming sunlight [@baldocchiSeasonalVariationEnergy1997].
-Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute light in the form sunflecks--around 10-80% of of photon flux density-- to shaded forest regions that receive 1/3 of their carbon  during photosynthesis through these short light exposures [@waySunflecksTreesForests2012].
+This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig.
+`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].
+This is seen in tropical forest floors that receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], and mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006], compared to needle-leaf forest floors that receive 17-34% of incoming sunlight [@baldocchiSeasonalVariationEnergy1997].
+Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute sunflecks to shaded forest regions that receive ~1/3 of their carbon during photosynthesis through these short light exposures [@waySunflecksTreesForests2012].
 Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
-
-These light regimes shape foliage at a vertical gradient, which in turn shapes the microenvironment.
-At a foliar level, resulting light-adaptation in the form of sun and shade leaf dichotomy also facilitates plant-adaptation to the light environment in the understory and the overstory, ultimately contributing to shaping the microenvironment, forest vertical-profile's thermal sensitivity and their response to warming  [@fauset_differences_2018; @millerOnlySunlitLeaves2021; @michaletzPlantThermoregulationEnergetics2015; *De Frenne et al.
-2021*]
 
@@
 Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
-Understory wind-speed is also substantially reduced with increased canopy height [@juckerCanopyStructureTopography2018].
 Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
@@
 
-The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davisMicroclimaticBufferingForests2019].
-Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively where they maintain cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relatively more than pine and boreal forests, and nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
-2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009].
-Similarly, when compared to neighboring tree crop agricultural plantations, intact tropical forests maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-On the other hand, open-forests with sparse stands, have lower buffering capacity with greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009].
-Therefore, they maintain higher daytime maximum temperature and light availability in the understory that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009].
-During the night, pine forests can maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a].
+The strength of this buffering varies is greater under dense-canopy forests than open-canopy forests [*Defrenne et al.
+2021*].
+Greater temperature offsets by dense forests allows them to maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
+2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+On the other hand, open-forests tend to maintain higher daytime maximum temperature in the understory that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].
 
-Buffering is also affected by canopy roughness, turbulent air flow and the canopy boundary layer.
-As a result, taller trees upto 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory.
-However, beyond a certain tree height, temperature and VPD offset plateaud [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].
+Seasonally, buffering capacity increases with water availablity [@davis_microclimatic_2019] and is greater in wet than dry season, in tropical and temperate forests [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
+2017; @mcgregor_tree_nodate].
+However, under colder conditions, dense forest canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-forests and nearby clearings [@zellweger_seasonal_2019; *Defrenne et al.
+2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate] where greater radiative heat loss in exposed areas results in lower $T_{min}$.
 
-Seasonally, the capacity to moderate buffering is greater with water availability.
-During the wet season, tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al.
-2017], have all shown to offset temperatures positively in the understory, with lower maximum air temperatures and higher relative humidity below canopies and understories than above forest canopies and gaps.
-However, during the dry season of a semi-deciduous tropical forest in Panama, maximum temperatures were similar in the upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016].
-Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].
-
-Under colder conditions, minimum $T_{air}$ [$T_{min}$] is also buffered by forest canopies, increasing negetive temperature offset in the understory.
-Across Europe [@zellweger_seasonal_2019; *Defrenne et al.
-2021*] and in the northwestern United States [@davis_microclimatic_2019] higher minimum temperatures under forest canopies have been observed relative to nearby clearings.
-Greater radiative heat loss in exposed areas results in lower $T_{min}$.
-However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].
-
-*(Add something about risk of freeze.)*
+Tree height also plays a role in buffering where trees upto 20m with higher LAI have shown to lower mean and max temperature and VPD in the understory.
+However, beyond a certain tree height, temperature and VPD offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
 
 ### Humidity and evaporative demand
-**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**
-- DETAILS - REFS
-- @mcgregor_tree_nodate
-
-Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates, which tends to decrease with increasing maximum temperature.
-In three European forests (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.
-RH moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
-
-RH, in combination with temperature, determines vapor pressure deficit ($VPD$), an important metric expressing the driving force of water loss from a leaf.
-Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
-This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
+Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] in the below-canopy compared low LAI forests and to nearby open areas.
+This trend is greater in wetter conditions, in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
+RH, in combination with temperature, determines vapor pressure deficit (VPD), an important metric expressing the driving force of water loss from a leaf.
+Tropical forests during the wet season tend to have higher VPD, therefore greater evaporative demand, in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 
@@
 ### Geographic and climatic factors
-
-With higher altitude, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions.
-In Switzerland forests, buffering capacity has shown to be stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a], partly because trees in these high-altitude forests are sparserly distributed [@vonarxSpatiotemporalEffectsForest2012a] or have open canopies with greater light interception through canopy layers [@rajsnerovaMorphologicalBiochemicalPhysiological2015].
-Less capacity to buffer climatic conditions in open, high altitude forests might have implications to forest regeneration and climate change [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].
-This is also observed in  timberline and temperate humid-mountain forests in China, where tree height decreased with altitude, and tree growth sensitivity (based on tree-rings) increased [@liangForestTypeHeight2019].
-
-The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019].
-Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
+The strength of buffering also varies substantially with topography, climate and distance to waterbodies [@zellweger_seasonal_2019] where buffering capacity decreases with altitude, moving to open-forests  [@vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019]  and increases with wetter than drier regions [**REF**] and proximity to water such as in riparian forests.
+[@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
 
@@
 **Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
-Leaves adapt to their biophysical environment.
-Maintaining optimum leaf temperature for metabolic optima in relation to their biophysical environment maximizes carbon assimilation in a given circumstance.
-This is done through the adaptation of leaf traits to the surrounding biophysical environment [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-Thus, $T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+
+Leaves adapt to their biophysical environment to maximize carbon assimilation by maintaining optimum leaf temperature in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+As a result, leaves can exhibit 'limited homeothermy' where they are cooler than air at high temperature and warmer than air at low temperature or 'megathermy' where they are drastically warmer than $T_{air}$.
+In some cases, 'poikilothermy', where $T_{leaf}$ can be equalled to $T_{air}$ [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+
+$T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 `r fig_leaf_T`).
 The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
-`r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].
-When leaves are cooler than air at high temperature and warmer than air at low temperature, they considered to exhibit limited homeothermy.
-When leaves are drastically warmer than $T_{air}$ such as sometimes in the upper canopy, this is considered megathermy.
-In some cases, $T_{leaf}$ can be equalled to $T_{air}$, showing poikiolothermic behavior [Cavaleri, 2020; @drakeNoEvidenceHomeostatic2020].
+2; @michaletzPlantThermoregulationEnergetics2015].
 
@@
 
-**(paragraph on radiation as a driver)**
-Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$.
-On the other hand, under conditions of constant radiation and water availability, evaporative cooling sustains $T_{leaf}$ under increased $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
+Leaf temperature (and thermoregulation) is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
+Stomatal conductance ($g_s$)  largely influences $T_{leaf}$ by facilitating heat dissipation, whereas leaf water content affects the rate of leaf warming [@kochDiurnalPatternsLeaf1994a; @fauset_differences_2018].
+With sufficient $g_s$ leaves can regulate their temperature in relation to their biophysical environment, but limitation to $g_s$, driven by irradiance, VPD or drought curbs heat dissipation, drastically increasing leaf temperature.
 
-Greater upper canopy wind speeds (Fig.
-`r fig_NEON_vertical`) result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a].
-Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{leaf}$ and $T_{air}$. 
-In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017].
-Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata.
-This tends to be greater in the understory than in the canopy [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]
+Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$.
+With maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, but $g_s$ limitation can cause solar radiation to substantially elevate $T_{leaf}$ greater than $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
+
+Greater upper canopy wind speed (Fig.
+`r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ by increasing boundary layer conductance ($g_a$) for exposed leaves [@daudetWindSpeedLeaf1999a].
+Therefore, under hot conditions, leaves exposed to higher wind speeds would be most effective at cooling with high $g_s$ [Drake et al.
+2018].
+At low wind conditions, small $g_a$ to heat transfer can increase $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017] where boundary layer resistance increases with thickness of air layer at the leaf surface.
+This tends to be greater in buffered leaves [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]
 
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
-Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
-Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*.
-Under hot conditions, leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al.
-2018].
-However, during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.
-This is caused by increase in air temperature and VPD driving greater leaf water demand than water transport, which limits stomatal conductance along with decreased leaf water content, therefore limiting heat dissapation [@kochDiurnalPatternsLeaf1994; @fauset_differences_2018].
-On the other hand, buffered leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
+While stomatal conductance is inherently greater in exposed leaves than buffered shaded leaves, factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--are also greater with exposure.
+As a result, exposed leaves can experience steeper $g_s$ limitation than buffered leaves, driving an increase in $T_{leaf}$, further influenced by irradiance and $T_{air}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
+On the other hand, due to inherently lower $g_s$ in shade leaves, they can remain cooler under buffered conditions and substantially warmer under exposed conditions.
 
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** 
-*Smaller leaves have higher $g_{a}$ to heat transfer, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].  
-While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018].
-Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation or $T_{air}$ with stomata closed [@leighInfluenceLeafSize2017].
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.**  Smaller leaves have higher $g_{a}$ to heat transfer, therefore remain within a few degrees of $T_{air}$,  *with coolest leaves at intermediate sizes (~10mm)* [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].
+Whereas, regardless of $g_s$, large leaves (with greater leaf width) can be significantly cooler than T_air under low radiation with stomata open, and significantly hotter under high radiation or $T_{air}$ with stomata closed  [@leighInfluenceLeafSize2017; @fauset_differences_2018].
 
 ### In-situ observations
-
 **Vertical profiles in microclimate (Fig.
@@
 `r fig_leaf_T`).**
-Canopy buffering capacity, as well vertical biophysical gradient, affect vertical distribution of leaf temperature.
-While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
+Canopy buffering capacity and biophysical gradient affect vertical distribution of leaf temperature.
 
-Particularly, in open forests with sparse distribution of trees and lower buffering capacity, as observed in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of either increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
-Greater heat stress and $T_{air}$ in the lower canopy is also observed in open Australian desert trees [@curtis_intracanopy_2019] and among timberline krummholz needle-leaf trees, both growing close to the ground [@hadleyInfluenceKrummholzMat1987], this is analogous to greenhouse effect caused under dense lower canopies.
-Similarly, in dry conditions, in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016].
-In other tropical dry conditions, inner canopy can be warmer than above canopy and air temperature due to conditions of greater relative humidity [@dietzVerticalPatternsDuration2007].
+Open-forests with lower LAI or buffering capacity, can have greater lower canopy $T_{leaf}$ than upper canopy, reflecting microclimate conditions of vertical air mixing, heating from the ground, light permeating through canopy layers and still air in the lower canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+In other conditions, increase in relative humidity can increase inner canopy $T_{leaf}$  relatively more than above canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
+Similarly, open trees growing close to the ground can experience greater heat stress and $T_{air}$ in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
 Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
@@
 
-At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Especially in closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in the upper canopies than lower canopies, caused by midday stomatal closure [Slot et al.
+2019].
+This is seen in tropical forests [@fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a], semi-deciduous tropical forest during the wet season [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
 
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Especially in closed-canopy forests with increased buffering capacity, large leaf to air temperature differences and high leaf temperatures are found in the upper canopies compared to lower canopies, caused by midday stomatal closure that reduces transpirational cooling [Slot et al.
-2019].
-This is seen in Atlantic, Brazil forests [@fauset_differences_2018], Puerto Rico tropical wet, moist and subtropical wet forest[@mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021], semi-deciduous tropical forest during the wet season in Panama [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
+Greater light input in exposed upper canopy leaves can exceed $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
+Across forest types, temperate deciduous upper canopy $T_{leaf}$ - $T_{air}$ was observed to be greater than upper canopies of tropical moist and wet [@mauTemperateTropicalForest2018a]
 
-In three forest types- US temperate deciduous, Puerto Rico tropical wet and tropical moist, $T_{leaf_{max}}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in the upper canopies of all three forests, with increasing $T_{leaf}$ - $T_{air}$  with height.
-$T_{leaf}$ - $T_{air}$ was observed to be greatest in temperate deciduous forest compared to tropical moist and wet.
-Greater light input in exposed upper canopy leaves can sometimes elevate leaf temperature $1-7{^\circ}C$ above air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]
+### Leaf traits
 
-**Stability of $T_{leaf}$ varies with fluctuation in biophysical conditions**
-Biophysical environment fluctuation dictates how stable $T_{leaf}$ is, where stable biophysical conditions, for eg.
-low wind speed and consistent $T_{air}$ can prolong $T_{leaf}$, consistently reducing stomatal conductance and photosynthesis while increasing $T_{leaf}$, and therefore reducing chances for leaf recovery.
-However, increase in biophysical fluctuations, for eg.
-wind speed, cloud mediated irradiance fluctuation and $T_{air}$  make $T_{leaf}$ unstable as seen in upper canopies where with available water unstable $T_{leaf}$ can be more manageable at peak irradiance and $T_{air}$ than in stable buffered canopy regions [**REF**] ~this understanding needs more help from co-authors, should this be included?
-(thermal time constant?)~
-
-## Leaf traits
-
-**Leaf thermoregulation largely depends on leaf traits.
+**Leaf thermoregulation is mediated by leaf traits.
 Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light and water demand, in order to optimize C uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
@@
 Whereas shade leaves adapt to diffused light and stable microenvironment through investing in light capture efficiency by expanding specific leaf area (SLA) [@zwienieckiHydraulicLimitationsImposed2004; @rozendaalPlasticityLeafTraits2006; @mathurPhotosyntheticEfficiencySun2018a].
-Respective leaf traits also play a role in heat dissapation where exposed sun leaves are more efficient in dissapting heat than buffered shade leaves [@vogelConvectiveCoolingLow1970].
-
-Particularly in dense forests, the upper, outer canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs.
-`r fig_NEON_vertical`, `r SIfigs_NEON_all`).
 Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) and carbon balance.
-It is unsuprising, then, that leaf traits vary dramatically across vertical gradients.
+Since canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs.
+`r fig_NEON_vertical`, `r SIfigs_NEON_all`) it is unsuprising, then, that leaf traits vary dramatically across vertical gradients.
 

### 2021-05-08 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig.
-`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].
-This is seen in tropical forest floors that receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], and mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006], compared to needle-leaf forest floors that receive 17-34% of incoming sunlight [@baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].Therefore, light reaching forest floors that is incident to canopy tops can range from approximately 1-2% in tropical forest floors  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate [@bartemucciOverstoryInfluencesLight2006], to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997].
 Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute sunflecks to shaded forest regions that receive ~1/3 of their carbon during photosynthesis through these short light exposures [@waySunflecksTreesForests2012].
@@
 
-The strength of this buffering varies is greater under dense-canopy forests than open-canopy forests [*Defrenne et al.
+The strength of this buffering is greater under dense-canopy forests than open-canopy forests [*Defrenne et al.
 2021*].
-Greater temperature offsets by dense forests allows them to maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
+Therefore, dense forests maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
@@
 **Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
-Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] in the below-canopy compared low LAI forests and to nearby open areas.
+Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] below-canopy compared to open forests and nearby open areas.
 This trend is greater in wetter conditions, in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
@@
 ### Geographic and climatic factors
-The strength of buffering also varies substantially with topography, climate and distance to waterbodies [@zellweger_seasonal_2019] where buffering capacity decreases with altitude, moving to open-forests  [@vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019]  and increases with wetter than drier regions [**REF**] and proximity to water such as in riparian forests.
+
+The strength of buffering decreases with altitude, moving to open-forests  [[@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], increases with wetter than drier regions [**REF**] and proximity to water such as in riparian forests.
 [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
@@
 
-Leaf temperature (and thermoregulation) is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
-Stomatal conductance ($g_s$)  largely influences $T_{leaf}$ by facilitating heat dissipation, whereas leaf water content affects the rate of leaf warming [@kochDiurnalPatternsLeaf1994a; @fauset_differences_2018].
-With sufficient $g_s$ leaves can regulate their temperature in relation to their biophysical environment, but limitation to $g_s$, driven by irradiance, VPD or drought curbs heat dissipation, drastically increasing leaf temperature.
+Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]where stomatal conductance ($g_s$)  largely influences $T_{leaf}$ by facilitating heat dissipation, and leaf water content affects the rate of leaf warming [@kochDiurnalPatternsLeaf1994a; @fauset_differences_2018].
+With sufficient $g_s$ leaves can regulate their temperature even in biophysical environment [**REF**], but limitation to $g_s$, driven by irradiance, $T_{air}$, VPD or drought curbs heat dissipation, drastically increasing leaf temperature.
 
@@
 
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.**  Smaller leaves have higher $g_{a}$ to heat transfer, therefore remain within a few degrees of $T_{air}$,  *with coolest leaves at intermediate sizes (~10mm)* [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].
-Whereas, regardless of $g_s$, large leaves (with greater leaf width) can be significantly cooler than T_air under low radiation with stomata open, and significantly hotter under high radiation or $T_{air}$ with stomata closed  [@leighInfluenceLeafSize2017; @fauset_differences_2018].
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.**  Smaller leaves remain within a few degrees of $T_{air}$,  *with coolest leaves at intermediate sizes (~10mm)* due to higher $g_{a}$ to heat transfer [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].
+Whereas, large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation or $T_{air}$ with stomata closed  [@leighInfluenceLeafSize2017; @fauset_differences_2018].
 
@@
 `r fig_leaf_T`).**
-Canopy buffering capacity and biophysical gradient affect vertical distribution of leaf temperature.
+Vertical distribution of leaf temperature is dependent on canopy buffering capacity and the biophysical gradient.
 
-Open-forests with lower LAI or buffering capacity, can have greater lower canopy $T_{leaf}$ than upper canopy, reflecting microclimate conditions of vertical air mixing, heating from the ground, light permeating through canopy layers and still air in the lower canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
-In other conditions, increase in relative humidity can increase inner canopy $T_{leaf}$  relatively more than above canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
+Particularly, open-forests with buffering capacity can have greater lower canopy $T_{leaf}$ than upper canopy, reflecting microclimate conditions of vertical air mixing, heating from the ground, light permeating through canopy layers and still air in the lower canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+In other conditions, increase in inner canopy relative humidity can increase $T_{leaf}$  relatively more than above canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
 Similarly, open trees growing close to the ground can experience greater heat stress and $T_{air}$ in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
@@
 
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Especially in closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in the upper canopies than lower canopies, caused by midday stomatal closure [Slot et al.
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** In closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in the upper canopies than lower canopies, due to midday stomatal closure [Slot et al.
 2019].
@@
 
-Greater light input in exposed upper canopy leaves can exceed $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
-Across forest types, temperate deciduous upper canopy $T_{leaf}$ - $T_{air}$ was observed to be greater than upper canopies of tropical moist and wet [@mauTemperateTropicalForest2018a]
+Greater light input in exposed upper canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
+Across forest types, temperate deciduous upper canopy $T_{leaf}$ - $T_{air}$ has shown to be greater than $T_{leaf}$ - $T_{air}$ of upper canopy tropical moist and wet [@mauTemperateTropicalForest2018a].
 
@@
 **Leaf thermoregulation is mediated by leaf traits.
-Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], shaped by the intensity and availability of light and water demand, in order to optimize C uptake, thermoregulate and balance metabolism, across forest vertical strata, and thus show different degrees foliar plasticity [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
+Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], in order to optimize C uptake.
+Therefore, are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) across forest vertical strata (Figs.
+`r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
 
@@
 shade leaves and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-Therefore, exposed sun leaves are typically smaller in size and width, thicker, hairier and have greater lobation, and buffer biophysical conditions for shade leaves.
-Whereas shade leaves adapt to diffused light and stable microenvironment through investing in light capture efficiency by expanding specific leaf area (SLA) [@zwienieckiHydraulicLimitationsImposed2004; @rozendaalPlasticityLeafTraits2006; @mathurPhotosyntheticEfficiencySun2018a].
-Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) and carbon balance.
 Since canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs.
@@
 
-While light based acclimation in leaf traits is well established, research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
-Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.However, the majority of relevant research has focused on exposure gradients near ground level, with by far the most common study type being comparisons of sun and shade leaves.
-While light is the primary axis along which leaf traits vary, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig.
+While light based acclimation in leaf traits is well established, research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.However, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig.
 `r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.
@@
 2009, 2010*).
-Under constant light conditions (photon flux density), increase in height (>35m) showed decrease in canopy stomatal conductance, along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
-Adaptations to increasing height and hydraulic contraints show that leaf size [Kaare H.
+Isolating height (>35m) under constant light conditions (photon flux density) canopy stomatal conductance decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
+Morphologically, leaf size [Kaare H.
 Jensen and Maciej A.
-Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) tends to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].
+Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have shown to decrease, while leaf mass per area (at greater than 20m) to increase  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].
 
@@
 `r fig_NEON_vertical`) on leaf form and function remains an important area for research, but is beyond the scope of the current review.
+Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.
 For the purposes of interpreting the results presented here, we note that patterns across light and height gradients age generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
-
-*(The following notes need to be condensed/ worked into the paragraph above or discussion of specific traits.
-No need to get very specific (listing species studied, traits)--what we want is generalization.
-The goal is to answer the question, "Do we expect light gradients to be equivalent to height gradients, or is there variation attributable to height that is not captured by light gradients?")*
 
@@
 
-Within-individuals exposed canopies substanially buffer inner and lower canopies from light and biophysical conditions.
-Thus leaf traits in canopy positions vary accordingly [@casasSunShadeLeaves2011].
-Several studies point towards common variations in leaf anatomical, structural and biochemical trait plasticity at a vertical intra-canopy gradient across individual trees [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-The differences among these traits contribute to leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+Within-individuals going from buffered and shaded canopy positions to exposed positions, leaf traits vary on the basis of biophysical demands [@casasSunShadeLeaves2011] and therefore exhibit commonly observed patterns in leaf anatomical, structural and biochemical traits at an intra-canopy gradient, pointed by several studies [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+These observations contribute to [understanding] leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 
-Additionally, given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016], such as by incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient, partly because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020].
-
-*(Krista: the above paragraaph is really interesting, and I like the idea of starting off the section by mentioning what controls the formation of leaf traits.
-However, I think it gets into more detail than needed.
-Also, is there work out there on how they're influenced by temperature/ water at the time of formation?
-I'd broaden to include the full microenviroment (without increasing length)*)
+However, while a majority of relevant research have focused on exposure gradients near ground level, with by far the most common study type being comparisons of sun and shade leaves, given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016], such as by incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient, partly because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020].
 
@@
 
-Along the vertical gradient, exposed leaves are better equipped to dissapate heat than shaded leaves, through adaptive traits such as smaller leaf size, leaf dissections--lobes and serration-- decrease effective leaf width and therefore increase the rate of heat convection per unit area between leaf and air.
-[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].
-Mechanisms such vertical leaf angles and temporary wilting reduce surface light exposure and lower leaf temperature in exposed canopies [Ishida et al.
-1999; @niinemetsAdjustmentFoliageStructure1998].
-Hydraulic traits, for example leaf vein and stomtal density, also help maintain a balance photosynthesis and thermoregulation, in the high light environment of exposed canopies [Brodribb et al.
-2010].
-On the other hand, leaf temperature increases with effective leaf width and leaf size [@fauset_differences_2018], along with thermal time constant that is longer due to the increase in boundary layer thickness and stomatal resistance, and lower convective cooling [@leighInfluenceLeafSize2017; Wright et al.
-2019; @fauset_differences_2018].
-Therefore, shade leaves that typically have greater leaf size and effective leaf width compared to sun leaves, remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017].
-Additionally, understory leaves tend to open their stomata wider (stomatal apeture) than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].
+Going from shaded to exposed canopy position, morphological adaptation to dissapate heat increases.
+Sun leaves are often thicker, smaller, hairier, with greater dissection--lobes and serration--[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al.
+2010], with more vertical leaf angles than shade leaves [@niinemetsAdjustmentFoliageStructure1998].
+All, collectively regulate $T_{leaf}$-photosynthetic balance.
+On the other hand, because shaded leaves have greater effective leaf width [@fauset_differences_2018], boundary layer thickness, stomatal resistance and therefore longer thermal time constant, they exhibit lower convective cooling [@leighInfluenceLeafSize2017; Wright et al.
+2019].
+Thus, shade leaves remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017]
 
@@
 
-Canopy nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks, and could mediate sun vs.
-shade leaf response to heatflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018].
-Therefore, during growing season, nitrogen per area [$N_{area}$], along with LMA, leaf soluble sugars, starch and non-structural carbons increase with height and light, typically greater in exposed canopy positions or sun leaves compared to shade [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
+Nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018].
+Therefore, during growing season, $N_{area}$, along with LMA, leaf soluble sugars, starch and non-structural carbons increase with exposure [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
 2014].
-Phosphorous per area follows a similar trend along the canopy gradient, but with increased growth at higher light P uptake, in some cases, cannot keep up and thus shows variable results [@poorterMetaanalysisPlantResponses2019a]
+$P_{area}$ follows a similar trend, but with increased growth at higher light P uptake in some cases cannot keep up and thus shows variable results [@poorterMetaanalysisPlantResponses2019a]
 
@@
 
-Proportional to vertical light gradients at an intra- and inter-species level, sun exposed leaves have greater pools of xanthophyll pigments, comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] than in shaded leaves.
-[$VAZ$]--an antioxidant scavenging function--protects against cellular damage by converting excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+Pools of xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$],an antioxidant scavenging function--plays a role in protecting against cellular damage by converting excess light into heat energy dissipation, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].Therefore, $VAZ$ increases proportional to vertical light gradients.
 
@@
 
-Deciduousness as a leaf trait is an adaptation to water stress therefore can influence biophysical gradient and leaf temperature distribution during leaf-loss [@rey-sanchez_spatial_2016].
-*Among species that can be deciduous, greater proportion of deciduous individuals in larger size classes that also tend to lose their leaves than understory species that are seldom deciduous in the tropics.
+Deciduousness as a leaf trait is an adaptation to water and heat stress.
+Therefore, can influence biophysical gradient and leaf temperature distribution during leaf-loss [@rey-sanchez_spatial_2016].
+*In the tropics, among species that can be deciduous, greater proportion of deciduous individuals are in larger size classes that tend to lose their leaves more than understory species that are seldom deciduous.
 Deciduous species are also more prominent in drier sites compared to wet sites [@conditQuantifyingDeciduousnessTropical2000b]*
@@
 
-### Interspecific varaiation
+### Interspecific variation
 
-Across tree-size ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show fundamental light adapted leaf traits in their biophysical environment, as observed and similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Isolating ontogeny, under constant light conditions of open-grown canopy and understory trees, for example in large gaps, differences in traits show significantly greater LMA, photosynthetic capcity per area, and stomatal conductance in taller canopy trees compared to saplings [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Across  multi-species canopy and understory, while fundamental adaptations to light and water are observed, their pasticity to light differs along with inherent traits such as leaf shape, size and width, stomtal conductance levels and therefore influence leaf temperatures.As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
-
-*from Martijn in Tleaves figure: One thing that might be more relevant in tropical than in seasonal temperate forests, is the issue I mentioned before about the difference between shade leaves of canopy trees and leaves of understory species.
-The understory species in the tropics (the ones that won't necessarily make it to the canopy, and can reproduce in the understory) appear to invest more in leaf longevity than in light harvesting per se, so even though they are in the shade, they do not have large leaves with low LMA.
-This would be a deviation on, or a special case of, the understory parameters you are using now.
-I don't think we necessarily need to incorporate that in this figure, but I think it would be useful to be able to say that for understory species with small, high LMA leaves the effects of windspeed, radiation, RH, etc, are greater/smaller than for the understory species parameters in the figure.
-DOes this make sense?
-Is this relevant at all for seasonal forests?
-The only thing that comes to mind is Ilex, but maybe there are other species?*
-
- *(KAT: a lot of this content should probably move up the the intro to this section)*
+Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], LMA, photosynthetic capcity per area, and stomatal conductance increases with height compared to saplings, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Across  multi-species canopy and understory pasticity to light and biophysical condition differs along with inherent traits such as leaf shape, size, $gs$ levels and therefore influences $T_{leaf$.
+As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
 

### 2021-05-08 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig.
-`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].Therefore, light reaching forest floors that is incident to canopy tops can range from approximately 1-2% in tropical forest floors  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate [@bartemucciOverstoryInfluencesLight2006], to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].Light reaching forest floors that is incident to the canopy top can range from approximately 1-2% in tropical forests  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate forests[@bartemucciOverstoryInfluencesLight2006] to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997].
 Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute sunflecks to shaded forest regions that receive ~1/3 of their carbon during photosynthesis through these short light exposures [@waySunflecksTreesForests2012].
@@
 
-The strength of this buffering is greater under dense-canopy forests than open-canopy forests [*Defrenne et al.
+The strength of this buffering is greater under dense-canopy than open-canopy forests [*Defrenne et al.
 2021*].
-Therefore, dense forests maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
+Where dense forests maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-On the other hand, open-forests tend to maintain higher daytime maximum temperature in the understory that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].
+On the other hand, open-forests maintain higher daytime maximum temperature in the understory closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  sometimes warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].
 
@@
 **Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
-Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] below-canopy compared to open forests and nearby open areas.
+Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] below-canopy than open forests and nearby open areas.
 This trend is greater in wetter conditions, in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
@@
 
-The strength of buffering decreases with altitude, moving to open-forests  [[@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], increases with wetter than drier regions [**REF**] and proximity to water such as in riparian forests.
+With altitude, the strength of buffering decreases, moving to open-forests  [[@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], increases in wetter than drier regions [**REF**] and with proximity to water, for eg.
+in riparian forests.
 [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
@@
 
-Leaves adapt to their biophysical environment to maximize carbon assimilation by maintaining optimum leaf temperature in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaves adapt to their biophysical environment to maximize carbon assimilation, by maintaining optimum leaf temperature in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
 As a result, leaves can exhibit 'limited homeothermy' where they are cooler than air at high temperature and warmer than air at low temperature or 'megathermy' where they are drastically warmer than $T_{air}$.
@@
 
-Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]where stomatal conductance ($g_s$)  largely influences $T_{leaf}$ by facilitating heat dissipation, and leaf water content affects the rate of leaf warming [@kochDiurnalPatternsLeaf1994a; @fauset_differences_2018].
-With sufficient $g_s$ leaves can regulate their temperature even in biophysical environment [**REF**], but limitation to $g_s$, driven by irradiance, $T_{air}$, VPD or drought curbs heat dissipation, drastically increasing leaf temperature.
+Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
+With sufficient stomatal conductance ($g_s$) leaves can regulate their temperature, by dissapating heat, even in drastic biophysical environment [**REF**] but when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, VPD or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
 
-Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$.
+Incoming solar radiation is a strong driver of $T_{leaf}$, relative to $T_{air}$.
 With maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, but $g_s$ limitation can cause solar radiation to substantially elevate $T_{leaf}$ greater than $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
@@
 
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** In closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in the upper canopies than lower canopies, due to midday stomatal closure [Slot et al.
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** In closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in upper than lower canopies, due to midday stomatal closure [Slot et al.
 2019].
@@
 Greater light input in exposed upper canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
-Across forest types, temperate deciduous upper canopy $T_{leaf}$ - $T_{air}$ has shown to be greater than $T_{leaf}$ - $T_{air}$ of upper canopy tropical moist and wet [@mauTemperateTropicalForest2018a].
+Additionally, the upper canopy of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than $T_{leaf}$ - $T_{air}$ of upper canopy tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
 
@@
 
-While light based acclimation in leaf traits is well established, research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.However, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig.
+While light-driven plasticity in leaf traits is well established, research on solely height-driven leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
+However, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig.
 `r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.
@@
 2009, 2010*).
-Isolating height (>35m) under constant light conditions (photon flux density) canopy stomatal conductance decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
+Isolating height (>35m) under constant light conditions, show that canopy stomatal conductance decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
 Morphologically, leaf size [Kaare H.
@@
 
-Within-individuals going from buffered and shaded canopy positions to exposed positions, leaf traits vary on the basis of biophysical demands [@casasSunShadeLeaves2011] and therefore exhibit commonly observed patterns in leaf anatomical, structural and biochemical traits at an intra-canopy gradient, pointed by several studies [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Within-individuals going from buffered [shaded] canopy positions to exposed, leaf traits vary on the basis of biophysical demands [@casasSunShadeLeaves2011] and therefore exhibit commonly observed patterns in leaf anatomical, structural and biochemical traits at an intra-canopy gradient, pointed by several studies [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
 These observations contribute to [understanding] leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
@@
 2019].
-Thus, shade leaves remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017]
+Which leads to cooler $T_{leaf}$ under buffered conditions, and substanially warmer $T_{leaf}$ under greater light or $T_{air}$ [@leighInfluenceLeafSize2017]
 
@@
 
-Nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018].
+Nitrogen distribution along the vertical gradient optimizes to prevent photoinhition in sun leaves during direct light and shade leaves during sunflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018].
 Therefore, during growing season, $N_{area}$, along with LMA, leaf soluble sugars, starch and non-structural carbons increase with exposure [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
 2014].
-$P_{area}$ follows a similar trend, but with increased growth at higher light P uptake in some cases cannot keep up and thus shows variable results [@poorterMetaanalysisPlantResponses2019a]
 
@@
 
-Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], LMA, photosynthetic capcity per area, and stomatal conductance increases with height compared to saplings, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Across  multi-species canopy and understory pasticity to light and biophysical condition differs along with inherent traits such as leaf shape, size, $gs$ levels and therefore influences $T_{leaf$.
+Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], LMA, photosynthetic capcity per area, and stomatal conductance increase with height compared to saplings, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Across  multi-species canopy and understory, pasticity to light and biophysical condition differs along with inherent traits such as leaf shape, size, $gs$ levels and therefore influences $T_{leaf}$.
 As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
@@
 
-While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well.
-During the day, max transpiration rate and stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
+While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by $gs$ [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]; and therefore their ability to photosynthesize as well.
+During the day, max transpiration rate and $gs$ increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
 During midday, stomtal closure occurs when evaporative demand exceeds water transport capacity in exposed leaves thereby increasing $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004].
-Thus, stomatal limitation is greater and steeper in exposed canopies than shaded leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
 
@@
 
-Leaf temperature beyond photosynthetic optimum temperature declines photosynthesis [**REF**].
-Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
+Leaf temperature beyond photosynthetic optimum temperature declines photosynthesis due it's peaked response to temperature.
+Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
 2017; @slotSituTemperatureResponse2017).
@@
 2015 ).
-
-Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis.
-At increasing temperatures, photosynthetic system II [PSII] is particularly vulnerable.
-Damage to PSII  can lead to decreased electron transport rates and photosynthetic failure.
-Irreversible damage to PSII is done at 40-60C [Baker 2008; @feeley_thermal_2020].
-Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought.
-[@niinemetsWorldwideAnalysisWithincanopy2015]
+Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].
+Leaves in tropical upper canopies have shown to experience air temperatures well above $40-60{^\circ}C$ on a regular basis, if $T_{leaf}$ is not regulated through $gs$ or adapated, heat stress can loss of canopy leaves [**REF**]
 
@@
 2019 ; @hernandezSimilarTemperatureDependence2020].
-Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020],as well in vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and no significant trend in temperate trees [@mau_temperate_2018].
+Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical profile in tropical forests [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
 

### 2021-05-11 vertical-thermal-review.Rmd
7500 words total without remote sensing, modeling and conclusion

@@
 
-*[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N.
-Smith [agreed.
-Again the importance of differentiating between within-species and among species trait variation.
-Also; steepness of the vertical gradient in light and VPD.
-In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*
-
 **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that optimum temperature $T_{opt}$ of $J_{max}$ increased with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
 For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
-2019 ; @hernandezSimilarTemperatureDependence2020].
-Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical profile in tropical forests [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
+2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
 
 *If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.*
-
-*(KAT: let's cut the portion on T50 down to one or at most two par:)*
 
@@
 
-Across forest species, typical values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al.
+Typical $T_{50}$ values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al.
 2017).
-$T_{50}$ and $T_{crit}$ decreases with elevation, as seen in tropical forests in Panama and Columbia and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017] and varies with leaf traits [Sastry et al.
+With elevation, $T_{50}$ and $T_{crit}$ decrease, as seen in tropical foress and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], and vary with leaf traits [Sastry et al.
 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020].
 Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$.
-On the other hand, leaves of species with greater construction costs, observed through LMA, had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020].
-However, higher photosynthetic heat tolerance does not necessary increase the breadth of leaf thermal safety margins.
-Rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$) [@perezPhotosyntheticHeatTolerances2020]
+On the other hand, leaves of species with greater construction costs, greater LMA, had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020].
+However, higher photosynthetic heat tolerance does not necessary increase the breadth of leaf thermal safety margin.
+Rather, across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$) [@perezPhotosyntheticHeatTolerances2020]
 
@@
 For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions.
-Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [@curtis_intracanopy_2019].
+Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, low-wind speed, $T_{air} and greater canopy [@curtis_intracanopy_2019].
 
@@
 
-In two forests, north-eastern oak dominated-deciduous forest in the US and conifer-dominated temperate New Zealand rainforest, across species $Q_{10}$ was found to be greater in upper-canopy leaves than in lower-canopy leaves, where values differed between species [@turnbullScalingFoliarRespiration2003].
-Similarly, in sweetgum (*Liquidambara styraciflua*), upper canopy sun leaves showed greater $Q_{10}$ than lower within-canopy shade leaves [@harleyEffectsLightTemperature1996].
-However, in other feild studies, $Q_{10}$ didnot vary across species vertical profiles and light exposure such as in a lowland tropical Australian forest study [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta [@bolstad_foliar_1999] $Q_{10}$.
-At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999].
+Where $Q_{10}$ is greater in exposed upper-canopy leaves than in shaded lower-canopy leaves in across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] but in other feild studies $Q_{10}$ doesnot vary across species vertical profiles in a lowland tropical Australian forest study [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta, with a slight tendency to increase down the canopy relative to upper-canopy leaves as T increases (*see their Fig.
+1*)[@bolstad_foliar_1999].
 (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
 May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*).
-Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig.
-1*).
-@bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
+@bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
 
-With temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum.
-Beyond this inflection point, fluxnet global data analysis shows that photosynthesis declines whereas respiration increases exponentially, therefore, moving from carbon sink to carbon source with increasing temperature, independent of water and irradiance [@duffyHowCloseAre2021]
+Additionally, a fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]
 
@@
 **Isoprene production**
-- high emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
-- The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
--  Within species, isporene production scales with light/ T; however regardless of light, isoprene emission increases with T such as in understory shaded leaves?
+High emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
+The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
+Within species, isporene production scales with light/ T; however regardless of light, isoprene emission increases with T such as in understory shaded leaves?
 
@@
 More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
-Many observations of larger trees suffering more during drought (Bennett et al. 2015).
-One likely mechanism behind this pattern is that the crowns of larger trees exist higher in the vertical profile and therefore are exposed to higher evaporative demand and solar radiation [@mcgregor_tree_nodate].
-Although drought is primarily a hydraulic problem, lack of water --> lower gs --> higher leaf T, so leaves face tradeoffs of water loss vs potentially damaging leafT.
+Biophysically, greater vapor pressure deficit, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al.
+2016].
+On the other hand, greater relative humidity can do the opposite, where it stifles evapotranspiration thereby increasing leaf temperatures, particularly in canopy tops where irradiance can have a greater effect [Fig.3;@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+However, soil water loss in conditions of drought effect larger trees more than small trees in less than a few days.
+Primarily because, greater biophysical evaporative demand in the canopy cannot meet long-distance water transport from the ground due to limited soil moisture [Norway spruce study, bennett; @mcgregor_tree_nodate].
 
-On the basis of phenology, temperate understory  trees tend to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al.
-2013], and, increase in annual temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renner, 2019].
-However, short-term May spring heat waves have shown to negetively impact partially-expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than in understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].
+On the basis of phenology, because understory trees leaf out earlier than canopy due to ontogenetic variation in tree size [Vitasse et al.
+2013], understory growth is more sensitive to annual warming than canopy - showing earlier spring onset and later autumn senescense, particularly in temperate deciduous forests [Zohnner and Renner, 2019].
+However, late spring short-term heat waves negatively impact partially- expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].
+
+If we expect 'shade' plants to have a different heat sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.
 
@@
 
-Additionally, increased exposure to light increases non structural carbon reserve (NSC), plant height, foliage robustness,  while also increasing evapotranspiration, decreasing hydraulic conductivity, increasing chances of photoinhibition in leaves should water availability become limited, however with available water overall sensitivity to stress decreases.
-Understory shaded/younger trees have comparatively less NSC reserve due to light limitation as a primary stressor [which differs among shade-tolerant/intolerant], decreased plant biomass, increased leaf biomass, and decreased evapotranpiration demands, as well as greater investment in light harvesting than in photosynthesis--with greater overall sensitivity to stress and reduced tolerance.
-Larger NSC reserves in trees with greater light and water availability allows them to tolerate greater number of stressors compared to understory/younger trees that are able to tolerate only a fewer stressors making them overall more sensitive to stress [such as temperature].
-Thus, overall stress sensitivity decreases and tolerance increases with increasing height, light and age.[@niinemetsResponsesForestTrees2010a]
+Additionally, interacting biophysical and biological stressors show that stress sensitivity is greater in buffered small trees, and decreases where tolerance increases with height, light with available water.
+Partly because large NSC reserves in taller trees due to available light allow them to tolerate greater number of stressors than buffered trees that are light and NSC limited [@niinemetsResponsesForestTrees2010a]
 
@@
 It is possible that an increased respiration load is more disadvantageous for understory plants because 1), I suspect that a greater fraction of their total biomass is life biomass that respires (as opposed dead heartwood in the stems of big trees), and 2), because they operate closer to their light compensation point, so if they lose more carbon in respiration, they would need  to make up for lost carbon in photosynthesis, but might lack sunlight to do so*
-
-Interaction between degrees of shade x ontogeny with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
 
@@
 
-*Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-Larger trees suffer disproportionally greater hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
-2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
-Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].*
-
 *While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
@@
 Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].*
-
-*The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
-Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018].*
-
- if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered
+Additionally, warming and heat waves can alter canopy and understory phenology and forest photosynthesis [@zohnerOngoingSeasonallyUneven2019].*
 
@@
 
-Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a].
+*The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+With larger trees suffering disproportionally greater hydraulic stress during drought, can contribute to increasing larger tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
+2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
+Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020], or increase in mortality of big trees (e.g.
+@senfCanopyMortalityHas2018a) would increase canopy roughnessare, also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].*
+
+*Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Open-forest might experience greater microclimate warming than forests, having implications for forest regeneration, particularly in high-altitude regions [@vonarxSpatiotemporalEffectsForest2012a].
 Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
 Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Each of these changes have potential feedbacks to climate change.
-
-increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness
+Each of these changes have potential feedbacks to climate change.*
 

### 2021-05-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 All, collectively regulate $T_{leaf}$-photosynthetic balance.
-On the other hand, because shaded leaves have greater effective leaf width [@fauset_differences_2018], boundary layer thickness, stomatal resistance and therefore longer thermal time constant, they exhibit lower convective cooling [@leighInfluenceLeafSize2017; Wright et al.
+Under homogeneous light conditions in open-canopy, trait plasticity of sun leaves are based on $T_{air_{max}}$ and $VPD$ stress of canopy oriention [@mediavillaFoliarPlasticityRelated2019a].
+Other hand, because shaded leaves have greater effective leaf width [@fauset_differencsunes_2018], boundary layer thickness, stomatal resistance and therefore longer thermal time constant, they exhibit lower convective cooling [@leighInfluenceLeafSize2017; Wright et al.
 2019].
@@
 During the day, max transpiration rate and $gs$ increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
-During midday, stomtal closure occurs when evaporative demand exceeds water transport capacity in exposed leaves thereby increasing $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004].
+During midday, stomtal closure occurs when evaporative demand exceeds water transport capacity in exposed leaves thereby increasing $T_{leaf}$, and decreasing photosynthesis [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
 
@@
 @slotSituTemperatureResponse2017; Smith et al.
-2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs.
-e.g.
+2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g.
 Sage & Kubien 2007; Varhammer et al.
@@
 2017).
-With elevation, $T_{50}$ and $T_{crit}$ decrease, as seen in tropical foress and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], and vary with leaf traits [Sastry et al.
+With elevation, $T_{50}$ and $T_{crit}$ decrease, as seen in tropical foress and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], and vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al.
 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020].
@@
 
-*While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020].
+While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
 Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
 Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
-Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+Plant communities in forest microclimates are already undergoing thermophilizatioN [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to microclimatic fauna [@scheffersMicrohabitatsReduceAnimal2013] .
 Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
+Additionally, warming in other cases is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, in closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
+
 Additionally, warming and heat waves can alter canopy and understory phenology and forest photosynthesis [@zohnerOngoingSeasonallyUneven2019].*
 
-*Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth).
+Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth).
 Differential mortality within the canopy structure will alter the profile of metabolism and hence emergent forest function.
 Such alteration to canopy structure may be detectable by lidar and thermal remote sensing.
-If we know how metabolism maps to the thermal profile, then the observed forest structural changes can inform prediction of forest function.*
+If we know how metabolism maps to the thermal profile, then the observed forest structural changes can inform prediction of forest function.
 
@@
 
-*The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
+Canopy structural composition and $LAI$ largely influence understory species composition and soil fertility [majasalmiImpactTreeCanopy2020], including buffering, which in-turn is influenced by the hydrological cycle [@davis_microclimatic_2019].
+Each of these are subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
+Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
 With larger trees suffering disproportionally greater hydraulic stress during drought, can contribute to increasing larger tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
@@
 Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020], or increase in mortality of big trees (e.g.
-@senfCanopyMortalityHas2018a) would increase canopy roughnessare, also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].*
+@senfCanopyMortalityHas2018a) would increase canopy roughnessare, also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
 
-*Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
 Open-forest might experience greater microclimate warming than forests, having implications for forest regeneration, particularly in high-altitude regions [@vonarxSpatiotemporalEffectsForest2012a].
@@
 Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Each of these changes have potential feedbacks to climate change.*
+Each of these changes have potential feedbacks to climate change.
 

### 2021-05-12 vertical-thermal-review.Rmd
still, 688 words over

@@
 This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig.
-`r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].Light reaching forest floors that is incident to the canopy top can range from approximately 1-2% in tropical forests  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate forests[@bartemucciOverstoryInfluencesLight2006] to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997].
-Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute sunflecks to shaded forest regions that receive ~1/3 of their carbon during photosynthesis through these short light exposures [@waySunflecksTreesForests2012].
+`r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].
+Light reaching forest floors that is incident to canopy tops ranges from ~1-2% in tropical forests  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate forests[@bartemucciOverstoryInfluencesLight2006] to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997].
+Sunflecks, caused by small canopy gaps and wind-induced canopy movements also contribute to the shaded forest regions that receive 50-80% of their light for photosynthesis during these short exposures [@waySunflecksTreesForests2012].
 Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
@@
 Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
-Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
+Consistent with these results, higher wind speeds at greater heights have been observed in both closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
 
@@
 
-The strength of this buffering is greater under dense-canopy than open-canopy forests [*Defrenne et al.
-2021*].
-Where dense forests maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
+The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than open-forests, nearby clearings [*Defrenne et al.
+2021*; @vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
@@
 
-Seasonally, buffering capacity increases with water availablity [@davis_microclimatic_2019] and is greater in wet than dry season, in tropical and temperate forests [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
+Seasonally, buffering capacity increases with water availablity [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry season [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
 2017; @mcgregor_tree_nodate].
-However, under colder conditions, dense forest canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-forests and nearby clearings [@zellweger_seasonal_2019; *Defrenne et al.
+However, under colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings [@zellweger_seasonal_2019; *Defrenne et al.
 2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate] where greater radiative heat loss in exposed areas results in lower $T_{min}$.
 
-Tree height also plays a role in buffering where trees upto 20m with higher LAI have shown to lower mean and max temperature and VPD in the understory.
-However, beyond a certain tree height, temperature and VPD offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+Tree height also plays a role in buffering where trees upto 20m with higher LAI can maintain lower mean and max temperature and vapor pressure deficit ($VPD$) in the understory.
+However, beyond a certain tree height, temperature and $VPD$ offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
 
@@
 **Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
-Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] below-canopy than open forests and nearby open areas.
-This trend is greater in wetter conditions, in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].
-RH, in combination with temperature, determines vapor pressure deficit (VPD), an important metric expressing the driving force of water loss from a leaf.
-Tropical forests during the wet season tend to have higher VPD, therefore greater evaporative demand, in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a].
+RH, in combination with temperature, determines $VPD$, an important metric expressing the driving force of water loss from a leaf; tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 
@@
 
-With altitude, the strength of buffering decreases, moving to open-forests  [[@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], increases in wetter than drier regions [**REF**] and with proximity to water, for eg.
+With altitude, the strength of buffering decreases, moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for eg.
 in riparian forests.
@@
 
-**Many of the vertical gradient biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
+**The combination of biophysical variables reviewed above drive leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn has a strong influence on leaf metabolism (see following section).**
+Leaves adapt to their biophysical environment to maximize carbon assimilation, by maintaining optimum $T_{leaf}$ in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+As a result, leaves can be cooler than air at high temperature and warmer than air at low temperature ('limited homeothermy') or can be drastically warmer than $T_{air}$ ('megathermy').
+In some cases, $T_{leaf}$ can be equalled to $T_{air}$ ('poikilothermy') [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
 
-Leaves adapt to their biophysical environment to maximize carbon assimilation, by maintaining optimum leaf temperature in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-As a result, leaves can exhibit 'limited homeothermy' where they are cooler than air at high temperature and warmer than air at low temperature or 'megathermy' where they are drastically warmer than $T_{air}$.
-In some cases, 'poikilothermy', where $T_{leaf}$ can be equalled to $T_{air}$ [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
-
-$T_{leaf}$ is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+$T_{leaf}$ itself is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
 `r fig_leaf_T`).
@@
 
-Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
-With sufficient stomatal conductance ($g_s$) leaves can regulate their temperature, by dissapating heat, even in drastic biophysical environment [**REF**] but when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, VPD or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
+Leaf thermoregulation is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
+Leaves can regulate their temperature by dissapating heat through stomatal conductance ($g_s$), even in drastic biophysical conditions [**REF**] but when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, VPD or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
 
@@
 
-Greater upper canopy wind speed (Fig.
-`r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ by increasing boundary layer conductance ($g_a$) for exposed leaves [@daudetWindSpeedLeaf1999a].
-Therefore, under hot conditions, leaves exposed to higher wind speeds would be most effective at cooling with high $g_s$ [Drake et al.
+Greater wind speed (Fig.
+`r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ by increasing boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a].
+Therefore, under hot conditions, canopy-leaves exposed to higher wind speeds would be most effective at cooling with high $g_s$ [Drake et al.
 2018].
@@
 As a result, exposed leaves can experience steeper $g_s$ limitation than buffered leaves, driving an increase in $T_{leaf}$, further influenced by irradiance and $T_{air}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
-On the other hand, due to inherently lower $g_s$ in shade leaves, they can remain cooler under buffered conditions and substantially warmer under exposed conditions.
+On the other hand, due to inherently lower $g_s$ in shade leaves, they can remain cooler under buffered conditions and substantially warmer under increased $T_{air}$ or irradiance.
 
@@
 `r fig_leaf_T`).**
-Vertical distribution of leaf temperature is dependent on canopy buffering capacity and the biophysical gradient.
-
-Particularly, open-forests with buffering capacity can have greater lower canopy $T_{leaf}$ than upper canopy, reflecting microclimate conditions of vertical air mixing, heating from the ground, light permeating through canopy layers and still air in the lower canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
-In other conditions, increase in inner canopy relative humidity can increase $T_{leaf}$  relatively more than above canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
-Similarly, open trees growing close to the ground can experience greater heat stress and $T_{air}$ in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
+Canopy buffering capacity largely influences vertical $T_{leaf}$ dynamics.
+Particularly, in open-forests with lower buffering capacity, $T_{leaf}$ is greater in lower-canopy than upper canopy, reflecting microclimate conditions of vertical air mixing, light permeating through canopy layers, still air in the lower canopy and heating from the ground, [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
+Greater $RH$ in inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
 Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
@@
 
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** In closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in upper than lower canopies, due to midday stomatal closure [Slot et al.
-2019].
-This is seen in tropical forests [@fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a], semi-deciduous tropical forest during the wet season [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].
-
-Greater light input in exposed upper canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
-Additionally, the upper canopy of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than $T_{leaf}$ - $T_{air}$ of upper canopy tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Closed-canopy forests, such as tropical and semi-deciduous tropical, and temperate forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies due to greater heat and water stress from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016].
+Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper-canopies susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
+Comparing forests, upper-canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper-canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
 
@@
 
-**Leaf thermoregulation is mediated by leaf traits.
-Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016], in order to optimize C uptake.
-Therefore, are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism (following section) across forest vertical strata (Figs.
+**Leaf thermoregulation largely depends leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016].
+Therefore, are critical in maintaining desirable $T_{leaf}$ and in shaping leaf metabolism (following section) across forest vertical strata (Figs.
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
 
-#### *(clarify light vs height - 1 paragraph)*
-
 Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004].
-Degrees of direct to diffused light and consequental water demands shape sun vs.
+Degrees of direct to diffused light and resultant water demands shape sun vs.
 shade leaves and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-Since canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs.
-`r fig_NEON_vertical`, `r SIfigs_NEON_all`) it is unsuprising, then, that leaf traits vary dramatically across vertical gradients.
-
-While light-driven plasticity in leaf traits is well established, research on solely height-driven leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.
-However, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig.
-`r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.
+Therefore, taller tree canopy-leaves dramatically vary from canopy-interior and understory due to the differences in biophysical conditions (Figs.
+`r fig_NEON_vertical`, `r SIfigs_NEON_all`).
+While leaf traits vary primarily on the axis of light, at greater tree-heights, both theory and empirical evidence indicate that other factors-for example, tree-height or exposure to evaporative demand (Fig.
+`r fig_NEON_vertical`)- are also important in shaping at least a subset of traits.
 
@@
 2009, 2010*).
-Isolating height (>35m) under constant light conditions, show that canopy stomatal conductance decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
+Isolating height (>35m) under constant light conditions show that, canopy $gs$ decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
 Morphologically, leaf size [Kaare H.
@@
 
-Disentangling the influences of hydraulic constraints, light, and other biophysical drivers that vary with height (Fig.
-`r fig_NEON_vertical`) on leaf form and function remains an important area for research, but is beyond the scope of the current review.
+Disentangling the influences of hydraulic constraints, light and other biophysical drivers that vary with height (Fig.
+`r fig_NEON_vertical`) on leaf traits remains an important area for research, but is beyond the scope of the current review.
 Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.
-For the purposes of interpreting the results presented here, we note that patterns across light and height gradients age generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
+For the purposes of interpreting the results presented here, we note that patterns across light and height gradients are generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
 
@@
 
-Within-individuals going from buffered [shaded] canopy positions to exposed, leaf traits vary on the basis of biophysical demands [@casasSunShadeLeaves2011] and therefore exhibit commonly observed patterns in leaf anatomical, structural and biochemical traits at an intra-canopy gradient, pointed by several studies [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-These observations contribute to [understanding] leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
-
-However, while a majority of relevant research have focused on exposure gradients near ground level, with by far the most common study type being comparisons of sun and shade leaves, given the vastly different light environments at the bottom of the canopy and that most leaves in the canopy are likely to be shade leaves, there is a strong research need to better characterize trait relationships and leaf responses through the whole vertical canopy [@keenanGlobalLeafTrait2016], such as by incorporating sun and shade in modelling leaf economic spectrum (LES) traits within-canopy gradient, partly because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020].
+Within-canopy gradient of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- and therefore exhibit patterns in leaf anatomical, structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+These trait-patterns contribute to [understanding] $T_{leaf}$ dynamics across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] where Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+The characterization of trait relationships and leaf responses through the whole vertical canopy is becoming a strong research need because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
 
@@
 All, collectively regulate $T_{leaf}$-photosynthetic balance.
-Under homogeneous light conditions in open-canopy, trait plasticity of sun leaves are based on $T_{air_{max}}$ and $VPD$ stress of canopy oriention [@mediavillaFoliarPlasticityRelated2019a].
-Other hand, because shaded leaves have greater effective leaf width [@fauset_differencsunes_2018], boundary layer thickness, stomatal resistance and therefore longer thermal time constant, they exhibit lower convective cooling [@leighInfluenceLeafSize2017; Wright et al.
+On the other hand, shaded leaves invest more in $SLA$ for light capture than in heat dissapation, therefore have lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
 2019].
-Which leads to cooler $T_{leaf}$ under buffered conditions, and substanially warmer $T_{leaf}$ under greater light or $T_{air}$ [@leighInfluenceLeafSize2017]
+Additionally, under homogeneous light conditions, sun-leaves in different canopy orientations are shaped on the basis of $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-Nitrogen distribution along the vertical gradient optimizes to prevent photoinhition in sun leaves during direct light and shade leaves during sunflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018].
-Therefore, during growing season, $N_{area}$, along with LMA, leaf soluble sugars, starch and non-structural carbons increase with exposure [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
+Nitrogen along the vertical gradient, is constantly re-mobilized from senescensed to growing leaves [@niinemetsPhotosynthesisResourceDistribution2007a] where $N_{area}$, with LMA, leaf soluble sugars, starch and non-structural carbons are greater [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
 2014].
+Additionally, $N_{area}$ optimizes to prevent photoinhition in sun leaves during direct light and shade leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].
 
@@
 
-Pools of xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$],an antioxidant scavenging function--plays a role in protecting against cellular damage by converting excess light into heat energy dissipation, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].Therefore, $VAZ$ increases proportional to vertical light gradients.
+Pools of xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function--plays a role in converting excess light into heat energy dissipation, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998], increases proportional light gradients.
+
+##### Isoprene production
+Isoprene production *capability* increases vertically with light and therefore facilitates high heat tolerance in emitting species' canopy positons [Sharky]
+
+### Interspecific variation
+
+Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], increase in LMA, photosynthetic capcity per area, and stomatal conductance with height are attributed to ontogenetical differences, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Across  multi-species canopy and understory, trait adaptation to spatio-temporal light and biophysical gradient varies largely with inherent specie-specific trait differences such as leaf shape, size, $gs$ levels and therefore influences individual $T_{leafs}$.
+As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
 
@@
 
-##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
-*Leaf lifespan generally increases with LMA.
-LMA and leaflifespan is greater with available light, and largely depends on the cost of carbon investment in a leaf, which varies across species sun vs.
-shade leaves [**double check**] Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
+##### *Leaf lifespan* 
+*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
 This is bound to affect metabolism, and may affect other aspects of their performance.*
@@
 Deciduousness as a leaf trait is an adaptation to water and heat stress.
-Therefore, can influence biophysical gradient and leaf temperature distribution during leaf-loss [@rey-sanchez_spatial_2016].
+Therefore, can influence biophysical gradient and $T_{leaf}$ distribution during leaf-loss [@rey-sanchez_spatial_2016].
 *In the tropics, among species that can be deciduous, greater proportion of deciduous individuals are in larger size classes that tend to lose their leaves more than understory species that are seldom deciduous.
@@
 
-*Spring and fall leaf phenology in temperate deciduous forests (Augsburger).* Leaf emergence in deciduous forests is increasing delayed with height-ontogeny where understory species leaf-out earlier than canopy trees [Vitasse et al.
+*Spring and fall leaf phenology in temperate deciduous forests (Augsburger).* Leaf emergence in deciduous forests is delayed with height-ontogeny where understory species leaf-out earlier than canopy trees [Vitasse et al.
 2013; @seiwaChangesLeafPhenology1999].
-
-### Interspecific variation
-
-Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], LMA, photosynthetic capcity per area, and stomatal conductance increase with height compared to saplings, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Across  multi-species canopy and understory, pasticity to light and biophysical condition differs along with inherent traits such as leaf shape, size, $gs$ levels and therefore influences $T_{leaf}$.
-As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
-
-##### Isoprene production
--  Isoprene production *capability* as a trait
 
@@
 
-While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by $gs$ [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]; and therefore their ability to photosynthesize as well.
+While exposed canopy sun-leaves are equipped to thermoregulate more efficiently than shaded-leaves, they are largely constrained by $gs$ [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]; and therefore their ability to photosynthesize as well.
 During the day, max transpiration rate and $gs$ increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
-During midday, stomtal closure occurs when evaporative demand exceeds water transport capacity in exposed leaves thereby increasing $T_{leaf}$, and decreasing photosynthesis [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
+During midday, stomtal closure in exposed leaves increases $T_{leaf}$, and decreases photosynthesis [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
 
@@
 
-**Photosynthesis is generally higher in sun leaves and exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al.
+**Photosynthesis is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al.
 2015 ; Slot et al.
@@
 
-Leaf temperature beyond photosynthetic optimum temperature declines photosynthesis due it's peaked response to temperature.
+$T_{leaf}$ beyond photosynthetic $T_{opt}$ declines photosynthesis due it's peaked response to temperature.
 Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
 2017; @slotSituTemperatureResponse2017).
-Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g.
+Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g.
 @slotSituTemperatureResponse2017; Smith et al.
@@
 Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].
-Leaves in tropical upper canopies have shown to experience air temperatures well above $40-60{^\circ}C$ on a regular basis, if $T_{leaf}$ is not regulated through $gs$ or adapated, heat stress can loss of canopy leaves [**REF**]
+Leaves in tropical upper canopies have shown to experience in-situ $T_{air}$ well above $40-60{^\circ}C$ on a regular basis.
+If $T_{leaf}$ is not regulated or adapated, heat stress can loss of canopy leaves [**REF**]
 
@@
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that optimum temperature $T_{opt}$ of $J_{max}$ increased with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
@@
 2020].
-Leaf thermal thresholds strongly reflect species adaptation to their microenvironment rather than macroclimate conditions [@feeley_thermal_2020; Slot et al.
+Leaf thermal tolerances strongly reflect species adaptation to their microenvironment rather than macroclimate conditions [@feeley_thermal_2020; Slot et al.
 2020; Perez and Feely 2020a] and phylogeny [Slot et al.
@@
 2017).
-With elevation, $T_{50}$ and $T_{crit}$ decrease, as seen in tropical foress and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], and vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al.
+With elevation, $T_{50}$ and $T_{crit}$ decrease, as seen in tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], and vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al.
 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020].
 Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$.
-On the other hand, leaves of species with greater construction costs, greater LMA, had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020].
+On the other hand, leaves of species with greater construction costs--greater LMA-- had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020].
 However, higher photosynthetic heat tolerance does not necessary increase the breadth of leaf thermal safety margin.
-Rather, across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$) [@perezPhotosyntheticHeatTolerances2020]
+Rather, @perezPhotosyntheticHeatTolerances2020 found that, across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$)
 
 **Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
-Thermal tolerance $T_{crit}$ and $T_{50}$ have been predominantly measured on sun leaves; thus there remains limited information on leaves with varying light exposures and vertical height gradient.
+Because $T_{crit}$ and $T_{50}$ over the years have been predominantly measured on sun leaves; there remains limited information on sun- and shade-leaves  and at vertical height gradient.
 
-For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, though the results weren't significant and both were measured near ground level [@slotPhotosyntheticHeatTolerance2019a].
-In tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions had higher heat tolerance (by $2{^\circ}C$) than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
-In both species photosynthetic heat tolerance increased seasonally from June to August [Knapp et al.
-1994]
+Knapp et al.
+1994 showed that in tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
+In both species photosynthetic heat tolerance increased seasonally from June to August.
+However, more recetly @slotPhotosyntheticHeatTolerance2019a found that though $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in moist tropical forest in Panama, the results weren't significant, with both measured near ground level.
 
 In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient.
-For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions.
-Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, low-wind speed, $T_{air} and greater canopy [@curtis_intracanopy_2019].
+For *Acacia Papyrocarpa* (Benth.) trees in an arid region in Australia, higher composite climate stress and indeed, $T_{50}$ was greater in the lower north-facing canopy than other canopy positions.
+Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, low-wind speed and greater $T_{air} [@curtis_intracanopy_2019].
 
@@
 
-**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
+**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves (Chen et al.
+2020), but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**
 
-While respiration tends to increase along the vertical gradient, with greater respiration in canopy leaves [@bolstad_foliar_1999; Weerasinghe et al.2014], temperature sensitivity of respiration--$Q_{10}$--shows conflicting findings with no definite trend.
-
-Where $Q_{10}$ is greater in exposed upper-canopy leaves than in shaded lower-canopy leaves in across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] but in other feild studies $Q_{10}$ doesnot vary across species vertical profiles in a lowland tropical Australian forest study [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta, with a slight tendency to increase down the canopy relative to upper-canopy leaves as T increases (*see their Fig.
+Across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, $Q_{10}$ is observed to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+However, in a lowland tropical Australian forest study, $Q_{10}$ doesnot vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increases (*see their Fig.
 1*)[@bolstad_foliar_1999].
@@
 
-Additionally, a fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]
+Further information from fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]
 
@@
 **Isoprene production**
-High emission rates occur in the mid-canopy, and even from very small statured trees and shrubs, at a site in the Brazilian Amazon (Taylor et al., in prep).
+High emission rates occur in the mid-canopy, and even from very small statured trees, at a site in the Brazilian Amazon (Taylor et al., in prep).
 The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
-Within species, isporene production scales with light/ T; however regardless of light, isoprene emission increases with T such as in understory shaded leaves?
+Within species, isporene production scales with light and temperature; together and independtly?
 
@@
 **Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
+More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
 There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)*
 
-More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
-Biophysically, greater vapor pressure deficit, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al.
+Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al.
 2016].
-On the other hand, greater relative humidity can do the opposite, where it stifles evapotranspiration thereby increasing leaf temperatures, particularly in canopy tops where irradiance can have a greater effect [Fig.3;@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-However, soil water loss in conditions of drought effect larger trees more than small trees in less than a few days.
-Primarily because, greater biophysical evaporative demand in the canopy cannot meet long-distance water transport from the ground due to limited soil moisture [Norway spruce study, bennett; @mcgregor_tree_nodate].
+On the other hand, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in canopy tops where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+However, soil water loss in conditions of drought effect larger trees more than small trees.
+Primarily because, greater biophysical evaporative demand in the canopy cannot meet long-distance water transport from ground-level limited soil moisture [Norway spruce study, bennett; @mcgregor_tree_nodate].
 
-On the basis of phenology, because understory trees leaf out earlier than canopy due to ontogenetic variation in tree size [Vitasse et al.
-2013], understory growth is more sensitive to annual warming than canopy - showing earlier spring onset and later autumn senescense, particularly in temperate deciduous forests [Zohnner and Renner, 2019].
+On the basis of phenology, because understory trees leaf out earlier than canopy [Vitasse et al.
+2013], understory growth is more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate deciduous forests [Zohnner and Renner, 2019].
 However, late spring short-term heat waves negatively impact partially- expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].
-
-If we expect 'shade' plants to have a different heat sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered.
 
@@
 ## Global change responses
- *(this section is currently just a bunch of text moved from elsewhere/ loose notes)*
+
+### canopy disturbance
+
+Canopy structural composition and $LAI$ largely influence understory species composition and soil fertility [@majasalmiImpactTreeCanopy2020], buffering capacity, which in-turn is influenced by the hydrological cycle [@davis_microclimatic_2019].
+Each of these are subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
+Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and decreases tree canopy circumference [@aussenacInteractionsForestStands2000a].
+With larger trees suffering disproportionally greater hydraulic stress during drought, can contribute to increasing larger tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
+2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
+Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020], or increase in mortality of big trees (e.g.
+@senfCanopyMortalityHas2018a) would increase canopy roughness, also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
+
+Increasing global temperatures can increase canopy $T_{leaf}$ sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
+Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Each of these changes have potential feedbacks to climate change.
 
 ### warming
-
-- Influence of increasing temperatures on species compositions and feedbacks to future forest microclimates and function?
 
@@
 Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
-Plant communities in forest microclimates are already undergoing thermophilizatioN [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to microclimatic fauna [@scheffersMicrohabitatsReduceAnimal2013] .
+Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to microclimatic fauna [@scheffersMicrohabitatsReduceAnimal2013] .
 Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-Additionally, warming in other cases is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, in closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
+Whereas warming in other cases is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
+Evidence shows that warming and heat waves can also drastically alter canopy and understory phenology and forest photosynthesis [@zohnerOngoingSeasonallyUneven2019].
 
-Additionally, warming and heat waves can alter canopy and understory phenology and forest photosynthesis [@zohnerOngoingSeasonallyUneven2019].*
-
-Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth).
+Additionally, warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth).
 Differential mortality within the canopy structure will alter the profile of metabolism and hence emergent forest function.
@@
 If we know how metabolism maps to the thermal profile, then the observed forest structural changes can inform prediction of forest function.
-
-### canopy disturbance
-
---Variation across horizontal biophysical gradients (climate, soils, etc.)(E.O)
-
-*(there's a lot of literature on canopy disturbance impacts--tap into that in this section)*
-
-Canopy structural composition and $LAI$ largely influence understory species composition and soil fertility [majasalmiImpactTreeCanopy2020], including buffering, which in-turn is influenced by the hydrological cycle [@davis_microclimatic_2019].
-Each of these are subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
-Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory trees to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].
-With larger trees suffering disproportionally greater hydraulic stress during drought, can contribute to increasing larger tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
-2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
-Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020], or increase in mortality of big trees (e.g.
-@senfCanopyMortalityHas2018a) would increase canopy roughnessare, also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-
-Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Open-forest might experience greater microclimate warming than forests, having implications for forest regeneration, particularly in high-altitude regions [@vonarxSpatiotemporalEffectsForest2012a].
-Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
-Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Each of these changes have potential feedbacks to climate change.
 

### 2021-05-13 vertical-thermal-review.Rmd
total so far, 7962 words

@@
 
-**The biophysical environment across the vertical gradient from the forest floor to the the top of the canopy is in large part determined by the structure of the forest itself.** 
+**The biophysical environment across the vertical gradient from the forest floor to the top of the canopy is in large part determined by the structure of the forest itself.** 
 Here, we define the biophysical environment to include the physical structure of the vegetation and associated physical conditions.
@@
 ### Foliage density
-**Canopy foliage, which is vertically structured in correspondence with the size, structure of trees in the forest, acts as the primary physical barrier between the atmosphere and the forest floor, producing a buffering effect on multiple aspects of the understory conditions (Fig.
-`r fig_NEON_vertical`).**
+**Canopy foliage, acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions and is critical in influencing-and is largely influenced by-the vertical biophysical gradient.(Fig.
+`r fig_NEON_vertical`).**The distribution of this foliage is structured in correspondence with the size, and varies across tree structures and forests.
 Leaf area density.... *description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references)*
@@
 `r fig_NEON_vertical`; @koike_leaf_2001].**
-Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor.
-This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig.
-`r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].
-Light reaching forest floors that is incident to canopy tops ranges from ~1-2% in tropical forests  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate forests[@bartemucciOverstoryInfluencesLight2006] to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997].
-Sunflecks, caused by small canopy gaps and wind-induced canopy movements also contribute to the shaded forest regions that receive 50-80% of their light for photosynthesis during these short exposures [@waySunflecksTreesForests2012].
-Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
+Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density.
+This gradient is most pronounced under mixed and broad-leaf forests where light reaching forest floors, incident to the canopy-top range from ~1-2% to 4-5%, respectively [Fig.
+`r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006] compared to needle-leaf forest floors that receive ~17-34% [@baldocchiSeasonalVariationEnergy1997].
+Sunflecks, caused by small canopy gaps and wind-induced canopy movements are an important light-source for shaded forest regions that rely on sunflecks for photosynthesis [@waySunflecksTreesForests2012].
+In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
 
@@
 **Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
-Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom.
-Consistent with these results, higher wind speeds at greater heights have been observed in both closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].
+Across all **#** forested NEON sites, and in closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019] wind speeds increase with height.
+With daily maximum wind speeds averaging up to **#** $m s^-1^$ higher at the top of the NEON vertical profile than at the bottom
 
@@
 
-The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than open-forests, nearby clearings [*Defrenne et al.
+The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, more than open-forests, nearby clearings [*Defrenne et al.
 2021*; @vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
@@
 2017; @mcgregor_tree_nodate].
-However, under colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings [@zellweger_seasonal_2019; *Defrenne et al.
-2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate] where greater radiative heat loss in exposed areas results in lower $T_{min}$.
+However, under colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings where greater radiative heat loss in results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
+2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate].
 
@@
 
-**The combination of biophysical variables reviewed above drive leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn has a strong influence on leaf metabolism (see following section).**
-Leaves adapt to their biophysical environment to maximize carbon assimilation, by maintaining optimum $T_{leaf}$ in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-As a result, leaves can be cooler than air at high temperature and warmer than air at low temperature ('limited homeothermy') or can be drastically warmer than $T_{air}$ ('megathermy').
-In some cases, $T_{leaf}$ can be equalled to $T_{air}$ ('poikilothermy') [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+**Vertical biophysical drivers reviewed above influence heat and water stress patterns, and leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn shapes leaf metabolism (see following section).**
+Along the complex biophysical dynamics, leaves adjust to their environment by maintaining optimum $T_{leaf}$ for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Therefore, leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy).
+In some cases, $T_{leaf}$ can be equalled to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
 
-$T_{leaf}$ itself is determined by the energy balance of a leaf, influenced by leaf traits, and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
-`r fig_leaf_T`).
-The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
+$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles  [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
+`r fig_leaf_T`) where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
 2; @michaletzPlantThermoregulationEnergetics2015].
@@
 
-Leaf thermoregulation is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
-Leaves can regulate their temperature by dissapating heat through stomatal conductance ($g_s$), even in drastic biophysical conditions [**REF**] but when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, VPD or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
+This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because stomatal conductance ($g_s$) facilitates heat dissipation, when water demand exceeds water transport, driven by irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
 
-Incoming solar radiation is a strong driver of $T_{leaf}$, relative to $T_{air}$.
-With maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, but $g_s$ limitation can cause solar radiation to substantially elevate $T_{leaf}$ greater than $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
 
-Greater wind speed (Fig.
-`r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ by increasing boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a].
-Therefore, under hot conditions, canopy-leaves exposed to higher wind speeds would be most effective at cooling with high $g_s$ [Drake et al.
-2018].
-At low wind conditions, small $g_a$ to heat transfer can increase $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017] where boundary layer resistance increases with thickness of air layer at the leaf surface.
-This tends to be greater in buffered leaves [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]
+High wind speed (Fig.
+`r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a].
+Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling [Drake et al.
+2018] but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]
 
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
-While stomatal conductance is inherently greater in exposed leaves than buffered shaded leaves, factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--are also greater with exposure.
-As a result, exposed leaves can experience steeper $g_s$ limitation than buffered leaves, driving an increase in $T_{leaf}$, further influenced by irradiance and $T_{air}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
-On the other hand, due to inherently lower $g_s$ in shade leaves, they can remain cooler under buffered conditions and substantially warmer under increased $T_{air}$ or irradiance.
+ Inherent leaf $gs$ is an important factor for $T_{leaf}$ which is typically greater across species exposed sun-leaves than buffered shade leaves.
+However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
+On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress.
 
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.**  Smaller leaves remain within a few degrees of $T_{air}$,  *with coolest leaves at intermediate sizes (~10mm)* due to higher $g_{a}$ to heat transfer [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].
-Whereas, large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation or $T_{air}$ with stomata closed  [@leighInfluenceLeafSize2017; @fauset_differences_2018].
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves tend to heat and cool faster due to higher $ga$, maintaining lower $T_{leaf}$-$T_{air}$ than large leaves that take longer to heat and cool due to low $ga$ [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].
 
 ### In-situ observations
-**Vertical profiles in microclimate (Fig.
-`r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed?
-and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig.
+**The principles outlined above, vertical microclimate profile (Fig.
+`r fig_NEON_vertical`), leaf traits (see following section), and evapotranspiration (where is this discussed?
+and do we say anything about evaporation?) together shape in-situ $T_{leaf}$ observations (Fig.
 `r fig_leaf_T`).**
-Canopy buffering capacity largely influences vertical $T_{leaf}$ dynamics.
-Particularly, in open-forests with lower buffering capacity, $T_{leaf}$ is greater in lower-canopy than upper canopy, reflecting microclimate conditions of vertical air mixing, light permeating through canopy layers, still air in the lower canopy and heating from the ground, [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+
+Particularly, in open-forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground can collectively increase $T_{leaf}$ [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
 Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
-Greater $RH$ in inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
+Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
 Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
@@
 
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Closed-canopy forests, such as tropical and semi-deciduous tropical, and temperate forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies due to greater heat and water stress from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016].
-Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper-canopies susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Closed-canopy forests, such as tropical and semi-deciduous tropical, and temperate forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016].
+Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
 Comparing forests, upper-canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper-canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
@@
 
-**Leaf thermoregulation largely depends leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016].
-Therefore, are critical in maintaining desirable $T_{leaf}$ and in shaping leaf metabolism (following section) across forest vertical strata (Figs.
+**Leaf thermoregulation is controlled by leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016].
+Therefore, are critical in maintaining desirable $T_{leaf}$ and shaping leaf metabolism (following section) across forest vertical strata (Figs.
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
 
-Leaf morphological adaptation to the local microenvironement takes place during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004].
-Degrees of direct to diffused light and resultant water demands shape sun vs.
-shade leaves and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
-Therefore, taller tree canopy-leaves dramatically vary from canopy-interior and understory due to the differences in biophysical conditions (Figs.
+Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004] where degrees of direct to diffused light and resultant water demands shape sun vs.
+shade leaves, and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
-While leaf traits vary primarily on the axis of light, at greater tree-heights, both theory and empirical evidence indicate that other factors-for example, tree-height or exposure to evaporative demand (Fig.
-`r fig_NEON_vertical`)- are also important in shaping at least a subset of traits.
+But at greater tree height, biological--tree-height--and biophysical--light, temperature, and evaporative demand (Fig. `r fig_NEON_vertical`)--also play a role in shaping leaf traits
 
@@
 
-Disentangling the influences of hydraulic constraints, light and other biophysical drivers that vary with height (Fig.
-`r fig_NEON_vertical`) on leaf traits remains an important area for research, but is beyond the scope of the current review.
-Our primary interest is in how leaf traits and function vary across the vertical gradient from the top of the canopy to the understory in forests.
-For the purposes of interpreting the results presented here, we note that patterns across light and height gradients are generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
+However, disentangling tree-height related biological and biophysical drivers (Fig.
+`r fig_NEON_vertical`) on leaf traits remains an important area for research, which is sparse, but is beyond the scope of the current review.
+Here, our primary interest is in how leaf traits and function vary across the vertical gradient from the canopy-top to the understory in forests.
+We note that patterns across light and height gradients are generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
 
@@
 
-Going from shaded to exposed canopy position, morphological adaptation to dissapate heat increases.
-Sun leaves are often thicker, smaller, hairier, with greater dissection--lobes and serration--[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al.
-2010], with more vertical leaf angles than shade leaves [@niinemetsAdjustmentFoliageStructure1998].
-All, collectively regulate $T_{leaf}$-photosynthetic balance.
-On the other hand, shaded leaves invest more in $SLA$ for light capture than in heat dissapation, therefore have lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
+Exposed-canopy leaves are better adapted to heat and water stress than shaded-leaves where  sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al.
+2010], with adaptive strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998].
+All, collectively regulate $T_{leaf}$-photosynthetic balance.On the other hand, shaded leaves invest more in light capture than in heat dissipation, therefore have lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
 2019].
-Additionally, under homogeneous light conditions, sun-leaves in different canopy orientations are shaped on the basis of $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
+Under homogeneous light conditions, sun-leaves in different canopy orientations are shaped more based on $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-Nitrogen along the vertical gradient, is constantly re-mobilized from senescensed to growing leaves [@niinemetsPhotosynthesisResourceDistribution2007a] where $N_{area}$, with LMA, leaf soluble sugars, starch and non-structural carbons are greater [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al.
-2014].
-Additionally, $N_{area}$ optimizes to prevent photoinhition in sun leaves during direct light and shade leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].
+The distribution of biochemical traits enable photosynthesis and metabolic processes.
+Particularly $N_{area}$, is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a] and is optmized to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].
 
@@
 
-Pools of xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function--plays a role in converting excess light into heat energy dissipation, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998], increases proportional light gradients.
+Xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function--are greater in sun-leaves where they enable heat and light tolerance by converting excess light to heat dissapation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+$VAZ$, and in general biochemical, constraints can result in lower temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a]
 
@@
 
-Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], increase in LMA, photosynthetic capcity per area, and stomatal conductance with height are attributed to ontogenetical differences, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Across  multi-species canopy and understory, trait adaptation to spatio-temporal light and biophysical gradient varies largely with inherent specie-specific trait differences such as leaf shape, size, $gs$ levels and therefore influences individual $T_{leafs}$.
+Across tree-size ontogeny of the same species, while fundamental trait plasticity to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], increase in LMA, photosynthetic capcity per area, and $gs$ with height are attributed to ontogenetical differences, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Across  multi-species canopy and understory, trait adaptation to spatio-temporal light and biophysical gradient varies greatly with inherent specie-specific trait differences.
 As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
@@
 
-$T_{leaf}$ beyond photosynthetic $T_{opt}$ declines photosynthesis due it's peaked response to temperature.
+$T_{leaf}$ beyond photosynthetic $T_{opt}$, declines photosynthesis due it's peaked response to temperature.
 Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
@@
 Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
@@
 
-Typical $T_{50}$ values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al.
+Across latitude and climate, typical $T_{50}$ values are ~45-50$^\circ$C (O' sullivan et al.
 2017).
@@
 On the other hand, leaves of species with greater construction costs--greater LMA-- had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020].
-However, higher photosynthetic heat tolerance does not necessary increase the breadth of leaf thermal safety margin.
-Rather, @perezPhotosyntheticHeatTolerances2020 found that, across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum leaf temperatures, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins (difference between maximum leaf temperature and $T_{50}$)
+However, higher photosynthetic heat tolerance does not necessarily increase the breadth of leaf thermal safety margin.
+Rather, @perezPhotosyntheticHeatTolerances2020 found that, across species sun leaves, higher photosynthetic heat tolerance experienced higher $T_{leaf_{max}}$, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf}$-$T_{50}$)
 
 **Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
-Because $T_{crit}$ and $T_{50}$ over the years have been predominantly measured on sun leaves; there remains limited information on sun- and shade-leaves  and at vertical height gradient.
+$T_{crit}$ and $T_{50}$ over the years have been predominantly measured on sun leaves, hence there is limited information on sun- and shade-leaves  and at vertical height gradient.
 
@@
 In both species photosynthetic heat tolerance increased seasonally from June to August.
-However, more recetly @slotPhotosyntheticHeatTolerance2019a found that though $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in moist tropical forest in Panama, the results weren't significant, with both measured near ground level.
+However, more recetly @slotPhotosyntheticHeatTolerance2019a found that though $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Paname tropical forest, the results weren't significant, with both measured near ground level.
 
@@
 For *Acacia Papyrocarpa* (Benth.) trees in an arid region in Australia, higher composite climate stress and indeed, $T_{50}$ was greater in the lower north-facing canopy than other canopy positions.
-Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, low-wind speed and greater $T_{air} [@curtis_intracanopy_2019].
+Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, low-wind speed and greater $T_{air}$ [@curtis_intracanopy_2019].
 
@@
 Across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, $Q_{10}$ is observed to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
-However, in a lowland tropical Australian forest study, $Q_{10}$ doesnot vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increases (*see their Fig.
+However, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
 1*)[@bolstad_foliar_1999].
@@
 However, soil water loss in conditions of drought effect larger trees more than small trees.
-Primarily because, greater biophysical evaporative demand in the canopy cannot meet long-distance water transport from ground-level limited soil moisture [Norway spruce study, bennett; @mcgregor_tree_nodate].
+Primarily because, limited soil-water cannot meet long-distance water transport to the canopy's biophysical evaporative demand [Norway spruce study, bennett; @mcgregor_tree_nodate].
 
 On the basis of phenology, because understory trees leaf out earlier than canopy [Vitasse et al.
-2013], understory growth is more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate deciduous forests [Zohnner and Renner, 2019].
+2013], understory growth is more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019].
 However, late spring short-term heat waves negatively impact partially- expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].

### 2021-05-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 These trait-patterns contribute to [understanding] $T_{leaf}$ dynamics across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] where Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
-The characterization of trait relationships and leaf responses through the whole vertical canopy is becoming a strong research need because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
+The characterization of trait relationships and responses through the whole vertical canopy is becoming a strong research need because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
 
@@
 
-The distribution of biochemical traits enable photosynthesis and metabolic processes.
-Particularly $N_{area}$, is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a] and is optmized to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].
-
-##### *(Carotenoids/Xanthophyll cycle pigments)*
-
-Xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function--are greater in sun-leaves where they enable heat and light tolerance by converting excess light to heat dissapation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-$VAZ$, and in general biochemical, constraints can result in lower temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a]
-
-##### Isoprene production
-Isoprene production *capability* increases vertically with light and therefore facilitates high heat tolerance in emitting species' canopy positons [Sharky]
+The distribution of biochemical traits enable photosynthesis and metabolic processes, and protect against heat and light damage in sun-leaves.
+$N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging and photoprotective function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+Constraints to biochemical traits, particularly $VAZ$ can curb temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a].
+Additionally, isoprene producing species are particularly advantageous in that it facilitates high heat tolerance in canopy positions[*REF*]
 
@@
 
-Across tree-size ontogeny of the same species, while fundamental trait plasticity to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], increase in LMA, photosynthetic capcity per area, and $gs$ with height are attributed to ontogenetical differences, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Across  multi-species canopy and understory, trait adaptation to spatio-temporal light and biophysical gradient varies greatly with inherent specie-specific trait differences.
-As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].
+Across forest species widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits.
+Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007].
+Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Therefore, in inter-specific trait variation, while general patterns of thermoregulation can be observed, some species might be more suspectable to heat stress at a vertical gradient than others [@fauset_differences_2018].
 

### 2021-05-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
- Inherent leaf $gs$ is an important factor for $T_{leaf}$ which is typically greater across species exposed sun-leaves than buffered shade leaves.
-However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
+ Inherent leaf $gs$ is an important factor for $T_{leaf}$ which is typically greater across species exposed sun-leaves due to stomata density than shade leaves.
+However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
 On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress.
@@
 Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Therefore, in inter-specific trait variation, while general patterns of thermoregulation can be observed, some species might be more suspectable to heat stress at a vertical gradient than others [@fauset_differences_2018].
+Therefore, while general patterns of thermoregulation can be observed across species, some might be more suspectable to heat stress at a vertical gradient than others [@fauset_differences_2018].
 
@@
 ##### *Leaf lifespan*
+
+New-leaf growth across the vertical gradient towards light constantly shades leaves underneath.
+Therefore, exposed-canopies have greater proportion of young leaves than shaded regions that have older leaves.
+Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger leaves that are devoloping traits than in mature leaves where traits have been developed.
+In such cases, change in light can result in dropping of leaves [Niinemetes, 2007].
 *Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
@@
 Deciduousness as a leaf trait is an adaptation to water and heat stress.
-Therefore, can influence biophysical gradient and $T_{leaf}$ distribution during leaf-loss [@rey-sanchez_spatial_2016].
-*In the tropics, among species that can be deciduous, greater proportion of deciduous individuals are in larger size classes that tend to lose their leaves more than understory species that are seldom deciduous.
-Deciduous species are also more prominent in drier sites compared to wet sites [@conditQuantifyingDeciduousnessTropical2000b]*
-
-*Spring and fall leaf phenology in temperate deciduous forests (Augsburger).* Leaf emergence in deciduous forests is delayed with height-ontogeny where understory species leaf-out earlier than canopy trees [Vitasse et al.
+Indeed, greater proportion of deciduous species are found in drier tropicaal regions than wetter,  and among species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b].
+Therefore, deciduous leaf-loss can influence biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
+Spatio-temporal leaf-emergence in deciduous-temperate forests also shapes biophysical gradient where leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
 2013; @seiwaChangesLeafPhenology1999].
@@
 
-### Stomatal conductance
+### Stomatal conductance [$gs]
 
-While exposed canopy sun-leaves are equipped to thermoregulate more efficiently than shaded-leaves, they are largely constrained by $gs$ [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]; and therefore their ability to photosynthesize as well.
-During the day, max transpiration rate and $gs$ increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010].
-During midday, stomtal closure in exposed leaves increases $T_{leaf}$, and decreases photosynthesis [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
+While $Gs$ mediates crucial leaf-atmosphere interaction across species, their exposed-sun-leaves, though equipped to thermoregulate efficiently are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
+Max transpiration rate and $gs$ is peaks in exposed-canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines through midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
 
@@
 2020).**
-This is primarily driven by the greater light available to sun leaves.
+This is primarily driven by the greater light available to sun-leaves.
 Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
 
-In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma.
+In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma.
 Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P (and water availability).
@@
 Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].
-Leaves in tropical upper canopies have shown to experience in-situ $T_{air}$ well above $40-60{^\circ}C$ on a regular basis.
-If $T_{leaf}$ is not regulated or adapated, heat stress can loss of canopy leaves [**REF**]
 
@@
 14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
+However, such a trend is not apparent among recent studies that compared temperature sensitivity of photosynthesis in sun and shade leaves.
+
 For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
-2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
+2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun- and shade-leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
+
+**Leaf thermal tolerance (Tcrit/ T50)**
+
+$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [Slot et al.
+2020].
+
+**Thermal tolerance (e.g., $T_{50}$) sparsely varies with exposure among recent studies, but there are no studies isolating effect of height.**
+
+150 years ago,  Sachs studied garden species $T_{50}$.
+Though, with little information on shade-leaves [*REF*].
+In 1994, Knapp et al.
+showed that in tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
+In both species photosynthetic heat tolerance increased seasonally from June to August.
+However, more recetly @slotPhotosyntheticHeatTolerance2019a found that though $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Panama tropical forest, the results weren't significant, with both measured near ground level.
+
+$T_{50}$ in *Acacia Papyrocarpa* (Benth.) trees in an arid region was observed to reflect microclimatic stress index along a vertical gradient where lower north-facing canopy showed higher composite climate stress and greater $T_{50}$ than other canopy positions, that correlated with low wind-speed greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
+
+We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
+I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot*
 
@@
 
-**Leaf thermal tolerance (Tcrit/ T50)**
-$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [Slot et al.
-2020].
-Leaf thermal tolerances strongly reflect species adaptation to their microenvironment rather than macroclimate conditions [@feeley_thermal_2020; Slot et al.
-2020; Perez and Feely 2020a] and phylogeny [Slot et al.
+Recent studies show that while leaf thermal tolerances across species reflects local microenvironmental conditions than regional macroclimate, [@feeley_thermal_2020; Slot et al.
+2020; Perez and Feely 2020a] $T_{50}$ does not vary with phylogeny [Slot et al.
 2020; Perez and Feeley 2020a].
 
-Across latitude and climate, typical $T_{50}$ values are ~45-50$^\circ$C (O' sullivan et al.
+Global $T_{50}$s across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
 2017).
-With elevation, $T_{50}$ and $T_{crit}$ decrease, as seen in tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], and vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al.
+Additionally, $T_{50}$ and $T_{crit}$, decrease with elevation in tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al.
 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020].
-Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$.
-On the other hand, leaves of species with greater construction costs--greater LMA-- had greater $T_{50}$ compared to species with low LMA [@slotLeafHeatTolerance2020].
-However, higher photosynthetic heat tolerance does not necessarily increase the breadth of leaf thermal safety margin.
-Rather, @perezPhotosyntheticHeatTolerances2020 found that, across species sun leaves, higher photosynthetic heat tolerance experienced higher $T_{leaf_{max}}$, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf}$-$T_{50}$)
-
-**Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
-$T_{crit}$ and $T_{50}$ over the years have been predominantly measured on sun leaves, hence there is limited information on sun- and shade-leaves  and at vertical height gradient.
-
-Knapp et al.
-1994 showed that in tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
-In both species photosynthetic heat tolerance increased seasonally from June to August.
-However, more recetly @slotPhotosyntheticHeatTolerance2019a found that though $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Paname tropical forest, the results weren't significant, with both measured near ground level.
-
-In other cases, leaf thermal tolerance reflects microclimatic stress along a vertical gradient.
-For *Acacia Papyrocarpa* (Benth.) trees in an arid region in Australia, higher composite climate stress and indeed, $T_{50}$ was greater in the lower north-facing canopy than other canopy positions.
-Greater $T_{50}$ in lower north-position also correlated with longer leaf thermal time constants, low-wind speed and greater $T_{air}$ [@curtis_intracanopy_2019].
-
-We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
-I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot*
+$T_{crit}$ is also recently found to be greater in species with large leaves, and slow heating and cooling time [higher thermal time constant] and $T_{50}$ is found to be greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA [@slotLeafHeatTolerance2020].
+@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance experienced higher $T_{leaf_{max}}$, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).
 
@@
 Across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, $Q_{10}$ is observed to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
-However, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], and across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
+In contrast, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
 1*)[@bolstad_foliar_1999].
@@
 Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al.
-2016].
-On the other hand, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in canopy tops where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-However, soil water loss in conditions of drought effect larger trees more than small trees.
-Primarily because, limited soil-water cannot meet long-distance water transport to the canopy's biophysical evaporative demand [Norway spruce study, bennett; @mcgregor_tree_nodate].
+2016].This can implication to canopies and gaps that reside at levels of greater $VPD$.
+Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+However, soil water loss during drought effects larger trees more than small trees because, limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canonopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
 
-On the basis of phenology, because understory trees leaf out earlier than canopy [Vitasse et al.
-2013], understory growth is more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019].
-However, late spring short-term heat waves negatively impact partially- expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis in reflushed leaves, more than understory saplings that are buffered from such conditions [Filewood and Thomas, 2013].
+Additionally, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019].
+However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves by more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis [and C uptake?] in reflushed leaves [Filewood and Thomas, 2013].
 
@@
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
-[Increase in humidity?]
-
-Additionally, interacting biophysical and biological stressors show that stress sensitivity is greater in buffered small trees, and decreases where tolerance increases with height, light with available water.
-Partly because large NSC reserves in taller trees due to available light allow them to tolerate greater number of stressors than buffered trees that are light and NSC limited [@niinemetsResponsesForestTrees2010a]
+Interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water.
+Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].
 
@@
 
-Canopy structural composition and $LAI$ largely influence understory species composition and soil fertility [@majasalmiImpactTreeCanopy2020], buffering capacity, which in-turn is influenced by the hydrological cycle [@davis_microclimatic_2019].
-Each of these are subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
-Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and decreases tree canopy circumference [@aussenacInteractionsForestStands2000a].
-With larger trees suffering disproportionally greater hydraulic stress during drought, can contribute to increasing larger tree mortality in forests around the world [@mcdowellMechanismsPlantSurvival2008; Brienen et al.
+The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are largely interconneced, resulting in feedbacks to another.
+Therefore, climate-driven changes, such as rising temperature, can have implications to forests at several scales.
+
+Canopy buffering capacity which is influenced by hydrological-cycle at large is crucial to microclimates [Davis et al.
+2019], where disturbance to canopy structural composition and $LAI$ influences understory species composition, soil fertility [*@majasalmiImpactTreeCanopy2020*] and species in microclimate conditions, particularly threatening flora [REF] fauna adapted to microrefugias [REF].
+Such disturbances to canopies are observed, in part, by frequent droughts, where less severe drought can decrease tree canopy circumference [@aussenacInteractionsForestStands2000a], and more severe droughts can contribute to increasing larger tree mortality in forests around the world [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
-Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020], or increase in mortality of big trees (e.g.
-@senfCanopyMortalityHas2018a) would increase canopy roughness, also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].
-
-Increasing global temperatures can increase canopy $T_{leaf}$ sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
-Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Hydraulic limitation in forests can also alters species thermoregulation [@sastry_leaf_2018].
+Increasing temperatures can increase canopy $T_{leaf}$ sometimes above ambient temperature, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
+Recent adavances in canopy studies point toward large loss of canopies in Europe caused by wind-induced disturbance and anthropogenic deforestation [Cornelius Senf, 2021; @senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020].
+Loss of big trees (e.g.@senfCanopyMortalityHas2018a ) would increase canopy roughness, and reductions in canopy cover also associated with reduced canopy structural complexity, can lead to altered biophysical gradient and microclimates with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 Each of these changes have potential feedbacks to climate change.
@@
 
-While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020].
-Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020].
-Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to microclimatic fauna [@scheffersMicrohabitatsReduceAnimal2013] .
+While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020] leading to understory plant communities already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to microclimatic fauna [@scheffersMicrohabitatsReduceAnimal2013] .
 Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
 Whereas warming in other cases is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
-Evidence shows that warming and heat waves can also drastically alter canopy and understory phenology and forest photosynthesis [@zohnerOngoingSeasonallyUneven2019].
+Warming and heat waves can also drastically alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
 
-Additionally, warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth).
+warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth).
 Differential mortality within the canopy structure will alter the profile of metabolism and hence emergent forest function.

### 2021-05-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Vertical biophysical drivers reviewed above influence heat and water stress patterns, and leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn shapes leaf metabolism (see following section).**
-Along the complex biophysical dynamics, leaves adjust to their environment by maintaining optimum $T_{leaf}$ for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Along the biophysical dynamics vertically, leaves adjust to their environment by maintaining optimum $T_{leaf}$, for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
 Therefore, leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy).
@@
 2009, 2010*).
-Isolating height (>35m) under constant light conditions show that, canopy $gs$ decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
+Isolating height (>35m) under constant light show that, canopy $gs$ decreases with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
 Morphologically, leaf size [Kaare H.
@@
 
-Within-canopy gradient of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- and therefore exhibit patterns in leaf anatomical, structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-These trait-patterns contribute to [understanding] $T_{leaf}$ dynamics across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] where Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+Within-canopy gradients of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- and therefore exhibit patterns in leaf anatomical, structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+These trait-patterns contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] where Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 The characterization of trait relationships and responses through the whole vertical canopy is becoming a strong research need because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
@@
 2019].
-Under homogeneous light conditions, sun-leaves in different canopy orientations are shaped more based on $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
+Under homogeneous light conditions, sun-leaves in different canopy orientations are shaped based on $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-New-leaf growth across the vertical gradient towards light constantly shades leaves underneath.
+New-leaf growth along the light gradient constantly shades leaves underneath.
 Therefore, exposed-canopies have greater proportion of young leaves than shaded regions that have older leaves.
@@
 
-### Stomatal conductance [$gs]
+### Stomatal conductance
 
-While $Gs$ mediates crucial leaf-atmosphere interaction across species, their exposed-sun-leaves, though equipped to thermoregulate efficiently are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
-Max transpiration rate and $gs$ is peaks in exposed-canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines through midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
+While $gs$ mediates crucial leaf-atmosphere interaction across species, exposed-sun-leaves, though equipped to thermoregulate efficiently are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
+Max transpiration rate and $gs$ peaks in exposed-canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines through midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
 
@@
 
-150 years ago,  Sachs studied garden species $T_{50}$.
-Though, with little information on shade-leaves [*REF*].
+150 years ago,  Sachs studied garden species $T_{50}$, though, with little information on shade-leaves [*REF*].
 In 1994, Knapp et al.
@@
 2017).
-Additionally, $T_{50}$ and $T_{crit}$, decrease with elevation in tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al.
+$T_{50}$ and $T_{crit}$, decrease with elevation in tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al.
 2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020].
-$T_{crit}$ is also recently found to be greater in species with large leaves, and slow heating and cooling time [higher thermal time constant] and $T_{50}$ is found to be greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA [@slotLeafHeatTolerance2020].
+$T_{crit}$ is also recently found to be greater in species with large leaves, and slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ is found to be greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA [@slotLeafHeatTolerance2020].
 @perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance experienced higher $T_{leaf_{max}}$, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).

### 2021-05-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
-Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
-The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
+Leading to, individually or in combination, can reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
 
@@
 ### Foliage density
-**Canopy foliage, acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions and is critical in influencing-and is largely influenced by-the vertical biophysical gradient.(Fig.
+**Canopy foliage, acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions and is critical in influencing--and is influenced by--the vertical biophysical gradient.(Fig.
 `r fig_NEON_vertical`).**The distribution of this foliage is structured in correspondence with the size, and varies across tree structures and forests.
@@
 Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density.
-This gradient is most pronounced under mixed and broad-leaf forests where light reaching forest floors, incident to the canopy-top range from ~1-2% to 4-5%, respectively [Fig.
+This gradient is most pronounced under broad-leaf and mixed forests where light reaching forest floors, incident to the canopy-top range from ~1-2% to 4-5%, respectively [Fig.
 `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006] compared to needle-leaf forest floors that receive ~17-34% [@baldocchiSeasonalVariationEnergy1997].
-Sunflecks, caused by small canopy gaps and wind-induced canopy movements are an important light-source for shaded forest regions that rely on sunflecks for photosynthesis [@waySunflecksTreesForests2012].
+Sunflecks, are another axis of light for shaded forest regions, that is caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
 In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
@@
 
-The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, more than open-forests, nearby clearings [*Defrenne et al.
+The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than open-forests, nearby clearings [*Defrenne et al.
 2021*; @vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
-2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-On the other hand, open-forests maintain higher daytime maximum temperature in the understory closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  sometimes warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].
+2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+Nighttime temperatures in pine forest understory, sometimes can be warmer than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].
 
@@
 2017; @mcgregor_tree_nodate].
-However, under colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings where greater radiative heat loss in results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
+In colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings where greater radiative heat loss results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
 2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate].
@@
 
+The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are highly interconneced, resulting in feedbacks to another.
+Therefore, climate-driven changes, such as rising temperature, can have implications to forests on several levels.
+
 ### canopy disturbance
 
-The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are largely interconneced, resulting in feedbacks to another.
-Therefore, climate-driven changes, such as rising temperature, can have implications to forests at several scales.
-
-Canopy buffering capacity which is influenced by hydrological-cycle at large is crucial to microclimates [Davis et al.
-2019], where disturbance to canopy structural composition and $LAI$ influences understory species composition, soil fertility [*@majasalmiImpactTreeCanopy2020*] and species in microclimate conditions, particularly threatening flora [REF] fauna adapted to microrefugias [REF].
-Such disturbances to canopies are observed, in part, by frequent droughts, where less severe drought can decrease tree canopy circumference [@aussenacInteractionsForestStands2000a], and more severe droughts can contribute to increasing larger tree mortality in forests around the world [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al.
+Canopy buffering is crucial to microclimate functioning, which is influenced by hydrological-cycle at large [Davis et al.
+2019].
+Therefore, frequent droughts causes canopy disturbance, where less severe drought can decrease tree canopy circumference [@aussenacInteractionsForestStands2000a], and more severe droughts shows increasing larger tree mortality in forests globally [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
 Hydraulic limitation in forests can also alters species thermoregulation [@sastry_leaf_2018].
-Increasing temperatures can increase canopy $T_{leaf}$ sometimes above ambient temperature, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Increasing temperatures shows increase in canopy $T_{leaf}$ above ambient temperature, and sometimes beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
-Recent adavances in canopy studies point toward large loss of canopies in Europe caused by wind-induced disturbance and anthropogenic deforestation [Cornelius Senf, 2021; @senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020].
-Loss of big trees (e.g.@senfCanopyMortalityHas2018a ) would increase canopy roughness, and reductions in canopy cover also associated with reduced canopy structural complexity, can lead to altered biophysical gradient and microclimates with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Recent adavances in canopy studies further shows large loss of canopies in Europe caused by wind-induced disturbance and anthropogenic deforestation [Cornelius Senf, 2021; @senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020].
+Loss of big trees (e.g.@senfCanopyMortalityHas2018a ) would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity and composition which changes understory species composition and soil fertility [*@majasalmiImpactTreeCanopy2020*].Leading to altered biophysical gradient with local heating effects, [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 Each of these changes have potential feedbacks to climate change.
@@
 
-While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020] leading to understory plant communities already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to microclimatic fauna [@scheffersMicrohabitatsReduceAnimal2013] .
+While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
+Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to fauna in microrefugias [@scheffersMicrohabitatsReduceAnimal2013] .
 Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-Whereas warming in other cases is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
-Warming and heat waves can also drastically alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
+Whereas, in other cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
+Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
 
-warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with Andean thermophilization (death is faster than recruitment and growth).
 Differential mortality within the canopy structure will alter the profile of metabolism and hence emergent forest function.

### 2021-05-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Tree height also plays a role in buffering where trees upto 20m with higher LAI can maintain lower mean and max temperature and vapor pressure deficit ($VPD$) in the understory.
-However, beyond a certain tree height, temperature and $VPD$ offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-
 ### Humidity and evaporative demand
@@
 Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a].
-RH, in combination with temperature, determines $VPD$, an important metric expressing the driving force of water loss from a leaf; tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+RH, in combination with temperature, determines vapor pressure deficit [$VPD$], an important metric expressing the driving force of water loss from a leaf; tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Increased tree height [upto 20m with higher LAI] has also shown to lower mean and max temperature and $VPD$ in the understory.
+Beyond a certain tree-height, the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
 

### 2021-05-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006] compared to needle-leaf forest floors that receive ~17-34% [@baldocchiSeasonalVariationEnergy1997].
-Sunflecks, are another axis of light for shaded forest regions, that is caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
+Sunflecks, are another axis of light for shaded forest regions, caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
 In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
@@
 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-Nighttime temperatures in pine forest understory, sometimes can be warmer than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].
 
-Seasonally, buffering capacity increases with water availablity [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry season [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
+Seasonally, buffering capacity increases with water availablity [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry seasons [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
 2017; @mcgregor_tree_nodate].
@@
 RH, in combination with temperature, determines vapor pressure deficit [$VPD$], an important metric expressing the driving force of water loss from a leaf; tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Increased tree height [upto 20m with higher LAI] has also shown to lower mean and max temperature and $VPD$ in the understory.
+Increased tree height [upto 20m with higher LAI] can to lower mean and max temperature and $VPD$ in the understory.
 Beyond a certain tree-height, the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
@@
 
-**Atmopsheric carbon dioxide (CO~2~) concentrations tend to be higher near ground level at night, but largely dissipate during the day [@koike_leaf_2001; @yang_spatial_1999].** 
+**Atmospheric carbon dioxide (CO~2~) concentrations tend to be higher near ground level at night, but largely dissipate during the day [@koike_leaf_2001; @yang_spatial_1999].** 
 Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
@@
 These trait-patterns contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] where Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
-The characterization of trait relationships and responses through the whole vertical canopy is becoming a strong research need because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
+The characterization of trait relationships and responses through the vertical canopy is a strong research need because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
 
@@
 2010], with adaptive strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998].
-All, collectively regulate $T_{leaf}$-photosynthetic balance.On the other hand, shaded leaves invest more in light capture than in heat dissipation, therefore have lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
+All, collectively regulate $T_{leaf}$-photosynthetic balance.On the other hand, shaded leaves invest more in light capture than in heat dissipation, with lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
 2019].

### 2021-05-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
#20

@@
 Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Sites include a tropical broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
-([issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)) ](NEON_height_profiles/figures/profile_all.png)
+Sites include a tropical broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).](NEON_height_profiles/figures/profile_all.png)
 

### 2021-05-17 vertical-thermal-review.Rmd
small edits, and knit in latest schematic

@@
 Across all **#** forested NEON sites, and in closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019] wind speeds increase with height.
-With daily maximum wind speeds averaging up to **#** $m s^-1^$ higher at the top of the NEON vertical profile than at the bottom
+With daily maximum wind speeds averaging up to **#** m s^-1^ higher at the top of the NEON vertical profile than at the bottom
 
@@
 
-**Vertical biophysical drivers reviewed above influence heat and water stress patterns, and leaf temperature $T_{leaf}$ distribution along the vertical gradient, which in turn shapes leaf metabolism (see following section).**
-Along the biophysical dynamics vertically, leaves adjust to their environment by maintaining optimum $T_{leaf}$, for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-Therefore, leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy).
+**Vertical biophysical drivers reviewed above influence heat and water stress patterns, thereby influencing leaf temperature $T_{leaf}$ distributions along the vertical gradient.**
+Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy).
 In some cases, $T_{leaf}$ can be equalled to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
@@
 **The principles outlined above, vertical microclimate profile (Fig.
-`r fig_NEON_vertical`), leaf traits (see following section), and evapotranspiration (where is this discussed?
-and do we say anything about evaporation?) together shape in-situ $T_{leaf}$ observations (Fig.
+`r fig_NEON_vertical`) and leaf traits (see following section) together shape in-situ $T_{leaf}$ observations (Fig.
 `r fig_leaf_T`).**
-
-Particularly, in open-forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground can collectively increase $T_{leaf}$ [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+In open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground can collectively increase $T_{leaf}$ [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
 Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
 Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$  relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
-Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig.
-`r fig_NEON_vertical`).
+Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
 
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** Closed-canopy forests, such as tropical and semi-deciduous tropical, and temperate forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016].
-Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
-Comparing forests, upper-canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper-canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
+**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
+Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016].
+Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$, making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
+Comparing forests, upper canopies of temperate deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
 
-### Leaf traits
+## Leaf traits
 
@@
 
-#### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
+### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
 

### 2021-05-17 vertical-thermal-review.Rmd
total 7821 words, -comments

@@
 Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004] where degrees of direct to diffused light and resultant water demands shape sun vs.
-shade leaves, and their necessary vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+shade leaves, and their vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
@@
 
-However, disentangling tree-height related biological and biophysical drivers (Fig.
-`r fig_NEON_vertical`) on leaf traits remains an important area for research, which is sparse, but is beyond the scope of the current review.
+Disentangling tree-height related biological and biophysical drivers (Fig.
+`r fig_NEON_vertical`) of leaf traits remains an important area for research, which is sparse, but is beyond the scope of the current review.
 Here, our primary interest is in how leaf traits and function vary across the vertical gradient from the canopy-top to the understory in forests.
@@
 
-Within-canopy gradients of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- and therefore exhibit patterns in leaf anatomical, structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Within-canopy gradients of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
 These trait-patterns contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] where Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
-The characterization of trait relationships and responses through the vertical canopy is a strong research need because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
+Further, there is a need for research to characterize trait relationships and responses through the vertical canopy, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
 
@@
 2010], with adaptive strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998].
-All, collectively regulate $T_{leaf}$-photosynthetic balance.On the other hand, shaded leaves invest more in light capture than in heat dissipation, with lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
+All, collectively regulate $T_{leaf}$-photosynthetic balance.
+Whereas, shaded leaves invest more in light capture than in heat dissipation, with lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
 2019].
-Under homogeneous light conditions, sun-leaves in different canopy orientations are shaped based on $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
+Under homogeneous light conditions, sun-leaves in canopy orientations are shaped based on canopy $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 The distribution of biochemical traits enable photosynthesis and metabolic processes, and protect against heat and light damage in sun-leaves.
-$N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging and photoprotective function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+$N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 Constraints to biochemical traits, particularly $VAZ$ can curb temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a].
-Additionally, isoprene producing species are particularly advantageous in that it facilitates high heat tolerance in canopy positions[*REF*]
+Additionally, isoprene-producing species are particularly advantageous in that they have higher heat tolerance in exposed-canopies [*REF*]
 
@@
 
-Across forest species widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
 However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits.
@@
 Therefore, exposed-canopies have greater proportion of young leaves than shaded regions that have older leaves.
-Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger leaves that are devoloping traits than in mature leaves where traits have been developed.
-In such cases, change in light can result in dropping of leaves [Niinemetes, 2007].
+Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger than mature leaves, due to devoloping traits that can acclimate.
+Whereas, change in light for mature leaves can result in leaves dropping [Niinemetes, 2007].
 *Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
@@
 Deciduousness as a leaf trait is an adaptation to water and heat stress.
-Indeed, greater proportion of deciduous species are found in drier tropicaal regions than wetter,  and among species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b].
-Therefore, deciduous leaf-loss can influence biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
+Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter.
+Among species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b] where, deciduous leaf-loss can influence biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
 Spatio-temporal leaf-emergence in deciduous-temperate forests also shapes biophysical gradient where leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
@@
 
-While $gs$ mediates crucial leaf-atmosphere interaction across species, exposed-sun-leaves, though equipped to thermoregulate efficiently are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
-Max transpiration rate and $gs$ peaks in exposed-canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines through midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
+While $gs$ mediates crucial leaf-atmosphere interactions across species, exposed-sun-leaves, though equipped with thermoregulatory effiency are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
+Where maximum transpiration rate and $gs$ peaks during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines during midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
 
@@
 **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
-Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-However, such a trend is not apparent among recent studies that compared temperature sensitivity of photosynthesis in sun and shade leaves.
-
-For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
-2019 ; @hernandezSimilarTemperatureDependence2020].Similar results were shown for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun- and shade-leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
+Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
+14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
+2019 ; @hernandezSimilarTemperatureDependence2020].
+Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], along the vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
 
@@
 
-150 years ago,  Sachs studied garden species $T_{50}$, though, with little information on shade-leaves [*REF*].
 In 1994, Knapp et al.
@@
 In both species photosynthetic heat tolerance increased seasonally from June to August.
-However, more recetly @slotPhotosyntheticHeatTolerance2019a found that though $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Panama tropical forest, the results weren't significant, with both measured near ground level.
+However, more recently @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Panama tropical forest, but results weren't significant, with both measured near ground level.
 
-$T_{50}$ in *Acacia Papyrocarpa* (Benth.) trees in an arid region was observed to reflect microclimatic stress index along a vertical gradient where lower north-facing canopy showed higher composite climate stress and greater $T_{50}$ than other canopy positions, that correlated with low wind-speed greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
+Leaf thermal tolerances across species might reflect local microenvironmental conditions than regional macroclimate [@feeley_thermal_2020; Slot et al.
+2020] where $T_{50}$ in *Acacia Papyrocarpa* (Benth.) trees in an Australian arid region showed higher composite climate stress and greater $T_{50}$ than other canopy positions, that correlated with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
 
@@
 
-Recent studies show that while leaf thermal tolerances across species reflects local microenvironmental conditions than regional macroclimate, [@feeley_thermal_2020; Slot et al.
-2020; Perez and Feely 2020a] $T_{50}$ does not vary with phylogeny [Slot et al.
-2020; Perez and Feeley 2020a].
+Global $T_{50}$s across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
+2017) and $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
 
-Global $T_{50}$s across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
-2017).
-$T_{50}$ and $T_{crit}$, decrease with elevation in tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017], vary with species [@krauseHightemperatureToleranceTropical2010a] and leaf traits [Sastry et al.
-2018, @zhang_photosynthetic_2012; @slotLeafHeatTolerance2020].
-$T_{crit}$ is also recently found to be greater in species with large leaves, and slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ is found to be greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA [@slotLeafHeatTolerance2020].
-@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance experienced higher $T_{leaf_{max}}$, and showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).
+Additionally, $T_{50}$ varies with species [@krauseHightemperatureToleranceTropical2010a]and leaf traits and does not vary with phylogeny [Slot et al.
+2021; Perez and Feeley, 2020a].
+Recently, @slotLeafHeatTolerance2020 found that $T_{crit}$ was greater among tropical species' large leaves, with slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA.
+@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance, experienced higher $T_{leaf_{max}}$, but showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).
 
@@
 @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
-
-Further information from fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021]
 
@@
 Whereas, in other cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
-Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
+Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].Further information from fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
 

### 2021-05-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
-But at greater tree height, biological--tree-height--and biophysical--light, temperature, and evaporative demand (Fig. `r fig_NEON_vertical`)--also play a role in shaping leaf traits
+At greater tree height, biological--tree-height--and biophysical--light, temperature, and evaporative demand (Fig.
+`r fig_NEON_vertical`)--also play a role in shaping leaf traits
 
@@
 2009, 2010*).
-Isolating height (>35m) under constant light show that, canopy $gs$ decreases with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
+Isolating height (>35m) from light shows that, canopy $gs$ decreases with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
 Morphologically, leaf size [Kaare H.
@@
 Disentangling tree-height related biological and biophysical drivers (Fig.
-`r fig_NEON_vertical`) of leaf traits remains an important area for research, which is sparse, but is beyond the scope of the current review.
+`r fig_NEON_vertical`) of leaf traits remains an important area for research, which is sparse, is beyond the scope of the current review.
 Here, our primary interest is in how leaf traits and function vary across the vertical gradient from the canopy-top to the understory in forests.
@@
 
-Within-canopy gradients of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-These trait-patterns contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] where Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
+Within-canopy gradients of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020] that contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
 Further, there is a need for research to characterize trait relationships and responses through the vertical canopy, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].

### 2021-05-17 vertical-thermal-review.Rmd
minor changes - vertical profiles
@NidhiVinod , please check if my change to the schematic figure caption works (#27 )

@@
 ![**Figure `r fig_schematic`.
-DRAFT schematic of a forest summarizing most important gradients--sort of a "graphical abstract".** [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
+Schematic summarizing vertical gradients in forests with dense canopies during daytime growing season conditions.** Patterns summarized here tend to be weaker, or sometimes reversed, under more open canopies, during seasons of deciduous leaf loss, or at nighttime.
+[Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
 ](schematics/vertical thermal sensitivity image.png)
@@
 ### Foliage density
-**Canopy foliage, acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions and is critical in influencing--and is influenced by--the vertical biophysical gradient.(Fig.
-`r fig_NEON_vertical`).**The distribution of this foliage is structured in correspondence with the size, and varies across tree structures and forests.
+**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions.
+It is critical in influencing -- and in turn is influenced by -- the vertical biophysical gradient (Fig.
+`r fig_NEON_vertical`).**
+The distribution of this foliage is structured in correspondence with the distribtuion of tree sizes within a forest.  
 Leaf area density.... *description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references)*
@@
 ### Air temperature 
-**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
+**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs.
+`r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
 Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019].
 
-The strength of this buffering is greater under dense-canopy than open-canopy forests; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than open-forests, nearby clearings [*Defrenne et al.
+The strength of this buffering is greater under dense than open canopies; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than open-forests, nearby clearings [*Defrenne et al.
 2021*; @vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+(*This sentence is too complex-- hard to understand*)
 
-Seasonally, buffering capacity increases with water availablity [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry seasons [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
+Seasonally, buffering capacity increases with water availability [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry seasons [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
 2017; @mcgregor_tree_nodate].
 In colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings where greater radiative heat loss results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
-2021*; @nakamuraForestsTheirCanopies2017; @mcgregor_tree_nodate].
+2021*; @nakamuraForestsTheirCanopies2017].
 

### 2021-05-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology and ecosystem ecology in several ways.**
+**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology and ecosystem ecology.**
+
 Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in understory versus canopy conditions.

### 2021-05-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
-Leading to, individually or in combination, can reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
+Leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
 
@@
 Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density.
-This gradient is most pronounced under broad-leaf and mixed forests where light reaching forest floors, incident to the canopy-top range from ~1-2% to 4-5%, respectively [Fig.
+This gradient is most pronounced under broad-leaf and mixed forests where light reaching forest floors, incident to canopy-tops range from ~1-2% to 4-5%, respectively [Fig.
 `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006] compared to needle-leaf forest floors that receive ~17-34% [@baldocchiSeasonalVariationEnergy1997].
-Sunflecks, are another axis of light for shaded forest regions, caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
+Sunflecks, offer another light source for shaded forest regions, caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
 In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
-Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019].
-
-The strength of this buffering is greater under dense than open canopies; with daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than open-forests, nearby clearings [*Defrenne et al.
-2021*; @vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al.
-2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+The $T_{air}$ difference between understory and canopy temperature vary based on multiple factors such as radiation, local topography, vertical air mixing and cloud cover [*Defrenne et al.
+2021*].Dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies, where dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al.
+2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
 (*This sentence is too complex-- hard to understand*)
 
-Seasonally, buffering capacity increases with water availability [@davisMicroclimaticBufferingForests2019] and is greater in wet than dry seasons [@rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
-2017; @mcgregor_tree_nodate].
-In colder conditions, dense-canopies maintain warmer higher minimum temperatures [$T_{min}$] in the understory relative to open-canopy forests and nearby clearings where greater radiative heat loss results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
+Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
+2017; @mcgregor_tree_nodate]whereas in colder conditions, dense-canopy understories experience higher minimum temperatures [$T_{min}$] relative to open-canopy understories and nearby clearings where greater radiative heat loss results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
 2021*; @nakamuraForestsTheirCanopies2017].

### 2021-05-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
-2017; @mcgregor_tree_nodate]whereas in colder conditions, dense-canopy understories experience higher minimum temperatures [$T_{min}$] relative to open-canopy understories and nearby clearings where greater radiative heat loss results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
+2017; @mcgregor_tree_nodate].
+Whereas in colder conditions, dense-canopy understories experience higher minimum temperatures [$T_{min}$] relative to open-canopy understories and nearby clearings where greater radiative heat loss results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
 2021*; @nakamuraForestsTheirCanopies2017].
@@
 Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a].
-RH, in combination with temperature, determines vapor pressure deficit [$VPD$], an important metric expressing the driving force of water loss from a leaf; tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Increased tree height [upto 20m with higher LAI] can to lower mean and max temperature and $VPD$ in the understory.
+RH, in combination with temperature, determines vapor pressure deficit [$VPD$]--the driving force of water loss from a leaf-- tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory.
 Beyond a certain tree-height, the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
@@
 
-**Vertical biophysical drivers reviewed above influence heat and water stress patterns, thereby influencing leaf temperature $T_{leaf}$ distributions along the vertical gradient.**
+**Biophysical drivers reviewed above influence vertical heat and water stress patterns, and leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
 Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
@@
 At greater tree height, biological--tree-height--and biophysical--light, temperature, and evaporative demand (Fig.
-`r fig_NEON_vertical`)--also play a role in shaping leaf traits
+`r fig_NEON_vertical`)--also play a role in shaping leaf traits.
 
@@
 Disentangling tree-height related biological and biophysical drivers (Fig.
-`r fig_NEON_vertical`) of leaf traits remains an important area for research, which is sparse, is beyond the scope of the current review.
+`r fig_NEON_vertical`) of leaf traits remains an important area for research that is sparse, but is beyond the scope of the current review.
 Here, our primary interest is in how leaf traits and function vary across the vertical gradient from the canopy-top to the understory in forests.
@@
 2019].
-Under homogeneous light conditions, sun-leaves in canopy orientations are shaped based on canopy $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
+Under homogeneous light conditions, sun-leaf differences in canopy orientations are shaped based on canopy $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 $N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-Constraints to biochemical traits, particularly $VAZ$ can curb temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a].
-Additionally, isoprene-producing species are particularly advantageous in that they have higher heat tolerance in exposed-canopies [*REF*]
+Constraints to biochemical traits, particularly $VAZ$ can curb increasing temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a].
+Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed-canopies, making these species particularly advantageous [*REF*].
 
@@
 Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger than mature leaves, due to devoloping traits that can acclimate.
-Whereas, change in light for mature leaves can result in leaves dropping [Niinemetes, 2007].
+Whereas, change in light for mature leaves can result in leaf abscission [Niinemetes, 2007].
 *Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
@@
 
-Deciduousness as a leaf trait is an adaptation to water and heat stress.
+Deciduousness as a trait is an adaptation to water and heat stress.
 Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter.
-Among species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b] where, deciduous leaf-loss can influence biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
+Among species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b].
+Hence, deciduous leaf-loss can influence biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
 Spatio-temporal leaf-emergence in deciduous-temperate forests also shapes biophysical gradient where leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
@@
 
-While $gs$ mediates crucial leaf-atmosphere interactions across species, exposed-sun-leaves, though equipped with thermoregulatory effiency are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
+$Gs$ mediates crucial leaf-atmosphere interactions across species.With exposed-sun-leaves, though equipped with thermoregulatory effiency are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
 Where maximum transpiration rate and $gs$ peaks during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines during midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
@@
 2015 ).
-Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].
+Leaf heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].
 
@@
 Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), based on evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+14.5 in Campbell and Norman 1998; *revisit this!*), based on earlier evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
@@
 showed that in tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
-In both species photosynthetic heat tolerance increased seasonally from June to August.
+In both species photosynthetic heat tolerance also increased from June to August.
 However, more recently @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Panama tropical forest, but results weren't significant, with both measured near ground level.
 
-Leaf thermal tolerances across species might reflect local microenvironmental conditions than regional macroclimate [@feeley_thermal_2020; Slot et al.
-2020] where $T_{50}$ in *Acacia Papyrocarpa* (Benth.) trees in an Australian arid region showed higher composite climate stress and greater $T_{50}$ than other canopy positions, that correlated with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
+Further, leaf thermal tolerances vary less based on macroclimate, but might reflect microenvironment more [@feeley_thermal_2020; Slot et al.
+2020].
+Particularly, in an Australian arid region, *Acacia Papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, that correlated with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
 
-We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
+*We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
 I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot*
 
-*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.*
+Across species, $T_{50}$ is explained less by phylogeny [Slot et al.
+2021; Perez and Feeley, 2020a] and more by leaf traits.
+@slotLeafHeatTolerance2020 found that $T_{crit}$ was greater among tropical species' large leaves with slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA.
+@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance, experienced higher $T_{leaf_{max}}$, but showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$)
 
-Global $T_{50}$s across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
-2017) and $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
-
-Additionally, $T_{50}$ varies with species [@krauseHightemperatureToleranceTropical2010a]and leaf traits and does not vary with phylogeny [Slot et al.
-2021; Perez and Feeley, 2020a].
-Recently, @slotLeafHeatTolerance2020 found that $T_{crit}$ was greater among tropical species' large leaves, with slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA.
-@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance, experienced higher $T_{leaf_{max}}$, but showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).
+*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
+2017), where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
 
@@
 
-Across species in north-eastern oak deciduous forest, conifer temperate rainforest and in *Liquidambara styraciflua* tree, $Q_{10}$ is observed to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
-In contrast, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
+While in a few studies $Q_{10}$ across species has shown to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
 1*)[@bolstad_foliar_1999].

### 2021-05-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), based on earlier evidence that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+14.5 in Campbell and Norman 1998; *revisit this!*), where earlier evidence showed that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
@@
 
-While in a few studies $Q_{10}$ across species has shown to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
-Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
+While in a few studies $Q_{10}$ across species shows to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
 1*)[@bolstad_foliar_1999].
@@
 Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-However, soil water loss during drought effects larger trees more than small trees because, limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canonopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
+However, soil water loss during drought effects larger trees more than small trees because, limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
 
 Additionally, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019].
-However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves by more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis [and C uptake?] in reflushed leaves [Filewood and Thomas, 2013].
+However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis [and C uptake?] in reflushed leaves [Filewood and Thomas, 2013].
 
@@
 The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are highly interconneced, resulting in feedbacks to another.
-Therefore, climate-driven changes, such as rising temperature, can have implications to forests on several levels.
+Therefore, climate-driven changes, such as rising temperature, can have implications to forests on many levels.
 
@@
 2019].
-Therefore, frequent droughts causes canopy disturbance, where less severe drought can decrease tree canopy circumference [@aussenacInteractionsForestStands2000a], and more severe droughts shows increasing larger tree mortality in forests globally [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al.
+Frequent droughts can thus cause canopy disturbances.While less severe drought can decrease tree canopy circumference [@aussenacInteractionsForestStands2000a], more severe droughts show increasing larger tree mortality in forests globally [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
 Hydraulic limitation in forests can also alters species thermoregulation [@sastry_leaf_2018].
-Increasing temperatures shows increase in canopy $T_{leaf}$ above ambient temperature, and sometimes beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Particularly, increasing temperatures shows increase in canopy $T_{leaf}$ above ambient temperature, and sometimes beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
-Recent adavances in canopy studies further shows large loss of canopies in Europe caused by wind-induced disturbance and anthropogenic deforestation [Cornelius Senf, 2021; @senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020].
-Loss of big trees (e.g.@senfCanopyMortalityHas2018a ) would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity and composition which changes understory species composition and soil fertility [*@majasalmiImpactTreeCanopy2020*].Leading to altered biophysical gradient with local heating effects, [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Recent adavances in canopy studies further show large loss of canopies in Europe caused by wind-induced disturbance and anthropogenic deforestation [Cornelius Senf, 2021; @senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020].
+Loss of big trees (e.g.@senfCanopyMortalityHas2018a ) would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity and composition which changes understory species composition and soil fertility [*@majasalmiImpactTreeCanopy2020*].Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 Each of these changes have potential feedbacks to climate change.
@@
 Whereas, in other cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
-Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].Further information from fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
+Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
+At a leaf metabolism level, fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
 

### 2021-05-19 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -15,6 +15,9 @@ library(ggthemes)
 library(moderndive)
 library(ggpubr)
 library(cowplot)
+library(data.table)
+library(formattable)
+library(gridExtra)
 
 ####################################
 #create data frames using HARV NEON data for environmental parameters. Leaf characteristic dimmension in m is measured from Sun and shade Red oak leaves
@@ -27,11 +30,11 @@ library(cowplot)
 # Create a dataframe for overstory and understory, for scenarios: normal and drought
 # 1. Short wave radiation, normal
 #understory
-sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800), 
+sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
                   rh = 0.97, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
                   gs = 2.0)
 #overstory
-sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800), 
+sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
                   rh = 0.91, wind = 2.88, tair = 298.1, pressure = 99, leaf.size = 0.04, 
                   gs = 4.0)
 
@@ -87,12 +90,12 @@ sw_u$tleaf <- sw_u$tleaf +273.15
 sw_u$u_tleaf_tair<- sw_u$tleaf - sw_u$tair
 
 # 1. Shortwave radiation, drought
-dr_sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800), 
+dr_sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
                   rh = 0.37, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
-                  gs = 1.0)
-dr_sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800), 
+                  gs = 0.1)
[…]

### 2021-05-19 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -159,7 +159,7 @@ a1 <-ggplot(data = sw)+
   
   scale_color_identity(name = "canopy position",
                        breaks = c("red", "blue"),
-                       labels = c("overstory", "understory"),
+                       labels = c("overstory, dashed-lines represent drought", "understory, dashed-lines represent drought"),
                        guide = "legend")+
   scale_linetype_manual(name = "canopy position",
                         labels = c("overstory", "understory", "drought overstory", "drought understory"),
@@ -580,27 +580,15 @@ table<- data.frame(biophysical = c( "sw", "tair", "ws", "gs", "ls", "rh"),
                    normalno = c(152, 296.1, 0.24, 0.1, 0.010, 0.37))
 
 
-colnames(table) <- c("biophysical\nconditions",
+colnames(table) <- c("biophysical\nconstants",
                  "normal\noverstory", "normal\nunderstory", "drought\noverstory", "drought\nunderstory")
 table1<-ggtexttable(table, rows = NULL, theme = ttheme("mBlue"))
+table1
 
-
-gc_text <- paste("Constants from HARV NEON, solid lines represent normal, dotted lines represent drought", sep = " ")
-tex1 <- ggparagraph(text = gc_text, face = "italic", size = 8, color = "black")
 #without transformation
 
 figure1<-ggarrange(a1, b1, e1, d1, c1, table1, ncol=3, nrow =2, common.legend = TRUE, align = c("v"), legend = "top")
 figure1
-
-annotate_figure(
-  figure1,
-  bottom = text_grob("constants from harvard NEON data 
-                    \n overstory- sw: 871, tair: 298.1 [k], ws: 2.88 ms-1, gs: 4.0 µmol m-2 s-1 Pa-1, leaf size: 0.04m, rh: 0.91, 
-                     \n understory- sw: 102, tair: 296.1 [k], ws: 0.24 ms-1, gs: 2.0 µmol m-2 s-1 Pa-1, leaf size: 0.10m, rh: 0.97
-                     \n drought overstory (dotted line)- sw: 1000, tair: 298.1 [k], ws: 2.88 ms-1, gs: 1.0 µmol m-2 s-1 Pa-1, leaf size: 0.04m, rh: 0.31, 
-                     \n drought understory (dotted line)- sw: 152, tair: 296.1 [k], ws: 0.24 ms-1, gs: 0.1 µmol m-2 s-1 Pa-1, leaf size: 0.10m, rh: 0.37 
[…]

### 2021-05-19 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -91,10 +91,10 @@ sw_u$u_tleaf_tair<- sw_u$tleaf - sw_u$tair
 
 # 1. Shortwave radiation, drought
 dr_sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
-                  rh = 0.37, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
+                  rh = 0.49, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
                   gs = 0.1)
 dr_sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
-                  rh = 0.31, wind = 2.88, tair = 298.1, pressure = 99, leaf.size = 0.04, 
+                  rh = 0.46, wind = 2.88, tair = 298.1, pressure = 99, leaf.size = 0.04, 
                   gs = 1.0)
 dr_sw_l$par <- dr_sw_l$par*0.5#sw
 dr_sw_u$par <- dr_sw_u$par*0.5#sw
@@ -164,7 +164,7 @@ a1 <-ggplot(data = sw)+
   scale_linetype_manual(name = "canopy position",
                         labels = c("overstory", "understory", "drought overstory", "drought understory"),
                         values = c( 1, 1, 2, 2), guide = "legend")+
-  ylab("TLeaf - Tair [k]*")+xlab("short wave radiation (W/m^2)")
+  ylab("TLeaf - Tair [k]")+xlab("short wave radiation")
 
 a1 
 
@@ -220,13 +220,13 @@ a1
 
 #understory
 ws_l<- data.frame(par = 203.69, 
-                  rh = 0.97, wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88), 
+                  rh = 0.97, wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.07, 3.42), 
                   tair = 296.1, pressure = 99, 
                   leaf.size = 0.10, 
                   gs = 2.0)
 #overstory
 ws_u<- data.frame(par = 1741.65, 
[…]

### 2021-05-19 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -573,7 +573,7 @@ e1
 #constructing a table for the plot
 
 
-table<- data.frame(biophysical = c( "swr", "ws", "rh", "ls", "gs", "tair"),
+table<- data.frame(biophysical = c( "swr (W/m-2)", "ws (m/s)", "rh (%)", "ls (m)", "gs (mu.mol/m2/s/Pa)", "tair (K)"),
                    normalno = c(871, 2.88, 0.91, 0.04, 4.0, 298.1),
                    normal_u = c(102, 0.24, 0.97, 0.10, 2.0, 296.1),
                    drought_o = c(1307, 2.88, 0.46, 0.04, 1.0, 298.1),

### 2021-05-19 vertical-thermal-review.Rmd
Including Song et al. 2020

@@
 
-This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because stomatal conductance ($g_s$) facilitates heat dissipation, when water demand exceeds water transport, driven by irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
+Stomatal conductance ($g_s$) is a key factor mediating $T_{leaf}$ [@songTreeSurfaceTemperature2020] because leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by for example: irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
 
@@
 **In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves tend to heat and cool faster due to higher $ga$, maintaining lower $T_{leaf}$-$T_{air}$ than large leaves that take longer to heat and cool due to low $ga$ [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].
+
+Lastly, $gs$ and leaf size together are most important factors in deciding $T_{leaf}$ - $T_{air}$ and temperature sensitivity, because $gs$ determines leaf biological-biophysical interaction and leaf size determines leaf heating and cooling rate.
+Therefore, smaller leaves with greater $gs$ have are less sensitive to high temperature [@songTreeSurfaceTemperature2020].
 
@@
 Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Therefore, while general patterns of thermoregulation can be observed across species, some might be more suspectable to heat stress at a vertical gradient than others [@fauset_differences_2018].
+Species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 
-While in a few studies $Q_{10}$ across species shows to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+While in a few studies $Q_{10}$ shows to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
 Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
@@
 @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
+- Response of respiration to leaf temperature [partially represented  by $E_{0}$] in Quercus rubra hardly varies with canopy position, but canopy-position effect is respiration is much greater in mesic sites, and seasonal effect is greater in drier sites [@xuSeasonalVariationTemperature2006]
 

### 2021-05-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2014*].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
 Leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
@@
 
-Stomatal conductance ($g_s$) is a key factor mediating $T_{leaf}$ [@songTreeSurfaceTemperature2020] because leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by for example: irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
+Stomatal conductance ($g_s$) is a key factor mediating $T_{leaf}$ [@songTreeSurfaceTemperature2020].Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by for example: irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
 
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
- Inherent leaf $gs$ is an important factor for $T_{leaf}$ which is typically greater across species exposed sun-leaves due to stomata density than shade leaves.
+ Inherent leaf $gs$ is typically greater across species exposed sun-leaves due to stomata density than shade leaves.
 However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
@@
 
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves tend to heat and cool faster due to higher $ga$, maintaining lower $T_{leaf}$-$T_{air}$ than large leaves that take longer to heat and cool due to low $ga$ [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves tend to heat and cool faster due to higher $ga$, maintaining lower $T_{leaf}$-$T_{air}$ than large leaves that take longer to heat and cool due to low $ga$, therefore can experience greater $T_{leaf}$ - $T_{air}$ than small leaves [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
 
-Lastly, $gs$ and leaf size together are most important factors in deciding $T_{leaf}$ - $T_{air}$ and temperature sensitivity, because $gs$ determines leaf biological-biophysical interaction and leaf size determines leaf heating and cooling rate.
-Therefore, smaller leaves with greater $gs$ have are less sensitive to high temperature [@songTreeSurfaceTemperature2020].
+Lastly, $gs$ and leaf size together, though donot solely explain mean $T_{leaf}$, are most important factors in deciding $T_{leaf}$ - $T_{air}$ and temperature sensitivity, because $gs$ determines leaf biological-biophysical interaction and leaf size determines leaf heating and cooling rate.
+Therefore, smaller leaves with greater $gs$ are less sensitive to high temperature [@songTreeSurfaceTemperature2020].
 
@@
 Comparing forests, upper canopies of temperate deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
+
+-"Leaves with adequate water supply are partly buffered against air-temperature
+variations, through a passive biophysical mechanism"[@dongBiophysicalHomoeostasisLeaf2017].
 
@@
 Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007].
-Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-Species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
+Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+Therefore $T_{leaf}$ varies across speci-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al.
-2016].This can implication to canopies and gaps that reside at levels of greater $VPD$.
+2016; @corlettImpactsWarmingTropical2011] .This can implication to canopies and gaps that reside at levels of greater $VPD$.
 Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
@@
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. 
-Interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water.
+-Whereas, in other cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
+-Decelarating tree growth rates predominantly in smaller trees, but not restricted to, have also been observed in tropical forests, associated with local climate patterns of mean daily minimum temperature, increasing temperature and rainfall regime [Feeley et al.
+2007]
+-Interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water.
 Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].
@@
 Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-Whereas, in other cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
 Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
 At a leaf metabolism level, fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
+
+Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011]
 

### 2021-05-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
-Leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
+Leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
 
@@
 
-**Biophysical drivers reviewed above influence vertical heat and water stress patterns, and leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
+**Biophysical drivers reviewed above influence leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
 Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
@@
 
-Stomatal conductance ($g_s$) is a key factor mediating $T_{leaf}$ [@songTreeSurfaceTemperature2020].Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by for example: irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
+This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
 
@@
 However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
-On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress.
+On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress [@songTreeSurfaceTemperature2020].
 
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves tend to heat and cool faster due to higher $ga$, maintaining lower $T_{leaf}$-$T_{air}$ than large leaves that take longer to heat and cool due to low $ga$, therefore can experience greater $T_{leaf}$ - $T_{air}$ than small leaves [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to higher $ga$ than larger leaves, therefore maintain lower $T_{leaf}$-$T_{air}$ than large leaves that experience greater $T_{leaf}$-$T_{air}$.
+With adequate water, species with smaller leaves and greater $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @leuzingerTreeSpeciesDiversity2007]
 
-Lastly, $gs$ and leaf size together, though donot solely explain mean $T_{leaf}$, are most important factors in deciding $T_{leaf}$ - $T_{air}$ and temperature sensitivity, because $gs$ determines leaf biological-biophysical interaction and leaf size determines leaf heating and cooling rate.
-Therefore, smaller leaves with greater $gs$ are less sensitive to high temperature [@songTreeSurfaceTemperature2020].
+Lastly, canopy structure and architecture influence canopy temperature [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
 
@@
 Comparing forests, upper canopies of temperate deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
-
--"Leaves with adequate water supply are partly buffered against air-temperature
-variations, through a passive biophysical mechanism"[@dongBiophysicalHomoeostasisLeaf2017].
+At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007]
 
@@
 Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits.
+However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds based on inherent specie-specific traits.
 Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007].
@@
 
-Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011]
+-Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011]
 

### 2021-05-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
-The $T_{air}$ difference between understory and canopy temperature vary based on multiple factors such as radiation, local topography, vertical air mixing and cloud cover [*Defrenne et al.
+The $T_{air}$ difference between understory and canopy vary based on multiple factors such as radiation, local topography, vertical air mixing and cloud cover [*Defrenne et al.
 2021*].Dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies, where dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al.
@@
 
-This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.
+This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
 
@@
 
-Lastly, canopy structure and architecture influence canopy temperature [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
+Lastly, canopy structure and architecture also influence canopy temperature [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
 
@@
 
-While in a few studies $Q_{10}$ shows to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+While a few studies show $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
 Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
@@
 @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
-- Response of respiration to leaf temperature [partially represented  by $E_{0}$] in Quercus rubra hardly varies with canopy position, but canopy-position effect is respiration is much greater in mesic sites, and seasonal effect is greater in drier sites [@xuSeasonalVariationTemperature2006]
+- Response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] in Quercus rubra hardly varies with canopy position, but canopy-position effect in respiration is much greater in mesic sites, and seasonal effect is greater in drier sites [@xuSeasonalVariationTemperature2006]
 
@@
 
-Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al.
-2016; @corlettImpactsWarmingTropical2011] .This can implication to canopies and gaps that reside at levels of greater $VPD$.
-Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011] .Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
 However, soil water loss during drought effects larger trees more than small trees because, limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
@@
 Additionally, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019].
-However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis [and C uptake?] in reflushed leaves [Filewood and Thomas, 2013].
+Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
+However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
 
@@
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
--Whereas, in other cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
--Decelarating tree growth rates predominantly in smaller trees, but not restricted to, have also been observed in tropical forests, associated with local climate patterns of mean daily minimum temperature, increasing temperature and rainfall regime [Feeley et al.
-2007]
--Interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water.
+Similarly, in tropical forests, decelarating tree growth rates predominantly in smaller trees, but not restricted to, have also been observed due to local climate patterns of mean daily minimum temperature, increasing temperature and rainfall regime, but exact reasons remain unresolved  [Feeley et al.
+2007].
+Further, interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water.
 Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].
+In other, cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
 
@@
 Each of these changes have potential feedbacks to climate change.
-
+Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011].
 ### warming
@@
 While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to fauna in microrefugias [@scheffersMicrohabitatsReduceAnimal2013] .
-Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020].
-Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
+Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to fauna in microrefugias [@scheffersMicrohabitatsReduceAnimal2013].
+Heatwave and temperature  can alter understory tree grwoth rates, canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
 At a leaf metabolism level, fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
-
--Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011]
 

### 2021-05-21 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -7,6 +7,7 @@
 # Sept-August 2021
 ####################################
 #install packages if needed
+library(latex2exp)
 library(magrittr)
 library(tealeaves)
 library(ggplot2)
@@ -161,10 +162,7 @@ a1 <-ggplot(data = sw)+
                        breaks = c("red", "blue"),
                        labels = c("overstory, dashed-lines represent drought", "understory, dashed-lines represent drought"),
                        guide = "legend")+
-  scale_linetype_manual(name = "canopy position",
-                        labels = c("overstory", "understory", "drought overstory", "drought understory"),
-                        values = c( 1, 1, 2, 2), guide = "legend")+
-  ylab("TLeaf - Tair [k]")+xlab("short wave radiation")
+  ylab(TeX("TLeaf - Tair ($K$)"))+xlab(TeX("short wave radiation W/m$^{2}$"))
 
 a1 
 
@@ -275,7 +273,7 @@ b1<-ggplot(ws)+
   geom_smooth(aes(x = wind, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = wind, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = wind, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
-  ylab("TLeaf - Tair [k]")+xlab("windspeed")
+  ylab(TeX("TLeaf - Tair ($K$)"))+xlab(TeX("windspeed ($m/s$)"))
 
 b1
 
@@ -308,13 +306,13 @@ b1
 
 #plot, transformed
 
-b<-ggplot(ws)+
-  geom_smooth(aes(x = wind, y = l_tla),  method = lm, se = FALSE, color = "blue")+
-  geom_smooth(aes(x = wind, y = u_tla),  method = lm, color = "red", se = FALSE)+
-  geom_smooth(aes(x = wind, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
-  geom_smooth(aes(x = wind, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
-  ylab("TLeaf - Tair*")+xlab("windspeed (ms-1)")+scale_y_continuous(breaks = c(-2, 0, 2))
-b
[…]

### 2021-05-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 (3) leaf temperature ($T_{leaf}$);
-(2) the leaf traits most strongly influence leaf temperature and metabolism;
+(2) the leaf traits that most strongly influence leaf temperature and metabolism;
 (4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
@@
 Additionally, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019].
-Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
+Increase in summer temperature in mesic temperate forests is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
 However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].

### 2021-05-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$ [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a].
-Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling [Drake et al.
-2018] but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]
+Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]
 
@@
 **In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to higher $ga$ than larger leaves, therefore maintain lower $T_{leaf}$-$T_{air}$ than large leaves that experience greater $T_{leaf}$-$T_{air}$.
-With adequate water, species with smaller leaves and greater $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @leuzingerTreeSpeciesDiversity2007]
+With adequate water, species with smaller leaves and greater $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007]
 

### 2021-05-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The $T_{air}$ difference between understory and canopy vary based on multiple factors such as radiation, local topography, vertical air mixing and cloud cover [*Defrenne et al.
-2021*].Dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies, where dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al.
+2021*].Dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies.
+Therefore, dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al.
 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
@@
 
-This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
+This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].During $gs$, evapotranspiration facilitates heat dissipation.
+But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
 
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$ [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a].
-Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006]
+Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]
+
+Leaf interiors are typically at maximum $RH$ [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$.
+Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018].
+
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to higher $ga$ than larger leaves, therefore maintain lower $T_{leaf}$-$T_{air}$ than large leaves that experience greater $T_{leaf}$-$T_{air}$.
+With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007]
 
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
- Inherent leaf $gs$ is typically greater across species exposed sun-leaves due to stomata density than shade leaves.
-However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
-On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress [@songTreeSurfaceTemperature2020].
-
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to higher $ga$ than larger leaves, therefore maintain lower $T_{leaf}$-$T_{air}$ than large leaves that experience greater $T_{leaf}$-$T_{air}$.
-With adequate water, species with smaller leaves and greater $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007]
+ Inherent leaf $gs$ is typically greater across species exposed sun-leaves due to stomata density than shade leaves [Table 1].
+However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
+On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
 
@@
 Comparing forests, upper canopies of temperate deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
-At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007]
+At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007].
 

### 2021-05-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf interiors are typically at maximum $RH$ [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$.
+Leaf interiors are typically $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$.
 Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018].
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
- Inherent leaf $gs$ is typically greater across species exposed sun-leaves due to stomata density than shade leaves [Table 1].
+ Inherent leaf $gs$ is generally greater across species exposed sun-leaves due to stomata density than shade leaves [Table 1].
 However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
@@
 2019].
-Under homogeneous light conditions, sun-leaf differences in canopy orientations are shaped based on canopy $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
+Under homogeneous light conditions, sun-leaf differences in canopy orientations are shaped by canopy $T_{air_{max}}$ and $VPD$ stress than solely by light [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger than mature leaves, due to devoloping traits that can acclimate.
-Whereas, change in light for mature leaves can result in leaf abscission [Niinemetes, 2007].
+Whereas, change in longterm-light for mature leaves can result in leaf abscission [Niinemetes, 2007].
 *Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
@@
 Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter.
-Among species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b].
+Among tropical species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b].
 Hence, deciduous leaf-loss can influence biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
-Spatio-temporal leaf-emergence in deciduous-temperate forests also shapes biophysical gradient where leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
+Spatio-temporal leaf-emergence in deciduous-temperate forests also shape biophysical gradient where leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
 2013; @seiwaChangesLeafPhenology1999].
@@
 
-$Gs$ mediates crucial leaf-atmosphere interactions across species.With exposed-sun-leaves, though equipped with thermoregulatory effiency are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
-Where maximum transpiration rate and $gs$ peaks during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines during midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
+Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
+Where evapotranspiration and $gs$ peaks during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines during midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
+While specie-specific differences in $T_{opt}$ of $gs$ can influence their photosynthetic heat tolerance; exactly how, remains understudied [@perezPhotosyntheticHeatTolerances2020].
 
@@
 Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], along the vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
+*We [-nidhi, for now] hypothesize that the phenotypic-plastic nature of leaf investment in area-to-mass traits across a vertical gradient stablizes photosynthetic $T_{opt}$.
+Where as, across species photosynthetic $T_{opt}$ might vary more*
 
-**Leaf thermal tolerance (Tcrit/ T50)**
+**Leaf photosynthetic thermal tolerance (Tcrit/ T50)**
 

### 2021-05-22 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density.
-This gradient is most pronounced under broad-leaf and mixed forests where light reaching forest floors, incident to canopy-tops range from ~1-2% to 4-5%, respectively [Fig.
-`r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006] compared to needle-leaf forest floors that receive ~17-34% [@baldocchiSeasonalVariationEnergy1997].
+This gradient is most pronounced under dense canopies, including broad-leaf and mixed forests, than open canopy forests such as needle-leaf forests [Fig.
+`r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 Sunflecks, offer another light source for shaded forest regions, caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
-The $T_{air}$ difference between understory and canopy vary based on multiple factors such as radiation, local topography, vertical air mixing and cloud cover [*Defrenne et al.
-2021*].Dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies.
+The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al.
+2021*].
+Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies.
 Therefore, dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al.
-2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
 (*This sentence is too complex-- hard to understand*)
@@
 Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a].
-RH, in combination with temperature, determines vapor pressure deficit [$VPD$]--the driving force of water loss from a leaf-- tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory.
-Beyond a certain tree-height, the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+$RH$, in combination with temperature, determines vapor pressure deficit [$VPD$]--the driving force of water loss from a leaf-- tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
 
@@
 
-This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].During $gs$, evapotranspiration facilitates heat dissipation.
+This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
+During $gs$, evapotranspiration facilitates heat dissipation.
 But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
@@
 `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a].
-Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]
+Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]
 
-Leaf interiors are typically $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$.
-Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018].
+Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$.
+Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
 
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to higher $ga$ than larger leaves, therefore maintain lower $T_{leaf}$-$T_{air}$ than large leaves that experience greater $T_{leaf}$-$T_{air}$.
-With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007]
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to smaller $ga$ than larger leaves that take longer due to large $ga$.
+Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves that can exprience greater $T_{leaf}$-$T_{air}$.
+With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures than large leaf species [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
 
@@
 The distribution of biochemical traits enable photosynthesis and metabolic processes, and protect against heat and light damage in sun-leaves.
-$N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+$N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018].
+Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 Constraints to biochemical traits, particularly $VAZ$ can curb increasing temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a].
@@
 2019 ; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], along the vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
-*We [-nidhi, for now] hypothesize that the phenotypic-plastic nature of leaf investment in area-to-mass traits across a vertical gradient stablizes photosynthetic $T_{opt}$.
-Where as, across species photosynthetic $T_{opt}$ might vary more*
+Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], along the vertical gradient of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+
+Interestingly, within canopy Japanese oak sun and shade leaves under respective peak light conditions were found to maintain similar rate of excess energy production.
+Excessive light energy not utilized can cause photoinhibition.Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain high photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018].
+Perhaps, differences in these thermal stablization strategies realizes similar $T_{opt}$ for photosynthesis in sun and shade leaves.
 

### 2021-05-22 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$, making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
-Comparing forests, upper canopies of temperate deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
-At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007].
+Comparing forests, upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
+At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
 

### 2021-05-23 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
 
@@
 
-Lastly, canopy structure and architecture also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
+Lastly, canopy structure and architecture adaptation can also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
 
@@
 Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
-Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
+Under other conditions $T_{leaf}$ tracks $T_{air}$, due to high evapotranspiration and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
 
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
-Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies from $gs$ limitation, than buffered lower-canopies [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016].
-Greater light input in exposed upper-canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$, making upper-canopies highly susceptible to heat stress [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a].
-Comparing forests, upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a].
+Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+Comparing forests, upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies[@mauTemperateTropicalForest2018a; @leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018] .
 At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
+Spatial distribution of canopy temperature also shows that tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
 
@@
 
-Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011] .Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-However, soil water loss during drought effects larger trees more than small trees because, limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
+Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+Accordingly, heat, associated with increase in temperature and VPD even under adequete water can cause greater tree hydraulic stress and altered and soil-water dynamics than milder drought, due to increased $T_{leaf}$ and $gs$ limitation [@ruehrWaterAvailabilityDominant2016].
+However, soil water loss during drought effects larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
+Further, heat with drought exacerbates tree stress [@ruehrWaterAvailabilityDominant2016].
 
-Additionally, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense, particularly in temperate-deciduous forests [Zohnner and Renner, 2019].
+Phenologically, understory growth can be more sensitive to annual warming than canopy in temperate-deciduous forests, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
 Increase in summer temperature in mesic temperate forests is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
 However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
+Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+.
 
@@
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
-Similarly, in tropical forests, decelarating tree growth rates predominantly in smaller trees, but not restricted to, have also been observed due to local climate patterns of mean daily minimum temperature, increasing temperature and rainfall regime, but exact reasons remain unresolved  [Feeley et al.
-2007].
+Similarly, in tropical forests, decelarating tree growth rates predominantly in smaller trees, but not restricted to, are associated with higher nighttime temperatures and change in rain regimes [*Clark 2010; 2013; Feeley et al.
+2007*].
 Further, interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water.
 Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].
-In other, cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
+In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
 
@@
 Particularly, increasing temperatures shows increase in canopy $T_{leaf}$ above ambient temperature, and sometimes beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+
+Increase in canopy temperature is also associated with GPP reduction in the tropics.
+Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011].
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
@@
 Each of these changes have potential feedbacks to climate change.
-Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011].
+
 ### warming

### 2021-05-23 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Where evapotranspiration and $gs$ peaks during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines during midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
-While specie-specific differences in $T_{opt}$ of $gs$ can influence their photosynthetic heat tolerance; exactly how, remains understudied [@perezPhotosyntheticHeatTolerances2020].
+While specie-specific differences in $T_{opt}$ of $gs$ can influence their photosynthetic heat tolerance indirectly; exactly how, remains understudied [@perezPhotosyntheticHeatTolerances2020].
 
@@
 
+Temperature can affect photosynthesis via direct and indirect pathways.
+Direct, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008].
 $T_{leaf}$ beyond photosynthetic $T_{opt}$, declines photosynthesis due it's peaked response to temperature.

### 2021-05-23 NEON_height_profiles/scripts/verticalProfileFigure.R
Update verticalProfileFigure.R

@@ -17,12 +17,15 @@ library(ggpubr)
 #1. Download and save neon data ####
 # this is filtered to only the columns needed for plots
 # (mean measurements per each 30 min observation) in order to not have massive files.
-# source("scripts/get_neon_data.R")
+source("scripts/get_neon_data.R")
 # sites <- c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF")
-# pbapply::pblapply(sites, function(st){
-#   full_data <- get_NEON(x=st)
-#   save(full_data, file=paste0(st, "test.Rdata"))
-# })
+sites <- c("DELA", "GUAN", "JERC", "LENO", "CLBJ", "SJER", "YELL",
+              "GRSM", "ORNL", "TREE", "UKFS", "SOAP", "TALL", "TEAK",
+              "BART", "BONA", "DEJU")
+pbapply::pblapply(sites, function(st){
+  full_data <- get_NEON(x=st)
+  save(full_data, file=paste0("data/", st, "test.Rdata"))
+})
 
 #1a. Get NEON coordinates ####
 # library(data.table)
@@ -36,21 +39,6 @@ library(ggpubr)
 data <- fread("forested_NEON_sites.csv")
 data <- data[include==1, ]
 sites <- data[,site]
-
-#only focus on 6 core sites
-sites <- c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF")
-clrs <- c("#3399FF", "gold", "grey", "red", "#66CCFF", "#003399")
-
-#bring in normalized height (from Marielle)
-# meta <- read_excel("data/site_data/TIS site metadata_20190403_forNOAA.xlsx", sheet=1)
-# meta <- data.table(meta)
-# meta <- meta[SITE %in% sites, 
-#              ][,`:=` (DistZaxsCnpy = as.numeric(DistZaxsCnpy),
-#                     maxCanHeightMarielle = c(32, 43, 39, 21, 53))]
-normHeightData <- 
-  data.table(site=c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF"),
-                         normH=c(32, 21, 23, 43, 39, 53))
-# HARV=32, OSBS: 21, PUUM: 23, SCBI: 43, SERC: 39, WREF: 53
 
 dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT", "PARPAR"),
                  id = c("DP1.00001.001", "DP1.00098.001",
[…]

### 2021-05-24 NEON_height_profiles/scripts/get_neon_data.R
Updated neon rdata files

@@ -45,9 +45,9 @@ date <- data.table("year" = c(rep(2015, 2), rep(2016, 2), rep(2017, 2),
 # alldt <- list()
 # plotlist <- list()
 
-sites <- c("BART","BONA","CLBJ","DEJU","DELA","GRSM","GUAN","HARV","JERC",
-           "LENO","MLBS","ORNL","OSBS","SCBI","SERC","SJER","SOAP","STEI",
-           "TALL","TEAK","TREE","UKFS","UNDE","WREF","YELL")
+# sites <- c("BART","BONA","CLBJ","DEJU","DELA","GRSM","GUAN","HARV","JERC",
+#            "LENO","MLBS","ORNL","OSBS","SCBI","SERC","SJER","SOAP","STEI",
+#            "TALL","TEAK","TREE","UKFS","UNDE","WREF","YELL")
 
 get_NEON <- function(x="SCBI"){
   # full <- list()
@@ -61,7 +61,7 @@ get_NEON <- function(x="SCBI"){
       loadByProduct(dpID=dp[,id][i], 
                     site=x,
                     package="basic", 
-                    check.size = FALSE, avg=30,
+                    check.size = FALSE, timeIndex=30,
                     nCores = 2)
     # startdate=paste0(date[,year][j], "-0",
     #                  date[,month][j]),

### 2021-05-24 NEON_height_profiles/scripts/verticalProfileFigure.R

@@ -22,9 +22,22 @@ source("scripts/get_neon_data.R")
 sites <- c("DELA", "GUAN", "JERC", "LENO", "CLBJ", "SJER", "YELL",
               "GRSM", "ORNL", "TREE", "UKFS", "SOAP", "TALL", "TEAK",
               "BART", "BONA", "DEJU")
+
+library(parallel)
+cl <- makeCluster(detectCores()-1)
+clusterEvalQ(cl, library(data.table))
+clusterEvalQ(cl, library(neonUtilities))
+clusterExport(cl, list("cl", "sites", "get_NEON", "dp", "date"))
+
+parLapply(cl, X=1:length(sites), function(X){
+  full_data <- get_NEON(x=sites[X])
+  save(full_data, file=paste0("data/", sites[X], "test.Rdata"))
+})
+stopCluster(cl)
+
 pbapply::pblapply(sites, function(st){
-  full_data <- get_NEON(x=st)
-  save(full_data, file=paste0("data/", st, "test.Rdata"))
+  
+  
 })
 
 #1a. Get NEON coordinates ####

### 2021-05-24 vertical-thermal-review.Rmd
editing biophysical section, knit in latest leaf T fig

@@
 
-**The biophysical environment across the vertical gradient from the forest floor to the top of the canopy is in large part determined by the structure of the forest itself.** 
-Here, we define the biophysical environment to include the physical structure of the vegetation and associated physical conditions.
+**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs.
+`r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest itself.**
 In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S.
-National Ecological Observatory Network [NEON; Appendix `r appendix_NEON_methods`, Schimel et al.
+National Ecological Observatory Network [NEON; `r fig_NEON_vertical`; Appendix `r appendix_NEON_methods`, Schimel et al.
 2007].
-
-### Foliage density
-**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions.
-It is critical in influencing -- and in turn is influenced by -- the vertical biophysical gradient (Fig.
-`r fig_NEON_vertical`).**
-The distribution of this foliage is structured in correspondence with the distribtuion of tree sizes within a forest.  
-Leaf area density.... *description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references)*
-Vertical profiles in leaf area density vary across forests, being...
-*summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
-They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
-In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
-In this review, we focus primarily on growing season conditions, including leaf phenology at the edges of the growing season, but excluding periods when no leaves are present in deciduous forests.
+While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment cn be more closely linked to the distance from the outer canopy than to height (Parker 1995).
 
@@
 
+### Foliage density
+**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions.
+It is critical in influencing -- and in turn is influenced by -- the vertical biophysical gradient (Fig.
+`r fig_NEON_vertical`).**
+The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest.
+Leaf area density.... *[description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references).]*
+Vertical profiles in leaf area density vary across forests, being...
+*summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
+They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
+In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
+In this review, we focus on growing season conditions unless otherwise noted.
+
 ### Light environments
 
-**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
-`r fig_NEON_vertical`; @koike_leaf_2001].**
-Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density.
-This gradient is most pronounced under dense canopies, including broad-leaf and mixed forests, than open canopy forests such as needle-leaf forests [Fig.
+**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the shape of profile modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
+`r fig_NEON_vertical`, @koike_leaf_2001].**
+This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests, than open canopy forests such as needle-leaf forests [Fig.
 `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
-Sunflecks, offer another light source for shaded forest regions, caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).
+Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases is solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
 
 ### Wind speeds 
-**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
-Across all **#** forested NEON sites, and in closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019] wind speeds increase with height.
-With daily maximum wind speeds averaging up to **#** m s^-1^ higher at the top of the NEON vertical profile than at the bottom
+**Wind speeds are also higher at the top of the canopy, owing to... [@juckerCanopyStructureTopography2018].**
+Across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites, wind speeds increase with height.
+Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs.
+`r fig_NEON_vertical`, `r SIfigs_NEON_all`).
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).
 
@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
-The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al.
-2021*].
-Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies.
-Therefore, dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al.
-2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-(*This sentence is too complex-- hard to understand*)
-
+The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
+Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al.
+2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
 Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
-2017; @mcgregor_tree_nodate].
-Whereas in colder conditions, dense-canopy understories experience higher minimum temperatures [$T_{min}$] relative to open-canopy understories and nearby clearings where greater radiative heat loss results in lower $T_{min}$ [@zellweger_seasonal_2019; *Defrenne et al.
-2021*; @nakamuraForestsTheirCanopies2017].
+2017; **double check these refs--at least one didn't belong**].
 
 ### Humidity and evaporative demand
-**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
-Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a].
-$RH$, in combination with temperature, determines vapor pressure deficit [$VPD$]--the driving force of water loss from a leaf-- tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+**Relative humidity, $RH$, is generally highest in the understory and decreases with height across the vertical profile [Fig. `r fig_NEON_vertical`; e.g., @mcgregor_tree_nodate].**
+Along with temperature, dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a].
+$RH$, in combination with temperature, determines vapor pressure deficit [$VPD$] -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
@@
 
-With altitude, the strength of buffering decreases, moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for eg.
-in riparian forests.
+**The buffering patterns described above vary geographically and across forests types.**
+The strength of buffering decreases with altitude (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) and moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests.
 [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]

### 2021-05-24 vertical-thermal-review.Rmd
worked on biophysical section, removed sub-headings

@@
 
-### Foliage density
 **Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions.
@@
 
-### Light environments
-
 **Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the shape of profile modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
@@
 
-### Wind speeds 
 **Wind speeds are also higher at the top of the canopy, owing to... [@juckerCanopyStructureTopography2018].**
-Across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites, wind speeds increase with height.
+Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites.
 Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs.
@@
 
-### Air temperature 
 **Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs.
@@
 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
-2017; **double check these refs--at least one didn't belong**].
+(*Noteably, this pattern can be reversed in open forests, where...[explain])*
 
-### Humidity and evaporative demand
 **Relative humidity, $RH$, is generally highest in the understory and decreases with height across the vertical profile [Fig. `r fig_NEON_vertical`; e.g., @mcgregor_tree_nodate].**
-Along with temperature, dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a].
-$RH$, in combination with temperature, determines vapor pressure deficit [$VPD$] -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
-
-### Carbon dioxide
+Dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a].
+In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Increased tree height (up to 20 m with higher LAI) can also lower mean and maximum $T_{air}$ and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+*( This last sentence needs work.)*
 
@@
 
-### Geographic and climatic factors
-
-**The buffering patterns described above vary geographically and across forests types.**
+**The buffering patterns described above vary geographically, seasonally, and across forests types [REFS; *Defrenne et al. 2021*].**
 The strength of buffering decreases with altitude (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) and moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests.
 [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
+Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
+2017; **double check these refs--at least one didn't belong**].
+*(Presumably, the above is because leaf area is greater in set seasons)*
 

### 2021-05-24 vertical-thermal-review.Rmd
bit of rearranging

@@
 
+**The biophysical drivers reviewed above influence leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
+Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy). 
+Often, however, $T_{leaf}$ remains similar to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+
 ### Biophysical drivers
 
-**Biophysical drivers reviewed above influence leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
-Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-Leaves can be cooler than air at high temperature and warmer than air at low temperature (limited homeothermy) or can be drastically warmer than $T_{air}$ (megathermy).
-In some cases, $T_{leaf}$ can be equalled to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
-
-$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles  [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig.
-`r fig_leaf_T`) where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
-2; @michaletzPlantThermoregulationEnergetics2015].
+$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
+`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016 @michaletzPlantThermoregulationEnergetics2015].
+This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
+During $gs$, evapotranspiration facilitates heat dissipation.
+But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
 
@@
 [ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/leaf_energy_balance.png)
-
-This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
-During $gs$, evapotranspiration facilitates heat dissipation.
-But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
-
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
 
@@
 Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$.
-Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
+Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissipation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
 

### 2021-05-24 vertical-thermal-review.Rmd
work on T_leaf

@@
 
-**The biophysical drivers reviewed above influence leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
-Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, a that is optimal for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
 Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
-Often, however, $T_{leaf}$ remains similar to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+Often, however, $T_{leaf}$ tracks $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020]. 
+Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
+`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].
 
@@
 
-$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
-`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016 @michaletzPlantThermoregulationEnergetics2015].
-This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].
-During $gs$, evapotranspiration facilitates heat dissipation.
-But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
+*(In general, we'll want to adjust this section to parallel the figure and tie in with the section above.)*
 
@@
 
-High wind speed (Fig.
-`r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a].
+High wind speed (Fig. `r fig_NEON_vertical`) and large $g_a$ decrease $T_{leaf}$ - $T_{air}$ [Fig. `r fig_leaf_T`b, @daudetWindSpeedLeaf1999a]. 
 Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]
@@
 
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to smaller $ga$ than larger leaves that take longer due to large $ga$.
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to smaller $g_a$ than larger leaves that take longer due to large $ga$.
 Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves that can exprience greater $T_{leaf}$-$T_{air}$.
 With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures than large leaf species [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+
+This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
+When stomatal conductance ($g_s$) is high, evapotranspiration facilitates heat dissipation. 
+However, when high irradiance, $T_{air}$, $VPD$ or drought cause leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
 

### 2021-05-24 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -162,7 +162,8 @@ a1 <-ggplot(data = sw)+
                        breaks = c("red", "blue"),
                        labels = c("overstory, dashed-lines represent drought", "understory, dashed-lines represent drought"),
                        guide = "legend")+
-  ylab(TeX("TLeaf - Tair ($K$)"))+xlab(TeX("short wave radiation W/m$^{2}$"))
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ ($K$)"))+xlab(TeX("short wave radiation (swr, W/m$^{2}$)"))+
+  theme_few()+theme(text = element_text(size = 14))
 
 a1 
 
@@ -244,11 +245,11 @@ ws_u$u_tla<- ws_u$tleaf - ws_u$tair
 #2. Windspeed, drought
 #understory
 dr_wsl<- data.frame(par = 305.54, rh = 0.49, 
-                         wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.07, 3.42), tair = 296.1, pressure = 99, leaf.size = 0.10, 
+                         wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.88), tair = 296.1, pressure = 99, leaf.size = 0.10, 
                          gs = 0.1)
 #overstory
-dr_wsu<- data.frame(par = 2614.47, 
-                         rh = 0.46, wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.07, 3.42), tair = 298.1, pressure = 99, leaf.size = 0.04, 
+dr_wsu<- data.frame(par = 2302.15, 
+                         rh = 0.46, wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.88), tair = 298.1, pressure = 99, leaf.size = 0.04, 
                          gs = 1.0)
 
 dr_wsl$par <- dr_wsl$par*0.5#par to sw
@@ -273,7 +274,8 @@ b1<-ggplot(ws)+
   geom_smooth(aes(x = wind, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = wind, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
[…]

### 2021-05-24 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -219,13 +219,13 @@ a1
 
 #understory
 ws_l<- data.frame(par = 203.69, 
-                  rh = 0.97, wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.07, 3.42), 
+                  rh = 0.97, wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.88), 
                   tair = 296.1, pressure = 99, 
                   leaf.size = 0.10, 
                   gs = 2.0)
 #overstory
 ws_u<- data.frame(par = 1741.65, 
-                  rh = 0.91, wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.07, 3.42), 
+                  rh = 0.91, wind = c(0, 0.24, 0.42, 0.88, 1.07, 1.42, 1.88, 2.07, 2.42, 2.88, 3.88), 
                   tair = 298.1, pressure = 99, 
                   leaf.size = 0.04, 
                   gs = 4.0)
@@ -582,12 +582,12 @@ table<- data.frame(biophysical = c( "swr", "ws", "rh", "ls", "gs", "tair"),
                    drought_o = c("1151", 2.88, 0.46, 0.04, "1.0", "298"),
                    normalno = c("153", 0.24, 0.49, 0.010, "0.1", "296"))
 
-colnames(table) <- c(" ",
+colnames(table) <- c("variable",
                  "ON", "UN", "OD", "UD")
 
 main.title <- "Biophysical Constants"
 
-table1<-ggtexttable(table1, rows = NULL, theme = ttheme("mBlue"))
+table1<-ggtexttable(table, rows = NULL, theme = ttheme("mBlue"))
 
 table1<-table1 %>%
   tab_add_title(text = main.title, face = "bold")

### 2021-05-27 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -172,58 +172,18 @@ a1<-ggplot(data = sw)+
   scale_linetype_manual(values = c(1,1,2,2), name = "canopy position")+
   theme( 
     legend.key=element_blank(),
-    legend.position = c(0, 1),
+    legend.position = c(0, 1.05),
     legend.justification = c("left", "top"), 
-    legend.text = element_text(size=11),
-    legend.key.width = unit(0.50, 'cm'),
+    legend.text = element_text(size=10),
+    legend.key.height = unit(0.4, 'cm'),
+    legend.key.width = unit(1, 'cm'),
     legend.key.size = unit(0.4, 'cm'),
-    legend.title = element_blank())
+    legend.title = element_blank(),
+    legend.background = element_blank(),
+    #legend.box.background = element_rect(color = "darkgrey"))
+    legend.box.background = element_blank())
 a1
-#transforming data for standardization of values, for slope at the midpoint of variable x
-#lower<-lm(sw$l_tla ~ sw$par)
-#coef(lower)
-#y<- 0.01280917*450 + 0.04960374
-#y
-#y= 7.542647
-#sw$l_tla<-sw$l_tla - 5.81373
 
-#upper<-lm(sw$u_tla ~ sw$par)
-#coef(upper)
-#y<- 0.005356825*450 + -0.451573590
-#y
-#y = 2.961966
-#sw$u_tla<-sw$u_tla - 1.958998
-
-#drought_lower<-lm(sw$dr_ltla ~ sw$par)
-#coef(drought_lower)
-#y<- 0.01504194*450 + -2.06317804
-#y
-#y = 7.214889
-#sw$dr_ltla<- sw$dr_ltla - 4.705695
-
-#drought_upper<-lm(sw$dr_utla ~ sw$par)
-#coef(drought_upper)
-#y<- 0.006067448*450 + -2.669403031
-#y
-#sw$dr_utla<- sw$dr_utla - 0.06094857
-
-# plot with tranformed data
-#a<-ggplot(data = sw)+
-#  geom_smooth(aes(x = par, y = l_tla, colour = "blue"), method = lm, se = FALSE, linetype = 1)+
-#  geom_smooth(aes(x = par, y = u_tla, colour = "red"), method = lm, se = FALSE, linetype = 1)+
[…]

### 2021-05-27 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -307,7 +307,7 @@ c1<-ggplot(gs)+
   geom_smooth(aes(x = gs, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = gs, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = gs, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ ($°C$)"))+xlab(TeX("stomatal conductance ($g_{s}$, $\\mu mol/m^2/s/Pa$)"))+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("stomatal conductance ($g_{s}$, $\\mu mol/m^2/s/Pa$)"))+
   theme_few()+theme(text = element_text(size = 14))+ylim(-5, 14)
 c1
 
@@ -367,7 +367,7 @@ d1<-ggplot(ls)+
   geom_smooth(aes(x = ls, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = ls, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = ls, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
-  ylab(TeX("T_{Leaf}$ - $T_{air}$ ($°C$)"))+xlab("leaf size (ls, m)")+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab("leaf size (ls, m)")+
   theme_few()+theme(text = element_text(size = 14))+ylim(-5, 14) 
 d1
 
@@ -432,7 +432,7 @@ e1<-ggplot(rh)+
   geom_smooth(aes(x = rh, y = u_tla),  method = lm, color = "red", se = FALSE)+
   geom_smooth(aes(x = rh, y = dr_ltla),  method = lm, color = "blue", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = rh, y = dr_utla),  method = lm, color = "red", se = FALSE, linetype = "dashed")+
-  ylab(TeX("$T_{Leaf}$-$T_{air}$ (°C)"))+ xlab("relative humidity (rh, %)")+
[…]

### 2021-05-28 vertical-thermal-review_SI.Rmd
Update vertical-thermal-review_SI.Rmd

--- 2021-01-26 vertical-thermal-review_SI.Rmd
@@
 METHODS ON TEALEAVES GO HERE.
+
+Energy balance for a normal overstory sun leaf and understory shade leaf were parameterized with biophysical variables from Harvard NEON data documented here: (https://github.com/EcoClimLab/vertical-thermal-review/blob/master/NEON_height_profiles/HARV_neon_stats.csv).
+Leaf trait measurements for effective leaf width was measured on ~Quercus rubra~ sun and shade leaf, stomatal conductance measurements were refered from Tleaves typical sun and shade measurements [@muir_tealeaves_2019] and as well @cavender-baresChangesDroughtResponse2000.
+
+For drought scenario, biophysical variables were parameterized similar to normal scenario using Harvard NEON data.
+Overstory drought PAR values reflect maximum observed Harvard NEON PAR and understory drought reflect 50% increased PAR of understory normal value.
+Leaf trait measurements for effective leaf width is the same as normal scenario, stomatal conductance was referred from @cavender-baresChangesDroughtResponse2000.
+In each visual, all variables are constant (Biophysical Constants table) except for the independent variable that represents minimum - maximum range.
 

### 2021-05-28 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Updated T_leaf plot

@@ -19,7 +19,7 @@ library(cowplot)
 library(data.table)
 library(formattable)
 library(gridExtra)
-
+library(RColorBrewer)
 ####################################
 #create data frames using HARV NEON data for environmental parameters. Leaf characteristic dimmension in m is measured from Sun and shade Red oak leaves
 #all variables are constant except for the independent variable that represents minimum- maximum range
@@ -154,10 +154,10 @@ dr_sw_u$dr_u_tleaf_tair<- dr_sw_u$tleaf - dr_sw_u$tair
 
 sw<- data.frame(l_tla = sw_l$l_tleaf_tair, u_tla = sw_u$u_tleaf_tair, dr_ltla = dr_sw_l$dr_l_tleaf_tair, dr_utla = dr_sw_u$dr_u_tleaf_tair, par = sw_l$par)
 
-l_tla<-data.frame(par = sw$par, tleaf_tair = sw$l_tla, canopy_position = "understory normal (UN)")
-u_tla<-data.frame(par = sw$par, tleaf_tair = sw$u_tla, canopy_position = "overstory normal (ON)")
-dr_ltla<-data.frame(par = sw$par, tleaf_tair = sw$dr_ltla, canopy_position = "understory drought (UD)")
-dr_utla<-data.frame(par = sw$par, tleaf_tair = sw$dr_utla, canopy_position = "overstory drought (OD)") 
+l_tla<-data.frame(par = sw$par, tleaf_tair = sw$l_tla, canopy_position = "understory normal")
+u_tla<-data.frame(par = sw$par, tleaf_tair = sw$u_tla, canopy_position = "overstory normal")
+dr_ltla<-data.frame(par = sw$par, tleaf_tair = sw$dr_ltla, canopy_position = "understory drought")
+dr_utla<-data.frame(par = sw$par, tleaf_tair = sw$dr_utla, canopy_position = "overstory drought") 
 
 sw<- data.frame(rbind(u_tla, l_tla, dr_utla, dr_ltla))
 
@@ -168,7 +168,7 @@ a1<-ggplot(data = sw)+
               method = lm, se = FALSE)+
   ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("short wave radiation (swr, W/m$^{2}$)"))+
   theme_few()+theme(text = element_text(size = 14))+ylim(-5, 14)+
[…]

### 2021-05-28 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -2,9 +2,9 @@
 # Purpose: Create leaf energy balance figure using HARV neon data parameters for a typical overstory
 # and understory, normal and drought scenarios
 # Tleaves R package created by Chris Muir, 2019
-# Developed by: Nidhi Vinod, contact Nidhi Vinod (nidrup@gmail.com)
-# R version 3.5.3 --First created in October 2020
-# Sept-August 2021
+# Developed by: Nidhi Vinod, contact Anderson-Teixeira (teixeirak@si.edu)
+# R version 3.5.3 --First created in October 2020, updated on 5/28/2021
+# Oct-August 2021
 ####################################
 #install packages if needed
 library(latex2exp)
@@ -13,23 +13,21 @@ library(tealeaves)
 library(ggplot2)
 library(dplyr)
 library(ggthemes)
-library(moderndive)
 library(ggpubr)
 library(cowplot)
-library(data.table)
-library(formattable)
 library(gridExtra)
 library(RColorBrewer)
 ####################################
-#create data frames using HARV NEON data for environmental parameters. Leaf characteristic dimmension in m is measured from Sun and shade Red oak leaves
-#all variables are constant except for the independent variable that represents minimum- maximum range
-#normal represents HARV neon environmental parameters as observed
-#Stomtal conductane(gs) values for typical condition obtained from Tleaves supplementary manual for range of minimum and maximum values
-#drought condition is representative of gs values from Cavendar-Bares and Bazzaz, 2000 and typically low (50% decrease from normal) relative humidity, mean max par for uppercanopy, 50% par for lower
-#and the rest of the variables are normal conditions
+#dataframe for biophysical parameters from NEON HARV data
+#normal scenario represents HARV NEON environmental parameters as observed. 
[…]

### 2021-05-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-(*Noteably, this pattern can be reversed in open forests, where...[explain])*
+Noteably, this pattern can be reversed in open canopy forests or sparse stands, particularly in xeric sites, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due turbulent air mixing and upwelling of soil thermal radition that is intercepted by the lower-canopy layer, analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
 
@@
 In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Increased tree height (up to 20 m with higher LAI) can also lower mean and maximum $T_{air}$ and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].
+Additionally, increase in canopy height and LAI, upto a certain height, can also lower maximum $T_{air}$ and VPD in the understory [@juckerCanopyStructureTopography2018].
 *( This last sentence needs work.)*
@@
 **The buffering patterns described above vary geographically, seasonally, and across forests types [REFS; *Defrenne et al. 2021*].**
-The strength of buffering decreases with altitude (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) and moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests.
+The strength of buffering decreases when (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) moving to open-forests in high elevation, and increases with canopy density and height in high latitudes [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018].
+Buffering capacity is also greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests.
 [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
-Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al.
-2017; **double check these refs--at least one didn't belong**].
+Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018**double check these refs--at least one didn't belong**].
 *(Presumably, the above is because leaf area is greater in set seasons)*
@@
 All, collectively regulate $T_{leaf}$-photosynthetic balance.
-Whereas, shaded leaves invest more in light capture than in heat dissipation, with lower convective cooling [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
+Whereas, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
 2019].

### 2021-06-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2007].
-While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment cn be more closely linked to the distance from the outer canopy than to height (Parker 1995).
+While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height (Parker 1995).
 
@@
 **Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions.
-It is critical in influencing -- and in turn is influenced by -- the vertical biophysical gradient (Fig.
+It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig.
 `r fig_NEON_vertical`).**
-The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest.
-Leaf area density.... *[description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references).]*
-Vertical profiles in leaf area density vary across forests, being...
-*summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
-They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
-In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
+ The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density ($LAD$) ie.
+leaf area per unit volume, increases with tree-height eventually forming the canopy layer.
+
+LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests.
+Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory for eg.
+in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
+bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985].
+In open canopy forests, such as needle-leaf boreal forests, LAD is predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016].SOil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs.
+open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
+In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019).
+In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*).
 In this review, we focus on growing season conditions unless otherwise noted.
@@
 `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
-Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases is solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
+Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
 
@@
 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Noteably, this pattern can be reversed in open canopy forests or sparse stands, particularly in xeric sites, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due turbulent air mixing and upwelling of soil thermal radition that is intercepted by the lower-canopy layer, analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
+Noteably, this pattern can be reversed in open canopy forests or sparse stands, particularly in xeric sites, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the canopy and the soil that is intercepted by lower-canopy layers--analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
 
@@
 **The buffering patterns described above vary geographically, seasonally, and across forests types [REFS; *Defrenne et al. 2021*].**
-The strength of buffering decreases when (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) moving to open-forests in high elevation, and increases with canopy density and height in high latitudes [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018].
+The strength of buffering in high elevation across with forests, in some cases decreases when (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) moving to open-forests , and increases with canopy density [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018].
 Buffering capacity is also greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests.
 [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
-Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018**double check these refs--at least one didn't belong**].
-*(Presumably, the above is because leaf area is greater in set seasons)*
+Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in set seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018**double check these refs--at least one didn't belong**].
 
@@
 
-High wind speed (Fig. `r fig_NEON_vertical`) and large $g_a$ decrease $T_{leaf}$ - $T_{air}$ [Fig. `r fig_leaf_T`b, @daudetWindSpeedLeaf1999a]. 
+High wind speed (Fig. `r fig_NEON_vertical`) and large leaf $g_a$ decrease $T_{leaf}$ - $T_{air}$ [Fig. `r fig_leaf_T`b, @daudetWindSpeedLeaf1999a]. 
 Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]
@@
 
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to smaller $g_a$ than larger leaves that take longer due to large $ga$.
+**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to greater $g_a$ than larger leaves that take longer due to lower $ga$.
 Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves that can exprience greater $T_{leaf}$-$T_{air}$.
@@
 Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-Comparing forests, upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies[@mauTemperateTropicalForest2018a; @leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018] .
+Comparing forests, upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018] .
 At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
-Spatial distribution of canopy temperature also shows that tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
+Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
 
@@
 Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004] where degrees of direct to diffused light and resultant water demands shape sun vs.
-shade leaves, and their vascular architecture [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
@@
 Exposed-canopy leaves are better adapted to heat and water stress than shaded-leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al.
-2010], with adaptive strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998].
+2010], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 All, collectively regulate $T_{leaf}$-photosynthetic balance.
@@
 The distribution of biochemical traits enable photosynthesis and metabolic processes, and protect against heat and light damage in sun-leaves.
-$N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018].
+$N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018].
 Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
@@
 Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger than mature leaves, due to devoloping traits that can acclimate.
-Whereas, change in longterm-light for mature leaves can result in leaf abscission [Niinemetes, 2007].
+Whereas, change in longterm-light for mature leaves can result in leaf abscission [Niinemetes, 2007; *REFS*].
 *Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
@@
 
-Deciduousness as a trait is an adaptation to water and heat stress.
+Deciduousness as a trait is an adaptation to converse water under stress.
 Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter.
-Among tropical species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000b].
-Hence, deciduous leaf-loss can influence biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
-Spatio-temporal leaf-emergence in deciduous-temperate forests also shape biophysical gradient where leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
-2013; @seiwaChangesLeafPhenology1999].
+Among tropical species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000c].
+In deciduous-temperate forests, spatio-temporal leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
+2013; @seiwaChangesLeafPhenology1999].Therefore, deciduous leaf-loss or -gain can alter biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
 
@@
 
-Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].
-Where evapotranspiration and $gs$ peaks during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010] and declines during midday and afternoon due to stomtal closure [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b].
-While specie-specific differences in $T_{opt}$ of $gs$ can influence their photosynthetic heat tolerance indirectly; exactly how, remains understudied [@perezPhotosyntheticHeatTolerances2020].
+Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
 
@@
 2019 ; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], along the vertical gradient of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and along the vertical gradient of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
 
@@
 In 1994, Knapp et al.
-showed that in tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
+showed that in tall grass prarie gallery forest in Kansas, $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
 In both species photosynthetic heat tolerance also increased from June to August.
@@
 
-Further, leaf thermal tolerances vary less based on macroclimate, but might reflect microenvironment more [@feeley_thermal_2020; Slot et al.
+Further, leaf thermal tolerances are more adapted to microenvironment than macroclimate [@feeley_thermal_2020; Slot et al.
 2020].
 Particularly, in an Australian arid region, *Acacia Papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, that correlated with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
+However, more research is required to fully understand the relationship between leaf thermal tolerance and microenvironments.
 
@@
 
-While a few studies show $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+While a few studies show $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
 Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
@@
 Further, heat with drought exacerbates tree stress [@ruehrWaterAvailabilityDominant2016].
+Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
 
@@
 However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
-Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-.
 
@@
 It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
-Similarly, in tropical forests, decelarating tree growth rates predominantly in smaller trees, but not restricted to, are associated with higher nighttime temperatures and change in rain regimes [*Clark 2010; 2013; Feeley et al.
+Similarly, in tropical forests, decelarating tree growth rates predominantly in smaller trees, but not restricted to, are associated with higher nighttime temperatures that increase nighttime metabolic respiration potentially supressing growth [*Clark 2010; 2013; Feeley et al.
 2007*].
-Further, interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water.
+Further, interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height.
 Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].
@@
 
+[*jumbled sentences to work on*]
 Canopy buffering is crucial to microclimate functioning, which is influenced by hydrological-cycle at large [Davis et al.
 2019].
-Frequent droughts can thus cause canopy disturbances.While less severe drought can decrease tree canopy circumference [@aussenacInteractionsForestStands2000a], more severe droughts show increasing larger tree mortality in forests globally [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al.
+Frequent and severe droughts show increasing larger tree mortality in forests globally [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
-Hydraulic limitation in forests can also alters species thermoregulation [@sastry_leaf_2018].
-Particularly, increasing temperatures shows increase in canopy $T_{leaf}$ above ambient temperature, and sometimes beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Hydraulic limitation in forests also alters species thermoregulation [@sastry_leaf_2018].
+Further, increase in canopy temperature is also associated with GPP reduction in the tropics.
+-Increasing temperatures increase in canopy $T_{leaf}$ sometimes beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
 
-Increase in canopy temperature is also associated with GPP reduction in the tropics.
-Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011].
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].

### 2021-06-04 vertical-thermal-review.Rmd
small edits in ecology section [un-jumbling my sentences]

@@
 Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Accordingly, heat, associated with increase in temperature and VPD even under adequete water can cause greater tree hydraulic stress and altered and soil-water dynamics than milder drought, due to increased $T_{leaf}$ and $gs$ limitation [@ruehrWaterAvailabilityDominant2016].
+Accordingly, heat, associated with increase in temperature and VPD even under adequete water can cause greater tree hydraulic stress and altered and soil-water dynamics than milder drought, particulary in temperate species [@ruehrWaterAvailabilityDominant2016].
 However, soil water loss during drought effects larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
@@
 
-Phenologically, understory growth can be more sensitive to annual warming than canopy in temperate-deciduous forests, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
-Increase in summer temperature in mesic temperate forests is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
-However, late spring, short-term heat waves in temperate forests can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
+Phenologically in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
+Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
+However, late spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
 
@@
 This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
+However, evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees, but not restricted to, perhaps because increase in nighttime metabolic respiration could supress growth [*Clark 2010; 2013; Feeley et al.
+2007*].
+It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
+Further, interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height.
+Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].
+In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
+
 However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
 `r fig_leaf_T`).
-It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
-Similarly, in tropical forests, decelarating tree growth rates predominantly in smaller trees, but not restricted to, are associated with higher nighttime temperatures that increase nighttime metabolic respiration potentially supressing growth [*Clark 2010; 2013; Feeley et al.
-2007*].
-Further, interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height.
-Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].
-In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
 
@@
 
-The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are highly interconneced, resulting in feedbacks to another.
-Therefore, climate-driven changes, such as rising temperature, can have implications to forests on many levels.
+The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are highly interconneced, resulting in feedback loops to another.
+Therefore, climate-driven changes have implications on forests at many levels.
 
-### canopy disturbance
+### canopy disturbance and global warming
 
-[*jumbled sentences to work on*]
-Canopy buffering is crucial to microclimate functioning, which is influenced by hydrological-cycle at large [Davis et al.
-2019].
-Frequent and severe droughts show increasing larger tree mortality in forests globally [@senfCanopyMortalityHas2018a; @mcdowellMechanismsPlantSurvival2008; Brienen et al.
-2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a].
-Hydraulic limitation in forests also alters species thermoregulation [@sastry_leaf_2018].
-Further, increase in canopy temperature is also associated with GPP reduction in the tropics.
--Increasing temperatures increase in canopy $T_{leaf}$ sometimes beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment.
+This buffering is influenced by the hydraulogical-cycle at large, therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts.
+Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Increase in such canopy $T_{leaf}$ reduces forest GPP, specifically in the tropics, where thermal limits are narrower than temperate forests.
 
+Canopy structure affects understory species composition and soil fertility which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
+Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, despite buffering dampening extreme effects, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+This imposes an increasing threat to fauna in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
+
+Warming shifts forest composition, from leaf metabolism to whole forest structure.
+At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
+This can also potentially alter species thermoregulation [@sastry_leaf_2018].
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
-Recent adavances in canopy studies further show large loss of canopies in Europe caused by wind-induced disturbance and anthropogenic deforestation [Cornelius Senf, 2021; @senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020].
-Loss of big trees (e.g.@senfCanopyMortalityHas2018a ) would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity and composition which changes understory species composition and soil fertility [*@majasalmiImpactTreeCanopy2020*].Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Each of these changes have potential feedbacks to climate change.
+Feedback loops from loss of big trees through natural disturbance or anthropogenic (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al.
+2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020].
+Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 
-### warming
+While climate change driven forest shifts are observed, certain forests are more resilient to warming and droughts than others, such as African rainforests [Amy Bennett's paper].
+Forest resiliency is also observed through increasing temperature acclimation [martijn's paper and others] and reversible recovery of canopy heat stress [Smith et al.
+2020] among other mechanisms [for eg.
+*REF*]
 
-While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to fauna in microrefugias [@scheffersMicrohabitatsReduceAnimal2013].
-Heatwave and temperature  can alter understory tree grwoth rates, canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019].
-At a leaf metabolism level, fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
-
-Differential mortality within the canopy structure will alter the profile of metabolism and hence emergent forest function.
-Such alteration to canopy structure may be detectable by lidar and thermal remote sensing.
+These shifts in forest structure at large may be detectable by lidar and thermal remote sensing.
 If we know how metabolism maps to the thermal profile, then the observed forest structural changes can inform prediction of forest function.

### 2021-06-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`).**
- The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density ($LAD$) ie.
+ The distribution of this foliage is structured in correspondence with the distribution of tree sizes, and successional stages within a forest, where leaf area density ($LAD$) ie.
 leaf area per unit volume, increases with tree-height eventually forming the canopy layer.
@@
 in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
-bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985].
-In open canopy forests, such as needle-leaf boreal forests, LAD is predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016].SOil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs.
+bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a].
+In open canopy forests, such as needle-leaf forests, LAD is predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs.
 open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
@@
 However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds based on inherent specie-specific traits.
-Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007].
+Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space, with different levels of trait plasticity [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
 Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
@@
 
-Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
+Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
+This diurnal pattern and resulting $A_{max}$ differs between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].
 

### 2021-06-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **The buffering patterns described above vary geographically, seasonally, and across forests types [REFS; *Defrenne et al. 2021*].**
-The strength of buffering in high elevation across with forests, in some cases decreases when (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) moving to open-forests , and increases with canopy density [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018].
+The strength of buffering in high elevation with forest type, in some cases decreases when (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) moving to open-forests , and increases with canopy density [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018].
 Buffering capacity is also greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests.
 [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
-Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in set seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018**double check these refs--at least one didn't belong**].
+Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018**double check these refs--at least one didn't belong**].
 
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
- Inherent leaf $gs$ is generally greater across species exposed sun-leaves due to stomata density than shade leaves [Table 1].
+ Inherent leaf $gs$ is generally greater across species exposed sun-leaves than shade leaves, due to stomata density [Table 1].
 However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
@@
 ### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
+Leaf traits vary within an indivual tree at a vertical gradient from exposed canopy regions to shaded regions (intraspecific variation), and as well across mutiple trees' exposed canopy to shaded canopy regions (inter-specific variation).
 
-Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
+Therefore, across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
@@
 
-Within-canopy gradients of individuals, leaf traits vary--going from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020] that contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.
-Further, there is a need for research to characterize trait relationships and responses through the vertical canopy, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
+Leaf traits vary--from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020] that contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity.
+Further, the need for research to characterize trait relationships and responses through the vertical canopy is becoming a need, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
 
@@
 
-Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds based on inherent specie-specific traits.
-Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space, with different levels of trait plasticity [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
+Widely, across forest species, fundamental trait adaptation to light gradients are similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], that further vary with levels of trait plasticity [REF], for eg.
+between  canopy multi-species, and understory species, trait plasticity can vary several-folds based on inherent specie-specific traits, even under similar biophysical conditions respectively.
+Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space, with different levels of plasticity [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
 Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
-Therefore $T_{leaf}$ varies across speci-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
+Therefore $T_{leaf}$ varies across specie-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 In deciduous-temperate forests, spatio-temporal leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
-2013; @seiwaChangesLeafPhenology1999].Therefore, deciduous leaf-loss or -gain can alter biophysical gradient and $T_{leaf}$ [@rey-sanchez_spatial_2016].
+2013; @seiwaChangesLeafPhenology1999].Therefore, deciduous leaf-loss or -gain can alter biophysical and $T_{leaf}$ gradient [@rey-sanchez_spatial_2016].
 
@@
 showed that in tall grass prarie gallery forest in Kansas, $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
-In both species photosynthetic heat tolerance also increased from June to August.
+In both species $T_{50}$ also increased from June to August.
 However, more recently @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Panama tropical forest, but results weren't significant, with both measured near ground level.
@@
 
-### canopy disturbance and global warming
+### canopy disturbance and warming
 
@@
 Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Increase in such canopy $T_{leaf}$ reduces forest GPP, specifically in the tropics, where thermal limits are narrower than temperate forests.
+Increase in such canopy $T_{leaf}$ reduces forest GPP, specifically in the tropics where thermal limits are narrower than temperate forests.
 

### 2021-06-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 While a few studies show $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
-Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
+Contrastingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
 1*)[@bolstad_foliar_1999].
@@
 @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
-- Response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] in Quercus rubra hardly varies with canopy position, but canopy-position effect in respiration is much greater in mesic sites, and seasonal effect is greater in drier sites [@xuSeasonalVariationTemperature2006]
+
+Further, @xuSeasonalVariationTemperature2006 showed that the response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] in Quercus rubra hardly varied with canopy position.
+However, considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect is greater in drier sites.
 
@@
 
-Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment.
-This buffering is influenced by the hydraulogical-cycle at large, therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts.
+Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al.
+2019].
+This buffering is influenced by the hydraulogical-cycle at large [Davis et al.
+2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al.
+2015].
 Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Increase in such canopy $T_{leaf}$ reduces forest GPP, specifically in the tropics where thermal limits are narrower than temperate forests.
+Increase in such canopy $T_{leaf}$ reduces forest GPP [Pau et al.
+2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].[**to include Teskey 2015; Breashers et al.
+2021**]
 
@@
 
-While climate change driven forest shifts are observed, certain forests are more resilient to warming and droughts than others, such as African rainforests [Amy Bennett's paper].
-Forest resiliency is also observed through increasing temperature acclimation [martijn's paper and others] and reversible recovery of canopy heat stress [Smith et al.
+While climate change driven forest shifts are observed, certain forests are more resilient to warming and droughts than others, such as African rainforests [Amy Bennett's new paper].
+Forest resiliency is also observed through increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al.
 2020] among other mechanisms [for eg.

### 2021-06-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Further, @xuSeasonalVariationTemperature2006 showed that the response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] in Quercus rubra hardly varied with canopy position.
-However, considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect is greater in drier sites.
+Further, @xuSeasonalVariationTemperature2006 showed that the response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] in Quercus rubra also hardly varied with canopy position.
+However, considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites.
 
@@
 Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Accordingly, heat, associated with increase in temperature and VPD even under adequete water can cause greater tree hydraulic stress and altered and soil-water dynamics than milder drought, particulary in temperate species [@ruehrWaterAvailabilityDominant2016].
+Accordingly, heat, associated with increase in temperature and VPD even under adequete water can cause greater tree hydraulic stress and altered soil-water dynamics than milder drought, particulary in temperate species [@ruehrWaterAvailabilityDominant2016].
 However, soil water loss during drought effects larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
@@
 
+-Shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses (e.g., Bode et al.
+https://doi.org/10.1016/j.rse.2014.01.028, Musselman et al.
+https://doi.org/10.1016/j.rse.2013.05.021, Davis et al.
+https://doi.org/10.1016/j.agrformet.2019.02.015)
+
 ![**Figure `r fig_model`.

### 2021-06-09 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests.
-Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory for eg.
-in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
+Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
 bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a].
-In open canopy forests, such as needle-leaf forests, LAD is predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs.
+In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
+Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs.
 open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
@@
 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Noteably, this pattern can be reversed in open canopy forests or sparse stands, particularly in xeric sites, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the canopy and the soil that is intercepted by lower-canopy layers--analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
+Noteably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers--analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
 
@@
 ### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
-Leaf traits vary within an indivual tree at a vertical gradient from exposed canopy regions to shaded regions (intraspecific variation), and as well across mutiple trees' exposed canopy to shaded canopy regions (inter-specific variation).
+Leaf trait comparison within an individual tree here is referred to as intraspecific variation, and comparison of traits between species of multiple trees here is referred to as inter-specific variation.
 
@@
 2019].
-Under homogeneous light conditions, sun-leaf differences in canopy orientations are shaped by canopy $T_{air_{max}}$ and $VPD$ stress than solely by light [@mediavillaFoliarPlasticityRelated2019a].
+Under homogeneous canopy light conditions, sun-leaf differences in canopy orientations are shaped by canopy $T_{air_{max}}$ and $VPD$ stress than solely by light [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-The distribution of biochemical traits enable photosynthesis and metabolic processes, and protect against heat and light damage in sun-leaves.
-$N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018].
+The vertical distribution of biochemical traits enables photosynthesis and metabolic processes, and protects against heat and light damage in exposed canopy leaves.
+Particualy, $N_{area}$ that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018].
 Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
@@
 
-Widely, across forest species, fundamental trait adaptation to light gradients are similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], that further vary with levels of trait plasticity [REF], for eg.
-between  canopy multi-species, and understory species, trait plasticity can vary several-folds based on inherent specie-specific traits, even under similar biophysical conditions respectively.
-Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space, with different levels of plasticity [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
-Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
-Therefore $T_{leaf}$ varies across specie-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
+Widely, across forest species, the pattern of fundamental trait adaptation to light gradients is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+However, trait plasticity can vary several-folds between  multi-species canopy and understory,  based on their inherent specie-specific traits and environmental conditions.
+Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
+Along the ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ are observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+Therefore, collectively, $T_{leaf}$ differs across specie-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 
-Deciduousness as a trait is an adaptation to converse water under stress.
+Deciduousness as a trait is an adaptation to conserve water under stress.
 Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter.
@@
 
-Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
-This diurnal pattern and resulting $A_{max}$ differs between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].
+Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steep in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
+This diurnal pattern and resulting $A_{max}$ varies between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].
 

### 2021-06-09 vertical-thermal-review.Rmd
editing respiration part

@@
 ### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
-Leaf trait comparison within an individual tree here is referred to as intraspecific variation, and comparison of traits between species of multiple trees here is referred to as inter-specific variation.
+Leaf trait comparison within an individual tree of the same species, here, is referred to as intraspecific variation, and comparison of traits between species of multiple trees is referred to as interspecific variation.
 
@@
 
-Leaf traits vary--from buffered [shaded] to exposed canopy positions [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020] that contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+Within-canopy leaf traits vary--from shaded canopy position to exposed positions [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020] that contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity.
 Further, the need for research to characterize trait relationships and responses through the vertical canopy is becoming a need, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
+
+On another note, along the ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ are observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
 
@@
 Widely, across forest species, the pattern of fundamental trait adaptation to light gradients is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-However, trait plasticity can vary several-folds between  multi-species canopy and understory,  based on their inherent specie-specific traits and environmental conditions.
+However, trait plasticity can vary several-folds between  multi-species canopy and in understory species,  based on their inherent specie-specific traits and environmental conditions.
 Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
-Along the ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ are observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
 Therefore, collectively, $T_{leaf}$ differs across specie-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
@@
 Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steep in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
-This diurnal pattern and resulting $A_{max}$ varies between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].
+However, $T_{opt}$ of $gs$ doesnot vary between sun and shade leaves [Slot et al.
+2019].
+The diurnal $gs$ pattern and resulting $A_{max}$ varies between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].
 
@@
 
-While a few studies show $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
-Contrastingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig.
-1*)[@bolstad_foliar_1999].
+Leaf respiration is lower during the day than the night, due to light-induced inhibition that is highest in high-irradiance and moderate-to-high-temperature and lowest at high-irradiance and low-temperature [@atkinLeafRespirationSnow2000; @atkinLeafRespirationSnow2000; @keenanWidespreadInhibitionDaytime2019; @zaragoza-castellsDoesGrowthIrradiance2007].
+Light-induced inhibition of leaf respiration influences $Q_{10}$ where $Q_{10}$ is greater in the in the dark, and decreases with increase in irradiance [@atkinLeafRespirationSnow2000].
+
+However, when it comes to sun and shade leaves, a few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] and other studies have observed no difference in $Q_{10}$ across species vertical profiles [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999], and at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig.
+1*).
 (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
@@
 @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
-
-Further, @xuSeasonalVariationTemperature2006 showed that the response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] in Quercus rubra also hardly varied with canopy position.
-However, considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites.
+Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] showed no variation in canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
 

### 2021-06-09 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf respiration is lower during the day than the night, due to light-induced inhibition that is highest in high-irradiance and moderate-to-high-temperature and lowest at high-irradiance and low-temperature [@atkinLeafRespirationSnow2000; @atkinLeafRespirationSnow2000; @keenanWidespreadInhibitionDaytime2019; @zaragoza-castellsDoesGrowthIrradiance2007].
-Light-induced inhibition of leaf respiration influences $Q_{10}$ where $Q_{10}$ is greater in the in the dark, and decreases with increase in irradiance [@atkinLeafRespirationSnow2000].
+Leaf respiration is lower during the day than the night, due to light-induced inhibition that is highest in high-irradiance and moderate-to-high-temperature and lowest at high-irradiance and low-temperature [@atkinLeafRespirationSnow2000; @keenanWidespreadInhibitionDaytime2019; @zaragoza-castellsDoesGrowthIrradiance2007].
+Light-induced inhibition of leaf respiration influences $Q_{10}$ where $Q_{10}$ is greatest in the dark, and decreases with increase in irradiance [@atkinLeafRespirationSnow2000].
 

### 2021-06-10 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-However, when it comes to sun and shade leaves, a few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] and other studies have observed no difference in $Q_{10}$ across species vertical profiles [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999], and at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig.
+However, when it comes to sun and shade leaves, a few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] and other studies have observed no difference in $Q_{10}$ across species vertical profiles and seedling sun vs.
+shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008], and at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig.
 1*).
@@
 Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Accordingly, heat, associated with increase in temperature and VPD even under adequete water can cause greater tree hydraulic stress and altered soil-water dynamics than milder drought, particulary in temperate species [@ruehrWaterAvailabilityDominant2016].
+Accordingly, in temperate species, heat stress caused due to increase in temperature and VPD even under adequete water caused greater tree hydraulic stress and altered soil-water dynamics than milder drought [@ruehrWaterAvailabilityDominant2016].
 However, soil water loss during drought effects larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
-Further, heat with drought exacerbates tree stress [@ruehrWaterAvailabilityDominant2016].
 Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
@@
 Increase in such canopy $T_{leaf}$ reduces forest GPP [Pau et al.
-2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].[**to include Teskey 2015; Breashers et al.
-2021**]
+2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].
+Further, warm-season associated with increased VPD, along with drought exacerbates tree stress at a regional scale, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013]
+[**to include Teskey 2015; Breashers et al. 2021**]
 

### 2021-06-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-The vertical distribution of biochemical traits enables photosynthesis and metabolic processes, and protects against heat and light damage in exposed canopy leaves.
-Particualy, $N_{area}$ that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018].
-Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-Constraints to biochemical traits, particularly $VAZ$ can curb increasing temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a].
+These traits protect leaves along the vertical profile from photo-damage and heat stress, and are crucial to the functioning of photosynthesis and metabolism.
+Biochemically, caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light to heat dissaptaion [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+Evidence also shows that $VAZ$ enables leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
+Along with $VAZ$, Narea optimizes to prevent photo-inhibition during direct light and sunflecks in sun- and shade leaves, respectively, and is typically greater in the direction of new leaf growth [@kitaoCanopyNitrogenDistribution2018].
 Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed-canopies, making these species particularly advantageous [*REF*].

### 2021-06-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf respiration is lower during the day than the night, due to light-induced inhibition that is highest in high-irradiance and moderate-to-high-temperature and lowest at high-irradiance and low-temperature [@atkinLeafRespirationSnow2000; @keenanWidespreadInhibitionDaytime2019; @zaragoza-castellsDoesGrowthIrradiance2007].
-Light-induced inhibition of leaf respiration influences $Q_{10}$ where $Q_{10}$ is greatest in the dark, and decreases with increase in irradiance [@atkinLeafRespirationSnow2000].
-
-However, when it comes to sun and shade leaves, a few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996] and other studies have observed no difference in $Q_{10}$ across species vertical profiles and seedling sun vs.
-shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008], and at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig.
+A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas Griffin et al.
+2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
+Other studies have observed no difference in $Q_{10}$ across species vertical profiles and seedling sun vs.
+shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008], although at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig.
 1*).

### 2021-06-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**Wind speeds are also higher at the top of the canopy, owing to... [@juckerCanopyStructureTopography2018].**
+**Wind speeds are also higher at the top of the canopy, owing to height of the canopy [@juckerCanopyStructureTopography2018].**
 Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites.
@@
 In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Additionally, increase in canopy height and LAI, upto a certain height, can also lower maximum $T_{air}$ and VPD in the understory [@juckerCanopyStructureTopography2018].
+Additionally, increase in canopy height and LAI contributes to lowering maximum $T_{air}$ and VPD in the understory [@juckerCanopyStructureTopography2018].
 *( This last sentence needs work.)*
@@
 **The buffering patterns described above vary geographically, seasonally, and across forests types [REFS; *Defrenne et al. 2021*].**
-The strength of buffering in high elevation with forest type, in some cases decreases when (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) moving to open-forests , and increases with canopy density [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018].
+The strength of buffering in high elevation, in some cases, decreases when (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) moving to open-forests, and increases with canopy density [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018].
 Buffering capacity is also greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests.
@@
 **However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
-Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 Comparing forests, upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018] .
@@
 ### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
-Leaf trait comparison within an individual tree of the same species, here, is referred to as intraspecific variation, and comparison of traits between species of multiple trees is referred to as interspecific variation.
+Leaf trait comparison within an individual tree of the same species, here, is referred to as within-canopy intraspecific variation, and comparison of traits between species of multiple trees is referred to as interspecific variation.
 
@@
 
-Within-canopy leaf traits vary--from shaded canopy position to exposed positions [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020] that contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+Within-canopy leaf traits vary--along the canopy light gradient, ie.
+in sun and shade leaves [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020] that contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity.
@@
 
-On another note, along the ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ are observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+Intraspecific traits also vary along the ontogeny of the same species in canopy and understory, for example, LMA, photosynthetic capacity [per area] and $gs$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
 
@@
 2019].
-Under homogeneous canopy light conditions, sun-leaf differences in canopy orientations are shaped by canopy $T_{air_{max}}$ and $VPD$ stress than solely by light [@mediavillaFoliarPlasticityRelated2019a].
+Under homogeneous canopy light conditions, sun-leaves in different canopy orientations are primarily shaped by differences in $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-These traits protect leaves along the vertical profile from photo-damage and heat stress, and are crucial to the functioning of photosynthesis and metabolism.
-Biochemically, caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light to heat dissaptaion [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-Evidence also shows that $VAZ$ enables leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
-Along with $VAZ$, Narea optimizes to prevent photo-inhibition during direct light and sunflecks in sun- and shade leaves, respectively, and is typically greater in the direction of new leaf growth [@kitaoCanopyNitrogenDistribution2018].
+Biochemical protection against foliage light and heat damage increases with irradiance.
+Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998], this is particularly important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
+N and P are crucial for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco.
+$N_{area}$ increases in the direction of new leaf growth, where it optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018].
 Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed-canopies, making these species particularly advantageous [*REF*].
-
-### Interspecific variation
-
-Widely, across forest species, the pattern of fundamental trait adaptation to light gradients is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-However, trait plasticity can vary several-folds between  multi-species canopy and in understory species,  based on their inherent specie-specific traits and environmental conditions.
-Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
-Therefore, collectively, $T_{leaf}$ differs across specie-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 
-New-leaf growth along the light gradient constantly shades leaves underneath.
+New-leaf growth in response to the light gradient constantly shades leaves underneath.
 Therefore, exposed-canopies have greater proportion of young leaves than shaded regions that have older leaves.
@@
 This is bound to affect metabolism, and may affect other aspects of their performance.*
+
+### Interspecific variation
+
+Widely, across forest species, the pattern of fundamental trait adaptation to light gradients is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+However, trait plasticity can vary several-folds between  multi-species canopy and in understory species, including at an intraspecific level,  based on their inherent specie-specific traits and responses to environmental conditions.
+Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
+Therefore,  $T_{leaf}$ differs at an intra- and inter-specific levels with traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 
-Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steep in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
+Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
 However, $T_{opt}$ of $gs$ doesnot vary between sun and shade leaves [Slot et al.
 2019].
-The diurnal $gs$ pattern and resulting $A_{max}$ varies between species, such as early- and late-successional [@ishiiExploringRelationshipsCanopy2004a], and between gymnosperms and angiosperms [@bonan_ecological_2016].
-
+The diurnal $gs$ pattern and resulting $A_{max}$ varies between species, for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
 ### Photosynthesis
@@
 Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), where earlier evidence showed that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+14.5 in Campbell and Norman 1998; *revisit this!*), where earlier evidence showed that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.

### 2021-06-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The diurnal $gs$ pattern and resulting $A_{max}$ varies between species, for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
+
 ### Photosynthesis
@@
 Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*), where earlier evidence showed that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+14.5 in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
@@
 
-Interestingly, within canopy Japanese oak sun and shade leaves under respective peak light conditions were found to maintain similar rate of excess energy production.
-Excessive light energy not utilized can cause photoinhibition.Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain high photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018].
+Interestingly, within canopy Japanese oak sun and shade leaves were found to maintain similar rate of excess energy production under respective peak light conditions.
+Excessive light energy not utilized can cause photoinhibition.
+Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain high photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018].
 Perhaps, differences in these thermal stablization strategies realizes similar $T_{opt}$ for photosynthesis in sun and shade leaves.
@@
 @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
-Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] showed no variation in canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
+Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] also showed no variation in canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
 
@@
 2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].
-Further, warm-season associated with increased VPD, along with drought exacerbates tree stress at a regional scale, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013]
+A combination of warm-season associated with increased VPD, along with drought exacerbates tree stress at a regional scale, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013]
 [**to include Teskey 2015; Breashers et al. 2021**]
@@
 
-While climate change driven forest shifts are observed, certain forests are more resilient to warming and droughts than others, such as African rainforests [Amy Bennett's new paper].
-Forest resiliency is also observed through increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al.
+In the face of global climate change, certain forests, such as African rainforests [Amy Bennett's new paper] are more resilient to warming and drought than other temperate and tropical forests.
+Additionally, increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al.
 2020] among other mechanisms [for eg.
-*REF*]
+*REF*], point towards rays of hope.
 

### 2021-06-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-In the face of global climate change, certain forests, such as African rainforests [Amy Bennett's new paper] are more resilient to warming and drought than other temperate and tropical forests.
-Additionally, increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al.
-2020] among other mechanisms [for eg.
-*REF*], point towards rays of hope.
+In the face of global climate change, resiliency among forests are also observed, such as African rainforests [Amy Bennett's new paper] that are more resilient to warming and drought than other forests.
+Increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al.
+2020] are other mechanisms [for eg.
+*REF*] that will help cope with warming.
 

### 2021-06-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Accordingly, in temperate species, heat stress caused due to increase in temperature and VPD even under adequete water caused greater tree hydraulic stress and altered soil-water dynamics than milder drought [@ruehrWaterAvailabilityDominant2016].
-However, soil water loss during drought effects larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].
-Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Biophysical conditions can alter tree-water dynamics and growth patterns.
+Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species.
+Recovery time from heat stress varies between species hydraulic traits [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013].
+Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Altered precipitation regimes causing drought is more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, @bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
+However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
+`r fig_leaf_T`).
+
+Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light due to greater non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a].
+This can also alter tree growth patterns.
+Particularly higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014].
+Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supresses growth [*Clark 2010; 2013; Feeley et al.
+2007*].
+.
+
+**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
+In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
+This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
+ It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
+In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
 
@@
 However, late spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
-
-**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
-In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
-This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
-However, evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees, but not restricted to, perhaps because increase in nighttime metabolic respiration could supress growth [*Clark 2010; 2013; Feeley et al.
-2007*].
-It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
-Further, interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height.
-Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].
-In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
-
-However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
-`r fig_leaf_T`).
 
@@
 
-### canopy disturbance and warming
+### canopy disturbance
 
@@
 
+###warming
 Warming shifts forest composition, from leaf metabolism to whole forest structure.
 At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
-This can also potentially alter species thermoregulation [@sastry_leaf_2018].
+However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [Keenan et al.
+2019].
+However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pamela templer et al.].
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
+
 Feedback loops from loss of big trees through natural disturbance or anthropogenic (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al.
@@
 2020] are other mechanisms [for eg.
-*REF*] that will help cope with warming.
+*REF*] that will help forests cope with warming.
 

### 2021-06-13 vertical-thermal-review.Rmd

@@
 Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species.
-Recovery time from heat stress varies between species hydraulic traits [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013].
+Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013].
 Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
@@
 
-Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light due to greater non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a].
-This can also alter tree growth patterns.
+Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a].
+Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [PW West, 2019].
+This difference can alter tree growth patterns in response to biophysical conditions.
 Particularly higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014].
@@
 Phenologically in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
+However, late spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
 Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
-However, late spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
 
@@
 
-Feedback loops from loss of big trees through natural disturbance or anthropogenic (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al.
+Feedback loops from loss of big trees through natural or anthropogenic disturbance (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al.
 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020].
@@
 
-In the face of global climate change, resiliency among forests are also observed, such as African rainforests [Amy Bennett's new paper] that are more resilient to warming and drought than other forests.
+In the face of global climate change, resiliency among forests also stand out, such as African rainforests [Amy Bennett's new paper] that are more resilient to warming and drought than other forests.
 Increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al.

### 2021-06-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2014*].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
-Leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities], leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
 
@@
 
-**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the shape of profile modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
+**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the shape of profile modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
 `r fig_NEON_vertical`, @koike_leaf_2001].**
@@
 
-**Wind speeds are also higher at the top of the canopy, owing to height of the canopy [@juckerCanopyStructureTopography2018].**
+**Wind speeds are also higher at the top of the canopy, owing to greater exposure to macroclimatic conditions with canopy height [@juckerCanopyStructureTopography2018].**
 Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites.
@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
-The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
+The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
 Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al.
 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Noteably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers--analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
+Noteably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
 
@@
 Often, however, $T_{leaf}$ tracks $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020]. 
-Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig.
+Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost to the environment (such as transpiration) [Fig.
 `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].
@@
 **In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to greater $g_a$ than larger leaves that take longer due to lower $ga$.
-Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves that can exprience greater $T_{leaf}$-$T_{air}$.
-With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures than large leaf species [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves.
+With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
 
-This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
+Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
 When stomatal conductance ($g_s$) is high, evapotranspiration facilitates heat dissipation. 
@@
 **(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
- Inherent leaf $gs$ is generally greater across species exposed sun-leaves than shade leaves, due to stomata density [Table 1].
-However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
+ Across species, inherent leaf $gs$ is generally greater in sun leaves than shade leaves, due to stomata density [Table 1].
+However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--including increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, resulting in increased in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
 On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
-
 Lastly, canopy structure and architecture adaptation can also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
@@
 `r fig_leaf_T`).**
-In open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground can collectively increase $T_{leaf}$ [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+In open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
 Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
@@
 Comparing forests, upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018] .
-At a specie-specific level, leaf boundary resistance is lower for confiers than broad-leaf trees, followed by higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies[@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
+At a specie-specific level, leaf boundary resistance is lower for conifers than broad-leaf species, resulting in higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
 Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
@@
 **Leaf thermoregulation is controlled by leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016].
-Therefore, are critical in maintaining desirable $T_{leaf}$ and shaping leaf metabolism (following section) across forest vertical strata (Figs.
+Therefore, leaf traits are critical in maintaining desirable $T_{leaf}$ and shaping leaf metabolism (following section) across forest vertical strata (Figs.
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
@@
 `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
-At greater tree height, biological--tree-height--and biophysical--light, temperature, and evaporative demand (Fig.
-`r fig_NEON_vertical`)--also play a role in shaping leaf traits.
+At greater tree height, both biological variables (tree-height) and biophysical variables (light, temperature, and evaporative demand )(Fig.
+`r fig_NEON_vertical`) also play a role in shaping leaf traits.
 
-In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m.
+In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained at the tops of(~100 m.
 tall *Sequoia sempervirens*; *Ambrose et al.
@@
 Jensen and Maciej A.
-Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have shown to decrease, while leaf mass per area (at greater than 20m) to increase  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].
+Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have been shown to decrease, and leaf mass per area (at greater than 20m) to increase, with height  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].
 
 Disentangling tree-height related biological and biophysical drivers (Fig.
-`r fig_NEON_vertical`) of leaf traits remains an important area for research that is sparse, but is beyond the scope of the current review.
+`r fig_NEON_vertical`) of leaf traits remains an important area for further research, but is beyond the scope of the current review.
 Here, our primary interest is in how leaf traits and function vary across the vertical gradient from the canopy-top to the understory in forests.
@@
 ### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
-Leaf trait comparison within an individual tree of the same species, here, is referred to as within-canopy intraspecific variation, and comparison of traits between species of multiple trees is referred to as interspecific variation.
+Leaf trait comparison within the same species, here, is referred to as within-canopy intraspecific variation, and comparison of traits across species is referred to as interspecific variation.
 
@@
 
-Within-canopy leaf traits vary--along the canopy light gradient, ie.
-in sun and shade leaves [@casasSunShadeLeaves2011]-- at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020] that contribute to [understanding] vertical $T_{leaf}$ dynamics [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+Within-canopy leaf traits vary along the canopy light gradient, ie.
+in sun and shade leaves [@casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+This variation is crucial to understand vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity.
-Further, the need for research to characterize trait relationships and responses through the vertical canopy is becoming a need, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
+Further, research to characterize trait relationships and responses through the vertical canopy is needed, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
 
@@
 
-Exposed-canopy leaves are better adapted to heat and water stress than shaded-leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al.
-2010], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*].
-All, collectively regulate $T_{leaf}$-photosynthetic balance.
-Whereas, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
+Exposed canopy leaves are better adapted to heat and water stress than shade leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*], showing greater , hydraulically efficiency than shade leaves [Brodribb et al.
+2010].
+All these traits, collectively regulate $T_{leaf}$-photosynthetic balance.
+On the other hand, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
 2019].
@@
 Biochemical protection against foliage light and heat damage increases with irradiance.
-Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998], this is particularly important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
+Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+This is particularly important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
 N and P are crucial for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco.
 $N_{area}$ increases in the direction of new leaf growth, where it optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018].
-Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed-canopies, making these species particularly advantageous [*REF*].
+Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species particularly advantageous [*REF*].
 
@@
 Therefore, exposed-canopies have greater proportion of young leaves than shaded regions that have older leaves.
-Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger than mature leaves, due to devoloping traits that can acclimate.
-Whereas, change in longterm-light for mature leaves can result in leaf abscission [Niinemetes, 2007; *REFS*].
+Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger than mature leaves, due to developing traits that can acclimate, whereas, change in longterm-light for mature leaves can result in leaf abscission [Niinemetes, 2007; *REFS*].
 *Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
@@
 Widely, across forest species, the pattern of fundamental trait adaptation to light gradients is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-However, trait plasticity can vary several-folds between  multi-species canopy and in understory species, including at an intraspecific level,  based on their inherent specie-specific traits and responses to environmental conditions.
-Spatio-temporal gradients, offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
+However, trait plasticity can vary several-folds between  multi-species canopy and in understory species,  based on their inherent specie-specific traits and responses to environmental conditions.
+Spatio-temporal gradients, offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
 Therefore,  $T_{leaf}$ differs at an intra- and inter-specific levels with traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
@@
 
-Deciduousness as a trait is an adaptation to conserve water under stress.
+Deciduousness as a trait in tropical forests is an adaptation to conserve water under stress.
 Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter.
@@
 
-Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
+Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sunleaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
 However, $T_{opt}$ of $gs$ doesnot vary between sun and shade leaves [Slot et al.
 2019].
-The diurnal $gs$ pattern and resulting $A_{max}$ varies between species, for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
+The diurnal $gs$ pattern and resulting $A_{max}$ varies between species; for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
 
@@
 
-In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma.
+In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubisico, and multiple layers of palisade parenchyma.
 Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P (and water availability).
@@
 Direct, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008].
-$T_{leaf}$ beyond photosynthetic $T_{opt}$, declines photosynthesis due it's peaked response to temperature.
+Photosynthesis has a peaked response to temperature.
 Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.

### 2021-06-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**Wind speeds are also higher at the top of the canopy, owing to greater exposure to macroclimatic conditions with canopy height [@juckerCanopyStructureTopography2018].**
+**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
 Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites.
@@
 ### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
-Leaf trait comparison within the same species, here, is referred to as within-canopy intraspecific variation, and comparison of traits across species is referred to as interspecific variation.
-
+Leaf trait variation within the same species, here, is referred to as within-canopy intraspecific variation, and comparison of traits across species is referred to as interspecific variation.
 Therefore, across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
@@
 in sun and shade leaves [@casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-This variation is crucial to understand vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This variation is crucial, to understand vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity.
@@
 
-Exposed canopy leaves are better adapted to heat and water stress than shade leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*], showing greater , hydraulically efficiency than shade leaves [Brodribb et al.
+Exposed canopy leaves are better adapted to heat and water stress than shade leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*], showing greater hydraulically efficiency than shade leaves [Brodribb et al.
 2010].
@@
 This difference can alter tree growth patterns in response to biophysical conditions.
-Particularly higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014].
+Particularly, higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014].
 Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supresses growth [*Clark 2010; 2013; Feeley et al.
@@
 
-###warming
+### warming
+
 Warming shifts forest composition, from leaf metabolism to whole forest structure.

### 2021-06-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].
-A combination of warm-season associated with increased VPD, along with drought exacerbates tree stress at a regional scale, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013]
 [**to include Teskey 2015; Breashers et al. 2021**]
@@
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
+Further, combination of warm-season associated with increased VPD, along with drought exacerbates tree stress, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013].
 

### 2021-06-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-Comparing forests, upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018] .
+Comparing upper canopies, temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however considering a broader range of forests, this might not be always the case (Fig.
+`r fig_NEON_vertical`).
 At a specie-specific level, leaf boundary resistance is lower for conifers than broad-leaf species, resulting in higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].

### 2021-06-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
-[ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/leaf_energy_balance.png)
+[ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/Updated T_leaf_png.png)
 
@@
 
-#### Morphological traits
+#### Morphological and physiological traits
 
@@
 
-#### Biochemical and physiological traits
+#### Biochemical traits
 
@@
 Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-This is particularly important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
-N and P are crucial for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco.
+This is also important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
+N and P are essential for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco.
 $N_{area}$ increases in the direction of new leaf growth, where it optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018].
@@
 In both species $T_{50}$ also increased from June to August.
-However, more recently @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species in Panama tropical forest, but results weren't significant, with both measured near ground level.
+However, more recently @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species, measured near ground level, in Panama tropical forest, but results weren't significant.
 
@@
 2021; Perez and Feeley, 2020a] and more by leaf traits.
-@slotLeafHeatTolerance2020 found that $T_{crit}$ was greater among tropical species' large leaves with slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves with greater construction costs--greater LMA--compared to species with low LMA.
-@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance, experienced higher $T_{leaf_{max}}$, but showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$)
+$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves that had greater LMA (greater construction costs), compared to species with low LMA [@slotLeafHeatTolerance2020].
+@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance, experienced higher $T_{leaf_{max}}$, but showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).
 
@@
 
-A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas Griffin et al.
-2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
+A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
 Other studies have observed no difference in $Q_{10}$ across species vertical profiles and seedling sun vs.
@@
 Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Altered precipitation regimes causing drought is more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, @bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
+Altered precipitation regimes causing drought is more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
 However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
@@
 
-Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a].
-Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [PW West, 2019].
-This difference can alter tree growth patterns in response to biophysical conditions.
+Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to multiple stressors [@niinemetsResponsesForestTrees2010a].
+Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [@IncreasingRespiratoryCosts].
+These difference can alter tree growth patterns in response to biophysical conditions.
 Particularly, higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014].
-Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supresses growth [*Clark 2010; 2013; Feeley et al.
+Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supress growth [*Clark 2010; 2013; Feeley et al.
 2007*].
@@
 Phenologically in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
-However, late spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
+However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
 Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
@@
 Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Increase in such canopy $T_{leaf}$ reduces forest GPP [Pau et al.
-2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].
+Increase in such canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].
 [**to include Teskey 2015; Breashers et al. 2021**]
@@
 At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
-However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [Keenan et al.
-2019].
-However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pamela templer et al.].
+However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [@keenanWidespreadInhibitionDaytime2019].
+However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Templer et al.].
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
-Further, combination of warm-season associated with increased VPD, along with drought exacerbates tree stress, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013].
+Further, combination of warm-season associated with increased VPD, along with drought exacerbates tree stress[@parkwilliamsTemperaturePotentDriver2013] leading to mortality [@gouldenCaliforniaForestDieoff2019], with implications for forest structure.
 
-Feedback loops from loss of big trees through natural or anthropogenic disturbance (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al.
-2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020].
+Feedback loops from loss of big trees through natural or anthropogenic disturbance (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020].
 Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 
-In the face of global climate change, resiliency among forests also stand out, such as African rainforests [Amy Bennett's new paper] that are more resilient to warming and drought than other forests.
-Increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al.
-2020] are other mechanisms [for eg.
-*REF*] that will help forests cope with warming.
+In the face of global climate change, resiliency among forests also stand out, such as African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021].
+Increasing temperature acclimation among tropical, temperate and boreal species [@slotGeneralPatternsAcclimation2015;@slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019; @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] are other mechanisms that will help forests cope with warming.
 

### 2021-06-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 In the face of global climate change, resiliency among forests also stand out, such as African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021].
-Increasing temperature acclimation among tropical, temperate and boreal species [@slotGeneralPatternsAcclimation2015;@slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019; @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] are other mechanisms that will help forests cope with warming.
+Increasing temperature acclimation among tropical, temperate and boreal species [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019; @cunninghamTemperateRainforestTrees2003; @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] are other mechanisms that will help forests cope with warming.
 

### 2021-06-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubisico, and multiple layers of palisade parenchyma.
-Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy).The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P (and water availability).
+In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubisco, and multiple layers of palisade parenchyma.
+Light saturation levels differ between leaves and canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P, and water availability.
 
-Temperature can affect photosynthesis via direct and indirect pathways.
-Direct, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to temperature.
-Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
+Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008].
+Photosynthesis has a peaked response to temperature, commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
 2017; @slotSituTemperatureResponse2017).
@@
 @slotSituTemperatureResponse2017; Smith et al.
-2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g.
+2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g.
 Sage & Kubien 2007; Varhammer et al.
 2015 ).
-Leaf heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].
+Leaf heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II], leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].
 
 **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
-Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
+Biophysically, we might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
 14.5 in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
@@
 
-Interestingly, within canopy Japanese oak sun and shade leaves were found to maintain similar rate of excess energy production under respective peak light conditions.
+Interestingly, within-canopy Japanese oak sun and shade leaves were found to maintain similar rates of excess energy production under respective peak light conditions.
 Excessive light energy not utilized can cause photoinhibition.
-Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain high photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018].
-Perhaps, differences in these thermal stablization strategies realizes similar $T_{opt}$ for photosynthesis in sun and shade leaves.
+Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain higher photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018].
+Perhaps, differences in these thermal stablization strategies realize similar $T_{opt}$ for photosynthesis in sun and shade leaves.
 
@@
 
-**Thermal tolerance (e.g., $T_{50}$) sparsely varies with exposure among recent studies, but there are no studies isolating effect of height.**
+**Thermal tolerance (e.g., $T_{50}$) seldom varies with exposure among recent studies, but there are no studies isolating effect of height.**
 
 In 1994, Knapp et al.
-showed that in tall grass prarie gallery forest in Kansas, $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions.
+showed that in tall grass prarie gallery forest in Kansas, $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions.
 In both species $T_{50}$ also increased from June to August.
-However, more recently @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun-exposed leaves for two of three species, measured near ground level, in Panama tropical forest, but results weren't significant.
+However, more recently @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three species, measured near ground level, in Panama tropical forest, but results were not significant.
 
-Further, leaf thermal tolerances are more adapted to microenvironment than macroclimate [@feeley_thermal_2020; Slot et al.
+Further, leaf thermal tolerances are more closely adapted to microenvironment than macroclimate [@feeley_thermal_2020; Slot et al.
 2020].
-Particularly, in an Australian arid region, *Acacia Papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, that correlated with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
+Particularly, in an Australian arid region, *Acacia papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
 However, more research is required to fully understand the relationship between leaf thermal tolerance and microenvironments.
@@
 2021; Perez and Feeley, 2020a] and more by leaf traits.
-$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves that had greater LMA (greater construction costs), compared to species with low LMA [@slotLeafHeatTolerance2020].
+$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves with greater LMA (greater construction costs), compared to species with low LMA [@slotLeafHeatTolerance2020].
 @perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance, experienced higher $T_{leaf_{max}}$, but showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).
@@
 **Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves (Chen et al.
-2020), but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**
+2020), but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**
 
 A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
-Other studies have observed no difference in $Q_{10}$ across species vertical profiles and seedling sun vs.
-shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008], although at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig.
+Other studies have observed no difference in $Q_{10}$ across species vertical profiles and in seedling sun vs.
+shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008].
+Although, @bolstad_foliar_1999 found that temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased, [@bolstad_foliar_1999](*see their Fig.
 1*).
 (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
-May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*).
-@bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
-Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] also showed no variation in canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
+May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) and also found evidence of acclimation, but there remained a declining trend of respiration with topographical elevation.
+Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] also showed no variation within canopy positions in Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
 
@@
 Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013].
-Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Altered precipitation regimes causing drought is more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
+Whereas, greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Altered precipitation regimes causing drought are more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
 However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
@@
 
-Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to multiple stressors [@niinemetsResponsesForestTrees2010a].
+Tolerance to multiple biophysical and biological stressors can differ among tree size classes, having been shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to multiple stressors [@niinemetsResponsesForestTrees2010a].
 Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [@IncreasingRespiratoryCosts].
 These difference can alter tree growth patterns in response to biophysical conditions.
-Particularly, higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014].
-Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supress growth [*Clark 2010; 2013; Feeley et al.
+Particularly, higher temperatures are shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014].
+Evidence in tropical forests also shows that higher nighttime temperatures are associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supress growth [*Clark 2010; 2013; Feeley et al.
 2007*].
@@
 
-**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
+**Recently emerging evidence indicates that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
 In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
@@
 
-Phenologically in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
-However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].
+Phenologically, in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
+However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves more so than in understory saplings that are buffered [@filewodImpactsSpringHeat2014].
 Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].

### 2021-06-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
-bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a].
+bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
 In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
 Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs.
-open canopy forests [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
+open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
 In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019).
@@
 Isolating height (>35m) from light shows that, canopy $gs$ decreases with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
-Morphologically, leaf size [Kaare H.
-Jensen and Maciej A.
-Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have been shown to decrease, and leaf mass per area (at greater than 20m) to increase, with height  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].
+Morphologically, leaf size [@jensenPhysicalLimitsLeaf2013b] and lobation [@kusiPlasticLeafMorphology2020] have been shown to decrease, and leaf mass per area (at greater than 20m) to increase, with height  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017; @marshallFoliageHeightInfluences2003].
 
@@
 
-#### Morphological and physiological traits
+#### Anatomical, Morphological and Physiological traits
 
-Exposed canopy leaves are better adapted to heat and water stress than shade leaves where sun-leaves are often thicker, smaller, hairier, with greater leaf dissections [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], with structural strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*], showing greater hydraulically efficiency than shade leaves [Brodribb et al.
-2010].
-All these traits, collectively regulate $T_{leaf}$-photosynthetic balance.
-On the other hand, shaded leaves have lower convective cooling strategies, and invest more in light capture [@fauset_differencsunes_2018; @leighInfluenceLeafSize2017; Wright et al.
-2019].
-Under homogeneous canopy light conditions, sun-leaves in different canopy orientations are primarily shaped by differences in $T_{air_{max}}$ and $VPD$ stress [@mediavillaFoliarPlasticityRelated2019a].
+Exposed canopy environment induces traits in sun leaves that are better adaptated to heat and water stress than shade leaves through, for instance, being smaller, hairier, with greater vein density and LMA [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], as well as strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+These traits [see Table 1] collectively dissapate heat and regulate $T_{leaf}$-photosynthetic balance.
+Whereas, shade leaves have traits that enable rapid light capture in the diffused-light environment with sunflecks, for example greater SLA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a].
+In open canopies where light is comparitevely homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, than solely by light [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-Biochemical protection against foliage light and heat damage increases with irradiance.
+Biochemical protection against foliage light and heat damage also increases with irradiance.
 Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
@@
 N and P are essential for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco.
-$N_{area}$ increases in the direction of new leaf growth, where it optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018].
-Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species particularly advantageous [*REF*].
+Particularly, $N_{area}$ increases in the direction of new leaf growth, where it also optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018].
+Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species more advantageous [*REF*].
 
@@
 Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sunleaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
-However, $T_{opt}$ of $gs$ doesnot vary between sun and shade leaves [Slot et al.
-2019].
 The diurnal $gs$ pattern and resulting $A_{max}$ varies between species; for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
@@
 Increase in such canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].
-[**to include Teskey 2015; Breashers et al. 2021**]
 
@@
 Further, combination of warm-season associated with increased VPD, along with drought exacerbates tree stress[@parkwilliamsTemperaturePotentDriver2013] leading to mortality [@gouldenCaliforniaForestDieoff2019], with implications for forest structure.
+[**to include Teskey 2015; Breashers et al.
+2021**]
 

### 2021-06-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019).
-In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (@conditQuantifyingDeciduousnessTropical2000c, Meakem, *temperate REFS*).
 In this review, we focus on growing season conditions unless otherwise noted.
@@
 **Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites.
+Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites.
 Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs.
@@
 
-Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, a that is optimal for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, a optimal for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
 Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
@@
 However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--including increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, resulting in increased in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
-On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
+On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ and thermal capacitance can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
 Lastly, canopy structure and architecture adaptation can also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
@@
 Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-Comparing upper canopies, temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than the upper canopies of tropical moist and wet forest species [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however considering a broader range of forests, this might not be always the case (Fig.
+Upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig.
 `r fig_NEON_vertical`).
@@
 
-Exposed canopy environment induces traits in sun leaves that are better adaptated to heat and water stress than shade leaves through, for instance, being smaller, hairier, with greater vein density and LMA [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], as well as strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*].
-These traits [see Table 1] collectively dissapate heat and regulate $T_{leaf}$-photosynthetic balance.
+Exposed canopy environment induces traits in sun leaves that are better adaptated to heat and water stress than shade leaves where, for instance, being smaller, hairier, with greater vein density and LMA [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], as well as strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance.
 Whereas, shade leaves have traits that enable rapid light capture in the diffused-light environment with sunflecks, for example greater SLA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a].
@@
 This is also important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
-N and P are essential for photosynthetic and metabolic processes along with $VAZ$, as they contribute to enzyme formation and function, such as in Rubisco.
-Particularly, $N_{area}$ increases in the direction of new leaf growth, where it also optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018].
+Elemental traits, N and P though differ with soil conditions, $N_{area}$ typically increases in the direction of new leaf growth, where it also optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018].
 Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species more advantageous [*REF*].
@@
 Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter.
-Among tropical species individuals that can be deciduous, greater proportion are in larger size classes than understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000c].
+Among tropical species individuals that can be deciduous, the fraction of deciduous trees tends to increase across tree size classes, with understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000c; Meakem].
 In deciduous-temperate forests, spatio-temporal leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
@@
 
-Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sunleaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
-The diurnal $gs$ pattern and resulting $A_{max}$ varies between species; for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
+Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves that are buffered within-canopy [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
+Even so, $T_{opt}$ of $gs$ appears to be similar in sun and shade leaves [Slot et al.
+2019].
+This influences leaf thermal sensitivity [Perez and Feeley, 2018], but exact mechanisms remain unclear.
+Inter-specifically, diurnal $gs$ pattern and resulting $A_{max}$ varies where, for instance, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
 
@@
 
-Further, leaf thermal tolerances are more closely adapted to microenvironment than macroclimate [@feeley_thermal_2020; Slot et al.
-2020].
+Leaf thermal tolerances are more closely adapted to microenvironment than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
 Particularly, in an Australian arid region, *Acacia papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
@@
 shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008].
-Although, @bolstad_foliar_1999 found that temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased, [@bolstad_foliar_1999](*see their Fig.
+Although, in @bolstad_foliar_1999, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased, [@bolstad_foliar_1999](*see their Fig.
 1*).
 (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
-May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) and also found evidence of acclimation, but there remained a declining trend of respiration with topographical elevation.
-Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] also showed no variation within canopy positions in Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
+May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) with evidence of acclimation and a declining trend of respiration with topographical elevation.
+Similary, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] did not differ within canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
 
@@
 More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
-There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)*
+There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors.
+Tolerance of more than one biological and biophyical stressors can increase with tree height and light, partly due to non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
+However, growth also declines as trees age.
+This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts].
+Vertical biophysical conditions are bound to have different influenceS on size class and growth rates due to these existing demographic differences.
 
-Biophysical conditions can alter tree-water dynamics and growth patterns.
 Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
 Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013].
-Whereas, greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Altered precipitation regimes causing drought are more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
+Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Altered precipitation regimes cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
 However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
@@
 
-Tolerance to multiple biophysical and biological stressors can differ among tree size classes, having been shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to multiple stressors [@niinemetsResponsesForestTrees2010a].
-Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [@IncreasingRespiratoryCosts].
-These difference can alter tree growth patterns in response to biophysical conditions.
-Particularly, higher temperatures are shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014].
-Evidence in tropical forests also shows that higher nighttime temperatures are associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supress growth [*Clark 2010; 2013; Feeley et al.
-2007*].
-.
+Biophysical conditions can alter growth rates, but can do so differently between size classes through influencing existing demographic differences.
+Particulary evidence in in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014].
+Whereas in tropical forests higher nighttime temperatures was associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees [*Clark 2010; 2013; Feeley et al.
+2007*], perhaps due to increase in nighttime metabolic respiration that could supress growth.
 
@@
 
-The complex interactions of the biophysical environment and biological--leaf traits, metabolic processes and distribution of trees--are highly interconneced, resulting in feedback loops to another.
-Therefore, climate-driven changes have implications on forests at many levels.
+The complex interwoven relations between the biophysical environment and biological factors--leaf traits, metabolic processes and distribution of trees-- are likely to produce amplifications and feedback loops in a warming world, with implications for forests on many levels.
 
@@
 
-Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al.
+Canopies are crucial to microclimatic functioning, as they buffer several aspects of forest microenvironments [Zelleweger et al.
 2019].
 This buffering is influenced by the hydraulogical-cycle at large [Davis et al.
-2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al.
+2019], therefore severe droughts place canopies at risk, particularly as larger trees suffer more during droughts[Bennett et al.
 2015].
-Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Increase in such canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than temperate forests [Perez and Feeley, 2018].
+Further, increasing global temperatures show that canopies of temperate and tropical forests are already functioning close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019].
+Such increase in canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018].
 
-Canopy structure affects understory species composition and soil fertility which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, despite buffering dampening extreme effects, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
-This imposes an increasing threat to fauna in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
+Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
+Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy.
+Even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+This imposes an increasing threat to flora and fauna in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
 

### 2021-06-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`).**
- The distribution of this foliage is structured in correspondence with the distribution of tree sizes, and successional stages within a forest, where leaf area density ($LAD$) ie.
+ The distribution of this foliage is structured in correspondence with the distribution of tree sizes, and successional stages within a forest, where leaf area density (LAD) ie.
 leaf area per unit volume, increases with tree-height eventually forming the canopy layer.
@@
 LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests.
-Dense canopy forests composed of broadleaf trees can have high $LAD$ in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
+Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
 bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
@@
 May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) with evidence of acclimation and a declining trend of respiration with topographical elevation.
-Similary, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] did not differ within canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
+Similary, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] did not differ within canopy positions of Quercus rubra, whereas in considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
 
@@
 There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors.
-Tolerance of more than one biological and biophyical stressors can increase with tree height and light, partly due to non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
+Tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
 However, growth also declines as trees age.
 This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts].
-Vertical biophysical conditions are bound to have different influenceS on size class and growth rates due to these existing demographic differences.
+Vertical biophysical conditions are bound to influence growth and whole tree functioning.
 
@@
 Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Altered precipitation regimes cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
+Altered precipitation regimes can cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
 However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
@@
 Further, increasing global temperatures show that canopies of temperate and tropical forests are already functioning close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-Such increase in canopy $T_{leaf}$ reduces forest GPP [@pauTropicalForestTemperature2018] in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018].
+This can have implications in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018].
+Such increase in canopy $T_{leaf}$ has also shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018].
 
 Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy.
-Even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
-This imposes an increasing threat to flora and fauna in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
+Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+This imposes an increasing threat to the biodiversity in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
 

### 2021-06-22 vertical-thermal-review.Rmd
misc edits

@@
 #SI figure numbers
-SIfigs_NEON_all = "S1-S#"
+SIfig_NEON_all = "S1"
 
 #SI figure captions
+SIfig_example_legend <- "Figure S1. Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)"
 
@@
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients. 
-*(@niinemetsWithincanopyVariationRate2004a highlights how biophysical and photosynthetic gradients vary along the vertical canopy profile, however our current knowledge of this topic is lacking.)*
+Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsWithincanopyVariationRate2004a].
 Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
@@
 (1) the biophysical environment;
-(3) leaf temperature ($T_{leaf}$);
-(2) the leaf traits that most strongly influence leaf temperature and metabolism;
+(2) leaf temperature ($T_{leaf}$);
+(3) the leaf traits that most strongly influence leaf temperature and metabolism;
 (4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
@@
 In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S.
-National Ecological Observatory Network [NEON; `r fig_NEON_vertical`; Appendix `r appendix_NEON_methods`, Schimel et al.
+National Ecological Observatory Network [NEON; Figs.
+`r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, Schimel et al.
 2007].
@@
 Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Sites include a tropical broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).](NEON_height_profiles/figures/profile_all.png)
+Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
+Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig.
+S1.](NEON_height_profiles/figures/profile_all.png)
 
@@
 `r fig_NEON_vertical`).**
- The distribution of this foliage is structured in correspondence with the distribution of tree sizes, and successional stages within a forest, where leaf area density (LAD) ie.
-leaf area per unit volume, increases with tree-height eventually forming the canopy layer.
-
-LAD along the vertical gradient is heterogeneous, with common patterns that vary across forests.
-Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory such as in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] whereas in temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
-bottom heavy foliage trees and sometimes by dense understories [@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
-In open canopy forests, such as needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [@lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
-Soil moisture conditions, topography, gap formations followed by disturbance, all alter foliage patterns and further influence the formation of relatively closed vs.
+The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest. 
+Leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests. 
+Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., e.g., SCBI, SERC and HARV, Fig.
+`r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
+In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
+bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig.
+`r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
+In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig.
+`r fig_NEON_vertical`a, @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
+Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs.
 open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
-In addition, seasonally dry and wet conditions, deciduousness and phenology contribute to the spatial loss and gain of leaves (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019).
-In this review, we focus on growing season conditions unless otherwise noted.
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019).
+In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.
 
@@
 Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs.
-`r fig_NEON_vertical`, `r SIfigs_NEON_all`).
+`r fig_NEON_vertical`, `r SIfig_NEON_all`).
 The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).
@@
 **Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs.
-`r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
+`r fig_NEON_vertical`, `r SIfig_NEON_all`).**
 The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
@@
 Therefore, leaf traits are critical in maintaining desirable $T_{leaf}$ and shaping leaf metabolism (following section) across forest vertical strata (Figs.
-`r fig_NEON_vertical`, `r SIfigs_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
+`r fig_NEON_vertical`, `r SIfig_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
 
@@
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
-`r fig_NEON_vertical`, `r SIfigs_NEON_all`).
+`r fig_NEON_vertical`, `r SIfig_NEON_all`).
 At greater tree height, both biological variables (tree-height) and biophysical variables (light, temperature, and evaporative demand )(Fig.
@@
 
+SIfig_example_legend
+
 \newpage

### 2021-06-22 vertical-thermal-review_SI.Rmd
misc edits

@@
 #SI figure numbers
-SIfig_example = "S1"
+SIfig_NEON_all = "S1"
 
 #SI figure captions
-SIfig_example_legend <- "Figure S1. ...."
+SIfig_example_legend <- "Figure S1. Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)"
 
@@
 
-# Appendix `r appendix_SCBI_treering`. `r appendix_SCBI_treering_title`
+# `r SIfig_example_legend`
+![**Figure `r SIfig_NEON_all`.
+Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)**.
+Shown are  height profiles in July mean ± 1 standard deviation for maximum photosyntehtically active ratiation (PAR), maximum wind speed, minimum humidity, maximum $T_{air}$, and maximum biological temperature, $T_{bio}$.
+Sites are grouped into the following categories: (A) (sub)subtropical and warm temperate broadleaf: ..., (B) temperate open/ savanna forests: ..., (C) temperate mesic broadleaf forests: ..., (D) temperate conifer forests: ..., (E) northern and boreal forests: ...
+[Issue #35](https://github.com/EcoClimLab/vertical-thermal-review/issues/35).
+](NEON_height_profiles/figures/profile_all_groups.png)
 
-\newpage
 ## References

### 2021-06-22 vertical-thermal-review.Rmd
text edits

@@
 `r fig_NEON_vertical`).**
-The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest. 
-Leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests. 
-Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., e.g., SCBI, SERC and HARV, Fig.
+The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs.
+`r fig_NEON_vertical`a-b, `r SIfig_NEON_all`).
+Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
@@
 
-**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the shape of profile modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig.
-`r fig_NEON_vertical`, @koike_leaf_2001].**
-This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests, than open canopy forests such as needle-leaf forests [Fig.
-`r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Figs.
+`r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].**
+This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
+`r fig_NEON_vertical`d) than open canopy forests such as needle-leaf forests [e.g., OSBS, PUUM, and WREF, Fig.
+`r fig_NEON_vertical`d, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

### 2021-06-22 vertical-thermal-review.Rmd
text edits

@@
 This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`d) than open canopy forests such as needle-leaf forests [e.g., OSBS, PUUM, and WREF, Fig.
-`r fig_NEON_vertical`d, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`d), than under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
+`r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
@@
 **Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites.
-Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs.
+Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all 22 forested NEON sites (Figs.
+`r fig_NEON_vertical`e, `r SIfig_NEON_all`).
+Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds generally averaged $\le$ **~1** m s^-1^ (Figs.
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).
@@
 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Noteably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [@banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
+Notably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.
+`r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
 

### 2021-06-22 vertical-thermal-review.Rmd
switched order of Tair and CO2/RH
this is to match order in figure 2.
also, the last paragraph refers most to Tair, so this improves the continuity

@@
 
+**Atmospheric concentrations carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
+Atmospheric CO~2~ concentrations tend to be higher near ground level at night, but largely dissipate during the day [@koike_leaf_2001; @yang_spatial_1999].
+Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
+
+Relative humidity, $RH$, tends to be higher in the understory and decrease with height across the vertical profile, although this trend is absent in open forests [Figs.
+`r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregor_tree_nodate].
+Dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a].
+In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Additionally, increase in canopy height and LAI contributes to lowering maximum $T_{air}$ and VPD in the understory [@juckerCanopyStructureTopography2018].
+*( This last sentence needs work.)*
+
 **Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs.
@@
 `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
-
-**Relative humidity, $RH$, is generally highest in the understory and decreases with height across the vertical profile [Fig. `r fig_NEON_vertical`; e.g., @mcgregor_tree_nodate].**
-Dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a].
-In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Additionally, increase in canopy height and LAI contributes to lowering maximum $T_{air}$ and VPD in the understory [@juckerCanopyStructureTopography2018].
-*( This last sentence needs work.)*
-
-**Atmospheric carbon dioxide (CO~2~) concentrations tend to be higher near ground level at night, but largely dissipate during the day [@koike_leaf_2001; @yang_spatial_1999].** 
-Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
 

### 2021-06-22 vertical-thermal-review.Rmd
finished edits of "Review of vertical gradients" section

@@
 
-**The buffering patterns described above vary geographically, seasonally, and across forests types [REFS; *Defrenne et al. 2021*].**
-The strength of buffering in high elevation, in some cases, decreases when (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) moving to open-forests, and increases with canopy density [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018].
-Buffering capacity is also greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests.
-[@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
-Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018**double check these refs--at least one didn't belong**].
+**The buffering patterns described above vary geographically, seasonally, and across forests types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, *Defrenne et al. 2021*].**
+The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture [**REF**], and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019].
+(*Nidhi, please make sure the references are appropriate here-- I rearranged the sentence structure.*)
+Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018; **double check these refs--at least one didn't belong**].
 

### 2021-06-25 vertical-thermal-review.Rmd
start at major revamp of leaf T section
- add par on leaf energy budget
- add par on how biophysical principle shape Tleaf across vert gradient
-propose cutting a lot of content

@NidhiVinod , I have to take a break now, but hope to come back to this soon.

@@
 
-Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, a optimal for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the other biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
 Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
-Often, however, $T_{leaf}$ tracks $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020]. 
-Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost to the environment (such as transpiration) [Fig.
-`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].
+Often, however, $T_{leaf}$ tracks $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
 
@@
 
-*(In general, we'll want to adjust this section to parallel the figure and tie in with the section above.)*
+**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input through radiation versus heat lost to the environment [Fig.
+`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
+The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave), and high radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig.
+`r fig_leaf_T`a).
+The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required convert water from liquid to vapour state cools the leaf. 
+Evaporation ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
+Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed (Fig. `r fig_leaf_T`b) and decreases with leaf size (Fig. `r fig_leaf_T`d);
+decreases with $g_s$ (Fig. `r fig_leaf_T`e);
+and decreases with with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).
 
@@
 
+**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in forests.**
+High radiation at the top of the vertical profile (Fig.
+`r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig.
+`r fig_leaf_T`).
+However, higher windspeeds and lower RH (Fig.
+`r fig_NEON_vertical`), along with adaptive leaf traits (see next section), give them lower sensitivity to intercepted radiation (Fig.
+`r fig_leaf_T`a) and higher rates of heat loss (Fig.
+`r fig_leaf_T`b-e).
+*(I want to add a few sentences on how this varies with environmental conditions-- drought, high humidity....)*
+
+~~~ (*I think we should cut everything between here and "In-situ observations", after harvesting the most important content for the section above.*)
+
 High wind speed (Fig. `r fig_NEON_vertical`) and large leaf $g_a$ decrease $T_{leaf}$ - $T_{air}$ [Fig. `r fig_leaf_T`b, @daudetWindSpeedLeaf1999a]. 
@@
 Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$.
-Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissipation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
+In contrast, when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissipation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
 

### 2021-06-26 vertical-thermal-review.Rmd
T_leaf section revamp: some organization
-moved a paragraph up
-etc

@@
 Evaporation ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
-Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed (Fig. `r fig_leaf_T`b) and decreases with leaf size (Fig. `r fig_leaf_T`d);
+Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
 decreases with $g_s$ (Fig. `r fig_leaf_T`e);
@@
 [ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/Updated T_leaf_png.png)
+
+**Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
+When stomatal conductance ($g_s$) is high, evapotranspiration facilitates heat dissipation. 
+However, when high irradiance, $T_{air}$, $VPD$ or drought cause leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
 
@@
 
-~~~ (*I think we should cut everything between here and "In-situ observations", after harvesting the most important content for the section above.*)
+(**Vertical gradients in $T_{leaf}-T_{air}$ are expected to vary with midrometeorlogical conditions.**)
 
-High wind speed (Fig. `r fig_NEON_vertical`) and large leaf $g_a$ decrease $T_{leaf}$ - $T_{air}$ [Fig. `r fig_leaf_T`b, @daudetWindSpeedLeaf1999a]. 
+~~~ (*I think we should cut everything between here and  next "~~~", after harvesting the most important content for the section above.*)
+
 Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]
@@
 
-Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
-When stomatal conductance ($g_s$) is high, evapotranspiration facilitates heat dissipation. 
-However, when high irradiance, $T_{air}$, $VPD$ or drought cause leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
+~~~
 

### 2021-06-26 vertical-thermal-review.Rmd
text edits to paragraph I moved

@@
 **Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
-When stomatal conductance ($g_s$) is high, evapotranspiration facilitates heat dissipation. 
-However, when high irradiance, $T_{air}$, $VPD$ or drought cause leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
+When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
+However, when inadequate water supply -- for example under drought -- causes leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads, but during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
 

### 2021-06-26 vertical-thermal-review.Rmd
worked on leafT section revamp

@@
 decreases with $g_s$ (Fig. `r fig_leaf_T`e);
-and decreases with with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).
+and decreases with with vapor pressure deficit, or increases with RH [Fig. `r fig_leaf_T`c].
 
@@
 When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
-However, when inadequate water supply -- for example under drought -- causes leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
+However, when inadequate water supply -- for example under drought -- causes leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
+`r fig_leaf_T`e, @fauset_differences_2018].
 Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads, but during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
+With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
 
@@
 `r fig_leaf_T`b-e).
-*(I want to add a few sentences on how this varies with environmental conditions-- drought, high humidity....)*
 
-(**Vertical gradients in $T_{leaf}-T_{air}$ are expected to vary with midrometeorlogical conditions.**)
+**Vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
+Under hot conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
+`r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020].
+Thus, understory leaves, which typically experience lower windspeeds (Fig. `r fig_NEON_vertical`), would retain more heat.
+The higher-RH conditions of the understory (Fig.
+`r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig.
+`r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
+Thus, understory leaves would have less ability to shed excess heat under hot and humid conditions.
+In contrast, under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
 
-~~~ (*I think we should cut everything between here and  next "~~~", after harvesting the most important content for the section above.*)
+~~~ **(KAT stopped here)**
 
-Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]
-
-Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$.
-In contrast, when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissipation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
-
-**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to greater $g_a$ than larger leaves that take longer due to lower $ga$.
-Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves.
-With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
-
-~~~
-
-**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
  Across species, inherent leaf $gs$ is generally greater in sun leaves than shade leaves, due to stomata density [Table 1].

### 2021-06-27 NEON_height_profiles/scripts/verticalProfileFigure.R
Updated script and figure for #35

@@ -12,6 +12,7 @@ library(data.table)
 library(readxl)
 library(ggplot2)
 library(ggpubr)
+library(patchwork)
 
 #
 #1. Download and save neon data ####
@@ -259,6 +260,7 @@ groupPlots <- lapply(1:length(sites), function(X){
   clrsFocus <- clrs[[X]]
   
   plots <- list()
+  allNEON <- NULL
   for(i in 1:length(dp[,name])){
     alldata <- NULL
     allTopHeight <- c()
@@ -312,7 +314,8 @@ groupPlots <- lapply(1:length(sites), function(X){
       
       #only keep July values (from edit in Jan 2021)
       whei <- whei[month_char=="July", 
-                   ][, `:=` (var = dp[,name][i], site=sitesFocus[j], col=clrsFocus[j])]
+                   ][, `:=` (var = dp[,name][i], site=sitesFocus[j], 
+                             col=clrsFocus[j], biome=names(sites[X]))]
       
       alldata <- rbind(alldata, whei)
       allTopHeight <- c(allTopHeight, topHeight)
@@ -332,6 +335,7 @@ groupPlots <- lapply(1:length(sites), function(X){
       setnames(alldata, old=c("max_mean", "max_sd"), 
                new=c("all_mean", "all_sd"))
     }
+    allNEON <- rbind(allNEON, alldata)
     
     plots[[i]] <- local({
       graph <- ggplot(alldata) +
@@ -348,24 +352,23 @@ groupPlots <- lapply(1:length(sites), function(X){
       }
       
       graph <- graph +
-        geom_point(aes(x = all_mean, y = plotHeight, color = site), 
+        geom_point(aes(x = all_mean, y = plotHeight, color = site, size=1.1), 
                    shape=19) +
-        geom_path(aes(x = all_mean, y = plotHeight, color = site)) +
+        geom_path(aes(x = all_mean, y = plotHeight, color = site), size=1.1) +
         ggplot2::geom_errorbarh(aes(xmin = all_mean - all_sd, 
                                     xmax = all_mean + all_sd, 
-                                    y=plotHeight, color = site, height=0.1)) +
+                                    y=plotHeight, color = site, height=0.1,
+                                    size=1)) +
         labs(x = "",
              y = "") +
-        theme_bw()
+        theme_bw() +
[…]

### 2021-06-28 NEON_height_profiles/scripts/verticalProfileFigure.R
Updated figure and script as per #35

@@ -362,7 +362,7 @@ groupPlots <- lapply(1:length(sites), function(X){
         labs(x = "",
              y = "") +
         theme_bw() +
-        theme(text = element_text(size=14))
+        theme(text = element_text(size=16))
       
       if(X==5){#only labels on bottom row for putting all 25 plots together
         if(i %in% c(1,2)){ #windspeed, RH
@@ -431,16 +431,21 @@ groupPlots <- lapply(1:length(sites), function(X){
 })
 
 
-#get canopy minus bottom values across aggregate groupings
+#get numbers for manuscript
 aggNeon <- rbindlist(lapply(groupPlots, `[[`, 2))
 
-ttt <- aggNeon[, .SD[c(1,.N)], by=.(site,var)][order(site,var,plotHeight)]
-tr <- ttt[, diff := all_mean - shift(all_mean, fill = first(all_mean)), 
+##max mean ground level value across sites
+ground <- aggNeon[, .SD[1], by=.(site,var)][order(var),]
+ground[ground[, .I[all_mean == max(all_mean)], by=var]$V1
+       ][,.(site,var,plotHeight,all_mean,biome)]
+
+##max difference in windspeed btwn top and bottom 
+aggDiff <- aggNeon[, .SD[c(1,.N)], by=.(site,var)][order(site,var,plotHeight)]
+tr <- aggDiff[, diff := all_mean - shift(all_mean, fill = first(all_mean)), 
     by = .(site,var)
     ][diff != 0,]
+tr[var=="windSpeedMean", .(site, plotHeight, diff)][order(-diff),]
 
-tr[,mean(all_mean), by=var] #mean canopy minus bottom for all sites
-tr[,mean(all_mean), by=.(biome, var)][order(var)] #same as above but per biome
 
 #continuing with the plotting here
 groupPlots <- lapply(groupPlots, `[[`, 1)

### 2021-06-28 vertical-thermal-review.Rmd
fill in numbers from @mcgregorian1
Thanks, @mcgregorian1 !

@@
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
-Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds generally averaged $\le$ **~1** m s^-1^ (Figs.
+Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs.
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).

### 2021-06-28 vertical-thermal-review.Rmd
working on Tleaf section

@@
 decreases with $g_s$ (Fig. `r fig_leaf_T`e);
-and decreases with with vapor pressure deficit, or increases with RH [Fig. `r fig_leaf_T`c].
+and decreases with with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).
 
@@
 
-**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in forests.**
+**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
 High radiation at the top of the vertical profile (Fig.
@@
 However, higher windspeeds and lower RH (Fig.
-`r fig_NEON_vertical`), along with adaptive leaf traits (see next section), give them lower sensitivity to intercepted radiation (Fig.
+`r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (see next section), give them lower sensitivity to intercepted radiation (Fig.
 `r fig_leaf_T`a) and higher rates of heat loss (Fig.
@@
 
-**Vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
+**Within closed-canopy forets, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
 Under hot conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
@@
 `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
-Thus, understory leaves would have less ability to shed excess heat under hot and humid conditions.
+Thus, understory leaves can remain cooler under lower radiation load, but their environment is not conducive to shedding excess heat, particularly under humid conditions, and their tendency towards lower leaf $g_s$ and thermal capacitance can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
 In contrast, under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
-
-~~~ **(KAT stopped here)**
-
- Across species, inherent leaf $gs$ is generally greater in sun leaves than shade leaves, due to stomata density [Table 1].
-However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--including increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, resulting in increased in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018).
-On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ and thermal capacitance can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
-Lastly, canopy structure and architecture adaptation can also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
+Because sun leaves are exposed to higher irradiance and $VPD$, their $g_s$-limitation is steeper that that of shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig.
+`r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
 
 ### In-situ observations
-**The principles outlined above, vertical microclimate profile (Fig.
-`r fig_NEON_vertical`) and leaf traits (see following section) together shape in-situ $T_{leaf}$ observations (Fig.
-`r fig_leaf_T`).**
+**Following the principles outlined above, vertical microclimate profile and leaf traits (see following section) together shape in-situ $T_{leaf}$ observations (Figs.
+`r fig_NEON_vertical`-`r fig_leaf_T`).**
+
 In open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].

### 2021-06-28 vertical-thermal-review.Rmd
switched paragraph order under "in-situ observations"
putting closed canopy paragraph first

@@
 `r fig_NEON_vertical`-`r fig_leaf_T`).**
-
-In open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
-Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
-Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
-Under other conditions $T_{leaf}$ tracks $T_{air}$, due to high evapotranspiration and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
-
-**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** 
+Canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.
 Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
@@
 Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
+
+In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
+Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
+Under other conditions $T_{leaf}$ tracks $T_{air}$, due to high evapotranspiration and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
 

### 2021-06-28 vertical-thermal-review.Rmd
working on in-situ observations section
@NidhiVinod , I'm confused by this section and hope I'm not making a mess of it. I'll open an issue on this.

@@
 ### In-situ observations
-**Following the principles outlined above, vertical microclimate profile and leaf traits (see following section) together shape in-situ $T_{leaf}$ observations (Figs.
+**Following the principles outlined above, vertical microclimate profile and leaf traits (see following section) together explain in-situ $T_{leaf}$ observations (Figs.
 `r fig_NEON_vertical`-`r fig_leaf_T`).**
-Canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.
-Closed-canopy forests, including tropical and temperate broadleaf forests, experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optimum for photosynthesis ($T_{opt}$) and $T_{air}$, from $gs$ limitation and irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-Upper canopies of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig.
-`r fig_NEON_vertical`).
-At a specie-specific level, leaf boundary resistance is lower for conifers than broad-leaf species, resulting in higher mean $T_{leaf}$ in broad-leaf trees than conifer canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
-Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
-
-In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is greater than upper canopy, due to vertical air mixing, greater light permeability, still air in the lower canopy and heating from the ground [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is commonly greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
+`r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
 Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
 Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
-Under other conditions $T_{leaf}$ tracks $T_{air}$, due to high evapotranspiration and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
+Under other conditions $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s#* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig.
+`r fig_NEON_vertical`).
+
+The extent of this buffering varies across forest types.
+*Upper canopies of temperate-deciduous species can have greater $T_{leaf} - T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig.
+`r fig_NEON_vertical`).*
+Because conifer leaves have lower $g_a$ than broad leaves, conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018]. 
+Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
 

### 2021-06-28 vertical-thermal-review.Rmd
small changes, knit

@@
 Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
-Under other conditions $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s#* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig.
+Under other conditions $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig.
 `r fig_NEON_vertical`).
@@
 Because conifer leaves have lower $g_a$ than broad leaves, conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
+
 Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].

### 2021-06-28 vertical-thermal-review.Rmd
started editing leaf traits section

@@
 However, higher windspeeds and lower RH (Fig.
-`r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (see next section), give them lower sensitivity to intercepted radiation (Fig.
+`r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig.
 `r fig_leaf_T`a) and higher rates of heat loss (Fig.
@@
 ### In-situ observations
-**Following the principles outlined above, vertical microclimate profile and leaf traits (see following section) together explain in-situ $T_{leaf}$ observations (Figs.
+**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations (Figs.
 `r fig_NEON_vertical`-`r fig_leaf_T`).**
@@
 
-**Leaf thermoregulation is controlled by leaf traits, that are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016].
-Therefore, leaf traits are critical in maintaining desirable $T_{leaf}$ and shaping leaf metabolism (following section) across forest vertical strata (Figs.
-`r fig_NEON_vertical`, `r SIfig_NEON_all`)[@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
-
-Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004] where degrees of direct to diffused light and resultant water demands shape sun vs.
-shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+**Leaf thermoregulation is controlled by leaf traits, which are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016] in order to maintain desirable $T_{leaf}$ for leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
+Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], where the ratio of direct to diffused light and resultant water demands cause leaves to develop characteristics of sun vs shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).
-At greater tree height, both biological variables (tree-height) and biophysical variables (light, temperature, and evaporative demand )(Fig.
-`r fig_NEON_vertical`) also play a role in shaping leaf traits.
 
-In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained at the tops of(~100 m.
+~~~ *(this paragraph breaks the flow, and I'd love to be able to drop it or perhaps dramatically shorten. It is important and interesting content, though)*
+Both biophysical variables (light, temperature, and evaporative demand; Fig. `r fig_NEON_vertical`) and biological variables (tree height) play a role in shaping leaf traits. 
+In particular, hydraulic constraints increase with height on a tree [@mcdowell_relationship_2011; @cobleLightAcclimationOptimizes2015; @couvreur_water_2018a] such that gas exchange is strongly constrained at the tops of very tall trees (e.g., ~100 m.
 tall *Sequoia sempervirens*; *Ambrose et al.
@@
 Morphologically, leaf size [@jensenPhysicalLimitsLeaf2013b] and lobation [@kusiPlasticLeafMorphology2020] have been shown to decrease, and leaf mass per area (at greater than 20m) to increase, with height  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017; @marshallFoliageHeightInfluences2003].
-
 Disentangling tree-height related biological and biophysical drivers (Fig.
@@
 We note that patterns across light and height gradients are generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
+~~~
 
-### *(clarify intra- vs inter-specific variation - 1 short paragraph)*
-Leaf trait variation within the same species, here, is referred to as within-canopy intraspecific variation, and comparison of traits across species is referred to as interspecific variation.
-Therefore, across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
+Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
@@
 
-Within-canopy leaf traits vary along the canopy light gradient, ie.
-in sun and shade leaves [@casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-This variation is crucial, to understand vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+Within-canopy leaf traits vary along the canopy light gradient, *i.e.,* in sun and shade leaves [@casasSunShadeLeaves2011], at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+This variation shapes vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity.

### 2021-06-28 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -20,9 +20,9 @@ library(RColorBrewer)
 ####################################
 #dataframe for biophysical parameters from NEON HARV data
 #normal scenario represents HARV NEON environmental parameters as observed. 
-#drought scenario represents similar HARV NEON parameters, with max PAR for overstory, and 50% increase in normal PAR for understory, and 50% decrease in normal relative humidity for both positions.
+#drought scenario represents similar HARV NEON parameters, with: 1) max PAR for overstory and 50% increase in normal PAR for understory, and 2) 50% decrease in normal relative humidity for both positions.
 #typical normal stomtal conductane(gs) values are referred from Tleaves supplementary manual for range of minimum and maximum values and obtained from Cavendar-Bares and Bazzaz, 2000.
-#drought condition ~gs values obtained from Cavendar-Bares and Bazzaz, 2000.  
+#drought condition ~gs values are set to a constant value of 0.01 umol/m^2/s/Pa for canopy and understory. 
 #respective leaf sizes (m) are measured from sun and shade red oak leaves 
 #all variables are constant except for the y-axis variables that represent minimum- maximum range
 ####################################
@@ -93,10 +93,10 @@ sw_u$u_tleaf_tair<- sw_u$tleaf - sw_u$tair
 # 1. Shortwave radiation, drought
 dr_sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
                   rh = 0.49, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
-                  gs = 0.1)
+                  gs = 0.01)
 dr_sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
                   rh = 0.46, wind = 2.88, tair = 298.1, pressure = 99, leaf.size = 0.04, 
-                  gs = 1.0)
+                  gs = 0.01)
[…]

### 2021-06-28 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -465,3 +465,4 @@ figure1<-ggarrange(a1, b1, e1, d1, c1, table1, ncol=3, nrow =2, align = c('hv'),
                    labels = c("(a)", "(b)", "(c)", "(d)", "(e)"))
 figure1
 
+#plot saved as .png

### 2021-06-28 vertical-thermal-review_SI.Rmd
updated SI methods for Tleaves

@@
 Energy balance for a normal overstory sun leaf and understory shade leaf were parameterized with biophysical variables from Harvard NEON data documented here: (https://github.com/EcoClimLab/vertical-thermal-review/blob/master/NEON_height_profiles/HARV_neon_stats.csv).
-Leaf trait measurements for effective leaf width was measured on ~Quercus rubra~ sun and shade leaf, stomatal conductance measurements were refered from Tleaves typical sun and shade measurements [@muir_tealeaves_2019] and as well @cavender-baresChangesDroughtResponse2000.
+Leaf trait measurements for leaf width, ie.
+measured as leaf characteristic dimmension, was measured on ~Quercus rubra~ sun and shade leaf, stomatal conductance measurements were referred from Tleaves typical sun and shade measurements [@muir_tealeaves_2019] and as well @cavender-baresChangesDroughtResponse2000.
 
@@
 Overstory drought PAR values reflect maximum observed Harvard NEON PAR and understory drought reflect 50% increased PAR of understory normal value.
-Leaf trait measurements for effective leaf width is the same as normal scenario, stomatal conductance was referred from @cavender-baresChangesDroughtResponse2000.
+Leaf trait measurements for leaf width is the same as normal scenario, stomatal conductance is kept constant at a minimum value of 0.01 umol/m^2/s/Pa for both positions.
 In each visual, all variables are constant (Biophysical Constants table) except for the independent variable that represents minimum - maximum range.

### 2021-06-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Within-canopy leaf traits vary along the canopy light gradient, *i.e.,* in sun and shade leaves [@casasSunShadeLeaves2011], at a leaf anatomical, structural and biochemical level [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
 This variation shapes vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. 
-Table `r table_leaf_traits`, summarizes relevant traits for thermal sensitivity.
-Further, research to characterize trait relationships and responses through the vertical canopy is needed, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
+Further research to characterize trait relationships and responses through the vertical canopy is needed, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
 

### 2021-06-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`-`r fig_leaf_T`).**
+Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig.
+`r fig_NEON_vertical`).
 In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
@@
 Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
-Under other conditions $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig.
-`r fig_NEON_vertical`).
 

### 2021-06-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al.
+Globally, temperatures have risen # $1.2^\circ$C since the 1850-1900 period [Rohde & Hausfater, 2020] and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al.
 2014*].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities], leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
+Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
 
@@
 Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
-Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
+Overstory trees form canopies that play a crucial role in moderating macro and microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
@@
 `r fig_NEON_vertical`).**
-The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs.
+Foliage distribution is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs.
 `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`).
@@
 **Atmospheric concentrations carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
-Atmospheric CO~2~ concentrations tend to be higher near ground level at night, but largely dissipate during the day [@koike_leaf_2001; @yang_spatial_1999].
-Given that differences are small during the day when photosynthesis is occurring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.
+Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated $CO_{2}$, but the elevated $CO_{2}$ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].Elevated $CO_{2}$ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are neglible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
 
@@
 Dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a].
-In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Additionally, increase in canopy height and LAI contributes to lowering maximum $T_{air}$ and VPD in the understory [@juckerCanopyStructureTopography2018].
+In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Maximum $T_{air} and VPD in the understory are further decreased with increasing canopy height and LAI [@juckerCanopyStructureTopography2018].
 *( This last sentence needs work.)*
@@
 
-**Within closed-canopy forets, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
+**Within closed-canopy forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
 Under hot conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
@@
 
-### In-situ observations
+### In situ observations
 **Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations (Figs.
@@
 `r fig_NEON_vertical`).
-In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is commonly greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
+In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high radiation [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ may be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
 `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
@@
 `r fig_NEON_vertical`).*
-Because conifer leaves have lower $g_a$ than broad leaves, conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
+Because needle-leaves have lower $g_a$ than broad leaves, conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
 
@@
 **Leaf thermoregulation is controlled by leaf traits, which are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016] in order to maintain desirable $T_{leaf}$ for leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
-Leaf morphology adapts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], where the ratio of direct to diffused light and resultant water demands cause leaves to develop characteristics of sun vs shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands cause leaves to develop characteristics of sun vs shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
@@
 This variation shapes vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. 
-Further research to characterize trait relationships and responses through the vertical canopy is needed, because LES theory was developed using mainly sun exposed (upper canopy) leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
+Further research to characterize trait relationships and responses through the vertical canopy is needed, because much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al.
+2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
 
-Intraspecific traits also vary along the ontogeny of the same species in canopy and understory, for example, LMA, photosynthetic capacity [per area] and $gs$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+Independent of light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
+For example, LMA, photosynthetic capacity [per area] and $gs$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
 
@@
 
-Exposed canopy environment induces traits in sun leaves that are better adaptated to heat and water stress than shade leaves where, for instance, being smaller, hairier, with greater vein density and LMA [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], as well as strategies such as leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves where, for instance, being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], and steeper leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance.
-Whereas, shade leaves have traits that enable rapid light capture in the diffused-light environment with sunflecks, for example greater SLA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a].
-In open canopies where light is comparitevely homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, than solely by light [@mediavillaFoliarPlasticityRelated2019a].
+Whereas, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks, for example lower LMA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a].
+In open canopies where light is comparitively homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-Biochemical protection against foliage light and heat damage also increases with irradiance.
-Caratenoids and xanthophyll pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--an antioxidant scavenging function--convert excess light into heat [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-This is also important for leaf acclimation to higher temperature [@kumarathungeAcclimationAdaptationComponents2019a].
-Elemental traits, N and P though differ with soil conditions, $N_{area}$ typically increases in the direction of new leaf growth, where it also optimizes to prevent photoinhibition in leaves vertically [@kitaoCanopyNitrogenDistribution2018].
-Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed canopies, making these species more advantageous [*REF*].
+Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory. 
+Caratenoids and xanthophyll cycle pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway  [Havaux, M.
+and Tardy, F., 1996].
+Elemental traits, such as N and P differ with soil conditions, but $N_{area}$ typically increases in the direction of new leaf growth, where its allocation is optimized to prevent photoinhibition in leaves [@kitaoCanopyNitrogenDistribution2018].
+Additionally, isoprenes facilitate higher heat tolerance in exposed canopies in species capable of emitting isoprenes, providing these species with a selective advantage in a warming world [*REF*].
 

### 2021-06-28 vertical-thermal-review.Rmd
updating references

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen # $1.2^\circ$C since the 1850-1900 period [Rohde & Hausfater, 2020] and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al.
+Globally, temperatures have risen # $1.2^\circ$C since the 1850-1900 period [Rohde & Hausfater, 2020] and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@rohdeBerkeleyEarthLand2020] and hotter droughts [*Trenberth et al.
 2014*].

### 2021-06-28 vertical-thermal-review.Rmd
incorporated  Martijn's edits and suggestions
still working on this

@@
 **Atmospheric concentrations carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
-Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated $CO_{2}$, but the elevated $CO_{2}$ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].Elevated $CO_{2}$ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are neglible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
+Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated $CO_{2}$ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
+Elevated $CO_{2}$ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are neglible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
 
@@
 
-**Thermal tolerance (e.g., $T_{50}$) seldom varies with exposure among recent studies, but there are no studies isolating effect of height.**
+**Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
 
 In 1994, Knapp et al.
-showed that in tall grass prarie gallery forest in Kansas, $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions.
+showed that $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions.
 In both species $T_{50}$ also increased from June to August.
-However, more recently @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three species, measured near ground level, in Panama tropical forest, but results were not significant.
+Within species, @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.
 
@@
 $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves with greater LMA (greater construction costs), compared to species with low LMA [@slotLeafHeatTolerance2020].
-@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that showed higher photosynthetic heat tolerance, experienced higher $T_{leaf_{max}}$, but showed low optimum temperature for C assimilation [$T_{opt}$] and therefore had a narrower breadth of thermal safety margins ($T_{leaf_{max}}$-$T_{50}$).
+@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that experienced higher maximum temperatures had higher photosynthetic heat tolerance.
 
@@
 shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008].
-Although, in @bolstad_foliar_1999, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased, [@bolstad_foliar_1999](*see their Fig.
-1*).
 (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
-May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*) with evidence of acclimation and a declining trend of respiration with topographical elevation.
-Similary, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] did not differ within canopy positions of Quercus rubra, whereas in considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].
+May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*).
+@xuSeasonalVariationTemperature2006 found that the canopy-position effect  gradient in respiration rates was much greater for Quercus rubra trees in mesic than in xeric sites, but the temperature sensitivity of respiration did not differ between canopy positions.
 

### 2021-06-29 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
Update leaf energy balance fig_normal_and_drought.R

@@ -433,7 +433,7 @@ e1<-ggplot(rh)+
   geom_smooth(aes(x = rh, y = dr_ltla),  method = lm, color = "#7b3294", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = rh, y = dr_utla),  method = lm, color = "#008837", se = FALSE, linetype = "dashed")+
   geom_hline(yintercept=0, linetype='dotted', col = 'black')+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+ xlab("relative humidity (rh, %)")+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+ xlab("relative humidity (rh)")+
   theme_few()+theme(text = element_text(size = 14))+ ylim (-5, 14) 
 e1
 

### 2021-06-30 vertical-thermal-review.Rmd
knit in new leaf energy balance figure, deleted older copies
#6
Update vertical-thermal-review.Rmd

@@
 In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Maximum $T_{air} and VPD in the understory are further decreased with increasing canopy height and LAI [@juckerCanopyStructureTopography2018].
+Maximum $T_{air}$ and VPD in the understory are further decreased with increasing canopy height and LAI [@juckerCanopyStructureTopography2018].
 *( This last sentence needs work.)*
@@
 Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
-[ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/Updated T_leaf_png.png)
+[ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/leaf_energy_fig3.png)
 

### 2021-06-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**Leaf thermoregulation is controlled by leaf traits, which are in-turn adapted to their environment [@michaletzEnergeticCarbonEconomic2016] in order to maintain desirable $T_{leaf}$ for leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
-Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands cause leaves to develop characteristics of sun vs shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optamize leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
+Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
@@
 
-Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves where, for instance, being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], and steeper leaf-angle and rolling [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves where, for instance, being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], and steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance.
@@
 
-New-leaf growth in response to the light gradient constantly shades leaves underneath.
-Therefore, exposed-canopies have greater proportion of young leaves than shaded regions that have older leaves.
-Acclimation to shade from sun-exposure, and vice-versa, is more possible in younger than mature leaves, due to developing traits that can acclimate, whereas, change in longterm-light for mature leaves can result in leaf abscission [Niinemetes, 2007; *REFS*].
-*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves.
-This is bound to affect metabolism, and may affect other aspects of their performance.*
+Leaf age affects vulnerability to heat stress independent of light environment [e.g.
+Marias et al.
+2017], and has a strong effect on metabolism (Kikuzawa & Lechowitz, Kitajima et al.).
+Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, they therefore operate closer to their photosynthetic optimum .
+However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy,  damage does not really accumulate [Miller et al.
+2021].
+The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is XXX  .
 
@@
 
-Deciduousness as a trait in tropical forests is an adaptation to conserve water under stress.
-Indeed, greater proportion of deciduous species are found in drier tropical regions than wetter.
-Among tropical species individuals that can be deciduous, the fraction of deciduous trees tends to increase across tree size classes, with understory species that are seldom deciduous [@conditQuantifyingDeciduousnessTropical2000c; Meakem].
-In deciduous-temperate forests, spatio-temporal leaf-emergence is delayed with height-ontogeny with understory species leafing-out earlier than canopy trees [Vitasse et al.
-2013; @seiwaChangesLeafPhenology1999].Therefore, deciduous leaf-loss or -gain can alter biophysical and $T_{leaf}$ gradient [@rey-sanchez_spatial_2016].
+Seasonal leaf-loss or -gain can alter biophysical and $T_{leaf}$ gradient in the forest [@rey-sanchez_spatial_2016].
+Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests.
+Within tropical species, the fraction of deciduous trees tends to increase across tree size classes, with understory species typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c; Meakem].
+In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing-out earlier than canopy trees [Vitasse et al.
+2013; @seiwaChangesLeafPhenology1999].
 
@@
 Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
-However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves and vertical gradient (Table `r table_leaf_metabolism`).
+However, as we detail below, very few studies have valued temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
 

### 2021-06-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2021].
-The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is XXX  .
+The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is XXX.
 
@@
 ![Table `r table_leaf_metabolism`. (*For latest version of table, see tables.pdf.*)](table2.png)
+
 ```{r eval=FALSE, echo=FALSE}
@@
 
-Exposed sun-leaves, though equipped with thermoregulatory traits, are more constrained by $gs$ than shade leaves that are buffered within-canopy [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] because although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
-Even so, $T_{opt}$ of $gs$ appears to be similar in sun and shade leaves [Slot et al.
-2019].
-This influences leaf thermal sensitivity [Perez and Feeley, 2018], but exact mechanisms remain unclear.
-Inter-specifically, diurnal $gs$ pattern and resulting $A_{max}$ varies where, for instance, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
+Maximum $gs$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves.
+However, because of the high evaporative demand and irradiance experienced by sun leaves,  mid-day stomatal depression is prevalent in sun leaves but not in shade leaves.
+$T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
+Higher maximum temperatures influences leaf thermal sensitivity tolerance [Perez and Feeley, 2018], but exact mechanisms remain unclear.
+$T_{opt}$ of $gs$  did not differ significantly between sun and shade leaves in three tropical tree species [Slot et al.
+2019], but sun leaves show a stronger decrease in gs to rising temperature than shade leaves (Hernandez et al.
+2020; Carter et al.
+2021 ).
 
@@
 
-Interestingly, within-canopy Japanese oak sun and shade leaves were found to maintain similar rates of excess energy production under respective peak light conditions.
-Excessive light energy not utilized can cause photoinhibition.
-Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain higher photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018].
-Perhaps, differences in these thermal stablization strategies realize similar $T_{opt}$ for photosynthesis in sun and shade leaves.
+@hernandezSimilarTemperatureDependence2020 speculated that the lack of sun-shade differences in Topt may reflect acclimation to similar temperatures of peak photosynthesis.
+Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of mid-day stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
+Shade leaves that do not experience mid-day depression continue to fix carbon even when afternoon air temperatures peak.
+Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.
 
@@
 showed that $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions.
-In both species $T_{50}$ also increased from June to August.
+In both species $T_{50}$ also tracked increased from June to August.
 Within species, @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.
@@
 Whereas in tropical forests higher nighttime temperatures was associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees [*Clark 2010; 2013; Feeley et al.
-2007*], perhaps due to increase in nighttime metabolic respiration that could supress growth.
+2007*], perhaps due to increase in nighttime respiratory carbon loss that reduced carbon availability for growth.
 
-**Recently emerging evidence indicates that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
+**Recently emerging evidence indicates that understory trees tend to exhibit greater growth declines with increasing $T_{air}$ under mesic conditions.** 
 In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].

### 2021-06-30 vertical-thermal-review.Rmd
attempted to organize Tleaf section
I added a seperate para keeping the one that's already there, to see if this flows better?

@@
 Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
-Overstory trees form canopies that play a crucial role in moderating macro and microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory.
+Overstory trees form canopies that play a crucial role in moderating forest macroclimates [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], as well as microclimates by buffering understories from extreme temperatures and other macroclimatic conditions [@zellweger_seasonal_2019].
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
@@
 Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**.
-Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+Shown are  height profiles in growing season (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
 Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
@@
 (*Nidhi, please make sure the references are appropriate here-- I rearranged the sentence structure.*)
-Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018; **double check these refs--at least one didn't belong**].
+Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; **double check these refs--at least one didn't belong**].
 
@@
 [ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/leaf_energy_fig3.png)
+
+**attempted to oraganize**
+
+~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig.
+`r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $ga$ ie.
+boundary layer conductance) and latent heat loss ( via $gs$ ie.
+Stomatal conductance and $E$ ie evaporation).
+
+$Ga$ and $Gs$ are dominant factors in $T_{leaf}$ flux.
+$Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig.
+`r fig_leaf_T`b).
+Therefore, high windspeeds in upper canopies such as in NEON sites with windspeeds ~4 ms (Fig.
+`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig.
+`r fig_leaf_T`b)[Bonan 2008].
+However, low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low.
+
+Latent heat loss through $gs$, cools the leaf and decreases $T_{leaf}$-$T_{air}$.
+However in conditions of low soil moisture during drought (Fig.
+`r fig_leaf_T`), greater temperature than $gs$ thermal optimum, and air dryness causing an increase in VPD, $gs$ becomes limited, thereby increasing $T_{leaf}$-$T_{air}$.
+Limitation to $gs$ can also exaggerate effects of biophysical factors that contribute to energy inputs such as, solar radiation, humidity and $T_{air}$ (Fig.
+`r fig_leaf_T`).
+
+Leaf size and physiology influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$.
+Smaller leaves have thinner boundary layer that enables coupling with $T_{air}$ through higher $ga$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig.
+`r fig_leaf_T`a).
+However, greater $gs$ in large leaves under low raditation, can be cooler than small leaves.
+Sun leaves, such as in upper canopies, tend to have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves.
+But exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions.
+However, because shade leaves have higher thermal capacitance and comparitely lower $gs$, $T_{leaf}$-$T_{air}$ can increase with slight increase in radiation (Fig.
+`r fig_leaf_T`a).~
 

### 2021-07-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $ga$ ie.
-boundary layer conductance) and latent heat loss ( via $gs$ ie.
+boundary layer conductance) and latent heat loss (via $gs$ ie.
 Stomatal conductance and $E$ ie evaporation).
@@
 
-Latent heat loss through $gs$, cools the leaf and decreases $T_{leaf}$-$T_{air}$.
+Latent heat loss through maximum $gs$, cools the leaf and decreases $T_{leaf}$-$T_{air}$, allowing coupling between leaf and air.
 However in conditions of low soil moisture during drought (Fig.
@@
 But exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions.
-However, because shade leaves have higher thermal capacitance and comparitely lower $gs$, $T_{leaf}$-$T_{air}$ can increase with slight increase in radiation (Fig.
-`r fig_leaf_T`a).~
+However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in heat flux, $T_{leaf}$-$T_{air}$ can become more elevated than sun leaves (Fig.
+`r fig_leaf_T`a).
+
+**Within forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
+$T_{leaf}$ generally tracks $T_{air}$, and in turn the vertical gradient of $T_{air}$ depends on canopy structure and $LAI$.
+Closed canopies, and high $LAI$ act as a sheild, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
+Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
+On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
+It is also important to note that NEON sites (in Fig.
+`r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$, measured using infrared sensor, reflects wide swath of temperatures.
+But recent advances in $T_{leaf}$ research that we highlight in the Modelling section will enhance our understanding of large scale $T_leaf$ vertical patterns.~
 

### 2021-07-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 However, low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low.
+Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig.
+`r fig_leaf_T`c) because reduced $ga$ stifles heat dissapation through $E$ and $gs$.
 
@@
 On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
-It is also important to note that NEON sites (in Fig.
-`r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$, measured using infrared sensor, reflects wide swath of temperatures.
-But recent advances in $T_{leaf}$ research that we highlight in the Modelling section will enhance our understanding of large scale $T_leaf$ vertical patterns.~
+It is also important to also note though NEON sites (in Fig.
+`r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures, but recent advances in $T_{leaf}$ modelling [see Scaling across Space and Time] will enhance our understanding of large scale $T_leaf$ vertical patterns.~~
 

### 2021-07-03 vertical-thermal-review.Rmd
deleted taller tree traits section , added citations
I figured maybe this would shorten the manuscript?

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen # $1.2^\circ$C since the 1850-1900 period [Rohde & Hausfater, 2020] and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@rohdeBerkeleyEarthLand2020] and hotter droughts [*Trenberth et al.
-2014*].
+Globally, temperatures have risen # $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014].
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
@@
 boundary layer conductance) and latent heat loss (via $gs$ ie.
-Stomatal conductance and $E$ ie evaporation).
+Stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].
 
@@
 `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig.
-`r fig_leaf_T`b)[Bonan 2008].
+`r fig_leaf_T`b)[@bonan_ecological_2016].
 However, low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low.
@@
 However in conditions of low soil moisture during drought (Fig.
-`r fig_leaf_T`), greater temperature than $gs$ thermal optimum, and air dryness causing an increase in VPD, $gs$ becomes limited, thereby increasing $T_{leaf}$-$T_{air}$.
+`r fig_leaf_T`), greater temperature than $gs$ thermal optimum, and air dryness causing an increase in VPD, $gs$ becomes limited, thereby increasing $T_{leaf}$-$T_{air}$ [@bonan_ecological_2016].
 Limitation to $gs$ can also exaggerate effects of biophysical factors that contribute to energy inputs such as, solar radiation, humidity and $T_{air}$ (Fig.
-`r fig_leaf_T`).
+`r fig_leaf_T`)[@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @dongBiophysicalHomoeostasisLeaf2017].
 
-Leaf size and physiology influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$.
+Leaf size and physiology [see next section] influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$.
 Smaller leaves have thinner boundary layer that enables coupling with $T_{air}$ through higher $ga$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig.
 `r fig_leaf_T`a).
-However, greater $gs$ in large leaves under low raditation, can be cooler than small leaves.
-Sun leaves, such as in upper canopies, tend to have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves.
-But exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions.
-However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in heat flux, $T_{leaf}$-$T_{air}$ can become more elevated than sun leaves (Fig.
-`r fig_leaf_T`a).
+Smaller leaves can also heat and cool faster and with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018].
+However, under low raditation, large leaves can be cooler than small leaves due to greater $gs$ [@leighThickLeavesAvoid2012;@bonan_ecological_2016].
+Sun leaves, such as in the upper canopies, tend to have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves.
+But exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021].
+However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in $T_{air}$, $T_{leaf}$-$T_{air}$ can become more elevated than sun leaves (Fig.
+`r fig_leaf_T`a)[@schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 
@@
 $T_{leaf}$ generally tracks $T_{air}$, and in turn the vertical gradient of $T_{air}$ depends on canopy structure and $LAI$.
-Closed canopies, and high $LAI$ act as a sheild, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
+Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
 Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
 On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
-It is also important to also note though NEON sites (in Fig.
-`r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures, but recent advances in $T_{leaf}$ modelling [see Scaling across Space and Time] will enhance our understanding of large scale $T_leaf$ vertical patterns.~~
+It is also important to note that, NEON sites (in Fig.
+`r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures.
+But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_leaf$ vertical patterns [@stillImagingCanopyTemperature].~~
 
@@
 `r fig_NEON_vertical`).
-In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high radiation [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ may be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
+In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ can be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
 `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
@@
 The extent of this buffering varies across forest types.
-*Upper canopies of temperate-deciduous species can have greater $T_{leaf} - T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig.
+*Available published data suggest that upper canopies of temperate-deciduous species can have greater $T_{leaf} - T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig.
 `r fig_NEON_vertical`).*
-Because needle-leaves have lower $g_a$ than broad leaves, conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
+Because needle-leaves have lower $g_a$ than broad leaves, needle-leaf conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
 
@@
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).
-
-~~~ *(this paragraph breaks the flow, and I'd love to be able to drop it or perhaps dramatically shorten. It is important and interesting content, though)*
-Both biophysical variables (light, temperature, and evaporative demand; Fig. `r fig_NEON_vertical`) and biological variables (tree height) play a role in shaping leaf traits. 
-In particular, hydraulic constraints increase with height on a tree [@mcdowell_relationship_2011; @cobleLightAcclimationOptimizes2015; @couvreur_water_2018a] such that gas exchange is strongly constrained at the tops of very tall trees (e.g., ~100 m.
-tall *Sequoia sempervirens*; *Ambrose et al.
-2009, 2010*).
-Isolating height (>35m) from light shows that, canopy $gs$ decreases with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
-Morphologically, leaf size [@jensenPhysicalLimitsLeaf2013b] and lobation [@kusiPlasticLeafMorphology2020] have been shown to decrease, and leaf mass per area (at greater than 20m) to increase, with height  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017; @marshallFoliageHeightInfluences2003].
-Disentangling tree-height related biological and biophysical drivers (Fig.
-`r fig_NEON_vertical`) of leaf traits remains an important area for further research, but is beyond the scope of the current review.
-Here, our primary interest is in how leaf traits and function vary across the vertical gradient from the canopy-top to the understory in forests.
-We note that patterns across light and height gradients are generally similar, at least in direction (Table 1), but likely to be modified by covarying biophysical constraints.
-~~~
 

### 2021-07-04 vertical-thermal-review.Rmd
added new citations, small edits

@@
 `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig.
-`r fig_leaf_T`b)[@bonan_ecological_2016].
-However, low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low.
+`r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004].
+Low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low.
+Therefore, along with windspeed $T_{leaf}$ can fluctuate throughout the day.
 Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig.
@@
 Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
-Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
+The intensity of incoming radiation and $VPD$ can also limit $gs$, therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
 On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
@@
 `r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures.
-But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_leaf$ vertical patterns [@stillImagingCanopyTemperature].~~
+But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature].~~
 
@@
 Caratenoids and xanthophyll cycle pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway  [Havaux, M.
-and Tardy, F., 1996].
-Elemental traits, such as N and P differ with soil conditions, but $N_{area}$ typically increases in the direction of new leaf growth, where its allocation is optimized to prevent photoinhibition in leaves [@kitaoCanopyNitrogenDistribution2018].
+This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].
+Elemental traits, such as N and P differ with soil conditions, but leaf nitrogen distribution within a canopy can be regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for shade leaves [@kitaoCanopyNitrogenDistribution2018].
 Additionally, isoprenes facilitate higher heat tolerance in exposed canopies in species capable of emitting isoprenes, providing these species with a selective advantage in a warming world [*REF*].
@@
 Marias et al.
-2017], and has a strong effect on metabolism (Kikuzawa & Lechowitz, Kitajima et al.).
+2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006].
 Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, they therefore operate closer to their photosynthetic optimum .
-However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy,  damage does not really accumulate [Miller et al.
-2021].
+However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy,  damage does not really accumulate [@millerOnlySunlitLeaves2021].
 The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is XXX.
@@
 Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests.
-Within tropical species, the fraction of deciduous trees tends to increase across tree size classes, with understory species typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c; Meakem].
-In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing-out earlier than canopy trees [Vitasse et al.
-2013; @seiwaChangesLeafPhenology1999].
+Within tropical species, the fraction of deciduous trees tends to increase across tree size classes, with understory species typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000].
+In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing-out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].
 

### 2021-07-05 vertical-thermal-review.Rmd
cut parts of Photosynthesis

@@
 
-~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig.
+~~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig.
 `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $ga$ ie.
 boundary layer conductance) and latent heat loss (via $gs$ ie.
-Stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].
+stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].
 
@@
 Leaf size and physiology [see next section] influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$.
-Smaller leaves have thinner boundary layer that enables coupling with $T_{air}$ through higher $ga$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig.
-`r fig_leaf_T`a).
-Smaller leaves can also heat and cool faster and with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018].
-However, under low raditation, large leaves can be cooler than small leaves due to greater $gs$ [@leighThickLeavesAvoid2012;@bonan_ecological_2016].
-Sun leaves, such as in the upper canopies, tend to have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves.
-But exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021].
-However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in $T_{air}$, $T_{leaf}$-$T_{air}$ can become more elevated than sun leaves (Fig.
+Smaller leaves have thinner boundary layer and higher $ga$, that enables coupling with $T_{air}$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig.
+`r fig_leaf_T` a).
+Therefore, smaller leaves with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018].
+Additionally, under low raditation, large leaves can be cooler than small leaves due to greater $gs$ [(Fig.
+`r fig_leaf_T` d; @leighThickLeavesAvoid2012;@bonan_ecological_2016].
+Sun leaves, such as in the upper canopies, have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves.
+However, exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021].
+However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in radiation or $T_{air}$, $T_{leaf}$-$T_{air}$ can easily become more elevated than sun leaves (Fig.
 `r fig_leaf_T`a)[@schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
@@
 **Within forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
-$T_{leaf}$ generally tracks $T_{air}$, and in turn the vertical gradient of $T_{air}$ depends on canopy structure and $LAI$.
+$T_{leaf}$ generally tracks $T_{air}$, where vertical gradient of $T_{air}$ depends on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
 Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
-The intensity of incoming radiation and $VPD$ can also limit $gs$, therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
-On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
-It is also important to note that, NEON sites (in Fig.
+The intensity of incoming radiation and $VPD$ in the upper canopy can limit $gs$.
+Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
+However, higher windspeeds and lower RH (Fig.
+`r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig.
+`r fig_leaf_T`a) and higher rates of heat loss (Fig.
+`r fig_leaf_T`b-e).
+On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
+
+NEON sites (in Fig.
 `r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures.
@@
 `r fig_NEON_vertical`).
-In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In closed-canopy forests, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ can be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
@@
 Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
-
-The extent of this buffering varies across forest types.
-*Available published data suggest that upper canopies of temperate-deciduous species can have greater $T_{leaf} - T_{air}$ than tropical moist and wet forest [@mauTemperateTropicalForest2018a] and sparse canopies [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018], however in considering a broader range of forests, results may differ (Fig.
-`r fig_NEON_vertical`).*
-Because needle-leaves have lower $g_a$ than broad leaves, needle-leaf conifer canopies have lower mean $T_{leaf}$ [@leuzingerTreeSpeciesDiversity2007; @pauTropicalForestTemperature2018].
 
@@
 
-In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubisco, and multiple layers of palisade parenchyma.
+In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions [Table 1; Table 2].
 Light saturation levels differ between leaves and canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P, and water availability.
@@
 Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to temperature, commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al.
-2017; @slotSituTemperatureResponse2017).
+Photosynthesis has a peaked response to temperature, commonly found to be maximized at the prevalent ambient growing season conditions [Tan et al.
+2017; @slotSituTemperatureResponse2017].
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g.
@@
 **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
-Biophysically, we might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5 in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al.
-2019 ; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and along the vertical gradient of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+Based on consistent positive relationships between growth temperature and $T_{Opt}$ across sites and seasons (e.g.
+Tan et al.
+2017; Kumarathunge et al.
+2019), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
+14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be moderatly, but not significantly higher than that of[ @slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
+Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot differ systematically between tropical sun- and shade-leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
 

### 2021-07-05 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013].
+Greater $VPD$ and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
 Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].

### 2021-07-05 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsWithincanopyVariationRate2004a].
+Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
 Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
@@
 Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al.
-2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
 Notably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.

### 2021-07-06 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].
-Elemental traits, such as N and P differ with soil conditions, but leaf nitrogen distribution within a canopy can be regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for shade leaves [@kitaoCanopyNitrogenDistribution2018].
-Additionally, isoprenes facilitate higher heat tolerance in exposed canopies in species capable of emitting isoprenes, providing these species with a selective advantage in a warming world [*REF*].
+
+Elemental traits, such as $N_{area}$ and $P_{area}$ contribute to high activities of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves, where $V_{cmax}$ and $J_{max}$ [see next section]-- are linearly related to within-canopy foliar $N_{area}$ and light gradient (Table 1)[@scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014].
+Evidence also shows that within canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
+Hence, we would expect elemental traits to influence vertical thermal sensitivity gradient.
+However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 suggested no relation between foliar thermal tolerance and total foliar concentrations of N and P, pointing towards further research to investigate elemental trait relation with vertical thermal sensivity.
+In species capable of emitting isoprenes, higher heat tolerance in exposed canopies is facilitated by isoprenes, providing these species with a selective advantage in a warming world [*REF*].
 

### 2021-07-06 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-07-05 vertical-thermal-review.Rmd
@@
 This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].
-Elemental traits, such as N and P differ with soil conditions, but leaf nitrogen distribution within a canopy can be regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for shade leaves [@kitaoCanopyNitrogenDistribution2018].
-Additionally, isoprenes facilitate higher heat tolerance in exposed canopies in species capable of emitting isoprenes, providing these species with a selective advantage in a warming world [*REF*].
+Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to high activities of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves, where $V_{cmax}$ and $J_{max}$ [see next section]-- are linearly related to within-canopy foliar $N_{area}$ and light gradient (Table 1)[@scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014].
+Evidence also shows that within canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
+Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity, this remains unclear as @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.
+Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes, providing these species with a selective advantage in a warming world [*REF*].
 

### 2021-07-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Marias et al.
-2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006].
-Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, they therefore operate closer to their photosynthetic optimum .
+2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
+Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum, with a few expections among species [@niinemetsLeafAgeDependent2016].
 However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, damage does not really accumulate [@millerOnlySunlitLeaves2021].
-The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is XXX.
+The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is that heat tolerance decreases with increasing leaf age [@zhouLeafageEffectsTemperature2015], along with photosynthesis and $gs$ [@@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
+In evergreen species, temperature responses of photosynthesis and in some cases temperature sensitivity of respiration ($Q_{10}$) [@zhouLeafageEffectsTemperature2015] have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+Additionally, because older leaves reach maximum $T_{leaf}$ earlier than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015].
 

### 2021-07-07 vertical-thermal-review.Rmd
added citations, removed a few words

--- 2021-07-06 vertical-thermal-review.Rmd
@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen # $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
-Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
+Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels $1.6 - 4.3^{\circ}C$ by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities 2021; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
+Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
+The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
 
@@
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
@@
 **Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all 22 forested NEON sites (Figs.
+Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all 22 forested NEON sites (Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
@@
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
-The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
-Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al.
-2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [@frenneForestMicroclimatesClimate2021].
+Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@frenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
 Notably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.
@@
 Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the other biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
-Often, however, $T_{leaf}$ tracks $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+Leaves can often be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy) [@doughtyAreTropicalForests2008].
+However, rarely is $T_{leaf}$ equal to $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
 
@@
 Low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low.
-Therefore, along with windspeed $T_{leaf}$ can fluctuate throughout the day.
 Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig.
-`r fig_leaf_T`c) because reduced $ga$ stifles heat dissapation through $E$ and $gs$.
+`r fig_leaf_T`c) because reduced $ga$ stifles $E$ and $gs$ that dissapate heat [@songTreeSurfaceTemperature2020].
 
@@
 $T_{leaf}$ generally tracks $T_{air}$, where vertical gradient of $T_{air}$ depends on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
-Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
-The intensity of incoming radiation and $VPD$ in the upper canopy can limit $gs$.
-Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
+Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in sub canopies and the understory microclimate.
+The intensity of incoming radiation and $VPD$ in the upper canopy can limit $gs$, therefore, we would expect a stronger $T_{leaf}$ gradient in closed canopies where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
 However, higher windspeeds and lower RH (Fig.
@@
 `r fig_leaf_T`a) and higher rates of heat loss (Fig.
-`r fig_leaf_T`b-e).
-On the other hand, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
+`r fig_leaf_T`b-e) [@millerOnlySunlitLeaves2021].
+On the other hand, open canopies, or lower canopy $LAI$ allow more turbulent air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
 
@@
 **Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optamize leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
-Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a].
+Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
@@
 Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-This variation shapes vertical $T_{leaf}$ patterns [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. 
-Further research to characterize trait relationships and responses through the vertical canopy is needed, because much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al.
+This variation shapes vertical $T_{leaf}$ patterns [Table 1; @fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. 
+Further, research to characterize trait relationships and responses through the vertical canopy is needed, because much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al.
 2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
@@
 
-Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves where, for instance, being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], and steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves, for instance, by being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], and steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance.
-Whereas, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks, for example lower LMA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013a].
+Whereas, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks, for example lower LMA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
 In open canopies where light is comparitively homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
@@
 This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].
-Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to high activities of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves, where $V_{cmax}$ and $J_{max}$ [see next section]-- are linearly related to within-canopy foliar $N_{area}$ and light gradient (Table 1)[@scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014].
+
+Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1; @scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014].
 Evidence also shows that within canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
@@
 Marias et al.
-2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006].
-Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, they therefore operate closer to their photosynthetic optimum .
+2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
+Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum.
 However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, damage does not really accumulate [@millerOnlySunlitLeaves2021].
-The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is XXX.
+The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is that heat tolerance decreases with increasing leaf age [@zhouLeafageEffectsTemperature2015; REFS], along with photosynthesis and $gs$ [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
+In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+Additionally, because older leaves reach maximum $T_{leaf}$ faster than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015].
 
@@
 
-Widely, across forest species, the pattern of fundamental trait adaptation to light gradients is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-However, trait plasticity can vary several-folds between  multi-species canopy and in understory species,  based on their inherent specie-specific traits and responses to environmental conditions.
+Generally, the pattern of fundamental trait adaptation to light gradients is across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+However, the range in trait values can be several times greater in multi-species canopy than in understory species,  based on their inherent species-specific traits and responses to environmental conditions.
 Spatio-temporal gradients, offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
-Therefore,  $T_{leaf}$ differs at an intra- and inter-specific levels with traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
+Therefore,  patterns in $T_{leaf}$ are underpinned by both intra- and inter-specific variation in traits.
+For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
 
@@
 
-**Photosynthesis is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al.
+**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al.
 2015 ; Slot et al.
@@
 2020).**
-This is primarily driven by the greater light available to sun-leaves.
-Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
+This is primarily driven by the greater light available to sun-leaves that have traits [Table 1; Table 2] allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions.
+Photosynthesis increases with irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
+Light saturation levels differ between leaves and whole canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.
 
-In response to the greater light availability, sun-leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions [Table 1; Table 2].
-Light saturation levels differ between leaves and canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P, and water availability.
-
-Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increase in $T_{leaf}$ from $gs$ closure that is driven by biophysical factors [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to temperature, commonly found to be maximized at the prevalent ambient growing season conditions [Tan et al.
+Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increased $VPD$ causing stomatal closure [@lloydEffectsRisingTemperatures2008].
+Photosynthesis has a peaked response to temperature, with the peak commonly found to correspond to the prevalent ambient growing season conditions [Tan et al.
 2017; @slotSituTemperatureResponse2017].
-Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing $T_{leaf}$ (e.g.
+Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g.
 @slotSituTemperatureResponse2017; Smith et al.
-2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; e.g.
-Sage & Kubien 2007; Varhammer et al.
-2015 ).
-Leaf heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II], leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020].
+2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; Sage & Kubien 2007; Varhammer et al.
+2015).
+Irresversible damage to photosystem II, leading to leaf necrosis and eventually leaf death occurs at $T_{leaf}$ of $40-60^{\circ}C$
+[Baker 2008; @feeley_thermal_2020].
 
 **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
-Based on consistent positive relationships between growth temperature and $T_{Opt}$ across sites and seasons (e.g.
+Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons (e.g.
 Tan et al.
@@
 14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
-However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be moderatly, but not significantly higher than that of[ @slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot differ systematically between tropical sun- and shade-leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be moderatly, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
+Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun- and shade-leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
 
-@hernandezSimilarTemperatureDependence2020 speculated that the lack of sun-shade differences in Topt may reflect acclimation to similar temperatures of peak photosynthesis.
+@hernandezSimilarTemperatureDependence2020 speculated that the lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis.
 Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of mid-day stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
@@
 In both species $T_{50}$ also tracked increased from June to August.
-Within species, @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.
+No studies have compared within species sun and shade leaves other than @slotPhotosyntheticHeatTolerance2019a who found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.
 
@@
 
-A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
-Other studies have observed no difference in $Q_{10}$ across species vertical profiles and in seedling sun vs.
+A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences in values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
+Other studies have observed no difference in $Q_{10}$ within species vertical profiles and in seedling sun vs.
 shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008].
@@
 Biophysical conditions can alter growth rates, but can do so differently between size classes through influencing existing demographic differences.
-Particulary evidence in in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014].
+Particulary.
+evidence in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014].
 Whereas in tropical forests higher nighttime temperatures was associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees [*Clark 2010; 2013; Feeley et al.
@@
 Phenologically, in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
-However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves more so than in understory saplings that are buffered [@filewodImpactsSpringHeat2014].
+However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves more severly than in understory saplings that are buffered [@filewodImpactsSpringHeat2014].
 Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].

### 2021-07-09 vertical-thermal-review.Rmd
trying to align text with issue #41

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels $1.6 - 4.3^{\circ}C$ by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}C$ by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities 2021; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
 Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
-The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].
+The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [REFS].
 
@@
 
-**The buffering patterns described above vary geographically, seasonally, and across forests types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, *Defrenne et al. 2021*].**
-The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture [**REF**], and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019].
-(*Nidhi, please make sure the references are appropriate here-- I rearranged the sentence structure.*)
-Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; **double check these refs--at least one didn't belong**].
+**The buffering patterns described above vary geographically, seasonally, and across forests types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, @frenneForestMicroclimatesClimate2021].**
+The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture, and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019].
+Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a].
 
@@
 In closed-canopy forests, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ can be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
+In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
 `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
@@
 
-Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves, for instance, by being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200], and steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves, for instance, by being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200;@mathurPhotosyntheticEfficiencySun2018a], and steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance.
-Whereas, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks, for example lower LMA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@mathurPhotosyntheticEfficiencySun2018a; @leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
+Whereas, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks, for example lower LMA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
 In open canopies where light is comparitively homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
 
-#### Biochemical traits
+#### Biochemical and elemental traits, and isoprenes
 
-Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory. 
+Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory [Table 1]. 
 Caratenoids and xanthophyll cycle pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
@@
 However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, damage does not really accumulate [@millerOnlySunlitLeaves2021].
-The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is that heat tolerance decreases with increasing leaf age [@zhouLeafageEffectsTemperature2015; REFS], along with photosynthesis and $gs$ [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
+The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is that heat tolerance decreases with increasing leaf age [@zhouLeafageEffectsTemperature2015; REFS], along with photosynthesis and $gs$ and with a few expections between species [see @niinemetsLeafAgeDependent2016].
 In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
-Additionally, because older leaves reach maximum $T_{leaf}$ faster than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015].
+Additionally, because older leaves reach maximum $T_{leaf}$ sooner than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015] **this needs some confirming by co-authors**.
 
@@
 Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests.
-Within tropical species, the fraction of deciduous trees tends to increase across tree size classes, with understory species typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000].
+Within tropical species, the fraction of deciduous trees tends to increase across tree size classes, with understory species typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
 In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing-out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].
@@
 
-Maximum $gs$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves.
-However, because of the high evaporative demand and irradiance experienced by sun leaves,  mid-day stomatal depression is prevalent in sun leaves but not in shade leaves.
-$T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006].
-Higher maximum temperatures influences leaf thermal sensitivity tolerance [Perez and Feeley, 2018], but exact mechanisms remain unclear.
-$T_{opt}$ of $gs$  did not differ significantly between sun and shade leaves in three tropical tree species [Slot et al.
-2019], but sun leaves show a stronger decrease in gs to rising temperature than shade leaves (Hernandez et al.
-2020; Carter et al.
-2021 ).
+Maximum $gs$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves [Table 2].
+However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is prevalent in sun leaves but not in shade leaves.
+$T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
+Higher maximum temperatures influences leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
+Additionally, $T_{opt}$ of $gs$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], but sun leaves show a stronger decrease in $gs$ to rising temperature than shade leaves [@hernandezSimilarTemperatureDependence2020; Carter et al.
+2021].
 
@@
 2020).**
-This is primarily driven by the greater light available to sun-leaves that have traits [Table 1; Table 2] allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions.
+This is primarily driven by the greater light available to sun-leaves, that also have traits [Table 1; Table 2] allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions.
 Photosynthesis increases with irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
@@
 2019), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
-However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be moderatly, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun- and shade-leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderatly, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
+Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
 
@@
 Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of mid-day stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
-Shade leaves that do not experience mid-day depression continue to fix carbon even when afternoon air temperatures peak.
+Shade leaves that do not experience mid-day depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021].
 Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.
@@
 
-$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [Slot et al.
-2020].
+$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [@slotLeafHeatTolerance2020].
 
@@
 In both species $T_{50}$ also tracked increased from June to August.
-No studies have compared within species sun and shade leaves other than @slotPhotosyntheticHeatTolerance2019a who found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.
+Within species, @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.
 
@@
 However, more research is required to fully understand the relationship between leaf thermal tolerance and microenvironments.
+
+*Further, if we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
+2017), where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
 
@@
 @perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that experienced higher maximum temperatures had higher photosynthetic heat tolerance.
-
-*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
-2017), where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
+Therefore, we would expect, at vertical gradient, leaves exposed to greater heat stress to have higher heat tolerance but further research is required to confirm this.
 
@@
 
-A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences in values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
+A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
 Other studies have observed no difference in $Q_{10}$ within species vertical profiles and in seedling sun vs.
@@
 May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*).
-@xuSeasonalVariationTemperature2006 found that the canopy-position effect gradient in respiration rates was much greater for Quercus rubra trees in mesic than in xeric sites, but the temperature sensitivity of respiration did not differ between canopy positions.
+@xuSeasonalVariationTemperature2006 found that the canopy-position effect gradient in respiration rates was much greater for Quercus rubra trees in mesic than in xeric sites, but $Q_{10}$ did not differ between canopy positions.
 
@@
 More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
-There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors.
-Tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
+There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors.Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
 However, growth also declines as trees age.

### 2021-07-10 vertical-thermal-review.Rmd
edits to Tleaf section
@NidhiVinod , we need to have a discussion to reconcile our two versions, #48

@@
 
-Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the other biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
-Leaves can often be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy) [@doughtyAreTropicalForests2008].
-However, rarely is $T_{leaf}$ equal to $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
+While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020], it is rarely exactly equal.
 
@@
 
-**attempted to oraganize**
+**Nidhi's rearrangement:**
 
 ~~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig.
-`r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $ga$ ie.
-boundary layer conductance) and latent heat loss (via $gs$ ie.
+`r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $g_a$; *i.e.*, boundary layer conductance) and $\lambda E$ (via $g_s$ ie.
 stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].
 
-$Ga$ and $Gs$ are dominant factors in $T_{leaf}$ flux.
+$g_a$ and $g_s$ are dominant factors in $T_{leaf}$ flux.
 $Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig.
@@
 
-Latent heat loss through maximum $gs$, cools the leaf and decreases $T_{leaf}$-$T_{air}$, allowing coupling between leaf and air.
+Latent heat loss through maximum $g_s$, cools the leaf and decreases $T_{leaf}$-$T_{air}$, allowing coupling between leaf and air.
 However in conditions of low soil moisture during drought (Fig.
@@
 
-Leaf size and physiology [see next section] influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$.
-Smaller leaves have thinner boundary layer and higher $ga$, that enables coupling with $T_{air}$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig.
+Leaf size and physiology [see next section] influence $g_a$, $g_s$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$.
+Smaller leaves have thinner boundary layer and higher $g_a$, that enables coupling with $T_{air}$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig.
 `r fig_leaf_T` a).
@@
 But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature].~~
+
+**Krista's version:**
 
@@
 ### In situ observations
-**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations (Figs.
-`r fig_NEON_vertical`-`r fig_leaf_T`).**
-Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig.
-`r fig_NEON_vertical`).
-In closed-canopy forests, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs.
+**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations.**
+Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy.
+In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
 `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
-Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
 Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
 
-Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
+It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems. 
+For instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018]. 
+Thus, the vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.
+`r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories).
+Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature].
 

### 2021-07-10 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems. 
-For instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018]. 
+It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems; for instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
 Thus, the vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.

### 2021-07-10 vertical-thermal-review.Rmd
quick pass through leaf traits
@NidhiVinod , in general this looks very good!

@@
 
-**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optamize leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
+**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
 Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
@@
 Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-This variation shapes vertical $T_{leaf}$ patterns [Table 1; @fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016]. 
-Further, research to characterize trait relationships and responses through the vertical canopy is needed, because much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al.
-2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016].
-
-Independent of light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
+~~This variation shapes vertical $T_{leaf}$ patterns [Table 1, @fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].~~
+As much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al.
+2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses through the vertical canopy.
+It has been established that independent of light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
 For example, LMA, photosynthetic capacity [per area] and $gs$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
@@
 
-#### Anatomical, Morphological and Physiological traits
+#### (Anatomical, Morphological and Physiological traits)
 
-Sun leaves in the exposed canopy have traits that help avoid heat and water stress than shade leaves, for instance, by being smaller, hairier, with higher vein density [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200;@mathurPhotosyntheticEfficiencySun2018a], and steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
-These traits [see Table 1] collectively aid heat dissapation and regulate $T_{leaf}$-photosynthetic balance.
-Whereas, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks, for example lower LMA, but are more likely to get overheated than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies where light is comparitively homogenous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
+Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, by being smaller, hairier, and with higher vein density than shade leaves [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+These traits (Table 1) collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
+In contrast, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks (e.g., lower LMA), but are more likely to get overheated than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
+In open canopies where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
 
-#### Biochemical and elemental traits, and isoprenes
+#### (Biochemical and elemental traits, and isoprenes)
 
-Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory [Table 1]. 
-Caratenoids and xanthophyll cycle pigments--violaxanthin, antheraxanthin and zeaxanthin [$VAZ$]--can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-This is also important for leaf acclimation to higher temperature if high temperature impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].
+Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
+Caratenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin ($VAZ$) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+This is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].
 
-Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1; @scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014].
-Evidence also shows that within canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunlecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
-Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity, this remains unclear as @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.
-Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes, providing these species with a selective advantage in a warming world [*REF*].
+Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014].
+Within-canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
+Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity,  @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.
 
-#### Leaf lifespan and Deciduousness ("Lifecycle / ecological traits? ")
+Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes, *providing these species with a selective advantage in a warming world* [*REF*].
 
-##### *Leaf lifespan*
+#### (Leaf lifecycles)
 
-Leaf age affects vulnerability to heat stress independent of light environment [e.g.
-Marias et al.
+Leaf age affects vulnerability to heat stress independent of light environment [e.g., Marias et al.
 2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
 Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum.
-However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, damage does not really accumulate [@millerOnlySunlitLeaves2021].
-The effect of leaf age on heat tolerance once leaves are fully expanded and mature, is that heat tolerance decreases with increasing leaf age [@zhouLeafageEffectsTemperature2015; REFS], along with photosynthesis and $gs$ and with a few expections between species [see @niinemetsLeafAgeDependent2016].
+However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, less damage  accumulates [@millerOnlySunlitLeaves2021].
+Once leaves leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions between species, @niinemetsLeafAgeDependent2016].
 In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
@@
 
-Generally, the pattern of fundamental trait adaptation to light gradients is across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Generally, the pattern of fundamental trait adaptation to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
 However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
-Spatio-temporal gradients, offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007; @ishiiExploringRelationshipsCanopy2004a].
-Therefore,  patterns in $T_{leaf}$ are underpinned by both intra- and inter-specific variation in traits.
-For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].
+Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a]. 
+Therefore, $T_{leaf}$ patterns witnin forest ecosystems are underpinned by both intra- and inter-specific variation in traits. 
+*For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].*
 
-##### *Deciduousness*
+##### (*Deciduousness* -- drop header, probably integrate this into paragraph above, or somehow re-organize the whole itnerspecific variation section)
 
-Seasonal leaf-loss or -gain can alter biophysical and $T_{leaf}$ gradient in the forest [@rey-sanchez_spatial_2016].
+Seasonal leaf-loss or gain can alter biophysical and $T_{leaf}$ gradient in the forest [@rey-sanchez_spatial_2016].
 Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests.

### 2021-07-10 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
+Leaf metabolism is strongly shaped by $T_{leaf}$, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
 However, as we detail below, very few studies have valued temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).

### 2021-07-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ---
-nocite: | 
- @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007;Carter and Cavaleri 2018; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; Hamerlynck and Knapp 1994; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al.
+nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007;Carter and Cavaleri 2018; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; Hamerlynck and Knapp 1994; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al.
 1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; Harris and Medina 2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @athertonSpatialVariationLeaf2017
@@
 
-A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
+A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in *Populas deltoides*, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
 Other studies have observed no difference in $Q_{10}$ within species vertical profiles and in seedling sun vs.
@@
 
-## Implications [ & future research directions?]
-
-*Elsa: If emphsaize future research directions, could move the remote sensing opportunities to this section and frame modeling section around opportunities to use models to further explore open questions, scale these findings in space and time, and influence predictions*
-*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
-
 Still, partitioning $ET$ into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
@@
 With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.
+
+## Implications [ & future research directions?]
+
+*Elsa: If emphsaize future research directions, could move the remote sensing opportunities to this section and frame modeling section around opportunities to use models to further explore open questions, scale these findings in space and time, and influence predictions*
+*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
 

### 2021-07-11 vertical-thermal-review.Rmd
remove not on @SlotMartijn 's data
I think it's late to try to include this, but if @SlotMartijn feels otherwise we can change course here. For now, I'm taking it out and will close the issue (#8)
Update vertical-thermal-review.Rmd

@@
 Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
-While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020], it is rarely exactly equal.
+While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$, it is rarely exactly equal [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
 
@@
 
-Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes, *providing these species with a selective advantage in a warming world* [*REF*].
+Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes [REF].
 
@@
 Once leaves leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions between species, @niinemetsLeafAgeDependent2016].
-In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
 Additionally, because older leaves reach maximum $T_{leaf}$ sooner than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015] **this needs some confirming by co-authors**.
@@
 Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].
-Therefore, $T_{leaf}$ patterns witnin forest ecosystems are underpinned by both intra- and inter-specific variation in traits. 
-*For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].*
-
-##### (*Deciduousness* -- drop header, probably integrate this into paragraph above, or somehow re-organize the whole itnerspecific variation section)
 
@@
 In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing-out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].
+
+Therefore, $T_{leaf}$ patterns within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
 
@@
 shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008].
-(*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so.
-May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*).
 @xuSeasonalVariationTemperature2006 found that the canopy-position effect gradient in respiration rates was much greater for Quercus rubra trees in mesic than in xeric sites, but $Q_{10}$ did not differ between canopy positions.
+
+-In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
 
@@
 
+*For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].*
+
+-Isoprenes *providing emitting species with a selective advantage in a warming world* [*REF*].
+
 These shifts in forest structure at large may be detectable by lidar and thermal remote sensing.

### 2021-07-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}C$ by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities 2021; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
+Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilitiesa].
 Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
@@
 
-Relative humidity, $RH$, tends to be higher in the understory and decrease with height across the vertical profile, although this trend is absent in open forests [Figs.
+Relative humidity, $RH$, tends to be higher in the understory and decreases with height across the vertical profile, although this trend is absent in open forests [Figs.
 `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregor_tree_nodate].
@@
 ~~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig.
-`r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $g_a$; *i.e.*, boundary layer conductance) and $\lambda E$ (via $g_s$ ie.
-stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].
+`r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $g_a$; *i.e.*, boundary layer conductance) and $\lambda E$ (via $g_s$ *ie.* stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].
 
@@
 Therefore, smaller leaves with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018].
-Additionally, under low raditation, large leaves can be cooler than small leaves due to greater $gs$ [(Fig.
-`r fig_leaf_T` d; @leighThickLeavesAvoid2012;@bonan_ecological_2016].
-Sun leaves, such as in the upper canopies, have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves.
+Additionally, under low raditation, large leaves can be cooler than small leaves due to greater $g_s$ [Fig.
+`r fig_leaf_T` d, @leighThickLeavesAvoid2012;@bonan_ecological_2016].
+Sun leaves, such as in the upper canopies, have greater $gs$, due to greater stomatal density (Table 1) and lower thermal capacitance due to lower water potential than shade leaves.
 However, exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021].
@@
 `r fig_leaf_T`b-e) [@millerOnlySunlitLeaves2021].
-On the other hand, open canopies, or lower canopy $LAI$ allow more turbulent air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013].
+On the other hand, open canopies, or lower canopy $LAI$ allow more turbulent air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].
 
@@
 `r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures.
-But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature].~~
+But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021].~~
 
@@
 Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy.
-In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high radiation load [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
@@
 `r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories).
-Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature].
+Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021].
 

### 2021-07-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-~~This variation shapes vertical $T_{leaf}$ patterns [Table 1, @fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].~~
 As much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al.
@@
 It has been established that independent of light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
-For example, LMA, photosynthetic capacity [per area] and $gs$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+For example, LMA, photosynthetic capacity (per area) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
 
@@
 
-Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, by being smaller, hairier, and with higher vein density than shade leaves [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, by being smaller, hairier, and with higher vein density than shade leaves [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits (Table 1) collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
@@
 Caratenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin ($VAZ$) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-This is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].
+This is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
 
-Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014].
+Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
 Within-canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
@@
 
-Maximum $gs$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves [Table 2].
+Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves (Table 2).
 However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is prevalent in sun leaves but not in shade leaves.
 $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-Higher maximum temperatures influences leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
-Additionally, $T_{opt}$ of $gs$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], but sun leaves show a stronger decrease in $gs$ to rising temperature than shade leaves [@hernandezSimilarTemperatureDependence2020; Carter et al.
-2021].
+Higher maximum temperatures and $g_s$ influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
+Additionally, $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], but sun leaves show a stronger decrease in $gs$ to rising temperature than shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
 

### 2021-07-12 vertical-thermal-review.Rmd
brand new experimental warming paper

@@
 For example, LMA, photosynthetic capacity (per area) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical; @niinemetsResponsesForestTrees2010a].
 
@@
 Higher maximum temperatures and $g_s$ influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
-Additionally, $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], but sun leaves show a stronger decrease in $gs$ to rising temperature than shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
+Additionally, $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], but sun leaves in the upper canopy show a stronger decrease in $gs$ to rising temperature than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
 

### 2021-07-12 vertical-thermal-review.Rmd
added carter et al. 2021 (new paper)

@@
 Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy.
-In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high radiation load [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high radiation loads [@carterExperimentalWarmingTropical; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
@@
 Tan et al.
-2017; Kumarathunge et al.
-2019), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
+2017; @kumarathungeAcclimationAdaptationComponents2019a), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
 14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].

### 2021-07-12 vertical-thermal-review.Rmd
fix up Tleaf alternate versions
#48

@@
 
-**Nidhi's rearrangement:**
+**version 1 (for rest of biophyscial drivers subsection-- coauthors, please let us know which is better):**
 
-~~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig.
+Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig.
 `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $g_a$; *i.e.*, boundary layer conductance) and $\lambda E$ (via $g_s$ *ie.* stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].
@@
 
-NEON sites (in Fig.
-`r fig_NEON_vertical`) might not exactly reflect the $T_{leaf}$ pattern we might expect, partly because $T_{bio}$ measured using infrared sensor, reflects a wide swath of biological temperatures.
-But, recent advances in $T_{leaf}$ modelling that we outline [see Scaling across Space and Time], will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021].~~
-
-**Krista's version:**
+**version 2 (for rest of biophyscial drivers subsection-- coauthors, please let us know which is better):**
 

### 2021-07-12 vertical-thermal-review.Rmd
small changes

@@
 Leaf metabolism is strongly shaped by $T_{leaf}$, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
-However, as we detail below, very few studies have valued temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+However, as we detail below, very few studies have evaluated temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
 
@@
 
-**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology and ecosystem ecology.**
-
-Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in understory versus canopy conditions.
+**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect whole-tree ecology, ecosystem ecology, and their thermal sensitivity (Fig.
+`r fig_schematic`).**
+Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby
 

### 2021-07-12 vertical-thermal-review.Rmd
moved deciduousness text from traits to ecology

@@
 
-Seasonal leaf-loss or gain can alter biophysical and $T_{leaf}$ gradient in the forest [@rey-sanchez_spatial_2016].
-Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests.
-Within tropical species, the fraction of deciduous trees tends to increase across tree size classes, with understory species typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing-out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].
-
 Therefore, $T_{leaf}$ patterns within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
@@
 
-**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect whole-tree ecology, ecosystem ecology, and their thermal sensitivity (Fig.
+**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect community composition, whole-tree ecology, and ecosystem ecology (Fig.
 `r fig_schematic`).**
-Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby
+
+### Size-structuring of communities
+
+Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
+
+Seasonal leaf-loss or gain can alter biophysical and $T_{leaf}$ gradient in the forest [@rey-sanchez_spatial_2016].
+Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests.
+Within tropical species, the fraction of deciduous trees tends to increase across tree size classes, with understory species typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing-out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].
 

### 2021-07-12 vertical-thermal-review.Rmd
a few changes
... nothing very productive...

@@
 However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
-Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].
+8Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].*
 
@@
 
+Numerous traits vary between understory and canopy species. (*this needs to be reconciled with interspecific portion of traits section*)
+
 Seasonal leaf-loss or gain can alter biophysical and $T_{leaf}$ gradient in the forest [@rey-sanchez_spatial_2016].
@@
 **Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
-More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
-There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)* Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors.Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
+There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)*
+Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors. 
+Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
 However, growth also declines as trees age. 

### 2021-07-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
-Sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum.
+*In evergreen species,* sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum.
 However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, less damage  accumulates [@millerOnlySunlitLeaves2021].

### 2021-07-12 vertical-thermal-review.Rmd
moved deciduousness content back to traits section, edited interspecific traits

@@
 
+The timing of seasonal leaf loss or gain is influenced by vertical microclimate gradients, and in turn can alter biophysical and $T_{leaf}$ gradients in the forest [@rey-sanchez_spatial_2016].
+Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests. 
+Within tropical deciduous species, the fraction of deciduous individuals tends to increase across tree size classes, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].
+
 ### Interspecific variation
 
+Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
 Generally, the pattern of fundamental trait adaptation to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+For instance, across tree size classes, there is an increase in the fractions of species that are isoprene-emitting (**REFS**) and deciduous [@conditQuantifyingDeciduousnessTropical2000c; @Meakem2018].
 However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
-8Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].*
+*Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].*
 
@@
 
-**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect community composition, whole-tree ecology, and ecosystem ecology (Fig.
+**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect whole-tree ecology, ecosystem ecology, and their temperature responses (Fig.
 `r fig_schematic`).**
-
-### Size-structuring of communities
-
-Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
-
-Numerous traits vary between understory and canopy species. (*this needs to be reconciled with interspecific portion of traits section*)
-
-Seasonal leaf-loss or gain can alter biophysical and $T_{leaf}$ gradient in the forest [@rey-sanchez_spatial_2016].
-Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests.
-Within tropical species, the fraction of deciduous trees tends to increase across tree size classes, with understory species typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing-out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].
 

### 2021-07-13 vertical-thermal-review.Rmd
working on ecology section

@@
 
-### Demography
+### Tree metabolism and growth
 
-**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
-There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)*
-Demographic differences between canopy trees and understory trees can also influence their their ability to tolerate multiple stressors. 
-Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
-However, growth also declines as trees age. 
-This is partly because autotrophic respiration increases with age that declines NPP, whereas due to increase in morphological and anatomical complexity, GPP also increases [@IncreasingRespiratoryCosts].
-Vertical biophysical conditions are bound to influence growth and whole tree functioning.
+**Vertical gradients affect tree metabolism and growth.**
+Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeira_sizerelated_2015].
+Specifically, increases are linked in part to increased leaf area and in part to increasing probability that the crown is in the canopy [@Muller-Landau_2006_EcologyLetters], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific metabolism (Table 2).
+Of the fixed C, a higher proportion is consumed by respiration, in part because of higher leaf respiration ($R_{dark}$, Table 2) and in part because of higher sapwood:leaf area ratio (**REFS** -- [@IncreasingRespiratoryCosts]?).
 
-Greater $VPD$ and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+- C allocation --> respiration, woody growth, NSCs, foliage turnover, belowground allocation, reproduction
+- we don't know a whole lot about how C allocation scales with tree size
+- woody growth increases with DBH to a point, then declines [@AT_inReview].
+- NSCs increase , making smaller trees more sensitive to biological and biophyical stressors [@niinemetsResponsesForestTrees2010a]
+
+**Vertical gradients also affect the climate sensitivity of metabolism and growth.**
+
+- Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
+
+Greater $VPD$ and low RH can be more stressful for trees than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
 Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
@@
 Phenologically, in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
-However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves more severly than in understory saplings that are buffered [@filewodImpactsSpringHeat2014].
+However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves more severly than in understory saplings that are buffered [@filewodImpactsSpringHeat2014].
 Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].

### 2021-07-13 vertical-thermal-review.Rmd
working on ecology section
@NidhiVinod , this is focusing on getting the structure right, with lots of details to be filled in/ some condensing required. please see what you think.

@@
 
+*all of the following should be condensed into a paragraph or maybe 2:*
+
+~~~
+
 - Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
@@
 
-**Recently emerging evidence indicates that understory trees tend to exhibit greater growth declines with increasing $T_{air}$ under mesic conditions.** 
+Recently emerging evidence indicates that understory trees tend to exhibit greater growth declines with increasing $T_{air}$ under mesic conditions.
 In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
@@
 
+~~~~
+
 ![**Figure `r fig_treerings`.
@@
 
+**Tree mortality risk is also size-structured.**
+
+- most common trend is higher risk of drought mortality for larger trees [@bennettLargerTreesSuffer2015; @stovall], including in open forests (**REFS**). *(However, there are exceptions...)*
+- also, greater risks of lightning, blow-down, etc.  (@Esquivel-Muelbert_2021) *(although this isn't temperature-related, so should get minimal attention, if at all)*
+
 ### C and water flux
 
-*(here, it would be good to add some content on contributions to C cycling across the vertical gradient.
-If Camille's paper is published ahead of this, that would be the best source.
-Meakem et al.
-2018, New Phyt has a similar analysis for 3 sites.)*
+**Canopy trees account for the majority of forest ecosystem metabolism-- i.e., transpiration, GPP, woody productivity ($ANPP_{woody}$).**
 
-Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases.
+- ET: "Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases.
 As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see @nakamuraForestsTheirCanopies2017), leaf traits, and photosynthesis .
@@
 2020), likely affecting vertical thermal gradients.
-In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout.
-
-Still, partitioning $ET$ into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
+In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout." *(condense this by quite a lot.
+I don't think all the info here is necessary)*
+- ET : "Still, partitioning $ET$ into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
 Eddy covariance techniques, sap flow measurements, and stable isotope analysis offer useful tools, although errors can be large (Williams et al 2004, Herbst et al.
@@
 2017 -not sure if this is equivalent)*.
-With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.
+" - *(revise this to focus on the observed patterns, with at most brief mention of the methods)*
+- GPP : ??
+- $ANPP_{woody}$: Piponiot et al., accepted, Meakem et al. 2018, New Phyt
+
+**Less clear is how thermal sensitivity varies across strata.**
+
+- ET: "With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions."
+- GPP : ??
+- $ANPP_{woody}$: if we scale tree-rings to the ecosystem level, we expect that woody productivity of canopy trees is more sensitive to drought, but perhaps less sensitive to wet heat.
+More research is needed here.
 

### 2021-07-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Generally, the pattern of fundamental trait adaptation to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-For instance, across tree size classes, there is an increase in the fractions of species that are isoprene-emitting (**REFS**) and deciduous [@conditQuantifyingDeciduousnessTropical2000c; @Meakem2018].
+For instance, across tree size classes, there is an increase in the fractions of species that are isoprene-emitting (**REFS**) and deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
 However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
-*Spatio-temporal gradients offer another axis for trait variation, for instance in early- and late-successional species [Ishii and Asano, 2007, @ishiiExploringRelationshipsCanopy2004a].*
 

### 2021-07-13 vertical-thermal-review.Rmd
working on ecology section

@@
 
-### Tree metabolism and growth
+### Tree metabolism, growth, and survival
 
 **Vertical gradients affect tree metabolism and growth.**
-Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeira_sizerelated_2015].
-Specifically, increases are linked in part to increased leaf area and in part to increasing probability that the crown is in the canopy [@Muller-Landau_2006_EcologyLetters], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific metabolism (Table 2).
-Of the fixed C, a higher proportion is consumed by respiration, in part because of higher leaf respiration ($R_{dark}$, Table 2) and in part because of higher sapwood:leaf area ratio (**REFS** -- [@IncreasingRespiratoryCosts]?).
+Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizeRelatedScaling2015].
+Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetaabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific metabolism (Table 2).
+Of the fixed C, a higher proportion is consumed by respiration, in part because of higher leaf respiration ($R_{dark}$, Table 2) and in part because of higher sapwood volume (??confirm??) [**REFS**; @IncreasingRespiratoryCosts?].
 
@@
 **Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
+Across biomes, including mesic forests, surface conductance and evapotranspiration tend to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+In terms of transpiration and photosynthesis, the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have lower rates during periods of high- relative to low- atmospheric demand -- be these  on the time scale of hours (**REFS?**), days [@anderson-teixeiraSizeRelatedScaling2015], seasons [@christoffersenLinkingHydraulicTraits2016], or years (**?
+REFS?**).
 
-*all of the following should be condensed into a paragraph or maybe 2:*
+- Taller trees in the Amazon had greater stomatal sensitivity to drought, resulting in greater reductions transpiration and photosynthesis [@garciaImportanceHydraulicStrategy2021].
+- @christoffersenLinkingHydraulicTraits2016 uses trait-based modeling to show that canopy trees have lower dry-season: wet-season transpiration. 
+- Smaller trees also have greater capacity to adjust traits realted carbon metabolism to drought [@bartholomewSmallTropicalForest2020]
+- *(it would be good to find a relevant study for open forests-- perhaps track down studies mentioned by Bev Law)*
+
+In contrast, greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+
+The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments and sometimes forest census data.
+While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions (**REFS--KAT**), only a relatively limited number of studies have directly examined drought- or temperature- sensitivities as a function of tree size.
+These have generally found greater sensitivity to low precipitation or high temperatures among large/ canopy trees [**Fig.
+`r fig_treerings`**, @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021] , corroborating evidence from forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
+Despite the potential for shorter trees in open forests to experience greater environmental stress [*Fig.
+`r fig_schematic`b*; **REFS**], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021].
+
+There are, however, noteable exceptions to this pattern: in eight mesic forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
+Similarly, in tropical forests higher night-time temperatures have been associated with decelerating tree growth rates predominantly in smaller trees *but also in taller trees* [*Clark 2010; 2013; Feeley et al.
+2007*].
+Potential mechanisms for these patterns include increases in nighttime respiratory carbon loss that disproportionately reduced carbon availability for growth among small trees, or that understory $g_s$  might be stifled under humid conditions, resulting in higher $T_{leaf}$ (Fig.
+`r fig_leaf_T`) and lower photosynthesis.
+It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. 
+Additional research is required to understand these mechanisms and to disentangle tree growth responses to hot-wet versus hot-dry conditions.
 
 ~~~ 
+misc content to incorpoarte:
 
-- Within species, tolerance of more than one biological and biophyical stressors can increase with tree height and light, enabled by non-structural carbon reserve [NSC], with smaller shaded trees being more sensitive due to lower NSC [@niinemetsResponsesForestTrees2010a].
+- evidence in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014].
 
-Greater $VPD$ and low RH can be more stressful for trees than soil moisture across biomes-including mesic forests, as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Whereas greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Altered precipitation regimes can cause droughts, that are more  detrimental to larger trees than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [@bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020].
-However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig.
-`r fig_leaf_T`).
-
-Biophysical conditions can alter growth rates, but can do so differently between size classes through influencing existing demographic differences.
-Particulary.
-evidence in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014].
-Whereas in tropical forests higher nighttime temperatures was associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees [*Clark 2010; 2013; Feeley et al.
-2007*], perhaps due to increase in nighttime respiratory carbon loss that reduced carbon availability for growth.
-
-Recently emerging evidence indicates that understory trees tend to exhibit greater growth declines with increasing $T_{air}$ under mesic conditions.
-In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
-This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
- It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.
-In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
+- In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
+ALso, tropical warming experiment showed increased growth (**Winter**)
 
@@
 
-**Tree mortality risk is also size-structured.**
+**Tree mortality risk from climate-related stress is also size-structured.**
 

### 2021-07-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-There are, however, noteable exceptions to this pattern: in eight mesic forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
+There are, however, noteable exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
+In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory].
 Similarly, in tropical forests higher night-time temperatures have been associated with decelerating tree growth rates predominantly in smaller trees *but also in taller trees* [*Clark 2010; 2013; Feeley et al.

### 2021-07-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 @mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; Ishida et al.
-1999; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; Harris and Medina 2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000
+1999; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000
 ---
@@
 
-Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes [REF].
+Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes [*REF*].
 
@@
 ---
-nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007;Carter and Cavaleri 2018; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; Hamerlynck and Knapp 1994; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al.
-1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; Harris and Medina 2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @athertonSpatialVariationLeaf2017
+nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al.
+1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @athertonSpatialVariationLeaf2017
 ---
@@
 **Vertical gradients affect tree metabolism and growth.**
-Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizeRelatedScaling2015].
-Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetaabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific metabolism (Table 2).
+Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
+Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific metabolism (Table 2).
 Of the fixed C, a higher proportion is consumed by respiration, in part because of higher leaf respiration ($R_{dark}$, Table 2) and in part because of higher sapwood volume (??confirm??) [**REFS**; @IncreasingRespiratoryCosts?].

### 2021-07-14 vertical-thermal-review.Rmd
added some citations

@@
 Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific metabolism (Table 2).
-Of the fixed C, a higher proportion is consumed by respiration, in part because of higher leaf respiration ($R_{dark}$, Table 2) and in part because of higher sapwood volume (??confirm??) [**REFS**; @IncreasingRespiratoryCosts?].
+Of the fixed C, a higher proportion is consumed by respiration, in part because of higher leaf respiration ($R_{dark}$, Table 2) and in part because of higher sapwood volume (??confirm??) [**REFS**; @westIncreasingRespiratoryCosts2020].
 
@@
 Across biomes, including mesic forests, surface conductance and evapotranspiration tend to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-In terms of transpiration and photosynthesis, the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have lower rates during periods of high- relative to low- atmospheric demand -- be these  on the time scale of hours (**REFS?**), days [@anderson-teixeiraSizeRelatedScaling2015], seasons [@christoffersenLinkingHydraulicTraits2016], or years (**?
+In terms of transpiration and photosynthesis, the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have lower rates during periods of high- relative to low- atmospheric demand -- be these  on the time scale of hours (**REFS?**), days [@anderson-teixeiraSizerelatedScalingTree2015], seasons [@christoffersenLinkingHydraulicTraits2016], or years (**?
 REFS?**).
@@
 
-In contrast, greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+In contrast, greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].
 
@@
 
-- most common trend is higher risk of drought mortality for larger trees [@bennettLargerTreesSuffer2015; @stovall], including in open forests (**REFS**). *(However, there are exceptions...)*
-- also, greater risks of lightning, blow-down, etc.  (@Esquivel-Muelbert_2021) *(although this isn't temperature-related, so should get minimal attention, if at all)*
+- most common trend is higher risk of drought mortality for larger trees [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019b], including in open forests (**REFS**).
+*(However, there are exceptions...)*
+- also, greater risks of lightning, blow-down, etc.  (@goraImplicationsSizedependentTree2021) *(although this isn't temperature-related, so should get minimal attention, if at all)*
 
@@
 - GPP : ??
-- $ANPP_{woody}$: Piponiot et al., accepted, Meakem et al. 2018, New Phyt
+- $ANPP_{woody}$: Piponiot et al., accepted, @meakemRoleTreeSize2018, New Phyt
 

### 2021-07-14 vertical-thermal-review.Rmd
added some citations

@@
 Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilitiesa].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilitiesa 2021].
 Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
@@
 Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs.
-open canopy forests [e.g., @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
+open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019).
@@
 2020).**
-This is primarily driven by the greater light available to sun-leaves, that also have traits [Table 1; Table 2] allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions.
+This is primarily driven by the greater light available to sun-leaves, that also have traits (Table 1, Table 2) allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions.
 Photosynthesis increases with irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
@@
 
-In 1994, Knapp et al.
-showed that $T_{50}$ measured on chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions was higher by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions.
-In both species $T_{50}$ also tracked increased from June to August.
-Within species, @slotPhotosyntheticHeatTolerance2019a found that $T_{50}$ was slightly lower for shade than sun leaves for two of three tree species measured in Panama.
+Leaf thermal tolerances are more closely adapted to microenvironment through leaf traits than macroclimate and phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
+Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ measured by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+At vertal gradient within canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
+Further, at a leaf trait level, $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species' leaves with greater LMA, compared to species with low LMA [@slotLeafHeatTolerance2020].
+Additionally, @perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that experienced higher maximum temperatures had higher photosynthetic heat tolerance.
+However, even with sun and shade leaf differences within canopy (Table 1)$T_{50}$ modestly varies, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a].
+Based on these evidences, we would expect, leaves exposed to greater heat stress in exposed canopies to have higher heat tolerance but further research is required to fully understand relationship between heat tolerance, light exposure and microclimates.
 
-Leaf thermal tolerances are more closely adapted to microenvironment than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
-Particularly, in an Australian arid region, *Acacia papyrocarpa* (Benth.) trees showed higher composite climate stress and greater $T_{50}$ in lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
-However, more research is required to fully understand the relationship between leaf thermal tolerance and microenvironments.
-
-*Further, if we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
+*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
 2017), where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
-
-*We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?).
-I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot*
-
-Across species, $T_{50}$ is explained less by phylogeny [Slot et al.
-2021; Perez and Feeley, 2020a] and more by leaf traits.
-$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time [higher thermal time constant] whereas $T_{50}$ was greater among species' leaves with greater LMA (greater construction costs), compared to species with low LMA [@slotLeafHeatTolerance2020].
-@perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that experienced higher maximum temperatures had higher photosynthetic heat tolerance.
-Therefore, we would expect, at vertical gradient, leaves exposed to greater heat stress to have higher heat tolerance but further research is required to confirm this.
 
@@
 
-**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves (Chen et al.
-2020), but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**
+**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [@chenLeafEconomicsSpectrum2020], but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**
 

### 2021-07-14 vertical-thermal-review.Rmd
cutting words in T50 section

@@
 Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increased $VPD$ causing stomatal closure [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to temperature, with the peak commonly found to correspond to the prevalent ambient growing season conditions [Tan et al.
-2017; @slotSituTemperatureResponse2017].
+Photosynthesis has a peaked response to temperature, with the peak commonly found to correspond to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017].
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g.

### 2021-07-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf thermal tolerances are more closely adapted to microenvironment through leaf traits than macroclimate and phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
-Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ measured by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
-At vertal gradient within canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
-Further, at a leaf trait level, $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species' leaves with greater LMA, compared to species with low LMA [@slotLeafHeatTolerance2020].
-Additionally, @perezPhotosyntheticHeatTolerances2020 also found that, across species sun leaves that experienced higher maximum temperatures had higher photosynthetic heat tolerance.
-However, even with sun and shade leaf differences within canopy (Table 1)$T_{50}$ modestly varies, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a].
-Based on these evidences, we would expect, leaves exposed to greater heat stress in exposed canopies to have higher heat tolerance but further research is required to fully understand relationship between heat tolerance, light exposure and microclimates.
-
 *If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
 2017), where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
+
+Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
+Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+Along a vertical gradient within canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
+Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
+$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2020].
+Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020].
+However, considering sun and shade leaf differences within canopy (Table 1) $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a].
+Based on this evidence, we would expect leaves exposed to greater heat stress within canopy to have higher heat tolerance, but further research is required to fully understand relationship between heat tolerance, light exposure and microclimates.
 

### 2021-07-14 vertical-thermal-review.Rmd
Q10, maybe there is too much info here
Not sure if the last sentence goes here or else where, or if it is even needed?

@@
 
-A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas @griffinCanopyPositionAffects2002 found that in *Populas deltoides*, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy.
-Other studies have observed no difference in $Q_{10}$ within species vertical profiles and in seedling sun vs.
-shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008].
-@xuSeasonalVariationTemperature2006 found that the canopy-position effect gradient in respiration rates was much greater for Quercus rubra trees in mesic than in xeric sites, but $Q_{10}$ did not differ between canopy positions.
+Temperature sensitivity of respiration ($Q_{10}$) along a within canopy gradient remains uncertain where $Q_{10}$ can be constant within vertical profiles and in seedling sun vs.
+shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], and greater upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+Additionally, lower canopy leaf respiration have also been reported to be more sensitive to changes in $T_{air}$ than upper canopy [@griffinCanopyPositionAffects2002].
+Recent observations also suggest that tropical upper canopy leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical], and this may vary based on leaf age and leaf traits where in evergreen species, temperature responses of photosynthesis and $Q_{10}$ were greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
 
--In evergreen species, temperature responses of photosynthesis and temperature sensitivity of respiration ($Q_{10}$) have also been observed to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+Further, we would expect expect $Q_{10}$ to also vary with environmental conditions based on canopy-position effect gradient in respiration rates for Quercus rubra trees being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006] but further research is required to discern $Q_{10}$.
 

### 2021-07-15 vertical-thermal-review.Rmd
combined the two Tleaf versions
please let me know if there are parts to cut over here

@@
 
-Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
 Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
@@
 
-**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input through radiation versus heat lost to the environment [Fig.
+**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig.
 `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
-The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave), and high radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig.
+The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave).
+High radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig.
 `r fig_leaf_T`a).
-The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required convert water from liquid to vapour state cools the leaf. 
-Evaporation ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
+The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf.
+Evapotranspiration ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
 Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
 decreases with $g_s$ (Fig. `r fig_leaf_T`e);
-and decreases with with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).
+and decreases with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).
 
@@
 [ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/leaf_energy_fig3.png)
-
-**version 1 (for rest of biophyscial drivers subsection-- coauthors, please let us know which is better):**
-
-Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig.
-`r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $g_a$; *i.e.*, boundary layer conductance) and $\lambda E$ (via $g_s$ *ie.* stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].
-
-$g_a$ and $g_s$ are dominant factors in $T_{leaf}$ flux.
-$Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig.
-`r fig_leaf_T`b).
-Therefore, high windspeeds in upper canopies such as in NEON sites with windspeeds ~4 ms (Fig.
-`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig.
-`r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004].
-Low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low.
-Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig.
-`r fig_leaf_T`c) because reduced $ga$ stifles $E$ and $gs$ that dissapate heat [@songTreeSurfaceTemperature2020].
-
-Latent heat loss through maximum $g_s$, cools the leaf and decreases $T_{leaf}$-$T_{air}$, allowing coupling between leaf and air.
-However in conditions of low soil moisture during drought (Fig.
-`r fig_leaf_T`), greater temperature than $gs$ thermal optimum, and air dryness causing an increase in VPD, $gs$ becomes limited, thereby increasing $T_{leaf}$-$T_{air}$ [@bonan_ecological_2016].
-Limitation to $gs$ can also exaggerate effects of biophysical factors that contribute to energy inputs such as, solar radiation, humidity and $T_{air}$ (Fig.
-`r fig_leaf_T`)[@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @dongBiophysicalHomoeostasisLeaf2017].
-
-Leaf size and physiology [see next section] influence $g_a$, $g_s$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$.
-Smaller leaves have thinner boundary layer and higher $g_a$, that enables coupling with $T_{air}$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig.
-`r fig_leaf_T` a).
-Therefore, smaller leaves with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018].
-Additionally, under low raditation, large leaves can be cooler than small leaves due to greater $g_s$ [Fig.
-`r fig_leaf_T` d, @leighThickLeavesAvoid2012;@bonan_ecological_2016].
-Sun leaves, such as in the upper canopies, have greater $gs$, due to greater stomatal density (Table 1) and lower thermal capacitance due to lower water potential than shade leaves.
-However, exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021].
-However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in radiation or $T_{air}$, $T_{leaf}$-$T_{air}$ can easily become more elevated than sun leaves (Fig.
-`r fig_leaf_T`a)[@schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
-
-**Within forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
-$T_{leaf}$ generally tracks $T_{air}$, where vertical gradient of $T_{air}$ depends on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
-Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in sub canopies and the understory microclimate.
-The intensity of incoming radiation and $VPD$ in the upper canopy can limit $gs$, therefore, we would expect a stronger $T_{leaf}$ gradient in closed canopies where upper canopy leaves can be much warmer than below canopies layers [@millerOnlySunlitLeaves2021].
-However, higher windspeeds and lower RH (Fig.
-`r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig.
-`r fig_leaf_T`a) and higher rates of heat loss (Fig.
-`r fig_leaf_T`b-e) [@millerOnlySunlitLeaves2021].
-On the other hand, open canopies, or lower canopy $LAI$ allow more turbulent air mixing and sunlight into below canopies, that can either neutralize $T_{leaf}$ gradient or in the case of 'canopy greenhouse effect' or even sunflecks, can elevate below canopy $T_{leaf}$-$T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].
-
-**version 2 (for rest of biophyscial drivers subsection-- coauthors, please let us know which is better):**
 
@@
 `r fig_leaf_T`).
-However, higher windspeeds and lower RH (Fig.
+However, higher windspeeds such as in NEON sites with windspeeds ~4 ms (Fig.
+`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig.
+`r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004].
+Collectively, windspeed, lower RH (Fig.
 `r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig.
@@
 `r fig_leaf_T`b-e).
+
+**Vertical gradients**
 
@@
 `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
-Thus, understory leaves can remain cooler under lower radiation load, but their environment is not conducive to shedding excess heat, particularly under humid conditions, and their tendency towards lower leaf $g_s$ and thermal capacitance can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].
+Thus, understory leaves or shade leaves in a canopy can remain cooler under lower radiation load, but their environment is not conducive to shedding excess heat, particularly under humid conditions, and their tendency towards lower leaf $g_s$ and thermal capacitance (capacity to hold heat) can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In contrast, under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
@@
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
+
+**Vertical $T_{leaf}$ gradients can also vary with canopy structure (Fig. `r fig_leaf_T`).**
+$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
+Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
+The intensity of incoming radiation and $VPD$ in the upper canopy can limit $g_s$.
+Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021].
+However, higher windspeeds and lower $RH$ (Fig.
+2), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower leaf sensitivity to intercepted radiation (Fig.
+3a) and higher increase rates of heat loss (Fig.
+3b-e).
+In contrast, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves, as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].
 
@@
 
-Further, we would expect expect $Q_{10}$ to also vary with environmental conditions based on canopy-position effect gradient in respiration rates for Quercus rubra trees being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006] but further research is required to discern $Q_{10}$.
+Further, we would expect expect $Q_{10}$ to also vary with environmental conditions based on the evidence of canopy-position effect gradient in respiration rates for Quercus rubra trees being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
 

### 2021-07-15 vertical-thermal-review.Rmd
a bit of progress on ecology section

@@
 Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
-Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific metabolism (Table 2).
-Of the fixed C, a higher proportion is consumed by respiration, in part because of higher leaf respiration ($R_{dark}$, Table 2) and in part because of higher sapwood volume (??confirm??) [**REFS**; @westIncreasingRespiratoryCosts2020].
+Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
+The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
+The proportion of GPP allocaated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].
+As a result, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021]
 
-- C allocation --> respiration, woody growth, NSCs, foliage turnover, belowground allocation, reproduction
-- we don't know a whole lot about how C allocation scales with tree size
-- woody growth increases with DBH to a point, then declines [@AT_inReview].
 - NSCs increase , making smaller trees more sensitive to biological and biophyical stressors [@niinemetsResponsesForestTrees2010a]

### 2021-07-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-The proportion of GPP allocaated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].
-As a result, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021]
-
-- NSCs increase , making smaller trees more sensitive to biological and biophyical stressors [@niinemetsResponsesForestTrees2010a]
+The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].
+NSC reserves also increase with tree size, such that larger trees have more buffer against biological and biophyical stressors [@niinemetsResponsesForestTrees2010a].
+As a result of increasing C allocation to these other functions, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021]
 

### 2021-07-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 NSC reserves also increase with tree size, such that larger trees have more buffer against biological and biophyical stressors [@niinemetsResponsesForestTrees2010a].
-As a result of increasing C allocation to these other functions, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021]
+As a result of increasing C allocation to these other functions, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021].
 

### 2021-07-17 vertical-thermal-review.Rmd
some changes before sharing

@@
 
-**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Figs.
+**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by leaf area index ($LAI$), canopy height, canopy structure, and across species and forest types [Figs.
 `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].**
@@
 **Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all 22 forested NEON sites (Figs.
+Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all 22 forested NEON sites (Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
@@
 In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Maximum $T_{air}$ and VPD in the understory are further decreased with increasing canopy height and LAI [@juckerCanopyStructureTopography2018].
+Maximum $T_{air}$ and $VPD$ in the understory are further decreased with increasing canopy height and $LAI$ [@juckerCanopyStructureTopography2018].
 *( This last sentence needs work.)*
@@
 
+**Vertical gradients**
+
 **These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
@@
 However, higher windspeeds such as in NEON sites with windspeeds ~4 ms (Fig.
-`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig.
+`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ 4$^\circ$C (Fig.
 `r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004].
 Collectively, windspeed, lower RH (Fig.
-`r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower sensitivity to intercepted radiation (Fig.
+`r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give upper canopy lower sensitivity to intercepted radiation (Fig.
 `r fig_leaf_T`a) and higher rates of heat loss (Fig.
@@
 
-**Vertical gradients**
+**Vertical $T_{leaf}$ gradients can also vary with canopy structure (Fig. `r fig_leaf_T`).**
+$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
+Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
+The intensity of incoming radiation and $VPD$ in the upper canopy can limit $g_s$.
+Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021].
+In contrast, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves, as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].
 
@@
 
-**Vertical $T_{leaf}$ gradients can also vary with canopy structure (Fig. `r fig_leaf_T`).**
-$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
-Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
-The intensity of incoming radiation and $VPD$ in the upper canopy can limit $g_s$.
-Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021].
-However, higher windspeeds and lower $RH$ (Fig.
-2), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give them lower leaf sensitivity to intercepted radiation (Fig.
-3a) and higher increase rates of heat loss (Fig.
-3b-e).
-In contrast, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves, as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].
-
 ### In situ observations
@@
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
-Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
+Greater $RH$ in inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
 
@@
 For example, LMA, photosynthetic capacity (per area) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
-Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical; @niinemetsResponsesForestTrees2010a].
+Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical 2021; @niinemetsResponsesForestTrees2010a].
 
@@
 2017; @kumarathungeAcclimationAdaptationComponents2019a), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5) in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderatly, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
 Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+In contrast, tropical experimental warming observations also show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy [@carterExperimentalWarmingTropical 2021].
 
@@
 shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], and greater upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
-Additionally, lower canopy leaf respiration have also been reported to be more sensitive to changes in $T_{air}$ than upper canopy [@griffinCanopyPositionAffects2002].
-Recent observations also suggest that tropical upper canopy leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical], and this may vary based on leaf age and leaf traits where in evergreen species, temperature responses of photosynthesis and $Q_{10}$ were greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+Additionally, lower canopy leaf respiration has also been reported to be more sensitive to changes in $T_{air}$ than upper canopy [@griffinCanopyPositionAffects2002].
+Recent observations also suggest that tropical upper canopy leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical], and this may vary based on leaf age and leaf traits where in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
 
-Further, we would expect expect $Q_{10}$ to also vary with environmental conditions based on the evidence of canopy-position effect gradient in respiration rates for Quercus rubra trees being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
+Further, we would expect $Q_{10}$ to also vary with environmental conditions based on the evidence of canopy-position effect gradient in respiration rates for *Quercus rubra* trees being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
 
@@
 Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
 This imposes an increasing threat to the biodiversity in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
@@
 However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [@keenanWidespreadInhibitionDaytime2019].
-However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Templer et al.].
+However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pam Templer et al.].
 Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].

### 2021-07-17 vertical-thermal-review.Rmd
some changes before sharing
small edits

@@
 In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig.
-`r fig_NEON_vertical`a, @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
+`r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
 Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs.
-open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016].
+open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019).
@@
 
-**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by leaf area index ($LAI$), canopy height, canopy structure, and across species and forest types [Figs.
+**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Figs.
 `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].**
@@
 **Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open savannas [@curtis_intracanopy_2019], including all 22 forested NEON sites (Figs.
+Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open forests and savannas [@curtis_intracanopy_2019; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
@@
 
-**Atmospheric concentrations carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
+**Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
 Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated $CO_{2}$ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
@@
 
-Relative humidity, $RH$, tends to be higher in the understory and decreases with height across the vertical profile, although this trend is absent in open forests [Figs.
+Relative humidity, RH, tends to be higher in the understory and decreases with height across the vertical profile, although this trend is absent in open forests [Figs.
 `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregor_tree_nodate].
-Dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a].
-In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Maximum $T_{air}$ and $VPD$ in the understory are further decreased with increasing canopy height and $LAI$ [@juckerCanopyStructureTopography2018].
+Dense-canopy forests maintain higher daily maximum RH below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
+In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Maximum $T_{air}$ and VPD in the understory are further decreased with increasing canopy height and LAI [@juckerCanopyStructureTopography2018].
 *( This last sentence needs work.)*
@@
 Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@frenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Notably, this pattern can be reversed in open canopy forests or sparse stands, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.
+Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.
 `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
@@
 
-**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig.
+**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based via biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig.
 `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
@@
 `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads, but during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
-With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
+With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
 
@@
 
-**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
+**These basic biophysical principles shape the propensities for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
 High radiation at the top of the vertical profile (Fig.
@@
 `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ 4$^\circ$C (Fig.
-`r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004].
+`r fig_leaf_T`b)[@bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 Collectively, windspeed, lower RH (Fig.
@@
 Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021].
-In contrast, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves, as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].
+In contrast, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat), as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].
 
@@
 `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
-Thus, understory leaves or shade leaves in a canopy can remain cooler under lower radiation load, but their environment is not conducive to shedding excess heat, particularly under humid conditions, and their tendency towards lower leaf $g_s$ and thermal capacitance (capacity to hold heat) can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
+Thus, understory leaves and within canopy shade leaves can remain cooler under lower radiation load, but their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In contrast, under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
@@
 ### In situ observations
-**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations.**
+**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in situ $T_{leaf}$ observations.**
 Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy.
@@
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
-Greater $RH$ in inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
+Greater $RH$ in inner canopy can also increase $T_{leaf}$ relatively more than upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
 
@@
 Caratenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin ($VAZ$) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-This is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
+The dissapation of excess light energy is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
 
@@
 2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
-*In evergreen species,* sun leaves have shorter lifespans than shade leaves, and thus upper canopy leaves are on average younger than subcanopy leaves, therefore they operate closer to their photosynthetic optimum.
+*In evergreen species,* sun leaves have shorter lifespans than shade leaves; upper canopy leaves are on average younger than subcanopy leaves, therefore operating closer to their photosynthetic optimum.
 However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, less damage  accumulates [@millerOnlySunlitLeaves2021].
-Once leaves leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions between species, @niinemetsLeafAgeDependent2016].
+Once leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions between species, @niinemetsLeafAgeDependent2016].
 Additionally, because older leaves reach maximum $T_{leaf}$ sooner than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015] **this needs some confirming by co-authors**.
@@
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g.
-@slotSituTemperatureResponse2017; Smith et al.
-2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; Sage & Kubien 2007; Varhammer et al.
+@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; Sage & Kubien 2007; Varhammer et al.
 2015).
@@
 14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderatly, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
+However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
 Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
-In contrast, tropical experimental warming observations also show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy [@carterExperimentalWarmingTropical 2021].
+In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy [@carterExperimentalWarmingTropical 2021].
+Further research is required to discern temperature sensitivity at a vertical gradient.
 
@@
 Temperature sensitivity of respiration ($Q_{10}$) along a within canopy gradient remains uncertain where $Q_{10}$ can be constant within vertical profiles and in seedling sun vs.
-shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], and greater upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], and greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
 Additionally, lower canopy leaf respiration has also been reported to be more sensitive to changes in $T_{air}$ than upper canopy [@griffinCanopyPositionAffects2002].
-Recent observations also suggest that tropical upper canopy leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical], and this may vary based on leaf age and leaf traits where in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+Recent observations also suggest that tropical upper canopy leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical]; this may vary based on leaf age and leaf traits, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
 
-Further, we would expect $Q_{10}$ to also vary with environmental conditions based on the evidence of canopy-position effect gradient in respiration rates for *Quercus rubra* trees being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
+Further, we would expect $Q_{10}$ to also vary with soil moisture based on the evidence of canopy-position effect gradients in respiration rates for *Quercus rubra* being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
 

### 2021-07-18 vertical-thermal-review.Rmd
updating the summary section

@@
 Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
-Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree demography and ecosystem function.
-Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves, particularly in forests with dense canopies.
-However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
-Resulting $T_{leaf}$'s exceed air temperature ($T_{air}$) under conditions of high solar radiation, low wind, and low stomatal conductance.
-Differences are most pronounced when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become dramatically elevated above both $T_{air}$ and understory $T_{leaf}$.
-While leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
-The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
-Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
-In the meantime, our findings imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.
+Here, we synthesize evidence as to how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function.
+In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and are therefore at risk of elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
+However, foliar traits relevant to shaping $T_{leaf}$ also vary strongly across height or light gradients, partially offsetting but not eliminating the propensity towards elevated $T_{leaf}$ in the upper canopy.
+Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher metabolism and growth at both individual and ecosystem levels.
+Yet, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
+The implications for tree performance are mixed: whereas larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and appear to experience relatively greater stress under hot and humid conditions.
+Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients and scaling through space and time.
+In the meantime, our findings highlight the vulnerability of large canopy trees to warming combined with drought, but also imply high vulnerability of understory trees when the buffering provided by large trees is broken, and potentially under hot and wet conditions.
 

### 2021-07-19 vertical-thermal-review.Rmd
small edit

@@
 Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilitiesa 2021].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilitiesa; @mcdowellPervasiveShiftsForest2020].
 Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
@@
 [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
-](schematics/vertical thermal sensitivity image.png)
+](schematics/Screen Shot 2021-07-19 at 1.10.17 PM.png)
 
@@
 Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**.
-Shown are  height profiles in growing season (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+Shown are  height profiles in growing season (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
 Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
@@
 Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated $CO_{2}$ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Elevated $CO_{2}$ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are neglible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
+Elevated $CO_{2}$ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
 
@@
 
-Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020].
+Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
 Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
@@
 `r fig_leaf_T`).
-However, higher windspeeds such as in NEON sites with windspeeds ~4 ms (Fig.
+However, higher windspeeds such as in NEON sites with windspeeds ~4 m s^-1^ (Fig.
 `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ 4$^\circ$C (Fig.
@@
 **Vertical $T_{leaf}$ gradients can also vary with canopy structure (Fig. `r fig_leaf_T`).**
-$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and $LAI$ [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
-Closed canopies, and high $LAI$ canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
-The intensity of incoming radiation and $VPD$ in the upper canopy can limit $g_s$.
+$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and LAI [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
+Closed canopies, and high LAI canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
+The intensity of incoming radiation and VPD in the upper canopy can limit $g_s$.
 Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021].
-In contrast, open canopies, or lower canopy $LAI$ allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat), as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b].
+In contrast, open canopies, or lower canopy LAI allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].
 
-**Within closed-canopy forests, vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
+**Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorological conditions (Fig. `r fig_leaf_T`).**
 Under hot conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
@@
 Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy.
-In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high radiation loads [@carterExperimentalWarmingTropical; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
@@
 
-Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, by being smaller, hairier, and with higher vein density than shade leaves [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, being smaller, hairier, and with higher vein density than shade leaves [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits (Table 1) collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
@@
 Caratenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin ($VAZ$) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-The dissapation of excess light energy is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
+The dissipation of excess light energy is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
 
@@
 2015).
-Irresversible damage to photosystem II, leading to leaf necrosis and eventually leaf death occurs at $T_{leaf}$ of $40-60^{\circ}C$
+rreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death occurs at $T_{leaf}$ of $40-60^{\circ}C$
 [Baker 2008; @feeley_thermal_2020].
@@
 
-*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C (O' sullivan et al.
-2017), where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
+*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C [@osullivanThermalLimitsLeaf2017], where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
 

### 2021-07-20 vertical-thermal-review.Rmd
update on gs and Topt

@@
 In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
+In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
 `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020 (in review)].
 Greater $RH$ in inner canopy can also increase $T_{leaf}$ relatively more than upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
@@
 Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves (Table 2).
-However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is prevalent in sun leaves but not in shade leaves.
+However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is more prevalent in sun leaves than shade leaves in closed forests.
 $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-Higher maximum temperatures and $g_s$ influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
-Additionally, $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], but sun leaves in the upper canopy show a stronger decrease in $gs$ to rising temperature than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
+In more open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer time duration than upper canopy, perhaps due to higher RH [@zweifelMiddayStomatalClosure2002].
+Additionally, while $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $gs$ to rising temperature than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
 
@@
 Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons (e.g.
-Tan et al.
-2017; @kumarathungeAcclimationAdaptationComponents2019a), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
+@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
 14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
@@
 Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
-In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy [@carterExperimentalWarmingTropical 2021].
-Further research is required to discern temperature sensitivity at a vertical gradient.
+In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater $g_s$ thermal sensitivity of upper canopy sun leaves [@carterExperimentalWarmingTropical 2021].
+Higher maximum temperatures, and $g_s$ influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
 
-@hernandezSimilarTemperatureDependence2020 speculated that the lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis.
+@hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis.
 Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of mid-day stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
@@
 Temperature sensitivity of respiration ($Q_{10}$) along a within canopy gradient remains uncertain where $Q_{10}$ can be constant within vertical profiles and in seedling sun vs.
-shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], and greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
+shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], and greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
 Additionally, lower canopy leaf respiration has also been reported to be more sensitive to changes in $T_{air}$ than upper canopy [@griffinCanopyPositionAffects2002].
-Recent observations also suggest that tropical upper canopy leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical]; this may vary based on leaf age and leaf traits, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical]; this may vary based on leaf age and leaf traits, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
 

### 2021-07-21 vertical-thermal-review.Rmd
small spelling edits

@@
 **Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open forests and savannas [@curtis_intracanopy_2019; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
+Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
@@
 **Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
-Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated $CO_{2}$ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Elevated $CO_{2}$ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
+Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
+Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
 
@@
 `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
 With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
@@
 Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021].
-In contrast, open canopies, or lower canopy LAI allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].
+In contrast, open canopies, or lower canopy LAI allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].
 
@@
 
-It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems; for instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].
+It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems; for instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021].
 Thus, the vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.
@@
 Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
-Within-canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
-Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity,  @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.
+Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
+Therefore, we would expect elemental traits to influence vertical foliar thermal sensitivity,  however, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.
 
@@
 The timing of seasonal leaf loss or gain is influenced by vertical microclimate gradients, and in turn can alter biophysical and $T_{leaf}$ gradients in the forest [@rey-sanchez_spatial_2016].
-Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests. 
+Deciduousness in tropical forests is an adaptation to conserve water that is more common in dry than wet tropical forests. 
 Within tropical deciduous species, the fraction of deciduous individuals tends to increase across tree size classes, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
@@
 Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
-In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater $g_s$ thermal sensitivity of upper canopy sun leaves [@carterExperimentalWarmingTropical 2021].
-Higher maximum temperatures, and $g_s$ influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
+In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in the upper canopy [@carterExperimentalWarmingTropical 2021].
+Higher maximum temperatures, and $g_s$ are known influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
 

### 2021-07-21 vertical-thermal-review.Rmd
small spelling edits

@@
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g.
-@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; Sage & Kubien 2007; Varhammer et al.
-2015).
+@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020 [temperate/boreal refs needed), and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015].
 rreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death occurs at $T_{leaf}$ of $40-60^{\circ}C$
@@
 NSC reserves also increase with tree size, such that larger trees have more buffer against biological and biophyical stressors [@niinemetsResponsesForestTrees2010a].
-As a result of increasing C allocation to these other functions, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021].
+As a result of increasing C allocation to these other functions, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021; @anderson-teixeiraCarbonCyclingMature2021].
 
@@
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory].
-Similarly, in tropical forests higher night-time temperatures have been associated with decelerating tree growth rates predominantly in smaller trees *but also in taller trees* [*Clark 2010; 2013; Feeley et al.
-2007*].
+Similarly, in tropical forests higher night-time temperatures have been associated with decelerating tree growth rates predominantly in smaller trees *but also in taller trees* [@clarkTropicalRainForest2003; @clarkAnnualWoodProduction2010; @feeleyDeceleratingGrowthTropical2007b; @vlamTemperatureRainfallStrongly2014*].
 Potential mechanisms for these patterns include increases in nighttime respiratory carbon loss that disproportionately reduced carbon availability for growth among small trees, or that understory $g_s$  might be stifled under humid conditions, resulting in higher $T_{leaf}$ (Fig.

### 2021-07-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-In more open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer time duration than upper canopy, perhaps due to higher RH [@zweifelMiddayStomatalClosure2002].
-Additionally, while $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $gs$ to rising temperature than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
+In more open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer time duration than upper canopy resulting in higher $T_{leaf}$, perhaps due to higher RH [Fig.
+`r fig_leaf_T`,@zweifelMiddayStomatalClosure2002].
+Additionally, while $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ to rising temperature than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
 
@@
 **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
-Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons (e.g.
-@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
+Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons [e.g.
+@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
 14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
@@
 Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
-In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in the upper canopy [@carterExperimentalWarmingTropical 2021].
-Higher maximum temperatures, and $g_s$ are known influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
+In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical 2021].
+Higher maximum temperatures, and $g_s$ are known to influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
 
@@
 
-$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [@slotLeafHeatTolerance2020].
+$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorophyll a fluorescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [@slotLeafHeatTolerance2020].
 
@@
 However, considering sun and shade leaf differences within canopy (Table 1) $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a].
-Based on this evidence, we would expect leaves exposed to greater heat stress within canopy to have higher heat tolerance, but further research is required to fully understand relationship between heat tolerance, light exposure and microclimates.
+Based on this evidence, we would expect leaves exposed to greater heat stress within the canopy to have higher heat tolerance, but further research is required to fully understand the relationship between heat tolerance, light exposure and microclimates.
 

### 2021-07-22 vertical-thermal-review.Rmd
incorporating Krista's google doc changes
will get back to google doc comments tomorrow

@@
 The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].
-NSC reserves also increase with tree size, such that larger trees have more buffer against biological and biophyical stressors [@niinemetsResponsesForestTrees2010a].
-As a result of increasing C allocation to these other functions, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021; @anderson-teixeiraCarbonCyclingMature2021].
+The proportion of GPP allocated to many of these increases with tree size, including ...
+foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].As a result of increasing C allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021; @anderson-teixeiraCarbonCyclingMature2021].
+However, NSC reserves increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].
 
 **Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
-Across biomes, including mesic forests, surface conductance and evapotranspiration tend to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-In terms of transpiration and photosynthesis, the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have lower rates during periods of high- relative to low- atmospheric demand -- be these  on the time scale of hours (**REFS?**), days [@anderson-teixeiraSizerelatedScalingTree2015], seasons [@christoffersenLinkingHydraulicTraits2016], or years (**?
+Across biomes, including mesic forests, surface conductance and evapotranspiration tend to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016], the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have greater reductions in whole-tree transpiration and photosynthesis during periods of high atmospheric demand - be these on the time scale of hours (**REFS?**), days [@anderson-teixeiraSizerelatedScalingTree2015], seasons [@christoffersenLinkingHydraulicTraits2016], or years (**?
 REFS?**).
@@
 
-There are, however, noteable exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
+There are, however, notable exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory].
@@
 `r fig_leaf_T`) and lower photosynthesis.
-It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. 
+It is also possible that higher temperatures reduce understory tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. 
 Additional research is required to understand these mechanisms and to disentangle tree growth responses to hot-wet versus hot-dry conditions.

### 2021-07-26 vertical-thermal-review.Rmd
fixed highlighted citation

@@
 
-**Authors (so far, not necessarily in final order):** 
+**Authors:** 
 Nidhi Vinod^1^,
@@
 
-*(currently too long for NewPhyt)*
+(200 word limit, no bullet points )
 
-Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
-Here, we synthesize evidence as to how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function.
+Clean version implementing the changes below (198 words):
+
+Rising temperatures are having profound effects on forests; however, it is not well understood how responses vary across forest strata.
+Here, we review how environmental conditions, leaf temperatures,  and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function.
 In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and are therefore at risk of elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
-However, foliar traits relevant to shaping $T_{leaf}$ also vary strongly across height or light gradients, partially offsetting but not eliminating the propensity towards elevated $T_{leaf}$ in the upper canopy.
+However, foliar traits relevant to shaping $T_{leaf}$ also vary strongly across height or light gradients, partially offsetting the propensity towards elevated $T_{leaf}$ in the upper canopy.
 Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher metabolism and growth at both individual and ecosystem levels.
-Yet, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
-The implications for tree performance are mixed: whereas larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and appear to experience relatively greater stress under hot and humid conditions.
-Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients and scaling through space and time.
-In the meantime, our findings highlight the vulnerability of large canopy trees to warming combined with drought, but also imply high vulnerability of understory trees when the buffering provided by large trees is broken, and potentially under hot and wet conditions.
+Yet, differences in thermal sensitivity and damage thresholds are modest.
+The implications for tree performance are mixed: whereas larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may be stifled under hot, humid conditions.
+As warming continues, large canopy trees will be particularly vulnerable to drought, whereas understory trees will suffer most when the buffering provided by large trees is broken.
 
@@
 Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
-The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [REFS].
+The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006a].
 
@@
 (2) leaf temperature ($T_{leaf}$);
-(3) the leaf traits that most strongly influence leaf temperature and metabolism;
+(3) the leaf traits that most strongly influence $T_{leaf}$ and metabolism;
 (4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
 (5) size-structuring of whole-tree and ecosystem function.
-*We then consider the implications for understanding forest responses to global change, including scaling across space and time.*
+*We then consider the implications for understanding forest responses to global change, including how these scale across space and time.*
 
@@
 open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019).
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019].
 In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.

### 2021-07-26 vertical-thermal-review.Rmd
including edits from google doc

@@
 open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019].
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019].
 In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.
 
-**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Figs.
+**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs.
 `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].**
 This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`d), than under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
+`r fig_NEON_vertical`d), then under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
 `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
@@
 **Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
+Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
@@
 Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
+Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
 
 Relative humidity, RH, tends to be higher in the understory and decreases with height across the vertical profile, although this trend is absent in open forests [Figs.
-`r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregor_tree_nodate].
+`r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
 Dense-canopy forests maintain higher daily maximum RH below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
 In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Maximum $T_{air}$ and VPD in the understory are further decreased with increasing canopy height and LAI [@juckerCanopyStructureTopography2018].
-*( This last sentence needs work.)*
 
@@
 
-**The buffering patterns described above vary geographically, seasonally, and across forests types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, @frenneForestMicroclimatesClimate2021].**
+**The buffering patterns described above vary geographically, seasonally, and across forest types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, @frenneForestMicroclimatesClimate2021].**
 The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture, and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019].
@@
 Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
-While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$, it is rarely exactly equal [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+While $T_{leaf}$ is rarely exactly equal to $T_{air}$ (poikilothermy), it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
 
@@
 `r fig_leaf_T`a).
-The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf.
+Generally, the dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf.
 Evapotranspiration ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
@@
 decreases with $g_s$ (Fig. `r fig_leaf_T`e);
-and decreases with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).
+and decreases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
 

### 2021-07-26 vertical-thermal-review.Rmd
including edits from google doc

@@
 When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
-However, when inadequate water supply -- for example under drought -- causes leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
+However, when leaf water demand (shaped by VPD and photosynthetic CO2 demand) exceeds the rate of supply, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, during $g_s$- limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
 With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

### 2021-07-26 vertical-thermal-review.Rmd
including edits from google doc

@@
 ![**Figure `r fig_schematic`.
-Schematic summarizing vertical gradients in forests with dense canopies during daytime growing season conditions.** Patterns summarized here tend to be weaker, or sometimes reversed, under more open canopies, during seasons of deciduous leaf loss, or at nighttime.
+Schematic summarizing vertical gradients in forests with dense canopies during daytime growing season conditions.** Patterns summarized here tend to be weaker, or sometimes reversed, under more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
 [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
@@
 Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a) -- a pattern common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
+`r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
 In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
@@
 This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`d), then under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
+`r fig_NEON_vertical`d), than under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
 `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
@@
 Dense-canopy forests maintain higher daily maximum RH below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
-In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].Especially when canopy height and leaf area index are large, maximum Tair and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
 
@@
 The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [@frenneForestMicroclimatesClimate2021].
-Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@frenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+Typically, dense canopies buffer understories from high maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@frenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
 Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.
@@
 `r fig_leaf_T`).
-However, higher windspeeds such as in NEON sites with windspeeds ~4 m s^-1^ (Fig.
-`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ 4$^\circ$C (Fig.
-`r fig_leaf_T`b)[@bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Collectively, windspeed, lower RH (Fig.
-`r fig_NEON_vertical`), along with adaptive leaf traits that reduce $g_a$ and increase $g_s$ (reviewed below), give upper canopy lower sensitivity to intercepted radiation (Fig.
+However, higher wind speeds (e.g., up to ~4 $ms^{-1}$ at NEON sites; Fig.
+`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ 4$^\circ$C [Fig.
+`r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Collectively, higher windspeed (Fig.
+`r fig_NEON_vertical`), and with adaptive leaf traits increase $g_a$ and $g_s$ (reviewed below) give upper canopy leaves lower sensitivity to intercepted radiation (Fig.
 `r fig_leaf_T`a) and higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
+~In contrast, greater RH and lower wind speeds in the understory may stifle ??E(through reduced gs and ga), thereby increasing Tleaf (Fig.
+3, Perez & Feeley, 2018; Tibbitts, 1979).~
 

### 2021-07-27 vertical-thermal-review.Rmd
incorporating google doc edits and citations

@@
 `r fig_leaf_T`).
-However, higher wind speeds (e.g., up to ~4 $ms^{-1}$ at NEON sites; Fig.
+However, higher wind speeds (e.g., up to ~ 4$ms^{-1}$ at NEON sites; Fig.
 `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ 4$^\circ$C [Fig.
 `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Collectively, higher windspeed (Fig.
+Collectively, higher wind speed (Fig.
 `r fig_NEON_vertical`), and with adaptive leaf traits increase $g_a$ and $g_s$ (reviewed below) give upper canopy leaves lower sensitivity to intercepted radiation (Fig.
@@
 `r fig_leaf_T`b-e).
-~In contrast, greater RH and lower wind speeds in the understory may stifle ??E(through reduced gs and ga), thereby increasing Tleaf (Fig.
-3, Perez & Feeley, 2018; Tibbitts, 1979).~
+In contrast, greater RH and lower wind speeds in the understory may stifle $\lambda E$ (through reduced $g_s$ and $g_a$), thereby increasing $T_{leaf}$ [Fig.
+`r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].
 
-**Vertical $T_{leaf}$ gradients can also vary with canopy structure (Fig. `r fig_leaf_T`).**
+**Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).**
 $T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and LAI [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
 Closed canopies, and high LAI canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
-The intensity of incoming radiation and VPD in the upper canopy can limit $g_s$.
-Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021].
-In contrast, open canopies, or lower canopy LAI allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].
+The intensity of incoming radiation and VPD in the upper canopy, combined with hydraulic challenges of water transport to the tops of tall trees [**REFS**], can cause $g_s$-limitation.
+Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than lower canopies layers [@millerOnlySunlitLeaves2021].
+In contrast, open canopies (lower canopy LAI) allow more vertical air mixing and sunlight into the understory, and this can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ of the lower canopy more than that of the upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].
 
 **Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorological conditions (Fig. `r fig_leaf_T`).**
-Under hot conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
+Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
+Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than that of shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig.
+`r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
+In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
 `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020].
-Thus, understory leaves, which typically experience lower windspeeds (Fig. `r fig_NEON_vertical`), would retain more heat.
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation load, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
+Understory leaves, which typically experience lower windspeeds (Fig. `r fig_NEON_vertical`), would retain more heat.
 The higher-RH conditions of the understory (Fig.
 `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig.
-`r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
-Thus, understory leaves and within canopy shade leaves can remain cooler under lower radiation load, but their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
-In contrast, under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
-Because sun leaves are exposed to higher irradiance and $VPD$, their $g_s$-limitation is steeper that that of shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig.
-`r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
+`r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].
 
 ### In situ observations
-**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in situ $T_{leaf}$ observations.**
-Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy.
+
+**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
+Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* and therefore differs little between understory and canopy.
 In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
-`r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016].
+`r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020 (in review)].
@@
 Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-As much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al.
-2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses through the vertical canopy.
+As much of our understanding of trait coordination is based on the leaf economics spectrum concept (@wrightWorldwideLeafEconomics2004) which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses through the vertical canopy.
 It has been established that independent of light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
@@
 
-Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, being smaller, hairier, and with higher vein density than shade leaves [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
-These traits (Table 1) collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
+Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, by being smaller, hairier, and with higher vein density than shade leaves [Table 1, @leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
 In contrast, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks (e.g., lower LMA), but are more likely to get overheated than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
@@
 
+Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
+Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
+Therefore, we would expect elemental traits to influence vertical foliar thermal sensitivity;  however, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.
+
 Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
 Caratenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin ($VAZ$) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-The dissipation of excess light energy is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
-
-Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
-Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
-Therefore, we would expect elemental traits to influence vertical foliar thermal sensitivity,  however, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.
-
-Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes [*REF*].
+The dissipation of excess light energy is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes [**REF**].
 
@@
 However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, less damage  accumulates [@millerOnlySunlitLeaves2021].
-Once leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions between species, @niinemetsLeafAgeDependent2016].
+Once leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions, @niinemetsLeafAgeDependent2016].
 Additionally, because older leaves reach maximum $T_{leaf}$ sooner than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015] **this needs some confirming by co-authors**.
@@
 The timing of seasonal leaf loss or gain is influenced by vertical microclimate gradients, and in turn can alter biophysical and $T_{leaf}$ gradients in the forest [@rey-sanchez_spatial_2016].
-Deciduousness in tropical forests is an adaptation to conserve water that is more common in dry than wet tropical forests. 
+Deciduousness in tropical forests is an adaptation to conserve water, and is more common in drier forests and in canopy species [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
 Within tropical deciduous species, the fraction of deciduous individuals tends to increase across tree size classes, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].
 
@@
 
-Therefore, $T_{leaf}$ patterns within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
+Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
 
@@
 Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves (Table 2).
-However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is more prevalent in sun leaves than shade leaves in closed forests.
+However, because of the high VPD and irradiance experienced by sun leaves,  mid-day stomatal depression is more prevalent in sun leaves than shade leaves in closed forests.
 $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
@@
 `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002].
-Additionally, while $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ to rising temperature than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
+The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ to rising $T_{air}$ than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
 
@@
 2020).**
-This is primarily driven by the greater light available to sun-leaves, that also have traits (Table 1, Table 2) allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions.
+This is primarily driven by the greater light available to sun-leaves, which also have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions (Table 1, Table 2).
 Photosynthesis increases with irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
@@
 
-Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increased $VPD$ causing stomatal closure [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to temperature, with the peak commonly found to correspond to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017].
+Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
+Photosynthesis has a peaked response to $T_{air}$, with the peak commonly found to correspond to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017].
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g.
 @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020 [temperate/boreal refs needed), and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015].
-rreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death occurs at $T_{leaf}$ of $40-60^{\circ}C$
-[Baker 2008; @feeley_thermal_2020].
 
@@
 Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons [e.g.
-@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig.
-14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
-In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical 2021].
-Higher maximum temperatures, and $g_s$ are known to influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
+@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
+Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical 2021].
+Higher maximum temperatures, and $g_s$ are known to influence leaf thermal sensitivity [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
 
@@
 
-**Leaf photosynthetic thermal tolerance (Tcrit/ T50)**
-
+**Leaf photosynthetic thermal tolerance**
+Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [Baker 2008; @feeley_thermal_2020]
 $T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorophyll a fluorescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [@slotLeafHeatTolerance2020].

### 2021-07-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Leaf photosynthetic thermal tolerance**
-Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [Baker 2008; @feeley_thermal_2020]
-$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorophyll a fluorescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [@slotLeafHeatTolerance2020].
-
-**Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
-
-*If we take geographic gradients as a reliable proxy, we'd expect temperature sensitivities to reflect the environment to which the trees are adapted/ acclimatized.* Global $T_{50s}$ across latitude and climate range from ~45-50$^\circ$C [@osullivanThermalLimitsLeaf2017], where $T_{50}$ and $T_{crit}$ decrease with elevation among tropical forests and across latitudes [@feeley_thermal_2020;  @slotLeafHeatTolerance2020; @osullivanThermalLimitsLeaf2017].
+Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020].Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2020].
+Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
 

### 2021-07-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
-Along a vertical gradient within canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019].
+Along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
 Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
-$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2020].
+$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotPhotosyntheticPlasticityTropical2021].
 Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020].
-However, considering sun and shade leaf differences within canopy (Table 1) $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a].
-Based on this evidence, we would expect leaves exposed to greater heat stress within the canopy to have higher heat tolerance, but further research is required to fully understand the relationship between heat tolerance, light exposure and microclimates.
+However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a].
+We would expect leaves exposed to greater heat stress within the canopy to have higher heat tolerance, but further research is required to fully understand the relationship between heat tolerance, light exposure and microclimates.
 
@@
 
-**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [@chenLeafEconomicsSpectrum2020], but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**
-
-Temperature sensitivity of respiration ($Q_{10}$) along a within canopy gradient remains uncertain where $Q_{10}$ can be constant within vertical profiles and in seedling sun vs.
-shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], and greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996].
-Additionally, lower canopy leaf respiration has also been reported to be more sensitive to changes in $T_{air}$ than upper canopy [@griffinCanopyPositionAffects2002].
+**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
+shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
 Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical]; this may vary based on leaf age and leaf traits, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].

### 2021-07-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
-shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
-Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical]; this may vary based on leaf age and leaf traits, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
-
+shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].This may vary based on forest type [**REF**], leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical].
 Further, we would expect $Q_{10}$ to also vary with soil moisture based on the evidence of canopy-position effect gradients in respiration rates for *Quercus rubra* being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
@@
 
-### Tree metabolism, growth, and survival
+### Tree metabolism and growth, and survival
 

### 2021-07-27 vertical-thermal-review.Rmd
de frenne citation fixed

@@
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
-The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [@frenneForestMicroclimatesClimate2021].
-Typically, dense canopies buffer understories from high maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@frenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [@defrenneForestMicroclimatesClimate2021].
+Typically, dense canopies buffer understories from high maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
 Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.

### 2021-07-27 vertical-thermal-review.Rmd
mission et al. 2007 added

@@
 The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [@defrenneForestMicroclimatesClimate2021].
-Typically, dense canopies buffer understories from high maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+Typically, dense canopies buffer understories from high maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
 Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.

### 2021-07-27 vertical-thermal-review.Rmd
updated

@@
 
-**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
-shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].This may vary based on forest type [**REF**], leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
-Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical].
+**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
+shade leaves [@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
+This may vary based on forest type [**REF**], leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical].
 Further, we would expect $Q_{10}$ to also vary with soil moisture based on the evidence of canopy-position effect gradients in respiration rates for *Quercus rubra* being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
@@
 **Vertical gradients affect tree metabolism and growth.**
-Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
-Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
+Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
+Specifically, increases are linked to increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
 The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
@@
 **Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
-Across biomes, including mesic forests, surface conductance and evapotranspiration tend to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016], the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have greater reductions in whole-tree transpiration and photosynthesis during periods of high atmospheric demand - be these on the time scale of hours (**REFS?**), days [@anderson-teixeiraSizerelatedScalingTree2015], seasons [@christoffersenLinkingHydraulicTraits2016], or years (**?
-REFS?**).
+Because $g_s$ tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016], the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have greater reductions in whole-tree transpiration and photosynthesis during periods of high atmospheric demand - be these on time scale of hours, days, or seasons [@christoffersenLinkingHydraulicTraits2016; Garcia et al.
+2021],consistent with the observation that  $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2).
+More active stomatal regulation of tall canopy trees (e.g., Mediavilla & Escudero 2004)-- often combined with greater effective rooting depth--may offset greater xylem embolism risk (Olson et al.
+2018, Garcia et al.
+2021, Chitra-Tarak et al.
+2021) and lower capacity to adjust traits related to carbon metabolism to drought (Bartholomew et al., 2020).
+This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height (Table 1; McDowell 2011).
 
@@
 
-In contrast, greater $RH$ can do the opposite, stifling evapotranspiration and thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].
-
-The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments and sometimes forest census data.
+The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments.
 While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions (**REFS--KAT**), only a relatively limited number of studies have directly examined drought- or temperature- sensitivities as a function of tree size.

### 2021-07-27 vertical-thermal-review.Rmd
updated

@@
 The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments.
-While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions (**REFS--KAT**), only a relatively limited number of studies have directly examined drought- or temperature- sensitivities as a function of tree size.
-These have generally found greater sensitivity to low precipitation or high temperatures among large/ canopy trees [**Fig.
+While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions (**REFS--KAT**), only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
+These have generally found greater sensitivity to low precipitation or high summer temperatures among larger, more exposed trees [**Fig.
 `r fig_treerings`**, @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021] , corroborating evidence from forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].

### 2021-07-27 vertical-thermal-review.Rmd
Ecology section updates

@@
 **Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
-Because $g_s$ tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and $VPD$, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016], the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have greater reductions in whole-tree transpiration and photosynthesis during periods of high atmospheric demand - be these on time scale of hours, days, or seasons [@christoffersenLinkingHydraulicTraits2016; Garcia et al.
-2021],consistent with the observation that  $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2).
-More active stomatal regulation of tall canopy trees (e.g., Mediavilla & Escudero 2004)-- often combined with greater effective rooting depth--may offset greater xylem embolism risk (Olson et al.
-2018, Garcia et al.
-2021, Chitra-Tarak et al.
-2021) and lower capacity to adjust traits related to carbon metabolism to drought (Bartholomew et al., 2020).
-This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height (Table 1; McDowell 2011).
-
-- Taller trees in the Amazon had greater stomatal sensitivity to drought, resulting in greater reductions transpiration and photosynthesis [@garciaImportanceHydraulicStrategy2021].
-- @christoffersenLinkingHydraulicTraits2016 uses trait-based modeling to show that canopy trees have lower dry-season: wet-season transpiration. 
-- Smaller trees also have greater capacity to adjust traits realted carbon metabolism to drought [@bartholomewSmallTropicalForest2020]
-- *(it would be good to find a relevant study for open forests-- perhaps track down studies mentioned by Bev Law)*
+Because $g_s$ tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016], the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have greater reductions in whole-tree transpiration and photosynthesis during periods of high atmospheric demand - be these on time scale of hours, days, or seasons [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021],consistent with the observation that  $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2).
+More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
+This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table 1, @mcdowellRelationshipTreeHeight2011].
 
 The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments.
-While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions (**REFS--KAT**), only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
+While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions [@frittsTreeRingsClimate1976], only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
 These have generally found greater sensitivity to low precipitation or high summer temperatures among larger, more exposed trees [**Fig.
-`r fig_treerings`**, @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021] , corroborating evidence from forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
-Despite the potential for shorter trees in open forests to experience greater environmental stress [*Fig.
-`r fig_schematic`b*; **REFS**], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021].
+`r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
+In addition to this lower drought resistance of growth, larger trees also frequently exhibit lower resilience [@mcgregorTreeHeightLeaf2021; **REF**] and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
+Despite the potential for shorter trees in open forests to experience greater environmental stress [**REFS**], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
 
-There are, however, notable exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
-In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory].
-Similarly, in tropical forests higher night-time temperatures have been associated with decelerating tree growth rates predominantly in smaller trees *but also in taller trees* [@clarkTropicalRainForest2003; @clarkAnnualWoodProduction2010; @feeleyDeceleratingGrowthTropical2007b; @vlamTemperatureRainfallStrongly2014*].
-Potential mechanisms for these patterns include increases in nighttime respiratory carbon loss that disproportionately reduced carbon availability for growth among small trees, or that understory $g_s$  might be stifled under humid conditions, resulting in higher $T_{leaf}$ (Fig.
-`r fig_leaf_T`) and lower photosynthesis.
-It is also possible that higher temperatures reduce understory tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. 
+There are, however, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
+In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2021].
+Potential mechanisms for these patterns include stifling of understory ??E under still, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig.
+3) and lower photosynthesis, increases nighttime respiratory carbon loss that disroportionately reduced carbon availability for growth among small trees accelerated asymmetric competition.
 Additional research is required to understand these mechanisms and to disentangle tree growth responses to hot-wet versus hot-dry conditions.
-
-~~~ 
-misc content to incorpoarte:
-
-- evidence in mesic temperate and subtropical eucalyptus forests shows that higher temperatures primarily reduced tree growth through altering water availability  [@bowmanWarmerWorldWill2014].
-
-- In other cases, warming is associated with taller understory trees with lower LMA that complete their lifecycle under closed-canopy temperate forests [@maesPlantFunctionalTrait2020].
-ALso, tropical warming experiment showed increased growth (**Winter**)
-
-Phenologically, in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].
-However, later in spring, short-term heat waves can negatively impact partially-expanded canopy leaves by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves more severly than in understory saplings that are buffered [@filewodImpactsSpringHeat2014].
-Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].
-
-*In this section we could perhaps also consider differences in biomass distribution between canopy trees an understory plants, and what that may mean for performance at high temperature.
-What I'm thinking is the temperature sensitivity of respiration that applies to leaves, stems, and roots, and even though thermal acclimation of leaf respiration is common, much less is known about acclimation of stem and root respiration, and incomplete acclimation still increases the respiration load at higher temperature.
-It is possible that an increased respiration load is more disadvantageous for understory plants because 1), I suspect that a greater fraction of their total biomass is life biomass that respires (as opposed dead heartwood in the stems of big trees), and 2), because they operate closer to their light compensation point, so if they lose more carbon in respiration, they would need  to make up for lost carbon in photosynthesis, but might lack sunlight to do so*
-
-~~~~
 
@@
 2020.
-I'M NOT SURE THIS IS WHAT WE WANT, BUT I THINK SOME SORT OF TREE-RING FIGURE WOULD BE GOOD.
 [ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/tree_rings_draft.png)
-
-**Tree mortality risk from climate-related stress is also size-structured.**
-
-- most common trend is higher risk of drought mortality for larger trees [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019b], including in open forests (**REFS**).
-*(However, there are exceptions...)*
-- also, greater risks of lightning, blow-down, etc.  (@goraImplicationsSizedependentTree2021) *(although this isn't temperature-related, so should get minimal attention, if at all)*
 
@@
 
-**Canopy trees account for the majority of forest ecosystem metabolism-- i.e., transpiration, GPP, woody productivity ($ANPP_{woody}$).**
+**Canopy trees account for the majority of forest ecosystem water and carbon cycling.**
 

### 2021-07-27 vertical-thermal-review.Rmd
Ecology section updates

@@
 **Canopy trees account for the majority of forest ecosystem water and carbon cycling.**
+While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
+Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration (Fig 5a).
+In terms of C cycling, it has been estimated that canopy strata contribute ???64% of net daytime $CO_{2}$ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
+Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees ???10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig.
+5b, Piponiot et al., accepted).
 
-- ET: "Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases.
-As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see @nakamuraForestsTheirCanopies2017), leaf traits, and photosynthesis .
-Linked to this local biophysical variation is the partitioning of evapotranspiration ($ET$) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al.
-2007, Staudt et al.
-2011).
-Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity ($GPP$, Banerjee & Linn 2018).
-It has been demonstrated that net carbon assimilation and transpiration can vary substantially for forest canopies with the same LAI but different leaf area density distributions (e.g., Katul et al 2004 , Banerjee & Linn 2018).
-Further, although transpiration tends to be the dominant component of ET, partitioning is more complex in open canopies where soil evaporation can become more dominant (Baldocchi et al.
-2004, Ma et al.
-2020), likely affecting vertical thermal gradients.
-In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout." *(condense this by quite a lot.
-I don't think all the info here is necessary)*
-- ET : "Still, partitioning $ET$ into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge.
-Eddy covariance techniques, sap flow measurements, and stable isotope analysis offer useful tools, although errors can be large (Williams et al 2004, Herbst et al.
-1996).
-To date, empirical analyses of ET vertical partitioning and interactions with temperature and vegetation structure have be concentrated in midlatitude ecosystems (e.g.
-Haverd et al 2009, 2011, Staudt et al.
-2011, Ringgaard et al.
-2014) *(KAT: but see Kunert et al.
-2017 -not sure if this is equivalent)*.
-" - *(revise this to focus on the observed patterns, with at most brief mention of the methods)*
-- GPP : ??
-- $ANPP_{woody}$: Piponiot et al., accepted, @meakemRoleTreeSize2018, New Phyt
+**Include ET figure** Figure 5.
+Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
+Panel (a) is from (Kunert et al.
+2017).
+I'd like to create a parallel figure for C fluxes.
 
-**Less clear is how thermal sensitivity varies across strata.**
-
-- ET: "With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions."
-- GPP : ??
-- $ANPP_{woody}$: if we scale tree-rings to the ecosystem level, we expect that woody productivity of canopy trees is more sensitive to drought, but perhaps less sensitive to wet heat.
-More research is needed here.
+**It is less clear how thermal sensitivity of water and C fluxes varies across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures-- at least under conditions of moderate to high VPD.
+A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an El Nino drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.
+It remains far less clear how thermal sensitivity is likely to vary across forest strata under wet conditions, but both physiological mechanisms (Fig.
+3) and tree ring evidence (Fig.
+4) suggests that understory trees may respond more negatively to hot, wet conditions.
+Further research is required to better understand the thermal sensitivity of forest ecosystem function across strata.
 
 ## Implications [ & future research directions?]
-
-*Elsa: If emphsaize future research directions, could move the remote sensing opportunities to this section and frame modeling section around opportunities to use models to further explore open questions, scale these findings in space and time, and influence predictions*
-*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*
 
@@
 Warming shifts forest composition, from leaf metabolism to whole forest structure.
-At a leaf metabolism level, fluxnet global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].
-However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosytems [@keenanWidespreadInhibitionDaytime2019].
+At a leaf metabolism level, global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled up to photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, thereby shifting the C balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
+However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosystems [@keenanWidespreadInhibitionDaytime2019].
 However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pam Templer et al.].
@@
 
-Feedback loops from loss of big trees through natural or anthropogenic disturbance (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020].
+Feedback loops from loss of big trees through natural or anthropogenic disturbance (e.g.
+@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020].
 Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 
-In the face of global climate change, resiliency among forests also stand out, such as African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021].
-Increasing temperature acclimation among tropical, temperate and boreal species [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019; @cunninghamTemperateRainforestTrees2003; @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] are other mechanisms that will help forests cope with warming.
+Variation in resistance and resilience among forests may drive different trajectories from the forests in the face of global climate change.
+For example, African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021].
+Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019; @cunninghamTemperateRainforestTrees2003; @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] are other mechanisms that will help forests cope with warming.
 

### 2021-07-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_leaf_T`).
-However, higher wind speeds (e.g., up to ~ 4$ms^{-1}$ at NEON sites; Fig.
-`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ 4$^\circ$C [Fig.
+However, higher wind speeds (e.g., up to ~ 4 $ms^{-1}$ at NEON sites; Fig.
+`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ $4^\circ$C [Fig.
 `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
@@
 Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than lower canopies layers [@millerOnlySunlitLeaves2021].
-In contrast, open canopies (lower canopy LAI) allow more vertical air mixing and sunlight into the understory, and this can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ of the lower canopy more than that of the upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].
+In contrast, open canopies (lower canopy LAI) allow more vertical air mixing and sunlight into the understory.
+This can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ of the lower canopy more than that of the upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].
 

### 2021-07-27 vertical-thermal-review.Rmd
removed the buffering across topography/geography
let me know if this is better to keep, and I can add this back

@@
 `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
-
-**The buffering patterns described above vary geographically, seasonally, and across forest types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, @frenneForestMicroclimatesClimate2021].**
-The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture, and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019].
-Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a].
 

### 2021-07-28 vertical-thermal-review.Rmd
Edits in introduction section

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen $1.2^\circ$C since the 1850-1900 period [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilitiesa 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilitiesa; @mcdowellPervasiveShiftsForest2020].
+Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
-The resulting feedbacks to the climate system, decreased C storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006a].
+The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
 
@@
 Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
-Overstory trees form canopies that play a crucial role in moderating forest macroclimates [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], as well as microclimates by buffering understories from extreme temperatures and other macroclimatic conditions [@zellweger_seasonal_2019].
-This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
+Overstory trees form canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including understory microclimates by buffering extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019].
+This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This vertical gradient shapes nearly every aspect of plant metabolism, demography, and ecology; however, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
 
@@
 (4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
-(5) size-structuring of whole-tree and ecosystem function.
-*We then consider the implications for understanding forest responses to global change, including how these scale across space and time.*
+(5) the size-structuring of whole-tree and ecosystem function.
+*We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.*
 
@@
 **The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs.
-`r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest itself.**
+`r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.**
 In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S.
@@
 Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**.
-Shown are  height profiles in growing season (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
 Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
@@
 `r fig_NEON_vertical`).**
-Foliage distribution is structured in correspondence with the distribution of tree sizes within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs.
+Foliage distribution is structured in correspondence with the distribution of tree sizes and tree architecture within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs.
 `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`).
@@
 In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
-bottom heavy foliage trees and sometimes by dense understories [e.g., SCBI, Fig.
+bottom heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig.
 `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
@@
 `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
-Soil moisture conditions, topography, and gap formations followed by disturbance all alter foliage patterns and further influence the formation of relatively closed vs.
+Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs.
 open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
@@
 This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`d), than under more open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
+`r fig_NEON_vertical`d), than under open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
 `r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
@@
 
-**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
+**Wind speeds are also higher at the top of the canopy, owing to the effect of turbulence buffering with canopy height [@juckerCanopyStructureTopography2018].**
 Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
-Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs.
+Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the vertical profile than the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs.
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for both water and heat loss (see next section).
 
 **Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
-Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
+Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
 Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
 
-Relative humidity, RH, tends to be higher in the understory and decreases with height across the vertical profile, although this trend is absent in open forests [Figs.
+Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Figs.
 `r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
-Dense-canopy forests maintain higher daily maximum RH below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
-In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].Especially when canopy height and leaf area index are large, maximum Tair and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
+Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
+In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Especially when canopy height and leaf area index (LAI) are large, maximum Tair and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
 
-**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs.
+**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs.
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
-The $T_{air}$ gradient along canopies and understories varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [@defrenneForestMicroclimatesClimate2021].
-Typically, dense canopies buffer understories from high maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.
+The $T_{air}$ gradient from the canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021].
+Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.
 `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
@@
 
-**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism (see following section) across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
+**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
 Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].

### 2021-07-28 vertical-thermal-review.Rmd
adding edits

@@
 
-Rising temperatures are having profound effects on forests; however, it is not well understood how responses vary across forest strata.
+Rising temperatures are having profound effects on forests; however, it is not well understood how responses vary vertically across forest strata.
 Here, we review how environmental conditions, leaf temperatures,  and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function.
@@
 The implications for tree performance are mixed: whereas larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may be stifled under hot, humid conditions.
-As warming continues, large canopy trees will be particularly vulnerable to drought, whereas understory trees will suffer most when the buffering provided by large trees is broken.
+As warming continues, large canopy trees will be particularly vulnerable to drought, whereas understory trees will suffer most when the buffering provided by large trees is lost.
 
@@
 
-**Here, we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
+**Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests.** 
 We focus on five key themes (Fig. `r fig_schematic`):
@@
 National Ecological Observatory Network [NEON; Figs.
-`r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, Schimel et al.
-2007].
+`r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
 While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height (Parker 1995).

### 2021-07-29 vertical-thermal-review.Rmd
Tleaf section edits

@@
 `r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
-While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height (Parker 1995).
+While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].
 
@@
 **Wind speeds are also higher at the top of the canopy, owing to the effect of turbulence buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
+Wind speeds increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
@@
 Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to 'canopy greenhouse effect' [e.g., OSBS, Figs.
+Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Figs.
 `r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
@@
 
-Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
-Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
+Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+Leaves can be cooler than the air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
 While $T_{leaf}$ is rarely exactly equal to $T_{air}$ (poikilothermy), it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
@@
 `r fig_leaf_T`a).
-Generally, the dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf.
-Evapotranspiration ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
-Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
+Generally, the dominant avenue of heat loss to the environment is latent heat flux, wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf.
+Latent heat flux ($\lambda E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
+Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a]; decreases with leaf size (Fig. `r fig_leaf_T`d);
 decreases with $g_s$ (Fig. `r fig_leaf_T`e);
-and decreases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
+and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
 
@@
 
-**Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
+**Leaf energy balance is a tradeoff between water-use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
 When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
-However, when leaf water demand (shaped by VPD and photosynthetic CO2 demand) exceeds the rate of supply, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
+However, when leaf water demand (influenced by VPD and photosynthetic CO~2~ demand) exceeds the rate of supply, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`e, @fauset_differences_2018].
@@
 
-**These basic biophysical principles shape the propensities for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
+**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
 High radiation at the top of the vertical profile (Fig.
@@
 Collectively, higher wind speed (Fig.
-`r fig_NEON_vertical`), and with adaptive leaf traits increase $g_a$ and $g_s$ (reviewed below) give upper canopy leaves lower sensitivity to intercepted radiation (Fig.
-`r fig_leaf_T`a) and higher rates of heat loss (Fig.
+`r fig_NEON_vertical`), and with adaptive leaf traits increase $g_a$ and $g_s$ (reviewed below) mediate the direct effects of radiation on upper canopy leaves (Fig.
+`r fig_leaf_T`a) and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
@@
 **Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).**
-$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and LAI [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021].
-Closed canopies, and high LAI canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate.
-The intensity of incoming radiation and VPD in the upper canopy, combined with hydraulic challenges of water transport to the tops of tall trees [**REFS**], can cause $g_s$-limitation.
+Closed canopies with high LAI act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
+The intensity of incoming radiation and VPD in the upper canopy, combined with hydraulic challenges of water transport to the tops of tall trees [**REFS**] can cause $g_s$-limitation.
 Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than lower canopies layers [@millerOnlySunlitLeaves2021].
-In contrast, open canopies (lower canopy LAI) allow more vertical air mixing and sunlight into the understory.
-This can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ of the lower canopy more than that of the upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].
+In contrast, open canopies with lower canopy LAI allow more vertical air mixing and sunlight into the understory.
+This mixing and light penetration can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy.
+The latter can result from the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].
 

### 2021-07-29 vertical-thermal-review.Rmd
incorporated edits until T50 section

@@
 **Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorological conditions (Fig. `r fig_leaf_T`).**
-Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).
-Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than that of shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig.
+Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
+Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig.
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
+In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling.
+Under lower wind conditions, the thickness of the air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
 `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation load, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 Understory leaves, which typically experience lower windspeeds (Fig. `r fig_NEON_vertical`), would retain more heat.
-The higher-RH conditions of the understory (Fig.
+Higher RH conditions in the understory (Fig.
 `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig.
@@
 **Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
-Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* and therefore differs little between understory and canopy.
-In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
+In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies.
+In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
@@
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020 (in review)].
-Greater $RH$ in inner canopy can also increase $T_{leaf}$ relatively more than upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
+Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
 
-It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems; for instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021].
+It is noteworthy that $T_{leaf}$ often differs substantially from branch and stem temperatures.
+For instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021].
 Thus, the vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.
@@
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
-However, when it comes to understanding and modeling forest ecosystem function, the relevant scale of variation is interspecific (e.g., *Lamour, Serbin in prep*).
+However, when it comes to understanding and modeling forest ecosystem function, relevant scale of variation is interspecific (e.g., *Lamour, Serbin in prep*).
 
@@
 
-Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-As much of our understanding of trait coordination is based on the leaf economics spectrum concept (@wrightWorldwideLeafEconomics2004) which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses through the vertical canopy.
-It has been established that independent of light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
+Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at an anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+As much of our understanding of trait coordination is based on the leaf economics spectrum concept (@wrightWorldwideLeafEconomics2004) which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the canopy.
+It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
 For example, LMA, photosynthetic capacity (per area) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
-Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical 2021; @niinemetsResponsesForestTrees2010a].
+Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
 
@@
 
-Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, by being smaller, hairier, and with higher vein density than shade leaves [Table 1, @leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, hairier, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
-In contrast, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks (e.g., lower LMA), but are more likely to get overheated than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
+In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but they are more likely to overheat than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
+In open canopies, where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
-Caratenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin ($VAZ$) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-The dissipation of excess light energy is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].Further, higher heat tolerance in exposed canopies is facilitated by isoprenes in species capable of emitting isoprenes [**REF**].
+Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+The dissipation of excess light energy is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
+Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019a; @monsonLeafIsopreneEmission2021].
 
@@
 
-The timing of seasonal leaf loss or gain is influenced by vertical microclimate gradients, and in turn can alter biophysical and $T_{leaf}$ gradients in the forest [@rey-sanchez_spatial_2016].
+The timing of seasonal leaf loss or gain is influenced by vertical microclimate gradients, and in turn can alter forest biophysical and $T_{leaf}$ gradients [@rey-sanchez_spatial_2016].
 Deciduousness in tropical forests is an adaptation to conserve water, and is more common in drier forests and in canopy species [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
@@
 Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
-Generally, the pattern of fundamental trait adaptation to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-For instance, across tree size classes, there is an increase in the fractions of species that are isoprene-emitting (**REFS**) and deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
 However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
-
 Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
@@
 Leaf metabolism is strongly shaped by $T_{leaf}$, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
-However, as we detail below, very few studies have evaluated temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
 
@@
 
-Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves (Table 2).
-However, because of the high VPD and irradiance experienced by sun leaves,  mid-day stomatal depression is more prevalent in sun leaves than shade leaves in closed forests.
+Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in sub-canopy or understory leaves (Table 2).
+However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.
 $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-In more open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer time duration than upper canopy resulting in higher $T_{leaf}$, perhaps due to higher RH [Fig.
+In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves resulting in higher $T_{leaf}$, perhaps due to higher RH [Fig.
 `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002].
-The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ to rising $T_{air}$ than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
+The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ to rising $T_{air}$ than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
 
@@
 2020).**
-This is primarily driven by the greater light available to sun-leaves, which also have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions (Table 1, Table 2).
-Photosynthesis increases with irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$.
-Light saturation levels differ between leaves and whole canopies with leaves saturating at lower light level than canopies; while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy.
+This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Table 1, Table 2).
 
-Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to $T_{air}$, with the peak commonly found to correspond to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017].
-Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g.
-@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020 [temperate/boreal refs needed), and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015].
+Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
+Photosynthesis has a peaked response to $T_{air}$, with the peak commonly found to corresponding to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017].
+Beyond the optimum, photosynthesis decreases as a result of stomatal closure [e.g.
+@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020, temperate/boreal refs needed], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015].
 
 **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
-Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons [e.g.
+Based on consistent positive relationships between growth temperature and temperature optimum of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
 @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
@@
 Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
-In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical 2021].
+In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
 Higher maximum temperatures, and $g_s$ are known to influence leaf thermal sensitivity [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
@@
 @hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis.
-Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of mid-day stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
-Shade leaves that do not experience mid-day depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021].
+Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of midday stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
+Shade leaves that do not experience midday depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021].
 Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.

### 2021-07-29 vertical-thermal-review.Rmd
small edits

@@
 Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-The dissipation of excess light energy is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
+The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
 Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019a; @monsonLeafIsopreneEmission2021].
@@
 This may vary based on forest type [**REF**], leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
-Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical].
-Further, we would expect $Q_{10}$ to also vary with soil moisture based on the evidence of canopy-position effect gradients in respiration rates for *Quercus rubra* being much greater in mesic than in xeric sites [@xuSeasonalVariationTemperature2006].
+Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021].
 
@@
 
-### Remote sensing
-
-An increasing availability of airborne and spaceborne LiDAR remote sensing data offers a promising opportunity for mapping vertical thermal gradients and vegetation structure.
-[Drone based FLIR cams, tower thermal cams (Pau et al 2018), ECOSTRESS].
-To evaluate large-scale ecosystem patterns, high resolution vertically resolved vegetation structure from LiDAR data can be analyzed in combination with collocated field measurements collected using towers (FluxNet REF), canopy cranes (@nakamuraForestsTheirCanopies2017), or by tree climbing (REF – e.g.
-Asner papers, Shenkin et al – Unding Jami’s work).
-Existing research efforts have already revealed incredibly heterogeneity in vegetation structure, diversity, traits, and properties both within and across ecosystems (REFS – Stovall, Schneider, …others).
-
 ### Representing Vertical Gradients in Models
@@
 Still, as models increase in complexity, model developments and improvements have direct tradeoffs with increased computational cost and potentially decreased interpretability of model output, highlighting the need to identify and parsimoniously represent the most essential drivers of forest ecosystem function.
-
--Shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses (e.g., Bode et al.
-https://doi.org/10.1016/j.rse.2014.01.028, Musselman et al.
-https://doi.org/10.1016/j.rse.2013.05.021, Davis et al.
-https://doi.org/10.1016/j.agrformet.2019.02.015)
 
@@
 Each plant perceives a unique environment, including incident light, temperature, and vapor pressure deficit that varies across local scales in association with the horizontal and vertical position of each cohort relative to other cohorts, in addition to patch level differences in topographic and edaphic conditions.
-Coexistence of different vegetation types emerges from different environmental niches, either along a successional gradient of light availability or vertical position in the canopy (Moorcroft et al., 2001; Smith et al., 2001; Purves & Pacala, 2008).
-
-Temperature is an atmospheric boundary condition driving the ED2 model (Longo et al 2019).
-Each patch in each grid cell is defined by a thermodynamic envelope, comprised of distinct thermodynamic systems for each soil layer, temporary surface water or snow layer, the aboveground component of each cohort, and the canopy air space (Longo et al 2019).
-The  atmosphere above and outside of the canopy air space is referred to as free air and is determined directly from boundary conditions in ED2.
-Patches exchange heat and mass with the free air  and lose water and associated energy through surface and sub-surface runoff.
-The net enthalpy flux of soil, temporary surface water, and vegetation thermodynamic systems in ED2 are exclusively due to associated water fluxes, whereas the eddy flux between the free air and the canopy air space includes both water transport and flux associated with the mixing of air with different temperatures between the canopy air space and free air (Longo et al 2019).
 

### 2021-07-30 vertical-thermal-review.Rmd
updates

@@
 This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-This vertical gradient shapes nearly every aspect of plant metabolism, demography, and ecology; however, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
@@
 
-Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, hairier, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, hairier, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
@@
 $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves resulting in higher $T_{leaf}$, perhaps due to higher RH [Fig.
+In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated lower canopy $T_{leaf}$, perhaps due to higher RH [Fig.
 `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002].
-The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ to rising $T_{air}$ than shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
+The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
 

### 2021-07-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004]**
+**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].**
 Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
@@
 
-#### (Biochemical and elemental traits, and isoprenes)
+#### (Biochemical and elemental traits, and isoprene)
 
@@
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
-Higher maximum temperatures, and $g_s$ are known to influence leaf thermal sensitivity [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
 

### 2021-07-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2021].
-Potential mechanisms for these patterns include stifling of understory ??E under still, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig.
+Potential mechanisms for these patterns include stifling of understory $\lambda E$ under still, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig.
 3) and lower photosynthesis, increases nighttime respiratory carbon loss that disroportionately reduced carbon availability for growth among small trees accelerated asymmetric competition.

### 2021-08-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated lower canopy $T_{leaf}$, perhaps due to higher RH [Fig.
+In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [Fig.
 `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002].

### 2021-08-01 vertical-thermal-review.Rmd
Leaf age section needs some more guidance

@@
 Overstory trees form canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including understory microclimates by buffering extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019].
-This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
 Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
@@
 Schematic summarizing vertical gradients in forests with dense canopies during daytime growing season conditions.** Patterns summarized here tend to be weaker, or sometimes reversed, under more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
-[Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
+Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy  [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
 ](schematics/Screen Shot 2021-07-19 at 1.10.17 PM.png)
@@
 When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
-However, when leaf water demand (influenced by VPD and photosynthetic CO~2~ demand) exceeds the rate of supply, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
+However, when leaf water demand (influenced by VPD and photosynthetic CO~2~ demand) exceeds the rate of supply, conservation of water ($g_s$-limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`e, @fauset_differences_2018].
@@
 Collectively, higher wind speed (Fig.
-`r fig_NEON_vertical`), and with adaptive leaf traits increase $g_a$ and $g_s$ (reviewed below) mediate the direct effects of radiation on upper canopy leaves (Fig.
+`r fig_NEON_vertical`), and with adaptive leaf traits increase $g_a$ and $g_s$ (reviewed below) mediate the direct effects of solar radiation on upper canopy leaves (Fig.
 `r fig_leaf_T`a) and result in higher rates of heat loss (Fig.
@@
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
-However, when it comes to understanding and modeling forest ecosystem function, relevant scale of variation is interspecific (e.g., *Lamour, Serbin in prep*).
+However, when it comes to understanding and modeling forest ecosystem function, given that species traits strongly shape the propensity for trees to inhabit understory vs.
+canopy positions, interspecific variation shapes ecosystem function is interspecific (e.g., *Lamour, Serbin in prep*).
 
@@
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but they are more likely to overheat than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies, where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
+In open canopies, where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and VPD stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-Leaf age affects vulnerability to heat stress independent of light environment [e.g., Marias et al.
-2017], and has a strong effect on metabolism [@kikuzawaSynthesisRelationshipsLeaf2006], including photosynthesis and respiration, and their temperature sensitivities [@zhouLeafageEffectsTemperature2015; @niinemetsLeafAgeDependent2016].
-*In evergreen species,* sun leaves have shorter lifespans than shade leaves; upper canopy leaves are on average younger than subcanopy leaves, therefore operating closer to their photosynthetic optimum.
-However, because leaf turnover rates of sun leaves in the upper canopy are faster than in the subcanopy, less damage  accumulates [@millerOnlySunlitLeaves2021].
-Once leaves are fully expanded and mature, heat tolerance decreases with leaf age [@zhouLeafageEffectsTemperature2015; REFS], as do $g_s$ and photosynthesis [with a few exceptions, @niinemetsLeafAgeDependent2016].
-Additionally, because older leaves reach maximum $T_{leaf}$ sooner than younger leaves and can occupy a large part of the canopy, this would affect vertical gradient of thermal sensitivity [@zhouLeafageEffectsTemperature2015] **this needs some confirming by co-authors**.
-
-The timing of seasonal leaf loss or gain is influenced by vertical microclimate gradients, and in turn can alter forest biophysical and $T_{leaf}$ gradients [@rey-sanchez_spatial_2016].
-Deciduousness in tropical forests is an adaptation to conserve water, and is more common in drier forests and in canopy species [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
-Within tropical deciduous species, the fraction of deciduous individuals tends to increase across tree size classes, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+The vertical structure of microenvironments likely influences forest photosynthetic capacity and $T_{leaf}$ gradients [@rey-sanchez_spatial_2016] via leaf phenology.
+For example, deciduousness is an adaptation to conserve water during dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile.
+Within tropical deciduous species, the fraction of seasonally deciduous individuals increases toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] via age-specific leaf function [@albertCrypticPhenologyPlants2019].
+Leaf age affects vulnerability to heat stress [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015] and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.
+For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021].
 

### 2021-08-01 vertical-thermal-review.Rmd
Leaf age section needs some more guidance
Working on Ecology section now

@@
 Within tropical deciduous species, the fraction of seasonally deciduous individuals increases toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] via age-specific leaf function [@albertCrypticPhenologyPlants2019].
-Leaf age affects vulnerability to heat stress [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015] and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.
+Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019].
 For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021].
+Leaf age affects vulnerability to heat stress and has a strong effect on metabolism  [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015; @kikuzawaSynthesisRelationshipsLeaf2006], and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.
 
@@
 Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to $T_{air}$, with the peak commonly found to corresponding to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017].
+Photosynthesis has a peaked response to $T_{air}$, with the peak commonly found to corresponding to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure [e.g.
-@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020, temperate/boreal refs needed], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015].
+@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; @fredeenTemperatureHumidityEffects1999], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
 

### 2021-08-01 vertical-thermal-review.Rmd
Leaf age section needs some more guidance
Working on Ecology section now

@@
 **Vertical gradients affect tree metabolism and growth.**
-Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
-Specifically, increases are linked to increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
+Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
+Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
 The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
@@
 **Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
-Because $g_s$ tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016], the taller trees that occupy canopy positions in relatively dense-canopy forests tend to have greater reductions in whole-tree transpiration and photosynthesis during periods of high atmospheric demand - be these on time scale of hours, days, or seasons [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021],consistent with the observation that  $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2).
+Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+Therefore, periods of high atmospheric demand - be these on time scale of hours , days, or seasons - tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
+This is consistent with the observation that  $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2).
 More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].

### 2021-08-02 vertical-thermal-review.Rmd
incorporating Ian's edits

@@
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
-Collectively, these impacts may lead to reduced forest C sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
+Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
 The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
@@
 **Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
-Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
+Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
 Overstory trees form canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including understory microclimates by buffering extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019].
@@
 `r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
-In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top vs.
-bottom heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig.
+In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs.
+bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig.
 `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
@@
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for both water and heat loss (see next section).
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for both water and heat loss (see section "Leaf temperature").
 
@@
 
-**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based via biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig.
+**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig.
 `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
@@
 However, when it comes to understanding and modeling forest ecosystem function, given that species traits strongly shape the propensity for trees to inhabit understory vs.
-canopy positions, interspecific variation shapes ecosystem function is interspecific (e.g., *Lamour, Serbin in prep*).
+canopy positions, ecosystem function is shaped by interspecific trait variation (e.g., *Lamour, Serbin in prep*).
 
@@
 
-Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in sub-canopy or understory leaves (Table 2).
+Maximum $g_s$ increases with light environment, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2).
 However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.
@@
 
-Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
+Temperature can affect photosynthesis via direct and indirect pathways: directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
 Photosynthesis has a peaked response to $T_{air}$, with the peak commonly found to corresponding to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
@@
 **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
-Based on consistent positive relationships between growth temperature and temperature optimum of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
+Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
 @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].

### 2021-08-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Therefore, periods of high atmospheric demand - be these on time scale of hours , days, or seasons - tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
+Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
 This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2).
@@
 While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions [@frittsTreeRingsClimate1976], only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
-These have generally found greater sensitivity to low precipitation or high summer temperatures among larger, more exposed trees [**Fig.
+These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [**Fig.
 `r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
-In addition to this lower drought resistance of growth, larger trees also frequently exhibit lower resilience [@mcgregorTreeHeightLeaf2021; **REF**] and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
+In addition to lower drought resistance of growth, larger trees also frequently exhibit lower ability to bounce back from stress, ie.
+resilience [@mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
 Despite the potential for shorter trees in open forests to experience greater environmental stress [**REFS**], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
@@
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2021].
-Potential mechanisms for these patterns include stifling of understory $\lambda E$ under still, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig.
-3) and lower photosynthesis, increases nighttime respiratory carbon loss that disroportionately reduced carbon availability for growth among small trees accelerated asymmetric competition.
+Potential mechanisms for these patterns include stifling of understory $\lambda E$ under low wind, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig.
+3) and lower photosynthesis.
+Furthermore, warming-induced increases in nighttime respiratory carbon loss may be particularly disadvantageous for understory trees that are operating closer to their light-compensation point than canopy trees.
 Additional research is required to understand these mechanisms and to disentangle tree growth responses to hot-wet versus hot-dry conditions.
@@
 ![**Figure `r fig_treerings`.
-DRAFT tree-rings figure.
-Current figure shows (a) drought responses as a function of height in a temperate broadleaf forest (SCBI), from McGregor et al.
-2020; (b) responses to growing season temperature in eight New England forests, from Rollinson et al.
-2020.
-[ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/tree_rings_draft.png)
+Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.
+In column  (a), large trees have a more negative response to high temperatures during the current or previous growing season.
+Shown are tree ring width (RW) responses to the most influential temperature variable at the site (Tmax or PET) of the largest versus smallest trees included in GLS model with a DBH x temperature interaction (Anderson-Teixeira et al.
+, in review) .
+In column (b), understory trees have more negative responses to high growing season temperatures across eight New England forests, (@rollinsonClimateSensitivityUnderstory2021) [ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/Figure_4_tree_rings.png)
 

### 2021-08-02 vertical-thermal-review.Rmd
Stepping into Warming section

@@
 The proportion of GPP allocated to many of these increases with tree size, including ...
-foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].As a result of increasing C allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021; @anderson-teixeiraCarbonCyclingMature2021].
+foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].
+As a result of increasing carbon allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraCarbonCyclingMature2021].
 However, NSC reserves increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].
@@
 This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2).
-More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
+More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
 This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table 1, @mcdowellRelationshipTreeHeight2011].
@@
 These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [**Fig.
-`r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021; @anderson-teixeiraJointEffectsClimate2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
+`r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
 In addition to lower drought resistance of growth, larger trees also frequently exhibit lower ability to bounce back from stress, ie.
 resilience [@mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
-Despite the potential for shorter trees in open forests to experience greater environmental stress [**REFS**], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
+Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
 
@@
 In column (a), large trees have a more negative response to high temperatures during the current or previous growing season.
-Shown are tree ring width (RW) responses to the most influential temperature variable at the site (Tmax or PET) of the largest versus smallest trees included in GLS model with a DBH x temperature interaction (Anderson-Teixeira et al.
-, in review) .
-In column (b), understory trees have more negative responses to high growing season temperatures across eight New England forests, (@rollinsonClimateSensitivityUnderstory2021) [ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/Figure_4_tree_rings.png)
+Shown are tree ring width (RW) responses to the most influential temperature variable at the site ($T_{max}$ or PET) of the largest versus smallest trees included in GLS model with a DBH x temperature interaction (Anderson-Teixeira et al., in review).
+In column (b), understory trees have more negative responses to high growing season temperatures across eight New England forests, (@rollinsonClimateSensitivityUnderstory2021).
+[ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/Figure_4_tree_rings.png)
 
@@
 Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration (Fig 5a).
-In terms of C cycling, it has been estimated that canopy strata contribute ???64% of net daytime $CO_{2}$ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
-Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees ???10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig.
+In terms of C cycling, it has been estimated that canopy strata contribute $\geq$64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
+Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$10cm DBH usually contributing $>$85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig.
 5b, Piponiot et al., accepted).
@@
 
-**It is less clear how thermal sensitivity of water and C fluxes varies across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures-- at least under conditions of moderate to high VPD.
+**It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures-- at least under conditions of moderate to high VPD.
 A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an El Nino drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.

### 2021-08-04 vertical-thermal-review.Rmd
abstract edits

@@
 
-Rising temperatures are having profound effects on forests; however, it is not well understood how responses vary vertically across forest strata.
-Here, we review how environmental conditions, leaf temperatures,  and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function.
-In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and are therefore at risk of elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
-However, foliar traits relevant to shaping $T_{leaf}$ also vary strongly across height or light gradients, partially offsetting the propensity towards elevated $T_{leaf}$ in the upper canopy.
-Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher metabolism and growth at both individual and ecosystem levels.
-Yet, differences in thermal sensitivity and damage thresholds are modest.
-The implications for tree performance are mixed: whereas larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may be stifled under hot, humid conditions.
-As warming continues, large canopy trees will be particularly vulnerable to drought, whereas understory trees will suffer most when the buffering provided by large trees is lost.
+Rising temperatures are profoundly influencing forests on many scales with potentially strong variation vertically across forest strata.
+Using published research and new analyses, we evaluate how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function.
+In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and can experience elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
+However, foliar traits that also vary strongly across height or light gradients can partially mitigate the elevation of $T_{leaf}$ in the upper canopy.
+Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels.
+Yet, differences in thermal sensitivity and damage thresholds across the gradient are modest.
+Under climate warming and increased drought, larger trees can face disproportionate stress, yet understory trees have fewer cooling mechanisms and thus may be impacted the most under hot, humid conditions, or when the buffering provided by large trees is lost.
 
@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise over pre-industrial levels of $1.6 - 4.3^{\circ}$C by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].

### 2021-08-05 vertical-thermal-review.Rmd
incorporating Marielle & Ty's edits

@@
 
-**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
+**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
 Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
@@
 (3) the leaf traits that most strongly influence $T_{leaf}$ and metabolism;
-(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
+(4) leaf metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
 (5) the size-structuring of whole-tree and ecosystem function.
@@
 `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.**
-In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from the U.S.
+In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S.
 National Ecological Observatory Network [NEON; Figs.
-`r fig_NEON_vertical`-`r SIfig_NEON_all`; Appendix `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
+`r fig_NEON_vertical`-`r SIfig_NEON_all`; Supporting Information Methods `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
 While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].
@@
 ![**Figure `r fig_NEON_vertical`.
-Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**.
-Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+Vertical gradients in the biophysical environment for six sites six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**.
+Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
 Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
@@
 open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019].
-In this review, we focus on growing season conditions (peak LAD) unless otherwise noted.
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019; @parkerConsequencesEnvironmentalHeterogeneity2019].
+In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.
 
-**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from canopy tops to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs.
+**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs.
 `r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].**
-This gradient is more pronounced under dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`d), than under open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
-`r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
+`r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
+`r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
 
-**Wind speeds are also higher at the top of the canopy, owing to the effect of turbulence buffering with canopy height [@juckerCanopyStructureTopography2018].**
-Wind speeds increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
+**Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
+Wind speeds increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
@@
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
-In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Especially when canopy height and leaf area index (LAI) are large, maximum Tair and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
+In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
 
@@
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).**
-The $T_{air}$ gradient from the canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021].
+The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021].
 Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
@@
 `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, during $g_s$- limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$-limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
 With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
@@
 Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
-In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies.
-In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies.
+In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
@@
 Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at an anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-As much of our understanding of trait coordination is based on the leaf economics spectrum concept (@wrightWorldwideLeafEconomics2004) which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the canopy.
+As much of our understanding of trait coordination is based on the leaf economics spectrum concept [@wrightWorldwideLeafEconomics2004] which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the canopy.
 It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
-For example, LMA, photosynthetic capacity (per area) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
 Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
@@
 
-The vertical structure of microenvironments likely influences forest photosynthetic capacity and $T_{leaf}$ gradients [@rey-sanchez_spatial_2016] via leaf phenology.
+The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology.
 For example, deciduousness is an adaptation to conserve water during dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile.
@@
 
-Maximum $g_s$ increases with light environment, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2).
+Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2).
 However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.
 $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [Fig.
+In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [OSBS, Fig.
 `r fig_leaf_T`,@zweifelMiddayStomatalClosure2002].
@@
 Temperature can affect photosynthesis via direct and indirect pathways: directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to $T_{air}$, with the peak commonly found to corresponding to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
+Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure [e.g.
@@
 However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
@@
 However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a].
-We would expect leaves exposed to greater heat stress within the canopy to have higher heat tolerance, but further research is required to fully understand the relationship between heat tolerance, light exposure and microclimates.
+The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
 
@@
 
-**Isoprene production**
-High emission rates occur in the mid-canopy, and even from very small statured trees, at a site in the Brazilian Amazon (Taylor et al., in prep).
-The high temperature carbon compensation point of sun leaves is higher in isoprene emitting species than in non-emitting species, so it is to be expected that other aspects of heat tolerance will be similarly affected by isoprenes (https://doi.org/10.1111/pce.13564).
-Within species, isporene production scales with light and temperature; together and independtly?
+**The vertical structuring of leaf isoprene emissions may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.** Isoprene gas emission from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsHighWithincanopyVariation2010].
+Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests[@taylorIsopreneEmissionStructures2018a].
+In contrast, a recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy region instead of the hotter and brighter upper canopy [@taylorNewFieldInstrument2021].
+This observation is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [@sharkeyIsopreneEmissionPlants2008], due to the sun's passage across a dynamically structured canopy surface.
+This review focuses primarily on variation in mean conditions across the vertical profile, but future work should seek to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function.
 

### 2021-08-05 vertical-thermal-review.Rmd
Ty's VOC section

@@
 
-**The vertical structuring of leaf isoprene emissions may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.** Isoprene gas emission from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsHighWithincanopyVariation2010].
-Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests[@taylorIsopreneEmissionStructures2018a].
+**The vertical structuring of leaf isoprene emissions may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.** Isoprene gas emission from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
+Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
 In contrast, a recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy region instead of the hotter and brighter upper canopy [@taylorNewFieldInstrument2021].

### 2021-08-05 vertical-thermal-review.Rmd
changed temperature from k to c

@@
 - However, when we ask the critical question of how the temperature sensitivities  of metabolism and woody growth varies across these different environments, we can't say much-- there are only a small number of studies, often showing insignificant differences or mixed results.
-- We lean towards the conclusion that while large canopy trees ware the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
+- We lean towards the conclusion that while large canopy trees are the most vulnerable to warming when water is limited, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
 - Critically, change to the vertical gradient--e.g., through increased mortality of canopy trees--will change processses across forest strata, with potentially dramatic consequences.

### 2021-08-05 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R

@@ -441,10 +441,10 @@ e1
 
 
 table<- data.frame(biophysical = c( "swr", "ws", "rh", "lw", "gs", "tair"),
-                   normalno = c("871", 2.88, 0.91, 0.04, "4.0", "298"),
-                   normal_u = c("102", 0.24, 0.97, 0.10, "2.0", "296"),
-                   drought_o = c("1151", 2.88, 0.46, 0.04, "0.01", "298"),
-                   normalno = c("153", 0.24, 0.49, 0.10, "0.01", "296"))
+                   normalno = c("871", 2.88, 0.91, 0.04, "4.0", "25"),
+                   normal_u = c("102", 0.24, 0.97, 0.10, "2.0", "23"),
+                   drought_o = c("1151", 2.88, 0.46, 0.04, "0.01", "25"),
+                   normalno = c("153", 0.24, 0.49, 0.10, "0.01", "23"))
 
 colnames(table) <- c("variable",
                  "overstory\nnormal", "understory\nnormal", "overstory\ndrought", "understory\ndrought")

### 2021-08-05 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Elsa M. Ordway^4^,
-Marielle N. Smith^5^,
-Tyeen Taylor,
-Lawren Sack,
+Marielle N. Smith^6^,
+Tyeen Taylor^7^,
+Lawren Sack^5^,
 Kristina J. Anderson-Teixeira^1,2^*
@@
 
-5. Department of Forestry, Michigan State University, East Lansing, MI 48824, USA
+5. Department of Ecology and Evolutionary Biology, UCLA, Los Angeles, CA 90039, USA
+
+6. Department of Forestry, Michigan State University, East Lansing, MI 48824, USA
+
+7. Department of Civil & Environmental Engineering, University of Michigan, Ann Arbor, MI 48109, USA
 
@@
 Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
-[ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/leaf_energy_fig3.png)
+[ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/fig3_leafenergy.png)
 

### 2021-08-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
@@
 (2) leaf temperature ($T_{leaf}$);
-(3) the leaf traits that most strongly influence $T_{leaf}$ and metabolism;
+(3) the leaf traits that most strongly influence $T_{leaf}$;
 (4) leaf metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
@@
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`).
-Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the vertical profile than the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs.
+Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Figs.
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).
@@
 
-**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from radiation minus heat lost to the environment [Fig.
+**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig.
 `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
@@
 Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than lower canopies layers [@millerOnlySunlitLeaves2021].
-In contrast, open canopies with lower canopy LAI allow more vertical air mixing and sunlight into the understory.
+In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory.
 This mixing and light penetration can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy.
@@
 **Vertical gradients affect tree metabolism and growth.**
-Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
+Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
 Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
@@
 
-Warming shifts forest composition, from leaf metabolism to whole forest structure.
-At a leaf metabolism level, global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled up to photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, thereby shifting the C balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-However, taken into account light-induced lower leaf respiration in the day, partionationing of C losses differ diurnally and seasonally across ecosystems [@keenanWidespreadInhibitionDaytime2019].
+Warming impacts forests, from leaf metabolism to whole forest structure.
+At a leaf metabolism level, global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled up to photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, thereby shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
+However, taken into account light-induced lower leaf respiration in the day, partionationing of carbon losses differ diurnally and seasonally across ecosystems [@keenanWidespreadInhibitionDaytime2019].
 However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pam Templer et al.].

### 2021-08-08 vertical-thermal-review.Rmd
scaling and modelling section

@@
 Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
@@
 `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
-The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave).
-High radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig.
-`r fig_leaf_T`a).
-Generally, the dominant avenue of heat loss to the environment is latent heat flux, wherein the loss of energy required to convert water from a liquid to a vapour state during evaporation and transpiration cools the leaf.
-Latent heat flux ($\lambda E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
-Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a]; decreases with leaf size (Fig. `r fig_leaf_T`d);
-decreases with $g_s$ (Fig. `r fig_leaf_T`e);
-and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
+High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig.
+`r fig_leaf_T`a).Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance ($g_a$), which is greater in smaller leaves (Fig.
+`r fig_leaf_T`d) and higher wind speeds [Fig.
+`r fig_leaf_T`b].
+Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is primarily regulated by stomatal conductance ($g_s$), but also by RH through a direct effect on the leaf-to-air vapor gradient (Fig.
+`r fig_leaf_T`c) and an indirect effect on $g_s$.
+Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a]; decreases with leaf size (Fig. `r fig_leaf_T`d);
+decreases with $g_s$ (Fig. `r fig_leaf_T`e); and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
 
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
 [ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/fig3_leafenergy.png)
@@
 Closed canopies with high LAI act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-The intensity of incoming radiation and VPD in the upper canopy, combined with hydraulic challenges of water transport to the tops of tall trees [**REFS**] can cause $g_s$-limitation.
-Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than lower canopies layers [@millerOnlySunlitLeaves2021].
+The intensity of incoming radiation and VPD in the upper canopy, combined with hydraulic challenges of water transport to the tops of tall trees [**REFS**] can cause $g_s$-limitation and an increase in $T_{leaf}$.
+Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than buffered lower canopy layers [@millerOnlySunlitLeaves2021].
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory.
@@
 
-It is noteworthy that $T_{leaf}$ often differs substantially from branch and stem temperatures.
-For instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021].
-Thus, the vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.
+It is noteworthy that leaf and branch temperatures often differ substantially.
+For instance, exposed tropical tree bark can be much warmer than leaves [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021; @johnstonWhatLiesVertical2020].
+Thus, vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.
 `r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories).
-Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021].
+Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021; @macleanMicroclimcMechanisticModel2021].
 
@@
 However, when it comes to understanding and modeling forest ecosystem function, given that species traits strongly shape the propensity for trees to inhabit understory vs.
-canopy positions, ecosystem function is shaped by interspecific trait variation (e.g., *Lamour, Serbin in prep*).
+canopy positions, ecosystem function is also shaped by interspecific trait variation (e.g., *Lamour, Serbin in prep*).
 
@@
 @mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; Ishida et al.
-1999; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000
+1999; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021
 ---
@@
 Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration (Fig 5a).
-In terms of C cycling, it has been estimated that canopy strata contribute $\geq$64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
-Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$10cm DBH usually contributing $>$85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig.
+In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
+Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig.
 5b, Piponiot et al., accepted).
 
-**Include ET figure** Figure 5.
-Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
+![**Figure 5.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
 Panel (a) is from (Kunert et al.
 2017).
-I'd like to create a parallel figure for C fluxes.
+I'd like to create a parallel figure for C fluxes.](schematics/ecosystem vertical partitioning/ecosystem partitioning_updated.png)
 
@@
 
-*Scaling our understanding across space and time requires remote sensing and models.*
-
 ### Representing Vertical Gradients in Models
 
-Models are needed to predict future ecosystem dynamics, which remain one of the largest sources of uncertainty in Earth System Model (ESM) projections of the future of global carbon cycling and climate change (*Friedlingstein et al.
-2006*).
-Dynamic global vegetation models (DGVMs) comprise the land surface models in ESMs used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks (Cao & Woodward, 1998; Foley et al., 1996; Sitch et al., 2003; Woodward & Lomas, 2004).
-DGVMs operate at a range of scales and have varying degrees of complexity, ranging from detailed individual-based models (*a.k.a.* forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal resolution (Christoffersen et al., 2016; Fischer et al., 2016; Fyllas et al., 2014; Sato et al.
-2007; Shuman et al.
-2014; Smith et al., 2001; Bugmann, 2001; Dietze & Latimer, 2011), to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer (Fig.
-`r fig_model`).
-This simplification results in greater computational efficiency and thus easier integration into ESMs (Arora & Boer, 2010; Bonan et al.
-2003; Cox, 2001; Krinner et al., 2005), although it comes at a cost of a lack of representation of important demographic processes, canopy gap formation, vertical light competition, competitive exclusion, and successional recovery from disturbance (Feeley et al., 2007; Hurtt et al.
-1998; Moorcroft et al., 2001; Smith et al.
-2001; Stark et al., 2012).
-Improved representation of vegetation demographic processes in ESMs, specifically including forest canopy architectural variation in both horizontal and vertical directions, has repeatedly been identified as a critical step toward reduced uncertainty  and more accurate characterization of biologically mediated feedbacks (Banerjee & Linn 2018, Evans, 2012; Moorcroft, 2006; Moorcroft, Hurtt, & Pacala, 2001; Purves & Pacala, 2008; Thomas, Brookshire, & Gerber, 2015).
-Still, as models increase in complexity, model developments and improvements have direct tradeoffs with increased computational cost and potentially decreased interpretability of model output, highlighting the need to identify and parsimoniously represent the most essential drivers of forest ecosystem function.
+Models are needed to predict future ecosystem dynamics, which remain one of the largest sources of uncertainty in Earth System Model (ESM) projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006].
+Dynamic global vegetation models (DGVMs), which comprise the land surface models in ESMs, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].
+DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
+This simplification is computationally more efficient, although it comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
+The computational middle-ground to representing vertical structure in DVGMs lies in cohort-based models (CBMs), which sit between the oversimplified vegetation dynamics in big-leaf models and the computational expense of individual-based models.
+CBMs represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].
 
-![**Figure `r fig_model`.
-DRAFT.
-Enrich this by showing 3-4 carefully selected scenarios (e.g., baseline, warming-mesic, warming-drought, canopy disturbance), and maybe a few key processes/ consequences (leaf T, C balance).** [Issue #42](https://github.com/EcoClimLab/vertical-thermal-review/issues/42).
-](schematics/model_schematic.png)
+**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.** Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks (Moorcroft, Hurtt, & Pacala, 2001; Banerjee & Linn 2018; Bonan et al.
+2021).
+In addition, this review finds that, although larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may also be stifled under hot, humid conditions.
+**A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.** Most models have been developed to capture dynamics in the canopy, or overstory, given the disproportionate role of these large trees in ecosystem dynamics.
+Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models.
+This is likely due in large part to the fact that observations are only beginning to shed light on the vulnerability of understory trees to changes in temperature and other abiotic conditions [e.g., @rollinsonClimateSensitivityUnderstory2020].
 
-**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.**
-*[how?]*
+Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g., seasonal shifts in vertical leaf area distributions, Smith et al.
+2019) and functions (e.g., thermal responses, as we present in this paper).
+At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function (e.g., de Pury & Farquhar, 1997; Bonan et al 2021).
+In addition, capturing the vertical gradients in NPP, GPP, and respiration, and subsequently the net ecosystem effects, requires improved characterization of the functional response of leaf-level processes to vertically varying abiotic conditions, and the role of traits and abiotic conditions in mediating responses to thermal sensitivity.
 
-**The computationally feasible approach to representing vertical structuring in DVGMs lies in Cohort-based models (CBMs), which sit in the middle of this continuum between the oversimplified vegetation dynamics in big-leaf models and the computational expense of individual-based models.** CBMs represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type (Haverd et al., 2013; Hurtt et al., 1998; Lischke et al.
-2006; Medvigy et al.
-2009; Moorcroft et al., 2001; Scherstjanoi et al.
-2014; Smith et al., 2014; Weng et al., 2015).
-*(KAT: I think all of the following content is great, but probably needs to be heavily tightened:)*
+![**Figure `r fig_model`.** Lidar and thermal remote sensing data can be used to measure vertical canopy structure and plant thermal signatures at increasingly large scales.
+(a-b) Integrating in situ and remote sensing measurements will support our ability to scale understanding of vertical gradients within and across ecosystems.
+(c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization.
+Vertically resolved in situ data could include any variable listed in Table 1 or 2.
+[Issue #42](https://github.com/EcoClimLab/vertical-thermal-review/issues/42).
+](schematics/fig_remote_sensing.png)
 
-To review the vertical light environment and thermodynamics of CBMs in more detail, we focus here on the Ecosystem Demography model version 2 (ED2), a model that is actively undergoing continued development and application across a range of ecosystems and landscapes (Moorcroft et al.
-2001, Longo et al 2019 Pt 1 & pt 2).
-ED2 is also the progenitor of the Functionally Assembled Terrestrial Ecosystem Simulator (FATES), the DGVM coupled with the Energy Exascale Earth System Model (E3SM) (Xu & Christoffersen 2017 ) .
-In ED2, cohorts are tracked across multiple size-classes or individuals within the same plant functional type (PFT).
-Each cohort can encounter multiple light environments within a single climatic grid cell depending on patch-level disturbance history and vertical light competition that modulates interactions between plant traits and resource acquisition.
-Heterogeneity in the light environment and canopy structure across patches within a grid cell gives rise to varying horizontal and vertical micro-environments that differ in temperature, humidity, soil moisture, and soil nutrient conditions (Fisher et al 2018).
-The energy, water, and carbon dioxide cycles are solved separately for each patch in ED2, while fluxes and storage associated with individual plants are solved for each cohort within patches (Longo et al 2019).
-As a result, vegetation structure and functional distributions are emergent properties in the model that depend on plant functional traits and their interactions with abiotic environmental conditions.
-Each plant perceives a unique environment, including incident light, temperature, and vapor pressure deficit that varies across local scales in association with the horizontal and vertical position of each cohort relative to other cohorts, in addition to patch level differences in topographic and edaphic conditions.
+**Scaling in situ data with remote sensing**
 
-Because most CBMs are spatially implicit within grid cells, direct and diffuse photosynthetically active radiation (PAR) is distributed to cohorts of differing heights using radiative transfer models that depend on designated rules associated with how crown architecture is represented, rather than via direct spatial competition determined by spatially explicit crown locations (Fisher et al.
-2010?).
-The flat-top crown method used in ED and ED2 stacks cohort-layers vertically, and a radiative transfer model is used to determine radiation absorbed by each layer at its midpoint, meaning that each cohort is shaded by all taller cohorts.
-This is distinct from other models that represent crown architecture using a perfect plasticity approximation (PPA) approach, which assumes that tree crowns completely fill canopy gaps through phototropism (stem leaning) and crown filling (Strigul et al.
-2008).
-Discrete canopy layers are comprised of self-organizing crowns that all receive the same incoming radiation.
-Once the canopy layer is filled with tree crowns, successive shorter trees inhabit the subsequent understory layer and are shaded by the trees in the canopy (Fisher et al 2010/2018?).
-Models using the PPA approach differ in number of understory layers (e.g.
-some only have one) (Fisher et al 2018).
-Though an improvement over big-leaf models, models that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g., seasonal shifts in vertical leaf area distributions, Smith et al.
-2019) and functions (e.g.
-thermal responses, as we present in this paper).
-Hence, multi-layered ecosystem models may be necessary for accurately projecting future forest function.
-(E.g.
-see de Pury & Farquhar, 1997).
+Remote sensing data provide a valuable means to scale between in situ observations and DGVMs.
+Specifically, the increasing availability of airborne and spaceborne LiDAR and thermal remote sensing data offer a promising opportunity for mapping vertical thermal gradients in combination with vegetation structure at locations with in situ data and across larger landscapes (Fig.
+6).
+Airborne and spaceborne lidar, as well as terrestrial laser scanning data yield detailed 3D reconstructions of whole tree and forest structure.
+These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor (Hulley et al.
+2019, Fisher et al.
+2020) or drone- and tower-based FLIR cameras.
+
+Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions (Disney et al.
+2019), evaluate spatial variation in vertical leaf area density profiles (Detto et al.
+2015), and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems (Musselman et al.
+2013).
+In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses (e.g., Davis et al.
+2019).
+Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures (Johnston et al.
+in review).
+(Pau et al.
+2018) used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than air temperatures or VPD.
+In a savanna system in the western U.S., (Johnston et al.
+in review) found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies.
+This finding was the opposite of what was expected and occurred in association with very high grass temperatures.
+The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of in situ measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
 

### 2021-08-08 vertical-thermal-review.Rmd
scaling and modelling section

@@
 Airborne and spaceborne lidar, as well as terrestrial laser scanning data yield detailed 3D reconstructions of whole tree and forest structure.
-These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor (Hulley et al.
-2019, Fisher et al.
-2020) or drone- and tower-based FLIR cameras.
+These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based FLIR cameras.
 
-Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions (Disney et al.
-2019), evaluate spatial variation in vertical leaf area density profiles (Detto et al.
-2015), and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems (Musselman et al.
-2013).
+Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
 In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses (e.g., Davis et al.

### 2021-08-09 vertical-thermal-review.Rmd
added citations for modelling

@@
 
-**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.** Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks (Moorcroft, Hurtt, & Pacala, 2001; Banerjee & Linn 2018; Bonan et al.
-2021).
+**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.** Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
 In addition, this review finds that, although larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may also be stifled under hot, humid conditions.
@@
 
-Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g., seasonal shifts in vertical leaf area distributions, Smith et al.
-2019) and functions (e.g., thermal responses, as we present in this paper).
-At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function (e.g., de Pury & Farquhar, 1997; Bonan et al 2021).
+Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019a] and functions (e.g., thermal responses, as we present in this paper).
+At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @depurySimpleScalingPhotosynthesis1997; @bonanMovingIncorrectUseful2021].
 In addition, capturing the vertical gradients in NPP, GPP, and respiration, and subsequently the net ecosystem effects, requires improved characterization of the functional response of leaf-level processes to vertically varying abiotic conditions, and the role of traits and abiotic conditions in mediating responses to thermal sensitivity.
@@
 Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
-In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses (e.g., Davis et al.
-2019).
-Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures (Johnston et al.
-in review).
-(Pau et al.
-2018) used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than air temperatures or VPD.
-In a savanna system in the western U.S., (Johnston et al.
-in review) found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies.
+In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
+Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020 in review].
+[@pauTropicalForestTemperature2018] used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than air temperatures or VPD.
+In a savanna system in the western U.S., [@johnstonWhatLiesVertical2020 in review] found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies.
 This finding was the opposite of what was expected and occurred in association with very high grass temperatures.

### 2021-08-09 NEON_height_profiles/scripts/verticalProfileFigure.R
Removed unnecessary code to archive

@@ -68,29 +68,11 @@ dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT", "PARPAR"),
 sites <- c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF")
 clrs <- c("#3399FF", "gold", "grey", "red", "#66CCFF", "#003399")
 
-#bring in normalized height (from Marielle)
-meta <- read_excel("data/site_data/TIS site metadata_20190403_forNOAA.xlsx", sheet=1)
-meta <- data.table(meta)
-meta <- meta[SITE %in% sites,
-             ][,`:=` (DistZaxsCnpy = as.numeric(DistZaxsCnpy),
-                    maxCanHeightMarielle = c(32, 43, 39, 21, 53))]
-normHeightData <- 
-  data.table(site=c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF"),
-             normH=c(32, 21, 23, 43, 39, 53))
-# HARV=32, OSBS: 21, PUUM: 23, SCBI: 43, SERC: 39, WREF: 53
-
 plots <- list()
 for(i in 1:length(dp[,name])){
   alldata <- NULL
-  allTopHeight <- c()
-  allNormHeight <- c()
   for(j in 1:length(sites)){
     load(paste0("data/neon_rdata/", sites[j], "test.Rdata"))
-    
-    #get max vertical height (for use in plots)
-    hei <- sapply(full_data, function(x){unique(x[["verticalPosition"]])})
-    topHeight <- max(unlist(hei, use.names = FALSE))
-    normHeight <- normHeightData[site == sites[j], ]
     
     var <- as.data.table(full_data[names(full_data) == dp[,name][i]])
     colnames(var) <- 
@@ -113,6 +95,10 @@ for(i in 1:length(dp[,name])){
     ## originally we were getting delta[var] as being var at height h - var 
     ## at lowest height. We have decided (as of Mar. 2021) to only do raw values.
     ## The delta var code is at the bottom of this script in the Archive section
+    
+    ## Similarly, we initially normalized the height across the site variables,
+    ## but now are using absolute height values. The normalized code is below in
+    ## the Archive section.
     
     ## only calculate raw mean values
     whei <- test[order(day, zOffset),
@@ -137,8 +123,6 @@ for(i in 1:length(dp[,name])){
                  ][, `:=` (var = dp[,name][i], site=sites[j], col=clrs[j])]
     
[…]

### 2021-08-09 vertical-thermal-review.Rmd
added citations for modelling

@@
 The intensity of incoming radiation and VPD in the upper canopy, combined with hydraulic challenges of water transport to the tops of tall trees [**REFS**] can cause $g_s$-limitation and an increase in $T_{leaf}$.
-Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than buffered lower canopy layers [@millerOnlySunlitLeaves2021].
+Therefore, we would expect a strong $T_{leaf}$ gradient  in closed canopies [@millerOnlySunlitLeaves2021].
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory.
@@
 Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
-Understory leaves, which typically experience lower windspeeds (Fig. `r fig_NEON_vertical`), would retain more heat.
-Higher RH conditions in the understory (Fig.
+ In addition to lower wind speeds (Fig.
+`r fig_NEON_vertical`), higher RH conditions in the understory (Fig.
 `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig.
@@
 However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.
-$T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
+In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
 In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [OSBS, Fig.
@@
 **It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures-- at least under conditions of moderate to high VPD.
+Increases in canopy $T_{leaf}$ are shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018].
 A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an El Nino drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.
@@
 This can have implications in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018].
-Such increase in canopy $T_{leaf}$ has also shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018].
 

### 2021-08-09 tree_rings/SCBI_figure.m
delete old tree-ring figure material
#29

@@ -1,136 +1,0 @@
-% Kristina Anderson-Teixeira
-% June 2020
-
-%% INITIATE
-clear all; clf; clc; clear; close all;
-
-%% READ IN DATA & pull out variables
-table=readtable('correlation_with_CRU_SCBI_1901_2016_climate_data.csv');
-column_names=table.Properties.VariableNames;
-%table.sp%table.month
-%table.variable
-%table.CP
-n_records=height(table);
-species=unique(table.sp);
-n_sp=length(species);
-months = {'curr.may' 'curr.jun' 'curr.jul' 'curr.aug'};
-n_mo=length(months);
-variables= {'tmp' 'tmn' 'tmx'};
-n_var=length(variables);
-
-%% PULL OUT RECORDS OF INTEREST
-
-%create matrices to store rearranged values (rows are species, in alphabetical order; columns months in order)
-tmp_und=NaN*ones(n_sp, n_mo);
-tmp_can=NaN*ones(n_sp, n_mo);
-tmn_und=NaN*ones(n_sp, n_mo);
-tmn_can=NaN*ones(n_sp, n_mo);
-tmx_und=NaN*ones(n_sp, n_mo);
-tmx_can=NaN*ones(n_sp, n_mo);
-
-
-for s=1:n_sp% cycle through species
-    for m= 1:n_mo % cycle through months 
-        
-        %tmp-und
-        ind=find (strcmp(char(string(species(s))),cellstr(table.sp))==1 & ...
-                   strcmp('subcanopy',cellstr(table.CP))==1 &...
-                   strcmp('tmp',cellstr(table.variable))==1 & ...
-                    strcmp(char(string(months(m))), cellstr(table.month))==1);
-        if isempty(ind)==0
-            tmp_und(s,m)= table.coef(ind);
-        end
-         
-        
-        %tmp-can
-        ind=find (strcmp(char(string(species(s))),cellstr(table.sp))==1 & ...
-                   strcmp('canopy',cellstr(table.CP))==1 &...
-                   strcmp('tmp',cellstr(table.variable))==1 & ...
-                    strcmp(char(string(months(m))), cellstr(table.month))==1);
-        if isempty(ind)==0
-            tmp_can(s,m)= table.coef(ind);
-        end
-        
-        %tmn-und
-        ind=find (strcmp(char(string(species(s))),cellstr(table.sp))==1 & ...
[…]

### 2021-08-10 vertical-thermal-review.Rmd
inserted caption for fig 4

@@
 Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@rollinsonClimateSensitivityUnderstory; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
@@
 @mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; Ishida et al.
-1999; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021
+1999; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013
 ---
@@
 nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al.
-1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @athertonSpatialVariationLeaf2017
+1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006;@athertonSpatialVariationLeaf2017
 ---
@@
 ![**Figure `r fig_treerings`.
-Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.
-In column (a), large trees have a more negative response to high temperatures during the current or previous growing season.
-Shown are tree ring width (RW) responses to the most influential temperature variable at the site ($T_{max}$ or PET) of the largest versus smallest trees included in GLS model with a DBH x temperature interaction (Anderson-Teixeira et al., in review).
-In column (b), understory trees have more negative responses to high growing season temperatures across eight New England forests, (@rollinsonClimateSensitivityUnderstory2021).
-[ISSUE # 29](https://github.com/EcoClimLab/vertical-thermal-review/issues/29).](tree_rings/Figure_4_tree_rings.png)
+Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species.
+Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.
+Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction.
+Other model terms are held constant at their mean.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable.
+From Anderson-Teixeira et al.
+(in revision).
+In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests.
+Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively.
+Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals.
+From @rollinsonClimateSensitivityUnderstory2020.
+](tree_rings/Figure_4_tree_rings.png)
 
@@
 While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
-Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration (Fig 5a).
-In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
+Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration (Fig 5a; Kunert et al.
+2017), and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
+In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime ~CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
 Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig.
@@
 Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013].
+Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @greiserHidingClimateCharacterizing2019].
 This imposes an increasing threat to the biodiversity in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].

### 2021-08-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Ian McGregor^3^,
-Elsa M. Ordway^4^,
+Elsa M. Ordway^4,5^,
 Marielle N. Smith^6^,
@@
 Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].

### 2021-08-18 vertical-thermal-review.Rmd
Lawren's edits

@@
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
-Collectively, these impacts may lead to reduced forest carbon sequestration, as may already be occurring in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
-The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016], yet this influence remains extremely uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
+Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already be occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
+The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
 
-**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
+**A great measure of uncertainity arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as because small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
 Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
-Overstory trees form canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including understory microclimates by buffering extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019].
+Overstory canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019].
 This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
@@
 (3) the leaf traits that most strongly influence $T_{leaf}$;
-(4) leaf metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
+(4) leaf metabolism and gas exchange, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
 (5) the size-structuring of whole-tree and ecosystem function.
@@
 Vertical gradients in the biophysical environment for six sites six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**.
-Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
+Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawai'i; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
 Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig.
@@
 `r fig_NEON_vertical`).**
-Foliage distribution is structured in correspondence with the distribution of tree sizes and tree architecture within a forest, where leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs.
+Foliage distribution is determined by the distribution of tree sizes and tree architecture within a forest, and leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs.
 `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`).
@@
 
-Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
 Leaves can be cooler than the air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
-While $T_{leaf}$ is rarely exactly equal to $T_{air}$ (poikilothermy), it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
 
@@
 
-**Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorological conditions (Fig. `r fig_leaf_T`).**
+**Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ varies with micrometeorological conditions (Fig. `r fig_leaf_T`).**
 Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
@@
  In addition to lower wind speeds (Fig.
-`r fig_NEON_vertical`), higher RH conditions in the understory (Fig.
+`r fig_NEON_vertical`), higher RH in the understory (Fig.
 `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig.
@@
 The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
+A study of tropical oaks found decreasing stomatal conductive capacity (based on morphology) with increasing site mean $T_{air}$ (Lin et al., 2021), a response likely linked to water conservation under increasing VPD.
+The extent to which maximum conductance (and its leaf cooling potential) increases with light may therefore depend on the specific thermal microenvironments arrayed across the profile
 

### 2021-08-19 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
-Leaves can be cooler than the air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
-While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].
+Leaves are typically cooler than the air at night, and during the day especially under cloudy skies, at high windspeeds and high $T_{air}$.
+Leaves are typically warmer than air when under full sunlight, especially under slow windspeeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
+While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020].
 
@@
 **Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig.
-`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
+`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].**
 High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig.
@@
 `r fig_leaf_T`b].
-Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is primarily regulated by stomatal conductance ($g_s$), but also by RH through a direct effect on the leaf-to-air vapor gradient (Fig.
-`r fig_leaf_T`c) and an indirect effect on $g_s$.
-Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a]; decreases with leaf size (Fig. `r fig_leaf_T`d);
+Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD; notably, $g_b$ is influenced by windspeed and $g_s$ respond due to stomatal closure under high VPD.
+Therefore, under full sunlight, $T_{leaf}-T_{air}$ decreases with $g_b$, increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a]; decreases with leaf size (Fig. `r fig_leaf_T`d);
 decreases with $g_s$ (Fig. `r fig_leaf_T`e); and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
@@
 Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$-limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
-With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling under high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
 
@@
 `r fig_leaf_T`b-e).
-In contrast, greater RH and lower wind speeds in the understory may stifle $\lambda E$ (through reduced $g_s$ and $g_a$), thereby increasing $T_{leaf}$ [Fig.
+In contrast, greater RH and lower wind speeds in the understory may stifle $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

### 2021-08-22 vertical-thermal-review.Rmd
more edits

@@
 **Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs.
-`r fig_NEON_vertical`b-d, `r SIfig_NEON_all`, @koike_leaf_2001].**
+`r fig_NEON_vertical`a-d, `r SIfig_NEON_all`, @koike_leaf_2001].**
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
@@
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for both water and heat loss (see section "Leaf temperature").
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").
 
@@
 High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig.
-`r fig_leaf_T`a).Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance ($g_a$), which is greater in smaller leaves (Fig.
-`r fig_leaf_T`d) and higher wind speeds [Fig.
-`r fig_leaf_T`b].
+`r fig_leaf_T`a).
+Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig.
+`r fig_leaf_T`d) and higher wind speeds (Fig.
+`r fig_leaf_T`b).
 Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD; notably, $g_b$ is influenced by windspeed and $g_s$ respond due to stomatal closure under high VPD.
-Therefore, under full sunlight, $T_{leaf}-T_{air}$ decreases with $g_b$, increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a]; decreases with leaf size (Fig. `r fig_leaf_T`d);
-decreases with $g_s$ (Fig. `r fig_leaf_T`e); and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
+Therefore, under full sunlight, $T_{leaf}-T_{air}$ decreases with $g_b$, which increases with wind speed [Fig.
+`r fig_leaf_T`b, @daudetWindSpeedLeaf1999a] and $g_s$ (Fig.
+`r fig_leaf_T`e); increases with leaf size (Fig.
+`r fig_leaf_T`d);
+decreases with and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
 
@@
 When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
-However, when leaf water demand (influenced by VPD and photosynthetic CO~2~ demand) exceeds the rate of supply, conservation of water ($g_s$-limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
+However, when leaf water demand (influenced by VPD and stomatal opening) exceeds the rate of supply, conservation of water ($g_s$-limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`e, @fauset_differences_2018].
 Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$-limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
-With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling under high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+With adequate water, species with smaller leaves (higher $g_b$) and larger $g_s$ will be most effective at cooling under high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
 
@@
 Collectively, higher wind speed (Fig.
-`r fig_NEON_vertical`), and with adaptive leaf traits increase $g_a$ and $g_s$ (reviewed below) mediate the direct effects of solar radiation on upper canopy leaves (Fig.
+`r fig_NEON_vertical`), and with adaptive leaf traits increase $g_b$ and $g_s$ (reviewed below) mediate the direct effects of solar radiation on upper canopy leaves (Fig.
 `r fig_leaf_T`a) and result in higher rates of heat loss (Fig.
@@
 In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling.
-Under lower wind conditions, the thickness of the air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig.
+Under lower wind conditions, the thickness of the air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_b$ and increasing $T_{leaf} - T_{air}$ [Fig.
 `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020].
@@
 In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
+In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
 `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
@@
 **Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].**
-Leaf morphology adjusts to the local microenvironment during leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of direct to diffused light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
+Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
@@
 However, when it comes to understanding and modeling forest ecosystem function, given that species traits strongly shape the propensity for trees to inhabit understory vs.
-canopy positions, ecosystem function is also shaped by interspecific trait variation (e.g., *Lamour, Serbin in prep*).
+canopy positions, ecosystem function is also shaped by interspecific trait variation [e.g., *Lamour, Serbin in prep*; @cavaleri_height_2010].
 
@@
 
-Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at an anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-As much of our understanding of trait coordination is based on the leaf economics spectrum concept [@wrightWorldwideLeafEconomics2004] which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the canopy.
-It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
-For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+As much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004] which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
+It has been established that independently of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
+For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
 Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
@@
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but they are more likely to overheat than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies, where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and VPD stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].
+In open canopies, where light is comparatively homogeneous, exposed leaves are shaped more by $T_{air_{max}}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 
-Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
+Elemental concentrations such as $N_{area}$ and $P_{area}$, contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021].
 Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018].
-Therefore, we would expect elemental traits to influence vertical foliar thermal sensitivity;  however, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.
+However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of N and P and foliar thermal tolerance.
 
@@
 The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology.
-For example, deciduousness is an adaptation to conserve water during dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile.
+For example, deciduousness is an adaptation to conserve water during cold or dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile.
 Within tropical deciduous species, the fraction of seasonally deciduous individuals increases toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
@@
 Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
-However, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
+For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
+Further, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
 Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.

### 2021-08-22 NEON_height_profiles/scripts/get_neon_data.R
Update get_neon_data.R

@@ -61,7 +61,7 @@ get_NEON <- function(x="SCBI"){
       loadByProduct(dpID=dp[,id][i], 
                     site=x,
                     package="basic", 
-                    check.size = FALSE, timeIndex=30,
+                    check.size = FALSE, avg=30,
                     nCores = 2)
     # startdate=paste0(date[,year][j], "-0",
     #                  date[,month][j]),

### 2021-08-22 vertical-thermal-review.Rmd
ecology edits

@@
 (4) leaf metabolism and gas exchange, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
-(5) the size-structuring of whole-tree and ecosystem function.
+(5) whole-tree and ecosystem ecology.
 *We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.*
@@
 
-## Size-structuring within the ecosystem
+## Whole-tree and ecosystem ecology
 
@@
 **Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
-Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture in forests not experiencing soil drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
 Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
@@
 `r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
-In addition to lower drought resistance of growth, larger trees also frequently exhibit lower ability to bounce back from stress, ie.
+In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress, ie.
 resilience [@mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
@@
 
-**It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures-- at least under conditions of moderate to high VPD.
+**It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be sensitive to high temperatures-- at least under conditions of moderate to high VPD.
 Increases in canopy $T_{leaf}$ are shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018].
 A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an El Nino drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.
-It remains far less clear how thermal sensitivity is likely to vary across forest strata under wet conditions, but both physiological mechanisms (Fig.
+It remains far less clear how thermal sensitivity varies across forest strata under wet conditions, but both physiological mechanisms (Fig.
 3) and tree ring evidence (Fig.
@@
 The complex interwoven relations between the biophysical environment and biological factors--leaf traits, metabolic processes and distribution of trees-- are likely to produce amplifications and feedback loops in a warming world, with implications for forests on many levels.
+
+### warming
+
+Warming impacts forests, from leaf metabolism to whole forest structure.
+Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$.
+Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima.
+Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
+Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures (e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
+A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [Miller et al.
+2021], especially if RH in the understory does not inhibit transpirational cooling.
+Thermal limits are especially narrow in tropical and mid-latitudes forests (O'Sullivan et al.
+2017).
+A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [O'Sullivan et al.
+2020].
+In the understory, warming is disproportionately affecting the less thermally-adapted functional types, resulting in associated plant community thermophilization (Zellweger et al., 2019; Duque et al., 2015; Greiser et al.
+2019).
+Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al.
+2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes (Scheffers et al., 2013; Greiser et al.
+2019).
 
@@
 This imposes an increasing threat to the biodiversity in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
-
-### warming
-
-Warming impacts forests, from leaf metabolism to whole forest structure.
-At a leaf metabolism level, global analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled up to photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, thereby shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-However, taken into account light-induced lower leaf respiration in the day, partionationing of carbon losses differ diurnally and seasonally across ecosystems [@keenanWidespreadInhibitionDaytime2019].
-However, nitrogen that is particularly essential for plant metabolic processes has also shown to decline in forests with warming [Pam Templer et al.].
-Temperature driven photosynthesis limitation is also reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominated by younger trees [@mcdowellPervasiveShiftsForest2020].
-Further, combination of warm-season associated with increased VPD, along with drought exacerbates tree stress[@parkwilliamsTemperaturePotentDriver2013] leading to mortality [@gouldenCaliforniaForestDieoff2019], with implications for forest structure.
-[**to include Teskey 2015; Breashers et al.
-2021**]
 

### 2021-08-22 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures (e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
-A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [Miller et al.
-2021], especially if RH in the understory does not inhibit transpirational cooling.
-Thermal limits are especially narrow in tropical and mid-latitudes forests (O'Sullivan et al.
-2017).
-A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [O'Sullivan et al.
-2020].
-In the understory, warming is disproportionately affecting the less thermally-adapted functional types, resulting in associated plant community thermophilization (Zellweger et al., 2019; Duque et al., 2015; Greiser et al.
-2019).
+A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021], especially if RH in the understory does not inhibit transpirational cooling.
+Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017].
+A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
+In the understory, warming is disproportionately affecting the less thermally-adapted functional types, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; Greiser et al.
+2019].
 Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al.
-2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes (Scheffers et al., 2013; Greiser et al.
-2019).
+2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; Greiser et al.
+2019].
 

### 2021-08-23 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].**
-Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
+Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
@@
 nocite: |
-@mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; Ishida et al.
-1999; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013
+@mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013
 ---
@@
 
-Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, hairier, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, more pubescent, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
 These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
@@
 nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al.
-1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006;@athertonSpatialVariationLeaf2017
+1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006;@athertonSpatialVariationLeaf2017; @sackHydrologyLeavesCoordination2003
 ---
@@
 - However, when we ask the critical question of how the temperature sensitivities  of metabolism and woody growth varies across these different environments, we can't say much-- there are only a small number of studies, often showing insignificant differences or mixed results.
-- We lean towards the conclusion that while large canopy trees are the most vulnerable to warming when water is limited, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
+- The preponderance of available data suggest that while large canopy trees are the most vulnerable to warming when water is limited, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
 - Critically, change to the vertical gradient--e.g., through increased mortality of canopy trees--will change processses across forest strata, with potentially dramatic consequences.

### 2021-08-23 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 nocite: |
-@mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013
+@mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988
 ---
@@
 
-Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, more pubescent, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*].
+Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, more pubescent, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].
 These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.

### 2021-08-24 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
updated with gs conversion and units

@@ -307,7 +307,7 @@ c1<-ggplot(gs)+
   geom_smooth(aes(x = gs, y = u_tla),  method = lm, color = "#008837", se = FALSE)+
   geom_smooth(aes(x = gs, y = dr_ltla),  method = lm, color = "#7b3294", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = gs, y = dr_utla),  method = lm, color = "#008837", se = FALSE, linetype = "dashed")+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("stomatal conductance ($g_{s}$, $\\mu mol/m^2/s/Pa$)"))+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("stomatal conductance ($g_{s}$, mmol $m^{-2} s^{-1}$)"))+
   geom_hline(yintercept=0, linetype='dotted', col = 'black')+
   theme_few()+theme(text = element_text(size = 14))+ylim(-5, 14)
 c1
@@ -441,10 +441,10 @@ e1
 
 
 table<- data.frame(biophysical = c( "swr", "ws", "rh", "lw", "gs", "tair"),
-                   normalno = c("871", 2.88, 0.91, 0.04, "4.0", "25"),
-                   normal_u = c("102", 0.24, 0.97, 0.10, "2.0", "23"),
-                   drought_o = c("1151", 2.88, 0.46, 0.04, "0.01", "25"),
-                   normalno = c("153", 0.24, 0.49, 0.10, "0.01", "23"))
+                   normalno = c("871", 2.88, 0.91, 0.04, "404", "25"),
+                   normal_u = c("102", 0.24, 0.97, 0.10, "202", "23"),
+                   drought_o = c("1151", 2.88, 0.46, 0.04, "1.01", "25"),
+                   normalno = c("153", 0.24, 0.49, 0.10, "1.01", "23"))
 
 colnames(table) <- c("variable",
                  "overstory\nnormal", "understory\nnormal", "overstory\ndrought", "understory\ndrought")

### 2021-08-24 vertical-thermal-review.Rmd
schematic 1 updates
Update vertical-thermal-review.Rmd

@@
 **Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to rise a $1.6 - 4.3^{\circ}$C above pre-industrial levels by 2100 (IPCC 2014), accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed an average of $1.5^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
-Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already be occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
+Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
 The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
@@
 Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
-Overstory canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme temperatures, solar insolation, wind, and other meteorological conditions [@zellweger_seasonal_2019].
+Overstory canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
 This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
@@
 Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy  [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
-](schematics/Screen Shot 2021-07-19 at 1.10.17 PM.png)
+](schematics/Schematic figure 1.png)
 
@@
 Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawai'i; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
-Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig.
+Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Fig.
 S1.](NEON_height_profiles/figures/profile_all.png)
@@
 `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
-In open canopy forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig.
+In open-canopied forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig.
 `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
@@
 open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @smithSeasonalDroughtrelatedChanges2019; @parkerConsequencesEnvironmentalHeterogeneity2019].
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019].
 In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.
@@
 ---
-nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; Ishida et al.
-1999; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006;@athertonSpatialVariationLeaf2017; @sackHydrologyLeavesCoordination2003
+nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006; @athertonSpatialVariationLeaf2017; @sackHydrologyLeavesCoordination2003
 ---
@@
 Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures (e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
+Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
 A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021], especially if RH in the understory does not inhibit transpirational cooling.
@@
 A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
-In the understory, warming is disproportionately affecting the less thermally-adapted functional types, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; Greiser et al.
-2019].
+In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
 Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al.
-2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; Greiser et al.
-2019].
+2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
 
@@
 
-Canopies are crucial to microclimatic functioning, as they buffer several aspects of forest microenvironments [Zelleweger et al.
-2019].
-This buffering is influenced by the hydraulogical-cycle at large [Davis et al.
-2019], therefore severe droughts place canopies at risk, particularly as larger trees suffer more during droughts[Bennett et al.
-2015].
-Further, increasing global temperatures show that canopies of temperate and tropical forests are already functioning close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019].
-This can have implications in the tropics, where thermal limits are narrower than in temperate forests [Perez and Feeley, 2018].
+Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019].
+This buffering is influenced by hydrology, as wetter sites and seasons allow for denser, and more buffered canopies  [@davisMicroclimaticBufferingForests2019].
+Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015].
+The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging (Laurance et al.
+2006; Miller et al.
+2004), although fire tends to predominantly affect smaller trees (Brando et al.
+2019).
+Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
 
-Canopy structure affects understory species composition and soil fertility, which has been shown to shift under macroclimatic warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-Macroclimatic warming will disproportionately affect the less warm-adapted functional types within the forest canopy, even with buffering of climatic extremes, thereby creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), with associated understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @greiserHidingClimateCharacterizing2019].
-This imposes an increasing threat to the biodiversity in microrefugias that are  otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013].
+In general, disturbances (e.g., drought, fire, fragmentation, and logging) reduce canopy cover, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020].
+This shift to hotter and drier microclimates often makes forests more susceptible to further disturbances, for example, increasing fire risk (Brando et al.
+2014; Aragao et al.
+2018).
+Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization'; Silverio et al.
+2013) and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 
-Feedback loops from loss of big trees through natural or anthropogenic disturbance (e.g.
-@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020].
-Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-
-Variation in resistance and resilience among forests may drive different trajectories from the forests in the face of global climate change.
+Variation in resistance and resilience among forests may drive different trajectories in the face of global climate change.
 For example, African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021].
-Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019; @cunninghamTemperateRainforestTrees2003; @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] are other mechanisms that will help forests cope with warming.
-
-*For example, species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020].*
-
--Isoprenes *providing emitting species with a selective advantage in a warming world* [*REF*].
+Species with smaller leaves and larger gs will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020] and isoprene emitting species will have a selective advantage in a warming world [@taylorCapacityEmitIsoprene2019].
+Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] will also help forests cope with warming.
 

### 2021-08-25 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ![**Figure `r fig_schematic`.
-Schematic summarizing vertical gradients in forests with dense canopies during daytime growing season conditions.** Patterns summarized here tend to be weaker, or sometimes reversed, under more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
+Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature, (3) leaf traits, (4) leaf metabolism, and (5) whole-tree and ecosystem ecology.
+Abbreviations are as follows: VPD: vapor pressure deficit;  Tleaf - Tair: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; temp sensitivity: temperature sensitivity; ET: evapotranspiration, GPP: gross primary productivity.** Here, we summarize conditions in dense canopies during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
 Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).

### 2021-08-25 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
-Overstory canopies that play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
+Overstory canopies play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
 This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
@@
 ![**Figure `r fig_NEON_vertical`.
-Vertical gradients in the biophysical environment for six sites six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**.
+Vertical gradients in the biophysical environment for six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**.
 Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawai'i; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV).
+Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawai'i; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
 Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Fig.
@@
 `r fig_NEON_vertical`).**
-Foliage distribution is determined by the distribution of tree sizes and tree architecture within a forest, and leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Figs.
+Foliage distribution is determined by the distribution of tree sizes and tree architecture within a forest, and leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient that are heterogeneous across forests (Figs.
 `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`).
@@
 `r fig_leaf_T`b-e).
-In contrast, greater RH and lower wind speeds in the understory may stifle $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig.
+In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

### 2021-08-26 vertical-thermal-review.Rmd
Marielle's citations and ET imag

@@
 Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed an average of $1.5^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
@@
 2017).
-I'd like to create a parallel figure for C fluxes.](schematics/ecosystem vertical partitioning/ecosystem partitioning_updated.png)
+I'd like to create a parallel figure for C fluxes.](schematics/ecosystem vertical partitioning/raw from Nobby/Fig_5_ET.png)
 
@@
 Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015].
-The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging (Laurance et al.
-2006; Miller et al.
-2004), although fire tends to predominantly affect smaller trees (Brando et al.
-2019).
+The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging [[@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019].
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
@@
 In general, disturbances (e.g., drought, fire, fragmentation, and logging) reduce canopy cover, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020].
-This shift to hotter and drier microclimates often makes forests more susceptible to further disturbances, for example, increasing fire risk (Brando et al.
-2014; Aragao et al.
-2018).
-Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization'; Silverio et al.
-2013) and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+This shift to hotter and drier microclimates often makes forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018].
+Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
 

### 2021-08-26 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 **Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
-Wind speeds increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018] to savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs.
-`r fig_NEON_vertical`e, `r SIfig_NEON_all`).
+Wind speeds increase with height across forest types, including closed and open-canopy forests  [Figs.
+`r fig_NEON_vertical`e, `r SIfig_NEON_all`; @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
 Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Figs.
 `r fig_NEON_vertical`, `r SIfig_NEON_all`).
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").
 
@@
 `r fig_leaf_T`e); increases with leaf size (Fig.
-`r fig_leaf_T`d);
-decreases with and increases with VPD, or increases with RH (Fig. `r fig_leaf_T`c).
+`r fig_leaf_T`d); and increases with RH (Fig.
+`r fig_leaf_T`c).
 

### 2021-08-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**A great measure of uncertainity arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as because small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
+**A great measure of uncertainity arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
 Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
@@
 Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawai'i; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
+Sites include a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
 Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Fig.
@@
 `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
-In open-canopied forests, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig.
+In open-canopied forests ie.
+forests with lower density of trees and open canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig.
 `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
@@
 
-**Leaf energy balance is a tradeoff between water-use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
+**Leaf energy balance is a tradeoff between water uptake and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
 When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
@@
 `r fig_leaf_T`).
-However, higher wind speeds (e.g., up to ~ 4 $ms^{-1}$ at NEON sites; Fig.
+However, higher wind speeds (e.g., up to ~ 4 $ms^{-1}$; Fig.
 `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ $4^\circ$C [Fig.
@@
 `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020 (in review)].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
 Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
-
-It is noteworthy that leaf and branch temperatures often differ substantially.
-For instance, exposed tropical tree bark can be much warmer than leaves [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021; @johnstonWhatLiesVertical2020].
-Thus, vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.
-`r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories).
-Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021; @macleanMicroclimcMechanisticModel2021].
 
@@
 Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
+For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
+However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
 Further, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
@@
 **Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
-Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) than by low soil moisture in forests not experiencing soil drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
 Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
@@
 
-**It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be sensitive to high temperatures-- at least under conditions of moderate to high VPD.
-Increases in canopy $T_{leaf}$ are shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018].
+**It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be sensitive to high temperatures-- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020].
+Further, increases in canopy $T_{leaf}$ are also shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018].
 A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an El Nino drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.
@@
 Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
-A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021], especially if RH in the understory does not inhibit transpirational cooling.
+A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.
 Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017].

### 2021-08-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
+**These basic biophysical principles shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
 High radiation at the top of the vertical profile (Fig.
-`r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig.
+`r fig_NEON_vertical`) implies that canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig.
 `r fig_leaf_T`).
-However, higher wind speeds (e.g., up to ~ 4 $ms^{-1}$; Fig.
-`r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ upto ~ $4^\circ$C [Fig.
+However, higher wind speeds (e.g., ~ 4 $ms^{-1}$; Fig.
+`r fig_NEON_vertical`) can reduce $T_{leaf}-T_{air}$ to ~ $4^\circ$C [Fig.
 `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
@@
 **Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).**
-Closed canopies with high LAI act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
+Closed canopies with high LAI act as an parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
 The intensity of incoming radiation and VPD in the upper canopy, combined with hydraulic challenges of water transport to the tops of tall trees [**REFS**] can cause $g_s$-limitation and an increase in $T_{leaf}$.

### 2021-08-28 vertical-thermal-review_SI.Rmd
Update vertical-thermal-review_SI.Rmd

@@
 
+S1 Note:
+
+It is noteworthy that leaf and branch temperatures often differ substantially.
+For instance, exposed tropical tree bark can be much warmer than leaves [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021; @johnstonWhatLiesVertical2020].
+Thus, vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.
+`r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories).
+Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021; @macleanMicroclimcMechanisticModel2021]
+
 METHODS ON NEON ANALYSIS GO HERE.

### 2021-08-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Closed canopies with high LAI act as an parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-The intensity of incoming radiation and VPD in the upper canopy, combined with hydraulic challenges of water transport to the tops of tall trees [**REFS**] can cause $g_s$-limitation and an increase in $T_{leaf}$.
-Therefore, we would expect a strong $T_{leaf}$ gradient  in closed canopies [@millerOnlySunlitLeaves2021].
-In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory.
+In contrast to closed canopies, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory.
 This mixing and light penetration can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy.
@@
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under hot and wet conditions, higher wind speeds and $g_a$ enable cooling.
-Under lower wind conditions, the thickness of the air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_b$ and increasing $T_{leaf} - T_{air}$ [Fig.
+In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling.
+Under lower wind conditions, however,  lower $g_b$ increases $T_{leaf} - T_{air}$ [Fig.
 `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020].

### 2021-08-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
-In open-canopied forests ie.
-forests with lower density of trees and open canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy layers [e.g., PUUM and WREF, Fig.
+In forests with more open upper canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig.
 `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
-Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns and further influence the formation of closed vs.
-open canopy forests [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
+Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019].
@@
 **Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
-Wind speeds increase with height across forest types, including closed and open-canopy forests  [Figs.
+Wind speeds increase with height across forest types [Figs.
 `r fig_NEON_vertical`e, `r SIfig_NEON_all`; @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
@@
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-In open forests, midday stomatal closure can occur in both upper canopy and lower canopy leaves, with stomata of lower canopy leaves remaining closed for a longer duration than upper canopy leaves, resulting in elevated $T_{leaf}$ [OSBS, Fig.
-`r fig_leaf_T`,@zweifelMiddayStomatalClosure2002].
 The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
-A study of tropical oaks found decreasing stomatal conductive capacity (based on morphology) with increasing site mean $T_{air}$ (Lin et al., 2021), a response likely linked to water conservation under increasing VPD.
-The extent to which maximum conductance (and its leaf cooling potential) increases with light may therefore depend on the specific thermal microenvironments arrayed across the profile
 
@@
 **Vertical gradients affect tree metabolism and growth.**
-Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
+Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
 Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).

### 2021-09-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`).**
-Foliage distribution is determined by the distribution of tree sizes and tree architecture within a forest, and leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient that are heterogeneous across forests (Figs.
+Foliage distribution within a forest is determined by the distribution of tree sizes and tree architecture, along the vertical gradient by leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns that are heterogeneous across forests (Figs.
 `r fig_NEON_vertical`a-b, `r SIfig_NEON_all`).
@@
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
-Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019a; @monsonLeafIsopreneEmission2021].
+Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].
 
@@
 **Leaf photosynthetic thermal tolerance**
-Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020].Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2020].
+Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020].
+Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2020].
 Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
@@
 Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
-Along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
+In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
 Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
@@
 shade leaves [@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
-This may vary based on forest type [**REF**], leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+This may vary based on forest type [**REF**], acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015], or leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
 Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021].
@@
 Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
-The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-The proportion of GPP allocated to many of these increases with tree size, including ...
-foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].
-As a result of increasing carbon allocation to these other functions, mass-specific - and sometimes even absolute - woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraCarbonCyclingMature2021].
-However, NSC reserves increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].
+The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
+Within a stand, woody growth rates , measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, Stephenson et al.
+2014].
+NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].
 
@@
 
-The interannual climate sensitivity of annual woody growth can be determined through analysis of tree-ring increments.
-While it is generally accepted that large, exposed trees display the greatest climate sensitivity and are therefore best for climate reconstructions [@frittsTreeRingsClimate1976], only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
+In turn, the climate sensitivity of woody growth tends to be greater in canopy trees.
+In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976].
+However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
 These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [**Fig.
 `r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
-In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress, ie.
-resilience [@mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
+In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [ie.
+resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
 Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
-
-There are, however, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
-In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2021].
+However, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
+In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
 Potential mechanisms for these patterns include stifling of understory $\lambda E$ under low wind, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig.
@@
 Furthermore, warming-induced increases in nighttime respiratory carbon loss may be particularly disadvantageous for understory trees that are operating closer to their light-compensation point than canopy trees.
-Additional research is required to understand these mechanisms and to disentangle tree growth responses to hot-wet versus hot-dry conditions.
+Additional research is required to understand the mechanisms underlying these intriguing differences, and to disentangle size-related tree growth responses to hot-wet versus hot-dry conditions.
 

### 2021-09-01 vertical-thermal-review.Rmd
quick pass at big-picture SI needs
issue #53

@@
 appendix_tealeaves_methods <- "S2"
-appendix_lit_review_methods <- "S3"
 
@@
 appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
-appendix_lit_review_methods_title<- "Methods for literature review"
 
@@
 
-Appendix `r appendix_lit_review_methods`. `r appendix_lit_review_methods_title`
-
 SIfig_example_legend

### 2021-09-01 vertical-thermal-review_SI.Rmd
quick pass at big-picture SI needs
issue #53

@@
 
+# MAIN MANUSCRIPT DISPLAY ITEMS  (limit 8)
+#figures
+fig_schematic = 1
+fig_NEON_vertical =  2
+fig_leaf_T = 3
+fig_treerings= 4 
+fig_model=5
+
+#tables
+table_leaf_traits=1
+table_leaf_metabolism=2
+
 #Appendices
@@
 appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
-appendix_lit_review_methods_title<- "Methods for literature review"
 
 #SI table numbers
-SItable_example = "S1"
+Tealeaves_parameters = "S1"
 
 #SI table captions
-SItable_example_legend <- "Table S1. ...."
+Tealeaves_parameters_legend <- "Table S1. Tealeaves parameters"
 
@@
 #SI figure captions
-SIfig_example_legend <- "Figure S1. Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)"
+SIfig_NEON_all_legend <- "Figure S1. Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)"
 
@@
 
-METHODS ON TEALEAVES GO HERE.
+Energy balance for a typical overstory sun leaf and understory shade leaf were modeled in the R package *Tleaves* [@muir_tealeaves_2019], parameterized for *Quercus rubra* L.
+leaves at Harvard Forest, MA, USA (**GEOGRAPHIC COORDINATES).
+Parameters are presented in Table `r Tealeaves_parameters`.
 
-Energy balance for a normal overstory sun leaf and understory shade leaf were parameterized with biophysical variables from Harvard NEON data documented here: (https://github.com/EcoClimLab/vertical-thermal-review/blob/master/NEON_height_profiles/HARV_neon_stats.csv).
-Leaf trait measurements for leaf width, ie.
-measured as leaf characteristic dimmension, was measured on ~Quercus rubra~ sun and shade leaf, stomatal conductance measurements were referred from Tleaves typical sun and shade measurements [@muir_tealeaves_2019] and as well @cavender-baresChangesDroughtResponse2000.
+*Describe where you came up with all the parameters:*
+
+- micromet from NEON (explain exactly what you used)
+- leaf characteristic dimension, was measured on *Quercus rubra* sun and shade leaf, 
+- stomatal conductance measurements were referred from Tleaves typical sun and shade measurements [@muir_tealeaves_2019] 
+- @cavender-baresChangesDroughtResponse2000.
 
@@
 \newpage
+# `r Tealeaves_parameters_legend `
 
-#  Appendix `r appendix_lit_review_methods`. `r appendix_lit_review_methods_title`
-
-METHODS ON LIT REVIEW GO HERE.
+Make a table from this:
+https://github.com/EcoClimLab/vertical-thermal-review/blob/master/NEON_height_profiles/HARV_neon_stats.csv
 
@@
 
-# `r SIfig_example_legend`
+# `r SIfig_NEON_all_legend`
 ![**Figure `r SIfig_NEON_all`.
@@
 
+\newpage
+
 ## References

### 2021-09-01 vertical-thermal-review_SI.Rmd
knit in tealeaves parameters table
@NidhiVinod , FYI, and a question: did you use all of these, or just a subset? And is there a table of other parameters?
select & order columns for table

@@
 
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 9)
+parameters_table <- read.csv("NEON_height_profiles/HARV_neon_stats.csv", stringsAsFactors = FALSE, check.names = FALSE)
+parameters_table <- parameters_table[c(9, 1, 3:6)] #select and order columns for inclusion
+ft <- flextable(parameters_table)
+ft <- autofit(ft)
+ft
+```
+
 \newpage

### 2021-09-01 vertical-thermal-review_SI.Rmd
Update vertical-thermal-review_SI.Rmd

--- 2021-09-01 vertical-thermal-review_SI.Rmd
@@
 
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 9)
+parameters_table <- read.csv("NEON_height_profiles/HARV_neon_stats.csv", stringsAsFactors = FALSE, check.names = FALSE)
+parameters_table <- parameters_table[c(9, 1, 3, 5)] #select and order columns for inclusion
+ft <- flextable(parameters_table)
+ft <- autofit(ft)
+ft
+```
+
 \newpage

### 2021-09-01 vertical-thermal-review_SI.Rmd
reinstate lit review methods appendix
issue #53

@@
 
-#Appendices
+# SI
+## appendices
+
 appendix_NEON_methods <- "S1"
 appendix_tealeaves_methods <- "S2"
+appendix_lit_review_methods <- "S3"
 
@@
 appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
+appendix_lit_review_methods_title<- "Methods for literature review"
 
@@
 \newpage
+# `r appendix_lit_review_methods_title`
+
+\newpage
 # `r Tealeaves_parameters_legend `

### 2021-09-01 vertical-thermal-review_SI.Rmd
copy in lit review methods text from word document
@NidhiVinod , let's treat this .Rmd as the new master for this text.

@@
 
+## identification of relevant studies
+
+Relevant studies were searched in the databases: ISI Web of Science, Smithsonian online library and Google Scholar, using the following key terms: (leaf traits OR foliar traits) AND (inter-canopy OR intra-canopy OR canopy height) AND (e.g.
+chlorophyll OR e.g.
+LMA OR stomatal conductance); (leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical); (leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
+
+in Google Scholar, 600 articles were screened, using the key terms—(leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
+After screening the title and reading the abstract, 185 articles were most relevant.
+Herbaceous plant studies and seedling studies were excluded.
+These articles were added into Zotero folders: Leaf traits;  Leaf Metabolism and Processes, accordingly, for further careful reading.
+
+Similar search was conducted with the Smithsonian online library with the key terms— (leaf traits OR foliar traits) AND (inter-canopy OR intra-canopy OR canopy height) AND (e.g.
+chlorophyll OR e.g.
+LMA OR stomatal conductance)— provided most relevant results among the set of keywords mentioned above, with the field refine function that included botany, ecology, biology, environmental sciences, and forestry.
+Out of 150 relevant articles screened, 26 were most relevant after screening the title and reading the abstract, out of which 22 were already acquired through Google scholar search, 4 new articles were added to the folder.
+
+For ISI Web of Science, search the key terms—(leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical)— yielded 410 relevant results, 37 were most relevant after screening the title and reading the abstract, out of which 24 were already acquired through the above process, 13 new articles were added to Zotero folders.
+
+Through the above process, 202 articles were acquired into Zotero folders.
+The articles were tagged after careful reading of each as ‘added to the table’, ‘irrelevant to the table, but relevant to the review’ (with a note on the reason), ‘irrelevant to the table and the review’ (with a note on the reason for exclusion).
+Articles that were not yet carefully read were tagged ‘yet to read’, which will be tagged as one of the above after reading.
+Articles shared by co-authors and references mentioned in other studies collectively were ~>32 studies.
+So far 40 relevant articles are added into the tables
+
+## criteria for inclusion
+
+**this is outlined [here](https://github.com/EcoClimLab/vertical-thermal-review/blob/master/Lit_Review/Methods%20for%20Systematic%20review%20for%20the%20tables.docx). Section needs to be written.**
+
 \newpage

### 2021-09-01 vertical-thermal-review.Rmd
added space for NEON table, fixing headers
issues #53 , #54

--- 2021-09-01 vertical-thermal-review.Rmd
@@
 
-appendix_NEON_methods_title<- "Methods for NEON vertical profiles"
-appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
-appendix_lit_review_methods_title<- "Methods for literature review"
+appendix_NEON_methods_title<- "Methods S1. Methods for NEON vertical profiles"
+appendix_tealeaves_methods_title <- "Methods S2. Methods for leaf energy balance modeling"
+appendix_lit_review_methods_title<- "Methods S3. Methods for literature review"
 
 #SI table numbers
-SItable_example = "S1"
+NEON_sites_legend = "S1"
+Tealeaves_parameters = "S2"
 
 #SI table captions
-SItable_example_legend <- "Table S1. ...."
+NEON_sites_legend = "Table S1. (Neon sites)"
+Tealeaves_parameters_legend <- "Table S2. (Tealeaves parameters)"
 
@@
 
-Appendix `r appendix_NEON_methods`. `r appendix_NEON_methods_title`
+`r appendix_NEON_methods_title`
 
-Appendix `r appendix_tealeaves_methods`. `r appendix_tealeaves_methods_title`
+`r appendix_tealeaves_methods_title`
 
-Appendix `r appendix_lit_review_methods`. `r appendix_lit_review_methods_title`
+`r appendix_lit_review_methods_title`
 
-SIfig_example_legend
+`r NEON_sites_legend`
+
+`r Tealeaves_parameters_legend`
+
+`r SIfig_NEON_all_legend`
 

### 2021-09-01 vertical-thermal-review_SI.Rmd
added space for NEON table, fixing headers
issues #53 , #54

@@
 
-appendix_NEON_methods_title<- "Methods for NEON vertical profiles"
-appendix_tealeaves_methods_title <- "Methods for leaf energy balance modeling"
-appendix_lit_review_methods_title<- "Methods for literature review"
+appendix_NEON_methods_title<- "Methods S1. Methods for NEON vertical profiles"
+appendix_tealeaves_methods_title <- "Methods S2. Methods for leaf energy balance modeling"
+appendix_lit_review_methods_title<- "Methods S3. Methods for literature review"
 
 #SI table numbers
-Tealeaves_parameters = "S1"
+NEON_sites_legend = "S1"
+Tealeaves_parameters = "S2"
 
 #SI table captions
-Tealeaves_parameters_legend <- "Table S1. Tealeaves parameters"
+NEON_sites_legend = "Table S1. (Neon sites)"
+Tealeaves_parameters_legend <- "Table S2. (Tealeaves parameters)"
 
@@
 
-## identification of relevant studies
+## Identification of relevant studies
 
@@
 
-## criteria for inclusion
+## Criteria for inclusion
 
 **this is outlined [here](https://github.com/EcoClimLab/vertical-thermal-review/blob/master/Lit_Review/Methods%20for%20Systematic%20review%20for%20the%20tables.docx). Section needs to be written.**
+
+\newpage
+ # `r NEON_sites_legend`
 

### 2021-09-01 vertical-thermal-review_SI.Rmd
first pass at table, knit into document
Co-Authored-By: Marielle N Smith <45404993+m-n-smith@users.noreply.github.com>

@@
 
-#  Appendix `r appendix_NEON_methods`. `r appendix_NEON_methods_title`
+#  `r appendix_NEON_methods_title`
 
@@
 
-#  Appendix `r appendix_tealeaves_methods`. `r appendix_tealeaves_methods_title`
+#  `r appendix_tealeaves_methods_title`
 
@@
 
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 9)
+NEON_table <- read.csv("NEON_height_profiles/NEON_sites_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+NEON_table <- NEON_table[c(1:6)] #select and order columns for inclusion
+ft <- flextable(NEON_table)
+ft <- autofit(ft)
+ft <- width(ft, 1, width = .7)
+ft <- width(ft, 2, width = .75)
+ft <- width(ft, 3, width = 1.5)
+ft <- width(ft, 4, width = .5)
+ft <- width(ft, 5, width = 1)
+ft <- width(ft, 6, width = 1.5)
+ft
+```
+
 \newpage
@@
 
-Make a table from this:
-https://github.com/EcoClimLab/vertical-thermal-review/blob/master/NEON_height_profiles/HARV_neon_stats.csv
+See [here](https://github.com/EcoClimLab/vertical-thermal-review/commit/3f624a119d72d388c707d21fb20666a2ad31848f#commitcomment-55784918) for rows to include.
 

### 2021-09-01 vertical-thermal-review.Rmd
a bit of cleanup

@@
 
-appendix_NEON_methods_title<- "Methods S1. Methods for NEON vertical profiles"
+appendix_NEON_methods_title<- "Methods S1. Methods for analyzing vertical gradients in the biophysical environment"
 appendix_tealeaves_methods_title <- "Methods S2. Methods for leaf energy balance modeling"
@@
 #SI table captions
-NEON_sites_legend = "Table S1. (Neon sites)"
+NEON_sites_legend = "Table S1. National Ecological Observatory Network (NEON) sites included in the analysis of vertical gradients of key biophysical characteristics"
 Tealeaves_parameters_legend <- "Table S2. (Tealeaves parameters)"

### 2021-09-01 vertical-thermal-review_SI.Rmd
a bit of cleanup

@@
 ---
-title: ''
+title: 'Supporting Information'
 output: word_document
@@
 
-#  `r appendix_NEON_methods_title`
-
 S1 Note:
@@
 
-METHODS ON NEON ANALYSIS GO HERE.
+#  `r appendix_NEON_methods_title`
+
+METHODS ON NEON ANALYSIS GO HERE. Google doc: https://docs.google.com/document/d/1HdPa7X5AMrEwfLjazGMU92uTYMc6pObh/edit#heading=h.gjdgxs
 

### 2021-09-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 csl: new-phytologist.csl 
-
+nocite: |
+  @mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017;      @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @roberts_stomatal_1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003
 ---
@@
 
-```{r eval=FALSE, echo=FALSE}
-# refs in this table need to go in the nocite field below (easiest in alphabetical order)
-```
----
-nocite: |
-@mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @cobleHowVerticalPatterns2016a; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995;@matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014b; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988
----
-
 #### (Anatomical, Morphological and Physiological traits)
@@
 However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
-
-![Table `r table_leaf_metabolism`. (*For latest version of table, see tables.pdf.*)](table2.png)
-
-```{r eval=FALSE, echo=FALSE}
-# refs in this table need to go in the nocite field below (easiest in alphabetical order)
-```
----
-nocite: | @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012a; @roberts_stomatal_1990; @dangProfilesPhotosyntheticallyActive1997; @marencoLeafTraitPlasticity2017; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @hernandezSimilarTemperatureDependence2020; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @mau_temperate_2018; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @cobleHowVerticalPatterns2016; @wykaResponsesLeafStructure2012; @rijkersEffectTreeHeight2000a; @weerasingheCanopyPositionAffects2014; @scartazzaInvestigatingEuropeanBeech2016; @millerOnlySunlitLeaves2021; @harrisChangesLeafProperties2013; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @fauset_differences_2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @turnbullScalingFoliarRespiration2003; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @kenzoHeightrelatedChangesLeaf2015; @harleyEffectsLightTemperature1996;@xuSeasonalVariationTemperature2006; @athertonSpatialVariationLeaf2017; @sackHydrologyLeavesCoordination2003
----
 

### 2021-09-01 vertical-thermal-review.Rmd
added citations for SI materials

@@
 Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Fig.
-S1.](NEON_height_profiles/figures/profile_all.png)
+S1.
+Further site information is given in Supporting Information Table `r NEON_sites_legend`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all.png)
 
@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Appendix `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
 [ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/fig3_leafenergy.png)
@@
 
-![Table `r table_leaf_traits`. (*For latest version of table, see tables.pdf.*)](table1.png)
+![Table `r table_leaf_traits`.
+Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
+(*For latest version of table, see tables.pdf.*)](table1.png)
+
+```{r eval=FALSE, echo=FALSE}
+# refs in this table need to go in the nocite field at top of document 
+```
 
@@
 However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+
+![Table `r table_leaf_metabolism`.
+Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
+(*For latest version of table, see tables.pdf.*)](table2.png)
+
+```{r eval=FALSE, echo=FALSE}
+# refs in this table need to go in the nocite field at top of document 
+```
 

### 2021-09-01 vertical-thermal-review.Rmd
properly cite all SI materials
@NidhiVinod , this is taken care of.

@@
 #SI table numbers
-NEON_sites_legend = "S1"
+NEON_sites = "S1"
 Tealeaves_parameters = "S2"
@@
 In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S.
-National Ecological Observatory Network [NEON; Figs.
-`r fig_NEON_vertical`-`r SIfig_NEON_all`; Supporting Information Methods `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
+National Ecological Observatory Network [NEON; Fig.
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`; Supporting Information Methods `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
 While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].
@@
 S1.
-Further site information is given in Supporting Information Table `r NEON_sites_legend`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all.png)
+Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all.png)
 
@@
 `r fig_NEON_vertical`).**
-Foliage distribution within a forest is determined by the distribution of tree sizes and tree architecture, along the vertical gradient by leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns that are heterogeneous across forests (Figs.
-`r fig_NEON_vertical`a-b, `r SIfig_NEON_all`).
+Foliage distribution within a forest is determined by the distribution of tree sizes and tree architecture, along the vertical gradient by leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns that are heterogeneous across forests (Fig.
+`r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
 Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig.
@@
 
-**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Figs.
-`r fig_NEON_vertical`a-d, `r SIfig_NEON_all`, @koike_leaf_2001].**
+**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Fig.
+`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].**
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Figs.
-`r fig_NEON_vertical`d, `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
@@
 **Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
-Wind speeds increase with height across forest types [Figs.
-`r fig_NEON_vertical`e, `r SIfig_NEON_all`; @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
-Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Figs.
-`r fig_NEON_vertical`, `r SIfig_NEON_all`).
+Wind speeds increase with height across forest types [Fig.
+`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`; @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
+Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Fig.
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
 The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").
@@
 
-Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Figs.
-`r fig_NEON_vertical`f, `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
+Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig.
+`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
@@
 
-**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs.
-`r fig_NEON_vertical`, `r SIfig_NEON_all`).**
+**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Fig.
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).**
 The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021].
 Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Figs.
-`r fig_NEON_vertical`g, `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
+Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Fig.
+`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
 
@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra*) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
-[ISSUE #6](https://github.com/EcoClimLab/vertical-thermal-review/issues/6).](leaf_energy_balance/fig3_leafenergy.png)
+Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+Model parameters for biophysical conditions are given in Supplementary Information Table `r Tealeaves_parameters`.](leaf_energy_balance/fig3_leafenergy.png)
 
@@
 In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs.
-`r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig.
+`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
@@
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
-This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Figs.
-`r fig_NEON_vertical`, `r SIfig_NEON_all`).
+This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Fig.
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
 

### 2021-09-01 vertical-thermal-review_SI.Rmd
Update vertical-thermal-review_SI.Rmd
properly cite all SI materials
@NidhiVinod , this is taken care of.

@@
 Sites are grouped into the following categories: (A) (sub)subtropical and warm temperate broadleaf: ..., (B) temperate open/ savanna forests: ..., (C) temperate mesic broadleaf forests: ..., (D) temperate conifer forests: ..., (E) northern and boreal forests: ...
+Further site information is given in Supporting Information Table `r NEON_sites_legend`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.
 [Issue #35](https://github.com/EcoClimLab/vertical-thermal-review/issues/35).

### 2021-09-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 nocite: |
-  @mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @roberts_stomatal_1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003
+  @mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003
 ---

### 2021-09-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-09-01 vertical-thermal-review.Rmd
@@
 nocite: |
-  @mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @roberts_stomatal_1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003
+  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003
 ---
@@
 In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies.
-In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig.
@@
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
+The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
 
@@
 
-**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al.
-2015 ; Slot et al.
-2019; Chen et al.
-2020).**
+**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table 2; @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].**
 This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Table 1, Table 2).
@@
 @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].
+However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
+Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
@@
 Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020].
-Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2020].
+Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
 Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
 
-Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
+Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021].
 Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
 In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
-Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021].
-$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotPhotosyntheticPlasticityTropical2021].
+Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021].
+$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2021].
 Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020].
-However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a].
+However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
 The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
@@
 Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
+Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
 A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.

### 2021-09-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies.
-In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig.
@@
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
-Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
+Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021a; @niinemetsResponsesForestTrees2010a].
 
@@
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
+The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
 
@@
 Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
-In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
+In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a].
 
@@
 **Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
-shade leaves [@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
+shade leaves [@carterExperimentalWarmingTropical2021a; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
 This may vary based on forest type [**REF**], acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015], or leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
-Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021].
+Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021a].
 

### 2021-09-02 vertical-thermal-review.Rmd
add word count table (required by NewPhyt)

--- 2021-09-02 vertical-thermal-review.Rmd
@@
 *corresponding author: teixeirak@si.edu; +1 540 635 6546
+
+Text | word count | other | n
+---|---|-----|--------------------
+Total word count (excluding summary, references and legends) | 	## |	No. of figures | 	4 (all colour)
+Summary|  	183 | 	No. of Tables | 	2  
+Introduction|	## |	No of Supporting Information files | 	##
+Review of vertical gradients |	##	||	 
+Implications |	##	 ||
+Conclusions |	##	||
+Acknowledgments | ##	||
 

### 2021-09-02 vertical-thermal-review.Rmd
word count

--- 2021-09-02 vertical-thermal-review.Rmd
@@
 *corresponding author: teixeirak@si.edu; +1 540 635 6546
+
+Text | word count | other | n
+---|---|-----|--------------------
+Total word count (excluding summary, references and legends) | 	8565 |	No. of figures | 	5 (all colour)
+Summary|  	183 | 	No. of Tables | 	2  
+Introduction|	497 |	No of Supporting Information files | 	6
+Review of vertical gradients |	5976	||	 
+Implications |	1753	 ||
+Conclusions |	156	||
+Acknowledgments | ##	||
 
@@
 
+## Acknowledgements
+Thanks to ...., the ForestGEO Ecosystems & Climate lab at SCBI for helpful discussion.
+Funding was provided by the Smithsonian Institution, ....
+
 ## SI files

### 2021-09-02 vertical-thermal-review.Rmd
knit in fig 5, implement auto numbering for all fig references

@@
 fig_treerings= 4 #not sure if we want to include one
-fig_model=5
+fig_ecosystem=5
+fig_model=6
 
@@
 ---|---|-----|--------------------
-Total word count (excluding summary, references and legends) | 	8565 |	No. of figures | 	5 (all colour)
+Total word count (excluding summary, references and legends) | 	8565 |	No. of figures | 	6 (all colour)
 Summary|  	183 | 	No. of Tables | 	2  
@@
 Sites include a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
-Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Fig.
-S1.
+Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Supporting Information Figure `r SIfig_NEON_all`.
 Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all.png)
@@
 Potential mechanisms for these patterns include stifling of understory $\lambda E$ under low wind, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig.
-3) and lower photosynthesis.
+`r fig_leaf_T`) and lower photosynthesis.
 Furthermore, warming-induced increases in nighttime respiratory carbon loss may be particularly disadvantageous for understory trees that are operating closer to their light-compensation point than canopy trees.
@@
 Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig.
-5b, Piponiot et al., accepted).
+`r fig_ecosystem`, Piponiot et al., accepted).
 
-![**Figure 5.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
+![**Figure `r fig_ecosystem`.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
 Panel (a) is from (Kunert et al.
 2017).
-I'd like to create a parallel figure for C fluxes.](schematics/ecosystem vertical partitioning/raw from Nobby/Fig_5_ET.png)
+I'd like to create a parallel figure for C fluxes.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 
@@
 It remains far less clear how thermal sensitivity varies across forest strata under wet conditions, but both physiological mechanisms (Fig.
-3) and tree ring evidence (Fig.
-4) suggests that understory trees may respond more negatively to hot, wet conditions.
+`r fig_leaf_T`) and tree ring evidence (Fig.
+`r fig_treerings`) suggests that understory trees may respond more negatively to hot, wet conditions.
 Further research is required to better understand the thermal sensitivity of forest ecosystem function across strata.
@@
 Specifically, the increasing availability of airborne and spaceborne LiDAR and thermal remote sensing data offer a promising opportunity for mapping vertical thermal gradients in combination with vegetation structure at locations with in situ data and across larger landscapes (Fig.
-6).
+`r fig_model`).
 Airborne and spaceborne lidar, as well as terrestrial laser scanning data yield detailed 3D reconstructions of whole tree and forest structure.

### 2021-09-02 vertical-thermal-review.Rmd
some clean-up

@@
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
+The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
 
@@
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
-Potential mechanisms for these patterns include stifling of understory $\lambda E$ under low wind, humid conditions, resulting in above-optimal $T_{leaf}$ (Fig.
-`r fig_leaf_T`) and lower photosynthesis.
-Furthermore, warming-induced increases in nighttime respiratory carbon loss may be particularly disadvantageous for understory trees that are operating closer to their light-compensation point than canopy trees.
 Additional research is required to understand the mechanisms underlying these intriguing differences, and to disentangle size-related tree growth responses to hot-wet versus hot-dry conditions.
@@
 While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
-Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration (Fig 5a; Kunert et al.
-2017), and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
+Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
 In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime ~CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
@@
 ![**Figure `r fig_ecosystem`.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
-Panel (a) is from (Kunert et al.
-2017).
-I'd like to create a parallel figure for C fluxes.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 42 tropical forests from Piponiot et al.
+(accepted).](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 

### 2021-09-02 vertical-thermal-review.Rmd
minor fixes to legends

@@
 Abbreviations are as follows: VPD: vapor pressure deficit;  Tleaf - Tair: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; temp sensitivity: temperature sensitivity; ET: evapotranspiration, GPP: gross primary productivity.** Here, we summarize conditions in dense canopies during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
-Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy [Issue #27](https://github.com/EcoClimLab/vertical-thermal-review/issues/27).
+Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy.
 ](schematics/Schematic figure 1.png)
@@
 In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime ~CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
-Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests (Fig.
-`r fig_ecosystem`, Piponiot et al., accepted).
+Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests [Fig.
+`r fig_ecosystem`b, @piponiot_sizerelated_inreview].
 
 ![**Figure `r fig_ecosystem`.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
-Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 42 tropical forests from Piponiot et al.
-(accepted).](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 42 tropical forests from @piponiot_sizerelated_inreview.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 
@@
 (c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization.
-Vertically resolved in situ data could include any variable listed in Table 1 or 2.
-[Issue #42](https://github.com/EcoClimLab/vertical-thermal-review/issues/42).
-](schematics/fig_remote_sensing.png)
+Vertically resolved in situ data could include any variable listed in Table 1 or 2.](schematics/fig_remote_sensing.png)
 

### 2021-09-02 vertical-thermal-review.Rmd
Table 1: fixed up formatting and knit into doc
@NidhiVinod , I tweaked the formatting in your word doc, printed as pdf, created .pngs, and knit those into the doc. I also copied in the references list. After all that nocite confusion, it turns out it will be easier to just put the references in the list in the .Rmd!

Co-Authored-By: Nidhi Vinod <62031681+NidhiVinod@users.noreply.github.com>

@@
 
-![Table `r table_leaf_traits`.
-Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
-(*For latest version of table, see tables.pdf.*)](table1.png)
+\newpage
+**Table `r table_leaf_traits`.** Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
 
+![](table1a.png)
+
+\newpage
+Table `r table_leaf_traits`, cont.
+
+![](table1b.png)
+
+*1.
+Mau et al.
+2018; 2.
+Coble and Cavaleri 2014; 3.
+Sack et al.
+2006; 4.
+Chin and Sillett 2019; 5.
+Wyka et al.
+2012; 6.
+Atherton et al.
+2017; 7.
+Kenzo et al.
+2015; 8.
+Kusi and Karasi 2020; 9.
+Dang et al.
+1997; 10.
+Gebauer et al.
+2015; 11.
+Marenco et al.
+2017; 12.Kafuti et al.
+2020; 13.
+Van Wittenberghe et al.
+2012; 14.
+Zhang et al.
+2019; 15.
+Weerasinghe et al.
+2014; 16.
+Oldham et al.
+2010; 17.
+Ichie et al.
+2016; 18.
+Gregoriou et al.
+2007; 19.
+Levizou et al.
+2005; 20.
+Liakoura 1997; 21.
+Fauset et al.
+2018; 22.
+Niinemets et al.
+1998, 23.
+Ishida et al.
+1998; 24.
+Millen and Clendon 1979; 25.
+Smith and Carter, 1988; 26.
+Hadley and Smith 1987; 27.
+Panditharathna et al.
+2008; 28.
+Baltzer and Thomas 2005; 29.
+Dietz et al.
+2007; 30.
+Coble et al.
+2016; 31.
+Scartazza et al.
+2016; 32.
+Duursma and Marshall, 2006; 33.
+Harley et al.
+1996;  34.
+Hernandez et al.
+2020; 35.
+Turnbull et al.
+2003; 36.
+Chen et al.
+2020; 37.
+van de Weg et al.
+2012; 38.
+M.A Cavaleri et al.
+2008;  39.
+Koniger et al.
+1995; 40.
+Mastubara et al.
+2009; 41.
+Harris and Medina 2013; 42.
+Hansen et al.
+2001; 43.
+Poorter et al.
+1995; 44.
+Coble et al.
+2017; 45.
+Niinemets et al.
+2004; 46.
+Poorter et al.
+2000
 ```{r eval=FALSE, echo=FALSE}
@@
 
-## Implications [ & future research directions?]
+## Implications
 
@@
 
-### warming
+### Warming
 
@@
 In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
-Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al.
-2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
+Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
 
-### canopy disturbance
+### Canopy disturbance
 
-Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019].
-This buffering is influenced by hydrology, as wetter sites and seasons allow for denser, and more buffered canopies  [@davisMicroclimaticBufferingForests2019].
+Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019], particularly in wetter environments with dense canopies [@davisMicroclimaticBufferingForests2019].
 Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015].

### 2021-09-03 vertical-thermal-review.Rmd
formatted and knit in table 2 (also tweaked table 1)
@NidhiVinod , FYI. I'm happy with this new system! (Again, it would be smoother to put the references in nocite directly in the table footnote section.)

Co-Authored-By: Nidhi Vinod <62031681+NidhiVinod@users.noreply.github.com>
content edits and cleanup of intro, Fig. 1 caption

@@
 Summary|  	183 | 	No. of Tables | 	2  
-Introduction|	497 |	No of Supporting Information files | 	6
+Introduction|	495 |	No of Supporting Information files | 	6
 Review of vertical gradients |	5976	||	 
@@
 ```{r eval=FALSE, echo=FALSE}
-#  I think the limit for New Phyt is 200 words. 
+#  200 word limit, no bullet points  
 ```
-
-(200 word limit, no bullet points )
-
-Clean version implementing the changes below (198 words):
 
@@
 
-**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
-Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed an average of $1.5^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
-Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020].
+Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.
+Global average temperatures have risen 1.2 $^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5 $^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+The net result of higher temperatures may be increased or decreased tree growth and forest carbon sequestration, with decreases being more commonly documented across the world's forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @Anderson-TeixeiraJointEffectsClimate2021].
 The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
 
-**A great measure of uncertainity arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
-Forests are vertically and horizontally stratified, containg trees of different sizes along various successional stages. 
-Overstory canopies play a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
+A great measure of uncertainty arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].
+Forests are vertically and horizontally stratified, with overstory canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
 This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
@@
 
-**Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests.** 
+Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests. 
 We focus on five key themes (Fig. `r fig_schematic`):
@@
 (5) whole-tree and ecosystem ecology.
-*We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.*
+We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.
 
 ![**Figure `r fig_schematic`.
-Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature, (3) leaf traits, (4) leaf metabolism, and (5) whole-tree and ecosystem ecology.
-Abbreviations are as follows: VPD: vapor pressure deficit;  Tleaf - Tair: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; temp sensitivity: temperature sensitivity; ET: evapotranspiration, GPP: gross primary productivity.** Here, we summarize conditions in dense canopies during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
+Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) whole-tree and ecosystem ecology.** Abbreviations are as follows: VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; T sensitivity: temperature sensitivity; ET: evapotranspiration.
+Here, we summarize conditions in dense canopies during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
 Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy.
@@
 ```{r eval=FALSE, echo=FALSE}
-# refs in this table need to go in the nocite field at top of document 
+# refs in this table currently in the nocite field at top of document 
 ```
@@
 
-![Table `r table_leaf_metabolism`.
+\newpage
+**Table `r table_leaf_metabolism`.** Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves.
 Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
-(*For latest version of table, see tables.pdf.*)](table2.png)
+
+![](table2a.png)
+
+\newpage
+
+Table `r table_leaf_metabolism`, cont.
+
+![](table2b.png)
+
+\newpage
+
+Table `r table_leaf_metabolism`, cont.
+
+![](table2c.png)
+
+*1.
+Kafuti et al.
+2020; 2.
+Van Wittenberghe et al.
+2012; 3.
+Roberts et al.
+1990; 4.
+Dang et al.
+1997; 5.
+Marenco et al.
+2017; 6.
+Ambrose et al.
+2015; 7.
+Zweifel et al.
+2001; 8.
+Slot et al.
+2019; 9.
+Hernandez et al.
+2020; 10.
+Urban et al.
+2007; 11.
+Carter and Cavaleri 2018; 12.
+Martin et al.
+1999; 13.
+Mau et al.
+2018;  14.
+Kosugi et al.
+2012; 15.
+Niinemets et al.
+2015; 16.
+Bachofen et al.
+2020; 17.
+Hamerlynck and Knapp 1994; 18.
+Coble et al.
+2017; 19.
+Wyka et al.
+2012; 20.
+Rijkerse et al.
+2000; 21.
+Ishida et al.
+1999; 22.
+Weerasinghe et al.
+2014; 23.
+Scartazza et al.
+2016; 24.
+Miller et al.
+2021; 25.
+Harris and Medina 2013; 26.
+Legner et al.
+2014; 27.
+Kitao et al.
+2012; 28.
+Fauset et al.
+2018; 29.
+Rey-Sanchez et al.
+2016; 30.
+Muller et al.
+2021; 31.
+Curtis et al.
+2019; 32.
+Mier et al.
+2001; 33.
+Turnbull et al.
+2003; 34.
+Araki et al.
+2017; 35.
+Bolstad et al.
+1999; 36.
+Kenzo et al.
+2015; 37.
+Harley et al.
+1996; 38.
+Xu and Griffin 2006; 39.
+Harley et al.
+1997; 40.
+Niinemets and Sun, 2014; 41.
+Sharkey and Monson, 2014; 42.
+Saimpraga et al.
+2013; 43.
+Atherton et al.
+2017; 44.
+Carter et al.
+2021
 
 ```{r eval=FALSE, echo=FALSE}
-# refs in this table need to go in the nocite field at top of document 
+# refs in this table currently in the nocite field at top of document 
 ```

### 2021-09-03 vertical-thermal-review_SI.Rmd
knit in latest NEON table, some cleanup

--- 2021-09-01 vertical-thermal-review_SI.Rmd
@@
 NEON_table <- read.csv("NEON_height_profiles/NEON_sites_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-NEON_table <- NEON_table[c(1:6)] #select and order columns for inclusion
+NEON_table <- NEON_table[c(2:6)] #select and order columns for inclusion
 ft <- flextable(NEON_table)
 ft <- autofit(ft)
-ft <- width(ft, 1, width = .7)
-ft <- width(ft, 2, width = .75)
-ft <- width(ft, 3, width = 1.5)
-ft <- width(ft, 4, width = .5)
-ft <- width(ft, 5, width = 1)
-ft <- width(ft, 6, width = 1.5)
+ft <- width(ft, 1, width = .75)
+ft <- width(ft, 2, width = 2)
+ft <- width(ft, 3, width = .5)
+ft <- width(ft, 4, width = 1.2)
+ft <- width(ft, 5, width = 2)
 ft
@@
 Shown are  height profiles in July mean ± 1 standard deviation for maximum photosyntehtically active ratiation (PAR), maximum wind speed, minimum humidity, maximum $T_{air}$, and maximum biological temperature, $T_{bio}$.
-Sites are grouped into the following categories: (A) (sub)subtropical and warm temperate broadleaf: ..., (B) temperate open/ savanna forests: ..., (C) temperate mesic broadleaf forests: ..., (D) temperate conifer forests: ..., (E) northern and boreal forests: ...
-[Issue #35](https://github.com/EcoClimLab/vertical-thermal-review/issues/35).
-](NEON_height_profiles/figures/profile_all_groups.png)
+Site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all_groups.png)
 
 \newpage
-
 ## References

### 2021-09-03 vertical-thermal-review.Rmd
editing "the biophysical environment"

@@
 
-**The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs.
-`r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.**
+The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs.
+`r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.
 In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S.
-National Ecological Observatory Network [NEON; Fig.
-`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`; Supporting Information Methods `r appendix_NEON_methods`, @schimelNEONHierarchicallyDesigned2007].
+National Ecological Observatory Network (NEON; Fig.
+`r fig_NEON_vertical`, Supporting Information Methods `r appendix_NEON_methods`, Supporting Information Figure `r SIfig_NEON_all`).
 While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].
@@
 ![**Figure `r fig_NEON_vertical`.
-Vertical gradients in the biophysical environment for six US forest sites with contrasting forest structures in the National Ecological Observatory Network (NEON)**.
-Height profiles from the growing season are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy, and July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Sites include a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
-Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Supporting Information Figure `r SIfig_NEON_all`.
-Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all.png)
+Vertical gradients in the biophysical environment for six US forest sites in the National Ecological Observatory Network (NEON)**.
+Height profiles are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy during the growing season, and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+Sites, which represent a variety of forest structures, include a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
+Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.
+Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Supporting Information Figure `r SIfig_NEON_all`.](NEON_height_profiles/figures/profile_all.png)
 
-**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions.
+Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions.
 It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig.
-`r fig_NEON_vertical`).**
-Foliage distribution within a forest is determined by the distribution of tree sizes and tree architecture, along the vertical gradient by leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns that are heterogeneous across forests (Fig.
+`r fig_NEON_vertical`).
+Leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig.
 `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
-Dense canopy forests composed of broadleaf trees can have high LAD in the upper canopy layers, as well as in the understory (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a) -- a pattern also common in tropical forests [@ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993].
-In temperate forests, although LAD is generally more concentrated in the upper-canopy, undualting patterns are formed by conifers and top- vs.
-bottom-heavy foliage trees, and sometimes by dense understories [e.g., SCBI, Fig.
-`r fig_NEON_vertical`a, @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002].
+Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest LAD in the upper canopy layers, but understory LAD is often relatively high in the understory as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
+`r fig_NEON_vertical`a, @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] *[@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]*.
 In forests with more open upper canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig.

### 2021-09-03 vertical-thermal-review_SI.Rmd
editing "the biophysical environment"
update table S2

@@
 #  `r appendix_NEON_methods_title`
+
+*be sure to cite @schimelNEONHierarchicallyDesigned2007.*
 
@@
 
-See [here](https://github.com/EcoClimLab/vertical-thermal-review/commit/3f624a119d72d388c707d21fb20666a2ad31848f#commitcomment-55784918) for rows to include.
-
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 set_flextable_defaults(font.family = "Arial", font.size = 9)
-parameters_table <- read.csv("NEON_height_profiles/HARV_neon_stats.csv", stringsAsFactors = FALSE, check.names = FALSE)
-parameters_table <- parameters_table[c(9, 1, 3, 5)] #select and order columns for inclusion
+parameters_table <- read.csv("NEON_height_profiles/HARV_neon_stats_handedit.csv", stringsAsFactors = FALSE, check.names = FALSE)
+parameters_table <- parameters_table[which(parameters_table$include==1),c(9, 1, 3)] #select and order columns for inclusion
 ft <- flextable(parameters_table)

### 2021-09-03 vertical-thermal-review_SI.Rmd
work on table S1

@@
 NEON_table <- read.csv("NEON_height_profiles/NEON_sites_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-NEON_table <- NEON_table[c(2:6)] #select and order columns for inclusion
+NEON_table <- NEON_table[c(2:4, 6:8)] #select and order columns for inclusion
 ft <- flextable(NEON_table)
@@
 ft <- width(ft, 1, width = .75)
-ft <- width(ft, 2, width = 2)
+ft <- width(ft, 2, width = 1.75)
 ft <- width(ft, 3, width = .5)
-ft <- width(ft, 4, width = 1.2)
-ft <- width(ft, 5, width = 2)
+ft <- width(ft, 4, width = .75)
+ft <- width(ft, 5, width = .8)
+ft <- width(ft, 6, width = 1.75)
 ft

### 2021-09-03 vertical-thermal-review_SI.Rmd
table S1: switched to using the nlcd_class_tower for forest type
issue #53

--- 2021-09-03 vertical-thermal-review_SI.Rmd
@@
 NEON_table <- read.csv("NEON_height_profiles/NEON_sites_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-NEON_table <- NEON_table[c(2:6)] #select and order columns for inclusion
+NEON_table <- NEON_table[c(2:4, 6:7,9)] #select and order columns for inclusion
 ft <- flextable(NEON_table)
@@
 ft <- width(ft, 1, width = .75)
-ft <- width(ft, 2, width = 2)
+ft <- width(ft, 2, width = 1.75)
 ft <- width(ft, 3, width = .5)
-ft <- width(ft, 4, width = 1.2)
-ft <- width(ft, 5, width = 2)
+ft <- width(ft, 4, width = .75)
+ft <- width(ft, 5, width = .8)
+ft <- width(ft, 6, width = 2)
 ft

### 2021-09-03 vertical-thermal-review_SI.Rmd
small fixes to table S1

--- 2021-09-03 vertical-thermal-review_SI.Rmd
@@
 ft <- width(ft, 1, width = .75)
-ft <- width(ft, 2, width = 2)
+ft <- width(ft, 2, width = 1.75)
 ft <- width(ft, 3, width = .5)
-ft <- width(ft, 4, width = 1.2)
-ft <- width(ft, 5, width = 2)
+ft <- width(ft, 4, width = .75)
+ft <- width(ft, 5, width = .8)
+ft <- width(ft, 6, width = 2)
 ft
 ```
+*Refers to forest type around the tower on which the vertical profile of micrometeorological instruments is mounted, as provided by NEON (Hongyan Luo, personal communication; [site descriptions on NEON website](https://www.neonscience.org/field-sites)).
 

### 2021-09-03 vertical-thermal-review.Rmd
change HARV site description
HARV is a mixed broadleaf-conifer forest, but the tower is in a broadleaf part.

@@
 Height profiles are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy during the growing season, and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Sites, which represent a variety of forest structures, include a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
+Sites, which represent a variety of forest structures, include a mixed northern hardwood forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
 Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.

### 2021-09-03 vertical-thermal-review_SI.Rmd
a bit oof cleanup, added TOC

@@
 ---
-title: 'Supporting Information'
-output: word_document
+title: 'Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications'
+output:
+  word_document:
+    toc: true
+    toc_depth: 2
 bibliography: Zotero_references.bib
@@
 
-**Title:** Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications
-
-**Authors:** 
-Nidhi Vinod^1^,
-Martijn Slot^2^,
-Ian McGregor^3^,
-Elsa M. Ordway^4^,
-Marielle N. Smith^5^,
-Tyeen Taylor,
-Lawren Sack,
-Kristina J. Anderson-Teixeira^1,2^*
-
 \newpage
-
-S1 Note:
-
-It is noteworthy that leaf and branch temperatures often differ substantially.
-For instance, exposed tropical tree bark can be much warmer than leaves [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021; @johnstonWhatLiesVertical2020].
-Thus, vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig.
-`r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories).
-Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021; @macleanMicroclimcMechanisticModel2021]
 

### 2021-09-03 vertical-thermal-review_SI.Rmd
reinstate note
issue #58

@@
 see [Issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) , [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)
+
+*Note on interpreting $T_{veg}$*:
+
+Vegetation temperature ($T_{veg}$) is measured using infrared sensors that integrate across both leaves and woody vegetation. 
+Leaf and branch temperatures often differ substantially; for instance, exposed tropical tree bark can be much warmer than leaves [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021; @johnstonWhatLiesVertical2020].
+Thus, vegetation temperature ($T_{veg}$) does not always equal or approximate $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (i.e., low-LAI ecosystems and understories).
 

### 2021-09-03 vertical-thermal-review.Rmd
remove table S2
closes #57

@@
 NEON_sites = "S1"
-Tealeaves_parameters = "S2"
 
@@
 NEON_sites_legend = "Table S1. National Ecological Observatory Network (NEON) sites included in the analysis of vertical gradients of key biophysical characteristics"
-Tealeaves_parameters_legend <- "Table S2. (Tealeaves parameters)"
 
@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
-Model parameters for biophysical conditions are given in Supplementary Information Table `r Tealeaves_parameters`.](leaf_energy_balance/fig3_leafenergy.png)
+Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.](leaf_energy_balance/fig3_leafenergy.png)
 
@@
 
-`r Tealeaves_parameters_legend`
-
 `r SIfig_NEON_all_legend`

### 2021-09-03 vertical-thermal-review_SI.Rmd
remove table S2
closes #57

@@
 Energy balance for a typical overstory sun leaf and understory shade leaf were modeled in the R package *Tleaves* [@muir_tealeaves_2019], parameterized for *Quercus rubra* L.
-leaves at Harvard Forest, MA, USA (**GEOGRAPHIC COORDINATES).
-Parameters are presented in Table `r Tealeaves_parameters`.
+leaves at Harvard Forest, MA, USA (42.5369, -72.17266).
 
@@
 \newpage
-# `r Tealeaves_parameters_legend `
-
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(flextable)
-set_flextable_defaults(font.family = "Arial", font.size = 9)
-parameters_table <- read.csv("NEON_height_profiles/HARV_neon_stats_handedit.csv", stringsAsFactors = FALSE, check.names = FALSE)
-parameters_table <- parameters_table[which(parameters_table$include==1),c(9, 1, 3)] #select and order columns for inclusion
-ft <- flextable(parameters_table)
-ft <- autofit(ft)
-ft
-```
-
-\newpage
 

### 2021-09-04 vertical-thermal-review.Rmd
table fix-up
@NidhiVinod , a few minor things:
- took units out of italics
- inserted spaces in units (e.g., mms-1 --> mm s-1)
- put in placeholder for footnote (in Rmd) that seems to have been dropped (on savanna trees)
updated tables
@NidhiVinod , you removed the VOC production section... was that intentional?

@@
 
+\newpage
+Table `r table_leaf_traits`, cont.
+
+![](table1c.png)
+
 *1.
@@
 Marenco et al.
-2017; 12.Kafuti et al.
+2017; 12.
+Kafuti et al.
 2020; 13.
@@
 Poorter et al.
-2000
+2000; 47.
+Taylor et al.
+2021; 48.
+Harley et al.
+1997; 49.
+Niinemets and Sun, 2014; 50.
+Sharkey and Monson, 2014; 51.
+Saimpraga et al.
+2013; 52.
+Zwieniecki et al.
+2004; 53.
+Sack and Scoffoni, 2013; 54.
+Ball et al., 1988
 ```{r eval=FALSE, echo=FALSE}
@@
 Xu and Griffin 2006; 39.
-Harley et al.
-1997; 40.
-Niinemets and Sun, 2014; 41.
-Sharkey and Monson, 2014; 42.
-Saimpraga et al.
-2013; 43.
 Atherton et al.
-2017; 44.
+2017; 40.
 Carter et al.
-2021
+2021; 41.
+Sack et al.
+2003
 
+** (**NIDHI, we previously had a footnote here. I believe it's the savanna site. I can't find that, though.**)
 ```{r eval=FALSE, echo=FALSE}

### 2021-09-04 vertical-thermal-review.Rmd
modest edits to biophysical gradients section
@NidhiVinod , this was mostly word-smithing, but I also added sentences at the end of a few paragraphs that I think will be helpful to guide the reader.

@@
 `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
-Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @bonan_ecological_2016; @hanberryOpenForestEcosystems2020a].
+Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020a].
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019].
@@
 
-**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Fig.
-`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].**
+Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Fig.
+`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
-`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
+The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "Leaf metabolism and thermal responses"), but also greater propensity towards elevated $T_{leaf}$ (see section "Leaf temperature").
 
-**Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].**
-Wind speeds increase with height across forest types [Fig.
-`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`; @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
+Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].
+This holds true across the range of forest types [Fig.
+`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
 Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Fig.
@@
 
-**Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
+Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
@@
 Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig.
-`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018].
+`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
@@
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
+The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section "Leaf temperature").
 
-**Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Fig.
-`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).**
+Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Fig.
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
 The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021].
@@
 Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Fig.
-`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
+`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @hadleyInfluenceKrummholzMat1987; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
+These usually-modest gradients in $T_{air}$ combine with the above-described gradients in other biophysical conditions to shape $T_{leaf}$.
 
@@
 
-Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
 Leaves are typically cooler than the air at night, and during the day especially under cloudy skies, at high windspeeds and high $T_{air}$.

### 2021-09-05 vertical-thermal-review.Rmd
updated tables
minor text updates

@@
 
-$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections) with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
-Leaves are typically cooler than the air at night, and during the day especially under cloudy skies, at high windspeeds and high $T_{air}$.
+$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf metabolism and thermal responses") with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high windspeeds, and high $T_{air}$). 
 Leaves are typically warmer than air when under full sunlight, especially under slow windspeeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
@@
 2000; 47.
-Taylor et al.
-2021; 48.
-Harley et al.
-1997; 49.
-Niinemets and Sun, 2014; 50.
-Sharkey and Monson, 2014; 51.
-Saimpraga et al.
-2013; 52.
 Zwieniecki et al.
-2004; 53.
-Sack and Scoffoni, 2013; 54.
+2004; 48.
+Sack and Scoffoni, 2013; 49.
 Ball et al., 1988
@@
 Sack et al.
-2003
+2003; 42.
+Taylor et al.
+2021; 43.
+Harley et al.
+1997; 44.
+Niinemets and Sun, 2014; 45.
+Sharkey and Monson, 2014; 46.
+Saimpraga et al.
+2013
 

### 2021-09-05 NEON_height_profiles/scripts/verticalProfileFigure.R
Changed xlabs and font size for #35

@@ -335,13 +335,16 @@ groupPlots <- lapply(1:length(sites), function(X){
         labs(x = "",
              y = "") +
         theme_bw() +
-        theme(text = element_text(size=16))
+        theme(text = element_text(size=20))
       
       if(X==5){#only labels on bottom row for putting all 25 plots together
-        if(i %in% c(1,2)){ #windspeed, RH
+        if(i==1){ #windspeed
+          graph <- graph + 
+            xlab(bquote("max Windspeed [" ~m~s^'-1'~"]"))
+        } else if(i==2){ #RH
           graph <- graph + 
             xlab(paste0(dp[,stat][i], " ", dp[,xlabs][i]))
-          } else if(i==3){ #tempSingle
+        } else if(i==3){ #tempSingle
           graph <- graph +
             xlab(expression(
               paste("max T"["air"], " [",degree,"C]")))
@@ -351,7 +354,7 @@ groupPlots <- lapply(1:length(sites), function(X){
               paste("max T"["bio"], " [",degree,"C]")))
         } else if(i==5){ #PAR
           graph <- graph + 
-            xlab("max PAR")
+            xlab(bquote("max PAR [" ~mu ~ mol^'-2' ~ s^'-1'~"]"))
         }
       }
       
@@ -369,7 +372,7 @@ groupPlots <- lapply(1:length(sites), function(X){
       if(i==4){ #biotemp - legend purposes
         graph <- graph + 
           theme(legend.position=c(0.8,0.7), 
-                legend.text=element_text(size=14)) +
+                legend.text=element_text(size=16)) +
           guides(size=FALSE)
       } else {
         graph <- graph + theme(legend.position="none")
@@ -391,7 +394,7 @@ groupPlots <- lapply(1:length(sites), function(X){
     plots[["tempSingleMean"]] + plots[["bioTempMean"]]
   p <- p + 
     plot_annotation(title=titleTx, 
-                    theme = theme(plot.title = element_text(size = 18))) +
+                    theme = theme(plot.title = element_text(size = 20))) +
     plot_layout(ncol = 5)
   
   # p <- ggarrange(NULL, plot_0, NULL, NULL, NULL,

### 2021-09-05 NEON_height_profiles/scripts/verticalProfileFigure.R
Increased tick sizes #35
@teixeirak does this work?

@@ -335,7 +335,8 @@ groupPlots <- lapply(1:length(sites), function(X){
         labs(x = "",
              y = "") +
         theme_bw() +
-        theme(text = element_text(size=20))
+        theme(text = element_text(size=20),
+              axis.ticks.length=unit(.25, "cm"))
       
       if(X==5){#only labels on bottom row for putting all 25 plots together
         if(i==1){ #windspeed
@@ -343,15 +344,15 @@ groupPlots <- lapply(1:length(sites), function(X){
             xlab(bquote("max Windspeed [" ~m~s^'-1'~"]"))
         } else if(i==2){ #RH
           graph <- graph + 
-            xlab(paste0(dp[,stat][i], " ", dp[,xlabs][i]))
+            xlab("min RH [ % ]")
         } else if(i==3){ #tempSingle
           graph <- graph +
             xlab(expression(
-              paste("max T"["air"], " [",degree,"C]")))
+              paste("max T"["air"], " [ ",degree,"C ]")))
         } else if(i==4){ #bioTempMean
           graph <- graph +
             xlab(expression(
-              paste("max T"["bio"], " [",degree,"C]")))
+              paste("max T"["bio"], " [ ",degree,"C ]")))
         } else if(i==5){ #PAR
           graph <- graph + 
             xlab(bquote("max PAR [" ~mu ~ mol^'-2' ~ s^'-1'~"]"))

### 2021-09-06 vertical-thermal-review.Rmd
T leaf section

@@
 
-**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig.
-`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].**
-High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig.
-`r fig_leaf_T`a).
-Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig.
-`r fig_leaf_T`d) and higher wind speeds (Fig.
-`r fig_leaf_T`b).
-Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD; notably, $g_b$ is influenced by windspeed and $g_s$ respond due to stomatal closure under high VPD.
-Therefore, under full sunlight, $T_{leaf}-T_{air}$ decreases with $g_b$, which increases with wind speed [Fig.
-`r fig_leaf_T`b, @daudetWindSpeedLeaf1999a] and $g_s$ (Fig.
-`r fig_leaf_T`e); increases with leaf size (Fig.
-`r fig_leaf_T`d); and increases with RH (Fig.
-`r fig_leaf_T`c).
+Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig.
+`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].
+High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a). 
+Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and higher wind speeds (Fig. `r fig_leaf_T`b). 
+Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD, where $g_b$ increases with wind speed and $g_s$ can decline at high VPD due to stomatal closure.
+Therefore, $T_{leaf}-T_{air}$ decreases with wind speed [Fig.
+`r fig_leaf_T`b, @daudetWindSpeedLeaf1999a], increases with RH (Fig.
+`r fig_leaf_T`c), increases with leaf size (Fig.
+`r fig_leaf_T`d), and decreases with $g_s$ (Fig.
+`r fig_leaf_T`e).
 

### 2021-09-06 vertical-thermal-review.Rmd
Tleaf section: FIg. 3 caption and some text

@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in $T_{leaf}-T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the tealeaves R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA), as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.](leaf_energy_balance/fig3_leafenergy.png)
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both normal and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+In each scenario, an the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
+swr = short wave radiation (W/m^2^); ws = windspeed (m/s); rh= relative humidity; lw = leaf width (cm); gs = stomatal conductance (mmol m^2^/s).
+](leaf_energy_balance/fig3_leafenergy.png)
 
-**Leaf energy balance is a tradeoff between water uptake and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
-When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
-However, when leaf water demand (influenced by VPD and stomatal opening) exceeds the rate of supply, conservation of water ($g_s$-limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
+Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
+With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) and larger $g_s$ [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+However, when leaf water demand (influenced by VPD and stomatal opening) exceeds the rate of supply, conservation of water ($g_s$ - limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$-limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
-With adequate water, species with smaller leaves (higher $g_b$) and larger $g_s$ will be most effective at cooling under high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$ - limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
 

### 2021-09-06 vertical-thermal-review.Rmd
added the note we had for Table 2
ie. **composite climatic stress variable from canopy temperature, vapour pressure deficit, and relative humidity
is higher in lower canopy

@@
 
-** (**NIDHI, we previously had a footnote here. I believe it's the savanna site. I can't find that, though.**)
+** composite climatic stress variable from canopy temperature, vapour pressure deficit, and relative humidity is higher in lower canopy
 ```{r eval=FALSE, echo=FALSE}

### 2021-09-06 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
-The net result of higher temperatures may be increased or decreased tree growth and forest carbon sequestration, with decreases being more commonly documented across the world's forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @Anderson-TeixeiraJointEffectsClimate2021].
+The net result of higher temperatures may be increased or decreased tree growth and forest carbon sequestration, with decreases being more commonly documented across the world's forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021].
 The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

### 2021-09-07 vertical-thermal-review_SI.Rmd
updating SI

@@
 
-- micromet from NEON (explain exactly what you used)
-- leaf characteristic dimension, was measured on *Quercus rubra* sun and shade leaf, 
-- stomatal conductance measurements were referred from Tleaves typical sun and shade measurements [@muir_tealeaves_2019] 
-- @cavender-baresChangesDroughtResponse2000.
+For normal scenario, micrometeorological data from Harvard NEON was used to parameterize biophysical constants, ie.
+the maximum mean of PAR, wind speed, RH and $T_{air}$ (Fig.3) for growing season months at vertical positions 60m (overstory) and 10m (understory).
+Leaf size parameters were selected for *Quercus rubra* sun and shade leaf based on personal observations (by Vinod) and verified as reasonable by @zwienieckiHydraulicLimitationsImposed2004.
+Stomatal conductance measurements were referred from *Tleaves* typical sun and shade measurements [@muir_tealeaves_2019] and @cavender-baresChangesDroughtResponse2000.
 
-For drought scenario, biophysical variables were parameterized similar to normal scenario using Harvard NEON data.
-Overstory drought PAR values reflect maximum observed Harvard NEON PAR and understory drought reflect 50% increased PAR of understory normal value.
-Leaf trait measurements for leaf width is the same as normal scenario, stomatal conductance is kept constant at a minimum value of 0.01 umol/m^2/s/Pa for both positions.
-In each visual, all variables are constant (Biophysical Constants table) except for the independent variable that represents minimum - maximum range.
+For the drought scenario, biophysical constants were used from Harvard NEON data as well.
+For overstory drought PAR values, we used maximum Harvard NEON PAR value at 60m, and for understory drought PAR we used 50% increased PAR of understory normal value.
+Leaf trait measurements for leaf width is the same as normal scenario, stomatal conductance is kept constant at a minimum value of 0.01 $\mu mol/m^2/s/Pa$ for both overstory and understory.
+In each visual, all biophysical variables are constant except for the independent variable that represents a minimum - maximum range.
 
@@
 
-in Google Scholar, 600 articles were screened, using the key terms—(leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
+in Google Scholar, 600 articles were screened, using the key terms (leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
 After screening the title and reading the abstract, 185 articles were most relevant.
@@
 
-Similar search was conducted with the Smithsonian online library with the key terms— (leaf traits OR foliar traits) AND (inter-canopy OR intra-canopy OR canopy height) AND (e.g.
+Similar search was conducted with the Smithsonian online library with the key terms (leaf traits OR foliar traits) AND (inter-canopy OR intra-canopy OR canopy height) AND (e.g.
 chlorophyll OR e.g.
-LMA OR stomatal conductance)— provided most relevant results among the set of keywords mentioned above, with the field refine function that included botany, ecology, biology, environmental sciences, and forestry.
+LMA OR stomatal conductance) provided most relevant results among the set of keywords mentioned above, with the field refine function that included botany, ecology, biology, environmental sciences, and forestry.
 Out of 150 relevant articles screened, 26 were most relevant after screening the title and reading the abstract, out of which 22 were already acquired through Google scholar search, 4 new articles were added to the folder.
 
-For ISI Web of Science, search the key terms—(leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical)— yielded 410 relevant results, 37 were most relevant after screening the title and reading the abstract, out of which 24 were already acquired through the above process, 13 new articles were added to Zotero folders.
+For ISI Web of Science, search the key terms (leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical) yielded 410 relevant results, 37 were most relevant after screening the title and reading the abstract, out of which 24 were already acquired through the above process, 13 new articles were added to Zotero folders.
 
 Through the above process, 202 articles were acquired into Zotero folders.
-The articles were tagged after careful reading of each as ‘added to the table’, ‘irrelevant to the table, but relevant to the review’ (with a note on the reason), ‘irrelevant to the table and the review’ (with a note on the reason for exclusion).
-Articles that were not yet carefully read were tagged ‘yet to read’, which will be tagged as one of the above after reading.
+The articles were tagged after careful reading of each as added to the table, irrelevant to the table, but relevant to the reviewed (with a note on the reason), irrelevant to the table and the review (with a note on the reason for exclusion).
+Articles that were not yet carefully read were tagged yet to read, which will be tagged as one of the above after reading.
 Articles shared by co-authors and references mentioned in other studies collectively were ~>32 studies.
-So far 40 relevant articles are added into the tables
+So far 70 relevant articles are added into the tables
 
@@
 **this is outlined [here](https://github.com/EcoClimLab/vertical-thermal-review/blob/master/Lit_Review/Methods%20for%20Systematic%20review%20for%20the%20tables.docx). Section needs to be written.**
+
+The studies included in the review were based on a global geographic scope.
+Ecosystem types included any studies with trees for mostly forests, but savannas were also considered.
+Information for leaf traits and metabolism included studies with independent variables: sun/shade vs height (where height is generally also sun/shade), and dependent variables of interest were based on variables known to influence leaf temperature ie.
+anatomical, biochemical, structural foliar traits, including gas exchange, metabolism and thermal sensitivity variables.
 

### 2021-09-07 vertical-thermal-review.Rmd
add table footnotes
@NidhiVinod , I added footnotes in the .Rmd and in the column headers. However, I think this might create a conflicted version with what you have, as at least one of the changes you said you made (DCA in metabolism table) is not captured in this table. So I gave it a new name. Please reconcile and then delete the extra files.

issue #60

@@
 
-*1.
+*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, $\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
+
+$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).
+
+$\ddagger$ 1.
 Mau et al.
@@
 
-*1.
+*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, $\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
+
+$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).
+
+$\ddagger$ 1.
 Kafuti et al.

### 2021-09-13 vertical-thermal-review.Rmd
knit in latest tables/figures, remove an abbreviation

@@
 `r fig_NEON_vertical`).
-Leaf area density (LAD; *i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig.
+Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig.
 `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
-Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest LAD in the upper canopy layers, but understory LAD is often relatively high in the understory as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
+Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high in the understory as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`a, @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] *[@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]*.
-In forests with more open upper canopies, including many needle-leaf forests, LAD can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig.
+In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig.
 `r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020a].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting LAD patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019].
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019].
 In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.
 
-Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by LAD, canopy height, canopy structure, and across species and forest types [Fig.
+Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig.
 `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].

### 2021-09-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Vertical gradients in the biophysical environment for six US forest sites in the National Ecological Observatory Network (NEON)**.
-Height profiles are shown for: (a) leaf area density, (b) proportion of sun leaves, (c) proportion light incident at the top of the canopy during the growing season, and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Sites, which represent a variety of forest structures, include a mixed northern hardwood forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
+Height profiles are shown for: growing season (a) leaf area density, (b) proportion of sun leaves, and (c) proportion light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
+Measurements extend from ground level (height = 0 m) to the top of the canopy (a-c, h) oor above (d-g) Sites, which represent a variety of forest structures, include a mixed northern hardwood forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
 Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.
@@
 
-### Biophysical drivers
+### Biophysical drivers of $T_{leaf}$
 
@@
 `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, but during $g_s$ - limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
+Therefore, at high solar radiation loads, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at maximum $g_s$, but during $g_s$ - limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
 
-**Vertical gradients**
+## Vertical gradients
 
-**These basic biophysical principles shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
+**The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
 High radiation at the top of the vertical profile (Fig.
@@
 `r fig_leaf_T`).
-However, higher wind speeds (e.g., ~ 4 $ms^{-1}$; Fig.
+However, higher wind speeds (e.g., ~ 4 m s^-1^; Fig.
 `r fig_NEON_vertical`) can reduce $T_{leaf}-T_{air}$ to ~ $4^\circ$C [Fig.
@@
 `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].
-
-### In situ observations
 

### 2021-09-13 vertical-thermal-review.Rmd
adopted organization in google doc
#48

@@
 
-**Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).**
-Closed canopies with high LAI act as an parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-In contrast to closed canopies, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory.
-This mixing and light penetration can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy.
-The latter can result from the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].
-
-**Within closed-canopy forests vertical gradients, the propensity for $T_{leaf}$ to be elevated above $T_{air}$ varies with micrometeorological conditions (Fig. `r fig_leaf_T`).**
 Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
@@
 `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat or drought stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
  In addition to lower wind speeds (Fig.
@@
 `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].
+
+**Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).**
+Closed canopies with high LAI act as an parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
+In contrast to closed canopies, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
+This mixing and light penetration can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy. 
+The latter can result from the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].
 

### 2021-09-13 vertical-thermal-review.Rmd
Tleaf section

@@
 
-**The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.**
+The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.
 High radiation at the top of the vertical profile (Fig.
@@
 `r fig_leaf_T`).
-However, higher wind speeds (e.g., ~ 4 m s^-1^; Fig.
-`r fig_NEON_vertical`) can reduce $T_{leaf}-T_{air}$ to ~ $4^\circ$C [Fig.
-`r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
-Collectively, higher wind speed (Fig.
-`r fig_NEON_vertical`), and with adaptive leaf traits increase $g_b$ and $g_s$ (reviewed below) mediate the direct effects of solar radiation on upper canopy leaves (Fig.
-`r fig_leaf_T`a) and result in higher rates of heat loss (Fig.
+However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
+In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf metabolism and thermal responses" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
@@
 
-Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
-Because sun leaves are exposed to higher irradiance and VPD, their $g_s$-limitation is steeper than shade leaves, resulting in greater increases in $T_{leaf}$ when $g_s$ is reduced [Fig.
+The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions.
+Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
+This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$-limitation than shade leaves [Fig.
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling.
-Under lower wind conditions, however,  lower $g_b$ increases $T_{leaf} - T_{air}$ [Fig.
-`r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities can result in greater $T_{leaf}$ under slight heat or drought stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
+In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
+`r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In addition to lower wind speeds (Fig.

### 2021-09-13 vertical-thermal-review.Rmd
edit Fig. 3 caption, as per issue #64
#64

@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent typical conditions for a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both normal and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
 In each scenario, an the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.

### 2021-09-13 vertical-thermal-review_SI.Rmd
changing tealeaves description as per issue #64
(also general cleanup)

@@
 
-For normal scenario, micrometeorological data from Harvard NEON was used to parameterize biophysical constants, ie.
-the maximum mean of PAR, wind speed, RH and $T_{air}$ (Fig.3) for growing season months at vertical positions 60m (overstory) and 10m (understory).
+Micrometeorological data from Harvard NEON was used to parameterize biophysical constants, using the mean oof maximum PAR, wind speed, and $T_{air}$ (Fig.3) for growing season months at vertical positions 60m (overstory) and 10m (understory).
 Leaf size parameters were selected for *Quercus rubra* sun and shade leaf based on personal observations (by Vinod) and verified as reasonable by @zwienieckiHydraulicLimitationsImposed2004.
+
+For the humid scenario, we used mean maximum RH.
 Stomatal conductance measurements were referred from *Tleaves* typical sun and shade measurements [@muir_tealeaves_2019] and @cavender-baresChangesDroughtResponse2000.
 
-For the drought scenario, biophysical constants were used from Harvard NEON data as well.
-For overstory drought PAR values, we used maximum Harvard NEON PAR value at 60m, and for understory drought PAR we used 50% increased PAR of understory normal value.
-Leaf trait measurements for leaf width is the same as normal scenario, stomatal conductance is kept constant at a minimum value of 0.01 $\mu mol/m^2/s/Pa$ for both overstory and understory.
+For the drought scenario, we reduced RH to 50% of the mean maximum.
+*For overstory drought PAR values, we used maximum Harvard NEON PAR value at 60m, and for understory drought PAR we used 50% increased PAR of understory normal value.* 
+We reduced stomatal conductance to a minimum value of 0.01 $\mu mol/m^2/s/Pa$ for both overstory and understory.
+
 In each visual, all biophysical variables are constant except for the independent variable that represents a minimum - maximum range.

### 2021-09-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
finished my review of Tleaf section
... but still need to address #65

@@
 
-## Vertical gradients
+### Vertical gradients
 
@@
 In addition to lower wind speeds (Fig.
-`r fig_NEON_vertical`), higher RH in the understory (Fig.
+`r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig.
 `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig.
-`r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021].
+`r fig_leaf_T`c, @tibbittsHumidityPlants1979b; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021; @dietzVerticalPatternsDuration2007].
+
+**Nidhi, please integrate the content below into a single paragraph. (I'm not familiar enough with the references do so confidently.)**
 
@@
 
-**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
-Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
-In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies.
-In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In forests with dense canopies, including tropical and temperate broadleaf forests, leaves in the upper canopy experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig.
@@
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
-Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
 

### 2021-09-13 vertical-thermal-review_SI.Rmd
Update vertical-thermal-review_SI.Rmd
issue #64

@@
 
-Energy balance for a typical overstory sun leaf and understory shade leaf were modeled in the R package *Tleaves* [@muir_tealeaves_2019], parameterized for *Quercus rubra* L.
+Energy balance for a typical overstory sun leaf and understory shade leaf were modeled in the R package *tealeaves* [@muir_tealeaves_2019], parameterized for *Quercus rubra* L.
 leaves at Harvard Forest, MA, USA (42.5369, -72.17266).
 
-*Describe where you came up with all the parameters:*
-
-Micrometeorological data from Harvard NEON was used to parameterize biophysical constants, using the mean oof maximum PAR, wind speed, and $T_{air}$ (Fig.3) for growing season months at vertical positions 60m (overstory) and 10m (understory).
+Micrometeorological data from the NEON tower at Harvard Forest (Fig.
+`r fig_NEON_vertical`, Supporting Information Methods `r appendix_NEON_methods`) was used to parameterize biophysical constants, using the mean of maximum PAR, wind speed, and $T_{air}$ for growing season months at vertical positions 60m (overstory) and 10m (understory).
 Leaf size parameters were selected for *Quercus rubra* sun and shade leaf based on personal observations (by Vinod) and verified as reasonable by @zwienieckiHydraulicLimitationsImposed2004.
 
-For the humid scenario, we used mean maximum RH.
+For the humid scenario, we used mean maximum RH from the NEON tower at Harvard Forest.
 Stomatal conductance measurements were referred from *Tleaves* typical sun and shade measurements [@muir_tealeaves_2019] and @cavender-baresChangesDroughtResponse2000.
 
-For the drought scenario, we reduced RH to 50% of the mean maximum.
-*For overstory drought PAR values, we used maximum Harvard NEON PAR value at 60m, and for understory drought PAR we used 50% increased PAR of understory normal value.* 
+For the drought scenario, we reduced RH to 50% of the mean maximum from the NEON tower at Harvard Forest.
+For drought PAR values, for the overstory we used the maximum PAR value observed at 60m height on the NEON tower at Harvard Forest, and for the understory we assumed 50% increased PAR relative to the mean daily maximum understory value.
 We reduced stomatal conductance to a minimum value of 0.01 $\mu mol/m^2/s/Pa$ for both overstory and understory.

### 2021-09-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).**
+Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).
 Closed canopies with high LAI act as an parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
@@
 
-**Leaf traits determine the capacity of leaves to thermoregulate [@michaletzEnergeticCarbonEconomic2016], and as such are critical to optimizing leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].**
+Leaf traits shape $T_{leaf}-T_{air}$ [@michaletzEnergeticCarbonEconomic2016] and leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].

### 2021-09-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf traits shape $T_{leaf}-T_{air}$ [@michaletzEnergeticCarbonEconomic2016] and leaf metabolism across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015; @zwienieckiHydraulicLimitationsImposed2004].
+Leaf traits shape $T_{leaf}$, leaf metabolism, and its thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015;@michaletzEnergeticCarbonEconomic2016;  @zwienieckiHydraulicLimitationsImposed2004].
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
@@
 Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
-The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table 1).
+The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
 However, when it comes to understanding and modeling forest ecosystem function, given that species traits strongly shape the propensity for trees to inhabit understory vs.
-canopy positions, ecosystem function is also shaped by interspecific trait variation [e.g., *Lamour, Serbin in prep*; @cavaleri_height_2010].
+canopy positions, ecosystem function is also shaped by interspecific trait variation [e.g., @cavaleri_height_2010].
 
@@
 Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-As much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004] which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
+Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
 It has been established that independently of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
-Therefore, ontogeny is bound to play a role in thermal sensitivity [@carterExperimentalWarmingTropical2021a; @niinemetsResponsesForestTrees2010a].
+Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021a; @niinemetsResponsesForestTrees2010a].
 
@@
 
-#### (Anatomical, Morphological and Physiological traits)
-
-Sun leaves have traits that reduce heat and water stress -- for instance, by being smaller, more pubescent, having higher vein density [Table 1, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].
+Sun leaves have anatomical, morphological, and physiological traits that reduce heat and water stress (Table `r table_leaf_traits`).
+They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ [*REFS*]. 
+Higher stomatal densities allow higher $g_s$.
+Higher cuticle thickness and trichome density reduce water loss, and higher vein densities... [*REFS*].
+Steeper leaf angles [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988] reduce radiation loads and thereby $T_{leaf}-T_{air}$.
+*...
+are more pubescent, having higher vein density [@leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].*
 These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance. 
-In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but they are more likely to overheat than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
+In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but their greater thermal capacitance makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
 In open canopies, where light is comparatively homogeneous, exposed leaves are shaped more by $T_{air_{max}}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].

### 2021-09-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
auto number tables

@@
 
-Elemental concentrations such as $N_{area}$ and $P_{area}$, contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021].
+Elemental concentrations such as $N_{area}$ and $P_{area}$ contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table `r table_leaf_traits`, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021].
 Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. 
@@
 
-Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
+Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
 Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
@@
 
-Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2).
+Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
 However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.
@@
 
-**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table 2; @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].**
-This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Table 1, Table 2).
+**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table `r table_leaf_metabolism`; @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].**
+This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).
 
@@
 Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020].
-However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
+However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
 The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
@@
 
-**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
+**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
 shade leaves [@carterExperimentalWarmingTropical2021a; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
@@
 Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
-Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
+Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table `r table_leaf_metabolism`).
 The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
@@
 Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
-This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table 2).
+This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`).
 More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
-This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table 1, @mcdowellRelationshipTreeHeight2011].
+This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
 
@@
 (c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization.
-Vertically resolved in situ data could include any variable listed in Table 1 or 2.](schematics/fig_remote_sensing.png)
+Vertically resolved in situ data could include any variable listed in Table `r table_leaf_traits` or 2.](schematics/fig_remote_sensing.png)
 

### 2021-09-13 vertical-thermal-review.Rmd
work on warming implications, + misc

@@
 
-#### (Biochemical and elemental traits, and isoprene)
-
 Elemental concentrations such as $N_{area}$ and $P_{area}$ contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table `r table_leaf_traits`, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021].
@@
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
-Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defences and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].
-
-#### (Leaf lifecycles)
+Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].
 
@@
 
-Warming impacts forests, from leaf metabolism to whole forest structure.
+As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves.
+These changes will impact leaf metabolism and ecosystem, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
+Our in-depth review sheds some light as to how responses are likely to vary across forest vertical gradients, yet numerous questions remain.
+
 Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$.
@@
 Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
-A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.
+
 Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. 
 A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
+
+Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$, they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
+A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.
+(*mention Rollinson, Fig. 4b*)
 In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
+
 Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].

### 2021-09-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
Tleaf section edits

@@
 In each scenario, an the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
-swr = short wave radiation (W/m^2^); ws = windspeed (m/s); rh= relative humidity; lw = leaf width (cm); gs = stomatal conductance (mmol m^2^/s).
+swr = short wave radiation (W/m^2^); ws = windspeed (m/s); rh= relative humidity; lw = leaf width (m); gs = stomatal conductance (mmol m^2^/s).
 ](leaf_energy_balance/fig3_leafenergy.png)
@@
 
-Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`).
-Closed canopies with high LAI act as an parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
-In contrast to closed canopies, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
-This mixing and light penetration can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy. 
-The latter can result from the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].
-
-In forests with dense canopies, including tropical and temperate broadleaf forests, leaves in the upper canopy experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
-In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig.
+Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig.
+`r fig_leaf_T`).
+Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory.
+This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig.
 `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
@@
 Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
-
 Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. 

### 2021-09-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
-In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime ~CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
-Large trees also dominate in terms of woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests [Fig.
+In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
+Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests [Fig.
 `r fig_ecosystem`b, @piponiot_sizerelated_inreview].

### 2021-09-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
issue #33

@@
 `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @hadleyInfluenceKrummholzMat1987; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
-These usually-modest gradients in $T_{air}$ combine with the above-described gradients in other biophysical conditions to shape $T_{leaf}$.
+These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$.
 
@@
 
-$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf metabolism and thermal responses") with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf metabolism and thermal responses"), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
 Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high windspeeds, and high $T_{air}$). 

### 2021-09-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
#65

@@
 
-**Nidhi, please integrate the content below into a single paragraph. (I'm not familiar enough with the references do so confidently.)**
-
-Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig.
-`r fig_leaf_T`).
+Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
@@
 This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig.
-`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987].
+`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
 The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].

### 2021-09-14 vertical-thermal-review.Rmd
finished reviewing traits section

@@
 The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
-For example, deciduousness is an adaptation to conserve water during cold or dry periods [@meakemRoleTreeSize2018], but is also structured across the vertical profile.
-Within tropical deciduous species, the fraction of seasonally deciduous individuals increases toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
+For example, in a temperate deciduous forest, spring bud-break tended to occur earlier among juvenile trees in the more sheltered understory, where temperature sums accumulated more rapidly, than among conspecific canopy trees [@augsbergerDifferencesLeafPhenology2003].
+Within tropical dry-deciduous species in Panama, the fraction of seasonally deciduous individuals increased toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
 Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019].

### 2021-09-14 vertical-thermal-review.Rmd
conductance section

@@
 Sun leaves have anatomical, morphological, and physiological traits that reduce heat and water stress (Table `r table_leaf_traits`).
-They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ [*REFS*]. 
+They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ and lower $T_{leaf}-T{air}$ [Fig. `r fig_leaf_T`, *REFS*]. 
 Higher stomatal densities allow higher $g_s$.
@@
 
-### Stomatal conductance
+### Conductance
 
+Leaf hydraulic, stomatal, and boundary layer conductance are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
+Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig.
+`r fig_NEON_vertical`, Table `r table_leaf_traits`), boundary layer conductance increases with height (Table `r table_leaf_metabolism`).
 Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
-However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.
+However, because water supply often cannot meat the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`).
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].

### 2021-09-14 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
updating Tleaf figure

@@ -4,7 +4,7 @@
 # Tleaves R package created by Chris Muir, 2019
 # Developed by: Nidhi Vinod, contact Anderson-Teixeira (teixeirak@si.edu)
 # R version 3.5.3 --First created in October 2020, updated on 5/28/2021
-# Oct-August 2021
+# Oct,2020-Oct,2021
 ####################################
 #install packages if needed
 library(latex2exp)
@@ -18,14 +18,7 @@ library(cowplot)
 library(gridExtra)
 library(RColorBrewer)
 ####################################
-#dataframe for biophysical parameters from NEON HARV data
-#normal scenario represents HARV NEON environmental parameters as observed. 
-#drought scenario represents similar HARV NEON parameters, with: 1) max PAR for overstory and 50% increase in normal PAR for understory, and 2) 50% decrease in normal relative humidity for both positions.
-#typical normal stomtal conductane(gs) values are referred from Tleaves supplementary manual for range of minimum and maximum values and obtained from Cavendar-Bares and Bazzaz, 2000.
-#drought condition ~gs values are set to a constant value of 0.01 umol/m^2/s/Pa for canopy and understory. 
-#respective leaf sizes (m) are measured from sun and shade red oak leaves 
-#all variables are constant except for the y-axis variables that represent minimum- maximum range
-####################################
+
 
 # create a dataframe for overstory and understory, for scenarios: normal and drought
 # 1. Short wave radiation, normal
@@ -152,8 +145,8 @@ dr_sw_u$dr_u_tleaf_tair<- dr_sw_u$tleaf - dr_sw_u$tair
 
 sw<- data.frame(l_tla = sw_l$l_tleaf_tair, u_tla = sw_u$u_tleaf_tair, dr_ltla = dr_sw_l$dr_l_tleaf_tair, dr_utla = dr_sw_u$dr_u_tleaf_tair, par = sw_l$par)
 
-l_tla<-data.frame(par = sw$par, tleaf_tair = sw$l_tla, canopy_position = "understory normal")
-u_tla<-data.frame(par = sw$par, tleaf_tair = sw$u_tla, canopy_position = "overstory normal")
+l_tla<-data.frame(par = sw$par, tleaf_tair = sw$l_tla, canopy_position = "understory humid")
+u_tla<-data.frame(par = sw$par, tleaf_tair = sw$u_tla, canopy_position = "overstory humid")
[…]

### 2021-09-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-Within a stand, woody growth rates , measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, Stephenson et al.
-2014].
+Within a stand, woody growth rates , measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, @stephensonRateTreeCarbon2014].
 NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

### 2021-09-14 vertical-thermal-review.Rmd
photosynthesis section

--- 2021-09-14 vertical-thermal-review.Rmd
@@
 
-**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table `r table_leaf_metabolism`; @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].**
-This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).
+Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
+This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under conditions of high light and sufficient water (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).
 
@@
 
-**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
+We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
 Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
@@
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a].
-
 @hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 

### 2021-09-14 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-**Leaf photosynthetic thermal tolerance**
 Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020]. 

### 2021-09-14 vertical-thermal-review.Rmd
photosynthesis section
Respiration section

@@
 
-**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
+Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].
+Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
 shade leaves [@carterExperimentalWarmingTropical2021a; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
-This may vary based on forest type [**REF**], acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015], or leaf traits and leaf age, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
+This may vary based on forest type, acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015], or leaf traits and leaf age. 
+In evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. 
 Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021a]. 
+Overall, however, we currently lack synthetic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
 

### 2021-09-14 vertical-thermal-review.Rmd
isoprenes

@@
 
-**The vertical structuring of leaf isoprene emissions may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.** Isoprene gas emission from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
+Isoprene gas emission from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
 Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
-In contrast, a recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy region instead of the hotter and brighter upper canopy [@taylorNewFieldInstrument2021].
+In contrast, vertical structuring of leaf isoprene emissions within forests may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
+A recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy region instead of the hotter and brighter upper canopy [@taylorNewFieldInstrument2021].
 This observation is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [@sharkeyIsopreneEmissionPlants2008], due to the sun's passage across a dynamically structured canopy surface.

### 2021-09-14 vertical-thermal-review.Rmd
tweak isoprene paragraph

@@
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
-Heat-sensitive emissions of isoprene gas from tree leaves enhance photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].
+Capacity for heat-sensitive emissions of isoprene gas from tree leaves enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].
 
@@
 
-Isoprene gas emission from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
+Emission of the VOC isoprene gas from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
 Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
@@
 This observation is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [@sharkeyIsopreneEmissionPlants2008], due to the sun's passage across a dynamically structured canopy surface.
-This review focuses primarily on variation in mean conditions across the vertical profile, but future work should seek to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function.
+Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of isoprenes in forest thermal sensitivity across the vertical gradient.
 

### 2021-09-14 vertical-thermal-review.Rmd
starting ecology section
@NidhiVinod , I have to pause now and give some attention to a deadline. Hoping to get back to this later today!

@@
 
-**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect whole-tree ecology, ecosystem ecology, and their temperature responses (Fig.
-`r fig_schematic`).**
+Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect whole-tree ecology, ecosystem ecology, and their temperature responses (Fig.
+`r fig_schematic`).
 
@@
 Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
-Specifically, increases are linked to increased leaf area and in part to an increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table `r table_leaf_metabolism`).
+Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
 The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).

### 2021-09-15 vertical-thermal-review.Rmd
bit of work on ecology section

@@
 
-**Vertical gradients affect tree metabolism and growth.**
-Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured as diameter at breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, and height, all of which scale with DBH [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
+Tree metabolism and growth and shaped by the positioning of their crowns within the vertical gradient.
+Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of whole-tree transpiration and photosynthesis [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
 Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
 The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-Within a stand, woody growth rates , measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, @stephensonRateTreeCarbon2014].
+Within a stand, woody growth rates, measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, @stephensonRateTreeCarbon2014].
 NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

### 2021-09-15 vertical-thermal-review.Rmd
small edits

@@
 Sun leaves have anatomical, morphological, and physiological traits that reduce heat and water stress (Table `r table_leaf_traits`).
-They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ and lower $T_{leaf}-T{air}$ [Fig. `r fig_leaf_T`, *REFS*]. 
-Higher stomatal densities allow higher $g_s$.
-Higher cuticle thickness and trichome density reduce water loss, and higher vein densities... [*REFS*].
-Steeper leaf angles [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988] reduce radiation loads and thereby $T_{leaf}-T_{air}$.
-*...
-are more pubescent, having higher vein density [@leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and a steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].*
-These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance. 
+They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ and lower $T_{leaf}-T{air}$ [Fig.
+`r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a].
+Higher stomatal and vein densities allow higher $g_s$ and facilitate higher transpiration respectively [@zwienieckiHydraulicLimitationsImposed2004].
+Higher cuticle thickness and trichome density reduce water loss.
+Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].
+These traits collectively aid heat dissipation and regulate T_leaf-photosynthetic balance.
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but their greater thermal capacitance makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].

### 2021-09-17 vertical-thermal-review.Rmd
rearranged refs for traits
move implications header up a level
fix table 1 title

@@
 \newpage
-**Table `r table_leaf_traits`.** Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
+**Table `r table_leaf_traits`.** Summary of observed variation in leaf traits across the vertical gradient and/or between sun and shade leaves.
+Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
 
@@
 Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].
-These traits collectively aid heat dissipation and regulate T_leaf-photosynthetic balance.
+These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but their greater thermal capacitance makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
@@
 
-## Implications
+# Implications
 

### 2021-09-19 vertical-thermal-review.Rmd
updating Fig 1, and change "whole-tree" to "tree"

@@
 Rising temperatures are profoundly influencing forests on many scales with potentially strong variation vertically across forest strata.
-Using published research and new analyses, we evaluate how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping whole-tree ecology and ecosystem function.
+Using published research and new analyses, we evaluate how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping tree ecology and ecosystem function.
 In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and can experience elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
@@
 Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.
-Global average temperatures have risen 1.2 $^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5 $^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 The net result of higher temperatures may be increased or decreased tree growth and forest carbon sequestration, with decreases being more commonly documented across the world's forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021].
-The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
+The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
 
@@
 (4) leaf metabolism and gas exchange, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
-(5) whole-tree and ecosystem ecology.
+(5) tree and ecosystem ecology.
 We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.
@@
 ![**Figure `r fig_schematic`.
-Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) whole-tree and ecosystem ecology.** Abbreviations are as follows: VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; T sensitivity: temperature sensitivity; ET: evapotranspiration.
+Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology.** Abbreviations are as follows: VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; T sensitivity: temperature sensitivity; ET: evapotranspiration.
 Here, we summarize conditions in dense canopies during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
@@
 
-## Whole-tree and ecosystem ecology
+## Tree and ecosystem ecology
 
-Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect whole-tree ecology, ecosystem ecology, and their temperature responses (Fig.
+Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect tree ecology, ecosystem ecology, and their temperature responses (Fig.
 `r fig_schematic`).
@@
 Tree metabolism and growth and shaped by the positioning of their crowns within the vertical gradient.
-Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of whole-tree transpiration and photosynthesis [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
+Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration and photosynthesis [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
 Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
@@
 Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in whole-tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
+Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
 This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`).

### 2021-09-19 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Vertical gradients in the biophysical environment for six US forest sites in the National Ecological Observatory Network (NEON)**.
-Height profiles are shown for: growing season (a) leaf area density, (b) proportion of sun leaves, and (c) proportion light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$.
-Measurements extend from ground level (height = 0 m) to the top of the canopy (a-c, h) oor above (d-g) Sites, which represent a variety of forest structures, include a mixed northern hardwood forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
+Height profiles are shown for: growing season (a) leaf area density, (b) proportion of sun leaves, and (c) proportion light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature ($T_{air}$), and (h) maximum biological temperature, ($T_{bio}$).
+Measurements extend from ground level (height = 0 m) to the top of the canopy (a-c, h) or above (d-g).
+Sites, which represent a variety of forest structures, include a mixed northern hardwood forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
 Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.
@@
 `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
-Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom (Fig.
+Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile (Fig.
 `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
@@
 The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021].
-Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
 Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Fig.
@@
 Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
-In each scenario, an the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
+In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
 swr = short wave radiation (W/m^2^); ws = windspeed (m/s); rh= relative humidity; lw = leaf width (m); gs = stomatal conductance (mmol m^2^/s).

### 2021-09-19 vertical-thermal-review.Rmd
traits table change-- "physiological" to "biophysical"

@@
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
 In addition to lower wind speeds (Fig.
@@
 
-Leaf traits shape $T_{leaf}$, leaf metabolism, and its thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015;@michaletzEnergeticCarbonEconomic2016;  @zwienieckiHydraulicLimitationsImposed2004].
+Leaf traits shape $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015;@michaletzEnergeticCarbonEconomic2016;  @zwienieckiHydraulicLimitationsImposed2004].
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].

### 2021-09-19 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Leaf traits shape $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015;@michaletzEnergeticCarbonEconomic2016;  @zwienieckiHydraulicLimitationsImposed2004].
-Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
+Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Fig.
@@
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but their greater thermal capacitance makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies, where light is comparatively homogeneous, exposed leaves are shaped more by $T_{air_{max}}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
+In open canopies, where light is comparatively homogeneous, exposed leaves are shaped more by maximum air temperature ($T_{air_{max}}$) and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
 
-Elemental concentrations such as $N_{area}$ and $P_{area}$ contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table `r table_leaf_traits`, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021].
+Elemental concentrations such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table `r table_leaf_traits`, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021].
 Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. 

### 2021-09-19 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. 
-However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of N and P and foliar thermal tolerance.
+However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of nitrogen and phosphorus and foliar thermal tolerance.
 
@@
 In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
-For example, in a temperate deciduous forest, spring bud-break tended to occur earlier among juvenile trees in the more sheltered understory, where temperature sums accumulated more rapidly, than among conspecific canopy trees [@augsbergerDifferencesLeafPhenology2003].
+For example, in a temperate deciduous forest, spring bud-break tended to occur earlier among juvenile trees in the more sheltered understory, where temperature sums accumulated more rapidly, than among conspecific canopy trees [@augspurgerDifferencesLeafPhenology2003].
 Within tropical dry-deciduous species in Panama, the fraction of seasonally deciduous individuals increased toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
@@
 
-Leaf hydraulic, stomatal, and boundary layer conductance are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
-Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig.
-`r fig_NEON_vertical`, Table `r table_leaf_traits`), boundary layer conductance increases with height (Table `r table_leaf_metabolism`).
+Leaf hydraulic, stomatal, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
+Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
 Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
-However, because water supply often cannot meat the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`).
+However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`).
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
@@
 Temperature can affect photosynthesis via direct and indirect pathways: directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
+Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure [e.g.
@@
 Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
-@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [@campbell_introduction_1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
+Similarly, $T_{opt}$ of RuBP carboxylation  ($Vc_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a].
@@
 Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
-In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
+In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
 Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
@@
 
-Tree metabolism and growth and shaped by the positioning of their crowns within the vertical gradient.
+Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
 Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration and photosynthesis [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
 Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
-The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-Within a stand, woody growth rates, measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006, @stephensonRateTreeCarbon2014].
+The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
+Within a stand, woody growth rates, measured as radial stem growth or biomass increment,  generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006; @stephensonRateTreeCarbon2014].
 NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].

### 2021-09-20 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
tried increasing table font size
Please let me know if the font still appears to be small

@@ -157,7 +157,7 @@ sw<- data.frame(rbind(u_tla, l_tla, dr_utla, dr_ltla))
 a1<-ggplot(data = sw)+
   geom_smooth(aes(x = par, y = tleaf_tair, color = canopy_position, linetype = canopy_position), 
               method = lm, se = FALSE)+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("short wave radiation (swr, W/m$^{2}$)"))+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("Short Wave Radiation (swr, W/m$^{2}$)"))+
   theme_few()+theme(text = element_text(size = 14))+ylim(-5, 14)+
   geom_hline(yintercept=0, linetype='dotted', color = 'black')+
   scale_colour_manual(values=c("#008837","#7b3294","#008837","#7b3294"), name = "canopy position") + 
@@ -238,7 +238,7 @@ b1<-ggplot(ws)+
   geom_smooth(aes(x = wind, y = u_tla),  method = lm, color = "#008837", se = FALSE)+
   geom_smooth(aes(x = wind, y = dr_ltla),  method = lm, color = "#7b3294", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = wind, y = dr_utla),  method = lm, color = "#008837", se = FALSE, linetype = "dashed")+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ ($°C$)"))+xlab(TeX("windspeed (ws, $m/s$)"))+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ ($°C$)"))+xlab(TeX("Wind Speed (ws, $m/s$)"))+
   geom_hline(yintercept=0, linetype='dotted', col = 'black')+
   theme_few()+theme(text = element_text(size = 14))+ ylim(-5, 14) 
 
@@ -300,7 +300,7 @@ c1<-ggplot(gs)+
   geom_smooth(aes(x = gs, y = u_tla),  method = lm, color = "#008837", se = FALSE)+
   geom_smooth(aes(x = gs, y = dr_ltla),  method = lm, color = "#7b3294", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = gs, y = dr_utla),  method = lm, color = "#008837", se = FALSE, linetype = "dashed")+
[…]

### 2021-09-20 vertical-thermal-review.Rmd
small spelling edits

@@
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
-In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves, which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+In combination with $T_{air}$, RH determines vapor pressure deficit (VPD)-- the driving force of water loss from leaves--which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
@@
 $T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf metabolism and thermal responses"), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
-Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high windspeeds, and high $T_{air}$). 
-Leaves are typically warmer than air when under full sunlight, especially under slow windspeeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
+Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
+Leaves are typically warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
 While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020].
@@
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
-swr = short wave radiation (W/m^2^); ws = windspeed (m/s); rh= relative humidity; lw = leaf width (m); gs = stomatal conductance (mmol m^2^/s).
+swr = short wave radiation (W/m^2^); ws = wind speed (m/s); rh= relative humidity; lw = leaf width (m); gs = stomatal conductance (mmol m^2^/s).
 ](leaf_energy_balance/fig3_leafenergy.png)
@@
 `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, at high solar radiation loads, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at maximum $g_s$, but during $g_s$ - limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
+Therefore, at high solar radiation loads, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at maximum $g_s$, but during $g_s$ limitation, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
 
@@
 Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
-This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$-limitation than shade leaves [Fig.
+This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$ limitation than shade leaves [Fig.
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
@@
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
@@
 `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a].
-Higher stomatal and vein densities allow higher $g_s$ and facilitate higher transpiration respectively [@zwienieckiHydraulicLimitationsImposed2004].
+Higher stomatal and vein densities allow higher $g_s$ and facilitate higher transpiration [@zwienieckiHydraulicLimitationsImposed2004].
 Higher cuticle thickness and trichome density reduce water loss.
@@
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
-Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
+Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
 
@@
 Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
-In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
+In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
 Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
@@
 Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
-This is consistent with the observation that $g_s$-limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`).
+This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`).
 More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
@@
 Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
-However, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
+However, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees exist.
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
@@
 
-Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$.
-Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+Leaf level $T_{opt}$ is closely related to ecosystem $T_{opt}$.
+Temperature limitations to leaf processes, such as VPD and $g_s$ limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
 At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima.
 Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
+Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperate and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
 Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. 
-A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
+A daytime temperature threshold of 32$^\circ$C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
 
 Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$, they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
-A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.
-(*mention Rollinson, Fig. 4b*)
+A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
 In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
@@
 Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015].
-The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging [[@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019].
+The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging [@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019].
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]

### 2021-09-21 vertical-thermal-review.Rmd
Pau et al. 2018 note from Elsa and Lawren included

@@
 Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
-In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
+In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LiDAR-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
 Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020 in review].
-[@pauTropicalForestTemperature2018] used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than air temperatures or VPD.
+@pauTropicalForestTemperature2018 used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
+However, the authors did not consider vertical light environment conditions even though light directly influences the deviation between canopy temperature and $T_{air}$.
 In a savanna system in the western U.S., [@johnstonWhatLiesVertical2020 in review] found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies.

### 2021-09-21 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
made axis font bigger

@@ -158,7 +158,7 @@ a1<-ggplot(data = sw)+
   geom_smooth(aes(x = par, y = tleaf_tair, color = canopy_position, linetype = canopy_position), 
               method = lm, se = FALSE)+
   ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("Short Wave Radiation (swr, W/m$^{2}$)"))+
-  theme_few()+theme(text = element_text(size = 14))+ylim(-5, 14)+
+  theme_few()+theme(text = element_text(size = 15))+ylim(-5, 14)+
   geom_hline(yintercept=0, linetype='dotted', color = 'black')+
   scale_colour_manual(values=c("#008837","#7b3294","#008837","#7b3294"), name = "canopy position") + 
   scale_linetype_manual(values = c(1,1,2,2), name = "canopy position")+
@@ -240,7 +240,7 @@ b1<-ggplot(ws)+
   geom_smooth(aes(x = wind, y = dr_utla),  method = lm, color = "#008837", se = FALSE, linetype = "dashed")+
   ylab(TeX("$T_{Leaf}$ - $T_{air}$ ($°C$)"))+xlab(TeX("Wind Speed (ws, $m/s$)"))+
   geom_hline(yintercept=0, linetype='dotted', col = 'black')+
-  theme_few()+theme(text = element_text(size = 14))+ ylim(-5, 14) 
+  theme_few()+theme(text = element_text(size = 15))+ ylim(-5, 14) 
 
 b1
 
@@ -302,7 +302,7 @@ c1<-ggplot(gs)+
   geom_smooth(aes(x = gs, y = dr_utla),  method = lm, color = "#008837", se = FALSE, linetype = "dashed")+
   ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("Stomatal Conductance ($g_{s}$, mmol $m^{-2} s^{-1}$)"))+
   geom_hline(yintercept=0, linetype='dotted', col = 'black')+
-  theme_few()+theme(text = element_text(size = 14))+ylim(-5, 14)
+  theme_few()+theme(text = element_text(size = 15))+ylim(-5, 14)
 c1
 
 #########################################################################################################
@@ -363,7 +363,7 @@ d1<-ggplot(ls)+
   geom_smooth(aes(x = ls, y = dr_utla),  method = lm, color = "#008837", se = FALSE, linetype = "dashed")+
[…]

### 2021-09-21 vertical-thermal-review.Rmd
Eleinis' edit included

@@
 ![**Figure `r fig_treerings`.
-Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species.
+Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species.
 Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.
@@
 
-Leaf level $T_{opt}$ is closely related to ecosystem $T_{opt}$.
-Temperature limitations to leaf processes, such as VPD and $g_s$ limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+Leaf level $T_{opt}$ is closely related to the $T_{opt}$ of ecosystem photosynthesis ($T_{optE}$).
+Temperature limitations to leaf processes, such as VPD and $g_s$ limitation, can modify $T_{optE}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
 At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima.

### 2021-09-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-## Tree and ecosystem ecology
+## Tree ecology
 
@@
 
-Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
+Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
 
@@
 Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015].
-The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging [@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019].
+The disproportionate impact of stress on large trees holds for other types of disturbance, including forest fragmentation and logging [@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019].
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
@@
 
-Remote sensing data provide a valuable means to scale between in situ observations and DGVMs.
-Specifically, the increasing availability of airborne and spaceborne LiDAR and thermal remote sensing data offer a promising opportunity for mapping vertical thermal gradients in combination with vegetation structure at locations with in situ data and across larger landscapes (Fig.
+Remote sensing data provide a valuable means to scale between *in situ* observations and DGVMs.
+Specifically, the increasing availability of airborne and spaceborne lidar and thermal remote sensing data offer a promising opportunity for mapping vertical thermal gradients in combination with vegetation structure at locations with in situ data and across larger landscapes (Fig.
 `r fig_model`).
@@
 Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
-In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LiDAR-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
+In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
 Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020 in review].

### 2021-09-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high in the understory as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a, @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993] *[@parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985; @ishiiExploringRelationshipsCanopy2004a; @parkerCanopyLightTransmittance2002]*.
+`r fig_NEON_vertical`a, @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985].
 In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig.
-`r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016; @ishiiExploringRelationshipsCanopy2004a].
+`r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016].
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020a].
@@
 With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) and larger $g_s$ [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
-However, when leaf water demand (influenced by VPD and stomatal opening) exceeds the rate of supply, conservation of water ($g_s$ - limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
+However, when leaf water demand (influenced by VPD and stomatal opening) exceeds the rate of supply, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`e, @fauset_differences_2018].
@@
 Sun leaves have anatomical, morphological, and physiological traits that reduce heat and water stress (Table `r table_leaf_traits`).
-They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ and lower $T_{leaf}-T{air}$ [Fig.
+They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ and lower $T_{leaf}-T_{air}$ [Fig.
 `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a].
 Higher stomatal and vein densities allow higher $g_s$ and facilitate higher transpiration [@zwienieckiHydraulicLimitationsImposed2004].
-Higher cuticle thickness and trichome density reduce water loss.
-Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].
+Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].
 These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
@@
 This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`).
-More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
+More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower the capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
 This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].

### 2021-09-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 For example, African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021].
-Species with smaller leaves and larger gs will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020] and isoprene emitting species will have a selective advantage in a warming world [@taylorCapacityEmitIsoprene2019].
+Species with smaller leaves and larger $g_s$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020] and isoprene emitting species will have a selective advantage in a warming world [@taylorCapacityEmitIsoprene2019].
 Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] will also help forests cope with warming.
@@
 This finding was the opposite of what was expected and occurred in association with very high grass temperatures.
-The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of in situ measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
+The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of *in situ* measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
 

### 2021-09-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Sun leaves have anatomical, morphological, and physiological traits that reduce heat and water stress (Table `r table_leaf_traits`).
-They are generally smaller, thicker, and more deeply lobed, resulting in higher $g_b$ and lower $T_{leaf}-T_{air}$ [Fig.
+Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to increased reflectance, lower wettability, and/or higher $g_b$, resulting in lower $T_{leaf}$-$T_{air}$ and/or reduced water stress (Table `r table_leaf_traits`).
+Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed, with greater cuticle thickness, and higher trichome density [Fig.
 `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a].
-Higher stomatal and vein densities allow higher $g_s$ and facilitate higher transpiration [@zwienieckiHydraulicLimitationsImposed2004].
-Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988].
-These traits collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance.
-In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA), but their greater thermal capacitance makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies, where light is comparatively homogeneous, exposed leaves are shaped more by maximum air temperature ($T_{air_{max}}$) and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
+Sun leaves tend to have higher photosynthetic rates per leaf area, related to higher nitrogen and chlorophyll concentrations per leaf area and lower intercellular CO~2~ concentrations.
+Further, sun leaves tend to have higher stomatal and vein densities would enable their higher transpiration rates, and that would facilitate effective cooling  [@zwienieckiHydraulicLimitationsImposed2004].
+Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], and, in addition to higher droplet contact angles, would have reduced wettability.
+Leaves in high light environments also exhibit greater leaf packing and clumping.
+In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance, larger leaf size and lower transpiration--makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
+In open canopies, where light is comparatively homogeneous, leaf traits may be more strongly related to maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].
 

### 2021-09-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Leaves in high light environments also exhibit greater leaf packing and clumping.
-In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance, larger leaf size and lower transpiration--makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
+In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
 In open canopies, where light is comparatively homogeneous, leaf traits may be more strongly related to maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].

### 2021-09-28 vertical-thermal-review.Rmd
update affiliations (Nidhi @ UCLA)
leaf traits paragraph
@NidhiVinod , I  put one line per sentence so its easier to see what I changed. But before I did that, I made two substantive edits. Please check. In particular, I'm not sure about the part between double asterisks. I thought this was driven by more stomatal closure.

Sun leaves tend to have higher photosynthetic rates per leaf area, made possible by higher nitrogen and chlorophyll concentrations per leaf area **and resulting in lower intercellular CO~2~ concentrations. **

In open canopies, where light is comparatively homogeneous, leaf

@@
 **Authors:** 
-Nidhi Vinod^1^,
-Martijn Slot^2^,
-Ian McGregor^3^,
-Elsa M. Ordway^4,5^,
+Nidhi Vinod^1,2^,
+Martijn Slot^3^,
+Ian McGregor^4^,
+Elsa M. Ordway^2,5^,
 Marielle N. Smith^6^,
 Tyeen Taylor^7^,
-Lawren Sack^5^,
-Kristina J. Anderson-Teixeira^1,2^*
+Lawren Sack^2^,
+Kristina J. Anderson-Teixeira^1,3^*
 
@@
 
-2. Smithsonian Tropical Research Institute; Panama, Republic of Panama
+2. Department of Ecology and Evolutionary Biology, UCLA, Los Angeles, CA 90039, USA
 
-3. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
+3. Smithsonian Tropical Research Institute; Panama, Republic of Panama
 
-4. Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA 02138, USA
+4. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
 
-5. Department of Ecology and Evolutionary Biology, UCLA, Los Angeles, CA 90039, USA
+5. Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA 02138, USA
 
@@
 `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a].
-Sun leaves tend to have higher photosynthetic rates per leaf area, related to higher nitrogen and chlorophyll concentrations per leaf area and lower intercellular CO~2~ concentrations.
+Sun leaves tend to have higher photosynthetic rates per leaf area, made possible by higher nitrogen and chlorophyll concentrations per leaf area and resulting in lower intercellular CO~2~ concentrations.
 Further, sun leaves tend to have higher stomatal and vein densities would enable their higher transpiration rates, and that would facilitate effective cooling  [@zwienieckiHydraulicLimitationsImposed2004].
@@
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies, where light is comparatively homogeneous, leaf traits may be more strongly related to maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].
+In open canopies, where light is comparatively homogeneous, leaf traits may be more shabed by maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].
 

### 2021-09-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
@NidhiVinod , please verify all these changes.

@@
 Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to increased reflectance, lower wettability, and/or higher $g_b$, resulting in lower $T_{leaf}$-$T_{air}$ and/or reduced water stress (Table `r table_leaf_traits`).
-Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed, with greater cuticle thickness, and higher trichome density [Fig.
+Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed, with greater cuticle thickness (protects against water loss), and higher trichome density (increases reflectance) [Fig.
 `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a].
 Sun leaves tend to have higher photosynthetic rates per leaf area, made possible by higher nitrogen and chlorophyll concentrations per leaf area and resulting in lower intercellular CO~2~ concentrations.
-Further, sun leaves tend to have higher stomatal and vein densities would enable their higher transpiration rates, and that would facilitate effective cooling  [@zwienieckiHydraulicLimitationsImposed2004].
-Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], and, in addition to higher droplet contact angles, would have reduced wettability.
+Further, sun leaves tend to have higher stomatal and vein densities, which enable higher $g_s$ and thereby facilitate effective cooling  [@zwienieckiHydraulicLimitationsImposed2004]. 
+Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], and also results in higher droplet contact angles and reduced wettability.
 Leaves in high light environments also exhibit greater leaf packing and clumping. 
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies, where light is comparatively homogeneous, leaf traits may be more shabed by maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].
+In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].
 

### 2021-09-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-### Tree metabolism and growth, and survival
+### Tree metabolism, growth, and survival
 
 Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
-Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration and photosynthesis [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015].
+Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
 Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).

### 2021-09-28 vertical-thermal-review.Rmd
edits to traits paragraph
@NidhiVinod , please check all this. I did a fair amount of rearranging.

@@
 
-Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to increased reflectance, lower wettability, and/or higher $g_b$, resulting in lower $T_{leaf}$-$T_{air}$ and/or reduced water stress (Table `r table_leaf_traits`).
-Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed, with greater cuticle thickness (protects against water loss), and higher trichome density (increases reflectance) [Fig.
-`r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a].
-Sun leaves tend to have higher photosynthetic rates per leaf area, made possible by higher nitrogen and chlorophyll concentrations per leaf area and resulting in lower intercellular CO~2~ concentrations.
-Further, sun leaves tend to have higher stomatal and vein densities, which enable higher $g_s$ and thereby facilitate effective cooling  [@zwienieckiHydraulicLimitationsImposed2004]. 
-Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], and also results in higher droplet contact angles and reduced wettability.
-Leaves in high light environments also exhibit greater leaf packing and clumping. 
+Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`).
+Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
+`r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], but with greater leaf packing and clumping (reduces $g_b$).
+Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
+Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see also "Leaf metabolism and thermal responses" section, @zwienieckiHydraulicLimitationsImposed2004].
+At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness (protects against water loss) and higher trichome density (reduces $g_b$; **REF?**).
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
@@
 However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of nitrogen and phosphorus and foliar thermal tolerance.
+*Sun leaves tend to have higher photosynthetic rates per leaf area, made possible by higher nitrogen and chlorophyll concentrations per leaf area and resulting in lower intercellular CO~2~ concentrations.*
 

### 2021-09-28 vertical-thermal-review.Rmd
revamped paragraph on biochemical traits related to photosynthesis
@NidhiVinod , please check all this

@@
 
-Elemental concentrations such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) contribute to high rates of leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table `r table_leaf_traits`, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014], and may be optimized throughout the vertical canopy gradient [@buckleyOptimalCarbonPartitioning2021].
-Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. 
-However, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between total foliar concentrations of nitrogen and phosphorus and foliar thermal tolerance.
-*Sun leaves tend to have higher photosynthetic rates per leaf area, made possible by higher nitrogen and chlorophyll concentrations per leaf area and resulting in lower intercellular CO~2~ concentrations.*
+Leaf biochemical traits also vary strongly across light and height gradients (Table `r table_leaf_traits`), both enabling and resulting from gradients in metabolism (see "Leaf metabolism and thermal responses" section, Table `r table_leaf_metabolism`).
+Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
+Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see "Leaf metabolism and thermal responses" section, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).
 

### 2021-09-29 vertical-thermal-review.Rmd
add acknowledgements
@NidhiVinod , FYI. I added the people who have contributed figure materials and lab members for feedback (I especially mentioned Eleinis because she put a lot into her review). Please add others as you feel appropriate.

@@
 ## Acknowledgements
-Thanks to ...., the ForestGEO Ecosystems & Climate lab at SCBI for helpful discussion.
+Thanks to Norbert Kunert, Camille Piponiot, and Andrew Davies for providing figure materials. 
+This manuscript benefited from feedback from Eleinis Ávila-Lovera, ... and the ForestGEO Ecosystems & Climate lab at SCBI.
 Funding was provided by the Smithsonian Institution, ....

### 2021-09-29 vertical-thermal-review.Rmd
rename gas exchange section (plus minor edits)

@@
 (3) the leaf traits that most strongly influence $T_{leaf}$;
-(4) leaf metabolism and gas exchange, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
+(4) leaf gas exchange, including stomatal and boundary layer conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
 (5) tree and ecosystem ecology.
@@
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "Leaf metabolism and thermal responses"), but also greater propensity towards elevated $T_{leaf}$ (see section "Leaf temperature").
+The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "Leaf gas exchange and its thermal sensitivity"), but also greater propensity towards elevated $T_{leaf}$ (see section "Leaf temperature").
 
@@
 
-$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf metabolism and thermal responses"), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf gas exchange and its thermal sensitivity"), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
 Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
@@
 However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
-In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf metabolism and thermal responses" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
+In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf gas exchange and its thermal sensitivity" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
@@
 Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
-Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see also "Leaf metabolism and thermal responses" section, @zwienieckiHydraulicLimitationsImposed2004].
+Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see also "Leaf gas exchange and its thermal sensitivity" section, @zwienieckiHydraulicLimitationsImposed2004].
 At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness (protects against water loss) and higher trichome density (reduces $g_b$; **REF?**).
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].
+In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum air temperature ($T_{air_{max}}$) and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
 
-Leaf biochemical traits also vary strongly across light and height gradients (Table `r table_leaf_traits`), both enabling and resulting from gradients in metabolism (see "Leaf metabolism and thermal responses" section, Table `r table_leaf_metabolism`).
+Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`).
 Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
 Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see "Leaf metabolism and thermal responses" section, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).
 
@@
 
-## Leaf metabolism and thermal responses
+## Leaf gas exchange and its thermal sensitivity
 
@@
 \newpage
-**Table `r table_leaf_metabolism`.** Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves.
+**Table `r table_leaf_metabolism`.** Summary of observed variation in Leaf gas exchange and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
 Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.

### 2021-09-29 vertical-thermal-review.Rmd
add numbering system to section headers
Update vertical-thermal-review.Rmd

@@
 
-# Introduction
+# I. Introduction
 
@@
 (3) the leaf traits that most strongly influence $T_{leaf}$;
-(4) leaf gas exchange, including stomatal and boundary layer conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
+(4) leaf gas exchange (including stomatal and boundary layer conductance, photosynthesis, respiration, and volatile organic compound emission) and its thermal sensitivity; and 
 (5) tree and ecosystem ecology.
@@
 
-# Review of vertical gradients
+# II. Review of vertical gradients
 
-## The biophysical environment
+## 1. The biophysical environment
 
@@
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "Leaf gas exchange and its thermal sensitivity"), but also greater propensity towards elevated $T_{leaf}$ (see section "Leaf temperature").
+The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "4.
+Leaf gas exchange and its thermal sensitivity"), but also greater propensity towards elevated $T_{leaf}$ (see section "2.
+Leaf temperature").
 
@@
 `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "2.
+Leaf temperature").
 
@@
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section "Leaf temperature").
+The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section "2. Leaf temperature").
 
@@
 
-## Leaf temperature
+## 2. Leaf temperature
 
@@
 
-## Leaf traits
+## 3. Leaf traits
 
@@
 
+*beta carotene and lutein are part of the caratenoids pool that responsible for light absorption, and also heat energy dissipation*
+
 Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
@@
 
-## Leaf gas exchange and its thermal sensitivity
+## 3. Leaf gas exchange and its thermal sensitivity
 
@@
 
-### VOC production
+### VOC emissions
 
@@
 
-## Tree ecology
+## 4. Tree and ecosystem ecology
 
@@
 
-# Implications
+# III.  Implications
 
@@
 
-## Conclusions
+## IV. Conclusions
 

### 2021-09-29 vertical-thermal-review.Rmd
update treatment of VOCs, also more detailed section headers
@NidhiVinod , FYI, and you also were going to add some text on monoterpenes.

@m-n-smith and @tyeentaylor , I'd appreciate it if you could review these changes to make sure I got everything right, and also fill in at least one ref ( section 1).

It will be easiest to see the changes if you just review this commit.

@@
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "4.
-Leaf gas exchange and its thermal sensitivity"), but also greater propensity towards elevated $T_{leaf}$ (see section "2.
-Leaf temperature").
+Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [**REF- Ty or Marielle**]
+The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see section 2.2).
 
@@
 `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "2.
-Leaf temperature").
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section 2.2).
 
@@
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section "2. Leaf temperature").
+The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section 2.2).
 
@@
 
-### Biophysical drivers of $T_{leaf}$
+### 2.1. Biophysical drivers of $T_{leaf}$
 
@@
 
-### Vertical gradients
+### 2.2 Vertical gradients in leaf temperature
 
@@
 
-### Intraspecific variation
+### 3.1. Intraspecific trait variation
 
@@
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
-Capacity for heat-sensitive emissions of isoprene gas from tree leaves enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021].
+Capacity for heat-sensitive emissions of volatile organic compounds (VOCs), including isoprenes and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; **REF ON MONOTERPENES?**].
+VOC emission capacity is a species-level trait [**REFS**], such that intrapspecific variation across across vertical gradients are not applicable, but frequency of this trait varies across the vertical gradient (see section 3.2), as do emission rates [see section 4.4, @taylorNewFieldInstrument2021].
 
@@
 
-### Interspecific variation
+### 3.2. Interspecific trait variation
 
 Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
-Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018]. 
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [mirroring the faction of individuals, @conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018], and a recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy [see section 4.4, @taylorNewFieldInstrument2021].
 However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
@@
 
-## 3. Leaf gas exchange and its thermal sensitivity
+## 4. Leaf gas exchange and its thermal sensitivity
 
@@
 
-### Conductance
+### 4.1. Conductance
 
@@
 
-### Photosynthesis
+### 4.2. Photosynthesis
 
@@
 
-### Respiration
+### 4.3. Respiration
 
@@
 
-### VOC emissions
+### 4.4. VOC emissions
 
-Emission of the VOC isoprene gas from tree leaves is associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
+Emission of the VOC isoprene gas from tree leaves is a temperature-sensitive process associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
 Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
 In contrast, vertical structuring of leaf isoprene emissions within forests may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
-A recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy region instead of the hotter and brighter upper canopy [@taylorNewFieldInstrument2021].
-This observation is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [@sharkeyIsopreneEmissionPlants2008], due to the sun's passage across a dynamically structured canopy surface.
+The observation that both the abundance of emitting species (Table ) and average emission rates were highest in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021] -- as opposed to the the hotter and brighter upper canopy-- is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
 Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of isoprenes in forest thermal sensitivity across the vertical gradient.
 
-## 4. Tree and ecosystem ecology
+## 5. Tree and ecosystem ecology
 
@@
 
-### Tree metabolism, growth, and survival
+### 5.1. Tree metabolism, growth, and survival
 
@@
 
-### C and water flux
+### 5.2. C and water flux
 

### 2021-09-29 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
@NidhiVinod , please verify that this is correct.

@@
 
-*beta carotene and lutein are part of the caratenoids pool that responsible for light absorption, and also heat energy dissipation*
-
 Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Carotenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin (VAZ) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].

### 2021-09-29 vertical-thermal-review.Rmd
update figure caption
Thanks, @eoway !

Co-Authored-By: Elsa Ordway <22920153+eoway@users.noreply.github.com>

@@
 (c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization.
-Vertically resolved in situ data could include any variable listed in Table `r table_leaf_traits` or 2.](schematics/fig_remote_sensing.png)
+Vertically resolved in situ data could include any variable listed in Table 1 or 2.
+Panel b shows drone lidar point cloud data colored with thermal data from a savanna woodland in Kruger National Park, South Africa illustrating vertical and horizontal variation in vegetation temperature.
+Image courtesy of the Harvard Animal-Landscape Observatory, Peter B.
+Boucher and Andrew B.
+Davies.](schematics/fig_remote_sensing.png)
 

### 2021-09-29 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
@tyeentaylor , @m-n-smith -- one more thing to check

And @NidhiVinod (and @tyeentaylor ?)- should we add something on monoterpenes?

@@
 Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of isoprenes in forest thermal sensitivity across the vertical gradient.
+In the meantime, our best understannding is that isoprenes are important for mid-canopy thermal regulation, but that other traits are more important in the most sun-exposed canopy leaves.
+
+*add something on monoterpenes?*
 
@@
 
-**Scaling in situ data with remote sensing**
+### Scaling in situ data with remote sensing
 
@@
 ## Acknowledgements
-Thanks to Norbert Kunert, Camille Piponiot, and Andrew Davies for providing figure materials. 
+Thanks to Norbert Kunert, Camille Piponiot, Peter B. Boucher and Andrew B. Davies for providing figure materials. 
 This manuscript benefited from feedback from Eleinis Ávila-Lovera, ... and the ForestGEO Ecosystems & Climate lab at SCBI.

### 2021-09-29 vertical-thermal-review.Rmd
refs for trichome and gb

@@
 Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
-`r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], but with greater leaf packing and clumping (reduces $g_b$).
+`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017; @mathurPhotosyntheticEfficiencySun2018a], but with greater leaf packing and clumping (reduces $g_b$).
 Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
 Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see also "Leaf gas exchange and its thermal sensitivity" section, @zwienieckiHydraulicLimitationsImposed2004].
-At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness (protects against water loss) and higher trichome density (reduces $g_b$; **REF?**).
-In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
+At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness (protects against water loss) and higher trichome density [reduces $g_b$ or increases boundary layer turbulence; @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016].
+In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
 In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum air temperature ($T_{air_{max}}$) and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].

### 2021-09-30 vertical-thermal-review.Rmd
added some summary sentences, worked on thermal tolerances paragraph

@@
 The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
+This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$  should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.
 
@@
 Shade leaves that do not experience midday depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021]. 
-Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.
+Evaluating this would require diurnal monitoring of net photosynthesis across a vertical profile through the canopy.
+In the meantime, it remains unresolved whether photosynthesis is more affected by high temperatures in canopy or understory leaves, and how this might vary across forest types and environmental conditions.
 
-Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020]. 
+At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020] (**KUNERT REF**).
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
-Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
-
-Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
-Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021].
+Thermal tolerances vary globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
+However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
+For example, *Quercus muehlenbergii* growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
 In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
-Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
-$T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2021].
 Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
@@
 shade leaves [@carterExperimentalWarmingTropical2021a; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
-This may vary based on forest type, acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015], or leaf traits and leaf age. 
-In evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015]. 
+This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015].
 Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021a]. 

### 2021-09-30 vertical-thermal-review.Rmd
improving ecology section

@@
 In contrast, vertical structuring of leaf isoprene emissions within forests may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
-The observation that both the abundance of emitting species (Table ) and average emission rates were highest in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021] -- as opposed to the the hotter and brighter upper canopy-- is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+The observation that both the abundance of emitting species (Table 1) and average emission rates were highest in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021] -- as opposed to the the hotter and brighter upper canopy-- is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
 Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of isoprenes in forest thermal sensitivity across the vertical gradient.
@@
 The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-Within a stand, woody growth rates, measured as radial stem growth or biomass increment, generally increase with tree size [e.g., @muller-landauTestingMetabolicEcology2006; @stephensonRateTreeCarbon2014].
-NSC reserves also increase with tree size, such that larger trees have relatively more buffer against biological and biophysical stressors [@niinemetsResponsesForestTrees2010a].
+Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russell_residence_2014].
+While woody growth rates, measured as radial stem growth, may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2021], patterns are consistent when comparing growth rates across individuals within a closed-canopy forest: growth rate consistently increases with tree size [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
+In contrast, in more open forests and for open-growth individual trees, growth rate declines with DHB [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2021], pointing to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.
 
-**Vertical gradients also affect the climate sensitivity of metabolism and growth.** 
+Vertical gradients also affect the climate sensitivity of metabolism and growth.
 Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Therefore, periods of high atmospheric demand--be these on time scales of hours, days, or seasons--tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
+Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
 This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`). 
-More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004]-- often combined with greater effective rooting depth--may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower the capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
+More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower the capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
 This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
@@
 Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable.
-From Anderson-Teixeira et al.
-(in revision).
+From @anderson-teixeiraJointEffectsClimate2021.
 In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests.

### 2021-09-30 vertical-thermal-review.Rmd
start each sentence on new line
@NidhiVinod , starting each sentence on a new line makes it easier to track changes

@@
 However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
-These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [**Fig.
-`r fig_treerings`**, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
-In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [ie.
-resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
+These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig.
+`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
+In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
 Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
@@
 
-**Canopy trees account for the majority of forest ecosystem water and carbon cycling.**
+Canopy trees account for the majority of forest ecosystem water and carbon cycling.
 While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
@@
 
-**It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.** We expect that transpiration, GPP, and $ANPP_{woody}$ should all be sensitive to high temperatures-- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020].
+It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.
+We expect that transpiration, GPP, and $ANPP_{woody}$ should all be sensitive to high temperatures-- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020]. 
 Further, increases in canopy $T_{leaf}$ are also shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 

### 2021-09-30 vertical-thermal-review.Rmd
improved ecology section
... but still want to deal with https://github.com/EcoClimLab/vertical-thermal-review/issues/70

@@
 Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
-In terms of C cycling, it has been estimated that canopy strata contribute $\geq$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
-Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of ANPPwoody and Mwoody across 25 globally distributed forests [Fig.
+In terms of C cycling, it has been estimated that canopy strata contribute $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
+Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
 `r fig_ecosystem`b, @piponiot_sizerelated_inreview].
@@
 ![**Figure `r fig_ecosystem`.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
-Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 42 tropical forests from @piponiot_sizerelated_inreview.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiot_sizerelated_inreview.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 
-It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but likely responses can be inferred based on the patterns and mechanisms reviewed above.
-We expect that transpiration, GPP, and $ANPP_{woody}$ should all be sensitive to high temperatures-- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020]. 
-Further, increases in canopy $T_{leaf}$ are also shown to reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 
-A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an El Nino drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.
+It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but probable responses can be inferred based on the patterns and mechanisms reviewed above.
+We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures in the upper canopy than in the understory -- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020].
+Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will drive the whole-ecosystem response, potentially with modest buffering by the understory. 
+Thus, for example, increases in canopy $T_{leaf}$ reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 
+A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an *El Niño* drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.
 It remains far less clear how thermal sensitivity varies across forest strata under wet conditions, but both physiological mechanisms (Fig.
@@
 
+- under moderate temperatures, the canopy is the advantageous place for trees in terms of being able to have higher photosynthesis and growth.
+However, as temperatures increase, it will be increasingly common that it is the more stressful place to be.
+
 ## Acknowledgements

### 2021-09-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-![**Figure `r fig_ecosystem`.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.
-Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiot_sizerelated_inreview.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+![**Figure `r fig_ecosystem`.
+Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiot_sizerelated_inreview.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 

### 2021-09-30 vertical-thermal-review.Rmd
new line for each sentence in global change section

@@
 
-Having established how physical conditions and biological form and function vary across vertical gradients, we now turn attention to the implications of these patterns for our understanding of how forest ecosystems may be impacted by global change, and our ability to project this across space and time.
+Having established how physical conditions and biological form and function vary across vertical gradients (Fig.
+`r schematic`), we now turn attention to the implications of these patterns for our understanding of how forest ecosystems may be impacted by global change, and our ability to project this across space and time.
 
@@
 
-The complex interwoven relations between the biophysical environment and biological factors--leaf traits, metabolic processes and distribution of trees-- are likely to produce amplifications and feedback loops in a warming world, with implications for forests on many levels.
+The complex interwoven relations between the biophysical environment and biological factors -- leaf traits, metabolic processes, and distribution of species across canopy strata -- are likely to produce amplifications and feedback loops in a warming world, with implications for forests on many levels.
 

### 2021-09-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
-Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
+Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought (Figs.
+`r schematic` - `r fig_leaf_T`), yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
+
 However, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees exist.

### 2021-09-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
@NidhiVinod , I'll be taking a 1-hr break now

--- 2021-09-30 vertical-thermal-review.Rmd
@@
 
-In turn, the climate sensitivity of woody growth tends to be greater in canopy trees. 
+In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
 In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976].
@@
 In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
-Despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], this pattern has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021].
-However, exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees exist.
+Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs.
+`r schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021], supporting the latter concept.
+The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
+
+While it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
+Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, and trees will exhibit drought responses even when even when soil moisture remains high (**REFS**).
+However, under conditions that are humid enough for canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].

### 2021-09-30 vertical-thermal-review.Rmd
edit ecology section

@@
 While it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
-Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, and trees will exhibit drought responses even when even when soil moisture remains high (**REFS**).
-However, under conditions that are humid enough for canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
+Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric], or when precipitation is statistically accounted for Fig.
+`r fig_treerings`,@anderson-teixeiraJointEffectsClimate2021].
+However, under conditions that are humid enough that canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].

### 2021-09-30 vertical-thermal-review_SI.Rmd
updating NEON methods in SI from docs

@@
 
-#  `r appendix_NEON_methods_title`
+##  `r appendix_NEON_methods_title`
 
-*be sure to cite @schimelNEONHierarchicallyDesigned2007.*
-
-METHODS ON NEON ANALYSIS GO HERE. Google doc: https://docs.google.com/document/d/1HdPa7X5AMrEwfLjazGMU92uTYMc6pObh/edit#heading=h.gjdgxs
+Google doc: https://docs.google.com/document/d/1HdPa7X5AMrEwfLjazGMU92uTYMc6pObh/edit#heading=h.gjdgxs
 
 see [Issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) , [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)
+
+# Study sites
+We analyzed vertical gradients in key biophysical variables from the National Ecological Observatory Network's Airborne Observatory Platform (NEON AOP) and meteorology towers at six focal NEON sites [Figs 2, S1, @schimelNEONHierarchicallyDesigned2007].
+The sites are well distributed across the United States and represent key forest types and structures (Table S1).
+
+# Estimating forest and light environments from lidar
+
+*Data selection*
+
+We downloaded classified lidar point clouds as .laz files from the NEON (National Ecological Observatory Network, 2020) data portal.
+For each site, we analyzed forest structure and light environments using lidar data for the 1 x 1 km plot where the meteorology tower is located.
+If a large proportion of the plot containing the tower was deforested, we used the closest forested pixel instead.
+We selected the most recent lidar survey available at each site (all conducted in 2019), which most closely temporally-corresponded to the meteorological time series.
+All sites were surveyed in growing season months (April-July), except for PUUM (January survey), an evergreen site where total leaf area seasonality is expected to be low.
+
+*Lidar data processing*
+
+Lidar data were processed according to methods detailed in @starkAmazonForestCarbon2012.
+Outliers on the z plane (height) were removed by applying an interquartile range (IQR) approach.
+The IQR is calculated as the third quartile (Q3) minus the first quartile (Q1); lidar pulse returns were restricted to > Q1-(1.5xIQR) and < Q3+(1.5xIQR).
+Next, pulse return counts were quantified, or 'voxelized' within all 2 x 2 x 1 m voxels in a plot.
+
+Digital terrain models (DTMs) were created for each plot using minimum quantile spline regression and interpolation (see Shao et al.
+2019) and used to normalize the voxelized lidar data with respect to the ground height.
+Leaf area density (LAD, $m^2$ $m^-3$) was estimated by applying the MacArthur-Horn transformation [@macarthurFoliageProfileVertical1969] to the voxelized pulse returns.
+We excluded LAD data at the bottom 5 m of the canopy, which are more uncertain due to limited pulse penetration within this region furthest away from the sensor [@starkAmazonForestCarbon2012]; for calibration purposes, we assumed the bottom 5 m to have a leaf area index (LAI) of 0.25 $m^2$ $m^{-2}$ [@shaoHighThroughputAssessment2019].
+
+We scaled LAD values by a factor of 5.5 so that they matched an independent estimate of LAI.
+Specifically, published estimates of LAI at SERC (average of 5.78 $m^2$ $m^{-2}$) derived from litterfall [@parkerCanopySurfaceStand2004], which were then in good agreement with our lidar-derived estimate (5.75 $m^2$ $m^-2$).
+Generating leaf area and light transmission profiles
+Mean LAD profiles for each site were calculated as the mean LAD at each height.
+Mean light transmission profiles (proportion of incident light) were estimated by applying a vertical light reduction model to LAD estimates [@starkAmazonForestCarbon2012].
+To estimate the proportion of sun leaves along the vertical profile, we isolated voxels in the upper canopy surface layer (1-2 m from the local canopy surface, likely to containing the majority of sun leaves), calculated the mean LAD profile for this surface layer, and divided it by the total mean LAD profile.
+
+# Analyzing biophysical gradients from micrometeorological data
+
+Micrometeorological data was downloaded for each site from NEON (neonUtilies R package) at 30 minute intervals for 2015-2020.
+To focus on the middle of the growing season, the data was constrained to be for the month of July each year.
+Per site, we calculated the mean maximum and minimum values of each variable by day and by sensor height.
+In Figure S1 we expanded on Figure XX by including 22 sites representing five forest structure types.
+
+All data processing, analyses, and graphing were conducted using R (R core team 2020).
 
@@
 
-#  `r appendix_tealeaves_methods_title`
+##  `r appendix_tealeaves_methods_title`
 

### 2021-09-30 vertical-thermal-review_SI.Rmd
just upload fig_6.jpg that Elsa sent, and suggested edits doc from Elsa into the archive to review

@@
 2019) and used to normalize the voxelized lidar data with respect to the ground height.
-Leaf area density (LAD, $m^2$ $m^-3$) was estimated by applying the MacArthur-Horn transformation [@macarthurFoliageProfileVertical1969] to the voxelized pulse returns.
+Leaf area density (LAD, $m^2$ $m^{-3}$) was estimated by applying the MacArthur-Horn transformation [@macarthurFoliageProfileVertical1969] to the voxelized pulse returns.
 We excluded LAD data at the bottom 5 m of the canopy, which are more uncertain due to limited pulse penetration within this region furthest away from the sensor [@starkAmazonForestCarbon2012]; for calibration purposes, we assumed the bottom 5 m to have a leaf area index (LAI) of 0.25 $m^2$ $m^{-2}$ [@shaoHighThroughputAssessment2019].

### 2021-09-30 vertical-thermal-review.Rmd
working on warming section

@@
 Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs.
-`r schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].
+`r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].
 Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021], supporting the latter concept.
@@
 Having established how physical conditions and biological form and function vary across vertical gradients (Fig.
-`r schematic`), we now turn attention to the implications of these patterns for our understanding of how forest ecosystems may be impacted by global change, and our ability to project this across space and time.
+`r fig_schematic`), we now turn attention to the implications of these patterns for our understanding of how forest ecosystems may be impacted by global change, and our ability to project this across space and time.
 
@@
 
-As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves.
-These changes will impact leaf metabolism and ecosystem, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
-Our in-depth review sheds some light as to how responses are likely to vary across forest vertical gradients, yet numerous questions remain.
+As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves (**CITE LATEST IPCC**).
+These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
+Our in-depth review sheds some light as to how responses are likely to vary across forest vertical gradients, yet important uncertainties remain.
 
-Leaf level $T_{opt}$ is closely related to the $T_{opt}$ of ecosystem photosynthesis ($T_{optE}$). 
-Temperature limitations to leaf processes, such as VPD and $g_s$ limitation, can modify $T_{optE}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
-At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. 
+To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, at least in part because their crowns are positioned in a more challenging microenvironment (**FIGS, REFS**).
+*Unknown: Is there a difference between soil  drought and atmospheric drought, with the most extreme case being a heat wave under well-watered conditions?*
+
+What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
+The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adapation to the local climate but do not vary consistently with height in the canopy (see section 4.2, Table 2).
 Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperate and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021].
-Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. 
-A daytime temperature threshold of 32$^\circ$C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
+While leaves display substantial plasticity to adapt to warmer temperatures (**REFS**), failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
+Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may often approach thermal tolerance thresholds, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021] (**KUNERT, in press**).
 

### 2021-09-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020] (**KUNERT REF**).
+At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReach].
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
@@
 Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
-Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may often approach thermal tolerance thresholds, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021] (**KUNERT, in press**).
+Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReach].
 
-Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$, they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
-A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
-In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
+Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
+A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018].
+However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to shed excess heat may prove disadvantageous. 
+While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), tree-ring evidence indicates that understory trees tend to exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
 
-Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
+*In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].*
 
 ### Canopy disturbance
+
+*Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].*
 

### 2021-09-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-### Canopy disturbance
+*Canopy trees seem more vulnerable to mortality from distinct disturbances, such as drought or heat waves, whereas  more buffered understory trees may be exposed to chronic stress that could negatively impact their  carbon balance (which already has less reserves; @niinemetsResponsesForestTrees2010].*
 
 *Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].*
+
+### Canopy disturbance
 

### 2021-09-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-09-30 vertical-thermal-review.Rmd
@@
 However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to shed excess heat may prove disadvantageous. 
+Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.
 While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), tree-ring evidence indicates that understory trees tend to exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
+Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the more buffered understory may be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].
 
+Thus, in synthesis, warming on its own...
 *In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].*
+
+*Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].*
 
@@
 
-*Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].*
-
 Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019], particularly in wetter environments with dense canopies [@davisMicroclimaticBufferingForests2019].
-Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015]. 
-The disproportionate impact of stress on large trees holds for other types of disturbance, including forest fragmentation and logging [@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019].
+Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015] and are at higher risk of experiencing lethal leaf $T_{leaf}$ during heat waves.
+Beyond the warming and drought, the disproportionate impact of stress on large trees holds for other types of disturbance, including forest fragmentation and logging [@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019].
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]

### 2021-09-30 vertical-thermal-review.Rmd
just upload fig_6 png

@@
 nocite: |
-  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003
+  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010
 ---
@@
 Baltzer and Thomas 2005; 29.
-Dietz et al.
-2007; 30.
 Coble et al.
+2016; 30.
+Scartazza et al.
 2016; 31.
-Scartazza et al.
-2016; 32.
-Duursma and Marshall, 2006; 33.
+Duursma and Marshall, 2006; 32.
 Harley et al.
-1996;  34.
+1996;  33.
 Hernandez et al.
-2020; 35.
+2020; 34.
 Turnbull et al.
-2003; 36.
+2003; 35.
 Chen et al.
-2020; 37.
+2020; 36.
 van de Weg et al.
-2012; 38.
+2012; 37.
 M.A Cavaleri et al.
-2008;  39.
+2008;  38.
 Koniger et al.
-1995; 40.
+1995; 39.
 Mastubara et al.
-2009; 41.
-Harris and Medina 2013; 42.
+2009; 40.
+Harris and Medina 2013; 41.
 Hansen et al.
-2001; 43.
+2001; 42.
 Poorter et al.
-1995; 44.
+1995; 43.
 Coble et al.
-2017; 45.
+2016; 44.
 Niinemets et al.
-2004; 46.
+2004; 45.
 Poorter et al.
-2000; 47.
+2000; 46.
 Zwieniecki et al.
-2004; 48.
-Sack and Scoffoni, 2013; 49.
-Ball et al., 1988
+2004; 47.
+Sack and Scoffoni, 2013; 48.
+Ball et al., 1988; 49.
+Taylor et al.
+2021; 50.
+Niinemets et al.
+2010
 ```{r eval=FALSE, echo=FALSE}

### 2021-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 While it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
-Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric], or when precipitation is statistically accounted for Fig.
+Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for Fig.
 `r fig_treerings`,@anderson-teixeiraJointEffectsClimate2021].

### 2021-10-01 vertical-thermal-review.Rmd
"finished" warming section
... including addition of some somewhat wild speculation. :-)

@NidhiVinod , see what you think.

@@
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraCarbonCyclingMature2021], supporting the latter concept.
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021], *although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [**REFS**]*.
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 
-To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, at least in part because their crowns are positioned in a more challenging microenvironment (**FIGS, REFS**).
-*Unknown: Is there a difference between soil  drought and atmospheric drought, with the most extreme case being a heat wave under well-watered conditions?*
+To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
+`r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2021].
+Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
+In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015].
+An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
+While the two are coupled over longer time scales [@humphrey_soil_2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016].
+Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.
 
@@
 
-Thus, in synthesis, warming on its own...
-*In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].*
-
-*Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].*
+Thus, in synthesis, warming temperatures will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
+We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, whereas understory trees may be increasingly prone to carbon starvation, particularly under humid conditions that make it more challenging to shed excess heat.
+Both groups -- but probably disproportionately the canopy trees -- are likely to experience increasing mortality, with the modes of mortality mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@gora_implications_2021].
+When the canopy trees die, this will open canopy gaps, as discussed in the following section.
 
@@
 Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] will also help forests cope with warming.
+
+*In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].*
+
+*Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].*
 

### 2021-10-01 vertical-thermal-review.Rmd
working on canopy disturbance

@@
 
-Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019], particularly in wetter environments with dense canopies [@davisMicroclimaticBufferingForests2019].
-Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015] and are at higher risk of experiencing lethal leaf $T_{leaf}$ during heat waves.
-Beyond the warming and drought, the disproportionate impact of stress on large trees holds for other types of disturbance, including forest fragmentation and logging [@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019].
-Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
-
-In general, disturbances (e.g., drought, fire, fragmentation, and logging) reduce canopy cover, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020].
-This shift to hotter and drier microclimates often makes forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018].
+As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
+Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @gora_implications_2021] that are expected to intensify with climate change (**IPCC REF**), and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
+Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
+Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020].
+While wetter forests can prove quite resilient to such changes [@millerReducedImpactLogging2011], this shift to hotter and drier microclimates can often make forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018].
 Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering from severe canopy disturbance [**REFS**, @tepley_vulnerability_2017; @miller_alternative_2019; @mcdowellPervasiveShiftsForest2020].
 
 Variation in resistance and resilience among forests may drive different trajectories in the face of global climate change. 
-For example, African rainforests, with canopy trees that are more resilient to warming and drought than other forests [@bennettResistanceAfricanTropical2021]. 
+For example, canopy trees of African rainforests appear to be more resilient to warming and drought than other tropical forests [@bennettResistanceAfricanTropical2021]. 
 Species with smaller leaves and larger $g_s$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020] and isoprene emitting species will have a selective advantage in a warming world [@taylorCapacityEmitIsoprene2019].
 Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] will also help forests cope with warming.
+
+Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
 

### 2021-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-While leaves display substantial plasticity to adapt to warmer temperatures (**REFS**), failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+While leaves display substantial plasticity to adapt to warmer temperatures  [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and to recover from canopy heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
 Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
@@
 Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering from severe canopy disturbance [**REFS**, @tepley_vulnerability_2017; @miller_alternative_2019; @mcdowellPervasiveShiftsForest2020].
+Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [**REFS**, @tepley_vulnerability_2017; @miller_alternative_2019; @mcdowellPervasiveShiftsForest2020].
 
@@
 Species with smaller leaves and larger $g_s$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020] and isoprene emitting species will have a selective advantage in a warming world [@taylorCapacityEmitIsoprene2019].
-Temperature acclimation [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and recovery from canopy heat stress [@smithEmpiricalEvidenceResilience2020] will also help forests cope with warming.
 
+Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
+In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
 
-*In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].*
-
-*Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].*
+*
 

### 2021-10-01 vertical-thermal-review.Rmd
"finished" canopy disturbance section

@@
 
-Variation in resistance and resilience among forests may drive different trajectories in the face of global climate change. 
-For example, canopy trees of African rainforests appear to be more resilient to warming and drought than other tropical forests [@bennettResistanceAfricanTropical2021]. 
-Species with smaller leaves and larger $g_s$ will be less vulnerable to temperature increase than species with larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020] and isoprene emitting species will have a selective advantage in a warming world [@taylorCapacityEmitIsoprene2019].
-
-Canopy disturbance along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
+Canopy disturbance poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
 In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
-
-*
-
-These shifts in forest structure at large may be detectable by lidar and thermal remote sensing.
-If we know how metabolism maps to the thermal profile, then the observed forest structural changes can inform prediction of forest function.
+If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g., **REFS**, @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy buffering and the potential state changes described above.
 
 ## Scaling across space and time
+
+As we have reviewed here,
 

### 2021-10-01 vertical-thermal-review.Rmd

@@
 
-As we have reviewed here,
+As we have reviewed here, vertical profiles in forests strongly shape forest dynamics and climate change responses.
+Ultimately, to achieve the important goal of understanding feedbacks between the world's forests and climate change, it is essential that these mechanisms be sufficiently represented in models and scaled spatially via remote sensing.
 

### 2021-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [**REFS**, @tepley_vulnerability_2017; @miller_alternative_2019; @mcdowellPervasiveShiftsForest2020].
+Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [**REFS**, @tepley_vulnerability_2017; @miller_alternative_2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].
 

### 2021-10-01 vertical-thermal-review.Rmd
new line for each sentence in modeling and scaling sections

@@
 
-Models are needed to predict future ecosystem dynamics, which remain one of the largest sources of uncertainty in Earth System Model (ESM) projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006].
+Future forest dynamics remain one of the largest sources of uncertainty in Earth System Model (ESM) projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006].
 Dynamic global vegetation models (DGVMs), which comprise the land surface models in ESMs, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].
@@
 
-**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.** Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
+**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.**
+Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
 In addition, this review finds that, although larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may also be stifled under hot, humid conditions.
-**A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.** Most models have been developed to capture dynamics in the canopy, or overstory, given the disproportionate role of these large trees in ecosystem dynamics.
+**A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.** 
+Most models have been developed to capture dynamics in the canopy, or overstory, given the disproportionate role of these large trees in ecosystem dynamics.
 Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models.

### 2021-10-01 vertical-thermal-review.Rmd
edit "Scaling across space and time " section
@NidhiVinod and @eoway , this commit shows my changes to the "Scaling across space and time" section. I changed very little content-wise, but made some edits to tie it better with the rest of the paper.

@@
 
-As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves (**CITE LATEST IPCC**).
+As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves (@ipccClimateChangeWidespread2021).
 These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
@@
 As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
-Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @gora_implications_2021] that are expected to intensify with climate change (**IPCC REF**), and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
+Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @gora_implications_2021] that are expected to intensify with climate change (@ipccClimateChangeWidespread2021), and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
@@
 
-Future forest dynamics remain one of the largest sources of uncertainty in Earth System Model (ESM) projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006].
-Dynamic global vegetation models (DGVMs), which comprise the land surface models in ESMs, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].
+Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006].
+Dynamic global vegetation models (DGVMs), which comprise the land surface models in Earth system models, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].
 DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
 This simplification is computationally more efficient, although it comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
-The computational middle-ground to representing vertical structure in DVGMs lies in cohort-based models (CBMs), which sit between the oversimplified vegetation dynamics in big-leaf models and the computational expense of individual-based models.
-CBMs represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].
+The computational middle-ground to representing vertical structure in DVGMs lies in cohort-based models, which represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].
+These sit between the oversimplified vegetation dynamics in big-leaf models, which do not represent any vertical stratification, and individual-based models, whose computational expense prevents them from being integrated into Earth system models.
 
-**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.**
+The findings of this review reinforce the notion that representing vertical structuring, as can be done using cohort-based-models, is essential to capturing forest dynamics under global change.
 Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
-In addition, this review finds that, although larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may also be stifled under hot, humid conditions.
-**A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.** 
-Most models have been developed to capture dynamics in the canopy, or overstory, given the disproportionate role of these large trees in ecosystem dynamics.
+Moreover, given the anticipated importance of mid-canopy and understory trees in ecosystem resilience to increasing mortality of canopy trees, it is absolutely essential that models separately represent these strata.
+
+A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.
+Most models have been developed to capture dynamics in the canopy, or overstory, given the disproportionate role of these large trees in ecosystem dynamics (Fig. `r fig_ecosystem`). 
 Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models. 
-This is likely due in large part to the fact that observations are only beginning to shed light on the vulnerability of understory trees to changes in temperature and other abiotic conditions [e.g., @rollinsonClimateSensitivityUnderstory2020].
+This is likely due in large part to the fact that observational and experimental studies required to resolve key patterns and underlying mechanisms remain somewhat sparse (see sections 4, 5).
+Pairing of models and observational studies will be important for further improving our mechanistic understanding of vertical gradients and their implications.
 
@@
 At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @depurySimpleScalingPhotosynthesis1997; @bonanMovingIncorrectUseful2021].
-In addition, capturing the vertical gradients in NPP, GPP, and respiration, and subsequently the net ecosystem effects, requires improved characterization of the functional response of leaf-level processes to vertically varying abiotic conditions, and the role of traits and abiotic conditions in mediating responses to thermal sensitivity.
+In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.
+`r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.
+`r `fig_leaf_T`) to vertically varying abiotic conditions (Fig.
+`fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.
+
+### Scaling in situ data with remote sensing
+
+Remote sensing data provide a valuable means to scale between *in situ* observations and DGVMs.
+Specifically, the increasing availability of airborne and spaceborne lidar and thermal remote sensing data offer a promising opportunity for mapping vertical thermal gradients in combination with vegetation structure at locations with *in situ* data and across larger landscapes (Fig.
+`r fig_model`).
+Airborne and spaceborne lidar, as well as terrestrial laser scanning data, yield detailed 3D reconstructions of whole tree and forest structure. 
+These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based Forward Looking Infrared (FLIR) cameras.
 
@@
 
-### Scaling in situ data with remote sensing
-
-Remote sensing data provide a valuable means to scale between *in situ* observations and DGVMs.
-Specifically, the increasing availability of airborne and spaceborne lidar and thermal remote sensing data offer a promising opportunity for mapping vertical thermal gradients in combination with vegetation structure at locations with in situ data and across larger landscapes (Fig.
-`r fig_model`).
-Airborne and spaceborne lidar, as well as terrestrial laser scanning data yield detailed 3D reconstructions of whole tree and forest structure.
-These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based FLIR cameras.
-
 Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
 In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
-Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020 in review].
-@pauTropicalForestTemperature2018 used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
+Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020].
+As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
 However, the authors did not consider vertical light environment conditions even though light directly influences the deviation between canopy temperature and $T_{air}$. 
-In a savanna system in the western U.S., [@johnstonWhatLiesVertical2020 in review] found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies.
-This finding was the opposite of what was expected and occurred in association with very high grass temperatures.
+In a savanna system in the western U.S., @johnstonWhatLiesVertical2020 found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding opposite of what was expected that occurred in association with very high grass temperatures.
 The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of *in situ* measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.

### 2021-10-01 vertical-thermal-review.Rmd
drafted conclusions
@NidhiVinod , see what you think!

@@
 
-- across vertical gradients, directional trends in biophysical environment and leaf traits are the rule, suggesting that the physiology and ecology that have these as their underpinnings should also vary
-- a number of studies confirm that this is the case.
-- However, when we ask the critical question of how the temperature sensitivities  of metabolism and woody growth varies across these different environments, we can't say much-- there are only a small number of studies, often showing insignificant differences or mixed results.
-- The preponderance of available data suggest that while large canopy trees are the most vulnerable to warming when water is limited, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
-- Critically, change to the vertical gradient--e.g., through increased mortality of canopy trees--will change processses across forest strata, with potentially dramatic consequences.
-(some of the literature on forest degredation or fragmentation could inform what happens when those gradients change)
+Across vertical gradients, directional trends in the biophysical environment and leaf traits are the rule, driving variation in the physiology and ecology that have these as their underpinnings (Fig.
+`r fig_schematic`).
+However, there remains a lot of uncertainty as to how the temperature sensitivities of metabolism and woody growth vary across these vertical gradients.
+The preponderance of available data suggest that while large canopy trees are the most vulnerable to warming when water is limited, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
 
-- under moderate temperatures, the canopy is the advantageous place for trees in terms of being able to have higher photosynthesis and growth.
-However, as temperatures increase, it will be increasingly common that it is the more stressful place to be.
+As the climate changes, patterns and processes across these vertical gradients are likely to shift as well.
+Under the historical temperatures to which tree species are acclimated, the canopy is an advantageous place for trees to have their crowns, as it affords higher photosynthesis and growth. 
+However, as temperatures increase, it is likely that exposed canopy positions will become physiologically stressful for increasing proportions of time.
+Ensuant increased mortality of canopy trees will create an increasing number of gaps, resulting in changes to understory conditions and community composition.
+Ultimately, mid- and understory trees communities will be critical to the resilience of forest ecosystems under changing climate, making improved understanding and model representation of their dynamics critical to understanding future forest dynamics.
+Integrating the patterns and mechanisms reviewed here, along with remote sensing data on forest structure and thermal environments, into cohort-based models that integrate with Earth system models will be critical to understanding and forecasting forest-climate feedbacks in the coming decades.
 
@@
 This manuscript benefited from feedback from Eleinis Ávila-Lovera, ... and the ForestGEO Ecosystems & Climate lab at SCBI.
-Funding was provided by the Smithsonian Institution, ....
+Funding was provided by the Smithsonian Institution.
 

### 2021-10-01 vertical-thermal-review.Rmd
revisiting abstract

@@
 
-Rising temperatures are profoundly influencing forests on many scales with potentially strong variation vertically across forest strata.
+Rising temperatures are profoundly influencing forests on many scales, with potentially strong variation vertically across forest strata. 
 Using published research and new analyses, we evaluate how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping tree ecology and ecosystem function.
@@
 Yet, differences in thermal sensitivity and damage thresholds across the gradient are modest.
-Under climate warming and increased drought, larger trees can face disproportionate stress, yet understory trees have fewer cooling mechanisms and thus may be impacted the most under hot, humid conditions, or when the buffering provided by large trees is lost.
+Under climate warming and increased drought, larger trees often face disproportionate stress, yet understory trees have fewer cooling mechanisms and thus may be impacted the most under hot, humid conditions, or when the buffering provided by large trees is lost.
+As the climate changes, integrating the patterns and mechanisms reviewed here, along withdata on forest structure and thermal environments, into models will be critical to forecasting forest-climate feedbacks.
 

### 2021-10-01 vertical-thermal-review.Rmd
reworked abstract, still too long

@@
 Rising temperatures are profoundly influencing forests on many scales, with potentially strong variation vertically across forest strata. 
-Using published research and new analyses, we evaluate how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping tree ecology and ecosystem function.
+Using published research and new analyses, we evaluate how microclimate, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping tree ecology and ecosystem function.
 In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and can experience elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
-However, foliar traits that also vary strongly across height or light gradients can partially mitigate the elevation of $T_{leaf}$ in the upper canopy.
-Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels.
-Yet, differences in thermal sensitivity and damage thresholds across the gradient are modest.
-Under climate warming and increased drought, larger trees often face disproportionate stress, yet understory trees have fewer cooling mechanisms and thus may be impacted the most under hot, humid conditions, or when the buffering provided by large trees is lost.
-As the climate changes, integrating the patterns and mechanisms reviewed here, along withdata on forest structure and thermal environments, into models will be critical to forecasting forest-climate feedbacks.
+However, foliar traits also vary strongly across height or light gradients and can partially mitigate the elevation of upper canopy $T_{leaf}$.
+Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity and damage thresholds across the gradient are modest.
+Tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$s, particularly under climate change.
+In contrast, understory trees benefit from a buffered microclimate but hae fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when the buffering provided by large trees is lost.
+As the climate changes, integrating the patterns and mechanisms reviewed here, along with data on forest structure and thermal environments, into models will be critical to forecasting forest-climate feedbacks.
 

### 2021-10-01 vertical-thermal-review.Rmd
reworked abstract
@NidhiVinod , I reworked the abstract. It's now exactly at the 200 word limit. Please see what you think.

--- 2021-10-01 vertical-thermal-review.Rmd
@@
 Total word count (excluding summary, references and legends) | 	8565 |	No. of figures | 	6 (all colour)
-Summary|  	183 | 	No. of Tables | 	2  
-Introduction|	495 |	No of Supporting Information files | 	6
+Summary|  	200 | 	No. of Tables | 	2  
+Introduction|	496 |	No of Supporting Information files | 	6
 Review of vertical gradients |	5976	||	 
@@
 
-Rising temperatures are profoundly influencing forests on many scales with potentially strong variation vertically across forest strata.
-Using published research and new analyses, we evaluate how environmental conditions, leaf temperatures, and foliar traits and metabolism vary across vertical gradients, shaping tree ecology and ecosystem function.
-In forests with dense canopies, canopy leaves are exposed to higher solar radiation and evaporative demand than understory leaves, and can experience elevated leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
-However, foliar traits that also vary strongly across height or light gradients can partially mitigate the elevation of $T_{leaf}$ in the upper canopy.
-Leaf metabolism generally increases with height across the vertical gradient, and tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels.
-Yet, differences in thermal sensitivity and damage thresholds across the gradient are modest.
-Under climate warming and increased drought, larger trees can face disproportionate stress, yet understory trees have fewer cooling mechanisms and thus may be impacted the most under hot, humid conditions, or when the buffering provided by large trees is lost.
+Rising temperatures are influencing forests on many scales, with potentially strong variation vertically across forest strata. 
+Using published research and new analyses, we evaluate how microclimate, leaf temperatures, and foliar traits and metabolism vary vertically in forests, shaping tree ecology and ecosystem function. 
+In forests with dense canopies, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
+However, foliar traits also vary across height or light gradients, partially mitigating the elevation of upper-canopy $T_{leaf}$.
+Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity and damage thresholds across the gradient appear modest.
+Scaling up, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$'s, particularly under climate change.
+In contrast, understory trees benefit from a buffered microclimate but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when the buffering provided by large trees is lost.
+As the climate changes, integrating the patterns and mechanisms reviewed here into models will be critical to forecasting forest-climate feedbacks.
 

### 2021-10-01 vertical-thermal-review_SI.Rmd
Update vertical-thermal-review_SI.Rmd
fixed headers & TOC

@@
     toc: true
-    toc_depth: 2
+    toc_depth: 1
 bibliography: Zotero_references.bib
@@
 fig_treerings= 4 
-fig_model=5
+fig_ecosystem=5
+fig_model=6
 
@@
 
-##  `r appendix_NEON_methods_title`
+#  `r appendix_NEON_methods_title`
 
@@
 
-# Study sites
+## Study sites
 We analyzed vertical gradients in key biophysical variables from the National Ecological Observatory Network's Airborne Observatory Platform (NEON AOP) and meteorology towers at six focal NEON sites [Figs 2, S1, @schimelNEONHierarchicallyDesigned2007].
@@
 
-# Estimating forest and light environments from lidar
+## Estimating forest and light environments from lidar
 
@@
 
-# Analyzing biophysical gradients from micrometeorological data
+## Analyzing biophysical gradients from micrometeorological data
 
@@
 
-##  `r appendix_tealeaves_methods_title`
+#  `r appendix_tealeaves_methods_title`
 
@@
 \newpage
-## References
+# References

### 2021-10-01 vertical-thermal-review_SI.Rmd
review/ "finalize" Methods S1

@@
 
-Google doc: https://docs.google.com/document/d/1HdPa7X5AMrEwfLjazGMU92uTYMc6pObh/edit#heading=h.gjdgxs
-
-see [Issue 2](https://github.com/EcoClimLab/vertical-thermal-review/issues/2) , [issue 20](https://github.com/EcoClimLab/vertical-thermal-review/issues/20)
-
 ## Study sites
@@
 Per site, we calculated the mean maximum and minimum values of each variable by day and by sensor height.
-In Figure S1 we expanded on Figure XX by including 22 sites representing five forest structure types.
+In Figure S1 we expanded on Figure `r fig_NEON_vertical` by including 22 sites representing five forest structure types.
 

### 2021-10-01 vertical-thermal-review.Rmd
schematic change: "leaf gas exchange"

@@
 This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
-Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
 Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

### 2021-10-01 vertical-thermal-review.Rmd
schematic change: "leaf gas exchange"

@@
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [**REF- Ty or Marielle**]
+Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012]
 The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see section 2.2).

### 2021-10-01 vertical-thermal-review.Rmd
edit abstract
@NidhiVinod , I changed metabolism to gas exchange

@@
 Rising temperatures are influencing forests on many scales, with potentially strong variation vertically across forest strata. 
-Using published research and new analyses, we evaluate how microclimate, leaf temperatures, and foliar traits and metabolism vary vertically in forests, shaping tree ecology and ecosystem function. 
+Using published research and new analyses, we evaluate how microclimate and leaf temperatures, traits, and gas exchange vary vertically in forests, shaping tree ecology and ecosystem function. 
 In forests with dense canopies, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.

### 2021-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2021-10-01 vertical-thermal-review.Rmd
@@
 `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
-`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`) [@chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [**REF- Ty or Marielle**]
+Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
 The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see section 2.2).
@@
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
-In combination with $T_{air}$, RH determines vapor pressure deficit (VPD)-- the driving force of water loss from leaves--which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+In combination with $T_{air}$, RH determines vapor pressure deficit (VPD)-- the driving force of water loss from leaves--which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
@@
 `r fig_leaf_T`).
-However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
+However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. 
 In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf gas exchange and its thermal sensitivity" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
@@
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b].
 In addition to lower wind speeds (Fig.
@@
 Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`).
-Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
+Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
 Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
@@
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
-Capacity for heat-sensitive emissions of volatile organic compounds (VOCs), including isoprenes and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; **REF ON MONOTERPENES?**].
-VOC emission capacity is a species-level trait [**REFS**], such that intrapspecific variation across across vertical gradients are not applicable, but frequency of this trait varies across the vertical gradient (see section 3.2), as do emission rates [see section 4.4, @taylorNewFieldInstrument2021].
+
+Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; **REF ON MONOTERPENES?**].
+Less than half of tree species express significant light-dependent emissions (REFS; Taylor et al., 2018), such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions (see section 3.2) and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
 

### 2021-10-01 vertical-thermal-review_SI.Rmd

@@
 
-## Identification of relevant studies
+To summarize vertical gradients in leaf traits and gas exchange (Tables 1-2), we search the literature for relevant studies and recorded results from all studies meeting our criteria for inclusion.
 
-Relevant studies were searched in the databases: ISI Web of Science, Smithsonian online library and Google Scholar, using the following key terms: (leaf traits OR foliar traits) AND (inter-canopy OR intra-canopy OR canopy height) AND (e.g.
-chlorophyll OR e.g.
-LMA OR stomatal conductance); (leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical); (leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
+The studies included in the review were based on a global geographic scope. 
+Ecosystem types included any studies with trees for mostly forests, but savannas were also considered. Herbaceous plant studies and seedling studies were excluded. 
+We targeted studies examining variation in foliar traits and metabolism across independent variables of sun/shade or height (where height is generally also sun/shade). 
+Dependent variables of interest included variables known to influence leaf temperature or metabolism: i.e., anatomical, biochemical, and structural foliar traits;  gas exchange, metabolism, and thermal sensitivity variables.
 
-in Google Scholar, 600 articles were screened, using the key terms (leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
-After screening the title and reading the abstract, 185 articles were most relevant.
-Herbaceous plant studies and seedling studies were excluded.
-These articles were added into Zotero folders: Leaf traits;  Leaf Metabolism and Processes, accordingly, for further careful reading.
+## Databases, search terms, and numbers of studies reviewed
+
+We searched for relevant studies using ISI Web of Science, Smithsonian online library, and Google Scholar, with the following key terms:
+
+- (leaf traits OR foliar traits) AND (inter-canopy OR intra-canopy OR canopy height) AND (e.g. chlorophyll OR e.g. LMA OR stomatal conductance)
+- (leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical)
+- (leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
+
+In Google Scholar, 600 articles were screened, using the key terms (leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
+After screening the title and reading the abstract, 185 articles were identified as most relevant and saved for further careful reading.
 
@@
 LMA OR stomatal conductance) provided most relevant results among the set of keywords mentioned above, with the field refine function that included botany, ecology, biology, environmental sciences, and forestry.
-Out of 150 relevant articles screened, 26 were most relevant after screening the title and reading the abstract, out of which 22 were already acquired through Google scholar search, 4 new articles were added to the folder.
+Out of 150 relevant articles screened, 26 were most relevant after screening the title and reading the abstract, out of which 22 were already acquired through Google scholar search, such that 4 new articles were added.
 
-For ISI Web of Science, search the key terms (leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical) yielded 410 relevant results, 37 were most relevant after screening the title and reading the abstract, out of which 24 were already acquired through the above process, 13 new articles were added to Zotero folders.
+For ISI Web of Science, a search of the key terms (leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical) yielded 410 relevant results.
+Of these, 37 were most relevant after screening the title and reading the abstract, out of which 24 were already acquired through the above process, such that 13 new articles were added.
 
-Through the above process, 202 articles were acquired into Zotero folders.
-The articles were tagged after careful reading of each as added to the table, irrelevant to the table, but relevant to the reviewed (with a note on the reason), irrelevant to the table and the review (with a note on the reason for exclusion).
-Articles that were not yet carefully read were tagged yet to read, which will be tagged as one of the above after reading.
-Articles shared by co-authors and references mentioned in other studies collectively were ~>32 studies.
-So far 70 relevant articles are added into the tables
+Through the above process, 202 articles were saved for careful evaluation. To this, articles shared by co-authors and references mentioned in other studies collectively added 32 studies.
 
-## Criteria for inclusion
+The articles were tagged after careful reading of each as added to the tables, irrelevant to the tables but relevant to the review (with a note on the reason), or irrelevant to the table and the review (with a note on the reason for exclusion).
 
-**this is outlined [here](https://github.com/EcoClimLab/vertical-thermal-review/blob/master/Lit_Review/Methods%20for%20Systematic%20review%20for%20the%20tables.docx). Section needs to be written.**
-
-The studies included in the review were based on a global geographic scope.
-Ecosystem types included any studies with trees for mostly forests, but savannas were also considered.
-Information for leaf traits and metabolism included studies with independent variables: sun/shade vs height (where height is generally also sun/shade), and dependent variables of interest were based on variables known to influence leaf temperature ie.
-anatomical, biochemical, structural foliar traits, including gas exchange, metabolism and thermal sensitivity variables.
+In total, following careful review, we identified **70** articles as relevant and included their results in Tables 1-2.
 

### 2021-10-01 vertical-thermal-review.Rmd
add key words
citations and ty, marielle's suggestions

@@
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
-
-Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; **REF ON MONOTERPENES?**].
-Less than half of tree species express significant light-dependent emissions (REFS; Taylor et al., 2018), such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions (see section 3.2) and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
+Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017b].
+Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018a], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions (see section 3.2) and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
 

### 2021-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 As the climate changes, integrating the patterns and mechanisms reviewed here into models will be critical to forecasting forest-climate feedbacks.
+
+**Key words**: forest; vertical gradients; microclimate; leaf temperature; leaf traits; gas exchange; ecosystem; climate change
 
@@
 Beyond the optimum, photosynthesis decreases as a result of stomatal closure [e.g.
-@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; @fredeenTemperatureHumidityEffects1999], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
+@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; @fredeenTemperatureHumidityEffects1999], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
 
@@
 Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
-@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [@campbell_introduction_1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].
+@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [@campbell_introduction_1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
 However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].

### 2021-10-01 vertical-thermal-review.Rmd
ty's ref

@@
 Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
@@
 
-Emission of the VOC isoprene gas from tree leaves is a temperature-sensitive process associated with photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008], and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
+Emission of the VOC isoprene gas from tree leaves is a temperature-sensitive process associated with photosynthetic thermal tolerance , and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
 Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
@@
 In the meantime, our best understannding is that isoprenes are important for mid-canopy thermal regulation, but that other traits are more important in the most sun-exposed canopy leaves.
+
+Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and are evidenced to play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
+In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009].
+Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both classes of VOC are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe (see Taylor et al,.
+2021 and references therein).
+Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments (Niinemets et al., 2010), and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests (Taylor et al., 2018).
+However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest (Taylor et al., 2021).
+Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface (REF).
+This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
+Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region (see section 1, Sharkey et al., 2008).
+Future work seeking to understand how temporal dynamics of leaf heating (see Leigh et al., 2017) vary through the canopy and influence leaf function will be important for resolving the role of VOCs  in forest thermal sensitivity across the vertical gradient.
+Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
 

### 2021-10-01 vertical-thermal-review.Rmd
citation update

@@
 Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and are evidenced to play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
-In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009].
+In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017b].
 Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both classes of VOC are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe (see Taylor et al,.

### 2021-10-01 vertical-thermal-review.Rmd
add author contributions
draft data availability statement
@nad

@@
 
+## Author Contributions
+NV and KAT planned and designed the research, with contributions from all authors.
+NV reviewed the literature.
+NV, IM, EMO and MNS contributed data and analyses.
+All authors contributed to writing and revising of the manuscript.
+
+## Data Availability
+No new data were created in this study.
+New analyses are based upon data available from the National Ecological Observatory Network (NEON; https://www.neonscience.org/).
+The R scripts used for new analyses presented here are available from the authors upon reasonable request.
+
 ## SI files

### 2021-10-01 vertical-thermal-review.Rmd
"final review" through line 298

@@
 Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high in the understory as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a, @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @terborghVerticalComponentPlant1985].
+`r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989].
 In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig.
-`r fig_NEON_vertical`a, @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001; @baldocchiSeasonalVariationEnergy1997; @bonan_ecological_2016].
+`r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001;  @bonan_ecological_2016].
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020a].
@@
 `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
-`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`) [@chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
 Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
-The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see section 2.2).
+The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
 
@@
 `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
-Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile (Fig.
-`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section 2.2).
+Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile.
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see sections 2.2, 4.1).
 
@@
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section 2.2).
+The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see sections 2.2, 4.1).
 
@@
 
-$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf gas exchange and its thermal sensitivity"), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
 Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
@@
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
-swr = short wave radiation (W/m^2^); ws = wind speed (m/s); rh= relative humidity; lw = leaf width (m); gs = stomatal conductance (mmol m^2^/s).
 ](leaf_energy_balance/fig3_leafenergy.png)
@@
 `r fig_leaf_T`).
-However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004]. 
-In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf gas exchange and its thermal sensitivity" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
+However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
+In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
@@
 This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig.
-`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
+`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
 The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].
@@
 This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Fig.
-`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
+`r fig_NEON_vertical`).
 
@@
 Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
-Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see also "Leaf gas exchange and its thermal sensitivity" section, @zwienieckiHydraulicLimitationsImposed2004].
-At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness (protects against water loss) and higher trichome density [reduces $g_b$ or increases boundary layer turbulence; @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016].
+Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004].
+At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [reduces $g_b$ or increases boundary layer turbulence, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016].
 In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
-In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum air temperature ($T_{air_{max}}$) and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
+In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum $T_{air}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
 
-Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`).
+Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
 Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
@@
 New analyses are based upon data available from the National Ecological Observatory Network (NEON; https://www.neonscience.org/).
-The R scripts used for new analyses presented here are available from the authors upon reasonable request.
+The R scripts used for new analyses are (*will be*) available via GitHub (https://github.com/EcoClimLab/vertical-thermal-review) and archived in Zenodo (DOI: [TBD]).
 

### 2021-10-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Emission of the VOC isoprene gas from tree leaves is a temperature-sensitive process associated with photosynthetic thermal tolerance , and within species, emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsEmissionFactorVolatile2010].
-Across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
-In contrast, vertical structuring of leaf isoprene emissions within forests may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
-The observation that both the abundance of emitting species (Table 1) and average emission rates were highest in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021] -- as opposed to the the hotter and brighter upper canopy-- is consistent with the hypothesis that temperature sensitive isoprene emissions enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
-Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of isoprenes in forest thermal sensitivity across the vertical gradient.
-In the meantime, our best understannding is that isoprenes are important for mid-canopy thermal regulation, but that other traits are more important in the most sun-exposed canopy leaves.
-
 Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and are evidenced to play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
 In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017b].
-Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both classes of VOC are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe (see Taylor et al,.
-2021 and references therein).
-Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments (Niinemets et al., 2010), and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests (Taylor et al., 2018).
-However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest (Taylor et al., 2021).
-Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface (REF).
+Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both classes of VOC are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
+Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
+However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021].
+Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [@simpragaVerticalCanopyGradient2013].
 This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
-Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region (see section 1, Sharkey et al., 2008).
-Future work seeking to understand how temporal dynamics of leaf heating (see Leigh et al., 2017) vary through the canopy and influence leaf function will be important for resolving the role of VOCs  in forest thermal sensitivity across the vertical gradient.
+Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of VOCs  in forest thermal sensitivity across the vertical gradient.
 Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.

### 2021-10-01 vertical-thermal-review.Rmd

@@
 Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
-For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [mirroring the faction of individuals, @conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018], and a recent inventory in an Amazon forest showed that both the abundance of emitting species and average emission rates were highest in the mid-canopy [see section 4.4, @taylorNewFieldInstrument2021].
+For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [mirroring the faction of individuals, @conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
 However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].

### 2021-10-01 vertical-thermal-review.Rmd
updating refs

@@
 
-*add something on monoterpenes?*
-
 ## 5. Tree and ecosystem ecology
@@
 The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russell_residence_2014].
-While woody growth rates, measured as radial stem growth, may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2021], patterns are consistent when comparing growth rates across individuals within a closed-canopy forest: growth rate consistently increases with tree size [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
-In contrast, in more open forests and for open-growth individual trees, growth rate declines with DHB [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2021], pointing to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.
+Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
+While woody growth rates, measured as radial stem growth, may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress], patterns are consistent when comparing growth rates across individuals within a closed-canopy forest: growth rate consistently increases with tree size [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
+In contrast, in more open forests and for open-growth individual trees, growth rate declines with DHB [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress], pointing to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.
 
@@
 While it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
-Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for Fig.
-`r fig_treerings`,@anderson-teixeiraJointEffectsClimate2021].
+Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.
+`r fig_treerings`, @anderson-teixeiraJointEffectsClimateinpress].
 However, under conditions that are humid enough that canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
@@
 To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
-`r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2021].
+`r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimateinpress].
 Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
@@
 An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
-While the two are coupled over longer time scales [@humphrey_soil_2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016].
+While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016].
 Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.
@@
 We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, whereas understory trees may be increasingly prone to carbon starvation, particularly under humid conditions that make it more challenging to shed excess heat.
-Both groups -- but probably disproportionately the canopy trees -- are likely to experience increasing mortality, with the modes of mortality mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@gora_implications_2021].
+Both groups -- but probably disproportionately the canopy trees -- are likely to experience increasing mortality, with the modes of mortality mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@goraImplicationsSizedependentTree2021].
 When the canopy trees die, this will open canopy gaps, as discussed in the following section.
@@
 As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
-Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @gora_implications_2021] that are expected to intensify with climate change (@ipccClimateChangeWidespread2021), and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
+Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChangeWidespread2021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].

### 2021-10-01 vertical-thermal-review.Rmd
more updates for refs

@@
 Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [**REFS**, @tepley_vulnerability_2017; @miller_alternative_2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].
+Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [**REFS**, @tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].
 
@@
 
-Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019a] and functions (e.g., thermal responses, as we present in this paper).
+Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, as we present in this paper).
 At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @depurySimpleScalingPhotosynthesis1997; @bonanMovingIncorrectUseful2021].

### 2021-10-01 vertical-thermal-review.Rmd
rearrange traits table
@NidhiVinod , please see if you like this before I make the images

--- 2021-10-01 vertical-thermal-review.Rmd
@@
 Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see section 4, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).
 

### 2021-10-01 vertical-thermal-review.Rmd
more updates for refs

--- 2021-10-01 vertical-thermal-review.Rmd
@@
 
-As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves (@ipccClimateChangeWidespread2021).
+As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves [@ipccClimateChangeWidespread2021].
 These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 

### 2021-10-01 vertical-thermal-review.Rmd
cleanup
@NidhiVinod , I deleted a few non-critical placeholders for refs (issue #73 ) . We could come back to these later

--- 2021-10-01 vertical-thermal-review.Rmd
@@
 Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017b].
-Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018a], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions (see section 3.2) and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
+Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018a], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
 
@@
 Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
-However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`).
+However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO_2 and carbon isotope discrimination discussed in section 3.1 (Table 1).
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
@@
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021], *although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [**REFS**]*.
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021].
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [**REFS**, @tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].
+Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].
 
@@
 In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
-If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g., **REFS**, @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy buffering and the potential state changes described above.
+If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g.,  @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy buffering and the potential state changes described above.
 
@@
 Thanks to Norbert Kunert, Camille Piponiot, Peter B. Boucher and Andrew B. Davies for providing figure materials. 
-This manuscript benefited from feedback from Eleinis Ávila-Lovera, ... and the ForestGEO Ecosystems & Climate lab at SCBI.
+This manuscript benefited from feedback from Eleinis Ávila-Lovera and the ForestGEO Ecosystems & Climate lab at SCBI.
 Funding was provided by the Smithsonian Institution.

### 2021-10-02 vertical-thermal-review.Rmd
changes needed to table headers/ footers

@@
 \newpage
-**Table `r table_leaf_traits`.** Summary of observed variation in leaf traits across the vertical gradient and/or between sun and shade leaves.
+**Table `r table_leaf_traits`.** Summary of observed variation in Leaf gas exchange and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
 Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
@@
 
-\newpage
-Table `r table_leaf_traits`, cont.
-
-![](table1c.png)
-
-*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, $\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
+*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, in response to height (H) or light (L).
+$\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
 
@@
 Smith and Carter, 1988; 26.
-Hadley and Smith 1987; 27.
-Panditharathna et al.
-2008; 28.
+Hadley and Smith 1987; 28.
 Baltzer and Thomas 2005; 29.
@@
 
-*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, $\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
+*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, in response to height (H) or light (L).
+$\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
 

### 2021-10-02 vertical-thermal-review.Rmd
ref updates

@@
 Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.
-Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
@@
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2021].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress].
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable.
-From @anderson-teixeiraJointEffectsClimate2021.
+From @anderson-teixeiraJointEffectsClimateinpress.
 In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests.
@@
 Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals.
-From @rollinsonClimateSensitivityUnderstory2020.
-](tree_rings/Figure_4_tree_rings.png)
+From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)
 
@@
 
-As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves [@ipccClimateChangeWidespread2021].
+As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves [@ipccClimateChange20212021].
 These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
@@
 As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
-Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChangeWidespread2021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
+Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
@@
 At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @depurySimpleScalingPhotosynthesis1997; @bonanMovingIncorrectUseful2021].
-In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.
-`r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.
-`r `fig_leaf_T`) to vertically varying abiotic conditions (Fig.
-`fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.
+In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.`r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.`r `fig_leaf_T`) to vertically varying abiotic conditions (Fig.`fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.
 

### 2021-10-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
+One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the canopy of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
+However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients.
 
@@
 Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs.
-`r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportThrough2018].
+`r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
 Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress].
@@
 Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable.
-From @anderson-teixeiraJointEffectsClimateinpress.
+From @anderson-teixeiraJointEffectsClimateinpress in press.
 In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests.

### 2021-10-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
@NidhiVinod , explanation of a few changes:
- deleted "Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021a]. " because the pattern is not strong or consistent across species, plus that would go better in the warming section.

@@
 However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
-The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
+The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
 
@@
 
-Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].
-Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs.
+Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$ or $E_0$, Table `r table_leaf_metabolism`) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].
+Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs.
 shade leaves [@carterExperimentalWarmingTropical2021a; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
-This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015].
-Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising $T_{air}$ [@carterExperimentalWarmingTropical2021a]. 
+This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021a].
 Overall, however, we currently lack synthetic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
@@
 Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [@simpragaVerticalCanopyGradient2013]. 
-This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average.
-Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. T
+emperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
 Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of VOCs  in forest thermal sensitivity across the vertical gradient.

### 2021-10-02 vertical-thermal-review.Rmd
minor clean-up

@@
 
-Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and are evidenced to play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
+Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
 In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017b].
@@
 Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [@simpragaVerticalCanopyGradient2013]. 
-This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. T
-emperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
+Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
 Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of VOCs  in forest thermal sensitivity across the vertical gradient.
@@
 Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
-While woody growth rates, measured as radial stem growth, may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress], patterns are consistent when comparing growth rates across individuals within a closed-canopy forest: growth rate consistently increases with tree size [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
-In contrast, in more open forests and for open-growth individual trees, growth rate declines with DHB [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress], pointing to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.
+Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress].
+In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
+This points to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.
 
@@
 This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`). 
-More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and lower the capacity to adjust traits related to carbon metabolism to drought [@bartholomewSmallTropicalForest2020].
+More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce adjustment of traits related to carbon metabolism during drought [@bartholomewSmallTropicalForest2020].
 This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
@@
 However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
-These have generally found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig.
-`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021], corroborating evidence from globally disurbed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
+These have most commonly found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig.
+`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimateinpress], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
 In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
@@
 Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable.
-From @anderson-teixeiraJointEffectsClimateinpress in press.
+From @anderson-teixeiraJointEffectsClimateinpress.
 In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests.

### 2021-10-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgeRelatedChanges2004].
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures in the upper canopy than in the understory -- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020].
-Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will drive the whole-ecosystem response, potentially with modest buffering by the understory. 
+Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will strongly influence the whole-ecosystem response, potentially with modest buffering by the understory. 
 Thus, for example, increases in canopy $T_{leaf}$ reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 
+Yet there is also evidence that GPP and ecosystem respiration are less sensitive to heat and drought stress in older forests than in young forests [@XuSeasonalVariabilityForest2020], perhaps in part because of more complex vertical structuring [@jonesManagingRedPine2019].
 A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an *El Niño* drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.

### 2021-10-02 vertical-thermal-review.Rmd
citations and

--- 2021-10-02 vertical-thermal-review.Rmd
@@
 Sharkey and Monson, 2014; 46.
-Saimpraga et al.
+Simpraga et al.
 2013
@@
 Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
-However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [@taylorNewFieldInstrument2021].
-Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [@simpragaVerticalCanopyGradient2013]. 
+However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_metabolism`, @taylorNewFieldInstrument2021].
+Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_metabolism`, @simpragaVerticalCanopyGradient2013].
 This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
@@
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures in the upper canopy than in the understory -- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020].
-Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will drive the whole-ecosystem response, potentially with modest buffering by the understory. 
+Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will strongly influence the whole-ecosystem response, potentially with modest buffering by the understory. 
 Thus, for example, increases in canopy $T_{leaf}$ reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 
+Yet there is also evidence that GPP and ecosystem respiration are less sensitive to heat and drought stress in older forests than in young forests [@xuSeasonalVariabilityForest2020], perhaps in part because of more complex vertical structuring [@jonesManagingRedPine2019].
 A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an *El Niño* drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.

### 2021-10-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReach].
+At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress in press].
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
@@
 Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
-Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress].
-In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
+Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress in press].
+In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress in press], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
 This points to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.
@@
 These have most commonly found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig.
-`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimateinpress], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
+`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimateinpress in press], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
 In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
@@
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress in press], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.
-`r fig_treerings`, @anderson-teixeiraJointEffectsClimateinpress].
+`r fig_treerings`, @anderson-teixeiraJointEffectsClimateinpress in press].
 However, under conditions that are humid enough that canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
@@
 Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
-`r fig_ecosystem`b, @piponiot_sizerelated_inreview].
+`r fig_ecosystem`b, @piponiotSizerelatedTreeBiomassinreview in review].
 
 ![**Figure `r fig_ecosystem`.
-Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiot_sizerelated_inreview.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotSizerelatedTreeBiomassinreview in review.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 
@@
 To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
-`r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimateinpress].
+`r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimateinpress in press].
 Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
@@
 Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
-Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReach].
+Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress in press].
 

### 2021-10-02 vertical-thermal-review_SI.Rmd
small edit in SI

@@
 
-Through the above process, 202 articles were saved for careful evaluation. To this, articles shared by co-authors and references mentioned in other studies collectively added 32 studies.
+Through the above process, 202 articles were saved for careful evaluation. To this, additional studies were 
+shared by co-authors.
 
@@
 
-In total, following careful review, we identified **70** articles as relevant and included their results in Tables 1-2.
+In total, following careful review, we identified **75** articles as relevant and included their results in Tables 1-2.
 

### 2021-10-02 vertical-thermal-review_SI.Rmd
small changes to SI

@@
 
-To summarize vertical gradients in leaf traits and gas exchange (Tables 1-2), we search the literature for relevant studies and recorded results from all studies meeting our criteria for inclusion.
+To summarize vertical gradients in leaf traits and gas exchange (Tables 1-2), we searched the literature for relevant studies and recorded results from all studies meeting our criteria for inclusion.
 
@@
 
-We searched for relevant studies using ISI Web of Science, Smithsonian online library, and Google Scholar, with the following key terms:
+We searched for relevant studies using Smithsonian online library, Google Scholar and ISI Web of Science, with the following key terms, resepecitvely:
 
@@
 
-In Google Scholar, 600 articles were screened, using the key terms (leaf temperature and metabolism OR leaf thermal sensitivity  OR leaf thermal tolerance OR leaf traits OR foliar traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy height OR canopy gradient OR canopy profile OR canopy position) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical).
-After screening the title and reading the abstract, 185 articles were identified as most relevant and saved for further careful reading.
-
-Similar search was conducted with the Smithsonian online library with the key terms (leaf traits OR foliar traits) AND (inter-canopy OR intra-canopy OR canopy height) AND (e.g.
-chlorophyll OR e.g.
-LMA OR stomatal conductance) provided most relevant results among the set of keywords mentioned above, with the field refine function that included botany, ecology, biology, environmental sciences, and forestry.
-Out of 150 relevant articles screened, 26 were most relevant after screening the title and reading the abstract, out of which 22 were already acquired through Google scholar search, such that 4 new articles were added.
-
-For ISI Web of Science, a search of the key terms (leaf* temperature* and metabolism OR leaf thermal* sensitivity  OR leaf thermal tolerance OR leaf* traits OR foliar* traits) AND (within-canopy OR intra-canopy OR sun shade OR canopy* height OR canopy gradient OR canopy profile) AND (temperate forests OR boreal forest OR conifer OR savanna OR tropical) yielded 410 relevant results.
-Of these, 37 were most relevant after screening the title and reading the abstract, out of which 24 were already acquired through the above process, such that 13 new articles were added.
-
-Through the above process, 202 articles were saved for careful evaluation. To this, additional studies were 
+Through the above process, 202 articles were saved with careful evaluation. To this, additional studies were 
 shared by co-authors.
-
-The articles were tagged after careful reading of each as added to the tables, irrelevant to the tables but relevant to the review (with a note on the reason), or irrelevant to the table and the review (with a note on the reason for exclusion).
 

### 2021-12-20 response-to-reviews.Rmd
Create response-to-reviews.Rmd

---
title: ''
output: pdf_document
bibliography: Zotero_references.bib #This can be generated by exporting BibTex file from Zotero
csl: new-phytologist.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Decision: Major revision advised

# Referee: 1

## General comment:

The article is easy to read and well structured, I appreciated reading it. It covers a broad area and explains relatively complex ideas in simple terms and efficiently. I also enjoyed the changes in scale (leaf scale to ecosystem scale). I think this paper will be very useful as a reference for the plant physiologist community and for modelers.

I have been really interested in the comparison between gradients in canopy of individual trees and canopy made of an assemblage of species (paragraph 3.1 and 3.2. However, the paragraph 3.2 is relatively short, and I am curious if it can be developed a little bit more (maybe not!).

I find it interesting to try and link the empirical and theoretical knowledge on vertical gradients in forests with what is done in DGCMs and what should be done to improve their accuracy. However, I think there is a big step between the review (part I and II) and the author’s view on the implications for models (Scaling across space and time). I wonder if it would be possible to link a little bit more the different parts, and maybe without going that much in the technical details (cohort, big leaf, multi layers, ...) which I think are sometimes imprecise or maybe wrong. Would it be possible to explain more simply what are the vertical gradients presented in the section II, that are misrepresented today in DGCMs? Maybe by explaining which of the ‘biophysical variables’ and ‘leaf traits’ that you listed, are considered constant when they should be variable vertically?

## Minor details that I hope will be of interest for the authors. Consider them as you see fit.

L30 I did not understand ‘Scaling up’ in this sentence.

Figure 1 is nice. I struggled with ‘Tleaf sensitivity to shortwave radiation’ but I think it is clear after, when reading the text. Consider using the same words in the paragraph where you detail it (line 268?) and adding the reference to the figure in the text. Or maybe change the term (‘thermal capacitance’ ?) if this is the part you are referring to in the text.

Figure 2: Homogenize the units (in parenthesis or in bracket) and add the unit for PAR.

I like your figures, but maybe consider homogenizing a little bit the styles for the plots between figures 2 3 and 4.

L205-206 Maybe add some references. How many degrees? In your Figure 3 there is not really a difference between Tleaf and Tair when SWR = 0. Given that the transpiration nearly stops at night (high humidity, low conductance) I would expect Tleaf – Tair to be very close to zero.

L219, repetition with L 216

Figure 3: TLeaf and Tleaf.

l258 under soil wet conditions?

L298-299 Far red to red light: I think this could be developed here or somewhere else. You covered the vertical variation in light quantity inside the canopy, but you could also report the vertical change in light quality (light spectrum) as the leaves absorb preferentially some wavelengths but reflects or transmit others.

L 322 consider homogenizing the photosynthetic capacity (AA) with the other notations in Table 2.

Table 1: Unit for PAR absorptance (should be % I think)

Paragraphs 3.1 and 3.2:
I thought that the comparison between intra species and interspecies gradients is very interesting. I wondered if there are more studies to expand it. One key article that I know is (Lloyd et al., 2010) which showed that within tree gradients and between tree gradients are similar for a lot of leaf traits in tropical forests. I think this is one of the reasons why DGCMs are relatively accurate even if the canopy description is quite simple. (The vertical gradients in photosynthetic parameters are scaled on Na gradients, which are the same intra and between species in this study).

L429 I am not fully sure if I understood this sentence, consider maybe developing the idea. Do you mean that the variation is higher in canopy made of multi species than on the canopy of single species? If so, I didn’t understand the comparison with understory species.

L470 A thought: Stomatal conductance is highly correlated to photosynthetic capacity (Wong et al., 1979) so the gradients in conductance are expected to follow the gradients in photosynthetic capacity. The water use efficiency (gs/A) or the slope parameter of conductance models (Medlyn et al., 2011) could change vertically inside the profiles. I don’t know if there are a lot of studies on the vertical variation of the water use efficiency or the slope parameter, but that could be something interesting to mention. See for example (Buckley, 2021).

I also think you could describe a little bit more the hydraulic constraints that increase with height. See for example (Koch et al., 2004). I think there are also other papers discussing the effect of height on water potential and stomatal conductance. You mention some aspects of it later, around L 636.

L471, I think the word conductance is missing.

L 513 Vcmax (c in subscript?)

L 609 It is a detail, but you sometimes use the word ‘canopy’ to describe the top of the vegetation, and sometimes you precise ‘top of canopy’. You also use canopy to describe all the vegetation from understory to the canopy, and you sometimes change definition from one sentence to another. Usually, it is easy to understand with the context of the sentence, but you might want to be consistent everywhere.

L 665 though

More details on my thoughts for the ‘Scaling across space and time’ paragraph:

Most DGCMs represent a vertical variability in the main physiological parameters (Vcmax, Jmax, Rdark, Na). See for example (Krinner et al., 2005; Clark et al., 2011; Oleson et al., 2013). They all use some form of an exponential decrease in the photosynthetic parameters from the canopy to the ground. Since a lot of parameters and variables depend on Vcmax (for example A and therefore gs), this allows the representation of the vertical variation of a lot of leaf traits and variables. Big leaf models have this capability (see for example Krinner et al. 2005 that you cited, Appendix, or Clark et al. 2011). I think it is not clear in the way you wrote this paragraph, and it looks like ‘big leaf’ models do not represent any vertical variation at all. Note also that it is possible to use multi-layer models with or without considering shaded and sunlit leaves (See for example Clark et al. 2011).

To me, an important remark is that most gradients in DGVMs are prescribed. More data would of course help to better understand and quantify the gradients, and to compare them with model representations (part ‘Scaling in situ data with remote sensing’ of your paper). But in the end, the mechanisms explaining those gradients are still not really known, in the sense that they can’t be efficiently modeled prognostically. This is due to the multiple sources of parameters variation that you highlighted (light environment, temperature, hydraulic, …). A lot of research is done using optimality frameworks to try and predict the photosynthetic and other vegetation traits based on environmental variables. See for example (Ali et al., 2016; Buckley, 2021). A question I have, is if including more precise cohorts as you suggest, would improve the accuracy of models given all the uncertainty in the parametrization of their traits.

Table S3: Including the words vertical or vertical gradient could have been useful.

I included below some more references, mostly in the tropics.

(Kitajima et al., 1997; Valladares et al., 1997, 2000; Thomas & Bazzaz, 1999; Carswell et al., 2000; Koch et al., 2004; Coste et al., 2005; Domingues et al., 2005; Lloyd et al., 2010; Van Goethem et al., 2014; Crous et al., 2020; Béland & Baldocchi, 2021)

Ali AA, Xu C, Rogers A, Fisher RA, Wullschleger SD, Massoud EC, Vrugt JA, Muss JD, McDowell NG, Fisher JB, et al. 2016. A global scale mechanistic model of photosynthetic capacity (LUNA V1.0). Geoscientific Model Development 9: 587–606.

Béland M, Baldocchi DD. 2021. Vertical structure heterogeneity in broadleaf forests: Effects on light interception and canopy photosynthesis. Agricultural and Forest Meteorology 307: 108525.

Buckley TN. 2021. Optimal carbon partitioning helps reconcile the apparent divergence between optimal and observed canopy profiles of photosynthetic capacity. New Phytologist 230: 2246–2260.

Carswell FE, Meir P, Wandelli EV, Bonates LCM, Kruijt B, Barbosa EM, Nobre AD, Grace J, Jarvis PG. 2000. Photosynthetic capacity in a central Amazonian rain forest. Tree Physiology 20: 179–186.

Clark DB, Mercado LM, Sitch S, Jones CD, Gedney N, Best MJ, Pryor M, Rooney GG, Essery RLH, Blyth E, et al. 2011. The Joint UK Land Environment Simulator (JULES), model description – Part 2: Carbon fluxes and vegetation dynamics. Geoscientific Model Development 4: 701–722.

Coste S, Roggy J-C, Imbert P, Born C, Bonal D, Dreyer E. 2005. Leaf photosynthetic traits of 14 tropical rain forest species in relation to leaf nitrogen concentration and shade tolerance. Tree physiology 25: 1127–1137.

Crous KY, Campany CE, Lopez Rodriguez RA, Cano FJ, Ellsworth DS. 2020. [In Press] Canopy position affects photosynthesis and anatomy in mature Eucalyptus trees in elevated CO2. Tree Physiology.

Domingues TF, Berry JA, Martinelli LA, Ometto JP, Ehleringer JR. 2005. Parameterization of canopy structure and leaf-level gas exchange for an eastern Amazonian tropical rain forest (Tapajos National Forest, Para, Brazil). Earth Interactions 9: 1–23.

Kitajima K, Mulkey SS, Wright SJ. 1997. Seasonal leaf phenotypes in the canopy of a tropical dry forest: photosynthetic characteristics and associated traits. Oecologia 109: 490–498.

Koch GW, Sillett SC, Jennings GM, Davis SD. 2004. The limits to tree height. Nature 428: 851–854.

Krinner G, Viovy N, Noblet-Ducoudré N de, Ogée J, Polcher J, Friedlingstein P, Ciais P, Sitch S, Prentice IC. 2005. A dynamic global vegetation model for studies of the coupled atmosphere-biosphere system. Global Biogeochemical Cycles 19.

Lloyd J, Patiño S, Paiva RQ, Nardoto GB, Quesada CA, Santos AJB, Baker TR, Brand WA, Hilke I, Gielmann H, et al. 2010. Optimisation of photosynthetic carbon gain and within-canopy gradients of associated foliar traits for Amazon forest trees. Biogeosciences 7: 1833–1859.

Medlyn BE, Duursma RA, Eamus D, Ellsworth DS, Prentice IC, Barton CVM, Crous KY, Angelis PD, Freeman M, Wingate L. 2011. Reconciling the optimal and empirical approaches to modelling stomatal conductance. Global Change Biology 17: 2134–2144.

Oleson K, Lawrence D, Bonan G, Drewniak B, Huang M, Koven C, Levis S, Li F, Riley W, Subin Z, et al. 2013. Technical description of version 4.5 of the Community Land Model (CLM). Boulder, Colorado, US: NCAR TECHNICAL NOTES.

Thomas SC, Bazzaz FA. 1999. Asymptotic height as a predictor of photosynthetic characteristics in Malaysian rain forest trees. Ecology 80: 1607–1622.
Valladares F, Allen MT, Pearcy RW. 1997. Photosynthetic responses to dynamic light under field conditions in six tropical rainforest shrubs occuring along a light gradient. Oecologia 111: 505–514.

Valladares F, Wright SJ, Lasso E, Kitajima K, Pearcy RW. 2000. Plastic phenotypic response to light of 16 congeneric shrubs from a Panamanian rainforest. Ecology 81: 1925–1936.

Van Goethem D, Potters G, De Smedt S, Gu L, Samson R. 2014. Seasonal, diurnal and vertical variation in photosynthetic parameters in Phyllostachys humilis bamboo plants. Photosynthesis Research 120: 331–346.

Wong SC, Cowan IR, Farquhar GD. 1979. Stomatal conductance correlates with photosynthetic capacity. Nature 282: 424–426.

# Referee: 2

A defining feature of forest canopies is their vertical variation in environment and morphology and physiology, originally described through the sun-shade leaf dichotomy or even earlier through vertical stratification diagrams as in Pearson (1971), dating further back such as the classic work of P. W. Richards (1952). Such observations were carried through to the forest meteorology community by Monteith, Jarvis and others in the 1980s (Monteith & Unsworth 1990), and there are well-known patterns in micrometeorology through canopies and first-principle understanding of their meaning for energy balance within canopies. I read this review keenly interested in what these known aspects would mean for the interactions amongst leaf traits and sensitivity to temperature, tie them back to some empirical findings of more recent vintage, and possibly to gain some insights into canopy dieback phenomena during heat events or with climate warming. In the end, the latter weren’t discussed but the authors are applauded for taking on such a huge topic. Fig. 1 is a useful entrée to the subject area.

Given the depth of what is known about within-canopy micrometeorology, I would have thought taking this on in a review wouldn’t be needed, or would be daunting and require too much detail to cover it all. As it is, I found the article launches into the empirical info too early in such an article. The article could be improved by going through the theory of how momentum and mass-transfers are attenuated through canopies and then show the empirical info from NEON in America. Overall there is too little on the theory end of things and too much 'case study' for this kind of article (see Monteith and Unsworth, and Gates to enhance the first-principle theory). As an aside, there need to be a number of improvements in Fig. 2, with lines made bolder and also height should be normalised to height relative to the top of the canopy given that trees were very different heights and meteorological masts went above the canopy to different degrees. Doing so would help make the authors’ point about differences in the Figure as well as convergence in certain properties inside canopies rather than everything being on a different y-scale.

Certainly, a discussion with a micrometeorologist or atmospheric scientist, if not already had, could improve the manuscript. Statements like l. 156: 'Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy' are an awkward read. Buffering? There could be much improvement by looking at it the other way: wind will blow as it does at the top of the canopy until encountering the plant canopy top as an aerodynamic drag element; then the additional leaf area entrains air movement and sweep-eject motion and eddies are attenuated through the canopy resulting in progressively lower windspeeds deeper into canopies. I don’t know how buffering enters into this, or what the authors mean here.

I was eager to see the part of the manuscript about photoprotection and also its relationship to heat damage. The opening sentence (l. 384) was obvious to the point of being painful to read.  Of course photoprotection is higher in the upper canopy: radiation levels are higher! Can this first sentence get revised to increase information content? This is one of the more exciting topics of the review, yet is handled only in a cursory fashion. If the authors could manage a bit more discussion of this topic it would be useful – instead they move (in the same paragraph) to VOC emissions, something I view as a separate capability of some plants. This is also confused in Fig. 1 where VAZ and VOC are placed on the same line. I’d like to see a more complete discussion of photoprotection and heat dissipation by NPQ first, and a separate paragraph about VOC. Also, there is nothing about critical temperature until late in the manuscript even though the authors skirt the issue in their text about photoprotection. This kind of thermal sensitivity is different from gas exchange (the section where the text about Tcrit occurs, l. 530-555), and should have been discussed earlier in the manuscript.

In the first part of Section 5.1, the authors conclude this paragraph stating the 'dominant role of vertical profiles in microclimate in shaping tree growth rates'. However, I believe the text has confused the very large effect of light with other aspects of forest microclimate, and they very much need to clarify this. If they believe other aspects of microclimate apart from light are drivers, then I believe this would be a minority opinion in the forest biology community. If they wish to speculate about other microclimate factors that cannot be disentangled from light, I suggest that they clearly indicate their opinion and speculation. It is perhaps an area for future research to disentangle light microclimate from aspects of vertical microclimate.
I am uncomfortable reading some of the major conclusions of the work here, such as speculation that large canopy trees are most vulnerable to warming when water is limited, but understory trees may be more vulnerable [to warming] when well-watered. This is not consistent with my knowledge of 'preponderance of available data'. Perhaps the authors implying that understory trees will succumb to warming earlier in a heat wave than large canopy trees? Is this based on the T50 evidence (l. 542-548) which is from 2 studies, or from one tree-ring study (l. 674-676), or both? An intriguing thought, but these studies and the corresponding author’s study have weaknesses and no clean experiment on this has been set up and done. I recommend that the more cautious language on l. 676-678 should in fact be repeated in the conclusions relevant to this point. It’s less stimulating and sober, but true, that it’s difficult to say so with strength of evidence and remains for further testing with good experimental design. An objective of such a review can stimulate further and better research.

## Details

There are a few dense sentences or run-ons that could be revised. Lines 192-196 and lines 201-204, for instance. Please revise.

l. 52: 'with decreases being more commonly documented across the world’s forests'

l. 128: '… but understory leaf area density is often relatively high in the understory as well' sounds redundant. Please reword.

l. 140 'Light … decreases from the canopy top to the forest floor' – why isn’t foliage clumping mentioned here amongst the other factors?

l. 148: 'Variability in the light environment decreases with height' – I don’t see this except perhaps at z = 0 and even so seems tenuous. Explain.

Fig. 1: In the diagram, stomatal density, leaf angle are separate categories yet VAZ and VOC are not. As far as I know though, there is no biochemical link between VAZ and VOC, so these should be separate shouldn't they?

Fig. 2 is a mix of computed and measured quantities. LAD and proportion of sun leaves are both computed quantities, and the basis of these should be clearer. Or at least state 'modelled LAD' and 'calculated proportion of sun leaves'.

l. 165-172: refixation of respired CO2 in the understory isn’t mentioned here, but there have been a number of studies, particularly involving 13C, that have looked at this. Please mention along with at least 1 reference.

l. 185: Buffering again. I supposed what is 'buffering' for the authors would conventionally be thought of differently by meteorologists. Please consult with one of them, and consider using different wording.

l. 187: Where mentioning radiation fluxes, could I please ask the authors to use the phrase 'and sensible heat emission'? I think this is really what the authors are referring to, so why not say so?

l. 188: 'dense canopies buffer understories' is one perspective, that there is far less radiant energy received and therefore understories stay cool is another. Can the authors please consider reflecting this perspective? I’ve studies canopies for a a score of years and I just don’t think 'buffering' is at play so much!

l. 199 should end with something like 'to shape Tleaf patterns within canopies' rather than just drop at Tleaf.

l. 206-207 'Leaves are typically warmer than air … '. That is not what is predicted in theory, proven by parts of the simulations in Fig. 3. Why not use 'Leaves can be warmer …'? That would be more consistent.

Figure 3: Variables in the inset table need to be stated. I can guess what they are, but this should be explicit.

Figure 3: Humid and drought is not a straight comparison. Well-watered vs. drought would be clearer as there are some assumptions about gs in these conditions that underlie the modelling being done here.

Figure 3: why not put a point that represents the leaf values for top and bottom of the canopy?

Table 1: What is here is not leaf gas exchange, and it seems that the authors have repeated the caption for Table 2 in error. Please fix this. Also don’t capitalise ‘leaf’.

Lines 354-372 on sun versus shade leaves and Table 1. Most of what is here was known a long time ago, say in Vogel 1968 (cited) and Boardman 1977 (not cited) but pertains to within-canopy rather than between sun and shade plants. I would suggest that this would be improved if the authors started from these old studies that most readers will recognise, and state how we’ve advanced in understanding and progressed toward within-canopy work rather than between sun-and-shade habitats.

l. 382: 'more frequent stomatal closure higher' is awkward and does not follow the comparative. Higher than what?

l. 383: d13C is NOT a concentration, it is an isotope ratio. Please revise.

l. 404-405: A cause-effect for an earlier, juvenile understory tree leaf-out is implied here, and I;m not sure that is demonstrated. There are other phenomena underlying this e.g. turgor, hydraulics, hormones, etc.

l. 475: 'because water supply often cannot meet the demands incurred …'. How do we know how frequent this is? Evidence?

l. 478: 'intracellular CO_2' needs to be fixed.

l. 479: 'In sun leaves, Tleaf thus further increases'. Further than what? Comparatives should have be clear what we’re comparing to.

l. 481 'is maximized' or is maximal? Please correct.

l. 485-487: It seems that VPD and its effect on gs is ignored here. Please clarify? The well-known stomatal closure at high VPD alone would mean higher Tair would decrease gs in the upper canopy (not 'canopy leaves' as stated) compared to the lower canopy. See Grossiord et al. (2020) Tansley Review. VPD only gets mentioned in the section on photosynthesis (l. 496). Please do so earlier.

l. 510-525: There is a long discussion about whether Topt for gas exchange or its components differs between overstory and understory. Here the authors aren’t clear, but such a comparison only matters if the same species is measured in different canopy layers.

The section on VOC emissions was interesting and informative.

l. 641: 'tends to be greater in canopy trees'. Greater than what? Please state the comparison.

l. 650: 'In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover'. Having studied large overstory trees and young understory saplings, I have a very, very difficult time with this statement especially portraying it as categorically true even if consistent with the 4 cited studies. Very few researchers have studies the same species in this context, and in this review manuscript the authors cannot scientifically lay the effect on canopy position and microclimate rather than age or other confounding factors. I must challenge their assertion and ask a higher level of evidence to make these statements, which should be qualified.

l. 946: '… how … temperature sensitivities of metabolism and woody growth vary across these vertical gradients'. I’m not sure why the authors are speculating on woody growth in this context, surely it would be extremely difficult to show that woody growth varied vertically in an individual? And that across-individual differences compound comparisons of individuals in different vertical strata? Shouldn’t these things be pointed out to readers?

\newpage
## References

### 2021-12-20 response-to-reviews.Rmd
knit, replace "l. " with "line" to avoid weird formatting

@@
 title: ''
-output: pdf_document
+output: word_document
 bibliography: Zotero_references.bib #This can be generated by exporting BibTex file from Zotero
@@
 
-l. 52: 'with decreases being more commonly documented across the world’s forests'
+line52: 'with decreases being more commonly documented across the world’s forests'
 
-l. 128: '… but understory leaf area density is often relatively high in the understory as well' sounds redundant. Please reword.
+line128: '… but understory leaf area density is often relatively high in the understory as well' sounds redundant. Please reword.
 
-l. 140 'Light … decreases from the canopy top to the forest floor' – why isn’t foliage clumping mentioned here amongst the other factors?
+line140 'Light … decreases from the canopy top to the forest floor' – why isn’t foliage clumping mentioned here amongst the other factors?
 
-l. 148: 'Variability in the light environment decreases with height' – I don’t see this except perhaps at z = 0 and even so seems tenuous. Explain.
+line148: 'Variability in the light environment decreases with height' – I don’t see this except perhaps at z = 0 and even so seems tenuous. Explain.
 
@@
 
-l.
-165-172: refixation of respired CO2 in the understory isn’t mentioned here, but there have been a number of studies, particularly involving 13C, that have looked at this.
+line165-172: refixation of respired CO2 in the understory isn’t mentioned here, but there have been a number of studies, particularly involving 13C, that have looked at this.
 Please mention along with at least 1 reference.
 
-l.
-185: Buffering again.
+line185: Buffering again.
 I supposed what is 'buffering' for the authors would conventionally be thought of differently by meteorologists.
@@
 
-l. 187: Where mentioning radiation fluxes, could I please ask the authors to use the phrase 'and sensible heat emission'? I think this is really what the authors are referring to, so why not say so?
+line187: Where mentioning radiation fluxes, could I please ask the authors to use the phrase 'and sensible heat emission'? I think this is really what the authors are referring to, so why not say so?
 
-l.
-188: 'dense canopies buffer understories' is one perspective, that there is far less radiant energy received and therefore understories stay cool is another.
+line188: 'dense canopies buffer understories' is one perspective, that there is far less radiant energy received and therefore understories stay cool is another.
 Can the authors please consider reflecting this perspective?
@@
 
-l. 199 should end with something like 'to shape Tleaf patterns within canopies' rather than just drop at Tleaf.
+line199 should end with something like 'to shape Tleaf patterns within canopies' rather than just drop at Tleaf.
 
-l.
-206-207 'Leaves are typically warmer than air … '.
+line206-207 'Leaves are typically warmer than air … '.
 That is not what is predicted in theory, proven by parts of the simulations in Fig.
@@
 
-l. 382: 'more frequent stomatal closure higher' is awkward and does not follow the comparative. Higher than what?
+line382: 'more frequent stomatal closure higher' is awkward and does not follow the comparative. Higher than what?
 
-l. 383: d13C is NOT a concentration, it is an isotope ratio. Please revise.
+line383: d13C is NOT a concentration, it is an isotope ratio. Please revise.
 
-l.
-404-405: A cause-effect for an earlier, juvenile understory tree leaf-out is implied here, and I;m not sure that is demonstrated.
+line404-405: A cause-effect for an earlier, juvenile understory tree leaf-out is implied here, and I;m not sure that is demonstrated.
 There are other phenomena underlying this e.g.
@@
 
-l. 475: 'because water supply often cannot meet the demands incurred …'. How do we know how frequent this is? Evidence?
+line475: 'because water supply often cannot meet the demands incurred …'. How do we know how frequent this is? Evidence?
 
-l. 478: 'intracellular CO_2' needs to be fixed.
+line478: 'intracellular CO_2' needs to be fixed.
 
-l. 479: 'In sun leaves, Tleaf thus further increases'. Further than what? Comparatives should have be clear what we’re comparing to.
+line479: 'In sun leaves, Tleaf thus further increases'. Further than what? Comparatives should have be clear what we’re comparing to.
 
-l. 481 'is maximized' or is maximal? Please correct.
+line481 'is maximized' or is maximal? Please correct.
 
-l.
-485-487: It seems that VPD and its effect on gs is ignored here.
+line485-487: It seems that VPD and its effect on gs is ignored here.
 Please clarify?
@@
 
-l.
-510-525: There is a long discussion about whether Topt for gas exchange or its components differs between overstory and understory.
+line510-525: There is a long discussion about whether Topt for gas exchange or its components differs between overstory and understory.
 Here the authors aren’t clear, but such a comparison only matters if the same species is measured in different canopy layers.
@@
 
-l. 641: 'tends to be greater in canopy trees'. Greater than what? Please state the comparison.
+line641: 'tends to be greater in canopy trees'. Greater than what? Please state the comparison.
 
-l.
-650: 'In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover'.
+line650: 'In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover'.
 Having studied large overstory trees and young understory saplings, I have a very, very difficult time with this statement especially portraying it as categorically true even if consistent with the 4 cited studies.
@@
 
-l.
-946: '… how … temperature sensitivities of metabolism and woody growth vary across these vertical gradients'.
+line946: '… how … temperature sensitivities of metabolism and woody growth vary across these vertical gradients'.
 I’m not sure why the authors are speculating on woody growth in this context, surely it would be extremely difficult to show that woody growth varied vertically in an individual?

### 2021-12-20 response-to-reviews.Rmd
added template for letter to editor, a few notes
@NidhiVinod , this .Rmd is ready to start filling in.

@@
 
-Decision: Major revision advised
+# (temporary section)
 
-# Referee: 1
+Nidhi, I like to use the following system for reviewer responses:
 
-## General comment:
+*italics = comments/ notes/ tentative responses that still require work*
+
+**bold= completed responses**
+
+For issues that require discussion, it's helpful to open an issue and link to it under the reviewer comment.
+
+\newpage
+
+Dear Editor:
+
+We are pleased to submit a revised version of our manuscript, "Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications" (NPH-TR-2021-38020), for consideration for publication in *New Phytologist*.
+We have addressed all points raised by the reviewers, as detailed below.
+The most significant changes include:
+
+- ...
+
+- ...
+
+Thank you for considering this revised version, and we look forward to your response.
+
+Sincerely,
+
+Nidhi Vinod & Kristina Anderson-Teixeira (on behalf of all coauthors)
+
+\newpage
+
+# Response to Reviewer(s)' Comments to Author:
+
+## Referee: 1
+
+### General comment:
 
@@
 
-## Minor details that I hope will be of interest for the authors. Consider them as you see fit.
+### Minor details that I hope will be of interest for the authors. Consider them as you see fit.
 
@@
 
-# Referee: 2
+## Referee: 2
 
@@
 
-## Details
+### Details
 
@@
 
+**Thank you.**
+
 line641: 'tends to be greater in canopy trees'. Greater than what? Please state the comparison.

### 2021-12-21 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 
+*(To ensure rapid handling of your manuscript we would be grateful if you could confirm, in the cover letter to the Editor, whether or not there are any important elements of the figures that need to be retained (eg.
+colours used, plant species drawn, layout and spacing, symbols, etc) and if you have specific instructions that you would like the illustrator to take on board.)*
+
 Thank you for considering this revised version, and we look forward to your response.
@@
 I think this paper will be very useful as a reference for the plant physiologist community and for modelers.
+
+**Thank you.**
 

### 2022-01-04 vertical-thermal-review.Rmd
Comment on Fig 3: Captalizing L in leaf
Figure 3: TLeaf and Tleaf.

@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.

### 2022-01-04 vertical-thermal-review.Rmd
l258 Comment on Tleaf section included
l258 under soil wet conditions?

@@
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
+In contrast, under hot and soil wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].

### 2022-01-05 vertical-thermal-review.Rmd
comments on L471, L513, L665
small spelling details addressed

@@
 
-Leaf hydraulic, stomatal, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
+Leaf hydraulic, stomatal conductance, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
 Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
@@
 However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of RuBP carboxylation  ($Vc_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
+Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a].
@@
 
-While it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
+Although it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
 Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.

### 2022-01-05 vertical-thermal-review.Rmd
Addressing comments on L192-196
There are a few dense sentences or run-ons that could be revised. Lines 192-196 and lines 201-204, for instance. Please revise. (included)

@@
 Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
-Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a 'canopy greenhouse effect' [e.g., OSBS, Fig.
+Notably, this pattern can be reversed in open forests.
+Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
+This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers [e.g., OSBS, Fig.
 `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @hadleyInfluenceKrummholzMat1987; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].

### 2022-01-05 vertical-thermal-review.Rmd
Addressing comments for lines 201-204
There are a few dense sentences or run-ons that could be revised. Lines 192-196 and lines 201-204, for instance. Please revise.

@@
 
-$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
-Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
+$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4).
+Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
 Leaves are typically warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].

### 2022-01-05 vertical-thermal-review.Rmd
addressing lines l. 187
l. 187: Where mentioning radiation fluxes, could I please ask the authors to use the phrase 'and sensible heat emission'? I think this is really what the authors are referring to, so why not say so?

@@
 `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
-The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation type [@defrenneForestMicroclimatesClimate2021].
+The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as local topography, vertical air mixing, cloud cover,  vegetation type and sensible heat emissions [@defrenneForestMicroclimatesClimate2021].
 Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].

### 2022-01-05 vertical-thermal-review.Rmd
comment on l.199
l. 199 should end with something like 'to shape Tleaf patterns within canopies' rather than just drop at Tleaf.

@@
 `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @hadleyInfluenceKrummholzMat1987; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
-These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$.
+These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns in canopies.
 

### 2022-01-05 vertical-thermal-review.Rmd
addressing comments on l. 206-207
l. 206-207 'Leaves are typically warmer than air … '. That is not what is predicted in theory, proven by parts of the simulations in Fig. 3. Why not use 'Leaves can be warmer …'? That would be more consistent.
table 1 comments addressed
Table 1: What is here is not leaf gas exchange, and it seems that the authors have repeated the caption for Table 2 in error. Please fix this. Also don’t capitalise ‘leaf’.

@@
 Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
-Leaves are typically warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
+Leaves can be warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
 While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020].
@@
 \newpage
-**Table `r table_leaf_traits`.** Summary of observed variation in Leaf gas exchange and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
+**Table `r table_leaf_traits`.** Summary of typically observed variation in leaf traits its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
 Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.

### 2022-01-05 vertical-thermal-review.Rmd
l. 382, l. 383
l. 382: 'more frequent stomatal closure higher' is awkward and does not follow the comparative. Higher than what?
l. 383: d13C is NOT a concentration, it is an isotope ratio. Please revise.

@@
 Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see section 4, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
 

### 2022-01-05 vertical-thermal-review.Rmd
l. 478
l. 478: 'intracellular CO_2' needs to be fixed.

@@
 Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
-However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO_2 and carbon isotope discrimination discussed in section 3.1 (Table 1).
+However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].

### 2022-01-05 vertical-thermal-review.Rmd
l. 481
l. 481 'is maximized' or is maximal? Please correct.

@@
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
+The temperature at which $g_s$ is maximal, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
 This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.

### 2022-01-05 response-to-reviews.Rmd
flag issue #90 comment
@NidhiVinod , this is an example of how I link issues to the .Rmd doc
Update response-to-reviews.Rmd
flag #91
Update response-to-reviews.Rmd
flag #92
Update response-to-reviews.Rmd
create #93

@@
 
+**Thank you for the review. We have added citations to the references mentioned here. We have also added some discussion of canopy dieback events(?)... [details].**
+
+*Nidhi, be sure to cite all the references listed here.*
+
+*It would be good to add some discussion of canopy dieback events. *[issue #93](https://github.com/EcoClimLab/vertical-thermal-review/issues/93)**
+
 Given the depth of what is known about within-canopy micrometeorology, I would have thought taking this on in a review wouldn’t be needed, or would be daunting and require too much detail to cover it all.
@@
 
+*[issue #92](https://github.com/EcoClimLab/vertical-thermal-review/issues/92)*
+
 In the first part of Section 5.1, the authors conclude this paragraph stating the 'dominant role of vertical profiles in microclimate in shaping tree growth rates'.
@@
 It is perhaps an area for future research to disentangle light microclimate from aspects of vertical microclimate.
+
+*[issue #91](https://github.com/EcoClimLab/vertical-thermal-review/issues/91)*
+
 I am uncomfortable reading some of the major conclusions of the work here, such as speculation that large canopy trees are most vulnerable to warming when water is limited, but understory trees may be more vulnerable [to warming] when well-watered.
@@
 An objective of such a review can stimulate further and better research.
+
+*[issue #90](https://github.com/EcoClimLab/vertical-thermal-review/issues/90)*
 

### 2022-01-05 response-to-reviews.Rmd
R2 major comments
# 93, # 94, #95

@@
 
-*It would be good to add some discussion of canopy dieback events. *[issue #93](https://github.com/EcoClimLab/vertical-thermal-review/issues/93)**
+*It would be good to add some discussion of canopy dieback events. ([issue #93](https://github.com/EcoClimLab/vertical-thermal-review/issues/93))*
 
@@
 
+*It would be good to bring in more theory. ([issue #94](https://github.com/EcoClimLab/vertical-thermal-review/issues/94))*
+
+*Fig 2: [issue #83](https://github.com/EcoClimLab/vertical-thermal-review/issues/83))*
+
 Certainly, a discussion with a micrometeorologist or atmospheric scientist, if not already had, could improve the manuscript.
@@
 I don’t know how buffering enters into this, or what the authors mean here.
+
+*[issue #95](https://github.com/EcoClimLab/vertical-thermal-review/issues/95)*
 

### 2022-01-05 vertical-thermal-review.Rmd
Revert "l258 Comment on Tleaf section included"
This reverts commit 05c4d244fd703e1f3de59340b4df5c84483aa8fc.

@NidhiVinod , doing this because I don't think we want to talk about soil conditions. I've always had atmospheric conditions in my mind as I've thought about this.

@@
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under hot and soil wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
+In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].

### 2022-01-07 response-to-reviews.Rmd
Update response-to-reviews.Rmd
small response updates

@@
 
+**Done**
+
 Figure 2: Homogenize the units (in parenthesis or in bracket) and add the unit for PAR.
+
+*[issue #83](https://github.com/EcoClimLab/vertical-thermal-review/issues/83)*
 
@@
 
+*Agree, but we're only changing the one variable at a time. Just need to make this clear.*
+
 L219, repetition with L 216
@@
 
+**Done**
+
 l258 under soil wet conditions?
+
+*doing this because I don't think we want to talk about soil conditions. I've always had atmospheric conditions in my mind as I've thought about this.*
 
@@
 Table 1: Unit for PAR absorptance (should be % I think)
+
+**The units in the table are in %nm i.e. % wavelength**
 
@@
 
+**Done**
+
 L 513 Vcmax (c in subscript?)
+
+**Done**
 
@@
 L 665 though
+
+**Done**
 
@@
 
+*[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
+
 To me, an important remark is that most gradients in DGVMs are prescribed.
@@
 A question I have, is if including more precise cohorts as you suggest, would improve the accuracy of models given all the uncertainty in the parametrization of their traits.
+
+*[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
 
@@
 
+*Important point that needs to be addressed.*
+
 line946: '… how … temperature sensitivities of metabolism and woody growth vary across these vertical gradients'.
@@
 
+*This requires rewording to clarify that we're not discussing intra-individual differences.*
+
 \newpage

### 2022-01-07 vertical-thermal-review.Rmd
small comment updates

@@
 It has been established that independently of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
-For example, leaf mass per area (LMA), photosynthetic capacity ($A_A$) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
 Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021a; @niinemetsResponsesForestTrees2010a].

### 2022-01-07 vertical-thermal-review.Rmd
L219, repetition with L 216 comment

@@
 Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and higher wind speeds (Fig. `r fig_leaf_T`b). 
-Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD, where $g_b$ increases with wind speed and $g_s$ can decline at high VPD due to stomatal closure.
+Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD, where although $g_b$ increases with wind speed, $g_s$ can decline at high VPD due to stomatal closure.
 Therefore, $T_{leaf}-T_{air}$ decreases with wind speed [Fig.

### 2022-01-07 vertical-thermal-review.Rmd
updates

@@
 Vertical gradients in the biophysical environment for six US forest sites in the National Ecological Observatory Network (NEON)**.
-Height profiles are shown for: growing season (a) leaf area density, (b) proportion of sun leaves, and (c) proportion light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature ($T_{air}$), and (h) maximum biological temperature, ($T_{bio}$).
+Height profiles are shown for: growing season (a) modelled leaf area density, (b) calculated proportion of sun leaves, and (c) proportion light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature ($T_{air}$), and (h) maximum biological temperature, ($T_{bio}$).
 Measurements extend from ground level (height = 0 m) to the top of the canopy (a-c, h) or above (d-g).

### 2022-01-07 response-to-reviews.Rmd
small comment updates
Update response-to-reviews.Rmd
updates
small updates

@@
 
+*Yes, night time conditions entail a shift in Tair, which in our case is constant*
+
 L219, repetition with L 216
+
+**Done**
 
@@
 L 322 consider homogenizing the photosynthetic capacity (AA) with the other notations in Table 2.
+
+**Done**
 
@@
 If so, I didn’t understand the comparison with understory species.
+
+*yes, I think this sentence is confusing now that I am reading it again, need to work on this*
 
@@
 
+**Done**
+
 line52: 'with decreases being more commonly documented across the world’s forests'
@@
 
+**Done**
+*Need to replace updated Fig 1 image in the .rmd once all changes are made*
+
 Fig.
@@
 Or at least state 'modelled LAD' and 'calculated proportion of sun leaves'.
+
+**Done**
 
@@
 
+**Done**
+
 line188: 'dense canopies buffer understories' is one perspective, that there is far less radiant energy received and therefore understories stay cool is another.
@@
 
+**Done**
+
 line206-207 'Leaves are typically warmer than air … '.
@@
 That would be more consistent.
+
+**Done**
 
@@
 
+**Done**
+*Need to add the updated table to .rmd*
+
 Lines 354-372 on sun versus shade leaves and Table 1.
@@
 line383: d13C is NOT a concentration, it is an isotope ratio. Please revise.
+
+**Done**
 
@@
 
+**Done**
+
 line479: 'In sun leaves, Tleaf thus further increases'. Further than what? Comparatives should have be clear what we’re comparing to.
@@
 line481 'is maximized' or is maximal? Please correct.
+
+**Done**
 

### 2022-01-07 vertical-thermal-review.Rmd
small updates

@@
 `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @hadleyInfluenceKrummholzMat1987; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
-These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns in canopies.
+These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.
 

### 2022-01-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
-Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
+Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$)[@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020].
 Leaves can be warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
@@
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 

### 2022-01-08 vertical-thermal-review.Rmd
@eoway please let me know if this is right?
Addressing comment:
L298-299 Far red to red light: I think this could be developed here or somewhere else. You covered the vertical variation in light quantity inside the canopy, but you could also report the vertical change in light quality (light spectrum) as the leaves absorb preferentially some wavelengths but reflects or transmit others.

@@
 
-Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig.
+Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types [Fig.
 `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
+Canopy foliage also influence the quality of light or the spectral characteristics along the gradient that shift from wavelengths in blue (400-500nm) in the top canopy layers to wavelengths in near infra red (700-1000nm) in the lower canopy layers and the forest floor.
+With red to far red ratio decreasing with height.
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.

### 2022-01-08 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
-Canopy foliage also influence the quality of light or the spectral characteristics along the gradient that shift from wavelengths in blue (400-500nm) in the top canopy layers to wavelengths in near infra red (700-1000nm) in the lower canopy layers and the forest floor.
-With red to far red ratio decreasing with height.
+Canopy foliage filter light along the gradient, therefore influencing the quality of light or the spectral characteristics.
+In the top canopy layers wavelengths range from 400-500nm in the blue range, whereas in the lower canopy layers and the forest floor, wavelengths are enriched in near infra red ranges (700-1000nm).
+With red to far red ratio decreasing with height [de Castro 2000].
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.

### 2022-01-08 response-to-reviews.Rmd
Update response-to-reviews.Rmd
Light paragraph

@@
 
+**Added the following references to L205-206:**
+**Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$)[@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020]**
+
 *Agree, but we're only changing the one variable at a time. Just need to make this clear.*
@@
 You covered the vertical variation in light quantity inside the canopy, but you could also report the vertical change in light quality (light spectrum) as the leaves absorb preferentially some wavelengths but reflects or transmit others.
+
+**Added the following lines:**
+**Canopy foliage absorb a large portion of incoming light, predominantly, and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+This entails a decrease in red to far red ratio of light as it reaches the forest floor, where understories are enriched in near infra red radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000].**
 

### 2022-01-08 vertical-thermal-review.Rmd
Light paragraph

--- 2022-01-07 vertical-thermal-review.Rmd
@@
 
-Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig.
+Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
+The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types [Fig.
 `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
+Canopy foliage absorb a large portion of PAR, and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+Along with decreasing PAR, this entails a decrease in red to far red ratio of light as it reaches the forest floor, where understories are enriched in near infra red radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000].
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
@@
 Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
-The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
+The implication of this light gradient is the variation of leaf traits along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
 

### 2022-01-08 response-to-reviews.Rmd
updates

--- 2022-01-07 response-to-reviews.Rmd
@@
 You covered the vertical variation in light quantity inside the canopy, but you could also report the vertical change in light quality (light spectrum) as the leaves absorb preferentially some wavelengths but reflects or transmit others.
+
+**Added the following lines:**
+**Canopy foliage absorb a large portion of PAR, and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+Along with decreasing PAR, this entails a decrease in red to far red ratio of light as it reaches the forest floor, where understories are enriched in near infra red radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000]**
 
@@
 
+*removed the second 'understory' word here:* 
+*'… but understory leaf area density is often relatively high in the understory as well' sounds redundant.*
+
 line140 'Light … decreases from the canopy top to the forest floor' – why isn’t foliage clumping mentioned here amongst the other factors?
+
+**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
+The profile shape of this light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types (Fig.
+2a-d, Supporting Information Figure S1, Koike et al., 2001).**
 

### 2022-01-08 vertical-thermal-review.Rmd
updates

@@
 `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
-Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high in the understory as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
+Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989].

### 2022-01-08 vertical-thermal-review.Rmd
trying to replace buffering in wind speed sentence

@@
 
-Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].
+Wind speeds are also higher at the top of the canopy, that are substantially reduced moving into lower canopy layers due to resistance caused by canopy foliage  [@juckerCanopyStructureTopography2018; @jiao-junWindProfilesTrees2004].
 This holds true across the range of forest types [Fig.

### 2022-01-11 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
"drought" to "dry" r code

@@ -147,8 +147,8 @@ sw<- data.frame(l_tla = sw_l$l_tleaf_tair, u_tla = sw_u$u_tleaf_tair, dr_ltla = 
 
 l_tla<-data.frame(par = sw$par, tleaf_tair = sw$l_tla, canopy_position = "understory humid")
 u_tla<-data.frame(par = sw$par, tleaf_tair = sw$u_tla, canopy_position = "overstory humid")
-dr_ltla<-data.frame(par = sw$par, tleaf_tair = sw$dr_ltla, canopy_position = "understory drought")
-dr_utla<-data.frame(par = sw$par, tleaf_tair = sw$dr_utla, canopy_position = "overstory drought") 
+dr_ltla<-data.frame(par = sw$par, tleaf_tair = sw$dr_ltla, canopy_position = "understory dry")
+dr_utla<-data.frame(par = sw$par, tleaf_tair = sw$dr_utla, canopy_position = "overstory dry") 
 
 sw<- data.frame(rbind(u_tla, l_tla, dr_utla, dr_ltla))
 

### 2022-01-11 response-to-reviews.Rmd
trying to replace buffering in wind speed sentence
Update response-to-reviews.Rmd

@@
 
+*Buffering was replaced with the following sentences:*
+*Wind speeds are also higher at the top of the canopy, that are substantially reduced moving into lower canopy layers due to resistance caused by canopy foliage  [@juckerCanopyStructureTopography2018; @jiao-junWindProfilesTrees2004].*
+
 line187: Where mentioning radiation fluxes, could I please ask the authors to use the phrase 'and sensible heat emission'? I think this is really what the authors are referring to, so why not say so?
@@
 drought would be clearer as there are some assumptions about gs in these conditions that underlie the modelling being done here.
+
+*This is either a semantics issue, or a case where the reviewer is expecting greater realism that what we're actually doing.
+You should make that clear in the caption (also addresses comment above), and perhaps use "humid" and "dry" (rather than drought).
+
+Also note that well-watered would refer to soil moisture, and require modeling water transport through roots and stem.
+Drought could refer to either soil dryness or atmospheric dryness.
+In my mind, it's always been the latter.*
+
+*changed to "drought" to "dry"
 

### 2022-01-11 response-to-reviews.Rmd
starting on issue #90

@@
 
+**We agree that more caution is warrented on comparing canopy vs understory responses to warming, particularly because understory responses are highly uncertain.
+We have reworded relevant text as follows:**
+
+- **"Implications: Global Change Responses : Warming" section, 4th par (previous lines 793-797: "While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), *limited tree-ring evidence* indicates that understory trees *can* exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig 4b, @rollinsonClimateSensitivityUnderstory2020]."**
+
 *[issue #90](https://github.com/EcoClimLab/vertical-thermal-review/issues/90)*

### 2022-01-11 vertical-thermal-review.Rmd
starting on issue #90

@@
 Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.
-While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), tree-ring evidence indicates that understory trees tend to exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
-Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the more buffered understory may be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].
+While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), limited tree-ring evidence indicates that understory trees can exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
+Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the more buffered understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].
 

### 2022-01-11 response-to-reviews.Rmd
more on issue #90
(interrupted-- will return)

@@
 
+- **"Implications: Global Change Responses : Warming" section, final par (previous lines 803-806: "We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, *while much larger uncertainty remains surrounding the resoponses of understory trees*."**
+
 *[issue #90](https://github.com/EcoClimLab/vertical-thermal-review/issues/90)*

### 2022-01-11 vertical-thermal-review.Rmd
more on issue #90
(interrupted-- will return)

@@
 Thus, in synthesis, warming temperatures will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
-We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, whereas understory trees may be increasingly prone to carbon starvation, particularly under humid conditions that make it more challenging to shed excess heat.
+We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, while much larger uncertainty remains surrounding the resoponses of understory trees.
 Both groups -- but probably disproportionately the canopy trees -- are likely to experience increasing mortality, with the modes of mortality mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@goraImplicationsSizedependentTree2021].

### 2022-01-12 response-to-reviews.Rmd
edit conclusions - issue #90

@@
 
+- **Conclusions section, first par (text specifically cited in the comment above): "The preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited.
+*While limited evidence suggests that understory trees may be more vulnerable to chronic warming stress under relatively mesic conditions*, more research is needed.**
+
 *[issue #90](https://github.com/EcoClimLab/vertical-thermal-review/issues/90)*

### 2022-01-12 vertical-thermal-review.Rmd
edit conclusions - issue #90

@@
 However, there remains a lot of uncertainty as to how the temperature sensitivities of metabolism and woody growth vary across these vertical gradients.
-The preponderance of available data suggest that while large canopy trees are the most vulnerable to warming when water is limited, understory trees may be more vulnerable under more mesic conditions, but more research is needed.
+The preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited.
+While limited evidence suggests that understory trees may be more vulnerable to chronic warming stress under relatively mesic conditions, more research is needed.
 

### 2022-01-12 vertical-thermal-review.Rmd
update citation
@NidhiVinod , I loaded updated citation to Zotero library

@@
 Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
-Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress in press].
-In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress in press], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
+Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2022].
+In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2022], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
 This points to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.
@@
 These have most commonly found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig.
-`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimateinpress in press], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
+`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimate2022], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
 In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
@@
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress in press], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2022], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.
-`r fig_treerings`, @anderson-teixeiraJointEffectsClimateinpress in press].
+`r fig_treerings`, @anderson-teixeiraJointEffectsClimate2022].
 However, under conditions that are humid enough that canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
@@
 Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable.
-From @anderson-teixeiraJointEffectsClimateinpress in press.
+From @anderson-teixeiraJointEffectsClimate2022.
 In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests.
@@
 To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
-`r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimateinpress in press].
+`r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
 Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.

### 2022-01-12 response-to-reviews.Rmd
addressed comment on l.650

@@
 
-*Important point that needs to be addressed.*
+**As the point about resilience was not critical, we have removed that part of the statement.
+The sentence now reads, "In addition to lower drought resistance of growth, larger trees frequently exhibit greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019]."**
 

### 2022-01-12 vertical-thermal-review.Rmd
addressed comment on l.650

@@
 `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimate2022], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
-In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
+In addition to lower drought resistance of growth, larger trees frequently exhibit greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
 Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs.

### 2022-01-12 vertical-thermal-review.Rmd
R2 line641

@@
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
-In combination with $T_{air}$, RH determines vapor pressure deficit (VPD)-- the driving force of water loss from leaves--which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
@@
 
-In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
+In turn, the drought sensitivity of woody growth tends to be greater in canopy trees than in smaller trees with less exposed crowns. 
 In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976].

### 2022-01-12 response-to-reviews.Rmd
R2 line641
respond to comments on ecology section

--- 2022-01-11 response-to-reviews.Rmd
@@
 
+- **Conclusions section, first par (text specifically cited in the comment above): "Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth.
+While the preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
+4)."**
+
 *[issue #90](https://github.com/EcoClimLab/vertical-thermal-review/issues/90)*
@@
 
+**This sentence has been revised to read, "In turn, the drought sensitivity of woody growth tends to be greater in canopy trees than in smaller trees with less exposed crowns."**
+
 line650: 'In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover'.
@@
 
-*This requires rewording to clarify that we're not discussing intra-individual differences.*
+**This has been reworded to read, "However, there remains a lot of uncertainty as to how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
+Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth.
+While the preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
+4)."**
 

### 2022-01-12 vertical-thermal-review.Rmd
respond to comments on ecology section

@@
 `r fig_schematic`).
-However, there remains a lot of uncertainty as to how the temperature sensitivities of metabolism and woody growth vary across these vertical gradients.
-The preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited.
-While limited evidence suggests that understory trees may be more vulnerable to chronic warming stress under relatively mesic conditions, more research is needed.
+However, there remains a lot of uncertainty as to how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
+Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth. 
+While the preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
+`r fig_treerings`).
 

### 2022-01-12 response-to-reviews.Rmd
fix issue #91

@@
 
+**We did not intend to argue that light was not the dominant factor, but the wording was misleading.
+This sentence has been revised to read, "This points to a dominant role of vertical profiles in the biophysical environment (Fig.
+2) in shaping tree growth rates within forests."**
 *[issue #91](https://github.com/EcoClimLab/vertical-thermal-review/issues/91)*

### 2022-01-12 response-to-reviews.Rmd
Update response-to-reviews.Rmd
#91
Update response-to-reviews.Rmd
removing flag for issue #90, as I think we're okay with that now.

@@
 2) in shaping tree growth rates within forests."**
-*[issue #91](https://github.com/EcoClimLab/vertical-thermal-review/issues/91)*
 
@@
 
-**We agree that more caution is warrented on comparing canopy vs understory responses to warming, particularly because understory responses are highly uncertain.
+**We agree that more caution is warranted on comparing canopy vs understory responses to warming, particularly because understory responses are highly uncertain.
 We have reworded relevant text as follows:**
@@
 4)."**
-
-*[issue #90](https://github.com/EcoClimLab/vertical-thermal-review/issues/90)*
 

### 2022-01-13 vertical-thermal-review.Rmd
fix issue #91
commiting to the change

@@
 Lawren Sack^2^,
+Tom Buckly^8^,
 Kristina J. Anderson-Teixeira^1,3^*
@@
 7. Department of Civil & Environmental Engineering, University of Michigan, Ann Arbor, MI 48109, USA
+
+8. Department of Plant Sciences; University of California; Davis, CA, 95616, USA
 
@@
 In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2022], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
-This points to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.
+This points to a dominant role of vertical profiles in the biophysical environment (Fig. `r fig_NEON_vertical`) in shaping tree growth rates within forests.
 

### 2022-01-14 vertical-thermal-review.Rmd
added a cool reference

@@
 These have most commonly found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig.
-`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimate2022], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
+`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimate2022; @heilmanEcologicalForecastingTree2022], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
 In addition to lower drought resistance of growth, larger trees frequently exhibit greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].

### 2022-01-14 vertical-thermal-review.Rmd
Fig 3 variables in the table stated
Figure 3: Variables in the inset table need to be stated. I can guess what they are, but this should be explicit.
red, NIR, PAR language updates

@@
 `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
-Canopy foliage absorb a large portion of PAR, and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
-Along with decreasing PAR, this entails a decrease in red to far red ratio of light as it reaches the forest floor, where understories are enriched in near infra red radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000].
+Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+Along with decreasing PAR, there is a decrease in the red to far red ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000].
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
@@
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
+Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf width (lw), stomatal conductance (gs), air temperature (tair).
 ](leaf_energy_balance/fig3_leafenergy.png)

### 2022-01-14 vertical-thermal-review.Rmd
added nm for red and far red

@@
 Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
-Along with decreasing PAR, there is a decrease in the red to far red ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000].
+Along with decreasing PAR, there is a decrease in the red (~685-690nm) to far red (~730-740nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000].
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.

### 2022-01-14 vertical-thermal-review.Rmd
removed L429 related to multi-species canopy
L470 comment

@@
 However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
-Further, the range in trait values can be several times greater in multi-species canopy than in understory species, based on their inherent species-specific traits and responses to environmental conditions.
 Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
@@
 Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
-However, because water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
+However, when water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves typically at noon, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
 In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
 The temperature at which $g_s$ is maximal, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
+
+Further, tree height also plays a role in determining the gradient of $g_s$ vertically, which has implications for $T_{leaf}$.
+With an increase in tree height, water travels a longer distance to meet the evaporative demands in sunlit canopy leaves.
+Longer water path length, higher irradiance and VPD imposes hydraulic constraints in sunlit canopy leaves [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
+Therefore, although typically $g_s$ increases with light, upper canopy leaves of taller trees have lower $g_s$ and leaf water potential to enable high leaf water-use efficiency [@kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a].
 This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.

### 2022-01-15 response-to-reviews.Rmd
Fig 3 variables in the table stated
Figure 3: Variables in the inset table need to be stated. I can guess what they are, but this should be explicit.
removed L429 related to multi-species canopy
minor edits

@@
 
+**We removed this sentence because yes, it is confusing, and the points addressed from comments above provide substantial information for inter- and intra-canopy leaf traits.**
+
 *yes, I think this sentence is confusing now that I am reading it again, need to work on this*
@@
 You mention some aspects of it later, around L 636.
+
+*need to add more water water use efficiency*
+
+**We added the following sentences related to hydraulic contraints with height:** 
+**Further, tree height also plays a role in determining the gradient of $g_s$ vertically, which has implications for $T_leaf$.
+With an increase in tree height, water travels a longer distance to meet the evaporative demands in sunlit canopy leaves.
+Longer water path length, high irradiance and VPD imposes hydraulic constraints in sunlit canopy leaves (Schäfer et al., 2000; Ambrose et al., 2010).
+Therefore, although typically $g_s$ increases with light, upper canopy leaves of taller trees have lower $g_s$ and leaf water potential to enable high leaf water-use efficiency (Rijkers et al., 2000; Kenzo et al., 2015).
+This, added to the tendency for sun leaves to have higher $T_leaf$, implies that high $T_air$ should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.**
 
@@
 
+**Done**
+
 Figure 3: Humid and drought is not a straight comparison.

### 2022-01-15 vertical-thermal-review.Rmd
minor edits

@@
 Longer water path length, higher irradiance and VPD imposes hydraulic constraints in sunlit canopy leaves [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
-Therefore, although typically $g_s$ increases with light, upper canopy leaves of taller trees have lower $g_s$ and leaf water potential to enable high leaf water-use efficiency [@kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a].
+Therefore, although typically $g_s$ increases with light, due to hydraulic constraints, upper canopy leaves of taller trees have lower $g_s$ and leaf water potential to enable high leaf water-use efficiency [@kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a].
 This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.

### 2022-01-15 response-to-reviews.Rmd
line165-172 comment on CO2 refixation

@@
 
+**Thank you for this suggestion, we added the following information:**
+*Vertical CO~2~ concentration gradients can influence gas exchange indicated by the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions.
+Particularly, understory seedlings above ground can have highest rates of CO~2~ refixation [Brooks et al.
+1997].*
+
 line185: Buffering again.
@@
 
-*changed to "drought" to "dry"
+**changed to "drought" to "dry"**
 

### 2022-01-15 vertical-thermal-review.Rmd
line165-172 comment on CO2 refixation

@@
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
+Vertical CO~2~ concentration gradients can influence gas exchange indicated by the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions [Table 1].
+Particularly, understory seedlings above ground can have highest rates of CO~2~ refixation [Brooks et al.
+1997].
 Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

### 2022-01-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2022-01-14 vertical-thermal-review.Rmd
@@
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
+Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions [Table 1].
+Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [Brooks et al.
+1997].
+However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
 
@@
 Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
-Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
-However, when water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves typically at noon, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
-In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is maximal, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
-
-Further, tree height also plays a role in determining the gradient of $g_s$ vertically, which has implications for $T_{leaf}$.
-With an increase in tree height, water travels a longer distance to meet the evaporative demands in sunlit canopy leaves.
-Longer water path length, higher irradiance and VPD imposes hydraulic constraints in sunlit canopy leaves [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010].
-Therefore, although typically $g_s$ increases with light, upper canopy leaves of taller trees have lower $g_s$ and leaf water potential to enable high leaf water-use efficiency [@kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a].
-This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.
+Maximum $g_s$ increases with light--– typically tracking photosynthetic capacity (Wong et al.
+1979)-– and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
+However, water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves with fully open stomata because of height-related constraints on water transport [ **e.g, Yoder et al.
+1994, Koch et al.
+2004, Sillet et al.
+2010**] and/or because of ncreased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations.
+These constraints tend to educe leaf water potential making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
+Stomatal depression duces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
+The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven y increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
+This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
 

### 2022-01-16 NEON_height_profiles/scripts/lad_profs_v_therm_plotting_for_ian_v1.R
Updated Figure 2 based on #83

@@ -43,7 +43,7 @@ for(i in 1:3){
   #create the plots and send to list
   plotsLAD[[i]] <- local({
     graph <- ggplot(df, aes(val, ht, color=site)) +
-      geom_path(size=0.7) + 
+      geom_path(size=1) + 
       ylim(0,80) + #80 to be standardized across all panels
       labs(x="", y = "") +
       theme_bw() +
@@ -52,16 +52,16 @@ for(i in 1:3){
     
     if(i==1){ #LAD
       graph <- graph +
-        xlab(expression(Leaf~area~density~(m^{2}~m^{-3}))) +
+        xlab(expression(Modelled~leaf~area~density~(m^{2}~m^{-3}))) +
         ylab("Height [m]")
     } else if(i==2){ # light profile
       graph <- graph +
-        xlab('Proportion incident light') +
+        xlab('Proportion of incident light') +
         theme(axis.text.y=element_blank(),
               axis.ticks.y=element_blank())
     } else if(i==3){ #sun leaves
       graph <- graph +
-        xlab("Proportion sun leaves") +
+        xlab("Calculated proportion of sun leaves") +
         theme(axis.text.y=element_blank(),
               axis.ticks.y=element_blank())
     } 

### 2022-01-17 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 *removed the second 'understory' word here:* 
-*'… but understory leaf area density is often relatively high in the understory as well' sounds redundant.*
+*"Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high as well,"*
 

### 2022-01-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Martijn Slot^3^,
-Ian McGregor^4^,
+Ian R. McGregor^4^,
 Elsa M. Ordway^2,5^,
@@
 Lawren Sack^2^,
-Tom Buckly^8^,
+Thomas N. Buckley^8^,
 Kristina J. Anderson-Teixeira^1,3^*
@@
 Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.
-Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
+Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014].
 These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
@@
 
-Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions.
+Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy.
 It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig.
@@
 Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
-Along with decreasing PAR, there is a decrease in the red (~685-690nm) to far red (~730-740nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000].
+Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
 This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
@@
 Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
-The implication of this light gradient is the variation of leaf traits along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
+The implications of this light gradient are that leaf traits vary along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
 
-Wind speeds are also higher at the top of the canopy, that are substantially reduced moving into lower canopy layers due to resistance caused by canopy foliage  [@juckerCanopyStructureTopography2018; @jiao-junWindProfilesTrees2004].
+Despite these complexities of within-canopy transport, wind speeds are generally much higher at the top of the canopy than within or beneath [@juckerCanopyStructureTopography2018; @jiao-junWindProfilesTrees2004].
 This holds true across the range of forest types [Fig.
@@
 Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile.
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see sections 2.2, 4.1).
+The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat (see sections 2.2, 4.1).
 
@@
 Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions [Table 1].
-Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [Brooks et al.
-1997].
+Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997].
 However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
@@
 
-Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Fig.
+Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances, attenuation of radiation and vertical transport of sensible heat by the canopy can buffer the lower canopy and understory from large diel swings in air temperature  (Fig.
 `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
 The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as local topography, vertical air mixing, cloud cover,  vegetation type and sensible heat emissions [@defrenneForestMicroclimatesClimate2021].
-Typically, dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting in maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
 Notably, this pattern can be reversed in open forests.
@@
 Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
-Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$)[@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020].
+Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$) [@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020].
 Leaves can be warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].

### 2022-01-17 vertical-thermal-review.Rmd
Adding some of Tom's suggestions

@@
 High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a). 
-Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and higher wind speeds (Fig. `r fig_leaf_T`b). 
-Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD, where although $g_b$ increases with wind speed, $g_s$ can decline at high VPD due to stomatal closure.
-Therefore, $T_{leaf}-T_{air}$ decreases with wind speed [Fig.
+Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and under higher wind speeds (Fig. `r fig_leaf_T`b). 
+Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD, where although $g_b$ increases with wind speed, $g_s$ generally declines as VPD increases due to stomatal closure [Darwin 1898, Mott and Parkhurst 1991].
+Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig.
 `r fig_leaf_T`b, @daudetWindSpeedLeaf1999a], increases with RH (Fig.
 `r fig_leaf_T`c), increases with leaf size (Fig.
-`r fig_leaf_T`d), and decreases with $g_s$ (Fig.
+`r fig_leaf_T`d), and decreases absolutely with $g_s$ (Fig.
 `r fig_leaf_T`e).
@@
 With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) and larger $g_s$ [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
-However, when leaf water demand (influenced by VPD and stomatal opening) exceeds the rate of supply, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig.
+However, when leaf water demand (determined by VPD and stomatal opening) exceeds the rate of supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`e, @fauset_differences_2018].
@@
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b].
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to shedding excess heat compared to upper-canopy leaves, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b].
 In addition to lower wind speeds (Fig.
@@
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do buffered lower-canopies, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 

### 2022-01-17 response-to-reviews.Rmd
fixing citations
connected by zotero from the new computer to github zotero_refrences.bib, so some citations are different

@@
 
-*Buffering was replaced with the following sentences:*
-*Wind speeds are also higher at the top of the canopy, that are substantially reduced moving into lower canopy layers due to resistance caused by canopy foliage  [@juckerCanopyStructureTopography2018; @jiao-junWindProfilesTrees2004].*
+**This passage now occurs after the new paragraph about turbulent transport within canopies.
+We have reworded the passage as, “Despite the complexities of turbulent transport in canopies, wind speeds are generally much higher at the top of the canopy (Jiao-jun et al., 2004; Jucker et al., 2018).”**
 
@@
 I’ve studies canopies for a a score of years and I just don’t think 'buffering' is at play so much!
+
+**We were using the term “buffer” in a functional or biological sense, meaning “protect,” rather than as a description of the underlying physics; i.e., the canopy protects the understory from large diel swings in both air temperature and radiation.
+We have reworded this passage to clarify: “Typically, diel temperature range is smaller beneath dense canopies than above, resulting in,  ~~dense canopies buffer understories from high maximum Tair more than open canopies, i.e., dense canopy understories can have~~ cooler maximum daytime Tair and warmer nighttime minimum Tair than open canopy understories or nearby clearings.”**
 

### 2022-01-17 vertical-thermal-review.Rmd
fixing citations
connected by zotero from the new computer to github zotero_refrences.bib, so some citations are different

@@
 Forests are vertically and horizontally stratified, with overstory canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
-This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
+This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
 Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
@@
 In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig.
-`r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020a; @lawLeafAreaDistribution2001;  @bonan_ecological_2016].
-Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020a].
+`r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;  @bonan_ecological_2016].
+Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020].
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019].
@@
 This holds true across the range of forest types [Fig.
-`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
+`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
 Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile.
@@
 `r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
-Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a].
+Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020].
 In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
@@
 $T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4).
-Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015a; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
 Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$) [@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020].
@@
 Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig.
-`r fig_leaf_T`b, @daudetWindSpeedLeaf1999a], increases with RH (Fig.
+`r fig_leaf_T`b, @daudetWindSpeedLeaf1999], increases with RH (Fig.
 `r fig_leaf_T`c), increases with leaf size (Fig.

### 2022-01-18 vertical-thermal-review.Rmd
update

@@
 nocite: |
-  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010
+  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010
 ---
@@
 Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
-With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) and larger $g_s$ [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) and larger $g_s$ [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
 However, when leaf water demand (determined by VPD and stomatal opening) exceeds the rate of supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, at high solar radiation loads, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at maximum $g_s$, but during $g_s$ limitation, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
+Therefore, at high solar radiation loads, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at maximum $g_s$, but during $g_s$ limitation, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].
 
@@
 In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig.
-`r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].
+`r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
 
@@
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to shedding excess heat compared to upper-canopy leaves, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b].
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to shedding excess heat compared to upper-canopy leaves, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
 In addition to lower wind speeds (Fig.
@@
 `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig.
-`r fig_leaf_T`c, @tibbittsHumidityPlants1979b; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021; @dietzVerticalPatternsDuration2007].
+`r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].
 
@@
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021a; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
@@
 `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
-The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013b].
+The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013].
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
@@
 
-Leaf traits shape $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015;@michaletzEnergeticCarbonEconomic2016;  @zwienieckiHydraulicLimitationsImposed2004].
+Leaf traits shape $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
@@
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
-Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021a; @niinemetsResponsesForestTrees2010a].
+Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
 
@@
 Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004].
-At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [reduces $g_b$ or increases boundary layer turbulence, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016].
-In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b].
+At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [reduces $g_b$ or increases boundary layer turbulence, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
+In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013].
 In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum $T_{air}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
@@
 Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
-Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
+Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
 Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
@@
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
-Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017b].
-Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018a], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
+Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
+Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
 
@@
 Stomatal depression duces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven y increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].
+The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven y increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
 This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
@@
 Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
-In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a].
+In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
 @hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 
@@
 Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs.
-shade leaves [@carterExperimentalWarmingTropical2021a; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
-This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021a].
+shade leaves [@carterExperimentalWarmingTropical2021; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
+This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
 Overall, however, we currently lack synthetic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
@@
 Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
-In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017b].
+In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
 Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both classes of VOC are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
-Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018a].
+Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
 However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_metabolism`, @taylorNewFieldInstrument2021].
@@
 Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
-A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018].
+A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
 However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to shed excess heat may prove disadvantageous. 

### 2022-01-18 vertical-thermal-review.Rmd
updating citations

@@
 
-Canopy disturbance poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
+Canopy disturbance poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013a; @greiserHidingClimateCharacterizing2019].
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
-In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
+In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020b; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
 If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g.,  @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy buffering and the potential state changes described above.

### 2022-01-18 vertical-thermal-review.Rmd
citation update

@@
 1979)-– and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
-However, water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves with fully open stomata because of height-related constraints on water transport [ **e.g, Yoder et al.
-1994, Koch et al.
-2004, Sillet et al.
-2010**] and/or because of ncreased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations.
+However, water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves with fully open stomata because of height-related constraints on water transport [ **e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010**] and/or because of ncreased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations.
 These constraints tend to educe leaf water potential making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).

### 2022-01-18 response-to-reviews.Rmd
adding responses to reviewer's comments

@@
 **Added the following lines:**
-**Canopy foliage absorb a large portion of PAR, and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
-Along with decreasing PAR, this entails a decrease in red to far red ratio of light as it reaches the forest floor, where understories are enriched in near infra red radiation and absorb light in wavelengths of 700-1000nm [de Castro 2000; @poorterLeafOpticalProperties2000]**
+**Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]**
 
@@
 **We removed this sentence because yes, it is confusing, and the points addressed from comments above provide substantial information for inter- and intra-canopy leaf traits.**
-
-*yes, I think this sentence is confusing now that I am reading it again, need to work on this*
 
@@
 
-**Done**
+**We revised the following lines:**
+
+- **Lines 192-196: Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a ‘canopy greenhouse effect’**
+
+- **Notably, this pattern can be reversed in open forests.
+Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
+This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers**
+
+- **Lines 201-204: $T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes (Drake et al., 2020; Perez & Feeley, 2020; Michaletz et al., 2015)**
+
+- **$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4).
+Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes **
 
@@
 
-*removed the second 'understory' word here:* 
-*"Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high as well,"*
+**We removed the second 'understory' word here:** 
+**Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high ~~in the understory~~ as well**
 
@@
 
-**Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
+**We added foliage clumping into the list of factors influencing light gradient here:**
+**"Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
 The profile shape of this light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types (Fig.
-2a-d, Supporting Information Figure S1, Koike et al., 2001).**
+2a-d, Supporting Information Figure S1, Koike et al., 2001)."**
 
@@
 
-**Done**
-*Need to replace updated Fig 1 image in the .rmd once all changes are made*
+**We agree to the suggestions. Therefore, to make it more clearer, we added photoprotection as one of the factors in Fig.1 instead of VAZ and VOC in the same line.**
 
@@
 **Thank you for this suggestion, we added the following information:**
-*Vertical CO~2~ concentration gradients can influence gas exchange indicated by the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions.
+*"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions.
 Particularly, understory seedlings above ground can have highest rates of CO~2~ refixation [Brooks et al.
-1997].*
+1997]."*
 
@@
 
-**Done**
+**Variables in the inset table are now described in the figure caption as: "Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf width (lw), stomatal conductance (gs), air temperature (tair)."**
 
@@
 *This is either a semantics issue, or a case where the reviewer is expecting greater realism that what we're actually doing.
-You should make that clear in the caption (also addresses comment above), and perhaps use "humid" and "dry" (rather than drought).
+You should make that clear in the caption (also addresses comment above), and perhaps use "humid" and "dry" (rather than drought).*
 
-Also note that well-watered would refer to soil moisture, and require modeling water transport through roots and stem.
+*Also note that well-watered would refer to soil moisture, and require modeling water transport through roots and stem.
 Drought could refer to either soil dryness or atmospheric dryness.
@@
 
-**changed to "drought" to "dry"**
+**changed to "drought" to "dry" in the figure to refer to atmospheric dryness.**
 

### 2022-01-18 vertical-thermal-review.Rmd
Adding Tom's micromet section
still working on citations here

@@
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+
+Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles.
+Vertical transport at the macroscopic scale of a plant canopy is dominated by turbulent transport rather than by molecular diffusion.
+Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients (eddy diffusivities, often symbolized K) that depend on both friction and buoyancy [@penmanWeatherWheatEssay1960].
+In this “K-theory” regime, wind speed declines logarithmically with proximity to the ground or canopy surface [@monteithPrinciplesEnvironmentalPhysics2013].
+Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
+However, K-theory often fails within canopies, making canopy wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; **Raupach 1986**; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux (e.g., Raupach 1987).
+Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above (Finnigan 1979, Baldocchi and Meyers 1991).
+These "sweep-eject" events can couple understory conditions more directly to conditions above the canopy than beneath.
+
 Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].

### 2022-01-18 vertical-thermal-review.Rmd
citations

@@
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
+Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
+The implications of this light gradient are that leaf traits vary along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
 
@@
 Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-However, K-theory often fails within canopies, making canopy wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; **Raupach 1986**; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
-For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux (e.g., Raupach 1987).
-Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above (Finnigan 1979, Baldocchi and Meyers 1991).
+However, K-theory often fails within canopies, making canopy wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @finniganTurbulencePlantCanopies; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux [e.g., @raupachLagrangianAnalysisScalar1987].
+Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991].
 These "sweep-eject" events can couple understory conditions more directly to conditions above the canopy than beneath.
-
-Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
-The implications of this light gradient are that leaf traits vary along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
 

### 2022-01-18 vertical-thermal-review.Rmd
citations

@@
 Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-However, K-theory often fails within canopies, making canopy wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @finniganTurbulencePlantCanopies; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+However, K-theory often fails within canopies, making canopy wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
 For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux [e.g., @raupachLagrangianAnalysisScalar1987].
@@
 Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and under higher wind speeds (Fig. `r fig_leaf_T`b). 
-Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD, where although $g_b$ increases with wind speed, $g_s$ generally declines as VPD increases due to stomatal closure [Darwin 1898, Mott and Parkhurst 1991].
+Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD, where although $g_b$ increases with wind speed, $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
 Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig.

### 2022-01-18 vertical-thermal-review.Rmd
Tom's edits

@@
 \newpage
-**Table `r table_leaf_traits`.** Summary of typically observed variation in leaf traits its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
+**Table `r table_leaf_traits`.** Summary of typically observed variation in leaf traits and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
 Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
@@
 Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`).
-Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
-`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017; @mathurPhotosyntheticEfficiencySun2018a], but with greater leaf packing and clumping (reduces $g_b$).
-Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
+Thus, sun leaves are generally smaller and thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
+`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017; @mathurPhotosyntheticEfficiencySun2018a], but with greater leaf packing and clumping (which reduces $g_b$).
+Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
 Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004].
@@
 
-Leaf hydraulic, stomatal conductance, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
-Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
-Maximum $g_s$ increases with light--– typically tracking photosynthetic capacity (Wong et al.
-1979)-– and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
-However, water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves with fully open stomata because of height-related constraints on water transport [ **e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010**] and/or because of ncreased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations.
-These constraints tend to educe leaf water potential making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
-Stomatal depression duces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven y increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
+Leaf hydraulic, stomatal and boundary layer conductances are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
+Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
+Maximum $g_s$ increases with light--typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979]--and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
+However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata because, of height-related constraints on water transport [ **e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010**] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations (Buckley et al.
+2014).
+These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
+Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
+The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
 This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
@@
 
-Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
+Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both predicted by optamization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
 This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under conditions of high light and sufficient water (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).

### 2022-01-18 response-to-reviews.Rmd
added Tom's response to reviwer's comments and removed thermal capacitance in the sentence

@@
 
-**Added the following references to L205-206:**
+**Tleaf can be well below Tair on clear nights despite low transpiration, due to radiative coupling with the very cold sky. Added the following references to L205-206:**
 **Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$)[@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020]**
@@
 **Added the following lines:**
-**Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+**Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
 Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]**
@@
 *[issue #92](https://github.com/EcoClimLab/vertical-thermal-review/issues/92)*
+**It is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.**
 
@@
 **We did not intend to argue that light was not the dominant factor, but the wording was misleading.
-This sentence has been revised to read, "This points to a dominant role of vertical profiles in the biophysical environment (Fig.
+This sentence has been revised to read, "This points to a dominant role of vertical profiles in the biophysical environment, particularly light (Fig.
 2) in shaping tree growth rates within forests."**
@@
 
-**We agree to the suggestions. Therefore, to make it more clearer, we added photoprotection as one of the factors in Fig.1 instead of VAZ and VOC in the same line.**
+**We agree with the suggestions. Therefore, to make it more clearer, we added photoprotection as one of the factors in Fig.1 instead of VAZ and VOC in the same line.**
 
@@
 
+**We have added four references showing that stomatal conductance is often limited by imbalance of water supply and evaporative demand for leaves in more elevated or more sunlit canopy positions [@yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010].**
+
 line478: 'intracellular CO_2' needs to be fixed.
@@
 
+**This sentence has been rewritten as, “Stomatal closure reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads.”**
+
 line481 'is maximized' or is maximal? Please correct.
 
-**Done**
+**We have replaced “maximal” with “greatest”.**
 
@@
 Please do so earlier.
+
+**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text underlined): “However, ~~because~~ water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al.
+1994, Koch et al.
+2004, Sillett et al 2010) and/or because of increased evaporative demand (VPD) caused by leaf warming in sunlit canopy locations (Buckley et al.
+2014)***.
+As a result, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO2 and carbon isotope discrimination discussed in section 3.1 (Table 1).
+Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in gs in response to rising Tair”, and clarified that “high Tair should decrease gs of ***upper canopy***leaves”.**
 

### 2022-01-18 vertical-thermal-review.Rmd
added Tom's response to reviwer's comments and removed thermal capacitance in the sentence

@@
 `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
-The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$ and greater thermal capacitance [capacity to hold heat, @schymanskiStomatalControlLeaf2013].
+The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
@@
 Maximum $g_s$ increases with light--typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979]--and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
-However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata because, of height-related constraints on water transport [ **e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010**] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations (Buckley et al.
+However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata because, of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations (Buckley et al.
 2014).

### 2022-01-18 response-to-reviews.Rmd
responding to some of the comments

@@
 
+**We specified "Scaling up" here as "Scaling up from leaves to trees,"**
+
 Figure 1 is nice.
@@
 
-**Tleaf can be well below Tair on clear nights despite low transpiration, due to radiative coupling with the very cold sky. Added the following references to L205-206:**
+**$T_{leaf}$ can be well below $T_{air}$ on clear nights despite low transpiration, due to radiative coupling with the very cold sky. Added the following references to L205-206:**
 **Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$)[@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020]**
@@
 
-**Added the following lines:**
+**Thank you for the suggestion, we added the following lines to Lines (xxx):**
 **Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
@@
 
+**We appreciate this comment, and have added mention of the fact that maximum $g_s$ increases with light in the canopy because it typically tracks photosynthetic capacity, per @wongStomatalConductanceCorrelates1979.
+There is a great deal of literature showing that the relative limitation of photosynthesis by stomata (gs/A, or intrinsic WUE) tends to be greater in the upper canopy, which results in lower intercellular CO~2~ and C isotope discrimination, as noted in the original manuscript (lines xxx).**
+
 I also think you could describe a little bit more the hydraulic constraints that increase with height.
@@
 
-*need to add more water water use efficiency*
-
-**We added the following sentences related to hydraulic contraints with height:** 
-**Further, tree height also plays a role in determining the gradient of $g_s$ vertically, which has implications for $T_leaf$.
-With an increase in tree height, water travels a longer distance to meet the evaporative demands in sunlit canopy leaves.
-Longer water path length, high irradiance and VPD imposes hydraulic constraints in sunlit canopy leaves (Schäfer et al., 2000; Ambrose et al., 2010).
-Therefore, although typically $g_s$ increases with light, upper canopy leaves of taller trees have lower $g_s$ and leaf water potential to enable high leaf water-use efficiency (Rijkers et al., 2000; Kenzo et al., 2015).
-This, added to the tendency for sun leaves to have higher $T_leaf$, implies that high $T_air$ should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.**
+**We have added mention of a few papers documenting the hydraulic constraints with height as they affect stomatal conductance [@yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010].**
 
@@
 *[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
+
+**Thank you for noting the need to clarify this point.
+We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text underlined here).
+We also deleted the clause that occurred two sentences later, that had said about big-leaf models: “...which do not represent any vertical stratification…”**
 
@@
 
+**We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines xxx-xxx), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles.
+We have also modified the subsequent paragraph about wind speeds to eliminate the use of the confusing term “buffering”.**
+
+**We had also used the word “buffering” in different context, where the intention was to describe how canopy attenuation (of radiation and wind) can reduce daily extremes of temperature in the lower canopy and understory.
+We have modified the text in those places to clarify.
+For example,**
+- **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line xxx)***(new text underlined)**
+- **“Air temperature (Tair) often shows little variation across the vertical gradient, but under certain circumstances, ***attenuation of radiation and vertical transport of sensible heat by the canopy can buffer the lower canopy and understory from large diurnal swings in air temperature***~~it can be significantly buffered by forest canopies~~"; (line xxx)**
+- **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum Tair more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime Tair and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line xxx)**
+
 I was eager to see the part of the manuscript about photoprotection and also its relationship to heat damage.
@@
 
-**We added foliage clumping into the list of factors influencing light gradient here:**
+**We added foliage clumping into the list of factors influencing vertical light gradient here:**
 **"Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
@@
 **We were using the term “buffer” in a functional or biological sense, meaning “protect,” rather than as a description of the underlying physics; i.e., the canopy protects the understory from large diel swings in both air temperature and radiation.
-We have reworded this passage to clarify: “Typically, diel temperature range is smaller beneath dense canopies than above, resulting in,  ~~dense canopies buffer understories from high maximum Tair more than open canopies, i.e., dense canopy understories can have~~ cooler maximum daytime Tair and warmer nighttime minimum Tair than open canopy understories or nearby clearings.”**
+We have reworded this passage to clarify: “Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum Tair more than open canopies, i.e., dense canopy understories can have~~ cooler maximum daytime Tair and warmer nighttime minimum Tair than open canopy understories or nearby clearings.”**
 

### 2022-01-19 vertical-thermal-review.Rmd
small update in the abstract. Krista please let me know if this sounds rights?

@@
 Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity and damage thresholds across the gradient appear modest.
-Scaling up, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$'s, particularly under climate change.
+Scaling up from leaves to trees, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$'s, particularly under climate change.
 In contrast, understory trees benefit from a buffered microclimate but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when the buffering provided by large trees is lost.

### 2022-01-19 vertical-thermal-review.Rmd
small edits

@@
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020].
-In combination with $T_{air}$, RH determines vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+In combination with $T_{air}$ and $T_{leaf}$, RH determines leaf-to-air vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
@@
 Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
-With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) and larger $g_s$ [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
 However, when leaf water demand (determined by VPD and stomatal opening) exceeds the rate of supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig.
@@
 
-Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both predicted by optamization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
+Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both predicted by optamization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
 This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under conditions of high light and sufficient water (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).
@@
 In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2022], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
-This points to a dominant role of vertical profiles in the biophysical environment (Fig. `r fig_NEON_vertical`) in shaping tree growth rates within forests.
+This points to a dominant role of vertical profiles in the biophysical environment, particularly light (Fig. `r fig_NEON_vertical`), in shaping tree growth rates within forests.
 
@@
 Dynamic global vegetation models (DGVMs), which comprise the land surface models in Earth system models, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].
-DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
+DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
 This simplification is computationally more efficient, although it comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].

### 2022-01-19 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-The findings of this review reinforce the notion that representing vertical structuring, as can be done using cohort-based-models, is essential to capturing forest dynamics under global change.
+The findings of this review reinforce the notion that representing vertical structuring, as can be done more thoroughly using cohort-based-models, is essential to capturing forest dynamics under global change.
 Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].

### 2022-01-22 response-to-reviews.Rmd
responding to comment on line382 by r2

@@
 *doing this because I don't think we want to talk about soil conditions. I've always had atmospheric conditions in my mind as I've thought about this.*
+*[suggestion: you could say ‘conditions conducive to stomatal opening’.
+If by ‘wet’ you meant high humidity, then the sentence is probably incorrect, because high humidity by itself would suppress evaporative cooling.]*
 
@@
 
+**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax).
+See lines X-X**
+*Need to add these lines: "Owing to difference...".*
+
 To me, an important remark is that most gradients in DGVMs are prescribed.
@@
 *[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
+
+*[This is essentially impossible to answer without doing the research. Perhaps you could just acknowledge that the value of different modeling approaches is difficult to judge a priori.]*
+
+*Agreed, although I think there is a recognition in the modeling community that we do need improved understanding of the mechanisms to better represent the level of vertical variation desired by some.
+I added a sentence: ("However, doing so will require improved understanding of the mechanisms controlling vertical gradients.") to acknowledge that and I also removed some of the language that emphasizes cohort based models as the only or best way to do this.*
 
@@
 
+*I don't understand what needs to be done with the sentence*
+
 line128: '… but understory leaf area density is often relatively high in the understory as well' sounds redundant. Please reword.
@@
 **We removed the second 'understory' word here:** 
-**Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high ~~in the understory~~ as well**
+**"Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high ~~in the understory~~ as well"**
 
@@
 **Thank you for this suggestion, we added the following information:**
-*"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions.
-Particularly, understory seedlings above ground can have highest rates of CO~2~ refixation [Brooks et al.
-1997]."*
+**"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions.
+Particularly, understory seedlings above ground benefit from elevated CO~2~, therefore showing highest rates of CO~2~ refixation [Brooks et al.
+1997]."**
 
@@
 line382: 'more frequent stomatal closure higher' is awkward and does not follow the comparative. Higher than what?
+
+**We agree that the comparison is not very clear here.
+We added "compared to shaded lower canopies" to the sentence: "Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) ***compared to shaded lower canopies,***result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`)**
 

### 2022-01-22 response-to-reviews.Rmd
responding to R2 line148, and editing the sentences

@@
 line148: 'Variability in the light environment decreases with height' – I don’t see this except perhaps at z = 0 and even so seems tenuous. Explain.
+**We revised the sentence to make it clearer:**
+- **Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
+Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
+
+- **Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation or PAR, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
 
@@
 **We agree that the comparison is not very clear here.
-We added "compared to shaded lower canopies" to the sentence: "Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) ***compared to shaded lower canopies,***result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`)**
+We added "compared to shaded lower canopies" to the sentence: "Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies ***compared to shaded lower canopies,***result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues**
 

### 2022-01-22 response-to-reviews.Rmd
small changes

@@
 line148: 'Variability in the light environment decreases with height' – I don’t see this except perhaps at z = 0 and even so seems tenuous. Explain.
+
 **We revised the sentence to make it clearer:**
-- **Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
+
+- **Previous sentences: Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
 Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
 
-- **Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation or PAR, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
+- **Changed sentences: Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation or PAR, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
 

### 2022-01-22 vertical-thermal-review.Rmd
small changes

--- 2022-01-19 vertical-thermal-review.Rmd
@@
 `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
-Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
-Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
-This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
+Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
+Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation or PAR, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
+This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
-Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
 The implications of this light gradient are that leaf traits vary along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
@@
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions [Table 1].
+Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ vertically, where both tend to be higher in the lower canopy positions (Table 1).
 Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997].
@@
 Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
 
@@
 Maximum $g_s$ increases with light--typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979]--and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
-However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata because, of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations (Buckley et al.
-2014).
+However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata, because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations [@buckleyStomatalOptimisationRelation2014].
 These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
@@
 
-At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress in press].
+At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
@@
 Canopy trees account for the majority of forest ecosystem water and carbon cycling.
-While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
+While studies partitioning transpiration across forest vertical gradients are rare, both evapo-transpiration (ET) and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
 Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
@@
 Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
-Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress in press].
+Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
 
@@
 As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
-Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting;  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
+Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
@@
 
-The findings of this review reinforce the notion that representing vertical structuring, as can be done using cohort-based-models, is essential to capturing forest dynamics under global change.
+Owing to differences in the representation of canopy layers, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes.
+In general, however, this variation is accounted for via light competition.
+Models partition incoming radiation and radiation within the canopy (i.e., direct vs.
+diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple canopy layers can analytically solve the 2-stream approximation for each layer.
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].
+In recent years, there has been a growing effort to incorporate vertical variation more directly in models, specifically as it pertains to leaf water potential and light absorption by leaves, along with a general increasing interest in making direct comparison between models and field measurements [@fisherPerspectivesFutureLand2020].
+Several recent model developments are at the cutting edge of representing vertical variation.
+Model developments by [Bonan et al., 2018; Chen et al., 2016; Longo et al., 2019] implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into canopy air space and the humidity of the airspace, modulated by turbulence processes within the canopy and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
+
+The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.
 Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
 Moreover, given the anticipated importance of mid-canopy and understory trees in ecosystem resilience to increasing mortality of canopy trees, it is absolutely essential that models separately represent these strata.
+However, doing so will require improved understanding of the mechanisms controlling vertical gradients.
 
@@
 
-Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, as we present in this paper).
-At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @depurySimpleScalingPhotosynthesis1997; @bonanMovingIncorrectUseful2021].
+Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers [e.g., sunlit and shaded portions; @depurySimpleScalingPhotosynthesis1997] may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, as we present in this paper).
+At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @bonanMovingIncorrectUseful2021].
 In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.`r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.`r `fig_leaf_T`) to vertically varying abiotic conditions (Fig.`fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.

### 2022-01-23 vertical-thermal-review.Rmd
added BoN for Ci into traits table

@@
 nocite: |
-  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010
+  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010; @brooksVerticalGradientsPhotosynthetic1997a
 ---
@@
 Niinemets et al.
-2010
+2010; 51.
+Brooks et al.
+1997
 ```{r eval=FALSE, echo=FALSE}

### 2022-01-23 NEON_height_profiles/scripts/lad_profs_v_therm_plotting_for_ian_v1.R
Updated scripts and figures per #83

@@ -36,15 +36,15 @@ for(i in 1:3){
     df$val <- df$lad
   }
   
-  #normalize the height
-  ## as of Feb. 2021, we are not normalizing the height
-  # df$ht.norm <- df$ht/(df$max.ht)
+  #normalize the height. If not normalizing, comment this out and change the 
+  ## y-values of the plot to be ht instead of ht.norm
+  df$ht.norm <- df$ht/(df$max.ht)
   
   #create the plots and send to list
   plotsLAD[[i]] <- local({
-    graph <- ggplot(df, aes(val, ht, color=site)) +
+    graph <- ggplot(df, aes(val, ht.norm, color=site)) +
       geom_path(size=1) + 
-      ylim(0,80) + #80 to be standardized across all panels
+      ylim(0,1.75) + #if not normalizing height, ylim=c(0,80). Otherwise, c(0,2)
       labs(x="", y = "") +
       theme_bw() +
       # theme_classic(16)  + 
@@ -52,7 +52,7 @@ for(i in 1:3){
     
     if(i==1){ #LAD
       graph <- graph +
-        xlab(expression(Modelled~leaf~area~density~(m^{2}~m^{-3}))) +
+        xlab(expression(Modelled~leaf~area~density~"[m"^{2}~"m"^{-3}~"]")) +
         ylab("Height [m]")
     } else if(i==2){ # light profile
       graph <- graph +

### 2022-01-23 NEON_height_profiles/scripts/verticalProfileFigure.R

@@ -68,21 +68,29 @@ dp <- data.table(data = c("2DWSD", "RH", "SAAT", "IRBT", "PARPAR"),
 sites <- c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF")
 clrs <- c("#3399FF", "gold", "grey", "red", "#66CCFF", "#003399")
 
+#bring in normalized height (from load("data/lad_n_light_profs_for_ian_v1.RData") in
+## the lad_profs_v_therm_plotting_for_ian.v1)
+normHeightData <- 
+  data.table(site=c("HARV", "OSBS", "PUUM", "SCBI", "SERC", "WREF"),
+             normH=c(32, 21, 23, 43, 39, 53))
+# HARV=32, OSBS: 21, PUUM: 23, SCBI: 43, SERC: 39, WREF: 53
+
 plots <- list()
 for(i in 1:length(dp[,name])){
-  alldata <- NULL
-  for(j in 1:length(sites)){
-    load(paste0("data/neon_rdata/", sites[j], "test.Rdata"))
+  # alldata <- NULL
+  # for(j in 1:length(sites)){
+  allDataList <- lapply(1:length(sites), function(X, varName=i){
+    load(paste0("data/neon_rdata/", sites[X], "test.Rdata"))
     
-    var <- as.data.table(full_data[names(full_data) == dp[,name][i]])
+    var <- as.data.table(full_data[names(full_data) == dp[,name][varName]])
     colnames(var) <- 
-      gsub(paste0(dp[,name][i], "."), "", 
+      gsub(paste0(dp[,name][varName], "."), "", 
            colnames(var))
     
     #get daily min/max by vertical position
     test <- var[, day := as.Date(startDateTime)
-                ][, .(day_max = max(get(dp[,name][i]), na.rm=TRUE),
-                      day_min = min(get(dp[,name][i]), na.rm=TRUE)),
+                ][, .(day_max = max(get(dp[,name][varName]), na.rm=TRUE),
+                      day_min = min(get(dp[,name][varName]), na.rm=TRUE)),
                   by = .(day, zOffset)
                   ][, `:=` (day_max = ifelse(day_max %in% c(-Inf, Inf), NA, 
                                              day_max),
@@ -92,14 +100,6 @@ for(i in 1:length(dp[,name])){
                               yr = year(day))
                       ][yr <= 2020, ]
     
-    ## originally we were getting delta[var] as being var at height h - var 
[…]

### 2022-01-23 response-to-reviews.Rmd
addressing comment on line510-525 by r2

@@
 
+**Thank you for noting this, we revised the sentence to make it clear, where we specify that majority of the studies are within species between sun and shade leaves of the same trees.
+However there are few studies comparing between overstory and understory, other than @carterExperimentalWarmingTropical2021**
+
+- **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, that are predominately within species between sun and shade leaves of the same trees reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes**
+
+- **We also made similar changes specifying within canopy sun and shade from lines (xx-xx)**
+
 The section on VOC emissions was interesting and informative.

### 2022-01-23 vertical-thermal-review.Rmd
addressing comment on line510-525 by r2

@@
 
-We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
+We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, that are predominately within species between sun and shade leaves of the same trees reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
 Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
@@
 However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
+Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], within-canopies of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].

### 2022-01-23 response-to-reviews.Rmd
Added NPQ lines to photoprotection and responded to comment on NPQ

@@
 
+**We seperated VAZ section from VOC, and provided further context to NPQ in the following lines:**
+
+- **Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species (ROS).
+However, at a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.
+However, beyond this PPFD threshold, NPQ rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
+The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].**
+
 *[issue #92](https://github.com/EcoClimLab/vertical-thermal-review/issues/92)*

### 2022-01-23 vertical-thermal-review.Rmd
seperating paragraphs: biochemical and VOC
Added NPQ lines to photoprotection and responded to comment on NPQ

@@
 Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition, and therefore also serve as thermoprotection and postpone heat damage [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching (NPQ) [@mathurPhotosyntheticEfficiencySun2018b].
+Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species (ROS).
+However, at a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.
+However, beyond this PPFD threshold, NPQ rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
+
 Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].

### 2022-01-23 response-to-reviews.Rmd
need to work on NPQ a little more

@@
 
-- **Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species (ROS).
+- **With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching (NPQ) [@mathurPhotosyntheticEfficiencySun2018b].
+Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species (ROS).
 However, at a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.

### 2022-01-23 NEON_height_profiles/scripts/lad_profs_v_therm_plotting_for_ian_v1.R
Updated scripts and figures for #83

@@ -48,22 +48,33 @@ for(i in 1:3){
       labs(x="", y = "") +
       theme_bw() +
       # theme_classic(16)  + 
-      scale_color_jco()
+      scale_color_jco() +
+      geom_hline(yintercept=1, linetype=2)
     
     if(i==1){ #LAD
       graph <- graph +
         xlab(expression(Modelled~leaf~area~density~"[m"^{2}~"m"^{-3}~"]")) +
-        ylab("Height [m]")
+        ylab("Height [m]") +
+        theme(axis.title.x = element_text(size=14),
+              axis.text.x = element_text(size=16),
+              axis.title.y = element_text(size=14),
+              axis.text.y = element_text(size=16))
+        
     } else if(i==2){ # light profile
       graph <- graph +
         xlab('Proportion of incident light') +
         theme(axis.text.y=element_blank(),
-              axis.ticks.y=element_blank())
+              axis.ticks.y=element_blank(),
+              axis.title.x = element_text(size=14),
+              axis.text.x = element_text(size=16))
+      
     } else if(i==3){ #sun leaves
       graph <- graph +
         xlab("Calculated proportion of sun leaves") +
         theme(axis.text.y=element_blank(),
-              axis.ticks.y=element_blank())
+              axis.ticks.y=element_blank(),
+              axis.title.x = element_text(size=14),
+              axis.text.x = element_text(size=16))
     } 
     graph <- graph + 
       theme(plot.title = element_text(size=16),

### 2022-01-23 NEON_height_profiles/scripts/verticalProfileFigure.R

@@ -157,11 +157,11 @@ for(i in 1:length(dp[,name])){
       graph +
       geom_point(aes(x = all_mean, y = plotHeight, color = site), 
                  shape=19, size=4) +
-      geom_line(aes(x = all_mean, y = plotHeight, color = site), size=1) +
+      geom_path(aes(x = all_mean, y = plotHeight, color = site), size=1) +
       ggplot2::geom_errorbarh(aes(xmin = all_mean - all_sd, 
                                   xmax = all_mean + all_sd, 
-                                  y=plotHeight, color = site, height=0.1),
-                              size=1) +
+                                  y=plotHeight, color = site, height=0.1)) +
+      geom_hline(yintercept=1, linetype=2) +
       labs(x = "",
            y = "") +
       theme_bw()
@@ -197,6 +197,10 @@ for(i in 1:length(dp[,name])){
     } else {
       graph <- graph + theme(legend.position="none")
     }
+    
+    graph <- graph +
+      theme(axis.title = element_text(size=14),
+            axis.text = element_text(size=16))
   })
 }
 
@@ -205,11 +209,13 @@ names(plots) <- dp[,name]
 ## bring in LAD profiles
 source("scripts/lad_profs_v_therm_plotting_for_ian_v1.R")
 
-p <- ggarrange(plotsLAD[["LAD"]], plotsLAD[["sun"]], plotsLAD[["lgt"]], 
-               plots[["PARMean"]], 
-               plots[["windSpeedMean"]], plots[["RHMean"]], 
-               plots[["tempSingleMean"]], plots[["bioTempMean"]],
-               nrow=2, ncol=4, labels=c("A", "B", "C", "D", "E", "F", "G", "H"))
+p <- cowplot::plot_grid(plotsLAD[["LAD"]], plotsLAD[["sun"]], plotsLAD[["lgt"]], 
+                        plots[["PARMean"]], 
+                        plots[["windSpeedMean"]], plots[["RHMean"]], 
+                        plots[["tempSingleMean"]], plots[["bioTempMean"]],
+                        nrow=2, ncol=4, align="hv",
+                        labels=c("A", "B", "C", "D", "E", "F", "G", "H"),
+                        label_x=0.13)
 
 png(paste0("figures/Fig2_normalized.png"), height=600, width=960)
 print(p)

### 2022-01-26 response-to-reviews.Rmd
added canopy die-back information
maybe I need to expand a little more o n it?

@@
 
-**Thank you for the review. We have added citations to the references mentioned here. We have also added some discussion of canopy dieback events(?)... [details].**
+**Thank you for the review.
+We have added citations to the references mentioned here:[@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
+We have also added some discussion of canopy dieback events here:**
+**Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can rapidly dry soils up, exacerbating effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].**
 

### 2022-01-26 vertical-thermal-review.Rmd
added canopy die-back information
maybe I need to expand a little more o n it?

@@
 Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
+Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can rapidly dry soils up, exacerbating effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].

### 2022-01-27 vertical-thermal-review.Rmd
small edits

@@
 Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
-Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation or PAR, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
+Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
 This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and dry conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
@@
 One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the canopy of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
-However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients.
+However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2020].
 
@@
 While studies partitioning transpiration across forest vertical gradients are rare, both evapo-transpiration (ET) and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
-Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
-In terms of C cycling, it has been estimated that canopy strata contribute $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007].
+Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
+In terms of C cycling, it has been estimated that canopy strata contribute $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007].
 Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
@@
 Canopy disturbance poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013a; @greiserHidingClimateCharacterizing2019].
-Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
+Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
 In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020b; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
@@
 
-Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers [e.g., sunlit and shaded portions; @depurySimpleScalingPhotosynthesis1997] may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, as we present in this paper).
+Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers [e.g., sunlit and shaded portions, @depurySimpleScalingPhotosynthesis1997] may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, as we present in this paper).
 At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @bonanMovingIncorrectUseful2021].

### 2022-01-27 vertical-thermal-review.Rmd
Including Bin et al. 2022

@@
 The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types [Fig.
-`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
+`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022].
 Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
@@
 Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig.
-`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
+`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021; @binLeafTraitExpression2022].
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020].
@@
 The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as local topography, vertical air mixing, cloud cover,  vegetation type and sensible heat emissions [@defrenneForestMicroclimatesClimate2021].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting in maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009].
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting in maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022].
 Notably, this pattern can be reversed in open forests.
@@
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
-This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Fig.
-`r fig_NEON_vertical`).
+This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions [Fig.
+`r fig_NEON_vertical`, @binLeafTraitExpression2022] and also due to differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
 
@@
 
-Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
 Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
 It has been established that independently of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
-For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].
+For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
 Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
@@
 Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
-Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012].
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022].
 For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [mirroring the faction of individuals, @conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
@@
 Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
-Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
+Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] photosynthesis, and intra-canopy trait variation [@binLeafTraitExpression2022].
 Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).

### 2022-01-27 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
changed leaf energy balance gs units

@@ -293,14 +293,15 @@ dr_gsu$u_tla<- dr_gsu$tleaf - dr_gsu$tair
 
 #create a data frame for Tleaf-Tair, for overstory and understory, in normal and drought scenario
 gs<-data.frame(l_tla = gs_l$l_tleaf_tair, u_tla = gs_u$u_tleaf_tair, dr_ltla = dr_gsl$l_tla, dr_utla = dr_gsu$u_tla, gs = gs_l$gs)
-
+gs$gs<-gs$gs/10#changing units from umol/m^2/s/Pa
+View(gs)
 #plot gs
 c1<-ggplot(gs)+
   geom_smooth(aes(x = gs, y = l_tla),  method = lm, se = FALSE, color = "#7b3294")+
   geom_smooth(aes(x = gs, y = u_tla),  method = lm, color = "#008837", se = FALSE)+
   geom_smooth(aes(x = gs, y = dr_ltla),  method = lm, color = "#7b3294", se = FALSE, linetype = "dashed")+
   geom_smooth(aes(x = gs, y = dr_utla),  method = lm, color = "#008837", se = FALSE, linetype = "dashed")+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("Stomatal Conductance ($g_{s}$, mmol $m^{-2} s^{-1}$)"))+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab(TeX("Stomatal Conductance ($g_{s}$, mol $m^{-2} s^{-1}$)"))+
   geom_hline(yintercept=0, linetype='dotted', col = 'black')+
   theme_few()+theme(text = element_text(size = 15))+ylim(-5, 14)
 c1
@@ -434,10 +435,10 @@ e1
 
 
 table<- data.frame(biophysical = c("swr", "ws", "rh", "lw", "gs", "tair"),
-                   normalno = c("871", 2.88, 0.91, 0.04, "404", "25"),
-                   normal_u = c("102", 0.24, 0.97, 0.10, "202", "23"),
-                   drought_o = c("1151", 2.88, 0.46, 0.04, "1.01", "25"),
-                   normalno = c("153", 0.24, 0.49, 0.10, "1.01", "23"))
+                   normalno = c("871", 2.88, 0.91, 0.04, "0.4", "25"),
+                   normal_u = c("102", 0.24, 0.97, 0.10, "0.2", "23"),
[…]

### 2022-01-29 vertical-thermal-review.Rmd
L 609 and throughout

@@
 `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
-Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
+Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers (throughout we use "overstory", "top" and "upper" canopies as synonyms), but understory leaf area density is often relatively high as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989].

### 2022-01-31 response-to-reviews.Rmd
L 609 and throughout
small edits

@@
 
-**We specified "Scaling up" here as "Scaling up from leaves to trees,"**
+**We specified "Scaling up" here as "Scaling up from leaves to trees,".**
 
@@
 Usually, it is easy to understand with the context of the sentence, but you might want to be consistent everywhere.
+
+**We agree with the suggestion. To make the useage of different synonyms clearer, we added a note in lines(xxx)"--(throughout we use "overstory", "top" and "upper" canopies as synonyms)--"**
 
@@
 **Thank you for this suggestion, we added the following information:**
-**"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions.
-Particularly, understory seedlings above ground benefit from elevated CO~2~, therefore showing highest rates of CO~2~ refixation [Brooks et al.
-1997]."**
+**"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~, where both tend to be higher in the lower canopy positions.
+Particularly, understory seedlings above ground benefit from elevated CO~2~, therefore showing highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997a]."**
 
@@
 
-**Variables in the inset table are now described in the figure caption as: "Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf width (lw), stomatal conductance (gs), air temperature (tair)."**
+**We added the description for the variables as part of the figure caption: Variables in the inset table are now described in the figure caption as: "Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf width (lw), stomatal conductance (gs), air temperature (tair)."**
 
@@
 **Thank you for noting this, we revised the sentence to make it clear, where we specify that majority of the studies are within species between sun and shade leaves of the same trees.
-However there are few studies comparing between overstory and understory, other than @carterExperimentalWarmingTropical2021**
+@carterExperimentalWarmingTropical2021 is one of the only studies comparing between overstory and understory.**
 

### 2022-01-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species (ROS).
-However, at a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.
-However, beyond this PPFD threshold, NPQ rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
+However, below a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.
+Beyond this PPFD threshold, NPQ rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].

### 2022-02-05 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity and damage thresholds across the gradient appear modest.
-Scaling up from leaves to trees, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$'s, particularly under climate change.
+Scaling up from leaves to trees, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$, particularly under climate change.
 In contrast, understory trees benefit from a buffered microclimate but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when the buffering provided by large trees is lost.
@@
 `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
-Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers (throughout we use "overstory", "top" and "upper" canopies as synonyms), but understory leaf area density is often relatively high as well, sometimes with undulating patterns [e.g., SCBI, SERC and HARV, Fig.
+Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers (throughout we use "overstory", "top" and "upper" canopies as synonyms), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989].
@@
 Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
-Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
+Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation, and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
 Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
@@
 
-Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles.
+Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature might also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles.
 Vertical transport at the macroscopic scale of a plant canopy is dominated by turbulent transport rather than by molecular diffusion.
@@
 The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as local topography, vertical air mixing, cloud cover,  vegetation type and sensible heat emissions [@defrenneForestMicroclimatesClimate2021].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting in maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022].
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022].
 Notably, this pattern can be reversed in open forests.
@@
 Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and under higher wind speeds (Fig. `r fig_leaf_T`b). 
-Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD, where although $g_b$ increases with wind speed, $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
+Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD:$g_b$ increases with wind speed, and $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
 Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig.
@@
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
-Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf width (lw), stomatal conductance (gs), air temperature (tair).
+Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), characteristic leaf width (clw), stomatal conductance (gs), air temperature (tair).
 ](leaf_energy_balance/fig3_leafenergy.png)
@@
 The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions.
-Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
+Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.
+`r fig_leaf_T`).
 This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$ limitation than shade leaves [Fig.
@@
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to shedding excess heat compared to upper-canopy leaves, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipate excess heat compared to upper-canopy leaves.
+Limitations in heat dissipation, in combination with their physiological propensities, may cause shade leaves to experience above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
 In addition to lower wind speeds (Fig.
@@
 The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the Tleaf of the foliage close to the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
 One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the canopy of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
@@
 Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition, and therefore also serve as thermoprotection and postpone heat damage [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
+Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and harmlessly dissipating excess excitation energy as heat to reduce photoinhibition, and therefore also serve as thermoprotection and postpone heat damage [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
 With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching (NPQ) [@mathurPhotosyntheticEfficiencySun2018b].
-Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species (ROS).
-However, below a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.
+Efficient light-use is optimized through this process while preventing the accumulation of excess light, an energetic imbalance that results in the generation of harmful reactive oxygen  species (ROS).
+However, below a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within a beech canopy (*Fagus sylvatica L.*), shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy leaves, reflecting rapid photosynthetic induction kinetics in shaded leaves as an acclimation to sunflecks.
 Beyond this PPFD threshold, NPQ rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
@@
 
-Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both predicted by optamization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
-This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under conditions of high light and sufficient water (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).
+Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both predicted by optimization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
 
@@
 Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
-Beyond the optimum, photosynthesis decreases as a result of stomatal closure [e.g.
-@slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; @fredeenTemperatureHumidityEffects1999], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
+Beyond the optimum, photosynthesis decreases as a result of stomatal closure [ @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; @fredeenTemperatureHumidityEffects1999], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
 
-We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, that are predominately within species between sun and shade leaves of the same trees reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
+We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, which  compare sun and shade leaves of the same trees, reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
 Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
@@
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
+Thermal thresholds for photosystem II functioning serve as good proxies for temperatures at which leaves get necrotic [@krauseHightemperatureToleranceTropical2010].
 Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021].
@@
 However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
-The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
+The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
 
@@
 In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
-Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both classes of VOC are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
+Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
 Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].

### 2022-02-05 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and dry conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) characteristic leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and dry conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.

### 2022-02-07 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 
+**Thank you for pointing this out.
+In line404-405, we are concentrated on interactions between microenvironments and function.
+We edited the text to emphasize the mechanisms for earlier bud break in the understory, including warmer temperatures, and the benefits of maximizing carbon gain before overstory closure reduces light availability [added ref.
+@leeSpringPhenologicalEscape2021].
+We edited the rest of the paragraph to more clearly emphasize mechanisms of feedbacks between vertically structured microenvironments, leaf phenology, and forest function.**
+
 line475: 'because water supply often cannot meet the demands incurred …'. How do we know how frequent this is? Evidence?

### 2022-02-07 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
-It has been established that independently of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
+Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
+It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
@@
 In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
-For example, in a temperate deciduous forest, spring bud-break tended to occur earlier among juvenile trees in the more sheltered understory, where temperature sums accumulated more rapidly, than among conspecific canopy trees [@augspurgerDifferencesLeafPhenology2003].
-Within tropical dry-deciduous species in Panama, the fraction of seasonally deciduous individuals increased toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-Even in evergreen forests, environmental variation alters leaf age distributions [@albertCrypticPhenologyPlants2019], which affects forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and heat stress via age-specific leaf function [@albertCrypticPhenologyPlants2019].
-For instance, in tropical evergreen forests, because leaf turnover rates of sun leaves in the upper canopy is faster than in the subcanopy, less damage accumulates [@millerOnlySunlitLeaves2021]. 
-Leaf age affects vulnerability to heat stress and has a strong effect on metabolism [@mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015; @kikuzawaSynthesisRelationshipsLeaf2006], and can be expected to similarly scale through vertically structured phenology to affect forest responses to heat anomalies.
+In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
+Early leaf-out allows some tree saplings to do the majority of their annual carbon fixation before the overstory closes and reduces light availability [@leeSpringPhenologicalEscape2021]. 
+Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].
+Leaf turnover is faster in the upper canopy than the understory, putatively due to more rapid accumulation of leaf damage [@osadaLeafDynamicsMaintenance2001; @millerOnlySunlitLeaves2021].
+Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; @mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015].
+Leaf phenology thereby both responds to and influences the vertical structure of forest microenvironments and function.
 

### 2022-02-07 response-to-reviews.Rmd
add instructions on figures to cover letter
#97, #96

@@
 
-*(To ensure rapid handling of your manuscript we would be grateful if you could confirm, in the cover letter to the Editor, whether or not there are any important elements of the figures that need to be retained (eg.
-colours used, plant species drawn, layout and spacing, symbols, etc) and if you have specific instructions that you would like the illustrator to take on board.)*
+Regarding production of the figures, below are some specific criteria:
+<!--To ensure rapid handling of your manuscript we would be grateful if you could confirm, in the cover letter to the Editor, whether or not there are any important elements of the figures that need to be retained (eg.
+colours used, plant species drawn, layout and spacing, symbols, etc) and if you have specific instructions that you would like the illustrator to take on board.-->
+
+- It will generally be fine to adjust colors, layout, and spacing on the figures.
+- We really like Figure 1 and would like to retain the background painting (by the lead author)
+- On Fig.
+4, please adjust colors on panel (a) to correspond to those in panel (b), where "min MIN" color should match "understory" and "max DBH" should match "canopy".
+The same color scheme can be applied to all panels in (a), and there is no need to keep the line type differences (both can be solid).
 

### 2022-02-07 vertical-thermal-review.Rmd
fix Piponiot reference
@NidhiVinod , I've synced the updated reference in Zotero.

@@
 Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
-`r fig_ecosystem`b, @piponiotSizerelatedTreeBiomassinreview in review].
+`r fig_ecosystem`b, @piponiotSizerelatedTreeBiomass2022].
 
 ![**Figure `r fig_ecosystem`.
-Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotSizerelatedTreeBiomassinreview in review.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotSizerelatedTreeBiomass2022.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 

### 2022-02-07 response-to-reviews.Rmd
working through response to reviews - R1

@@
 
+*[issue #99](https://github.com/EcoClimLab/vertical-thermal-review/issues/99)*
+
 I find it interesting to try and link the empirical and theoretical knowledge on vertical gradients in forests with what is done in DGCMs and what should be done to improve their accuracy.
@@
 Maybe by explaining which of the ‘biophysical variables’ and ‘leaf traits’ that you listed, are considered constant when they should be variable vertically?
+
+*[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
 
@@
 
-*[issue #83](https://github.com/EcoClimLab/vertical-thermal-review/issues/83)*
+**Done.**
 
 I like your figures, but maybe consider homogenizing a little bit the styles for the plots between figures 2 3 and 4.
+
+**This will be done by the New Phytologist editorial staff (because it is a Tansley review). We will work with them to ensure that all the figures look good.**
 
@@
 
-**$T_{leaf}$ can be well below $T_{air}$ on clear nights despite low transpiration, due to radiative coupling with the very cold sky. Added the following references to L205-206:**
-**Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$)[@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020]**
+**We revised and added references to this sentence in question, which now reads: **
+**"Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020]."**
 
-*Agree, but we're only changing the one variable at a time. Just need to make this clear.*
+**Figure 3 shows the effect of each variable while the others are held constant (as specified in the caption). Thus, it does not represent the full suite of night-time conditions.**
 
-*Yes, night time conditions entail a shift in Tair, which in our case is constant*
+*Nidhi, is this all correct??.*
 

### 2022-02-07 vertical-thermal-review.Rmd
working through response to reviews - R1

@@
 Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.
-Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Supporting Information Figure `r SIfig_NEON_all`.](NEON_height_profiles/figures/profile_all.png)
+Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Supporting Information Figure `r SIfig_NEON_all`.](NEON_height_profiles/figures/Fig2_normalized.png)
 
@@
 Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015a; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
-Therefore, leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$) [@rey-sanchez_spatial_2016; @vogelLeavesLowestHighest2009; @cavaleriColdbloodedForestsWarming2020].
+Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].
 Leaves can be warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].

### 2022-02-07 response-to-reviews.Rmd
a bit of work on R1 comments

@@
 
-*doing this because I don't think we want to talk about soil conditions. I've always had atmospheric conditions in my mind as I've thought about this.*
-*[suggestion: you could say ‘conditions conducive to stomatal opening’.
-If by ‘wet’ you meant high humidity, then the sentence is probably incorrect, because high humidity by itself would suppress evaporative cooling.]*
+**We have reworded this sentence to read, "In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
+3b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]."**
 
@@
 
-**Thank you for the suggestion, we added the following lines to Lines (xxx):**
-**Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
-Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]**
+**Thank you for the suggestion, we added the following here:"Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]."**
 

### 2022-02-07 vertical-thermal-review.Rmd
a bit of work on R1 comments

@@
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
+In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].

### 2022-02-08 response-to-reviews.Rmd
R1-R2 responses
@NidhiVinod , I added a few responses and comments. There are a few that are not yet complete, but I assume will be. (An aside: I try to avoid putting in line numbers, as it's annoying to have to go back and add those. Easier to specify where you put the sentence and quote the sentence.)

@@
 
-- ...
+- We have brought in an additional author, Thomas Buckley, to round out the expertise of the author team and help address some of the reviewer comments (particularly R2).
 
@@
 
+*[issue #99](https://github.com/EcoClimLab/vertical-thermal-review/issues/99)*
+
 L429 I am not fully sure if I understood this sentence, consider maybe developing the idea.
@@
 
-**We removed this sentence because yes, it is confusing, and the points addressed from comments above provide substantial information for inter- and intra-canopy leaf traits.**
+**We removed this sentence because we agreed that it was confusing, and the points addressed from comments above provide substantial information for inter- and intra-canopy leaf traits.**
 
@@
 
+*(incomplete)*
+
 L 665 though
@@
 
-*[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
-
 **Thank you for noting the need to clarify this point.
-We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text underlined here).
+We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text italicized here).
 We also deleted the clause that occurred two sentences later, that had said about big-leaf models: “...which do not represent any vertical stratification…”**
@@
 
-*[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
+**We agree that there is a recognition in the modeling community that we do need improved understanding of the mechanisms to better represent the level of vertical variation desired by some.
+We added the following sentence to acknowledge that (LXX-XX): " ...
+".
+We also removed some of the language that emphasizes cohort based models as the only or best way to do this.**
 
-*[This is essentially impossible to answer without doing the research. Perhaps you could just acknowledge that the value of different modeling approaches is difficult to judge a priori.]*
-
-*Agreed, although I think there is a recognition in the modeling community that we do need improved understanding of the mechanisms to better represent the level of vertical variation desired by some.
-I added a sentence: ("However, doing so will require improved understanding of the mechanisms controlling vertical gradients.") to acknowledge that and I also removed some of the language that emphasizes cohort based models as the only or best way to do this.*
+**We note that addressing the question poosed here is essentially impossible to answer without doing the research. The value of different modeling approaches is difficult to judge a priori.**
 
@@
 
+*(Nidhi, what do you think of the following?)*
+
+**Thanks for the suggestion.
+We agree, but at the same time are confident that the search terms we used covered the majority of the relevant literature.
+Adding additional search terms could always turn up additional papers, but we feel that we have reached a point of diminishing returns, and unfortunately don't have time to expand the search (lead author has moved on from the institution where she performed the work to a PhD program).**
+
 I included below some more references, mostly in the tropics.
+
+**Thank you for the references. We have reviewed these and incorporated as appropriate.** *(Nidhi, assuing this is true?)*
 
@@
 
+**Thank you for the review!**
+
 ## Referee: 2
@@
 **Thank you for the review.
-We have added citations to the references mentioned here:[@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
-We have also added some discussion of canopy dieback events here:**
-**Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can rapidly dry soils up, exacerbating effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].**
+We have added citations to the references mentioned here [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].**
 
-*Nidhi, be sure to cite all the references listed here.*
-
-*It would be good to add some discussion of canopy dieback events. ([issue #93](https://github.com/EcoClimLab/vertical-thermal-review/issues/93))*
+**We have also added the following sentence on canopy dieback to the canopy disturance section:**
+**"Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021]."**
 
@@
 
-*It would be good to bring in more theory. ([issue #94](https://github.com/EcoClimLab/vertical-thermal-review/issues/94))*
+**Regarding incorporation of more theory, please see our response to the comment below.**
 
-*Fig 2: [issue #83](https://github.com/EcoClimLab/vertical-thermal-review/issues/83))*
+**We have made the suggested improvements to figure 2: lines are bolder, and height normalized to height at the top of the canopy.**
 
@@
 
-*[issue #95](https://github.com/EcoClimLab/vertical-thermal-review/issues/95)*
-
 **We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines xxx-xxx), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles.
@@
 For example,**
+
 - **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line xxx)***(new text underlined)**
-- **“Air temperature (Tair) often shows little variation across the vertical gradient, but under certain circumstances, ***attenuation of radiation and vertical transport of sensible heat by the canopy can buffer the lower canopy and understory from large diurnal swings in air temperature***~~it can be significantly buffered by forest canopies~~"; (line xxx)**
+- **“Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances, ***attenuation of radiation and vertical transport of sensible heat by the canopy can buffer the lower canopy and understory from large diurnal swings in air temperature***~~it can be significantly buffered by forest canopies~~"; (line xxx)**
 - **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum Tair more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime Tair and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line xxx)**
@@
 
-**We seperated VAZ section from VOC, and provided further context to NPQ in the following lines:**
+*[issue #92](https://github.com/EcoClimLab/vertical-thermal-review/issues/92)*
 
-- **With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching (NPQ) [@mathurPhotosyntheticEfficiencySun2018b].
-Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species (ROS).
-However, at a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.
-However, beyond this PPFD threshold, NPQ rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
+**We have revised the first sentence to read: "Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.
+We also choose to keep this rather obvious sentence because we feel that basic topic sentences are important for helping to guide readers from a diversity of sub-disciplines through the manuscript.**
+
+**We separated VAZ section from VOC, and provided further context to NPQ in the following lines:**
+
+- **With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching [@mathurPhotosyntheticEfficiencySun2018b].
+Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species.
+However, at a photon flux density of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher non-photochemical quenching rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.
+However, beyond this photon flux density threshold, non-photochemical quenching rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].**
-
-*[issue #92](https://github.com/EcoClimLab/vertical-thermal-review/issues/92)*
-**It is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.**
 

### 2022-02-08 vertical-thermal-review.Rmd
R1-R2 responses
@NidhiVinod , I added a few responses and comments. There are a few that are not yet complete, but I assume will be. (An aside: I try to avoid putting in line numbers, as it's annoying to have to go back and add those. Easier to specify where you put the sentence and quote the sentence.)

@@
 Vertical gradients in the biophysical environment for six US forest sites in the National Ecological Observatory Network (NEON)**.
-Height profiles are shown for: growing season (a) modelled leaf area density, (b) calculated proportion of sun leaves, and (c) proportion light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature ($T_{air}$), and (h) maximum biological temperature, ($T_{bio}$).
-Measurements extend from ground level (height = 0 m) to the top of the canopy (a-c, h) or above (d-g).
+Height profiles, normalized relative to the top of the canopy, are shown for: growing season (a) modelled leaf area density, (b) calculated proportion of sun leaves, and (c) proportion light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature ($T_{air}$), and (h) maximum biological temperature, ($T_{bio}$).
+Measurements extend from ground level (normalized height = 0 m) to the top of the canopy (horizontal dashed line at normalized height = 1) or above (d-g).
 Sites, which represent a variety of forest structures, include a mixed northern hardwood forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).
 Further site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.
-Vertical profiles in micrometeorological variables (d-h) at all six NEON sites are shown in Supporting Information Figure `r SIfig_NEON_all`.](NEON_height_profiles/figures/Fig2_normalized.png)
+Vertical profiles in micrometeorological variables (d-h) at all forested NEON sites are shown in Supporting Information Figure `r SIfig_NEON_all`.](NEON_height_profiles/figures/Fig2_normalized.png)
 
@@
 
-Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
+Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
 Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and harmlessly dissipating excess excitation energy as heat to reduce photoinhibition, and therefore also serve as thermoprotection and postpone heat damage [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching (NPQ) [@mathurPhotosyntheticEfficiencySun2018b].
-Efficient light-use is optimized through this process while preventing the accumulation of excess light, an energetic imbalance that results in the generation of harmful reactive oxygen  species (ROS).
-However, below a photon flux density (PPFD) of $200\mu molm^{-2}s^{-1}$, within a beech canopy (*Fagus sylvatica L.*), shaded lower canopy leaves have been observed to have higher NPQ rates than upper canopy leaves, reflecting rapid photosynthetic induction kinetics in shaded leaves as an acclimation to sunflecks.
-Beyond this PPFD threshold, NPQ rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
+With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching[@mathurPhotosyntheticEfficiencySun2018b].
+Efficient light-use is optimized through this process while preventing the accumulation of excess light, an energetic imbalance that results in the generation of harmful reactive oxygen species. 
+However, below a photon flux density of $200\mu molm^{-2}s^{-1}$, within a beech canopy (*Fagus sylvatica L.*), shaded lower canopy leaves have been observed to have higher non-photochemical quenching rates than upper canopy leaves, reflecting rapid photosynthetic induction kinetics in shaded leaves as an acclimation to sunflecks.
+Beyond this photon flux density threshold, non-photochemical quenching rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a]. 
 The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
@@
 Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
-Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can rapidly dry soils up, exacerbating effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
+Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].

### 2022-02-08 vertical-thermal-review.Rmd
marielle's suggestion

@@
 Marielle N. Smith^6^,
-Tyeen Taylor^7^,
+Tyeen C. Taylor^7^,
 Lawren Sack^2^,
@@
 Vertical gradients in the biophysical environment for six US forest sites in the National Ecological Observatory Network (NEON)**.
-Height profiles, normalized relative to the top of the canopy, are shown for: growing season (a) modelled leaf area density, (b) calculated proportion of sun leaves, and (c) proportion light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature ($T_{air}$), and (h) maximum biological temperature, ($T_{bio}$).
+Height profiles, normalized relative to the top of the canopy, are shown for: growing season (a) modelled leaf area density, (b) estimated proportion of sun leaves, and (c) proportion of light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature ($T_{air}$), and (h) maximum biological temperature, ($T_{bio}$).
 Measurements extend from ground level (normalized height = 0 m) to the top of the canopy (horizontal dashed line at normalized height = 1) or above (d-g).
+Variables shown in panels a-c are derived from lidar, and those shown in panels d-h from micrometeorological tower data.
 Sites, which represent a variety of forest structures, include a mixed northern hardwood forest (Harvard Forest, MA; HARV), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), a tropical montane broadleaf evergreen forest (Pu'u Maka'ala Natural Area Reserve, Hawai'i; PUUM), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a coniferous forest (Wind River Experimental Forests, WA; WREF).

### 2022-02-08 response-to-reviews.Rmd
standardize use of term "canopy"

@@
 
-**We agree with the suggestion. To make the useage of different synonyms clearer, we added a note in lines(xxx)"--(throughout we use "overstory", "top" and "upper" canopies as synonyms)--"**
-
-*(incomplete)*
+**Thanks for pointing this out.
+We have revisited the use of the term "canopy" throughout the entire manuscript and employed with consistent terminology, using "canopy" to refer only to the layer formed by dominant trees, and using a term such as to "vertical profile" to refer to the entire gradient from understory to the canopy.**
 

### 2022-02-08 vertical-thermal-review.Rmd
standardize use of term "canopy"

@@
 
-A great measure of uncertainty arises due to the differential impacts of rising temperatures on trees of differential size and canopy position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].
+A great measure of uncertainty arises due to the differential impacts of rising temperatures on trees of differential size and crown position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].
 Forests are vertically and horizontally stratified, with overstory canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
@@
 
-Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests. 
+Here, we review how the biophysical environment and plant form and function vary across the vertical gradient in forests. 
 We focus on five key themes (Fig. `r fig_schematic`):
@@
 Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology.** Abbreviations are as follows: VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; T sensitivity: temperature sensitivity; ET: evapotranspiration.
-Here, we summarize conditions in dense canopies during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open canopies, when canopy trees are seasonally deciduous, or at nighttime.
-Arrows indicate direction of increase, with double-pointed arrows indicating that observations have shown increases both higher and lower in the canopy.
-](schematics/Schematic figure 1.png)
+Here, we summarize conditions in closed-canopy forests during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime.
+Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.](schematics/Schematic figure 1.png)
 
@@
 `r fig_NEON_vertical`, Supporting Information Methods `r appendix_NEON_methods`, Supporting Information Figure `r SIfig_NEON_all`).
-While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer canopy than to height [@lowmanForestCanopies1995].
+While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer edge of vegetation than to height [@lowmanForestCanopies1995].
 
@@
 `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
-Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers (throughout we use "overstory", "top" and "upper" canopies as synonyms), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
+Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989].
-In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., OSBS, PUUM, and WREF, Fig.
+In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig.
 `r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;  @bonan_ecological_2016].
@@
 Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
-The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types [Fig.
+The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
 `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022].
-Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
+Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers.
 Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation, and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
@@
 This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
+`r fig_NEON_vertical`d), than in forests with more open upper canopies, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
@@
 Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature might also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles.
-Vertical transport at the macroscopic scale of a plant canopy is dominated by turbulent transport rather than by molecular diffusion. 
+Vertical transport at the macroscopic scale of a forest canopy is dominated by turbulent transport rather than by molecular diffusion. 
 Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients (eddy diffusivities, often symbolized K) that depend on both friction and buoyancy [@penmanWeatherWheatEssay1960].
@@
 Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-However, K-theory often fails within canopies, making canopy wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+However, K-theory often fails within canopies, making forest verticcal wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
 For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux [e.g., @raupachLagrangianAnalysisScalar1987].
@@
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ vertically, where both tend to be higher in the lower canopy positions (Table 1).
+Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
 Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
@@
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see sections 2.2, 4.1).
+The implication is that upper canopy leaves face higher $g_b$ and associated potential for water and heat loss (see sections 2.2, 4.1).
 
@@
 The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as local topography, vertical air mixing, cloud cover,  vegetation type and sensible heat emissions [@defrenneForestMicroclimatesClimate2021].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than open canopy understories or nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022].
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022].
 Notably, this pattern can be reversed in open forests. 
@@
 High radiation at the top of the vertical profile (Fig.
-`r fig_NEON_vertical`) implies that canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig.
+`r fig_NEON_vertical`) implies that upper canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig.
 `r fig_leaf_T`).
@@
 The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions.
-Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.
+Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the upper canopy compared to the understory (Fig.
 `r fig_leaf_T`).
@@
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
-This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ in the lower canopy relative to the upper canopy [Fig.
+This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [Fig.
 `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
-The latter can result from a combination of still air in the lower canopy, and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
+The latter can result from a combination of still air at lower heights (Fig.
+`r fig_NEON_vertical`e), and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the Tleaf of the foliage close to the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
-One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the canopy of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
+One  of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
 However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2020].
@@
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
-This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions [Fig.
-`r fig_NEON_vertical`, @binLeafTraitExpression2022] and also due to differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
+This dichotomy in leaves is observed along the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions [Fig.
+`r fig_NEON_vertical`, @binLeafTraitExpression2022], and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
 
@@
 
-Within-canopy, leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
-It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
+Leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
+Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.
+It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy. 
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
@@
 Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
-Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
+Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
 Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
@@
 Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and harmlessly dissipating excess excitation energy as heat to reduce photoinhibition, and therefore also serve as thermoprotection and postpone heat damage [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching[@mathurPhotosyntheticEfficiencySun2018b].
+With saturating irradiance experienced by upper canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching[@mathurPhotosyntheticEfficiencySun2018b].
 Efficient light-use is optimized through this process while preventing the accumulation of excess light, an energetic imbalance that results in the generation of harmful reactive oxygen species. 
-However, below a photon flux density of $200\mu molm^{-2}s^{-1}$, within a beech canopy (*Fagus sylvatica L.*), shaded lower canopy leaves have been observed to have higher non-photochemical quenching rates than upper canopy leaves, reflecting rapid photosynthetic induction kinetics in shaded leaves as an acclimation to sunflecks.
+However, below a photon flux density of $200\mu molm^{-2}s^{-1}$, within a beech (*Fagus sylvatica L.*) forest, shaded lower canopy leaves have been observed to have higher non-photochemical quenching rates than upper canopy leaves, reflecting rapid photosynthetic induction kinetics in shaded leaves as an acclimation to sunflecks.
 Beyond this photon flux density threshold, non-photochemical quenching rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a]. 
@@
 Leaf metabolism is strongly shaped by $T_{leaf}$, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
-However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical canopy gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical forest gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
 
@@
 
-** composite climatic stress variable from canopy temperature, vapour pressure deficit, and relative humidity is higher in lower canopy
+** composite climatic stress variable integrating temperature, vapour pressure deficit, and relative humidity is higher in lower canopy
 ```{r eval=FALSE, echo=FALSE}
@@
 Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
+The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than do shade leaves in lower vertical positions [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
 This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
@@
 Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], within-canopies of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
-In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
+In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the top of the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
 @hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 
@@
 Shade leaves that do not experience midday depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021]. 
-Evaluating this would require diurnal monitoring of net photosynthesis across a vertical profile through the canopy.
-In the meantime, it remains unresolved whether photosynthesis is more affected by high temperatures in canopy or understory leaves, and how this might vary across forest types and environmental conditions.
+Evaluating this would require diurnal monitoring of net photosynthesis across a forest vertical profile.
+In the meantime, it remains unresolved whether photosynthesis is more affected by high temperatures in upper canopy or understory leaves, and how this might vary across forest types and environmental conditions.
 
@@
 However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
-For example, *Quercus muehlenbergii* growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
-In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
-Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
-However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
+For example, *Quercus muehlenbergii* Engelm.
+growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx.
+growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+In an Australian savanna, crowns of *Acacia papyrocarpa* Benth.
+showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
+Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
+However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].
 The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
@@
 
-Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$ or $E_0$, Table `r table_leaf_metabolism`) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].
+Similar to photosynthesis, respiration tends to be higher in upper-canopy sun leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$ or $E_0$, Table `r table_leaf_metabolism`) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].
 Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs.
@@
 Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
-Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary through the canopy and influence leaf function will be important for resolving the role of VOCs  in forest thermal sensitivity across the vertical gradient.
+Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
 Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
@@
 Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
-This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height in the canopy (Table `r table_leaf_metabolism`). 
+This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height across the vertical profile (Table `r table_leaf_metabolism`). 
 More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce adjustment of traits related to carbon metabolism during drought [@bartholomewSmallTropicalForest2020].
@@
 Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
-In terms of C cycling, it has been estimated that canopy strata contribute $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007].
+In terms of C cycling, it has been estimated that the canopy strata contributes $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007].
 Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
@@
 
-The complex interwoven relations between the biophysical environment and biological factors -- leaf traits, metabolic processes, and distribution of species across canopy strata -- are likely to produce amplifications and feedback loops in a warming world, with implications for forests on many levels.
+The complex interwoven relations between the biophysical environment and biological factors -- leaf traits, metabolic processes, and distribution of species across forest strata -- are likely to produce amplifications and feedback loops in a warming world, with implications for forests on many levels.
 
@@
 What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
-The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adapation to the local climate but do not vary consistently with height in the canopy (see section 4.2, Table 2).
+The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adapation to the local climate but do not vary consistently with height in the vertical profile (see section 4.2, Table 2).
 Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-While leaves display substantial plasticity to adapt to warmer temperatures  [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and to recover from canopy heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+While leaves display substantial plasticity to adapt to warmer temperatures [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
 Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
@@
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
-Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020].
+Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the forest [@starkReframingTropicalSavannization2020].
 While wetter forests can prove quite resilient to such changes [@millerReducedImpactLogging2011], this shift to hotter and drier microclimates can often make forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018].
@@
 In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020b; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
-If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g.,  @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy buffering and the potential state changes described above.
+If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g.,  @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy structure and the potential state changes described above.
 
@@
 
-Owing to differences in the representation of canopy layers, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes.
+Owing to differences in the representation of forest vertical strata, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes. 
 In general, however, this variation is accounted for via light competition. 
-Models partition incoming radiation and radiation within the canopy (i.e., direct vs.
+Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
 diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple canopy layers can analytically solve the 2-stream approximation for each layer.
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer.
 Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].
@@
 Several recent model developments are at the cutting edge of representing vertical variation.
-Model developments by [Bonan et al., 2018; Chen et al., 2016; Longo et al., 2019] implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into canopy air space and the humidity of the airspace, modulated by turbulence processes within the canopy and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
+Model developments by [Bonan et al., 2018; Chen et al., 2016; Longo et al., 2019] implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into the vegetated air space and the humidity of the airspace, modulated by turbulence processes within the forest and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
 
 The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.
-Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
+Improved representation of vertical variation in forest structure and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
 Moreover, given the anticipated importance of mid-canopy and understory trees in ecosystem resilience to increasing mortality of canopy trees, it is absolutely essential that models separately represent these strata.
@@
 A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.
-Most models have been developed to capture dynamics in the canopy, or overstory, given the disproportionate role of these large trees in ecosystem dynamics (Fig. `r fig_ecosystem`). 
+Most models have been developed to capture dynamics in the canopy, given the disproportionate role of these large trees in ecosystem dynamics (Fig. `r fig_ecosystem`). 
 Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models. 
@@
 
-![**Figure `r fig_model`.** Lidar and thermal remote sensing data can be used to measure vertical canopy structure and plant thermal signatures at increasingly large scales.
+![**Figure `r fig_model`.** Lidar and thermal remote sensing data can be used to measure vertical forest structure and plant thermal signatures at increasingly large scales.
 (a-b) Integrating in situ and remote sensing measurements will support our ability to scale understanding of vertical gradients within and across ecosystems.

### 2022-02-08 vertical-thermal-review_SI.Rmd
standardize use of term "canopy"

@@
 Leaf area density (LAD, $m^2$ $m^{-3}$) was estimated by applying the MacArthur-Horn transformation [@macarthurFoliageProfileVertical1969] to the voxelized pulse returns.
-We excluded LAD data at the bottom 5 m of the canopy, which are more uncertain due to limited pulse penetration within this region furthest away from the sensor [@starkAmazonForestCarbon2012]; for calibration purposes, we assumed the bottom 5 m to have a leaf area index (LAI) of 0.25 $m^2$ $m^{-2}$ [@shaoHighThroughputAssessment2019].
+We excluded LAD data at the bottom 5 m of the vertical profile, which are more uncertain due to limited pulse penetration within this region furthest away from the sensor [@starkAmazonForestCarbon2012]; for calibration purposes, we assumed the bottom 5 m to have a leaf area index (LAI) of 0.25 $m^2$ $m^{-2}$ [@shaoHighThroughputAssessment2019].
 

### 2022-02-08 vertical-thermal-review.Rmd
overstory --> canopy
Just to simplify terminology

@@
 A great measure of uncertainty arises due to the differential impacts of rising temperatures on trees of differential size and crown position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].
-Forests are vertically and horizontally stratified, with overstory canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
+Forests are vertically and horizontally stratified, with canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
 This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
@@
 In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
-Early leaf-out allows some tree saplings to do the majority of their annual carbon fixation before the overstory closes and reduces light availability [@leeSpringPhenologicalEscape2021]. 
+Early leaf-out allows some tree saplings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
 Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
@@
 Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
-In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
+In addition, shading by the canopy can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
 Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020].

### 2022-02-08 response-to-reviews.Rmd
review/ reduce use of "buffer"/"buffering"/"buffered"

@@
 
-**We had also used the word “buffering” in different context, where the intention was to describe how canopy attenuation (of radiation and wind) can reduce daily extremes of temperature in the lower canopy and understory.
-We have modified the text in those places to clarify.
-For example,**
+**We had also used the words “buffering”, "buffer", or "buffered" in a different context, where the intention was to describe how canopy attenuation (of radiation and wind) can reduce daily extremes of temperature in the lower canopy and understory.
+We have modified the text in those places to clarify, generally reducing use of the terms.
+However, the term has been widely used in recent publications on this theme, and we therefore did not feel it was necessary to eliminate it's usage.
+Examples of modified sentences include the following:**
 

### 2022-02-08 vertical-thermal-review.Rmd
review/ reduce use of "buffer"/"buffering"/"buffered"

@@
 Scaling up from leaves to trees, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$, particularly under climate change.
-In contrast, understory trees benefit from a buffered microclimate but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when the buffering provided by large trees is lost.
+In contrast, understory trees experience fewer extreme high $T_{leaf}$'s but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when exposed to harsher conditions through canopy disturbance.
 As the climate changes, integrating the patterns and mechanisms reviewed here into models will be critical to forecasting forest-climate feedbacks.
@@
 In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig.
-`r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;  @bonan_ecological_2016].
+`r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020].
@@
 Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
-A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
+Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
 However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to shed excess heat may prove disadvantageous. 
@@
 While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), limited tree-ring evidence indicates that understory trees can exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
-Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the more buffered understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].
+Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].
 

### 2022-02-08 response-to-reviews.Rmd
R2 comments

@@
 
-**Thank you.**
+**Thank you for the review!**
 
@@
 
-**Thank you for the review!**
-
 ## Referee: 2
@@
 
-*I don't understand what needs to be done with the sentence*
+**We do not understand this comment, but have verified that this sentence is correct.**
 
@@
 
-*This is either a semantics issue, or a case where the reviewer is expecting greater realism that what we're actually doing.
-You should make that clear in the caption (also addresses comment above), and perhaps use "humid" and "dry" (rather than drought).*
-
-*Also note that well-watered would refer to soil moisture, and require modeling water transport through roots and stem.
-Drought could refer to either soil dryness or atmospheric dryness.
-In my mind, it's always been the latter.*
-
-**changed to "drought" to "dry" in the figure to refer to atmospheric dryness.**
+**We changed to "humid" to "moist" (as opposed to "well-watered"). We chose the term "moist" because it conveys both high soil moisture and atmospheric humidity, which is what we aim to capture.**
 

### 2022-02-08 vertical-thermal-review.Rmd
R2 comments

@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) characteristic leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both humid and dry conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) characteristic leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.

### 2022-02-08 response-to-reviews.Rmd
finished reviewing response to reviews

@@
 
-- ...
+- Revised three figures (Figs. 1-3) based on reviewer comments
 
@@
 **"Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
-The profile shape of this light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types (Fig.
+The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types (Fig.
 2a-d, Supporting Information Figure S1, Koike et al., 2001)."**
@@
 
+*https://github.com/EcoClimLab/vertical-thermal-review/issues/84*
+
 Table 1: What is here is not leaf gas exchange, and it seems that the authors have repeated the caption for Table 2 in error. Please fix this. Also don’t capitalise ‘leaf’.
@@
 **Done**
-*Need to add the updated table to .rmd*
 
@@
 I would suggest that this would be improved if the authors started from these old studies that most readers will recognise, and state how we’ve advanced in understanding and progressed toward within-canopy work rather than between sun-and-shade habitats.
+
+*(see Lawren's comments in google doc)*
 
@@
 
-**We have added four references showing that stomatal conductance is often limited by imbalance of water supply and evaporative demand for leaves in more elevated or more sunlit canopy positions [@yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010].**
+**We have added references showing that stomatal conductance is often limited by imbalance of water supply and evaporative demand for leaves in more elevated or more sunlit canopy positions [@yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010].**
 
@@
 
-**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text underlined): “However, ~~because~~ water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al.
+**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al.
 1994, Koch et al.

### 2022-02-09 vertical-thermal-review.Rmd
shaved a few words off the abstract
203--> 200 words.

@@
 
-```{r eval=FALSE, echo=FALSE}
-#  200 word limit, no bullet points  
-```
+<!--  200 word limit, no bullet points-->
 
@@
 Using published research and new analyses, we evaluate how microclimate and leaf temperatures, traits, and gas exchange vary vertically in forests, shaping tree ecology and ecosystem function. 
-In forests with dense canopies, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
+In closed-canopy forests, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
 However, foliar traits also vary across height or light gradients, partially mitigating the elevation of upper-canopy $T_{leaf}$.
 Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity and damage thresholds across the gradient appear modest.
-Scaling up from leaves to trees, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$, particularly under climate change.
+Scaling from leaves to trees, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$, particularly under climate change.
 In contrast, understory trees experience fewer extreme high $T_{leaf}$'s but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when exposed to harsher conditions through canopy disturbance.

### 2022-02-09 vertical-thermal-review.Rmd

@@
 Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the forest [@starkReframingTropicalSavannization2020].
-While wetter forests can prove quite resilient to such changes [@millerReducedImpactLogging2011], this shift to hotter and drier microclimates can often make forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018].
+Such changes often result in increased growth of smaller trees, which benefit from increased light [@hoganDroughtInterannualVariability2019;@bennettLargerTreesSuffer2015], and wetter forests can prove quite resilient to canopy disturbance [@millerReducedImpactLogging2011].
+However, this shift to hotter and drier microclimates makes some forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018].
 Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].

### 2022-02-10 vertical-thermal-review.Rmd
small updates

@@
 Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].
-Leaf turnover is faster in the upper canopy than the understory, putatively due to more rapid accumulation of leaf damage [@osadaLeafDynamicsMaintenance2001; @millerOnlySunlitLeaves2021].
+Leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
 Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; @mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015].

### 2022-02-10 response-to-reviews.Rmd
Update response-to-reviews.Rmd
added a response

@@
 
+**Thank you for the comment. We included a section that explains how vertical gradients are represented in models:**
+
+- **Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
+diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer.
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**
+
+**Given that models vary greatly in how they capture vertical light gradients, what biophysical variables are included and how those variables are parameterized to vary (or not) across vertical gradients, we feel that it is beyond the scope of this paper to comprehensively list or explain which of those variables vary vertically across different models.**
+
 *[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
@@
 
-**The units in the table are in %nm i.e. % wavelength**
+**Done**
 

### 2022-02-10 response-to-reviews.Rmd
yes, I think this is correct
reply to *Nidhi, is this all correct??.*
updated another response

@@
 
-*Nidhi, is this all correct??.*
-
 L219, repetition with L 216
@@
 **We appreciate this comment, and have added mention of the fact that maximum $g_s$ increases with light in the canopy because it typically tracks photosynthetic capacity, per @wongStomatalConductanceCorrelates1979.
-There is a great deal of literature showing that the relative limitation of photosynthesis by stomata (gs/A, or intrinsic WUE) tends to be greater in the upper canopy, which results in lower intercellular CO~2~ and C isotope discrimination, as noted in the original manuscript (lines xxx).**
+There is a great deal of literature showing that the relative limitation of photosynthesis by stomata (gs/A, or intrinsic WUE) tends to be greater in the upper canopy, which results in lower intercellular CO~2~ and C isotope discrimination, as noted in the original manuscript (lines 478).**
 
@@
 
-**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax).
-See lines X-X**
-*Need to add these lines: "Owing to difference...".*
+**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax).**
+
+- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes.
+In general, however, this variation is accounted for via light competition.
+Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
+diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer.
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**
 

### 2022-02-10 response-to-reviews.Rmd
Thanks Krista, yes, I agree. I think I tried adding vertical gradient but the terms I used there gave the best and most robust results
*(Nidhi, what do you think of the following?)*

@@
 
-**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax).**
+**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax):**
 
@@
 **We agree that there is a recognition in the modeling community that we do need improved understanding of the mechanisms to better represent the level of vertical variation desired by some.
-We added the following sentence to acknowledge that (LXX-XX): " ...
-".
+We added the following sentence to acknowledge that: "However, doing so will require improved understanding of the mechanisms controlling vertical gradients.".
 We also removed some of the language that emphasizes cohort based models as the only or best way to do this.**
@@
 Table S3: Including the words vertical or vertical gradient could have been useful.
-
-*(Nidhi, what do you think of the following?)*
 

### 2022-02-10 response-to-reviews.Rmd

@@
 
-- **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line xxx)***(new text underlined)**
+- **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line xxx)***(new text italicized)**
 - **“Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances, ***attenuation of radiation and vertical transport of sensible heat by the canopy can buffer the lower canopy and understory from large diurnal swings in air temperature***~~it can be significantly buffered by forest canopies~~"; (line xxx)**
-- **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum Tair more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime Tair and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line xxx)**
+- **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line xxx)**
 
@@
 
-- **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, that are predominately within species between sun and shade leaves of the same trees reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes**
+- **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, which  compare sun and shade leaves of the same trees, reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes**
 

### 2022-02-10 vertical-thermal-review.Rmd

@@
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
-This dichotomy in leaves is observed along the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions [Fig.
+Therefore, a continuum of sun and shade leaves are observed throughout the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions [Fig.
 `r fig_NEON_vertical`, @binLeafTraitExpression2022], and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].

### 2022-02-11 vertical-thermal-review.Rmd
Included changes from KAT.doc into .rmd
new version of the manuscript with structuring

@@
 
-### 3.1. Intraspecific trait variation
+Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022].
+However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
+Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
 
@@
 
+## 3.1 Leaf anatomy and morphology
+
 Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`).
@@
 
+## 3.2 Metabolic capacity and efficiency
+
 Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
@@
 Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
+
+## 3.3 – Light absorption and reflectance
+
+## 3.#. Biochemical protection against foliage light and heat damage
 
@@
 
+## 3.# Thermal tolerance
+
+At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
+Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
+Thermal thresholds for photosystem II functioning serve as good proxies for temperatures at which leaves get necrotic [@krauseHightemperatureToleranceTropical2010].
+Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021].
+Thermal tolerances vary globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
+However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
+For example, *Quercus muehlenbergii* Engelm.
+growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx.
+growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+In an Australian savanna, crowns of *Acacia papyrocarpa* Benth.
+showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
+Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
+However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].
+The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
+
+"Thermal thresholds for leaves have also been hypothesized to vary with leaf temperature, and thus be higher in sun-exposed upper canopy leaves than in understory leaves.
+The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II (PSII) functioning using chlorophyll fluorescence techniques (Krause et al.
+2012).
+Considering sun and shade leaf differences within canopy (Table 1), the temperature at which PSII efficiency declines, Tcrit, and at which it has decreased by 50%, T50, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama (Slot et al., 2019).
+In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019).
+These patterns of higher heat tolerance in the warmest leaves are consistent with interspecific patterns across temperature gradients (O’Sullivan et al.
+2018; Feeley et al., 2020; Slot et al., 2021a), and experienced maximum temperatures (Perez & Feeley, 2020).
+However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences [sorry, bit of a blabla sentence to wrap up]."
+
+## 3.# Leaf phenology
+
 The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
@@
 Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of species that are deciduous (Condit et al., 2000; Meakem et al., 2018).
 Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].
@@
 Leaf phenology thereby both responds to and influences the vertical structure of forest microenvironments and function.
-
-### 3.2. Interspecific trait variation
-
-Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
-Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022].
-For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [mirroring the faction of individuals, @conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
-However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
-Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
 
@@
 In the meantime, it remains unresolved whether photosynthesis is more affected by high temperatures in upper canopy or understory leaves, and how this might vary across forest types and environmental conditions.
-
-At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
-Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
-Thermal thresholds for photosystem II functioning serve as good proxies for temperatures at which leaves get necrotic [@krauseHightemperatureToleranceTropical2010].
-Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021].
-Thermal tolerances vary globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
-However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
-For example, *Quercus muehlenbergii* Engelm.
-growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx.
-growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
-In an Australian savanna, crowns of *Acacia papyrocarpa* Benth.
-showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
-Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
-However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].
-The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
 

### 2022-02-11 vertical-thermal-review.Rmd
Ty's edits

@@
 Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`).
-Carotenoids (e.g., beta carotene and lutein), including xanthophyll cycle pigments (i.e., violaxanthin, antheraxanthin and zeaxanthin, VAZ), can play a role in antioxidant scavenging and harmlessly dissipating excess excitation energy as heat to reduce photoinhibition, and therefore also serve as thermoprotection and postpone heat damage [@niinemetsPhotosynthesisResourceDistribution2007; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998].
-With saturating irradiance experienced by upper canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching[@mathurPhotosyntheticEfficiencySun2018b].
-Efficient light-use is optimized through this process while preventing the accumulation of excess light, an energetic imbalance that results in the generation of harmful reactive oxygen species. 
-However, below a photon flux density of $200\mu molm^{-2}s^{-1}$, within a beech (*Fagus sylvatica L.*) forest, shaded lower canopy leaves have been observed to have higher non-photochemical quenching rates than upper canopy leaves, reflecting rapid photosynthetic induction kinetics in shaded leaves as an acclimation to sunflecks.
-Beyond this photon flux density threshold, non-photochemical quenching rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a]. 
-The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].
+Heat and associated VPD can indirectly impact the photosynthetic machinery by reducing the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
+Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species (ROS) [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
+A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
+This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
+Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
+Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10˚C) during sunflecks [@waySunflecksTreesForests2012].
+In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
 

### 2022-02-11 response-to-reviews.Rmd
responding to line404-405

@@
 
-- **Conclusions section, first par (text specifically cited in the comment above): "Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth.
+- **Conclusions section, first part (text specifically cited in the comment above): "Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth.
 While the preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
@@
 **Thank you for pointing this out.
-In line404-405, we are concentrated on interactions between microenvironments and function.
+In line 404-405, we are concentrated on interactions between microenvironments and function.
 We edited the text to emphasize the mechanisms for earlier bud break in the understory, including warmer temperatures, and the benefits of maximizing carbon gain before overstory closure reduces light availability [added ref.

### 2022-02-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of species that are deciduous (Condit et al., 2000; Meakem et al., 2018).
+Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
 Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].

### 2022-02-11 response-to-reviews.Rmd
NPQ section

@@
 
-**We separated VAZ section from VOC, and provided further context to NPQ in the following lines:**
+**We separated VAZ section from VOC, and provided further context to NPQ in the following paragraph:**
 
-- **With saturating irradiance in canopy leaves, unused energy from photosystem II antennae is further dissipated as heat energy through non-photochemical quenching [@mathurPhotosyntheticEfficiencySun2018b].
-Efficient light-use is optimized through this process while preventing the accumulation of excess light, provoking the over production of harmful reactive oxygen species.
-However, at a photon flux density of $200\mu molm^{-2}s^{-1}$, within beech canopy, shaded lower canopy leaves have been observed to have higher non-photochemical quenching rates than upper canopy reflecting rapid photosynthetic kinetics in shaded leaves as an acclimation to sunflecks.
-However, beyond this photon flux density threshold, non-photochemical quenching rates of upper canopy increase with an increase in light intensity [@scartazzaInvestigatingEuropeanBeech2016a].
-The dissipation of excess light energy is also important for leaf acclimation to higher temperatures if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].**
+- **Heat and associated VPD can indirectly impact the photosynthetic machinery by reducing the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
+Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species (ROS) [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
+A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
+This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
+Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
+Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table 1, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10˚C) during sunflecks [@waySunflecksTreesForests2012].
+In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
 

### 2022-02-11 vertical-thermal-review.Rmd
working on Tcrit paragraph

@@
 
+Thermal thresholds for leaves have also been hypothesized to vary with leaf temperature, and thus be higher in sun-exposed upper canopy leaves than in understory leaves.
+The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II (PSII) functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010].
+Considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), the temperature at which PSII efficiency declines, $T_{crit}$, and at which it has decreased by 50%, $T_{50}$, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
 At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
@@
 
-"Thermal thresholds for leaves have also been hypothesized to vary with leaf temperature, and thus be higher in sun-exposed upper canopy leaves than in understory leaves.
-The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II (PSII) functioning using chlorophyll fluorescence techniques (Krause et al.
-2012).
-Considering sun and shade leaf differences within canopy (Table 1), the temperature at which PSII efficiency declines, Tcrit, and at which it has decreased by 50%, T50, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama (Slot et al., 2019).
-In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019).
+ "In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019).
 These patterns of higher heat tolerance in the warmest leaves are consistent with interspecific patterns across temperature gradients (O’Sullivan et al.

### 2022-02-11 vertical-thermal-review.Rmd
merged two paragraphs in introduction to traits section

@@
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
-However, when it comes to understanding and modeling forest ecosystem function, given that species traits strongly shape the propensity for trees to inhabit understory vs.
-canopy positions, ecosystem function is also shaped by interspecific trait variation [e.g., @cavaleri_height_2010].
-
-Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes.
-Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022].
-However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
-Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010]; however, counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
+Overall, when it comes to understanding and modeling forest ecosystem function, many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010].
 

### 2022-02-11 vertical-thermal-review.Rmd
switched paragraph order
Update vertical-thermal-review.Rmd
add Lloyd 2010 references

@@
 
-Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
-The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
-Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010]; however, counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
-Overall, when it comes to understanding and modeling forest ecosystem function, many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010].
-
 Leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
@@
 Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
+
+Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
+The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloyd_optimisation_2010]; however, vertical trends vary across trees [@lloyd_optimisation_2010], and counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
+Overall, when it comes to understanding and modeling forest ecosystem function, many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010; @lloyd_optimisation_2010].
 

### 2022-02-11 response-to-reviews.Rmd
working on Tcrit paragraph
Update response-to-reviews.Rmd

@@
 
-- We have brought in an additional author, Thomas Buckley, to round out the expertise of the author team and help address some of the reviewer comments (particularly R2).
+- We have brought in an additional author, Thomas Buckley, to round out the expertise of the author team and help address some of the reviewer comments (particularly Reviewer 2).
 
-- Revised three figures (Figs. 1-3) based on reviewer comments
+- We did some modest reorganization, including moving the paragraph on $T_{crit}$ to the traits section (suggested by Reviewer 2) and reorganizing the traits section around trait groupings, as opposed to intra-specific/ interspecific.
+This rearrangement of the traits section better parallels Table 1 and the following section on leaf gas exchange.
+
+- We revised three figures (Figs. 1-3) based on reviewer comments
 
@@
 
-*[issue #99](https://github.com/EcoClimLab/vertical-thermal-review/issues/99)*
+**We ended up removing former section 3.2 because we reorganized the traits section around trait groupings, as opposed to intra-specific/ interspecific.
+This rearrangement of the traits section better parallels Table 1 and the following section on leaf gas exchange.
+We integrated the content that was previously in this section throughout the traits section, as appropriate.**
+
+**Unfortunately, we've been unable to find many studies examining vertical gradients at the interspecific level, and have included those of which we are aware.**
 
@@
 
-*[issue #99](https://github.com/EcoClimLab/vertical-thermal-review/issues/99)*
+**We ended up removing former section 3.2 because we reorganized the traits section around trait groupings, as opposed to intra-specific/ interspecific.
+This rearrangement of the traits section better parallels Table 1 and the following section on leaf gas exchange.
+We integrated the content that was previously in this section throughout the traits section, as appropriate.**
+
+**Unfortunately, we've been unable to find many studies examining vertical gradients at the interspecific level, and have included those of which we are aware.**
+
+**Thanks for pointing out the Lloyd reference, which we had previously missed. We have added citations to this (e.g., *2nd* paragraph of the traits section), and have also added it to our tables).**
+
+*https://github.com/EcoClimLab/vertical-thermal-review/issues/101*
 
@@
 In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
+
+**We also moved thermal sensitivity paragraph to the earlier part of the manuscript, the paragph after the discussion on VOC.**
 

### 2022-02-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
#102

@@
 
-Leaf traits shape $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].
+Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].
+The strongest driver of this variation is the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022].
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
@@
 `r fig_NEON_vertical`, @binLeafTraitExpression2022], and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
-
-Leaf traits vary along the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020].
-Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.
-It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy. 
+Yet, independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy. 
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
 Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
+Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.
 

### 2022-02-11 vertical-thermal-review.Rmd
small adjustments to first traits paragraph
#102

@@
 Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].
-The strongest driver of this variation is the vertical light gradient [*i.e.,* sun and shade leaves, @casasSunShadeLeaves2011; @binLeafTraitExpression2022].
+The primary driver of this variation is the vertical light gradient, with strong differentiation between sun and shade leaves [@casasSunShadeLeaves2011; @binLeafTraitExpression2022].
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
@@
 Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
-Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], so further research is needed to characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.
+Yet, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], and further research is needed to fully characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.
 

### 2022-02-11 vertical-thermal-review.Rmd
updating lloyd citation

--- 2022-02-11 vertical-thermal-review.Rmd
@@
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
-Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloyd_optimisation_2010]; however, vertical trends vary across trees [@lloyd_optimisation_2010], and counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
-Overall, when it comes to understanding and modeling forest ecosystem function, many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010; @lloyd_optimisation_2010].
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010]; however, vertical trends vary across trees [@lloydOptimisationPhotosyntheticCarbon2010], and counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
+Overall, when it comes to understanding and modeling forest ecosystem function, many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010].
 

### 2022-02-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
starting to edit thermal tolerance section: insert new line per sentence

@@
 Brooks et al.
-1997
+1997; 52.
+Curtis et al.
+2019; 53.
+Slot et al.
+2019; 54.
+Hamerlynck and Knapp 1994.
 ```{r eval=FALSE, echo=FALSE}
@@
 
- "In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019).
+In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019).
 These patterns of higher heat tolerance in the warmest leaves are consistent with interspecific patterns across temperature gradients (O’Sullivan et al.
 2018; Feeley et al., 2020; Slot et al., 2021a), and experienced maximum temperatures (Perez & Feeley, 2020).
-However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences [sorry, bit of a blabla sentence to wrap up]."
+However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.
 
@@
 2021; 31.
-Curtis et al.
-2019; 32.
 Mier et al.
-2001; 33.
+2001; 32.
 Turnbull et al.
-2003; 34.
+2003; 33.
 Araki et al.
-2017; 35.
+2017; 34.
 Bolstad et al.
-1999; 36.
+1999; 35.
 Kenzo et al.
-2015; 37.
+2015; 36.
 Harley et al.
-1996; 38.
-Xu and Griffin 2006; 39.
+1996; 37.
+Xu and Griffin 2006; 38.
 Atherton et al.
-2017; 40.
+2017; 39.
 Carter et al.
-2021; 41.
+2021; 40.
 Sack et al.
-2003; 42.
+2003; 41.
 Taylor et al.
-2021; 43.
+2021; 42.
 Harley et al.
-1997; 44.
-Niinemets and Sun, 2014; 45.
-Sharkey and Monson, 2014; 46.
+1997; 43.
+Niinemets and Sun, 2014; 44.
+Sharkey and Monson, 2014; 45.
 Simpraga et al.
-2013
+2013; 46.
+Slot et al.
+2021
 

### 2022-02-11 vertical-thermal-review.Rmd
starting to rearrange thermal tolerance section

@@
 
-Thermal thresholds for leaves have also been hypothesized to vary with leaf temperature, and thus be higher in sun-exposed upper canopy leaves than in understory leaves. 
+<!--short par giving background on thermal tolerance-->
 The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II (PSII) functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010].
-Considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), the temperature at which PSII efficiency declines, $T_{crit}$, and at which it has decreased by 50%, $T_{50}$, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
 At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
-Thermal thresholds for photosystem II functioning serve as good proxies for temperatures at which leaves get necrotic [@krauseHightemperatureToleranceTropical2010].
+
+Thermal thresholds for leaves have been hypothesized to vary with leaf temperature, and thus to be higher in sun-exposed upper canopy leaves than in understory leaves.
+
+Considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), the temperature at which PSII efficiency declines, $T_{crit}$, and at which it has decreased by 50%, $T_{50}$, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
+
 Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021].

### 2022-02-11 vertical-thermal-review.Rmd
worked on thermal tolerance section (still in progress)

@@
 <!--short par giving background on thermal tolerance-->
-The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II (PSII) functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010].
+The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
 At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
-
-Thermal thresholds for leaves have been hypothesized to vary with leaf temperature, and thus to be higher in sun-exposed upper canopy leaves than in understory leaves.
-
-Considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), the temperature at which PSII efficiency declines, $T_{crit}$, and at which it has decreased by 50%, $T_{50}$, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019].
-
-Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021].
-Thermal tolerances vary globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
+Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->.
+Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020].
+Thermal tolerances globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
 However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
@@
 growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+
+Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be higher in sun-exposed upper canopy leaves than in understory leaves. 
 In an Australian savanna, crowns of *Acacia papyrocarpa* Benth.
 showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
-Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
 However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].

### 2022-02-11 vertical-thermal-review.Rmd
finished revising thermal tolerances section

@@
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
-Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->.
-Across species, sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020].
-Thermal tolerances globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
+Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and with higher thermal tolerance among leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
+$T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
 However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
-For example, *Quercus muehlenbergii* Engelm.
+<!--For example, *Quercus muehlenbergii* Engelm.
 growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx.
 growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+(removed because I had trouble finding this in the paper, plus it's not really needed, but feel free to reinstate) -->
 
+<!--par on thermal tolerance across vertical gradients-->
 Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be higher in sun-exposed upper canopy leaves than in understory leaves. 
-In an Australian savanna, crowns of *Acacia papyrocarpa* Benth.
+Indeed, along a vertical gradient within a very open canopy in Australia, crowns of *Acacia papyrocarpa* Benth.
 showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
 However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].
-The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
-
-In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019).
-These patterns of higher heat tolerance in the warmest leaves are consistent with interspecific patterns across temperature gradients (O’Sullivan et al.
-2018; Feeley et al., 2020; Slot et al., 2021a), and experienced maximum temperatures (Perez & Feeley, 2020).
+This small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
 However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.

### 2022-02-11 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 **Thanks for pointing out the Lloyd reference, which we had previously missed. We have added citations to this (e.g., *2nd* paragraph of the traits section), and have also added it to our tables).**
-
-*https://github.com/EcoClimLab/vertical-thermal-review/issues/101*
 
@@
 
-*[issue #92](https://github.com/EcoClimLab/vertical-thermal-review/issues/92)*
-
-**We have revised the first sentence to read: "Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.
+**We have revised the first sentence here to read: "Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.
 We also choose to keep this rather obvious sentence because we feel that basic topic sentences are important for helping to guide readers from a diversity of sub-disciplines through the manuscript.**
@@
 
-**We also moved thermal sensitivity paragraph to the earlier part of the manuscript, the paragph after the discussion on VOC.**
+**We also moved thermal sensitivity content to the traits section (now with its own sub-heading).**
 
@@
 
-*https://github.com/EcoClimLab/vertical-thermal-review/issues/84*
+**We have added points along each line to indicate the constants applied in the other modeling scenarios. These dots represent our best estimates of "typical" values.**
+
+*(this remains to be done: https://github.com/EcoClimLab/vertical-thermal-review/issues/84)*
 

### 2022-02-11 vertical-thermal-review.Rmd
tweak abstract to match new organization

@@
 In closed-canopy forests, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
-However, foliar traits also vary across height or light gradients, partially mitigating the elevation of upper-canopy $T_{leaf}$.
-Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity and damage thresholds across the gradient appear modest.
+However, foliar traits also vary across height or light gradients, partially mitigating and protecting against the elevation of upper-canopy $T_{leaf}$.
+Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity across the gradient appear modest.
 Scaling from leaves to trees, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$, particularly under climate change.

### 2022-02-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
NPQ section update

@@
 2019; 54.
-Hamerlynck and Knapp 1994.
+Hamerlynck and Knapp 1994; 55.
+Lloyd et al.
+2010; 56.
+Augspurger and Bartlett, 2003; 57.
+Osada et al.
+2001; 58.
+Meakem et al.
+2018; 59.
+Condit et al.
+2000
 ```{r eval=FALSE, echo=FALSE}
@@
 Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`).
-Heat and associated VPD can indirectly impact the photosynthetic machinery by reducing the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
+By reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
 Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species (ROS) [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].

### 2022-02-11 vertical-thermal-review.Rmd
proofreading/ small edits

@@
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
-The implications of this light gradient are that leaf traits vary along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
+The implications of this light gradient are that leaf traits vary along the vertical gradient (Table 1), and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
 
@@
 `r fig_leaf_T`c), increases with leaf size (Fig.
-`r fig_leaf_T`d), and decreases absolutely with $g_s$ (Fig.
+`r fig_leaf_T`d), and $T_{leaf}$ decreases absolutely with $g_s$ (Fig.
 `r fig_leaf_T`e).
@@
 Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) characteristic leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
-In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
+In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values illustrated by dots and given in the table of biophysical constants.
 Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), characteristic leaf width (clw), stomatal conductance (gs), air temperature (tair).
@@
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipate excess heat compared to upper-canopy leaves.
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves.
 Limitations in heat dissipation, in combination with their physiological propensities, may cause shade leaves to experience above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
@@
 The latter can result from a combination of still air at lower heights (Fig.
-`r fig_NEON_vertical`e), and the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the Tleaf of the foliage close to the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
+`r fig_NEON_vertical`e), and sunflecks or the 'canopy greenhouse effect', as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the $T_{leaf}$ of the foliage close to the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
 One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
@@
 Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
-Therefore, a continuum of sun and shade leaves are observed throughout the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions [Fig.
-`r fig_NEON_vertical`, @binLeafTraitExpression2022], and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
-Yet, independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy. 
+Therefore, a continuum of sun and shade leaves are observed throughout the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions (Fig.
+`r fig_NEON_vertical`), and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
+Traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy, independent of the light environment. 
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
@@
 By reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
-Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species (ROS) [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
+Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
 A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].

### 2022-02-11 vertical-thermal-review.Rmd
incorporating some of R1's citations as necessary

@@
 nocite: |
-  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010; @brooksVerticalGradientsPhotosynthetic1997a
+  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010; @brooksVerticalGradientsPhotosynthetic1997a; @carswellPhotosyntheticCapacityCentral2000
 ---

### 2022-02-11 vertical-thermal-review.Rmd
adding R1's citation
small changes

@@
 `r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
-Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkReframingTropicalSavannization2020; @hanberryOpenForestEcosystems2020].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019].
+Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017].
 In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.
@@
 The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
-`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022].
+`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022; @belandVerticalStructureHeterogeneity2021].
 Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers.
@@
 `r fig_NEON_vertical`d), than in forests with more open upper canopies, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
-`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 The implications of this light gradient are that leaf traits vary along the vertical gradient (Table 1), and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
@@
 What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
-The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adapation to the local climate but do not vary consistently with height in the vertical profile (see section 4.2, Table 2).
+The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to the local climate but do not vary consistently with height in the vertical profile (see section 4.2, Table 2).
 Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
@@
 
-As discussed above, rising temperatures and severe droughts place the canopy trees at particularly elevated risk of mortality. 
+As discussed above, rising temperatures and severe droughts place canopy trees at particularly elevated risk of mortality. 
 Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
@@
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
-Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the forest [@starkReframingTropicalSavannization2020].
+Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the forest [@juckerCanopyStructureTopography2018; @starkReframingTropicalSavannization2020].
 Such changes often result in increased growth of smaller trees, which benefit from increased light [@hoganDroughtInterannualVariability2019;@bennettLargerTreesSuffer2015], and wetter forests can prove quite resilient to canopy disturbance [@millerReducedImpactLogging2011].
@@
 Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but has dramatically reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @starkReframingTropicalSavannization2020; @mcdowellPervasiveShiftsForest2020].
+Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may have reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020].
 
-Canopy disturbance poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013a; @greiserHidingClimateCharacterizing2019].
+Canopy disturbance poses an increasing threat to the biodiversity of understory species that are otherwise buffered from climatic extremes [@scheffersMicrohabitatsReduceAnimal2013a; @greiserHidingClimateCharacterizing2019].
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
@@
 Several recent model developments are at the cutting edge of representing vertical variation.
-Model developments by [Bonan et al., 2018; Chen et al., 2016; Longo et al., 2019] implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into the vegetated air space and the humidity of the airspace, modulated by turbulence processes within the forest and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
+Model developments by [**Bonan et al., 2018; Chen et al., 2016; Longo et al., 2019**] implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into the vegetated air space and the humidity of the airspace, modulated by turbulence processes within the forest and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
 
@@
 Ensuant increased mortality of canopy trees will create an increasing number of gaps, resulting in changes to understory conditions and community composition.
-Ultimately, mid- and understory trees communities will be critical to the resilience of forest ecosystems under changing climate, making improved understanding and model representation of their dynamics critical to understanding future forest dynamics.
+Ultimately, mid- and understory tree communities will be critical to the resilience of forest ecosystems under changing climate, making improved understanding and model representation of their dynamics critical to understanding future forest dynamics.
 Integrating the patterns and mechanisms reviewed here, along with remote sensing data on forest structure and thermal environments, into cohort-based models that integrate with Earth system models will be critical to understanding and forecasting forest-climate feedbacks in the coming decades.

### 2022-02-11 response-to-reviews.Rmd
I have incorporated some of these citations, not all
because some of them are repetative in terms of traits
co2 section edits

@@
 
-**Thank you for the references. We have reviewed these and incorporated as appropriate.** *(Nidhi, assuing this is true?)*
+**Thank you for the references. We have reviewed these and incorporated as appropriate.**
 
@@
 **Thank you for this suggestion, we added the following information:**
-**"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~, where both tend to be higher in the lower canopy positions.
+**"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~, where both tend to be higher in the lower canopy positions.
 Particularly, understory seedlings above ground benefit from elevated CO~2~, therefore showing highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997a]."**

### 2022-02-11 vertical-thermal-review.Rmd
co2 section edits

@@
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
+Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
 Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 

### 2022-02-11 response-to-reviews.Rmd
small npq edits

@@
 
-- **Heat and associated VPD can indirectly impact the photosynthetic machinery by reducing the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
+- **Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
 Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species (ROS) [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
@@
 **"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~, where both tend to be higher in the lower canopy positions.
-Particularly, understory seedlings above ground benefit from elevated CO~2~, therefore showing highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997a]."**
+Particularly, understory seedlings above ground benefit from elevated CO~2~, therefore showing highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]."**
 

### 2022-02-11 vertical-thermal-review.Rmd
small npq edits

@@
 Thus, sun leaves are generally smaller and thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
-`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017; @mathurPhotosyntheticEfficiencySun2018a], but with greater leaf packing and clumping (which reduces $g_b$).
+`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces $g_b$).
 Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
@@
 Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-By reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
+Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
 Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].

### 2022-02-11 vertical-thermal-review.Rmd
marielle's citation

@@
 Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may have reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020].
+Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may have reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020; @floresFloodplainsAchillesHeel2017].
 

### 2022-02-12 response-to-reviews.Rmd
move isoprenes to traits
#102

@@
 
-- We did some modest reorganization, including moving the paragraph on $T_{crit}$ to the traits section (suggested by Reviewer 2) and reorganizing the traits section around trait groupings, as opposed to intra-specific/ interspecific.
+- We did some modest reorganization, including moving the content on $T_{crit}$ and VOCs to the traits section (the former suggested by Reviewer 2) and reorganizing the traits section around trait groupings, as opposed to intra-specific/ interspecific.
 This rearrangement of the traits section better parallels Table 1 and the following section on leaf gas exchange.

### 2022-02-12 vertical-thermal-review.Rmd
move isoprenes to traits
#102

@@
 (2) leaf temperature ($T_{leaf}$);
-(3) the leaf traits that most strongly influence $T_{leaf}$;
-(4) leaf gas exchange (including stomatal and boundary layer conductance, photosynthesis, respiration, and volatile organic compound emission) and its thermal sensitivity; and 
+(3) the leaf traits that most strongly influence $T_{leaf}$ and thermal tolerance;
+(4) leaf gas exchange (including stomatal and boundary layer conductance, photosynthesis, and respiration) and its thermal sensitivity; and 
 (5) tree and ecosystem ecology.
@@
 
+<!--VOCs-->
 Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
-Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [see section 4.4, @taylorNewFieldInstrument2021].
+Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
+
+Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
+In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
+Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
+Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
+However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_metabolism`, @taylorNewFieldInstrument2021].
+Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_metabolism`, @simpragaVerticalCanopyGradient2013].
+This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
+Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
+Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
 
@@
 Overall, however, we currently lack synthetic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
-
-### 4.4. VOC emissions
-
-Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
-In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
-Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
-Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
-However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_metabolism`, @taylorNewFieldInstrument2021].
-Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_metabolism`, @simpragaVerticalCanopyGradient2013].
-This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
-Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
-Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
-Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
 

### 2022-02-12 vertical-thermal-review.Rmd
form 2 new VOC paragraphs

@@
 Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10˚C) during sunflecks [@waySunflecksTreesForests2012].
+In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10$^\circ$C) during sunflecks [@waySunflecksTreesForests2012].
 In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
@@
 Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
-Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
-
 Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
@@
 Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
+
+Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
 Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].

### 2022-02-12 vertical-thermal-review.Rmd
wordsmithing VOC paragraphs
#102

@@
 Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
-Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005].
 In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
@@
 
-Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
+Less than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile is influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
 Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
-However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_metabolism`, @taylorNewFieldInstrument2021].
-Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_metabolism`, @simpragaVerticalCanopyGradient2013].
+However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_traits`, @taylorNewFieldInstrument2021].
+Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_traits`, @simpragaVerticalCanopyGradient2013].
 This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 

### 2022-02-12 vertical-thermal-review.Rmd
updated the numbering for tables in .rmd

@@
 nocite: |
-  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010; @brooksVerticalGradientsPhotosynthetic1997a; @carswellPhotosyntheticCapacityCentral2000
+  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010; @brooksVerticalGradientsPhotosynthetic1997; @carswellPhotosyntheticCapacityCentral2000
 ---
@@
 $\ddagger$ 1.
+Kafuti et al.
+2020; 2.
+Van Wittenberghe et al.
+2012; 3.
+Roberts et al.
+1990; 4.
+Dang et al.
+1997; 5.
+Marenco et al.
+2017; 6.
+Ambrose et al.
+2015; 7.
+Zweifel et al.
+2001; 8.
+Slot et al.
+2019; 9.
+Hernandez et al.
+2020; 10.
+Urban et al.
+2007; 11.
+Carter and Cavaleri 2018; 12.
+Martin et al.
+1999; 13.
 Mau et al.
-2018; 2.
-Coble and Cavaleri 2014; 3.
+2018;  14.
+Kosugi et al.
+2012; 15.
+Niinemets et al.
+2015; 16.
+Bachofen et al.
+2020; 17.
+Hamerlynck and Knapp 1994; 18.
+Coble et al.
+2017; 19.
+Wyka et al.
+2012; 20.
+Rijkerse et al.
+2000; 21.
+Ishida et al.
+1999; 22.
+Weerasinghe et al.
+2014; 23.
+Scartazza et al.
+2016; 24.
+Miller et al.
+2021; 25.
+Harris and Medina 2013; 26.
+Legner et al.
+2014; 27.
+Kitao et al.
+2012; 28.
+Fauset et al.
+2018; 29.
+Rey-Sanchez et al.
+2016; 30.
+Muller et al.
+2021; 31.
+Mier et al.
+2001; 32.
+Turnbull et al.
+2003; 33.
+Araki et al.
+2017; 34.
+Bolstad et al.
+1999; 35.
+Kenzo et al.
+2015; 36.
+Harley et al.
+1996; 37.
+Xu and Griffin 2006; 38.
+Atherton et al.
+2017; 39.
+Carter et al.
+2021; 40.
 Sack et al.
-2006; 4.
-Chin and Sillett 2019; 5.
-Wyka et al.
-2012; 6.
-Atherton et al.
-2017; 7.
-Kenzo et al.
-2015; 8.
-Kusi and Karasi 2020; 9.
-Dang et al.
-1997; 10.
-Gebauer et al.
-2015; 11.
-Marenco et al.
-2017; 12.
-Kafuti et al.
-2020; 13.
-Van Wittenberghe et al.
-2012; 14.
-Zhang et al.
-2019; 15.
-Weerasinghe et al.
-2014; 16.
-Oldham et al.
-2010; 17.
-Ichie et al.
-2016; 18.
-Gregoriou et al.
-2007; 19.
-Levizou et al.
-2005; 20.
-Liakoura 1997; 21.
-Fauset et al.
-2018; 22.
-Niinemets et al.
-1998, 23.
-Ishida et al.
-1998; 24.
-Millen and Clendon 1979; 25.
-Smith and Carter, 1988; 26.
-Hadley and Smith 1987; 28.
-Baltzer and Thomas 2005; 29.
-Coble et al.
-2016; 30.
-Scartazza et al.
-2016; 31.
-Duursma and Marshall, 2006; 32.
-Harley et al.
-1996;  33.
-Hernandez et al.
-2020; 34.
-Turnbull et al.
-2003; 35.
-Chen et al.
-2020; 36.
-van de Weg et al.
-2012; 37.
-M.A Cavaleri et al.
-2008;  38.
-Koniger et al.
-1995; 39.
-Mastubara et al.
-2009; 40.
-Harris and Medina 2013; 41.
-Hansen et al.
-2001; 42.
-Poorter et al.
-1995; 43.
-Coble et al.
-2016; 44.
-Niinemets et al.
-2004; 45.
-Poorter et al.
-2000; 46.
-Zwieniecki et al.
-2004; 47.
-Sack and Scoffoni, 2013; 48.
-Ball et al., 1988; 49.
-Taylor et al.
-2021; 50.
-Niinemets et al.
-2010; 51.
-Brooks et al.
-1997; 52.
-Curtis et al.
-2019; 53.
+2003; 41.
 Slot et al.
-2019; 54.
-Hamerlynck and Knapp 1994; 55.
-Lloyd et al.
-2010; 56.
-Augspurger and Bartlett, 2003; 57.
-Osada et al.
-2001; 58.
-Meakem et al.
-2018; 59.
-Condit et al.
+2021; 42.
+Carswell et al.
 2000
@@
 $\ddagger$ 1.
+Mau et al.
+2018; 2.
+Coble and Cavaleri 2014; 3.
+Sack et al.
+2006; 4.
+Chin and Sillett 2019; 5.
+Wyka et al.
+2012; 6.
+Atherton et al.
+2017; 7.
+Kenzo et al.
+2015; 8.
+Kusi and Karasi 2020; 9.
+Dang et al.
+1997; 10.
+Gebauer et al.
+2015; 11.
+Marenco et al.
+2017; 12.
 Kafuti et al.
-2020; 2.
+2020; 13.
 Van Wittenberghe et al.
-2012; 3.
-Roberts et al.
-1990; 4.
-Dang et al.
-1997; 5.
-Marenco et al.
-2017; 6.
-Ambrose et al.
-2015; 7.
-Zweifel et al.
-2001; 8.
+2012; 14.
+Zhang et al.
+2019; 15.
+Weerasinghe et al.
+2014; 16.
+Oldham et al.
+2010; 17.
+Ichie et al.
+2016; 18.
+Gregoriou et al.
+2007; 19.
+Levizou et al.
+2005; 20.
+Liakoura 1997; 21.
+Fauset et al.
+2018; 22.
+Niinemets et al.
+1998, 23.
+Ishida et al.
+1998; 24.
+Millen and Clendon 1979; 25.
+Smith and Carter, 1988; 26.
+Hadley and Smith 1987; 28.
+Baltzer and Thomas 2005; 29.
+Coble et al.
+2016; 30.
+Scartazza et al.
+2016; 31.
+Duursma and Marshall, 2006; 32.
+Harley et al.
+1996;  33.
+Hernandez et al.
+2020; 34.
+Turnbull et al.
+2003; 35.
+Chen et al.
+2020; 36.
+van de Weg et al.
+2012; 37.
+M.A Cavaleri et al.
+2008;  38.
+Koniger et al.
+1995; 39.
+Mastubara et al.
+2009; 40.
+Harris and Medina 2013; 41.
+Hansen et al.
+2001; 42.
+Poorter et al.
+1995; 43.
+Coble et al.
+2016; 44.
+Niinemets et al.
+2004; 45.
+Poorter et al.
+2000; 46.
+Zwieniecki et al.
+2004; 47.
+Sack and Scoffoni, 2013; 48.
+Ball et al., 1988; 49.
+Taylor et al.
+2021; 50.
+Niinemets et al.
+2010; 51.
+Brooks et al.
+1997; 52.
+Curtis et al.
+2019; 53.
 Slot et al.
-2019; 9.
-Hernandez et al.
-2020; 10.
-Urban et al.
-2007; 11.
-Carter and Cavaleri 2018; 12.
-Martin et al.
-1999; 13.
-Mau et al.
-2018;  14.
-Kosugi et al.
-2012; 15.
-Niinemets et al.
-2015; 16.
-Bachofen et al.
-2020; 17.
-Hamerlynck and Knapp 1994; 18.
-Coble et al.
-2017; 19.
-Wyka et al.
-2012; 20.
-Rijkerse et al.
-2000; 21.
-Ishida et al.
-1999; 22.
-Weerasinghe et al.
-2014; 23.
-Scartazza et al.
-2016; 24.
-Miller et al.
-2021; 25.
-Harris and Medina 2013; 26.
-Legner et al.
-2014; 27.
-Kitao et al.
-2012; 28.
-Fauset et al.
-2018; 29.
-Rey-Sanchez et al.
-2016; 30.
-Muller et al.
-2021; 31.
-Mier et al.
-2001; 32.
-Turnbull et al.
-2003; 33.
-Araki et al.
-2017; 34.
-Bolstad et al.
-1999; 35.
-Kenzo et al.
-2015; 36.
+2019; 54.
+Hamerlynck and Knapp 1994; 55.
+Lloyd et al.
+2010; 56.
+Augspurger and Bartlett, 2003; 57.
+Osada et al.
+2001; 58.
+Meakem et al.
+2018; 59.
+Condit et al.
+2000; 60.
 Harley et al.
-1996; 37.
-Xu and Griffin 2006; 38.
-Atherton et al.
-2017; 39.
-Carter et al.
-2021; 40.
-Sack et al.
-2003; 41.
-Taylor et al.
-2021; 42.
-Harley et al.
-1997; 43.
-Niinemets and Sun, 2014; 44.
-Sharkey and Monson, 2014; 45.
+1997; 61.
+Niinemets and Sun, 2014; 62.
+Sharkey and Monson, 2014; 63.
 Simpraga et al.
-2013; 46.
-Slot et al.
-2021
+2013
 

### 2022-02-12 vertical-thermal-review.Rmd
various small edits (proofreading)

@@
 
-Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
+Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
 Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
@@
 In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
-Early leaf-out allows some tree saplings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
+Early leaf-out allows some tree seedlings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
 Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
@@
 However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata, because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations [@buckleyStomatalOptimisationRelation2014].
-These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
-Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
+These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.2 (Table 1).
+Stomatal depression reduces transpirational cooling (Fig.
+`r fig_leaf_T`), thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
 The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than do shade leaves in lower vertical positions [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
@@
 Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in relatively dense-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
-This is consistent with the observation that $g_s$ limitation with $T_{air}$ increases with height across the vertical profile (Table `r table_leaf_metabolism`). 
+Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in closed-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
+This is consistent with the observation that $g_s$ limitation increases with height across the vertical profile (section 4.1, Table `r table_leaf_metabolism`). 
 More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce adjustment of traits related to carbon metabolism during drought [@bartholomewSmallTropicalForest2020].
@@
 `r fig_treerings`, @anderson-teixeiraJointEffectsClimate2022].
-However, under conditions that are humid enough that canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
+However, under well-watered conditions conducive to high $g_s$ in canopy trees, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].

### 2022-02-12 vertical-thermal-review.Rmd
misc small edits

@@
 Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will strongly influence the whole-ecosystem response, potentially with modest buffering by the understory. 
-Thus, for example, increases in canopy $T_{leaf}$ reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 
+Thus, for example, increases in canopy temperature reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 
 Yet there is also evidence that GPP and ecosystem respiration are less sensitive to heat and drought stress in older forests than in young forests [@xuSeasonalVariabilityForest2020], perhaps in part because of more complex vertical structuring [@jonesManagingRedPine2019].
@@
 
-As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased $T_{max}$s, with extreme increases during heat waves [@ipccClimateChange20212021].
+As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased maximum temperatures, with extreme increases during heat waves [@ipccClimateChange20212021].
 These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
@@
 Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
-In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015].
+In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019].
 An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 

### 2022-02-12 NEON_height_profiles/scripts/verticalProfileFigure.R
Updated figure for #83

@@ -170,7 +170,7 @@ for(i in 1:length(dp[,name])){
       graph <- graph + 
         xlab(paste0(dp[,stat][i], " ", dp[,xlabs][i]))
       if(i==1){
-        graph <- graph + ylab("Height [m]")
+        graph <- graph + ylab("Normalized height")
       }
     } else if(i==3){ #tempSingle
       graph <- graph +

### 2022-02-13 vertical-thermal-review.Rmd
update table 2
#100
update table 1 references
update table 1
#100
added a rought leaf optical property section
Krista, I think your suggestion actually is better. Maybe this would be better as #3.2?

@@
 
-*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, in response to height (H) or light (L).
-$\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
+\newpage
+Table `r table_leaf_traits`, cont.
 
-$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).
-
-$\ddagger$ 1.
-Kafuti et al.
-2020; 2.
-Van Wittenberghe et al.
-2012; 3.
-Roberts et al.
-1990; 4.
-Dang et al.
-1997; 5.
-Marenco et al.
-2017; 6.
-Ambrose et al.
-2015; 7.
-Zweifel et al.
-2001; 8.
-Slot et al.
-2019; 9.
-Hernandez et al.
-2020; 10.
-Urban et al.
-2007; 11.
-Carter and Cavaleri 2018; 12.
-Martin et al.
-1999; 13.
-Mau et al.
-2018;  14.
-Kosugi et al.
-2012; 15.
-Niinemets et al.
-2015; 16.
-Bachofen et al.
-2020; 17.
-Hamerlynck and Knapp 1994; 18.
-Coble et al.
-2017; 19.
-Wyka et al.
-2012; 20.
-Rijkerse et al.
-2000; 21.
-Ishida et al.
-1999; 22.
-Weerasinghe et al.
-2014; 23.
-Scartazza et al.
-2016; 24.
-Miller et al.
-2021; 25.
-Harris and Medina 2013; 26.
-Legner et al.
-2014; 27.
-Kitao et al.
-2012; 28.
-Fauset et al.
-2018; 29.
-Rey-Sanchez et al.
-2016; 30.
-Muller et al.
-2021; 31.
-Mier et al.
-2001; 32.
-Turnbull et al.
-2003; 33.
-Araki et al.
-2017; 34.
-Bolstad et al.
-1999; 35.
-Kenzo et al.
-2015; 36.
-Harley et al.
-1996; 37.
-Xu and Griffin 2006; 38.
-Atherton et al.
-2017; 39.
-Carter et al.
-2021; 40.
-Sack et al.
-2003; 41.
-Slot et al.
-2021; 42.
-Carswell et al.
-2000
-```{r eval=FALSE, echo=FALSE}
-# refs in this table currently in the nocite field at top of document 
-```
-
-## 3.1 Leaf anatomy and morphology
-
-Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`).
-Thus, sun leaves are generally smaller and thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
-`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces $g_b$).
-Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
-Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004].
-At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [reduces $g_b$ or increases boundary layer turbulence, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
-In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013].
-In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum $T_{air}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
-
-## 3.2 Metabolic capacity and efficiency
-
-Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
-Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
-Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
-
-## 3.3 – Light absorption and reflectance
-
-## 3.#. Biochemical protection against foliage light and heat damage
-
-Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
-Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
-A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
-This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
-Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
-Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10$^\circ$C) during sunflecks [@waySunflecksTreesForests2012].
-In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
-
-<!--VOCs-->
-Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
-In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
-Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
-
-Less than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile is influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
-Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
-However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_traits`, @taylorNewFieldInstrument2021].
-Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_traits`, @simpragaVerticalCanopyGradient2013].
-This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
-Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
-Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
-Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
-
-## 3.# Thermal tolerance
-
-<!--short par giving background on thermal tolerance-->
-The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
-At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
-Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
-Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and with higher thermal tolerance among leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
-$T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
-However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
-<!--For example, *Quercus muehlenbergii* Engelm.
-growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx.
-growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
-(removed because I had trouble finding this in the paper, plus it's not really needed, but feel free to reinstate) -->
-
-<!--par on thermal tolerance across vertical gradients-->
-Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be higher in sun-exposed upper canopy leaves than in understory leaves. 
-Indeed, along a vertical gradient within a very open canopy in Australia, crowns of *Acacia papyrocarpa* Benth.
-showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
-However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].
-This small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
-However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.
-
-## 3.# Leaf phenology
-
-The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
-In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
-In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
-Early leaf-out allows some tree seedlings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
-Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
-Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].
-Leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
-Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; @mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015].
-Leaf phenology thereby both responds to and influences the vertical structure of forest microenvironments and function.
-
-## 4. Leaf gas exchange and its thermal sensitivity
-
-Leaf metabolism is strongly shaped by $T_{leaf}$, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
-However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical forest gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
-
-\newpage
-**Table `r table_leaf_metabolism`.** Summary of observed variation in Leaf gas exchange and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
-Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
-
-![](table2a.png)
-
-\newpage
-
-Table `r table_leaf_metabolism`, cont.
-
-![](table2b.png)
-
-\newpage
-
-Table `r table_leaf_metabolism`, cont.
-
-![](table2c.png)
+![](table1c.png)
 
@@
 2013
+```{r eval=FALSE, echo=FALSE}
+# refs in this table currently in the nocite field at top of document 
+```
+
+## 3.1 Leaf anatomy and morphology
+
+Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`).
+Thus, sun leaves are generally smaller and thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
+`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces $g_b$).
+Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
+Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004].
+At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [reduces $g_b$ or increases boundary layer turbulence, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
+In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013].
+In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum $T_{air}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
+
+## 3.2 Metabolic capacity and efficiency
+
+Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
+Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
+Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
+
+## 3.3 – Leaf optical properties
+<!---need to cut parts of this paragraph or merge into 3.1?-->
+Leaf optical properties of leaves are influenced by a combination of anatomical, morphological, biochemical traits outlined above and water content, that vary vertically.
+Leaf light absorptance is maximized along the vertical light gradient and is balanced with leaf temperature regulation.
+Yet, leaf absorptance is only modestly greater in sun leaves.
+In the understory or shade regions where PAR is more limiting, leaf light adsorption efficiency is greater than in sun leaves, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilites, which realizes to be similar chlorophyll concentration per unit area as in sun leaves.
+Therefore, shade leaves can have comparable light harvesting capabilities as in sun leaves, and higher absorption effiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
+On the other hand, in higher light environment, there is less need for light absorption efficiency, but greater need to channel higher availablity of light into deeper cell layers through "light pipes" or palisade layers that increase path length for light penetration, contributing to higher photosynthetic capacity.
+At the same time, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or biochemically dissipating heat through caretenoid concentrations [Table 1, @knappVariabilityLeafOptical1998; @leeCorrelatesLeafOptical1990].
+Thicker sun leaves also reduce light transmittance or light scattered after entering the leaf than shade leaves [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
+Therefore, while canopy leaves tend to invest more energy in metabolic activity and heat dissipation, understory leaves invest in light capture.
+
+## 3.#. Biochemical protection against foliage light and heat damage
+
+Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
+Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
+Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
+A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
+This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
+Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
+Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10$^\circ$C) during sunflecks [@waySunflecksTreesForests2012].
+In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
+
+<!--VOCs-->
+Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
+In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
+Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
+
+Less than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile is influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
+Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
+However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_traits`, @taylorNewFieldInstrument2021].
+Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_traits`, @simpragaVerticalCanopyGradient2013].
+This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
+Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
+Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
+
+## 3.# Thermal tolerance
+
+<!--short par giving background on thermal tolerance-->
+The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
+At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
+Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
+Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and with higher thermal tolerance among leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
+$T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
+However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
+<!--For example, *Quercus muehlenbergii* Engelm.
+growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx.
+growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+(removed because I had trouble finding this in the paper, plus it's not really needed, but feel free to reinstate) -->
+
+<!--par on thermal tolerance across vertical gradients-->
+Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be higher in sun-exposed upper canopy leaves than in understory leaves. 
+Indeed, along a vertical gradient within a very open canopy in Australia, crowns of *Acacia papyrocarpa* Benth.
+showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
+However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].
+This small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
+However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.
+
+## 3.# Leaf phenology
+
+The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
+In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
+In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
+Early leaf-out allows some tree seedlings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
+Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
+Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].
+Leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
+Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; @mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015].
+Leaf phenology thereby both responds to and influences the vertical structure of forest microenvironments and function.
+
+## 4. Leaf gas exchange and its thermal sensitivity
+
+Leaf metabolism is strongly shaped by $T_{leaf}$, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
+However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical forest gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+
+\newpage
+**Table `r table_leaf_metabolism`.** Summary of observed variation in Leaf gas exchange and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
+Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
+
+![](table2a.png)
+
+\newpage
+
+Table `r table_leaf_metabolism`, cont.
+
+![](table2b.png)
+
+\newpage
+
+Table `r table_leaf_metabolism`, cont.
+
+![](table2c.png)
+
+*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, in response to height (H) or light (L).
+$\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
+
+$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).
+
+$\ddagger$ 1.
+Kafuti et al.
+2020; 2.
+Van Wittenberghe et al.
+2012; 3.
+Roberts et al.
+1990; 4.
+Dang et al.
+1997; 5.
+Marenco et al.
+2017; 6.
+Ambrose et al.
+2015; 7.
+Zweifel et al.
+2001; 8.
+Slot et al.
+2019; 9.
+Hernandez et al.
+2020; 10.
+Urban et al.
+2007; 11.
+Carter and Cavaleri 2018; 12.
+Martin et al.
+1999; 13.
+Mau et al.
+2018;  14.
+Kosugi et al.
+2012; 15.
+Niinemets et al.
+2015; 16.
+Bachofen et al.
+2020; 17.
+Hamerlynck and Knapp 1994; 18.
+Coble et al.
+2017; 19.
+Wyka et al.
+2012; 20.
+Rijkerse et al.
+2000; 21.
+Ishida et al.
+1999; 22.
+Weerasinghe et al.
+2014; 23.
+Scartazza et al.
+2016; 24.
+Miller et al.
+2021; 25.
+Harris and Medina 2013; 26.
+Legner et al.
+2014; 27.
+Kitao et al.
+2012; 28.
+Fauset et al.
+2018; 29.
+Rey-Sanchez et al.
+2016; 30.
+Muller et al.
+2021; 31.
+Mier et al.
+2001; 32.
+Turnbull et al.
+2003; 33.
+Araki et al.
+2017; 34.
+Bolstad et al.
+1999; 35.
+Kenzo et al.
+2015; 36.
+Harley et al.
+1996; 37.
+Xu and Griffin 2006; 38.
+Atherton et al.
+2017; 39.
+Carter et al.
+2021; 40.
+Sack et al.
+2003; 41.
+Slot et al.
+2021; 42.
+Carswell et al.
+2000
 

### 2022-02-13 vertical-thermal-review.Rmd
leap optical properties section

@@
 Yet, leaf absorptance is only modestly greater in sun leaves.
-In the understory or shade regions where PAR is more limiting, leaf light adsorption efficiency is greater than in sun leaves, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilites, which realizes to be similar chlorophyll concentration per unit area as in sun leaves.
+In the understory or shade regions where PAR is more limiting, leaf light absorption efficiency is greater than in sun leaves, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilites, which realizes to be similar chlorophyll concentration per unit area as in sun leaves.
 Therefore, shade leaves can have comparable light harvesting capabilities as in sun leaves, and higher absorption effiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].

### 2022-02-13 vertical-thermal-review.Rmd
small changes, need to come back to this

--- 2022-02-12 vertical-thermal-review.Rmd
@@
 2013
+```{r eval=FALSE, echo=FALSE}
+# refs in this table currently in the nocite field at top of document 
+```
+
+## 3.1 Leaf anatomy and morphology
+
+Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`).
+Thus, sun leaves are generally smaller and thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
+`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces $g_b$).
+Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
+Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004].
+At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [reduces $g_b$ or increases boundary layer turbulence, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
+In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013].
+In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum $T_{air}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
+
+## 3.2 Metabolic capacity and efficiency
+
+Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
+Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
+Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
+
+## 3.3 – Leaf optical properties
+<!---need to cut parts of this paragraph or merge into 3.1?-->
+Leaf optical properties of leaves are influenced by a combination of anatomical, morphological, and biochemical traits outlined above, that vary vertically.
+Leaf light absorptance is maximized along the vertical light gradient and is balanced with leaf temperature regulation, where excess light can impair photosynthetic machinery (see below).
+Yet, leaf absorptance is only modestly greater in sun leaves or mid-canopy leaves than shade leaves.
+In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat through caretenoid concentrations [Table 1, @knappVariabilityLeafOptical1998; @leeCorrelatesLeafOptical1990].
+Further, there is less a need for light absorption efficiency, but greater need to channel the higher availablity of light into deeper cell layers through "light pipes" or palisade layers that increase path length for light penetration, contributing to higher photosynthetic capacity.
+The thickness of sun leaves also reduce light transmittance or light scattered after entering the leaf than shade leaves [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
+However, in the understory or shade regions where PAR is more limiting, leaf light absorption efficiency is greater, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilites, which realizes to be similar chlorophyll concentration per unit area as in sun leaves.
+Therefore, shade leaves can have comparable light harvesting capabilities as in sun leaves, and higher absorption effiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
+Therefore, while canopy leaves tend to invest more energy in metabolic activity and heat dissipation, understory leaves invest in light capture.
+
+## 3.#. Biochemical protection against foliage light and heat damage
+
+Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
+Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
+Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
+A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
+This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
+Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
+Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10$^\circ$C) during sunflecks [@waySunflecksTreesForests2012].
+In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
+
+<!--VOCs-->
+Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
+In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
+Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
+
+Less than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile is influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
+Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
+However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_traits`, @taylorNewFieldInstrument2021].
+Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_traits`, @simpragaVerticalCanopyGradient2013].
+This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
+Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
+Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
+
+## 3.# Thermal tolerance
+
+<!--short par giving background on thermal tolerance-->
+The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
+At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
+Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
+Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and with higher thermal tolerance among leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
+$T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
+However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
+<!--For example, *Quercus muehlenbergii* Engelm.
+growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx.
+growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
+(removed because I had trouble finding this in the paper, plus it's not really needed, but feel free to reinstate) -->
+
+<!--par on thermal tolerance across vertical gradients-->
+Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be higher in sun-exposed upper canopy leaves than in understory leaves. 
+Indeed, along a vertical gradient within a very open canopy in Australia, crowns of *Acacia papyrocarpa* Benth.
+showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
+However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].
+This small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
+However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.
+
+## 3.# Leaf phenology
+
+The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
+In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
+In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
+Early leaf-out allows some tree seedlings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
+Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
+Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].
+Leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
+Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; @mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015].
+Leaf phenology thereby both responds to and influences the vertical structure of forest microenvironments and function.
+
+## 4. Leaf gas exchange and its thermal sensitivity
+
+Leaf metabolism is strongly shaped by $T_{leaf}$, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
+However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical forest gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+
+\newpage
+**Table `r table_leaf_metabolism`.** Summary of observed variation in Leaf gas exchange and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
+Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
+
+![](table2a.png)
+
+\newpage
+
+Table `r table_leaf_metabolism`, cont.
+
+![](table2b.png)
+
+\newpage
+
+Table `r table_leaf_metabolism`, cont.
+
+![](table2c.png)
+
+*Responses across height and light gradients are summarized, with up and down arrows indicating significant increasing or decreasing trends, respectively, in response to height (H) or light (L).
+$\approx$ indicates lack of any notable directional variation, and $\approx \uparrow$ or $\approx \downarrow$ indicate non-significant or mixed trends (e.g., significant in some but not all species studied).
+
+$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).
+
+$\ddagger$ 1.
+Kafuti et al.
+2020; 2.
+Van Wittenberghe et al.
+2012; 3.
+Roberts et al.
+1990; 4.
+Dang et al.
+1997; 5.
+Marenco et al.
+2017; 6.
+Ambrose et al.
+2015; 7.
+Zweifel et al.
+2001; 8.
+Slot et al.
+2019; 9.
+Hernandez et al.
+2020; 10.
+Urban et al.
+2007; 11.
+Carter and Cavaleri 2018; 12.
+Martin et al.
+1999; 13.
+Mau et al.
+2018;  14.
+Kosugi et al.
+2012; 15.
+Niinemets et al.
+2015; 16.
+Bachofen et al.
+2020; 17.
+Hamerlynck and Knapp 1994; 18.
+Coble et al.
+2017; 19.
+Wyka et al.
+2012; 20.
+Rijkerse et al.
+2000; 21.
+Ishida et al.
+1999; 22.
+Weerasinghe et al.
+2014; 23.
+Scartazza et al.
+2016; 24.
+Miller et al.
+2021; 25.
+Harris and Medina 2013; 26.
+Legner et al.
+2014; 27.
+Kitao et al.
+2012; 28.
+Fauset et al.
+2018; 29.
+Rey-Sanchez et al.
+2016; 30.
+Muller et al.
+2021; 31.
+Mier et al.
+2001; 32.
+Turnbull et al.
+2003; 33.
+Araki et al.
+2017; 34.
+Bolstad et al.
+1999; 35.
+Kenzo et al.
+2015; 36.
+Harley et al.
+1996; 37.
+Xu and Griffin 2006; 38.
+Atherton et al.
+2017; 39.
+Carter et al.
+2021; 40.
+Sack et al.
+2003; 41.
+Slot et al.
+2021; 42.
+Carswell et al.
+2000
 

### 2022-02-13 vertical-thermal-review.Rmd
moved up optical properties, numbered traits sections
#103, #106

@@
 
-## 3.2 Metabolic capacity and efficiency
-
-Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
-Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
-Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
-
-## 3.3 – Leaf optical properties
+## 3.2. Leaf optical properties
 <!---need to cut parts of this paragraph or merge into 3.1?-->
@@
 
-## 3.#. Biochemical protection against foliage light and heat damage
+## 3.3. Metabolic capacity and efficiency
+
+Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
+Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
+Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
+
+## 3.4. Biochemical protection against foliage light and heat damage
 
@@
 
-## 3.# Thermal tolerance
+## 3.5. Thermal tolerance
 
@@
 
-## 3.# Leaf phenology
+## 3.6. Leaf phenology
 

### 2022-02-13 vertical-thermal-review.Rmd
verify section numbering following rearrangement
#106

@@
 However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata, because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations [@buckleyStomatalOptimisationRelation2014].
-These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.2 (Table 1).
+These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.3 (Table 1).
 Stomatal depression reduces transpirational cooling (Fig.

### 2022-02-13 vertical-thermal-review.Rmd
revised T50 paragraph
closes #105

@@
 <!--par on thermal tolerance across vertical gradients-->
-Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be higher in sun-exposed upper canopy leaves than in understory leaves. 
-Indeed, along a vertical gradient within a very open canopy in Australia, crowns of *Acacia papyrocarpa* Benth.
-showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing portions of their crowns than in other crown positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019].
-However, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama (Table `r table_leaf_traits`), $T_{50}$ varied modestly, being slightly lower for shade than sun leaves for two of three tree species [@slotPhotosyntheticHeatTolerance2019].
-This small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
+Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be greatest for the leaves that face the greatest radiation and highest temperatures along the vertical gradient.
+Indeed, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama, $T_{50}$ was 0.7 - 1.4 $^\circ$C lower for shade than sun leaves for two of three tree species [Table `r table_leaf_traits`, @slotPhotosyntheticHeatTolerance2019].
+Similarly, along a vertical gradient within a very open canopy in Australia, where crowns of *Acacia papyrocarpa* Benth.
+faced the greatest composite climate stress -- including low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation -- in the lower, north-facing portions of their crowns, $T_{50}$ was highest in these positions, *albiet* by <1 $^\circ$C [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
+These modest difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments imply that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
 However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.

### 2022-02-13 vertical-thermal-review.Rmd
revised T_air paragraph
@NidhiVinod , I felt that this paragraph needed to be a little more decisive on T_air trends (didn't like the "often shows little variation"). Please check to make sure this looks good, and perhaps adjust placement of some of the references between the 2nd and 3rd sentences (maybe 1st as well).

This adds one of the citations in issue #79

@@
 
-Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances, attenuation of radiation and vertical transport of sensible heat by the canopy can buffer the lower canopy and understory from large diel swings in air temperature  (Fig.
-`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
-The $T_{air}$ gradient from the upper canopy to the understory varies with factors such as local topography, vertical air mixing, cloud cover,  vegetation type and sensible heat emissions [@defrenneForestMicroclimatesClimate2021].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022].
-Notably, this pattern can be reversed in open forests. 
+Vertical gradients in air temperature ($T_{air}$) across forest profiles are variable (Fig.
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), being dependent upon factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021].
+It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTEMPSubcanopyMicroclimate2021].
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTEMPSubcanopyMicroclimate2021].
+Notably, this pattern can be reversed in open forests (Supporting Information Figure `r SIfig_NEON_all`). 
 Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.

### 2022-02-13 vertical-thermal-review.Rmd
further revision of T_air paragraph
Adds the second reference in issue #79.

@@
 It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTEMPSubcanopyMicroclimate2021].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTEMPSubcanopyMicroclimate2021].
-Notably, this pattern can be reversed in open forests (Supporting Information Figure `r SIfig_NEON_all`). 
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTEMPSubcanopyMicroclimate2021; @meeussenMicroclimaticEdgeInterior2021].
+Notably, this pattern becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgeInterior2021].
 Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.

### 2022-02-13 vertical-thermal-review.Rmd
a bit more on Tair paragraph

@@
 Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTEMPSubcanopyMicroclimate2021; @meeussenMicroclimaticEdgeInterior2021].
-Notably, this pattern becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgeInterior2021].
+Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgeInterior2021].
 Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.

### 2022-02-13 vertical-thermal-review.Rmd
new line for each sentence
@NidhiVinod , this makes it easier to see what changes.
editing leaf optical properties

@@
 <!---need to cut parts of this paragraph or merge into 3.1?-->
-Leaf optical properties of leaves are influenced by a combination of anatomical, morphological, and biochemical traits outlined above, that vary vertically.
-Leaf light absorptance is maximized along the vertical light gradient and is balanced with leaf temperature regulation, where excess light can impair photosynthetic machinery (see below).
+Leaf optical properties of leaves are influenced by a combination of anatomical, morphological, and biochemical traits that vary vertically, as outlined above.
+Leaf light absorptance is maximized along the vertical light gradient and is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see below).
 Yet, leaf absorptance is only modestly greater in sun leaves or mid-canopy leaves than shade leaves.
 In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat through caretenoid concentrations [Table 1, @knappVariabilityLeafOptical1998; @leeCorrelatesLeafOptical1990].
-Further, there is less a need for light absorption efficiency, but greater need to channel the higher availablity of light into deeper cell layers through "light pipes" or palisade layers that increase path length for light penetration, contributing to higher photosynthetic capacity.
-The thickness of sun leaves also reduce light transmittance or light scattered after entering the leaf than shade leaves [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
-However, in the understory or shade regions where PAR is more limiting, leaf light absorption efficiency is greater, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilites, which realizes to be similar chlorophyll concentration per unit area as in sun leaves.
-Therefore, shade leaves can have comparable light harvesting capabilities as in sun leaves, and higher absorption effiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
-Therefore, while canopy leaves tend to invest more energy in metabolic activity and heat dissipation, understory leaves invest in light capture.
+Further, there is less need for light absorption efficiency, but greater need to channel the higher availability of light into deeper cell layers through "light pipes" or palisade layers that increase path length for light penetration, contributing to higher photosynthetic capacity.
+The thickness of sun leaves also reduces light transmittance or light scattered after entering the leaf more than in shade leaves [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
+However, in the understory or shade regions where PAR is more limiting, leaf light absorption efficiency is greater, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilities, resulting in similar chlorophyll concentration per unit area as in sun leaves.
+Therefore, shade leaves can have comparable light harvesting capabilities as in sun leaves, and higher absorption efficiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
+Thus, while canopy leaves tend to invest more energy in metabolic activity and heat dissipation, understory leaves invest in light capture.
 

### 2022-02-13 vertical-thermal-review.Rmd
Tleaf par

@@
 Leaf optical properties of leaves are influenced by a combination of anatomical, morphological, and biochemical traits that vary vertically, as outlined above.
-Leaf light absorptance is maximized along the vertical light gradient and is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see below).
+Along the vertical light gradient, leaf light absorptance is maximized and is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see below).
 Yet, leaf absorptance is only modestly greater in sun leaves or mid-canopy leaves than shade leaves. 

### 2022-02-13 vertical-thermal-review.Rmd
leaf optical properties- conceptual change for review
@NidhiVinod , light absorptance can't be both maximized and balanced with Tleaf regulation. Please confirm that this conceptual change is ok, or edit if not.

--- 2022-02-13 vertical-thermal-review.Rmd
@@
 Leaf optical properties of leaves are influenced by a combination of anatomical, morphological, and biochemical traits that vary vertically, as outlined above.
-Leaf light absorptance is maximized along the vertical light gradient and is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see below).
-Yet, leaf absorptance is only modestly greater in sun leaves or mid-canopy leaves than shade leaves.
+Along the vertical light gradient, leaf light absorptance is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see below), and is only modestly greater in sun leaves or mid-canopy leaves than shade leaves.
 In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat through caretenoid concentrations [Table 1, @knappVariabilityLeafOptical1998; @leeCorrelatesLeafOptical1990].

### 2022-02-13 vertical-thermal-review.Rmd
updated T_air citations

@@
 Vertical gradients in air temperature ($T_{air}$) across forest profiles are variable (Fig.
-`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), being dependent upon factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021].
-It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTEMPSubcanopyMicroclimate2021].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTEMPSubcanopyMicroclimate2021; @meeussenMicroclimaticEdgeInterior2021].
-Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgeInterior2021].
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), being dependent upon factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019].
+It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021].
+Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
 Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
 This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers [e.g., OSBS, Fig.
-`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @hadleyInfluenceKrummholzMat1987; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015].
+`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @hadleyInfluenceKrummholzMat1987; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015; @mullerEvidenceEfficientNonevaporative2021].
 These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.

### 2022-02-13 leaf_energy_balance/Fig3_Tleaf.R
added typical values

@@ -1,0 +1,484 @@
+####################################
+# Purpose: Create leaf energy balance figure using HARV neon data parameters for a typical overstory
+# and understory, normal and drought scenarios
+# Tleaves R package created by Chris Muir, 2019
+# Developed by: Nidhi Vinod, contact Anderson-Teixeira (teixeirak@si.edu)
+# R version 3.5.3 --First created in October 2020, updated on 5/28/2021
+# Oct,2020-Oct,2021
+####################################
+#install packages if needed
+library(latex2exp)
+library(magrittr)
+library(tealeaves)
+library(ggplot2)
+library(dplyr)
+library(ggthemes)
+library(ggpubr)
+library(cowplot)
+library(gridExtra)
+library(RColorBrewer)
+####################################
+
+
+# create a dataframe for overstory and understory, for scenarios: normal and drought
+# 1. Short wave radiation, normal
+#understory
+sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
+                  rh = 0.97, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
+                  gs = 2.0)
+#overstory
+sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
+                  rh = 0.91, wind = 2.88, tair = 298.1, pressure = 99, leaf.size = 0.04, 
+                  gs = 4.0)
+
+sw_l$par <- sw_l$par*0.5#sw
+sw_u$par <- sw_u$par*0.5#sw
+
+#apply Tleaves function
+
+get_tleaf <- function(sw, rh, wind, tair, pressure, leaf.size, gs) {
+  
+  # Set the environmental parameters (Tair, RH, PAR, and wind speed)
+  enviro_par <- make_enviropar(
+    replace = list(
+      T_air = set_units(c(tair), "K"),
+      RH = set_units(c(rh)),
+      S_sw = set_units(c(sw), "W/m^2"),
+      wind = set_units(c(wind), "m/s"),
+      P = set_units(c(pressure), "kPa")
+      
[…]

### 2022-02-13 vertical-thermal-review.Rmd

@@
 Along the vertical light gradient, leaf light absorptance is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see below), and is only modestly greater in sun leaves or mid-canopy leaves than shade leaves.
-In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat through caretenoid concentrations [Table 1, @knappVariabilityLeafOptical1998; @leeCorrelatesLeafOptical1990].
+In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat through caretenoid concentrations [Table `r table_leaf_traits`, @knappVariabilityLeafOptical1998; @leeCorrelatesLeafOptical1990].
 Further, there is less need for light absorption efficiency, but greater need to channel the higher availability of light into deeper cell layers through "light pipes" or palisade layers that increase path length for light penetration, contributing to higher photosynthetic capacity.

### 2022-02-13 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) characteristic leaf width, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf characteristic dimension, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values illustrated by dots and given in the table of biophysical constants.
-Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), characteristic leaf width (clw), stomatal conductance (gs), air temperature (tair).
+Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), cleaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair).
 ](leaf_energy_balance/fig3_leafenergy.png)

### 2022-02-15 vertical-thermal-review.Rmd
Lawren's paragraph added
Stil working on this

@@
 ## 3. Leaf traits 
+<!--Lawren's paragraph for R2 comments on having older citations and connecting to newer work-->
+The vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade19688] and this tendency grew given the parallel differences observed as for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
+More recent work has included a focus not only on shade, but also on leaf height in the canopy as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy  [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
+For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
+Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`; @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; **Niinemets et al., 2015a**].
+Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
+Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (cite Niinemets papers on this, as well as Japanese papers on N allocation throughout the canopy).
+Yet, while these studies have clarified many details of leaf differences throughout the canopy,  no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
+Within-canopy, leaf traits vary along the vertical light gradient (i.e., sun and shade leaves, @casasSunShadeLeaves2011;  @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; Niinemets et al., 2015a; @chenLeafEconomicsSpectrum2020].
+Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], using sun leaves [@keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020], and canopy shade is known to alter these trait relationships (Osnas et al.
+2018 PNAS), and thus further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
+It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
+Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity.
 
+<!--older paragph-->
 Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].

### 2022-02-17 vertical-thermal-review.Rmd
new line for each sentence
bold citations that need to be inserted

@@
 Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (cite Niinemets papers on this, as well as Japanese papers on N allocation throughout the canopy).
+Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (**cite Niinemets papers on this, as well as Japanese papers on N allocation throughout the canopy**).
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
-Within-canopy, leaf traits vary along the vertical light gradient (i.e., sun and shade leaves, @casasSunShadeLeaves2011;  @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; Niinemets et al., 2015a; @chenLeafEconomicsSpectrum2020].
-Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], using sun leaves [@keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020], and canopy shade is known to alter these trait relationships (Osnas et al.
-2018 PNAS), and thus further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
+Within-canopy, leaf traits vary along the vertical light gradient (i.e., sun and shade leaves, @casasSunShadeLeaves2011;  @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; **Niinemets et al., 2015a**; @chenLeafEconomicsSpectrum2020].
+Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], using sun leaves [@keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020], and canopy shade is known to alter these trait relationships (**Osnas et al.
+2018 PNAS**), and thus further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
 It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.

### 2022-02-17 vertical-thermal-review.Rmd
some rearrangment

@@
 <!--Lawren's paragraph for R2 comments on having older citations and connecting to newer work-->
-The vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].
+This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
 Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade19688] and this tendency grew given the parallel differences observed as for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
@@
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
-Within-canopy, leaf traits vary along the vertical light gradient (i.e., sun and shade leaves, @casasSunShadeLeaves2011;  @binLeafTraitExpression2022] at a biochemical, anatomical and structural level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; **Niinemets et al., 2015a**; @chenLeafEconomicsSpectrum2020].
-Much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], using sun leaves [@keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020], and canopy shade is known to alter these trait relationships (**Osnas et al.
+Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], using sun leaves [@keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020], and canopy shade is known to alter these trait relationships (**Osnas et al.
 2018 PNAS**), and thus further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
@@
 <!--older paragph-->
-Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].
+
 The primary driver of this variation is the vertical light gradient, with strong differentiation between sun and shade leaves [@casasSunShadeLeaves2011; @binLeafTraitExpression2022].

### 2022-02-17 vertical-thermal-review.Rmd
one sentence per line
rearrangements reconciling 2 traits intro paragraphs
@NidhiVinod , this commit shows most of the changes I've made here (along with the few previous commits). There is a little bit of content (lines 274.5-275 that is not really captured in the new paragraph, but I think that's okay.  I'm going to delete the whole old paragraph on the next commit, but you can come back to this commit in case you want to "rescue" anything that I'm deleting.

@@
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
-Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], using sun leaves [@keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020], and canopy shade is known to alter these trait relationships (**Osnas et al.
-2018 PNAS**), and thus further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
-It has been established that independent of the light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy.
-Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity.
+Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020], and as canopy shade is known to alter these trait relationships (**Osnas et al.
+2018 PNAS**), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
 
 <!--older paragph-->
-
-The primary driver of this variation is the vertical light gradient, with strong differentiation between sun and shade leaves [@casasSunShadeLeaves2011; @binLeafTraitExpression2022].
-Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
+~~Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].~~
+~~The primary driver of this variation is the vertical light gradient, with strong differentiation between sun and shade leaves [@casasSunShadeLeaves2011; @binLeafTraitExpression2022].~~
+~~Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], ~~ and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
 Therefore, a continuum of sun and shade leaves are observed throughout the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions (Fig.
 `r fig_NEON_vertical`), and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
-Traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy, independent of the light environment. 
-For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].
-Yet, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], and further research is needed to fully characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.
+~~Traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy, independent of the light environment.~~ 
+~~For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
+~~
+~~Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].~~
+~~Yet, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], and further research is needed to fully characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.~~
 
-Across the vertical gradient, traits vary (1) across sun and shade leaves within individuals, (2) across canopy and understory individuals of the same species, and (3) across canopy and understory species.
+Across the vertical gradient, traits vary (1) within individuals, with height and across shaded and sunny positions [e.g., @example_within-individual_references], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a], and (3) across canopy and understory species [e.g., @example_interspecific_references].
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).

### 2022-02-17 vertical-thermal-review.Rmd
delete older paragraph

@@
 2018 PNAS**), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
-
-<!--older paragph-->
-~~Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].~~
-~~The primary driver of this variation is the vertical light gradient, with strong differentiation between sun and shade leaves [@casasSunShadeLeaves2011; @binLeafTraitExpression2022].~~
-~~Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], ~~ and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016].
-Therefore, a continuum of sun and shade leaves are observed throughout the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions (Fig.
-`r fig_NEON_vertical`), and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
-~~Traits also vary within species along the ontogenetic trajectory from the understory to the upper canopy, independent of the light environment.~~ 
-~~For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-~~
-~~Therefore, ontogeny is bound to play a role in shaping leaf traits and thermal sensitivity [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a].~~
-~~Yet, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species [@wrightWorldwideLeafEconomics2004], which was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], and further research is needed to fully characterize trait relationships and responses across the full range of microenvironments spanning the forest vertical gradient.~~
 

### 2022-02-17 vertical-thermal-review.Rmd
proofreading
(figures 1-3, traits section through 3.3)

@@
 ![**Figure `r fig_schematic`.
-Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology.** Abbreviations are as follows: VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; T sensitivity: temperature sensitivity; ET: evapotranspiration.
+Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology.** Abbreviations are as follows: $T_{air}$: air temperature, VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; T sensitivity: temperature sensitivity; ET: evapotranspiration.
 Here, we summarize conditions in closed-canopy forests during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime.
@@
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf characteristic dimension, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under both moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
-In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values illustrated by dots and given in the table of biophysical constants.
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf characteristic dimension, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under hypothetical moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
+Dots along each line correspond to the biophysical constant assigned for the independent variable in the other scenarios, and therefore represent the modeled value for $T_{Leaf}-T_{air}$ with all parameters set to the biophysical constants given in the table.
 Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), cleaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair).
@@
 This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
-Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade19688] and this tendency grew given the parallel differences observed as for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
-More recent work has included a focus not only on shade, but also on leaf height in the canopy as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
+Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade19688], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
+More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
@@
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
-Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020], and as canopy shade is known to alter these trait relationships (**Osnas et al.
+Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
+As canopy shade is known to alter these trait relationships (**Osnas et al.
 2018 PNAS**), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
 
-Across the vertical gradient, traits vary (1) within individuals, with height and across shaded and sunny positions [e.g., @example_within-individual_references], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a], and (3) across canopy and understory species [e.g., @example_interspecific_references].
+Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @example_within-individual_references], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a], and (3) across understory and canopy species [e.g., @example_interspecific_references].
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
@@
 
+<!--refs in this table are in the nocite field at top of document-->
 $\ddagger$ 1.
@@
 2013
-```{r eval=FALSE, echo=FALSE}
-# refs in this table currently in the nocite field at top of document 
-```
+
+\newpage
 
@@
 <!---need to cut parts of this paragraph or merge into 3.1?-->
-Leaf optical properties of leaves are influenced by a combination of anatomical, morphological, and biochemical traits that vary vertically, as outlined above.
-Along the vertical light gradient, leaf light absorptance is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see below), and is only modestly greater in sun leaves or mid-canopy leaves than shade leaves.
-In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat through caretenoid concentrations [Table `r table_leaf_traits`, @knappVariabilityLeafOptical1998; @leeCorrelatesLeafOptical1990].
+Leaf optical properties are influenced by a combination of anatomical, morphological, and biochemical traits that vary vertically, as outlined in sections 3.1 and 3.3.
+Along the vertical light gradient, leaf light absorptance is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see sections 3.4 and 3.5), and is only modestly greater in sun leaves or mid-canopy leaves than shade leaves (Table `r table_leaf_traits`).
+In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat through caretenoid concentrations [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
 Further, there is less need for light absorption efficiency, but greater need to channel the higher availability of light into deeper cell layers through "light pipes" or palisade layers that increase path length for light penetration, contributing to higher photosynthetic capacity.
 The thickness of sun leaves also reduces light transmittance or light scattered after entering the leaf more than in shade leaves [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
-However, in the understory or shade regions where PAR is more limiting, leaf light absorption efficiency is greater, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilities, resulting in similar chlorophyll concentration per unit area as in sun leaves.
-Therefore, shade leaves can have comparable light harvesting capabilities as in sun leaves, and higher absorption efficiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
+However, in the understory or shade regions where PAR is more limiting, leaf light absorption efficiency is greater, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilities, resulting in similar chlorophyll concentration per unit area as in sun leaves (Table `r table_leaf_traits`).
+Therefore, shade leaves can have light harvesting capabilities comparable to those of sun leaves, with higher absorption efficiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
 Thus, while canopy leaves tend to invest more energy in metabolic activity and heat dissipation, understory leaves invest in light capture.
@@
 Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
-Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
+Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile contribute to the reater PAR absorptance efficiency of shade leaves (Table `r table_leaf_traits`).
 Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
@@
 ** composite climatic stress variable integrating temperature, vapour pressure deficit, and relative humidity is higher in lower canopy
-```{r eval=FALSE, echo=FALSE}
-# refs in this table currently in the nocite field at top of document 
-```
+
+\newpage
 

### 2022-02-17 vertical-thermal-review.Rmd
proofread through end of traits

@@
 `r fig_leaf_T`b-e).
-In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig.
+In contrast, at lower heights along the vertical profile, greater RH and lower wind speeds may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Leaves in these normally-shaded locations experience rapid light and temperature surges (can exceed 10$^\circ$C) during sunflecks [@waySunflecksTreesForests2012].
 
@@
 Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
+Further, by reducing photosynthesis, heat and VPD-associated stomatal closure can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@ref <!--Nidhi, I feel like this sentence needs a reference-->].
 Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
@@
 This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
-Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
+Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
 Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10$^\circ$C) during sunflecks [@waySunflecksTreesForests2012].
-In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, in order to maximize carbon uptake and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
 
@@
 Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
-Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
+Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
 Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
@@
 The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
-At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
+At very high $T_{leaf}$ (~40-60$^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
 Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
@@
 Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be greatest for the leaves that face the greatest radiation and highest temperatures along the vertical gradient.
-Indeed, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama, $T_{50}$ was 0.7 - 1.4 $^\circ$C lower for shade than sun leaves for two of three tree species [Table `r table_leaf_traits`, @slotPhotosyntheticHeatTolerance2019].
+Indeed, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama, $T_{50}$ was 0.7 - 1.4$^\circ$C lower for shade than sun leaves for two of three tree species [Table `r table_leaf_traits`, @slotPhotosyntheticHeatTolerance2019].
 Similarly, along a vertical gradient within a very open canopy in Australia, where crowns of *Acacia papyrocarpa* Benth.
-faced the greatest composite climate stress -- including low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation -- in the lower, north-facing portions of their crowns, $T_{50}$ was highest in these positions, *albiet* by <1 $^\circ$C [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
+faced the greatest composite climate stress (low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation) in the lower, north-facing portions of their crowns, $T_{50}$ was highest in these positions, *albiet* by <1$^\circ$C [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
 These modest difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments imply that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
@@
 
-The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
+Leaf phenology likely influences the vertical structure of microenvironments, $T_{leaf}$, and forest photosynthetic capacity [@rey-sanchez_spatial_2016]. 
 In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
 In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
-Early leaf-out allows some tree seedlings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
+Early leaf-out gives tree saplings and seedlings a window for disproportionately high photosynthesis -- amounting to the majority of annual carbon fixation for some seedlings -- before canopy foliage and reduces light availability [@augspurgerDifferencesLeafPhenology2003; @leeSpringPhenologicalEscape2021].
 Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
+Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of deciduous species [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
 Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].
@@
 ** composite climatic stress variable integrating temperature, vapour pressure deficit, and relative humidity is higher in lower canopy
-
 \newpage

### 2022-02-17 vertical-thermal-review.Rmd
1st abbreviation

--- 2022-02-17 vertical-thermal-review.Rmd
@@
 Forests are vertically and horizontally stratified, with canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
-This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
+This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
 Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
@@
 Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-However, K-theory often fails within canopies, making forest verticcal wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+However, K-theory often fails within canopies, making forest vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
 For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux [e.g., @raupachLagrangianAnalysisScalar1987].

### 2022-02-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.
+Atmospheric concentrations of CO~2~ and water vapor can also vary across the vertical gradient.
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
 Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
-Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
+Particularly, understory seedlings benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
 However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

### 2022-02-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that upper canopy leaves face higher $g_b$ and associated potential for water and heat loss (see sections 2.2, 4.1).
+The implication is that upper canopy leaves have higher $g_b$ and associated potential for water and heat loss (see sections 2.2, 4.1).
 
@@
 `r fig_leaf_T`).
-This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$ limitation than shade leaves [Fig.
+This is because sun leaves are exposed to higher irradiance and VPD, and therefore have stronger $g_s$ limitation than shade leaves [Fig.
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].

### 2022-02-17 vertical-thermal-review.Rmd
leaf optical, just few more edits

@@
 Along the vertical light gradient, leaf light absorptance is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see sections 3.4 and 3.5), and is only modestly greater in sun leaves or mid-canopy leaves than shade leaves (Table `r table_leaf_traits`).
-In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat through caretenoid concentrations [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
+In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat by caretenoids [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
 Further, there is less need for light absorption efficiency, but greater need to channel the higher availability of light into deeper cell layers through "light pipes" or palisade layers that increase path length for light penetration, contributing to higher photosynthetic capacity.
-The thickness of sun leaves also reduces light transmittance or light scattered after entering the leaf more than in shade leaves [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
-However, in the understory or shade regions where PAR is more limiting, leaf light absorption efficiency is greater, enhanced by shade leaves maintaining higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, higher nitrogen, and chlorophyll per unit biomass that exemplifies optical absorption capabilities, resulting in similar chlorophyll concentration per unit area as in sun leaves (Table `r table_leaf_traits`).
+The thickness of sun leaves also means that  light transmittance is lower than in shade leaves [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
+In contrast, where PAR is more limiting, leaf light absorption efficiency is greater, as shade leaves maintain higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, and higher concentrations of nitrogen, and chlorophyll (Table `r table_leaf_traits`).
+Thus, while canopy leaves tend to invest more energy in metabolic activity and heat dissipation, understory leaves invest in light capture.
 Therefore, shade leaves can have light harvesting capabilities comparable to those of sun leaves, with higher absorption efficiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
-Thus, while canopy leaves tend to invest more energy in metabolic activity and heat dissipation, understory leaves invest in light capture.
 
@@
 Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
-Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
+Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthesis [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
 Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile contribute to the reater PAR absorptance efficiency of shade leaves (Table `r table_leaf_traits`).

### 2022-02-17 vertical-thermal-review.Rmd
minor edits

--- 2022-02-17 vertical-thermal-review.Rmd
@@
 Forests are vertically and horizontally stratified, with canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
-This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
+This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and CO~2~ concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
 Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
@@
 Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-However, K-theory often fails within canopies, making forest vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+However, K-theory often fails within canopies, making forest verticcal wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
 For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux [e.g., @raupachLagrangianAnalysisScalar1987].
@@
 
-Atmospheric concentrations of CO~2~ and water vapor can also vary across the vertical gradient.
+Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
 Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
-Particularly, understory seedlings benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
+Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
 However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
@@
 However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical forest gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+\newpage
 
-\newpage
 **Table `r table_leaf_metabolism`.** Summary of observed variation in Leaf gas exchange and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves.
@@
 
-** composite climatic stress variable integrating temperature, vapour pressure deficit, and relative humidity is higher in lower canopy
-\newpage
+** composite climatic stress variable integrating temperature, vapour pressure deficit, and relative humidity is higher in lower canopy \newpage
 
@@
 Vertical gradients also affect the climate sensitivity of metabolism and growth.
-Stomatal conductance tends to be more strongly limited by high atmospheric demand (high $T_{air}$ and VPD, low RH) even before soil water becomes limited in forests not experiencing drought [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in tree transpiration and photosynthesis in tall trees that occupy canopy positions in closed-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
+Stomatal conductance can be strongly limited by high atmospheric demand (low RH, high $T_{air}$ and VPD) even when soil water is plentiful [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in transpiration and photosynthesis in tall trees that occupy canopy positions in closed-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
 This is consistent with the observation that $g_s$ limitation increases with height across the vertical profile (section 4.1, Table `r table_leaf_metabolism`). 
 More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce adjustment of traits related to carbon metabolism during drought [@bartholomewSmallTropicalForest2020].
-This results in decreasing carbon isotope discrimination, indicative of an increasingly conservative hydraulic strategy, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
+This results in decreasing carbon isotope discrimination, indicative of greater $g_s$ limitation, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
 

### 2022-02-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2022-02-17 vertical-thermal-review.Rmd
@@
 <!--VOCs-->
-Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
+The capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
 In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
@@
 At very high $T_{leaf}$ (~40-60$^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
-Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021].
+Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [e.g., @slotLeafHeatTolerance2021].
 Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and with higher thermal tolerance among leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].

### 2022-02-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
-However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
+However, they appear to be are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
 <!--For example, *Quercus muehlenbergii* Engelm.
@@
 Indeed, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama, $T_{50}$ was 0.7 - 1.4$^\circ$C lower for shade than sun leaves for two of three tree species [Table `r table_leaf_traits`, @slotPhotosyntheticHeatTolerance2019].
-Similarly, along a vertical gradient within a very open canopy in Australia, where crowns of *Acacia papyrocarpa* Benth.
+Similarly, along a vertical gradient within a very open canopy in Australia, where trees of *Acacia papyrocarpa* Benth.
 faced the greatest composite climate stress (low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation) in the lower, north-facing portions of their crowns, $T_{50}$ was highest in these positions, *albiet* by <1$^\circ$C [Table `r table_leaf_traits`, @curtis_intracanopy_2019].

### 2022-02-17 vertical-thermal-review.Rmd
adding missing citation

@@
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`; @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; **Niinemets et al., 2015a**].
+Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`; @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
 Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (**cite Niinemets papers on this, as well as Japanese papers on N allocation throughout the canopy**).
+Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (@buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007).
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
 Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
-As canopy shade is known to alter these trait relationships (**Osnas et al.
-2018 PNAS**), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
+As canopy shade is known to alter these trait relationships (@osnasDivergentDriversLeaf2018), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
 

### 2022-02-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
-Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade19688], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
+Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
 More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].

### 2022-02-17 vertical-thermal-review.Rmd
updated piponiot citation

@@
 Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (@buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007).
+Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007).
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
@@
 
-Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @example_within-individual_references], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a], and (3) across understory and canopy species [e.g., @example_interspecific_references].
+Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @binLeafTraitExpression2022], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a], and (3) across understory and canopy species [e.g., @lloydOptimisationPhotosyntheticCarbon2010].
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
@@
 Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
-`r fig_ecosystem`b, @piponiotSizerelatedTreeBiomass2022].
+`r fig_ecosystem`b, @piponiotDistributionBiomassDynamicsinpress].
 
 ![**Figure `r fig_ecosystem`.
-Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotSizerelatedTreeBiomass2022.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotDistributionBiomassDynamicsinpress.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 

### 2022-02-17 vertical-thermal-review.Rmd
small edit

@@
 However, K-theory often fails within canopies, making forest vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
-For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux [e.g., @raupachLagrangianAnalysisScalar1987].
+For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat moving vertically upwards through the canopy [e.g., @raupachLagrangianAnalysisScalar1987].
 Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991].
@@
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
+Vertical CO~2~ concentration gradients can influence the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
 Particularly, understory seedlings benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 

### 2022-02-17 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 
-*(see Lawren's comments in google doc)*
+**Thank you for the suggestion. We have added additional information highlighting older work and recent advances in the paragraph below**
+
+**This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
+More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
+For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
+Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`; @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
+Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007).
+Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
+Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
+As canopy shade is known to alter these trait relationships (@osnasDivergentDriversLeaf2018), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.**
 

### 2022-02-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2022], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2022], although there is also evidence that short trees in young stands can be more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 Although it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
-Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.
+High $T_{air}$ is often associated with high VPD and atmospheric drought, likely explaining negative growth responses to $T_{air}$ -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.
 `r fig_treerings`, @anderson-teixeiraJointEffectsClimate2022].

### 2022-02-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Further, by reducing photosynthesis, heat and VPD-associated stomatal closure can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@ref <!--Nidhi, I feel like this sentence needs a reference-->].
+Further, by reducing photosynthesis, heat and VPD-associated stomatal closure can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007] <!--Nidhi, I feel like this sentence needs a reference.
+{maybe this would be better as a topic sentence without a ref?
+but there is one Ty suggested-->].
 Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].

### 2022-02-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
-Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
+Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play similar role in sun-exposed canopy leaves.
 

### 2022-02-18 vertical-thermal-review.Rmd
@teixeirak, I added the isoprene * in the caption but please let me know if this seems okay?

@@
 ![**Figure `r fig_schematic`.
-Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology.** Abbreviations are as follows: $T_{air}$: air temperature, VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area, VAZ: violaxanthin, antheraxanthin and zeaxanthin concentrations, VOC: volatile organic compounds; T sensitivity: temperature sensitivity; ET: evapotranspiration.
+Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology.** Abbreviations are as follows: $T_{air}$: air temperature, VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area; T sensitivity: temperature sensitivity; ET: evapotranspiration.
 Here, we summarize conditions in closed-canopy forests during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime.
-Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.](schematics/Schematic figure 1.png)
+Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.
+*Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`)](schematics/Schematic figure 1.png)
 

### 2022-02-18 vertical-thermal-review.Rmd
vertical lines and grey shading should be removed from Fig.4
#107
proofreading: Global change responses section

@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable.
 From @anderson-teixeiraJointEffectsClimate2022.
@@
 
-To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
+To the extent that warming is coupled to drought, we expect that the tallest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
 `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
@@
 Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
-Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
+Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
 
@@
 
-As discussed above, rising temperatures and severe droughts place canopy trees at particularly elevated risk of mortality. 
-Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change[@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
+Rising temperatures and severe droughts place canopy trees at particularly elevated risk of mortality. 
+Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
 Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].

### 2022-02-18 leaf_energy_balance/Fig3_Tleaf.R
Update Fig3_Tleaf.R

@@ -379,7 +379,7 @@ d1<-ggplot(ls)+
   geom_point(aes(x=0.04, y=7.5), colour="#008837")+
   geom_point(aes(x=0.1, y=2.79), colour="#7b3294")+
   
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab("leaf characteristic dimension (lcd, m)")+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (°C)"))+xlab("Leaf Characteristic Dimension (lcd, m)")+
   geom_hline(yintercept=0, linetype='dotted', col = 'black')+
   theme_few()+theme(text = element_text(size = 15))+ylim(-5, 14) 
 d1

### 2022-02-18 leaf_energy_balance/Fig3_Tleaf.R
Update Fig3_Tleaf.R

@@ -4,7 +4,7 @@
 # Tleaves R package created by Chris Muir, 2019
 # Developed by: Nidhi Vinod, contact Anderson-Teixeira (teixeirak@si.edu)
 # R version 3.5.3 --First created in October 2020, updated on 5/28/2021
-# Oct,2020-Oct,2021
+# Oct,2020-Feb,2022
 ####################################
 #install packages if needed
 library(latex2exp)

### 2022-02-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006].
+Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006; @aroraCarbonConcentrationCarboon2020].
 Dynamic global vegetation models (DGVMs), which comprise the land surface models in Earth system models, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].
@@
 diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].
 In recent years, there has been a growing effort to incorporate vertical variation more directly in models, specifically as it pertains to leaf water potential and light absorption by leaves, along with a general increasing interest in making direct comparison between models and field measurements [@fisherPerspectivesFutureLand2020].

### 2022-02-18 vertical-thermal-review.Rmd
change to check
@NidhiVinod , @eoway , @TomBuckleyLab -- Is this edit correct? The intent is to tie the modeling section back to earlier parts of the review, and I think this is consistent with content that Tom added in section 1, but I want to have this verified by people who know the subject.  You can see this if you follow the link to the commit in GitHub, or here's the modified sentence:  ("This simplification is computationally more efficient, although it **does not always capture observed vertical profiles (see section 1)** and comes at a cost of a lack of representation of important dem

@@
 DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
-This simplification is computationally more efficient, although it comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
+This simplification is computationally more efficient, although it does not always capture observed vertical profiles (see section 1) and comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
 The computational middle-ground to representing vertical structure in DVGMs lies in cohort-based models, which represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].

### 2022-02-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
https://github.com/EcoClimLab/vertical-thermal-review/commit/0758011e3e2967347cc55fe3a36128d73da52bda

--- 2022-02-18 vertical-thermal-review.Rmd
@@
 DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
-This simplification is computationally more efficient, although it comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
+This simplification is computationally more efficient, although it does not always capture observed vertical profiles (sections 1-4; for example, vertical shifts in the balance between stomatal conductance and photosynthetic capacity, see sections 4.1-4.2) and comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
 The computational middle-ground to representing vertical structure in DVGMs lies in cohort-based models, which represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].

### 2022-02-18 vertical-thermal-review.Rmd
proofreading: Scaling across space and time & conclusions

@@
 `r fig_NEON_vertical`e), and sunflecks or the 'canopy greenhouse effect', as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the $T_{leaf}$ of the foliage close to the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the $T_{leaf}$ of the foliage close to the ground [Fig.
+`r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
 One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
@@
 diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
 Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].
@@
 Several recent model developments are at the cutting edge of representing vertical variation.
-Model developments by [**Bonan et al., 2018; Chen et al., 2016; Longo et al., 2019**] implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into the vegetated air space and the humidity of the airspace, modulated by turbulence processes within the forest and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
+Model developments by @Chen_et_al_2016, @Bonan_et_al_2018, and @Longo_et_al_2019 implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into the vegetated air space and the humidity of the airspace, modulated by turbulence processes within the forest and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
 
@@
 
-Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers [e.g., sunlit and shaded portions, @depurySimpleScalingPhotosynthesis1997] may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, as we present in this paper).
+Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers [e.g., sunlit and shaded portions, @depurySimpleScalingPhotosynthesis1997] may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, Table 1, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, Table 2).
 At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @bonanMovingIncorrectUseful2021].
-In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.`r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.`r `fig_leaf_T`) to vertically varying abiotic conditions (Fig.`fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.
+In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.
+`r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.
+`r fig_leaf_T`) to vertically varying abiotic conditions (Fig.
+`r fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.
 
@@
 Airborne and spaceborne lidar, as well as terrestrial laser scanning data, yield detailed 3D reconstructions of whole tree and forest structure. 
-These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based Forward Looking Infrared (FLIR) cameras.
+These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based Forward Looking Infrared cameras.
 
@@
 Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020].
-As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
+As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based Forward Looking Infrared camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
 However, the authors did not consider vertical light environment conditions even though light directly influences the deviation between canopy temperature and $T_{air}$. 
-In a savanna system in the western U.S., @johnstonWhatLiesVertical2020 found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding opposite of what was expected that occurred in association with very high grass temperatures.
+In a savanna system in the western U.S., @johnstonWhatLiesVertical2020 found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
+`r SIfig_NEON_all`).
 The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of *in situ* measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
@@
 ## Acknowledgements
-Thanks to Norbert Kunert, Camille Piponiot, Peter B. Boucher and Andrew B. Davies for providing figure materials. 
+Thanks to Valentine Herrmann, Norbert Kunert, Camille Piponiot, Peter B. Boucher and Andrew B. Davies for providing figure materials. 
 This manuscript benefited from feedback from Eleinis Ávila-Lovera and the ForestGEO Ecosystems & Climate lab at SCBI.

### 2022-02-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
fix related to #110

closes #109

@@
 One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
-However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2020].
+However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2020].
 
@@
 As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based Forward Looking Infrared camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
-However, the authors did not consider vertical light environment conditions even though light directly influences the deviation between canopy temperature and $T_{air}$. 
 In a savanna system in the western U.S., @johnstonWhatLiesVertical2020 found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.

### 2022-02-18 response-to-reviews.Rmd
vertical lines and grey shading should be removed from Fig.4
#107
update Fig. 4
Here's a much improved version of Fig. 4, with many thanks to @ValentineHerr !

Co-Authored-By: Valentine Herrmann <19673377+ValentineHerr@users.noreply.github.com>

#107

@@
 
-- We revised three figures (Figs. 1-3) based on reviewer comments
+- We revised three figures (Figs. 1-4) based on reviewer comments, including some standardization of style, but have not perfected the styling because the journal illustrators will fix up all figures.
 
@@
 - We really like Figure 1 and would like to retain the background painting (by the lead author)
-- On Fig.
-4, please adjust colors on panel (a) to correspond to those in panel (b), where "min MIN" color should match "understory" and "max DBH" should match "canopy".
-The same color scheme can be applied to all panels in (a), and there is no need to keep the line type differences (both can be solid).
 

### 2022-02-18 vertical-thermal-review.Rmd
update Fig. 4 caption

@@
 Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species.
-Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.
+Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET over various seasonal windows, where "c" and "p" represent months in the current and previous year, respectively) for the maximum and minimum tails of the DBH distribution included in the analysis.
 Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction.

### 2022-02-19 vertical-thermal-review.Rmd
add Nunes et al. 2022 references
@NidhiVinod , FYI.

Nunes et al. 2022  is a really cool study!

@@
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017].
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].
 In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.
@@
 Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of deciduous species [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
-Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019].
+Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019;@tangLightdrivenGrowthAmazon2017;@nunesForestFragmentationImpacts2022].
+For instance, in a central Amazonian forest, the upper canopies of both interior and edge forests shed leaves when maximum daily $T_{air}$ exceeded ~35$^\circ$C [@nunesForestFragmentationImpacts2022].
 Leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
@@
 However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical forest gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+
 \newpage
@@
 Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in transpiration and photosynthesis in tall trees that occupy canopy positions in closed-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
-This is consistent with the observation that $g_s$ limitation increases with height across the vertical profile (section 4.1, Table `r table_leaf_metabolism`). 
-More active stomatal regulation of tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce adjustment of traits related to carbon metabolism during drought [@bartholomewSmallTropicalForest2020].
+This is consistent with observations that both dry season leaf loss (section 3.6, Table `r table_leaf_traits`) and $g_s$ limitation (section 4.1, Table `r table_leaf_metabolism`) increases with height across the vertical profile.
+More active regulation of transpiration by tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce the need for adjustment of traits related to carbon metabolism during drought [@bartholomewSmallTropicalForest2020].
 This results in decreasing carbon isotope discrimination, indicative of greater $g_s$ limitation, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
@@
 It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but probable responses can be inferred based on the patterns and mechanisms reviewed above.
-We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures in the upper canopy than in the understory -- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020].
+We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures in the upper canopy than in the understory -- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020; @nunesForestFragmentationImpacts2022].
 Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will strongly influence the whole-ecosystem response, potentially with modest buffering by the understory. 
 Thus, for example, increases in canopy temperature reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 
-Yet there is also evidence that GPP and ecosystem respiration are less sensitive to heat and drought stress in older forests than in young forests [@xuSeasonalVariabilityForest2020], perhaps in part because of more complex vertical structuring [@jonesManagingRedPine2019].
+Yet there is also evidence that GPP and ecosystem respiration are less sensitive to heat and drought stress in older forests than in young forests [@xuSeasonalVariabilityForest2020], perhaps in part because of more complex vertical structuring [@jonesManagingRedPine2019;@nunesForestFragmentationImpacts2022].
 A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an *El Niño* drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.
@@
 Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
-In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019].
+In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022].
 An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
@@
 Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the forest [@juckerCanopyStructureTopography2018; @starkReframingTropicalSavannization2020].
-Such changes often result in increased growth of smaller trees, which benefit from increased light [@hoganDroughtInterannualVariability2019;@bennettLargerTreesSuffer2015], and wetter forests can prove quite resilient to canopy disturbance [@millerReducedImpactLogging2011].
+Such changes often result in increased growth of smaller trees, which benefit from increased light [@hoganDroughtInterannualVariability2019;@bennettLargerTreesSuffer2015; @nunesForestFragmentationImpacts2022], and wetter forests can prove quite resilient to canopy disturbance [@millerReducedImpactLogging2011].
 However, this shift to hotter and drier microclimates makes some forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018].

### 2022-02-19 vertical-thermal-review_SI.Rmd
small changes from marielle

--- 2021-10-02 vertical-thermal-review_SI.Rmd
@@
 Leaf area density (LAD, $m^2$ $m^{-3}$) was estimated by applying the MacArthur-Horn transformation [@macarthurFoliageProfileVertical1969] to the voxelized pulse returns.
-We excluded LAD data at the bottom 5 m of the canopy, which are more uncertain due to limited pulse penetration within this region furthest away from the sensor [@starkAmazonForestCarbon2012]; for calibration purposes, we assumed the bottom 5 m to have a leaf area index (LAI) of 0.25 $m^2$ $m^{-2}$ [@shaoHighThroughputAssessment2019].
+We excluded LAD data at the bottom 5 m of the vertical profile, due to limited pulse penetration within this region furthest away from the sensor [@starkAmazonForestCarbon2012]; for calibration purposes, we assumed the bottom 5 m to have a leaf area index (LAI) of 0.25 $m^2 m^{-2}$ [@shaoHighThroughputAssessment2019].
 
 We scaled LAD values by a factor of 5.5 so that they matched an independent estimate of LAI.
-Specifically, published estimates of LAI at SERC (average of 5.78 $m^2$ $m^{-2}$) derived from litterfall [@parkerCanopySurfaceStand2004], which were then in good agreement with our lidar-derived estimate (5.75 $m^2$ $m^-2$).
-Generating leaf area and light transmission profiles
-Mean LAD profiles for each site were calculated as the mean LAD at each height.
+Specifically, published estimates of LAI at SERC (average of 5.78 $m^2$ $m^{-2}$) derived from litterfall [@parkerPortableLIDARSystem2004], which were then in good agreement with our lidar-derived estimate (5.75 $m^2 m^-2$).
+
+*Generating leaf area and light transmission profiles.*
+
+Site-level LAD profiles were calculated as the mean LAD at each height.
 Mean light transmission profiles (proportion of incident light) were estimated by applying a vertical light reduction model to LAD estimates [@starkAmazonForestCarbon2012].
-To estimate the proportion of sun leaves along the vertical profile, we isolated voxels in the upper canopy surface layer (1-2 m from the local canopy surface, likely to containing the majority of sun leaves), calculated the mean LAD profile for this surface layer, and divided it by the total mean LAD profile.
+To estimate the proportion of sun leaves along the vertical profile, we isolated voxels in the upper canopy surface layer (1-2 m from the local canopy surface, likely contain the majority of sun leaves), calculated the mean LAD profile for this surface layer, and divided it by the total mean LAD profile.
 
@@
 
-For the humid scenario, we used mean maximum RH from the NEON tower at Harvard Forest.
+For the moist scenario, we used mean maximum RH from the NEON tower at Harvard Forest.
 Stomatal conductance measurements were referred from *Tleaves* typical sun and shade measurements [@muir_tealeaves_2019] and @cavender-baresChangesDroughtResponse2000.

### 2022-02-19 vertical-thermal-review_SI.Rmd
spelling

@@
 Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)**.
-Shown are  height profiles in July mean ± 1 standard deviation for maximum photosyntehtically active ratiation (PAR), maximum wind speed, minimum humidity, maximum $T_{air}$, and maximum biological temperature, $T_{bio}$.
+Shown are  height profiles in July mean ± 1 standard deviation for maximum photosynthetically active radiation (PAR), maximum wind speed, minimum humidity, maximum $T_{air}$, and maximum biological temperature, $T_{bio}$.
 Site information is given in Supporting Information Table `r NEON_sites`, and analysis details in Supporting Information Methods `r appendix_NEON_methods`.](NEON_height_profiles/figures/profile_all_groups.png)

### 2022-02-19 vertical-thermal-review.Rmd
Johnston citation

@@
 This holds true across the range of forest types [Fig.
-`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
+`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalInpress].
 Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile.
@@
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the $T_{leaf}$ of the foliage close to the ground [Fig.
-`r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
+`r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @jjohnstonWhatLiesVerticalInpress].
 One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
-However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2020].
+However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVerticalInpress].
 
@@
 In addition, shading by the canopy can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
-Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020].
+Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVerticalInpress].
 As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based Forward Looking Infrared camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
-In a savanna system in the western U.S., @johnstonWhatLiesVertical2020 found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
+In a savanna system in the western U.S., @johnstonWhatLiesVerticalInpress found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
 `r SIfig_NEON_all`).

### 2022-02-19 vertical-thermal-review.Rmd
citation update

--- 2022-02-19 vertical-thermal-review.Rmd
@@
 This holds true across the range of forest types [Fig.
-`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020].
+`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalinpress].
 Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile.
@@
 Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the $T_{leaf}$ of the foliage close to the ground [Fig.
-`r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020].
+`r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
 One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
-However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2020].
+However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVerticalinpress].
 
@@
 In addition, shading by the canopy can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
-Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020].
+Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVerticalinpress].
 As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based Forward Looking Infrared camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
-In a savanna system in the western U.S., @johnstonWhatLiesVertical2020 found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
+In a savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
 `r SIfig_NEON_all`).

### 2022-02-19 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ## 3. Leaf traits
-<!--Lawren's paragraph for R2 comments on having older citations and connecting to newer work-->
+
 Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].

### 2022-02-20 vertical-thermal-review.Rmd
citations

@@
 Several recent model developments are at the cutting edge of representing vertical variation.
-Model developments by @Chen_et_al_2016, @Bonan_et_al_2018, and @Longo_et_al_2019 implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into the vegetated air space and the humidity of the airspace, modulated by turbulence processes within the forest and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
+Model developments by @chenColumnCanopyAirTurbulent2019, @bonanModelingCanopyinducedTurbulence2018, and @longoBiophysicsEcologyBiogeochemistry2019 implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into the vegetated air space and the humidity of the airspace, modulated by turbulence processes within the forest and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
 

### 2022-02-20 vertical-thermal-review.Rmd
updating word count

@@
 ---|---|-----|--------------------
-Total word count (excluding summary, references and legends) | 	8565 |	No. of figures | 	6 (all colour)
+Total word count (excluding summary, references and legends) | 	9560 |	No. of figures | 	6 (all colour)
 Summary|  	200 | 	No. of Tables | 	2  
-Introduction|	496 |	No of Supporting Information files | 	6
-Review of vertical gradients |	5976	||	 
-Implications |	1753	 ||
-Conclusions |	156	||
+Introduction|	391 |	No of Supporting Information files | 	6
+Review of vertical gradients |	6484	||	 
+Implications |	2408	 ||
+Conclusions |	276	||
 Acknowledgments | ##	||
@@
 
-Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006; @aroraCarbonConcentrationCarboon2020].
+Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006; @aroraCarbonConcentrationCarbon2020].
 Dynamic global vegetation models (DGVMs), which comprise the land surface models in Earth system models, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].

### 2022-02-20 vertical-thermal-review.Rmd
acknowledgments word count

@@
 Conclusions |	276	||
-Acknowledgments | ##	||
+Acknowledgments | 43	||
 

### 2022-02-20 vertical-thermal-review.Rmd
made leaf optics shorter

@@
 ## 3.2. Leaf optical properties
-<!---need to cut parts of this paragraph or merge into 3.1?-->
-Leaf optical properties are influenced by a combination of anatomical, morphological, and biochemical traits that vary vertically, as outlined in sections 3.1 and 3.3.
-Along the vertical light gradient, leaf light absorptance is balanced with $T_{leaf}$ regulation, where excess light can impair photosynthetic machinery (see sections 3.4 and 3.5), and is only modestly greater in sun leaves or mid-canopy leaves than shade leaves (Table `r table_leaf_traits`).
-In higher light environment, PAR absorptance and thus heat load is reduced by reflecting light through leaf surface modifications as mentioned above, such as trichome and cuticle layer, or through biochemically dissipating heat by caretenoids [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
-Further, there is less need for light absorption efficiency, but greater need to channel the higher availability of light into deeper cell layers through "light pipes" or palisade layers that increase path length for light penetration, contributing to higher photosynthetic capacity.
-The thickness of sun leaves also means that  light transmittance is lower than in shade leaves [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
-In contrast, where PAR is more limiting, leaf light absorption efficiency is greater, as shade leaves maintain higher SLA, denser layers of chloroplasts, shorter, fewer layers of conical palisade cells, and higher concentrations of nitrogen, and chlorophyll (Table `r table_leaf_traits`).
-Thus, while canopy leaves tend to invest more energy in metabolic activity and heat dissipation, understory leaves invest in light capture.
-Therefore, shade leaves can have light harvesting capabilities comparable to those of sun leaves, with higher absorption efficiency per unit biomass [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
+Leaf optical properties are influenced by a combination of anatomical, morphological, and biochemical traits, all of which vary throughout the canopy, as outlined in sections 3.1 and 3.3.
+Higher up in the canopy, a modestly greater leaf light absorptance can impair photosynthetic machinery and would be in part mitigated by $T_{leaf}$ regulation (see sections 3.4 and 3.5; Table `r table_leaf_traits`).
+For leaves in higher light environments, light absorptance and thus heat load can be reduced by leaf surface modifications that increase reflectance, as mentioned above, such as trichome and cuticle, and heat may be dissipated biochemically  by carotenoids, including xanthophylls [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
+Further, sun leaves allocate toward a thicker lamina with more layers of palisade cells, which act as “light pipes”, and channel the abundant light into deeper cell layers, contributing to higher photosynthetic capacity  [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
+In contrast, where shade leaves develop a  higher SLA, gaining in light absorption efficiency per mass investment, as well as denser layers of chloroplasts, and a greater proportion of spongy mesophyll, more effective for capturing diffuse light (Table `r table_leaf_traits`) [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
 

### 2022-02-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ---|---|-----|--------------------
-Total word count (excluding summary, references and legends) | 	9560 |	No. of figures | 	6 (all colour)
+Total word count (excluding summary, references and legends) | 	9559 |	No. of figures | 	6 (all colour)
 Summary|  	200 | 	No. of Tables | 	2  
@@
 Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.
-*Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`)](schematics/Schematic figure 1.png)
+*Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`).](schematics/Schematic figure 1.png)
 
@@
 Dots along each line correspond to the biophysical constant assigned for the independent variable in the other scenarios, and therefore represent the modeled value for $T_{Leaf}-T_{air}$ with all parameters set to the biophysical constants given in the table.
-Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), cleaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair).
+Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair).
 ](leaf_energy_balance/fig3_leafenergy.png)
@@
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`; @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
 Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007).
+Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
 Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
-As canopy shade is known to alter these trait relationships (@osnasDivergentDriversLeaf2018), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
+As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
 

### 2022-02-20 vertical-thermal-review.Rmd
update small changes

@@
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
-Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
+Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
 As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.
@@
 Leaf optical properties are influenced by a combination of anatomical, morphological, and biochemical traits, all of which vary throughout the canopy, as outlined in sections 3.1 and 3.3.
-Higher up in the canopy, a modestly greater leaf light absorptance can impair photosynthetic machinery and would be in part mitigated by $T_{leaf}$ regulation (see sections 3.4 and 3.5; Table `r table_leaf_traits`).
+Higher up in the canopy, a modestly greater leaf light absorptance can impair photosynthetic machinery and would be in part be mitigated by $T_{leaf}$ regulation (see sections 3.4 and 3.5; Table `r table_leaf_traits`).
 For leaves in higher light environments, light absorptance and thus heat load can be reduced by leaf surface modifications that increase reflectance, as mentioned above, such as trichome and cuticle, and heat may be dissipated biochemically  by carotenoids, including xanthophylls [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
@@
 {maybe this would be better as a topic sentence without a ref?
-but there is one Ty suggested-->].
+but there is one Ty suggested-->.
 Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].

### 2022-02-20 response-to-reviews.Rmd
Update response-to-reviews.Rmd
small things

@@
 
-*[issue #82](https://github.com/EcoClimLab/vertical-thermal-review/issues/82)*
-
 ### Minor details that I hope will be of interest for the authors. Consider them as you see fit.
@@
 
-**This will be done by the New Phytologist editorial staff (because it is a Tansley review). We will work with them to ensure that all the figures look good.**
+**This will be done by the New Phytologist illustrators (because it is a Tansley review). We will work with them to ensure that all the figures look good.**
 
@@
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`; @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table 1, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
 Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].

### 2022-02-20 vertical-thermal-review.Rmd
updated domingues citation in table

@@
 nocite: |
-  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010; @brooksVerticalGradientsPhotosynthetic1997; @carswellPhotosyntheticCapacityCentral2000
+  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016a; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016a; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003; @niinemetsHighWithincanopyVariation2010; @brooksVerticalGradientsPhotosynthetic1997; @carswellPhotosyntheticCapacityCentral2000; @dominguesParameterizationCanopyStructure2005
 ---
@@
 Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989].
+`r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005].
 In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig.
@@
 Simpraga et al.
-2013
+2013; 64.
+Domingues et al.
+2005
 

### 2022-02-20 vertical-thermal-review.Rmd
removed duplicate citation

@@
 In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
-`r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
+`r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
 Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves. 
@@
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
 Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
@@
 
-Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @binLeafTraitExpression2022], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a], and (3) across understory and canopy species [e.g., @lloydOptimisationPhotosyntheticCarbon2010].
+Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @binLeafTraitExpression2022], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010], and (3) across understory and canopy species [e.g., @lloydOptimisationPhotosyntheticCarbon2010].
 The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
@@
 Thus, sun leaves are generally smaller and thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
-`r fig_leaf_T`, @vogelSunLeavesShade1968a; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces $g_b$).
+`r fig_leaf_T`, @vogelSunLeavesShade1968; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces $g_b$).
 Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
@@
 Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
-diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
+diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
 Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.

### 2022-02-21 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], within-canopies of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
+Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], nor within-canopies of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the top of the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
-@hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 
+@hernandezSimilarTemperatureDependence2020 speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 
 Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of midday stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
@@
 This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
-Overall, however, we currently lack synthetic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
+Overall, however, we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
 
@@
 Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
-However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to shed excess heat may prove disadvantageous. 
+However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
 Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.
@@
 
-Owing to differences in the representation of forest vertical strata, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes. 
+Owing to differences in the representation of forest vertical strata, DVGMs vary in their capacity to incorporate vertical variation in leaf traits and physiological processes. 
 In general, however, this variation is accounted for via light competition. 
@@
 diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
 Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].
@@
 Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models. 
-This is likely due in large part to the fact that observational and experimental studies required to resolve key patterns and underlying mechanisms remain somewhat sparse (see sections 4, 5).
+This is likely due in large part to the fact that observational and experimental studies required to resolve key patterns and underlying mechanisms remain sparse (see sections 4, 5).
 Pairing of models and observational studies will be important for further improving our mechanistic understanding of vertical gradients and their implications.
@@
 As the climate changes, patterns and processes across these vertical gradients are likely to shift as well.
-Under the historical temperatures to which tree species are acclimated, the canopy is an advantageous place for trees to have their crowns, as it affords higher photosynthesis and growth. 
+Under the historical temperatures to which tree species are adapted, the canopy is an advantageous place for trees to have their crowns, as it affords higher photosynthesis and growth. 
 However, as temperatures increase, it is likely that exposed canopy positions will become physiologically stressful for increasing proportions of time.

### 2022-02-22 response-to-reviews.Rmd
small updates

@@
 
-# (temporary section)
-
-Nidhi, I like to use the following system for reviewer responses:
-
-*italics = comments/ notes/ tentative responses that still require work*
-
-**bold= completed responses**
-
-For issues that require discussion, it's helpful to open an issue and link to it under the reviewer comment.
-
-\newpage
-
 Dear Editor:
@@
 
-- We have brought in an additional author, Thomas Buckley, to round out the expertise of the author team and help address some of the reviewer comments (particularly Reviewer 2).
+- We have brought in an additional author, Thomas N. Buckley, to round out the expertise of the author team and help address some of the reviewer comments (particularly Reviewer 2).
 
@@
 - **Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
-diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
+diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
 Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer.
@@
 Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
-diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a].
+diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
 Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer.
@@
 
-- **Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
-Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species (ROS) [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
+- **Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table 1).
+Further, by reducing photosynthesis, heat and VPD-associated stomatal closure can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007] <!--Nidhi, I feel like this sentence needs a reference.
+{maybe this would be better as a topic sentence without a ref?
+but there is one Ty suggested-->.
+Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
 A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
 This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
-Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
+Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
 Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table 1, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10˚C) during sunflecks [@waySunflecksTreesForests2012].
-In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
+In contrast, in order to maximize carbon uptake and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
 
@@
 
-- **Notably, this pattern can be reversed in open forests.
+- **Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open.
 Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
@@
 
-**We agree with the suggestions. Therefore, to make it more clearer, we added photoprotection as one of the factors in Fig.1 instead of VAZ and VOC in the same line.**
+**We agree with the suggestions.
+Therefore, to make it more clearer, we added photoprotection as one of the factors in Fig.1 and isoprene emission in a separate line instead of VAZ and VOC in the same line.**
 
@@
 
-**Thank you for this suggestion, we added the following information:**
-**"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~, where both tend to be higher in the lower canopy positions.
-Particularly, understory seedlings above ground benefit from elevated CO~2~, therefore showing highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]."**
+**Vertical CO~2~ concentration gradients can influence the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
+Particularly, understory seedlings benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]."**
 
@@
 **We were using the term “buffer” in a functional or biological sense, meaning “protect,” rather than as a description of the underlying physics; i.e., the canopy protects the understory from large diel swings in both air temperature and radiation.
-We have reworded this passage to clarify: “Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum Tair more than open canopies, i.e., dense canopy understories can have~~ cooler maximum daytime Tair and warmer nighttime minimum Tair than open canopy understories or nearby clearings.”**
+We have reworded this passage to clarify: “Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~ cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.”**
 
@@
 
-**We added the description for the variables as part of the figure caption: Variables in the inset table are now described in the figure caption as: "Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf width (lw), stomatal conductance (gs), air temperature (tair)."**
+**We added the description for the variables as part of the figure caption: Variables in the inset table are now described in the figure caption as: "Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair)."**
 
@@
 
-*(this remains to be done: https://github.com/EcoClimLab/vertical-thermal-review/issues/84)*
-
 Table 1: What is here is not leaf gas exchange, and it seems that the authors have repeated the caption for Table 2 in error. Please fix this. Also don’t capitalise ‘leaf’.
@@
 
-**Thank you for the suggestion. We have added additional information highlighting older work and recent advances in the paragraph below**
+**Thank you for the suggestion. We have added additional information highlighting older work and recent advances in the paragraph below:**
 
-**This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+**Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table 1, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
+This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
 Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
@@
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table 1, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table 1, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
 Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007).
+Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
-Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
-As canopy shade is known to alter these trait relationships (@osnasDivergentDriversLeaf2018), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.**
+Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
+As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.**
 
@@
 
-**Thank you for noting this, we revised the sentence to make it clear, where we specify that majority of the studies are within species between sun and shade leaves of the same trees.
+**Thank you for noting this, we revised the sentence to make it clear (please see below), where we specify that majority of the studies are within species between sun and shade leaves of the same trees.
 @carterExperimentalWarmingTropical2021 is one of the only studies comparing between overstory and understory.**
@@
 - **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, which  compare sun and shade leaves of the same trees, reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes**
-
-- **We also made similar changes specifying within canopy sun and shade from lines (xx-xx)**
 

### 2022-02-22 response-to-reviews.Rmd
line numbers

@@
 
-**We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines xxx-xxx), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles.
+**We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines 169-192), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles.
 We have also modified the subsequent paragraph about wind speeds to eliminate the use of the confusing term “buffering”.**
@@
 
-- **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line xxx)***(new text italicized)**
-- **“Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances, ***attenuation of radiation and vertical transport of sensible heat by the canopy can buffer the lower canopy and understory from large diurnal swings in air temperature***~~it can be significantly buffered by forest canopies~~"; (line xxx)**
-- **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line xxx)**
+- **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line 127)***(new text italicized)**
+- **“It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature" (line 229)**
+- **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line 232)**
 

### 2022-02-22 vertical-thermal-review.Rmd

@@
 ---|---|-----|--------------------
-Total word count (excluding summary, references and legends) | 	9559 |	No. of figures | 	6 (all colour)
+Total word count (excluding summary, references and legends) | 	9495 |	No. of figures | 	6 (all colour)
 Summary|  	200 | 	No. of Tables | 	2  
-Introduction|	391 |	No of Supporting Information files | 	6
-Review of vertical gradients |	6484	||	 
+Introduction|	393 |	No of Supporting Information files | 	6
+Review of vertical gradients |	6374	||	 
 Implications |	2408	 ||

### 2022-03-01 vertical-thermal-review.Rmd
redo word count
@NidhiVinod , I just redid a careful word count using Word on my Mac, and it's much higher than what we had, matching what Holly said. I think you did the most recent one, but I know I did some word counts on the first submission that agreed with what you were getting -- or at least so I thought. But Holly is right that we're way over. How did we get so far off with the word count?

@@
 ---|---|-----|--------------------
-Total word count (excluding summary, references and legends) | 	9495 |	No. of figures | 	6 (all colour)
+Total word count (excluding summary, references and legends) | 	11854 |	No. of figures | 	6 (all colour)
 Summary|  	200 | 	No. of Tables | 	2  
-Introduction|	393 |	No of Supporting Information files | 	6
-Review of vertical gradients |	6374	||	 
-Implications |	2408	 ||
-Conclusions |	276	||
-Acknowledgments | 43	||
+Introduction|	498 |	No of Supporting Information files | 	6
+Review of vertical gradients |	8192	||	 
+Implications |	2842	 ||
+Conclusions |	278	||
+Acknowledgments | 44	||
 

### 2022-03-23 vertical-thermal-review.Rmd
another fix to word count
commenting out citations

@@
 ---|---|-----|--------------------
-Total word count (excluding summary, references and legends) | 	11854 |	No. of figures | 	6 (all colour)
+Total word count (excluding summary, references and legends) | 	11810 |	No. of figures | 	6 (all colour)
 Summary|  	200 | 	No. of Tables | 	2  
@@
 It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [@missonPartitioningForestCarbon2007; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021].
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007-->; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019-->; @zellweger_seasonal_2019; <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->].
 Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].

### 2022-03-23 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers [e.g., OSBS, Fig.
-`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, @hadleyInfluenceKrummholzMat1987; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015; @mullerEvidenceEfficientNonevaporative2021].
+`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, <!--@hadleyInfluenceKrummholzMat1987-->; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015-->; @mullerEvidenceEfficientNonevaporative2021].
 These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.
@@
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018].
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999-->; @doughtyAreTropicalForests2008; <!--@fauset_differences_2018-->; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; <!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
 This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [Fig.
-`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
+`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, <!--@hadleyInfluenceKrummholzMat1987-->; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
 The latter can result from a combination of still air at lower heights (Fig.

### 2022-03-23 vertical-thermal-review.Rmd
commenting out citations

@@
 Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs.
-shade leaves [@carterExperimentalWarmingTropical2021; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
+shade leaves [<!--@carterExperimentalWarmingTropical2021-->; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006-->], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
 This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].

### 2022-03-27 vertical-thermal-review.Rmd
moving content to SI

@@
 `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
-Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005].
-In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig.
-`r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
-Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].

### 2022-03-27 vertical-thermal-review.Rmd
moving yellow to Notes_S1

@@
 #Appendices
+
+Notes_S1<- "Notes S1"
+
 appendix_NEON_methods <- "S1"
@@
 Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig.
-`r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
+`r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`, `r Notes_S1`).
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].
@@
 
-Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
-The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
-`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022; @belandVerticalStructureHeterogeneity2021].
+Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor (`r Notes_S1`).
 Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers.

### 2022-03-27 vertical-thermal-review_SI.Rmd
moving content to SI
moving yellow to Notes_S1

@@
 
+Notes_S1<- "Notes S1"
+
 appendix_NEON_methods <- "S1"
@@
 
+# `r Notes_S1`
+
+## Note S1: Vertical variation in forest structure across forests types
+
+Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
+`r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005].
+In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig.
+`r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
+Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
+
+The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
+`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022; @belandVerticalStructureHeterogeneity2021].
+
+\newpage
 #  `r appendix_NEON_methods_title`

### 2022-03-29 vertical-thermal-review.Rmd
cutting words from Tom's suggestion
Please let me know what you think of these changes Krista!

--- 2022-03-23 vertical-thermal-review.Rmd
@@
 
-Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.
-Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed +1.5$^\circ$C in the next 20 years [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014].
-These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
-The net result of higher temperatures may be increased or decreased tree growth and forest carbon sequestration, with decreases being more commonly documented across the world's forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021].
-The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
+Forest responses and feedbacks to climate change will critically influence Earth's furture climate.
+Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach +1.5$^\circ$C by 2040 [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014].
+These changes are expected to affect tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+Warming may increase or decreased tree growth and forest carbon sequestration on balance, with decreases being more common globally [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021].
+The resulting feedbacks to climate and carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
 
-A great measure of uncertainty arises due to the differential impacts of rising temperatures on trees of differential size and crown position, especially as small understory trees exist in microenvironments that are substantially buffered by more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].
-Forests are vertically and horizontally stratified, with canopies playing a crucial role in moderating forest climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
-This creates a vertical stratification of the biophysical environment such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
-Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+Much uncertainty arises because effects of warming on trees depend on size and crown position. 
+Forests are vertically and horizontally stratified, and canopies moderate climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
+The resulting vertical stratification of the biophysical environment such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations that influences leaf traits, thermoregulation and metabolism, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
+Although these vertical gradients strongly influence plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
 Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
 
-Here, we review how the biophysical environment and plant form and function vary across the vertical gradient in forests. 
-We focus on five key themes (Fig. `r fig_schematic`):
+Here, we review vertical gradients in the biophysical environment and plant form and function in forests, focusing on five key themes (Fig. `r fig_schematic`):
 (1) the biophysical environment;
@@
 (5) tree and ecosystem ecology.
-We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.
+We then consider the implications for understanding forest responses to global change and how they scale across space and time.
 
@@
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
+
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].
@@
 Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
-The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
-`r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @binLeafTraitExpression2022; @belandVerticalStructureHeterogeneity2021].
 Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers.
@@
 It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007-->; @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019-->; @zellweger_seasonal_2019; <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->].
+Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->].
 Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
@@
 This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers [e.g., OSBS, Fig.
-`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, <!--@hadleyInfluenceKrummholzMat1987-->; @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015-->; @mullerEvidenceEfficientNonevaporative2021].
+`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, <!--@hadleyInfluenceKrummholzMat1987--> @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015-->; @mullerEvidenceEfficientNonevaporative2021].
 These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.

### 2022-03-29 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies across the vertical gradient from the forest floor to the top of the canopy (Figs.
-`r fig_schematic`, `r fig_NEON_vertical`), with physical conditions in large part determined by the structure of the forest.
-In this section, we supplement a review of the existing literature with a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S.
+The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies vertically from the forest floor to the top of the canopy (Figs.
+`r fig_schematic`, `r fig_NEON_vertical`), with physical conditions largely determined by forest structure.
+In this section, we review of the existing literature and provide a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S.
 National Ecological Observatory Network (NEON; Fig.
 `r fig_NEON_vertical`, Supporting Information Methods `r appendix_NEON_methods`, Supporting Information Figure `r SIfig_NEON_all`).
-While the focus here is on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer edge of vegetation than to height [@lowmanForestCanopies1995].
+Although we focus on vertical gradients, it is important to note that in heterogeneous canopies with high gap fractions and large variation in tree height, or at forest edges, the biophysical environment can be more closely linked to the distance from the outer edge of vegetation than to height [@lowmanForestCanopies1995].
 
@@
 
-Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy.
-It is critical in influencing -- and is influenced by -- the vertical biophysical gradient (Fig.
+Canopy foliage is the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy.
+It strongly influences -- and is influenced by -- the vertical biophysical gradient (Fig.
 `r fig_NEON_vertical`).
-Leaf area density (*i.e.*, leaf area per unit volume) patterns along the vertical gradient are heterogeneous across forests (Fig.
-`r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`).
-Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
+Vertical patterns in leaf area density (*i.e.*, leaf area per unit volume) differ across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`). 
+In tropical and temperate forests with dense broadleaf canopies, leaf area density is generally highest in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
 `r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005].
-In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be predominantly accumulated in the lower canopy or understory layers [e.g., PUUM and WREF, Fig.
+In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be greatest in the lower canopy or understory [e.g., PUUM and WREF, Fig.
 `r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
-
 In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].
@@
 
-Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
-Canopy foliage absorb a large portion of the direct light in PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy and understory layers.
-Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories receive diffused light enriched in near infrared radiation, and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
+The intensity of visible and photosynthetically active radiation (PAR, 400-700nm) decreases from the canopy top to the forest floor, with a profile whose shape is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @belandVerticalStructureHeterogeneity2021; @koike_leaf_2001; @binLeafTraitExpression2022].
+Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters light, thereby altering the spectrum of PAR creceived in the lower canopy and understory layers. 
+The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along qith total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation (700-1000 nm)  [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
 Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
@@
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
-The implications of this light gradient are that leaf traits vary along the vertical gradient (Table 1), and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
+This light gradient drives variation in leaf traits (Table 1), photosynthetic capacity (see section 4.2) and leaf temperature (see sections 2.2, 4.2).
 
 Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature might also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles.
-Vertical transport at the macroscopic scale of a forest canopy is dominated by turbulent transport rather than by molecular diffusion. 
-Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients (eddy diffusivities, often symbolized K) that depend on both friction and buoyancy [@penmanWeatherWheatEssay1960].
-In this “K-theory” regime, wind speed declines logarithmically with proximity to the ground or canopy surface [@monteithPrinciplesEnvironmentalPhysics2013].
+Vertical transport in forest canopies is dominated by turbulent transport. 
+Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients (eddy diffusivities, often symbolized K) [@penmanWeatherWheatEssay1960]-- such that wind speed declines logarithmically with proximity to the canopy surface [@monteithPrinciplesEnvironmentalPhysics2013].
 Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-However, K-theory often fails within canopies, making forest vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
-For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat moving vertically upwards through the canopy [e.g., @raupachLagrangianAnalysisScalar1987].
+Within canopies, however, transport often fails to mimic diffusion – e.g., transport of momentum [@shawSecondaryWindSpeed1977] or heat [e.g., @raupachLagrangianAnalysisScalar1987] may occur in the direction opposite to that predicted by gradients of wind speed or temperature, respectively – making vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
 Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991].
@@
 `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalinpress].
-Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile.
-The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat (see sections 2.2, 4.1).
+Thus, upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat exchange (see sections 2.2, 4.1).
 

### 2022-03-29 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Atmospheric concentrations of CO~2~ and water vapor can also vary across the vertical gradient.
+Atmospheric concentrations of CO~2~ and water vapor can also vary vertically.
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Vertical CO~2~ concentration gradients can influence the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
-Particularly, understory seedlings benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
-However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.
+Vertical CO~2~ concentration gradients can drive gradients in the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and in rates of refixation of respired CO~2~ vertically, both of which tend to be higher at lower heights (Table 1).
+Particularly, understory seedlings benefit from elevated CO~2~ by having higher rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
+However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations concentrations likely have little effect on energy balance and metabolism across the vertical gradient.
 
@@
 
-Vertical gradients in air temperature ($T_{air}$) across forest profiles are variable (Fig.
-`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), being dependent upon factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019].
-It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
+Vertical gradients in air temperature ($T_{air}$) differ widely (Fig.
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`) and depend on factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019].
+Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
 Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->].
 Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
-Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
-This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers [e.g., OSBS, Fig.
-`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b, <!--@hadleyInfluenceKrummholzMat1987--> @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015-->; @mullerEvidenceEfficientNonevaporative2021].
+Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$, due to turbulent air mixing and interception of thermal radiative flux from the soil and the canopy by lower-canopy layers [e.g., OSBS, Fig.
+`r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b; <!--@hadleyInfluenceKrummholzMat1987--> @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015--> @mullerEvidenceEfficientNonevaporative2021].
 These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.

### 2022-03-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 $T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4).
-Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015a; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020].
+Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for photosynthesis and metabolism [@michaletzPlantThermoregulationEnergetics2015a <!--@perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020-->].
 Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].
-Leaves can be warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
-While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020].
+Leaves can be warmer than air in full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
+$T_{leaf}$ is usually within a few degrees of $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
 

### 2022-03-30 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig.
+Fundamentally, $T_{leaf}$ is determined by the energy balance and can be estimated based on biophysical principles: $T_{leaf}-T_{air}$ depends on energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig.
 `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].
-High $R_n$ loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a). 
+High $R_n$ loads can elevate $T_{leaf}$ above $T_{air}$ (Fig. `r fig_leaf_T`a). 
 Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and under higher wind speeds (Fig. `r fig_leaf_T`b). 
-Latent heat flux ($\lambda E$) through transpiration has a strong cooling effect, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD:$g_b$ increases with wind speed, and $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
+Latent heat flux ($\lambda E$) through transpiration cools the leaf, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD:$g_b$ increases with wind speed, and $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
 Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig.
@@
 Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
-With adequate water, high $\lambda E$ can facilitate heat dissipation, particularly for leaves with smaller size (higher $g_b$) [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
-However, when leaf water demand (determined by VPD and stomatal opening) exceeds the rate of supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig.
-`r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, at high solar radiation loads, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at maximum $g_s$, but during $g_s$ limitation, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].
+With adequate water, high $\lambda E$ can dissipate excess heat, particularly for smaller leaves (higher $g_b$) [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+However, when transpiration exceeds water supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
+Therefore, at high solar radiation loads, latent cooling can maintain $T_{leaf}$ closer to $T_{air}$, but if stomatal opening is limited, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].
 
@@
 
-The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.
-High radiation at the top of the vertical profile (Fig.
-`r fig_NEON_vertical`) implies that upper canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig.
+The bbiophysical principles outlined above determine how $T_{leaf}$ is above $T_{air}$ across vertical gradients in closed-canopy forests.
+High radiation in the upper canopy (Fig.
+`r fig_NEON_vertical`) implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig.
 `r fig_leaf_T`).
-However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
+However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce leaf-air temperature differences [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
 In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
-In contrast, at lower heights along the vertical profile, greater RH and lower wind speeds may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig.
+In contrast, in the lower canopy, greater RH and lower wind speeds may limit $\lambda E$ (through reduced VPD and $g_b$), increasing $T_{leaf}$ [Fig.
 `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Leaves in these normally-shaded locations experience rapid light and temperature surges (can exceed 10$^\circ$C) during sunflecks [@waySunflecksTreesForests2012].
+Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks (>10$^\circ$C)  [@waySunflecksTreesForests2012].
 
-The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions.
-Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the upper canopy compared to the understory (Fig.
+Elevation of $T_{leaf}$ above $T_{air}$ is not uniform across time and space, but varies with micrometeorological conditions.
+During drought (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, $T_{leaf} - T_{air}$ will tend to be elevated in the upper canopy compared to the understory (Fig.
 `r fig_leaf_T`).
@@
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
+In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, limit evaporative cooling  [Fig.
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
 Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves. 
-Limitations in heat dissipation, in combination with their physiological propensities, may cause shade leaves to experience above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
+Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal $T_{leaf}$ under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
 In addition to lower wind speeds (Fig.
 `r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig.
-`r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig.
+`r fig_NEON_vertical`) also inhibits evaporative cooling by reducinf VPD, thus increasing $T_{leaf}$-$T_{air}$ [Fig.
 `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].
@@
 Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
-Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999-->; @doughtyAreTropicalForests2008; <!--@fauset_differences_2018-->; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; <!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
-In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
-This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [Fig.
-`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, <!--@hadleyInfluenceKrummholzMat1987-->; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
+Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most incoming radiation and preventing vertical air mixing in the understory.
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than lower-canopy leaves, in some cases exceeding photosynthetic temperature optima ($T_{opt}$) because of $g_s$ limitation and high radiation [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999-->; @doughtyAreTropicalForests2008; <!--@fauset_differences_2018-->; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; <!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
+In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory, which can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [Fig.
+`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`; <!--@hadleyInfluenceKrummholzMat1987--> @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
 The latter can result from a combination of still air at lower heights (Fig.
 `r fig_NEON_vertical`e), and sunflecks or the 'canopy greenhouse effect', as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the $T_{leaf}$ of the foliage close to the ground [Fig.
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing $T_{leaf}$ of foliage near the ground [Fig.
 `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
-One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
-However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVerticalinpress].
+One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense subalpine stand in the alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
+However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVerticalinpress].
 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
-This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
-Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
-More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
+Anatomical, structural and biochemical leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
+Vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
+More recent work has extended the focus from light alone to leaf height in the canopy, as important in determining differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
-Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
+Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [Table `r table_leaf_traits`, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
+Overall, leaves may optimize their biochemistry, anatomy and structure for their local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @binLeafTraitExpression2022], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010], and (3) across understory and canopy species [e.g., @lloydOptimisationPhotosyntheticCarbon2010].
-The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
-Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010]; however, vertical trends vary across trees [@lloydOptimisationPhotosyntheticCarbon2010], and counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
-Overall, when it comes to understanding and modeling forest ecosystem function, many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010].
+Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @binLeafTraitExpression2022], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010], and (3) among understory and canopy species [e.g., @lloydOptimisationPhotosyntheticCarbon2010].
+Most studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010]; however, vertical trends vary across trees [@lloydOptimisationPhotosyntheticCarbon2010], and counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
+Many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010].
 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`).
-Thus, sun leaves are generally smaller and thicker, with higher leaf mass per area, and are more deeply lobed [Fig.
+Sun leaves have anatomical, morphological, and physiological traits that reduce $T_{leaf}$ through higher conductance ($g_b$ or maximal $g_s$) and/or reflectance (Table `r table_leaf_traits`). 
+Thus, sun leaves are generally smaller and thicker, with higher LMA, and are more deeply lobed [Fig.
 `r fig_leaf_T`, @vogelSunLeavesShade1968; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces $g_b$).
-Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
-Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004].
-At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [reduces $g_b$ or increases boundary layer turbulence, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
-In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013].
-In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum $T_{air}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
+Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, also decreasing radiation load.
+Further, sun leaves tend to have higher stomatal and vein densities, which fascilitate evaporative cooling by enabling higher $g_s$ [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004]. 
+Yet sun leaves also have adaptations to limit water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [which may reduce or increase $g_b$; @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
+In contrast, shade leaves have traits that maximize light capture (e.g., lower LMA and greater light absorptance per unit biomass), but larger leaves and lower transpiration make them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013].
+In open canopies, where light is comparatively homogeneous, leaf traits may be shaped more by maximum $T_{air}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 ## 3.2. Leaf optical properties
-Leaf optical properties are influenced by a combination of anatomical, morphological, and biochemical traits, all of which vary throughout the canopy, as outlined in sections 3.1 and 3.3.
-Higher up in the canopy, a modestly greater leaf light absorptance can impair photosynthetic machinery and would be in part be mitigated by $T_{leaf}$ regulation (see sections 3.4 and 3.5; Table `r table_leaf_traits`).
-For leaves in higher light environments, light absorptance and thus heat load can be reduced by leaf surface modifications that increase reflectance, as mentioned above, such as trichome and cuticle, and heat may be dissipated biochemically  by carotenoids, including xanthophylls [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
-Further, sun leaves allocate toward a thicker lamina with more layers of palisade cells, which act as “light pipes”, and channel the abundant light into deeper cell layers, contributing to higher photosynthetic capacity  [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
-In contrast, where shade leaves develop a  higher SLA, gaining in light absorption efficiency per mass investment, as well as denser layers of chloroplasts, and a greater proportion of spongy mesophyll, more effective for capturing diffuse light (Table `r table_leaf_traits`) [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
+Leaf optical properties are influenced by anatomical, morphological, and biochemical traits, that vary throughout the canopy, as outlined in sections 3.1 and 3.3.
+High in the canopy, high light absorptance can lead to photoinhibition and would be in part be mitigated by $T_{leaf}$ regulation (see sections 3.4 and 3.5; Table `r table_leaf_traits`). 
+For leaves in higher light environments, light absorptance and thus heat load can be reduced by leaf surface modifications that increase reflectance, such as trichome and cuticle, and heat may be dissipated biochemically  by carotenoids, including xanthophylls [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
+Further, sun leaves tend to be thicker, with more palisade layers, which act as “light pipes”, and channel abundant light into deeper cell layers, enhancing photosynthetic capacity  [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
+In contrast, shade leaves have lower LMA and absorb more light per unit mass investment, as well as denser chloroplasts layers and a greater proportion of spongy mesophyll, more effective for capturing diffuse light (Table `r table_leaf_traits`) [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
 
@@
 Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
-Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthesis [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
-Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile contribute to the reater PAR absorptance efficiency of shade leaves (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues (Table `r table_leaf_traits`).
+Sun leaves have higher concentrations area-based concentrations of elements like nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical for respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthesis [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
+Increases in chlorophyll a/b ratios with height reflect greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile increase PAR absorptance efficiency of shade leaves (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, reduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table `r table_leaf_traits`).
 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 
-Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Further, by reducing photosynthesis, heat and VPD-associated stomatal closure can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007] <!--Nidhi, I feel like this sentence needs a reference.
+Biochemical protection against light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
+Further, by reducing photosynthesis, VPD-associated stomatal closure can reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007] <!--Nidhi, I feel like this sentence needs a reference.
 {maybe this would be better as a topic sentence without a ref?
 but there is one Ty suggested-->.
-Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
-A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
-This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
+Accumulation of excess light energy causes overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
+A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by excess light [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
+This form of NPQ entails interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
 Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
 Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, in order to maximize carbon uptake and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 <!--VOCs-->
-The capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
+Heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017].
 In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017].
-Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].
+Due to its much higher emission rate and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by many angiosperms and gymnosperms across all biomes [see @taylorNewFieldInstrument2021 and references therein].
 
-Less than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile is influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
-Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
-However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_traits`, @taylorNewFieldInstrument2021].
-Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_traits`, @simpragaVerticalCanopyGradient2013].
+Fewer than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that vertical variation in emissions is influenced by both vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
+Within species, isoprene emission tends to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species are more abundant in hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
+However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy of an Amazonian forest [Table `r table_leaf_traits`, @taylorNewFieldInstrument2021].
+Similarly, within European beech crowns, monoterpene emissions were highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_traits`, @simpragaVerticalCanopyGradient2013]. 
 This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
-Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
-Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
-Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play similar role in sun-exposed canopy leaves.
+Temperature-sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+Future work to understand how temporal dynamics of $T_{leaf}$ [see @leighInfluenceLeafSize2017] vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
+Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play similar roles in upper-canopy leaves.
 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 <!--short par giving background on thermal tolerance-->
-The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
-At very high $T_{leaf}$ (~40-60$^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
-Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [e.g., @slotLeafHeatTolerance2021].
-Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and with higher thermal tolerance among leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
+Upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning via using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
+Very high $T_{leaf}$ (~40-60$^{\circ}$C) irreversibly damages photosystem II, leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
+Thermal tolerance is described in terms of the $T_{leaf}s$ at which photosystem II efficiency starts to decrease ($T_{crit}$) is reduced by 50%, $T_{50}$ [e.g., @slotLeafHeatTolerance2021]. 
+Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and being greater in leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
 $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
-However, they appear to be are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
+However, they appear more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
 <!--For example, *Quercus muehlenbergii* Engelm.
@@
 <!--par on thermal tolerance across vertical gradients-->
-Based on their variation with the temperatures experienced by leaves, thermal tolerance metrics are hypothesized to be greatest for the leaves that face the greatest radiation and highest temperatures along the vertical gradient.
+Thermal tolerance metrics are hypothesized to be greatest for the leaves with the greatest radiation loads and highest temperatures along the vertical gradient. 
 Indeed, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama, $T_{50}$ was 0.7 - 1.4$^\circ$C lower for shade than sun leaves for two of three tree species [Table `r table_leaf_traits`, @slotPhotosyntheticHeatTolerance2019].
-Similarly, along a vertical gradient within a very open canopy in Australia, where trees of *Acacia papyrocarpa* Benth.
-faced the greatest composite climate stress (low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation) in the lower, north-facing portions of their crowns, $T_{50}$ was highest in these positions, *albiet* by <1$^\circ$C [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
-These modest difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments imply that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
+Similarly, within a very open canopy in *Acacia papyrocarpa* Benth.
+canopy, $T_{50}$ was highest (*albiet* by <1$^\circ$C) in the lower, north-facing leaves, which faced the greatest composite climate stress (low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation) [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
+These modest differences in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments imply that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
 However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.
@@
 In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
-Early leaf-out gives tree saplings and seedlings a window for disproportionately high photosynthesis -- amounting to the majority of annual carbon fixation for some seedlings -- before canopy foliage and reduces light availability [@augspurgerDifferencesLeafPhenology2003; @leeSpringPhenologicalEscape2021].
-Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
-Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of deciduous species [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
-Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019;@tangLightdrivenGrowthAmazon2017;@nunesForestFragmentationImpacts2022].
+Early leaf-out gives saplings and seedlings a window for disproportionately high photosynthesis -- contributing the majority of annual carbon fixation for some seedlings -- before canopy foliage and reduces light availability [@augspurgerDifferencesLeafPhenology2003; @leeSpringPhenologicalEscape2021].
+Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals is greater in larger trees in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
+Similarly, fractions of deciduous species increase with tree size class [@conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018].
+Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct vertical patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019;@tangLightdrivenGrowthAmazon2017;@nunesForestFragmentationImpacts2022].
 For instance, in a central Amazonian forest, the upper canopies of both interior and edge forests shed leaves when maximum daily $T_{air}$ exceeded ~35$^\circ$C [@nunesForestFragmentationImpacts2022].
-Leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
-Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; @mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015].
+Beacsue leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
+Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; <!--@mariasImpactsLeafAge2017;--> @zhouLeafageEffectsTemperature2015].
 Leaf phenology thereby both responds to and influences the vertical structure of forest microenvironments and function.

### 2022-03-31 vertical-thermal-review.Rmd
Tom's edits

@@
 
-Leaf hydraulic, stomatal and boundary layer conductances are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
+Leaf hydraulic, stomatal and boundary layer conductances are all critical for regulating $T_{leaf}$ via latent heat loss (Fig. `r fig_leaf_T`). 
 Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
 Maximum $g_s$ increases with light--typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979]--and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
-However, water supply cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves with fully open stomata, because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or because of increased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations [@buckleyStomatalOptimisationRelation2014].
-These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.3 (Table 1).
+However, water supply cannot meet the demands caused by the high radiation and $g_b$ in sun leaves with fully open stomata, because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014].
+These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and $\delta^{13}C$ discussed in section 3.3 (Table 1).
 Stomatal depression reduces transpirational cooling (Fig.
-`r fig_leaf_T`), thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than do shade leaves in lower vertical positions [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
+`r fig_leaf_T`), thus amplifying the warming of sun leaves by high radiation [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
+The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than lower-canopy shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
 This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
@@
 
-Temperature can affect photosynthesis via direct and indirect pathways: directly, by altering photosynthetic enzyme activity and the electron transport chain, and indirectly through increased VPD causing stomatal closure [@lloydEffectsRisingTemperatures2008].
+Temperature can affect photosynthesis directly, by altering photosynthetic enzyme activity and the electron transport chain,  and indirectly, by increasing VPD and closing stomata [@lloydEffectsRisingTemperatures2008].
 Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
-Beyond the optimum, photosynthesis decreases as a result of stomatal closure [ @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; @fredeenTemperatureHumidityEffects1999], and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
+Beyond the optimum, photosynthesis decreases due to stomatal closure [ @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; <!--@fredeenTemperatureHumidityEffects1999-->] and biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; <!--@sageTemperatureResponseC32007;--> @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
 
-We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, which  compare sun and shade leaves of the same trees, reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
+We have very little evidence about how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, which  compare sun and shade leaves of the same trees, reveal no pronounced overall trend with height in photosynthetic temperature optima  (Table `r table_leaf_metabolism`).
 Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
-@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [@campbell_introduction_1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
-However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
+@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019], one might expect sun leaves to have a stronger temperature-dependence and higher $T_{opt}$ than shade leaves [@campbell_introduction_1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+However, such a trend is not apparent among recent studies, where for three species in Panama, $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
 Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], nor within-canopies of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
@@
 Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs.
-shade leaves [<!--@carterExperimentalWarmingTropical2021-->; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006-->], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
+shade leaves [<!--@carterExperimentalWarmingTropical2021-->; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007 <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006-->], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
 This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
@@
 
-Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
-Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] photosynthesis, and intra-canopy trait variation [@binLeafTraitExpression2022].
+Tree metabolism and growth are shaped by crown location in the vertical gradient.
+Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which strongly predicts tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] photosynthesis, and intra-canopy trait variation [@binLeafTraitExpression2022].
 Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).

### 2022-03-31 vertical-thermal-review.Rmd
more edits to cutting

@@
 Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
-The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
+Photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
+Among these, the best-studied process is aboveground woody growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling because woody tissues have a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
 Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2022].
@@
 Vertical gradients also affect the climate sensitivity of metabolism and growth.
-Stomatal conductance can be strongly limited by high atmospheric demand (low RH, high $T_{air}$ and VPD) even when soil water is plentiful [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
-Therefore, periods of high atmospheric demand -- be these on time scales of hours, days, or seasons -- tend to cause greater reductions in transpiration and photosynthesis in tall trees that occupy canopy positions in closed-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
-This is consistent with observations that both dry season leaf loss (section 3.6, Table `r table_leaf_traits`) and $g_s$ limitation (section 4.1, Table `r table_leaf_metabolism`) increases with height across the vertical profile.
-More active regulation of transpiration by tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce the need for adjustment of traits related to carbon metabolism during drought [@bartholomewSmallTropicalForest2020].
-This results in decreasing carbon isotope discrimination, indicative of greater $g_s$ limitation, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
+Stomatal conductance can be strongly limited by high VPD even when soil water is plentiful [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016].
+Therefore, periods of high VPD, whether on time scales of hours, days, or seasons, tend to reduce transpiration and photosynthesis more in tall trees that occupy canopy positions in closed-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
+This is consistent with observations that both dry season leaf loss (section 3.6, Table `r table_leaf_traits`) and $g_s$ limitation (section 4.1, Table `r table_leaf_metabolism`) increase with height across the vertical profile.
+More active regulation of transpiration by tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce the need to adjust carbon metabolism traits during drought [@bartholomewSmallTropicalForest2020].
+This reduces carbon isotope discrimination, indicative of greater $g_s$ limitation, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
 
-In turn, the drought sensitivity of woody growth tends to be greater in canopy trees than in smaller trees with less exposed crowns. 
-In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976].
-However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
-These have most commonly found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig.
-`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimate2022; @heilmanEcologicalForecastingTree2022], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015].
-In addition to lower drought resistance of growth, larger trees frequently exhibit greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
-Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs.
-`r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2022], although there is also evidence that short trees in young stands can be more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
+In turn, the drought sensitivity of woody growth tends to be greater in upper-canopy trees than in smaller trees with less exposed crowns. 
+Dendrochronologists generally agree tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth is most sensitive to interannual climate variation [@frittsTreeRingsClimate1976].
+However, few studies have directly examined drought- or temperature-sensitivities as a function of tree size.
+Most have found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig.
+`r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimate2022; @heilmanEcologicalForecastingTree2022], corroborating evidence from global forest censuses that drought reduces growth more in large trees [@bennettLargerTreesSuffer2015].
+Drought also causes greater mortality in larger trees [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
+Mechanistically, this is likely driven in part by larger trees having crown in microenvironments that are more challenging during drought [Figs.
+`r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet height itself also creates disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2022], although some evidence indicates that short trees in young stands can be more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
 
-Although it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
+Although drought sensitivity clearly increases with crown height and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies along the vertical gradient.
 High $T_{air}$ is often associated with high VPD and atmospheric drought, likely explaining negative growth responses to $T_{air}$ -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.
 `r fig_treerings`, @anderson-teixeiraJointEffectsClimate2022].
-However, under well-watered conditions conducive to high $g_s$ in canopy trees, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts.
+However, under well-watered conditions conducive to high $g_s$, we do not necessarily expect higher thermal sensitivity of photosynthesis (see section 4.2) or woody growth in upper-canopy trees than their understory counterparts.
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
@@
 Canopy trees account for the majority of forest ecosystem water and carbon cycling.
-While studies partitioning transpiration across forest vertical gradients are rare, both evapo-transpiration (ET) and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
-Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
+While studies partitioning transpiration across forest vertical gradients are rare, both evapo-transpiration (ET) and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, where the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011].
+Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed ~93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
 In terms of C cycling, it has been estimated that the canopy strata contributes $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007].
-Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
+Large trees also dominate woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
 `r fig_ecosystem`b, @piponiotDistributionBiomassDynamicsinpress].
@@
 Thus, for example, increases in canopy temperature reduce forest GPP in the tropics [@pauTropicalForestTemperature2018]. 
-Yet there is also evidence that GPP and ecosystem respiration are less sensitive to heat and drought stress in older forests than in young forests [@xuSeasonalVariabilityForest2020], perhaps in part because of more complex vertical structuring [@jonesManagingRedPine2019;@nunesForestFragmentationImpacts2022].
-A rare example of a study comparing the climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an *El Niño* drought in Panama, while the smallest size classes had higher productivity during the drought, likely because of increased light in the understory.
+Yet evidence also suggests GPP and ecosystem respiration are less sensitive to heat and drought stress in older forests than in young forests [@xuSeasonalVariabilityForest2020], perhaps in part because of more complex vertical structuring [@jonesManagingRedPine2019;@nunesForestFragmentationImpacts2022].
+A rare example of a study comparing climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an *El Niño* drought in Panama, while the smallest trees had higher productivity during the drought, likely because of increased light in the understory.
 It remains far less clear how thermal sensitivity varies across forest strata under wet conditions, but both physiological mechanisms (Fig.
 `r fig_leaf_T`) and tree ring evidence (Fig.
-`r fig_treerings`) suggests that understory trees may respond more negatively to hot, wet conditions.
+`r fig_treerings`) suggest that understory trees may respond more negatively to hot, wet conditions.
 Further research is required to better understand the thermal sensitivity of forest ecosystem function across strata.
@@
 Having established how physical conditions and biological form and function vary across vertical gradients (Fig.
-`r fig_schematic`), we now turn attention to the implications of these patterns for our understanding of how forest ecosystems may be impacted by global change, and our ability to project this across space and time.
+`r fig_schematic`), we now turn to the implications of these patterns for understanding how forest ecosystems may be impacted by global change, and our ability to project this across space and time.
 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
-Our in-depth review sheds some light as to how responses are likely to vary across forest vertical gradients, yet important uncertainties remain.
+Our in-depth review sheds some light on how responses are likely to vary across forest vertical gradients, yet important uncertainties remain.
 
-To the extent that warming is coupled to drought, we expect that the tallest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
+To the extent that warming is coupled to drought, we expect that the tallest trees to usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
 `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
-Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
+Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
 In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022].
@@
 While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016].
-Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.
+Given the mechanisms reviewed here, we might expect that VPD in particular skews the stress more towards the exposed canopy trees.
 
-What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
-The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to the local climate but do not vary consistently with height in the vertical profile (see section 4.2, Table 2).
-Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021].
-While leaves display substantial plasticity to adapt to warmer temperatures [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+It remains poorly understood how responses to warming will vary across vertical gradients under mesic conditions. 
+Leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to local climate but do not vary consistently with height (see section 4.2, Table 2).
+Beyond the optimal temperature, photosynthesis declines whereas respiration continues to increase, shifting the carbon balance from source to sink, independent of water and irradiance [@duffyHowCloseAre2021].
+While leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will reduce carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
 Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
-Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
+Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
 
@@
 However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
-Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.
-While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), limited tree-ring evidence indicates that understory trees can exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
+<!--Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
+While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (see section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
 Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].
 
-Thus, in synthesis, warming temperatures will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
-We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, while much larger uncertainty remains surrounding the resoponses of understory trees.
-Both groups -- but probably disproportionately the canopy trees -- are likely to experience increasing mortality, with the modes of mortality mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@goraImplicationsSizedependentTree2021].
-When the canopy trees die, this will open canopy gaps, as discussed in the following section.
+Thus, in synthesis, warming will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
+We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$, while greater uncertainty remains surrounding the resoponses of understory trees.
+Both groups -- but probably disproportionately the upper-canopy trees -- are likely to experience increasing mortality, with patterns mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@goraImplicationsSizedependentTree2021].
+Death of canopy trees will open canopy gaps, as discussed in the following section.
 

### 2022-03-31 vertical-thermal-review.Rmd
warming section edits

@@
 Rising temperatures and severe droughts place canopy trees at particularly elevated risk of mortality. 
-Moreover, large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging operations [e.g., @millerReducedImpactLogging2011].
-Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
+Large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging [e.g., @millerReducedImpactLogging2011].
+Increasing severity and frequency of heat waves, accompanied with increases in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
-Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases incident radiation levels and temperatures within the forest [@juckerCanopyStructureTopography2018; @starkReframingTropicalSavannization2020].
-Such changes often result in increased growth of smaller trees, which benefit from increased light [@hoganDroughtInterannualVariability2019;@bennettLargerTreesSuffer2015; @nunesForestFragmentationImpacts2022], and wetter forests can prove quite resilient to canopy disturbance [@millerReducedImpactLogging2011].
+Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases radiation and temperatures within the forest [@juckerCanopyStructureTopography2018; @starkReframingTropicalSavannization2020].
+Such changes often enhance growth of smaller trees, which benefit from increased light [@hoganDroughtInterannualVariability2019;@bennettLargerTreesSuffer2015; @nunesForestFragmentationImpacts2022], and wetter forests can prove quite resilient to canopy disturbance [@millerReducedImpactLogging2011].
 However, this shift to hotter and drier microclimates makes some forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018].
-Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Such dynamics are likely to be amplified by warming temperatures, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may have reduced probability of recovering and persisting when affected by severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020; @floresFloodplainsAchillesHeel2017].
+Severe degradation can cause dramatic ecological state changes (e.g., the transition of tropical forests more open, savanna-like vegetation) and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
+Such dynamics will likely to be amplified by warming, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may not recover and persist after severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020; @floresFloodplainsAchillesHeel2017].
 
@@
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020b; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
-If such compositional shifts towards more thermally-adapted species fail to keep up with the pace of warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g.,  @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy structure and the potential state changes described above.
+In the understory, warming disproportionately affects less thermally-adapted plant species, causing thermophilization of the plant community [@zellwegerForestMicroclimateDynamics2020b; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
+If compositional shifts towards more thermally-adapted species fail to keep pace with warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g.,  @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy structure and the potential state changes described above.
 
@@
 As we have reviewed here, vertical profiles in forests strongly shape forest dynamics and climate change responses.
-Ultimately, to achieve the important goal of understanding feedbacks between the world's forests and climate change, it is essential that these mechanisms be sufficiently represented in models and scaled spatially via remote sensing.
+Ultimately, to achieve the important goal of understanding feedbacks between the world's forests and climate change, these mechanisms must be sufficiently represented in models and scaled spatially via remote sensing.
 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_schematic`, `r fig_NEON_vertical`), with physical conditions largely determined by forest structure.
-In this section, we review of the existing literature and provide a new analysis of data on vegetation structure and vertical microclimate profiles from focal sites within the U.S.
+In this section, we review of the existing literature and analyze data on vegetation structure and vertical microclimate profiles from focal sites within the U.S.
 National Ecological Observatory Network (NEON; Fig.
@@
 
-Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of the future of global carbon cycling and climate change [@friedlingsteinClimateCarbonCycle2006; @aroraCarbonConcentrationCarbon2020].
+Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of global carbon cycling [@friedlingsteinClimateCarbonCycle2006; @aroraCarbonConcentrationCarbon2020].
 Dynamic global vegetation models (DGVMs), which comprise the land surface models in Earth system models, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].
-DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, capturing spatial variability in the light environment and microclimates at high vertical and horizontal spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
-This simplification is computationally more efficient, although it does not always capture observed vertical profiles (sections 1-4; for example, vertical shifts in the balance between stomatal conductance and photosynthetic capacity, see sections 4.1-4.2) and comes at a cost of a lack of representation of important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
-The computational middle-ground to representing vertical structure in DVGMs lies in cohort-based models, which represent vegetation as cohorts of individual plants, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].
-These sit between the oversimplified vegetation dynamics in big-leaf models, which do not represent any vertical stratification, and individual-based models, whose computational expense prevents them from being integrated into Earth system models.
+DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, which capturing spatial variability in light environment and microclimates at high 3D spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure to single vegetation layer, implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
+This simplification is computationally more efficient, but does not always capture observed vertical profiles (sections 1-4; for example, vertical shifts in the balance between stomatal conductance and photosynthetic capacity, see sections 4.1-4.2) and cannot capture important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
+The computational middle-ground lies in cohort-based models, which represent vegetation as cohorts of individuals, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].
 
@@
 In general, however, this variation is accounted for via light competition. 
-Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
+Models partition radiation above and within the forest vertical profile (i.e., direct vs.
 diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
 Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].
-In recent years, there has been a growing effort to incorporate vertical variation more directly in models, specifically as it pertains to leaf water potential and light absorption by leaves, along with a general increasing interest in making direct comparison between models and field measurements [@fisherPerspectivesFutureLand2020].
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].
+Recent years have seen growing efforts to incorporate vertical variation more directly in models, specifically involving leaf water potential and light absorption, along with a increasing interest in confronting models directly with field measurements [@fisherPerspectivesFutureLand2020].
 Several recent model developments are at the cutting edge of representing vertical variation.
-Model developments by @chenColumnCanopyAirTurbulent2019, @bonanModelingCanopyinducedTurbulence2018, and @longoBiophysicsEcologyBiogeochemistry2019 implement vertical gradients of irradiance, water content, leaf temperature, and the feedback between the evaporation of water into the vegetated air space and the humidity of the airspace, modulated by turbulence processes within the forest and a roughness layer that extends to roughly twice the height of the canopy [@fisherPerspectivesFutureLand2020].
+Model developments by @chenColumnCanopyAirTurbulent2019, @bonanModelingCanopyinducedTurbulence2018, and @longoBiophysicsEcologyBiogeochemistry2019 implement vertical gradients of irradiance, water content, $T_{leaf},$ and humidification of canopy air by transpiration, modulated by turbulence within the forest and a roughness layer that extends to roughly twice the canopy height [@fisherPerspectivesFutureLand2020].
 
 The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.
-Improved representation of vertical variation in forest structure and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
-Moreover, given the anticipated importance of mid-canopy and understory trees in ecosystem resilience to increasing mortality of canopy trees, it is absolutely essential that models separately represent these strata.
-However, doing so will require improved understanding of the mechanisms controlling vertical gradients.
-
-A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.
-Most models have been developed to capture dynamics in the canopy, given the disproportionate role of these large trees in ecosystem dynamics (Fig. `r fig_ecosystem`). 
+Improved representation of vertical variation in forest structure and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as important for reducing uncertainty and accuratly characterizating of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021].
+Moreover, given the anticipated importance of mid-canopy and understory trees in ecosystem resilience given increasing mortality of canopy trees, it is essential that models separately represent these strata.
+A key question is whether existing models adequately represent the processes that underpin understory and large tree responses to thermal stress.
+Most models have been developed to capture dynamics in the upper canopy, given the disproportionate role of large trees in ecosystem dynamics (Fig. `r fig_ecosystem`). 
 Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models. 
-This is likely due in large part to the fact that observational and experimental studies required to resolve key patterns and underlying mechanisms remain sparse (see sections 4, 5).
-Pairing of models and observational studies will be important for further improving our mechanistic understanding of vertical gradients and their implications.
+This is due in large part to the paucity of observational and experimental studies needed to resolve key patterns and underlying mechanisms (see sections 4, 5).
+Pairing of models with observational studies is key to further improving our mechanistic understanding of vertical gradients and their implications.
 
 Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers [e.g., sunlit and shaded portions, @depurySimpleScalingPhotosynthesis1997] may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, Table 1, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, Table 2).
-At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @bonanMovingIncorrectUseful2021].
-In addition, capturing the vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.
+Multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @bonanMovingIncorrectUseful2021].
+In addition, capturing vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.
 `r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.
@@
 Airborne and spaceborne lidar, as well as terrestrial laser scanning data, yield detailed 3D reconstructions of whole tree and forest structure. 
-These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based Forward Looking Infrared cameras.
+These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based Forward Looking Infrared (FLIR) cameras.
 
@@
 
-Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
-In addition, shading by the canopy can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
+Detailed structural information from lidar data can be used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
+Canopy shading can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
 Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVerticalinpress].
-As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based Forward Looking Infrared camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
+As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
 In a savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
 `r SIfig_NEON_all`).
-The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of *in situ* measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
+The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of *in situ* measurements reviewed above to explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
 
@@
 `r fig_schematic`).
-However, there remains a lot of uncertainty as to how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
+However, uncertainty remains as to how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
 Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth. 

### 2022-03-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Remote sensing data provide a valuable means to scale between *in situ* observations and DGVMs.
-Specifically, the increasing availability of airborne and spaceborne lidar and thermal remote sensing data offer a promising opportunity for mapping vertical thermal gradients in combination with vegetation structure at locations with *in situ* data and across larger landscapes (Fig.
+Specifically, the increasing availability of airborne and spaceborne lidar and thermal remote sensing data offer a promising opportunity for mapping vertical thermal gradients and vegetation structure across scales (Fig.
 `r fig_model`).
-Airborne and spaceborne lidar, as well as terrestrial laser scanning data, yield detailed 3D reconstructions of whole tree and forest structure. 
-These data can be leveraged in combination with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based Forward Looking Infrared (FLIR) cameras.
+Airborne, spaceborne and terrestrial laser scanning data yield detailed 3D reconstructions of whole tree and forest structure. 
+These data can be combined with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based Forward Looking Infrared (FLIR) cameras.
 
@@
 Canopy shading can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
-Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVerticalinpress].
-As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
-In a savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
+Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although few applications exist at finer-scales needed to understand vertical variation in canopy temperatures [@johnstonWhatLiesVerticalinpress].
+As an example of the type of insight possible with this approach, @pauTropicalForestTemperature2018 used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest GPP was more strongly associated with canopy temperature than $T_{air}$ or VPD.
+In a a western-U.S.
+savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies, together with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
 `r SIfig_NEON_all`).
-The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of *in situ* measurements reviewed above to explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
+The growing availability of such data makes this an opportune time to link *in situ* measurements with fine- and landscape-scale measurements to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
 
@@
 `r fig_schematic`).
-However, uncertainty remains as to how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
-Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth. 
-While the preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
+However, uncertainty remains about how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
+Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth. 
+While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
 `r fig_treerings`).
 
-As the climate changes, patterns and processes across these vertical gradients are likely to shift as well.
-Under the historical temperatures to which tree species are adapted, the canopy is an advantageous place for trees to have their crowns, as it affords higher photosynthesis and growth. 
-However, as temperatures increase, it is likely that exposed canopy positions will become physiologically stressful for increasing proportions of time.
-Ensuant increased mortality of canopy trees will create an increasing number of gaps, resulting in changes to understory conditions and community composition.
-Ultimately, mid- and understory tree communities will be critical to the resilience of forest ecosystems under changing climate, making improved understanding and model representation of their dynamics critical to understanding future forest dynamics.
-Integrating the patterns and mechanisms reviewed here, along with remote sensing data on forest structure and thermal environments, into cohort-based models that integrate with Earth system models will be critical to understanding and forecasting forest-climate feedbacks in the coming decades.
+As climate changes, patterns and processes across vertical gradients will likely to shift as well.
+In the historical climates to which trees adapted, the canopy was an advantageous place for photosynthesis and growth. 
+However, as temperatures increase, it is likely that exposed canopy positions will more often be physiologically stressful.
+Ensuant increased mortality of canopy trees will create ever more gaps, changing understory conditions and community composition.
+Ultimately, mid- and understory tree communities will be critical to the resilience of forest ecosystems under changing climate, making improved understanding and model representation of their dynamics essential to understanding future forest dynamics.
+Integrating the patterns and mechanisms reviewed here, along with remote sensing data on forest structure and thermal environments, into cohort-based models that integrate with Earth system models will be crucial to understanding and forecasting forest-climate feedbacks in the coming decades.
 

### 2022-03-31 vertical-thermal-review.Rmd
cutting citations

@@
 Forest responses and feedbacks to climate change will critically influence Earth's furture climate.
-Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach +1.5$^\circ$C by 2040 [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChange20212021] and hotter droughts [@trenberthGlobalWarmingChanges2014].
-These changes are expected to affect tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach +1.5$^\circ$C by 2040 [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves and hotter droughts [@meehlMoreIntenseMore2004; @ipccClimateChange20212021].
+These changes are expected to affect tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; <!--@corlettImpactsWarmingTropical2011-->; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 Warming may increase or decreased tree growth and forest carbon sequestration on balance, with decreases being more common globally [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021].
@@
 In tropical and temperate forests with dense broadleaf canopies, leaf area density is generally highest in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a,@terborghVerticalComponentPlant1985; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005].
+`r fig_NEON_vertical`a,<!--@terborghVerticalComponentPlant1985;--> @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005].
 In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be greatest in the lower canopy or understory [e.g., PUUM and WREF, Fig.
@@
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [@parkerStructuralPhenologyLeaf2004; @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [<!--@parkerStructuralPhenologyLeaf2004;--> @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].
 In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.

### 2022-03-31 response-to-reviews.Rmd
updating the responses

@@
 
-- **Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
+- **Models partition radiation above and within the forest vertical profile (i.e., direct vs.
 diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].
+**
 

### 2022-03-31 response-to-reviews.Rmd
small update

@@
 **We revised and added references to this sentence in question, which now reads: **
-**"Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020]."**
+**Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].**
 
@@
 
-**We have reworded this sentence to read, "In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig.
-3b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]."**
+**We have reworded this sentence to read, "In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig.
+`r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017]."**
 

### 2022-03-31 vertical-thermal-review.Rmd
small update

@@
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, limit evaporative cooling  [Fig.
+In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig.
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].

### 2022-03-31 vertical-thermal-review.Rmd
integrating a few suggestions from  ..._KAT shortening suggestions.docx
#113

--- 2022-03-31 vertical-thermal-review.Rmd
@@
 
+<!--This first paragraph is all good content, but mostly pretty basic background knowledge that could be cut if needed.-->
 Forest responses and feedbacks to climate change will critically influence Earth's furture climate.
@@
 As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased maximum temperatures, with extreme increases during heat waves [@ipccClimateChange20212021].
-These changes will impact leaf and ecosystem gas exchange with the atmosphere, and, over longer time frames, forest structure, composition, and leaf trait distributions. 
-Our in-depth review sheds some light on how responses are likely to vary across forest vertical gradients, yet important uncertainties remain.
-
 To the extent that warming is coupled to drought, we expect that the tallest trees to usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
 `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
-Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
 In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022].
@@
 
+<!--the following two paragraphs could be shortened/ condensed into one-->
 It remains poorly understood how responses to warming will vary across vertical gradients under mesic conditions. 
@@
 
+<!--consider deleting this whole paragraph-->
 Thus, in synthesis, warming will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
@@
 Airborne, spaceborne and terrestrial laser scanning data yield detailed 3D reconstructions of whole tree and forest structure. 
-These data can be combined with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based Forward Looking Infrared (FLIR) cameras.
+These data can be combined with thermal remote sensing data from the spaceborne ECOSTRESS sensor [@hulleyNewECOSTRESSMODIS2019; @fisherECOSTRESSNASANext2020] or drone- and tower-based infrared cameras.
 
@@
 Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although few applications exist at finer-scales needed to understand vertical variation in canopy temperatures [@johnstonWhatLiesVerticalinpress].
-As an example of the type of insight possible with this approach, @pauTropicalForestTemperature2018 used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest GPP was more strongly associated with canopy temperature than $T_{air}$ or VPD.
+As an example of the type of insight possible with this approach, @pauTropicalForestTemperature2018 used data from a tower-based infrared camera in combination with eddy-covariance data and found that tropical forest GPP was more strongly associated with canopy temperature than $T_{air}$ or VPD.
 In a a western-U.S.
-savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies, together with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
+savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the tops of tree crowns, together with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
 `r SIfig_NEON_all`).

### 2022-03-31 response-to-reviews.Rmd
updating reviewer's responses

@@
 
-**Thank you for the suggestion, we added the following here:"Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers.
-Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]."**
+**Thank you for the suggestion, we added the following here:"Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters light, thereby altering the spectrum of PAR creceived in the lower canopy and understory layers.
+The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along qith total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation (700-1000 nm)  [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]."**
 
@@
 **Thank you for noting the need to clarify this point.
-We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text italicized here).
+We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure to a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text italicized here).
 We also deleted the clause that occurred two sentences later, that had said about big-leaf models: “...which do not represent any vertical stratification…”**
@@
 
-- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes.
-In general, however, this variation is accounted for via light competition.
-Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs.
-diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**
+- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their capacity to incorporate vertical variation in leaf traits and physiological processes.In general, however, this variation is accounted for via light competition.
+Models partition radiation above and within the forest vertical profile (i.e., direct vs.
+diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].**
 
@@
 **We agree that there is a recognition in the modeling community that we do need improved understanding of the mechanisms to better represent the level of vertical variation desired by some.
-We added the following sentence to acknowledge that: "However, doing so will require improved understanding of the mechanisms controlling vertical gradients.".
+We added the following sentence to acknowledge that: "Moreover, given the anticipated importance of mid-canopy and understory trees in ecosystem resilience given increasing mortality of canopy trees, it is essential that models separately represent these strata.".
 We also removed some of the language that emphasizes cohort based models as the only or best way to do this.**
@@
 **We have also added the following sentence on canopy dieback to the canopy disturance section:**
-**"Increase in the severity and frequency of heat waves, accompanied with an increase in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021]."**
+**"Increasing severity and frequency of heat waves, accompanied with increases in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021]."**
 
@@
 
-- **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line 127)***(new text italicized)**
-- **“It is most commonly the case that attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature" (line 229)**
-- **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line 232)**
+- **“Canopy foliage is the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions experienced above the canopy.”; (line xxx)***(new text italicized)**
+- **“Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019]" (line 229)**
+- **“Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting from,***~~lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings.” (line 232)**
 

### 2022-03-31 vertical-thermal-review.Rmd
updating reviewer's responses

--- 2022-03-31 vertical-thermal-review.Rmd
@@
 In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig.
-`r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020].
+`r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017 <!-- @songTreeSurfaceTemperature2020-->].
 Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves. 

### 2022-03-31 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 
-**We have revised the first sentence here to read: "Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.
+**We have revised the first sentence here to read: "Biochemical protection against light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.
 We also choose to keep this rather obvious sentence because we feel that basic topic sentences are important for helping to guide readers from a diversity of sub-disciplines through the manuscript.**
@@
 
-- **Biochemical protection against foliage light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table 1).
-Further, by reducing photosynthesis, heat and VPD-associated stomatal closure can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007] <!--Nidhi, I feel like this sentence needs a reference.
-{maybe this would be better as a topic sentence without a ref?
-but there is one Ty suggested-->.
-Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
-A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
-This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
-Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
-Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table 1, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, in order to maximize carbon uptake and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
+- **Further, by reducing photosynthesis, VPD-associated stomatal closure can reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].Accumulation of excess light energy causes overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by excess light [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].This form of NPQ entails interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
+Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
 
@@
 
-- **"Implications: Global Change Responses : Warming" section, 4th par (previous lines 793-797: "While it is currently difficult to predict whether canopy or understory photosynthesis is likely to be more severely affected by higher $T_{air}$ (see section 4.2), *limited tree-ring evidence* indicates that understory trees *can* exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig 4b, @rollinsonClimateSensitivityUnderstory2020]."**
+- **"Implications: Global Change Responses : Warming" section, 4th par (previous lines 793-797: "While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (see section 4.2), *limited tree-ring evidence* indicates that understory trees *can* exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig 4b, @rollinsonClimateSensitivityUnderstory2020]."**
 
@@
 
-- **Conclusions section, first part (text specifically cited in the comment above): "Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth.
-While the preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
+- **Conclusions section, first part (text specifically cited in the comment above): "Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth.
+While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
 4)."**

### 2022-03-31 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 
-- **Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open.
-Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$.
-This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers**
+- **Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
+Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$, due to turbulent air mixing and interception of thermal radiative flux from the soil and the canopy by lower-canopy layers.**
 
@@
 - **$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4).
-Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes **
+Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for photosynthesis and metabolism [@michaletzPlantThermoregulationEnergetics2015a <!--@perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020-->].
+**
 
@@
 **We removed the second 'understory' word here:** 
-**"Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high ~~in the understory~~ as well"**
+**"In tropical and temperate forests with dense broadleaf canopies, leaf area density is generally highest in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high ~~in the understory~~ as well"**
 
@@
 **We added foliage clumping into the list of factors influencing vertical light gradient here:**
-**"Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
-The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types (Fig.
-2a-d, Supporting Information Figure S1, Koike et al., 2001)."**
+**"The intensity of visible and photosynthetically active radiation (PAR, 400-700nm) decreases from the canopy top to the forest floor, with a profile whose shape is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
+2a-d, Supporting Information Figure S1, @belandVerticalStructureHeterogeneity2021; @koike_leaf_2001; @binLeafTraitExpression2022]."**
 

### 2022-03-31 response-to-reviews.Rmd
updated comments

@@
 
-**We have reworded this sentence to read, "In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig.
-`r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017]."**
+**We have reworded this sentence to read, "In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig.3 b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017]."**
 
@@
 - **Further, by reducing photosynthesis, VPD-associated stomatal closure can reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].Accumulation of excess light energy causes overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by excess light [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].This form of NPQ entails interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
-Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
+Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table 1, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
 
@@
 
-- **Changed sentences: Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation or PAR, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
+- **Changed sentences: Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
 
@@
 
-**Vertical CO~2~ concentration gradients can influence the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and refixation of respired CO~2~ vertically, where both tend to be higher at lower heights (Table 1).
-Particularly, understory seedlings benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]."**
+**Vertical CO~2~ concentration gradients can drive gradients in the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and in rates of refixation of respired CO~2~ vertically, both of which tend to be higher at lower heights (Table 1).Particularly, understory seedlings benefit from elevated CO~2~ by having higher rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]."**
 
@@
 **This passage now occurs after the new paragraph about turbulent transport within canopies.
-We have reworded the passage as, “Despite the complexities of turbulent transport in canopies, wind speeds are generally much higher at the top of the canopy (Jiao-jun et al., 2004; Jucker et al., 2018).”**
+We have reworded the passage as, “Despite these complexities of within-canopy transport, wind speeds are generally much higher at the top of the canopy than within or beneath [@juckerCanopyStructureTopography2018; @jiao-junWindProfilesTrees2004].”**
 
@@
 **We were using the term “buffer” in a functional or biological sense, meaning “protect,” rather than as a description of the underlying physics; i.e., the canopy protects the understory from large diel swings in both air temperature and radiation.
-We have reworded this passage to clarify: “Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~ cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.”**
+We have reworded this passage to clarify: “Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting from***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~ lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings ”**
 
@@
 
-**Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table 1, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
-This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
-Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
-More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
+**Anatomical, structural and biochemical leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table 1, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].Vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].More recent work has extended the focus from light alone to leaf height in the canopy, as important in determining differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table 1, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
-Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
+Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [Table 1, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
+Overall, leaves may optimize their biochemistry, anatomy and structure for their local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
 Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
@@
 **We agree that the comparison is not very clear here.
-We added "compared to shaded lower canopies" to the sentence: "Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies ***compared to shaded lower canopies,***result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C isotopic ratios in leaf tissues**
+We added "compared to shaded lower canopies" to the sentence: "Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies ***compared to shaded lower canopies,***rreduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table 1).**
 
@@
 
-**This sentence has been rewritten as, “Stomatal closure reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads.”**
+**This sentence has been rewritten as, “Stomatal depression reduces transpirational cooling (Fig. 3), thus amplifying the warming of sun leaves by high radiation.”**
 
@@
 As a result, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO2 and carbon isotope discrimination discussed in section 3.1 (Table 1).
-Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in gs in response to rising Tair”, and clarified that “high Tair should decrease gs of ***upper canopy***leaves”.**
+Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$”, and clarified that “high Tair should decrease gs of ***upper canopy***leaves”.**
 

### 2022-03-31 response-to-reviews.Rmd
deleted the suggested para

@@
 
-**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al.
+**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ cannot meet the demands caused by the high radiation and $g_b$ in sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al.
 1994, Koch et al.
-2004, Sillett et al 2010) and/or because of increased evaporative demand (VPD) caused by leaf warming in sunlit canopy locations (Buckley et al.
-2014)***.
-As a result, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO2 and carbon isotope discrimination discussed in section 3.1 (Table 1).
-Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$”, and clarified that “high Tair should decrease gs of ***upper canopy***leaves”.**
+2004, Sillett et al 2010) and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014]***.
+These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO~2~ and $\delta^{13}C$ discussed in section 3.3 (Table 1).
+Stomatal depression reduces transpirational cooling (Fig.
+`r fig_leaf_T`), thus amplifying the warming of sun leaves by high radiation.
+~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$”, and clarified that “high $T_{air}$ should decrease $g_s$ of ***upper canopy***leaves”.**
 

### 2022-03-31 vertical-thermal-review.Rmd
deleted the suggested para

@@
 
-<!--consider deleting this whole paragraph-->
-Thus, in synthesis, warming will affect trees across the vertical gradient, but the stress will be of a different nature at different heights.
-We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$, while greater uncertainty remains surrounding the resoponses of understory trees.
-Both groups -- but probably disproportionately the upper-canopy trees -- are likely to experience increasing mortality, with patterns mirroring the current tendency for canopy trees to be more prone to climate-related disturbances including drought while understory trees are more prone to competition and carbon starvation [@goraImplicationsSizedependentTree2021].
-Death of canopy trees will open canopy gaps, as discussed in the following section.
-
 ### Canopy disturbance

### 2022-03-31 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 
-**This sentence has been revised to read, "In turn, the drought sensitivity of woody growth tends to be greater in canopy trees than in smaller trees with less exposed crowns."**
+**This sentence has been revised to read, "In turn, the drought sensitivity of woody growth tends to be greater in upper-canopy trees than in smaller trees with less exposed crowns."**
 
@@
 **As the point about resilience was not critical, we have removed that part of the statement.
-The sentence now reads, "In addition to lower drought resistance of growth, larger trees frequently exhibit greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019]."**
+The sentence now reads, "Drought also causes greater mortality in larger trees [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019]."**
 

### 2022-03-31 vertical-thermal-review.Rmd
update word count
@NidhiVinod , FYI
#113

@@
 ---|---|-----|--------------------
-Total word count (excluding summary, references and legends) | 	11810 |	No. of figures | 	6 (all colour)
+Total word count (excluding summary, references and legends) | 	10604 |	No. of figures | 	6 (all colour)
 Summary|  	200 | 	No. of Tables | 	2  
-Introduction|	498 |	No of Supporting Information files | 	6
-Review of vertical gradients |	8192	||	 
-Implications |	2842	 ||
-Conclusions |	278	||
+Introduction|	416 |	No of Supporting Information files | 	6
+Review of vertical gradients |	7541	||	 
+Implications |	2399	 ||
+Conclusions |	248	||
 Acknowledgments | 44	||

### 2022-03-31 response-to-reviews.Rmd

@@
 Stomatal depression reduces transpirational cooling (Fig.
-`r fig_leaf_T`), thus amplifying the warming of sun leaves by high radiation.
+3), thus amplifying the warming of sun leaves by high radiation.
 ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$”, and clarified that “high $T_{air}$ should decrease $g_s$ of ***upper canopy***leaves”.**
@@
 
-**This has been reworded to read, "However, there remains a lot of uncertainty as to how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
-Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth.
-While the preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
+**This has been reworded to read, "However, uncertainty remains about how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
+Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth.
+While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
 4)."**

### 2022-03-31 vertical-thermal-review.Rmd
moved 2.1 to notes

@@
 
-### 2.1. Biophysical drivers of $T_{leaf}$
-
-Fundamentally, $T_{leaf}$ is determined by the energy balance and can be estimated based on biophysical principles: $T_{leaf}-T_{air}$ depends on energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig.
-`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].
-High $R_n$ loads can elevate $T_{leaf}$ above $T_{air}$ (Fig. `r fig_leaf_T`a). 
-Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. `r fig_leaf_T`d) and under higher wind speeds (Fig. `r fig_leaf_T`b). 
-Latent heat flux ($\lambda E$) through transpiration cools the leaf, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD:$g_b$ increases with wind speed, and $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
-Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig.
-`r fig_leaf_T`b, @daudetWindSpeedLeaf1999], increases with RH (Fig.
-`r fig_leaf_T`c), increases with leaf size (Fig.
-`r fig_leaf_T`d), and $T_{leaf}$ decreases absolutely with $g_s$ (Fig.
-`r fig_leaf_T`e).
-
 ![**Figure `r fig_leaf_T`.
@@
 
-Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
-With adequate water, high $\lambda E$ can dissipate excess heat, particularly for smaller leaves (higher $g_b$) [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
-However, when transpiration exceeds water supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
-Therefore, at high solar radiation loads, latent cooling can maintain $T_{leaf}$ closer to $T_{air}$, but if stomatal opening is limited, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].
+### 2.1 Vertical gradients in leaf temperature
 
-### 2.2 Vertical gradients in leaf temperature
-
-The bbiophysical principles outlined above determine how $T_{leaf}$ is above $T_{air}$ across vertical gradients in closed-canopy forests.
+Biophysical principles (`r Notes_S1`) determine how elevated $T_{leaf}$ is above $T_{air}$ across vertical gradients in closed-canopy forests.
 High radiation in the upper canopy (Fig.
@@
 
+`r Notes_S1`
+
 `r appendix_NEON_methods_title`

### 2022-03-31 vertical-thermal-review_SI.Rmd
moved 2.1 to notes

--- 2022-02-19 vertical-thermal-review_SI.Rmd
@@
 # SI
+##Notes
+
+Notes_S1<- "Notes S1"
+
 ## appendices
@@
 ```
+
+\newpage
+
+# `r Notes_S1`: Biophysical drivers of $T_{leaf}$
+
+Fundamentally, $T_{leaf}$ is determined by the energy balance and can be estimated based on biophysical principles: $T_{leaf}-T_{air}$ depends on energy input from net radiation ($R_n$, including shortwave and longwave) minus heat lost to the environment [Fig.
+`r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019].
+High $R_n$ loads can elevate $T_{leaf}$ above $T_{air}$ (Fig. 3a). 
+Sensible heat flux between leaf and air is regulated by leaf boundary layer conductance, which is greater in smaller leaves (Fig. 3d) and under higher wind speeds (Fig. 3b). 
+Latent heat flux ($\lambda E$) through transpiration cools the leaf, and is determined by stomatal and boundary layer conductances ($g_s$ and $g_b$) and VPD:$g_b$ increases with wind speed, and $g_s$ generally declines as VPD increases due to stomatal closure [@darwinIXObservationsStomata1898; @mottStomatalResponsesHumidity1991].
+Therefore, $T_{leaf}-T_{air}$ decreases in magnitude with wind speed [Fig.
+3b, @daudetWindSpeedLeaf1999], increases with RH (Fig.
+`r fig_leaf_T`c), increases with leaf size (Fig.
+`r fig_leaf_T`d), and $T_{leaf}$ decreases absolutely with $g_s$ (Fig.
+3e).
+
+Under hot and dry conditions, leaves face a trade-off between $T_{leaf}$ regulation and water conservation [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
+With adequate water, high $\lambda E$ can dissipate excess heat, particularly for smaller leaves (higher $g_b$) [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021a; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].
+However, when transpiration exceeds water supply, stomata close to conserve water, increasing $T_{leaf}$ [Fig. 3e, @fauset_differences_2018].
+Therefore, at high solar radiation loads, latent cooling can maintain $T_{leaf}$ closer to $T_{air}$, but if stomatal opening is limited, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].
 

### 2022-04-01 vertical-thermal-review.Rmd
further work on Tleaf section

@@
 
-Notes_S1<- "Notes S1"
+Notes_leafTbiophysics<- "S1"
 
@@
 
-$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4).
+$T_{leaf}$ is strongly tied, and usually within a few degrees of, $T_{air}$ [@cavaleriColdbloodedForestsWarming2020], but can be substantially warmer or cooler under certain conditions (Fig.
+`r fig_leaf_T`; Note `r Notes_leafTbiophysics`).
+Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (Note `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig.
+`r fig_schematic`).
 Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for photosynthesis and metabolism [@michaletzPlantThermoregulationEnergetics2015a <!--@perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020-->].
-Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].
-Leaves can be warmer than air in full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
-$T_{leaf}$ is usually within a few degrees of $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].
 
@@
 
-### 2.1 Vertical gradients in leaf temperature
-
-Biophysical principles (`r Notes_S1`) determine how elevated $T_{leaf}$ is above $T_{air}$ across vertical gradients in closed-canopy forests.
+Biophysical principles (Note `r Notes_leafTbiophysics`) shape $T_{leaf}-T_{air}$ across vertical gradients in closed-canopy forests.
 High radiation in the upper canopy (Fig.
@@
 
-`r Notes_S1`
+`r Notes_leafTbiophysics`
 

### 2022-04-01 vertical-thermal-review_SI.Rmd
further work on Tleaf section

@@
 Therefore, at high solar radiation loads, latent cooling can maintain $T_{leaf}$ closer to $T_{air}$, but if stomatal opening is limited, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].
+
+Leaves can be substantially warmer or cooler under certain conditions. 
+They can be  warmer than air in full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
+Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].
 

### 2022-04-01 vertical-thermal-review.Rmd
replace reference
@NidhiVinod , here I replace a commentary on a paper with the paper itself.

@@
 
-$T_{leaf}$ is strongly tied, and usually within a few degrees of, $T_{air}$ [@cavaleriColdbloodedForestsWarming2020], but can be substantially warmer or cooler under certain conditions (Fig.
+$T_{leaf}$ is strongly tied, and usually within a few degrees of, $T_{air}$ [@drakeNoEvidenceHomeostatic2020], but can be substantially warmer or cooler under certain conditions (Fig.
 `r fig_leaf_T`; Note `r Notes_leafTbiophysics`).

### 2022-04-01 vertical-thermal-review.Rmd
remove thermoregulation sentence
@NidhiVinod , I'm removing this because the debate seems to be out on this (see last sentence of Cavaleri 2020), but in any case it's not really necessary to get into that here.
merge 1st and 2nd par of Tleaf section

@@
 `r fig_schematic`).
-Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for photosynthesis and metabolism [@michaletzPlantThermoregulationEnergetics2015a <!--@perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020-->].
-
-![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf characteristic dimension, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under hypothetical moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
-In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
-Dots along each line correspond to the biophysical constant assigned for the independent variable in the other scenarios, and therefore represent the modeled value for $T_{Leaf}-T_{air}$ with all parameters set to the biophysical constants given in the table.
-Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair).
-](leaf_energy_balance/fig3_leafenergy.png)
-
-Biophysical principles (Note `r Notes_leafTbiophysics`) shape $T_{leaf}-T_{air}$ across vertical gradients in closed-canopy forests.
 High radiation in the upper canopy (Fig.
@@
 Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks (>10$^\circ$C)  [@waySunflecksTreesForests2012].
+
+![**Figure `r fig_leaf_T`.
+Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf characteristic dimension, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under hypothetical moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
+Dots along each line correspond to the biophysical constant assigned for the independent variable in the other scenarios, and therefore represent the modeled value for $T_{Leaf}-T_{air}$ with all parameters set to the biophysical constants given in the table.
+Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair).
+](leaf_energy_balance/fig3_leafenergy.png)
 

### 2022-04-01 vertical-thermal-review.Rmd
wordsmithing Tleaf section

@@
 
-$T_{leaf}$ is strongly tied, and usually within a few degrees of, $T_{air}$ [@drakeNoEvidenceHomeostatic2020], but can be substantially warmer or cooler under certain conditions (Fig.
-`r fig_leaf_T`; Note `r Notes_leafTbiophysics`).
-Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (Note `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig.
+$T_{leaf}$ is strongly tied, and usually within a few degrees of, $T_{air}$ [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [@rey-sanchez_spatial_2016;@doughtyAreTropicalForests2008;@vogelLeavesLowestHighest2009].
+Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (Fig.
+`r fig_leaf_T`; Note `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig.
 `r fig_schematic`).
@@
 `r fig_NEON_vertical`) implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig.
-`r fig_leaf_T`).
-However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce leaf-air temperature differences [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
+`r fig_leaf_T`a).
+However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
 In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.

### 2022-04-01 vertical-thermal-review_SI.Rmd
wordsmithing Tleaf section

--- 2022-03-31 vertical-thermal-review_SI.Rmd
@@
 Therefore, at high solar radiation loads, latent cooling can maintain $T_{leaf}$ closer to $T_{air}$, but if stomatal opening is limited, solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021a].
+
+Leaves can be substantially warmer or cooler under certain conditions. 
+They can be  warmer than air in full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
+Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016].
 

### 2022-04-01 vertical-thermal-review.Rmd
update word count
@NidhiVinod , I set up to automatically populate the table so that it's easier to update the word count without counting the whole document

@@
 
+#Word count
+w_intro=414
+w_biophys=1582
+w_tleaf=758
+w_traits= 2477
+w_met=1000
+w_ecol=1352
+w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
+w_implications=2399
+w_conclusions=248
+w_total=w_intro+w_vertgradients+w_implications+w_conclusions
+
 ```
@@
 ---|---|-----|--------------------
-Total word count (excluding summary, references and legends) | 	10604 |	No. of figures | 	6 (all colour)
+Total word count (excluding summary, references and legends) | 	`r w_total` |	No. of figures | 	6 (all colour)
 Summary|  	200 | 	No. of Tables | 	2  
-Introduction|	416 |	No of Supporting Information files | 	6
-Review of vertical gradients |	7541	||	 
-Implications |	2399	 ||
-Conclusions |	248	||
+Introduction|	`r w_intro` |	No of Supporting Information files | 	6
+Review of vertical gradients |	`r w_vertgradients`	||	 
+Implications |	`r w_implications`	 ||
+Conclusions |	`r w_conclusions`	||
 Acknowledgments | 44	||
@@
 
-Elevation of $T_{leaf}$ above $T_{air}$ is not uniform across time and space, but varies with micrometeorological conditions. 
-During drought (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, $T_{leaf} - T_{air}$ will tend to be elevated in the upper canopy compared to the understory (Fig.
+Elevation of $T_{leaf}$ above $T_{air}$ varies with micrometeorological conditions. 
+During drought (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, $T_{leaf} - T_{air}$ will tend to be higher in the upper canopy compared to the understory (Fig.
 `r fig_leaf_T`).
@@
 In addition to lower wind speeds (Fig.
-`r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig.
-`r fig_NEON_vertical`) also inhibits evaporative cooling by reducinf VPD, thus increasing $T_{leaf}$-$T_{air}$ [Fig.
+`r fig_NEON_vertical`), higher RH (lower VPD) in the understory and inner canopy (Fig.
+`r fig_NEON_vertical`) also inhibits evaporative cooling, thus increasing $T_{leaf}-T_{air}$ [Fig.
 `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].
 
-Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
+Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure. 
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most incoming radiation and preventing vertical air mixing in the understory.

### 2022-04-01 vertical-thermal-review.Rmd
biophysical section: insert subheadings + wordsmithing
@NidhiVinod , I think adding subheadings here is helpful for breaking up this long section

@@
 w_intro=414
-w_biophys=1582
+w_biophys=1580
 w_tleaf=758
@@
 
+### 1.1 Foliage
+
 Canopy foliage is the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy.
@@
 
+### 1.2 Light
+
 The intensity of visible and photosynthetically active radiation (PAR, 400-700nm) decreases from the canopy top to the forest floor, with a profile whose shape is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
@@
 
-Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature might also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles.
-Vertical transport in forest canopies is dominated by turbulent transport. 
-Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients (eddy diffusivities, often symbolized K) [@penmanWeatherWheatEssay1960]-- such that wind speed declines logarithmically with proximity to the canopy surface [@monteithPrinciplesEnvironmentalPhysics2013].
+### 1.3 Turbulent transport and wind
+Vertical transport in forest canopies is dominated by turbulent transport, making vertical profiles of wind speed, water vapor, CO~2~, and temperature more difficult to predict and to generalize on the basis of theoretical first principles compared to radiation profiles.
+Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients [eddy diffusivities, @penmanWeatherWheatEssay1960]-- such that wind speed declines logarithmically with proximity to the canopy surface [@monteithPrinciplesEnvironmentalPhysics2013].
 Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-Within canopies, however, transport often fails to mimic diffusion – e.g., transport of momentum [@shawSecondaryWindSpeed1977] or heat [e.g., @raupachLagrangianAnalysisScalar1987] may occur in the direction opposite to that predicted by gradients of wind speed or temperature, respectively – making vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
-Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991].
+Within canopies, however, transport often fails to mimic diffusion – e.g., transport of momentum [@shawSecondaryWindSpeed1977] or heat [@raupachLagrangianAnalysisScalar1987] may occur in the direction opposite to that predicted by gradients of wind speed or temperature, respectively – making vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991].
 These "sweep-eject" events can couple understory conditions more directly to conditions above the canopy than beneath.
@@
 
-Atmospheric concentrations of CO~2~ and water vapor can also vary vertically.
+### 1.4 Atmospheric concentrations of CO~2~ and water vapor
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
@@
 
+### 1.5 Air temperature 
 Vertical gradients in air temperature ($T_{air}$) differ widely (Fig.

### 2022-04-01 vertical-thermal-review.Rmd
small changes

@@
 Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
-Typically, diel temperature range is smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->].
+Diel temperature range is typically smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->].
 Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
@@
 Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
-Photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
+Photosynthate production is allocated among functions including respiration, aboveground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
 Among these, the best-studied process is aboveground woody growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling because woody tissues have a long residence time in the ecosystem [@russellResidenceTimesDecay2014].

### 2022-04-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which strongly predicts tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] photosynthesis, and intra-canopy trait variation [@binLeafTraitExpression2022].
-Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
+Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where leaves receive more light and have higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
 Photosynthate production is allocated among functions including respiration, aboveground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).

### 2022-04-01 vertical-thermal-review.Rmd
merge two paragraphs in warming section
#113

@@
 w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
-w_implications=2399
+w_implications=2333
 w_conclusions=248
@@
 
-<!--the following two paragraphs could be shortened/ condensed into one-->
 It remains poorly understood how responses to warming will vary across vertical gradients under mesic conditions. 
-Leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adaptation to local climate but do not vary consistently with height (see section 4.2, Table 2).
-Beyond the optimal temperature, photosynthesis declines whereas respiration continues to increase, shifting the carbon balance from source to sink, independent of water and irradiance [@duffyHowCloseAre2021].
-While leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will reduce carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
-Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019].
-Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
-
-Although understory leaves are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), they will also be affected by higher $T_{air}$, and in some ways are likely to prove more sensitive.
+Leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], yet failure to fully acclimate will reduce carbon sequestration at leaf, tree, and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+As growing season $T_{air}$ increasingly equals or exceeds photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], exposed crowns are likely most vulnerable.
+Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and $T_{crit}$ [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
+In contrast, understory leaves and trees are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), but in some ways are likely to prove more sensitive to higher $T_{air}$.
 Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
@@
 While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (see section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
-Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, which have lower NSC reserves [@niinemetsResponsesForestTrees2010].
+Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].
 

### 2022-04-01 vertical-thermal-review.Rmd
fix section 2 references

@@
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
-This light gradient drives variation in leaf traits (Table 1), photosynthetic capacity (see section 4.2) and leaf temperature (see sections 2.2, 4.2).
+This light gradient drives variation in leaf traits (Table 1), photosynthetic capacity (see section 4.2) and leaf temperature (see sections 2, 4.2).
 
@@
 `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalinpress].
-Thus, upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat exchange (see sections 2.2, 4.1).
+Thus, upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat exchange (see sections 2, 4.1).
 
@@
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that upper canopy leaves have higher $g_b$ and associated potential for water and heat loss (see sections 2.2, 4.1).
+The implication is that upper canopy leaves have higher $g_b$ and associated potential for water and heat loss (see sections 2, 4.1).
 
@@
 Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
 

### 2022-04-01 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
-w_implications=2333
+w_imp_gc=1102
+w_imp_modeling_scaling=1231
+w_implications=w_imp_gc+w_imp_modeling_scaling
 w_conclusions=248
@@
 To the extent that warming is coupled to drought, we expect that the tallest trees to usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
-`r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimate2022].
+`r fig_schematic`- `r fig_treerings`].
 In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022].

### 2022-04-01 vertical-thermal-review.Rmd
"see section"-->"section"

@@
 w_intro=414
-w_biophys=1580
-w_tleaf=758
-w_traits= 2477
+w_biophys=1576
+w_tleaf=752
+w_traits= 2471
 w_met=1000
-w_ecol=1352
+w_ecol=1353
 w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
-w_imp_gc=1102
-w_imp_modeling_scaling=1231
+w_imp_gc=1088
+w_imp_modeling_scaling=1229
 w_implications=w_imp_gc+w_imp_modeling_scaling
@@
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
-This light gradient drives variation in leaf traits (Table 1), photosynthetic capacity (see section 4.2) and leaf temperature (see sections 2, 4.2).
+This light gradient drives variation in leaf traits (Table 1), photosynthetic capacity (section 4.2) and leaf temperature (sections 2, 4.2).
 
@@
 `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalinpress].
-Thus, upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat exchange (see sections 2, 4.1).
+Thus, upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat exchange (sections 2, 4.1).
 
@@
 Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that upper canopy leaves have higher $g_b$ and associated potential for water and heat loss (see sections 2, 4.1).
+The implication is that upper canopy leaves have higher $g_b$ and associated potential for water and heat loss (sections 2, 4.1).
 
@@
 However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
-In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
+In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
@@
 Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, also decreasing radiation load.
-Further, sun leaves tend to have higher stomatal and vein densities, which fascilitate evaporative cooling by enabling higher $g_s$ [see section 4.1, @zwienieckiHydraulicLimitationsImposed2004]. 
+Further, sun leaves tend to have higher stomatal and vein densities, which fascilitate evaporative cooling by enabling higher $g_s$ [section 4.1, @zwienieckiHydraulicLimitationsImposed2004]. 
 Yet sun leaves also have adaptations to limit water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [which may reduce or increase $g_b$; @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
@@
 Leaf optical properties are influenced by anatomical, morphological, and biochemical traits, that vary throughout the canopy, as outlined in sections 3.1 and 3.3.
-High in the canopy, high light absorptance can lead to photoinhibition and would be in part be mitigated by $T_{leaf}$ regulation (see sections 3.4 and 3.5; Table `r table_leaf_traits`). 
+High in the canopy, high light absorptance can lead to photoinhibition and would be in part be mitigated by $T_{leaf}$ regulation (sections 3.4 and 3.5; Table `r table_leaf_traits`). 
 For leaves in higher light environments, light absorptance and thus heat load can be reduced by leaf surface modifications that increase reflectance, such as trichome and cuticle, and heat may be dissipated biochemically  by carotenoids, including xanthophylls [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
@@
 
-Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see section 4, Table `r table_leaf_metabolism`).
+Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (section 4, Table `r table_leaf_metabolism`).
 Sun leaves have higher concentrations area-based concentrations of elements like nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical for respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthesis [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
 Increases in chlorophyll a/b ratios with height reflect greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile increase PAR absorptance efficiency of shade leaves (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (see section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, reduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table `r table_leaf_traits`).
+Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, reduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table `r table_leaf_traits`).
 
@@
 Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
 
@@
 This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
-Temperature-sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008].
+Temperature-sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [section 1, @sharkeyIsopreneEmissionPlants2008].
 Future work to understand how temporal dynamics of $T_{leaf}$ [see @leighInfluenceLeafSize2017] vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
@@
 `r fig_treerings`, @anderson-teixeiraJointEffectsClimate2022].
-However, under well-watered conditions conducive to high $g_s$, we do not necessarily expect higher thermal sensitivity of photosynthesis (see section 4.2) or woody growth in upper-canopy trees than their understory counterparts.
+However, under well-watered conditions conducive to high $g_s$, we do not necessarily expect higher thermal sensitivity of photosynthesis (section 4.2) or woody growth in upper-canopy trees than their understory counterparts.
 In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
@@
 <!--Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
-While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (see section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
+While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
 Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].
@@
 DGVMs operate at a range of scales and vary in complexity, from detailed individual-based models (i.e., forest gap models), which represent vegetation at the level of individual plants, which capturing spatial variability in light environment and microclimates at high 3D spatial resolution [@shugartGapModelsTheir2018], to big-leaf models that reduce 3D vegetation structure to single vegetation layer, implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically [@bonanDynamicGlobalVegetation2003; @krinnerDynamicGlobalVegetation2005].
-This simplification is computationally more efficient, but does not always capture observed vertical profiles (sections 1-4; for example, vertical shifts in the balance between stomatal conductance and photosynthetic capacity, see sections 4.1-4.2) and cannot capture important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
+This simplification is computationally more efficient, but does not always capture observed vertical profiles (sections 1-4; for example, vertical shifts in the balance between stomatal conductance and photosynthetic capacity, sections 4.1-4.2) and cannot capture important demographic processes and vertical light competition [@hurttTerrestrialModelsGlobal1998; @krinnerDynamicGlobalVegetation2005; @smithRepresentationVegetationDynamics2001].
 The computational middle-ground lies in cohort-based models, which represent vegetation as cohorts of individuals, grouped together based on properties including size, age, and functional type [@fisherVegetationDemographicsEarth2018].
@@
 Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models. 
-This is due in large part to the paucity of observational and experimental studies needed to resolve key patterns and underlying mechanisms (see sections 4, 5).
+This is due in large part to the paucity of observational and experimental studies needed to resolve key patterns and underlying mechanisms (sections 4, 5).
 Pairing of models with observational studies is key to further improving our mechanistic understanding of vertical gradients and their implications.

### 2022-04-01 vertical-thermal-review.Rmd
condensing intro and modeling sections
48 words over!

#113

@@
 #Word count
-w_intro=414
+w_intro=374
 w_biophys=1576
@@
 w_imp_gc=1088
-w_imp_modeling_scaling=1229
+w_imp_modeling_scaling=1186
 w_implications=w_imp_gc+w_imp_modeling_scaling
@@
 
-<!--This first paragraph is all good content, but mostly pretty basic background knowledge that could be cut if needed.-->
-Forest responses and feedbacks to climate change will critically influence Earth's furture climate.
 Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach +1.5$^\circ$C by 2040 [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves and hotter droughts [@meehlMoreIntenseMore2004; @ipccClimateChange20212021].
-These changes are expected to affect tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; <!--@corlettImpactsWarmingTropical2011-->; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
-Warming may increase or decreased tree growth and forest carbon sequestration on balance, with decreases being more common globally [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020; @anderson-teixeiraCarbonCyclingMature2021].
-The resulting feedbacks to climate and carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].
+These changes are expected to affect tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; <!--@corlettImpactsWarmingTropical2011;--> @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+The resulting feedbacks to climate and carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet future forest dynamics remain one of the largest sources of uncertainty in Earth system model climate change projections [@friedlingsteinClimateCarbonCycle2006; @krauseLargeUncertaintyCarbon2018; @aroraCarbonConcentrationCarbon2020].
 
-Much uncertainty arises because effects of warming on trees depend on size and crown position. 
-Forests are vertically and horizontally stratified, and canopies moderate climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
-The resulting vertical stratification of the biophysical environment such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations that influences leaf traits, thermoregulation and metabolism, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
-Although these vertical gradients strongly influence plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
-Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecosystem dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
+Future forest dynamics will depend on the impacts of climate change across trees varying in height and crown position. 
+Forests are vertically stratified, and canopies moderate climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
+Vertical gradients in biophysical variables such as temperature, light, wind, humidity and carbon dioxide (CO~2~) concentrations influence leaf temperatures, traits, and metabolism, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016a].
+Although forest vertical stratification strongly influences plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest dynamics, biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.
 
@@
 (3) the leaf traits that most strongly influence $T_{leaf}$ and thermal tolerance;
-(4) leaf gas exchange (including stomatal and boundary layer conductance, photosynthesis, and respiration) and its thermal sensitivity; and 
+(4) leaf gas exchange and its thermal sensitivity; and 
 (5) tree and ecosystem ecology.
@@
 
-Future forest dynamics remain one of the largest sources of uncertainty in Earth system model projections of global carbon cycling [@friedlingsteinClimateCarbonCycle2006; @aroraCarbonConcentrationCarbon2020].
 Dynamic global vegetation models (DGVMs), which comprise the land surface models in Earth system models, are used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks [@foleyIntegratedBiosphereModel1996; @sitchEvaluationEcosystemDynamics2003; @woodwardVegetationDynamicsSimulating2004].
@@
 Most models have been developed to capture dynamics in the upper canopy, given the disproportionate role of large trees in ecosystem dynamics (Fig. `r fig_ecosystem`). 
-Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models. 
-This is due in large part to the paucity of observational and experimental studies needed to resolve key patterns and underlying mechanisms (sections 4, 5).
+Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models, in large part to the paucity of observational and experimental studies needed to resolve key patterns and underlying mechanisms (sections 4, 5).
 Pairing of models with observational studies is key to further improving our mechanistic understanding of vertical gradients and their implications.
@@
 Multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @bonanMovingIncorrectUseful2021].
-In addition, capturing vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects (Figs.
-`r fig_treerings` - `r fig_ecosystem`), requires improved characterization of the functional response of leaf-level processes (Table `r table_leaf_metabolism`, Fig.
-`r fig_leaf_T`) to vertically varying abiotic conditions (Fig.
-`r fig_NEON_vertical`), and the role of traits (Table `r table_leaf_traits`) in mediating responses to thermal sensitivity.
+In addition, capturing vertical gradients in ET, GPP, respiration, and woody growth, and subsequently the net ecosystem effects, requires improved characterization of the functional response of leaf-level processes to vertically varying abiotic conditions, and the role of traits in mediating responses to thermal sensitivity.
 

### 2022-04-01 vertical-thermal-review.Rmd
cleanup/ condensing : NOW <10K words!
removed some excessive citations to tables/ figures, "e.g."s in front of references...

#113

@NidhiVinod , we're now at 9993 words! It would still be good to do some more cleanup/ condensing, proofread, but (1) it's short enough, and (2) we can submit on Monday!

@@
 w_intro=374
-w_biophys=1576
-w_tleaf=752
-w_traits= 2471
+w_biophys=1548
+w_tleaf=736
+w_traits= 2460
 w_met=1000
@@
 Canopy foliage is the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions from large fluctuations in conditions experienced above the canopy.
-It strongly influences -- and is influenced by -- the vertical biophysical gradient (Fig.
-`r fig_NEON_vertical`).
+It strongly influences -- and is influenced by -- the vertical biophysical gradient.
 Vertical patterns in leaf area density (*i.e.*, leaf area per unit volume) differ across forests (Fig. `r fig_NEON_vertical`a-b, Supporting Information Figure `r SIfig_NEON_all`). 
-In tropical and temperate forests with dense broadleaf canopies, leaf area density is generally highest in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often high as well, sometimes causing undulating patterns with height [e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a,<!--@terborghVerticalComponentPlant1985;--> @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @parkerVerticalProfileCanopy1989; @dominguesParameterizationCanopyStructure2005].
-In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be greatest in the lower canopy or understory [e.g., PUUM and WREF, Fig.
-`r fig_NEON_vertical`a, @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
+In tropical and temperate forests with dense broadleaf canopies, leaf area density is generally highest in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often high as well, sometimes causing undulating patterns with height [<!--e.g., SCBI, SERC and HARV, Fig.
+`r fig_NEON_vertical`a,--><!--@terborghVerticalComponentPlant1985;--> @parkerVerticalProfileCanopy1989; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993;  @dominguesParameterizationCanopyStructure2005].
+In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be greatest in the lower canopy or understory [<!--e.g., PUUM and WREF, Fig.
+`r fig_NEON_vertical`a,--> @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
 Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
@@
 Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters light, thereby altering the spectrum of PAR creceived in the lower canopy and understory layers. 
-The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along qith total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation (700-1000 nm)  [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
+The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along qith total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation [700-1000 nm; @decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
 Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
-This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`d), than in forests with more open upper canopies, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig.
+This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests, than in forests with more open upper canopies, including many conifer forests [Fig.
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
-This light gradient drives variation in leaf traits (Table 1), photosynthetic capacity (section 4.2) and leaf temperature (sections 2, 4.2).
+This light gradient drives variation in leaf temperature (section 2), traits (section 3), and photosynthesis (section 4.2).
 
@@
 `r fig_leaf_T`a).
-However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
+However, higher wind speeds reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
 In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
-In contrast, in the lower canopy, greater RH and lower wind speeds may limit $\lambda E$ (through reduced VPD and $g_b$), increasing $T_{leaf}$ [Fig.
-`r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks (>10$^\circ$C)  [@waySunflecksTreesForests2012].
+In contrast, in the lower canopy, greater RH and lower wind speeds may limit $\lambda E$ (through reduced VPD and $g_b$), increasing $T_{leaf}$ [@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks [>10$^\circ$C; @waySunflecksTreesForests2012].
 
@@
 Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal $T_{leaf}$ under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
-In addition to lower wind speeds (Fig.
-`r fig_NEON_vertical`), higher RH (lower VPD) in the understory and inner canopy (Fig.
-`r fig_NEON_vertical`) also inhibits evaporative cooling, thus increasing $T_{leaf}-T_{air}$ [Fig.
+In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also inhibits evaporative cooling, thus increasing $T_{leaf}-T_{air}$ [Fig.
 `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].
 
-Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure. 
+Vertical $T_{leaf}$ gradients also vary with canopy structure (Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`). 
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than lower-canopy leaves, in some cases exceeding photosynthetic temperature optima ($T_{opt}$) because of $g_s$ limitation and high radiation [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999-->; @doughtyAreTropicalForests2008; <!--@fauset_differences_2018-->; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; <!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
-In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory, which can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [Fig.
-`r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`; <!--@hadleyInfluenceKrummholzMat1987--> @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
-The latter can result from a combination of still air at lower heights (Fig.
-`r fig_NEON_vertical`e), and sunflecks or the 'canopy greenhouse effect', as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing $T_{leaf}$ of foliage near the ground [Fig.
-`r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than lower-canopy leaves, in some cases exceeding photosynthetic temperature optima ($T_{opt}$) because of $g_s$ limitation and high radiation [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999;-->@doughtyAreTropicalForests2008; <!--@fauset_differences_2018;--> @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021 <!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
+In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory, which can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [<!--@hadleyInfluenceKrummholzMat1987--> @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
+The latter can result from a combination of still air at lower heights, and sunflecks or the 'canopy greenhouse effect' [section 1.5, @schymanskiStomatalControlLeaf2013; @hardwickRelationshipLeafArea2015], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing $T_{leaf}$ of foliage near the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
 One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense subalpine stand in the alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
@@
 Anatomical, structural and biochemical leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
-Vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+Vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
 Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
-More recent work has extended the focus from light alone to leaf height in the canopy, as important in determining differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
+More recent work has extended the focus from light alone to leaf height, as the latter shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [Table `r table_leaf_traits`, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
 Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
@@
 
-Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @binLeafTraitExpression2022], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010], and (3) among understory and canopy species [e.g., @lloydOptimisationPhotosyntheticCarbon2010].
+Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [@zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @binLeafTraitExpression2022], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010], and (3) among understory and canopy species [@lloydOptimisationPhotosyntheticCarbon2010].
 Most studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
-Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010]; however, vertical trends vary across trees [@lloydOptimisationPhotosyntheticCarbon2010], and counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
-Many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010].
+Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010]; however, vertical trends vary across trees [@lloydOptimisationPhotosyntheticCarbon2010], and counter-gradient variation can exist [@luskWhyAreEvergreen2008].
+Many leaf physiological parameters can be estimated based on foliage height profiles [@cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010].
 

### 2022-04-02 vertical-thermal-review.Rmd
proofread intro

@@
 #Word count
-w_intro=374
+w_intro=369
 w_biophys=1548
@@
 Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach +1.5$^\circ$C by 2040 [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves and hotter droughts [@meehlMoreIntenseMore2004; @ipccClimateChange20212021].
-These changes are expected to affect tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; <!--@corlettImpactsWarmingTropical2011;--> @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
+These changes are affecting tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; <!--@corlettImpactsWarmingTropical2011;--> @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
 The resulting feedbacks to climate and carbon storage, and changes in albedo and hydrology will in turn impact the future trajectory of climate change [@bonan_ecological_2016], yet future forest dynamics remain one of the largest sources of uncertainty in Earth system model climate change projections [@friedlingsteinClimateCarbonCycle2006; @krauseLargeUncertaintyCarbon2018; @aroraCarbonConcentrationCarbon2020].
 
-Future forest dynamics will depend on the impacts of climate change across trees varying in height and crown position. 
+Future forest dynamics will depend on how climate change impacts trees of varying height and crown position. 
 Forests are vertically stratified, and canopies moderate climatic conditions [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], including buffering understory microclimates from extreme meteorological conditions [@zellweger_seasonal_2019].
@@
 
-Here, we review vertical gradients in the biophysical environment and plant form and function in forests, focusing on five key themes (Fig. `r fig_schematic`):
+Here, we review vertical gradients in the biophysical environment and plant form and function in forests, focusing on five themes (Fig. `r fig_schematic`):
 (1) the biophysical environment;
@@
 ![**Figure `r fig_schematic`.
-Schematic summarizing vertical gradients in forests according to (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology.** Abbreviations are as follows: $T_{air}$: air temperature, VPD: vapor pressure deficit;  $T_{leaf}$ - $T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area; T sensitivity: temperature sensitivity; ET: evapotranspiration.
-Here, we summarize conditions in closed-canopy forests during growing season and daytime conditions; patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime.
-Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.
+Schematic summarizing typical vertical gradients in (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology within closed-canopy forests during daytime growing season conditions.** Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.
+Patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime.
+Abbreviations are as follows: $T_{air}$: air temperature, VPD: vapor pressure deficit;  $T_{leaf} - T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area; T sensitivity: temperature sensitivity; ET: evapotranspiration.
 *Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`).](schematics/Schematic figure 1.png)

### 2022-04-02 vertical-thermal-review.Rmd
misc cleanup

@@
 
-The biophysical environment, defined here to include the physical structure of the vegetation and associated physical conditions, varies vertically from the forest floor to the top of the canopy (Figs.
-`r fig_schematic`, `r fig_NEON_vertical`), with physical conditions largely determined by forest structure.
-In this section, we review of the existing literature and analyze data on vegetation structure and vertical microclimate profiles from focal sites within the U.S.
+The biophysical environment, defined here to include the physical structure of the vegetation and associated micrometeorological conditions, varies vertically from the forest floor to the top of the canopy (Figs.
+`r fig_schematic`, `r fig_NEON_vertical`), with micrometeorological conditions largely determined by forest structure.
+In this section, we review of the existing literature and analyze data on vegetation structure and vertical microclimate profiles from forested sites within the U.S.
 National Ecological Observatory Network (NEON; Fig.
@@
 In tropical and temperate forests with dense broadleaf canopies, leaf area density is generally highest in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often high as well, sometimes causing undulating patterns with height [<!--e.g., SCBI, SERC and HARV, Fig.
-`r fig_NEON_vertical`a,--><!--@terborghVerticalComponentPlant1985;--> @parkerVerticalProfileCanopy1989; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993;  @dominguesParameterizationCanopyStructure2005].
+`r fig_NEON_vertical`a,--><!--@terborghVerticalComponentPlant1985;-->@parkerVerticalProfileCanopy1989; @ashtonComparisonsStructureMixed1992; @koikeCanopyStructureTropical1993; @dominguesParameterizationCanopyStructure2005].
 In forests with more open upper canopies, including many needle-leaf forests, leaf area density can be greatest in the lower canopy or understory [<!--e.g., PUUM and WREF, Fig.
-`r fig_NEON_vertical`a,--> @baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
-Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [e.g., @bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
-In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [<!--@parkerStructuralPhenologyLeaf2004;--> @parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].
-In this review, we focus on growing season conditions (peak leaf area index, LAI) unless otherwise noted.
+`r fig_NEON_vertical`a,-->@baldocchiSeasonalVariationEnergy1997; @hanberryOpenForestEcosystems2020; @lawLeafAreaDistribution2001;@bonan_ecological_2016].
+Soil moisture conditions, topography, and gap formations following disturbances all alter foliage patterns [@bonan_ecological_2016; @almeidaContrastingFireDamage2016; @starkAmazonForestCarbon2012; @hanberryOpenForestEcosystems2020].
+In addition, seasonally dry and wet conditions, deciduousness, and phenology contribute to temporally shifting leaf area density patterns [<!--@parkerStructuralPhenologyLeaf2004;-->@parkerConsequencesEnvironmentalHeterogeneity2019; @smithSeasonalDroughtrelatedChanges2019; @tangLightdrivenGrowthAmazon2017; @nunesForestFragmentationImpacts2022].
+In this review, we focus on growing season conditions unless otherwise noted.
 
@@
 The intensity of visible and photosynthetically active radiation (PAR, 400-700nm) decreases from the canopy top to the forest floor, with a profile whose shape is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
-`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @belandVerticalStructureHeterogeneity2021; @koike_leaf_2001; @binLeafTraitExpression2022].
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @belandVerticalStructureHeterogeneity2021;  @binLeafTraitExpression2022].
 Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters light, thereby altering the spectrum of PAR creceived in the lower canopy and understory layers. 
-The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along qith total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation [700-1000 nm; @decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
+The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along qith total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation [700-1000 nm, @decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
 Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
@@
 ### 1.3 Turbulent transport and wind
-Vertical transport in forest canopies is dominated by turbulent transport, making vertical profiles of wind speed, water vapor, CO~2~, and temperature more difficult to predict and to generalize on the basis of theoretical first principles compared to radiation profiles.
+Vertical transport in forest canopies is dominated by turbulent transport, making vertical profiles of wind speed, water vapor, CO~2~, and temperature more difficult to predict and to generalize on the basis of theoretical first principles than radiation profiles.
 Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients [eddy diffusivities, @penmanWeatherWheatEssay1960]-- such that wind speed declines logarithmically with proximity to the canopy surface [@monteithPrinciplesEnvironmentalPhysics2013].
-Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-Within canopies, however, transport often fails to mimic diffusion – e.g., transport of momentum [@shawSecondaryWindSpeed1977] or heat [@raupachLagrangianAnalysisScalar1987] may occur in the direction opposite to that predicted by gradients of wind speed or temperature, respectively – making vertical wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [@raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
+Within canopies, however, transport often fails to mimic diffusion – e.g., transport of momentum [@shawSecondaryWindSpeed1977] or heat [@raupachLagrangianAnalysisScalar1987] may occur in the direction opposite to that predicted by gradients of wind speed or temperature, respectively – making vertical wind profiles difficult to generalize from first principles [@denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
 Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991].
@@
 This holds true across the range of forest types [Fig.
-`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalinpress].
-Thus, upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat exchange (sections 2, 4.1).
+`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; <!--@mcgregorTreeHeightLeaf2021;-->@juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalinpress].
+As a result, upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat exchange (sections 2, 4.1).
 
@@
 Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig.
-`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021; @binLeafTraitExpression2022].
+`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021; @binLeafTraitExpression2022].
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020].
 In combination with $T_{air}$ and $T_{leaf}$, RH determines leaf-to-air vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
-Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
+Especially when canopy height and leaf area index (LAI) are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
 The implication is that upper canopy leaves have higher $g_b$ and associated potential for water and heat loss (sections 2, 4.1).
@@
 Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
-Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
+Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
 Overall, leaves may optimize their biochemistry, anatomy and structure for their local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
@@
 Rising temperatures and severe droughts place canopy trees at particularly elevated risk of mortality. 
-Large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging [e.g., @millerReducedImpactLogging2011].
+Large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting,  @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging [@millerReducedImpactLogging2011].
 Increasing severity and frequency of heat waves, accompanied with increases in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].

### 2022-04-02 vertical-thermal-review.Rmd
revise paragraph on CO2 gradients
#114

@@
 ### 1.4 Atmospheric concentrations of CO~2~ and water vapor
-Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999].
-Vertical CO~2~ concentration gradients can drive gradients in the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and in rates of refixation of respired CO~2~ vertically, both of which tend to be higher at lower heights (Table 1).
-Particularly, understory seedlings benefit from elevated CO~2~ by having higher rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]. 
-However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations concentrations likely have little effect on energy balance and metabolism across the vertical gradient.
+Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day such that differences are negligible for most of the day when photosynthesis is occurring [@koike_leaf_2001; @yang_spatial_1999; @brooksVerticalGradientsPhotosynthetic1997].
+While understory seedlings may benefit modestly from higher CO~2~ concentrations near the ground during some parts of the day [e.g., 6% of C fixation from recently respired CO~2~; @brooksVerticalGradientsPhotosynthetic1997], gradients in CO~2~ concentrations concentrations likely have little effect on energy balance and metabolism of trees across the vertical gradient [@brooksVerticalGradientsPhotosynthetic1997].
 

### 2022-04-02 vertical-thermal-review.Rmd
biophys section proofread

@@
 w_intro=369
-w_biophys=1548
+w_biophys=1515
 w_tleaf=736
@@
 Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig.
-`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021; @binLeafTraitExpression2022].
+`r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, @juckerCanopyStructureTopography2018; <!--@mcgregorTreeHeightLeaf2021;-->@binLeafTraitExpression2022].
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020].
@@
 ### 1.5 Air temperature 
-Vertical gradients in air temperature ($T_{air}$) differ widely (Fig.
-`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`) and depend on factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019].
+Vertical gradients in air temperature ($T_{air}$) vary widely (Fig.
+`r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), depending on factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019].
 Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].

### 2022-04-03 vertical-thermal-review.Rmd
proofread Tleaf

@@
 w_biophys=1515
-w_tleaf=736
+w_tleaf=749
 w_traits= 2460
@@
 
-$T_{leaf}$ is strongly tied, and usually within a few degrees of, $T_{air}$ [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [@rey-sanchez_spatial_2016;@doughtyAreTropicalForests2008;@vogelLeavesLowestHighest2009].
-Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (Fig.
+$T_{leaf}$ is strongly tied to, and usually within a few degrees of, $T_{air}$ [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Note `r Notes_leafTbiophysics`,@doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009].
+Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance ($g_s$; Fig.
 `r fig_leaf_T`; Note `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig.
 `r fig_schematic`).
-High radiation in the upper canopy (Fig.
-`r fig_NEON_vertical`) implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig.
-`r fig_leaf_T`a).
+High radiation in the upper canopy implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`a).
 However, higher wind speeds reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
-In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (sections 3, 4.1) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
+In addition, adaptive leaf traits that increase $g_s$ and $g_b$, such as small size and high stomatal density (Note `r Notes_leafTbiophysics`; sections 3, 4.1), mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
-In contrast, in the lower canopy, greater RH and lower wind speeds may limit $\lambda E$ (through reduced VPD and $g_b$), increasing $T_{leaf}$ [@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Leaves in these normally-shaded locations experience rapid light and temperature surges during sunflecks [>10$^\circ$C; @waySunflecksTreesForests2012].
+In contrast, in the lower canopy, greater RH (reduced VPD) and lower wind speeds (reduced $g_b$) may limit leaf cooling via latent heat flux, increasing $T_{leaf}$ [@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Leaves in these normally-shaded locations can experience rapid light and temperature surges during sunflecks [>10$^\circ$C, @waySunflecksTreesForests2012].
 
@@
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than lower-canopy leaves, in some cases exceeding photosynthetic temperature optima ($T_{opt}$) because of $g_s$ limitation and high radiation [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999;-->@doughtyAreTropicalForests2008; <!--@fauset_differences_2018;--> @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021 <!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
+Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than lower-canopy leaves, in some cases exceeding photosynthetic temperature optima ($T_{opt}$) because of $g_s$ limitation and high radiation [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999;-->@doughtyAreTropicalForests2008; <!--@fauset_differences_2018;--> @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021<!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
 In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory, which can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [<!--@hadleyInfluenceKrummholzMat1987--> @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
-The latter can result from a combination of still air at lower heights, and sunflecks or the 'canopy greenhouse effect' [section 1.5, @schymanskiStomatalControlLeaf2013; @hardwickRelationshipLeafArea2015], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing $T_{leaf}$ of foliage near the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
-One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense subalpine stand in the alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
+The latter can result from a combination of still air at lower heights, sunflecks or the 'canopy greenhouse effect' (section 1.5), and the tendency for shade leaves to have lower $g_s$ [section 3, @schymanskiStomatalControlLeaf2013; @hardwickRelationshipLeafArea2015].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing $T_{leaf}$ near the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
+One of the few studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense subalpine stand in the alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
 However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVerticalinpress].

### 2022-04-04 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001; @belandVerticalStructureHeterogeneity2021;  @binLeafTraitExpression2022].
-Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters light, thereby altering the spectrum of PAR creceived in the lower canopy and understory layers. 
-The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along qith total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation [700-1000 nm, @decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
+Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters light, thereby altering the spectrum of PAR received in the lower canopy and understory layers. 
+The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along with total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation [700-1000 nm, @decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
 Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
 This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests, than in forests with more open upper canopies, including many conifer forests [Fig.
-`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; <!--@aussenacInteractionsForestStands2000a-->; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 This light gradient drives variation in leaf temperature (section 2), traits (section 3), and photosynthesis (section 4.2).
@@
 Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day such that differences are negligible for most of the day when photosynthesis is occurring [@koike_leaf_2001; @yang_spatial_1999; @brooksVerticalGradientsPhotosynthetic1997].
-While understory seedlings may benefit modestly from higher CO~2~ concentrations near the ground during some parts of the day [e.g., 6% of C fixation from recently respired CO~2~; @brooksVerticalGradientsPhotosynthetic1997], gradients in CO~2~ concentrations concentrations likely have little effect on energy balance and metabolism of trees across the vertical gradient [@brooksVerticalGradientsPhotosynthetic1997].
+While understory seedlings may benefit modestly from higher CO~2~ concentrations near the ground during some parts of the day [e.g., 6% of C fixation from recently respired CO~2~, @brooksVerticalGradientsPhotosynthetic1997], gradients in CO~2~ concentrations concentrations likely have little effect on energy balance and metabolism of trees across the vertical gradient [@brooksVerticalGradientsPhotosynthetic1997].
 

### 2022-04-04 vertical-thermal-review.Rmd
I thought this was a mistake, because the words are repeating

@@
 
-Anatomical, structural and biochemical leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
+Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
 Vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].

### 2022-04-04 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 Further, sun leaves tend to have higher stomatal and vein densities, which fascilitate evaporative cooling by enabling higher $g_s$ [section 4.1, @zwienieckiHydraulicLimitationsImposed2004]. 
-Yet sun leaves also have adaptations to limit water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [which may reduce or increase $g_b$; @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
+Yet sun leaves also have adaptations to limit water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [which may reduce or increase $g_b$, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
 In contrast, shade leaves have traits that maximize light capture (e.g., lower LMA and greater light absorptance per unit biomass), but larger leaves and lower transpiration make them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013].
@@
 For instance, in a central Amazonian forest, the upper canopies of both interior and edge forests shed leaves when maximum daily $T_{air}$ exceeded ~35$^\circ$C [@nunesForestFragmentationImpacts2022].
-Beacsue leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
+Because leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
 Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; <!--@mariasImpactsLeafAge2017;--> @zhouLeafageEffectsTemperature2015].
@@
 Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
-Beyond the optimum, photosynthesis decreases due to stomatal closure [ @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020; <!--@fredeenTemperatureHumidityEffects1999-->] and biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; <!--@sageTemperatureResponseC32007;--> @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
+Beyond the optimum, photosynthesis decreases due to stomatal closure [ @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020 <!--@fredeenTemperatureHumidityEffects1999-->] and biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; <!--@sageTemperatureResponseC32007;--> @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
 
@@
 Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs.
-shade leaves [<!--@carterExperimentalWarmingTropical2021-->; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007 <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006-->], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
+shade leaves [<!--@carterExperimentalWarmingTropical2021--> @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007 <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006-->], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
 This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].

### 2022-04-04 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 
-**Vertical CO~2~ concentration gradients can drive gradients in the ratio of leaf intercellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$) calculated from $\delta^{13}C$, and in rates of refixation of respired CO~2~ vertically, both of which tend to be higher at lower heights (Table 1).Particularly, understory seedlings benefit from elevated CO~2~ by having higher rates of CO~2~ refixation [@brooksVerticalGradientsPhotosynthetic1997]."**
+**Thanks for the suggestion.
+We have included the following sentence with the refrence: While understory seedlings may benefit modestly from higher CO~2~ concentrations near the ground during some parts of the day [e.g., 6% of C fixation from recently respired CO~2~, @brooksVerticalGradientsPhotosynthetic1997]."**
 

### 2022-04-04 vertical-thermal-review.Rmd
proofread traits section

@@
 w_tleaf=749
-w_traits= 2460
+w_traits= 2391
 w_met=1000
@@
 Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
-Vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+Vertical variation in leaf structure and composition has long been recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
 Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
-More recent work has extended the focus from light alone to leaf height, as the latter shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
+More recent work has extended the focus from light alone to height in the vertical profile, as the latter shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
@@
 Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemistry, anatomy and structure for their local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
-Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
+Overall, leaf biochemistry, anatomy and structure may be optimized to local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
+However, we lack a a cohesive framework for integrating the many differences in leaf traits throughout the canopy.
 Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
@@
 ## 3.2. Leaf optical properties
-Leaf optical properties are influenced by anatomical, morphological, and biochemical traits, that vary throughout the canopy, as outlined in sections 3.1 and 3.3.
-High in the canopy, high light absorptance can lead to photoinhibition and would be in part be mitigated by $T_{leaf}$ regulation (sections 3.4 and 3.5; Table `r table_leaf_traits`). 
-For leaves in higher light environments, light absorptance and thus heat load can be reduced by leaf surface modifications that increase reflectance, such as trichome and cuticle, and heat may be dissipated biochemically  by carotenoids, including xanthophylls [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
-Further, sun leaves tend to be thicker, with more palisade layers, which act as “light pipes”, and channel abundant light into deeper cell layers, enhancing photosynthetic capacity  [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
-In contrast, shade leaves have lower LMA and absorb more light per unit mass investment, as well as denser chloroplasts layers and a greater proportion of spongy mesophyll, more effective for capturing diffuse light (Table `r table_leaf_traits`) [@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000; @leeCorrelatesLeafOptical1990].
+Leaf optical properties are influenced by anatomical, morphological, and biochemical traits that vary throughout the canopy (sections 3.1, 3.3).
+High in the canopy, high light absorptance can lead to photoinhibition and would be in part be mitigated by $T_{leaf}$ regulation (sections 3.4, 3.5; Table `r table_leaf_traits`). 
+For leaves in higher light environments, light absorptance and thus heat load can be reduced by leaf surface modifications (e.g., trichomes, cuticle) that increase reflectance, and heat may be dissipated biochemically  by carotenoids, including xanthophylls [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
+Further, sun leaves tend to be thicker, with more palisade layers, which act as “light pipes” that channel abundant light into deeper cell layers, enhancing photosynthetic capacity [@poorterLeafOpticalProperties1995; @leeCorrelatesLeafOptical1990].
+In contrast, shade leaves have lower LMA and absorb more light per unit mass investment, as well as denser chloroplasts layers and a greater proportion of spongy mesophyll, more effective for capturing diffuse light [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990;@poorterLeafOpticalProperties1995; @poorterLeafOpticalProperties2000].
 
@@
 
-Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (section 4, Table `r table_leaf_metabolism`).
-Sun leaves have higher concentrations area-based concentrations of elements like nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical for respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthesis [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
-Increases in chlorophyll a/b ratios with height reflect greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the forest vertical profile increase PAR absorptance efficiency of shade leaves (Table `r table_leaf_traits`).
-Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies (section 4, Table `r table_leaf_metabolism`) compared to shaded lower canopies, reduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table `r table_leaf_traits`).
+Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (section 4).
+Sun leaves have higher area-based concentrations of elements like nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical for respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthesis [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
+Increases in chlorophyll a/b ratios with height reflect greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights increase PAR absorptance efficiency of shade leaves (Table `r table_leaf_traits`).
+Higher photosynthetic rates (area-based) and more frequent stomatal closure in sun exposed canopies (section 4), reduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table `r table_leaf_traits`).
 
@@
 
-Biochemical protection against light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
-Further, by reducing photosynthesis, VPD-associated stomatal closure can reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007] <!--Nidhi, I feel like this sentence needs a reference.
-{maybe this would be better as a topic sentence without a ref?
-but there is one Ty suggested-->.
+Biochemical protection against light and heat damage increases with peak radiation loads and thus tends to be higher in the upper canopy, which is subject to higher $T_{leaf}$ and hydraulic limitations, than in the understory (Table `r table_leaf_traits`).
+More frequent stomatal closure in upper canoopy leaves (section 4) reduces their the capacity to use light energy for photochemistry, thereby requiring a high capacity to dissipate excess light energy and protect against photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].
 Accumulation of excess light energy causes overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
@@
 This form of NPQ entails interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
-Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
-Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
+Leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
 In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
@@
 Temperature-sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [section 1, @sharkeyIsopreneEmissionPlants2008].
-Future work to understand how temporal dynamics of $T_{leaf}$ [see @leighInfluenceLeafSize2017] vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
+Future work to understand how temporal dynamics of $T_{leaf}$ vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
 Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play similar roles in upper-canopy leaves.
@@
 <!--short par giving background on thermal tolerance-->
+Very high $T_{leaf}$ (~40-60$^{\circ}$C) irreversibly damages photosystem II, leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
 Upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning via using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
-Very high $T_{leaf}$ (~40-60$^{\circ}$C) irreversibly damages photosystem II, leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
 Thermal tolerance is described in terms of the $T_{leaf}s$ at which photosystem II efficiency starts to decrease ($T_{crit}$) is reduced by 50%, $T_{50}$ [e.g., @slotLeafHeatTolerance2021]. 
@@
 <!--par on thermal tolerance across vertical gradients-->
-Thermal tolerance metrics are hypothesized to be greatest for the leaves with the greatest radiation loads and highest temperatures along the vertical gradient. 
+Thermal tolerance is hypothesized to be greatest for the leaves with the greatest radiation loads and highest temperatures along the vertical gradient. 
 Indeed, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama, $T_{50}$ was 0.7 - 1.4$^\circ$C lower for shade than sun leaves for two of three tree species [Table `r table_leaf_traits`, @slotPhotosyntheticHeatTolerance2019].
-Similarly, within a very open canopy in *Acacia papyrocarpa* Benth.
-canopy, $T_{50}$ was highest (*albiet* by <1$^\circ$C) in the lower, north-facing leaves, which faced the greatest composite climate stress (low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation) [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
+Similarly, within a very open canopy in *Acacia papyrocarpa* canopy, $T_{50}$ was highest (*albiet* by <1$^\circ$C) in the lower, north-facing leaves, which faced the greatest composite climate stress (low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation) [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
 These modest differences in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments imply that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
@@
 
-Leaf phenology likely influences the vertical structure of microenvironments, $T_{leaf}$, and forest photosynthetic capacity [@rey-sanchez_spatial_2016]. 
+Leaf phenology influences the vertical structure of microenvironments, $T_{leaf}$, and forest photosynthetic capacity [@rey-sanchez_spatial_2016]. 
 In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
-In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
+In temperate climates, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
 Early leaf-out gives saplings and seedlings a window for disproportionately high photosynthesis -- contributing the majority of annual carbon fixation for some seedlings -- before canopy foliage and reduces light availability [@augspurgerDifferencesLeafPhenology2003; @leeSpringPhenologicalEscape2021].
@@
 For instance, in a central Amazonian forest, the upper canopies of both interior and edge forests shed leaves when maximum daily $T_{air}$ exceeded ~35$^\circ$C [@nunesForestFragmentationImpacts2022].
-Because leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
+In evergreen forests, leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
 Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; <!--@mariasImpactsLeafAge2017;--> @zhouLeafageEffectsTemperature2015].

### 2022-04-04 vertical-thermal-review.Rmd
proofread/ cleanup gas exchange section

@@
 w_traits= 2391
-w_met=1000
+w_met=940
 w_ecol=1353
@@
 
-Leaf metabolism is strongly shaped by $T_{leaf}$, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
-However, as we detail below, few studies have evaluated temperature responses of leaf metabolic rates along a vertical forest gradient, or compared sun and shade leaves (Table `r table_leaf_metabolism`).
+Leaf metabolism is strongly shaped by environmental drivers, $T_{leaf}$, and traits, all of which vary across vertical forest gradients (sections 1-3), Fig. `r fig_schematic`). 
+However, as we detail below, few studies have evaluated how temperature responses of leaf metabolic rates are shaped by the varying environmental conditions across vertical forest gradients (Table `r table_leaf_metabolism`).
 
@@
 Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
-Maximum $g_s$ increases with light--typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979]--and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
-However, water supply cannot meet the demands caused by the high radiation and $g_b$ in sun leaves with fully open stomata, because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014].
-These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and $\delta^{13}C$ discussed in section 3.3 (Table 1).
+Maximum $g_s$ increases with light -- typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979] -- and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
+However, water supply cannot meet the demands caused by the high radiation and $g_b$ in sun leaves with fully open stomata, in part because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014].
+These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and $\delta^{13}C$ (section 3.3, Table `r table_leaf_traits`).
 Stomatal depression reduces transpirational cooling (Fig.
-`r fig_leaf_T`), thus amplifying the warming of sun leaves by high radiation [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
+`r fig_leaf_T`e), thus amplifying the warming of sun leaves by high radiation [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
 The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than lower-canopy shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
@@
 
-Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both predicted by optimization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
+Photosynthetic capacity is generally higher in exposed canopy positions -- a fact that is both predicted by optimization theory [@fieldAllocatingLeafNitrogen1983; @hiroseMaximizingDailyCanopy1987] and observed in numerous field studies (Table `r table_leaf_metabolism`)<!-- @niinemetsPhotosynthesisResourceDistribution2007; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020]-->.
 
 Temperature can affect photosynthesis directly, by altering photosynthetic enzyme activity and the electron transport chain,  and indirectly, by increasing VPD and closing stomata [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to $T_{air}$, with the peak commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
+Photosynthesis has a peaked response to $T_{air}$, with the optimum commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
 Beyond the optimum, photosynthesis decreases due to stomatal closure [ @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020 <!--@fredeenTemperatureHumidityEffects1999-->] and biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; <!--@sageTemperatureResponseC32007;--> @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
@@
 In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the top of the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
-@hernandezSimilarTemperatureDependence2020 speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 
+The observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis [@hernandezSimilarTemperatureDependence2020]. 
 Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of midday stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
@@
 
-Similar to photosynthesis, respiration tends to be higher in upper-canopy sun leaves [Table `r table_leaf_metabolism`, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$ or $E_0$, Table `r table_leaf_metabolism`) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; @weerasingheCanopyPositionAffects2014].
+Similar to photosynthesis, respiration tends to be higher in upper-canopy sun leaves, but its temperature sensitivity ($Q_{10}$ or $E_0$) shows no definite trend along the vertical gradient.
 Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs.
-shade leaves [<!--@carterExperimentalWarmingTropical2021--> @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007 <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006-->], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002].
+shade leaves<!--[<!--@carterExperimentalWarmingTropical2021-> @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007 <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006->]-->, greater in upper-canopy leaves<!--[@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]-->, or greater in the lower canopy <!--[@griffinCanopyPositionAffects2002]--> (Table `r table_leaf_metabolism`).
 This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
@@
 Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table `r table_leaf_metabolism`, e.g., @krinnerDynamicGlobalVegetation2005].
 Recent years have seen growing efforts to incorporate vertical variation more directly in models, specifically involving leaf water potential and light absorption, along with a increasing interest in confronting models directly with field measurements [@fisherPerspectivesFutureLand2020].
@@
 
-Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers [e.g., sunlit and shaded portions, @depurySimpleScalingPhotosynthesis1997] may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, Table 1, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, Table 2).
+Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers [e.g., sunlit and shaded portions, @depurySimpleScalingPhotosynthesis1997] may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, Table `r table_leaf_traits`, @smithSeasonalDroughtrelatedChanges2019] and functions (e.g., thermal responses, Table `r table_leaf_metabolism`).
 Multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @bonanMovingIncorrectUseful2021].
@@
 (c) In turn, these efforts can be used to directly inform ecosystem model development and parameterization.
-Vertically resolved in situ data could include any variable listed in Table 1 or 2.
+Vertically resolved in situ data could include any variable listed in Table `r table_leaf_traits` or 2.
 Panel b shows drone lidar point cloud data colored with thermal data from a savanna woodland in Kruger National Park, South Africa illustrating vertical and horizontal variation in vegetation temperature.

### 2022-04-04 vertical-thermal-review.Rmd
metabolism --> gas exchange
we use these pretty interchangeably, but in some cases gas exchange is more suitable.

@@
 w_tleaf=749
-w_traits= 2391
-w_met=940
+w_traits= 2392
+w_met=941
 w_ecol=1353
@@
 ![**Figure `r fig_schematic`.
-Schematic summarizing typical vertical gradients in (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf metabolism, and (5) tree and ecosystem ecology within closed-canopy forests during daytime growing season conditions.** Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.
+Schematic summarizing typical vertical gradients in (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf gas exchange, and (5) tree and ecosystem ecology within closed-canopy forests during daytime growing season conditions.** Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.
 Patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime.
@@
 
-Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
+Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, gas exchange, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
 Vertical variation in leaf structure and composition has long been recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
@@
 
-Leaf metabolism is strongly shaped by environmental drivers, $T_{leaf}$, and traits, all of which vary across vertical forest gradients (sections 1-3), Fig. `r fig_schematic`). 
+Leaf gas exchange is strongly shaped by environmental drivers, $T_{leaf}$, and traits, all of which vary across vertical forest gradients (sections 1-3), Fig. `r fig_schematic`). 
 However, as we detail below, few studies have evaluated how temperature responses of leaf metabolic rates are shaped by the varying environmental conditions across vertical forest gradients (Table `r table_leaf_metabolism`).

### 2022-04-04 vertical-thermal-review.Rmd
cleanup/ proofread ecology section

@@
 w_met=941
-w_ecol=1353
+w_ecol=1357
 w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
@@
 
-Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect tree ecology, ecosystem ecology, and their temperature responses (Fig.
+Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism scale up to affect ecological rates and their temperature responses at tree and ecosystem levels (Fig.
 `r fig_schematic`).
@@
 Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which strongly predicts tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] photosynthesis, and intra-canopy trait variation [@binLeafTraitExpression2022].
-Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where leaves receive more light and have higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
+Specifically, increases are linked to greater leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where leaves receive more light and have higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
 Photosynthate production is allocated among functions including respiration, aboveground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
@@
 Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2022].
-In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimate2022], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015].
-This points to a dominant role of vertical profiles in the biophysical environment, particularly light (Fig. `r fig_NEON_vertical`), in shaping tree growth rates within forests.
+In open forests and for open-growth individual trees, growth rate declines with DBH, whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015;@anderson-teixeiraJointEffectsClimate2022].
+This points to a dominant role of vertical profiles in the biophysical environment, particularly light, in shaping tree growth rates within forests.
 
@@
 In turn, the drought sensitivity of woody growth tends to be greater in upper-canopy trees than in smaller trees with less exposed crowns. 
-Dendrochronologists generally agree tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth is most sensitive to interannual climate variation [@frittsTreeRingsClimate1976].
+Dendrochronologists generally agree that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth is most sensitive to interannual climate variation [@frittsTreeRingsClimate1976].
 However, few studies have directly examined drought- or temperature-sensitivities as a function of tree size.
@@
 Drought also causes greater mortality in larger trees [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
-Mechanistically, this is likely driven in part by larger trees having crown in microenvironments that are more challenging during drought [Figs.
+Mechanistically, this is likely driven in part by larger trees having their crowns in microenvironments that are more challenging during drought [Figs.
 `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet height itself also creates disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].
-Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2022], although some evidence indicates that short trees in young stands can be more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
+Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2022].
+However, evidence that short trees in young stands can be more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004;@wangDiscrepancyGrowthResilience2022] reinforces the importance of exposure in shaping drought sensitivity.
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.
@@
 In terms of C cycling, it has been estimated that the canopy strata contributes $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007].
-Large trees also dominate woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
+Large trees also dominate woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
 `r fig_ecosystem`b, @piponiotDistributionBiomassDynamicsinpress].
@@
 It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but probable responses can be inferred based on the patterns and mechanisms reviewed above.
-We expect that transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures in the upper canopy than in the understory -- at least under conditions of moderate to high VPD [@grossiordPlantResponsesRising2020; @nunesForestFragmentationImpacts2022].
+We expect that under conditions of moderate to high VPD, transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures in the upper canopy than in the understory [@grossiordPlantResponsesRising2020; @nunesForestFragmentationImpacts2022].
 Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will strongly influence the whole-ecosystem response, potentially with modest buffering by the understory. 

### 2022-04-04 vertical-thermal-review.Rmd
finish proofreading/ cleanup !
@NidhiVinod , all yours now!

@@
 w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
-w_imp_gc=1088
-w_imp_modeling_scaling=1186
+w_imp_gc=1071
+w_imp_modeling_scaling=1173
 w_implications=w_imp_gc+w_imp_modeling_scaling
-w_conclusions=248
+w_conclusions=247
 w_total=w_intro+w_vertgradients+w_implications+w_conclusions
@@
 As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased maximum temperatures, with extreme increases during heat waves [@ipccClimateChange20212021].
-To the extent that warming is coupled to drought, we expect that the tallest trees to usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs.
-`r fig_schematic`- `r fig_treerings`].
+To the extent that warming is coupled to drought, we expect that tall trees with exposed crowns will usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment (section 5.1, Figs.
+`r fig_schematic`- `r fig_treerings`).
 In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022].
-An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
+It remains unclear how these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
 While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016].
-Given the mechanisms reviewed here, we might expect that VPD in particular skews the stress more towards the exposed canopy trees.
+Given the mechanisms reviewed here, we might expect that high VPD in particular skews the stress more towards the exposed canopy trees.
 
-It remains poorly understood how responses to warming will vary across vertical gradients under mesic conditions. 
+It remains uncertain how responses to warming will vary across vertical gradients under mesic conditions. 
 Leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], yet failure to fully acclimate will reduce carbon sequestration at leaf, tree, and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
@@
 <!--Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
-While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [section 5.1, Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
+While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
 Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].
@@
 
-Rising temperatures and severe droughts place canopy trees at particularly elevated risk of mortality. 
+Increasing severity and frequency of heat waves, accompanied with increases in VPD and ET, place canopy trees at particularly elevated risk of mortality (section 5.1), potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
 Large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting, @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging [@millerReducedImpactLogging2011].
-Increasing severity and frequency of heat waves, accompanied with increases in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
 Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
@@
 diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
 Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table `r table_leaf_metabolism`, e.g., @krinnerDynamicGlobalVegetation2005].
 Recent years have seen growing efforts to incorporate vertical variation more directly in models, specifically involving leaf water potential and light absorption, along with a increasing interest in confronting models directly with field measurements [@fisherPerspectivesFutureLand2020].
-Several recent model developments are at the cutting edge of representing vertical variation.
-Model developments by @chenColumnCanopyAirTurbulent2019, @bonanModelingCanopyinducedTurbulence2018, and @longoBiophysicsEcologyBiogeochemistry2019 implement vertical gradients of irradiance, water content, $T_{leaf},$ and humidification of canopy air by transpiration, modulated by turbulence within the forest and a roughness layer that extends to roughly twice the canopy height [@fisherPerspectivesFutureLand2020].
+Recent model developments at the cutting edge of representing vertical variation [@chenColumnCanopyAirTurbulent2019;@bonanModelingCanopyinducedTurbulence2018; @longoBiophysicsEcologyBiogeochemistry2019] implement vertical gradients of irradiance, water content, $T_{leaf}$, and humidification of canopy air by transpiration, modulated by turbulence within the forest and a roughness layer that extends to roughly twice the canopy height [@fisherPerspectivesFutureLand2020].
 
@@
 In a a western-U.S.
-savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the tops of tree crowns, together with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig.
+savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found lower foliage temperatures at the tops of tree crowns than in the understory, which was influenced by very high grass temperatures, consistent with the principles outlined in sections 1-2 (Fig.
 `r SIfig_NEON_all`).
@@
 In the historical climates to which trees adapted, the canopy was an advantageous place for photosynthesis and growth. 
-However, as temperatures increase, it is likely that exposed canopy positions will more often be physiologically stressful.
+However, as temperatures increase, it is likely that exposed canopy positions will become increasingly physiologically stressful.
 Ensuant increased mortality of canopy trees will create ever more gaps, changing understory conditions and community composition.
@@
 
-`r Notes_leafTbiophysics`
+Note `r Notes_leafTbiophysics`. Biophysical drivers of T~leaf~
 

### 2022-04-04 response-to-reviews.Rmd
updated r1's comment

@@
 
-**We revised and added references to this sentence in question, which now reads: **
+**We moved Biophysical Drivers of $T_{leaf}$ section (previously section 2) to supplementary material notes as we had to cut words to meet the required word limit.
+Although, we revised and added references to this sentence in question, which now reads: **
 **Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].**

### 2022-04-04 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 
-- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their capacity to incorporate vertical variation in leaf traits and physiological processes.In general, however, this variation is accounted for via light competition.
+- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their capacity to incorporate vertical variation in leaf traits and physiological processes.
+In general, however, this variation is accounted for via light competition.
 Models partition radiation above and within the forest vertical profile (i.e., direct vs.
-diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].**
+diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
+Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table `r table_leaf_metabolism`, e.g., @krinnerDynamicGlobalVegetation2005].**
 

### 2022-04-04 response-to-reviews.Rmd
small spelling edits

@@
 
-- **Further, by reducing photosynthesis, VPD-associated stomatal closure can reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].Accumulation of excess light energy causes overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by excess light [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].This form of NPQ entails interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed (Table 2), and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
-Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table 1, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (see section 2.2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
+- **More frequent stomatal closure in upper canopy leaves (section 4) reduces their capacity to use light energy for photochemistry, thereby requiring a high capacity to dissipate excess light energy and protect against photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].
+Accumulation of excess light energy causes overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
+A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by excess light [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
+This form of NPQ entails interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
+Leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
 
@@
 **We did not intend to argue that light was not the dominant factor, but the wording was misleading.
-This sentence has been revised to read, "This points to a dominant role of vertical profiles in the biophysical environment, particularly light (Fig.
-2) in shaping tree growth rates within forests."**
+This sentence has been revised to read, "This points to a dominant role of vertical profiles in the biophysical environment, particularly light, in shaping tree growth rates within forests."**
 
@@
 
-- **"Implications: Global Change Responses : Warming" section, final par (previous lines 803-806: "We expect that the tallest trees will be increasingly prone to hydraulic failure and damaging or lethal $T_{leaf}$'s, *while much larger uncertainty remains surrounding the resoponses of understory trees*."**
+- **"Implications: Global Change Responses : Warming" section, final par (previous lines 803-806: "To the extent that warming is coupled to drought, we expect that tall trees with exposed crowns will usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment.
+*In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability* [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022]."**
 

### 2022-04-04 vertical-thermal-review.Rmd
small spelling edits

@@
 Biochemical protection against light and heat damage increases with peak radiation loads and thus tends to be higher in the upper canopy, which is subject to higher $T_{leaf}$ and hydraulic limitations, than in the understory (Table `r table_leaf_traits`).
-More frequent stomatal closure in upper canoopy leaves (section 4) reduces their the capacity to use light energy for photochemistry, thereby requiring a high capacity to dissipate excess light energy and protect against photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].
+More frequent stomatal closure in upper canopy leaves (section 4) reduces their capacity to use light energy for photochemistry, thereby requiring a high capacity to dissipate excess light energy and protect against photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].
 Accumulation of excess light energy causes overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].

### 2022-04-04 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 - **Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
-Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$, due to turbulent air mixing and interception of thermal radiative flux from the soil and the canopy by lower-canopy layers.**
+Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$, due to turbulent air mixing and interception of thermal radiative flux from the soil and the canopy by lower-canopy layers.
+**
 
@@
 
-- **$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4).
-Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for photosynthesis and metabolism [@michaletzPlantThermoregulationEnergetics2015a <!--@perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020-->].
+- **$T_{leaf}$ is strongly tied to, and usually within a few degrees of, $T_{air}$ [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Note S1, @doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009].
+Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance ($g_s$; Fig.
+3; Note S1), all of which vary across forest vertical gradients (Fig.
+1).
 **
@@
 
-**Anatomical, structural and biochemical leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table 1, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, metabolism, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].Vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
-Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].More recent work has extended the focus from light alone to leaf height in the canopy, as important in determining differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
+**Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, gas exchange, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
+Vertical variation in leaf structure and composition has long been recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
+Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
+More recent work has extended the focus from light alone to height in the vertical profile, as the latter shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
 For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [Table 1, @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
-Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaves may optimize their biochemistry, anatomy and structure for their local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
-Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
+Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
+Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
+Overall, leaf biochemistry, anatomy and structure may be optimized to local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
+However, we lack a a cohesive framework for integrating the many differences in leaf traits throughout the canopy.
 Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].

### 2022-04-04 response-to-reviews.Rmd
updated the responses

@@
 
-**We do not understand this comment, but have verified that this sentence is correct.**
+**We agree that this sentence seems confusing. Therefore, we removed the sentence.**
 
@@
 
-- **Changed sentences: Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
+- **Changed sentences: Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
+**
 
@@
 **We were using the term “buffer” in a functional or biological sense, meaning “protect,” rather than as a description of the underlying physics; i.e., the canopy protects the understory from large diel swings in both air temperature and radiation.
-We have reworded this passage to clarify: “Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting from***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~ lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings ”**
+We have reworded this passage to clarify: “***Diel temperature range is typically smaller beneath dense canopies than above, resulting from***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~ $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings.”**
 
@@
 
-**We agree that the comparison is not very clear here.
-We added "compared to shaded lower canopies" to the sentence: "Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies ***compared to shaded lower canopies,***rreduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table 1).**
+**We agree that the comparison is not very clear here. We reworded the sentences to make it clearer:** 
+**Increases in chlorophyll a/b ratios with height reflect greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights increase PAR absorptance efficiency of shade leaves (Table 1).
+Higher photosynthetic rates (area-based) and more frequent stomatal closure in sun exposed canopies (section 4), reduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table 1).**
 
@@
 
-**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ cannot meet the demands caused by the high radiation and $g_b$ in sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al.
+**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ cannot meet the demands caused by the high radiation and $g_b$ in sun leaves ***with fully open stomata, in part because of height-related constraints on water transport (e.g., Yoder et al.
 1994, Koch et al.

### 2022-04-04 response-to-reviews.Rmd
citation fix

@@
 Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table `r table_leaf_metabolism`, e.g., @krinnerDynamicGlobalVegetation2005].**
+Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].**
 
@@
 This form of NPQ entails interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
-Leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
+Leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table 1, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
 In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
@@
 
-**Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, gas exchange, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
+**Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table 1, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, gas exchange, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
 Vertical variation in leaf structure and composition has long been recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].

### 2022-04-04 vertical-thermal-review.Rmd
citation

@@
 Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimate2022].
-However, evidence that short trees in young stands can be more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004;@wangDiscrepancyGrowthResilience2022] reinforces the importance of exposure in shaping drought sensitivity.
+However, evidence that short trees in young stands can be more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004;@wangDiscrepancyGrowthResilience2022a] reinforces the importance of exposure in shaping drought sensitivity.
 The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

### 2022-05-16 journal submissions/2022_04_02/response-to-reviews.Rmd
move previous response to review files to folder for that submission

--- 2022-04-04 response-to-reviews.Rmd
@@
 This rearrangement of the traits section better parallels Table 1 and the following section on leaf gas exchange.
+
+- We shortened the manuscript, with a current word count of 9814.
 
@@
 
-**We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines 169-192), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles.
+**We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines 157-184), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles.
 We have also modified the subsequent paragraph about wind speeds to eliminate the use of the confusing term “buffering”.**
@@
 
-- **“Canopy foliage is the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions experienced above the canopy.”; (line xxx)***(new text italicized)**
-- **“Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019]" (line 229)**
-- **“Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting from,***~~lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings.” (line 232)**
+- **“Canopy foliage is the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions experienced above the canopy.”; (line 120)***(new text italicized)**
+- **“Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019]" (line 210)**
+- **“Diel temperature range is typically smaller beneath dense canopies than above, resulting from,lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings.” (line 232)**
 
@@
 
-**We have revised the first sentence here to read: "Biochemical protection against light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.
+**We have revised the first sentence here to read: "Biochemical protection against light and heat damage increases with peak radiation loads and thus tends to be higher in the upper canopy, which is subject to higher $T_{leaf}$ and hydraulic limitations, than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.
 We also choose to keep this rather obvious sentence because we feel that basic topic sentences are important for helping to guide readers from a diversity of sub-disciplines through the manuscript.**

### 2022-05-16 response-to-reviews.Rmd
load new reviewer comments into response to review file
@NidhiVinod , FYI.

--- 2022-04-04 response-to-reviews.Rmd
@@
 
-- We have brought in an additional author, Thomas N. Buckley, to round out the expertise of the author team and help address some of the reviewer comments (particularly Reviewer 2).
-
-- We did some modest reorganization, including moving the content on $T_{crit}$ and VOCs to the traits section (the former suggested by Reviewer 2) and reorganizing the traits section around trait groupings, as opposed to intra-specific/ interspecific.
-This rearrangement of the traits section better parallels Table 1 and the following section on leaf gas exchange.
-
-- We revised three figures (Figs. 1-4) based on reviewer comments, including some standardization of style, but have not perfected the styling because the journal illustrators will fix up all figures.
+...
 
@@
 
-### General comment:
+Comments to the Author
+I am satisfied with the corrections and the responses to the comments. I like the new structure, the writting is very precise while covering broad areas and the paper very interesting.
 
-The article is easy to read and well structured, I appreciated reading it.
-It covers a broad area and explains relatively complex ideas in simple terms and efficiently.
-I also enjoyed the changes in scale (leaf scale to ecosystem scale).
-I think this paper will be very useful as a reference for the plant physiologist community and for modelers.
+**Thank you.**
 
-**Thank you for the review!**
+Some minor points :
 
-I have been really interested in the comparison between gradients in canopy of individual trees and canopy made of an assemblage of species (paragraph 3.1 and 3.2.
-However, the paragraph 3.2 is relatively short, and I am curious if it can be developed a little bit more (maybe not!).
+-PAR units in Fig 2d
 
-**We ended up removing former section 3.2 because we reorganized the traits section around trait groupings, as opposed to intra-specific/ interspecific.
-This rearrangement of the traits section better parallels Table 1 and the following section on leaf gas exchange.
-We integrated the content that was previously in this section throughout the traits section, as appropriate.**
+-Fig 3 caption: TLeaf and some other discreptances with rest of the paper abreviations or format (tair, gs)
 
-**Unfortunately, we've been unable to find many studies examining vertical gradients at the interspecific level, and have included those of which we are aware.**
+-l304 Not sure about the structure of the sentence and causality. Gravity and hydraulic pathlength (due to height) could mitigate the effect of light alone ?
 
-I find it interesting to try and link the empirical and theoretical knowledge on vertical gradients in forests with what is done in DGCMs and what should be done to improve their accuracy.
-However, I think there is a big step between the review (part I and II) and the author’s view on the implications for models (Scaling across space and time).
-I wonder if it would be possible to link a little bit more the different parts, and maybe without going that much in the technical details (cohort, big leaf, multi layers, ...) which I think are sometimes imprecise or maybe wrong.
-Would it be possible to explain more simply what are the vertical gradients presented in the section II, that are misrepresented today in DGCMs?
-Maybe by explaining which of the ‘biophysical variables’ and ‘leaf traits’ that you listed, are considered constant when they should be variable vertically?
+-Table 2, gb unit, mmol m-2 s-1 ?
 
-**Thank you for the comment. We included a section that explains how vertical gradients are represented in models:**
-
-- **Models partition radiation above and within the forest vertical profile (i.e., direct vs.
-diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005].
-**
-
-**Given that models vary greatly in how they capture vertical light gradients, what biophysical variables are included and how those variables are parameterized to vary (or not) across vertical gradients, we feel that it is beyond the scope of this paper to comprehensively list or explain which of those variables vary vertically across different models.**
-
-### Minor details that I hope will be of interest for the authors. Consider them as you see fit.
-
-L30 I did not understand ‘Scaling up’ in this sentence.
-
-**We specified "Scaling up" here as "Scaling up from leaves to trees,".**
-
-Figure 1 is nice.
-I struggled with ‘Tleaf sensitivity to shortwave radiation’ but I think it is clear after, when reading the text.
-Consider using the same words in the paragraph where you detail it (line 268?) and adding the reference to the figure in the text.
-Or maybe change the term (‘thermal capacitance’ ?) if this is the part you are referring to in the text.
-
-**Done**
-
-Figure 2: Homogenize the units (in parenthesis or in bracket) and add the unit for PAR.
-
-**Done.**
-
-I like your figures, but maybe consider homogenizing a little bit the styles for the plots between figures 2 3 and 4.
-
-**This will be done by the New Phytologist illustrators (because it is a Tansley review). We will work with them to ensure that all the figures look good.**
-
-L205-206 Maybe add some references.
-How many degrees?
-In your Figure 3 there is not really a difference between Tleaf and Tair when SWR = 0.
-Given that the transpiration nearly stops at night (high humidity, low conductance) I would expect Tleaf – Tair to be very close to zero.
-
-**We moved Biophysical Drivers of $T_{leaf}$ section (previously section 2) to supplementary material notes as we had to cut words to meet the required word limit.
-Although, we revised and added references to this sentence in question, which now reads: **
-**Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].**
-
-**Figure 3 shows the effect of each variable while the others are held constant (as specified in the caption). Thus, it does not represent the full suite of night-time conditions.**
-
-L219, repetition with L 216
-
-**Done**
-
-Figure 3: TLeaf and Tleaf.
-
-**Done**
-
-l258 under soil wet conditions?
-
-**We have reworded this sentence to read, "In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig.3 b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017]."**
-
-L298-299 Far red to red light: I think this could be developed here or somewhere else.
-You covered the vertical variation in light quantity inside the canopy, but you could also report the vertical change in light quality (light spectrum) as the leaves absorb preferentially some wavelengths but reflects or transmit others.
-
-**Thank you for the suggestion, we added the following here:"Canopy foliage absorbs a large portion of PAR (400-700nm), and selectively filters light, thereby altering the spectrum of PAR creceived in the lower canopy and understory layers.
-The ratio of red (~685-690 nm) to far red (~730-740 nm) light declines along qith total PAR with increasing depth in the canopy, and understories receive diffuse light enriched in near infrared radiation (700-1000 nm)  [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]."**
-
-L 322 consider homogenizing the photosynthetic capacity (AA) with the other notations in Table 2.
-
-**Done**
-
-Table 1: Unit for PAR absorptance (should be % I think)
-
-**Done**
-
-Paragraphs 3.1 and 3.2:
-I thought that the comparison between intra species and interspecies gradients is very interesting.
-I wondered if there are more studies to expand it.
-One key article that I know is (Lloyd et al., 2010) which showed that within tree gradients and between tree gradients are similar for a lot of leaf traits in tropical forests.
-I think this is one of the reasons why DGCMs are relatively accurate even if the canopy description is quite simple.
-(The vertical gradients in photosynthetic parameters are scaled on Na gradients, which are the same intra and between species in this study).
-
-**We ended up removing former section 3.2 because we reorganized the traits section around trait groupings, as opposed to intra-specific/ interspecific.
-This rearrangement of the traits section better parallels Table 1 and the following section on leaf gas exchange.
-We integrated the content that was previously in this section throughout the traits section, as appropriate.**
-
-**Unfortunately, we've been unable to find many studies examining vertical gradients at the interspecific level, and have included those of which we are aware.**
-
-**Thanks for pointing out the Lloyd reference, which we had previously missed. We have added citations to this (e.g., *2nd* paragraph of the traits section), and have also added it to our tables).**
-
-L429 I am not fully sure if I understood this sentence, consider maybe developing the idea.
-Do you mean that the variation is higher in canopy made of multi species than on the canopy of single species?
-If so, I didn’t understand the comparison with understory species.
-
-**We removed this sentence because we agreed that it was confusing, and the points addressed from comments above provide substantial information for inter- and intra-canopy leaf traits.**
-
-L470 A thought: Stomatal conductance is highly correlated to photosynthetic capacity (Wong et al., 1979) so the gradients in conductance are expected to follow the gradients in photosynthetic capacity.
-The water use efficiency (gs/A) or the slope parameter of conductance models (Medlyn et al., 2011) could change vertically inside the profiles.
-I don’t know if there are a lot of studies on the vertical variation of the water use efficiency or the slope parameter, but that could be something interesting to mention.
-See for example (Buckley, 2021).
-
-**We appreciate this comment, and have added mention of the fact that maximum $g_s$ increases with light in the canopy because it typically tracks photosynthetic capacity, per @wongStomatalConductanceCorrelates1979.
-There is a great deal of literature showing that the relative limitation of photosynthesis by stomata (gs/A, or intrinsic WUE) tends to be greater in the upper canopy, which results in lower intercellular CO~2~ and C isotope discrimination, as noted in the original manuscript (lines 478).**
-
-I also think you could describe a little bit more the hydraulic constraints that increase with height.
-See for example (Koch et al., 2004).
-I think there are also other papers discussing the effect of height on water potential and stomatal conductance.
-You mention some aspects of it later, around L 636.
-
-**We have added mention of a few papers documenting the hydraulic constraints with height as they affect stomatal conductance [@yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010].**
-
-L471, I think the word conductance is missing.
-
-**Done**
-
-L 513 Vcmax (c in subscript?)
-
-**Done**
-
-L 609 It is a detail, but you sometimes use the word ‘canopy’ to describe the top of the vegetation, and sometimes you precise ‘top of canopy’.
-You also use canopy to describe all the vegetation from understory to the canopy, and you sometimes change definition from one sentence to another.
-Usually, it is easy to understand with the context of the sentence, but you might want to be consistent everywhere.
-
-**Thanks for pointing this out.
-We have revisited the use of the term "canopy" throughout the entire manuscript and employed with consistent terminology, using "canopy" to refer only to the layer formed by dominant trees, and using a term such as to "vertical profile" to refer to the entire gradient from understory to the canopy.**
-
-L 665 though
-
-**Done**
-
-More details on my thoughts for the ‘Scaling across space and time’ paragraph:
-
-Most DGCMs represent a vertical variability in the main physiological parameters (Vcmax, Jmax, Rdark, Na).
-See for example (Krinner et al., 2005; Clark et al., 2011; Oleson et al., 2013).
-They all use some form of an exponential decrease in the photosynthetic parameters from the canopy to the ground.
-Since a lot of parameters and variables depend on Vcmax (for example A and therefore gs), this allows the representation of the vertical variation of a lot of leaf traits and variables.
-Big leaf models have this capability (see for example Krinner et al.
-2005 that you cited, Appendix, or Clark et al.
-2011).
-I think it is not clear in the way you wrote this paragraph, and it looks like ‘big leaf’ models do not represent any vertical variation at all.
-Note also that it is possible to use multi-layer models with or without considering shaded and sunlit leaves (See for example Clark et al.
-2011).
-
-**Thank you for noting the need to clarify this point.
-We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure to a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text italicized here).
-We also deleted the clause that occurred two sentences later, that had said about big-leaf models: “...which do not represent any vertical stratification…”**
-
-**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax):**
-
-- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their capacity to incorporate vertical variation in leaf traits and physiological processes.
-In general, however, this variation is accounted for via light competition.
-Models partition radiation above and within the forest vertical profile (i.e., direct vs.
-diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018].
-Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table `r table_leaf_metabolism`, e.g., @krinnerDynamicGlobalVegetation2005].**
-
-To me, an important remark is that most gradients in DGVMs are prescribed.
-More data would of course help to better understand and quantify the gradients, and to compare them with model representations (part ‘Scaling in situ data with remote sensing’ of your paper).
-But in the end, the mechanisms explaining those gradients are still not really known, in the sense that they can’t be efficiently modeled prognostically.
-This is due to the multiple sources of parameters variation that you highlighted (light environment, temperature, hydraulic, …).
-A lot of research is done using optimality frameworks to try and predict the photosynthetic and other vegetation traits based on environmental variables.
-See for example (Ali et al., 2016; Buckley, 2021).
-A question I have, is if including more precise cohorts as you suggest, would improve the accuracy of models given all the uncertainty in the parametrization of their traits.
-
-**We agree that there is a recognition in the modeling community that we do need improved understanding of the mechanisms to better represent the level of vertical variation desired by some.
-We added the following sentence to acknowledge that: "Moreover, given the anticipated importance of mid-canopy and understory trees in ecosystem resilience given increasing mortality of canopy trees, it is essential that models separately represent these strata.".
-We also removed some of the language that emphasizes cohort based models as the only or best way to do this.**
-
-**We note that addressing the question poosed here is essentially impossible to answer without doing the research. The value of different modeling approaches is difficult to judge a priori.**
-
-Table S3: Including the words vertical or vertical gradient could have been useful.
-
-**Thanks for the suggestion.
-We agree, but at the same time are confident that the search terms we used covered the majority of the relevant literature.
-Adding additional search terms could always turn up additional papers, but we feel that we have reached a point of diminishing returns, and unfortunately don't have time to expand the search (lead author has moved on from the institution where she performed the work to a PhD program).**
-
-I included below some more references, mostly in the tropics.
-
-**Thank you for the references. We have reviewed these and incorporated as appropriate.**
-
-(Kitajima et al., 1997; Valladares et al., 1997, 2000; Thomas & Bazzaz, 1999; Carswell et al., 2000; Koch et al., 2004; Coste et al., 2005; Domingues et al., 2005; Lloyd et al., 2010; Van Goethem et al., 2014; Crous et al., 2020; Béland & Baldocchi, 2021)
-
-Ali AA, Xu C, Rogers A, Fisher RA, Wullschleger SD, Massoud EC, Vrugt JA, Muss JD, McDowell NG, Fisher JB, et al.
-2016.
-A global scale mechanistic model of photosynthetic capacity (LUNA V1.0).
-Geoscientific Model Development 9: 587–606.
-
-Béland M, Baldocchi DD. 2021. Vertical structure heterogeneity in broadleaf forests: Effects on light interception and canopy photosynthesis. Agricultural and Forest Meteorology 307: 108525.
-
-Buckley TN. 2021. Optimal carbon partitioning helps reconcile the apparent divergence between optimal and observed canopy profiles of photosynthetic capacity. New Phytologist 230: 2246–2260.
-
-Carswell FE, Meir P, Wandelli EV, Bonates LCM, Kruijt B, Barbosa EM, Nobre AD, Grace J, Jarvis PG. 2000. Photosynthetic capacity in a central Amazonian rain forest. Tree Physiology 20: 179–186.
-
-Clark DB, Mercado LM, Sitch S, Jones CD, Gedney N, Best MJ, Pryor M, Rooney GG, Essery RLH, Blyth E, et al.
-2011.
-The Joint UK Land Environment Simulator (JULES), model description – Part 2: Carbon fluxes and vegetation dynamics.
-Geoscientific Model Development 4: 701–722.
-
-Coste S, Roggy J-C, Imbert P, Born C, Bonal D, Dreyer E.
-2005.
-Leaf photosynthetic traits of 14 tropical rain forest species in relation to leaf nitrogen concentration and shade tolerance.
-Tree physiology 25: 1127–1137.
-
-Crous KY, Campany CE, Lopez Rodriguez RA, Cano FJ, Ellsworth DS. 2020. [In Press] Canopy position affects photosynthesis and anatomy in mature Eucalyptus trees in elevated CO2. Tree Physiology.
-
-Domingues TF, Berry JA, Martinelli LA, Ometto JP, Ehleringer JR.
-2005.
-Parameterization of canopy structure and leaf-level gas exchange for an eastern Amazonian tropical rain forest (Tapajos National Forest, Para, Brazil).
-Earth Interactions 9: 1–23.
-
-Kitajima K, Mulkey SS, Wright SJ. 1997. Seasonal leaf phenotypes in the canopy of a tropical dry forest: photosynthetic characteristics and associated traits. Oecologia 109: 490–498.
-
-Koch GW, Sillett SC, Jennings GM, Davis SD. 2004. The limits to tree height. Nature 428: 851–854.
-
-Krinner G, Viovy N, Noblet-Ducoudré N de, Ogée J, Polcher J, Friedlingstein P, Ciais P, Sitch S, Prentice IC.
-2005.
-A dynamic global vegetation model for studies of the coupled atmosphere-biosphere system.
-Global Biogeochemical Cycles 19.
-
-Lloyd J, Patiño S, Paiva RQ, Nardoto GB, Quesada CA, Santos AJB, Baker TR, Brand WA, Hilke I, Gielmann H, et al.
-2010.
-Optimisation of photosynthetic carbon gain and within-canopy gradients of associated foliar traits for Amazon forest trees.
-Biogeosciences 7: 1833–1859.
-
-Medlyn BE, Duursma RA, Eamus D, Ellsworth DS, Prentice IC, Barton CVM, Crous KY, Angelis PD, Freeman M, Wingate L.
-2011.
-Reconciling the optimal and empirical approaches to modelling stomatal conductance.
-Global Change Biology 17: 2134–2144.
-
-Oleson K, Lawrence D, Bonan G, Drewniak B, Huang M, Koven C, Levis S, Li F, Riley W, Subin Z, et al.
-2013.
-Technical description of version 4.5 of the Community Land Model (CLM).
-Boulder, Colorado, US: NCAR TECHNICAL NOTES.
-
-Thomas SC, Bazzaz FA. 1999. Asymptotic height as a predictor of photosynthetic characteristics in Malaysian rain forest trees. Ecology 80: 1607–1622.
-Valladares F, Allen MT, Pearcy RW. 1997. Photosynthetic responses to dynamic light under field conditions in six tropical rainforest shrubs occuring along a light gradient. Oecologia 111: 505–514.
-
-Valladares F, Wright SJ, Lasso E, Kitajima K, Pearcy RW. 2000. Plastic phenotypic response to light of 16 congeneric shrubs from a Panamanian rainforest. Ecology 81: 1925–1936.
-
-Van Goethem D, Potters G, De Smedt S, Gu L, Samson R.
-2014.
-Seasonal, diurnal and vertical variation in photosynthetic parameters in Phyllostachys humilis bamboo plants.
-Photosynthesis Research 120: 331–346.
-
-Wong SC, Cowan IR, Farquhar GD. 1979. Stomatal conductance correlates with photosynthetic capacity. Nature 282: 424–426.
+-I found the x axes labels in Fig 4.a difficult to understand, there is maybe a more intuitive way to write them. Note that PET is not defined. °C for the units.
 
@@
 
-A defining feature of forest canopies is their vertical variation in environment and morphology and physiology, originally described through the sun-shade leaf dichotomy or even earlier through vertical stratification diagrams as in Pearson (1971), dating further back such as the classic work of P.
-W.
-Richards (1952).
-Such observations were carried through to the forest meteorology community by Monteith, Jarvis and others in the 1980s (Monteith & Unsworth 1990), and there are well-known patterns in micrometeorology through canopies and first-principle understanding of their meaning for energy balance within canopies.
-I read this review keenly interested in what these known aspects would mean for the interactions amongst leaf traits and sensitivity to temperature, tie them back to some empirical findings of more recent vintage, and possibly to gain some insights into canopy dieback phenomena during heat events or with climate warming.
-In the end, the latter weren’t discussed but the authors are applauded for taking on such a huge topic.
-Fig.
-1 is a useful entrée to the subject area.
-
-**Thank you for the review.
-We have added citations to the references mentioned here [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].**
-
-**We have also added the following sentence on canopy dieback to the canopy disturance section:**
-**"Increasing severity and frequency of heat waves, accompanied with increases in VPD and ET, can exacerbate effects of drought on predominantly canopy trees, potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021]."**
-
-Given the depth of what is known about within-canopy micrometeorology, I would have thought taking this on in a review wouldn’t be needed, or would be daunting and require too much detail to cover it all.
-As it is, I found the article launches into the empirical info too early in such an article.
-The article could be improved by going through the theory of how momentum and mass-transfers are attenuated through canopies and then show the empirical info from NEON in America.
-Overall there is too little on the theory end of things and too much 'case study' for this kind of article (see Monteith and Unsworth, and Gates to enhance the first-principle theory).
-As an aside, there need to be a number of improvements in Fig.
-2, with lines made bolder and also height should be normalised to height relative to the top of the canopy given that trees were very different heights and meteorological masts went above the canopy to different degrees.
-Doing so would help make the authors’ point about differences in the Figure as well as convergence in certain properties inside canopies rather than everything being on a different y-scale.
-
-**Regarding incorporation of more theory, please see our response to the comment below.**
-
-**We have made the suggested improvements to figure 2: lines are bolder, and height normalized to height at the top of the canopy.**
-
-Certainly, a discussion with a micrometeorologist or atmospheric scientist, if not already had, could improve the manuscript.
-Statements like l.
-156: 'Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy' are an awkward read.
-Buffering?
-There could be much improvement by looking at it the other way: wind will blow as it does at the top of the canopy until encountering the plant canopy top as an aerodynamic drag element; then the additional leaf area entrains air movement and sweep-eject motion and eddies are attenuated through the canopy resulting in progressively lower windspeeds deeper into canopies.
-I don’t know how buffering enters into this, or what the authors mean here.
-
-**We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines 169-192), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles.
-We have also modified the subsequent paragraph about wind speeds to eliminate the use of the confusing term “buffering”.**
-
-**We had also used the words “buffering”, "buffer", or "buffered" in a different context, where the intention was to describe how canopy attenuation (of radiation and wind) can reduce daily extremes of temperature in the lower canopy and understory.
-We have modified the text in those places to clarify, generally reducing use of the terms.
-However, the term has been widely used in recent publications on this theme, and we therefore did not feel it was necessary to eliminate it's usage.
-Examples of modified sentences include the following:**
-
-- **“Canopy foliage is the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions experienced above the canopy.”; (line xxx)***(new text italicized)**
-- **“Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019]" (line 229)**
-- **“Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting from,***~~lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings.” (line 232)**
-
-I was eager to see the part of the manuscript about photoprotection and also its relationship to heat damage.
-The opening sentence (l.
-384) was obvious to the point of being painful to read.
-Of course photoprotection is higher in the upper canopy: radiation levels are higher!
-Can this first sentence get revised to increase information content?
-This is one of the more exciting topics of the review, yet is handled only in a cursory fashion.
-If the authors could manage a bit more discussion of this topic it would be useful – instead they move (in the same paragraph) to VOC emissions, something I view as a separate capability of some plants.
-This is also confused in Fig.
-1 where VAZ and VOC are placed on the same line.
-I’d like to see a more complete discussion of photoprotection and heat dissipation by NPQ first, and a separate paragraph about VOC.
-Also, there is nothing about critical temperature until late in the manuscript even though the authors skirt the issue in their text about photoprotection.
-This kind of thermal sensitivity is different from gas exchange (the section where the text about Tcrit occurs, l.
-530-555), and should have been discussed earlier in the manuscript.
-
-**We have revised the first sentence here to read: "Biochemical protection against light and heat damage increases with peak radiation loads, and thus tends to be higher in the upper canopy than in the understory (Table 1)." We note that it is not necessarily obvious that mid- or lower-canopy leaves should have less photoprotection, given that they may experience nearly the same peak radiation loads, during sunflecks, as leaves in the upper canopy, despite generally having less capacity to dissipate light by photochemistry.
-We also choose to keep this rather obvious sentence because we feel that basic topic sentences are important for helping to guide readers from a diversity of sub-disciplines through the manuscript.**
-
-**We separated VAZ section from VOC, and provided further context to NPQ in the following paragraph:**
-
-- **More frequent stomatal closure in upper canopy leaves (section 4) reduces their capacity to use light energy for photochemistry, thereby requiring a high capacity to dissipate excess light energy and protect against photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].
-Accumulation of excess light energy causes overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
-A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by excess light [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
-This form of NPQ entails interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
-Leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].**
-
-**We also moved thermal sensitivity content to the traits section (now with its own sub-heading).**
-
-In the first part of Section 5.1, the authors conclude this paragraph stating the 'dominant role of vertical profiles in microclimate in shaping tree growth rates'.
-However, I believe the text has confused the very large effect of light with other aspects of forest microclimate, and they very much need to clarify this.
-If they believe other aspects of microclimate apart from light are drivers, then I believe this would be a minority opinion in the forest biology community.
-If they wish to speculate about other microclimate factors that cannot be disentangled from light, I suggest that they clearly indicate their opinion and speculation.
-It is perhaps an area for future research to disentangle light microclimate from aspects of vertical microclimate.
-
-**We did not intend to argue that light was not the dominant factor, but the wording was misleading.
-This sentence has been revised to read, "This points to a dominant role of vertical profiles in the biophysical environment, particularly light, in shaping tree growth rates within forests."**
-
-I am uncomfortable reading some of the major conclusions of the work here, such as speculation that large canopy trees are most vulnerable to warming when water is limited, but understory trees may be more vulnerable [to warming] when well-watered.
-This is not consistent with my knowledge of 'preponderance of available data'.
-Perhaps the authors implying that understory trees will succumb to warming earlier in a heat wave than large canopy trees?
-Is this based on the T50 evidence (l.
-542-548) which is from 2 studies, or from one tree-ring study (l.
-674-676), or both?
-An intriguing thought, but these studies and the corresponding author’s study have weaknesses and no clean experiment on this has been set up and done.
-I recommend that the more cautious language on l.
-676-678 should in fact be repeated in the conclusions relevant to this point.
-It’s less stimulating and sober, but true, that it’s difficult to say so with strength of evidence and remains for further testing with good experimental design.
-An objective of such a review can stimulate further and better research.
-
-**We agree that more caution is warranted on comparing canopy vs understory responses to warming, particularly because understory responses are highly uncertain.
-We have reworded relevant text as follows:**
-
-- **"Implications: Global Change Responses : Warming" section, 4th par (previous lines 793-797: "While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (see section 4.2), *limited tree-ring evidence* indicates that understory trees *can* exhibit greater reductions in growth during unusually hot growing seasons [section 5.1, Fig 4b, @rollinsonClimateSensitivityUnderstory2020]."**
-
-- **"Implications: Global Change Responses : Warming" section, final par (previous lines 803-806: "To the extent that warming is coupled to drought, we expect that tall trees with exposed crowns will usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment.
-*In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability* [@bennettLargerTreesSuffer2015;@hoganDroughtInterannualVariability2019; @nunesForestFragmentationImpacts2022]."**
-
-- **Conclusions section, first part (text specifically cited in the comment above): "Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth.
-While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
-4)."**
-
-### Details
-
-There are a few dense sentences or run-ons that could be revised. Lines 192-196 and lines 201-204, for instance. Please revise.
-
-**We revised the following lines:**
-
-- **Lines 192-196: Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a ‘canopy greenhouse effect’**
-
-- **Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
-Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$, due to turbulent air mixing and interception of thermal radiative flux from the soil and the canopy by lower-canopy layers.
-**
-
-- **Lines 201-204: $T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes (Drake et al., 2020; Perez & Feeley, 2020; Michaletz et al., 2015)**
-
-- **$T_{leaf}$ is strongly tied to, and usually within a few degrees of, $T_{air}$ [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Note S1, @doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009].
-Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance ($g_s$; Fig.
-3; Note S1), all of which vary across forest vertical gradients (Fig.
-1).
-**
-
-line52: 'with decreases being more commonly documented across the world’s forests'
-
-**We do not understand this comment, but have verified that this sentence is correct.**
-
-line128: '… but understory leaf area density is often relatively high in the understory as well' sounds redundant. Please reword.
-
-**We removed the second 'understory' word here:** 
-**"In tropical and temperate forests with dense broadleaf canopies, leaf area density is generally highest in the canopy layer (i.e., that formed by the crowns of dominant trees), but understory leaf area density is often relatively high ~~in the understory~~ as well"**
-
-line140 'Light … decreases from the canopy top to the forest floor' – why isn’t foliage clumping mentioned here amongst the other factors?
-
-**We added foliage clumping into the list of factors influencing vertical light gradient here:**
-**"The intensity of visible and photosynthetically active radiation (PAR, 400-700nm) decreases from the canopy top to the forest floor, with a profile whose shape is modified by leaf area density, leaf clumping, canopy height, and vertical structure across species and forest types [Fig.
-2a-d, Supporting Information Figure S1, @belandVerticalStructureHeterogeneity2021; @koike_leaf_2001; @binLeafTraitExpression2022]."**
-
-line148: 'Variability in the light environment decreases with height' – I don’t see this except perhaps at z = 0 and even so seems tenuous. Explain.
-
-**We revised the sentence to make it clearer:**
-
-- **Previous sentences: Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012].
-Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
-
-- **Changed sentences: Mid-canopies and understories experience a highly dynamic light environment due to sunflecks, or brief increases in direct solar radiation, caused by small canopy gaps, wind-induced canopy movements or the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].**
-
-Fig.
-1: In the diagram, stomatal density, leaf angle are separate categories yet VAZ and VOC are not.
-As far as I know though, there is no biochemical link between VAZ and VOC, so these should be separate shouldn't they?
-
-**We agree with the suggestions.
-Therefore, to make it more clearer, we added photoprotection as one of the factors in Fig.1 and isoprene emission in a separate line instead of VAZ and VOC in the same line.**
-
-Fig.
-2 is a mix of computed and measured quantities.
-LAD and proportion of sun leaves are both computed quantities, and the basis of these should be clearer.
-Or at least state 'modelled LAD' and 'calculated proportion of sun leaves'.
-
-**Done**
-
-line165-172: refixation of respired CO2 in the understory isn’t mentioned here, but there have been a number of studies, particularly involving 13C, that have looked at this.
-Please mention along with at least 1 reference.
-
-**Thanks for the suggestion.
-We have included the following sentence with the refrence: While understory seedlings may benefit modestly from higher CO~2~ concentrations near the ground during some parts of the day [e.g., 6% of C fixation from recently respired CO~2~, @brooksVerticalGradientsPhotosynthetic1997]."**
-
-line185: Buffering again.
-I supposed what is 'buffering' for the authors would conventionally be thought of differently by meteorologists.
-Please consult with one of them, and consider using different wording.
-
-**This passage now occurs after the new paragraph about turbulent transport within canopies.
-We have reworded the passage as, “Despite these complexities of within-canopy transport, wind speeds are generally much higher at the top of the canopy than within or beneath [@juckerCanopyStructureTopography2018; @jiao-junWindProfilesTrees2004].”**
-
-line187: Where mentioning radiation fluxes, could I please ask the authors to use the phrase 'and sensible heat emission'? I think this is really what the authors are referring to, so why not say so?
-
-**Done**
-
-line188: 'dense canopies buffer understories' is one perspective, that there is far less radiant energy received and therefore understories stay cool is another.
-Can the authors please consider reflecting this perspective?
-I’ve studies canopies for a a score of years and I just don’t think 'buffering' is at play so much!
-
-**We were using the term “buffer” in a functional or biological sense, meaning “protect,” rather than as a description of the underlying physics; i.e., the canopy protects the understory from large diel swings in both air temperature and radiation.
-We have reworded this passage to clarify: “Typically, ***diel temperature range is smaller beneath dense canopies than above, resulting from***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~ lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings ”**
-
-line199 should end with something like 'to shape Tleaf patterns within canopies' rather than just drop at Tleaf.
-
-**Done**
-
-line206-207 'Leaves are typically warmer than air … '.
-That is not what is predicted in theory, proven by parts of the simulations in Fig.
-3.
-Why not use 'Leaves can be warmer …'?
-That would be more consistent.
-
-**Done**
-
-Figure 3: Variables in the inset table need to be stated. I can guess what they are, but this should be explicit.
-
-**We added the description for the variables as part of the figure caption: Variables in the inset table are now described in the figure caption as: "Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair)."**
-
-Figure 3: Humid and drought is not a straight comparison.
-Well-watered vs.
-drought would be clearer as there are some assumptions about gs in these conditions that underlie the modelling being done here.
-
-**We changed to "humid" to "moist" (as opposed to "well-watered"). We chose the term "moist" because it conveys both high soil moisture and atmospheric humidity, which is what we aim to capture.**
-
-Figure 3: why not put a point that represents the leaf values for top and bottom of the canopy?
-
-**We have added points along each line to indicate the constants applied in the other modeling scenarios. These dots represent our best estimates of "typical" values.**
-
-Table 1: What is here is not leaf gas exchange, and it seems that the authors have repeated the caption for Table 2 in error. Please fix this. Also don’t capitalise ‘leaf’.
-
-**Done**
-
-Lines 354-372 on sun versus shade leaves and Table 1.
-Most of what is here was known a long time ago, say in Vogel 1968 (cited) and Boardman 1977 (not cited) but pertains to within-canopy rather than between sun and shade plants.
-I would suggest that this would be improved if the authors started from these old studies that most readers will recognise, and state how we’ve advanced in understanding and progressed toward within-canopy work rather than between sun-and-shade habitats.
-
-**Thank you for the suggestion. We have added additional information highlighting older work and recent advances in the paragraph below:**
-
-**Anatomical, structural and biochemical leaf traits vary vertically across forest strata [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping vertical profiles of leaf temperature, gas exchange, and thermal sensitivity [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004].
-Vertical variation in leaf structure and composition has long been recognized, with sun and shade leaf traits distinguished for over a century [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
-Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
-More recent work has extended the focus from light alone to height in the vertical profile, as the latter shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
-For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
-Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
-Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
-Overall, leaf biochemistry, anatomy and structure may be optimized to local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].
-However, we lack a a cohesive framework for integrating the many differences in leaf traits throughout the canopy.
-Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004; @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020].
-As canopy shade is known to alter these trait relationships [@osnasDivergentDriversLeaf2018], further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.**
-
-line382: 'more frequent stomatal closure higher' is awkward and does not follow the comparative. Higher than what?
-
-**We agree that the comparison is not very clear here.
-We added "compared to shaded lower canopies" to the sentence: "Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure in sun exposed canopies ***compared to shaded lower canopies,***rreduce intercellular CO~2~ concentrations and increase leaf $\delta^{13}$C (Table 1).**
-
-line383: d13C is NOT a concentration, it is an isotope ratio. Please revise.
-
-**Done**
-
-line404-405: A cause-effect for an earlier, juvenile understory tree leaf-out is implied here, and I;m not sure that is demonstrated.
-There are other phenomena underlying this e.g.
-turgor, hydraulics, hormones, etc.
-
-**Thank you for pointing this out.
-In line 404-405, we are concentrated on interactions between microenvironments and function.
-We edited the text to emphasize the mechanisms for earlier bud break in the understory, including warmer temperatures, and the benefits of maximizing carbon gain before overstory closure reduces light availability [added ref.
-@leeSpringPhenologicalEscape2021].
-We edited the rest of the paragraph to more clearly emphasize mechanisms of feedbacks between vertically structured microenvironments, leaf phenology, and forest function.**
-
-line475: 'because water supply often cannot meet the demands incurred …'. How do we know how frequent this is? Evidence?
-
-**We have added references showing that stomatal conductance is often limited by imbalance of water supply and evaporative demand for leaves in more elevated or more sunlit canopy positions [@yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010].**
-
-line478: 'intracellular CO_2' needs to be fixed.
-
-**Done**
-
-line479: 'In sun leaves, Tleaf thus further increases'. Further than what? Comparatives should have be clear what we’re comparing to.
-
-**This sentence has been rewritten as, “Stomatal depression reduces transpirational cooling (Fig. 3), thus amplifying the warming of sun leaves by high radiation.”**
-
-line481 'is maximized' or is maximal? Please correct.
-
-**We have replaced “maximal” with “greatest”.**
-
-line485-487: It seems that VPD and its effect on gs is ignored here.
-Please clarify?
-The well-known stomatal closure at high VPD alone would mean higher Tair would decrease gs in the upper canopy (not 'canopy leaves' as stated) compared to the lower canopy.
-See Grossiord et al.
-(2020) Tansley Review.
-VPD only gets mentioned in the section on photosynthesis (l.
-496).
-Please do so earlier.
-
-**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ cannot meet the demands caused by the high radiation and $g_b$ in sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al.
-1994, Koch et al.
-2004, Sillett et al 2010) and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014]***.
-These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO~2~ and $\delta^{13}C$ discussed in section 3.3 (Table 1).
-Stomatal depression reduces transpirational cooling (Fig.
-3), thus amplifying the warming of sun leaves by high radiation.
-~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$”, and clarified that “high $T_{air}$ should decrease $g_s$ of ***upper canopy***leaves”.**
-
-line510-525: There is a long discussion about whether Topt for gas exchange or its components differs between overstory and understory.
-Here the authors aren’t clear, but such a comparison only matters if the same species is measured in different canopy layers.
-
-**Thank you for noting this, we revised the sentence to make it clear (please see below), where we specify that majority of the studies are within species between sun and shade leaves of the same trees.
-@carterExperimentalWarmingTropical2021 is one of the only studies comparing between overstory and understory.**
-
-- **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, which  compare sun and shade leaves of the same trees, reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes**
-
-The section on VOC emissions was interesting and informative.
+Comments to the Author
+The authors seek to tackle a fairly complex topic in this review article.
+Whilst the topic of vertical variation in forest canopies is fairly specialised, they have pulled together quite a large body of literature and the review cuts across the fields of micrometeorology, physiology and ecology for forests, and many will want to read and perhaps cite the review.
 
@@
 
-line641: 'tends to be greater in canopy trees'. Greater than what? Please state the comparison.
+The authors have done a good job of addressing the reviewers' comments from my inspection.
+This reviewer apologises for needing to raise additional points that were not raised in the previous reviews of the article, but some conceptual aspects and some of the writing are perhaps flawed or require clarification.
+There are some additions done in the review process that have confused things, or a few aspects that were not picked up in the initial review.
+The authors and readers both benefit from a clearer picture emerging in this review.
+There are still a number of places where the authors could improve and one cannot resist letting these past at this point.
 
-**This sentence has been revised to read, "In turn, the drought sensitivity of woody growth tends to be greater in upper-canopy trees than in smaller trees with less exposed crowns."**
+The words 'disproportionate' or 'disproportionately' are used l.
+33, 35, 504, 648, 814, 838.
+For instance, l.
+35 'disproportionately impacted under hot, humid conditions': disproportionate to what?
+The implication is disproportionate to canopy trees, but there is almost no evidence of this.
+The authors need to be clear what the comparative is in each case that a form of 'disproportionate' is used, especially in their abstract.
 
-line650: 'In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover'.
-Having studied large overstory trees and young understory saplings, I have a very, very difficult time with this statement especially portraying it as categorically true even if consistent with the 4 cited studies.
-Very few researchers have studies the same species in this context, and in this review manuscript the authors cannot scientifically lay the effect on canopy position and microclimate rather than age or other confounding factors.
-I must challenge their assertion and ask a higher level of evidence to make these statements, which should be qualified.
+In fact, the abstract reads like the authors cannot decide whether tall trees '...
+are disproportionately vulnerable to drought and damaging 𝑇𝑙𝑒𝑎𝑓' (l.
+33) or understory trees (l.
+35).
+It cannot be both if one is comparing them.
+In this aspect the abstract comes off indecisive or unclear.
+This is an element in the review as well (see below).
 
-**As the point about resilience was not critical, we have removed that part of the statement.
-The sentence now reads, "Drought also causes greater mortality in larger trees [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019]."**
+I have issue with the ecosystem level being mentioned in the Abstract.
+There was little true 'ecosystem level' assessment done in the manuscript, and most of the 'ecosystem level' inferences are based on these statements which are indirect:
 
-line946: '… how … temperature sensitivities of metabolism and woody growth vary across these vertical gradients'.
-I’m not sure why the authors are speculating on woody growth in this context, surely it would be extremely difficult to show that woody growth varied vertically in an individual?
-And that across-individual differences compound comparisons of individuals in different vertical strata?
-Shouldn’t these things be pointed out to readers?
+l.744 'Because canopy trees dominate these fluxes (Fig. 5), their responses will strongly influence the whole-ecosystem response'
 
-**This has been reworded to read, "However, uncertainty remains about how temperature sensitivity of foliar gas exchange varies across these vertical gradients.
-Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth.
-While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
-4)."**
+l.898 'disproportionate role of large trees in ecosystem dynamics'
+
+So perhaps the authors can greatly soften their ecosystem level assessment. To their credit, and what truly should be emphasised, is the further research needed on ecosystem function (l. 757).
+
+l.32 tall trees have higher absolute metabolic capacity and growth at both individual *and ecosystem levels* (please remove the text between the asterices).
+
+I make a major point of revisions around this idea (l.
+35-36) that 'understory trees may be disproportionately impacted under hot, humid conditions'.
+What evidence is there?
+There is speculation about what might happen in sunflecks (l.
+800-801) but little direct evidence, and there is Fig.
+4b which is one study with smoothed results that are not convincing (see above).
+Yet this statement is contrary to statements made in the actual review, like l.
+200-201 to the effect that upper canopy is warming with higher VPD than understory, l.
+232 that there are higher leaf-air temperature difference in the upper canopy than understory, l.
+275-276 that leaves in the upper canopy have higher leaf-air temperature difference and maximum temperatures than understory, also l.
+775-776.
+If so many arguments have been made for a harsher environment and higher leaf temperatures in the upper canopy than understory, it comes off as illogical to reverse and conclude on the basis of Fig.
+4b which is flimsy evidence that, as stated in the abstract l.
+35-36: 'understory trees may be disproportionately impacted under hot, humid conditions'.
+I would suggest that the authors clean up their contentions and not come off so contrary to their own evidence and logic.
+Certainly this titillating idea of understory tree damage is not to be highlighted in the abstract if so much evidence to the contrary is presented in the review.
+I would like to see direct observations of the impacts that they suggest for the understory, but in the absence of strong arguments the balance of the review would support upper canopy vulnerability rather than the understory.
+This is a major aspect for the authors to clarify.
+
+These are wording issues where the writing can improve:
+
+l.187: 'but the elevated CO2 quickly dissipates during the day such that differences are negligible for most of the day when photosynthesis is occurring'.
+Please specify 'CO2 concentration differences' where it says differences, to increase clarity.
+
+l.200 'upper canopy leaves have higher 𝑔𝑏 and associated potential for water and heat loss' compared to what??? Please clarify.
+
+l.256: The statement "During drought (hot and dry, with higher-than-average solar radiation) ..." is worded in a confusing manner.
+Meteorological 'Drought' is conventionally defined as a prolonged shortage in the supply of water to land or plants, but with no link made to temperature or cloudiness.
+It is fine if the authors associate drought with high temperatures and high radiation, but the parenthetical should emphasise this association, whereas it could be read to imply a one-for-one relationship which is incorrect.
+
+I suggest revising as 'associated with conditions that are hot ...'.
+
+l.264-265 'their environment is less conducive to dissipating excess heat compared to upper-canopy leaves.' What evidence?
+This is a logical argument and should either be backed with references or labelled as a logical outcome of the conditions they have stated.
+
+l.269 How does lower VPD in the understory inhibit evaporative cooling, if VPD is the major driver for evaporation?
+I think I understand how this can be counter-intuitive, but the reasoning isn't explained so it all becomes opaque to the reader.
+Please explain very briefly.
+
+l.313 'information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, ...'.
+This was not in the previous version that I can see, and I don't see the basis for inclusion of this statement.
+As well, use of the term 'memory' seems awkward if not entirely wrong (even reading between quotes).
+Please revise.
+
+l.579 'implies that high 𝑇𝑎𝑖𝑟 should decrease 𝑔𝑠 of upper canopy leaves more than understory leaves'.
+The authors here are ignoring what they’ve said previously, that high Tair is associated with high VPD (l.
+199-200), and that VPD is conclusively a much stronger driver of gs than Tleaf (Leuning 1995, Grossiord).
+I think the sentence should be corrected and include VPD.
+I'm not yet aware of a good study that has decoupled VPD and Tleaf to prove that Tleaf intrinsically causes reduction in gs.
+So the authors cannot conclude this.
+
+l.629 'we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile'.
+Other than that it varies according to the prevailing temperature difference?
+I think the temperature difference between upper and lower canopy positions, the theme of this review, should be mentioned in regard to this.
+
+l.806-807 'trees in the understory might be more negatively affected by chronic stress from warming 𝑇𝑎𝑖𝑟'.
+This is based on such very thin evidence, almost no evidence.
+Nor do I think that the  authors are appropriately highlighting a lack of knowledge very well by their speculation (which is not based on data).
+They are inferring what would happen with climate warming on the basis of historical evidence, highly indirect in Fig.
+4B which only shows modelled results, and when there are highly overlapping CI's.
+I hardly find it a basis for a strong statement landing in the abstract, especially when it is contradicted by other statements throughout the review.
 

### 2022-06-09 response-to-reviews.Rmd
first pass through R1 comments
@NidhiVinod , FYI

@@
 
+*https://github.com/EcoClimLab/vertical-thermal-review/issues/118*
+
 -Fig 3 caption: TLeaf and some other discreptances with rest of the paper abreviations or format (tair, gs)
+
+*Nidhi*
 
@@
 
+**The full sentence reads: "More recent work has extended the focus from light alone to *height* in the vertical profile, as *the latter* shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy." Thus, we're saying that *height* shapes differences in leaf traits that caould mitigate effects of gravity and hydraulic pathlength.
+We believe the sentence to be correct as written.**
+
 -Table 2, gb unit, mmol m-2 s-1 ?
 
+*Nidhi*
+
 -I found the x axes labels in Fig 4.a difficult to understand, there is maybe a more intuitive way to write them. Note that PET is not defined. °C for the units.
+
+*https://github.com/EcoClimLab/vertical-thermal-review/issues/119*
 

### 2022-06-09 vertical-thermal-review.Rmd
edit Fig. 4 caption to clarify x-axes in panel a
#119

@@
 Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species.
-Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET over various seasonal windows, where "c" and "p" represent months in the current and previous year, respectively) for the maximum and minimum tails of the DBH distribution included in the analysis.
+Lines represent responses of trees at the maximum and minimum tails of the DBH distribution included in the analysis to the most influential temperature-related variable at the site.
+These variables are maximum temperature ($T_{max}$) or potential evapotranspiration (PET) over a seasonal window spanning from the noted beginning to end month, where "c" and "p" represent months in the current and previous calendar year, respectively.
 Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction.

### 2022-06-09 vertical-thermal-review.Rmd
additional edit to Fig. 4 caption
#119

@@
 ![**Figure `r fig_treerings`.
-Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species.
+Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures than did small trees of the same species.
 Lines represent responses of trees at the maximum and minimum tails of the DBH distribution included in the analysis to the most influential temperature-related variable at the site.

### 2022-06-09 vertical-thermal-review.Rmd
remove Rollinson panel in Fig. 4, make associated changes to text
#120

--- 2022-04-04 vertical-thermal-review.Rmd
@@
 Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity across the gradient appear modest.
-Scaling from leaves to trees, tall trees have higher absolute metabolic capacity and growth at both individual and ecosystem levels, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$, particularly under climate change.
+Scaling from leaves to trees, tall trees have higher absolute metabolic capacity and growth, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$, particularly under climate change.
 In contrast, understory trees experience fewer extreme high $T_{leaf}$'s but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when exposed to harsher conditions through canopy disturbance.
@@
 However, under well-watered conditions conducive to high $g_s$, we do not necessarily expect higher thermal sensitivity of photosynthesis (section 4.2) or woody growth in upper-canopy trees than their understory counterparts.
-In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
+Indeed, there are cases where tree-ring records show more negative or less positive $T_{air}$-responses of understory trees compared to canopy dominants [@anderson-teixeiraJointEffectsClimate2022; @rollinsonClimateSensitivityUnderstory2020].
 Additional research is required to understand the mechanisms underlying these intriguing differences, and to disentangle size-related tree growth responses to hot-wet versus hot-dry conditions.
@@
 ![**Figure `r fig_treerings`.
-Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species.
-Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET over various seasonal windows, where "c" and "p" represent months in the current and previous year, respectively) for the maximum and minimum tails of the DBH distribution included in the analysis.
-Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction.
+Examples of tree-ring evidence for greater interannual temperature sensitivity of annual growth for large than for small trees.** Across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high maximum temperature ($T_{max}$) or potential evapotranspiration (PET) than did small trees of the same species.
+Independent variables are the most influential temperature-related variable at the site ($T_{max}$ or PET over a seasonal window spanning from the noted beginning to end month, where "c" and "p" represent months in the current and previous calendar year, respectively).
+Colored lines represent responses  of trees at the maximum and minimum tails of the DBH distribution included in the analysis to the temperature variable in a model including a DBH - temperature interaction.
 Other model terms are held constant at their mean.
 Transparent ribbons indicate 95% confidence intervals.
-From @anderson-teixeiraJointEffectsClimate2022.
-In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests.
-Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively.
-Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals.
-From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)
+From @anderson-teixeiraJointEffectsClimate2022.](tree_rings/pre_temp_groups_dbh_interactions_for_NIDHI.png)
 
@@
 It remains far less clear how thermal sensitivity varies across forest strata under wet conditions, but both physiological mechanisms (Fig.
-`r fig_leaf_T`) and tree ring evidence (Fig.
-`r fig_treerings`) suggest that understory trees may respond more negatively to hot, wet conditions.
+`r fig_leaf_T`) and some tree ring evidence [@anderson-teixeiraJointEffectsClimate2022; @rollinsonClimateSensitivityUnderstory2020] raise the possibility that understory trees may be at a relative disadvantage under hot, humid conditions.
 Further research is required to better understand the thermal sensitivity of forest ecosystem function across strata.
@@
 <!--Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
-While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [Fig `r fig_treerings`b, @rollinsonClimateSensitivityUnderstory2020].
+While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [@rollinsonClimateSensitivityUnderstory2020].
 Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].
@@
 Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth. 
-While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig.
-`r fig_treerings`).
+While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions [@rollinsonClimateSensitivityUnderstory2020].
 

### 2022-06-09 vertical-thermal-review.Rmd
replace a bunch of latex symbols with alternative
@NidhiVinod , FYI. .Rmd now supports at least some special characters, and I've found that the latex symbols look bad in the final online version.

@@
 Using published research and new analyses, we evaluate how microclimate and leaf temperatures, traits, and gas exchange vary vertically in forests, shaping tree ecology and ecosystem function. 
-In closed-canopy forests, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature ($T_{leaf}$), particularly when transpirational cooling is curtailed by limited stomatal conductance.
-However, foliar traits also vary across height or light gradients, partially mitigating and protecting against the elevation of upper-canopy $T_{leaf}$.
+In closed-canopy forests, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature (*T~leaf~*), particularly when transpirational cooling is curtailed by limited stomatal conductance.
+However, foliar traits also vary across height or light gradients, partially mitigating and protecting against the elevation of upper-canopy *T~leaf~*.
 Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity across the gradient appear modest.
-Scaling from leaves to trees, tall trees have higher absolute metabolic capacity and growth, yet are disproportionately vulnerable to drought and damaging $T_{leaf}$, particularly under climate change.
-In contrast, understory trees experience fewer extreme high $T_{leaf}$'s but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when exposed to harsher conditions through canopy disturbance.
+Scaling from leaves to trees, tall trees have higher absolute metabolic capacity and growth, yet are disproportionately vulnerable to drought and damaging *T~leaf~*, particularly under climate change.
+In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when exposed to harsher conditions through canopy disturbance.
 As the climate changes, integrating the patterns and mechanisms reviewed here into models will be critical to forecasting forest-climate feedbacks.
@@
 
-Global average temperatures have risen 1.2$^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach +1.5$^\circ$C by 2040 [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves and hotter droughts [@meehlMoreIntenseMore2004; @ipccClimateChange20212021].
+Global average temperatures have risen 1.2°C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach +1.5°C by 2040 [@ipccClimateChange20212021], accompanied by increasing severity and frequency of heat waves and hotter droughts [@meehlMoreIntenseMore2004; @ipccClimateChange20212021].
 These changes are affecting tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021] by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; <!--@corlettImpactsWarmingTropical2011;--> @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020].
@@
 (1) the biophysical environment;
-(2) leaf temperature ($T_{leaf}$);
-(3) the leaf traits that most strongly influence $T_{leaf}$ and thermal tolerance;
+(2) leaf temperature (*T~leaf~*);
+(3) the leaf traits that most strongly influence *T~leaf~* and thermal tolerance;
 (4) leaf gas exchange and its thermal sensitivity; and 
@@
 ![**Figure `r fig_schematic`.
-Schematic summarizing typical vertical gradients in (1) biophysical environment, (2) leaf temperature ($T_{leaf}$), (3) leaf traits, (4) leaf gas exchange, and (5) tree and ecosystem ecology within closed-canopy forests during daytime growing season conditions.** Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.
+Schematic summarizing typical vertical gradients in (1) biophysical environment, (2) leaf temperature (*T~leaf~*), (3) leaf traits, (4) leaf gas exchange, and (5) tree and ecosystem ecology within closed-canopy forests during daytime growing season conditions.** Arrows indicate direction of increase, with double-pointed arrows indicating that the direction of the trend is uncertain or inconsistent.
 Patterns tend to be weaker, or sometimes reversed, in more open forests, when canopy trees are seasonally deciduous, or at nighttime.
-Abbreviations are as follows: $T_{air}$: air temperature, VPD: vapor pressure deficit;  $T_{leaf} - T_{air}$: leaf-to-air temperature difference; LMA: leaf mass per area; T sensitivity: temperature sensitivity; ET: evapotranspiration.
+Abbreviations are as follows: *T~air~*: air temperature, VPD: vapor pressure deficit;  *T~leaf~* - *T~air~*: leaf-to-air temperature difference; LMA: leaf mass per area; T sensitivity: temperature sensitivity; ET: evapotranspiration.
 *Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`).](schematics/Schematic figure 1.png)
@@
 Vertical gradients in the biophysical environment for six US forest sites in the National Ecological Observatory Network (NEON)**.
-Height profiles, normalized relative to the top of the canopy, are shown for: growing season (a) modelled leaf area density, (b) estimated proportion of sun leaves, and (c) proportion of light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature ($T_{air}$), and (h) maximum biological temperature, ($T_{bio}$).
+Height profiles, normalized relative to the top of the canopy, are shown for: growing season (a) modelled leaf area density, (b) estimated proportion of sun leaves, and (c) proportion of light incident to the top of the canopy (as fraction relative to to top of canopy), and for July mean ± 1 standard deviation for (d) maximum photosynthetically active radiation (PAR), (e) maximum wind speed, (f) minimum humidity (min RH), (g) maximum air temperature (*T~air~*), and (h) maximum biological temperature, (*T~bio~*).
 Measurements extend from ground level (normalized height = 0 m) to the top of the canopy (horizontal dashed line at normalized height = 1) or above (d-g).
@@
 `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; <!--@mcgregorTreeHeightLeaf2021;-->@juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalinpress].
-As a result, upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for exchange of both mass and sensible heat exchange (sections 2, 4.1).
+As a result, upper canopy leaves have substantially higher boundary layer conductance (*g~b~*) and therefore greater potential for exchange of both mass and sensible heat exchange (sections 2, 4.1).
 
@@
 Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020].
-In combination with $T_{air}$ and $T_{leaf}$, RH determines leaf-to-air vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
-Especially when canopy height and leaf area index (LAI) are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that upper canopy leaves have higher $g_b$ and associated potential for water and heat loss (sections 2, 4.1).
+In combination with *T~air~* and *T~leaf~*, RH determines leaf-to-air vapor pressure deficit (VPD) -- the driving force of water loss from leaves -- which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+Especially when canopy height and leaf area index (LAI) are large, maximum *T~air~* and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
+The implication is that upper canopy leaves have higher *g~b~* and associated potential for water and heat loss (sections 2, 4.1).
 
 ### 1.5 Air temperature 
-Vertical gradients in air temperature ($T_{air}$) vary widely (Fig.
+Vertical gradients in air temperature (*T~air~*) vary widely (Fig.
 `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`), depending on factors such as vegetation type, local topography, vertical air mixing, cloud cover, sensible heat emissions, and times of day and year [@defrenneForestMicroclimatesClimate2021; @zellweger_seasonal_2019].
 Commonly, attenuation of radiation and vertical transport of sensible heat by the canopy buffers the lower canopy and understory from large diel and seasonal swings in air temperature [@defrenneForestMicroclimatesClimate2021; @haesenForestTempSubcanopyMicroclimate2021a; @zellweger_seasonal_2019].
-Diel temperature range is typically smaller beneath dense canopies than above, resulting from lower maximum daytime $T_{air}$ and warmer nighttime minimum $T_{air}$ than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->].
-Notably, buffering against high maximum daytime $T_{air}$ becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
-Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$, due to turbulent air mixing and interception of thermal radiative flux from the soil and the canopy by lower-canopy layers [e.g., OSBS, Fig.
+Diel temperature range is typically smaller beneath dense canopies than above, resulting from lower maximum daytime *T~air~* and warmer nighttime minimum *T~air~* than beneath open canopies or in nearby clearings [<!-- @missonPartitioningForestCarbon2007--> @defrenneForestMicroclimatesClimate2021; @vonarxSpatiotemporalEffectsForest2012a; <!--@defrenneGlobalBufferingTemperatures2019; @davisMicroclimaticBufferingForests2019--> @zellweger_seasonal_2019 <!--@ramboCanopyMicroclimateResponse2009; @binLeafTraitExpression2022; @haesenForestTempSubcanopyMicroclimate2021a; @meeussenMicroclimaticEdgetointeriorGradients2021-->].
+Notably, buffering against high maximum daytime *T~air~* becomes weaker and can be reversed as canopies become more open [Supporting Information Figure `r SIfig_NEON_all`,@curtis_intracanopy_2019; @meeussenMicroclimaticEdgetointeriorGradients2021].
+Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum *T~air~* can be warmer than canopy *T~air~*, due to turbulent air mixing and interception of thermal radiative flux from the soil and the canopy by lower-canopy layers [e.g., OSBS, Fig.
 `r fig_NEON_vertical`g, Supporting Information Figure `r SIfig_NEON_all`b; <!--@hadleyInfluenceKrummholzMat1987--> @banerjeeExplainingConvectorEffect2017; @curtis_intracanopy_2019;<!--@zweifelMiddayStomatalClosure2002; @ramboCanopyMicroclimateResponse2009; @hardwickRelationshipLeafArea2015--> @mullerEvidenceEfficientNonevaporative2021].
-These usually-modest gradients in $T_{air}$ combine with the above-described gradients in micrometeorological conditions to shape $T_{leaf}$ patterns within canopies.
+These usually-modest gradients in *T~air~* combine with the above-described gradients in micrometeorological conditions to shape *T~leaf~* patterns within canopies.
 
@@
 
-$T_{leaf}$ is strongly tied to, and usually within a few degrees of, $T_{air}$ [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Note `r Notes_leafTbiophysics`,@doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009].
-Deviation of $T_{leaf}$ from $T_{air}$ is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance ($g_s$; Fig.
+*T~leaf~* is strongly tied to, and usually within a few degrees of, *T~air~* [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Note `r Notes_leafTbiophysics`,@doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009].
+Deviation of *T~leaf~* from *T~air~* is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (*g~s~*; Fig.
 `r fig_leaf_T`; Note `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig.
 `r fig_schematic`).
-High radiation in the upper canopy implies that upper canopy leaves often have higher $T_{leaf}-T_{air}$, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`a).
-However, higher wind speeds reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
-In addition, adaptive leaf traits that increase $g_s$ and $g_b$, such as small size and high stomatal density (Note `r Notes_leafTbiophysics`; sections 3, 4.1), mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
+High radiation in the upper canopy implies that upper canopy leaves often have higher *T~leaf~* - *T~air~*, and thus are often warmer than understory leaves (Fig. `r fig_leaf_T`a).
+However, higher wind speeds reduce *T~leaf~* - *T~air~* [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
+In addition, adaptive leaf traits that increase *g~s~* and *g~b~*, such as small size and high stomatal density (Note `r Notes_leafTbiophysics`; sections 3, 4.1), mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
-In contrast, in the lower canopy, greater RH (reduced VPD) and lower wind speeds (reduced $g_b$) may limit leaf cooling via latent heat flux, increasing $T_{leaf}$ [@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
-Leaves in these normally-shaded locations can experience rapid light and temperature surges during sunflecks [>10$^\circ$C, @waySunflecksTreesForests2012].
+In contrast, in the lower canopy, greater RH (reduced VPD) and lower wind speeds (reduced *g~b~*) may limit leaf cooling via latent heat flux, increasing *T~leaf~* [@perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979].
+Leaves in these normally-shaded locations can experience rapid light and temperature surges during sunflecks [>10°C, @waySunflecksTreesForests2012].
 
 ![**Figure `r fig_leaf_T`.
-Theoretical expectations for variation in the difference between leaf and air temperatures, $T_{Leaf}-T_{air}$, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf characteristic dimension, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under hypothetical moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
+Theoretical expectations for variation in the difference between leaf and air temperatures, *T~leaf~* - *T~air~*, in response to (a) shortwave radiation, (b) wind speed, (c) relative humidity, (d) leaf characteristic dimension, and (e) stomatal conductance.** Leaf temperatures were modeled using the  *tealeaves* R package of @muir_tealeaves_2019 parameterized to represent a broadleaf species (*Quercus rubra* L.) in a mesic temperate forest (Harvard Forest, Massachusetts, USA) under hypothetical moist and drought conditions, as detailed in Supplementary Information Methods `r appendix_tealeaves_methods`.
 In each scenario, the independent variable was allowed to vary while other parameters were held constant at the values given in the table of biophysical constants.
-Dots along each line correspond to the biophysical constant assigned for the independent variable in the other scenarios, and therefore represent the modeled value for $T_{Leaf}-T_{air}$ with all parameters set to the biophysical constants given in the table.
+Dots along each line correspond to the biophysical constant assigned for the independent variable in the other scenarios, and therefore represent the modeled value for *T~leaf~* - *T~air~* with all parameters set to the biophysical constants given in the table.
 Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair).
@@
 
-Elevation of $T_{leaf}$ above $T_{air}$ varies with micrometeorological conditions. 
-During drought (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, $T_{leaf} - T_{air}$ will tend to be higher in the upper canopy compared to the understory (Fig.
+Elevation of *T~leaf~* above *T~air~* varies with micrometeorological conditions. 
+During drought (hot and dry, with higher-than-average solar radiation), when *g~s~* is limiting, *T~leaf~* - *T~air~* will tend to be higher in the upper canopy compared to the understory (Fig.
 `r fig_leaf_T`).
-This is because sun leaves are exposed to higher irradiance and VPD, and therefore have stronger $g_s$ limitation than shade leaves [Fig.
+This is because sun leaves are exposed to higher irradiance and VPD, and therefore have stronger *g~s~* limitation than shade leaves [Fig.
 `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018].
-In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig.
+In contrast, under conditions conducive to stomatal opening, higher wind speeds and *g~b~* enable cooling in the upper canopy, whereas lower wind and *g~b~* in the understory limit evaporative cooling  [Fig.
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017 <!-- @songTreeSurfaceTemperature2020-->].
 Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves. 
-Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal $T_{leaf}$ under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
-In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also inhibits evaporative cooling, thus increasing $T_{leaf}-T_{air}$ [Fig.
+Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal *T~leaf~* under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
+In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also inhibits evaporative cooling, thus increasing *T~leaf~* - *T~air~* [Fig.
 `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].
 
-Vertical $T_{leaf}$ gradients also vary with canopy structure (Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`). 
+Vertical *T~leaf~* gradients also vary with canopy structure (Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`). 
 Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most incoming radiation and preventing vertical air mixing in the understory.
-Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than lower-canopy leaves, in some cases exceeding photosynthetic temperature optima ($T_{opt}$) because of $g_s$ limitation and high radiation [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999;-->@doughtyAreTropicalForests2008; <!--@fauset_differences_2018;--> @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021<!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
-In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory, which can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [<!--@hadleyInfluenceKrummholzMat1987--> @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
-The latter can result from a combination of still air at lower heights, sunflecks or the 'canopy greenhouse effect' (section 1.5), and the tendency for shade leaves to have lower $g_s$ [section 3, @schymanskiStomatalControlLeaf2013; @hardwickRelationshipLeafArea2015].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing $T_{leaf}$ near the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
-One of the few studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense subalpine stand in the alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
+Therefore, in these forests, leaves in the upper canopy can experience greater *T~leaf~* - *T~air~* and higher maximum *T~leaf~* than lower-canopy leaves, in some cases exceeding photosynthetic temperature optima (*T~opt~*) because of *g~s~* limitation and high radiation [@carterExperimentalWarmingTropical2021; <!--@niinemetsShapeLeafPhotosynthetic1999;-->@doughtyAreTropicalForests2008; <!--@fauset_differences_2018;--> @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021<!--@rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018-->].
+In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory, which can either neutralize a *T~leaf~* gradient or elevate *T~leaf~* - *T~air~* at lower heights relative to the upper canopy [<!--@hadleyInfluenceKrummholzMat1987--> @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002].
+The latter can result from a combination of still air at lower heights, sunflecks or the 'canopy greenhouse effect' (section 1.5), and the tendency for shade leaves to have lower *g~s~* [section 3, @schymanskiStomatalControlLeaf2013; @hardwickRelationshipLeafArea2015].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing *T~leaf~* near the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
+One of the few studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5°C cooler temperatures lower in the vertical profile of a dense subalpine stand in the alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
 However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVerticalinpress].
@@
 More recent work has extended the focus from light alone to height in the vertical profile, as the latter shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
-For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
+For example, leaf mass per area (LMA), photosynthetic capacity (*A~max~* per area) and *g~s~* have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
 Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
@@
 
-Sun leaves have anatomical, morphological, and physiological traits that reduce $T_{leaf}$ through higher conductance ($g_b$ or maximal $g_s$) and/or reflectance (Table `r table_leaf_traits`). 
+Sun leaves have anatomical, morphological, and physiological traits that reduce *T~leaf~* through higher conductance (*g~b~* or maximal *g~s~*) and/or reflectance (Table `r table_leaf_traits`). 
 Thus, sun leaves are generally smaller and thicker, with higher LMA, and are more deeply lobed [Fig.
-`r fig_leaf_T`, @vogelSunLeavesShade1968; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces $g_b$).
-Steeper leaf angles reduce mid-day radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, also decreasing radiation load.
-Further, sun leaves tend to have higher stomatal and vein densities, which fascilitate evaporative cooling by enabling higher $g_s$ [section 4.1, @zwienieckiHydraulicLimitationsImposed2004]. 
-Yet sun leaves also have adaptations to limit water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [which may reduce or increase $g_b$, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
+`r fig_leaf_T`, @vogelSunLeavesShade1968; @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @leighInfluenceLeafSize2017], but with greater leaf packing and clumping (which reduces *g~b~*).
+Steeper leaf angles reduce mid-day radiation loads and thereby decrease *T~leaf~* - *T~air~* [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, also decreasing radiation load.
+Further, sun leaves tend to have higher stomatal and vein densities, which fascilitate evaporative cooling by enabling higher *g~s~* [section 4.1, @zwienieckiHydraulicLimitationsImposed2004]. 
+Yet sun leaves also have adaptations to limit water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [which may reduce or increase *g~b~*, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
 In contrast, shade leaves have traits that maximize light capture (e.g., lower LMA and greater light absorptance per unit biomass), but larger leaves and lower transpiration make them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013].
-In open canopies, where light is comparatively homogeneous, leaf traits may be shaped more by maximum $T_{air}$ and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
+In open canopies, where light is comparatively homogeneous, leaf traits may be shaped more by maximum *T~air~* and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].
 
@@
 Leaf optical properties are influenced by anatomical, morphological, and biochemical traits that vary throughout the canopy (sections 3.1, 3.3).
-High in the canopy, high light absorptance can lead to photoinhibition and would be in part be mitigated by $T_{leaf}$ regulation (sections 3.4, 3.5; Table `r table_leaf_traits`). 
+High in the canopy, high light absorptance can lead to photoinhibition and would be in part be mitigated by *T~leaf~* regulation (sections 3.4, 3.5; Table `r table_leaf_traits`). 
 For leaves in higher light environments, light absorptance and thus heat load can be reduced by leaf surface modifications (e.g., trichomes, cuticle) that increase reflectance, and heat may be dissipated biochemically  by carotenoids, including xanthophylls [Table `r table_leaf_traits`, @leeCorrelatesLeafOptical1990; @knappVariabilityLeafOptical1998].
@@
 Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (section 4).
-Sun leaves have higher area-based concentrations of elements like nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical for respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthesis [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
+Sun leaves have higher area-based concentrations of elements like nitrogen (*N~area~*) and phosphorus (*P~area~*) that are critical for respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthesis [@scartazzaInvestigatingEuropeanBeech2016a; @niinemetsPhotosyntheticAcclimationSimultaneous2004; @weerasingheCanopyPositionAffects2014].
 Increases in chlorophyll a/b ratios with height reflect greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights increase PAR absorptance efficiency of shade leaves (Table `r table_leaf_traits`).
@@
 
-Biochemical protection against light and heat damage increases with peak radiation loads and thus tends to be higher in the upper canopy, which is subject to higher $T_{leaf}$ and hydraulic limitations, than in the understory (Table `r table_leaf_traits`).
+Biochemical protection against light and heat damage increases with peak radiation loads and thus tends to be higher in the upper canopy, which is subject to higher *T~leaf~* and hydraulic limitations, than in the understory (Table `r table_leaf_traits`).
 More frequent stomatal closure in upper canopy leaves (section 4) reduces their capacity to use light energy for photochemistry, thereby requiring a high capacity to dissipate excess light energy and protect against photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007].
@@
 Leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
-In contrast, to maximize photosynthesis and minimize damage during $T_{leaf}$ surges caused by sunflecks (section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
+In contrast, to maximize photosynthesis and minimize damage during *T~leaf~* surges caused by sunflecks (section 2), shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light than sun-acclimated upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].
 
@@
 Temperature-sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [section 1, @sharkeyIsopreneEmissionPlants2008].
-Future work to understand how temporal dynamics of $T_{leaf}$ vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
+Future work to understand how temporal dynamics of *T~leaf~* vary with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient.
 Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play similar roles in upper-canopy leaves.
@@
 <!--short par giving background on thermal tolerance-->
-Very high $T_{leaf}$ (~40-60$^{\circ}$C) irreversibly damages photosystem II, leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
+Very high *T~leaf~* (~40-60°C) irreversibly damages photosystem II, leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress].
 Upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II functioning via using chlorophyll fluorescence techniques [@krauseHightemperatureToleranceTropical2010]. 
-Thermal tolerance is described in terms of the $T_{leaf}s$ at which photosystem II efficiency starts to decrease ($T_{crit}$) is reduced by 50%, $T_{50}$ [e.g., @slotLeafHeatTolerance2021]. 
-Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and being greater in leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
-$T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
+Thermal tolerance is described in terms of the *T~leaf~*s at which photosystem II efficiency starts to decrease (*T~crit~*) is reduced by 50%, *T~50~* [e.g., @slotLeafHeatTolerance2021]. 
+Thermal tolerances vary across species, being more closely linked to leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]<!--; for example, among tropical species, *T~crit~* and *T~50~* were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]-->, and being greater in leaves that experience higher maximum temperatures [@perezPhotosyntheticHeatTolerances2020].
+*T~50~* and *T~crit~* decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
 However, they appear more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
 <!--For example, *Quercus muehlenbergii* Engelm.
-growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* Michx.
+growing in xeric, sunnier conditions showed higher *T~50~* by 2°C than *Quercus macrocarpa* Michx.
 growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a].
@@
 Thermal tolerance is hypothesized to be greatest for the leaves with the greatest radiation loads and highest temperatures along the vertical gradient. 
-Indeed, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama, $T_{50}$ was 0.7 - 1.4$^\circ$C lower for shade than sun leaves for two of three tree species [Table `r table_leaf_traits`, @slotPhotosyntheticHeatTolerance2019].
-Similarly, within a very open canopy in *Acacia papyrocarpa* canopy, $T_{50}$ was highest (*albiet* by <1$^\circ$C) in the lower, north-facing leaves, which faced the greatest composite climate stress (low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation) [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
-These modest differences in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments imply that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
+Indeed, considering sun and shade leaf differences across the vertical profile of a closed-canopy forest in Panama, *T~50~* was 0.7 - 1.4°C lower for shade than sun leaves for two of three tree species [Table `r table_leaf_traits`, @slotPhotosyntheticHeatTolerance2019].
+Similarly, within a very open canopy in *Acacia papyrocarpa* canopy, *T~50~* was highest (*albiet* by <1°C) in the lower, north-facing leaves, which faced the greatest composite climate stress (low wind speed, greater radiation and *T~air~*, and lower rates of heat dissipation) [Table `r table_leaf_traits`, @curtis_intracanopy_2019].
+These modest differences in *T~50~* coupled with larger differences in *T~leaf~* across thermal microenvironments imply that more thermally tolerant sun leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].
 However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences.
@@
 
-Leaf phenology influences the vertical structure of microenvironments, $T_{leaf}$, and forest photosynthetic capacity [@rey-sanchez_spatial_2016]. 
+Leaf phenology influences the vertical structure of microenvironments, *T~leaf~*, and forest photosynthetic capacity [@rey-sanchez_spatial_2016]. 
 In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
@@
 Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct vertical patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019;@tangLightdrivenGrowthAmazon2017;@nunesForestFragmentationImpacts2022].
-For instance, in a central Amazonian forest, the upper canopies of both interior and edge forests shed leaves when maximum daily $T_{air}$ exceeded ~35$^\circ$C [@nunesForestFragmentationImpacts2022].
+For instance, in a central Amazonian forest, the upper canopies of both interior and edge forests shed leaves when maximum daily *T~air~* exceeded ~35°C [@nunesForestFragmentationImpacts2022].
 In evergreen forests, leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
@@
 
-Leaf gas exchange is strongly shaped by environmental drivers, $T_{leaf}$, and traits, all of which vary across vertical forest gradients (sections 1-3), Fig. `r fig_schematic`). 
+Leaf gas exchange is strongly shaped by environmental drivers, *T~leaf~*, and traits, all of which vary across vertical forest gradients (sections 1-3), Fig. `r fig_schematic`). 
 However, as we detail below, few studies have evaluated how temperature responses of leaf metabolic rates are shaped by the varying environmental conditions across vertical forest gradients (Table `r table_leaf_metabolism`).
@@
 
-Leaf hydraulic, stomatal and boundary layer conductances are all critical for regulating $T_{leaf}$ via latent heat loss (Fig. `r fig_leaf_T`). 
-Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
-Maximum $g_s$ increases with light -- typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979] -- and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
-However, water supply cannot meet the demands caused by the high radiation and $g_b$ in sun leaves with fully open stomata, in part because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014].
+Leaf hydraulic, stomatal and boundary layer conductances are all critical for regulating *T~leaf~* via latent heat loss (Fig. `r fig_leaf_T`). 
+Due to higher wind, lower RH, and smaller leaf sizes in the upper canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), *g~b~* increases with height (Table `r table_leaf_metabolism`).
+Maximum *g~s~* increases with light -- typically tracking photosynthetic capacity [@wongStomatalConductanceCorrelates1979] -- and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`).
+However, water supply cannot meet the demands caused by the high radiation and *g~b~* in sun leaves with fully open stomata, in part because of height-related constraints on water transport [e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010] and/or increased leaf-air VPD caused by leaf warming in sunlit canopy locations [@buckleyStomatalConductanceOptimized2014].
 These constraints tend to reduce leaf water potential, making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and $\delta^{13}C$ (section 3.3, Table `r table_leaf_traits`).
@@
 `r fig_leaf_T`e), thus amplifying the warming of sun leaves by high radiation [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006].
-The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven by increased VPD) than lower-canopy shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
-This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
+The temperature at which *g~s~* is greatest, *T~opt~* of *g~s~*, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in *g~s~* in response to rising *T~air~* (driven by increased VPD) than lower-canopy shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
+This, added to the tendency for sun leaves to have higher *T~leaf~*, implies that high *T~air~* should decrease *g~s~* of upper canopy leaves more than understory leaves, particularly when water availability is limited.
 
@@
 Temperature can affect photosynthesis directly, by altering photosynthetic enzyme activity and the electron transport chain,  and indirectly, by increasing VPD and closing stomata [@lloydEffectsRisingTemperatures2008].
-Photosynthesis has a peaked response to $T_{air}$, with the optimum commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
+Photosynthesis has a peaked response to *T~air~*, with the optimum commonly corresponding to the prevalent ambient growing season temperature [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017; @doughtyAreTropicalForests2008].
 Beyond the optimum, photosynthesis decreases due to stomatal closure [ @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020; @grossiordPlantResponsesRising2020 <!--@fredeenTemperatureHumidityEffects1999-->] and biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019; <!--@sageTemperatureResponseC32007;--> @varhammarPhotosyntheticTemperatureResponses2015; @sharkeyEffectsModerateHeat2005].
@@
 We have very little evidence about how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, which  compare sun and shade leaves of the same trees, reveal no pronounced overall trend with height in photosynthetic temperature optima  (Table `r table_leaf_metabolism`).
-Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g.
-@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019], one might expect sun leaves to have a stronger temperature-dependence and higher $T_{opt}$ than shade leaves [@campbell_introduction_1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
-However, such a trend is not apparent among recent studies, where for three species in Panama, $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
-Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{cmax}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], nor within-canopies of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
-In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the top of the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].
-The observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis [@hernandezSimilarTemperatureDependence2020]. 
+Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis (*T~opt~*) across sites and seasons [e.g.
+@tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019], one might expect sun leaves to have a stronger temperature-dependence and higher *T~opt~* than shade leaves [@campbell_introduction_1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004].
+However, such a trend is not apparent among recent studies, where for three species in Panama, *T~opt~* for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020].
+Similarly, *T~opt~* of RuBP carboxylation  (*V~cmax~*) and regeneration rates (*J~max~*) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], nor within-canopies of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018].
+In contrast, tropical experimental warming observations showed that *T~opt~*, and *T~opt~* of *J~max~* decrease from the understory to the top of the canopy, potentially linked to greater thermal sensitivity of *g~s~* in upper canopy leaves [@carterExperimentalWarmingTropical2021].
+The observed lack of sun-shade differences in *T~opt~* may reflect acclimation to similar temperatures of peak photosynthesis [@hernandezSimilarTemperatureDependence2020]. 
 Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of midday stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous.
@@
 
-Similar to photosynthesis, respiration tends to be higher in upper-canopy sun leaves, but its temperature sensitivity ($Q_{10}$ or $E_0$) shows no definite trend along the vertical gradient.
+Similar to photosynthesis, respiration tends to be higher in upper-canopy sun leaves, but its temperature sensitivity (*Q~10~* or *E~0~*) shows no definite trend along the vertical gradient.
 Specifically, the temperature sensitivity of respiration can be constant within vertical profiles and in seedling sun vs.
 shade leaves<!--[<!--@carterExperimentalWarmingTropical2021-> @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007 <!--@zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006->]-->, greater in upper-canopy leaves<!--[@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]-->, or greater in the lower canopy <!--[@griffinCanopyPositionAffects2002]--> (Table `r table_leaf_metabolism`).
-This variation may be attributable to forest type, leaf traits and age [e.g., greater $Q_{10}$ in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases $Q_{10}$ [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
+This variation may be attributable to forest type, leaf traits and age [e.g., greater *Q~10~* in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases *Q~10~* [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
 Overall, however, we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
@@
 Therefore, periods of high VPD, whether on time scales of hours, days, or seasons, tend to reduce transpiration and photosynthesis more in tall trees that occupy canopy positions in closed-canopy forests [@christoffersenLinkingHydraulicTraits2016; @garciaImportanceHydraulicStrategy2021].
-This is consistent with observations that both dry season leaf loss (section 3.6, Table `r table_leaf_traits`) and $g_s$ limitation (section 4.1, Table `r table_leaf_metabolism`) increase with height across the vertical profile.
+This is consistent with observations that both dry season leaf loss (section 3.6, Table `r table_leaf_traits`) and *g~s~* limitation (section 4.1, Table `r table_leaf_metabolism`) increase with height across the vertical profile.
 More active regulation of transpiration by tall canopy trees [e.g., @mediavillaStomatalResponsesDrought2004] -- often combined with greater effective rooting depth -- may offset greater xylem embolism risk [@olsonPlantHeightHydraulic2018; @garciaImportanceHydraulicStrategy2021; @chitra-tarakHydraulicallyvulnerableTreesSurvive2021] and reduce the need to adjust carbon metabolism traits during drought [@bartholomewSmallTropicalForest2020].
-This reduces carbon isotope discrimination, indicative of greater $g_s$ limitation, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
+This reduces carbon isotope discrimination, indicative of greater *g~s~* limitation, with increasing tree height [Table `r table_leaf_traits`, @mcdowellRelationshipTreeHeight2011].
 
@@
 
-Although drought sensitivity clearly increases with crown height and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies along the vertical gradient.
-High $T_{air}$ is often associated with high VPD and atmospheric drought, likely explaining negative growth responses to $T_{air}$ -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.
+Although drought sensitivity clearly increases with crown height and hence with DBH, it remains unclear how growth sensitivity to high *T~air~*, independent of drought, varies along the vertical gradient.
+High *T~air~* is often associated with high VPD and atmospheric drought, likely explaining negative growth responses to *T~air~* -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig.
 `r fig_treerings`, @anderson-teixeiraJointEffectsClimate2022].
-However, under well-watered conditions conducive to high $g_s$, we do not necessarily expect higher thermal sensitivity of photosynthesis (section 4.2) or woody growth in upper-canopy trees than their understory counterparts.
-Indeed, there are cases where tree-ring records show more negative or less positive $T_{air}$-responses of understory trees compared to canopy dominants [@anderson-teixeiraJointEffectsClimate2022; @rollinsonClimateSensitivityUnderstory2020].
+However, under well-watered conditions conducive to high *g~s~*, we do not necessarily expect higher thermal sensitivity of photosynthesis (section 4.2) or woody growth in upper-canopy trees than their understory counterparts.
+Indeed, there are cases where tree-ring records show more negative or less positive *T~air~*-responses of understory trees compared to canopy dominants [@anderson-teixeiraJointEffectsClimate2022; @rollinsonClimateSensitivityUnderstory2020].
 Additional research is required to understand the mechanisms underlying these intriguing differences, and to disentangle size-related tree growth responses to hot-wet versus hot-dry conditions.
@@
 ![**Figure `r fig_treerings`.
-Examples of tree-ring evidence for greater interannual temperature sensitivity of annual growth for large than for small trees.** Across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high maximum temperature ($T_{max}$) or potential evapotranspiration (PET) than did small trees of the same species.
-Independent variables are the most influential temperature-related variable at the site ($T_{max}$ or PET over a seasonal window spanning from the noted beginning to end month, where "c" and "p" represent months in the current and previous calendar year, respectively).
+Examples of tree-ring evidence for greater interannual temperature sensitivity of annual growth for large than for small trees.** Across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high maximum temperature (*T~max~*) or potential evapotranspiration (PET) than did small trees of the same species.
+Independent variables are the most influential temperature-related variable at the site (*T~max~* or PET over a seasonal window spanning from the noted beginning to end month, where "c" and "p" represent months in the current and previous calendar year, respectively).
 Colored lines represent responses  of trees at the maximum and minimum tails of the DBH distribution included in the analysis to the temperature variable in a model including a DBH - temperature interaction.
@@
 Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed ~93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018].
-In terms of C cycling, it has been estimated that the canopy strata contributes $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007].
-Large trees also dominate woody aboveground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig.
+In terms of C cycling, it has been estimated that the canopy strata contributes ≥64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007].
+Large trees also dominate woody aboveground net primary productivity (*ANPP~woody~*) and mortality (*M~woody~*), with trees ≥10cm DBH usually contributing > 85% of *ANPP~woody~* and *M~woody~* across 25 globally distributed forests [Fig.
 `r fig_ecosystem`b, @piponiotDistributionBiomassDynamicsinpress].
@@
 It is less clear how thermal sensitivity of water and carbon fluxes vary across strata, but probable responses can be inferred based on the patterns and mechanisms reviewed above.
-We expect that under conditions of moderate to high VPD, transpiration, GPP, and $ANPP_{woody}$ should all be more sensitive to high temperatures in the upper canopy than in the understory [@grossiordPlantResponsesRising2020; @nunesForestFragmentationImpacts2022].
+We expect that under conditions of moderate to high VPD, transpiration, GPP, and *ANPP~woody~* should all be more sensitive to high temperatures in the upper canopy than in the understory [@grossiordPlantResponsesRising2020; @nunesForestFragmentationImpacts2022].
 Because canopy trees dominate these fluxes (Fig. `r fig_ecosystem`), their responses will strongly influence the whole-ecosystem response, potentially with modest buffering by the understory. 
@@
 Yet evidence also suggests GPP and ecosystem respiration are less sensitive to heat and drought stress in older forests than in young forests [@xuSeasonalVariabilityForest2020], perhaps in part because of more complex vertical structuring [@jonesManagingRedPine2019;@nunesForestFragmentationImpacts2022].
-A rare example of a study comparing climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that $M_{woody}$ was more strongly elevated among large than small trees during an *El Niño* drought in Panama, while the smallest trees had higher productivity during the drought, likely because of increased light in the understory.
+A rare example of a study comparing climate sensitivity of C fluxes across size classes [@meakemRoleTreeSize2018] showed that *M~woody~* was more strongly elevated among large than small trees during an *El Niño* drought in Panama, while the smallest trees had higher productivity during the drought, likely because of increased light in the understory.
 It remains far less clear how thermal sensitivity varies across forest strata under wet conditions, but both physiological mechanisms (Fig.
@@
 
-As climate change progresses, we anticipate increases in both mean daytime and nighttime $T_{air}$, as well as increased maximum temperatures, with extreme increases during heat waves [@ipccClimateChange20212021].
+As climate change progresses, we anticipate increases in both mean daytime and nighttime *T~air~*, as well as increased maximum temperatures, with extreme increases during heat waves [@ipccClimateChange20212021].
 To the extent that warming is coupled to drought, we expect that tall trees with exposed crowns will usually be hardest-hit, particularly in severe drought, partly because their crowns are positioned in a more challenging microenvironment (section 5.1, Figs.
@@
 Leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], yet failure to fully acclimate will reduce carbon sequestration at leaf, tree, and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
-As growing season $T_{air}$ increasingly equals or exceeds photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], exposed crowns are likely most vulnerable.
-Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and $T_{crit}$ [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
-In contrast, understory leaves and trees are unlikely to face the same absolute extremes of $T_{leaf}$ (unless exposed to sunflecks), but in some ways are likely to prove more sensitive to higher $T_{air}$.
+As growing season *T~air~* increasingly equals or exceeds photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], exposed crowns are likely most vulnerable.
+Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and *T~crit~* [@osullivanThermalLimitsLeaf2017], *T~leaf~* of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
+In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~* (unless exposed to sunflecks), but in some ways are likely to prove more sensitive to higher *T~air~*.
 Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
-However, with $T_{air}$ more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
-<!--Moreover, increased nighttime $T_{air}$ may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
+However, with *T~air~* more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
+<!--Moreover, increased nighttime *T~air~* may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
 While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [@rollinsonClimateSensitivityUnderstory2020].
-Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming $T_{air}$, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].
+Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming *T~air~*, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].
 
@@
 Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer.
-Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table `r table_leaf_metabolism`, e.g., @krinnerDynamicGlobalVegetation2005].
+Thus, even in single-layer models, key physiological parameters like *V~cmax~* and *J~max~* vary, decreasing with increasing cumulative LAI or lower light conditions [Table `r table_leaf_metabolism`, e.g., @krinnerDynamicGlobalVegetation2005].
 Recent years have seen growing efforts to incorporate vertical variation more directly in models, specifically involving leaf water potential and light absorption, along with a increasing interest in confronting models directly with field measurements [@fisherPerspectivesFutureLand2020].
-Recent model developments at the cutting edge of representing vertical variation [@chenColumnCanopyAirTurbulent2019;@bonanModelingCanopyinducedTurbulence2018; @longoBiophysicsEcologyBiogeochemistry2019] implement vertical gradients of irradiance, water content, $T_{leaf}$, and humidification of canopy air by transpiration, modulated by turbulence within the forest and a roughness layer that extends to roughly twice the canopy height [@fisherPerspectivesFutureLand2020].
+Recent model developments at the cutting edge of representing vertical variation [@chenColumnCanopyAirTurbulent2019;@bonanModelingCanopyinducedTurbulence2018; @longoBiophysicsEcologyBiogeochemistry2019] implement vertical gradients of irradiance, water content, *T~leaf~*, and humidification of canopy air by transpiration, modulated by turbulence within the forest and a roughness layer that extends to roughly twice the canopy height [@fisherPerspectivesFutureLand2020].
 
@@
 Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although few applications exist at finer-scales needed to understand vertical variation in canopy temperatures [@johnstonWhatLiesVerticalinpress].
-As an example of the type of insight possible with this approach, @pauTropicalForestTemperature2018 used data from a tower-based infrared camera in combination with eddy-covariance data and found that tropical forest GPP was more strongly associated with canopy temperature than $T_{air}$ or VPD.
+As an example of the type of insight possible with this approach, @pauTropicalForestTemperature2018 used data from a tower-based infrared camera in combination with eddy-covariance data and found that tropical forest GPP was more strongly associated with canopy temperature than *T~air~* or VPD.
 In a a western-U.S.

### 2022-06-09 response-to-reviews.Rmd
Update response-to-reviews.Rmd
work on R2 responses, issue #120
#120

@@
 
+**We have modified these statements as follows:**
+
+**l.33 - ...**
+
+**l.35 - "In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be strongly impacted by warming under some conditions, particularly when exposed to a harsher microenvironment through canopy disturbance."**
+
+**l.504 - ...**
+
+**l.648 - ...**
+
+**l.814 - ...**
+
+**l.838 - ...**
+
 In fact, the abstract reads like the authors cannot decide whether tall trees '...
@@
 This is an element in the review as well (see below).
+
+**We modified this sentence to read, "In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be strongly impacted by warming under some conditions, particularly when exposed to a harsher microenvironment through canopy disturbance."**
 
@@
 
+*response required*
+
 l.32 tall trees have higher absolute metabolic capacity and growth at both individual *and ecosystem levels* (please remove the text between the asterices).
+
+**Done.**
 
@@
 This is a major aspect for the authors to clarify.
+
+**We have greatly de-emphasized the discussion of the potential for stronger negative effects of rising temperatures on understory trees, including the following:**
+
+**(1) removed Fig. 4b [from @rollinsonClimateSensitivityUnderstory2020];**
+
+**(2) modified the abstract sentence to read, "In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be strongly impacted by warming under some conditions, particularly when exposed to a harsher microenvironment through canopy disturbance."**
+
+**(3) changed statement summarizing @rollinsonClimateSensitivityUnderstory2020 to a more generic statement: "Indeed, there are cases where tree-ring records show more negative or less positive $T_{air}$-responses of understory trees compared to canopy dominants [@anderson-teixeiraJointEffectsClimate2022; @rollinsonClimateSensitivityUnderstory2020]."**
+
+**(4) changed the statement, "...and tree ring evidence (Fig.
+4) suggest that understory trees may respond more negatively to hot, wet conditions." to read "...and some tree ring evidence [@anderson-teixeiraJointEffectsClimate2022; @rollinsonClimateSensitivityUnderstory2020] raise the possibility that understory trees may be at a relative disadvantage under hot, humid conditions."**
+
+*...*
+
+*https://github.com/EcoClimLab/vertical-thermal-review/issues/120*
 

### 2022-06-09 vertical-thermal-review.Rmd
work on R2 responses, issue #120
#120

@@
 Scaling from leaves to trees, tall trees have higher absolute metabolic capacity and growth, yet are disproportionately vulnerable to drought and damaging *T~leaf~*, particularly under climate change.
-In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be disproportionately impacted under hot, humid conditions, or when exposed to harsher conditions through canopy disturbance.
+In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be strongly impacted by warming under some conditions, particularly when exposed to a harsher microenvironment through canopy disturbance.
 As the climate changes, integrating the patterns and mechanisms reviewed here into models will be critical to forecasting forest-climate feedbacks.

### 2022-06-09 vertical-thermal-review.Rmd
R2 wording changes

@@
 ### 1.4 Atmospheric concentrations of CO~2~ and water vapor
-Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day such that differences are negligible for most of the day when photosynthesis is occurring [@koike_leaf_2001; @yang_spatial_1999; @brooksVerticalGradientsPhotosynthetic1997].
+Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day such that CO~2~ concentration differences are negligible for most of the day when photosynthesis is occurring [@koike_leaf_2001; @yang_spatial_1999; @brooksVerticalGradientsPhotosynthetic1997].
 While understory seedlings may benefit modestly from higher CO~2~ concentrations near the ground during some parts of the day [e.g., 6% of C fixation from recently respired CO~2~, @brooksVerticalGradientsPhotosynthetic1997], gradients in CO~2~ concentrations concentrations likely have little effect on energy balance and metabolism of trees across the vertical gradient [@brooksVerticalGradientsPhotosynthetic1997].
@@
 Especially when canopy height and leaf area index (LAI) are large, maximum *T~air~* and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
-The implication is that upper canopy leaves have higher *g~b~* and associated potential for water and heat loss (sections 2, 4.1).
+The implication is that upper canopy leaves have higher *g~b~* and associated potential for water and heat loss compared to understory leaves (sections 2, 4.1).
 
@@
 Elevation of *T~leaf~* above *T~air~* varies with micrometeorological conditions. 
-During drought (hot and dry, with higher-than-average solar radiation), when *g~s~* is limiting, *T~leaf~* - *T~air~* will tend to be higher in the upper canopy compared to the understory (Fig.
+During drought (associated with conditions that are hot and dry, with higher-than-average solar radiation), when *g~s~* is limiting, *T~leaf~* - *T~air~* will tend to be higher in the upper canopy compared to the understory (Fig.
 `r fig_leaf_T`).

### 2022-06-09 vertical-thermal-review.Rmd
misc

@@
 
-1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA
+1. Conservation Ecology Center; Smithsonian National Zoo & Conservation Biology Institute; Front Royal, VA 22630, USA
 
@@
 Thanks to Valentine Herrmann, Norbert Kunert, Camille Piponiot, Peter B. Boucher and Andrew B. Davies for providing figure materials. 
-This manuscript benefited from feedback from Eleinis Ávila-Lovera and the ForestGEO Ecosystems & Climate lab at SCBI.
+This manuscript benefited from feedback from Eleinis Ávila-Lovera, the ForestGEO Ecosystems & Climate lab at NZCBI, and two anonymous reviewers.
 Funding was provided by the Smithsonian Institution.

### 2022-06-09 response-to-reviews.Rmd
R2 wording changes
Update response-to-reviews.Rmd
closes #119

@@
 
-*https://github.com/EcoClimLab/vertical-thermal-review/issues/119*
+**We have fixed the axes labels to list units as °C and updated the caption to define PET and better explain the x-axis labels.**
 
@@
 There are still a number of places where the authors could improve and one cannot resist letting these past at this point.
+
+**Thank you for your additional comments.**
 
@@
 
+**Revised as suggested.**
+
 l.200 'upper canopy leaves have higher 𝑔𝑏 and associated potential for water and heat loss' compared to what??? Please clarify.
+
+**Done. The sentence now reads: "The implication is that upper canopy leaves have higher *g~b~* and associated potential for water and heat loss compared to understory leaves (sections 2, 4.1)."**
 
@@
 I suggest revising as 'associated with conditions that are hot ...'.
+
+**Revised as suggested.**
 

### 2022-06-09 response-to-reviews.Rmd
edit discussion of small trees
#120

@@
 
-*...*
+**(5) revised discussion in the "Warming" section to read: "In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~*, and protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
+However, with *T~air~* more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous.
+While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can dramatically reduce the growth of understory trees [@rollinsonClimateSensitivityUnderstory2020], and these small trees may be at higher risk of carbon starvation because they have lower NSC reserves [@niinemetsResponsesForestTrees2010].
+Thus, despite their buffered microclimate, it is possible that understory trees could be more adversely impacted by warming temperatures than their canopy counterparts under some conditions."**
 
-*https://github.com/EcoClimLab/vertical-thermal-review/issues/120*
+**(6) in the conclusions section, removed the statement  "...[understory trees] might be more vulnerable to chronic warming stress under relatively mesic conditions [@rollinsonClimateSensitivityUnderstory2020]"**
 
@@
 
+**We have revised the second half of the paragraph containing this statement as follows: "In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~*, and protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
+However, with *T~air~* more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous.
+While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can dramatically reduce the growth of understory trees [@rollinsonClimateSensitivityUnderstory2020], and these small trees may be at higher risk of carbon starvation because they have lower NSC reserves [@niinemetsResponsesForestTrees2010].
+Thus, despite their buffered microclimate, it is possible that understory trees could be more adversely impacted by warming temperatures than their canopy counterparts under some conditions."**
+
+**We have modified the abstract sentence to read, "In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be strongly impacted by warming under some conditions, particularly when exposed to a harsher microenvironment through canopy disturbance."**
+
 \newpage

### 2022-06-09 vertical-thermal-review.Rmd
update Fig. 4
#119
edit discussion of small trees
#120

@@
 
-1. Conservation Ecology Center; Smithsonian National Zoo & Conservation Biology Institute; Front Royal, VA 22630, USA
+1. Conservation Ecology Center; Smithsonian's National Zoo & Conservation Biology Institute; Front Royal, VA 22630, USA
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-From @anderson-teixeiraJointEffectsClimate2022.](tree_rings/pre_temp_groups_dbh_interactions_for_NIDHI.png)
+From @anderson-teixeiraJointEffectsClimate2022.](tree_rings/tree_ring_figure.png)
 
@@
 Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and *T~crit~* [@osullivanThermalLimitsLeaf2017], *T~leaf~* of sun-exposed leaves may approach tolerance thresholds during heat waves, causing photosynthetic decline or even leaf death [@osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress].
-In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~* (unless exposed to sunflecks), but in some ways are likely to prove more sensitive to higher *T~air~*.
-Protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
+In contrast, understory leaves and trees are unlikely to face the same absolute extremes of *T~leaf~*, and protection from thermal stress associated with high direct radiation might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018a].
 However, with *T~air~* more frequently equaling or exceeding photosynthetic optima, occasional exposure to sunflecks coupled with lower capacity to dissipate excess heat may prove disadvantageous. 
 <!--Moreover, increased nighttime *T~air~* may disproportionately increase respiration relative to photosynthesis, thereby negatively affecting the carbon balance.-->
-While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can reduce growth more for understory trees [@rollinsonClimateSensitivityUnderstory2020].
-Thus, while canopy trees are probably more vulnerable to mortality from distinct heat-related disturbances, such as drought or heat waves, trees in the understory might be more negatively affected by chronic stress from warming *T~air~*, which may reduce growth and increase the risk of carbon starvation in these small trees, whose lower NSC reserves make them more vulnerable [@niinemetsResponsesForestTrees2010].
+While it is currently difficult to predict whether canopy or understory photosynthesis will be more severely affected by warming (section 4.2), limited tree-ring evidence indicates that unusually hot growing seasons can dramatically reduce the growth of understory trees [@rollinsonClimateSensitivityUnderstory2020], and these small trees may be at higher risk of carbon starvation because they have lower NSC reserves [@niinemetsResponsesForestTrees2010].
+Thus, despite their buffered microclimate, it is possible that understory trees could be more adversely impacted by warming temperatures than their canopy counterparts under some conditions.
 
@@
 Similarly, much remains unknown about how crown exposure influences the temperature sensitivity of woody stem growth. 
-While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions [@rollinsonClimateSensitivityUnderstory2020].
+While most available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees.
 

### 2022-06-10 response-to-reviews.Rmd
Update response-to-reviews.Rmd
remove use of "disproportionately" (R2)

@@
 
-*Nidhi*
+*https://github.com/EcoClimLab/vertical-thermal-review/issues/121*
 
@@
 
-**l.33 - ...**
+**l.33 - "Scaling from leaves to trees, canopy trees have higher absolute metabolic capacity and growth, yet are more vulnerable to drought and damaging *T~leaf~* than their smaller counterparts, particularly under climate change."**
 
@@
 
-**l.504 - ...**
+**l.504 - "Early leaf-out gives saplings and seedlings a window for high photosynthesis before they are shaded by canopy leaves-- contributing the majority of annual carbon fixation for some seedlings -- before canopy foliage and reduces light availability [@augspurgerDifferencesLeafPhenology2003; @leeSpringPhenologicalEscape2021].
+"**
 
-**l.648 - ...**
+**l.648 - "Among these, the best-studied process is aboveground woody growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but, is critically important to long-term forest dynamics and carbon cycling because woody tissues have a long residence time in the ecosystem [@russellResidenceTimesDecay2014]."**
 
-**l.814 - ...**
+**l.814 - "Large trees are also the most vulnerable to other climate-related disturbances [e.g., wind, lighting, @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging [@millerReducedImpactLogging2011]."**
 
-**l.838 - ...**
+**l.838 - "In the understory, warming reduces the abundance of less thermally-adapted plant species, causing thermophilization of the plant community [@zellwegerForestMicroclimateDynamics2020b; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019]."**
 
@@
 
+*response required*
+
 l.269 How does lower VPD in the understory inhibit evaporative cooling, if VPD is the major driver for evaporation?
@@
 Please explain very briefly.
+
+*response required*
 
@@
 Please revise.
+
+*response required*
 
@@
 
+*response required*
+
 l.629 'we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile'.
@@
 I think the temperature difference between upper and lower canopy positions, the theme of this review, should be mentioned in regard to this.
+
+*response required*
 

### 2022-06-10 vertical-thermal-review.Rmd
remove use of "disproportionately" (R2)

@@
 Leaf metabolism generally increases with height across the vertical gradient, yet differences in thermal sensitivity across the gradient appear modest.
-Scaling from leaves to trees, tall trees have higher absolute metabolic capacity and growth, yet are disproportionately vulnerable to drought and damaging *T~leaf~*, particularly under climate change.
+Scaling from leaves to trees, canopy trees have higher absolute metabolic capacity and growth, yet are more vulnerable to drought and damaging *T~leaf~* than their smaller counterparts, particularly under climate change.
 In contrast, understory trees experience fewer extreme high *T~leaf~*'s but have fewer cooling mechanisms and thus may be strongly impacted by warming under some conditions, particularly when exposed to a harsher microenvironment through canopy disturbance.
@@
 In temperate climates, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003].
-Early leaf-out gives saplings and seedlings a window for disproportionately high photosynthesis -- contributing the majority of annual carbon fixation for some seedlings -- before canopy foliage and reduces light availability [@augspurgerDifferencesLeafPhenology2003; @leeSpringPhenologicalEscape2021].
+Early leaf-out gives saplings and seedlings a window for high photosynthesis before they are shaded by canopy leaves-- contributing the majority of annual carbon fixation for some seedlings -- before canopy foliage and reduces light availability [@augspurgerDifferencesLeafPhenology2003; @leeSpringPhenologicalEscape2021].
 Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals is greater in larger trees in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c].
@@
 Photosynthate production is allocated among functions including respiration, aboveground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-Among these, the best-studied process is aboveground woody growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling because woody tissues have a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
+Among these, the best-studied process is aboveground woody growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but, is critically important to long-term forest dynamics and carbon cycling because woody tissues have a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
 Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2022].
@@
 Increasing severity and frequency of heat waves, accompanied with increases in VPD and ET, place canopy trees at particularly elevated risk of mortality (section 5.1), potentially causing large scale canopy die-back [@matusickSuddenForestCanopy2013; @teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities2021].
-Large trees are also disproportionately impacted by other climate-related disturbances [e.g., wind, lighting, @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging [@millerReducedImpactLogging2011].
-Forest fragmentation also disproportionately kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
+Large trees are also the most vulnerable to other climate-related disturbances [e.g., wind, lighting, @goraImplicationsSizedependentTree2021] that are expected to intensify with climate change [@ipccClimateChange20212021], and they are also targeted by selective logging [@millerReducedImpactLogging2011].
+Forest fragmentation also kills large trees by making them more vulnerable to wind, desiccation, and liana infestation [@lauranceRainForestFragmentation2006].
 Thus, canopies are becoming increasingly prone to disturbance and gap formation, which in turn increases radiation and temperatures within the forest [@juckerCanopyStructureTopography2018; @starkReframingTropicalSavannization2020].
@@
 Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020].
-In the understory, warming disproportionately affects less thermally-adapted plant species, causing thermophilization of the plant community [@zellwegerForestMicroclimateDynamics2020b; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
+In the understory, warming reduces the abundance of less thermally-adapted plant species, causing thermophilization of the plant community [@zellwegerForestMicroclimateDynamics2020b; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019].
 If compositional shifts towards more thermally-adapted species fail to keep pace with warming, the ecosystem-level resilience to canopy disturbance that is often provided by smaller trees [e.g., @millerReducedImpactLogging2011] will be destroyed, resulting in breakdown of canopy structure and the potential state changes described above.

### 2022-06-10 response-to-reviews.Rmd
response on ecosystem in abstract (R2)
@NidhiVinod , FYI. I believe I've now dealt with all of the most difficult reviewer comments. There are ~5 wording changes requested by R2 and a few small technical corrections from R1 remaining. I'm stopping for now and expect that you can easily handle the rest, but please ping me if you need help.

@@
 
-*response required*
+**We have reworded the abstract sentence to place less emphasis on ecosystems: "Using published research and new analyses, we evaluate how microclimate and leaf temperatures, traits, and gas exchange vary vertically in forests, shaping tree and ecosystem ecology." This wording parallels the section headers in the main manuscript.
+While we agree that there is little information on how ecosystem function varies across vertical profiles, this was included within the scope of our review, and we dedicate a 450-word subsection (sec.
+5.2) plus a figure to this theme.**
 

### 2022-06-10 vertical-thermal-review.Rmd
response on ecosystem in abstract (R2)
@NidhiVinod , FYI. I believe I've now dealt with all of the most difficult reviewer comments. There are ~5 wording changes requested by R2 and a few small technical corrections from R1 remaining. I'm stopping for now and expect that you can easily handle the rest, but please ping me if you need help.

@@
 Rising temperatures are influencing forests on many scales, with potentially strong variation vertically across forest strata. 
-Using published research and new analyses, we evaluate how microclimate and leaf temperatures, traits, and gas exchange vary vertically in forests, shaping tree ecology and ecosystem function. 
+Using published research and new analyses, we evaluate how microclimate and leaf temperatures, traits, and gas exchange vary vertically in forests, shaping tree and ecosystem ecology. 
 In closed-canopy forests, upper-canopy leaves are exposed to the highest solar radiation and evaporative demand, which can elevate leaf temperature (*T~leaf~*), particularly when transpirational cooling is curtailed by limited stomatal conductance.

### 2022-06-10 NEON_height_profiles/scripts/verticalProfileFigure.R
Fixed Fig 2 units per #118

@@ -182,7 +182,7 @@ for(i in 1:length(dp[,name])){
           paste("max T"["bio"], " [",degree,"C]")))
     } else if(i==5){ #PAR
       graph <- graph + 
-        xlab(bquote("max PAR [" ~mu ~ mol^'-2' ~ s^'-1'~"]"))
+        xlab(bquote("max PAR [" ~mu ~mol ~m^'-2' ~ s^'-1'~"]"))
     }
     
     ## remove y-axis from all but windspeed

### 2022-06-18 vertical-thermal-review.Rmd
updating responses to comments

@@
 Thus, leaves develop differently according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of branch ontogenetic stage [@carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
-Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, including a “memory” of previous conditions in that location, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
+Plasticity throughout the canopy may emerge when development of new leaves is influenced by information from adjacent mature leaves or apical meristems, and may also acclimate during and after expansion to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
 Overall, leaf biochemistry, anatomy and structure may be optimized to local conditions [@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007].

### 2022-06-18 vertical-thermal-review.Rmd
@teixeirak updated this sentence
Not sure if R2 read it as just respiration instead of Q10 but I added this sentence, let me know if it makes sense to the comment:

R2's comment: we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile'. Other than that it varies according to the prevailing temperature difference? I think the temperature difference between upper and lower canopy positions, the theme of this review, should be mentioned in regard to this

@@
 This variation may be attributable to forest type, leaf traits and age [e.g., greater *Q~10~* in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases *Q~10~* [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
-Overall, however, we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
+Overall, although *Q~10~* is influenced by the temperature difference between upper and lower canopy positions, we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
 

### 2022-06-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

--- 2022-06-18 vertical-thermal-review.Rmd
@@
 This variation may be attributable to forest type, leaf traits and age [e.g., greater *Q~10~* in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases *Q~10~* [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
-Overall, however, we currently lack mechanistic understanding of how and why the temperature sensitivity of respiration varies across the forest vertical profile.
+Overall, although *Q~10~* is influenced by the temperature difference between upper and lower canopy positions, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.
 

### 2022-06-18 vertical-thermal-review.Rmd
addressing comment about gs

@@
 The temperature at which *g~s~* is greatest, *T~opt~* of *g~s~*, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in *g~s~* in response to rising *T~air~* (driven by increased VPD) than lower-canopy shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
-This, added to the tendency for sun leaves to have higher *T~leaf~*, implies that high *T~air~* should decrease *g~s~* of upper canopy leaves more than understory leaves, particularly when water availability is limited.
+Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per *se* in nature, causing *g~s~* to decline when the leaf warms.
 

### 2022-06-18 response-to-reviews.Rmd
updating responses to comments
Update response-to-reviews.Rmd
here is the response to the reviewer
addressing comment about gs

@@
 
-*Nidhi*
+**Done**
 
@@
 
-*response required*
+**Revised as suggested.**
 
@@
 
+**We have removed the quoted sentence as it was redundant with that which had preceded it, but had confusingly couched the message of the preceding sentence as if it were a speculation rather than an empirical fact.
+The preceding sentence, which we have retained, read:**
+*“…sun leaves in the upper canopy show a stronger decrease in* *g~s~* *in response to rising* *T~air~* *(driven by increased VPD) than lower-canopy shade leaves (Hernández et al., 2020; Carter et al., 2021)”.*
+**That is an empirical result.
+The fact that the effect of *T~air~* on *g~s~* in that result is driven by VPD rather than somehow directly by *T~air~* itself is indicated by the fact that increasing temperature itself apparently *opens* stomata, rather than closing them.
+This is also germane to the reviewer’s final comment about decoupling VPD and *T~leaf~*.
+There are in fact six published studies that have decoupled VPD and *T~leaf~*, and they all suggest temperature per se causes an increase, not a reduction, in stomatal conductance.
+We have added the following sentence to the end of the paragraph to clarify this matter, citing three of the studies out of six due to word count limit: “Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@schulzeRoleAirHumidity1974; @hallRegulationWaterLoss1975; @hallStomatalResponseEnvironment1975; @fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per se in nature, causing *g~s~* to decline when the leaf warms.**
+
 *response required*
@@
 
-*response required*
+**Thanks for suggesting. We included the following sentence:**
+**Overall, although *Q~10~* is influenced by the temperature difference between upper and lower canopy positions, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.**
 

### 2022-06-18 response-to-reviews.Rmd
responded to the comment

@@
 There are in fact six published studies that have decoupled VPD and *T~leaf~*, and they all suggest temperature per se causes an increase, not a reduction, in stomatal conductance.
-We have added the following sentence to the end of the paragraph to clarify this matter, citing three of the studies out of six due to word count limit: “Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@schulzeRoleAirHumidity1974; @hallRegulationWaterLoss1975; @hallStomatalResponseEnvironment1975; @fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per se in nature, causing *g~s~* to decline when the leaf warms.**
-
-*response required*
+We have added the following sentence to the end of the paragraph to clarify this matter, citing three of the six studies due to limit on word count: “Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@schulzeRoleAirHumidity1974; @hallRegulationWaterLoss1975; @hallStomatalResponseEnvironment1975; @fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per se in nature, causing *g~s~* to decline when the leaf warms.**
 

### 2022-06-18 response-to-reviews.Rmd
Update response-to-reviews.Rmd

@@
 **Thanks for suggesting. We included the following sentence:**
-**Overall, although *Q~10~* is influenced by the temperature difference between upper and lower canopy positions, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.**
+**Overall, although the temperature difference between upper and lower canopy positions is important driver of *Q~10~*, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.**
 

### 2022-06-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

@@
 `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017 <!-- @songTreeSurfaceTemperature2020-->].
-Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves. 
+Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves, because wind speeds and hence *g~b~* are generally smaller in the lower canopy [@robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999].
 Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal *T~leaf~* under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
@@
 This variation may be attributable to forest type, leaf traits and age [e.g., greater *Q~10~* in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases *Q~10~* [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
-Overall, although *Q~10~* is influenced by the temperature difference between upper and lower canopy positions, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.
+Overall, although the temperature difference between upper and lower canopy positions is important driver of *Q~10~*, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.
 

### 2022-06-18 response-to-reviews.Rmd
comment on l.269

@@
 
-*response required*
+**We added an explanation to l.264-265 with references:**
+**"Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves, *because wind speeds and hence boundary layer conductance are generally smaller in the lower canopy [@robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999].***
 
@@
 
-*response required*
+**Lower VPD inhibits evaporative cooling precisely because VPD is the major driver for evaporation: Transpiration rate generally increases as VPD increases, and decreases as VPD decreases.
+Therefore lower VPD will mean less transpiration and hence less evaporative cooling.
+Perhaps the word *‘inhibits’* confused the issue; we have replaced *‘inhibits’* with *‘reduces’*.
+We have also modified the sentence to read, ‘In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also reduces evaporative cooling (because evaporation rate generally increases with VPD), thus increasing Tleaf – Tair …’.” (phrase “because evaporation rate generally increases with VPD” added)**
 

### 2022-06-18 vertical-thermal-review.Rmd
comment on l.269

@@
 Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal *T~leaf~* under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
-In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also inhibits evaporative cooling, thus increasing *T~leaf~* - *T~air~* [Fig.
+In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also reduces evaporative cooling (because evaporation rate generally increases with VPD), thus increasing *T~leaf~* - *T~air~* [Fig.
 `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].

### 2022-06-18 response-to-reviews.Rmd
I have addressed all the comments now

@@
 Perhaps the word *‘inhibits’* confused the issue; we have replaced *‘inhibits’* with *‘reduces’*.
-We have also modified the sentence to read, ‘In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also reduces evaporative cooling (because evaporation rate generally increases with VPD), thus increasing Tleaf – Tair …’.” (phrase “because evaporation rate generally increases with VPD” added)**
+We have also modified the sentence to read, ‘In addition to lower wind speeds, higher RH (lower VPD) in the understory and inner canopy also reduces evaporative cooling (because evaporation rate generally increases with VPD), thus increasing *T~leaf~* – *T~air~* …’.” (phrase “because evaporation rate generally increases with VPD” added)**
 

### 2022-06-27 vertical-thermal-review.Rmd
small updates

@@
 Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [@raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988].
-Within canopies, however, transport often fails to mimic diffusion – e.g., transport of momentum [@shawSecondaryWindSpeed1977] or heat [@raupachLagrangianAnalysisScalar1987] may occur in the direction opposite to that predicted by gradients of wind speed or temperature, respectively – making vertical wind profiles difficult to generalize from first principles [@denmeadScalarTransportPlant1987; @RaupachM.R1989Ttip; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
+Within canopies, however, transport often fails to mimic diffusion – e.g., transport of momentum [@shawSecondaryWindSpeed1977] or heat [@raupachLagrangianAnalysisScalar1987] may occur in the direction opposite to that predicted by gradients of wind speed or temperature, respectively – making vertical wind profiles difficult to generalize from first principles [@denmeadScalarTransportPlant1987; @raupachApplyingLagrangianFluid1989; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007].
 Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or depleted air from beneath and within the canopy and flushing them to the atmosphere above [@finniganTurbulenceWavingWheat1979; @baldocchiTraceGasExchange1991].

### 2022-06-27 response-to-reviews.Rmd
rewording on Q10
@nidhivinod,  here's a little wording change related tot his commit: https://github.com/EcoClimLab/vertical-thermal-review/commit/f4f03b53812ef7fbcf96cd3359c5c31321aab11c

It's important to distinguish here between temperature sensitivity of respiration (e.g., Q10) and respiration (determined based on T and other factors). I believe the reviewer's comment was referring to respiration, as opposed to Q10.  Also, I broadened wording from Q10 to "temperature sensitivity" because T sensitivity can also be characterized as E_o.

@@
 **Thanks for suggesting. We included the following sentence:**
-**Overall, although the temperature difference between upper and lower canopy positions is important driver of *Q~10~*, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.**
+**Overall, although the temperature difference between upper and lower canopy positions is important driver of respiration, we currently lack mechanistic understanding of how and why its temperature sensitivity varies across the forest vertical profile.**
 

### 2022-06-27 vertical-thermal-review.Rmd
rewording on Q10
@nidhivinod,  here's a little wording change related tot his commit: https://github.com/EcoClimLab/vertical-thermal-review/commit/f4f03b53812ef7fbcf96cd3359c5c31321aab11c

It's important to distinguish here between temperature sensitivity of respiration (e.g., Q10) and respiration (determined based on T and other factors). I believe the reviewer's comment was referring to respiration, as opposed to Q10.  Also, I broadened wording from Q10 to "temperature sensitivity" because T sensitivity can also be characterized as E_o.

@@
 This variation may be attributable to forest type, leaf traits and age [e.g., greater *Q~10~* in younger leaves, @zhouLeafageEffectsTemperature2015], or acclimation to high temperature that decreases *Q~10~* [@slotGeneralPatternsAcclimation2015; @carterExperimentalWarmingTropical2021].
-Overall, although the temperature difference between upper and lower canopy positions is important driver of *Q~10~*, we currently lack mechanistic understanding of how and why *Q~10~* varies across the forest vertical profile.
+Overall, although the temperature difference between upper and lower canopy positions is important driver of respiration, we currently lack mechanistic understanding of how and why its temperature sensitivity varies across the forest vertical profile.
 

### 2022-06-28 vertical-thermal-review.Rmd
final check, update word count
@NidhiVinod , FYI. Ready to submit once we deal with issue #115

@@
 #Word count
-w_intro=369
-w_biophys=1515
-w_tleaf=749
-w_traits= 2392
-w_met=941
-w_ecol=1357
+w_intro=367
+w_biophys=1519
+w_tleaf=794
+w_traits= 2390
+w_met=991
+w_ecol=1361
 w_vertgradients=w_biophys+w_tleaf+w_traits+w_met+w_ecol
-w_imp_gc=1071
+w_imp_gc=1042
 w_imp_modeling_scaling=1173
 w_implications=w_imp_gc+w_imp_modeling_scaling
-w_conclusions=247
+w_conclusions=231
 w_total=w_intro+w_vertgradients+w_implications+w_conclusions
@@
 Total word count (excluding summary, references and legends) | 	`r w_total` |	No. of figures | 	6 (all colour)
-Summary|  	200 | 	No. of Tables | 	2  
+Summary|  	199 | 	No. of Tables | 	2  
 Introduction|	`r w_intro` |	No of Supporting Information files | 	6
@@
 This light gradient is more pronounced in dense canopies, including broad-leaf and mixed forests, than in forests with more open upper canopies, including many conifer forests [Fig.
-`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; <!--@aussenacInteractionsForestStands2000a-->; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
+`r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; <!--@aussenacInteractionsForestStands2000a--> @tymenQuantifyingMicroenvironmentalVariation2017; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].
 This light gradient drives variation in leaf temperature (section 2), traits (section 3), and photosynthesis (section 4.2).
@@
 Dots along each line correspond to the biophysical constant assigned for the independent variable in the other scenarios, and therefore represent the modeled value for *T~leaf~* - *T~air~* with all parameters set to the biophysical constants given in the table.
-Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf characteristic dimension (lcd), stomatal conductance (gs), air temperature (tair).
+Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (*RH*), leaf characteristic dimension (lcd), stomatal conductance (*g~s~*), air temperature (*T~air~*).
 ](leaf_energy_balance/fig3_leafenergy.png)
@@
 Steeper leaf angles reduce mid-day radiation loads and thereby decrease *T~leaf~* - *T~air~* [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, also decreasing radiation load.
-Further, sun leaves tend to have higher stomatal and vein densities, which fascilitate evaporative cooling by enabling higher *g~s~* [section 4.1, @zwienieckiHydraulicLimitationsImposed2004]. 
+Further, sun leaves tend to have higher stomatal and vein densities, which facilitate evaporative cooling by enabling higher *g~s~* [section 4.1, @zwienieckiHydraulicLimitationsImposed2004]. 
 Yet sun leaves also have adaptations to limit water stress, particularly in drier climates, including greater cuticle thickness and higher trichome density [which may reduce or increase *g~b~*, @schreuderModelledInfluencesNonexchanging2001; @ichieEcologicalDistributionLeaf2016a].
@@
 The temperature at which *g~s~* is greatest, *T~opt~* of *g~s~*, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in *g~s~* in response to rising *T~air~* (driven by increased VPD) than lower-canopy shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].
-Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per *se* in nature, causing *g~s~* to decline when the leaf warms.
+Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature *per se* in nature, causing *g~s~* to decline when the leaf warms.
 
@@
 Photosynthate production is allocated among functions including respiration, aboveground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs).
-Among these, the best-studied process is aboveground woody growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but, is critically important to long-term forest dynamics and carbon cycling because woody tissues have a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
+Among these, the best-studied process is aboveground woody growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021], but is critically important to long-term forest dynamics and carbon cycling because woody tissues have a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
 Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimate2022].
@@
 Transparent ribbons indicate 95% confidence intervals.
-From @anderson-teixeiraJointEffectsClimate2022.](tree_rings/tree_ring_figure.png)
+Redrawn from @anderson-teixeiraJointEffectsClimate2022.](tree_rings/tree_ring_figure.png)
 
@@
 Severe degradation can cause dramatic ecological state changes (e.g., the transition of tropical forests more open, savanna-like vegetation) and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
-Such dynamics will likely to be amplified by warming, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may not recover and persist after severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020; @floresFloodplainsAchillesHeel2017].
+Such dynamics are likely to be amplified by warming, such that climate change is pushing some of the world's forests into alternative stable states wherein forest can persist as long as the canopy remains largely intact, but may not recover and persist after severe canopy disturbance [@tepleyVulnerabilityForestLoss2017; @millerAlternativeStableEquilibria2019; @mcdowellPervasiveShiftsForest2020; @floresFloodplainsAchillesHeel2017].
 
@@
 
-As climate changes, patterns and processes across vertical gradients will likely to shift as well.
+As climate changes, patterns and processes across vertical gradients will likely shift as well.
 In the historical climates to which trees adapted, the canopy was an advantageous place for photosynthesis and growth. 

### 2022-06-29 vertical-thermal-review_SI.Rmd
small fix to SI
@teixeirak i added the citations here

@@
 
-# `r Notes_S1`: Biophysical drivers of $T_{leaf}$
+# Note `r Notes_leafTbiophysics`: Biophysical drivers of T~leaf~
 
@@
 
-Micrometeorological data was downloaded for each site from NEON (neonUtilies R package) at 30 minute intervals for 2015-2020.
+Micrometeorological data was downloaded for each site from NEON (neonUtilies R package) at 30 minute intervals for 2015-2020 [@https://doi.org/10.48443/77n6-eh42;  @https://doi.org/10.48443/0tyt-ky71;  @https://doi.org/10.48443/7rs6-ff56;  @https://doi.org/10.48443/8a01-0677;  @https://doi.org/10.48443/w9nf-k476].
 To focus on the middle of the growing season, the data was constrained to be for the month of July each year.

### 2022-06-29 vertical-thermal-review.Rmd
Johnston in press is published

@@
 This holds true across the range of forest types [Fig.
-`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; <!--@mcgregorTreeHeightLeaf2021;-->@juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVerticalinpress].
+`r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016; <!--@mcgregorTreeHeightLeaf2021;-->@juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2022a].
 As a result, upper canopy leaves have substantially higher boundary layer conductance (*g~b~*) and therefore greater potential for exchange of both mass and sensible heat exchange (sections 2, 4.1).
@@
 The latter can result from a combination of still air at lower heights, sunflecks or the 'canopy greenhouse effect' (section 1.5), and the tendency for shade leaves to have lower *g~s~* [section 3, @schymanskiStomatalControlLeaf2013; @hardwickRelationshipLeafArea2015].
-Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing *T~leaf~* near the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVerticalinpress].
+Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing *T~leaf~* near the ground [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2022a].
 One of the few studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5°C cooler temperatures lower in the vertical profile of a dense subalpine stand in the alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018].
-However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVerticalinpress].
+However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2022a].
 
@@
 Canopy shading can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
-Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although few applications exist at finer-scales needed to understand vertical variation in canopy temperatures [@johnstonWhatLiesVerticalinpress].
+Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although few applications exist at finer-scales needed to understand vertical variation in canopy temperatures [@johnstonWhatLiesVertical2022a].
 As an example of the type of insight possible with this approach, @pauTropicalForestTemperature2018 used data from a tower-based infrared camera in combination with eddy-covariance data and found that tropical forest GPP was more strongly associated with canopy temperature than *T~air~* or VPD.
 In a a western-U.S.
-savanna system in the western U.S., @johnstonWhatLiesVerticalinpress found lower foliage temperatures at the tops of tree crowns than in the understory, which was influenced by very high grass temperatures, consistent with the principles outlined in sections 1-2 (Fig.
+savanna system in the western U.S., @johnstonWhatLiesVertical2022a found lower foliage temperatures at the tops of tree crowns than in the understory, which was influenced by very high grass temperatures, consistent with the principles outlined in sections 1-2 (Fig.
 `r SIfig_NEON_all`).

### 2022-06-29 vertical-thermal-review_SI.Rmd
added docs to new journal submission folder

@@
 Vegetation temperature ($T_{veg}$) is measured using infrared sensors that integrate across both leaves and woody vegetation. 
-Leaf and branch temperatures often differ substantially; for instance, exposed tropical tree bark can be much warmer than leaves [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021; @johnstonWhatLiesVertical2020].
+Leaf and branch temperatures often differ substantially; for instance, exposed tropical tree bark can be much warmer than leaves [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021; @johnstonWhatLiesVertical2022a].
 Thus, vegetation temperature ($T_{veg}$) does not always equal or approximate $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (i.e., low-LAI ecosystems and understories).

### 2022-08-03 journal submissions/2022_29_06/response-to-reviews.Rmd
Create response-to-reviews.Rmd

--- 2022-06-27 response-to-reviews.Rmd
@@
 We are pleased to submit a revised version of our manuscript, "Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications" (NPH-TR-2021-38020), for consideration for publication in *New Phytologist*.
-We have addressed all points raised by the reviewers, as detailed below.
-The most significant changes include:
-
-...
+We have addressed all points raised by the reviewers, as detailed in our responses to the reviewers.
+The most significant change was de-emphasizing the discussion of the potential for stronger negative effects of rising temperatures on understory trees, including removal of former Fig.
+4b and modifying the text throughout to give a more consistent message (as requested by reviewer #2).
 
@@
 - It will generally be fine to adjust colors, layout, and spacing on the figures.
-- We really like Figure 1 and would like to retain the background painting (by the lead author)
+- We would like to retain the background painting in Figure 1 (by the lead author)
 
@@
 Comments to the Author
+
 I am satisfied with the corrections and the responses to the comments. I like the new structure, the writting is very precise while covering broad areas and the paper very interesting.
@@
 
-*https://github.com/EcoClimLab/vertical-thermal-review/issues/118*
+**Fixed.**
 
@@
 
-*https://github.com/EcoClimLab/vertical-thermal-review/issues/121*
+**Fixed.**
 
@@
 
-**The full sentence reads: "More recent work has extended the focus from light alone to *height* in the vertical profile, as *the latter* shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy." Thus, we're saying that *height* shapes differences in leaf traits that caould mitigate effects of gravity and hydraulic pathlength.
+**The full sentence reads: "More recent work has extended the focus from light alone to *height* in the vertical profile, as *the latter* shapes differences in leaf structure and function that can partially mitigate the effects of gravity and hydraulic pathlength on leaves higher up in the canopy." Thus, we're saying that *height* shapes differences in leaf traits that could mitigate effects of gravity and hydraulic pathlength.
 We believe the sentence to be correct as written.**
@@
 This is also germane to the reviewer’s final comment about decoupling VPD and *T~leaf~*.
-There are in fact six published studies that have decoupled VPD and *T~leaf~*, and they all suggest temperature per se causes an increase, not a reduction, in stomatal conductance.
-We have added the following sentence to the end of the paragraph to clarify this matter, citing three of the six studies due to limit on word count: “Available data also suggest that high *T~air~* can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with *T~air~* to keep VPD constant [@schulzeRoleAirHumidity1974; @hallRegulationWaterLoss1975; @hallStomatalResponseEnvironment1975; @fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], though it appears that the negative effect of temperature on *g~s~*, mediated by VPD, generally overrides the positive effect of temperature per se in nature, causing *g~s~* to decline when the leaf warms.**
+There are in fact six published studies that have decoupled VPD and *T~leaf~* [@schulzeRoleAirHumidity1974; @hallRegulationWaterLoss1975; @hallStomatalResponseEnvironment1975; @fredeenTemperatureHumidityEffects1999a; @mottStomatalResponsesHumidity2010; @urbanStomatalConductanceIncreases2017a], and they all suggest temperature per se causes an increase, not a reduction, in stomatal conductance.
+We have added the following sentence to the end of the paragraph to clarify this matter, citing three of the six studies due to limit on word count: "Available data also suggest that high Tair can directly drive stomatal opening, if vapor pressure is experimentally adjusted in tandem with Tair to keep VPD constant (Fredeen & Sage, 1999; Mott & Peak, 2010; Urban et al., 2017), though it appears that the negative effect of temperature on gs, mediated by VPD, generally overrides the positive effect of temperature per se in nature, causing gs to decline when the leaf warms."**
 

### 2022-08-04 vertical-thermal-review.Rmd
work on editorial requests
@NidhiVinod , FYI

@@
 **Authors:**
-Nidhi Vinod^1,2^,
-Martijn Slot^3^,
-Ian R. McGregor^4^,
-Elsa M. Ordway^2,5^,
-Marielle N. Smith^6^,
-Tyeen C. Taylor^7^,
-Lawren Sack^2^,
-Thomas N. Buckley^8^,
-Kristina J. Anderson-Teixeira^1,3^*
+
+Nidhi Vinod^1,2^
+
+Martijn Slot^3^
+
+Ian R. McGregor^4^ (Orcid ID: 0000-0002-5763-021X)
+
+Elsa M. Ordway^2,5^
+
+Marielle N. Smith^6^
+
+Tyeen C. Taylor^7^
+
+Lawren Sack^2^
+
+Thomas N. Buckley^8^
+
+Kristina J. Anderson-Teixeira^1,3^* ( Orcid ID : 0000-0001-8461-9713)
 
@@
 
-3. Smithsonian Tropical Research Institute; Panama, Republic of Panama
+3. Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panama City, Panama
 
@@
 
-*T~leaf~* is strongly tied to, and usually within a few degrees of, *T~air~* [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Note `r Notes_leafTbiophysics`,@doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009].
+*T~leaf~* is strongly tied to, and usually within a few degrees of, *T~air~* [@drakeNoEvidenceHomeostatic2020;@rey-sanchez_spatial_2016], but can be substantially warmer or cooler under certain conditions [Notes `r Notes_leafTbiophysics`,@doughtyAreTropicalForests2008;@rey-sanchez_spatial_2016;@vogelLeavesLowestHighest2009].
 Deviation of *T~leaf~* from *T~air~* is influenced by other micrometeorological drivers and by leaf traits and stomatal conductance (*g~s~*; Fig.
-`r fig_leaf_T`; Note `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig.
+`r fig_leaf_T`; Notes `r Notes_leafTbiophysics`), all of which vary across forest vertical gradients (Fig.
 `r fig_schematic`).
@@
 However, higher wind speeds reduce *T~leaf~* - *T~air~* [Fig. `r fig_leaf_T`b, @niinemetsPhotosyntheticAcclimationSimultaneous2004; @bonan_ecological_2016]. 
-In addition, adaptive leaf traits that increase *g~s~* and *g~b~*, such as small size and high stomatal density (Note `r Notes_leafTbiophysics`; sections 3, 4.1), mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
+In addition, adaptive leaf traits that increase *g~s~* and *g~b~*, such as small size and high stomatal density (Notes `r Notes_leafTbiophysics`; sections 3, 4.1), mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig.
 `r fig_leaf_T`b-e).
@@
 
-Note `r Notes_leafTbiophysics`. Biophysical drivers of T~leaf~
+Notes `r Notes_leafTbiophysics`. Biophysical drivers of T~leaf~
 

### 2022-08-04 NEON_height_profiles/scripts/verticalProfileFigure.R
Updated figure and added comment #123

@@ -401,7 +401,7 @@ groupPlots <- lapply(1:length(sites), function(X){
   
   # void <- ggplot() + theme_void()
   
-  if(X==1) titleTx <- "(A) (Sub) Subtropical and warm temperate broadleaf forests"
+  if(X==1) titleTx <- "(A) (Sub)tropical and warm temperate broadleaf forests"
   if(X==2) titleTx <- "(B) Temperate open forests / savanna"
   if(X==3) titleTx <- "(C) Temperate mesic broadleaf forests"
   if(X==4) titleTx <- "(D) Temperate conifer forests"
@@ -423,6 +423,7 @@ groupPlots <- lapply(1:length(sites), function(X){
   return(list(p, allNEON))
 })
 
+# NB the warnings from the code above are ok.
 
 #get numbers for manuscript
 aggNeon <- rbindlist(lapply(groupPlots, `[[`, 2))
@@ -458,7 +459,7 @@ q <- ggarrange(groupPlots[["subBroad"]], groupPlots[["savOpenTemp"]],
                groupPlots[["tempBroad"]], groupPlots[["tempCon"]],
                groupPlots[["northBor"]],
                nrow=5, ncol=1)
-png(paste0("figures/profile_all_groups.png"), 
+png(paste0("figures/profile_all_groups1.png"), 
     height=1800, width=1500)
 print(q)
 dev.off()

### 2022-08-04 vertical-thermal-review.Rmd
progress on editorial requests
a few updates

@@
 
-Martijn Slot^3^
+Martijn Slot^3^ (ORCID: 0000-0002-5558-1792; Twitter: @martijn_slot; Facebook: martijn.slot.52)
 
-Ian R. McGregor^4^ (Orcid ID: 0000-0002-5763-021X)
+Ian R. McGregor^4^ (ORCID: 0000-0002-5763-021X; Twitter: @ecoterrestrian)
 
-Elsa M. Ordway^2,5^
+Elsa M. Ordway^2,5^ (ORCID: 0000-0002-7720-1754; Twitter: @ElsaOrdway)
 
@@
 
-Thomas N. Buckley^8^
+Thomas N. Buckley^8^ (ORCID: 0000-0001-7610-7136; Twitter: @TomBuckleyLab)
 
-Kristina J. Anderson-Teixeira^1,3^* ( Orcid ID : 0000-0001-8461-9713)
+Kristina J. Anderson-Teixeira^1,3^* (Orcid ID : 0000-0001-8461-9713; Twitter: @K_A_Teixeira)
 
@@
 
-3. Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panama City, Panama
+3. Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panama City, Panama
 
@@
 This manuscript benefited from feedback from Eleinis Ávila-Lovera, the ForestGEO Ecosystems & Climate lab at NZCBI, and two anonymous reviewers.
-Funding was provided by the Smithsonian Institution.
+Funding was provided by the Smithsonian Institution, NSF grant #1951244 to LS and TNB, and USDA-NIFA Hatch project grant #1016439 to TNB.
 

### 2022-08-04 vertical-thermal-review.Rmd
misc updates

@@
 
-Martijn Slot^3^ (ORCID: 0000-0002-5558-1792; Twitter: @martijn_slot; Facebook: martijn.slot.52)
+Martijn Slot^3^ (ORCID: 0000-0002-5558-1792; Twitter: martijn_slot; Facebook: martijn.slot.52)
 
-Ian R. McGregor^4^ (ORCID: 0000-0002-5763-021X; Twitter: @ecoterrestrian)
+Ian R. McGregor^4^ (ORCID: 0000-0002-5763-021X; Twitter: ecoterrestrian)
 
-Elsa M. Ordway^2,5^ (ORCID: 0000-0002-7720-1754; Twitter: @ElsaOrdway)
+Elsa M. Ordway^2,5^ (ORCID: 0000-0002-7720-1754; Twitter: ElsaOrdway)
 
@@
 
-Thomas N. Buckley^8^ (ORCID: 0000-0001-7610-7136; Twitter: @TomBuckleyLab)
+Thomas N. Buckley^8^ (ORCID: 0000-0001-7610-7136; Twitter: TomBuckleyLab)
 
-Kristina J. Anderson-Teixeira^1,3^* (Orcid ID : 0000-0001-8461-9713; Twitter: @K_A_Teixeira)
+Kristina J. Anderson-Teixeira^1,3^* (Orcid ID : 0000-0001-8461-9713; Twitter: K_A_Teixeira)
 

### 2022-08-05 NEON_height_profiles/scripts/verticalProfileFigure.R
Updated xax labels 2a #124
Also made bracket spacing consistent when there are superscripts to units

@@ -166,12 +166,13 @@ for(i in 1:length(dp[,name])){
            y = "") +
       theme_bw()
     
-    if(i %in% c(1,2)){ #windspeed, RH
+    if(i ==1){ #windspeed
+      graph <- graph + 
+        xlab(bquote("max Windspeed ["~m ~s^'-1'~"]")) + 
+        ylab("Normalized height")
+    } else if(i==2){ #RH
       graph <- graph + 
         xlab(paste0(dp[,stat][i], " ", dp[,xlabs][i]))
-      if(i==1){
-        graph <- graph + ylab("Normalized height")
-      }
     } else if(i==3){ #tempSingle
       graph <- graph +
         xlab(expression(
@@ -182,7 +183,7 @@ for(i in 1:length(dp[,name])){
           paste("max T"["bio"], " [",degree,"C]")))
     } else if(i==5){ #PAR
       graph <- graph + 
-        xlab(bquote("max PAR [" ~mu ~mol ~m^'-2' ~ s^'-1'~"]"))
+        xlab(bquote("max PAR ["~mu ~mol ~m^'-2' ~ s^'-1'~"]"))
     }
     
     ## remove y-axis from all but windspeed
@@ -204,6 +205,7 @@ for(i in 1:length(dp[,name])){
   })
 }
 
+# NB the warnings from the code above are ok.
 names(plots) <- dp[,name]
 
 ## bring in LAD profiles

### 2022-08-10 vertical-thermal-review_SI.Rmd

@@
 
+**Authors:** 
+Nidhi Vinod,
+Martijn Slot,
+Ian R. McGregor,
+Elsa M. Ordway,
+Marielle N. Smith,
+Tyeen C. Taylor,
+Lawren Sack,
+Thomas N. Buckley,
+Kristina J. Anderson-Teixeira
+
+**Article acceptance date:** 31 July 2022.
+
 \newpage
 
-# Note `r Notes_leafTbiophysics`: Biophysical drivers of T~leaf~
+# Notes `r Notes_leafTbiophysics`: Biophysical drivers of T~leaf~
 
@@
 Vertical gradients in micrometeorological conditions for all forested sites in the National Ecological Observatory Network (NEON)**.
+Sites are grouped into five forest types: (sub)tropical and warm temperate broadleaf deciduous forests (A), temperate open forests and savannas (B), temperate mesic broadleaf forests (C), temperate conifer forests (D), and northern and boreal forests (E).
 Shown are  height profiles in July mean ± 1 standard deviation for maximum photosynthetically active radiation (PAR), maximum wind speed, minimum humidity, maximum $T_{air}$, and maximum biological temperature, $T_{bio}$.

### 2022-08-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
small citation updates

@@
 
-Marielle N. Smith^6^
+Marielle N. Smith^6,7^ (ORCID: 0000-0003-2323-331X; Twitter: marielle_smith1)
 
-Tyeen C. Taylor^7^
+Tyeen C. Taylor^8^ (ORCID: 0000-0002-0926-098X)
 
@@
 
-Thomas N. Buckley^8^ (ORCID: 0000-0001-7610-7136; Twitter: TomBuckleyLab)
+Thomas N. Buckley^9^ (ORCID: 0000-0001-7610-7136; Twitter: TomBuckleyLab)
 
@@
 
-7. Department of Civil & Environmental Engineering, University of Michigan, Ann Arbor, MI 48109, USA
+7. School of Natural Sciences, College of Environmental Sciences and Engineering, Bangor University, Bangor LL57 2DG, United Kingdom
 
-8. Department of Plant Sciences; University of California; Davis, CA, 95616, USA
+8. Department of Civil & Environmental Engineering, University of Michigan, Ann Arbor, MI 48109, USA
+
+9. Department of Plant Sciences; University of California; Davis, CA, 95616, USA
 
@@
 
-Fewer than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that vertical variation in emissions is influenced by both vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
+Fewer than half of tree species express significant light-dependent isoprene emissions [@kesselmeierBiogenicVolatileOrganic1999; @taylorIsopreneEmissionStructures2018], such that vertical variation in emissions is influenced by both vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
 Within species, isoprene emission tends to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species are more abundant in hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018].
@@
 It remains uncertain how responses to warming will vary across vertical gradients under mesic conditions. 
-Leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], yet failure to fully acclimate will reduce carbon sequestration at leaf, tree, and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
+Leaves display substantial plasticity to adapt to warmer temperatures [<!--@slotGeneralPatternsAcclimation2015;--> @slotPhotosyntheticPlasticityTropical2021a; @wayJustRightTemperature2019;@cunninghamTemperateRainforestTrees2003] and to recover from heat stress [@smithEmpiricalEvidenceResilience2020], yet failure to fully acclimate will reduce carbon sequestration at leaf, tree, and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021].
 As growing season *T~air~* increasingly equals or exceeds photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], exposed crowns are likely most vulnerable.
@@
 Funding was provided by the Smithsonian Institution, NSF grant #1951244 to LS and TNB, and USDA-NIFA Hatch project grant #1016439 to TNB.
+MNS was supported by NSF DEB grant (no. 1950080) to Michigan State University.
 

### 2022-09-02 vertical-thermal-review.Rmd
small changes

@@
 
-$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).
+$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer); TrS = tropical savanna.
 
@@
 
-$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN= boreal needleleaf (conifer).
+$\dagger$ Forest types are coded as follows: TrB = tropical broadleaf; TeB = temperate broadleaf; TeN = temperate needleleaf (conifer); BoN = boreal needleleaf (conifer).
 
@@
 Large trees also dominate woody aboveground net primary productivity (*ANPP~woody~*) and mortality (*M~woody~*), with trees ≥10cm DBH usually contributing > 85% of *ANPP~woody~* and *M~woody~* across 25 globally distributed forests [Fig.
-`r fig_ecosystem`b, @piponiotDistributionBiomassDynamicsinpress].
+`r fig_ecosystem`b, @piponiotDistributionBiomassDynamics2022].
 
 ![**Figure `r fig_ecosystem`.
-Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotDistributionBiomassDynamicsinpress.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotDistributionBiomassDynamics2022.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
 

### 2022-09-02 Lit_Review/Zotero_references/files/112/hotjar-1678608.js
move Lit_Review to private repo
@NidhiVinod , please make sure I didn't miss anything that should be published. Note that we cannot publish pdfs of other studies.

@@ -1,4 +1,0 @@
-window.hjSiteSettings = window.hjSiteSettings || {"site_id":1678608,"r":0.46287797678523845,"rec_value":1.0,"state_change_listen_mode":"automatic","record":false,"continuous_capture_enabled":false,"recording_capture_keystrokes":true,"anonymize_digits":true,"anonymize_emails":true,"suppress_all":false,"suppress_text":false,"suppress_location":false,"user_attributes_enabled":false,"legal_name":"Annual Reviews","privacy_policy_url":"https://www.annualreviews.org/page/about/privacy","deferred_page_contents":[{"id":18602031,"targeting":[{"component":"url","match_operation":"simple","pattern":"https://www.annualreviews.org/toc/psych/63/1","negate":false},{"component":"device","match_operation":"exact","pattern":"tablet","negate":false}]}],"record_targeting_rules":[{"component":"url","match_operation":"contains","pattern":"/doi/full/","negate":false},{"component":"url","match_operation":"contains","pattern":"/doi/abs/","negate":false},{"component":"url","match_operation":"contains","pattern":"/journal/","negate":false},{"component":"url","match_operation":"simple","pattern":"https://www.annualreviews.org/","negate":false},{"component":"url","match_operation":"contains","pattern":"/toc/","negate":false},{"component":"url","match_operation":"contains","pattern":"/loi/","negate":false},{"component":"url","match_operation":"contains","pattern":"/action/doSearch","negate":false}],"feedback_widgets":[],"forms":[],"heatmaps":[{"id":5892533,"created_epoch_time":1588809022,"targeting":[{"component":"url","match_operation":"simple","pattern":"https://www.annualreviews.org/toc/psych/63/1","negate":false}],"selector_version":2,"capture_type":"SNAPSHOT_10000"},{"id":5892531,"created_epoch_time":1588808939,"targeting":[{"component":"url","match_operation":"simple","pattern":"https://www.annualreviews.org/journal/psych","negate":false}],"selector_version":2,"capture_type":"SNAPSHOT_10000"},{"id":5892525,"created_epoch_time":1588808851,"targeting":[{"component":"url","match_operation":"simple","pattern":"https://www.an
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/259/choice.js

@@ -1,18 +1,0 @@
-'use strict';(function(){(function(){var cmpScriptElement=document.createElement('script');var firstScript=document.getElementsByTagName('script')[0];cmpScriptElement.async=true;cmpScriptElement.type='text/javascript';var cmpVersion='https://quantcast.mgr.consensu.org/tcfv2/cmp2.js';cmpScriptElement.src=cmpVersion;firstScript.parentNode.insertBefore(cmpScriptElement,firstScript);})();(function(){var css=""
-+" .qc-cmp-button.qc-cmp-secondary-button:hover { "
-+"   background-color: #368bd6 !important; "
-+"   border-color: transparent !important; "
-+" } "
-+" .qc-cmp-button.qc-cmp-secondary-button:hover { "
-+"   color: #ffffff !important; "
-+" } "
-+" .qc-cmp-button.qc-cmp-secondary-button { "
-+"   color: #368bd6 !important; "
-+" } "
-+" .qc-cmp-button.qc-cmp-secondary-button { "
-+"   background-color: #eee !important; "
-+"   border-color: transparent !important; "
-+" } "
-+""
-+"";var stylesElement=document.createElement('style');var re=new RegExp('&quote;','g');css=css.replace(re,'"');stylesElement.type='text/css';if(stylesElement.styleSheet){stylesElement.styleSheet.cssText=css;}else{stylesElement.appendChild(document.createTextNode(css));}
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/259/custom-202005191922.js

@@ -1,216 +1,0 @@
-$(document).ready(function() {
-
-	$('[data-toggle="tooltip"]').tooltip(
-		{animation: false}
-	);
-	
-	$('#topnav').affix({
-		offset: { top: $('#topnav').offset().top }
-	});
-	
-	$('#navwrapper').css('top', $('header').height());
-	
-    // For the Second level Dropdown menu, highlight the parent 
-    $( ".dropdown-menu" )
-    .mouseenter(function() {
-        $(this).parent('li').addClass('active-gray');
-    })
-    .mouseleave(function() {
-        $(this).parent('li').removeClass('active-gray');
-    });
-
-   	$('[data-tab]').hover(function() {
-	   	var tab_id = $(this).data('tab');
-	   	var block = tab_id.substring(1,tab_id.lastIndexOf('_'));
-	   	$('.tab-pane').each(function() {
-			if ($(this).attr('id').indexOf(block) == 0) {
-				$(this).removeClass('active');
-			}
-		});
-	   	$(tab_id).addClass('active');
-	});
-
-	$('[data-tab]').click(function(e) {
-		var supportsTouch = 'ontouchstart' in window || navigator.msMaxTouchPoints;
-		if (supportsTouch) e.preventDefault();
-	   	var tab_id = $(this).data('tab');
-	   	var block = tab_id.substring(1,tab_id.lastIndexOf('_'));
-	   	$('.tab-pane').each(function() {
-			if ($(this).attr('id').indexOf(block) == 0) {
-				$(this).removeClass('active');
-			}
-		});
-	   	$(tab_id).addClass('active');
-		var clicked = $(this).data('clicked');
-	   	if (clicked) {
-		   	var url = $(this).attr('href');
-		   	window.location.href = url;
-	   	} else {
-		   	$(this).data('clicked',true);
-	   	}
-	});
-
-	$('.dropdown-submenu a').on('click',function(e) {
-		var supportsTouch = 'ontouchstart' in window || navigator.msMaxTouchPoints;
-		var url = $(this).data('url');
-		$(this).attr('href', url);
-		if (!supportsTouch && e.which == 1) {
-			window.location.href = url;
-		} else if (supportsTouch && $(window).width() < 728) {
-			window.location.href = url;
-		};
-	});
-
-	$('.newsticker').css({ display: "block" });
-	$('.newsticker').newsTicker({
-		max_rows:		1,
-		row_height:		20,
-		duration:		5000,
-		speed:			250,
-		prevButton:		$('.backward'),
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/259/dpg.js

@@ -1,1 +1,0 @@
-!function(t,e){"function"==typeof define&&define.amd?define([],e):t.libGlobalName=e()}(this,function(){var t,$,e,r,n;!function(p){function f(t,e){return r.call(t,e)}function a(t,e){var r,n,o,i,a,s,c,d,u,l,p,f=e&&e.split("/"),h=v.map,g=h&&h["*"]||{};if(t){for(a=(t=t.split("/")).length-1,v.nodeIdCompat&&S.test(t[a])&&(t[a]=t[a].replace(S,"")),"."===t[0].charAt(0)&&f&&(t=f.slice(0,f.length-1).concat(t)),u=0;u<t.length;u++)if("."===(p=t[u]))t.splice(u,1),u-=1;else if(".."===p){if(0===u||1===u&&".."===t[2]||".."===t[u-1])continue;0<u&&(t.splice(u-1,2),u-=2)}t=t.join("/")}if((f||g)&&h){for(u=(r=t.split("/")).length;0<u;u-=1){if(n=r.slice(0,u).join("/"),f)for(l=f.length;0<l;l-=1)if((o=h[f.slice(0,l).join("/")])&&(o=o[n])){i=o,s=u;break}if(i)break;!c&&g&&g[n]&&(c=g[n],d=u)}!i&&c&&(i=c,s=d),i&&(r.splice(0,s,i),t=r.join("/"))}return t}function h(e,r){return function(){var t=n.call(arguments,0);return"string"!=typeof t[0]&&1===t.length&&t.push(null),c.apply(p,t.concat([e,r]))}}function g(e){return function(t){w[e]=t}}function m(t){if(f(_,t)){var e=_[t];delete _[t],A[t]=!0,i.apply(p,e)}if(!f(w,t)&&!f(A,t))throw new Error("No "+t);return w[t]}function s(t){var e,r=t?t.indexOf("!"):-1;return-1<r&&(e=t.substring(0,r),t=t.substring(r+1,t.length)),[e,t]}var i,c,y,b,w={},_={},v={},A={},r=Object.prototype.hasOwnProperty,n=[].slice,S=/\.js$/;y=function(t,e){var r,n,o=s(t),i=o[0];return t=o[1],i&&(r=m(i=a(i,e))),i?t=r&&r.normalize?r.normalize(t,(n=e,function(t){return a(t,n)})):a(t,e):(i=(o=s(t=a(t,e)))[0],t=o[1],i&&(r=m(i))),{f:i?i+"!"+t:t,n
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/259/gtm.js

@@ -1,251 +1,0 @@
-
-// Copyright 2012 Google Inc. All rights reserved.
-(function(){
-
-var data = {
-"resource": {
-  "version":"15",
-  
-  "macros":[{
-      "function":"__e"
-    },{
-      "function":"__gas",
-      "vtp_cookieDomain":"sciencedaily.com",
-      "vtp_doubleClick":false,
-      "vtp_setTrackerName":false,
-      "vtp_useDebugVersion":false,
-      "vtp_fieldsToSet":["list",["map","fieldName","anonymizeIp","value","true"]],
-      "vtp_useHashAutoLink":false,
-      "vtp_decorateFormsAutoLink":false,
-      "vtp_enableLinkId":false,
-      "vtp_enableEcommerce":false,
-      "vtp_trackingId":"UA-31969-1",
-      "vtp_enableUaRlsa":false,
-      "vtp_enableUseInternalVersion":false
-    },{
-      "function":"__u",
-      "vtp_component":"URL"
-    },{
-      "function":"__u",
-      "vtp_component":"HOST"
-    },{
-      "function":"__u",
-      "vtp_component":"PATH"
-    },{
-      "function":"__f",
-      "vtp_component":"URL"
-    },{
-      "function":"__e"
-    }],
-  "tags":[{
-      "function":"__ua",
-      "once_per_event":true,
-      "vtp_overrideGaSettings":false,
-      "vtp_trackType":"TRACK_PAGEVIEW",
-      "vtp_gaSettings":["macro",1],
-      "vtp_enableUaRlsa":false,
-      "vtp_enableUseInternalVersion":false,
-      "vtp_enableFirebaseCampaignData":true,
-      "tag_id":2
-    },{
-      "function":"__html",
-      "once_per_event":true,
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/273/5981609_003.js

@@ -1,12 +1,0 @@
-/**
- * HubSpot Cookie Banner Code Copyright 2020 HubSpot, Inc.  http://www.hubspot.com
- */
-var _hsp = window['_hsp'] = window['_hsp'] || [];
-_hsp.push(['setCookiePolicy', [{"domain":null,"path":"/hubspot-analytics-default-policy","label":null,"enabled":false,"privacyPolicy":0,"privacyHideDecline":false,"privacyDefault":true,"privacyPolicyWording":null,"privacyAcceptWording":null,"privacyDismissWording":null,"privacyDisclaimerWording":null,"privacyBannerAccentColor":null,"privacyBannerType":"TOP"}]]);
-_hsp.push(['addCookieDomain', '.mongabay.org']);
-_hsp.push(['addCookieDomain', '.hsforms.com']);
-_hsp.push(['addCookieDomain', '.hs-sites.com']);
-_hsp.push(['addCookieDomain', '.hubspot.com']);
-_hsp.push(['addCookieDomain', '.mongabay.com']);
-!function(e){var i={};function t(n){if(i[n])return i[n].exports;var o=i[n]={i:n,l:!1,exports:{}};e[n].call(o.exports,o,o.exports,t);o.l=!0;return o.exports}t.m=e;t.c=i;t.d=function(e,i,n){t.o(e,i)||Object.defineProperty(e,i,{enumerable:!0,get:n})};t.r=function(e){"undefined"!=typeof Symbol&&Symbol.toStringTag&&Object.defineProperty(e,Symbol.toStringTag,{value:"Module"});Object.defineProperty(e,"__esModule",{value:!0})};t.t=function(e,i){1&i&&(e=t(e));if(8&i)return e;if(4&i&&"object"==typeof e&&e&&e.__esModule)return e;var n=Object.create(null);t.r(n);Object.defineProperty(n,"default",{enumerable:!0,value:e});if(2&i&&"string"!=typeof e)for(var o in e)t.d(n,o,function(i){return e[i]}.bind(null,o));return n};t.n=function(e){var i=e&&e.__esModule?function(){return e.default}:function(){return e};t.d(i,"a",i);return i};t.o=function(e,i){return Object.prototype.hasOwnProperty.call(e,i)};t.p="//static.hsappstatic.net/cookie-banner/static-1.282/";t(t.s=2)}([function(e,i){e.exports="div#hs-eu-cookie-confirmation{background:#fff;height:auto;left:0;position:abs
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/273/loader-v2.js

@@ -1,101 +1,0 @@
-(function() {
-
-    var __hs_cta_json = {"css":"a#cta_button_5981609_new {\n    -webkit-font-smoothing: antialiased !important;\ncursor: pointer !important;\n-moz-user-select: none !important;\n-webkit-user-select: none !important;\n-o-user-select: none !important;\nuser-select: none !important;\ndisplay: inline-block !important;\nfont-weight: normal !important;\ntext-align: center !important;\ntext-decoration: none !important;\n-moz-transition: all .4s ease !important;\n-webkit-transition: all .4s ease !important;\n-o-transition: all .4s ease !important;\nbackground: rgb(53,116,227) !important;\nborder-radius: 6px !important;\nborder-width: 0px !important;\ncolor: rgb(255,255,255) !important;\nfont-family: sans-serif !important;\nheight: auto !important;\ntransition: all .4s ease !important;\npadding: 6px 18px !important;\ntext-shadow: none !important;\nwidth: auto !important;\nfont-size: 24px !important;\nline-height: 1.5em !important;\n\n    \n  }\n\na#cta_button_5981609_new:hover {\nbackground: rgb(58,127,249) !important;\ncolor: rgb(255,255,255) !important;\n}\n\na#cta_button_5981609_new:active, #cta_button_5981609_new:active:hover {\nbackground: rgb(42,92,181) !important;\ncolor: rgb(244,244,244) !important;\n}\n\n","css_links":[],"image_html":"<a id=\"cta_button_5981609_c3b67c76-5664-432c-abf2-0900ffcd0cae\" class=\"cta_button\" href=\"https://donate.mongabay.org/cs/c/?cta_guid=c3b67c76-5664-432c-abf2-0900ffcd0cae&placement_guid=fb0aab00-e110-4bff-ad96-0e49b955b758&portal_id=5981609&canon=https%3A%2F%2Fnews.mongabay.com%2F2019%2F04%2Fshade-or-sun-forest-structure-affects-tree-responses-to-amazon-drought%2F&redirect_url=APefjpENJptfupzmLxORxGCLmlMctndX-VDCeZRl_cDSamnRlajHAJKVLaF8gJMYAO8nNdI1ZV0kuSt0AkuMXDfCHdb9_4D5RYjlMDsLFsV93TUqsZW3mTorodxzS3jPn1gDWMoM6zG4&click=f9fc08
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/273/scripts.js

@@ -1,1 +1,0 @@
-function captionsAlign(){const e=document.querySelectorAll("figure"),o=document.querySelector("div#main.col-lg-8").clientWidth;e.forEach(function(e){const t=e.childNodes;let n=0;t.forEach(function(e){switch(e.nodeName){case"A":n=e.firstChild.clientWidth;break;case"IMG":n=e.clientWidth;break;case"FIGCAPTION":const t=o-n;o/n>1.3&&e.setAttribute("style","position:absolute;bottom:0;right:0;padding:0 0 0 1em;text-align:left;width:"+t-30+"px")}})})}document.addEventListener("DOMContentLoaded",function(){captionsAlign(),document.addEventListener("resize",function(){captionsAlign()})}),addToHomescreen(),jQuery(function(){"use strict";jQuery(".responsive-title").fitText(1.2,{minFontSize:"22px",maxFontSize:"45px"}),jQuery(".navbar-toggler, #backdrop").on("click touch",function(){jQuery(".navbar-toggler").toggleClass("opened"),jQuery(".navbar-collapse").toggleClass("show"),jQuery("#backdrop").toggleClass("show")}),jQuery(".sharemobile").on("click touch",function(){jQuery(".header .social").toggleClass("show")}),jQuery(".follow-mobile-button").on("click touch",function(){jQuery("#follow-modal").toggleClass("show")}),jQuery(".follow-modal-close").on("click touch",function(){jQuery("#follow-modal").toggleClass("show")}),jQuery(".app-store-close").on("click touch",function(){jQuery(".app-store-notify").toggleClass("hide")}),"function"==typeof iFrameResize&&iFrameResize({log:!1,initCallback:function(){jQuery(window).trigger("resize.px.parallax")},checkOrigin:!1,heightCalculationMethod:"taggedElement"},"#consevidence")});

### 2022-09-02 Lit_Review/Zotero_references/files/30/session.js

@@ -1,12 +1,0 @@
-/*
- * session.js: Payment Session integration support.
- *
- * Copyright (c) 2015 MasterCard. All rights reserved.
- */
-
-var PaymentSession =(function() {
-    !function a(b,c,d){function e(g,h){if(!c[g]){if(!b[g]){var i="function"==typeof require&&require;if(!h&&i)return i(g,!0);if(f)return f(g,!0);var j=new Error("Cannot find module '"+g+"'");throw j.code="MODULE_NOT_FOUND",j}var k=c[g]={exports:{}};b[g][0].call(k.exports,function(a){var c=b[g][1][a];return e(c?c:a)},k,k.exports,a,b,c,d)}return c[g].exports}for(var f="function"==typeof require&&require,g=0;g<d.length;g++)e(d[g]);return e}({1:[function(a,b,c){b.exports=function(){function b(a,b,c,d){if(a.callbacks){var f=a.callbacks[b];if("function"==typeof f)return void f(c)}d&&e.log(JSON.stringify(c))}function c(a,c,d){b(a,c,d,!1)}function d(a,c,d){b(a,c,d,!0)}var e=a("./log.js")();return{invoke:c,invokeOrLog:d}}},{"./log.js":2}],2:[function(a,b,c){b.exports=function(a){"use strict";function b(b){return function(c,d){if(("debug"!==b||a)&&console&&"function"==typeof console[b]){var e=c;"undefined"!=typeof d&&(e=e+" "+JSON.stringify(d)),console[b](e)}}}return{log:b("log"),debug:b("debug"),error:b("error")}}},{}],3:[function(a,b,c){b.exports=function(){function a(a){var b=0,c=a.length;if(0===c)return 0;for(var d=0;c>d;d++)b=(b<<5)-b+a.charCodeAt(d),b|=0;return b}function b(){var a=navigator.userAgent;return a.indexOf("MSIE")>=0||a.indexOf("Trident")>=0&&a.indexOf("rv:11")>=0}function c(){var a=navigator.userAgent;return a.indexOf("Safari")>=0?a.indexOf("Chrome")>=0||a.indexOf("CriOS")>=0?!1:!0:!1}function d(){return e()&&c()&&f()>=8?!0:!1}function e(){return navigator.userAgent.match(/(iPod|iPhone|iPad)/)?!0:!1}function f(
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/307/event.js

@@ -1,7 +1,0 @@
-﻿/*<!--*/(function() {var qss="&cb="+Math.floor(99999999999*Math.random());try{qss+="&ref="+encodeURIComponent(document.referrer)}catch(e){}try{qss+="&sc_r="+encodeURIComponent(screen.width+"x"+screen.height)}catch(e){}try{qss+="&sc_d="+encodeURIComponent(screen.colorDepth.toString())}catch(e){}var callDis=function(e,t,n,o){function c(){callDisInternal(e,t,n,o)}"complete"===document.readyState?setTimeout(c):window.addEventListener?window.addEventListener("load",c,!1):window.attachEvent("onload",c)},disCalled=!1,callDisInternal=function(e,t,n,o){if(!disCalled){disCalled=!0;var c=(n?"https:":"")+"//"+t+"/dis/dis.aspx",r=document.createElement("iframe");if(r.width=r.height="0",r.style.display="none",r.title="Criteo DIS iframe",void 0!==o&&""!==o&&(document.body.appendChild(r),null!=r.contentDocument))return r.contentDocument.open(),r.contentDocument.write(o),void r.contentDocument.close();r.src=(c+"?p="+e+qss).substring(0,2e3);var i=document.getElementById("criteoTagsContainer");i?(i.appendChild(r),"undefined"!=typeof criteo_q&&"function"==typeof criteo_q.removeLater&&criteo_q.removeLater(r)):criteo_q.push({event:"appendTag",element:r})}};
-qss += '&idcpy=ed44653a-2b51-413e-a922-1ee2b5cf833c';
-callDis(60881, 'widget.us.criteo.com', true, '<!-- default response -->');
-})();
-
-
-/*-->*/

### 2022-09-02 Lit_Review/Zotero_references/files/324/gtm.js

@@ -1,898 +1,0 @@
-
-// Copyright 2012 Google Inc. All rights reserved.
-(function(){
-
-var data = {
-"resource": {
-  "version":"14",
-  
-  "macros":[{
-      "function":"__e"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"event_type"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"supplier_tag"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"object_type"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"siteid"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"authentication_method"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"authzrequired"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"account_id"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"account_list"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"result_click"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"authnips"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"doi"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"jsessionid"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"turnaway_id"
-    },{
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/326/cookie-consent.js

@@ -1,1 +1,0 @@
-!function(n){var e={};function o(t){if(e[t])return e[t].exports;var i=e[t]={i:t,l:!1,exports:{}};return n[t].call(i.exports,i,i.exports,o),i.l=!0,i.exports}o.m=n,o.c=e,o.d=function(n,e,t){o.o(n,e)||Object.defineProperty(n,e,{enumerable:!0,get:t})},o.r=function(n){"undefined"!=typeof Symbol&&Symbol.toStringTag&&Object.defineProperty(n,Symbol.toStringTag,{value:"Module"}),Object.defineProperty(n,"__esModule",{value:!0})},o.t=function(n,e){if(1&e&&(n=o(n)),8&e)return n;if(4&e&&"object"==typeof n&&n&&n.__esModule)return n;var t=Object.create(null);if(o.r(t),Object.defineProperty(t,"default",{enumerable:!0,value:n}),2&e&&"string"!=typeof n)for(var i in n)o.d(t,i,function(e){return n[e]}.bind(null,i));return t},o.n=function(n){var e=n&&n.__esModule?function(){return n.default}:function(){return n};return o.d(e,"a",e),e},o.o=function(n,e){return Object.prototype.hasOwnProperty.call(n,e)},o.p="",o(o.s=0)}([function(n,e){var o,t=document;o=function(){var n=t.createElement("div");n.className="qc-cmp-deprecation-messaging",t&&t.body&&t.body.appendChild&&t.body.appendChild(n)},"complete"===t.readyState||"interactive"===t.readyState?setTimeout(o,1):t.addEventListener("DOMContentLoaded",o);var i=window.canAskForCookieConsent&&window.canAskForCookieConsent();if(window.cookieConsentConfig={gdprApplies:i},i)if(window.__tcfapi){var r={7:"https://www.google-analytics.com"};window.__tcfapi("addEventListener",2,(function(n,e){var o=[];if(e)switch(n.eventStatus){case"useractioncomplete":case"tcloaded":window.quantcastUiFallback=!0,!1!==n.purpose.consents[1]&&!1!==n.purpose.consents[3]||Object.keys(r).forEach((function(n){return o.push(n)})),o.length?window.loadDeferredObjects(o):window.__tcfapi("getNonIABVendorConsents",2,(function(n,e){e&&n.gdprApplies?(Object.keys(r).forEach((function(e){!1===n.nonIabVendorConsents[e]&&o.push(r[e])})),window.loa
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/326/lite.js

@@ -1,8 +1,0 @@
-!function(e){var t={};function n(r){if(t[r])return t[r].exports;var o=t[r]={i:r,l:!1,exports:{}};return e[r].call(o.exports,o,o.exports,n),o.l=!0,o.exports}n.m=e,n.c=t,n.d=function(e,t,r){n.o(e,t)||Object.defineProperty(e,t,{enumerable:!0,get:r})},n.r=function(e){"undefined"!=typeof Symbol&&Symbol.toStringTag&&Object.defineProperty(e,Symbol.toStringTag,{value:"Module"}),Object.defineProperty(e,"__esModule",{value:!0})},n.t=function(e,t){if(1&t&&(e=n(e)),8&t)return e;if(4&t&&"object"==typeof e&&e&&e.__esModule)return e;var r=Object.create(null);if(n.r(r),Object.defineProperty(r,"default",{enumerable:!0,value:e}),2&t&&"string"!=typeof e)for(var o in e)n.d(r,o,function(t){return e[t]}.bind(null,o));return r},n.n=function(e){var t=e&&e.__esModule?function(){return e.default}:function(){return e};return n.d(t,"a",t),t},n.o=function(e,t){return Object.prototype.hasOwnProperty.call(e,t)},n.p="https://c5.rgstatic.net/",n(n.s=742)}({1:function(e,t,n){"use strict";function r(e,t,n){return t in e?Object.defineProperty(e,t,{value:n,enumerable:!0,configurable:!0,writable:!0}):e[t]=n,e}function o(e,t){var n=Object.keys(e);if(Object.getOwnPropertySymbols){var r=Object.getOwnPropertySymbols(e);t&&(r=r.filter((function(t){return Object.getOwnPropertyDescriptor(e,t).enumerable}))),n.push.apply(n,r)}return n}function i(e){for(var t=1;t<arguments.length;t++){var n=null!=arguments[t]?arguments[t]:{};t%2?o(Object(n),!0).forEach((function(t){r(e,t,n[t])})):Object.getOwnPropertyDescriptors?Object.defineProperties(e,Object.getOwnPropertyDescriptors(n)):o(Object(n)).forEach((function(t){Object.defineProperty(e,t,Object.getOwnPropertyDescriptor(n,t))}))}return e}n.d(t,"a",(function(){return i}))},103:function(e,t,n){"use strict"
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/335/api.js

@@ -1,1 +1,0 @@
-/* PLEASE DO NOT COPY AND PASTE THIS CODE. */(function(){var w=window,C='___grecaptcha_cfg',cfg=w[C]=w[C]||{},N='grecaptcha';var gr=w[N]=w[N]||{};gr.ready=gr.ready||function(f){(cfg['fns']=cfg['fns']||[]).push(f);};(cfg['render']=cfg['render']||[]).push('explicit');(cfg['onload']=cfg['onload']||[]).push('onloadCallback');w['__google_recaptcha_client']=true;var d=document,po=d.createElement('script');po.type='text/javascript';po.async=true;po.src='https://www.gstatic.com/recaptcha/releases/QVh-Tz10ahidjrORgXOS1oB0/recaptcha__en.js';var e=d.querySelector('script[nonce]'),n=e&&(e['nonce']||e.getAttribute('nonce'));if(n){po.setAttribute('nonce',n);}var s=d.getElementsByTagName('script')[0];s.parentNode.insertBefore(po, s);})();

### 2022-09-02 Lit_Review/Zotero_references/files/335/article-page.js

@@ -1,487 +1,0 @@
-//--------DOCUMENTATION--------
-//This widget has 2 modes for the view: 1) a condensed version, and 2) a full version.  By default, only the condensed 
-//version is shown, and the full version is called via AJAX from within the short version.
-
-//    Dependencies:
-//1. CDN Chartist.CSS <link rel="stylesheet" href="bower_components/chartist/dist/chartist.min.css">
-//2. CDN Chartist.js  <script src="bower_components/chartist/dist/chartist.min.js">
-//3. ArticleLevelMetrics.js --> which should ship over in the SharedWidgets NuGet Package
-//4. _article-level-metrics.scss --> default presentation for the widget. should ship over with NuGet Package.
-var ArtMetric = {};
-
-ArtMetric.fullVersion = function ($dest, routeUrl) {
-    // Only do AJAX call if it hasn't been done before.
-    //console.log(routeUrl);
-    if ($dest.find('.widget-ArticleLevelMetrics').length < 1) {
-        // Code from https://github.com/gionkunz/chartist-js/issues/1015
-        var ctHtmlLabels = function (options) {
-            return function (chart) {
-                chart.on('draw', function (context) {
-                    if (context.type === 'label') {
-                        var redo = context.element._node.childNodes[0];
-                        redo.innerHTML = redo.innerText;
-                    }
-                });
-            }
-        };
-
-        var baseSiteUrl = $("#hfSiteURL");
-
-        if (typeof baseSiteUrl != "undefined" && typeof baseSiteUrl.val() != "undefined" && baseSiteUrl != '') {
-            routeUrl = "//" + baseSiteUrl.val() + routeUrl;
-        }
-
-        $.ajax({
-            url: routeUrl,
-            datatype: "HTML",
-            success: function (html) {
-                $dest.prepend(html);
-
-                setTimeout(function () {
-                    if (typeof window.ChartistData !== "undefined") {
-                        // Chartist Options
-                        var opts = {
-                            lineSmooth: false,
-                            showArea: true,
-                            low: 0,
-                            plugins: [ctHtmlLabels()]
-                        };
-
-                        var responsiveOpts = [
-                            ['screen and (max-width: 1024px)', {
-                                axisX: {
-                                    labelInterpolationFnc: function (value) {
-                                        return value.slice(0, 3);
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/335/gpt.js

@@ -1,6 +1,0 @@
-(function(E){var window=this;/*
-
- Copyright The Closure Library Authors.
- SPDX-License-Identifier: Apache-2.0
-*/
-var aa=function(a){var b=0;return function(){return b<a.length?{done:!1,value:a[b++]}:{done:!0}}},ba="function"==typeof Object.defineProperties?Object.defineProperty:function(a,b,c){if(a==Array.prototype||a==Object.prototype)return a;a[b]=c.value;return a},ca=function(a){a=["object"==typeof globalThis&&globalThis,a,"object"==typeof window&&window,"object"==typeof self&&self,"object"==typeof global&&global];for(var b=0;b<a.length;++b){var c=a[b];if(c&&c.Math==Math)return c}throw Error("Cannot find global object");},da=ca(this),ea="function"===typeof Symbol&&"symbol"===typeof Symbol("x"),m={},ha={},n=function(a,b){var c=ha[b];if(null==c)return a[b];c=a[c];return void 0!==c?c:a[b]},q=function(a,b,c){if(b)a:{var d=a.split(".");a=1===d.length;var e=d[0],f;!a&&e in m?f=m:f=da;for(e=0;e<d.length-1;e++){var g=d[e];if(!(g in f))break a;f=f[g]}d=d[d.length-1];c=ea&&"es6"===c?f[d]:null;b=b(c);null!=b&&(a?ba(m,d,{configurable:!0,writable:!0,value:b}):b!==c&&(ha[d]=ea?da.Symbol(d):"$jscp$"+d,d=ha[d],ba(f,d,{configurable:!0,writable:!0,value:b})))}};q("Symbol",function(a){if(a)return a;var b=function(e,f){this.a=e;ba(this,"description",{configurable:!0,writable:!0,value:f})};b.prototype.toString=function(){return this.a};var c=0,d=function(e){if(this instanceof d)throw new TypeError("Symbol is not a constructor");return new b("jscomp_symbol_"+(e||"")+"_"+c++,e)};return d},"es6");q("Symbol.iterator",function(a){if(a)return a;a=(0,m.Symbol)("Symbol.iterator");for(var b="Array Int8Array Uint8Array Uint8ClampedArray Int16Array Uint16Array Int32Array Uint32Array Float32Array Float64Array".split(" "),c=0;c<b.length;c++){var d=da[b[c]];"function"===typeof d&&"function"!=typeof d.p
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/337/session.js

@@ -1,12 +1,0 @@
-/*
- * session.js: Payment Session integration support.
- *
- * Copyright (c) 2015 MasterCard. All rights reserved.
- */
-
-var PaymentSession =(function() {
-    !function a(b,c,d){function e(g,h){if(!c[g]){if(!b[g]){var i="function"==typeof require&&require;if(!h&&i)return i(g,!0);if(f)return f(g,!0);var j=new Error("Cannot find module '"+g+"'");throw j.code="MODULE_NOT_FOUND",j}var k=c[g]={exports:{}};b[g][0].call(k.exports,function(a){var c=b[g][1][a];return e(c?c:a)},k,k.exports,a,b,c,d)}return c[g].exports}for(var f="function"==typeof require&&require,g=0;g<d.length;g++)e(d[g]);return e}({1:[function(a,b,c){b.exports=function(){function b(a,b,c,d){if(a.callbacks){var f=a.callbacks[b];if("function"==typeof f)return void f(c)}d&&e.log(JSON.stringify(c))}function c(a,c,d){b(a,c,d,!1)}function d(a,c,d){b(a,c,d,!0)}var e=a("./log.js")();return{invoke:c,invokeOrLog:d}}},{"./log.js":2}],2:[function(a,b,c){b.exports=function(a){"use strict";function b(b){return function(c,d){if(("debug"!==b||a)&&console&&"function"==typeof console[b]){var e=c;"undefined"!=typeof d&&(e=e+" "+JSON.stringify(d)),console[b](e)}}}return{log:b("log"),debug:b("debug"),error:b("error")}}},{}],3:[function(a,b,c){b.exports=function(){function a(a){var b=0,c=a.length;if(0===c)return 0;for(var d=0;c>d;d++)b=(b<<5)-b+a.charCodeAt(d),b|=0;return b}function b(){var a=navigator.userAgent;return a.indexOf("MSIE")>=0||a.indexOf("Trident")>=0&&a.indexOf("rv:11")>=0}function c(){var a=navigator.userAgent;return a.indexOf("Safari")>=0?a.indexOf("Chrome")>=0||a.indexOf("CriOS")>=0?!1:!0:!1}function d(){return e()&&c()&&f()>=8?!0:!1}function e(){return navigator.userAgent.match(/(iPod|iPhone|iPad)/)?!0:!1}function f(
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/74/js__NKAtuj0DQYbxgN6smAsbWhIflwXe6AV73nHYxbYqhQw__psO4qinGYZGz.js

@@ -1,3 +1,0 @@
-(function(a){Drupal.behaviors.myCustomJS={attach:function(t,h){a(".highwire-markup .em-addr",t).each(function(){var t=a(this).html().replace("\{at\}","@"),t="<a href='mailto:"+t+"'>"+t+"</a>";a(this).html(t)})}}})(jQuery);(function(t){Drupal.progressBar=function(r,i,s,e){var a=this;this.id=r;this.method=s||'GET';this.updateCallback=i;this.errorCallback=e;this.element=t('<div class="progress" aria-live="polite"></div>').attr('id',r);this.element.html('<div class="bar"><div class="filled"></div></div><div class="percentage"></div><div class="message">&nbsp;</div>')};Drupal.progressBar.prototype.setProgress=function(r,i){if(r>=0&&r<=100){t('div.filled',this.element).css('width',r+'%');t('div.percentage',this.element).html(r+'%')};t('div.message',this.element).html(i);if(this.updateCallback){this.updateCallback(r,i,this)}};Drupal.progressBar.prototype.startMonitoring=function(t,r){this.delay=r;this.uri=t;this.sendPing()};Drupal.progressBar.prototype.stopMonitoring=function(){clearTimeout(this.timer);this.uri=null};Drupal.progressBar.prototype.sendPing=function(){if(this.timer){clearTimeout(this.timer)};if(this.uri){var r=this;t.ajax({type:this.method,url:this.uri,data:'',dataType:'json',success:function(t){if(t.status==0){r.displayError(t.data);return};r.setProgress(t.percentage,t.message);r.timer=setTimeout(function(){r.sendPing()},r.delay)},error:function(t){r.displayError(Drupal.ajaxError(t,r.uri))}})}};Drupal.progressBar.prototype.displayError=function(r){var i=t('<div class="messages error"></div>').html(r);t(this.element).before(i).hide();if(this.errorCallback){this.errorCallback(this)}}})(jQuery);(function(t){Drupal.behaviors.articleAuthorPopup={attach:function(o,e){var i=!!jQuery.fn.qtip;if(!i){if(window.console){console.error("HW's qTip behavior is called without the qTip library loaded. Please investigate.")};return};var a=(e.instances!=undefined)?t.parseJSON(e.instances):{};t("body",o).delegate(".has-author-tooltip span.highwire-citation-author:not(.hasTooltip):not(.noTooltip)","mouseenter",f
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/74/js__on06AwGisxYIeuMzIiAIgSFUOCTqgag02UtyiOm39oA__0Y58bI7A_OKq.js

@@ -1,28 +1,0 @@
-(function(i){Drupal.behaviors.highwire_citation_export={attach:function(a,o){var t=!1;i('.highwire-citation-export-link-ajax',a).click(function(){i('.highwire-citation-export-link-ajax-popup').dialog({modal:!0,draggable:!1,maxWidth:700});return!1})}}}(jQuery));(function(i){Drupal.behaviors.minipanel_dialog_link={attach:function(n,a){var t=function(){var n=i(window).width();if(n>=Drupal.settings.highwire.modal_window_width){return Drupal.settings.highwire.modal_window_width}
-else{return'90%'}};i('.minipanel-dialog-link-trigger',n,a).once('minipanel-dialog-link-trigger',function(){if(i(this).attr('title')!='<blank>'){var a=i(this).attr('title')}
-else{var a=''};var n=i(this).parent().parent().find('.minipanel-dialog-link-mini');n.dialog({'modal':!0,'draggable':!1,'title':a,'resizable':!1,'autoOpen':!1});i(this).click(function(){n.dialog({'width':t()});n.dialog('open');return!1})})}}}(jQuery));(function(i){Drupal.behaviors.highwire_shareitajax={attach:function(r,e){var h=function(){var a=i(window).width();if(a>=Drupal.settings.highwire.share_modal_width){return Drupal.settings.highwire.share_modal_width}
-else{return'90%'}},t=Drupal.settings.highwire.share_modal_title,n=!1,a='',l='';i('.highwire_clipboard_link_ajax',r,e).click(function(){a=i(this).attr('id');i(this).parent().find('.highwire_clipboard_form_ajax_'+a).dialog({'modal':!0,'draggable':!1,'width':h(),'title':t,'resizable':!1});i('.highwire_clipboard_form_ajax_'+a).dialog('open');return!1})}}}(jQuery));/*!
- * clipboard.js v1.5.12
- * https://zenorocha.github.io/clipboard.js
- *
- * Licensed MIT © Zeno Rocha
- */
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/74/loader.js

@@ -1,1 +1,0 @@
-var fc_css=document.createElement("link");fc_css.setAttribute("rel","stylesheet");fc_css.setAttribute("media","screen");fc_css.setAttribute("href","https://cdn.foxycart.com/pnas.ecommerce.highwire.org/responsive_styles.1594668239.css");var fc_script=document.createElement("script");window.jQuery&&(1<=window.jQuery.fn.jquery.match(/(\d+)\.(\d+)/)[1]&&7<window.jQuery.fn.jquery.match(/(\d+)\.(\d+)/)[2]||2<=window.jQuery.fn.jquery.match(/(\d+)\.(\d+)/)[1])?fc_script.src="https://cdn.foxycart.com/pnas.ecommerce.highwire.org/foxycart.jsonp.sidecart.min.1594666795.js":fc_script.src="https://cdn.foxycart.com/pnas.ecommerce.highwire.org/foxycart.jsonp.sidecart.with-jquery.min.1594666795.js";function fc_loader(){document.getElementsByTagName("body")[0].appendChild(fc_script);document.getElementsByTagName("body")[0].appendChild(fc_css);}document.readyState==="complete"?fc_loader():window.addEventListener?window.addEventListener("load",fc_loader,!1):window.attachEvent?window.attachEvent("onload",fc_loader):window.onload=fc_loader;

### 2022-09-02 Lit_Review/Zotero_references/files/74/widget.js

@@ -1,11 +1,0 @@
-document.CROSSMARK={VERIFICATION:"eyJ0eXAiOiJKV1QiLCJhbGciOiJIUzI1NiJ9.eyJpYXQiOjE1OTQ5OTQwNTksInZlciI6IjIuMC4xMDcifQ.uAeFMQm3UrvsjpaJXrGItMFLv_XMNSlqaGcJIj83JPQ",ENDPOINT:"https://crossmark.crossref.org/dialog",SCRIPT_VERSION:"2.0.107",STYLESHEET_URL:"https://crossmark-cdn.crossref.org/widget/v2.0/style.css",LOGO_URL:"https://crossmark-cdn.crossref.org/images/logo-crossmark.svg"};
-document.CROSSMARK.getDoiMetaTags=function(){var b=[],d=document.querySelectorAll("meta"),a;for(a in d)if(d.hasOwnProperty(a)){var c=d[a];if((c.name||"").toLowerCase().match(/^dc\.identifier/)){var e=(c.getAttribute("scheme")||"").toLowerCase();c.getAttribute("content")&&c.getAttribute("content").match(/(10\.\d+\/.*$)/)&&(""===e||"doi"===e)&&b.push(c)}}return b};
-document.CROSSMARK.getDoi=function(){var b=document.CROSSMARK.getDoiMetaTags()[0];if(0===b.length)return null;var b=b?b.getAttribute("content").replace(/^(info:doi\/|doi:)/,""):null,d=b.match(/(10\.\d+\/.*$)/);null!==d&&(b=d[0]);return b};document.CROSSMARK.buildQueryString=function(b){var d=[],a;for(a in b)b.hasOwnProperty(a)&&d.push(encodeURIComponent(a)+"\x3d"+encodeURIComponent(b[a]));return"?"+d.join("\x26")};document.CROSSMARK.touchStarted=!1;document.CROSSMARK.touchArea=null;
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/83/gtm.js

@@ -1,891 +1,0 @@
-
-// Copyright 2012 Google Inc. All rights reserved.
-(function(){
-
-var data = {
-"resource": {
-  "version":"14",
-  
-  "macros":[{
-      "function":"__e"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"event_type"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"supplier_tag"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"object_type"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"siteid"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"authentication_method"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"authzrequired"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"account_id"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"account_list"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"result_click"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"authnips"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"doi"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"jsessionid"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"turnaway_id"
-    },{
[…]

### 2022-09-02 Lit_Review/Zotero_references/files/99/gtm.js

@@ -1,2096 +1,0 @@
-
-// Copyright 2012 Google Inc. All rights reserved.
-(function(w,g){w[g]=w[g]||{};w[g].e=function(s){return eval(s);};})(window,'google_tag_manager');(function(){
-
-var data = {
-"resource": {
-  "version":"679",
-  
-  "macros":[{
-      "function":"__v",
-      "vtp_name":"gtm.elementId",
-      "vtp_dataLayerVersion":1
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":false,
-      "vtp_name":"OnetrustActiveGroups"
-    },{
-      "function":"__jsm",
-      "vtp_javascript":["template","(function(){return\"string\"===typeof ",["escape",["macro",1],8,16],"\u0026\u00260\u003C",["escape",["macro",1],8,16],".length?-1!==",["escape",["macro",1],8,16],".indexOf(\",2,\"):!1})();"]
-    },{
-      "function":"__e"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":true,
-      "vtp_defaultValue":"",
-      "vtp_name":"Event Category"
-    },{
-      "function":"__jsm",
-      "convert_true_to":"true",
-      "convert_false_to":"false",
-      "vtp_javascript":["template","(function(){return\"string\"===typeof ",["escape",["macro",1],8,16],"\u0026\u00260\u003C",["escape",["macro",1],8,16],".length?-1!==",["escape",["macro",1],8,16],".indexOf(\",4,\"):!1})();"]
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":true,
-      "vtp_defaultValue":"N",
-      "vtp_name":"Open Access"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":true,
-      "vtp_defaultValue":"N",
-      "vtp_name":"Labs"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":true,
-      "vtp_defaultValue":"",
-      "vtp_name":"content.book.bookProductType"
-    },{
-      "function":"__v",
-      "vtp_dataLayerVersion":2,
-      "vtp_setDefaultValue":true,
-      "vtp_defaultValue":"",
[…]

### 2022-09-02 vertical-thermal-review.Rmd
added lawren's
Update vertical-thermal-review.Rmd
entered my orcid ID
moved table pngs to tables folder
#126
organize figures in "figures" folder, move extraneous files to private
New "figures" folder contains figures for which little to new analysis was done.

@@
 
-Nidhi Vinod^1,2^
+Nidhi Vinod^1,2^ (ORCID: 0000-0003-2726-4105; Twitter: NidhiVinod4)
 
@@
 
-Lawren Sack^2^
+Lawren Sack^2^ (ORCID: 0000-0002-7009-7202; Twitter: lawrensack)
 
@@
 Abbreviations are as follows: *T~air~*: air temperature, VPD: vapor pressure deficit;  *T~leaf~* - *T~air~*: leaf-to-air temperature difference; LMA: leaf mass per area; T sensitivity: temperature sensitivity; ET: evapotranspiration.
-*Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`).](schematics/Schematic figure 1.png)
+*Isoprene emission has also been observed to peak in mid-canopy (Table `r table_leaf_traits`).](figures/Schematic figure 1.png)
 
@@
 
-![](table1a.png)
+![](tables/table1a.png)
 
@@
 
-![](table1b.png)
+![](tables/table1b.png)
 
@@
 
-![](table1c.png)
+![](tables/table1c.png)
 
@@
 
-\newpage
+** composite climatic stress variable integrating temperature, vapour pressure deficit, and relative humidity is higher in lower canopy \newpage
 
@@
 
-![](table2a.png)
+![](tables/table2a.png)
 
@@
 
-![](table2b.png)
+![](tables/table2b.png)
 
@@
 
-![](table2c.png)
+![](tables/table2c.png)
 
@@
 
-** composite climatic stress variable integrating temperature, vapour pressure deficit, and relative humidity is higher in lower canopy \newpage
-
 ### 4.1. Conductance
@@
 Transparent ribbons indicate 95% confidence intervals.
-Redrawn from @anderson-teixeiraJointEffectsClimate2022.](tree_rings/tree_ring_figure.png)
+Redrawn from @anderson-teixeiraJointEffectsClimate2022.](figures/tree_ring_figure.png)
 
@@
 ![**Figure `r fig_ecosystem`.
-Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotDistributionBiomassDynamics2022.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
+Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotDistributionBiomassDynamics2022.](figures/ecosystem vertical partitioning/ecosystem partitioning.png)
 
@@
 Boucher and Andrew B.
-Davies.](schematics/fig_remote_sensing.png)
+Davies.](figures/fig_remote_sensing.png)
 
@@
 NV, IM, EMO and MNS contributed data and analyses.
-All authors contributed to writing and revising of the manuscript.
+MS, TCT, LS, TNB contributed to writing and revising of the manuscript.
 

### 2022-09-30 vertical-thermal-review.Rmd

@@
 New analyses are based upon data available from the National Ecological Observatory Network (NEON; https://www.neonscience.org/).
-The R scripts used for new analyses are (*will be*) available via GitHub (https://github.com/EcoClimLab/vertical-thermal-review) and archived in Zenodo (DOI: [TBD]).
+The R scripts used for new analyses are available via GitHub (https://github.com/EcoClimLab/vertical-thermal-review) and archived in Zenodo (DOI: 10.5281/zenodo.7050582).
 

### 2022-09-30 NEON_height_profiles/scripts/verticalProfileFigure.R
saved NEON figure as pdf

@@ -218,10 +218,11 @@ p <- cowplot::plot_grid(plotsLAD[["LAD"]], plotsLAD[["sun"]], plotsLAD[["lgt"]],
                         nrow=2, ncol=4, align="hv",
                         labels=c("A", "B", "C", "D", "E", "F", "G", "H"),
                         label_x=0.13)
-
+#p
 png(paste0("figures/Fig2_normalized.png"), height=600, width=960)
 print(p)
 dev.off()
+dev.copy2pdf(file = "Fig2_normalized.pdf", useDingbats=FALSE, width = 13, height = 8)
 
 ##########################################################################
 ## Make plots for SI ####

### 2022-09-30 leaf_energy_balance/Figure3_Tleaf.R
Create Figure3_Tleaf.R

@@ -1,0 +1,483 @@
+####################################
+# Purpose: Create leaf energy balance figure using HARV neon data parameters for a typical overstory
+# and understory, normal and drought scenarios
+# Tleaves R package created by Chris Muir, 2019
+# Developed by: Nidhi Vinod, contact Anderson-Teixeira (teixeirak@si.edu)
+# R version 3.5.3 --First created in October 2020, updated on 5/28/2021
+# Oct,2020-Oct,2021
+####################################
+#install packages if needed
+library(latex2exp)
+library(magrittr)
+library(tealeaves)
+library(ggplot2)
+library(dplyr)
+library(ggthemes)
+library(ggpubr)
+library(cowplot)
+library(gridExtra)
+library(RColorBrewer)
+####################################
+
+
+# create a dataframe for overstory and understory, for scenarios: normal and drought
+# 1. Short wave radiation, normal
+#understory
+sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
+                  rh = 0.97, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
+                  gs = 2.0)
+#overstory
+sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
+                  rh = 0.91, wind = 2.88, tair = 298.1, pressure = 99, leaf.size = 0.04, 
+                  gs = 4.0)
+
+sw_l$par <- sw_l$par*0.5#sw
+sw_u$par <- sw_u$par*0.5#sw
+
+#apply Tleaves function
+
+get_tleaf <- function(sw, rh, wind, tair, pressure, leaf.size, gs) {
+  
+  # Set the environmental parameters (Tair, RH, PAR, and wind speed)
+  enviro_par <- make_enviropar(
+    replace = list(
+      T_air = set_units(c(tair), "K"),
+      RH = set_units(c(rh)),
+      S_sw = set_units(c(sw), "W/m^2"),
+      wind = set_units(c(wind), "m/s"),
+      P = set_units(c(pressure), "kPa")
+      
[…]

### 2022-09-30 leaf_energy_balance/Figure3_Tleaf.R
updatedpdf for tleaf

@@ -157,7 +157,7 @@ sw<- data.frame(rbind(u_tla, l_tla, dr_utla, dr_ltla))
 a1<-ggplot(data = sw)+
   geom_smooth(aes(x = par, y = tleaf_tair, color = canopy_position, linetype = canopy_position), 
               method = lm, se = FALSE)+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (째C)"))+xlab(TeX("Short Wave Radiation (swr, Wm$^{-2}$)"))+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (캜)"))+xlab(TeX("Short Wave Radiation (swr, Wm$^{-2}$)"))+
   theme_few()+theme(text = element_text(size = 15))+ylim(-5, 14)+
   geom_hline(yintercept=0, linetype='dotted', color = 'black')+
   scale_colour_manual(values=c("#008837","#7b3294","#008837","#7b3294"), name = "canopy position") + 
@@ -246,7 +246,7 @@ b1<-ggplot(ws)+
   geom_point(aes(x=2.88, y=8.9), colour="#008837")+
   geom_point(aes(x=0.24, y=2.8), colour="#7b3294")+
   geom_point(aes(x=0.24, y=1.4), colour="#7b3294")+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ ($째C$)"))+xlab(TeX("Wind Speed (ws, $ms^{-1}$)"))+
+  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (캜)"))+xlab(TeX("Wind Speed (ws, $ms^{-1}$)"))+
   geom_hline(yintercept=0, linetype='dotted', col = 'black')+
   theme_few()+theme(text = element_text(size = 15))+ ylim(-5, 14) 
 
@@ -313,7 +313,7 @@ c1<-ggplot(gs)+
   geom_point(aes(x=0.2, y=1.4), colour="#7b3294")+
   geom_point(aes(x=0.001, y=8.1), colour="#008837")+
   geom_point(aes(x=0.001, y=1.82), colour="#7b3294")+
-  ylab(TeX("$T_{Leaf}$ - $T_{air}$ (째C)"))+xlab(TeX("Stomatal Conductance ($g_{s}$, mol $m^{-2} s^{-1}$)"))+
[…]

### 2022-09-30 leaf_energy_balance/Figure3_Tleaf.R
Update Figure3_Tleaf.R

@@ -480,6 +480,6 @@ figure1<-ggarrange(a1, b1, e1, d1, c1, table1, ncol=3, nrow =2, align = c('hv'),
                    labels = c("(a)", "(b)", "(c)", "(d)", "(e)"))
 figure1
 
-dev.copy2pdf(file="FigureTleaf2.pdf", useDingbats=FALSE, width=16, height=8)
+dev.copy2pdf(file="FigureTleaf2.pdf", useDingbats=FALSE, width=17, height=10)
 #plot saved as .png
 

### 2022-09-30 leaf_energy_balance/leaf energy balance fig_normal_and_drought.R
moved files to private

@@ -1,462 +1,0 @@
-####################################
-# Purpose: Create leaf energy balance figure using HARV neon data parameters for a typical overstory
-# and understory, normal and drought scenarios
-# Tleaves R package created by Chris Muir, 2019
-# Developed by: Nidhi Vinod, contact Anderson-Teixeira (teixeirak@si.edu)
-# R version 3.5.3 --First created in October 2020, updated on 5/28/2021
-# Oct,2020-Oct,2021
-####################################
-#install packages if needed
-library(latex2exp)
-library(magrittr)
-library(tealeaves)
-library(ggplot2)
-library(dplyr)
-library(ggthemes)
-library(ggpubr)
-library(cowplot)
-library(gridExtra)
-library(RColorBrewer)
-####################################
-
-
-# create a dataframe for overstory and understory, for scenarios: normal and drought
-# 1. Short wave radiation, normal
-#understory
-sw_l<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
-                  rh = 0.97, wind = 0.24, tair = 296.1, pressure = 99, leaf.size = 0.10, 
-                  gs = 2.0)
-#overstory
-sw_u<- data.frame(par = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800, 2000, 2200), 
-                  rh = 0.91, wind = 2.88, tair = 298.1, pressure = 99, leaf.size = 0.04, 
-                  gs = 4.0)
-
-sw_l$par <- sw_l$par*0.5#sw
-sw_u$par <- sw_u$par*0.5#sw
-
-#apply Tleaves function
-
-get_tleaf <- function(sw, rh, wind, tair, pressure, leaf.size, gs) {
-  
-  # Set the environmental parameters (Tair, RH, PAR, and wind speed)
-  enviro_par <- make_enviropar(
-    replace = list(
-      T_air = set_units(c(tair), "K"),
-      RH = set_units(c(rh)),
-      S_sw = set_units(c(sw), "W/m^2"),
-      wind = set_units(c(wind), "m/s"),
-      P = set_units(c(pressure), "kPa")
[…]

### 2022-09-30 leaf_energy_balance/Figure3_Tleaf.R
updating the code

@@ -480,6 +480,6 @@ figure1<-ggarrange(a1, b1, e1, d1, c1, table1, ncol=3, nrow =2, align = c('hv'),
                    labels = c("(a)", "(b)", "(c)", "(d)", "(e)"))
 figure1
 
-dev.copy2pdf(file="FigureTleaf2.pdf", useDingbats=FALSE, width=17, height=10)
+dev.copy2pdf(file="FigureTleaf2.pdf", useDingbats=FALSE, width=17, height=9)
 #plot saved as .png
 
